### 2014-09-22 heavy-ms.tex
Initial commit

\documentclass[10pt]{article}
\usepackage{geometry}
\geometry{letterpaper}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
%\usepackage{pdfpages}
%\usepackage{lscape}
%\usepackage{pdflscape}
%\usepackage{setspace}
\usepackage{booktabs}

\usepackage{longtable}
\usepackage{multirow}
\usepackage{array}
\usepackage{ragged2e}
\usepackage{tabu} % for spacing between rows in longtable
\setlength{\tabulinesep}{9pt}

\mathchardef\mhyphen="2D
%\usepackage{url}
%\urlstyle{rm}

\textheight 20.5cm

\usepackage[round]{natbib}
\bibliographystyle{apalike}
\bibpunct{(}{)}{;}{a}{}{;}

\title{The evidence for heavy tails in global\\population dynamics}
\author{Sean Anderson \ldots}
%\date{}

%\pagestyle{empty}
%\setcounter{secnumdepth}{-2}

\begin{document}

\input{values}

%\begin{spacing}{1.1}
%\doublespacing
%\onehalfspacing
\maketitle

\section*{Abstract}

\section*{Introduction}

Increasing realization about the importance of extreme events in the environment (REFs)

and of the importance of ecological surprises and black swans (REFs)

Papers to work in:

\citep{inchausti2002,halley2002,inchausti2001}

\citep{jentsch2007}

\citep{ward2007}

\citep{garcia-carreras2011}
\citep{sornette2009}

\citep{nunez2012}

\citep{thompson2013}
\citep{beaugrand2012}
\citep{pine-iii2009}

\citep{doak2008}

\citep{smale2013}

\citep{easterling2000}
\citep{scheffer2003}
\citep{katz2005}

\citep{taleb2007}

\citep{vasseur2014}

\citep{vert-pre2013}
\citep{lindenmayer2010}

\ldots

citation with good justification of gompertz and other references: \citep{herrandoprez2014}

\section*{Methods}

First, I tried fitting this as a Gompertz model and estimating the density dependence parameter. The Gompertz was by far the most favoured model in \citet{brook2006} with the GPDD.

I could see an argument being made that we're just seeing them because of autocorrelation in the residuals. Autocorrelation plays a big role in \citet{inchausti2002} justifying why they think they're seeing increasing CV of time series in the GPDD with time and an increasing CV could create heavy tails. They just model the spectral properties and CV of the abundance time series --- not a population dynamics model. I tried estimating a coefficient for the first-order autocorrelation of the residuals. There is some autocorrelation it picks up sometimes, but interestingly this doesn't change things that much with respect to heavy tail results. Nothing systematic at least (Another important reference on spectral analysis and the GPDD: \citet{garcia-carreras2011}.)

\subsection*{Modelling framework}

The autoregressive Student-t Gompertz model is:
\begin{align*}
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}_\nu(\phi \epsilon_{t-1}, \sigma_\mathrm{proc}^2),
\end{align*}

\noindent where $x_t$ is the $\ln$ abundance at time $t$ and $\nu$ is the Student-t distribution degrees of freedom parameter. If $\nu$ is small (say 2--20) the distribution has heavy tails. As $\nu$ approaches infinity the distribution approaches the normal distribution. The parameter $b$ is the density dependence parameter. The model is density independent if $b = 1$ and maximally density dependent if $b = 0$. The parameter $\phi$ represent the first-order autoregressive correlation of the residuals $\epsilon$. $\sigma_\mathrm{proc}$ is the process error standard deviation. $\lambda$ represents the population growth rate per time step.

The mostly weakly-informative priors I'm using are:
\begin{align*}
b &\sim \mathrm{Uniform}(-1.2, 1.2)\\
\lambda &\sim \mathrm{Normal}(0, 20^2)\\
\nu &\sim \mathrm{Exponential}(0.01)\\
\phi &\sim \mathrm{Truncated\mhyphen Normal}(0, 1, \mathrm{min.} = -0.99, \mathrm{max.} = 0.99)\\
\sigma_\mathrm{proc} &\sim \mathrm{Half\mhyphen Cauchy} (5).
\end{align*}

Priors justification:

\begin{itemize}
\item $b$ is bounded just past stationary so we can detect if they are non-stationary will keeping the sampler from wandering off

\item $\lambda$ prior (variance = 400) is basically uninformative within the range of expected values for population growth... it allows even a X probability at a value of X

\item $\nu$ is based on \citet{fernandez1998}; they chose a more informative 0.1 value, we chose a less informative 0.01 and justify it based on performance in supplemental figure X (gives X probability of value less than 10) but constrains the sampling somewhat since above 40 or 50 the shape of the t distribution is almost identical and data are not usually informative at these data quantities... sampler would head off to infinity otherwise

\item $\phi$ given a standard deviation of 1 given prior analyses which suggest autocorrelation in these datasets is minimal; prior is weak enough to allow high or low values, but given little information will stay near our expectation; similar approach in \citep{thorson2014a}

\item $\sigma_\mathrm{proc}$ can be justified based on \citet{gelman2006c} and the expected range of this variable in nature from previous studies \citep[e.g.][]{connors2014}
\end{itemize}

Using Stan 2.4.0 \citep{stan-manual2014}, and R 3.1.1 \citep{r2014} I'm starting with 4 chains and 2000  iterations with the first 1000 as warmup (i.e.\ 4000 total samples). If rhat is greater than 1.05 for any parameter or the minimum effective sample size is less than 200 for any parameter then I double both the total iterations and warmup and run again. These conditions were met by 8000 iterations (16000 total samples) in all cases and in almost all cases (X\%) within the initial 2000 iteration case (i.e.\ a total of 8000 samples).

Then, I again tried allowing for a specified level of observation error. I'm still running that now for just the mammals. From before, I believe this will reduce the probability of heavy tails a bit for many of them, but the vast majority that had heavy tails would still be considered heavy tailed. This does reduce substantially many of the bird heavy tails though.

(See the simulation results... I don't think this has the intended benefit and can probably be left out of the main results.)

When I incorporate observation error, the model is:
\begin{align*}
U_t &= \lambda + b U_{t-1} + \epsilon_t\\
x_t &\sim \mathrm{Normal}(U_t, \sigma_\mathrm{obs}^2)\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}_\nu(\phi \epsilon_{t-1}, \sigma_\mathrm{proc}^2),
\end{align*}

\noindent
where $U$ represents the unobserved state vector, and $\sigma_\mathrm{obs}$ represents the standard deviation of observation error (on a log scale), which was set at 0.2.

%Also, we may not be crazy seeing the heavier tails for heavier longer lived animals. There's some precedence for that in a paper that found more red noise (longer term fluctuations dominating the variance and therefore making the CV grow with time) in large-bodied animals. They attributed it to a mismatch between dynamics and the observation scale. They used a different life-history dataset too. But, there are other systematic differences with the body size-nu relationship that make this sketchy for us. E.g.\ taxonomic orders and interval frequency of data collection vary systematically and in clumps with body size (not shown in the below figure). And basically the only stuff with very \textit{low} probability of heavy tails are the the small ones with generation times less than a year and small collection intervals. It's basically a step function. We may not be able to tease that all apart... at least not in this paper.

\subsection*{Data selection}

Filtering rules for the GPDD:
We removed all populations with zeros (can't be fit with a Gompertz), all populations with uneven sampling intervals (end date minus start date for each step), and all populations with population estimates listed as ``harvest'' as well as those listed as from two fur trading sources (give details).
Then, we removed any populations with 5 or more identical values in a row since these suggest either recording error or extrapolation between two observations.
Then, we filled in all missing time steps with NAs and imputed single missing values with the geometric mean of the previous and following values (linear on the log scale that the Gompertz operates on). This leaves NAs that were recorded in successive time steps.
Then, we went through each population time series and found the longest window of abundance with at least 25 time steps. If there were any with multiple windows of identical length, we took the most recent window.
Finally, we removed main\_id 20531, which is a duplicate of 10139 (a heron population) and main\_id 10008, which had obvious errors in the data recording (a water vole population). We provide a supplemental figure of all the time series included in our analysis and indicate which values were imputed (\percImputedPops\% of populations had at least one point imputed and only \percImputedPoints\% of the total observations were imputed) (Fig.\ SX).

\subsection*{Simulation testing the model}

Throughout all of this --- show that the model, if anything, under-predicts heavy tails but with sufficient data is unbiased.

2 parts: how many samples from the true population t distribution do you need to detect low nu? And, given that you have a set of deviations in which nu is detectable (effective nu is within 0.5 CV of true nu), can the more complex Gompertz still capture this?

First part: We drew from t distributions with different nu values and mean of 0, scale of 1. We started with 1600 samples and then fitted again at the first 800, 400, 200,  100, 50, 25. Each time we recorded the nu posterior.

Second part: To generate a series of process deviations with an effective nu approximately equal to the true nu, we generated process deviation sets repeatedly and estimated the mean, scale, and nu values each time. We recorded when the estimated nu was within 0.5 CVs of the true nu and used this set of random seed values in our Gompertz simulation. We then fit AR1 Gompertz models to the simulated datasets with all parameters (except nu) set near the median values estimated in the GPDD.

\section*{Results}

Main text figures:

\begin{enumerate}
\item illustrate $\nu$ and show simulated Gompertz time series with these levels of $\nu$ (include probability density at some extreme deviation in the caption or on figure for different $\nu$ values

\item a figure showing $\nu$ estimates within classes and  orders, ordered by decreasing $\nu$ estimate to illustrate percentage of populations, grouped by order in a phylogenetically related way include some silhouettes

\item 5--10 example time series with heavy tails from the GPDD; highlight the ``black swans'' with, say, red dots (perhaps also highlighting if they are up or down swings); pick some \textit{very} heavy tails and some moderately heavy tails. Show the $\nu$ estimate somewhere in each panel

\item something looking at hypotheses for what's associated with heavy tails: data collection method, mid date of time series (or when it started?), time series length, AR1, density dependence, lambda, body size for mammals, IUCN status? Most likely a grid of scatter plots coloured by taxonomic class.... or rows by taxonomic class and colours by order. (also ratio of generation time to collection window) (possibly, some of these will go in supplement with interesting ones in main paper).

\end{enumerate}

Main text tables:

\begin{enumerate}
\item Table summarizing hypotheses for why we might see heavy tails
\item Pick some example populations (say the ones in the time series example plot) and show the nu estimate, reference, and a column digging into any known caveats or reasons
\end{enumerate}

Supplemental figures:

\begin{enumerate}

\item illustration of how the exponential prior slightly constrains the sampling of $\nu$\ldots illustrating what the prior probability of heavy tails is given uninformative data (about 5\% probability $\nu < 10$.

\item plots of the priors overlayed with estimates across populations for all parameters

\item cross plot of $\nu$ estimates from the models: no density dependence, add density dependence, add AR1, add assumed observation error

\item time series plots of all heavy-tailed populations

\item simulation testing of detecting nu given reduced sampling from true distribution; show repeated estimates of nu with different sample sizes, and show a few example samples with different observation window lengths and highlight the deviations that are beyond the 0.001 and 0.999 probabilities for a normal distribution.

\item simulation testing of the Gompertz model with process error deviations fixed to have an effective $\nu$ estimate at the true level --- this asks how well is the Gompertz model able to partition the various parameters in a precise and unbiased way given that we know the heavy tails are there (also include a scenario with a massive black swan to show how it behaves there)

\end{enumerate}

\section*{Discussion}

- why we think we see them in some taxa but not others (observation scale dynamics mismatch; observation errors...)

Ways forward:

- hierarchical modelling of nu and other parameters

- better datasets for specific taxa (e.g.\ fish recruitment)

- joint prior on scale and nu parameters

- diving much deeper into what heavy tails are associated with

- do heavy tails have conservation importance?

- can we forecast their probability and develop correlates?

- how can we detect them relatively quickly after they happen?

- how often are heavy tails real? observation error, recording errors...

- what does this mean for how we model time series (should we consider using the t as a default even if we fix nu?)

- what does this mean for conservation management (probability of crash  might be higher than our typical models tell us)

\bibliography{/Users/seananderson/Dropbox/tex/jshort.bib,/Users/seananderson/Dropbox/tex/ref3.bib}

\clearpage

\section{Tables}

\LTcapwidth=\textwidth
\bibpunct{}{}{;}{a}{}{;}

\begin{small}
\begin{longtable}{>{\RaggedRight}m{2.0cm}>{\RaggedRight}p{3.0cm}>{\RaggedRight}p{7.0cm}>{\RaggedRight}p{2.0cm}}

\caption{TODO}\\

\toprule
Time series & Species & Black swan description & Reference \\
\midrule

\includegraphics[width=2cm]{sparks/6528} &
Shag (\textit{Phalacrocorax aristotelis}), Northumberland &
Shortage of nest sites reduced productivity; red-tide event in 1968 caused extreme mortality; no longer a nest shortage; population rapidly increased. &
\citep{potts1980}\\

%\includegraphics[width=2cm]{sparks/9675} &
%Carrot fly (\textit{Psila rosae}, Finland &
%TODO &
%\citep{markkula1965}\\

%\includegraphics[width=2cm]{sparks/10139} &
%Grey heron (\textit{Ardea cinerea}), UK &
%TODO &
%\citep{stafford1971}\\

\includegraphics[width=2cm]{sparks/1235} &
Wren (\textit{Troglodytes troglodytes}), UK &
TODO &
REF \\

%\includegraphics[width=2cm]{sparks/20580} &
%Chamois, \textit{Rupicapra rupicapra}, Switzerland &
%TODO &
%\citep{brook2006a}\\

\includegraphics[width=2cm]{sparks/5019} &
Barbary macaque, TODO &
TODO &
REF\\

\bottomrule
\label{tab:sparks}
\end{longtable}
\end{small}

\clearpage

\include{stat-table}
%\bibpunct{}{}{;}{a}{}{;}
%\begin{footnotesize}
%
%\renewcommand*{\arraystretch}{1.5}
%\begin{longtable}{>{\RaggedRight}p{3.5cm}>{\RaggedRight}p{8.2cm}>{\RaggedRight}p{2.2cm}}
%
%\caption{Description of black swans in GPDD}\\
%\toprule
%
%\midrule
%
%\bottomrule
%\label{tab:causes}
%\end{longtable}
%\end{footnotesize}
%\bibpunct{(}{)}{;}{a}{}{;}

\section{Figures}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.6\textwidth]{t-nu.pdf}
\caption{A figure illustrating $\nu$. Maybe put some example (Gompertz) time series to the right with the same seed and different $\nu$ values?}
\label{default}
\end{center}
\end{figure}

\clearpage

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{gomp-nu-ar1.pdf}
\caption{Estimate of $\nu$ from the Gompertz model (with estimated AR1 autocorrelation of the residuals but no observation error). Shown are median (points), quartiles (mid line segments) and 90\% quantiles (widest line segments) of the posterior.}
\label{fig:nu-coefs}
\end{center}
\end{figure}

\clearpage

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{gomp-ar1-p20-dot-order.pdf}
\caption{Probability $\nu<20$, i.e.\ one way of asking what the probability is of heavy tails. Dots represent populations. Something like this could be made more pretty and sorted in some appropriate way\ldots maybe with little silhouettes for the animals. Looks pretty similar for $\nu < 10$, with the probabilities a bit lower. There is some vertical jitter added. Could do this and have another subpanel that breaks down the mammals by family, since there's the most interesting stuff happening there and there are lots of them... or just colour the dots by family within each panel.}
\label{fig:dots}
\end{center}
\end{figure}

%\clearpage
%
%\begin{figure}[htbp]
%\begin{center}
%\includegraphics[width=\textwidth]{gomp-ar1-p20-hist.pdf}
%\caption{Histogram of the same data as Figure~\ref{fig:dots} but lumped by class. Probably can't justify including this as well as the previous two figures unless there's some way we can make this more informative.}
%\label{default}
%\end{center}
%\end{figure}

%\clearpage

%\begin{figure}[htbp]
%\begin{center}
%\includegraphics[width=0.9\textwidth]{effect-of-ar1-on-nu.pdf}
%\caption{The effect of assuming no autocorrelation of residuals (vertical axis) vs.\ estimating autocorrelation of residuals (horizontal axis). I don't see a systematic effect. These are medians and interquartile ranges of the posterior. The colour shows $\phi$ --- the parameter estimating the lag-1 correlation of the residuals. The diagonal line illustrates 1:1. A few of these look strange (ones with $\phi$ near zero but far from the 1:1 line). It's possible the chains haven't converged sufficiently in these cases in one of the model versions. I will check what's going on and run more iterations in a final run if needed.}
%\label{default}
%\end{center}
%\end{figure}

\clearpage

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{nu-covariates.pdf}
\caption{(Also see the next figure.) A number of hypotheses for what could be driving $\nu$ values. From top left to bottom right: $\phi$ (AR1 coefficient), $\log \sigma_\mathrm{proc}$ (log process error standard deviation), $N$ (time series length), $b$ (density dependence parameter in the Gompertz model; 1 is density independent and 0 is maximally density dependent), $\lambda$ (growth parameter from Gompertz model), $\log M$ (log adult body mass in grams; just for mammals and from the PanTHERIA database). All dots show median estimates and vertical line segments show interquartile range of posterior for the $\nu$ values. I don't generally see a whole lot going on here besides taxonomic differences but I'm open to some eye fuzzing and other ideas. If we want to, I was playing with a hierarchical beta regression with different levels for the taxonomic orders explaining the variation in p(nu $<$ 10) or these data\ldots but hypothesis generation with some visuals might be enough for this paper.}
\label{default}
\end{center}
\end{figure}

\clearpage

%\begin{figure}[htbp]
%\begin{center}
%\includegraphics[width=\textwidth]{nu-covariates-p.pdf}
%\caption{Same as previous figure but the y axis is p(nu $<$ 20), which I think might be better to use\ldots, in reality this kind of coarse dichotomy I think is more realistic. If p(nu $<$ 20) is high (say above 0.5) then there's a good chance the tails are heavy. Otherwise, we just don't know.}
%\label{default}
%\end{center}
%\end{figure}

%\begin{figure}[htbp]
%\begin{center}
%\includegraphics[width=0.7\textwidth]{p10-mammals-cross.pdf}
%\caption{Probability that nu $<$ 10 vs.\ log10(body mass) and log10(years at sexual maturity).}
%\label{default}
%\end{center}
%\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{ts-gpdd-heavy-eg-log10-no-fur.pdf}
\caption{Time series for populations with p(nu $<$ 10) $>$ 0.5. Ordered by increasing p(nu$<$10) (this value is shown pasted onto the front of each panel label). These are shown on a log10 y axis. There are a few funny time series we may want to look into and consider rules to remove. Colours show taxonomic class. This is for our understanding and a version of this could be a supplemental figure. But, we could take maybe 6 to 8 of these and use them as illustrations in the paper with the heavy tailed deviations highlighted. Could also probably dig into the literature on some of these and summarize in a table if we narrowed it down a bit. Lots of minks, beavers, and bears.}
\label{default}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{t-dist-sampling-sim-prior-exp0point01.pdf}
\includegraphics[width=\textwidth]{t-dist-sampling-sim-sigma-prior-exp0point01.pdf}
\caption{TODO --- also maybe show the different exponential priors.}
\label{default}
\end{center}
\end{figure}

\clearpage

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=1.1\textwidth]{sim-gompertz.pdf}
\caption{\textbf{TODO Caption doesn't match new figure.} Results from a simulation test of the estimation model. Dashed horizontal lines are the true population value. True values were chosen to represent approximately the median values as estimated from the GPDD. ``Iteration'' refers to a stochastic draw from the true population distribution and a model fitting. The panels show true values of $\nu = $ 2.1, 5, 15, and 200 (very heavy, heavy, slightly heavy, and not heavy tailed). The panels from top to bottom show nu (red), process error standard deviation (blue), growth parameter (green), density dependence parameter (purple), and autocorrelation parameter (orange). The larger blocks of panels show four scenarios: (1) N = 100, (2) N = 50, (3) N = 50 with ignored observation standard deviation of 0.2, and (4) a model where the model also allows for observation standard deviation of 0.2
In general, this looks fairly good, with the caveat that if the tails are heavy, the model tends to overestimate process error standard deviation and underestimate just how heavy the tails are --- it's a bit biased towards calling heavy tails not as heavy as they are. Also, density independence gets consistently underestimated (purple). Observation error erodes the heavy-tail signal unless the tails are very heavy but doesn't tend to create false positives. Trying to allow for assumed observation error doesn't do much good besides fixing some very subtle bias in the process error estimate when the tails aren't heavy. We'd need a larger observation error to see this clearly. Line segments are 90\% quantiles of the posterior and the dots are the medians.}
\label{default}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{check-sim-box.pdf}
\caption{Same data as previous figure but here represented as probability that nu is less than 10. Basically, this is looking at sensitivity/specificity. So, if you take something like a p = 0.5 that nu $<$ 10 as a cutoff, then you very rarely if ever get a false positive. However, the method doesn't \textit{always} detect heavy tails when they're there\ldots although in our data there are often tails that are much heavier than the nu = 2.5 used here.}
\label{default}
\end{center}
\end{figure}

% \begin{figure}[htbp]
%\begin{center}
%\includegraphics[width=0.6\textwidth]{effect-of-obs-on-nu.pdf}
%\caption{Effect of assuming that observation error is normally distributed (on a log scale) with standard deviation of 0.2.}
%\label{default}
%\end{center}
%\end{figure}
%

%\end{spacing}
\end{document}

little evidence for benefit of including annual climate or climate extremes with these data:
\citep{gregory2010}
and little evidence for allee effect with them

the natural logorithm versino of the logistic is often convenient for parameter estimation
\citep{valpine2002} (and tthey cite a couple people too)

### 2014-09-22 stat-table.tex
Initial commit

% latex table generated in R 3.1.1 by xtable 1.7-3 package
% Fri Sep 19 11:41:03 2014
\begin{table}[ht]
\centering
\begin{tabular}{lrrrrrrr}
  \hline
taxonomic\_class & n & n\_orders & n\_species & median\_N & max\_N & n\_interpolated & n\_zero\_sub \\ 
  \hline
Aves & 227 &  15 & 113 &  26 & 117 &  68 &  32 \\ 
  Insecta & 182 &   7 &  91 &  25 &  87 &  26 &  55 \\ 
  Mammalia & 125 &   8 &  51 &  28 & 106 &  18 &  21 \\ 
  Osteichthyes & 109 &   6 &  36 &  26 &  67 &  14 &   4 \\ 
  Crustacea &  13 &   2 &   2 &  28 &  33 &   1 &   0 \\ 
  Chondrichtyhes &   1 &   1 &   1 &  20 &  20 &   1 &   0 \\ 
  Gastropoda &   1 &   1 &   1 &  21 &  21 &   0 &   0 \\ 
   \hline
\end{tabular}
\end{table}

### 2014-09-22 0-make-data.R

@@ -1,0 +1,365 @@
+# this file makes the gpdd and mammals data
+# some of the gpdd processing happens in another folder but gets sources from
+# here
+# note that there's a shell script in the gpdd folder to turn the original
+# Access database files into the .csvs
+# not running it here because I haven't put the appropriate binary in R's
+# path
+
+library(dplyr)
+
+# GPDD:
+setwd("gpdd")
+source("make-gpdd.R")
+setwd("..")
+
+# Data filtering (TODO note)
+gpdd <- readRDS("gpdd.rds")
+# gpdd <- subset(gpdd, taxonomic_phylum %in%
+#     c("Chordata", "Arthropoda", "Mollusca", "Annelida", "Echinodermata"))
+#
+# for checking later:
+gpdd$population_untransformed_original <- gpdd$population_untransformed
+
+# There are a couple that need this
+# and tripped me up before:
+gpdd <- arrange(gpdd, main_id, series_step)
+
+gpdd <- subset(gpdd, sampling_protocol != "Harvest")
+
+# additional plan based on Brook et al. 2006:
+# - fill in single zeros with lowest value
+# - turn multiple zeros into multiple NAs but don't necessarily remove; the windowing procedure might find a usable length
+# - allow down to 15 time steps not just 25
+# - don't exclude based on tax. class
+# - only keep those with full species name
+# - USE THE UNTRANSFORMED DATA!!!!!
+# - take 10^ of all those columns with negative values or "log" in notes
+# - bring in the correlates data from Brook e al. 2006
+
+
+
+# hidden harvests: the fur trade stats:
+gpdd <- plyr::ddply(gpdd, "main_id", function(y) {
+  fur <- grepl("harvest", y$ref[1])
+  if(!fur) y
+})
+gpdd <- plyr::ddply(gpdd, "main_id", function(y) {
+  fur <- grepl("commerce", y$ref[1])
+  if(!fur) y
+})
+
+# now a big one: some have missing time steps
+# first go through and expand missing rows to NAs:
+gpdd <- plyr::ddply(gpdd, "main_id", function(y) {
+  steps_df <- data.frame(series_step =
+      min(y$series_step, na.rm = TRUE):max(y$series_step, na.rm = TRUE),
[…]

### 2014-09-22 1-compile-models.R

@@ -1,0 +1,207 @@
+# this file compiles the Stan models, nothing more
+
+library(rstan)
+
+stan_model <-
+'data {
+  int<lower=0> N; // rows of data
+  vector[N] y; // vector to hold observations
+  real<lower=0> nu_rate; // rate parameter for nu exponential prior
+  real b_lower;
+  real b_upper;
+}
+parameters {
+  real lambda;
+  real<lower=b_lower, upper=b_upper> b;
+  real<lower=0> sigma_proc;
+  real<lower=2> nu;
+}
+model {
+  nu ~ exponential(nu_rate);
+  lambda ~ normal(0, 10);
+  //b ~ normal(b_mu, b_sigma);
+  sigma_proc ~ cauchy(0, 5);
+  for (i in 2:N) {
+    y[i] ~ student_t(nu, lambda + b * y[i-1], sigma_proc);
+  }
+}
+'
+
+stan_gomp <- stan_model(model_code = stan_model)
+saveRDS(stan_gomp, file = "stan-gomp.rds")
+
+# with autocorrelation:
+
+stan_model <-
+  'data {
+  int<lower=0> N; // rows of data
+  vector[N] y; // vector to hold observations
+  real<lower=0> nu_rate; // rate parameter for nu exponential prior
+  real b_lower;
+  real b_upper;
+}
+parameters {
+  real lambda;
+  real<lower=b_lower, upper=b_upper> b;
+  real<lower=0> sigma_proc;
+  real<lower=2> nu;
+  real<lower=-0.99, upper=0.99> phi;
+}
+model {
+  nu ~ exponential(nu_rate);
+  lambda ~ normal(0, 10);
+  sigma_proc ~ cauchy(0, 5);
+  phi ~ normal(0, 0.5);
+  for (i in 3:N) {
+    y[i] ~ student_t(nu,
+            (lambda + b * y[i-1]) + phi * (y[i-1] - (lambda + b * y[i-2])),
+            sigma_proc);
+  }
+}
+'
+stan_gomp_ar1 <- stan_model(model_code = stan_model)
+saveRDS(stan_gomp_ar1, file = "stan-gomp-ar1.rds")
+
+# with fixed observation error:
+
+stan_model <-
+  'data {
+  int<lower=0> N; // rows of data
+  vector[N] y; // vector to hold observations
+  real<lower=0> nu_rate; // rate parameter for nu exponential prior
[…]

### 2014-09-22 2-run-model-logistic.R

@@ -1,0 +1,46 @@
+# this file runs the models with a logistic growth model
+
+stan_logistic <- readRDS("stan-logistic.rds")
+gpdd <- readRDS("gpdd-clean.rds")
+library(rstan)
+
+main_id_vec <- unique(gpdd$main_id)
+
+out_gomp <- plyr::d_ply(subset(gpdd, main_id %in% main_id_vec), "main_id",
+  function(x) {
+
+    max_rhat <- 999
+    min_neff <- 0
+    iterations<- 2000
+    warmup <- 1000
+    file_base <- "/global/scratch/anderson/heavy/logistic-5.0/sm-"
+
+    if(!file.exists(paste0(file_base, unique(x$main_id), ".rds"))) {
+      print("already done")
+      print(unique(x$main_id))
+      # TODO scale the data to max of 1 and calculate the log-diff rate column here
+      x$r_obs <- c(NA, diff(log(x$population_untransformed)))
+      x$population_untrans_scaled <-
+        x$population_untransformed / max(x$population_untransformed)
+      N <- nrow(x) - 1
+
+      while((max_rhat > 1.05 | min_neff < 200) & iterations <= 4000) {
+        sm <- sampling(stan_logistic,
+          data = list(N = N, r_obs = x$r_obs[-1], K_upper = 2, r_upper = 20,
+            nu_rate = 0.01, Nt = x$population_untrans_scaled[-N]),
+          pars = c("r", "K", "nu", "sigma_proc"), iter = iterations,
+          chains = 4, warmup = warmup)
+
+        # check:
+        max_rhat <- max(summary(sm)$summary[, "Rhat"])
+        min_neff <- min(summary(sm)$summary[, "n_eff"])
+
+        warmup <- warmup * 2
+        iterations <- iterations * 2
+      }
+      saveRDS(sm, file = paste0(file_base, unique(x$main_id), ".rds"))
+      sink(paste0(file_base, unique(x$main_id), ".txt"))
+      print(sm)
+      sink()
+    }
+  })

### 2014-09-22 2-run-models-ar1-obs.R

@@ -1,0 +1,46 @@
+# this file runs the Gompertz model with AR1 residual coefficient
+# but also adds an assumed quantity of observation error
+# it does not estimate the observation error
+
+stan_gomp_ar1_obs <- readRDS("stan-gomp-ar1-obs.rds")
+gpdd <- readRDS("gpdd-clean.rds")
+library(rstan)
+
+# Mammals only:
+main_id_vec <- unique(subset(gpdd, taxonomic_class == "Mammalia")$main_id)
+#main_id_vec <- sample(main_id_vec, 4)
+
+out_gomp <- plyr::d_ply(subset(gpdd, main_id %in% main_id_vec), "main_id",
+  function(x) {
+    max_rhat <- 999
+    min_neff <- 0
+    iterations <- 2000
+    warmup <- 1000
+    if(!file.exists(paste0("/global/scratch/anderson/heavy/ar1-obs0.2/sm-",
+      unique(x$main_id), ".rds"))) {
+      while((max_rhat > 1.1 | min_neff < 100) & iterations <= 4000) {
+        sm <- sampling(stan_gomp_ar1_obs,
+          data = list(N = nrow(x), y = log(x$population_untransformed),
+            nu_rate = 0.01,
+            b_lower = -1, b_upper = 2, sigma_obs = 0.2),
+          pars = c("lambda", "sigma_proc", "nu", "b", "phi"), iter = iterations,
+          chains = 4, warmup = warmup)
+
+        # check:
+        max_rhat <- max(summary(sm)$summary[, "Rhat"])
+        min_neff <- min(summary(sm)$summary[, "n_eff"])
+
+        warmup <- warmup * 2
+        iterations <- iterations * 2
+      }
+
+      if(file.exists("/global/scratch/anderson/heavy/ar1-obs0.2/")) {
+        saveRDS(sm, file = paste0("/global/scratch/anderson/heavy/ar1-obs0.2/sm-",
+          unique(x$main_id), ".rds"))
+        sink(paste0("/global/scratch/anderson/heavy/ar1-obs0.2/sm-",
+          unique(x$main_id), ".txt"))
+        print(sm)
+        sink()
+      }
+    }
+  })

### 2014-09-22 2-run-models-ar1.R

@@ -1,0 +1,49 @@
+# this file runs the models with an AR1 coefficient on the residuals
+
+stan_gomp_ar1 <- readRDS("stan-gomp-ar1.rds")
+gpdd <- readRDS("gpdd-clean.rds")
+library(rstan)
+
+#main_id_vec <- unique(gpdd$main_id)
+
+out_gomp <- plyr::d_ply(subset(gpdd, main_id %in% main_id_vec), "main_id",
+  function(x) {
+
+    max_rhat <- 999
+    min_neff <- 0
+    iterations<- 1000
+    warmup <- 500
+    file_base <- "/global/scratch/anderson/heavy/ar1-5.0/sm-"
+
+    if(!file.exists(paste0(file_base, unique(x$main_id), ".rds"))) {
+      print("already done")
+      print(unique(x$main_id))
+      while((max_rhat > 1.05 | min_neff < 200) & iterations <= 4000) {
+
+        sm <- sampling(stan_gomp_ar1,
+          data = list(N = nrow(x), y = log(x$population_untransformed),
+            nu_rate = 0.01,
+            b_lower = -1, b_upper = 2),
+          pars = c("lambda", "sigma_proc", "nu", "b", "phi"), iter = iterations,
+          chains = 4, warmup = warmup)
+
+        # check:
+        max_rhat <- max(summary(sm)$summary[, "Rhat"])
+        min_neff <- min(summary(sm)$summary[, "n_eff"])
+
+        warmup <- warmup * 2
+        iterations <- iterations * 2
+      }
+      saveRDS(sm, file = paste0(file_base, unique(x$main_id), ".rds"))
+      sink(paste0(file_base, unique(x$main_id), ".txt"))
+      print(sm)
+      sink()
+    }
+  })
+
+# trash ones we no longer want:
+#f <- list.files("/global/scratch/anderson/heavy/ar1-4.0", pattern = "*\\.rds")
+#fn <- sub("sm-([0-9]+)\\.rds", "\\1", f)
+#fn[!fn %in% gpdd$main_id]
+#file.rename(
+

### 2014-09-22 2-run-models-base.R

@@ -1,0 +1,49 @@
+# this file runs the models *without* AR1 coefficient on the residuals
+
+stan_gomp <- readRDS("stan-gomp.rds")
+gpdd <- readRDS("gpdd-clean.rds")
+library(rstan)
+
+main_id_vec <- unique(gpdd$main_id)
+
+out_gomp <- plyr::d_ply(subset(gpdd, main_id %in% main_id_vec), "main_id",
+  function(x) {
+
+    max_rhat <- 999
+    min_neff <- 0
+    iterations<- 1000
+    warmup <- 500
+    file_base <- "/global/scratch/anderson/heavy/base-5.0/sm-"
+
+    if(!file.exists(paste0(file_base, unique(x$main_id), ".rds"))) {
+      print("already done")
+      print(unique(x$main_id))
+      while((max_rhat > 1.05 | min_neff < 200) & iterations <= 4000) {
+
+        sm <- sampling(stan_gomp,
+          data = list(N = nrow(x), y = log(x$population_untransformed),
+            nu_rate = 0.01,
+            b_lower = -1, b_upper = 2),
+          pars = c("lambda", "sigma_proc", "nu", "b"), iter = iterations,
+          chains = 4, warmup = warmup)
+
+        # check:
+        max_rhat <- max(summary(sm)$summary[, "Rhat"])
+        min_neff <- min(summary(sm)$summary[, "n_eff"])
+
+        warmup <- warmup * 2
+        iterations <- iterations * 2
+      }
+      saveRDS(sm, file = paste0(file_base, unique(x$main_id), ".rds"))
+      sink(paste0(file_base, unique(x$main_id), ".txt"))
+      print(sm)
+      sink()
+    }
+  })
+
+# trash ones we no longer want:
+#f <- list.files("/global/scratch/anderson/heavy/ar1-4.0", pattern = "*\\.rds")
+#fn <- sub("sm-([0-9]+)\\.rds", "\\1", f)
+#fn[!fn %in% gpdd$main_id]
+#file.rename(
+

### 2014-09-22 3.0-test-t-sampling.R

@@ -1,0 +1,62 @@
+# this file tests the combination of N samples and degree of heavy tailedness
+# in terms of the frequency that nu is estimated correctly
+# this is part 1 of 2 simulation testing checks
+
+library(rstan)
+stan_t <- readRDS("stan-t.rds")
+
+# take a basic t- dist
+# and try and re-capture it with our prior at different nu values
+# use a ton of data (2000 points, 500 points, 100 points, 50 points)
+# show how probability of noticing goes down
+set.seed(123)
+
+N_vec <- c(1600, 800, 400, 200, 100, 50, 25)
+sigma_proc <- 1
+nu_vec <- c(2, 5, 10, 1e6)
+reps <- 1:20
+show_plot <- FALSE
+
+# test:
+#N_vec <- 50
+#nu_vec <- 100
+#reps <- 1:2
+
+out <- plyr::ldply(reps, function(k) {
+  out_N <- plyr::ldply(seq_along(nu_vec), function(i) {
+
+    y <- metRology::rt.scaled(max(N_vec), df = nu_vec[i], mean = 0, sd = sigma_proc)
+
+    if(show_plot) {
+      plot(y, ylim = c(-6, 6), xaxs = "i", log = "x")
+      abline(v = N_vec, col = "red", lwd = 2)
+      rect(1, -100, max(N_vec) + 10, qnorm(0.001), col = "#00000030", border = NA)
+      rect(1, 100, max(N_vec) + 10, qnorm(0.999), col = "#00000030", border = NA)
+      #rect(1, -100, max(N_vec) + 10, qnorm(0.01), col = "#00000030", border = NA)
+      #rect(1, 100, max(N_vec) + 10, qnorm(0.99), col = "#00000030", border = NA)
+    }
+
+    out_nu <-  plyr::ldply(seq_along(N_vec), function(j) {
+      sm <- sampling(stan_t,
+        data = list(N = N_vec[j], y = y[1:N_vec[j]], nu_rate = 0.05),
+        pars = c("nu", "sigma_proc"), iter = 1000,
+        chains = 4, warmup = 500)
+      max_rhat <- max(summary(sm)$summary[, "Rhat"])
+      min_neff <- min(summary(sm)$summary[, "n_eff"])
+      e <- extract(sm, pars = "nu")[[1]]
+      e_sig <- extract(sm, pars = "sigma_proc")[[1]]
[…]

### 2014-09-22 3.1-plot-sample-test.R

@@ -1,0 +1,29 @@
+# take the output from 3-test-t-sampling.R and plot it
+
+sample_t <- readRDS("sample-t-sim-check.rds")
+
+library(ggplot2)
+p <- ggplot(sample_t, aes(iter, 1/med_nu)) + geom_pointrange(aes(ymin = 1/l_nu, ymax = 1/u_nu)) +
+  facet_grid(nu_true~N) +
+  scale_y_continuous(breaks = c(0, 0.1, 0.2, 0.333, 0.5),
+    labels = c("Infinity", "10", "5", "3", "2")) +
+  geom_hline(aes(yintercept = 1/nu_true), col = "red") +
+  xlab("Iteration") + ylab(expression(t-distribution~degrees~of~freedom~(nu)))
+
+  ggsave("t-dist-sampling-sim-prior-exp0.01.pdf", width = 8, height = 5)
+
+p <- ggplot(sample_t, aes(iter, med_sigma)) + geom_pointrange(aes(ymin = l_sigma, ymax = u_sigma)) +
+  facet_grid(nu_true~N) +
+  geom_hline(aes(yintercept = 1), col = "red") +
+  xlab("Iteration") + ylab("Scale parameter")
+
+  ggsave("t-dist-sampling-sim-sigma-prior-exp0.01.pdf", width = 8, height = 5)
+
+sample_t <- readRDS("sample-t-sim-check-prior-exp-0.05.rds")
+p <- ggplot(sample_t, aes(iter, 1/med_nu)) + geom_pointrange(aes(ymin = 1/l_nu, ymax = 1/u_nu)) +
+  facet_grid(nu_true~N) +
+  scale_y_continuous(breaks = c(0, 0.1, 0.2, 0.333, 0.5),
+    labels = c("Infinity", "10", "5", "3", "2")) +
+  geom_hline(aes(yintercept = 1/nu_true), col = "red") +
+  xlab("Iteration") + ylab(expression(t-distribution~degrees~of~freedom~(nu)))
+  ggsave("t-dist-sampling-sim-prior-exp0.05.pdf", width = 8, height = 5)

### 2014-09-22 3.2-get-effect-nu-seeds.R

@@ -1,0 +1,52 @@
+library(rstan)
+stan_t <- readRDS("stan-t.rds")
+
+get_effective_nu_seeds <- function(nu_true = 5, cv = 0.5, N = 50,
+  seed_N = 20, start_seed = 0, sigma_proc = 0.45) {
+
+  N <- N
+  sigma_proc <- sigma_proc
+  nu <- nu_true
+  i <- start_seed
+  cv <- cv
+  j <- 1
+  seeds <- rep(0, length = seed_N)
+  nu_ests <- rep(0, length = seed_N)
+
+  while(seeds[length(seeds)] == 0) {
+    nu_close <- FALSE
+    while(nu_close == FALSE) {
+      i <- i + 1
+      set.seed(i)
+      y <- metRology::rt.scaled(N, df = nu, mean = 0, sd = 0.45)
+      sm <- sampling(stan_t,
+        data = list(N = N, y = y, nu_rate = 0.01),
+        pars = c("sigma_proc", "nu", "mu"), iter = 800,
+        chains = 4, warmup = 400)
+      med_nu_hat <- median(extract(sm, pars = "nu")[[1]])
+      # 88 is a very weird draw with a giant deviation as the 2nd value
+      # which creates lambda plots that mess up all the axes
+      # so, skipping seed 88 for nu = 5, N = 50
+      if(med_nu_hat > (nu - cv) & med_nu_hat < (nu + cv) & i != 88) {
+        nu_close <- TRUE
+        seeds[j] <- i # a valid seed
+        nu_ests[j] <- med_nu_hat # capture the estimated nu
+        j <- j + 1
+      } else {
+        message(paste0("nu hat was ", med_nu_hat))
+      }
+    }
+  }
+  return(list(seeds = seeds, nu_ests = nu_ests))
+}
+
+nu_2_seeds_N50 <- get_effective_nu_seeds(seed_N = 20, nu_true = 2)
+nu_5_seeds_N50 <- get_effective_nu_seeds(seed_N = 20, nu_true = 5)
+
+nu_2_seeds_N100 <- get_effective_nu_seeds(seed_N = 20, nu_true = 2, N = 100)
+nu_5_seeds_N100 <- get_effective_nu_seeds(seed_N = 20, nu_true = 5, N = 100)
+
+save(nu_2_seeds_N50, nu_5_seeds_N50, nu_2_seeds_N100, nu_5_seeds_N100,
+  file = "nu_effective_seeds.rda")
+
+# now go remove all N50 iterations 2 and re-run

### 2014-09-22 3.3-test-gomp-models.R

@@ -1,0 +1,135 @@
+# this file runs some simulation tests of the Stan model
+# I have set the parameter values to ballpark median values from the GPDD
+# this file tests the ability for the Gompertz to perform well
+# when we feed it process deviations that have been selected to have the true
+# population value of nu (within a certain CV)
+
+library(rstan)
+stan_gomp_ar1_obs <- readRDS("stan-gomp-ar1-obs.rds")
+stan_gomp_ar1 <- readRDS("stan-gomp-ar1.rds")
+
+sim_gomp <- function(id = 99, phi = -0.15, nu = 10, sigma_proc = 0.45,
+  sigma_obs_true = 0.001, sigma_obs_assumed = 0.001, N = 50, b = 0.65,
+  lambda = 1.3, y1 = 5.0, iterations = 2000, warmup = 1000, plot = FALSE,
+  max_iter = 8000, chains = 4, seed) {
+
+  file_base <- paste0("/global/scratch/anderson/heavy/sim2/sm-",
+    id, "-sigma_obs_true-", sigma_obs_true, "-sigma_obs_assumed-",
+    sigma_obs_assumed, "-N-", N, "-b-", b, "-sigma_proc-", sigma_proc,
+    "-nu-", nu, "-y1-", y1, "-lambda-", lambda, "-phi-", phi)
+
+  # simulate data:
+  y <- vector(length = N)
+  set.seed(seed)
+  proc_error <- metRology::rt.scaled(N, df = nu, mean = 0, sd = sigma_proc)
+  proc_error_ar1 <- vector(length = N)
+  proc_error_ar1[1] <- proc_error[1]
+  for(i in 2:N) {
+    proc_error_ar1[i] <- proc_error[i] + phi * proc_error[i-1]
+  }
+  for(i in 2:N) {
+    y[i] <- lambda + b * y[i-1] + proc_error_ar1[i-1]
+  }
+  y <- rnorm(N, mean = y, sd = sigma_obs_true)
+
+  if(plot == TRUE) plot(y)
+
+  # fit model:
+  min_neff <- 0
+  max_rhat <- 999
+
+  if(!file.exists(paste0(file_base, ".rds"))) {
+    while((max_rhat > 1.05 | min_neff < 200) & iterations <= max_iter) {
+      if(sigma_obs_assumed <= 0.01) { # use the faster Stan model without obs error
[…]

### 2014-09-22 3.4-plot-test-gomp-models.R

@@ -1,0 +1,126 @@
+# plot the simulation output from 3.3-test-gomp-models.R
+
+check_nu <- readRDS("check_nu.rds")
+
+check_nu <- mutate(check_nu, inverse_nu = 1/nu, inverse_nu_l = 1/nu_l,
+  inverse_nu_u = 1/nu_u, inverse_nu_true = 1/nu_true)
+
+id.vars <- c("id", "nu_true", "N", "sigma_obs_true", "sigma_obs_assumed")
+measure.vars <- c("inverse_nu", "b", "sigma_proc", "lambda", "phi")
+check_nu_est <- reshape2::melt(check_nu, id.vars = id.vars,
+  measure.vars = measure.vars)
+check_nu_l <- reshape2::melt(check_nu, id.vars = id.vars,
+  measure.vars = paste0(measure.vars, "_l"))
+check_nu_u <- reshape2::melt(check_nu, id.vars = id.vars,
+  measure.vars = paste0(measure.vars, "_u"))
+
+check_nu_l <- plyr::rename(check_nu_l, c("value" = "l"))
+check_nu_u <- plyr::rename(check_nu_u, c("value" = "u"))
+check_nu_est <- plyr::rename(check_nu_est, c("value" = "est"))
+
+check_nu_long <- data.frame(check_nu_l, est = check_nu_est[,"est"],
+  u = check_nu_u[,"u"])
+rm(check_nu_l, check_nu_u, check_nu_est)
+check_nu_long$variable <- as.character(check_nu_long$variable)
+check_nu_long$variable <- sub("([a-z]*)_l", "\\1",
+  check_nu_long$variable)
+
+## now bring in the true values for plotting horizontal lines:
+measure.vars <- paste0(measure.vars, "_true")
+check_nu_true <- reshape2::melt(check_nu, id.vars = id.vars,
+  measure.vars = measure.vars)
+check_nu_true <- plyr::rename(check_nu_true, c("variable" = "true_variable",
+    "value" = "true_value"))
+check_nu_long <- data.frame(check_nu_long,
+  true_value = check_nu_true[, "true_value"])
+check_nu_long$true_variable <- NULL # not needed
+
+limits <- data.frame(variable = c("inverse_nu", "sigma_proc", "lambda", "b", "phi"), ylim_l = c(0, 0, 0, -0.25, -1), ylim_u = c(0.5, 1.35, 12, 1, 1), stringsAsFactors = FALSE)
[…]

### 2014-09-22 4-extract-models.R

@@ -1,0 +1,70 @@
+# this file extracts posterior and diagnostic data from all the model runs
+
+gpdd <- readRDS("gpdd-clean.rds")
+
+library(rstan)
+library(dplyr)
+
+extract_model <- function(id, get_phi = FALSE,
+  folder = "/global/scratch/anderson/heavy/", subfolder = "base") {
+
+  sm <- readRDS(paste0(folder, subfolder, "/sm-", id, ".rds"))
+  samp <- extract(sm)
+  q <- lapply(samp, quantile, probs = c(0.05, 0.25, 0.5, 0.75, 0.95))
+  q <- do.call("rbind", q)
+
+  get_q <- function(par) {
+    x <- q[par, ]
+    names(x) <- paste0(par, "_", names(x))
+    as.data.frame(t(x))
+  }
+
+  lambda <- get_q("lambda")
+  nu <- get_q("nu")
+  b <- get_q("b")
+  sigma_proc <- get_q("sigma_proc")
+
+  if(get_phi) {
+    phi <- get_q("phi")
+  }
+
+  nu_samples <- samp$nu
+  p10 <- length(nu_samples[nu_samples < 10]) / length(nu_samples)
+  p20 <- length(nu_samples[nu_samples < 20]) / length(nu_samples)
+
+  sm_summ <- summary(sm)$summary
+  max_rhat <- max(sm_summ[, "Rhat"])
+  min_neff <- min(sm_summ[, "n_eff"])
+  nu_rhat <- sm_summ["nu", "Rhat"]
+  nu_neff <- sm_summ["nu", "n_eff"]
+
+  iter <- sm@stan_args[[1]]$iter
+
+  out <- data.frame(lambda, nu, b, sigma_proc)
+
+  if(get_phi) {
+    out <- data.frame(out, phi)
+  }
+
+  out <- data.frame(out, p10 = p10, p20 = p20, max_rhat = max_rhat,
+    min_neff = min_neff, nu_rhat = nu_rhat, nu_neff = nu_neff, iter = iter)
+
+  out$main_id <- id
+  names(out) <- sub("\\.", "", names(out))
+  out
+
+}
+
+ids <- unique(gpdd$main_id)
+#mammals_ids <- gpdd %>% filter(taxonomic_class == "Mammalia") %>%
+  #group_by(main_id) %>% summarise(id = main_id[1]) %>% select(main_id)
[…]

### 2014-09-22 5-shape-data.R

@@ -1,0 +1,111 @@
+# this file reads in the posterior data, filter it a bit, and joins it with
+# the original gpdd and mammal life history databases
+# it then does some reshaping for plotting purposes
+
+library(dplyr)
+
+gpdd <- readRDS("gpdd-clean.rds")
+gomp_hat_base <- readRDS("gomp_hat_base.rds")
+gomp_hat_ar1 <- readRDS("gomp_hat_ar1-4.0.rds")
+gomp_hat_ar1_obs <- readRDS("gomp_hat_ar1_obs0.2.rds")
+mammals <- readRDS("mammals.rds")
+
+gomp_hat_ar1 <- subset(gomp_hat_ar1,
+  main_id %in% gpdd$main_id)
+gomp_hat_base <- subset(gomp_hat_base,
+  main_id %in% gpdd$main_id)
+gomp_hat_ar1_obs <- subset(gomp_hat_ar1_obs,
+  main_id %in% gpdd$main_id)
+
+# back to reshaping
+
+# make some joings for looking at the effect of observation error
+# and estimating AR1 parameter:
+temp <- gomp_hat_base
+names(temp) <- paste0(names(temp), "_base")
+temp <- plyr::rename(temp, c("main_id_base" = "main_id"))
+ar1_vs_base <- plyr::join(gomp_hat_ar1, temp)
+rm(temp)
+
+temp <- gomp_hat_ar1
+names(temp) <- paste0(names(temp), "_ar1_no_obs")
+temp <- plyr::rename(temp, c("main_id_ar1_no_obs" = "main_id"))
+ar1_no_obs_vs_obs <- plyr::join(gomp_hat_ar1_obs, temp)
+rm(temp)
+
+# join in taxonomic data from the gpdd:
+lookup <- gpdd[,c("main_id", "common_name", "taxonomic_class", "taxon_name",
+  "taxonomic_order", "taxonomic_family", "dataset_length")]
+lookup <- lookup[!duplicated(lookup), ]
+
+gomp_hat_base <- plyr::join(gomp_hat_base, lookup)
+gomp_hat_ar1 <- plyr::join(gomp_hat_ar1, lookup)
+ar1_vs_base <- plyr::join(ar1_vs_base, lookup)
+ar1_no_obs_vs_obs <- plyr::join(ar1_no_obs_vs_obs, lookup)
+
+# for the paper:
[…]

### 2014-09-22 6-plot-eg-ts-gpdd.R

@@ -1,0 +1,25 @@
+heavy <- filter(gomp_hat_base, p10 > .5) %>%
+#   filter(fur == FALSE) %>%
+  select(main_id, common_name, p10, p20, nu_5, nu_25, nu_50, nu_75, nu_95)
+
+#gpdd_heavy <- filter(gpdd, main_id %in% heavy$main_id)
+#gpdd_heavy <- filter(gpdd_heavy, !grepl("harvest", ref))
+#gpdd_heavy <- filter(gpdd_heavy, !grepl("commerce", ref))
+
+
+heavy <-
+  plyr::join(heavy, gpdd) %>%
+  arrange(main_id, p10) %>%
+  mutate(id_name = paste(round(p10, 2), common_name, main_id))
+
+library(ggplot2)
+p <- ggplot(heavy, aes(series_step, population_untransformed, colour = taxonomic_class)) + geom_line() + facet_wrap(~id_name, scales = "free")
+ggsave("ts-gpdd-heavy-eg-base.pdf", width = 15, height = 12)
+
+p <- ggplot(heavy, aes(series_step, log10(population_untransformed), colour = taxonomic_class)) + geom_line() + facet_wrap(~id_name, scales = "free", ncol = 4)
+ggsave("ts-gpdd-heavy-eg-log10-base.pdf", width = 15, height = 17)
+
+heavy_table <- plyr::ddply(heavy, "main_id", function(x) x[1,]) %>%
+  arrange(id_name)
+
+write.csv(heavy_table, file = "heavy.csv", row.names = FALSE)

### 2014-09-22 6-plot-gomp-samples.R

@@ -1,0 +1,84 @@
+# this file makes a variety of plots of the model posteriors
+
+library(ggplot2)
+
+gomp_hat_ar1 <- filter(gomp_hat_ar1, taxonomic_class != "Crustacea")
+gomp_hat_ar1 <- filter(gomp_hat_base, !taxonomic_class %in% c("Crustacea", "Chondrichtyhes", "Gastropoda"))
+
+p <- ggplot(ar1_vs_base, aes(1/nu_50, 1/nu_50_base, col = phi_50)) + geom_point(cex = 3.5, alpha = 0.8) + geom_segment(aes(x = 1/nu_50, xend = 1/nu_50, y = 1/nu_25_base, yend = 1/nu_75_base), lwd = 0.5, alpha = 0.2) + geom_segment(aes(x = 1/nu_25, xend = 1/nu_75, y = 1/nu_50_base, yend = 1/nu_50_base), lwd = 0.5, alpha = 0.2)+ facet_wrap(~taxonomic_class) + geom_abline(intercept = 0, slope = 1, lty = 3) + xlab(expression(nu~with~AR1)) + ylab(expression(nu~without~AR1)) + coord_fixed() + theme_bw() +  scale_y_continuous(breaks = c(0, 0.1, 0.2, 0.333, 0.5), labels = c("Infinity", "10", "5", "3", "2")) +  scale_x_continuous(breaks = c(0, 0.1, 0.2, 0.333, 0.5), labels = c("Infinity", "10", "5", "3", "2")) + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank()) + labs(colour = expression(phi)) + scale_colour_gradient2(low = scales::muted("blue"), high = scales::muted("red"), mid = "grey")
+ggsave("effect-of-ar1-on-nu.pdf", width = 8, height = 8)
+# not very different!
+# switching to AR1 to base just in name for now...
+
+# and the effect of assumed observation error?
+p <- ggplot(ar1_no_obs_vs_obs, aes(1/nu_50, 1/nu_50_ar1_no_obs, colour = taxonomic_order)) + geom_point() + geom_abline(intercept = 0, slope = 1) + xlab(expression(1/nu~with~assumed~observation~error)) + ylab(expression(1/nu~without~observation~error)) + coord_fixed() + theme_bw()
+ggsave("effect-of-obs-on-nu.pdf", width = 7, height = 7)
+
[…]

### 2014-09-22 6-plot-p20-order.R

@@ -1,0 +1,8 @@
+# make a publication plot of p(nu < 20) by taxonomic order and class
+
+# plot_nu_panel <- function
+
+x <- filter(gomp_hat_ar1, taxonomic_class == "Mammalia") %>%
+  arrange(taxonomic_class, order_p20) %>%
+
+

### 2014-09-22 6-plot-sparks.R

@@ -1,0 +1,68 @@
+# make little time series sparklines (copyright Tufte) of the heavy-tailed
+# populations for a table
+
+
+#source("5-shape-data.R")
+
+heavy <- filter(gomp_hat_base, nu_50 < 20) %>%
+  select(main_id, common_name, p10)
+
+heavy <-
+  plyr::join(heavy, gpdd) %>%
+  arrange(main_id, p10) %>%
+  mutate(id_name = paste(round(p10, 2), common_name, main_id))
+
+get_gomp_res <- function(id_show) {
+  p <- subset(gomp_hat_base, main_id == id_show)
+  pop <- subset(gpdd, main_id == id_show)$population_untransformed
+  res <- rep(NA, length(pop))
+  for(i in 3:length(pop)) {
+    res[i] <- log(pop)[i] - (p$lambda_50 + p$b_50 * log(pop[i-1]))
+      #p$phi_50 * (p$lambda_50 + p$b_50 * log(pop[i-2]))
+  }
+  l <- qnorm(0.0001, 0, sd = p$sigma_proc_50)
+  u <- qnorm(0.9999, 0, sd = p$sigma_proc_50)
+  return(list(res = res, l = l, u = u))
+}
+
+make_spark <- function(x) {
+  res <- get_gomp_res(x$main_id[1])
+  bsw_l <- as.numeric(na.omit(seq_along(res$res)[res$res<res$l]))
+  bsw_u <- as.numeric(na.omit(seq_along(res$res)[res$res>res$u]))
+  pdf(paste0("sparks/", x$main_id[1], ".pdf"), width = 2.5, height = 1.2)
+  par(mar = c(0.5,0.2,0.5,0.2), oma = c(0,0,1.0,0), cex = 0.8)
+  par(xpd = NA)
+  with(x, plot(seq_along(population_untransformed), population, type = "l", axes = FALSE,
+      xlab = "", ylab = "", yaxs = "i", xaxs = "i"))
+  if(length(bsw_l) > 0)
+    points(bsw_l, x$population_untransformed[bsw_l], col = "red", pch = 20, cex = 3)
+  if(length(bsw_u) > 0)
+    points(bsw_u, x$population_untransformed[bsw_u], col = "blue", pch = 20, cex = 3)
+  dev.off()
+}
+
+#make_spark(subset(heavy, main_id == 10139))
+
+plyr::d_ply(heavy, "main_id", function(y) {
[…]

### 2014-09-22 6-plot-t-nu-eg.R

@@ -1,0 +1,35 @@
+# Check how student-t changes with nu
+# to justify tailing off prior around 100 or 200
+pdf("t-nu.pdf", width = 5, height = 4.5)
+cols <- c(rev(RColorBrewer::brewer.pal(5, "YlOrRd"))[-5], "grey30")
+lwd <- 2.0
+par(mfrow = c(2, 1))
+par(mar = c(0, 4, 0, 1), oma = c(4, 0, 1, 0), las = 1)
+par(xpd = FALSE)
+par(cex = 0.8, tck = -0.025, mgp = c(2, 0.6, 0))
+x <- seq(-9.5, 9.5, length.out = 500)
+plot(x, dt(x, 2), type = "l", xlab = "Standard deviation",
+  ylab = "", yaxs = "i", lty = 1, , lwd = lwd,
+  xlim = c(-8, 8), ylim = c(0, 0.42), col = cols[1], xaxt = "n", yaxt = "n")
+axis(2, at = c(0, 0.2, 0.4))
+lines(x, dt(x, 3), lty = 1, col = cols[2], lwd = lwd)
+lines(x, dt(x, 10), lty = 1, col = cols[3], lwd = lwd)
+lines(x, dt(x, 50), lty = 1, col = cols[4], lwd = lwd)
+lines(x, dt(x, 1e9), lty = 1, col = cols[5], lwd = lwd)
+legend("topright", legend = c(expression(nu==2), expression(nu==3),
+  expression(nu==10), expression(nu==50), expression(nu==infinity~(Normal))),
+  col = cols, bty = "n", lty = c(1, 1, 1, 1, 1),
+  lwd = c(lwd, lwd, lwd, lwd, lwd))
+# zoomed panel:
+plot(x, dt(x, 2), type = "l", xlab = "Standard deviation",
+  ylab = "", yaxs = "i", lty = 1, , lwd = lwd,
+  xlim = c(-8, 8), ylim = c(0, 0.025), col = cols[1], yaxt = "n")
+axis(2, at = c(0, 0.02))
+lines(x, dt(x, 3), lty = 1, col = cols[2], lwd = lwd)
+lines(x, dt(x, 10), lty = 1, col = cols[3], lwd = lwd)
+lines(x, dt(x, 50), lty = 1, col = cols[4], lwd = lwd)
+lines(x, dt(x, 1e6), lty = 1, col = cols[5], lwd = lwd)
+par(xpd = NA)
[…]

### 2014-09-22 7-values-for-paper.R

@@ -1,0 +1,28 @@
+# this file produces some values for inclusion of the text of the associated
+# paper
+#
+library(dplyr)
+
+write_tex <- function(x, macro, ...) {
+  out <- paste0("\\newcommand{\\", macro, "}{", x, "}")
+  cat(out, file = zz)
+  cat("\n", file = zz)
+  #writeLines(out, con = "values.tex", ...)
+}
+
+# what's the median and mean nu in the exponential prior?
+#x <- rexp(1e7, 0.01)
+#median(x)
+#mean(x)
+
+# how much was imputed?
+gpdd <- readRDS("gpdd-clean.rds")
+perc_imputed <- gpdd %>% group_by(main_id) %>% summarise(has_imputed = ifelse(max(interpolated) == 1, TRUE, FALSE)) %>% summarise(percent_imputed = sum(has_imputed)/length(has_imputed)) %>% round(2) * 100
+perc_imputed_pops <- as.character(perc_imputed$percent_imputed)
+
+perc_imputed_points <- round(sum(gpdd$interpolated)/nrow(gpdd) * 100, 1)
+
+zz <- file("values.tex", "w")
+write_tex(perc_imputed_pops, "percImputedPops")
+write_tex(perc_imputed_points, "percImputedPoints")
+close(zz)

### 2014-09-22 figure-out-phi.R

@@ -1,0 +1,86 @@
+gpdd <- readRDS("gpdd-clean.rds")
+library(rstan)
+x <- subset(gpdd, main_id == 20527)
+pop <- x$population_untransformed
+y <- diff(log(x$population_untransformed))
+par(mfrow = c(2, 1))
+plot(log(pop), type = "o")
+plot(y, type = "o")
+stan_gomp_ar1 <- readRDS("stan-gomp-ar1.rds")
+stan_gomp <- readRDS("stan-gomp.rds")
+warmup <- 400
+iterations <- 800
+
+
+###
+# OR simulate something
+phi_sim <- 0.9
+eps_sim <- arima.sim(n = 100, list(ar = phi_sim, 1, 1))
+lamb_sim <- 10
+b_sim <- 0
+y1 <- 1
+pop <- rep(NA, 100)
+pop[1] <- y1
+for(i in 2:100) {
+  pop[i] <- lamb_sim + b_sim * pop[i-1] + eps_sim[i]
+}
+pop <- exp(pop)
+par(mfrow = c(1, 1))
+plot(log(pop), type = "l")
+x <- data.frame(population_untransformed = pop)
+###
+
+sm <- sampling(stan_gomp,
+  data = list(N = nrow(x), y = log(x$population_untransformed),
+    nu_rate = 0.01,
+    b_lower = -1, b_upper = 2),
+  pars = c("lambda", "sigma_proc", "nu", "b"), iter = iterations,
+  chains = 3, warmup = warmup)
+
+sm_ar1 <- sampling(stan_gomp_ar1,
+  data = list(N = nrow(x), y = log(x$population_untransformed),
+    nu_rate = 0.01,
+    b_lower = -1, b_upper = 2),
+  pars = c("lambda", "sigma_proc", "nu", "b", "phi"), iter = iterations,
+  chains = 3, warmup = warmup)
+
+e <- extract(sm)
+p <- lapply(e, median)
+
+lambda <- p$lambda
+sigma_proc <- p$sigma_proc
+nu <- p$nu
+b <- p$b
+phi <- 0
+res <- rep(NA, length(pop))
+
+for(i in 3:length(pop)) {
+  res[i] <- log(pop)[i] - (
+    lambda + b * log(pop[i-1]) +
+    phi * (log(pop[i-1]) - (lambda + b * log(pop[i-2])))
+    )
+}
[…]

### 2014-09-22 figure-out-pop-model.R

@@ -1,0 +1,40 @@
+# play with gompertz and logistic growth model parameterizations
+# figure out what's the same and what's different and how they
+# model growth
+
+logistic1 <- function(Nt, r, K, theta = 1) {
+  r * (1 - (Nt/K)^theta)
+}
+
+#logistic2 <- function(Nt, r, K) {
+  #r * Nt * (1 - Nt/K)
+#}
+
+# both return log(abundance_{t + 1})
+logistic3 <- function(Nt, r, K) {
+  log(Nt) + r * (1 - Nt/K)
+}
+
+gompertz1 <- function(Nt, lambda, b) {
+  lambda + b * log(Nt)
+}
+
+gompertz2 <- function(Nt, lambda, b) {
+  log(Nt) + lambda + b * log(Nt)
+}
+
+N <- 50
+logx <- log(seq(1, 20))
+r <- 0.8
+K <- 10
+
+par(mfrow = c(2, 1))
+plot(exp(logx), log(exp(logistic3(exp(logx), 0.9, 5.29))/exp(logx)))
+plot(logx, log(exp(gompertz1(exp(logx), 0.5, 0.6))/exp(logx)))
+#plot(logx, log(exp(gompertz2(exp(logx), 0.5, 0.6-1))/exp(logx)))
+
+#exp(1.0/0.6)
+
+par(mfrow = c(2, 1))
+plot(exp(logx), log(exp(logistic3(exp(logx), 0.5, 5.29))/exp(logx)), main = "logistic")
+plot(exp(logx), log(exp(gompertz1(exp(logx), 0.5, 0.6))/exp(logx)), main = "gompertz")

### 2014-09-22 gpdd/extract-gpdd.sh

@@ -1,0 +1,10 @@
+# mdb-tables gpdd.mdb
+mdb-export gpdd.mdb DATA > data.csv
+mdb-export gpdd.mdb TAXON > taxon.csv
+mdb-export gpdd.mdb DATASOURCE > datasource.csv
+mdb-export gpdd.mdb VERSION > version.csv
+mdb-export gpdd.mdb TIMEPERIOD > timeperiod.csv
+mdb-export gpdd.mdb RestrictedDatasets > restricted_datasets.csv
+mdb-export gpdd.mdb LOCATION > location.csv
+mdb-export gpdd.mdb BIOTOPE > biotope.csv
+mdb-export gpdd.mdb MAIN > main.csv

### 2014-09-22 gpdd/make-gpdd.R

@@ -1,0 +1,77 @@
+# system("extract_gpdd.sh")
+
+dat <- read.csv("data.csv", stringsAsFactors = FALSE,
+  strip.white = TRUE)
+taxon <- read.csv("taxon.csv", stringsAsFactors = FALSE,
+  strip.white = TRUE)
+datasource <- read.csv("datasource.csv", stringsAsFactors = FALSE,
+  strip.white = TRUE)
+version <- read.csv("version.csv", stringsAsFactors = FALSE,
+  strip.white = TRUE)
+timeperiod <- read.csv("timeperiod.csv", stringsAsFactors = FALSE,
+  strip.white = TRUE)
+restricted_datasets <- read.csv("restricted_datasets.csv",
+  stringsAsFactors = FALSE, strip.white = TRUE)
+location <- read.csv("location.csv", stringsAsFactors = FALSE,
+  strip.white = TRUE)
+biotope <- read.csv("biotope.csv", stringsAsFactors = FALSE,
+  strip.white = TRUE)
+main <- read.csv("main.csv", stringsAsFactors = FALSE,
+  strip.white = TRUE)
+
+underscore_names <- function(x) {
+  y <- gsub("([a-z])([A-Z])", "\\1_\\2", names(x))
+  tolower(y)
+}
+
+names(dat) <- underscore_names(dat)
+names(taxon) <- underscore_names(taxon)
+names(datasource) <- underscore_names(datasource)
+names(version) <- underscore_names(version)
+names(timeperiod) <- underscore_names(timeperiod)
+names(restricted_datasets) <- underscore_names(restricted_datasets)
+names(location) <- underscore_names(location)
+names(biotope) <- underscore_names(biotope)
+names(main) <- underscore_names(main)
+
+dat$decimal_year_begin <- as.numeric(dat$decimal_year_begin)
+dat$decimal_year_end <- as.numeric(dat$decimal_year_end)
+dat$population <- as.numeric(dat$population)
+dat$population_untransformed <- as.numeric(dat$population_untransformed)
+
+location$longitude_degrees <- as.numeric(location$longitude_degrees)
+location$longitude_minutes <- as.numeric(location$longitude_minutes)
+location$latitude_degrees <- as.numeric(location$latitude_degrees)
+location$latitude_minutes <- as.numeric(location$latitude_minutes)
+location$lat_dd <- as.numeric(location$lat_dd)
[…]

### 2014-09-22 ids2.R

@@ -1,0 +1,4 @@
+gpdd <- readRDS("gpdd-clean.rds")
+ids <- unique(gpdd$main_id)
+main_id_vec <- ids[201:400]
+source("2-run-models-ar1.R")

### 2014-09-22 ids3.R

@@ -1,0 +1,4 @@
+gpdd <- readRDS("gpdd-clean.rds")
+ids <- unique(gpdd$main_id)
+main_id_vec <- ids[401:length(ids)]
+source("2-run-models-ar1.R")

### 2014-09-22 logistic-ids1.R

@@ -1,0 +1,4 @@
+gpdd <- readRDS("gpdd-clean.rds")
+ids <- unique(gpdd$main_id)
+main_id_vec <- ids[1:200]
+source("2-run-models-ar1.R")

### 2014-09-25 anderson-etal-blackswan-timeseries.tex
Blanket update to code and paper

--- 2014-09-22 heavy-ms.tex
@@
-\documentclass[10pt]{article}
+\documentclass[11pt]{article}
 \usepackage{geometry}
+\geometry{verbose,letterpaper,tmargin=2.54cm,bmargin=2.54cm,lmargin=2.54cm,rmargin=2.54cm}
 \geometry{letterpaper}
@@
 %\usepackage{pdflscape}
-%\usepackage{setspace}
+\usepackage{setspace}
 \usepackage{booktabs}
+\usepackage{lineno}
 
@@
 \mathchardef\mhyphen="2D
-%\usepackage{url}
-%\urlstyle{rm}
 
-\textheight 20.5cm
+\textheight 22.0cm
 
@@
 
-\title{The evidence for heavy tails in global\\population dynamics}
-\author{Sean Anderson \ldots}
-%\date{}
+\title{Black swans in ecological time series}
+\author{
+Sean C. Anderson$^{1\ast}$ \and
+Trevor A. Branch$^2$ \and
+Andrew B. Cooper$^3$ \and
+Nicholas K. Dulvy$^1$
+}
+\date{}
 
-%\pagestyle{empty}
-%\setcounter{secnumdepth}{-2}
+% remove numbers in front of sections:
+\makeatletter
+\renewcommand\@seccntformat[1]{}
+\makeatother
 
 \begin{document}
+\input{values} % R output
 
-\input{values}
+%\begin{spacing}{1.4}
+%\doublespacing
+\onehalfspacing
 
-%\begin{spacing}{1.1}
-%\doublespacing
-%\onehalfspacing
 \maketitle
+\thispagestyle{empty}
 
-\section*{Abstract}
+\noindent
+\textit{Authors are listed alphabetically (ABCD!) and open to reordering.}
 
-\section*{Introduction}
+\noindent
+\textsuperscript{1}Earth to Ocean Research Group, Department of Biological Sciences, Simon Fraser University, Burnaby BC, V5A 1S6, Canada
+
+\noindent
+\textsuperscript{2}School of Aquatic and Fishery Sciences, University of Washington, Box 355020, Seattle, WA 98195, USA
+
+\noindent
+\textsuperscript{3}School of Resource and Environmental Management, Simon Fraser University, Burnaby, BC, V5A 1S6, Canada
+
+\noindent
+\textsuperscript{*}Corresponding author: Sean C.
+Anderson; Earth to Ocean Research Group, Department of Biological Sciences, Simon Fraser University, Burnaby BC, V5A 1S6; Phone: 1-778-782-3989; E-mail: sean\_anderson@sfu.ca
+
+\clearpage
+
+\setcounter{page}{1}
+\section{Abstract}
+
+\section{Introduction}
 
 Increasing realization about the importance of extreme events in the environment (REFs)
-
 and of the importance of ecological surprises and black swans (REFs)
@@
 
-citation with good justification of gompertz and other references: \citep{herrandoprez2014}
+\section{Methods}
 
-\section*{Methods}
+\subsection{Modelling framework}
 
-First, I tried fitting this as a Gompertz model and estimating the density dependence parameter. The Gompertz was by far the most favoured model in \citet{brook2006} with the GPDD.
-
-I could see an argument being made that we're just seeing them because of autocorrelation in the residuals.
-Autocorrelation plays a big role in \citet{inchausti2002} justifying why they think they're seeing increasing CV of time series in the GPDD with time and an increasing CV could create heavy tails.
-They just model the spectral properties and CV of the abundance time series --- not a population dynamics model.
-I tried estimating a coefficient for the first-order autocorrelation of the residuals.
-There is some autocorrelation it picks up sometimes, but interestingly this doesn't change things that much with respect to heavy tail results.
-Nothing systematic at least (Another important reference on spectral analysis and the GPDD: \citet{garcia-carreras2011}.)
-
-\subsection*{Modelling framework}
-
-The autoregressive Student-t Gompertz model is:
+The Student-t Gompertz model is:
 \begin{align*}
 x_t &= \lambda + b x_{t-1} + \epsilon_t\\
-\epsilon_t &\sim \mathrm{Student\mhyphen t}_\nu(\phi \epsilon_{t-1}, \sigma_\mathrm{proc}^2),
+%\epsilon_t &\sim \mathrm{Student\mhyphen t}_\nu(\phi \epsilon_{t-1}, \sigma_\mathrm{proc}^2),
+\epsilon_t &\sim \mathrm{Student\mhyphen t}_\nu(0, \sigma_\mathrm{proc}),
 \end{align*}
 
-\noindent where $x_t$ is the $\ln$ abundance at time $t$ and $\nu$ is the Student-t distribution degrees of freedom parameter.
-If $\nu$ is small (say 2--20) the distribution has heavy tails.
-As $\nu$ approaches infinity the distribution approaches the normal distribution.
-The parameter $b$ is the density dependence parameter.
+\noindent where $x_t$ is the $\ln$ abundance at time $t$.
 The model is density independent if $b = 1$ and maximally density dependent if $b = 0$.
-The parameter $\phi$ represent the first-order autoregressive correlation of the residuals $\epsilon$.
-$\sigma_\mathrm{proc}$ is the process error standard deviation.
-$\lambda$ represents the population growth rate per time step.
+The parameter $\lambda$ represents the expected per capita growth rate at $\ln$ abundance = 0.
+The process noise is represented by a Student-t distribution that is centered at 0, has a scale parameter of $\sigma_\mathrm{proc}$, and has a degrees of freedom parameter of $\nu$.
+If $\nu$ is small ($\lesssim 10$) the distribution has heavy tails.
+As $\nu$ approaches infinity the distribution approaches the normal distribution (Fig~\ref{fig:didactic}).
 
-The mostly weakly-informative priors I'm using are:
-\begin{align*}
-b &\sim \mathrm{Uniform}(-1.2, 1.2)\\
-\lambda &\sim \mathrm{Normal}(0, 20^2)\\
-\nu &\sim \mathrm{Exponential}(0.01)\\
-\phi &\sim \mathrm{Truncated\mhyphen Normal}(0, 1, \mathrm{min.} = -0.99, \mathrm{max.} = 0.99)\\
-\sigma_\mathrm{proc} &\sim \mathrm{Half\mhyphen Cauchy} (5).
-\end{align*}
+We fit alternative models that allowed for autocorrelation of the residuals, allowed for observation error, and allowed the functional form of the population dynamics to be represented by a Ricker-logistic population model (Supporting Material).
 
-Priors justification:
+\section{Results}
 
-\begin{itemize}
-\item $b$ is bounded just past stationary so we can detect if they are non-stationary will keeping the sampler from wandering off
-
-\item $\lambda$ prior (variance = 400) is basically uninformative within the range of expected values for population growth... it allows even a X probability at a value of X
-
-\item $\nu$ is based on \citet{fernandez1998}; they chose a more informative 0.1 value, we chose a less informative 0.01 and justify it based on performance in supplemental figure X (gives X probability of value less than 10) but constrains the sampling somewhat since above 40 or 50 the shape of the t distribution is almost identical and data are not usually informative at these data quantities...
-sampler would head off to infinity otherwise
-
-\item $\phi$ given a standard deviation of 1 given prior analyses which suggest autocorrelation in these datasets is minimal; prior is weak enough to allow high or low values, but given little information will stay near our expectation; similar approach in \citep{thorson2014a}
-
-\item $\sigma_\mathrm{proc}$ can be justified based on \citet{gelman2006c} and the expected range of this variable in nature from previous studies \citep[e.g.][]{connors2014}
-\end{itemize}
-
-Using Stan 2.4.0 \citep{stan-manual2014}, and R 3.1.1 \citep{r2014} I'm starting with 4 chains and 2000  iterations with the first 1000 as warmup (i.e.\ 4000 total samples).
-If rhat is greater than 1.05 for any parameter or the minimum effective sample size is less than 200 for any parameter then I double both the total iterations and warmup and run again.
-These conditions were met by 8000 iterations (16000 total samples) in all cases and in almost all cases (X\%) within the initial 2000 iteration case (i.e.\ a total of 8000 samples).
-
-Then, I again tried allowing for a specified level of observation error.
-I'm still running that now for just the mammals.
-From before, I believe this will reduce the probability of heavy tails a bit for many of them, but the vast majority that had heavy tails would still be considered heavy tailed.
-This does reduce substantially many of the bird heavy tails though.
-
-(See the simulation results... I don't think this has the intended benefit and can probably be left out of the main results.)
-
-When I incorporate observation error, the model is:
-\begin{align*}
-U_t &= \lambda + b U_{t-1} + \epsilon_t\\
-x_t &\sim \mathrm{Normal}(U_t, \sigma_\mathrm{obs}^2)\\
-\epsilon_t &\sim \mathrm{Student\mhyphen t}_\nu(\phi \epsilon_{t-1}, \sigma_\mathrm{proc}^2),
-\end{align*}
-
-\noindent
-where $U$ represents the unobserved state vector, and $\sigma_\mathrm{obs}$ represents the standard deviation of observation error (on a log scale), which was set at 0.2.
-
-%Also, we may not be crazy seeing the heavier tails for heavier longer lived animals.
-There's some precedence for that in a paper that found more red noise (longer term fluctuations dominating the variance and therefore making the CV grow with time) in large-bodied animals.
-They attributed it to a mismatch between dynamics and the observation scale.
-They used a different life-history dataset too.
-But, there are other systematic differences with the body size-nu relationship that make this sketchy for us.
-E.g.\ taxonomic orders and interval frequency of data collection vary systematically and in clumps with body size (not shown in the below figure).
-And basically the only stuff with very \textit{low} probability of heavy tails are the the small ones with generation times less than a year and small collection intervals.
-It's basically a step function.
-We may not be able to tease that all apart...
-at least not in this paper.
-
-\subsection*{Data selection}
-
-Filtering rules for the GPDD:
-We removed all populations with zeros (can't be fit with a Gompertz), all populations with uneven sampling intervals (end date minus start date for each step), and all populations with population estimates listed as ``harvest'' as well as those listed as from two fur trading sources (give details).
-Then, we removed any populations with 5 or more identical values in a row since these suggest either recording error or extrapolation between two observations.
-Then, we filled in all missing time steps with NAs and imputed single missing values with the geometric mean of the previous and following values (linear on the log scale that the Gompertz operates on).
-This leaves NAs that were recorded in successive time steps.
-Then, we went through each population time series and found the longest window of abundance with at least 25 time steps.
-If there were any with multiple windows of identical length, we took the most recent window.
-Finally, we removed main\_id 20531, which is a duplicate of 10139 (a heron population) and main\_id 10008, which had obvious errors in the data recording (a water vole population).
-We provide a supplemental figure of all the time series included in our analysis and indicate which values were imputed (\percImputedPops\% of populations had at least one point imputed and only \percImputedPoints\% of the total observations were imputed) (Fig.\ SX).
-
-\subsection*{Simulation testing the model}
-
-Throughout all of this --- show that the model, if anything, under-predicts heavy tails but with sufficient data is unbiased.
-
-2 parts: how many samples from the true population t distribution do you need to detect low nu?
-And, given that you have a set of deviations in which nu is detectable (effective nu is within 0.5 CV of true nu), can the more complex Gompertz still capture this?
-
-First part: We drew from t distributions with different nu values and mean of 0, scale of 1.
-We started with 1600 samples and then fitted again at the first 800, 400, 200,  100, 50, 25.
-Each time we recorded the nu posterior.
-
-Second part: To generate a series of process deviations with an effective nu approximately equal to the true nu, we generated process deviation sets repeatedly and estimated the mean, scale, and nu values each time.
-We recorded when the estimated nu was within 0.5 CVs of the true nu and used this set of random seed values in our Gompertz simulation.
-We then fit AR1 Gompertz models to the simulated datasets with all parameters (except nu) set near the median values estimated in the GPDD.
-
-\section*{Results}
-
-Main text figures:
-
-\begin{enumerate}
-\item illustrate $\nu$ and show simulated Gompertz time series with these levels of $\nu$ (include probability density at some extreme deviation in the caption or on figure for different $\nu$ values
-
-\item a figure showing $\nu$ estimates within classes and  orders, ordered by decreasing $\nu$ estimate to illustrate percentage of populations, grouped by order in a phylogenetically related way include some silhouettes
-
-\item 5--10 example time series with heavy tails from the GPDD; highlight the ``black swans'' with, say, red dots (perhaps also highlighting if they are up or down swings); pick some \textit{very} heavy tails and some moderately heavy tails.
-Show the $\nu$ estimate somewhere in each panel
-
-\item something looking at hypotheses for what's associated with heavy tails: data collection method, mid date of time series (or when it started?), time series length, AR1, density dependence, lambda, body size for mammals, IUCN status?
-Most likely a grid of scatter plots coloured by taxonomic class....
-or rows by taxonomic class and colours by order.
-(also ratio of generation time to collection window) (possibly, some of these will go in supplement with interesting ones in main paper).
-
-\end{enumerate}
-
-Main text tables:
-
-\begin{enumerate}
-\item Table summarizing hypotheses for why we might see heavy tails
-\item Pick some example populations (say the ones in the time series example plot) and show the nu estimate, reference, and a column digging into any known caveats or reasons
-\end{enumerate}
-
-Supplemental figures:
-
-\begin{enumerate}
-
-\item illustration of how the exponential prior slightly constrains the sampling of $\nu$\ldots illustrating what the prior probability of heavy tails is given uninformative data (about 5\% probability $\nu < 10$.
-
-\item plots of the priors overlayed with estimates across populations for all parameters
-
-\item cross plot of $\nu$ estimates from the models: no density dependence, add density dependence, add AR1, add assumed observation error
-
-\item time series plots of all heavy-tailed populations
-
-\item simulation testing of detecting nu given reduced sampling from true distribution; show repeated estimates of nu with different sample sizes, and show a few example samples with different observation window lengths and highlight the deviations that are beyond the 0.001 and 0.999 probabilities for a normal distribution.
-
-\item simulation testing of the Gompertz model with process error deviations fixed to have an effective $\nu$ estimate at the true level --- this asks how well is the Gompertz model able to partition the various parameters in a precise and unbiased way given that we know the heavy tails are there (also include a scenario with a massive black swan to show how it behaves there)
-
-\end{enumerate}
-
-\section*{Discussion}
+\section{Discussion}
 
@@
 
-\bibliography{/Users/seananderson/Dropbox/tex/jshort.bib,/Users/seananderson/Dropbox/tex/ref3.bib}
+\bibliography{/Users/seananderson/Dropbox/tex/jshort,/Users/seananderson/Dropbox/tex/ref3}
 
@@
 
-\caption{TODO}\\
+\caption{Example population dynamic black swans from the Global Population Dynamics Database and descriptions of their causes.
+Red and blue dots indicate downward and upward events that have a probability of occurring of 0.001\% or less if the population dynamics were explained by a Gompertz model with normally distributed process noise.}\\
 
 \toprule
-Time series & Species & Black swan description & Reference \\
+Time series (log scale) & Population & Black swan description & Reference \\
 \midrule
@@
 \includegraphics[width=2cm]{sparks/6528} &
-Shag (\textit{Phalacrocorax aristotelis}), Northumberland &
-Shortage of nest sites reduced productivity; red-tide event in 1968 caused extreme mortality; no longer a nest shortage; population rapidly increased. &
+Shag,
+\textit{Phalacrocorax aristotelis},
+UK &
+Shortage of nest sites reduced productivity; red-tide event in 1968 caused extreme mortality; no longer a nest shortage; population rapidly increased &
 \citep{potts1980}\\
+
+\includegraphics[width=2cm]{sparks/10007} &
+Water vole,
+\textit{Arvicola terrestris},
+UK &
+Short-term population cycles from predator interactions combined with longterm environmental cycle caused sharp downswing  &
+\citep{saucy1994}\\
+
+\includegraphics[width=2cm]{sparks/7115} &
+Fur seal,
+\textit{Arctocephalus pusillus},
+South Africa &
+Strong decreases in harvesting, loss of predators, and diamond mining regulations reducing human traffic caused sharp upswings  &
+\citep{shaughnessy1982}\\
+
+\includegraphics[width=2cm]{sparks/10113} &
+Willow grouse,
+\textit{Lagopus lagopus},
+UK &
+Parasite and predation effects interacted to cause low years  &
+\citep{dobson1995}\\
+
+\includegraphics[width=2cm]{sparks/10162} &
+Red grouse,
+\textit{Lagopus lagopus scoticus},
+UK &
+Good environmental conditions produced high numbers and vulnerable populations; bad conditions and overcrowding combined to create crashes  &
+\citep{mackenzie1952}\\
+
+\includegraphics[width=2cm]{sparks/1235} &
+Wren,
+\textit{Troglodytes troglodytes},
+UK &
+Severe winters where food was submerged in snow caused population crash &
+\citep{newton1998} \\
+
+\includegraphics[width=2cm]{sparks/20579} &
+Grey heron,
+\textit{Ardea cinerea},
+UK &
+Severe winters in 1929, 1940--1942, and 1962--1963; 1963 event so severe that recovery took three times as long as expected &
+\citep{stafford1971} \\
+
+%\includegraphics[width=2cm]{sparks/20580} &
+%Chamois, \textit{Rupicapra rupicapra}, Switzerland &
+% &
+%\citep{brook2006a}\\
+
+%\includegraphics[width=2cm]{sparks/5019} &
+%Barbary macaque,  &
+% &
+%REF\\
 
@@
 %Carrot fly (\textit{Psila rosae}, Finland &
-%TODO &
+% &
 %\citep{markkula1965}\\
@@
 %Grey heron (\textit{Ardea cinerea}), UK &
-%TODO &
+% &
 %\citep{stafford1971}\\
-
-\includegraphics[width=2cm]{sparks/1235} &
-Wren (\textit{Troglodytes troglodytes}), UK &
-TODO &
-REF \\
-
-%\includegraphics[width=2cm]{sparks/20580} &
-%Chamois, \textit{Rupicapra rupicapra}, Switzerland &
-%TODO &
-%\citep{brook2006a}\\
-
-\includegraphics[width=2cm]{sparks/5019} &
-Barbary macaque, TODO &
-TODO &
-REF\\
 
@@
 
+% reset citation style:
+\bibpunct{(}{)}{;}{a}{}{;}
+
 \clearpage
-
-\include{stat-table}
-%\bibpunct{}{}{;}{a}{}{;}
-%\begin{footnotesize}
-%
-%\renewcommand*{\arraystretch}{1.5}
-%\begin{longtable}{>{\RaggedRight}p{3.5cm}>{\RaggedRight}p{8.2cm}>{\RaggedRight}p{2.2cm}}
-%
-%\caption{Description of black swans in GPDD}\\
-%\toprule
-%
-%\midrule
-%
-%\bottomrule
-%\label{tab:causes}
-%\end{longtable}
-%\end{footnotesize}
-%\bibpunct{(}{)}{;}{a}{}{;}
 
@@
 \begin{center}
-\includegraphics[width=0.6\textwidth]{t-nu.pdf}
-\caption{A figure illustrating $\nu$. Maybe put some example (Gompertz) time series to the right with the same seed and different $\nu$ values?}
-\label{default}
+\includegraphics[width=\textwidth]{t-nu-eg.pdf}
+\caption{An illustration of fitting population dynamics models that allow for heavy tails, represented by the Student-t degrees of freedom parameter $\nu$.
+  (a, b) The probability density for t distributions with a scale parameter of 1 and different values of $\nu$.
+  Small values of $\nu$ create heavy tails.
+  As $\nu$ approaches infinity the distribution approaches the normal distribution.
+  For example, the at $\nu = 2$, the probability of drawing a value less than -5 is 1.8\% whereas the probability of drawing such a value from a normal distribution is almost impossible: $2.9\cdot10^{-5}$\%.
+  (c--e) Simulated population dynamics from a Gompertz model with process noise drawn from t distributions with different values of $\nu$.
+  (f--h) Estimates of $\nu$ from models fit to the times series in panels c--e.
+  Shown are the posterior samples (histograms), median and interquartile range of the posterior (IQR) (dots and line segments), and the exponential prior on $\nu$ (dashed lines).
+  Colour shading behind panels f--h illustrates the region of heavy tails.
+  Coloured dots in panels c and d represent points with less than a 1 in 1000 chance of occurring in a normal distribution.}
+\label{fig:didactic}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=\textwidth]{gomp-nu-ar1.pdf}
-\caption{Estimate of $\nu$ from the Gompertz model (with estimated AR1 autocorrelation of the residuals but no observation error).
+\includegraphics[width=\textwidth]{gomp-nu-base.pdf}
+\caption{
+  Estimate of $\nu$ from the base Gompertz model (no autocorrelation or observation error).
   Shown are median (points), quartiles (mid line segments) and 90\% quantiles (widest line segments) of the posterior.}
@@
 \begin{center}
-\includegraphics[width=\textwidth]{gomp-ar1-p20-dot-order.pdf}
-\caption{Probability $\nu<20$, i.e.\ one way of asking what the probability is of heavy tails.
+\includegraphics[width=\textwidth]{gomp-base-p10-dot-order.pdf}
+\caption{
+  Probability $\nu<20$, i.e.\ one way of asking what the probability is of heavy tails.
   Dots represent populations.
@@
 
-%\clearpage
-%
-%\begin{figure}[htbp]
-%\begin{center}
-%\includegraphics[width=\textwidth]{gomp-ar1-p20-hist.pdf}
-%\caption{Histogram of the same data as Figure~\ref{fig:dots} but lumped by class.
-Probably can't justify including this as well as the previous two figures unless there's some way we can make this more informative.}
-%\label{default}
-%\end{center}
-%\end{figure}
+\clearpage
 
-%\clearpage
-
-%\begin{figure}[htbp]
-%\begin{center}
-%\includegraphics[width=0.9\textwidth]{effect-of-ar1-on-nu.pdf}
-%\caption{The effect of assuming no autocorrelation of residuals (vertical axis) vs.\ estimating autocorrelation of residuals (horizontal axis).
-I don't see a systematic effect.
-These are medians and interquartile ranges of the posterior.
-The colour shows $\phi$ --- the parameter estimating the lag-1 correlation of the residuals.
-The diagonal line illustrates 1:1.
-A few of these look strange (ones with $\phi$ near zero but far from the 1:1 line).
-It's possible the chains haven't converged sufficiently in these cases in one of the model versions.
-I will check what's going on and run more iterations in a final run if needed.}
-%\label{default}
-%\end{center}
-%\end{figure}
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=\textwidth]{nu-covariates-base.pdf}
+\caption{
+  (Also see the next figure.)
+  A number of hypotheses for what could be driving $\nu$ values.
+From top left to bottom right: $\phi$ (AR1 coefficient), $\log \sigma_\mathrm{proc}$ (log process error standard deviation), $N$ (time series length), $b$ (density dependence parameter in the Gompertz model; 1 is density independent and 0 is maximally density dependent), $\lambda$ (growth parameter from Gompertz model), $\log M$ (log adult body mass in grams; just for mammals and from the PanTHERIA database).
+  All dots show median estimates and vertical line segments show interquartile range of posterior for the $\nu$ values.
+  I don't generally see a whole lot going on here besides taxonomic differences but I'm open to some eye fuzzing and other ideas.
+  If we want to, I was playing with a hierarchical beta regression with different levels for the taxonomic orders explaining the variation in p(nu $<$ 10) or these data\ldots but hypothesis generation with some visuals might be enough for this paper.}
+\label{fig:correlates}
+\end{center}
+\end{figure}
 
@@
 \begin{center}
-\includegraphics[width=\textwidth]{nu-covariates.pdf}
-\caption{(Also see the next figure.) A number of hypotheses for what could be driving $\nu$ values.
-From top left to bottom right: $\phi$ (AR1 coefficient), $\log \sigma_\mathrm{proc}$ (log process error standard deviation), $N$ (time series length), $b$ (density dependence parameter in the Gompertz model; 1 is density independent and 0 is maximally density dependent), $\lambda$ (growth parameter from Gompertz model), $\log M$ (log adult body mass in grams; just for mammals and from the PanTHERIA database).
-All dots show median estimates and vertical line segments show interquartile range of posterior for the $\nu$ values.
-I don't generally see a whole lot going on here besides taxonomic differences but I'm open to some eye fuzzing and other ideas.
-If we want to, I was playing with a hierarchical beta regression with different levels for the taxonomic orders explaining the variation in p(nu $<$ 10) or these data\ldots but hypothesis generation with some visuals might be enough for this paper.}
-\label{default}
+\includegraphics[width=\textwidth]{brook-etal-correlates.pdf}
+\caption{
+  \citet{brook2006a} correlates plotted against nu estimates.
+}
+\label{fig:brook}
 \end{center}
@@
 
-%\begin{figure}[htbp]
-%\begin{center}
-%\includegraphics[width=\textwidth]{nu-covariates-p.pdf}
-%\caption{Same as previous figure but the y axis is p(nu $<$ 20), which I think might be better to use\ldots, in reality this kind of coarse dichotomy I think is more realistic.
+\begin{centering}
+\LARGE
+Supporting material\\[1.5em]
+\end{centering}
+
+\section{Data selection}
+
+TODO this paragraph is old:
+
+Filtering rules for the GPDD:
+We removed all populations with zeros NO LONGER TRUE (can't be fit with a Gompertz), all populations with uneven sampling intervals (end date minus start date for each step), and all populations with population estimates listed as ``harvest'' as well as those listed as from two fur trading sources (give details).
+Then, we removed any populations with 5 or more identical values in a row since these suggest either recording error or extrapolation between two observations.
+Then, we filled in all missing time steps with NAs and imputed single missing values with the geometric mean of the previous and following values (linear on the log scale that the Gompertz operates on).
+This leaves NAs that were recorded in successive time steps.
+Then, we went through each population time series and found the longest window of abundance with at least 25 time steps.
+If there were any with multiple windows of identical length, we took the most recent window.
+Finally, we removed main\_id 20531, which is a duplicate of 10139 (a heron population).
+We provide a supplemental figure of all the time series included in our analysis and indicate which values were imputed (\percImputedPops\% of populations had at least one point imputed and only \percImputedPoints\% of the total observations were imputed) (Fig.\ SX).
+
+Go over zero rules.
+
+\section{Details on heavy-tailed Gompertz probability model}
+
+For the Gompertz, the mostly weakly-informative priors I'm using are:
+\begin{align*}
+b &\sim \mathrm{Uniform}(-1, 2)\\
+\lambda &\sim \mathrm{Normal}(0, 10^2)\\
+\nu &\sim \mathrm{Exponential}(0.01)\\
+\phi &\sim \mathrm{Truncated\mhyphen Normal}(0, 1, \mathrm{min.} = -0.99, \mathrm{max.} = 0.99)\\
+\sigma_\mathrm{proc} &\sim \mathrm{Half\mhyphen Cauchy} (5).
+\end{align*}
+
+See Figure~\ref{fig:priors}.
+
+Priors justification: (TODO this is old)
+
+\begin{itemize}
+\item $b$ is bounded just past stationary so we can detect if they are non-stationary will keeping the sampler from wandering off too far
+
+\item $\lambda$ prior (variance = 100) is basically uninformative within the range of expected values for population growth... it allows even a X probability at a value of X
+
+\item $\nu$ is based on \citet{fernandez1998}; they chose a more informative 0.1 value, we chose a less informative 0.01 and justify it based on performance in supplemental figure X (gives X probability of value less than 10) but constrains the sampling somewhat since above 40 or 50 the shape of the t distribution is almost identical and data are not usually informative at these data quantities...
+sampler would head off to infinity otherwise
+
+\item $\phi$ given a standard deviation of 1 given prior analyses which suggest autocorrelation in these datasets is minimal; prior is weak enough to allow high or low values, but given little information will stay near our expectation; similar approach in \citep{thorson2014a}
+
+\item $\sigma_\mathrm{proc}$ can be justified based on \citet{gelman2006c} and the expected range of this variable in nature from previous studies \citep[e.g.][]{connors2014}
+\end{itemize}
+
+Using Stan 2.4.0 \citep{stan-manual2014}, and R 3.1.1 \citep{r2014} I'm starting with 4 chains and 1000 iterations with the first 500 as warmup (i.e.\ 2000 total samples).
+If rhat is greater than 1.05 for any parameter or the minimum effective sample size is less than 200 for any parameter then I double both the total iterations and warmup and run again.
+These conditions were met by 8000 iterations (16000 total samples) in all cases and in almost all cases (X\%) within the initial 1000 iteration case (i.e.\ a total of 2000 samples).
+
+\section{Simulation testing the model}
+
+Throughout all of this --- show that the model, if anything, under-predicts heavy tails but with sufficient data is unbiased.
+
+2 parts: how many samples from the true population t distribution do you need to detect low nu?
+And, given that you have a set of deviations in which nu is detectable (effective nu is within 0.5 CV of true nu), can the more complex Gompertz still capture this?
+
+First part: We drew from t distributions with different nu values and mean of 0, scale of 1.
+We started with 1600 samples and then fitted again at the first 800, 400, 200,  100, 50, 25.
+Each time we recorded the nu posterior.
+
+Second part: To generate a series of process deviations with an effective nu approximately equal to the true nu, we generated process deviation sets repeatedly and estimated the mean, scale, and nu values each time.
+We recorded when the estimated nu was within 0.5 CVs of the true nu and used this set of random seed values in our Gompertz simulation.
+We then fit AR1 Gompertz models to the simulated datasets with all parameters (except nu) set near the median values estimated in the GPDD.
+
+\section{Alternative population models}
+
+\subsection{Ricker-logistic}
+
+I also fitted a Ricker-logistic model:
+\begin{align*}
+x_t &= x_{t-1} + r_{\mathrm{max}}\left(1 - \frac{N_{t-1}}{K}\right) + b x_{t-1} + \epsilon_t\\
+%\epsilon_t &\sim \mathrm{Student\mhyphen t}_\nu(\phi \epsilon_{t-1}, \sigma_\mathrm{proc}^2),
+\epsilon_t &\sim \mathrm{Student\mhyphen t}_\nu(0, \sigma_\mathrm{proc}^2),
+\end{align*}
+
+Priors: For the Ricker-logistic $K$ uniform between 0 and twice the maximum observed abundance (\citet{clark2010} was less conservative at 0 and maximum observed).
+$r_\mathrm{max}$ uniform between 0 and 20 as in \citet{clark2010}.
+Same priors on $\nu$ and process standard deviation.
+
+\subsection{Autocorrelated residuals}
+
+\subsection{Assumed observation error}
+
+When I incorporate observation error, the model is:
+\begin{align*}
+U_t &= \lambda + b U_{t-1} + \epsilon_t\\
+x_t &\sim \mathrm{Normal}(U_t, \sigma_\mathrm{obs}^2)\\
+\epsilon_t &\sim \mathrm{Student\mhyphen t}_\nu(0, \sigma_\mathrm{proc}^2),
+\end{align*}
+
+\noindent
+where $U$ represents the unobserved state vector, and $\sigma_\mathrm{obs}$ represents the standard deviation of observation error (on a log scale), which was set at 0.3.
+
+%\end{spacing}
+
+\renewcommand{\thetable}{S\arabic{table}}
+\setcounter{table}{0}
+
+\include{stat-table}
+
+\renewcommand{\thefigure}{S\arabic{figure}}
+\setcounter{figure}{0}
+
+% \begin{figure}[htbp]
+% \begin{center}
+% \includegraphics[width=\textwidth]{nu-covariates-p-base.pdf}
+% \caption{Same as previous figure but the y axis is p(nu $<$ 20), which I think might be better to use\ldots, in reality this kind of coarse dichotomy I think is more realistic.
 If p(nu $<$ 20) is high (say above 0.5) then there's a good chance the tails are heavy.
 Otherwise, we just don't know.}
-%\label{default}
-%\end{center}
-%\end{figure}
+% \label{default}
+% \end{center}
+% \end{figure}
 
@@
 \includegraphics[width=\textwidth]{ts-gpdd-heavy-eg-log10-no-fur.pdf}
-\caption{Time series for populations with p(nu $<$ 10) $>$ 0.5.
-Ordered by increasing p(nu$<$10) (this value is shown pasted onto the front of each panel label).
-These are shown on a log10 y axis.
-There are a few funny time series we may want to look into and consider rules to remove.
-Colours show taxonomic class.
-This is for our understanding and a version of this could be a supplemental figure.
-But, we could take maybe 6 to 8 of these and use them as illustrations in the paper with the heavy tailed deviations highlighted.
-Could also probably dig into the literature on some of these and summarize in a table if we narrowed it down a bit.
-Lots of minks, beavers, and bears.}
-\label{default}
+\caption{Time series for populations with p($\nu < 10$) $>$ 0.5 using the base Gompertz population model.
+Panels are ordered by increasing p($\nu < 10$).
+Vertical axes are on a log10 scale.
+Colours indicate taxonomic classes.
+TODO add dots for upswings and downswings as in Table 1.
+The labels on each panel indicate p($\nu < 10$) $>$ 0.5, the common name for the species, and the GPDD ID number.}
+\label{fig:heavy-ts}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=\textwidth]{t-dist-sampling-sim-prior-exp0point01.pdf}
-\includegraphics[width=\textwidth]{t-dist-sampling-sim-sigma-prior-exp0point01.pdf}
-\caption{TODO --- also maybe show the different exponential priors.}
-\label{default}
+\includegraphics[width=0.9\textwidth]{t-dist-sampling-sim-prior-exp0point01.pdf}
+\includegraphics[width=0.9\textwidth]{t-dist-sampling-sim-sigma-prior-exp0point01.pdf}
+\caption{
+  Testing the ability to estimate $\nu$ (top panels) and the scale parameter of the process error (bottom panels) for a given number of samples (columns) drawn from a distribution with a given true $\nu$ value (rows).
+The red lines indicate the true population value.
+When a small number of samples are drawn there may not be samples sufficiently far into the tails to recapture the true $\nu$ value; however, heavy tails are still distinguished from normal tails in most cases, even with only 25 or 50 samples.
+TODO switch to nu = black swan, 3, 5, and normal.
+}
+\label{fig:sim-nu}
 \end{center}
@@
 \includegraphics[width=1.1\textwidth]{sim-gompertz.pdf}
-\caption{\textbf{TODO Caption doesn't match new figure.} Results from a simulation test of the estimation model.
-Dashed horizontal lines are the true population value.
-True values were chosen to represent approximately the median values as estimated from the GPDD.
-``Iteration'' refers to a stochastic draw from the true population distribution and a model fitting.
-The panels show true values of $\nu = $ 2.1, 5, 15, and 200 (very heavy, heavy, slightly heavy, and not heavy tailed).
-The panels from top to bottom show nu (red), process error standard deviation (blue), growth parameter (green), density dependence parameter (purple), and autocorrelation parameter (orange).
-The larger blocks of panels show four scenarios: (1) N = 100, (2) N = 50, (3) N = 50 with ignored observation standard deviation of 0.2, and (4) a model where the model also allows for observation standard deviation of 0.2
-In general, this looks fairly good, with the caveat that if the tails are heavy, the model tends to overestimate process error standard deviation and underestimate just how heavy the tails are --- it's a bit biased towards calling heavy tails not as heavy as they are.
-Also, density independence gets consistently underestimated (purple).
-Observation error erodes the heavy-tail signal unless the tails are very heavy but doesn't tend to create false positives.
-Trying to allow for assumed observation error doesn't do much good besides fixing some very subtle bias in the process error estimate when the tails aren't heavy.
-We'd need a larger observation error to see this clearly.
-Line segments are 90\% quantiles of the posterior and the dots are the medians.}
-\label{default}
+\caption{Simulation testing the Gompertz estimation model when the process error is chosen so that $\nu$ can be estimated close to the true value outside the full population model (``effective $\nu$'' within a CV of 0.5 of specified $\nu$).
+  The simulation was run across population $\nu$ values (columns) and different scenarios (colours): (1) 100 time steps and no observation error, (2) 50 time steps and no observation error, (3) 50 time steps and observation error drawn from $\mathrm{Normal} (0, 0.3^2)$ but ignored, and (4) 50 time steps with observation error in which the quantity of observation error was assumed known.
+  Dashed horizontal lines show the true population values;
+  these true values were chosen to represent approximately the median values as estimated from the GPDD.
+  Individual dots and lines represent a stochastic draws from the true population distribution and a model fitting.
+  The panels show true values of $\nu$ =  3, 5, normal (very heavy, heavy, and not heavy tailed).
+  The panels from top to bottom show $1/\nu$, process noise scale parameter $\sigma_\mathrm{proc}$, growth rate parameter $\lambda$, and the density dependence parameter $b$.
+  Dots and line segments are medians and 80\% credible intervals of the posterior.}
+\label{fig:sim-gompertz}
 \end{center}
@@
 \includegraphics[width=\textwidth]{check-sim-box.pdf}
-\caption{Same data as previous figure but here represented as probability that nu is less than 10.
-Basically, this is looking at sensitivity/specificity.
-So, if you take something like a p = 0.5 that nu $<$ 10 as a cutoff, then you very rarely if ever get a false positive.
-However, the method doesn't \textit{always} detect heavy tails when they're there\ldots although in our data there are often tails that are much heavier than the nu = 2.5 used here.}
-\label{default}
+\caption{The same simulations results as the previous figure but here represented as probability that $\nu < 10$ (i.e.\ the probability of heavy tails).
+Within each scenario the dots represent stochastic draws from the true population distributions.}
+\label{fig:sim-prob}
 \end{center}
@@
 
-% \begin{figure}[htbp]
-%\begin{center}
-%\includegraphics[width=0.6\textwidth]{effect-of-obs-on-nu.pdf}
-%\caption{Effect of assuming that observation error is normally distributed (on a log scale) with standard deviation of 0.2.}
-%\label{default}
-%\end{center}
-%\end{figure}
-%
+\clearpage
 
-%\end{spacing}
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=0.9\textwidth]{gomp-vs-logistic-nu.pdf}
+\caption{Gompertz vs.\ Ricker logistic $\nu$ (nu) estimates.
+Line segments show 50\% credible intervals; dots show median estimates.
+The vast majority of heavy-tailed estimates are similar for the two population models.
+In a few instances one model estimates heavy tails where the other does not.
+This happens more often for the Ricker-logistic than the Gompertz model.}
+\label{fig:gomp-vs-ricker}
+\end{center}
+\end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=0.8\textwidth]{effect-of-ar1-on-nu.pdf}
+\caption{The effect of assuming no autocorrelation of residuals (vertical axis) vs.\ estimating autocorrelation of residuals (horizontal axis).
+Shown are medians and 50\% credible intervals of the posterior.
+The colours indicate $\phi$ --- the parameter estimating the lag-1 correlation of the residuals.
+The diagonal line illustrates a one-to-one relationship.
+There are a few populations in which the heavy-tail estimate changes when autocorrelation is incorporated, but a systematic difference.}
+\label{fig:ar1}
+\end{center}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=0.9\textwidth]{priors-gomp-base.pdf} \caption{Probability
+  density of the Bayesian priors for the base Gompertz model. Per capita growth
+  rate at log(abundance) = 0: $\lambda \sim \mathrm{Normal}(0, 10^2)$; scale
+  parameter of t-distribution process error: $\sigma_\mathrm{proc} \sim \mathrm{Half\mhyphen
+    Cauchy} (5)$; t-distribution degrees of freedom parameter: $\nu \sim
+  \mathrm{Exponential}(0.01)$. Not shown is $b$, the density dependence
+  parameter: $b \sim \mathrm{Uniform}(-1, 2)$.}
+
+\label{fig:priors}
+\end{center}
+\end{figure}
+
 \end{document}
+
+1 - there is evidence of heavy tails/black swans... in ecological time series... but they are rare
+
+- taxonomically widespread
+- we don't find compelling... intrinsic obvious biological reason
+- commonly extrinsic extreme events and interactions - climate and severe winters, cycles of natural enemies (parasites + predators),
+- as climate gets more variable, extremes more frequent and greater magnitude (and skewed), might expect more of these
+- and as our time series get longer and censusing gets more accurate... more ability to detect these
+- calls for policy that is be robust to the possibility of black swans
+- lead with ecological application
+
+grouse are hunted - shags have erupted
+
+2 - method
+- probabilistic
+- you can get away with 'these kinds of mistakes if you allow for heavy-tailed process errors'
+-  first with a population dynamics model
+- para on why novel
+
+hook - to what degree to extreme environmental events..
+population dynamics can often be connected to
+will be increase in...
+and then, however it remains unclear the degree to which animal populations also exhibit (too neutral, something like suffer) from extreme events...
+
+ecology letters
+
+- argue for why we do what we're doing
+
+some we were able to track down
+
+- digg into Barbary macaque
+
+decreasing median - order plot
+
+4 rows one column
+
+add little years on spark lines at beginning and end at bottom?
+
+switch to nu = 20 (or 10?) in first figure
+l
+combine fig 2 and 3 into new fig 2
+
+new fig3 is cross plots
+b, lambda, sigma_proc, N,
+log10length, log10Lifespan or min age at maturity,
+
+add 'black swan' simulation in which single value chosen at deviation = 8 or something like that
+
+new supp fig that shows cross plot of accounting or not for observation error
+
+show boxes of misclassification in cross plot figure of various model estimates
+
+line up black swans, say with birds in UK - NAO
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+citation with good justification of gompertz and other references: \citep{herrandoprez2014}
+
+The Gompertz was by far the most favoured model in \citet{brook2006} with the GPDD.
+
+I could see an argument being made that we're just seeing them because of autocorrelation in the residuals.
+Autocorrelation plays a big role in \citet{inchausti2002} justifying why they think they're seeing increasing CV of time series in the GPDD with time and an increasing CV could create heavy tails.
+They just model the spectral properties and CV of the abundance time series --- not a population dynamics model.
+I tried estimating a coefficient for the first-order autocorrelation of the residuals.
+There is some autocorrelation it picks up sometimes, but interestingly this doesn't change things that much with respect to heavy tail results.
+Nothing systematic at least (Another important reference on spectral analysis and the GPDD: \citet{garcia-carreras2011}.)
 
@@
 
-the natural logorithm versino of the logistic is often convenient for parameter estimation
-\citep{valpine2002} (and tthey cite a couple people too)
+the natural logorithm version of the logistic is often convenient for parameter estimation
+\citep{valpine2002} (and they cite a couple people too)
+
+cite Limited evidence for the demographic Allee effect from numerous species across taxa
+Stephen D. Gregory1,4, Corey J. A. Bradshaw2,3, Barry W. Brook2, and Franck Courchamp1
+Read More: http://www.esajournals.org.proxy.lib.sfu.ca/doi/full/10.1890/09-1128.1
+
+###
+
+\begin{enumerate}
+
+\item illustration of how the exponential prior slightly constrains the sampling of $\nu$\ldots illustrating what the prior probability of heavy tails is given uninformative data (about 5\% probability $\nu < 10$.
+
+\item plots of the priors overlayed with estimates across populations for all parameters
+
+\item cross plot of $\nu$ estimates from the models: no density dependence, add density dependence, add AR1, add assumed observation error
+
+\item time series plots of all heavy-tailed populations
+
+\item simulation testing of detecting nu given reduced sampling from true distribution; show repeated estimates of nu with different sample sizes, and show a few example samples with different observation window lengths and highlight the deviations that are beyond the 0.001 and 0.999 probabilities for a normal distribution.
+
+\item simulation testing of the Gompertz model with process error deviations fixed to have an effective $\nu$ estimate at the true level --- this asks how well is the Gompertz model able to partition the various parameters in a precise and unbiased way given that we know the heavy tails are there (also include a scenario with a massive black swan to show how it behaves there)
+
+\end{enumerate}
+
+%Also, we may not be crazy seeing the heavier tails for heavier longer lived animals.
+There's some precedence for that in a paper that found more red noise (longer term fluctuations dominating the variance and therefore making the CV grow with time) in large-bodied animals.
+They attributed it to a mismatch between dynamics and the observation scale.
+They used a different life-history dataset too.
+But, there are other systematic differences with the body size-nu relationship that make this sketchy for us.
+E.g.\ taxonomic orders and interval frequency of data collection vary systematically and in clumps with body size (not shown in the below figure).
+And basically the only stuff with very \textit{low} probability of heavy tails are the the small ones with generation times less than a year and small collection intervals.
+It's basically a step function.
+We may not be able to tease that all apart...
+at least not in this paper.

### 2014-09-25 3.2-get-effect-nu-seeds.R

@@ -40,13 +40,11 @@ get_effective_nu_seeds <- function(nu_true = 5, cv = 0.5, N = 50,
   return(list(seeds = seeds, nu_ests = nu_ests))
 }
 
-nu_2_seeds_N50 <- get_effective_nu_seeds(seed_N = 20, nu_true = 2)
+nu_3_seeds_N50 <- get_effective_nu_seeds(seed_N = 20, nu_true = 3)
 nu_5_seeds_N50 <- get_effective_nu_seeds(seed_N = 20, nu_true = 5)
 
-nu_2_seeds_N100 <- get_effective_nu_seeds(seed_N = 20, nu_true = 2, N = 100)
+nu_3_seeds_N100 <- get_effective_nu_seeds(seed_N = 20, nu_true = 3, N = 100)
 nu_5_seeds_N100 <- get_effective_nu_seeds(seed_N = 20, nu_true = 5, N = 100)
 
-save(nu_2_seeds_N50, nu_5_seeds_N50, nu_2_seeds_N100, nu_5_seeds_N100,
+save(nu_3_seeds_N50, nu_5_seeds_N50, nu_3_seeds_N100, nu_5_seeds_N100,
   file = "nu_effective_seeds.rda")
-
-# now go remove all N50 iterations 2 and re-run

### 2014-09-25 3.3-test-gomp-models.R

@@ -3,32 +3,31 @@
 # this file tests the ability for the Gompertz to perform well
 # when we feed it process deviations that have been selected to have the true
 # population value of nu (within a certain CV)
+#
+# Note that phi is being ignored now
 
 library(rstan)
-stan_gomp_ar1_obs <- readRDS("stan-gomp-ar1-obs.rds")
-stan_gomp_ar1 <- readRDS("stan-gomp-ar1.rds")
+#stan_gomp_ar1_obs <- readRDS("stan-gomp-ar1-obs.rds")
+#stan_gomp_ar1 <- readRDS("stan-gomp-ar1.rds")
+stan_gomp_obs <- readRDS("stan-gomp-obs.rds")
+stan_gomp <- readRDS("stan-gomp.rds")
 
-sim_gomp <- function(id = 99, phi = -0.15, nu = 10, sigma_proc = 0.45,
-  sigma_obs_true = 0.001, sigma_obs_assumed = 0.001, N = 50, b = 0.65,
-  lambda = 1.3, y1 = 5.0, iterations = 2000, warmup = 1000, plot = FALSE,
+sim_gomp <- function(id = 99, nu = 10, sigma_proc = 0.65,
+  sigma_obs_true = 0.001, sigma_obs_assumed = 0.001, N = 50, b = 0.50,
+  lambda = 1.5, y1 = 3, iterations = 2000, warmup = 1000, plot = FALSE,
   max_iter = 8000, chains = 4, seed) {
 
-  file_base <- paste0("/global/scratch/anderson/heavy/sim2/sm-",
+  file_base <- paste0("/global/scratch/anderson/heavy/sim3/sm-",
     id, "-sigma_obs_true-", sigma_obs_true, "-sigma_obs_assumed-",
     sigma_obs_assumed, "-N-", N, "-b-", b, "-sigma_proc-", sigma_proc,
-    "-nu-", nu, "-y1-", y1, "-lambda-", lambda, "-phi-", phi)
+    "-nu-", nu, "-y1-", y1, "-lambda-", lambda)
 
   # simulate data:
   y <- vector(length = N)
   set.seed(seed)
   proc_error <- metRology::rt.scaled(N, df = nu, mean = 0, sd = sigma_proc)
-  proc_error_ar1 <- vector(length = N)
-  proc_error_ar1[1] <- proc_error[1]
   for(i in 2:N) {
[…]

### 2014-09-25 4-extract-models.R

@@ -6,7 +6,8 @@ library(rstan)
 library(dplyr)
 
 extract_model <- function(id, get_phi = FALSE,
-  folder = "/global/scratch/anderson/heavy/", subfolder = "base") {
+  folder = "/global/scratch/anderson/heavy/", subfolder = "base",
+  type = "gompertz") {
 
   sm <- readRDS(paste0(folder, subfolder, "/sm-", id, ".rds"))
   samp <- extract(sm)
@@ -19,10 +20,21 @@ extract_model <- function(id, get_phi = FALSE,
     as.data.frame(t(x))
   }
 
-  lambda <- get_q("lambda")
-  nu <- get_q("nu")
-  b <- get_q("b")
-  sigma_proc <- get_q("sigma_proc")
+  if(type[1] == "gompertz") {
+    lambda <- get_q("lambda")
+    nu <- get_q("nu")
+    b <- get_q("b")
+    sigma_proc <- get_q("sigma_proc")
+    sigma_proc <- get_q("sigma_proc")
+    out <- data.frame(lambda, nu, b, sigma_proc)
+  }
+  if(type[1] == "logistic") {
+    r <- get_q("r")
+    nu <- get_q("nu")
+    K <- get_q("K")
+    sigma_proc <- get_q("sigma_proc")
+    out <- data.frame(r, nu, K, sigma_proc)
+  }
 
   if(get_phi) {
     phi <- get_q("phi")
@@ -40,7 +52,6 @@ extract_model <- function(id, get_phi = FALSE,
 
   iter <- sm@stan_args[[1]]$iter
 
-  out <- data.frame(lambda, nu, b, sigma_proc)
 
   if(get_phi) {
     out <- data.frame(out, phi)
@@ -63,8 +74,11 @@ ids <- unique(gpdd$main_id)
   #get_phi = FALSE)
 gomp_hat_ar1 <- plyr::ldply(ids, extract_model, subfolder = "ar1-5.0",
   get_phi = TRUE)
+gomp_hat_logistic <- plyr::ldply(ids, extract_model, type = "logistic",
+  subfolder = "logistic-5.0", get_phi = FALSE)
 #gomp_hat_ar1_obs <- plyr::ldply(mammals_ids$main_id, extract_model,
 #  subfolder = "ar1-obs0.2", get_phi = TRUE)
[…]

### 2014-09-25 5-shape-data.R

@@ -6,18 +6,23 @@ library(dplyr)
 
 gpdd <- readRDS("gpdd-clean.rds")
 gomp_hat_base <- readRDS("gomp_hat_base.rds")
+gomp_hat_logistic <- readRDS("gomp_hat_logistic-5.0.rds")
 gomp_hat_ar1 <- readRDS("gomp_hat_ar1-4.0.rds")
-gomp_hat_ar1_obs <- readRDS("gomp_hat_ar1_obs0.2.rds")
+#gomp_hat_ar1_obs <- readRDS("gomp_hat_ar1_obs0.2.rds")
 mammals <- readRDS("mammals.rds")
+brook <- read.csv("brook-etal.csv", stringsAsFactors = FALSE)
 
 gomp_hat_ar1 <- subset(gomp_hat_ar1,
   main_id %in% gpdd$main_id)
 gomp_hat_base <- subset(gomp_hat_base,
   main_id %in% gpdd$main_id)
-gomp_hat_ar1_obs <- subset(gomp_hat_ar1_obs,
-  main_id %in% gpdd$main_id)
+#gomp_hat_ar1_obs <- subset(gomp_hat_ar1_obs,
+  #main_id %in% gpdd$main_id)
 
 # back to reshaping
+# add pgr to gpdd:
+gpdd <- gpdd %>% group_by(main_id) %>%
+  mutate(pgr = c(diff(log(population_untransformed)), NA))
 
 # make some joings for looking at the effect of observation error
 # and estimating AR1 parameter:
@@ -26,12 +31,12 @@ names(temp) <- paste0(names(temp), "_base")
 temp <- plyr::rename(temp, c("main_id_base" = "main_id"))
 ar1_vs_base <- plyr::join(gomp_hat_ar1, temp)
 rm(temp)
-
-temp <- gomp_hat_ar1
-names(temp) <- paste0(names(temp), "_ar1_no_obs")
-temp <- plyr::rename(temp, c("main_id_ar1_no_obs" = "main_id"))
-ar1_no_obs_vs_obs <- plyr::join(gomp_hat_ar1_obs, temp)
-rm(temp)
+#
+# temp <- gomp_hat_ar1
+# names(temp) <- paste0(names(temp), "_ar1_no_obs")
+# temp <- plyr::rename(temp, c("main_id_ar1_no_obs" = "main_id"))
+# ar1_no_obs_vs_obs <- plyr::join(gomp_hat_ar1_obs, temp)
[…]

### 2014-09-25 6-plot-gomp-samples.R

@@ -3,12 +3,13 @@
 library(ggplot2)
 
 gomp_hat_ar1 <- filter(gomp_hat_ar1, taxonomic_class != "Crustacea")
-gomp_hat_ar1 <- filter(gomp_hat_base, !taxonomic_class %in% c("Crustacea", "Chondrichtyhes", "Gastropoda"))
+#gomp_hat_ar1 <- filter(gomp_hat_base, !taxonomic_class %in% c("Crustacea", "Chondrichtyhes", "Gastropoda"))
 
-p <- ggplot(ar1_vs_base, aes(1/nu_50, 1/nu_50_base, col = phi_50)) + geom_point(cex = 3.5, alpha = 0.8) + geom_segment(aes(x = 1/nu_50, xend = 1/nu_50, y = 1/nu_25_base, yend = 1/nu_75_base), lwd = 0.5, alpha = 0.2) + geom_segment(aes(x = 1/nu_25, xend = 1/nu_75, y = 1/nu_50_base, yend = 1/nu_50_base), lwd = 0.5, alpha = 0.2)+ facet_wrap(~taxonomic_class) + geom_abline(intercept = 0, slope = 1, lty = 3) + xlab(expression(nu~with~AR1)) + ylab(expression(nu~without~AR1)) + coord_fixed() + theme_bw() +  scale_y_continuous(breaks = c(0, 0.1, 0.2, 0.333, 0.5), labels = c("Infinity", "10", "5", "3", "2")) +  scale_x_continuous(breaks = c(0, 0.1, 0.2, 0.333, 0.5), labels = c("Infinity", "10", "5", "3", "2")) + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank()) + labs(colour = expression(phi)) + scale_colour_gradient2(low = scales::muted("blue"), high = scales::muted("red"), mid = "grey")
-ggsave("effect-of-ar1-on-nu.pdf", width = 8, height = 8)
-# not very different!
-# switching to AR1 to base just in name for now...
[…]

### 2014-09-25 6-plot-prior.R

@@ -1,0 +1,164 @@
+# Try estimating nu with a weakly informative prior in Stan
+# See how the ability to detect heavy tails deteriorates with sample size,
+# observation error and the degree of heavy-tailedness.
+
+# Check priors:
+plot_prior <- function(x, y, xlab) {
+  plot(x, y, type = "l", xlab = xlab, ylab = "Probability density",
+    ylim = c(0, max(y)*1.04), yaxs = "i", xaxs = "i")
+}
+
+pdf("priors-gomp-base.pdf", width = 8, height = 4)
+par(mfrow = c(1, 3), cex = 0.8)
+x <- seq(-10, 10, length.out = 200)
+plot_prior(x, dnorm(x, 0, 10), expression(lambda))
+x <- seq(0, 10, length.out = 200)
+plot_prior(x, dcauchy(x, 0, 5), expression(sigma[proc]))
+x <- seq(2, 300, length.out = 200)
+plot_prior(x, dexp(x, 0.01), expression(nu))
+dev.off()
+
+# How much probability mass is below 10 and above 2 in the prior?
+x <- rexp(2e6, rate = 0.01)
+#hist(x)
+round(length(x[x<10 & x >2])/length(x), 2)
+round(length(x[x>10])/length(x), 2)
+round(length(x[x<50 & x >2])/length(x), 2)
+round(length(x[x>50])/length(x), 2)
+# TODO check if this is actually shifted to > 2 in Stan
+
+# # Compile Stan model:
+# mstan2.obs.nu <-
+# "data {
+#   int<lower=0> N; // rows of data
+#   vector[N] y; // vector to hold observations
+#   real<lower=0> sigma_obs; // observation error standard deviation
+#   real nu_rate; // rate parameter for nu exponential prior
+# }
+# parameters {
+#   vector[N] t; // state vector of `true` values
+#   real a; // intercept (mu)
+#   real<lower=2> nu; // student-t degrees of freedom
+#   real<lower=0> sigma_proc; // process noise standard deviation
+# }
+# model {
+#   // priors:
+#   a ~ normal(0, 5);
+#   sigma_proc ~ cauchy(0, 5);
+#   nu ~ exponential(nu_rate);
[…]

### 2014-09-25 6-plot-sparks.R

@@ -16,7 +16,7 @@ get_gomp_res <- function(id_show) {
   p <- subset(gomp_hat_base, main_id == id_show)
   pop <- subset(gpdd, main_id == id_show)$population_untransformed
   res <- rep(NA, length(pop))
-  for(i in 3:length(pop)) {
+  for(i in 2:length(pop)) {
     res[i] <- log(pop)[i] - (p$lambda_50 + p$b_50 * log(pop[i-1]))
       #p$phi_50 * (p$lambda_50 + p$b_50 * log(pop[i-2]))
   }
@@ -32,8 +32,8 @@ make_spark <- function(x) {
   pdf(paste0("sparks/", x$main_id[1], ".pdf"), width = 2.5, height = 1.2)
   par(mar = c(0.5,0.2,0.5,0.2), oma = c(0,0,1.0,0), cex = 0.8)
   par(xpd = NA)
-  with(x, plot(seq_along(population_untransformed), population, type = "l", axes = FALSE,
-      xlab = "", ylab = "", yaxs = "i", xaxs = "i"))
+  with(x, plot(seq_along(population_untransformed), population_untransformed, type = "l", axes = FALSE,
+      xlab = "", ylab = "", yaxs = "i", xaxs = "i", log = "y"))
   if(length(bsw_l) > 0)
     points(bsw_l, x$population_untransformed[bsw_l], col = "red", pch = 20, cex = 3)
   if(length(bsw_u) > 0)

### 2014-09-25 6-plot-t-nu-eg.R

@@ -1,35 +1,203 @@
 # Check how student-t changes with nu
 # to justify tailing off prior around 100 or 200
-pdf("t-nu.pdf", width = 5, height = 4.5)
+#pdf("t-nu.pdf", width = 5, height = 4.5)
+
+library(rstan)
+
+bs_cols <- c(rev(RColorBrewer::brewer.pal(5, "YlOrRd"))[-5], "grey30")
+
+add_label <- function(xfrac = 0.1, yfrac = 0.1, label = "", pos = 4, ...) {
+  u <- par("usr")
+  x <- u[1] + xfrac * (u[2] - u[1])
+  y <- u[4] - yfrac * (u[4] - u[3])
+  text(x, y, label, pos = pos, ...)
+}
+
+box.col <- "grey20"
+pdf("t-nu-eg.pdf", width = 7.5, height = 4)
+par(col.axis = "grey20", col = "grey20")
+l <- rbind(c(1, 1, 1, 3, 3, 6, 6), c(1, 1, 1, 4,4, 7,7), c(2,2,2, 5,5, 8,8))
+layout(l)
 cols <- c(rev(RColorBrewer::brewer.pal(5, "YlOrRd"))[-5], "grey30")
 lwd <- 2.0
-par(mfrow = c(2, 1))
-par(mar = c(0, 4, 0, 1), oma = c(4, 0, 1, 0), las = 1)
+#par(mfrow = c(2, 1))
+par(mar = c(0, 2, 0, 0), oma = c(3.5, 2.5, 1.5, 0.5), las = 1)
 par(xpd = FALSE)
 par(cex = 0.8, tck = -0.025, mgp = c(2, 0.6, 0))
 x <- seq(-9.5, 9.5, length.out = 500)
-plot(x, dt(x, 2), type = "l", xlab = "Standard deviation",
+plot(x, dt(x, 2), type = "l", xlab = "",
   ylab = "", yaxs = "i", lty = 1, , lwd = lwd,
   xlim = c(-8, 8), ylim = c(0, 0.42), col = cols[1], xaxt = "n", yaxt = "n")
-axis(2, at = c(0, 0.2, 0.4))
+text(-7.5, 0.4, "(a)")
+mtext("Student-t distribution", side = 3, line = 0, cex = 0.9)
+axis(2, at = c(0, 0.2, 0.4), col = box.col)
 lines(x, dt(x, 3), lty = 1, col = cols[2], lwd = lwd)
[…]

### 2014-09-25 anderson-etal-blackswan-timeseries.tex
Many edits to ms

@@
 
+% Palatino
+\usepackage[full]{textcomp}
+     \usepackage[osf]{newpxtext} % osf for text, not math
+     \usepackage{cabin} % sans serif
+     \usepackage[varqu,varl]{inconsolata} % sans serif typewriter
+     \usepackage[bigdelims,vvarbb]{newpxmath} % bb from STIX
+     \usepackage[cal=boondoxo]{mathalfa} % mathcal
+
+ % Linux Libertine:
+%  \usepackage{textcomp}
+%     \usepackage[sb]{libertine}
+%     \usepackage[varqu,varl]{inconsolata}% sans serif typewriter
+%     \usepackage[libertine,bigdelims,vvarbb]{newtxmath} % bb from STIX
+%     \usepackage[cal=boondoxo]{mathalfa} % mathcal
+%     \useosf % osf for text, not math
+%     \usepackage[supstfm=libertinesups,%
+%       supscaled=1.2,%
+%       raised=-.13em]{superiors}
+
 \mathchardef\mhyphen="2D
@@
 
-\title{Black swans in ecological time series}
+\title{Black swans in ecological time series\\Black swans in population dynamics\\Heavy tailed ecological time series}
 \author{
@@
 \setcounter{page}{1}
+
 \section{Abstract}
+- introduce black swans and heavy tails, perhaps use the climate and increasing climate variability hook
+
+- we find evidence of heavy tails/black swans in ecological time series, but they are rare
+
+- taxonomically widespread although most frequent in birds, followed by mammals, insects
+
+- we don't find compelling intrinsic biological (e.g.\ lifehistory) correlates
+
+- common causes were extrinsic extreme events or multiple extreme events in a row: climate and severe winters, cycles of natural enemies (parasites and predators) and interactions between these
+
+- also mention the advance of our method: probabilistic in quantifying heavy tailedness, incorporates population dynamics, can allow for autocorrelation and observation error, and incorporate prior information
+
+- conclude with ecological, modelling, policy, and management implications
 
@@
 
+\citep{valpine2002} 
+\citep{gregory2010}
+\citet{garcia-carreras2011}
+\citet{brook2006}
+\citep{herrandoprez2014}
+
 \ldots
@@
 
-\subsection{Modelling framework}
+The data: briefly introduce the GPDD and how we filtered it. Expand in the supplement. Mention the breakdown of taxonomy and interpolation (Table~S1).
 
@@
 \begin{align*}
+x_t &= \ln N_t\\
 x_t &= \lambda + b x_{t-1} + \epsilon_t\\
-%\epsilon_t &\sim \mathrm{Student\mhyphen t}_\nu(\phi \epsilon_{t-1}, \sigma_\mathrm{proc}^2),
-\epsilon_t &\sim \mathrm{Student\mhyphen t}_\nu(0, \sigma_\mathrm{proc}),
+\epsilon_t &\sim \mathrm{Student\mhyphen t}_\nu(0, \sigma),
 \end{align*}
@@
 \noindent where $x_t$ is the $\ln$ abundance at time $t$.
-The model is density independent if $b = 1$ and maximally density dependent if $b = 0$.
-The parameter $\lambda$ represents the expected per capita growth rate at $\ln$ abundance = 0.
-The process noise is represented by a Student-t distribution that is centered at 0, has a scale parameter of $\sigma_\mathrm{proc}$, and has a degrees of freedom parameter of $\nu$.
-If $\nu$ is small ($\lesssim 10$) the distribution has heavy tails.
+The model is density independent if $b = 1$, maximally density dependent if $b = 0$, and inversely density dependence if $b < 0$.
+The parameter $\lambda$ represents the expected population growth rate at $x_t = 0$.
+The process noise, $\epsilon_t$, is modelled as a Student-t distribution that is centered at $0$, has a scale parameter of $\sigma_\mathrm{proc}$, and has a degrees of freedom parameter of $\nu$.
+If $\nu$ is small ($\lesssim 10$) the distribution has much heavier tails than a normal distribution.
+For example, at $\nu = 2$, the probability of drawing a value less than -5 is 1.8\%, whereas the probability of drawing such a value from a normal distribution is nearly zero ($2.9\cdot10^{-5}$\%).
 As $\nu$ approaches infinity the distribution approaches the normal distribution (Fig~\ref{fig:didactic}).
+By estimating the value of $\nu$ we can quantity the degree of heavy-tailedness in the process noise jumps from $\ln N_t$ to $\ln N_{t+1}$.
 
-We fit alternative models that allowed for autocorrelation of the residuals, allowed for observation error, and allowed the functional form of the population dynamics to be represented by a Ricker-logistic population model (Supporting Material).
+We fit alternative models that allowed for autocorrelation of the residuals, allowed for observation error, and allowed the functional form of the population dynamics to be represented as Ricker-logistic (Supporting Material).
+
+Mention that we simulation tested the model.
+
+Mention the priors (Fig.~\ref{fig:priors}) and expand on their justification in the supplement.
+
+$\nu$ prior: \citep{fernandez1998}
+$\sigma$ prior: \citep{gelman2006c}
+consider alternative inverse prior on $\nu$ as in \citep{gelman2014}
+
+mention data used for life-history hypothesis comparisons from \citet{brook2006a}
+
+justify why we only present these as hypotheses but do not formally model them
 
 \section{Results}
+
+- quantify how frequent and strong heavy tails were (Fig.~\ref{fig:nu-coefs})
+
+- plot showing all p($\nu < 10$) $> 0.5$ (Fig.~\ref{fig:heavy-ts})
+
+- across how many different species
+
+- break down by taxonomy  (Fig.~\ref{fig:nu-coefs} inset panels)
+
+- no obvious intrinsic biological drivers (Fig.~\ref{fig:correlates})
+
+- mention the common causes (Table~\ref{tab:sparks})
+
+- touch on the supplemental figures --- simulation results (Fig.~\ref{fig:sim-nu}, Fig.~\ref{fig:sim-gompertz}, Fig.~\ref{fig:sim-prob})
+
+- alternative models: (Fig.~\ref{fig:ar1}, Fig.~\ref{fig:gomp-vs-ricker})
 
@@
 
-- diving much deeper into what heavy tails are associated with
+- formally investigating what heavy tails are associated with
 
@@
 
-- what does this mean for conservation management (probability of crash  might be higher than our typical models tell us)
+- what does this mean for conservation management (crashes might be more frequent and greater in magnitude than our typical models tell us) --- importance of establishing polices that are robust to this
 
@@
 \caption{Example population dynamic black swans from the Global Population Dynamics Database and descriptions of their causes.
-Red and blue dots indicate downward and upward events that have a probability of occurring of 0.001\% or less if the population dynamics were explained by a Gompertz model with normally distributed process noise.}\\
+Red and blue dots indicate downward and upward events that have a probability of occurring of 0.001\% or less if the population dynamics were explained by a Gompertz model with normally distributed process noise.
+These populations are a non-random sample that we were able to verify in the primary literature.}\\
 
@@
   As $\nu$ approaches infinity the distribution approaches the normal distribution.
-  For example, the at $\nu = 2$, the probability of drawing a value less than -5 is 1.8\% whereas the probability of drawing such a value from a normal distribution is almost impossible: $2.9\cdot10^{-5}$\%.
+  For example, at $\nu = 2$, the probability of drawing a value less than -5 is 1.8\%, whereas the probability of drawing such a value from a normal distribution is nearly zero ($2.9\cdot10^{-5}$\%).
   (c--e) Simulated population dynamics from a Gompertz model with process noise drawn from t distributions with different values of $\nu$.  
+  Coloured dots in panels c and d represent jumpts with less than a 1 in 1000 chance of occurring in a normal distribution.
   (f--h) Estimates of $\nu$ from models fit to the times series in panels c--e.
   Shown are the posterior samples (histograms), median and interquartile range of the posterior (IQR) (dots and line segments), and the exponential prior on $\nu$ (dashed lines).
-  Colour shading behind panels f--h illustrates the region of heavy tails.
-  Coloured dots in panels c and d represent points with less than a 1 in 1000 chance of occurring in a normal distribution.}
+  Colour shading behind panels f--h illustrates the region of heavy tails.}
 \label{fig:didactic}
@@
 \begin{center}
-\includegraphics[width=\textwidth]{gomp-nu-base.pdf}
+\includegraphics[width=0.47\textwidth]{nu-coefs-2.pdf}
 \caption{
@@
 
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=\textwidth]{gomp-base-p10-dot-order.pdf}
-\caption{
-  Probability $\nu<20$, i.e.\ one way of asking what the probability is of heavy tails.
-  Dots represent populations.
-  Something like this could be made more pretty and sorted in some appropriate way\ldots maybe with little silhouettes for the animals.
-  Looks pretty similar for $\nu < 10$, with the probabilities a bit lower.
-  There is some vertical jitter added.
+%\begin{figure}[htbp]
+%\begin{center}
+%\includegraphics[width=\textwidth]{gomp-base-p10-dot-order.pdf}
+%\caption{
+%  Probability $\nu<20$, i.e.\ one way of asking what the probability is of heavy tails.
+%  Dots represent populations.
+%  Something like this could be made more pretty and sorted in some appropriate way\ldots maybe with little silhouettes for the animals.
+%  Looks pretty similar for $\nu < 10$, with the probabilities a bit lower.
+%  There is some vertical jitter added.
 Could do this and have another subpanel that breaks down the mammals by family, since there's the most interesting stuff happening there and there are lots of them...
 or just colour the dots by family within each panel.}
-\label{fig:dots}
-\end{center}
-\end{figure}
+%\label{fig:dots}
+%\end{center}
+%\end{figure}
 
@@
 \begin{center}
-\includegraphics[width=\textwidth]{nu-covariates-base.pdf}
+\includegraphics[width=0.6\textwidth]{nu-covariates-base.pdf}
+\includegraphics[width=0.7\textwidth]{brook-etal-correlates.pdf}
 \caption{
-  (Also see the next figure.)
+  (\textit{To be redone with 6 panels: 3 parameters, data point number, log length, minimum age or lifespan.})
   A number of hypotheses for what could be driving $\nu$ values.
-From top left to bottom right: $\phi$ (AR1 coefficient), $\log \sigma_\mathrm{proc}$ (log process error standard deviation), $N$ (time series length), $b$ (density dependence parameter in the Gompertz model; 1 is density independent and 0 is maximally density dependent), $\lambda$ (growth parameter from Gompertz model), $\log M$ (log adult body mass in grams; just for mammals and from the PanTHERIA database).
-  All dots show median estimates and vertical line segments show interquartile range of posterior for the $\nu$ values.
-  I don't generally see a whole lot going on here besides taxonomic differences but I'm open to some eye fuzzing and other ideas.
-  If we want to, I was playing with a hierarchical beta regression with different levels for the taxonomic orders explaining the variation in p(nu $<$ 10) or these data\ldots but hypothesis generation with some visuals might be enough for this paper.}
+From top left to bottom right: $\phi$ (AR1 coefficient), $\log \sigma_\mathrm{proc}$ (log process error standard deviation), $N$ (time series length), $b$ (density dependence parameter in the Gompertz model; 1 is density independent and 0 is maximally density dependent), $\lambda$ (growth parameter from Gompertz model).
+  All dots show median estimates and vertical line segments show 50\% credible intervals of posterior for the $\nu$ values.}
 \label{fig:correlates}
-\end{center}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=\textwidth]{brook-etal-correlates.pdf}
-\caption{
-  \citet{brook2006a} correlates plotted against nu estimates.
-}
-\label{fig:brook}
 \end{center}
@@
 
-TODO this paragraph is old:
+To TODO \ldots
 
-Filtering rules for the GPDD:
-We removed all populations with zeros NO LONGER TRUE (can't be fit with a Gompertz), all populations with uneven sampling intervals (end date minus start date for each step), and all populations with population estimates listed as ``harvest'' as well as those listed as from two fur trading sources (give details).
-Then, we removed any populations with 5 or more identical values in a row since these suggest either recording error or extrapolation between two observations.
-Then, we filled in all missing time steps with NAs and imputed single missing values with the geometric mean of the previous and following values (linear on the log scale that the Gompertz operates on).
-This leaves NAs that were recorded in successive time steps.
-Then, we went through each population time series and found the longest window of abundance with at least 25 time steps.
-If there were any with multiple windows of identical length, we took the most recent window.
-Finally, we removed main\_id 20531, which is a duplicate of 10139 (a heron population).
+\begin{enumerate}
+
+\item To remove populations with unreliable population indices that could be strongly confounded with economics and sampling effort, we removed all populations with a sampling protocol listed as \texttt{"harvest"} as well populations with the words \texttt{"harvest"} or \texttt{"fur"} in the cited reference title.
+\item We filled in all missing time steps with \texttt{NA} values and imputed single missing values with the geometric mean of the previous and following values.
+We chose a geometric mean to be linear on the log scale that the Gompertz and Ricker-logistic models were fit on.
+\item We filled in single recorded values of zero with the lowest non-zero value in the time series \citep[following][]{brook2006a}.
+This assumes that single values of zero result from abundance being low enough that censusing missed present individuals.
+We turned multiple zero values in a row into \texttt{NA} values.
+This implies that multiple zero values were either censusing errors or caused by emigration.
+Regardless, our population  models were fit on a multiplicative (log) scale and so could not account for zero abundance.
+\item We removed all populations with four or more identical values in a row since these suggest either recording error or extrapolation between two observations.
+\item We removed all populations without at least four unique values \citep[following][]{brook2006a}.
+\item We then went through each population time series and found the longest unbroken  window of abundance (no \texttt{NA}s) with at least 20 time steps.
+This is a longer window that used in some previous analyses \citep[e.g.][]{brook2006a}, but since our model attempts to capture the shape of the distribution tails our model requires more data.
+If there were any populations with multiple windows of identical length, we took the most recent window.
+\item Finally, we removed GPDD Main ID \texttt{20531}, which we noticed was a duplicate of \texttt{10139} (a heron population). 
+\end{enumerate}
+
 We provide a supplemental figure of all the time series included in our analysis and indicate which values were imputed (\percImputedPops\% of populations had at least one point imputed and only \percImputedPoints\% of the total observations were imputed) (Fig.\ SX).
+Table~S1 shows the final taxonomic breakdown after filtering as well as the number of populations with interpolated values.
 
-Go over zero rules.
-
-\section{Details on heavy-tailed Gompertz probability model}
+\section{Details on the heavy-tailed Gompertz probability model}
 
@@
 \nu &\sim \mathrm{Exponential}(0.01)\\
-\phi &\sim \mathrm{Truncated\mhyphen Normal}(0, 1, \mathrm{min.} = -0.99, \mathrm{max.} = 0.99)\\
 \sigma_\mathrm{proc} &\sim \mathrm{Half\mhyphen Cauchy} (5).
@@
 
-I also fitted a Ricker-logistic model:
+We also fitted a Ricker-logistic model:
 \begin{align*}
 x_t &= x_{t-1} + r_{\mathrm{max}}\left(1 - \frac{N_{t-1}}{K}\right) + b x_{t-1} + \epsilon_t\\
-%\epsilon_t &\sim \mathrm{Student\mhyphen t}_\nu(\phi \epsilon_{t-1}, \sigma_\mathrm{proc}^2),
-\epsilon_t &\sim \mathrm{Student\mhyphen t}_\nu(0, \sigma_\mathrm{proc}^2),
+\epsilon_t &\sim \mathrm{Student\mhyphen t}_\nu(0, \sigma),
 \end{align*}
+\noindent
+where  $r_\mathrm{max}$ represents the maximum population growth rate that is obtained when $N$ (abundance) $= 0$.
+The parameter $K$ represents the carrying capacity and, as before, $x_t$ represents the $\ln$ transformed abundance at time $t$.
+The Ricker-logistic model assumes a linear decrease in population growth rate ($x_t / x_{t-1}$) with increases in abundance ($N_t$).
+In contrast, the Gompertz model assumes a linear decrease in population growth rate with increases in $\ln$ abundance ($x_t$) (REF).
 
-Priors: For the Ricker-logistic $K$ uniform between 0 and twice the maximum observed abundance (\citet{clark2010} was less conservative at 0 and maximum observed).
-$r_\mathrm{max}$ uniform between 0 and 20 as in \citet{clark2010}.
-Same priors on $\nu$ and process standard deviation.
+To fit the Ricker-logistic models ,we chose a prior on $K$ uniform between 0 and twice the maximum observed abundance (\citet{clark2010} chose 0 and maximum observed, which is less conservative).
+We set the prior on $r_\mathrm{max}$ as uniform between 0 and 20 as in \citet{clark2010}.
+We used the same priors on $\nu$ and $\sigma$ as in the Gompertz model.
 
@@
 
+We considered a version of the Gompertz model in which an autoregressive parameter was fit to the process noise residuals:
+\begin{align*}
+x_t &= \lambda + b x_{t-1} + \epsilon_t\\
+\epsilon_t &\sim \mathrm{Student\mhyphen t}_\nu(\phi \epsilon_{t-1}, \sigma).
+\end{align*}
+\noindent
+In addition to the parameters in the original Gompertz model, an additional parameter, $\phi$, is estimated that represents the similarity of subsequent process noise residuals.
+Based on the results of previous analyses with the GPDD \citep[e.g.][]{connors2014} and the chosen priors in previous analyses \citep[e.g.][]{thorson2014a} and to greatly speed up chain convergence when running our model across all populations, we placed a weakly informative prior on $\phi$ that assumed the greatest probability density near zero with the reduced possibility of $\phi$ being near $-1$ or $1$.
+Specifically, we chose $\phi \sim \mathrm{Truncated\mhyphen Normal}(0, 1, \mathrm{min.} = -0.99, \mathrm{max.} = 0.99)$.
+
 \subsection{Assumed observation error}
 
-When I incorporate observation error, the model is:
+We considered a version of the base Gompertz model that allowed for a specified level of observation error:
 \begin{align*}
 U_t &= \lambda + b U_{t-1} + \epsilon_t\\
-x_t &\sim \mathrm{Normal}(U_t, \sigma_\mathrm{obs}^2)\\
-\epsilon_t &\sim \mathrm{Student\mhyphen t}_\nu(0, \sigma_\mathrm{proc}^2),
+x_t &\sim \mathrm{Normal}(U_t, \sigma_\mathrm{obs})\\
+\epsilon_t &\sim \mathrm{Student\mhyphen t}_\nu(0, \sigma_\mathrm{proc}),
 \end{align*}
@@
 \noindent
-where $U$ represents the unobserved state vector, and $\sigma_\mathrm{obs}$ represents the standard deviation of observation error (on a log scale), which was set at 0.3.
+where $U$ represents the unobserved state vector, and $\sigma_\mathrm{obs}$ represents the standard deviation of observation error (on a log scale), which was set at $0.3$. TODO justify $0.3$.
 
@@
 \begin{center}
-\includegraphics[width=0.9\textwidth]{priors-gomp-base.pdf} \caption{Probability
+\includegraphics[width=\textwidth]{priors-gomp-base.pdf} \caption{Probability
   density of the Bayesian priors for the base Gompertz model. Per capita growth
@@
   parameter: $b \sim \mathrm{Uniform}(-1, 2)$.}
+  \label{fig:priors}
+\end{center}
+\end{figure}
 
-\label{fig:priors}
+\clearpage
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=\textwidth]{all-clean-ts-3.pdf} \caption{All filtered times series.
+The abundances are shown on a log vertical axis.
+Colours indicate taxonomic classes.
+Open black circles indicate abundance values that were interpolated.
+Closed black circles indicate single abundance values that were recorded as 0 but were set to the next lowest observed abundance.}
+    \label{fig:all-ts}
 \end{center}

### 2014-09-25 stat-table.tex
Many edits to ms

@@
 % latex table generated in R 3.1.1 by xtable 1.7-3 package
-% Fri Sep 19 11:41:03 2014
+% Thu Sep 25 13:22:15 2014
 \begin{table}[ht]
 \centering
-\begin{tabular}{lrrrrrrr}
-  \hline
-taxonomic\_class & n & n\_orders & n\_species & median\_N & max\_N & n\_interpolated & n\_zero\_sub \\ 
-  \hline
-Aves & 227 &  15 & 113 &  26 & 117 &  68 &  32 \\ 
-  Insecta & 182 &   7 &  91 &  25 &  87 &  26 &  55 \\ 
-  Mammalia & 125 &   8 &  51 &  28 & 106 &  18 &  21 \\ 
-  Osteichthyes & 109 &   6 &  36 &  26 &  67 &  14 &   4 \\ 
-  Crustacea &  13 &   2 &   2 &  28 &  33 &   1 &   0 \\ 
-  Chondrichtyhes &   1 &   1 &   1 &  20 &  20 &   1 &   0 \\ 
-  Gastropoda &   1 &   1 &   1 &  21 &  21 &   0 &   0 \\ 
-   \hline
+\caption{Summary statistics for the filtered Global Population Dynamics Database time series arranged by taxonomic class.
+Columns are: number of populations, number of taxonomic orders, numbers of species, median time series length, total number of interpolated time steps, and total number of substituted zeros.}
+\begin{tabular}{lrrrrrr}
+  \toprule
+Taxonomic class & Populations & Orders & Species & Median length & Interpolated pts & Zeros pts \\ 
+  \midrule
+Aves & 227 &  15 & 113 &  26 &  68 &  32 \\ 
+  Insecta & 182 &   7 &  91 &  25 &  26 &  55 \\ 
+  Mammalia & 125 &   8 &  51 &  28 &  18 &  21 \\ 
+  Osteichthyes & 109 &   6 &  36 &  26 &  14 &   4 \\ 
+  Crustacea &  13 &   2 &   2 &  28 &   1 &   0 \\ 
+  Chondrichtyhes &   1 &   1 &   1 &  20 &   1 &   0 \\ 
+  Gastropoda &   1 &   1 &   1 &  21 &   0 &   0 \\ 
+   \bottomrule
 \end{tabular}

### 2014-09-25 anderson-etal-blackswan-timeseries.tex
Touch up fig 2

@@
 \caption{
-  (\textit{To be redone with 6 panels: 3 parameters, data point number, log length, minimum age or lifespan.})
+  (\textit{\textbf{To be redone with 6 panels: 3 parameters, data point number, log length, minimum age or lifespan.}})
   A number of hypotheses for what could be driving $\nu$ values.

### 2014-09-25 6-plot-nu-coefs.R

@@ -13,7 +13,7 @@ add_label <- function(xfrac = -0.02, yfrac = -0.04, label = "", pos = 4, ...) {
 set.seed(1) # for jittering
 pdf("nu-coefs-2.pdf", width = 3.8, height = 9.5)
 
-par(mfrow = c(4, 1), mar = c(1.3,0,0,0), oma = c(2.2, 3.5, 1, .8), tck = -0.02, mgp = c(2, 0.5, 0), col.lab = "grey20", col.axis = "grey20", col = "grey20")
+par(mfrow = c(4, 1), mar = c(1.5,0,0,0), oma = c(2.2, 3.5, 1, .8), tck = -0.02, mgp = c(2, 0.5, 0), col.axis = "grey25", col = "grey25")
 par(cex = 0.9)
 
 for(i in 1:length(classes)) {
@@ -56,7 +56,7 @@ points(x$sort_id_perc, 1/x$nu_50, pch = 21, col = "grey10", lwd = 0.5, bg = "bla
 if(unique(x$taxonomic_class) == "Aves") {
   cutoff <- 4
 } else {
-  cutoff <- 1
+  cutoff <- 0
 }
  x <- x %>% group_by(taxonomic_order) %>%
   mutate(n_pops = length(p10)) %>%
@@ -75,6 +75,7 @@ if(unique(x$taxonomic_class) == "Aves") {
  abline(h = as.numeric(as.factor(x$taxonomic_order)), col = "grey93", lwd = 0.7, lty = 1)
    par(xpd = NA)
  points(x$p10, jitter(x$p10_order, amount = 0.25), pch = 21, bg = "grey80", col = "grey40", cex = 0.7)
+ #box(col = "grey80")
    par(xpd = FALSE)
  },
  x = c(60, 95),

### 2014-09-26 anderson-etal-blackswan-timeseries.tex
Edits to ms

--- 2014-09-25 anderson-etal-blackswan-timeseries.tex
@@
 
-\title{Black swans in ecological time series\\Black swans in population dynamics\\Heavy tailed ecological time series}
+\title{Black swans in ecological time series\\Black swans in population dynamics\\Heavy-tailed ecological time series}
 \author{
@@
 
+\section{Acknowledgements}
+
+Silhouettes: \texttt{phylopic.org}: rabbit by Sarah Werning, grey heron by Ardea cinerea, hoverfly by Gareth Monger. All under Creative Commons Attribution 3.0 Unported license.
+
+GPDD
+
+Funding
+
 \bibliography{/Users/seananderson/Dropbox/tex/jshort,/Users/seananderson/Dropbox/tex/ref3}
@@
 \begin{center}
-\includegraphics[width=0.47\textwidth]{nu-coefs-2.pdf}
+\includegraphics[width=0.46\textwidth]{nu-coefs-2.pdf}
 \caption{
-  Estimate of $\nu$ from the base Gompertz model (no autocorrelation or observation error).
-  Shown are median (points), quartiles (mid line segments) and 90\% quantiles (widest line segments) of the posterior.}
-\label{fig:nu-coefs}
+Estimates of how heavy-tailed population dynamics are for \nuCoefPopN\ populations of birds, mammals, insects, and fishes.
+Small values of $\nu$ ($\lesssim 10$) suggest heavy-tailed dynamics; larger values of $\nu$ suggest approximately normal-tailed dynamics.
+Vertical points and line segments represent posterior medians and 50\% / 90\% credible intervals for individual populations.
+Inset plots show probability that $\nu < 10$ (probability of heavy tails) for populations arranged by taxonomic order.
+Taxonomic orders are sorted by decreasing mean p($\nu < 10$).
+Taxonomic orders with three or fewer populations in panel a are omitted for space.}\label{fig:nu-coefs}
 \end{center}
@@
 \caption{
-  (\textit{To be redone with 6 panels: 3 parameters, data point number, log length, minimum age or lifespan.})
+  (\textit{\textbf{To be redone in base graphics with 6 panels: 3 parameters, data point number, log length, minimum age or lifespan. Include silhouettes}})
   A number of hypotheses for what could be driving $\nu$ values.
@@
 \item To remove populations with unreliable population indices that could be strongly confounded with economics and sampling effort, we removed all populations with a sampling protocol listed as \texttt{"harvest"} as well populations with the words \texttt{"harvest"} or \texttt{"fur"} in the cited reference title.
+\item Populations with negative abundance values were assumed to be log values and transformed by taking $10$ to the power of the recorded abundance.
+In many cases this was noted for the population, but not in all cases.
+We inspected each of these  \totalAssumedLog\ time series to make sure our assumption made sense (Fig.~\ref{fig:log10-assumed}).
 \item We filled in all missing time steps with \texttt{NA} values and imputed single missing values with the geometric mean of the previous and following values.
@@
 \noindent
-In addition to the parameters in the original Gompertz model, an additional parameter, $\phi$, is estimated that represents the similarity of subsequent process noise residuals.
+In addition to the parameters in the original Gompertz model, an additional parameter, $\phi$, is estimated that represents the relationship between of subsequent process noise residuals.
 Based on the results of previous analyses with the GPDD \citep[e.g.][]{connors2014} and the chosen priors in previous analyses \citep[e.g.][]{thorson2014a} and to greatly speed up chain convergence when running our model across all populations, we placed a weakly informative prior on $\phi$ that assumed the greatest probability density near zero with the reduced possibility of $\phi$ being near $-1$ or $1$.
@@
 \begin{center}
-\includegraphics[width=\textwidth]{all-clean-ts-3.pdf} \caption{All filtered times series.
-The abundances are shown on a log vertical axis.
+\includegraphics[width=\textwidth]{log10-assumed.pdf} \caption{Populations in which the abundance was specified as or assumed to be recorded as log10 values.
+The numbers at the front of each panel label are the GPDD IDs.}
+    \label{fig:log10-assumed}
+\end{center}
+\end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=\textwidth]{all-clean-ts-3.pdf} \caption{All filtered times series used in our analysis.
+The abundances are shown on a log10 vertical axis.
 Colours indicate taxonomic classes.
 Open black circles indicate abundance values that were interpolated.
-Closed black circles indicate single abundance values that were recorded as 0 but were set to the next lowest observed abundance.}
+Closed black circles indicate single abundance values that were recorded as zero but were set to the next lowest observed abundance.}
     \label{fig:all-ts}

### 2014-09-26 7-values-for-paper.R
Output more values for the ms

@@ -22,7 +22,16 @@ perc_imputed_pops <- as.character(perc_imputed$percent_imputed)
 
 perc_imputed_points <- round(sum(gpdd$interpolated)/nrow(gpdd) * 100, 1)
 
+fig2_pop_n <- filter(gpdd, taxonomic_class %in% c("Aves", "Mammalia", "Insecta", "Osteichthyes")) %>% summarise(n = length(unique(main_id))) %>% summarise(N = sum(n))
+fig2_pop_n <- fig2_pop_n$N
+
+
+temp <- gpdd %>% group_by(main_id) %>% summarise(assume_log10 = assumed_log10[1]) %>% summarise(total_assumed_log10 = sum(assume_log10))
+total_assumed_log10 <- temp$total_assumed_log10
+
 zz <- file("values.tex", "w")
 write_tex(perc_imputed_pops, "percImputedPops")
 write_tex(perc_imputed_points, "percImputedPoints")
+write_tex(fig2_pop_n, "nuCoefPopN")
+write_tex(total_assumed_log10, "totalAssumedLog")
 close(zz)

### 2014-09-28 anderson-etal-blackswan-timeseries.tex
Ms additions

@@
 \citep{herrandoprez2014}
+
+\citep{sibly2005, ziebarth2010}
 
@@
 \begin{center}
-\includegraphics[width=0.6\textwidth]{nu-covariates-base.pdf}
-\includegraphics[width=0.7\textwidth]{brook-etal-correlates.pdf}
+\includegraphics[width=0.9\textwidth]{correlates.pdf}
 \caption{
-  (\textit{\textbf{To be redone in base graphics with 6 panels: 3 parameters, data point number, log length, minimum age or lifespan. Include silhouettes}})
-  A number of hypotheses for what could be driving $\nu$ values.
-From top left to bottom right: $\phi$ (AR1 coefficient), $\log \sigma_\mathrm{proc}$ (log process error standard deviation), $N$ (time series length), $b$ (density dependence parameter in the Gompertz model; 1 is density independent and 0 is maximally density dependent), $\lambda$ (growth parameter from Gompertz model).
-  All dots show median estimates and vertical line segments show 50\% credible intervals of posterior for the $\nu$ values.}
+Potential correlates of heavy-tailed behaviour (indicated by low $\nu$ values).
+Shown are (a--c) parameters from the Gompertz heavy-tailed population model, (b) number of time steps, (c) body length, and (d) lifespan.
+For the Gompertz parameters, $\sigma$ refers to the scale parameter of the Student-t process-noise distribution, $\lambda$ refers to the population growth rate at an abundance of one, $b$ refers to the density dependence parameter ($1$ is maximally density independent, $0$ is maximally density dependent, and $<0$ is inversely density dependent).
+For parameter values, points and lines represent medians and 50\% credible intervals of the posterior.
+Circles representing four populations are filled with white and represent a shark, crustaceans, and one gastropod.
+}
 \label{fig:correlates}
@@
 \item To remove populations with unreliable population indices that could be strongly confounded with economics and sampling effort, we removed all populations with a sampling protocol listed as \texttt{"harvest"} as well populations with the words \texttt{"harvest"} or \texttt{"fur"} in the cited reference title.
+\item We removed all populations rated as $< 2$ in the GPDD quality assessment (on a scale of $1$ to $5$, with $1$ being the lowest quality data) \citep[following][]{sibly2005, ziebarth2010}
 \item Populations with negative abundance values were assumed to be log values and transformed by taking $10$ to the power of the recorded abundance.
@@
 \item We removed all populations without at least four unique values \citep[following][]{brook2006a}.
-\item We then went through each population time series and found the longest unbroken  window of abundance (no \texttt{NA}s) with at least 20 time steps.
-This is a longer window that used in some previous analyses \citep[e.g.][]{brook2006a}, but since our model attempts to capture the shape of the distribution tails our model requires more data.
+\item We then wrote an algorithm to find the longest unbroken window of abundance (no \texttt{NA}s) with at least 20 time steps in each population time series.
 If there were any populations with multiple windows of identical length, we took the most recent window.
+This is a longer window than used in some previous analyses \citep[e.g.][]{brook2006a}, but since our model attempts to capture the shape of the distribution tails, our model requires more data.
 \item Finally, we removed GPDD Main ID \texttt{20531}, which we noticed was a duplicate of \texttt{10139} (a heron population). 
@@
 
-We provide a supplemental figure of all the time series included in our analysis and indicate which values were imputed (\percImputedPops\% of populations had at least one point imputed and only \percImputedPoints\% of the total observations were imputed) (Fig.\ SX).
-Table~S1 shows the final taxonomic breakdown after filtering as well as the number of populations with interpolated values.
+We provide a supplemental figure of all the time series included in our analysis and indicate which values were interpolated (\percImputedPops\% of populations had at least one point interpolated and only \percImputedPoints\% of the total observations were interpolated) (Fig.\ SX).
+Table~S1 shows the final taxonomic breakdown and the number of populations with interpolated values.
 
@@
 \nu &\sim \mathrm{Exponential}(0.01)\\
-\sigma_\mathrm{proc} &\sim \mathrm{Half\mhyphen Cauchy} (5).
+\sigma &\sim \mathrm{Half\mhyphen Cauchy} (5).
 \end{align*}
@@
 at least not in this paper.
+
+black swans:
+
+Pederson, N., Dyer, J.M., McEwan, R.W., Hessl, A.E., Mock, C.J., Orwig, D.A.
+et al.
+(in press).
+The legacy of episodic climatic events in shaping temperate, broadleaf forests.
+Ecol.
+Monogr., 10.1890/13– 1025.1.
+Cohn, J.P. (2000). Saving the Salton Sea: researchers work to understand its problems and provide possible solutions. Bioscience, 50, 295–301.
+
+Gelman 2006: "Large but finite values of A represent prior distributions which we call “weakly informative” because, even in the tail, they have a gentle slope (unlike, for example, a half-normal distribution) and can let the data dominate if the likelihood is strong in that region.
+"

### 2014-09-28 anderson-etal-blackswan-timeseries.tex
Add notes to ms and Stan code example

@@
 \item To remove populations with unreliable population indices that could be strongly confounded with economics and sampling effort, we removed all populations with a sampling protocol listed as \texttt{"harvest"} as well populations with the words \texttt{"harvest"} or \texttt{"fur"} in the cited reference title.
+\item We removed all populations with uneven sampling intervals. I.e\ we removed populations that didn't have a constant difference between the ``decimal year begin'' and ``decimal year end'' columns.
 \item We removed all populations rated as $< 2$ in the GPDD quality assessment (on a scale of $1$ to $5$, with $1$ being the lowest quality data) \citep[following][]{sibly2005, ziebarth2010}
@@
 
+\clearpage
+
+\noindent
+Example Stan code for heavy-tailed Gompertz model with AR1 correlated residuals without observation error.
+
+\begin{verbatim}
+data {
+  int<lower=0> N;        // number of data rows
+  vector[N] y;           // vector to hold ln abundance observations
+  real<lower=0> nu_rate; // rate parameter for nu exponential prior
+}
+parameters {
+  real lambda;                 // Gompertz growth rate parameter
+  real<lower=-1, upper=2> b;   // Gompertz density dependence parameter
+  real<lower=0> sigma_proc;    // process noise standard deviation
+  real<lower=2> nu;            // t-distribution degrees of freedom
+  real<lower=-1, upper=1> phi; // AR1 parameter
+}
+transformed parameters {
+  vector[N] epsilon;     // error terms
+  epsilon[1] <- 0;
+  for (i in 2:N) {
+    epsilon[i] <- y[i] - (lambda + b * y[i - 1])
+                       - (phi * epsilon[i - 1]);
+  }
+}
+model {
+  // priors:
+  nu ~ exponential(nu_rate);
+  lambda ~ normal(0, 10);
+  sigma_proc ~ cauchy(0, 2.5);
+  phi ~ normal(0, 1);
+  // data model:
+  for (i in 2:N) {
+    y[i] ~ student_t(nu,
+                     lambda + b * y[i - 1]
+                     + phi * epsilon[i - 1],
+                     sigma_proc);
+  }
+}
+\end{verbatim}
+
 \end{document}
@@
 "
+
+TODO - see this paper: http://onlinelibrary.wiley.com/doi/10.1046/j.1365-2656.2003.00738.x/full Population-level mechanisms for reddened spectra in ecological time series H.
+Resit Akçakaya1,*, John M.
+Halley1,2 andPablo Inchausti1,3

### 2014-09-30 anderson-etal-blackswan-timeseries.tex
Additions to ms

@@
 
+\usepackage{ragged2e}
+\setlength\RaggedRightParindent{\parindent} % default value of this parameter is `0pt`
+\RaggedRight % if invoked in preamble, entire document is set Ragged-Right
+
 \usepackage{longtable}
@@
 \usepackage{array}
-\usepackage{ragged2e}
 \usepackage{tabu} % for spacing between rows in longtable
@@
 
+\section{Main message (25 words)}
+
+Black swans are present, taxonomically widespread, but rare in population dynamics.
+Intrinsic drivers are unclear; extreme climate, predation, parasites, and their interactions are common causes.
+
 \section{Abstract}
+
 - introduce black swans and heavy tails, perhaps use the climate and increasing climate variability hook
@@
 
-Increasing realization about the importance of extreme events in the environment (REFs)
-and of the importance of ecological surprises and black swans (REFs)
+\begin{enumerate}
+  \item What are black swans and why do we care?
 
+%\bigskip
+\item Why might we expect to see  heavy tails in ecological timeseries?
+
+%\bigskip
+\item What is the existing evidence for heavy tails in ecological time series?
+
+%\bigskip
+\item Our research questions and main conclusions.
+
+\end{enumerate}
+
+%Increasing realization about the importance of extreme events in the environment (REFs)
+%and of the importance of ecological surprises and black swans (REFs)
+
+\bigskip
 Papers to work in:
@@
 \section{Methods}
+
+\begin{enumerate}
+  \item Short overview paragraph
+  \item The data: global population dynamics database
+  \item The model: Gompertz population model
+  \item Alternative models and sensitivity analyses
+  \end{enumerate}
 

### 2014-10-02 anderson-etal-blackswan-timeseries.tex
Bankruptcy commit

--- 2014-09-28 anderson-etal-blackswan-timeseries.tex
@@
-\documentclass[11pt]{article}
+\documentclass[12pt]{article}
 \usepackage{geometry}
@@
 \usepackage{amsfonts}
-%\usepackage{pdfpages}
-%\usepackage{lscape}
-%\usepackage{pdflscape}
 \usepackage{setspace}
@@
 \usepackage{lineno}
+\usepackage{ragged2e}
 
@@
 
-% Palatino
-\usepackage[full]{textcomp}
-     \usepackage[osf]{newpxtext} % osf for text, not math
-     \usepackage{cabin} % sans serif
-     \usepackage[varqu,varl]{inconsolata} % sans serif typewriter
-     \usepackage[bigdelims,vvarbb]{newpxmath} % bb from STIX
+ % Linux Libertine:
+  \usepackage{textcomp}
+     \usepackage[sb]{libertine}
+     \usepackage[varqu,varl]{inconsolata}% sans serif typewriter
+     \usepackage[libertine,bigdelims,vvarbb]{newtxmath} % bb from STIX
      \usepackage[cal=boondoxo]{mathalfa} % mathcal
-
- % Linux Libertine:
-%  \usepackage{textcomp}
-%     \usepackage[sb]{libertine}
-%     \usepackage[varqu,varl]{inconsolata}% sans serif typewriter
-%     \usepackage[libertine,bigdelims,vvarbb]{newtxmath} % bb from STIX
-%     \usepackage[cal=boondoxo]{mathalfa} % mathcal
-%     \useosf % osf for text, not math
-%     \usepackage[supstfm=libertinesups,%
-%       supscaled=1.2,%
-%       raised=-.13em]{superiors}
+     \useosf % osf for text, not math
+     \usepackage[supstfm=libertinesups,%
+       supscaled=1.2,%
+       raised=-.13em]{superiors}
 
@@
 
-\title{Black swans in ecological time series\\Black swans in population dynamics\\Heavy-tailed ecological time series}
+\title{Black swans in ecological time series\\Black swans in population dynamics\\Heavy-tailed ecological time series\\Black swans in ecological dynamics}
 \author{
@@
 \noindent
-\textsuperscript{1}Earth to Ocean Research Group, Department of Biological Sciences, Simon Fraser University, Burnaby BC, V5A 1S6, Canada
-
+\textsuperscript{1}Earth to ocean research group, department of biological sciences, Simon Fraser University, Burnaby BC, V5A 1S6, Canada
 \noindent
@@
 
+\noindent
+\textit{Main message, to be deleted (25 words)}: Black swans are present \& taxonomically widespread but rare in population dynamics.
+Extreme climate, predation, parasites, and their interactions are common causes; intrinsic drivers are unclear.
+
+%When they do occur they tend to be driven by ... little evidence....
+
 \section{Abstract}
-- introduce black swans and heavy tails, perhaps use the climate and increasing climate variability hook
 
-- we find evidence of heavy tails/black swans in ecological time series, but they are rare
+Black swans are statistically improbable events that have profound implications when they occur.
+Such events are known to occur in financial, social, and environmental systems; the magnitude of environmental extremes is projected to increase with climate change.
+But, the prevalence and magnitude of black swans in ecological population dynamics is unknown.
+Here, we develop a probability model that allows us to estimate the degree of heavy-tailedness in ecological process noise.
+We apply our model to \NPops\ time series from around the world across \NOrders\ taxonomic orders and seven classes.
+We find strong evidence of black swans, but they are rare, occurring in \overallMinPerc--\overallMaxPerc\% of populations and most frequently for birds (\AvesRangePerc\%) followed by mammals (\MammaliaRangePerc\%), insects (\InsectaRangePerc\%), and fishes (\OsteichthyesRangePerc\%).
+When they occur, they tend to be driven by climate and severe winters, cycles of natural enemies (parasites and predators) and interactions between these elements.
+We find little evidence of intrinsic life-history correlates.
+Our results suggest [FOR ECOLOGICAL MODELLING]
+and the importance of establishing management strategies that are robust to extreme events \ldots.
 
-- taxonomically widespread although most frequent in birds, followed by mammals, insects
+%lso mention the advance of our method: probabilistic in quantifying heavy tailedness, incorporates population dynamics, can allow for autocorrelation and observation error, and incorporate prior information
 
-- we don't find compelling intrinsic biological (e.g.\ lifehistory) correlates
-
-- common causes were extrinsic extreme events or multiple extreme events in a row: climate and severe winters, cycles of natural enemies (parasites and predators) and interactions between these
-
-- also mention the advance of our method: probabilistic in quantifying heavy tailedness, incorporates population dynamics, can allow for autocorrelation and observation error, and incorporate prior information
-
-- conclude with ecological, modelling, policy, and management implications
+%- conclude with ecological, modelling, policy, and management implications
 
@@
 
-Increasing realization about the importance of extreme events in the environment (REFs)
-and of the importance of ecological surprises and black swans (REFs)
+\begin{enumerate}
+\item Black swans are statistically improbable events whose occurrence has major ramifications.
+One of the most profound black swans in ecology was the asteroid marking the mass extinctions at the K-T boundary.
+Today, it is the extremeness of climate that is expected to cause the greatest societal damage.
+While extremes in the physical environment are present and widely understood, it remains unclear the extent to which ecological systems buffer or suffer from black swans.
 
+\item What is the existing evidence for heavy tails in ecological time series?
+
+\item There are two main possible reasons for why we may find little evidence of ecological black swans. (1) don't exist, (2) a paucity of appropriate methods or length/quality of time series.
+
+\item Here we develop \ldots our research questions and main conclusions.
+
+\end{enumerate}
+
+%Increasing realization about the importance of extreme events in the environment (REFs)
+%and of the importance of ecological surprises and black swans (REFs)
+
+\bigskip
 Papers to work in:
@@
 
-The data: briefly introduce the GPDD and how we filtered it. Expand in the supplement. Mention the breakdown of taxonomy and interpolation (Table~S1).
+\begin{enumerate}
+  \item Short overview paragraph
+  \item The data: global population dynamics database; briefly introduce the GPDD and how we filtered it. Expand in the supplement. Mention the breakdown of taxonomy and interpolation (Table~S1).
+  \item The heavy-tailed Gompertz population model. Mention the simulation testing but don't go into details here.
+  \item Mention the alternative population models we fit.
+  \item Modelling the possible correlates (mixed effects models with binomial or beta distributions and logit links; nested random effects for taxonomic class and taxonomic order).
+Life-history data from \citet{brook2006a}.
+  \end{enumerate}
 
@@
 x_t &= \lambda + b x_{t-1} + \epsilon_t\\
-\epsilon_t &\sim \mathrm{Student\mhyphen t}_\nu(0, \sigma),
+\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma),
 \end{align*}
@@
 As $\nu$ approaches infinity the distribution approaches the normal distribution (Fig~\ref{fig:didactic}).
-By estimating the value of $\nu$ we can quantity the degree of heavy-tailedness in the process noise jumps from $\ln N_t$ to $\ln N_{t+1}$.
+By estimating the value of $\nu$ we can quantity how heavy-tailed the process noise deviations are.
 
@@
 
-Mention that we simulation tested the model.
+%Mention that we simulation tested the model.
 
-Mention the priors (Fig.~\ref{fig:priors}) and expand on their justification in the supplement.
+%Mention the priors (Fig.~\ref{fig:priors}) and expand on their justification in the supplement.
 
-$\nu$ prior: \citep{fernandez1998}
-$\sigma$ prior: \citep{gelman2006c}
-consider alternative inverse prior on $\nu$ as in \citep{gelman2014}
+%$\nu$ prior: \citep{fernandez1998}
+%$\sigma$ prior: \citep{gelman2006c}
+%consider alternative inverse prior on $\nu$ as in \citep{gelman2014}
 
-mention data used for life-history hypothesis comparisons from \citet{brook2006a}
+%justify why we only present these as hypotheses but do not formally model them
 
-justify why we only present these as hypotheses but do not formally model them
+\bigskip
+Supporting material to be noted:
+
+\begin{enumerate}
+
+  \item Table showing the taxonomic breakdown, number of interpolated points, etc (Table~S1).
+
+  \item Plot showing time series we assumed were recorded as log transformed  (Fig.~\ref{fig:log10-assumed}).
+
+  \item Plot showing all the time series we used with interpolated and zero imputed values (Fig.~\ref{fig:all-ts}).
+
+  \item Plot showing the priors (Fig.~\ref{fig:priors}).
+\end{enumerate}
 
@@
 
-- quantify how frequent and strong heavy tails were (Fig.~\ref{fig:nu-coefs})
+Main figures and tables:
 
-- plot showing all p($\nu < 10$) $> 0.5$ (Fig.~\ref{fig:heavy-ts})
+\begin{enumerate}
+  \item Fig.~\ref{fig:didactic} illustrates the method: it shows the t-distribution tails, example simulated time series, and model fits to those time series.
 
-- across how many different species
+  \item Fig.~\ref{fig:nu-coefs} shows the posterior distributions of the estimated $\nu$ values. These are split by taxonomic class and order.
 
-- break down by taxonomy  (Fig.~\ref{fig:nu-coefs} inset panels)
+  \item Fig.~\ref{fig:correlates} shows possible biological and time-series-property correlates of heavy-tailed behaviour.
 
-- no obvious intrinsic biological drivers (Fig.~\ref{fig:correlates})
+  \item Table.~\ref{tab:sparks} selects populations that were categorized as heavy-tailed and digs into the causes. These are a non-random sample that I was able to verify in the literature.
 
-- mention the common causes (Table~\ref{tab:sparks})
+\end{enumerate}
 
-- touch on the supplemental figures --- simulation results (Fig.~\ref{fig:sim-nu}, Fig.~\ref{fig:sim-gompertz}, Fig.~\ref{fig:sim-prob})
+%- quantify how frequent and strong heavy tails were (Fig.~\ref{fig:nu-coefs})
 
-- alternative models: (Fig.~\ref{fig:ar1}, Fig.~\ref{fig:gomp-vs-ricker})
+Supporting material to be noted quickly:
+
+\begin{enumerate}
+
+  \item Simulation testing: the ability to recover $\nu$ when randomly sampling from various distributions (Fig.~\ref{fig:sim-nu}); heavy-tailed Gompertz model performance and confidence interval coverage given the process noise has deviations that are known to have effective $\nu$ equal to true $\nu$ (Fig.~\ref{fig:sim-gompertz}); boxplots of the same output (Fig.~\ref{fig:sim-gompertz-boxplots}); probability that $\nu < 10$ for the Gompertz simulation testing (Fig.~\ref{fig:sim-prob}).
+
+  \item Time series of all heavy tailed populations (Fig.~\ref{fig:heavy-ts}).
+
+  \item The effect of alternative population dynamics models on $\nu$ estimates (Fig.~\ref{fig:alt}).
+
+  \item Coefficients from modelling covariates of the probability of heavy tails (Fig.~\ref{fig:correlate-coefs}).
+
+  \item Example Stan code for heavy-tailed Gompertz model with AR1 residuals and observation error (a model including everything).
+
+  \item GPDD IDs used.
+
+\end{enumerate}
+
+Modelling result: an increase of 1 time step of data (given that you are ballpark around the mean --- 30 time steps --- to start with) results in an approximately 1\% increase in the expected probability of observing heavy tails.
+(Using Gelman's `divide by 4' rule for interpreting logistic regression coefficients.)
 
@@
 
-- why we think we see them in some taxa but not others (observation scale dynamics mismatch; observation errors...)
-
-Ways forward:
-
-- hierarchical modelling of nu and other parameters
-
-- better datasets for specific taxa (e.g.\ fish recruitment)
-
-- joint prior on scale and nu parameters
-
-- formally investigating what heavy tails are associated with
-
-- do heavy tails have conservation importance?
-
-- can we forecast their probability and develop correlates?
-
-- how can we detect them relatively quickly after they happen?
-
-- how often are heavy tails real? observation error, recording errors...
-
-- what does this mean for how we model time series (should we consider using the t as a default even if we fix nu?)
-
-- what does this mean for conservation management (crashes might be more frequent and greater in magnitude than our typical models tell us) --- importance of establishing polices that are robust to this
+\begin{enumerate}
+  \item Summary of our findings
+  \item How do our results mesh with previous related analyses?
+  \item Why might we expect to see heavy tails in ecological time-series? (mixture of normals or extreme drivers)
+  \item Are the observed frequencies by taxonomic class real or an observational phenomena?
+  \item Ways forward
+  \item Policy implications
+\end{enumerate}
 
@@
 
+Funding: SCA: SFU Graduate Fellowship, NKD: NSERC SFU \ldots, ABC: \ldots, TAB: \ldots
+
+Earth to Ocean's research group for helpful discussions
+
+Global Population Dynamics Database
+
 Silhouettes: \texttt{phylopic.org}: rabbit by Sarah Werning, grey heron by Ardea cinerea, hoverfly by Gareth Monger. All under Creative Commons Attribution 3.0 Unported license.
 
-GPDD
-
-Funding
+Compute Canada's WestGrid high-performance computing resources
 
@@
 
-\clearpage
+%\clearpage
 
@@
 \begin{center}
-\includegraphics[width=0.46\textwidth]{nu-coefs-2.pdf}
+\includegraphics[width=0.42\textwidth]{nu-coefs-2.pdf}
 \caption{
@@
 \end{figure}
-
-\clearpage
-
-%\begin{figure}[htbp]
-%\begin{center}
-%\includegraphics[width=\textwidth]{gomp-base-p10-dot-order.pdf}
-%\caption{
-%  Probability $\nu<20$, i.e.\ one way of asking what the probability is of heavy tails.
-%  Dots represent populations.
-%  Something like this could be made more pretty and sorted in some appropriate way\ldots maybe with little silhouettes for the animals.
-%  Looks pretty similar for $\nu < 10$, with the probabilities a bit lower.
-%  There is some vertical jitter added.
-Could do this and have another subpanel that breaks down the mammals by family, since there's the most interesting stuff happening there and there are lots of them...
-or just colour the dots by family within each panel.}
-%\label{fig:dots}
-%\end{center}
-%\end{figure}
 
@@
 
-Using Stan 2.4.0 \citep{stan-manual2014}, and R 3.1.1 \citep{r2014} I'm starting with 4 chains and 1000 iterations with the first 500 as warmup (i.e.\ 2000 total samples).
+We fit our models with Stan 2.4.0 \citep{stan-manual2014}, and R 3.1.1 \citep{r2014}.
+I'm starting with 4 chains and 2000 iterations with the first 1000 as warmup (i.e.\ 4000 total samples).
 If rhat is greater than 1.05 for any parameter or the minimum effective sample size is less than 200 for any parameter then I double both the total iterations and warmup and run again.
-These conditions were met by 8000 iterations (16000 total samples) in all cases and in almost all cases (X\%) within the initial 1000 iteration case (i.e.\ a total of 2000 samples).
+This continued up to 8000 iterations (16000 total samples) \ldots
 
@@
 The parameter $K$ represents the carrying capacity and, as before, $x_t$ represents the $\ln$ transformed abundance at time $t$.
-The Ricker-logistic model assumes a linear decrease in population growth rate ($x_t / x_{t-1}$) with increases in abundance ($N_t$).
+The Ricker-logistic model assumes a linear decrease in population growth rate ($x_t - x_{t-1}$) with increases in abundance ($N_t$).
 In contrast, the Gompertz model assumes a linear decrease in population growth rate with increases in $\ln$ abundance ($x_t$) (REF).
 
-To fit the Ricker-logistic models ,we chose a prior on $K$ uniform between 0 and twice the maximum observed abundance (\citet{clark2010} chose 0 and maximum observed, which is less conservative).
+To fit the Ricker-logistic models, we chose a prior on $K$ uniform between zero and twice the maximum observed abundance (\citet{clark2010} chose uniform between zero and maximum observed, which is less conservative).
 We set the prior on $r_\mathrm{max}$ as uniform between 0 and 20 as in \citet{clark2010}.
@@
 Based on the results of previous analyses with the GPDD \citep[e.g.][]{connors2014} and the chosen priors in previous analyses \citep[e.g.][]{thorson2014a} and to greatly speed up chain convergence when running our model across all populations, we placed a weakly informative prior on $\phi$ that assumed the greatest probability density near zero with the reduced possibility of $\phi$ being near $-1$ or $1$.
-Specifically, we chose $\phi \sim \mathrm{Truncated\mhyphen Normal}(0, 1, \mathrm{min.} = -0.99, \mathrm{max.} = 0.99)$.
+Specifically, we chose $\phi \sim \mathrm{Truncated\mhyphen Normal}(0, 1, \mathrm{min.} = 1, \mathrm{max.} = 1)$.
 
@@
 \noindent
-where $U$ represents the unobserved state vector, and $\sigma_\mathrm{obs}$ represents the standard deviation of observation error (on a log scale), which was set at $0.3$. TODO justify $0.3$.
+where $U$ represents the unobserved state vector, and $\sigma_\mathrm{obs}$ represents the standard deviation of observation error (on a log scale), which was set at $0.2$. TODO justify $0.2$.
 
@@
 \setcounter{figure}{0}
-
-% \begin{figure}[htbp]
-% \begin{center}
-% \includegraphics[width=\textwidth]{nu-covariates-p-base.pdf}
-% \caption{Same as previous figure but the y axis is p(nu $<$ 20), which I think might be better to use\ldots, in reality this kind of coarse dichotomy I think is more realistic.
-If p(nu $<$ 20) is high (say above 0.5) then there's a good chance the tails are heavy.
-Otherwise, we just don't know.}
-% \label{default}
-% \end{center}
-% \end{figure}
-
-%\begin{figure}[htbp]
-%\begin{center}
-%\includegraphics[width=0.7\textwidth]{p10-mammals-cross.pdf}
-%\caption{Probability that nu $<$ 10 vs.\ log10(body mass) and log10(years at sexual maturity).}
-%\label{default}
-%\end{center}
-%\end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=\textwidth]{ts-gpdd-heavy-eg-log10-no-fur.pdf}
-\caption{Time series for populations with p($\nu < 10$) $>$ 0.5 using the base Gompertz population model.
-Panels are ordered by increasing p($\nu < 10$).
-Vertical axes are on a log10 scale.
-Colours indicate taxonomic classes.
-TODO add dots for upswings and downswings as in Table 1.
-The labels on each panel indicate p($\nu < 10$) $>$ 0.5, the common name for the species, and the GPDD ID number.}
-\label{fig:heavy-ts}
-\end{center}
-\end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=0.9\textwidth]{t-dist-sampling-sim-prior-exp0point01.pdf}
-\includegraphics[width=0.9\textwidth]{t-dist-sampling-sim-sigma-prior-exp0point01.pdf}
-\caption{
-  Testing the ability to estimate $\nu$ (top panels) and the scale parameter of the process error (bottom panels) for a given number of samples (columns) drawn from a distribution with a given true $\nu$ value (rows).
-The red lines indicate the true population value.
-When a small number of samples are drawn there may not be samples sufficiently far into the tails to recapture the true $\nu$ value; however, heavy tails are still distinguished from normal tails in most cases, even with only 25 or 50 samples.
-TODO switch to nu = black swan, 3, 5, and normal.
-}
-\label{fig:sim-nu}
-\end{center}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=1.1\textwidth]{sim-gompertz.pdf}
-\caption{Simulation testing the Gompertz estimation model when the process error is chosen so that $\nu$ can be estimated close to the true value outside the full population model (``effective $\nu$'' within a CV of 0.5 of specified $\nu$).
-  The simulation was run across population $\nu$ values (columns) and different scenarios (colours): (1) 100 time steps and no observation error, (2) 50 time steps and no observation error, (3) 50 time steps and observation error drawn from $\mathrm{Normal} (0, 0.3^2)$ but ignored, and (4) 50 time steps with observation error in which the quantity of observation error was assumed known.
-  Dashed horizontal lines show the true population values;
-  these true values were chosen to represent approximately the median values as estimated from the GPDD.
-  Individual dots and lines represent a stochastic draws from the true population distribution and a model fitting.
-  The panels show true values of $\nu$ =  3, 5, normal (very heavy, heavy, and not heavy tailed).
-  The panels from top to bottom show $1/\nu$, process noise scale parameter $\sigma_\mathrm{proc}$, growth rate parameter $\lambda$, and the density dependence parameter $b$.
-  Dots and line segments are medians and 80\% credible intervals of the posterior.}
-\label{fig:sim-gompertz}
-\end{center}
-\end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=\textwidth]{check-sim-box.pdf}
-\caption{The same simulations results as the previous figure but here represented as probability that $\nu < 10$ (i.e.\ the probability of heavy tails).
-Within each scenario the dots represent stochastic draws from the true population distributions.}
-\label{fig:sim-prob}
-\end{center}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=0.9\textwidth]{gomp-vs-logistic-nu.pdf}
-\caption{Gompertz vs.\ Ricker logistic $\nu$ (nu) estimates.
-Line segments show 50\% credible intervals; dots show median estimates.
-The vast majority of heavy-tailed estimates are similar for the two population models.
-In a few instances one model estimates heavy tails where the other does not.
-This happens more often for the Ricker-logistic than the Gompertz model.}
-\label{fig:gomp-vs-ricker}
-\end{center}
-\end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=0.8\textwidth]{effect-of-ar1-on-nu.pdf}
-\caption{The effect of assuming no autocorrelation of residuals (vertical axis) vs.\ estimating autocorrelation of residuals (horizontal axis).
-Shown are medians and 50\% credible intervals of the posterior.
-The colours indicate $\phi$ --- the parameter estimating the lag-1 correlation of the residuals.
-The diagonal line illustrates a one-to-one relationship.
-There are a few populations in which the heavy-tail estimate changes when autocorrelation is incorporated, but a systematic difference.}
-\label{fig:ar1}
-\end{center}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=\textwidth]{priors-gomp-base.pdf} \caption{Probability
-  density of the Bayesian priors for the base Gompertz model. Per capita growth
-  rate at log(abundance) = 0: $\lambda \sim \mathrm{Normal}(0, 10^2)$; scale
-  parameter of t-distribution process error: $\sigma_\mathrm{proc} \sim \mathrm{Half\mhyphen
-    Cauchy} (5)$; t-distribution degrees of freedom parameter: $\nu \sim
-  \mathrm{Exponential}(0.01)$. Not shown is $b$, the density dependence
-  parameter: $b \sim \mathrm{Uniform}(-1, 2)$.}
-  \label{fig:priors}
-\end{center}
-\end{figure}
-
-\clearpage
 
@@
 
+\clearpage
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=\textwidth]{priors-gomp-base.pdf} \caption{Probability
+  density of the Bayesian priors for the base Gompertz model. From left to right: (1) per capita growth
+  rate at $\ln$(abundance) = $0$: $\lambda \sim \mathrm{Normal}(0, 10^2)$; (2) scale
+  parameter of t-distribution process noise: $\sigma \sim \mathrm{Half\mhyphen
+    Cauchy} (2.5)$; (3) t-distribution degrees of freedom parameter: $\nu \sim
+  \mathrm{Exponential}(0.01)$. (4) AR1 correlation coefficient of residuals: $\phi \sim \mathrm{Normal}(0, 1)[-1, 1]$. Not shown is $b$, the density dependence
+  parameter: $b \sim \mathrm{Uniform}(-1, 2)$.}
+  \label{fig:priors}
+\end{center}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=0.8\textwidth]{t-dist-sampling-sim-prior-exp0point01.pdf}
+\includegraphics[width=0.8\textwidth]{t-dist-sampling-sim-sigma-prior-exp0point01.pdf}
+\caption{
+  Testing the ability to estimate $\nu$ (top panels) and the scale parameter of the process error (bottom panels) for a given number of samples (columns) drawn from a distribution with a given true $\nu$ value (rows).
+The red lines indicate the true population value.
+When a small number of samples are drawn there may not be samples sufficiently far into the tails to recapture the true $\nu$ value; however, heavy tails are still distinguished from normal tails in most cases, even with only 25 or 50 samples.
+TODO switch to nu = black swan, 3, 5, and normal.
+}
+\label{fig:sim-nu}
+\end{center}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=\textwidth]{sim-gompertz.pdf}
+\caption{Simulation testing the Gompertz estimation model when the process error draws are chosen so that $\nu$ can be estimated close to the true value outside the full population model (``effective $\nu$'' within a CV of 0.2 of specified $\nu$).
+  The simulation was run across population $\nu$ values (columns) and different scenarios (colours): (1) 100 time steps and no observation error, (2) 50 time steps and no observation error, (3) 50 time steps and observation error drawn from $\mathrm{Normal} (0, 0.2^2)$ but ignored, and (4) 50 time steps with observation error in which the quantity of observation error was assumed known.
+  Dashed horizontal lines show the true population values;
+  these true values were chosen to represent approximately the median values as estimated from the GPDD.
+  Individual dots and lines represent a stochastic draws from the true population distribution and a model fitting (median and 80\% credible intervals).
+  %The panels show true values of $\nu$ =  3, 5, normal (very heavy, heavy, and not heavy tailed).
+  The panels from top to bottom show $1/\nu$, process noise scale parameter $\sigma$, growth rate parameter $\lambda$, and the density dependence parameter $b$.}
+\label{fig:sim-gompertz}
+\end{center}
+\end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=\textwidth]{sim-gompertz-boxplots.pdf}
+\caption{The distribution of median estimates from the same simulation as described in Fig.~\ref{fig:sim-gompertz}.
+The boxes show the interquartile range.
+The whiskers extend to $1.5$ times the interquartile range and outliers are shown as dots.}
+\label{fig:sim-gompertz-boxplots}
+\end{center}
+\end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=\textwidth]{check-sim-box.pdf}
+\caption{The probability that $\nu < 10$ (i.e.\ the approximate probability of heavy tails) for the same simulation scenarios as shown in Fig.~\ref{fig:sim-gompertz} and Fig.~\ref{fig:sim-gompertz-boxplots}.
+Within each scenario the dots represent stochastic draws from the true population distributions combined with model fits.}
+\label{fig:sim-prob}
+\end{center}
+\end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=0.9\textwidth]{ts-gpdd-heavy-eg-log10-base.pdf}
+\caption{Time series for populations with p($\nu < 10$) $>$ 0.5 using the base Gompertz population model.
+Panels are ordered by increasing p($\nu < 10$).
+Vertical axes are on a log10 scale.
+Colours indicate taxonomic classes.
+TODO add dots for upswings and downswings as in Table 1.
+The labels on each panel indicate p($\nu < 10$) $>$ 0.5, the common name for the species, and the GPDD ID number.}
+\label{fig:heavy-ts}
+\end{center}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=\textwidth]{gomp-comparison.pdf}
+\caption{Estimates of $\nu$ from alternative models plotted against the base Gompertz model estimates of $\nu$.
+Shown are medians of the posterior (dots) and 50\% credible intervals (segments).
+The diagonal line indicates a one-to-one relationship.
+Different colours indicate various taxonomic classes.
+The grey-shaded regions indicate regions of disagreement if $\nu = 10$ is taken as a threshold of heavy-tailed dynamics.
+The Gompertz observation error model assumes a fixed standard deviation of observation error of $0.2$ on the log scale.}
+\label{fig:alt}
+\end{center}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=0.7\textwidth]{admb-coefs}
+\includegraphics[width=0.8\textwidth]{length-vs-prob-tails}
+\caption{Potential correlates of heavy-tailed dynamics.
+  (Top panels) Dots, thick lines, and thin lines represent fixed-effect estimates and their 50\% and 95\% confidence intervals for the covariates listed on the left: $\log \sigma$, $b$, $\lambda$, $\log$(lifespan), $\log$(data set length).
+All covariates were standardized by subtracting their mean and dividing by their standard deviation.
+  The beta distribution model (left panel) uses p$(\nu < 10)$ as the response (probability of heavy tails).
+  The binomial distribution model (right panel) uses p$(\nu < 10) > 0.5$ vs.\ p$(\nu < 10) \le 0.5$ as the response (heavy vs.\ not heavy tails).
+  The models were fitted as mixed effects models with beta or binomial distributions, logit links, and multilevel intercepts for taxonomic class and order nested within class using the \texttt{glmmADMB} \texttt{R} package \citep{fournier2012,glmmadmb}.
+  (Bottom panels) Model predictions for the strongest covariate --- time series length --- with all other predictors set to zero (their mean value when standardized).
+Dots represent observations for individual populations.
+Lines and shaded regions represent estimates and 50\% and 95\% confidence intervals.
+}
+    \label{fig:correlate-coefs}
+\end{center}
+\end{figure}
+
+% \clearpage
+%
+% \begin{figure}[htbp]
+% \begin{center}
+% \includegraphics[width=0.8\textwidth]{length-vs-prob-tails}
+% \caption{Model predictions. TODO}
+%     \label{fig:length-predictions}
+% \end{center}
+% \end{figure}
+
+\clearpage
+
+\noindent
+Example Stan code for a heavy-tailed Gompertz model with AR1 correlated residuals and a specified level of observation error:
+
+\begin{spacing}{1.15}
+\begin{footnotesize}
+\begin{verbatim}
+data {
+  int<lower=3> N;              // number of observations
+  vector[N] y;                 // vector to hold ln abundance observations
+  real<lower=0> nu_rate;       // rate parameter for nu exponential prior
+}
+parameters {
+  real lambda;                 // Gompertz growth rate parameter
+  real<lower=-1, upper=2> b;   // Gompertz density dependence parameter
+  real<lower=0> sigma_proc;    // process noise scale parameter
+  real<lower=2> nu;            // t-distribution degrees of freedom
+  real<lower=-1, upper=1> phi; // AR1 parameter
+  vector[N] U;                 // unobserved states
+  real<lower=0> sigma_obs;     // specified observation error SD
+}
+transformed parameters {
+  vector[N] epsilon;           // error terms
+  epsilon[1] <- 0;
+  for (i in 2:N) {
+    epsilon[i] <- U[i] - (lambda + b * U[i - 1])
+                       - (phi * epsilon[i - 1]);
+  }
+}
+model {
+  // priors:
+  nu ~ exponential(nu_rate);
+  lambda ~ normal(0, 10);
+  sigma_proc ~ cauchy(0, 2.5);
+  phi ~ normal(0, 1);
+  // data model:
+  for (i in 2:N) {
+    U[i] ~ student_t(nu,
+                     lambda + b * U[i - 1]
+                     + phi * epsilon[i - 1],
+                     sigma_proc);
+  }
+  y ~ normal(U, sigma_obs);
+}
+\end{verbatim}
+\end{footnotesize}
+
+\clearpage
+
+\noindent
+The GPDD IDs used in our analysis:
+
+\begin{footnotesize}
+\noindent
+{\tt
+\input{mainids}
+}
+\end{footnotesize}
+\end{spacing}
 \end{document}
-
-1 - there is evidence of heavy tails/black swans... in ecological time series... but they are rare
-
-- taxonomically widespread
-- we don't find compelling... intrinsic obvious biological reason
-- commonly extrinsic extreme events and interactions - climate and severe winters, cycles of natural enemies (parasites + predators),
-- as climate gets more variable, extremes more frequent and greater magnitude (and skewed), might expect more of these
-- and as our time series get longer and censusing gets more accurate... more ability to detect these
-- calls for policy that is be robust to the possibility of black swans
-- lead with ecological application
-
-grouse are hunted - shags have erupted
-
-2 - method
-- probabilistic
-- you can get away with 'these kinds of mistakes if you allow for heavy-tailed process errors'
--  first with a population dynamics model
-- para on why novel
-
-hook - to what degree to extreme environmental events..
-population dynamics can often be connected to
-will be increase in...
-and then, however it remains unclear the degree to which animal populations also exhibit (too neutral, something like suffer) from extreme events...
-
-ecology letters
-
-- argue for why we do what we're doing
-
-some we were able to track down
-
-- digg into Barbary macaque
-
-decreasing median - order plot
-
-4 rows one column
-
-add little years on spark lines at beginning and end at bottom?
-
-switch to nu = 20 (or 10?) in first figure
-l
-combine fig 2 and 3 into new fig 2
-
-new fig3 is cross plots
-b, lambda, sigma_proc, N,
-log10length, log10Lifespan or min age at maturity,
-
-add 'black swan' simulation in which single value chosen at deviation = 8 or something like that
-
-new supp fig that shows cross plot of accounting or not for observation error
-
-show boxes of misclassification in cross plot figure of various model estimates
-
-line up black swans, say with birds in UK - NAO
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-citation with good justification of gompertz and other references: \citep{herrandoprez2014}
-
-The Gompertz was by far the most favoured model in \citet{brook2006} with the GPDD.
-
-I could see an argument being made that we're just seeing them because of autocorrelation in the residuals.
-Autocorrelation plays a big role in \citet{inchausti2002} justifying why they think they're seeing increasing CV of time series in the GPDD with time and an increasing CV could create heavy tails.
-They just model the spectral properties and CV of the abundance time series --- not a population dynamics model.
-I tried estimating a coefficient for the first-order autocorrelation of the residuals.
-There is some autocorrelation it picks up sometimes, but interestingly this doesn't change things that much with respect to heavy tail results.
-Nothing systematic at least (Another important reference on spectral analysis and the GPDD: \citet{garcia-carreras2011}.)
-
-little evidence for benefit of including annual climate or climate extremes with these data:
-\citep{gregory2010}
-and little evidence for allee effect with them
-
-the natural logorithm version of the logistic is often convenient for parameter estimation
-\citep{valpine2002} (and they cite a couple people too)
-
-cite Limited evidence for the demographic Allee effect from numerous species across taxa
-Stephen D. Gregory1,4, Corey J. A. Bradshaw2,3, Barry W. Brook2, and Franck Courchamp1
-Read More: http://www.esajournals.org.proxy.lib.sfu.ca/doi/full/10.1890/09-1128.1
-
-###
-
-\begin{enumerate}
-
-\item illustration of how the exponential prior slightly constrains the sampling of $\nu$\ldots illustrating what the prior probability of heavy tails is given uninformative data (about 5\% probability $\nu < 10$.
-
-\item plots of the priors overlayed with estimates across populations for all parameters
-
-\item cross plot of $\nu$ estimates from the models: no density dependence, add density dependence, add AR1, add assumed observation error
-
-\item time series plots of all heavy-tailed populations
-
-\item simulation testing of detecting nu given reduced sampling from true distribution; show repeated estimates of nu with different sample sizes, and show a few example samples with different observation window lengths and highlight the deviations that are beyond the 0.001 and 0.999 probabilities for a normal distribution.
-
-\item simulation testing of the Gompertz model with process error deviations fixed to have an effective $\nu$ estimate at the true level --- this asks how well is the Gompertz model able to partition the various parameters in a precise and unbiased way given that we know the heavy tails are there (also include a scenario with a massive black swan to show how it behaves there)
-
-\end{enumerate}
-
-%Also, we may not be crazy seeing the heavier tails for heavier longer lived animals.
-There's some precedence for that in a paper that found more red noise (longer term fluctuations dominating the variance and therefore making the CV grow with time) in large-bodied animals.
-They attributed it to a mismatch between dynamics and the observation scale.
-They used a different life-history dataset too.
-But, there are other systematic differences with the body size-nu relationship that make this sketchy for us.
-E.g.\ taxonomic orders and interval frequency of data collection vary systematically and in clumps with body size (not shown in the below figure).
-And basically the only stuff with very \textit{low} probability of heavy tails are the the small ones with generation times less than a year and small collection intervals.
-It's basically a step function.
-We may not be able to tease that all apart...
-at least not in this paper.
-
-black swans:
-
-Pederson, N., Dyer, J.M., McEwan, R.W., Hessl, A.E., Mock, C.J., Orwig, D.A.
-et al.
-(in press).
-The legacy of episodic climatic events in shaping temperate, broadleaf forests.
-Ecol.
-Monogr., 10.1890/13– 1025.1.
-Cohn, J.P. (2000). Saving the Salton Sea: researchers work to understand its problems and provide possible solutions. Bioscience, 50, 295–301.
-
-Gelman 2006: "Large but finite values of A represent prior distributions which we call “weakly informative” because, even in the tail, they have a gentle slope (unlike, for example, a half-normal distribution) and can let the data dominate if the likelihood is strong in that region.
-"

### 2014-10-02 1-compile-models.R

@@ -19,7 +19,6 @@ parameters {
 model {
   nu ~ exponential(nu_rate);
   lambda ~ normal(0, 10);
-  //b ~ normal(b_mu, b_sigma);
   sigma_proc ~ cauchy(0, 2.5);
   for (i in 2:N) {
     y[i] ~ student_t(nu, lambda + b * y[i-1], sigma_proc);
@@ -450,3 +449,31 @@ model {
 stan_gomp_obs_equal <- stan_model(model_code = stan_model)
 saveRDS(stan_gomp_obs_equal, file = "stan-gomp-obs-equal.rds")
 
+
+# uniform prior on sigma:
+
+stan_model <-
+  'data {
+  int<lower=0> N; // rows of data
+  vector[N] y; // vector to hold observations
+  real<lower=0> nu_rate; // rate parameter for nu exponential prior
+  real b_lower;
+  real b_upper;
+}
+parameters {
+  real lambda;
+  real<lower=b_lower, upper=b_upper> b;
+  real<lower=0, upper=20> sigma_proc;
+  real<lower=2> nu;
+}
+model {
+  nu ~ exponential(nu_rate);
+  lambda ~ normal(0, 10);
+  for (i in 2:N) {
+    y[i] ~ student_t(nu, lambda + b * y[i-1], sigma_proc);
+  }
+}
+'
+
+stan_gomp_uniform <- stan_model(model_code = stan_model)
+saveRDS(stan_gomp_uniform, file = "stan-gomp-uniform.rds")

### 2014-10-02 2-run-model-logistic.R

@@ -13,7 +13,7 @@ stan_pars <- c("r", "K", "nu", "sigma_proc")
 fit_gpdd_model(gpdd_dat = gpdd, model = model,
   sub_folder = id, stan_dat = stan_dat, pars = stan_pars)
 
-out <- plyr::ldply(unique(gpdd$main_id)[1], extract_model,
+out <- plyr::ldply(unique(gpdd$main_id), extract_model,
   sub_folder = id, get_phi = FALSE, type = "logistic")
 
 saveRDS(out, file = paste0(id, "-hat.rds"))

### 2014-10-02 2-run-models-obs-0.2.R

@@ -1,0 +1,22 @@
+# This file runs the Gompertz models with an AR1 coefficient on the residuals
+# and low assumed observation error
+
+source("1.5-compile-fit-function.R")
+source("1.6-extract-function.R")
+
+model <- readRDS("stan-gomp-obs.rds")
+gpdd <- readRDS("gpdd-clean.rds")
+id <- "gomp-obs-0.2"
+
+stan_dat <- "list(N = nrow(x), y = log(x$population_untransformed), nu_rate = 0.01, b_lower = -1, b_upper = 2, sigma_obs = 0.2)"
+
+stan_pars <- c("lambda", "sigma_proc", "nu", "b")
+
+fit_gpdd_model(gpdd_dat = gpdd, model = model,
+  sub_folder = id, stan_dat = stan_dat, iterations = 2000, warmup = 1000,
+  max_iterations = 4000, pars = stan_pars)
+
+out <- plyr::ldply(unique(gpdd$main_id), extract_model,
+  sub_folder = id, get_phi = FALSE, type = "gompertz")
+
+saveRDS(out, file = paste0(id, "-hat.rds"))

### 2014-10-02 2-run-models-rate.R

@@ -14,7 +14,7 @@ fit_gpdd_model(gpdd_dat = gpdd, model = model,
   sub_folder = id, stan_dat = stan_dat, pars = stan_pars,
   iterations = 2000, warmup = 1000)
 
-out <- plyr::ldply(unique(gpdd$main_id)[1], extract_model,
+out <- plyr::ldply(unique(gpdd$main_id), extract_model,
   sub_folder = id, get_phi = FALSE, type = "rate")
 
 saveRDS(out, file = paste0(id, "-hat.rds"))

### 2014-10-02 3.0-test-t-sampling.R

@@ -13,7 +13,7 @@ set.seed(123)
 
 N_vec <- c(1600, 800, 400, 200, 100, 50, 25)
 sigma_proc <- 1
-nu_vec <- c(2, 5, 10, 1e6)
+nu_vec <- c(1, 3, 5, 10, 1e6)
 reps <- 1:20
 show_plot <- FALSE
 
@@ -39,8 +39,8 @@ out <- plyr::ldply(reps, function(k) {
     out_nu <-  plyr::ldply(seq_along(N_vec), function(j) {
       sm <- sampling(stan_t,
         data = list(N = N_vec[j], y = y[1:N_vec[j]], nu_rate = 0.05),
-        pars = c("nu", "sigma_proc"), iter = 1000,
-        chains = 4, warmup = 500)
+        pars = c("nu", "sigma_proc"), iter = 2000,
+        chains = 4, warmup = 1000)
       max_rhat <- max(summary(sm)$summary[, "Rhat"])
       min_neff <- min(summary(sm)$summary[, "n_eff"])
       e <- extract(sm, pars = "nu")[[1]]

### 2014-10-02 3.1-plot-sample-test.R

@@ -3,6 +3,10 @@
 sample_t <- readRDS("sample-t-sim-check.rds")
 
 library(ggplot2)
+theme_set(theme_bw())
+library(dplyr)
+
+sample_t <- filter(sample_t, nu_true > 1)
 p <- ggplot(sample_t, aes(iter, 1/med_nu)) + geom_pointrange(aes(ymin = 1/l_nu, ymax = 1/u_nu)) +
   facet_grid(nu_true~N) +
   scale_y_continuous(breaks = c(0, 0.1, 0.2, 0.333, 0.5),
@@ -10,20 +14,20 @@ p <- ggplot(sample_t, aes(iter, 1/med_nu)) + geom_pointrange(aes(ymin = 1/l_nu, 
   geom_hline(aes(yintercept = 1/nu_true), col = "red") +
   xlab("Iteration") + ylab(expression(t-distribution~degrees~of~freedom~(nu)))
 
-  ggsave("t-dist-sampling-sim-prior-exp0.01.pdf", width = 8, height = 5)
+  ggsave("t-dist-sampling-sim-prior-exp0point01.pdf", width = 8, height = 5)
 
 p <- ggplot(sample_t, aes(iter, med_sigma)) + geom_pointrange(aes(ymin = l_sigma, ymax = u_sigma)) +
   facet_grid(nu_true~N) +
   geom_hline(aes(yintercept = 1), col = "red") +
   xlab("Iteration") + ylab("Scale parameter")
 
-  ggsave("t-dist-sampling-sim-sigma-prior-exp0.01.pdf", width = 8, height = 5)
+  ggsave("t-dist-sampling-sim-sigma-prior-exp0point01.pdf", width = 8, height = 5)
 
-sample_t <- readRDS("sample-t-sim-check-prior-exp-0.05.rds")
-p <- ggplot(sample_t, aes(iter, 1/med_nu)) + geom_pointrange(aes(ymin = 1/l_nu, ymax = 1/u_nu)) +
-  facet_grid(nu_true~N) +
-  scale_y_continuous(breaks = c(0, 0.1, 0.2, 0.333, 0.5),
-    labels = c("Infinity", "10", "5", "3", "2")) +
-  geom_hline(aes(yintercept = 1/nu_true), col = "red") +
-  xlab("Iteration") + ylab(expression(t-distribution~degrees~of~freedom~(nu)))
[…]

### 2014-10-02 3.2-get-effect-nu-seeds.R

@@ -1,8 +1,8 @@
 library(rstan)
 stan_t <- readRDS("stan-t.rds")
 
-get_effective_nu_seeds <- function(nu_true = 5, cv = 0.5, N = 50,
-  seed_N = 20, start_seed = 0, sigma_proc = 0.45) {
+get_effective_nu_seeds <- function(nu_true = 5, cv = 0.2, N = 50,
+  seed_N = 20, start_seed = 0, sigma_proc = 0.5) {
 
   N <- N
   sigma_proc <- sigma_proc

### 2014-10-02 3.3-test-gomp-models.R

@@ -17,7 +17,7 @@ sim_gomp <- function(id = 99, nu = 10, sigma_proc = 0.65,
   lambda = 1.5, y1 = 3, iterations = 2000, warmup = 1000, plot = FALSE,
   max_iter = 8000, chains = 4, seed) {
 
-  file_base <- paste0("/global/scratch/anderson/heavy/sim3/sm-",
+  file_base <- paste0("/global/scratch/anderson/heavy/sim4/sm-",
     id, "-sigma_obs_true-", sigma_obs_true, "-sigma_obs_assumed-",
     sigma_obs_assumed, "-N-", N, "-b-", b, "-sigma_proc-", sigma_proc,
     "-nu-", nu, "-y1-", y1, "-lambda-", lambda)
@@ -84,7 +84,7 @@ sim_gomp <- function(id = 99, nu = 10, sigma_proc = 0.65,
   out
 }
 
-iters <- 1:10
+iters <- 1:20
 
 load("nu_effective_seeds.rda")
 nu_norm_seeds <- 1:200
@@ -109,26 +109,26 @@ check_nu_N100 <- rbind(check_nu_1, check_nu_2, check_nu_4)
 # and with 50 data points and 0.3 observation error (but not estimated) to see
 # how this obscures the signal
 check_nu_1 <- plyr::ldply(iters, function(i) sim_gomp(i, nu = 3,
-    seed = nu_3_seeds_N50$seeds[i], sigma_obs_true = 0.3,
+    seed = nu_3_seeds_N50$seeds[i], sigma_obs_true = 0.2,
   sigma_obs_assumed = 0.001))
 check_nu_2 <- plyr::ldply(iters, function(i) sim_gomp(i, nu = 5,
-    seed = nu_5_seeds_N50$seeds[i], sigma_obs_true = 0.3,
+    seed = nu_5_seeds_N50$seeds[i], sigma_obs_true = 0.2,
   sigma_obs_assumed = 0.001))
 check_nu_4 <- plyr::ldply(iters, function(i) sim_gomp(i, nu = 1e9,
-    seed = nu_norm_seeds[i], sigma_obs_true = 0.3,
+    seed = nu_norm_seeds[i], sigma_obs_true = 0.2,
   sigma_obs_assumed = 0.001))
-check_nu_sigma_obs0.3_ignored <- rbind(check_nu_1, check_nu_2, check_nu_4)
[…]

### 2014-10-02 3.4-plot-test-gomp-models.R

@@ -2,11 +2,15 @@
 
 check_nu <- readRDS("check_nu.rds")
 
+load("nu_effective_seeds.rda")
+
+effective_df <- data.frame(N = c(50, 50, 50, 100, 100, 100), nu_true = c(3, 5, 1e9, 3, 5, 1e9), nu_effective = c(median(nu_3_seeds_N50$nu_ests), median(nu_5_seeds_N50$nu_ests), 1e9, median(nu_3_seeds_N100$nu_ests), median(nu_5_seeds_N100$nu_ests), 1e9))
+
 check_nu <- mutate(check_nu, inverse_nu = 1/nu, inverse_nu_l = 1/nu_l,
   inverse_nu_u = 1/nu_u, inverse_nu_true = 1/nu_true)
 
 id.vars <- c("id", "nu_true", "N", "sigma_obs_true", "sigma_obs_assumed")
-measure.vars <- c("inverse_nu", "b", "sigma_proc", "lambda", "phi")
+measure.vars <- c("inverse_nu", "b", "sigma_proc", "lambda")
 check_nu_est <- reshape2::melt(check_nu, id.vars = id.vars,
   measure.vars = measure.vars)
 check_nu_l <- reshape2::melt(check_nu, id.vars = id.vars,
@@ -35,22 +39,27 @@ check_nu_long <- data.frame(check_nu_long,
   true_value = check_nu_true[, "true_value"])
 check_nu_long$true_variable <- NULL # not needed
 
-limits <- data.frame(variable = c("inverse_nu", "sigma_proc", "lambda", "b", "phi"), ylim_l = c(0, 0, 0, -0.25, -1), ylim_u = c(0.5, 1.35, 12, 1, 1), stringsAsFactors = FALSE)
+check_nu_long <- left_join(check_nu_long, effective_df)
+check_nu_long$true_value[check_nu_long$variable == "inverse_nu"] <-
+  1/check_nu_long$nu_effective[check_nu_long$variable == "inverse_nu"]
+
+limits <- data.frame(variable = c("inverse_nu", "sigma_proc", "lambda", "b"), ylim_l = c(0, 0, 0, -0.25), ylim_u = c(0.5, 1.35, 12, 1), stringsAsFactors = FALSE)
 check_nu_long <- plyr::join(check_nu_long, limits)
 
 # and sort levels for plotting:
 check_nu_long$variable <- factor(check_nu_long$variable,
-  levels = c("inverse_nu", "sigma_proc", "lambda", "b", "phi"))
[…]

### 2014-10-02 4-extract-models.R

@@ -1,84 +1,85 @@
+# this file is no longer needed
 # this file extracts posterior and diagnostic data from all the model runs
-
-gpdd <- readRDS("gpdd-clean.rds")
-
-library(rstan)
-library(dplyr)
-
-extract_model <- function(id, get_phi = FALSE,
-  folder = "/global/scratch/anderson/heavy/", subfolder = "base",
-  type = "gompertz") {
-
-  sm <- readRDS(paste0(folder, subfolder, "/sm-", id, ".rds"))
-  samp <- extract(sm)
-  q <- lapply(samp, quantile, probs = c(0.05, 0.25, 0.5, 0.75, 0.95))
-  q <- do.call("rbind", q)
-
-  get_q <- function(par) {
-    x <- q[par, ]
-    names(x) <- paste0(par, "_", names(x))
-    as.data.frame(t(x))
-  }
-
-  if(type[1] == "gompertz") {
-    lambda <- get_q("lambda")
-    nu <- get_q("nu")
-    b <- get_q("b")
-    sigma_proc <- get_q("sigma_proc")
-    sigma_proc <- get_q("sigma_proc")
-    out <- data.frame(lambda, nu, b, sigma_proc)
-  }
-  if(type[1] == "logistic") {
-    r <- get_q("r")
-    nu <- get_q("nu")
-    K <- get_q("K")
-    sigma_proc <- get_q("sigma_proc")
-    out <- data.frame(r, nu, K, sigma_proc)
-  }
-
-  if(get_phi) {
-    phi <- get_q("phi")
-  }
-
-  nu_samples <- samp$nu
-  p10 <- length(nu_samples[nu_samples < 10]) / length(nu_samples)
-  p20 <- length(nu_samples[nu_samples < 20]) / length(nu_samples)
-
-  sm_summ <- summary(sm)$summary
-  max_rhat <- max(sm_summ[, "Rhat"])
-  min_neff <- min(sm_summ[, "n_eff"])
-  nu_rhat <- sm_summ["nu", "Rhat"]
-  nu_neff <- sm_summ["nu", "n_eff"]
-
-  iter <- sm@stan_args[[1]]$iter
-
-
-  if(get_phi) {
-    out <- data.frame(out, phi)
-  }
-
-  out <- data.frame(out, p10 = p10, p20 = p20, max_rhat = max_rhat,
[…]

### 2014-10-02 5-modelling.R

@@ -1,0 +1,67 @@
+# try some figure making
+
+library(ggplot2)
+library(mgcv)
+newdata <- data.frame(log10_Len = seq(min(gomp_hat_base$log10_Len, na.rm = TRUE), max(gomp_hat_base$log10_Len, na.rm = TRUE), length.out = 100), log10_Lifesp = seq(min(gomp_hat_base$log10_Lifesp, na.rm = TRUE), max(gomp_hat_base$log10_Lifesp, na.rm = TRUE), length.out = 100))
+
+m1 <- gam(1/nu_50 ~ s(log10_Len), data = gomp_hat_base)
+
+ggplot(gomp_hat_base, aes(log10_Len, p10, colour = taxonomic_class)) + geom_point()
+
+m2 <- gam(p10 ~ s(log10_Len), data = gomp_hat_base, family = betar(link = "logit"))
+plot(m2)
+newdata$p <- predict(m2, newdata = newdata, type = "response")
+plot(newdata$log10_Len, newdata$p)
+
+m3 <- gamm(p10 ~ s(log10_Len), data = gomp_hat_base, family = betar(link = "logit"), random = list(taxonomic_class=~1, taxonomic_order=~1), method = "REML")
+plot(m3$gam)
+
+ggplot(gomp_hat_base, aes(log10_Lifesp, p10, colour = taxonomic_class)) + geom_point()
+m4 <- gamm(p10 ~ s(log10_Lifesp), data = gomp_hat_base, family = betar(link = "logit"), random = list(taxonomic_class=~1, taxonomic_order=~1), method = "REML")
+plot(m4$gam)
+newdata$log10_Lifesp_pred <- predict(m4$gam, newdata = newdata, type = "response")
+with(newdata, plot(log10_Lifesp, log10_Lifesp_pred))
+
+m5 <- glm(p10 ~ log10_Lifesp, data = gomp_hat_base, family = quasibinomial)
+newdata$m5 <- predict(m5, newdata = newdata, type = "response")
+plot(newdata$log10_Lifesp, newdata$m5)
+
+
+library(gbm)
+m6 <- gbm(p10~log10_Lifesp + as.factor(taxonomic_class), data = gomp_hat_base)
+
+plot.gbm(m6, i.var = 1)
+
+m7 <- gbm(p10~b_50 + as.factor(taxonomic_class), data = gomp_hat_base)
+plot(m7)
+
+#library(gamm4)
[…]

### 2014-10-02 5-shape-data.R

@@ -5,24 +5,27 @@
 library(dplyr)
 
 gpdd <- readRDS("gpdd-clean.rds")
-gomp_hat_base <- readRDS("gomp_hat_base.rds")
-gomp_hat_logistic <- readRDS("gomp_hat_logistic-5.0.rds")
-gomp_hat_ar1 <- readRDS("gomp_hat_ar1-5.0.rds")
+
+gomp_hat_base <- readRDS("gomp-base-hat.rds")
+gomp_hat_logistic <- readRDS("logistic-hat.rds")
+gomp_hat_ar1 <- readRDS("gomp-ar1-hat.rds")
+gomp_hat_obs_equal <- readRDS("gomp-obs-equal-hat.rds")
+gomp_hat_obs_0.2 <- readRDS("gomp-obs-0.2-hat.rds")
+gomp_hat_rate <- readRDS("rate-hat.rds")
 #gomp_hat_ar1_obs <- readRDS("gomp_hat_ar1_obs0.2.rds")
+
 mammals <- readRDS("mammals.rds")
 brook <- read.csv("brook-etal.csv", stringsAsFactors = FALSE)
 
-gomp_hat_ar1 <- subset(gomp_hat_ar1,
-  main_id %in% gpdd$main_id)
-gomp_hat_base <- subset(gomp_hat_base,
-  main_id %in% gpdd$main_id)
+# gomp_hat_ar1 <- subset(gomp_hat_ar1,
+#   main_id %in% gpdd$main_id)
+# gomp_hat_base <- subset(gomp_hat_base,
+#   main_id %in% gpdd$main_id)
 #gomp_hat_ar1_obs <- subset(gomp_hat_ar1_obs,
   #main_id %in% gpdd$main_id)
 
 gpdd$dataset_length <- NULL
-gomp_hat_base$dataset_length <- NULL
 gpdd.temp <- gpdd %>% group_by(main_id) %>% summarise(dataset_length = n())
-gomp_hat_base <- plyr::join(gomp_hat_base, gpdd.temp)
 gpdd <- plyr::join(gpdd, gpdd.temp)
 
 # back to reshaping
@@ -30,7 +33,7 @@ gpdd <- plyr::join(gpdd, gpdd.temp)
 gpdd <- gpdd %>% group_by(main_id) %>%
   mutate(pgr = c(diff(log(population_untransformed)), NA))
 
-# make some joings for looking at the effect of observation error
[…]

### 2014-10-02 5.5-modelling.R

@@ -1,0 +1,173 @@
+library(lme4)
+
+#gomp_hat_base$log10_Lifesp_scaled <- scale(gomp_hat_base$log10_Lifesp)
+gomp_hat_base$log_Lifesp_scaled <- scale(log(gomp_hat_base$Lifesp))
+gomp_hat_base$log_sigma_proc_50_scaled <- scale(log(gomp_hat_base$sigma_proc_50))
+gomp_hat_base$b_50_scaled <- scale(gomp_hat_base$b_50)
+gomp_hat_base$lambda_50_scaled <- scale(gomp_hat_base$lambda_50)
+gomp_hat_base$log_dataset_length_scaled <- scale(log(gomp_hat_base$dataset_length))
+gomp_hat_base$log_Len_scaled <- scale(log(gomp_hat_base$Len))
+gomp_hat_base$heavy <- ifelse(gomp_hat_base$p10 > 0.5, 1, 0)
+
+# models <<- list()
+# i <<- 0
+# get_lmer_coefs <- function(param) {
+#   i <<- i + 1
+#   m1 <- glmer(heavy ~ gomp_hat_base[, param] + (1 | taxonomic_class / taxonomic_order), family = binomial, data = gomp_hat_base)
+#   models[[i]] <- m1
+#   est <- fixef(m1)[[2]]
+#   ci.50 <- tryCatch(confint(m1, parm = 4, level = 0.5), warning = function(w) c(NA, NA))
+#   ci.90 <- tryCatch(confint(m1, parm = 4, level = 0.90), warning = function(w) c(NA, NA))
+#   ret <- data.frame(est = est, l.50 = ci.50[1], u.50 = ci.50[2], l.90 = ci.90[1],
+#     u.90 = ci.90[2])
+#   print(ret)
+#   ret
+# }
+#
+# out <- data.frame(predictor =
+#   names(gomp_hat_base)[grep("scaled", names(gomp_hat_base))],
+#   stringsAsFactors = FALSE)
+#
+# out <- plyr::ddply(out, "predictor", function(x) {
+#   print(x$predictor)
+#   get_lmer_coefs(x$predictor)
+# })
+#
+# out$pos <- 1:6
+#
+# par(mar = c(4, 10, .5, .5), cex = 0.8)
+# plot(1, 1, xlim = c(-1, 1), ylim = c(1,6), type = "n", axes = FALSE,
+#   xlab = "Scaled coefficient", ylab = "")
[…]

### 2014-10-02 6-plot-alt-models.R

@@ -1,0 +1,63 @@
+cols_df <- data.frame(col =
+    c(RColorBrewer::brewer.pal(4, "Set3")[c(3, 4, 1, 2)],
+      rep("#FFFFFF", 3)),
+  taxonomic_class = c("Aves", "Mammalia", "Insecta", "Osteichthyes",
+    "Chondrichtyhes", "Crustacea", "Gastropoda"),
+  stringsAsFactors = FALSE)
+
+comp_panel <- function(xdat, ydat, xlab, ylab) {
+
+
+  not_converged <- which(ydat$max_rhat > 1.05)
+  if(length(not_converged) > 0) {
+    xdat <- xdat[-not_converged, ]
+    ydat <- ydat[-not_converged, ]
+  }
+
+  # colours:
+
+  xdat <- inner_join(xdat, lookup[,c("main_id", "taxonomic_class")])
+  #   ydat <- inner_join(ydat, lookup[,c("main_id", "taxonomic_class")])
+  xdat <- inner_join(xdat, cols_df)
+  #   ydat <- inner_join(ydat, cols_df)
+
+  set.seed(1)
+  scramble <- sample(1:nrow(xdat))
+  xdat <- xdat[scramble, ]
+  ydat <- ydat[scramble, ]
+
+  par(xpd = NA)
+  plot(1/xdat$nu_50, 1/ydat$nu_50, yaxs = "i", xaxs = "i",
+    xlim = c(0, 0.5), ylim = c(0, 0.5), xlab = xlab, ylab = ylab,
+    axes = FALSE, pch = 21, bg = xdat$col, fg = "grey40",
+    type = "n")
+  rect(0.1, 0, 0.5, 0.1, col = "grey90", border = "grey60")
+  rect(0, 0.1, 0.1, 0.5, col = "grey90", border = "grey60")
+  segments(1/xdat$nu_25, 1/ydat$nu_50, 1/xdat$nu_75, 1/ydat$nu_50,
+    col = "#00000025", lwd = 0.6)
+  segments(1/xdat$nu_50, 1/ydat$nu_25, 1/xdat$nu_50, 1/ydat$nu_75,
+    col = "#00000025", lwd = 0.6)
+  points(1/xdat$nu_50, 1/ydat$nu_50, pch = 21, bg = xdat$col,
+    fg = "grey40", cex = 1.1)
+  box()
+  ticks <- c(2, 3, 5, 10, Inf)
[…]

### 2014-10-02 6-plot-correlates.R

@@ -60,7 +60,7 @@ make_panel <- function(x, xl = NULL, xu = NULL, log = "", yaxis = FALSE,
 }
 
 pdf("correlates.pdf", width = 7, height = 4.55)
-par(mfrow = c(2, 3), mar = c(2.5,0,0,0), oma = c(.8, 4.2,2.9, .5),
+par(mfrow = c(2, 3), mar = c(2.5,0,0,0), oma = c(.8, 4.2,1.1, 5.4),
   tck = -0.02, mgp = c(2, 0.5, 0), col.axis = "grey25", col = "grey25")
 par(cex = 0.9)
 
@@ -80,9 +80,13 @@ with(gomp_hat_base_corr,
   make_panel(x = lambda_50, xl = lambda_25, xu = lambda_75,
     label = paste("Gompertz ", lambda)))
 
+with(gomp_hat_base_corr,
+  make_panel(x = b_50, xl = b_25, xu = b_75, label = paste("Gompertz ", b)))
+# text(-0.5, 1/10, "Density\ndependent", cex = 0.8, col = "grey50")
+
 add_sil <- function(x, y, file, tax_class, width_mult = 1, height_mult = 1) {
-  x <- x+5
-  y <- y + 0.02
+  #x <- x+5
+  y <- y + -0.1
   bg <- subset(cols_df, taxonomic_class == tax_class)$col
   library(grImport)
   par(xpd = NA)
@@ -90,20 +94,20 @@ add_sil <- function(x, y, file, tax_class, width_mult = 1, height_mult = 1) {
   p@paths$path@rgb <- "grey37"
   if(file == "chinook")
     p@paths[[2]]@rgb <- "grey37"  # different vector structure
-  grImport::picture(p, x, y, x + 3*width_mult, y+0.1*height_mult)
-  rect(x-2, y+0.05, x-2+ 1.2, y + 0.08, col = bg, lwd = 0.8, border = "grey50")
-  text(x-3.4, y-0.025, tax_class, col = "grey40", cex = 0.9, pos=4)
+  grImport::picture(p, x, y, x + 0.28*width_mult, y+0.1*height_mult)
+  rect(x-0.2, y+0.05, x-0.2 + 0.14, y + 0.08, col = bg, lwd = 0.8, border = "grey50")
+  text(x-0.315, y-0.025, tax_class, col = "grey40", cex = 0.9, pos=4)
   par(xpd = FALSE)
 }
 
[…]

### 2014-10-02 6-plot-eg-ts-gpdd.R

@@ -13,13 +13,18 @@ heavy <-
   mutate(id_name = paste(round(p10, 2), common_name, main_id))
 
 library(ggplot2)
+theme_set(theme_bw())
+
 p <- ggplot(heavy, aes(series_step, population_untransformed, colour = taxonomic_class)) + geom_line() + facet_wrap(~id_name, scales = "free")
 ggsave("ts-gpdd-heavy-eg-base.pdf", width = 15, height = 12)
 
-p <- ggplot(heavy, aes(series_step, log10(population_untransformed), colour = taxonomic_class)) + geom_line() + facet_wrap(~id_name, scales = "free", ncol = 4)
+p <- ggplot(heavy, aes(series_step, log10(population_untransformed), colour = taxonomic_class)) + geom_line() + facet_wrap(~id_name, scales = "free", ncol = 4) + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + geom_point()
 ggsave("ts-gpdd-heavy-eg-log10-base.pdf", width = 15, height = 17)
 
 heavy_table <- plyr::ddply(heavy, "main_id", function(x) x[1,]) %>%
   arrange(id_name)
 
 write.csv(heavy_table, file = "heavy.csv", row.names = FALSE)
+
+p<-ggplot(subset(gpdd, assumed_log10 == TRUE), aes(series_step, log10(population_untransformed))) + geom_point() + geom_line() + facet_wrap(~label, scales = "free_x") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
+ggsave("log10-assumed.pdf", width = 13, height = 13)

### 2014-10-02 6-plot-nu-coefs.R

@@ -24,38 +24,41 @@ par(cex = 0.9)
 
 for(i in 1:length(classes)) {
 
-xaxt  <- ifelse(i == 4, "s", "n")
+  xaxt  <- ifelse(i == 4, "s", "n")
 
-x <- subset(gomp_hat_base, taxonomic_class == classes[i])
+  x <- subset(gomp_hat_base, taxonomic_class == classes[i])
 
-ticks <- c(2, 3, 5, 10, Inf)
+  ticks <- c(2, 3, 5, 10, Inf)
 
-
-plot(1, 1, type = "n", xlim = c(-0.2, 100.2), ylim = c(0, 0.5), xlab = "", ylab = "", yaxt = "n", xaxs = "i", yaxs = "i", xaxt = xaxt, las = 1)
-#axis(2, at = seq(0, 100, 50), las = 1)
-par(xpd = NA)
-add_label(label = paste0("(", letters[i], ") ", classes[i]))
-if(i == 1) {
-  mtext("Heavy tails", side = 2, line = 1.8, adj = 0.90, col = "grey55")
-  mtext("Normal tails", side = 2, line = 1.8, adj = -0.09, col = "grey55")
-}
-par(xpd = FALSE)
-axis(2, las = 1, at = 1/ticks, labels = ticks)
+  plot(1, 1, type = "n", xlim = c(-0.2, 100.2), ylim = c(0, 0.5), xlab = "", ylab = "", yaxt = "n", xaxs = "i", yaxs = "i", xaxt = xaxt, las = 1)
+  #axis(2, at = seq(0, 100, 50), las = 1)
+  par(xpd = NA)
+  add_label(label = paste0("(", letters[i], ") ", classes[i]))
+  if(i == 1) {
+    mtext("Heavy tails", side = 2, line = 1.8, adj = 0.90, col = "grey55")
+    mtext("Normal tails", side = 2, line = 1.8, adj = -0.09, col = "grey55")
+  }
+  par(xpd = FALSE)
+  axis(2, las = 1, at = 1/ticks, labels = ticks)
 
   #locs <- 1/ticks
   #for(j in 1:length(cols)) {
-    #rect(0, locs[j], 100, locs[j+1], border = NA,
-      #col = paste0(cols[j], "35"))
+  #rect(0, locs[j], 100, locs[j+1], border = NA,
+  #col = paste0(cols[j], "35"))
   #}
 
[…]

### 2014-10-02 6-plot-prior.R

@@ -3,19 +3,21 @@
 # observation error and the degree of heavy-tailedness.
 
 # Check priors:
-plot_prior <- function(x, y, xlab) {
+plot_prior <- function(x, y, xlab, xlim = NULL) {
   plot(x, y, type = "l", xlab = xlab, ylab = "Probability density",
-    ylim = c(0, max(y)*1.04), yaxs = "i", xaxs = "i")
+    ylim = c(0, max(y)*1.04), yaxs = "i", xaxs = "i", xlim = xlim)
 }
 
-pdf("priors-gomp-base.pdf", width = 8, height = 4)
-par(mfrow = c(1, 3), cex = 0.8)
+pdf("priors-gomp-base.pdf", width = 10, height = 3.7)
+par(mfrow = c(1, 4), cex = 0.8)
 x <- seq(-10, 10, length.out = 200)
 plot_prior(x, dnorm(x, 0, 10), expression(lambda))
 x <- seq(0, 10, length.out = 200)
-plot_prior(x, dcauchy(x, 0, 5), expression(sigma[proc]))
+plot_prior(x, dcauchy(x, 0, 2.5), expression(sigma[proc]))
 x <- seq(2, 300, length.out = 200)
 plot_prior(x, dexp(x, 0.01), expression(nu))
+x <- seq(-1.1, 1.1, length.out = 200)
+plot_prior(x, dnorm(x, 0, 1) * dunif(x, -1, 1), expression(phi), xlim = c(-1.1, 1.1))
 dev.off()
 
 # How much probability mass is below 10 and above 2 in the prior?

### 2014-10-02 6-plot-t-nu-eg.R

@@ -33,10 +33,10 @@ mtext("Student-t distribution", side = 3, line = 0, cex = 0.9)
 axis(2, at = c(0, 0.2, 0.4), col = box.col)
 lines(x, dt(x, 3), lty = 1, col = cols[2], lwd = lwd)
 lines(x, dt(x, 5), lty = 1, col = cols[3], lwd = lwd)
-lines(x, dt(x, 50), lty = 1, col = cols[4], lwd = lwd)
+lines(x, dt(x, 20), lty = 1, col = cols[4], lwd = lwd)
 lines(x, dt(x, 1e9), lty = 1, col = cols[5], lwd = lwd)
 legend("topright", legend = c(expression(nu==2), expression(nu==3),
-  expression(nu==5), expression(nu==50), expression(Normal)),
+  expression(nu==5), expression(nu==20), expression(Normal)),
   col = cols, bty = "n", lty = c(1, 1, 1, 1, 1),
   lwd = c(lwd, lwd, lwd, lwd, lwd))
 # zoomed panel:
@@ -47,7 +47,7 @@ text(-7.5, 0.022, "(b)")
 axis(2, at = c(0, 0.02), col = box.col)
 lines(x, dt(x, 3), lty = 1, col = cols[2], lwd = lwd)
 lines(x, dt(x, 5), lty = 1, col = cols[3], lwd = lwd)
-lines(x, dt(x, 50), lty = 1, col = cols[4], lwd = lwd)
+lines(x, dt(x, 20), lty = 1, col = cols[4], lwd = lwd)
 lines(x, dt(x, 1e6), lty = 1, col = cols[5], lwd = lwd)
 par(xpd = NA)
 mtext("Probability density", side = 2, outer = TRUE, line = 1.2, las = 0, cex = 0.9)

### 2014-10-02 7-values-for-paper.R

@@ -25,6 +25,38 @@ perc_imputed_points <- round(sum(gpdd$interpolated)/nrow(gpdd) * 100, 1)
 fig2_pop_n <- filter(gpdd, taxonomic_class %in% c("Aves", "Mammalia", "Insecta", "Osteichthyes")) %>% summarise(n = length(unique(main_id))) %>% summarise(N = sum(n))
 fig2_pop_n <- fig2_pop_n$N
 
+NPops <- length(unique(gpdd$main_id))
+NOrders <- length(unique(gpdd$taxonomic_order))
+NClasses <- length(unique(gpdd$taxonomic_class))
+
+
+# p_base <- round(length(which(gomp_hat_base$p10 > 0.5)) / NPops * 100)
+# p_obs <- round(length(which(gomp_hat_obs_0.2$p10 > 0.5)) / NPops * 100)
+# p_logistic <- round(length(which(gomp_hat_logistic$p10 > 0.5)) / NPops * 100)
+# p_rate <- round(length(which(gomp_hat_rate$p10 > 0.5)) / NPops * 100)
+# MinPNu10 <- min(p_base, p_obs, p_logistic)
+# MaxPNu10 <- max(p_base, p_obs, p_logistic)
+
+cols_include <- c("taxonomic_class", "main_id", "p10", "nu_50", "type",
+  "max_rhat", "min_neff")
+gomp_hat_base$ type <- "base"
+gomp_hat_ar1$type <- "ar1"
+gomp_hat_logistic$type <- "logistic"
+gomp_hat_obs_0.2$type <- "obs"
+
+gtemp <- rbind(
+  gomp_hat_base[,cols_include],
+  gomp_hat_ar1[,cols_include],
+  gomp_hat_logistic[,cols_include],
+  gomp_hat_obs_0.2[,cols_include])
+
+# gtemp %>% filter(max_rhat < 1.05) %>% group_by(taxonomic_class, type) %>% summarise(n = n(), h =  length(which(p10 > 0.5)), p = round(100 * h / n))
+
+pheavy_class <- gtemp %>% filter(max_rhat < 1.05) %>% group_by(taxonomic_class, type) %>% summarise(n = n(), h =  length(which(p10 > 0.5)), p = round(100 * h / n)) %>% group_by(taxonomic_class) %>% summarise(min_p = min(p), max_p = max(p)) %>% filter(min_p > 0)
+
+
+
[…]

### 2014-10-02 anderson-etal-blackswan-timeseries.tex
Edits to ms

@@
 \usepackage{ragged2e}
+\usepackage{color}
 
@@
 
-The Student-t Gompertz model is:
-\begin{align*}
-x_t &= \ln N_t\\
-x_t &= \lambda + b x_{t-1} + \epsilon_t\\
-\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma),
-\end{align*}
+The heavy-tailed Gompertz model is:
+\begin{equation*}
+%\begin{align*}
+\ln N_t = x_t = \mathrm{Student\mhyphen t}( {\color{red} \nu }, \lambda + b x_{t-1}, \sigma),
+%\end{align*}
+\end{equation*}
 
-\noindent where $x_t$ is the $\ln$ abundance at time $t$.
+\noindent where $x_t$ is the $\ln$ abundance $N$ at time $t$.
 The model is density independent if $b = 1$, maximally density dependent if $b = 0$, and inversely density dependence if $b < 0$.
 The parameter $\lambda$ represents the expected population growth rate at $x_t = 0$.
-The process noise, $\epsilon_t$, is modelled as a Student-t distribution that is centered at $0$, has a scale parameter of $\sigma_\mathrm{proc}$, and has a degrees of freedom parameter of $\nu$.
+The process noise is modelled as a Student-t distribution with a scale parameter of $\sigma_\mathrm{proc}$, and a degrees of freedom parameter of $\nu$ (highlighted in red).
 If $\nu$ is small ($\lesssim 10$) the distribution has much heavier tails than a normal distribution.
@@
 Taxonomic orders are sorted by decreasing mean p($\nu < 10$).
-Taxonomic orders with three or fewer populations in panel a are omitted for space.}\label{fig:nu-coefs}
+Taxonomic orders with three or fewer populations in panel a are omitted for space.
+Red and orange colours highlight populations with a high probability of heavy-tailed dynamics.
+}
+\label{fig:nu-coefs}
 \end{center}
@@
 \nu &\sim \mathrm{Exponential}(0.01)\\
-\sigma &\sim \mathrm{Half\mhyphen Cauchy} (5).
+\sigma &\sim \mathrm{Half\mhyphen Cauchy} (2.5).
 \end{align*}
@@
 Based on the results of previous analyses with the GPDD \citep[e.g.][]{connors2014} and the chosen priors in previous analyses \citep[e.g.][]{thorson2014a} and to greatly speed up chain convergence when running our model across all populations, we placed a weakly informative prior on $\phi$ that assumed the greatest probability density near zero with the reduced possibility of $\phi$ being near $-1$ or $1$.
-Specifically, we chose $\phi \sim \mathrm{Truncated\mhyphen Normal}(0, 1, \mathrm{min.} = 1, \mathrm{max.} = 1)$.
+Specifically, we chose $\phi \sim \mathrm{Truncated\mhyphen Normal}(0, 1, \mathrm{min.} = -1, \mathrm{max.} = 1)$.
 
@@
     Cauchy} (2.5)$; (3) t-distribution degrees of freedom parameter: $\nu \sim
-  \mathrm{Exponential}(0.01)$. (4) AR1 correlation coefficient of residuals: $\phi \sim \mathrm{Normal}(0, 1)[-1, 1]$. Not shown is $b$, the density dependence
+  \mathrm{Exponential}(0.01)$.
+(4) AR1 correlation coefficient of residuals: $\phi \sim \mathrm{Truncated \mhyphen Normal}(0, 1, \mathrm{min.} = -1, \mathrm{max.} = 1)$.
+Not shown is $b$, the density dependence
   parameter: $b \sim \mathrm{Uniform}(-1, 2)$.}

### 2014-10-02 anderson-etal-blackswan-timeseries.tex
Switch to markdown ms; add to ms

--- 2014-10-02 anderson-etal-blackswan-timeseries.tex
@@
 
+\setlength\parskip{0.16in}
+\setlength\parindent{0in}
+\usepackage{titlesec}
+\titlespacing\section{0pt}{12pt plus 4pt minus 2pt}{-6pt plus 2pt minus 2pt}
+\titlespacing\subsection{0pt}{12pt plus 4pt minus 2pt}{-8pt plus 2pt minus 2pt}
+
 \begin{document}
@@
 
-\noindent
-\textit{Authors are listed alphabetically (ABCD!) and open to reordering.}
+\input{ms}
 
-\noindent
-\textsuperscript{1}Earth to ocean research group, department of biological sciences, Simon Fraser University, Burnaby BC, V5A 1S6, Canada
-\noindent
-\textsuperscript{2}School of Aquatic and Fishery Sciences, University of Washington, Box 355020, Seattle, WA 98195, USA
-
-\noindent
-\textsuperscript{3}School of Resource and Environmental Management, Simon Fraser University, Burnaby, BC, V5A 1S6, Canada
-
-\noindent
-\textsuperscript{*}Corresponding author: Sean C.
-Anderson; Earth to Ocean Research Group, Department of Biological Sciences, Simon Fraser University, Burnaby BC, V5A 1S6; Phone: 1-778-782-3989; E-mail: sean\_anderson@sfu.ca
+\bibliography{/Users/seananderson/Dropbox/tex/jshort,/Users/seananderson/Dropbox/tex/ref3}
+%\bibliography{/Users/seananderson/Dropbox/tex/jshort,refs}
 
 \clearpage
+\section{Tables}
 
-\setcounter{page}{1}
-
-\noindent
-\textit{Main message, to be deleted (25 words)}: Black swans are present \& taxonomically widespread but rare in population dynamics.
-Extreme climate, predation, parasites, and their interactions are common causes; intrinsic drivers are unclear.
-
-%When they do occur they tend to be driven by ... little evidence....
-
-\section{Abstract}
-
-Black swans are statistically improbable events that have profound implications when they occur.
-Such events are known to occur in financial, social, and environmental systems; the magnitude of environmental extremes is projected to increase with climate change.
-But, the prevalence and magnitude of black swans in ecological population dynamics is unknown.
-Here, we develop a probability model that allows us to estimate the degree of heavy-tailedness in ecological process noise.
-We apply our model to \NPops\ time series from around the world across \NOrders\ taxonomic orders and seven classes.
-We find strong evidence of black swans, but they are rare, occurring in \overallMinPerc--\overallMaxPerc\% of populations and most frequently for birds (\AvesRangePerc\%) followed by mammals (\MammaliaRangePerc\%), insects (\InsectaRangePerc\%), and fishes (\OsteichthyesRangePerc\%).
-When they occur, they tend to be driven by climate and severe winters, cycles of natural enemies (parasites and predators) and interactions between these elements.
-We find little evidence of intrinsic life-history correlates.
-Our results suggest [FOR ECOLOGICAL MODELLING]
-and the importance of establishing management strategies that are robust to extreme events \ldots.
-
-%lso mention the advance of our method: probabilistic in quantifying heavy tailedness, incorporates population dynamics, can allow for autocorrelation and observation error, and incorporate prior information
-
-%- conclude with ecological, modelling, policy, and management implications
-
-\section{Introduction}
-
-\begin{enumerate}
-\item Black swans are statistically improbable events whose occurrence has major ramifications.
-One of the most profound black swans in ecology was the asteroid marking the mass extinctions at the K-T boundary.
-Today, it is the extremeness of climate that is expected to cause the greatest societal damage.
-While extremes in the physical environment are present and widely understood, it remains unclear the extent to which ecological systems buffer or suffer from black swans.
-
-\item What is the existing evidence for heavy tails in ecological time series?
-
-\item There are two main possible reasons for why we may find little evidence of ecological black swans. (1) don't exist, (2) a paucity of appropriate methods or length/quality of time series.
-
-\item Here we develop \ldots our research questions and main conclusions.
-
-\end{enumerate}
-
-%Increasing realization about the importance of extreme events in the environment (REFs)
-%and of the importance of ecological surprises and black swans (REFs)
-
-\bigskip
-Papers to work in:
-
-\citep{inchausti2002,halley2002,inchausti2001}
-
-\citep{jentsch2007}
-
-\citep{ward2007}
-
-\citep{garcia-carreras2011}
-\citep{sornette2009}
-
-\citep{nunez2012}
-
-\citep{thompson2013}
-\citep{beaugrand2012}
-\citep{pine-iii2009}
-
-\citep{doak2008}
-
-\citep{smale2013}
-
-\citep{easterling2000}
-\citep{scheffer2003}
-\citep{katz2005}
-
-\citep{taleb2007}
-
-\citep{vasseur2014}
-
-\citep{vert-pre2013}
-\citep{lindenmayer2010}
-
-\citep{valpine2002}
-\citep{gregory2010}
-\citet{garcia-carreras2011}
-\citet{brook2006}
-\citep{herrandoprez2014}
-
-\citep{sibly2005, ziebarth2010}
-
-\ldots
-
-\section{Methods}
-
-\begin{enumerate}
-  \item Short overview paragraph
-  \item The data: global population dynamics database; briefly introduce the GPDD and how we filtered it. Expand in the supplement. Mention the breakdown of taxonomy and interpolation (Table~S1).
-  \item The heavy-tailed Gompertz population model. Mention the simulation testing but don't go into details here.
-  \item Mention the alternative population models we fit.
-  \item Modelling the possible correlates (mixed effects models with binomial or beta distributions and logit links; nested random effects for taxonomic class and taxonomic order).
-Life-history data from \citet{brook2006a}.
-  \end{enumerate}
-
-The Student-t Gompertz model is:
-\begin{align*}
-x_t &= \ln N_t\\
-x_t &= \lambda + b x_{t-1} + \epsilon_t\\
-\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma),
-\end{align*}
-
-\noindent where $x_t$ is the $\ln$ abundance at time $t$.
-The model is density independent if $b = 1$, maximally density dependent if $b = 0$, and inversely density dependence if $b < 0$.
-The parameter $\lambda$ represents the expected population growth rate at $x_t = 0$.
-The process noise, $\epsilon_t$, is modelled as a Student-t distribution that is centered at $0$, has a scale parameter of $\sigma_\mathrm{proc}$, and has a degrees of freedom parameter of $\nu$.
-If $\nu$ is small ($\lesssim 10$) the distribution has much heavier tails than a normal distribution.
-For example, at $\nu = 2$, the probability of drawing a value less than -5 is 1.8\%, whereas the probability of drawing such a value from a normal distribution is nearly zero ($2.9\cdot10^{-5}$\%).
-As $\nu$ approaches infinity the distribution approaches the normal distribution (Fig~\ref{fig:didactic}).
-By estimating the value of $\nu$ we can quantity how heavy-tailed the process noise deviations are.
-
-We fit alternative models that allowed for autocorrelation of the residuals, allowed for observation error, and allowed the functional form of the population dynamics to be represented as Ricker-logistic (Supporting Material).
-
-%Mention that we simulation tested the model.
-
-%Mention the priors (Fig.~\ref{fig:priors}) and expand on their justification in the supplement.
-
-%$\nu$ prior: \citep{fernandez1998}
-%$\sigma$ prior: \citep{gelman2006c}
-%consider alternative inverse prior on $\nu$ as in \citep{gelman2014}
-
-%justify why we only present these as hypotheses but do not formally model them
-
-\bigskip
-Supporting material to be noted:
-
-\begin{enumerate}
-
-  \item Table showing the taxonomic breakdown, number of interpolated points, etc (Table~S1).
-
-  \item Plot showing time series we assumed were recorded as log transformed  (Fig.~\ref{fig:log10-assumed}).
-
-  \item Plot showing all the time series we used with interpolated and zero imputed values (Fig.~\ref{fig:all-ts}).
-
-  \item Plot showing the priors (Fig.~\ref{fig:priors}).
-\end{enumerate}
-
-\section{Results}
-
-Main figures and tables:
-
-\begin{enumerate}
-  \item Fig.~\ref{fig:didactic} illustrates the method: it shows the t-distribution tails, example simulated time series, and model fits to those time series.
-
-  \item Fig.~\ref{fig:nu-coefs} shows the posterior distributions of the estimated $\nu$ values. These are split by taxonomic class and order.
-
-  \item Fig.~\ref{fig:correlates} shows possible biological and time-series-property correlates of heavy-tailed behaviour.
-
-  \item Table.~\ref{tab:sparks} selects populations that were categorized as heavy-tailed and digs into the causes. These are a non-random sample that I was able to verify in the literature.
-
-\end{enumerate}
-
-%- quantify how frequent and strong heavy tails were (Fig.~\ref{fig:nu-coefs})
-
-Supporting material to be noted quickly:
-
-\begin{enumerate}
-
-  \item Simulation testing: the ability to recover $\nu$ when randomly sampling from various distributions (Fig.~\ref{fig:sim-nu}); heavy-tailed Gompertz model performance and confidence interval coverage given the process noise has deviations that are known to have effective $\nu$ equal to true $\nu$ (Fig.~\ref{fig:sim-gompertz}); boxplots of the same output (Fig.~\ref{fig:sim-gompertz-boxplots}); probability that $\nu < 10$ for the Gompertz simulation testing (Fig.~\ref{fig:sim-prob}).
-
-  \item Time series of all heavy tailed populations (Fig.~\ref{fig:heavy-ts}).
-
-  \item The effect of alternative population dynamics models on $\nu$ estimates (Fig.~\ref{fig:alt}).
-
-  \item Coefficients from modelling covariates of the probability of heavy tails (Fig.~\ref{fig:correlate-coefs}).
-
-  \item Example Stan code for heavy-tailed Gompertz model with AR1 residuals and observation error (a model including everything).
-
-  \item GPDD IDs used.
-
-\end{enumerate}
-
-Modelling result: an increase of 1 time step of data (given that you are ballpark around the mean --- 30 time steps --- to start with) results in an approximately 1\% increase in the expected probability of observing heavy tails.
-(Using Gelman's `divide by 4' rule for interpreting logistic regression coefficients.)
-
-\section{Discussion}
-
-\begin{enumerate}
-  \item Summary of our findings
-  \item How do our results mesh with previous related analyses?
-  \item Why might we expect to see heavy tails in ecological time-series? (mixture of normals or extreme drivers)
-  \item Are the observed frequencies by taxonomic class real or an observational phenomena?
-  \item Ways forward
-  \item Policy implications
-\end{enumerate}
-
-\section{Acknowledgements}
-
-Funding: SCA: SFU Graduate Fellowship, NKD: NSERC SFU \ldots, ABC: \ldots, TAB: \ldots
-
-Earth to Ocean's research group for helpful discussions
-
-Global Population Dynamics Database
-
-Silhouettes: \texttt{phylopic.org}: rabbit by Sarah Werning, grey heron by Ardea cinerea, hoverfly by Gareth Monger. All under Creative Commons Attribution 3.0 Unported license.
-
-Compute Canada's WestGrid high-performance computing resources
-
-\bibliography{/Users/seananderson/Dropbox/tex/jshort,/Users/seananderson/Dropbox/tex/ref3}
+\input{sparks.tex}
 
 \clearpage
-
-\section{Tables}
-
-\LTcapwidth=\textwidth
-\bibpunct{}{}{;}{a}{}{;}
-
-\begin{small}
-\begin{longtable}{>{\RaggedRight}m{2.0cm}>{\RaggedRight}p{3.0cm}>{\RaggedRight}p{7.0cm}>{\RaggedRight}p{2.0cm}}
-
-\caption{Example population dynamic black swans from the Global Population Dynamics Database and descriptions of their causes.
-Red and blue dots indicate downward and upward events that have a probability of occurring of 0.001\% or less if the population dynamics were explained by a Gompertz model with normally distributed process noise.
-These populations are a non-random sample that we were able to verify in the primary literature.}\\
-
-\toprule
-Time series (log scale) & Population & Black swan description & Reference \\
-\midrule
-
-\includegraphics[width=2cm]{sparks/6528} &
-Shag,
-\textit{Phalacrocorax aristotelis},
-UK &
-Shortage of nest sites reduced productivity; red-tide event in 1968 caused extreme mortality; no longer a nest shortage; population rapidly increased &
-\citep{potts1980}\\
-
-\includegraphics[width=2cm]{sparks/10007} &
-Water vole,
-\textit{Arvicola terrestris},
-UK &
-Short-term population cycles from predator interactions combined with longterm environmental cycle caused sharp downswing  &
-\citep{saucy1994}\\
-
-\includegraphics[width=2cm]{sparks/7115} &
-Fur seal,
-\textit{Arctocephalus pusillus},
-South Africa &
-Strong decreases in harvesting, loss of predators, and diamond mining regulations reducing human traffic caused sharp upswings  &
-\citep{shaughnessy1982}\\
-
-\includegraphics[width=2cm]{sparks/10113} &
-Willow grouse,
-\textit{Lagopus lagopus},
-UK &
-Parasite and predation effects interacted to cause low years  &
-\citep{dobson1995}\\
-
-\includegraphics[width=2cm]{sparks/10162} &
-Red grouse,
-\textit{Lagopus lagopus scoticus},
-UK &
-Good environmental conditions produced high numbers and vulnerable populations; bad conditions and overcrowding combined to create crashes  &
-\citep{mackenzie1952}\\
-
-\includegraphics[width=2cm]{sparks/1235} &
-Wren,
-\textit{Troglodytes troglodytes},
-UK &
-Severe winters where food was submerged in snow caused population crash &
-\citep{newton1998} \\
-
-\includegraphics[width=2cm]{sparks/20579} &
-Grey heron,
-\textit{Ardea cinerea},
-UK &
-Severe winters in 1929, 1940--1942, and 1962--1963; 1963 event so severe that recovery took three times as long as expected &
-\citep{stafford1971} \\
-
-%\includegraphics[width=2cm]{sparks/20580} &
-%Chamois, \textit{Rupicapra rupicapra}, Switzerland &
-% &
-%\citep{brook2006a}\\
-
-%\includegraphics[width=2cm]{sparks/5019} &
-%Barbary macaque,  &
-% &
-%REF\\
-
-%\includegraphics[width=2cm]{sparks/9675} &
-%Carrot fly (\textit{Psila rosae}, Finland &
-% &
-%\citep{markkula1965}\\
-
-%\includegraphics[width=2cm]{sparks/10139} &
-%Grey heron (\textit{Ardea cinerea}), UK &
-% &
-%\citep{stafford1971}\\
-
-\bottomrule
-\label{tab:sparks}
-\end{longtable}
-\end{small}
-
-% reset citation style:
-\bibpunct{(}{)}{;}{a}{}{;}
-
-%\clearpage
-
 \section{Figures}
 
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=\textwidth]{t-nu-eg.pdf}
-\caption{An illustration of fitting population dynamics models that allow for heavy tails, represented by the Student-t degrees of freedom parameter $\nu$.
-  (a, b) The probability density for t distributions with a scale parameter of 1 and different values of $\nu$.
-  Small values of $\nu$ create heavy tails.
-  As $\nu$ approaches infinity the distribution approaches the normal distribution.
-  For example, at $\nu = 2$, the probability of drawing a value less than -5 is 1.8\%, whereas the probability of drawing such a value from a normal distribution is nearly zero ($2.9\cdot10^{-5}$\%).
-  (c--e) Simulated population dynamics from a Gompertz model with process noise drawn from t distributions with different values of $\nu$.
-  Coloured dots in panels c and d represent jumpts with less than a 1 in 1000 chance of occurring in a normal distribution.
-  (f--h) Estimates of $\nu$ from models fit to the times series in panels c--e.
-  Shown are the posterior samples (histograms), median and interquartile range of the posterior (IQR) (dots and line segments), and the exponential prior on $\nu$ (dashed lines).
-  Colour shading behind panels f--h illustrates the region of heavy tails.}
-\label{fig:didactic}
-\end{center}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=0.42\textwidth]{nu-coefs-2.pdf}
-\caption{
-Estimates of how heavy-tailed population dynamics are for \nuCoefPopN\ populations of birds, mammals, insects, and fishes.
-Small values of $\nu$ ($\lesssim 10$) suggest heavy-tailed dynamics; larger values of $\nu$ suggest approximately normal-tailed dynamics.
-Vertical points and line segments represent posterior medians and 50\% / 90\% credible intervals for individual populations.
-Inset plots show probability that $\nu < 10$ (probability of heavy tails) for populations arranged by taxonomic order.
-Taxonomic orders are sorted by decreasing mean p($\nu < 10$).
-Taxonomic orders with three or fewer populations in panel a are omitted for space.}\label{fig:nu-coefs}
-\end{center}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=0.9\textwidth]{correlates.pdf}
-\caption{
-Potential correlates of heavy-tailed behaviour (indicated by low $\nu$ values).
-Shown are (a--c) parameters from the Gompertz heavy-tailed population model, (b) number of time steps, (c) body length, and (d) lifespan.
-For the Gompertz parameters, $\sigma$ refers to the scale parameter of the Student-t process-noise distribution, $\lambda$ refers to the population growth rate at an abundance of one, $b$ refers to the density dependence parameter ($1$ is maximally density independent, $0$ is maximally density dependent, and $<0$ is inversely density dependent).
-For parameter values, points and lines represent medians and 50\% credible intervals of the posterior.
-Circles representing four populations are filled with white and represent a shark, crustaceans, and one gastropod.
-}
-\label{fig:correlates}
-\end{center}
-\end{figure}
-
-\clearpage
+\input{figs.tex}
 
@@
 
-\section{Data selection}
+\input{supp}
 
-To TODO \ldots
+\input{supp-figs}
 
-\begin{enumerate}
-
-\item To remove populations with unreliable population indices that could be strongly confounded with economics and sampling effort, we removed all populations with a sampling protocol listed as \texttt{"harvest"} as well populations with the words \texttt{"harvest"} or \texttt{"fur"} in the cited reference title.
-\item We removed all populations with uneven sampling intervals. I.e\ we removed populations that didn't have a constant difference between the ``decimal year begin'' and ``decimal year end'' columns.
-\item We removed all populations rated as $< 2$ in the GPDD quality assessment (on a scale of $1$ to $5$, with $1$ being the lowest quality data) \citep[following][]{sibly2005, ziebarth2010}
-\item Populations with negative abundance values were assumed to be log values and transformed by taking $10$ to the power of the recorded abundance.
-In many cases this was noted for the population, but not in all cases.
-We inspected each of these  \totalAssumedLog\ time series to make sure our assumption made sense (Fig.~\ref{fig:log10-assumed}).
-\item We filled in all missing time steps with \texttt{NA} values and imputed single missing values with the geometric mean of the previous and following values.
-We chose a geometric mean to be linear on the log scale that the Gompertz and Ricker-logistic models were fit on.
-\item We filled in single recorded values of zero with the lowest non-zero value in the time series \citep[following][]{brook2006a}.
-This assumes that single values of zero result from abundance being low enough that censusing missed present individuals.
-We turned multiple zero values in a row into \texttt{NA} values.
-This implies that multiple zero values were either censusing errors or caused by emigration.
-Regardless, our population  models were fit on a multiplicative (log) scale and so could not account for zero abundance.
-\item We removed all populations with four or more identical values in a row since these suggest either recording error or extrapolation between two observations.
-\item We removed all populations without at least four unique values \citep[following][]{brook2006a}.
-\item We then wrote an algorithm to find the longest unbroken window of abundance (no \texttt{NA}s) with at least 20 time steps in each population time series.
-If there were any populations with multiple windows of identical length, we took the most recent window.
-This is a longer window than used in some previous analyses \citep[e.g.][]{brook2006a}, but since our model attempts to capture the shape of the distribution tails, our model requires more data.
-\item Finally, we removed GPDD Main ID \texttt{20531}, which we noticed was a duplicate of \texttt{10139} (a heron population).
-\end{enumerate}
-
-We provide a supplemental figure of all the time series included in our analysis and indicate which values were interpolated (\percImputedPops\% of populations had at least one point interpolated and only \percImputedPoints\% of the total observations were interpolated) (Fig.\ SX).
-Table~S1 shows the final taxonomic breakdown and the number of populations with interpolated values.
-
-\section{Details on the heavy-tailed Gompertz probability model}
-
-For the Gompertz, the mostly weakly-informative priors I'm using are:
-\begin{align*}
-b &\sim \mathrm{Uniform}(-1, 2)\\
-\lambda &\sim \mathrm{Normal}(0, 10^2)\\
-\nu &\sim \mathrm{Exponential}(0.01)\\
-\sigma &\sim \mathrm{Half\mhyphen Cauchy} (5).
-\end{align*}
-
-See Figure~\ref{fig:priors}.
-
-Priors justification: (TODO this is old)
-
-\begin{itemize}
-\item $b$ is bounded just past stationary so we can detect if they are non-stationary will keeping the sampler from wandering off too far
-
-\item $\lambda$ prior (variance = 100) is basically uninformative within the range of expected values for population growth... it allows even a X probability at a value of X
-
-\item $\nu$ is based on \citet{fernandez1998}; they chose a more informative 0.1 value, we chose a less informative 0.01 and justify it based on performance in supplemental figure X (gives X probability of value less than 10) but constrains the sampling somewhat since above 40 or 50 the shape of the t distribution is almost identical and data are not usually informative at these data quantities...
-sampler would head off to infinity otherwise
-
-\item $\phi$ given a standard deviation of 1 given prior analyses which suggest autocorrelation in these datasets is minimal; prior is weak enough to allow high or low values, but given little information will stay near our expectation; similar approach in \citep{thorson2014a}
-
-\item $\sigma_\mathrm{proc}$ can be justified based on \citet{gelman2006c} and the expected range of this variable in nature from previous studies \citep[e.g.][]{connors2014}
-\end{itemize}
-
-We fit our models with Stan 2.4.0 \citep{stan-manual2014}, and R 3.1.1 \citep{r2014}.
-I'm starting with 4 chains and 2000 iterations with the first 1000 as warmup (i.e.\ 4000 total samples).
-If rhat is greater than 1.05 for any parameter or the minimum effective sample size is less than 200 for any parameter then I double both the total iterations and warmup and run again.
-This continued up to 8000 iterations (16000 total samples) \ldots
-
-\section{Simulation testing the model}
-
-Throughout all of this --- show that the model, if anything, under-predicts heavy tails but with sufficient data is unbiased.
-
-2 parts: how many samples from the true population t distribution do you need to detect low nu?
-And, given that you have a set of deviations in which nu is detectable (effective nu is within 0.5 CV of true nu), can the more complex Gompertz still capture this?
-
-First part: We drew from t distributions with different nu values and mean of 0, scale of 1.
-We started with 1600 samples and then fitted again at the first 800, 400, 200,  100, 50, 25.
-Each time we recorded the nu posterior.
-
-Second part: To generate a series of process deviations with an effective nu approximately equal to the true nu, we generated process deviation sets repeatedly and estimated the mean, scale, and nu values each time.
-We recorded when the estimated nu was within 0.5 CVs of the true nu and used this set of random seed values in our Gompertz simulation.
-We then fit AR1 Gompertz models to the simulated datasets with all parameters (except nu) set near the median values estimated in the GPDD.
-
-\section{Alternative population models}
-
-\subsection{Ricker-logistic}
-
-We also fitted a Ricker-logistic model:
-\begin{align*}
-x_t &= x_{t-1} + r_{\mathrm{max}}\left(1 - \frac{N_{t-1}}{K}\right) + b x_{t-1} + \epsilon_t\\
-\epsilon_t &\sim \mathrm{Student\mhyphen t}_\nu(0, \sigma),
-\end{align*}
-\noindent
-where  $r_\mathrm{max}$ represents the maximum population growth rate that is obtained when $N$ (abundance) $= 0$.
-The parameter $K$ represents the carrying capacity and, as before, $x_t$ represents the $\ln$ transformed abundance at time $t$.
-The Ricker-logistic model assumes a linear decrease in population growth rate ($x_t - x_{t-1}$) with increases in abundance ($N_t$).
-In contrast, the Gompertz model assumes a linear decrease in population growth rate with increases in $\ln$ abundance ($x_t$) (REF).
-
-To fit the Ricker-logistic models, we chose a prior on $K$ uniform between zero and twice the maximum observed abundance (\citet{clark2010} chose uniform between zero and maximum observed, which is less conservative).
-We set the prior on $r_\mathrm{max}$ as uniform between 0 and 20 as in \citet{clark2010}.
-We used the same priors on $\nu$ and $\sigma$ as in the Gompertz model.
-
-\subsection{Autocorrelated residuals}
-
-We considered a version of the Gompertz model in which an autoregressive parameter was fit to the process noise residuals:
-\begin{align*}
-x_t &= \lambda + b x_{t-1} + \epsilon_t\\
-\epsilon_t &\sim \mathrm{Student\mhyphen t}_\nu(\phi \epsilon_{t-1}, \sigma).
-\end{align*}
-\noindent
-In addition to the parameters in the original Gompertz model, an additional parameter, $\phi$, is estimated that represents the relationship between of subsequent process noise residuals.
-Based on the results of previous analyses with the GPDD \citep[e.g.][]{connors2014} and the chosen priors in previous analyses \citep[e.g.][]{thorson2014a} and to greatly speed up chain convergence when running our model across all populations, we placed a weakly informative prior on $\phi$ that assumed the greatest probability density near zero with the reduced possibility of $\phi$ being near $-1$ or $1$.
-Specifically, we chose $\phi \sim \mathrm{Truncated\mhyphen Normal}(0, 1, \mathrm{min.} = 1, \mathrm{max.} = 1)$.
-
-\subsection{Assumed observation error}
-
-We considered a version of the base Gompertz model that allowed for a specified level of observation error:
-\begin{align*}
-U_t &= \lambda + b U_{t-1} + \epsilon_t\\
-x_t &\sim \mathrm{Normal}(U_t, \sigma_\mathrm{obs})\\
-\epsilon_t &\sim \mathrm{Student\mhyphen t}_\nu(0, \sigma_\mathrm{proc}),
-\end{align*}
-
-\noindent
-where $U$ represents the unobserved state vector, and $\sigma_\mathrm{obs}$ represents the standard deviation of observation error (on a log scale), which was set at $0.2$. TODO justify $0.2$.
-
-%\end{spacing}
-
-\renewcommand{\thetable}{S\arabic{table}}
-\setcounter{table}{0}
-
-\include{stat-table}
-
-\renewcommand{\thefigure}{S\arabic{figure}}
-\setcounter{figure}{0}
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=\textwidth]{log10-assumed.pdf} \caption{Populations in which the abundance was specified as or assumed to be recorded as log10 values.
-The numbers at the front of each panel label are the GPDD IDs.}
-    \label{fig:log10-assumed}
-\end{center}
-\end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=\textwidth]{all-clean-ts-3.pdf} \caption{All filtered times series used in our analysis.
-The abundances are shown on a log10 vertical axis.
-Colours indicate taxonomic classes.
-Open black circles indicate abundance values that were interpolated.
-Closed black circles indicate single abundance values that were recorded as zero but were set to the next lowest observed abundance.}
-    \label{fig:all-ts}
-\end{center}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=\textwidth]{priors-gomp-base.pdf} \caption{Probability
-  density of the Bayesian priors for the base Gompertz model. From left to right: (1) per capita growth
-  rate at $\ln$(abundance) = $0$: $\lambda \sim \mathrm{Normal}(0, 10^2)$; (2) scale
-  parameter of t-distribution process noise: $\sigma \sim \mathrm{Half\mhyphen
-    Cauchy} (2.5)$; (3) t-distribution degrees of freedom parameter: $\nu \sim
-  \mathrm{Exponential}(0.01)$. (4) AR1 correlation coefficient of residuals: $\phi \sim \mathrm{Normal}(0, 1)[-1, 1]$. Not shown is $b$, the density dependence
-  parameter: $b \sim \mathrm{Uniform}(-1, 2)$.}
-  \label{fig:priors}
-\end{center}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=0.8\textwidth]{t-dist-sampling-sim-prior-exp0point01.pdf}
-\includegraphics[width=0.8\textwidth]{t-dist-sampling-sim-sigma-prior-exp0point01.pdf}
-\caption{
-  Testing the ability to estimate $\nu$ (top panels) and the scale parameter of the process error (bottom panels) for a given number of samples (columns) drawn from a distribution with a given true $\nu$ value (rows).
-The red lines indicate the true population value.
-When a small number of samples are drawn there may not be samples sufficiently far into the tails to recapture the true $\nu$ value; however, heavy tails are still distinguished from normal tails in most cases, even with only 25 or 50 samples.
-TODO switch to nu = black swan, 3, 5, and normal.
-}
-\label{fig:sim-nu}
-\end{center}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=\textwidth]{sim-gompertz.pdf}
-\caption{Simulation testing the Gompertz estimation model when the process error draws are chosen so that $\nu$ can be estimated close to the true value outside the full population model (``effective $\nu$'' within a CV of 0.2 of specified $\nu$).
-  The simulation was run across population $\nu$ values (columns) and different scenarios (colours): (1) 100 time steps and no observation error, (2) 50 time steps and no observation error, (3) 50 time steps and observation error drawn from $\mathrm{Normal} (0, 0.2^2)$ but ignored, and (4) 50 time steps with observation error in which the quantity of observation error was assumed known.
-  Dashed horizontal lines show the true population values;
-  these true values were chosen to represent approximately the median values as estimated from the GPDD.
-  Individual dots and lines represent a stochastic draws from the true population distribution and a model fitting (median and 80\% credible intervals).
-  %The panels show true values of $\nu$ =  3, 5, normal (very heavy, heavy, and not heavy tailed).
-  The panels from top to bottom show $1/\nu$, process noise scale parameter $\sigma$, growth rate parameter $\lambda$, and the density dependence parameter $b$.}
-\label{fig:sim-gompertz}
-\end{center}
-\end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=\textwidth]{sim-gompertz-boxplots.pdf}
-\caption{The distribution of median estimates from the same simulation as described in Fig.~\ref{fig:sim-gompertz}.
-The boxes show the interquartile range.
-The whiskers extend to $1.5$ times the interquartile range and outliers are shown as dots.}
-\label{fig:sim-gompertz-boxplots}
-\end{center}
-\end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=\textwidth]{check-sim-box.pdf}
-\caption{The probability that $\nu < 10$ (i.e.\ the approximate probability of heavy tails) for the same simulation scenarios as shown in Fig.~\ref{fig:sim-gompertz} and Fig.~\ref{fig:sim-gompertz-boxplots}.
-Within each scenario the dots represent stochastic draws from the true population distributions combined with model fits.}
-\label{fig:sim-prob}
-\end{center}
-\end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=0.9\textwidth]{ts-gpdd-heavy-eg-log10-base.pdf}
-\caption{Time series for populations with p($\nu < 10$) $>$ 0.5 using the base Gompertz population model.
-Panels are ordered by increasing p($\nu < 10$).
-Vertical axes are on a log10 scale.
-Colours indicate taxonomic classes.
-TODO add dots for upswings and downswings as in Table 1.
-The labels on each panel indicate p($\nu < 10$) $>$ 0.5, the common name for the species, and the GPDD ID number.}
-\label{fig:heavy-ts}
-\end{center}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=\textwidth]{gomp-comparison.pdf}
-\caption{Estimates of $\nu$ from alternative models plotted against the base Gompertz model estimates of $\nu$.
-Shown are medians of the posterior (dots) and 50\% credible intervals (segments).
-The diagonal line indicates a one-to-one relationship.
-Different colours indicate various taxonomic classes.
-The grey-shaded regions indicate regions of disagreement if $\nu = 10$ is taken as a threshold of heavy-tailed dynamics.
-The Gompertz observation error model assumes a fixed standard deviation of observation error of $0.2$ on the log scale.}
-\label{fig:alt}
-\end{center}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=0.7\textwidth]{admb-coefs}
-\includegraphics[width=0.8\textwidth]{length-vs-prob-tails}
-\caption{Potential correlates of heavy-tailed dynamics.
-  (Top panels) Dots, thick lines, and thin lines represent fixed-effect estimates and their 50\% and 95\% confidence intervals for the covariates listed on the left: $\log \sigma$, $b$, $\lambda$, $\log$(lifespan), $\log$(data set length).
-All covariates were standardized by subtracting their mean and dividing by their standard deviation.
-  The beta distribution model (left panel) uses p$(\nu < 10)$ as the response (probability of heavy tails).
-  The binomial distribution model (right panel) uses p$(\nu < 10) > 0.5$ vs.\ p$(\nu < 10) \le 0.5$ as the response (heavy vs.\ not heavy tails).
-  The models were fitted as mixed effects models with beta or binomial distributions, logit links, and multilevel intercepts for taxonomic class and order nested within class using the \texttt{glmmADMB} \texttt{R} package \citep{fournier2012,glmmadmb}.
-  (Bottom panels) Model predictions for the strongest covariate --- time series length --- with all other predictors set to zero (their mean value when standardized).
-Dots represent observations for individual populations.
-Lines and shaded regions represent estimates and 50\% and 95\% confidence intervals.
-}
-    \label{fig:correlate-coefs}
-\end{center}
-\end{figure}
-
-% \clearpage
-%
-% \begin{figure}[htbp]
-% \begin{center}
-% \includegraphics[width=0.8\textwidth]{length-vs-prob-tails}
-% \caption{Model predictions. TODO}
-%     \label{fig:length-predictions}
-% \end{center}
-% \end{figure}
-
-\clearpage
-
-\noindent
-Example Stan code for a heavy-tailed Gompertz model with AR1 correlated residuals and a specified level of observation error:
-
-\begin{spacing}{1.15}
-\begin{footnotesize}
-\begin{verbatim}
-data {
-  int<lower=3> N;              // number of observations
-  vector[N] y;                 // vector to hold ln abundance observations
-  real<lower=0> nu_rate;       // rate parameter for nu exponential prior
-}
-parameters {
-  real lambda;                 // Gompertz growth rate parameter
-  real<lower=-1, upper=2> b;   // Gompertz density dependence parameter
-  real<lower=0> sigma_proc;    // process noise scale parameter
-  real<lower=2> nu;            // t-distribution degrees of freedom
-  real<lower=-1, upper=1> phi; // AR1 parameter
-  vector[N] U;                 // unobserved states
-  real<lower=0> sigma_obs;     // specified observation error SD
-}
-transformed parameters {
-  vector[N] epsilon;           // error terms
-  epsilon[1] <- 0;
-  for (i in 2:N) {
-    epsilon[i] <- U[i] - (lambda + b * U[i - 1])
-                       - (phi * epsilon[i - 1]);
-  }
-}
-model {
-  // priors:
-  nu ~ exponential(nu_rate);
-  lambda ~ normal(0, 10);
-  sigma_proc ~ cauchy(0, 2.5);
-  phi ~ normal(0, 1);
-  // data model:
-  for (i in 2:N) {
-    U[i] ~ student_t(nu,
-                     lambda + b * U[i - 1]
-                     + phi * epsilon[i - 1],
-                     sigma_proc);
-  }
-  y ~ normal(U, sigma_obs);
-}
-\end{verbatim}
-\end{footnotesize}
-
-\clearpage
-
-\noindent
-The GPDD IDs used in our analysis:
-
-\begin{footnotesize}
-\noindent
-{\tt
-\input{mainids}
-}
-\end{footnotesize}
-\end{spacing}
 \end{document}

### 2014-10-02 figs.tex
Switch to markdown ms; add to ms

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{t-nu-eg.pdf}
\caption{An illustration of fitting population dynamics models that allow for heavy tails, represented by the Student-t degrees of freedom parameter $\nu$.
  (a, b) The probability density for t distributions with a scale parameter of 1 and different values of $\nu$.
  Small values of $\nu$ create heavy tails.
  As $\nu$ approaches infinity the distribution approaches the normal distribution.
  For example, at $\nu = 2$, the probability of drawing a value less than -5 is 1.8\%, whereas the probability of drawing such a value from a normal distribution is nearly zero ($2.9\cdot10^{-5}$\%).
  (c--e) Simulated population dynamics from a Gompertz model with process noise drawn from t distributions with different values of $\nu$.
  Coloured dots in panels c and d represent jumpts with less than a 1 in 1000 chance of occurring in a normal distribution.
  (f--h) Estimates of $\nu$ from models fit to the times series in panels c--e.
  Shown are the posterior samples (histograms), median and interquartile range of the posterior (IQR) (dots and line segments), and the exponential prior on $\nu$ (dashed lines).
  Colour shading behind panels f--h illustrates the region of heavy tails.}
\label{fig:didactic}
\end{center}
\end{figure}

\clearpage

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.42\textwidth]{nu-coefs-2.pdf}
\caption{
Estimates of how heavy-tailed population dynamics are for \nuCoefPopN\ populations of birds, mammals, insects, and fishes.
Small values of $\nu$ ($\lesssim 10$) suggest heavy-tailed dynamics; larger values of $\nu$ suggest approximately normal-tailed dynamics.
Vertical points and line segments represent posterior medians and 50\% / 90\% credible intervals for individual populations.
Inset plots show probability that $\nu < 10$ (probability of heavy tails) for populations arranged by taxonomic order.
Taxonomic orders are sorted by decreasing mean p($\nu < 10$).
Taxonomic orders with three or fewer populations in panel a are omitted for space.
Red and orange colours highlight populations with a high probability of heavy-tailed dynamics.
}
\label{fig:nu-coefs}
\end{center}
\end{figure}

\clearpage

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.9\textwidth]{correlates.pdf}
\caption{
Potential correlates of heavy-tailed behaviour (indicated by low $\nu$ values). Shown are (a--c) parameters from the Gompertz heavy-tailed population model, (b) number of time steps, (c) body length, and (d) lifespan.
For the Gompertz parameters, $\sigma$ refers to the scale parameter of the Student-t process-noise distribution, $\lambda$ refers to the population growth rate at an abundance of one, $b$ refers to the density dependence parameter ($1$ is maximally density independent, $0$ is maximally density dependent, and $<0$ is inversely density dependent).
For parameter values, points and lines represent medians and 50\% credible intervals of the posterior.
Circles representing four populations are filled with white and represent a shark, crustaceans, and one gastropod.
}
\label{fig:correlates}
\end{center}
\end{figure}

\clearpage

### 2014-10-02 sparks.tex
Switch to markdown ms; add to ms

\LTcapwidth=\textwidth
\bibpunct{}{}{;}{a}{}{;}

\begin{small}
\begin{longtable}{>{\RaggedRight}m{2.0cm}>{\RaggedRight}p{3.0cm}>{\RaggedRight}p{7.0cm}>{\RaggedRight}p{2.0cm}}

\caption{Example population dynamic black swans from the Global Population Dynamics Database and descriptions of their causes. Red and blue dots indicate downward and upward events that have a probability of occurring of 0.001\% or less if the population dynamics were explained by a Gompertz model with normally distributed process noise. These populations are a non-random sample that we were able to verify in the primary literature.}\\

\toprule
Time series (log scale) & Population & Black swan description & Reference \\
\midrule

\includegraphics[width=2cm]{sparks/6528} &
Shag,
\textit{Phalacrocorax aristotelis},
UK &
Shortage of nest sites reduced productivity; red-tide event in 1968 caused extreme mortality; no longer a nest shortage; population rapidly increased &
\citep{potts1980}\\

\includegraphics[width=2cm]{sparks/10007} &
Water vole,
\textit{Arvicola terrestris},
UK &
Short-term population cycles from predator interactions combined with longterm environmental cycle caused sharp downswing  &
\citep{saucy1994}\\

\includegraphics[width=2cm]{sparks/7115} &
Fur seal,
\textit{Arctocephalus pusillus},
South Africa &
Strong decreases in harvesting, loss of predators, and diamond mining regulations reducing human traffic caused sharp upswings  &
\citep{shaughnessy1982}\\

\includegraphics[width=2cm]{sparks/10113} &
Willow grouse,
\textit{Lagopus lagopus},
UK &
Parasite and predation effects interacted to cause low years  &
\citep{dobson1995}\\

\includegraphics[width=2cm]{sparks/10162} &
Red grouse,
\textit{Lagopus lagopus scoticus},
UK &
Good environmental conditions produced high numbers and vulnerable populations; bad conditions and overcrowding combined to create crashes  &
\citep{mackenzie1952}\\

\includegraphics[width=2cm]{sparks/1235} &
Wren,
\textit{Troglodytes troglodytes},
UK &
Severe winters where food was submerged in snow caused population crash &
\citep{newton1998} \\

\includegraphics[width=2cm]{sparks/20579} &
Grey heron,
\textit{Ardea cinerea},
UK &
Severe winters in 1929, 1940--1942, and 1962--1963; 1963 event so severe that recovery took three times as long as expected &
\citep{stafford1971} \\

%\includegraphics[width=2cm]{sparks/20580} &
%Chamois, \textit{Rupicapra rupicapra}, Switzerland &
% &
%\citep{brook2006a}\\

%\includegraphics[width=2cm]{sparks/5019} &
%Barbary macaque,  &
% &
%REF\\

%\includegraphics[width=2cm]{sparks/9675} &
%Carrot fly (\textit{Psila rosae}, Finland &
% &
%\citep{markkula1965}\\

%\includegraphics[width=2cm]{sparks/10139} &
%Grey heron (\textit{Ardea cinerea}), UK &
% &
%\citep{stafford1971}\\

\bottomrule
\label{tab:sparks}
\end{longtable}
\end{small}

% reset citation style:
\bibpunct{(}{)}{;}{a}{}{;}

### 2014-10-02 supp-figs.tex
Switch to markdown ms; add to ms

\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{figure}{0}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{log10-assumed.pdf} \caption{Populations in which the abundance was specified as or assumed to be recorded as log10 values. The numbers at the front of each panel label are the GPDD IDs.}
    \label{fig:log10-assumed}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{all-clean-ts-3.pdf} \caption{All filtered times series used in our analysis. The abundances are shown on a log10 vertical axis. Colours indicate taxonomic classes. Open black circles indicate abundance values that were interpolated. Closed black circles indicate single abundance values that were recorded as zero but were set to the next lowest observed abundance.}
    \label{fig:all-ts}
\end{center}
\end{figure}

\clearpage

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{priors-gomp-base.pdf} \caption{Probability
  density of the Bayesian priors for the base Gompertz model. From left to right: (1) per capita growth
  rate at $\ln$(abundance) = $0$: $\lambda \sim \mathrm{Normal}(0, 10^2)$; (2) scale
  parameter of t-distribution process noise: $\sigma \sim \mathrm{Half\mhyphen
    Cauchy} (2.5)$; (3) t-distribution degrees of freedom parameter: $\nu \sim
  \mathrm{Exponential}(0.01)$. (4) AR1 correlation coefficient of residuals: $\phi \sim \mathrm{Truncated \mhyphen Normal}(0, 1, \mathrm{min.} = -1, \mathrm{max.} = 1)$. Not shown is $b$, the density dependence
  parameter: $b \sim \mathrm{Uniform}(-1, 2)$.}
  \label{fig:priors}
\end{center}
\end{figure}

\clearpage

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.8\textwidth]{t-dist-sampling-sim-prior-exp0point01.pdf}
\includegraphics[width=0.8\textwidth]{t-dist-sampling-sim-sigma-prior-exp0point01.pdf}
\caption{
  Testing the ability to estimate $\nu$ (top panels) and the scale parameter of the process error (bottom panels) for a given number of samples (columns) drawn from a distribution with a given true $\nu$ value (rows). The red lines indicate the true population value. When a small number of samples are drawn there may not be samples sufficiently far into the tails to recapture the true $\nu$ value; however, heavy tails are still distinguished from normal tails in most cases, even with only 25 or 50 samples. TODO switch to nu = black swan, 3, 5, and normal.
}
\label{fig:sim-nu}
\end{center}
\end{figure}

\clearpage

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{sim-gompertz.pdf}
\caption{Simulation testing the Gompertz estimation model when the process error draws are chosen so that $\nu$ can be estimated close to the true value outside the full population model (``effective $\nu$'' within a CV of 0.2 of specified $\nu$).
  The simulation was run across population $\nu$ values (columns) and different scenarios (colours): (1) 100 time steps and no observation error, (2) 50 time steps and no observation error, (3) 50 time steps and observation error drawn from $\mathrm{Normal} (0, 0.2^2)$ but ignored, and (4) 50 time steps with observation error in which the quantity of observation error was assumed known.
  Dashed horizontal lines show the true population values;
  these true values were chosen to represent approximately the median values as estimated from the GPDD.
  Individual dots and lines represent a stochastic draws from the true population distribution and a model fitting (median and 80\% credible intervals).
  %The panels show true values of $\nu$ =  3, 5, normal (very heavy, heavy, and not heavy tailed).
  The panels from top to bottom show $1/\nu$, process noise scale parameter $\sigma$, growth rate parameter $\lambda$, and the density dependence parameter $b$.}
\label{fig:sim-gompertz}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{sim-gompertz-boxplots.pdf}
\caption{The distribution of median estimates from the same simulation as described in Fig.~\ref{fig:sim-gompertz}. The boxes show the interquartile range. The whiskers extend to $1.5$ times the interquartile range and outliers are shown as dots.}
\label{fig:sim-gompertz-boxplots}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{check-sim-box.pdf}
\caption{The probability that $\nu < 10$ (i.e.\ the approximate probability of heavy tails) for the same simulation scenarios as shown in Fig.~\ref{fig:sim-gompertz} and Fig.~\ref{fig:sim-gompertz-boxplots}. Within each scenario the dots represent stochastic draws from the true population distributions combined with model fits.}
\label{fig:sim-prob}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.9\textwidth]{ts-gpdd-heavy-eg-log10-base.pdf}
\caption{Time series for populations with p($\nu < 10$) $>$ 0.5 using the base Gompertz population model. Panels are ordered by increasing p($\nu < 10$). Vertical axes are on a log10 scale. Colours indicate taxonomic classes. TODO add dots for upswings and downswings as in Table 1. The labels on each panel indicate p($\nu < 10$) $>$ 0.5, the common name for the species, and the GPDD ID number.}
\label{fig:heavy-ts}
\end{center}
\end{figure}

\clearpage

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{gomp-comparison.pdf}
\caption{Estimates of $\nu$ from alternative models plotted against the base Gompertz model estimates of $\nu$. Shown are medians of the posterior (dots) and 50\% credible intervals (segments). The diagonal line indicates a one-to-one relationship. Different colours indicate various taxonomic classes. The grey-shaded regions indicate regions of disagreement if $\nu = 10$ is taken as a threshold of heavy-tailed dynamics. The Gompertz observation error model assumes a fixed standard deviation of observation error of $0.2$ on the log scale.}
\label{fig:alt}
\end{center}
\end{figure}

\clearpage

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.7\textwidth]{admb-coefs}
\includegraphics[width=0.8\textwidth]{length-vs-prob-tails}
\caption{Potential correlates of heavy-tailed dynamics.
  (Top panels) Dots, thick lines, and thin lines represent fixed-effect estimates and their 50\% and 95\% confidence intervals for the covariates listed on the left: $\log \sigma$, $b$, $\lambda$, $\log$(lifespan), $\log$(data set length). All covariates were standardized by subtracting their mean and dividing by their standard deviation.
  The beta distribution model (left panel) uses p$(\nu < 10)$ as the response (probability of heavy tails).
  The binomial distribution model (right panel) uses p$(\nu < 10) > 0.5$ vs.\ p$(\nu < 10) \le 0.5$ as the response (heavy vs.\ not heavy tails).
  The models were fitted as mixed effects models with beta or binomial distributions, logit links, and multilevel intercepts for taxonomic class and order nested within class using the \texttt{glmmADMB} \texttt{R} package \citep{fournier2012,glmmadmb}.
  (Bottom panels) Model predictions for the strongest covariate --- time series length --- with all other predictors set to zero (their mean value when standardized). Dots represent observations for individual populations. Lines and shaded regions represent estimates and 50\% and 95\% confidence intervals.
}
    \label{fig:correlate-coefs}
\end{center}
\end{figure}

% \clearpage
%
% \begin{figure}[htbp]
% \begin{center}
% \includegraphics[width=0.8\textwidth]{length-vs-prob-tails}
% \caption{Model predictions. TODO}
%     \label{fig:length-predictions}
% \end{center}
% \end{figure}

\clearpage

\noindent
Example Stan code for a heavy-tailed Gompertz model with AR1 correlated residuals and a specified level of observation error:

\begin{spacing}{1.15}
\begin{footnotesize}
\begin{verbatim}
data {
  int<lower=3> N;              // number of observations
  vector[N] y;                 // vector to hold ln abundance observations
  real<lower=0> nu_rate;       // rate parameter for nu exponential prior
}
parameters {
  real lambda;                 // Gompertz growth rate parameter
  real<lower=-1, upper=2> b;   // Gompertz density dependence parameter
  real<lower=0> sigma_proc;    // process noise scale parameter
  real<lower=2> nu;            // t-distribution degrees of freedom
  real<lower=-1, upper=1> phi; // AR1 parameter
  vector[N] U;                 // unobserved states
  real<lower=0> sigma_obs;     // specified observation error SD
}
transformed parameters {
  vector[N] epsilon;           // error terms
  epsilon[1] <- 0;
  for (i in 2:N) {
    epsilon[i] <- U[i] - (lambda + b * U[i - 1])
                       - (phi * epsilon[i - 1]);
  }
}
model {
  // priors:
  nu ~ exponential(nu_rate);
  lambda ~ normal(0, 10);
  sigma_proc ~ cauchy(0, 2.5);
  phi ~ normal(0, 1);
  // data model:
  for (i in 2:N) {
    U[i] ~ student_t(nu,
                     lambda + b * U[i - 1]
                     + phi * epsilon[i - 1],
                     sigma_proc);
  }
  y ~ normal(U, sigma_obs);
}
\end{verbatim}
\end{footnotesize}

\clearpage

\noindent
The GPDD IDs used in our analysis:

\begin{footnotesize}
\noindent
{\tt
\input{mainids}
}
\end{footnotesize}
\end{spacing}

### 2014-10-06 figs.tex
Substantial additions to text, supp text, and figures

@@
 \begin{center}
-\includegraphics[width=0.42\textwidth]{nu-coefs-2.pdf}
+\includegraphics[width=0.44\textwidth]{nu-coefs-2.pdf}
 \caption{
@@
 \begin{center}
-\includegraphics[width=0.9\textwidth]{correlates.pdf}
+%\includegraphics[width=0.9\textwidth]{correlates.pdf}
+\includegraphics[width=0.9\textwidth]{correlates-p10.pdf}
 \caption{
-Potential correlates of heavy-tailed behaviour (indicated by low $\nu$ values).
+Potential correlates of heavy-tailed behaviour (indicated by a high probability that $\nu < 10$).
 Shown are (a--c) parameters from the Gompertz heavy-tailed population model, (b) number of time steps, (c) body length, and (d) lifespan.
 For the Gompertz parameters, $\sigma$ refers to the scale parameter of the Student-t process-noise distribution, $\lambda$ refers to the population growth rate at an abundance of one, $b$ refers to the density dependence parameter ($1$ is maximally density independent, $0$ is maximally density dependent, and $<0$ is inversely density dependent).
-For parameter values, points and lines represent medians and 50\% credible intervals of the posterior.
-Circles representing four populations are filled with white and represent a shark, crustaceans, and one gastropod.
+%For parameter values, points and lines represent medians and 50\% credible intervals of the posterior.
+Circles representing a limited number of sharks, crustaceans, and gastropods are filled in white.
+Median and 90\% credible interval posterior predictions of a beta regression multilevel model are shown in panels a and d where there was a high probability the slope coefficient was different from zero (Supporting Material).
 }

### 2014-10-06 supp-figs.tex
Substantial additions to text, supp text, and figures

@@
 
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=\textwidth]{log10-assumed.pdf} \caption{Populations in which the abundance was specified as or assumed to be recorded as log10 values.
+%\begin{figure}[htbp]
+%\begin{center}
+%\includegraphics[width=\textwidth]{log10-assumed.pdf} \caption{Populations in which the abundance was specified as or assumed to be recorded as log10 values.
 The numbers at the front of each panel label are the GPDD IDs.}
-    \label{fig:log10-assumed}
-\end{center}
-\end{figure}
+    %\label{fig:log10-assumed}
+%\end{center}
+%\end{figure}
 
@@
 \includegraphics[width=\textwidth]{priors-gomp-base.pdf} \caption{Probability
-  density of the Bayesian priors for the base Gompertz model. From left to right: (1) per capita growth
+  density of the Bayesian priors for the Gompertz models. From left to right: (1) per capita growth
   rate at $\ln$(abundance) = $0$: $\lambda \sim \mathrm{Normal}(0, 10^2)$; (2) scale
   parameter of t-distribution process noise: $\sigma \sim \mathrm{Half\mhyphen
-    Cauchy} (2.5)$; (3) t-distribution degrees of freedom parameter: $\nu \sim
-  \mathrm{Exponential}(0.01)$.
-(4) AR1 correlation coefficient of residuals: $\phi \sim \mathrm{Truncated \mhyphen Normal}(0, 1, \mathrm{min.} = -1, \mathrm{max.} = 1)$.
-Not shown is $b$, the density dependence
-  parameter: $b \sim \mathrm{Uniform}(-1, 2)$.}
+    Cauchy} (0, 2.5)$; (3) t-distribution degrees of freedom parameter: $\nu \sim
+  \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2)$; (4) AR1 correlation coefficient of residuals: $\phi \sim \mathrm{Truncated \mhyphen Normal}(0, 1, \mathrm{min.} = -1, \mathrm{max.} = 1)$.
+Not shown is $b$, the density dependence parameter: $b \sim \mathrm{Uniform}(-1, 2)$.}
   \label{fig:priors}
@@
 \begin{center}
-\includegraphics[width=0.7\textwidth]{admb-coefs}
-\includegraphics[width=0.8\textwidth]{length-vs-prob-tails}
-\caption{Potential correlates of heavy-tailed dynamics.
-  (Top panels) Dots, thick lines, and thin lines represent fixed-effect estimates and their 50\% and 95\% confidence intervals for the covariates listed on the left: $\log \sigma$, $b$, $\lambda$, $\log$(lifespan), $\log$(data set length).
-All covariates were standardized by subtracting their mean and dividing by their standard deviation.
-  The beta distribution model (left panel) uses p$(\nu < 10)$ as the response (probability of heavy tails).
-  The binomial distribution model (right panel) uses p$(\nu < 10) > 0.5$ vs.\ p$(\nu < 10) \le 0.5$ as the response (heavy vs.\ not heavy tails).
-  The models were fitted as mixed effects models with beta or binomial distributions, logit links, and multilevel intercepts for taxonomic class and order nested within class using the \texttt{glmmADMB} \texttt{R} package \citep{fournier2012,glmmadmb}.
-  (Bottom panels) Model predictions for the strongest covariate --- time series length --- with all other predictors set to zero (their mean value when standardized).
+%\includegraphics[width=0.7\textwidth]{admb-coefs}
+%\includegraphics[width=0.8\textwidth]{length-vs-prob-tails}
+\includegraphics[width=0.5\textwidth]{stan-beta-correlates}
+\caption{
+  %Potential correlates of heavy-tailed dynamics.
+  %(Top panels) Dots, thick lines, and thin lines represent fixed-effect estimates and their 50\% and 95\% confidence intervals for the covariates listed on the left: $\log \sigma$, $b$, $\lambda$, $\log$(lifespan), $\log$(data set length).
+  Main effect posterior densitities for potential correlates of heavy-tailed dynamics.
+  The response variable was Pr($\nu < 10$) (approximately the probability of heavy tails).
+  The beta regression model was fit on a logit scale with random intercepts for taxonomic class, taxonomic order, and species.
+  All covariates were standardized by subtracting their mean and dividing by twice their standard deviation.
+Therefore, the coefficient values represent the expected effect of a change in two standard deviations of the raw input variable on the logit-transformed response variable.
+  %The beta distribution model (left panel) uses p$(\nu < 10)$ as the response (probability of heavy tails).
+  %The binomial distribution model (right panel) uses p$(\nu < 10) > 0.5$ vs.\ p$(\nu < 10) \le 0.5$ as the response (heavy vs.\ not heavy tails).
+  %The models were fitted as mixed effects models with beta or binomial distributions, logit links, and multilevel intercepts for taxonomic class and order nested within class using the \texttt{glmmADMB} \texttt{R} package \citep{fournier2012,glmmadmb}.
+  %(Bottom panels) Model predictions for the strongest covariate --- time series length --- with all other predictors set to zero (their mean value when standardized).
 Dots represent observations for individual populations.
@@
     \label{fig:correlate-coefs}
+\end{center}
+\end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=0.75\textwidth]{perc-prob-increase-with-n}
+\caption{
+Expected percent increase in the probability of detecting heavy tails, Pr($\nu < 10$), with an additional ten time steps of data.
+These values are based on the curve shown in Fig.~\ref{fig:correlates}d, which is based on the coefficient in Fig.~\ref{fig:correlate-coefs}.
+For example, we have about a 12\% increased probability of observing heavy tails in a time series that has 60 time steps vs.\ a time series with 50 time steps.
+}
+\label{fig:perc-inc-p}
 \end{center}
@@
 \clearpage
+\noindent
+Stan code for the multilevel beta regression:
+\begin{footnotesize}
+\verbatiminput{betareg4.stan}
+\end{footnotesize}
+
+\clearpage
 

### 2014-10-09 anderson-etal-blackswan-timeseries.tex
Shrink space around equations
Add verbatim package
A whole lot of paper writing/editing

@@
 \usepackage{color}
+\usepackage{verbatim}
+\usepackage[sectionbib]{chapterbib}
+
+\usepackage{amsmath}
+% \makeatletter
+% \g@addto@macro\normalsize{%
+%   \setlength\abovedisplayskip{-20pt}
+%   \setlength\belowdisplayskip{0pt}
+%   \setlength\abovedisplayshortskip{-20pt}
+%   \setlength\belowdisplayshortskip{0pt}
+% }
+% \makeatother
 
@@
 
-\usepackage[round]{natbib}
-\bibliographystyle{apalike}
+\usepackage[round,sectionbib]{natbib}
 \bibpunct{(}{)}{;}{a}{}{;}
@@
 \usepackage{titlesec}
-\titlespacing\section{0pt}{12pt plus 4pt minus 2pt}{-6pt plus 2pt minus 2pt}
-\titlespacing\subsection{0pt}{12pt plus 4pt minus 2pt}{-8pt plus 2pt minus 2pt}
+\titlespacing\section{0pt}{6pt plus 4pt minus 2pt}{-6pt plus 2pt minus 2pt}
+\titlespacing\subsection{0pt}{0pt plus 4pt minus 2pt}{-8pt plus 2pt minus 2pt}
 
 \begin{document}
+
+\maketitle
+
 \input{values} % R output
 
+\emph{Authors are listed alphabetically (ABCD!) and open to reordering.}
+
+\textsuperscript{1}Earth to ocean research group, department of biological sciences, Simon Fraser University, Burnaby BC, V5A 1S6, Canada
+
+\textsuperscript{2}School of Aquatic and Fishery Sciences, University of Washington, Box 355020, Seattle, WA 98195, USA
+
+\textsuperscript{3}School of Resource and Environmental Management, Simon Fraser University, Burnaby, BC, V5A 1S6, Canada
+
+\textsuperscript{*}Corresponding author: Sean C.
+Anderson; Earth to Ocean Research Group, Department of Biological Sciences, Simon Fraser University, Burnaby BC, V5A 1S6; Phone: 1-778-782-3989; E-mail: sean\_anderson@sfu.ca
+
 %\begin{spacing}{1.4}
-%\doublespacing
-\onehalfspacing
+\doublespacing
+%\onehalfspacing
 
-\maketitle
 \thispagestyle{empty}
 
-\input{ms}
+\linenumbers
 
-\bibliography{/Users/seananderson/Dropbox/tex/jshort,/Users/seananderson/Dropbox/tex/ref3}
-%\bibliography{/Users/seananderson/Dropbox/tex/jshort,refs}
-
-\clearpage
-\section{Tables}
-
-\input{sparks.tex}
+\include{ms}
 
@@
 
-\input{figs.tex}
+\input{figs}
 
-\begin{centering}
-\LARGE
-Supporting material\\[1.5em]
-\end{centering}
+\include{supp}
 
-\input{supp}
+\renewcommand{\thetable}{S\arabic{table}}
+\setcounter{table}{0}
 
-\input{supp-figs}
+\include{stat-table}
+
+\include{supp-figs}
 

### 2014-10-09 supp-figs.tex
A whole lot of paper writing/editing

@@
 \begin{center}
+\includegraphics[width=0.27\textwidth]{order-level-estimates.pdf}
+\caption{
+Taxonomic-order-level posterior estimates of probabilty that $\nu < 10$ from a beta regression multilevel model accounting for the effect of time-series length.
+Estimates are at the geometric mean of time series length across all the data (approximately 27 time steps).
+Colour shading refers to the taxonomic classes as indicated in Fig.~\ref{fig:nu-coefs} and Fig.~\ref{fig:correlates} (yellow for fishes, green for insects, purple for birds, and red for mammals).
+Short vertical line segments indicate median posterior estimates.
+Long dotted vertical line throughout the figure indicates the median expected Pr($\nu < 10$) from the prior distribution.
+}
+\label{fig:order-estimates}
+\end{center}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[htbp]
+\begin{center}
 \includegraphics[width=\textwidth]{gomp-comparison.pdf}
@@
 \begin{center}
-\includegraphics[width=0.75\textwidth]{perc-prob-increase-with-n}
+\includegraphics[width=0.70\textwidth]{perc-prob-increase-with-n}
 \caption{

### 2014-10-16 ms.tex
Switch back to .tex files; add to SOM

%!TEX root = anderson-etal-blackswan-timeseries.tex

\noindent
\emph{Main message, to be deleted (25 words)}: Black swans are present \& taxonomically widespread but rare in population dynamics. Extreme climate, predation, parasites, and their interactions are common causes; intrinsic drivers are unclear.

\section{Abstract}

Black swans are statistically improbable events that nonetheless occur --- often with profound implications. While extremes in the physical environment, such as monsoons and heat waves, are widely studied and increasing in magnitude and frequency, it remains unclear the extent to which ecological populations buffer or suffer from extremes. Here, we develop a probability model to estimate the degree of heavy-tailedness (presence of black swans) in ecological process noise. We apply our model to \NPops~time series from around the world across \NOrders~taxonomic orders and seven classes. We find strong evidence of black swans, but they are rare, occurring in \overallMinPerc--\overallMaxPerc\% of populations; most frequently for birds (\AvesRangePerc\%) followed by mammals (\MammaliaRangePerc\%), insects (\InsectaRangePerc\%), and fishes (\OsteichthyesRangePerc\%). When they occur, they tend to be driven by climate and severe winters, cycles of parasites and predators, and interactions between these elements. They are more frequently detected for populations with longer time series and lower levels of process noise. We find little evidence of intrinsic life-history covariates. \emph{close with modelling and policy implications}

\section{Introduction}

Black swans are unexpected extreme events with dramatic consequences. One of the most striking black swans in ecology was the asteroid marking the mass extinctions of the K-T boundary (REF). Today, it is the extremeness of climate --- particularly in concert with shifts in mean temperature --- that is expected to cause the greatest societal damage (REF).

However, while extremes in the physical environment are present and widely accepted, it remains unclear the extent to which ecological systems buffer or suffer from black swans.

Existing evidence for black swans or heavy tails in population dynamics is thus far limited. Considerable work has focussed on heavy tails in physical environmental processes in, for example, rain fall (REF), wave height (REF), earth quake magnitude, climate temperature (REF). time to extinction, plankton time series, GPDD work found little evidence
subsequent work suggested, but...
Ward applied a data-intense method and found evidence for...

There are two key reasons why we may find little evidence of ecological black swans. First, they might not exist. Indeed, the vast majority of population dynamics model fitting and risk forecasting assumes that population dynamics are not heavy tailed (e.g.~REF, REF, REF). Alternatively, population dynamics might be heavy tailed, but a paucity of statistical methods and shortness of available time series has limited our ability to detect them. For example, a commonly applied probability distribution to detect the shape of time series tails (the generalized extreme value distribution) requires lumping data into the most extreme event per unit time (e.g.~heaviest rainfall per year) and therefore requires a considerable quantity of data (REF). Previous work into the shape of population dynamic tails, has focused on model comparison to identify parsimonious explanations from a set of particular hypotheses, but does not quantify the magnitude of heavy tails on a continuous scale or put the conclusions in a probabilistic framework (REF). Furthermore, there is evidence that information criteria are poor at distinguishing subtle differences in population dynamics for the relatively short time series that are most readily available (REF).

Or bring back in Ward - data intensive, can't be applied to most individual time series

plankton --- hideous distribution to fit, even harder in population dynamics framework -and what about seasonal cycles, density dependence?

GPDD - model competition and finding most parsimonious explanation from set of hypotheses --- but only \emph{very} heavy from not heavy

GEV --- super data intensive

tipping point and regime shift methods --- say something\ldots{} not appropriate for most of the data available

Here we develop a probabilist framework for identifying ecological black swans in population dynamic process noise --- the stochastic jumps from time step to time step. Our framework quantifies both the magnitude and probability of heavy-tailed dynamics, allows for a range of population dynamic models, can incorporate observation uncertainty, and is applicable to the lengths of abundance time series most commonly available. We apply our model to hundreds of populations from around the world across four taxonomic classes to address three questions: (1) how frequent and strong are black swans in population dynamics across taxonomy, (2) what are some verified causes of black swans in ecological time series, and (3) are there characteristics of time series or intrinsic life-history characteristics associated with heavy-tailed dynamics. We find a high probability of ecological black swans across a wide range of taxa but in a limited number of populations. When they occurred, extreme climate, predation, parasites, and their interactions were common documented causes. We found little evidence that intrinsic life-history characteristics were related to the probability of black swans.

Papers to work in: \citep{inchausti2002, halley2002, inchausti2001} \citep{jentsch2007} \citep{ward2007} \citep{garcia-carreras2011} \citep{sornette2009} \citep{nunez2012} \citep{thompson2013} \citep{beaugrand2012} \citep{pine-iii2009} \citep{doak2008} Extreme rain and monsoons over India increasing \citep{goswami2006} \citep{smale2013} \citep{easterling2000} \citep{scheffer2003} \citep{katz2005} \citep{taleb2007} \citep{vasseur2014} \citep{vert-pre2013} \citep{lindenmayer2010} \citep{valpine2002} \citep{gregory2010} \citet{garcia-carreras2011} \citeauthor{brook2006} \citetext{\citeyear{brook2006}; \citealp{herrandoprez2014}} \citep{sibly2005, ziebarth2010}

\section{Methods}

To obtain estimates of the probability and magnitude of black swans, we fitted population dynamic models to hundreds of abundance time series from around the world. The key addition of our models, compared to typical population models, was estimating the shape of the process noise tails --- how extreme the largest jumps from time step to time step were. We then compared how these estimates related to taxonomy, time series properties, and life-history characteristics, and for a subset of populations, we documented the established causes of ecological black swans.

\subsection{Time-series data}

We obtained abundance time series from the Global Population Dynamics Database (GPDD) \citep{gpdd2010}. The GPDD contains nearly 5000 time series of abundance across $\sim$ 1000 species and 100 taxonomic orders. To derive a high-quality subset of populations that were suitable for our analyses, we filtered the data (Supporting Material). Our filtering procedure removed populations from less reliable data sources, removed those without sufficient data for our models, and interpolated some missing values. Our final dataset contained \NPops~populations across \NOrders~taxonomic orders and seven taxonomic classes with a median of XX time steps (range of XX--XX) (Table S1, Fig.~\ref{fig:all-ts}).

\subsection{Population models}

Our main analysis focusses on the commonly applied phenomenological Gompertz population dynamics model. In previous analyses of the GPDD, the Gompertz has proved to be the most frequently parsimonious model \citep{brook2006}. We extended the traditional Gompertz model by allowing the process noise to be drawn from a Student-t distribution with degrees of freedom parameter $\nu$. If $\nu$ is small ($\lesssim 10$) the t distribution has much heavier tails than a normal distribution. For example, at $\nu = 2$, the probability of drawing a value more than five standard deviations below the mean is $1.8$\%, whereas the probability of drawing such a value from a normal distribution is nearly zero ($2.9\cdot10^{-5}$\%). As $\nu$ approaches infinity the distribution approaches the normal distribution (Fig.~\ref{fig:didactic}). By estimating the value of $\nu$ we can quantity how heavy-tailed the process noise deviations are.

The heavy-tailed Gompertz model is: 
\begin{align*}
x_t &= \ln N_t\\
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma),
\end{align*}
where $x_t$ is the $\ln$ abundance ($N$) at time $t$. The model is density independent if $b = 1$, maximally density dependent if $b = 0$, and inversely density dependent if $b < 0$. The parameter $\lambda$ represents the expected $\ln$ number of new individuals at time $t + 1$ if $N_t = 1$ (per capita population growth rate). The process noise is modelled as a Student-t distribution with degrees of freedom parameter $\nu$, a mean of $0$, and a scale parameter of $\sigma$.

We chose weakly informative priors to incorporate our understanding of plausible population dynamics \citep[Supporting Material]{gelman2014}. For $\nu$, we chose an exponential prior with rate parameter of $0.01$ truncated at values above two (Fig.~\ref{fig:priors}), a slightly less informative prior than that suggested by \citet{fernandez1998}. This prior gives only a 7.7\% probability that $\nu < 10$ but constrains the sampling sufficiently to avoid wandering off towards infinity --- above approximately $\nu = 20$ the t distribution is so similar to the normal distribution that time series of the length considered here are unlikely to be informative about the precise value of $\nu$ (Fig.~\ref{fig:didactic}).

We estimated all models in a Bayesian framework using Stan \citep{stan-manual2014}. Stan samples from the posterior distribution with a version of Hamiltonian Markov chain Monte Carlo called the No-U-Turn Sampler \citep{hoffman2014}. We assured that chains had sufficiently converged and the sampler had obtained sufficient independent samples from the posterior ($\widehat{R} < 1.05$, $n_\mathrm{eff} > 200$; Supporting Material). We used simulated data to test how how easily we could detect $\nu$ given different sample sizes and to ensure we could recover unbiased parameter estimates from the Gompertz model (Supporting Material, Fig.~\ref{fig:sim-nu}, \ref{fig:sim-gompertz}, \ref{fig:sim-gompertz-boxplots}).

We fit four alternative population models to test if they systematically changed our conclusions. (1) Autocorrelation has been suggested as a reason for increased observed variability of abundance time series through time, which could create apparent heavy tails \citep{inchausti2002}; therefore, we fit a model that modelled the serial correlation in the residuals. (2) Previous work has focussed modelled growth rates without accounting for density dependence \citep{segura2013}; therefore, we fit a simpler model in which we assumed density independence. (3) Observation error could bias parameter estimates \citep{knape2012} or mask our ability to detect heavy tails \citep{ward2007}. However, simultaneously estimating observation error and process noise in a state space framework is a challenging computational problem \citep[e.g.][]{dennis2006, knape2008} even with longer time series and without estimating the shape of the process error tails; therefore, we fit a model where we assumed a level of observation error ($0.2$ standard deviations on a log scale). (4) The Gompertz model assumes that population growth rate declines linearly with log abundance. We also fit an alternative model, the Ricker-logistic model, which assumes that population growth rate declines linearly with abundance itself (Supporting Material).

\subsection{Covariates of population dynamic black swans}

We investigated possible covariates of heavy-tailed population dynamics visually and through multilevel modelling. We plotted characteristics of the time series ($\sigma$, $\lambda$, $b$, and time-series length) along with two life-history characteristics (body length and maximum lifespan obtained from \citet{brook2006a}) against our estimates of $\nu$. We formally investigated these relationships by fitting beta regression multilevel models to the probability that $\nu$ was less than $10$ (i.e.~approximately the probability of heavy tails). For covariates that were derived from Gompertz model parameter estimates, we incorporated standard deviations around the means. To account for broad patterns of phylogenetic relatedness, we allowed for varying intercepts at the taxonomic class, order, and species level (Supporting Material).

Finally, we investigated a sample of populations that our method categorized as having a high probability of heavy tails (Pr$(\nu < 10) > 0.5$). Where possible, we found the documented causes of ecological black swans in the primary data source cited in the GPDD or in other literature describing the population. These populations were chosen haphazardly with the purpose of generating hypotheses about the causes of population dynamic black swans.

\section{Results}

We found strong but rare evidence for black-swan dynamics. Defining black-swan dynamics as a greater than $0.5$ probability that $\nu < 10$, we observed black swans most frequently for birds (X\%) followed by mammals (X\%), insects (X\%), and fishes (X\%) (Fig.~\ref{fig:nu-coefs}, \ref{fig:heavy-ts}). Black swans were taxonomically widespread, occurring for at least one population in X\% of the taxonomic orders recorded. Accounting for time series length and partially pooling inference across taxonomic class and order with a multilevel model, there was stronger evidence for black swans in insect populations than is apparent in Fig.~\ref{fig:nu-coefs} --- four of 10 orders with the highest median probability of heavy tails were insect orders --- however, there was considerable uncertainty in these estimates (Fig.~\ref{fig:order-estimates}).

The majority of our heavy-tailed estimates were robust to alternative population models and observation error (Fig.~\ref{fig:alt}). Including an autocorrelation structure in the residuals, modelling population growth rates as a random walk with drift, or modelling the population dynamics as Ricker-logistic did not systematically alter our conclusions. Allowing for a fixed quantity of observation error decreased our estimates of black-swan dynamics, increasing the median estimate of $\nu$ from $<10$ to $>10$ in X of XX populations although the majority of $\nu$ estimates remained qualitatively similar (Fig.~\ref{fig:alt})

Across populations, the probability of observing black swan dynamics was positively related to time-series length and negatively related to magnitude of process noise ($\sigma$) but not clearly related to population growth rate ($\lambda$), density dependence ($b$), or maximum lifespan (Fig.~\ref{fig:correlates}, \ref{fig:correlate-coefs}). Longer time-series length was the strongest covariate of observing black swan dynamics (as was similarly observed in simulation testing, Fig.~\ref{fig:sim-nu}). For example, we were about 15\% TODO more likely to observe heavy tails in a population with 40 time steps than in one with 30 time steps (Fig.~\ref{fig:perc-inc-p}). However, the absolute change in probability with increased time series length was small ($0.X$ vs. $0.X$ in the previous example, Fig.~\ref{fig:correlates}).

We assembled the documented causes of black swan population dynamics in seven populations with high probabilities of heavy tails (Table 1). The majority of documented events were downward black swans and the majority involved a confluence of events. For example, a shortage of nest sites for the shag in the UK was thought to reduce population productivity and make the population vulnerable when a red-tide hit in 1968. The population experienced a black swan downswing, but the new availability of nest sites resulted in a rapid population upswing \citep{potts1980}. As another example, an interaction between population cycles caused by predation combined with cycles caused by the environment are thought to have caused a downward black swan for a water vole population \citep{saucy1994}. Other black swans were the result of a sequence of extreme climate events on their own. For example, severe winters in 1929, 1940--1942, and 1962--1963 were associated with black swan downswings in grey heron abundance in the UK \citep{stafford1971}. Our analysis finds that the last event was a combination of two black swan events in a row and population recovery took three times longer than predicted \citep{stafford1971}.

\section{Discussion}

We found strong evidence that the multiplicative process noise jumps from time step to time step were greater than we'd expect under a normal distribution and a Gompertz or Ricker-logistic model in XX--XX\% of populations assessed. We detected black swans more frequently for longer time series and for populations with less typical magnitude of process error, as we might expect. However, we failed to find strong evidence that black swans were associated with density dependence, population growth rate, or lifespan. In cases that we verified, black swans were often a result of the interaction between elements of extreme climate, predators and parasite cycles, and strong changes in human pressures. Our empirical results, sensitivity tests, and simulation tests suggest that estimating the tail shape of process error is a viable method of detecting heavy-tailed population dynamics. The presence of black swans highlights the importance of developing management strategies that detect quickly, respond to, and are robust to extremes in population dynamics --- particularly as the frequency and magnitude of climatic extremes increase over the next century.

Our results clarify, but do not disagree with previous related analyses. Previous work with an older version of the GPDD assessed heavy-tailed behaviour but focused on identifying if the lognormal was the most frequently parsimonious distribution of population abundance \citep{halley2002}. For heavy-tailed distributions, \citet{halley2002} fit the extremely heavy-tailed Cauchy distribution and the four-parameter Levy stable distribution and found little information criteria support for these distributions in longer time series. However, black swan events are by definition rare, and the majority of time series we studied are short for these purposes. Therefore, we would not expect to observe black swans in a large proportion of populations. Furthermore, by probabilistically quantifying process noise tail behaviour on a continuous spectrum, and allowing for non-stationary time series and density dependence, our analysis allows for a more nuanced description of the evidence of ecological black swans. In an earlier study, \citet{keitt1998} described heavy (power law) tails in breeding bird population abundance. However, this finding was challenged by \citet{allen2001}, who showed that mixing data across species could create falsely generate heavy tails. This work directly feeds into our use of the t distribution to represent heavy tails, since a t distribution can be represented by a mixture of normal distributions (with inverse-gamma-distributed variances).

We might expect to observe black swan dynamics in ecological time series because of unmodelled intrinsic properties of populations or extrinsic forces acting on populations. A t distribution can be formed through a mixture of normal distributions \citep[in which the variances are inverse-gamma distributed,][]{gelman2014}. Therefore, if we miss some underlying mixture of processes we might expect to observe heavy tails \citep{allen2001}. That process might be an aggregation of populations across space, or population diversity, or some intrinsic change in population variability through time. Extrinsic forces could also cause black swan dynamics (REF). These forces could be extreme themselves. For example, extreme climate, predation from or competition with other species experiencing black swans, or sharp changes in human pressure such hunting, fishing, or habitat destruction might cause black swans. Alternatively, the interaction of multiple ``normal'' forces could give rise to black swan ecological dynamics. This could occur if the interaction nature is synergistic \citep[e.g.][]{kirby2009} or even if non-synergistic interactions experience a rare alignment \citep{denny2009}.

There a number of important caveats when considering the generality of our results. First, the GPDD data represent a taxonomically and geographically biased sample of populations. The longer time series we focus on are dominated by commercially and recreationally important species and a disproportionate number of populations are located in the United Kingdom. Although we expect that we would find strong but rare evidence for black swans for other taxa around the world, the common forces driving those black swans (e.g.~severe winters in Table 1) likely differ. Second, in a large database with disparate datasources, we cannot ensure the quality of every datapoint. Recording mistakes undoubtedly comprise some of the black swans we detected. Conversely, recording mistakes could mask heavy tails if initial observations are discarded or altered as assumed recording errors. Third, the temporal scale of observation and population dynamics vary considerably across populations in the GPDD and this likely influences the detection of heavy tails. For example, if we make frequent observations relative to generation time (e.g.~for many large bodied mammals) we will average across generations and perhaps not observe the tails. Furthermore, if we record population censuses infrequently relative to generation time (e.g.~many insects in the GPDD) the recorded data may average across extreme and less-extreme events and dampen observed black swan dynamics.

Our results suggest a number of important future issues to address regarding ecological black swans. First, given that black swans can and do occur, can we forecast the probability of black swans in space and time? Furthermore, what management policies allow us to detect them quickly after they happen? Secondly, can we isolate the components of ecological dynamics that experience black swans events by moving from phenomenological models such as the Gompertz to mechanistic models, that take into account, for example, recruitment? Thirdly, the only way we may gain further insight into the mechanisms and covariates of ecological black swans may be to focus on specific geographic and taxonomic subsets of data where longer time series are available \citep[e.g.][]{segura2013}. Finally, if a sizeable fraction of populations experience heavy tails in population process noise, what is the impact of allowing for black swans in forecasts of ecological risk? Risk can be defined as the probability of a specific magnitude of undesired event occurring (REF). Therefore, financial analysts are acutely concerned with the shape of distribution tails (REF). It is likely that a comparable focus in ecology would alter our perception of forecasts of, for example, extinction risk.

Policy implications

\section{Acknowledgements}

We thank members of the Earth to Ocean research group for helpful discussions and comments. We are grateful to the contributors and maintainers of the Global Population Dynamics Database and to Compute Canada's WestGrid high-performance computing resources. Three silhouette images were obtained from \texttt{phylopic.org}: rabbit (Sarah Werning), hoverfly (Gareth Monger), and bird (Jean-Raphaël Guillaumin {[}photography{]} and T. Michael Keesey {[}vectorization{]}) under a Creative Commons Attribution 3.0 Unported license. Funding was provided by a Simon Fraser University Graduate Fellowship (SCA), NSERC (NKD, ABC), the Canada Research Chairs Program (NKD), and TAB TODO?.

\bibliographystyle{ecologyletters}
\bibliography{/Users/seananderson/Dropbox/tex/jshort,/Users/seananderson/Dropbox/tex/ref3}

\clearpage

\section{Tables}

\input{sparks.tex}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{t-nu-eg.pdf}
\caption{
An illustration of fitting population dynamics models that allow for heavy tails, represented by the Student-t degrees of freedom parameter $\nu$.
(a, b) The probability density for t distributions with a scale parameter of 1 and different values of $\nu$.
Small values of $\nu$ create heavy tails.
As $\nu$ approaches infinity the distribution approaches the normal distribution.
For example, at $\nu = 2$, the probability of drawing a value less than -5 is 1.8\%, whereas the probability of drawing such a value from a normal distribution is nearly zero ($2.9\cdot10^{-5}$\%).
(c--e) Simulated population dynamics from a Gompertz model with process noise drawn from t distributions with different values of $\nu$.
Coloured dots in panels c and d represent jumpts with less than a 1 in 1000 chance of occurring in a normal distribution.
(f--h) Estimates of $\nu$ from models fit to the times series in panels c--e.
Shown are the posterior samples (histograms), median and interquartile range of the posterior (IQR) (dots and line segments), and the exponential prior on $\nu$ (dashed lines).
Colour shading behind panels f--h illustrates the region of heavy tails.}
\label{fig:didactic}
\end{center}
\end{figure}

\clearpage

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.44\textwidth]{nu-coefs-2.pdf}
\caption{
Estimates of how heavy-tailed population dynamics are for \nuCoefPopN\ populations of birds, mammals, insects, and fishes.
Small values of $\nu$ ($\lesssim 10$) suggest heavy-tailed dynamics; larger values of $\nu$ suggest approximately normal-tailed dynamics.
Vertical points and line segments represent posterior medians and 50\% / 90\% credible intervals for individual populations.
Inset plots show probability that $\nu < 10$ (probability of heavy tails) for populations arranged by taxonomic order.
Taxonomic orders are sorted by decreasing mean p($\nu < 10$).
Taxonomic orders with three or fewer populations in panel a are omitted for space.
Red and orange colours highlight populations with a high probability of heavy-tailed dynamics.
}
\label{fig:nu-coefs}
\end{center}
\end{figure}

\clearpage

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.9\textwidth]{correlates-p10.pdf}
\caption{
Potential correlates of heavy-tailed behaviour (indicated by a high probability that $\nu < 10$).
Shown are (a--c) parameters from the Gompertz heavy-tailed population model, (b) number of time steps, (c) body length, and (d) lifespan.
For the Gompertz parameters, $\sigma$ refers to the scale parameter of the Student-t process-noise distribution, $\lambda$ refers to the population growth rate at an abundance of one, $b$ refers to the density dependence parameter ($1$ is maximally density independent, $0$ is maximally density dependent, and $<0$ is inversely density dependent).
%For parameter values, points and lines represent medians and 50\% credible intervals of the posterior.
Circles representing a limited number of sharks, crustaceans, and gastropods are filled in white.
Median and 90\% credible interval posterior predictions of a beta regression multilevel model are shown in panels a and d where there was a high probability the slope coefficient was different from zero (Supporting Material).
}
\label{fig:correlates}
\end{center}
\end{figure}

### 2014-10-16 supp.tex
Switch back to .tex files; add to SOM

%!TEX root = anderson-etal-blackswan-timeseries.tex

\begin{centering}
\LARGE
Supporting Material for TODO INSERT TITLE HERE\\[1.5em]
\end{centering}

\section{Data selection}

We followed the following data selection and quality-control rules with the GPDD:

\begin{enumerate}
\item
  To remove populations with unreliable population indices that could be strongly confounded with economics and sampling effort, we removed all populations with a sampling protocol listed as \texttt{harvest} as well populations with the words \texttt{harvest} or \texttt{fur} in the cited reference title.
\item
  We removed all populations with uneven sampling intervals. I.e.~we removed populations that didn't have a constant difference between the ``decimal year begin'' and ``decimal year end'' columns.
\item
  We removed all populations rated as $< 2$ in the GPDD quality assessment (on a scale of $1$ to $5$, with $1$ being the lowest quality data) \citep[following][]{sibly2005, ziebarth2010}
\item
  Populations with negative abundance values were assumed to be log values and transformed by taking $10$ to the power of the recorded abundance. In many cases this was noted for the population, but not in all cases. We inspected each of these \totalAssumedLog~time series to make sure our assumption made sense.
\item
  We filled in all missing time steps with \texttt{NA} values and imputed single missing values with the geometric mean of the previous and following values. We chose a geometric mean to be linear on the log scale that the Gompertz and Ricker-logistic models were fitted on.
\item
  We filled in single recorded values of zero with the lowest non-zero value in the time series \citep[following][]{brook2006a}. This assumes that single values of zero result from abundance being low enough that censusing missed present individuals. We turned multiple zero values in a row into \texttt{NA} values. This implies that multiple zero values were either censusing errors or caused by emigration. Regardless, our population models were fitted on a multiplicative (log) scale and so could not account for zero abundance.
\item
  We removed all populations with four or more identical values in a row since these suggest either recording error or extrapolation between two observations.
\item
  We removed all populations without at least four unique values \citep[following][]{brook2006a}.
\item
  We then wrote an algorithm to find the longest unbroken window of abundance (no \texttt{NA}s) with at least 20 time steps in each population time series. If there were any populations with multiple windows of identical length, we took the most recent window. This is a longer window than used in some previous analyses \citep[e.g.][]{brook2006a}, but since our model attempts to capture the shape of the distribution tails, our model requires more data.
\item
  Finally, we removed GPDD Main ID \texttt{20531}, which we noticed was a duplicate of \texttt{10139} (a heron population).
\end{enumerate}

We provide a supplemental figure of all the time series included in our analysis and indicate which values were interpolated (\percImputedPops\% of populations had at least one point interpolated and only \percImputedPoints\% of the total observations were interpolated) (Fig.~\ref{fig:all-ts}). Table S1 shows the final taxonomic breakdown and the number of populations with interpolated values.

\section{Details on the heavy-tailed Gompertz probability model}

For the Gompertz model, our weakly-informative priors (Fig.~\ref{fig:priors}) were: 
\begin{align*}
b &\sim \mathrm{Uniform}(-1, 2)\\
\lambda &\sim \mathrm{Normal}(0, 10^2)\\
\sigma &\sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)\\
\nu &\sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2). \end{align*}
Our prior on $b$ was uninformative between values of $-1$ and $2$. We would not expect values of $b$ with levels of inverse density dependence as low as $-1$, nor would we expect values above $1$. We allowed values of $b$ above $1$ to allow for somewhat non-stationary time series of growth rates. The estimates of $b$ were well within these bounds. Our prior on $\lambda$ was very weakly informative within the range of expected values for population growth and is similar to priors suggested by \citet{gelman2008d} for intercepts of regression models. Our prior on $\sigma$ follows \citet{gelman2006c} and \citet{gelman2008d} and is based on our expected range of the value in nature from previous studies \citep[e.g.][]{connors2014}. In our testing of a subsample of populations, our parameter estimates were not qualitatively changed by switching to an uninformative uniform prior on $\sigma$, but the weakly informative prior substantially sped up chain convergence.

Our prior on $\nu$ is based on \citet{fernandez1998}; they chose a more informative exponential rate parameter of 0.1. We chose a less informative rate parameter of 0.01 and truncated the distribution at two, since at $\nu < 2$ the variance of the t distribution is undefined. This prior gives only a 7.7\% probability $\nu < 10$ but constrains the sampling sufficiently to avoid wandering off towards infinity --- above approximately $\nu = 20$ the t distribution is so similar to the normal distribution (Fig.~\ref{fig:didactic}) that time series of the length considered here are unlikely to be sufficiently informative about the precise value of $\nu$. In the scenario where the data are not heavy tailed (e.g.~Fig.~\ref{fig:didactic}e, h) the posterior will approximately match the prior (median $= 71$, mean $= 102$) and not be flagged as likely heavy tailed using the metrics we used in our paper (e.g.~Pr($\nu < 10$) \textgreater{} 0.5).

We fitted our models with Stan 2.4.0 \citep{stan-manual2014}, and R 3.1.1 \citep{r2014}. We began with four chains and 2000 iterations, discarding the first 1000 as warm up (i.e.~4000 total samples). If $\hat{R}$ (a measure of chain convergence) was greater than 1.05 for any parameter or the minimum effective sample size, $n_\mathrm{eff}$, (a measure of the effective number of uncorrelated samples) for any parameter was less than 200, we doubled both the total iterations and warm up period and sampled from the model again. These thresholds are in excess of the minimums recommended by \citet{gelman2006a} of $\hat{R} < 1.1$ and effective sample size $> 100$ for reliable point estimates and confidence intervals. In the majority of cases our minimum thresholds were greatly exceeded. We continued this procedure up to 8000 iterations (16000 total samples) by which all chains were deemed to have sufficiently converged.

\section{Simulation testing the model}

We performed two types of simulation testing. First, we tested how easily the Student-t distribution $\nu$ parameter could be recovered given different true values of $\nu$ and sample sizes. Second, we tested the ability of the heavy-tailed Gompertz model to obtain unbiased parameter estimates of $\nu$ given that a set of process deviations was provided in which the ``effective $\nu$'' value was close to the true $\nu$ value.

We separated our simulation into these two components to avoid confounding two issues. (1) With smaller sample sizes, there may not be a stochastic draw from the tails of a distribution. In that case, no model, no matter how perfect a model, will be able to detect the shape of the tails. (2) Models (particularly complex models) may return biased parameter estimates if there are conceptual, computational, or coding errors in the models. Our first simulation tested the first issue and our second simulation tested the latter. In general, out simulations show that, if anything, our model under predicts the magnitude and frequency of heavy tailed events --- especially given the length of the time series in the GPDD.

\subsection{Estimating $\nu$ simulations}

First, we tested the ability to estimate $\nu$ given different true values of $\nu$ and sample sizes. We took stochastic draws from t distributions with different $\nu$ values ($\nu = 3, 5, 10,$ and $10^6$ [$\approx$ normal]), with central tendency parameters of $0$, and scale parameters of $1$. We started with 1600 stochastic draws and then fitted again at the first 800, 400, 200, 100, 50, and 25 draws. Each time we recorded the posterior samples of $\nu$.

We found that we could consistently and precisely recover median posterior estimates of $\nu$ near the true value of $\nu$ with large samples ($\ge 200$) (Fig.~\ref{fig:sim-gompertz} upper panels). At smaller samples the we could still usually distinguish ``heavy'' from ``not-heavy'' tails, but the model tended to underestimate how heavy the tails were. At the same time, when the model underestimated how heavy-tailed the distribution was, it tended to overestimate how large the scale parameter was (Fig.~\ref{fig:sim-gompertz} lower panels).

\subsection{Heavy-tailed Gompertz model simulations}

%2 parts: how many samples from the true population t distribution do you need to detect low nu? And, given that you have a set of deviations in which nu is detectable (effective nu is within 0.5 CV of true nu), can the more complex Gompertz still capture this?

In the second part of our simulation testing, we tested the ability of the heavy-tailed Gompertz model to obtain unbiased parameter estimates given that process deviations were provided in which our simpler model (in the previous section) could accurately estimate the value of $\nu$. To generate these process errors, we repeatedly generated candidate process deviations and estimated the central tendency, scale, and $\nu$ values each time. We recorded when the estimated nu was within $0.2$ CVs (coefficient of variations) of the true $\nu$ value and used this set of random seed values in our Gompertz simulation. We then fitted our Gompertz models to the simulated datasets with all parameters (except $\nu$) set near the median values estimated in the GPDD. We repeated this with $50$ and $100$ samples without observation error, $50$ samples with observation error ($\sigma_\mathrm{obs} = 0.2$), and $50$ samples with the same observation error and a Gompertz model that allowed for (correctly) specified observation error.

Our results show...

Fig.~\ref{fig:sim-gompertz}
Fig.~\ref{fig:sim-gompertz-boxplots}
Fig.~\ref{fig:sim-prob}

%Simulation testing: the ability to recover $\nu$ when randomly sampling from various distributions (Fig.~\ref{fig:sim-nu}); heavy-tailed Gompertz model performance and confidence interval coverage given the process noise has deviations that are known to have effective $\nu$ equal to true $\nu$ (Fig.~\ref{fig:sim-gompertz}); boxplots of the same output (Fig.~\ref{fig:sim-gompertz-boxplots}); probability that $\nu < 10$ for the Gompertz simulation testing (Fig.~\ref{fig:sim-prob}).

\section{Alternative population models}

We fit four alternative population models to the time-series data to check how they would influence our conclusions. Our alternative models allowed for autocorrelation in the residuals, assumed no density dependence, allowed for observation error, or assumed a Ricker-logistic functional form. The ranges of percentages of black swans by taxonomic class cited in the abstract are based on lower and upper limits across our main Gompertz model and these four alternative models.

\subsection{Autocorrelated residuals}

We considered a version of the Gompertz model in which an autoregressive parameter was fitted to the process noise residuals: 
\begin{align*}
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, \phi \epsilon_{t-1}, \sigma).
\end{align*}
In addition to the parameters in the original Gompertz model, we estimate an additional parameter $\phi$, which represents the relationship between subsequent residuals. Based on the results of previous analyses with the GPDD \citep[e.g.][]{connors2014} and the chosen priors in previous analyses \citep[e.g.][]{thorson2014a} and to greatly speed up chain convergence when running our model across all populations, we placed a weakly informative prior on $\phi$ that assumed the greatest probability density near zero with the reduced possibility of $\phi$ being near $-1$ or $1$. Specifically, we chose $\phi \sim \mathrm{Truncated\mhyphen Normal}(0, 1, \mathrm{min.} = -1, \mathrm{max.} = 1)$. The MCMC chains for a small number of these models (XX) did not converge according to our criteria ($\widehat{R} < 1.05, n_\mathrm{eff} > 200$) after 8000 iterations of four chains. We do not show these models in Fig.~\ref{fig:alt}.

\subsection{Assumed density independence}\label{assumed-density-independence}

We fit a simplified version of the Gompertz model in which the density dependence parameter $b$ was fixed at $1$ (density independent). This is equivalent to fitting a random walk model (with drift) to the $\ln$ abundances or assuming the growth rates are stationary. The model was as follows:

\begin{align*}
x_t &= \lambda + x_{t-1} + \epsilon_t\\
\epsilon &\sim \mathrm{Student\mhyphen t}(\nu, \epsilon, \sigma).
\end{align*}
We fit this model for three reasons: (1) it is computationally simpler and so provides a check that our more complicated full Gompertz model was obtaining ballpark reasonable estimates of $\nu$, (2) it provides a test of whether density dependence was systematically affecting our perception of heavy tails, (3) it matches how some previous authors have modelled heavy tails without accounting for density dependence \citep{segura2013}.

\subsection{Assumed observation error}

Observation error can bias parameter estimates \citep[e.g.][]{knape2012} and is known to affect the ability to detect extreme events \citep{ward2007}. However, simultaneously estimating observation and process error in state-space is a challenging problem and is known to result in identifiability issues with the Gompertz population model \citep{knape2008}. Furthermore, our model estimates an additional parameter --- the shape of the process error distribution tails --- potentially making identifiability and computational issues even greater. Therefore, we considered a version of the base Gompertz model that allowed for a fixed level of observation error:
\begin{align*}
U_t &= \lambda + b U_{t-1} + \epsilon_t\\
x_t &\sim \mathrm{Normal}(U_t, \sigma_\mathrm{obs})\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma_\mathrm{proc}),
\end{align*}
where $U$ represents the unobserved state vector, and $\sigma_\mathrm{obs}$ represents the standard deviation of observation error (on a log scale). We set $\sigma_\mathrm{obs}$ to $0.2$, which represents the upper end of values often used in simulation analyses \citep[e.g.][]{valpine2002, thorson2014b}.

\subsection{Ricker-logistic}

We also fitted a Ricker-logistic model:
\begin{align*}
x_t &= x_{t-1} + r_{\mathrm{max}}\left(1 - \frac{N_{t-1}}{K}\right) + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma),
\end{align*}
where $r_\mathrm{max}$ represents the maximum population growth rate that is obtained when $N$ (abundance) $= 0$. The parameter $K$ represents the carrying capacity and, as before, $x_t$ represents the $\ln$ transformed abundance at time $t$. The Ricker-logistic model assumes a linear decrease in population growth rate ($x_t - x_{t-1}$) with increases in abundance ($N_t$). In contrast, the Gompertz model assumes a linear decrease in population growth rate with increases in $\ln$ abundance ($x_t$) (REF).

To fit the Ricker-logistic models, we chose a prior on $K$ uniform between zero and twice the maximum observed abundance (\citet{clark2010} chose uniform between zero and maximum observed, which is more informative). We set the prior on $r_\mathrm{max}$ as uniform between 0 and 20 as in \citet{clark2010}. We used the same priors on $\nu$ and $\sigma$ as in the Gompertz model.

\section{Modelling covariates of heavy-tailed dynamics}

We fitted a multilevel beta regression model to the predicted probability of heavy tails, Pr($\nu < 10$), to investigate potential covariates of heavy-tailed dynamics. The beta distribution is useful when the response data range on a continuous scale between zero and one. We used a logit link function as is typically used in logistic regression. The model was as follows: 
\begin{align*}
\mathrm{Pr}(\nu_i < 0.5) &\sim \mathrm{Beta}(A_i, B_i)\\
\mu_i &= \mathrm{logit}^{-1}(\alpha 
  + \alpha^\mathrm{class}_{j[i]}
  + \alpha^\mathrm{order}_{k[i]} 
  + \alpha^\mathrm{species}_{l[i]}
  + X_i \beta),
  \: \text{for } i = 1, \dots, 617\\
A_i &= \phi_\mathrm{disp} \mu_i\\
B_i &= \phi_\mathrm{disp} (1 - \mu_i)\\
\alpha^\mathrm{class}_j &\sim 
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{class}}), 
  \: \text{for } j = 1, \dots, 6\\
\alpha^\mathrm{order}_k &\sim 
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{order}}), 
  \: \text{for } k = 1, \dots, 38\\
\alpha^\mathrm{species}_l &\sim 
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{species}}),
  \: \text{for } l = 1, \dots, 301,\\
\end{align*}
where $A$ and $B$ represent the beta distribution shape parameters; $\mu_i$ represents the predicted value for population $i$, class $j$, order $k$, and species $l$; $\phi_\mathrm{disp}$ represents the dispersion parameter; and $X_i$ represents a vector of predictors for population $i$ with associated $\beta$ parameters. The intercepts are allowed to vary from the overall intercept $\alpha$ by taxonomic classes ($\alpha^\mathrm{class}_j$), taxonomic orders ($\alpha^\mathrm{order}_k$), and species ($\alpha^\mathrm{species}_l$) with standard deviations $\sigma_{\alpha \; \mathrm{class}}$, $\sigma_{\alpha \; \mathrm{order}}$, and $\sigma_{\alpha \; \mathrm{species}}$. Where possible, we also allowed for error distributions around the predictors by incorporating the standard deviation of the posterior samples for the Gompertz parameters $\lambda$, $b$, and $\sigma$ around the mean point value (not shown in the above equation). We log transformed $\sigma$, time-series length, and lifespan to match the way they are visually represented in Fig.~\ref{fig:correlates}. All input variables were standardized by subtracting their mean and dividing by two standard deviations \citep{gelman2008c} to make their coefficients comparable in magnitude.

We incorporated weakly informative priors into our model: $\mathrm{Cauchy}(0, 10)$ on the global intercept $\alpha$, $\mathrm{Half\mhyphen Cauchy}(0, 2.5)$ on all standard deviation parameters, $\mathrm{Half\mhyphen Cauchy}(0, 10)$ on the dispersion parameter $\phi_\mathrm{disp}$, and $\mathrm{Cauchy}(0, 2.5)$ on all other parameters \citep{gelman2006c,gelman2008d}. Compared to normal priors, the Cauchy priors concentrate more probability density at reasonably parameter values while allowing for higher probability density far into the tails, thereby allowing the data to dominate the posterior more strongly if it disagrees with the prior. Our conclusions were not qualitatively changed by using uniform priors. We fitted our models with 5000 total iterations per chain, 2500 warm-up iterations, four chains, and discarding every second sample to save memory. We checked for chain convergence visually and with the same criteria as before ($\widehat{R} < 1.05$ and $n_\mathrm{eft} >200$ for all parameters).

To derive taxonomic-order-level estimates of the probability of heavy tails accounting for time-series length (Fig \ref{fig:order-estimates}), we fitted a separate multilevel model with the same structure but with only time-series length as a predictor. (In this case, we did not want to control for intrinsic population characteristics such as density dependence.) We obtained order-level estimates by adding the posteriors for $\alpha$, $\alpha^\mathrm{class}_j$, and $\alpha^\mathrm{order}_k$.

\bibliographystyle{ecologyletters}
\bibliography{/Users/seananderson/Dropbox/tex/jshort,/Users/seananderson/Dropbox/tex/ref3}

### 2014-10-17 ms.tex
Edit ms

@@
 Here we develop a probabilist framework for identifying ecological black swans in population dynamic process noise --- the stochastic jumps from time step to time step.
-Our framework quantifies both the magnitude and probability of heavy-tailed dynamics, allows for a range of population dynamic models, can incorporate observation uncertainty, and is applicable to the lengths of abundance time series most commonly available.
+Our framework quantifies both the magnitude and probability of heavy-tailed dynamics, allows for a range of population dynamic models, can incorporate observation uncertainty, and is applicable to commonly available lengths of abundance time series.
 We apply our model to hundreds of populations from around the world across four taxonomic classes to address three questions: (1) how frequent and strong are black swans in population dynamics across taxonomy, (2) what are some verified causes of black swans in ecological time series, and (3) are there characteristics of time series or intrinsic life-history characteristics associated with heavy-tailed dynamics.
@@
 To obtain estimates of the probability and magnitude of black swans, we fitted population dynamic models to hundreds of abundance time series from around the world.
-The key addition of our models, compared to typical population models, was estimating the shape of the process noise tails --- how extreme the largest jumps from time step to time step were.
+The key addition of our models, compared to typical population models, was estimating the shape of the process noise tails.
 We then compared how these estimates related to taxonomy, time series properties, and life-history characteristics, and for a subset of populations, we documented the established causes of ecological black swans.
@@
 We obtained abundance time series from the Global Population Dynamics Database (GPDD) \citep{gpdd2010}.
-The GPDD contains nearly 5000 time series of abundance across $\sim$ 1000 species and 100 taxonomic orders.
+The GPDD contains nearly 5000 time series of abundance across $\sim$1000 species and $\sim$100 taxonomic orders.
 To derive a high-quality subset of populations that were suitable for our analyses, we filtered the data (Supporting Material).
-Our filtering procedure removed populations from less reliable data sources, removed those without sufficient data for our models, and interpolated some missing values.
+Our filtering removed populations from less reliable data sources, removed those without sufficient data for our models, and interpolated some missing values.
 Our final dataset contained \NPops~populations across \NOrders~taxonomic orders and seven taxonomic classes with a median of XX time steps (range of XX--XX) (Table S1, Fig.~\ref{fig:all-ts}).
@@
 In previous analyses of the GPDD, the Gompertz has proved to be the most frequently parsimonious model \citep{brook2006}.
-We extended the traditional Gompertz model by allowing the process noise to be drawn from a Student-t distribution with degrees of freedom parameter $\nu$.
+We extended the traditional Gompertz model by allowing the process noise to be drawn from a Student-t distribution with degrees of freedom parameter $\nu$ (Fig.~\ref{fig:didactic}).
 If $\nu$ is small ($\lesssim 10$) the t distribution has much heavier tails than a normal distribution.
 For example, at $\nu = 2$, the probability of drawing a value more than five standard deviations below the mean is $1.8$\%, whereas the probability of drawing such a value from a normal distribution is nearly zero ($2.9\cdot10^{-5}$\%).
-As $\nu$ approaches infinity the distribution approaches the normal distribution (Fig.~\ref{fig:didactic}).
-By estimating the value of $\nu$ we can quantity how heavy-tailed the process noise deviations are.
+As $\nu$ approaches infinity the distribution approaches the normal distribution.
+By estimating the value of $\nu$ we can quantity how heavy-tailed the process noise deviations are (Fig.~\ref{fig:didactic}).
 
@@
 We assured that chains had sufficiently converged and the sampler had obtained sufficient independent samples from the posterior ($\widehat{R} < 1.05$, $n_\mathrm{eff} > 200$; Supporting Material).
-We used simulated data to test how how easily we could detect $\nu$ given different sample sizes and to ensure we could recover unbiased parameter estimates from the Gompertz model (Supporting Material, Fig.~\ref{fig:sim-nu}, \ref{fig:sim-gompertz}, \ref{fig:sim-gompertz-boxplots}).
+We used simulated data to test how easily we could detect $\nu$ given different sample sizes and to ensure we could recover unbiased parameter estimates from the Gompertz model (Supporting Material, Fig.~\ref{fig:sim-nu}, \ref{fig:sim-gompertz}, \ref{fig:sim-gompertz-boxplots}).
 
@@
 
-We found strong evidence that the multiplicative process noise jumps from time step to time step were greater than we'd expect under a normal distribution and a Gompertz or Ricker-logistic model in XX--XX\% of populations assessed.
-We detected black swans more frequently for longer time series and for populations with less typical magnitude of process error, as we might expect.
+We found strong evidence that the multiplicative process noise jumps from time step to time step were greater than we would expect under a normal distribution and a Gompertz or Ricker-logistic model in XX--XX\% of populations assessed.
+We detected black swans more frequently for longer time series and for populations with a smaller magnitude of process error, as we might expect.
 However, we failed to find strong evidence that black swans were associated with density dependence, population growth rate, or lifespan.
@@
 Our empirical results, sensitivity tests, and simulation tests suggest that estimating the tail shape of process error is a viable method of detecting heavy-tailed population dynamics.
-The presence of black swans highlights the importance of developing management strategies that detect quickly, respond to, and are robust to extremes in population dynamics --- particularly as the frequency and magnitude of climatic extremes increase over the next century.
+The presence of black swans highlights the importance of developing management strategies that detect quickly, respond to, and are robust to extremes in population dynamics --- particularly as the frequency and magnitude of climatic extremes increase over the next century (REF).
 
 Our results clarify, but do not disagree with previous related analyses.
-Previous work with an older version of the GPDD assessed heavy-tailed behaviour but focused on identifying if the lognormal was the most frequently parsimonious distribution of population abundance \citep{halley2002}.
+Previous work with an older version of the GPDD assessed heavy-tailed behaviour but focused on identifying if the lognormal distribution was the most frequently parsimonious model of population abundance \citep{halley2002}.
 For heavy-tailed distributions, \citet{halley2002} fit the extremely heavy-tailed Cauchy distribution and the four-parameter Levy stable distribution and found little information criteria support for these distributions in longer time series.
-However, black swan events are by definition rare, and the majority of time series we studied are short for these purposes.
+However, black swan events are by definition rare, and the majority of time series in the GPDD are short for these purposes.
 Therefore, we would not expect to observe black swans in a large proportion of populations.
@@
 However, this finding was challenged by \citet{allen2001}, who showed that mixing data across species could create falsely generate heavy tails.
-This work directly feeds into our use of the t distribution to represent heavy tails, since a t distribution can be represented by a mixture of normal distributions (with inverse-gamma-distributed variances).
+This rebutal supports our use of the t distribution to represent heavy tails, since a t distribution can be represented by a mixture of normal distributions \citep[with inverse-gamma-distributed variances,][]{gelman2014}.
 
 We might expect to observe black swan dynamics in ecological time series because of unmodelled intrinsic properties of populations or extrinsic forces acting on populations.
-A t distribution can be formed through a mixture of normal distributions \citep[in which the variances are inverse-gamma distributed,][]{gelman2014}.
-Therefore, if we miss some underlying mixture of processes we might expect to observe heavy tails \citep{allen2001}.
+Since a t-distribution can be formed by a mixture of normal distributions, we could observe heavy tails if we miss some underlying mixture of intrinsic processes \citep{allen2001}.
 That process might be an aggregation of populations across space, or population diversity, or some intrinsic change in population variability through time.
@@
 These forces could be extreme themselves.
-For example, extreme climate, predation from or competition with other species experiencing black swans, or sharp changes in human pressure such hunting, fishing, or habitat destruction might cause black swans.
-Alternatively, the interaction of multiple ``normal'' forces could give rise to black swan ecological dynamics.
+For example, extreme climate, predation from or competition with other species experiencing black swans, or sharp changes in human pressure such hunting, fishing, or habitat destruction might cause black swans (REFS).
+Alternatively, the interaction of multiple ``normal'' extrinsic forces could give rise to black swan ecological dynamics.
 This could occur if the interaction nature is synergistic \citep[e.g.][]{kirby2009} or even if non-synergistic interactions experience a rare alignment \citep{denny2009}.
 
-There a number of important caveats when considering the generality of our results.
+There a number of caveats when considering the generality of our results.
 First, the GPDD data represent a taxonomically and geographically biased sample of populations.
-The longer time series we focus on are dominated by commercially and recreationally important species and a disproportionate number of populations are located in the United Kingdom.
-Although we expect that we would find strong but rare evidence for black swans for other taxa around the world, the common forces driving those black swans (e.g.~severe winters in Table 1) likely differ.
+Also, the longer time series we focus on are dominated by commercially and recreationally important species and a disproportionate number of populations are located in the United Kingdom.
+Although we expect that we would find strong but rare evidence for black swans in other large taxonomic or geographic samples, the common forces driving those black swans (e.g.~severe winters in Table~1) likely differ.
 Second, in a large database with disparate datasources, we cannot ensure the quality of every datapoint.
@@
 Secondly, can we isolate the components of ecological dynamics that experience black swans events by moving from phenomenological models such as the Gompertz to mechanistic models, that take into account, for example, recruitment?
-Thirdly, the only way we may gain further insight into the mechanisms and covariates of ecological black swans may be to focus on specific geographic and taxonomic subsets of data where longer time series are available \citep[e.g.][]{segura2013}.
+We expect that greater insight into the mechanisms and covariates of ecological black swans may be best obtained through specific geographic and taxonomic subsets of data where longer time series are available \citep[e.g.][]{segura2013}.
 Finally, if a sizeable fraction of populations experience heavy tails in population process noise, what is the impact of allowing for black swans in forecasts of ecological risk?
-Risk can be defined as the probability of a specific magnitude of undesired event occurring (REF).
-Therefore, financial analysts are acutely concerned with the shape of distribution tails (REF).
-It is likely that a comparable focus in ecology would alter our perception of forecasts of, for example, extinction risk.
+Financial analysts are acutely concerned with the shape of the tail distribution of financial returns since these impact estimates of risk --- the probability of a specific magnitude of undesired event occurring (REF).
+It is possible that a comparable focus in ecology would alter our perception of ecological forecasts such as extinction risk.
 
@@
 We are grateful to the contributors and maintainers of the Global Population Dynamics Database and to Compute Canada's WestGrid high-performance computing resources.
-Three silhouette images were obtained from \texttt{phylopic.org}: rabbit (Sarah Werning), hoverfly (Gareth Monger), and bird (Jean-Raphaël Guillaumin {[}photography{]} and T.
-Michael Keesey {[}vectorization{]}) under a Creative Commons Attribution 3.0 Unported license.
-Funding was provided by a Simon Fraser University Graduate Fellowship (SCA), NSERC (NKD, ABC), the Canada Research Chairs Program (NKD), and TAB TODO?.
+Three silhouette images were obtained from \texttt{phylopic.org} under a Creative Commons Attribution 3.0 Unported license: rabbit (Sarah Werning), hoverfly (Gareth Monger), bird (Jean-Raphaël Guillaumin {[}photography{]} and T.
+Michael Keesey {[}vectorization{]}).
+Funding was provided by a Simon Fraser University Graduate Fellowship (SCA), the Natural Sciences and Engineering Research Council of Canada (NKD, ABC), the Canada Research Chairs Program (NKD), and XX-TODO (TAB).
 

### 2014-10-17 supp.tex
SOM tweaks

@@
 \end{enumerate}
-
+\noindent
 We provide a supplemental figure of all the time series included in our analysis and indicate which values were interpolated (\percImputedPops\% of populations had at least one point interpolated and only \percImputedPoints\% of the total observations were interpolated) (Fig.~\ref{fig:all-ts}).
@@
 We took stochastic draws from t distributions with different $\nu$ values ($\nu = 3, 5, 10,$ and $10^6$ [$\approx$ normal]), with central tendency parameters of $0$, and scale parameters of $1$.
-We started with 1600 stochastic draws and then fitted again at the first 800, 400, 200, 100, 50, and 25 draws.
+We started with 1600 stochastic draws and then fitted the models again at the first 800, 400, 200, 100, 50, and 25 draws.
 Each time we recorded the posterior samples of $\nu$.
@@
 The MCMC chains for a small number of these models (XX) did not converge according to our criteria ($\widehat{R} < 1.05, n_\mathrm{eff} > 200$) after 8000 iterations of four chains.
-We do not show these models in Fig.~\ref{fig:alt}.
+We did not include these models in Fig.~\ref{fig:alt}.
 

### 2014-10-17 anderson-etal-blackswan-timeseries.tex
Adding to the introduction and discussion;
Edits to the SOM too
Switch back to .tex files; add to SOM
Add a potential title

@@
 
-\title{Black swans in ecological time series\\Black swans in population dynamics\\Heavy-tailed ecological time series\\Black swans in ecological dynamics}
+\title{Black swans in ecological time series\\Black swans in population dynamics\\Heavy-tailed ecological time series\\Black swans in ecological dynamics\\Black swan ecological dynamics\\ \ldots}
 \author{
@@
 
-\setlength\parskip{0.16in}
-\setlength\parindent{0in}
+%\setlength\parskip{0.19in}
+%\setlength\parindent{0in}
 \usepackage{titlesec}
-\titlespacing\section{0pt}{6pt plus 4pt minus 2pt}{-6pt plus 2pt minus 2pt}
-\titlespacing\subsection{0pt}{0pt plus 4pt minus 2pt}{-8pt plus 2pt minus 2pt}
+%\titlespacing\section{0pt}{6pt plus 4pt minus 2pt}{-6pt plus 2pt minus 2pt}
+%\titlespacing\subsection{0pt}{0pt plus 4pt minus 2pt}{-8pt plus 2pt minus 2pt}
 
@@
 \maketitle
+\doublespacing
+\onehalfspacing
 
@@
 
+\noindent
 \emph{Authors are listed alphabetically (ABCD!) and open to reordering.}
 
+\noindent
 \textsuperscript{1}Earth to ocean research group, department of biological sciences, Simon Fraser University, Burnaby BC, V5A 1S6, Canada
 
+\noindent
 \textsuperscript{2}School of Aquatic and Fishery Sciences, University of Washington, Box 355020, Seattle, WA 98195, USA
 
+\noindent
 \textsuperscript{3}School of Resource and Environmental Management, Simon Fraser University, Burnaby, BC, V5A 1S6, Canada
 
+\noindent
 \textsuperscript{*}Corresponding author: Sean C.
@@
 %\begin{spacing}{1.4}
-\doublespacing
+
 %\onehalfspacing
@@
 \include{ms}
-
-\clearpage
-\section{Figures}
-
-\input{figs}
 

### 2014-10-17 supp.tex
Add to SOM

@@
 
-Our results show...
+Our results show that the more-complicated Gompertz model can still recapture the true value of $\nu$ when the process noise was chosen so that appropriate tail events were present (Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}, red and green symbols in the top rows).
+Likewise, the other Gompertz parameters were estimated without any systematic bias (Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}, red and green symbols).
+The addition of observation error caused the model to tend to underestimate the degree of heavy-tailedness, overestimate the magnitude of process noise, somewhat overestimate $\lambda$, and overestimate density dependence (blue symbols in Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}).
+The overestimation of density dependence with observation error is a known phenomenon \citep{knape2012}.
+Fitting a model with correctly specified observation error made marginal improvements to model bias (purple symbols in Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}).
 
-Fig.~\ref{fig:sim-gompertz}
-Fig.~\ref{fig:sim-gompertz-boxplots}
-Fig.~\ref{fig:sim-prob}
+When converting the posterior distributions of $\nu$ into Pr($\nu < 10$) the models performed well.
+Without observation error, and using a probability of $0.5$ as a threshold, the model would have miscategorized only one of $40$ simulations at $\nu = 5$ across a sample size of $50$ or $100$ data points (Fig.~\ref{fig:sim-prob}, upper panels).
+The model would have correctly categorized all cases where the process noise was not heavy tailed (indicated as ``$\nu =$ infinity'' in Fig.~\ref{fig:sim-prob}) and all cases where $\nu = 3$.
+With $0.2$ standard deviations of observation error, the model still categorized XX\% of cases as heavy tailed when $\nu = 5$ and all cases where $\nu = 3$.
+Allowing for observation error made little improvement to the detection of heavy tails (Fig.~\ref{fig:sim-prob}, lower-left vs.\ lower-right panel).
+Therefore, we chose to focus on the simpler model without observation error in the main text.
+
+%Fig.~\ref{fig:sim-gompertz}
+%Fig.~\ref{fig:sim-gompertz-boxplots}
+%Fig.~\ref{fig:sim-prob}
 

### 2014-10-17 ms.tex
Spell check and clean up TeX

@@
 
-Our main analysis focusses on the commonly applied phenomenological Gompertz population dynamics model.
+Our main analysis focuses on the commonly applied phenomenological Gompertz population dynamics model.
 In previous analyses of the GPDD, the Gompertz has proved to be the most frequently parsimonious model \citep{brook2006}.
@@
 However, this finding was challenged by \citet{allen2001}, who showed that mixing data across species could create falsely generate heavy tails.
-This rebutal supports our use of the t distribution to represent heavy tails, since a t distribution can be represented by a mixture of normal distributions \citep[with inverse-gamma-distributed variances,][]{gelman2014}.
+This rebuttal supports our use of the t distribution to represent heavy tails, since a t distribution can be represented by a mixture of normal distributions \citep[with inverse-gamma-distributed variances,][]{gelman2014}.
 
@@
 (c--e) Simulated population dynamics from a Gompertz model with process noise drawn from t distributions with different values of $\nu$.
-Coloured dots in panels c and d represent jumpts with less than a 1 in 1000 chance of occurring in a normal distribution.
+Coloured dots in panels c and d represent jumps with less than a 1 in 1000 chance of occurring in a normal distribution.
 (f--h) Estimates of $\nu$ from models fit to the times series in panels c--e.

### 2014-10-17 supp.tex
Spell check and clean up TeX

@@
 \LARGE
-Supporting Material for TODO INSERT TITLE HERE\\[1.5em]
+Supporting Material\\[1.0em]
 \end{centering}
@@
 
-We followed the following data selection and quality-control rules with the GPDD:
+We applied the following data selection and quality-control rules to the Global Population Dynamics Database (GPDD):
 
@@
 
-Our prior on $\nu$ is based on \citet{fernandez1998}; they chose a more informative exponential rate parameter of 0.1.
-We chose a less informative rate parameter of 0.01 and truncated the distribution at two, since at $\nu < 2$ the variance of the t distribution is undefined.
-This prior gives only a 7.7\% probability $\nu < 10$ but constrains the sampling sufficiently to avoid wandering off towards infinity --- above approximately $\nu = 20$ the t distribution is so similar to the normal distribution (Fig.~\ref{fig:didactic}) that time series of the length considered here are unlikely to be sufficiently informative about the precise value of $\nu$.
-In the scenario where the data are not heavy tailed (e.g.~Fig.~\ref{fig:didactic}e, h) the posterior will approximately match the prior (median $= 71$, mean $= 102$) and not be flagged as likely heavy tailed using the metrics we used in our paper (e.g.~Pr($\nu < 10$) \textgreater{} 0.5).
+Our prior on $\nu$ is based on \citet{fernandez1998}; they chose a more informative exponential rate parameter of $0.1$.
+We chose a less informative rate parameter of $0.01$ and truncated the distribution at $2$, since at $\nu < 2$ the variance of the t distribution is undefined.
+This prior gives only a $7.7$\% probability $\nu < 10$ but constrains the sampling sufficiently to avoid wandering off towards infinity --- above approximately $\nu = 20$ the t distribution is so similar to the normal distribution (Fig.~\ref{fig:didactic}) that time series of the length considered here are unlikely to be sufficiently informative about the precise value of $\nu$.
+In the scenario where the data are not heavy tailed (e.g.~Fig.~\ref{fig:didactic}e, h) the posterior will approximately match the prior (median $= 71$, mean $= 102$) and not be flagged as likely heavy tailed using the metrics we used in our paper (e.g.~Pr$(\nu < 10) > 0.5$).
 
 We fitted our models with Stan 2.4.0 \citep{stan-manual2014}, and R 3.1.1 \citep{r2014}.
-We began with four chains and 2000 iterations, discarding the first 1000 as warm up (i.e.~4000 total samples).
-If $\hat{R}$ (a measure of chain convergence) was greater than 1.05 for any parameter or the minimum effective sample size, $n_\mathrm{eff}$, (a measure of the effective number of uncorrelated samples) for any parameter was less than 200, we doubled both the total iterations and warm up period and sampled from the model again.
+We began with four chains and $2000$ iterations, discarding the first $1000$ as warm up (i.e.~4000 total samples).
+If $\hat{R}$ (a measure of chain convergence) was greater than $1.05$ for any parameter or the minimum effective sample size, $n_\mathrm{eff}$, (a measure of the effective number of uncorrelated samples) for any parameter was less than $200$, we doubled both the total iterations and warm up period and sampled from the model again.
 These thresholds are in excess of the minimums recommended by \citet{gelman2006a} of $\hat{R} < 1.1$ and effective sample size $> 100$ for reliable point estimates and confidence intervals.
 In the majority of cases our minimum thresholds were greatly exceeded.
-We continued this procedure up to 8000 iterations (16000 total samples) by which all chains were deemed to have sufficiently converged.
+We continued this procedure up to $8000$ iterations ($16000$ total samples) by which all chains were deemed to have sufficiently converged.
 
@@
 We took stochastic draws from t distributions with different $\nu$ values ($\nu = 3, 5, 10,$ and $10^6$ [$\approx$ normal]), with central tendency parameters of $0$, and scale parameters of $1$.
-We started with 1600 stochastic draws and then fitted the models again at the first 800, 400, 200, 100, 50, and 25 draws.
+We started with $1600$ stochastic draws and then fitted the models again at the first $800, 400, 200, 100, 50,$ and $25$ draws.
 Each time we recorded the posterior samples of $\nu$.
@@
 
-%2 parts: how many samples from the true population t distribution do you need to detect low nu?
-And, given that you have a set of deviations in which nu is detectable (effective nu is within 0.5 CV of true nu), can the more complex Gompertz still capture this?
-
 In the second part of our simulation testing, we tested the ability of the heavy-tailed Gompertz model to obtain unbiased parameter estimates given that process deviations were provided in which our simpler model (in the previous section) could accurately estimate the value of $\nu$.
 To generate these process errors, we repeatedly generated candidate process deviations and estimated the central tendency, scale, and $\nu$ values each time.
-We recorded when the estimated nu was within $0.2$ CVs (coefficient of variations) of the true $\nu$ value and used this set of random seed values in our Gompertz simulation.
+We recorded when $\hat{\nu}$ was within $0.2$ CVs (coefficient of variations) of the true $\nu$ value and used this set of random seed values in our Gompertz simulation.
 We then fitted our Gompertz models to the simulated datasets with all parameters (except $\nu$) set near the median values estimated in the GPDD.
-We repeated this with $50$ and $100$ samples without observation error, $50$ samples with observation error ($\sigma_\mathrm{obs} = 0.2$), and $50$ samples with the same observation error and a Gompertz model that allowed for (correctly) specified observation error.
+We repeated this with $50$ and $100$ samples without observation error, $50$ samples with observation error ($\sigma_\mathrm{obs} = 0.2$), and $50$ samples with the same observation error and a Gompertz model that allowed for correctly specified observation error magnitude.
 
@@
 Therefore, we chose to focus on the simpler model without observation error in the main text.
-
-%Fig.~\ref{fig:sim-gompertz}
-%Fig.~\ref{fig:sim-gompertz-boxplots}
-%Fig.~\ref{fig:sim-prob}
-
-%Simulation testing: the ability to recover $\nu$ when randomly sampling from various distributions (Fig.~\ref{fig:sim-nu}); heavy-tailed Gompertz model performance and confidence interval coverage given the process noise has deviations that are known to have effective $\nu$ equal to true $\nu$ (Fig.~\ref{fig:sim-gompertz}); boxplots of the same output (Fig.~\ref{fig:sim-gompertz-boxplots}); probability that $\nu < 10$ for the Gompertz simulation testing (Fig.~\ref{fig:sim-prob}).
 

### 2014-10-17 supp.tex
Trivial edits

@@
 \item
-  We then wrote an algorithm to find the longest unbroken window of abundance (no \texttt{NA}s) with at least 20 time steps in each population time series.
+  We then wrote an algorithm to find the longest unbroken window of abundance (no \texttt{NA}s) with at least $20$ time steps in each population time series.
 If there were any populations with multiple windows of identical length, we took the most recent window.

### 2014-10-18 anderson-etal-blackswan-timeseries.tex
Clean up main .tex file
Wrap ms and (very) minor edits

--- 2014-10-09 anderson-etal-blackswan-timeseries.tex
@@
 \usepackage{geometry}
-\geometry{verbose,letterpaper,tmargin=2.54cm,bmargin=2.54cm,lmargin=2.54cm,rmargin=2.54cm}
+\geometry{verbose, letterpaper, tmargin = 2.54cm, bmargin = 2.54cm,
+  lmargin = 2.54cm, rmargin = 2.54cm}
 \geometry{letterpaper}
@@
 \usepackage[sectionbib]{chapterbib}
-
 \usepackage{amsmath}
-% \makeatletter
-% \g@addto@macro\normalsize{%
-%   \setlength\abovedisplayskip{-20pt}
-%   \setlength\belowdisplayskip{0pt}
-%   \setlength\abovedisplayshortskip{-20pt}
-%   \setlength\belowdisplayshortskip{0pt}
-% }
-% \makeatother
-
 \usepackage{longtable}
@@
 
-\title{Black swans in ecological time series\\Black swans in population dynamics\\Heavy-tailed ecological time series\\Black swans in ecological dynamics}
+\title{Black swans in ecological time series\\
+  Black swans in population dynamics\\
+  Heavy-tailed ecological time series\\
+  Black swans in ecological dynamics\\
+  Black swan ecological dynamics\\
+  \ldots}
 \author{
@@
 
-\setlength\parskip{0.16in}
-\setlength\parindent{0in}
-\usepackage{titlesec}
-\titlespacing\section{0pt}{6pt plus 4pt minus 2pt}{-6pt plus 2pt minus 2pt}
-\titlespacing\subsection{0pt}{0pt plus 4pt minus 2pt}{-8pt plus 2pt minus 2pt}
-
 \begin{document}
@@
 \maketitle
+\doublespacing
+%\onehalfspacing
 
@@
 
+\noindent
 \emph{Authors are listed alphabetically (ABCD!) and open to reordering.}
 
-\textsuperscript{1}Earth to ocean research group, department of biological sciences, Simon Fraser University, Burnaby BC, V5A 1S6, Canada
+\noindent
+\textsuperscript{1}Earth to ocean research group,
+Department of biological sciences,
+Simon Fraser University, Burnaby BC, V5A 1S6, Canada
 
-\textsuperscript{2}School of Aquatic and Fishery Sciences, University of Washington, Box 355020, Seattle, WA 98195, USA
+\noindent
+\textsuperscript{2}School of Aquatic and Fishery Sciences,
+University of Washington, Box 355020, Seattle, WA 98195, USA
 
-\textsuperscript{3}School of Resource and Environmental Management, Simon Fraser University, Burnaby, BC, V5A 1S6, Canada
+\noindent
+\textsuperscript{3}School of Resource and Environmental Management,
+Simon Fraser University, Burnaby, BC, V5A 1S6, Canada
 
-\textsuperscript{*}Corresponding author: Sean C.
-Anderson; Earth to Ocean Research Group, Department of Biological Sciences, Simon Fraser University, Burnaby BC, V5A 1S6; Phone: 1-778-782-3989; E-mail: sean\_anderson@sfu.ca
-
-%\begin{spacing}{1.4}
-\doublespacing
-%\onehalfspacing
+\noindent
+\textsuperscript{*}Corresponding author: Sean C. Anderson;
+Earth to Ocean Research Group,
+Department of Biological Sciences,
+Simon Fraser University,
+Burnaby BC, V5A 1S6;
+Phone: 1-778-782-3989;
+E-mail: sean\_anderson@sfu.ca
 

### 2014-10-18 ms.tex
Wrap ms and (very) minor edits

@@
 \noindent
-\emph{Main message, to be deleted (25 words)}: Black swans are present \& taxonomically widespread but rare in population dynamics.
-Extreme climate, predation, parasites, and their interactions are common causes; intrinsic drivers are unclear.
+\emph{Main message, to be deleted (25 words)}: Black swans are present \&
+taxonomically widespread but rare in population dynamics. Extreme climate,
+predation, parasites, and their interactions are common causes; intrinsic
+drivers are unclear.
 
@@
 
-Black swans are statistically improbable events that nonetheless occur --- often with profound implications.
-While extremes in the physical environment, such as monsoons and heat waves, are widely studied and increasing in magnitude and frequency, it remains unclear the extent to which ecological populations buffer or suffer from extremes.
-Here, we develop a probability model to estimate the degree of heavy-tailedness (presence of black swans) in ecological process noise.
-We apply our model to \NPops~time series from around the world across \NOrders~taxonomic orders and seven classes.
-We find strong evidence of black swans, but they are rare, occurring in \overallMinPerc--\overallMaxPerc\% of populations; most frequently for birds (\AvesRangePerc\%) followed by mammals (\MammaliaRangePerc\%), insects (\InsectaRangePerc\%), and fishes (\OsteichthyesRangePerc\%).
-When they occur, they tend to be driven by climate and severe winters, cycles of parasites and predators, and interactions between these elements.
-They are more frequently detected for populations with longer time series and lower levels of process noise.
-We find little evidence of intrinsic life-history covariates.
-\emph{close with modelling and policy implications}
+Black swans are statistically improbable events that nonetheless occur --- often
+with profound implications. While extremes in the physical environment,such as
+monsoons and heat waves, are widely studied and increasing in magnitude and
+frequency, it remains unclear the extent to which ecological populations buffer
+or suffer from extremes. Here, we develop a probability model to estimate the
+degree of heavy-tailedness (presence of black swans) in ecological process
+noise. We apply our model to \NPops~time series from around the world across
+\NOrders~taxonomic orders and seven classes. We find strong evidence of black
+swans, but they are rare, occurring in \overallMinPerc--\overallMaxPerc\%of
+populations; most frequently for birds (\AvesRangePerc\%) followed by mammals
+(\MammaliaRangePerc\%), insects (\InsectaRangePerc\%), and
+fishes(\OsteichthyesRangePerc\%). When they occur, they tend to be driven by
+climate and severe winters, cycles of parasites and predators, and interactions
+between these elements. They are more frequently detected for populations with
+longer time series and lower levels of process noise. We find little evidence
+of intrinsic life-history covariates. TODO close with modelling and policy
+implications
 
@@
 
-Black swans are unexpected extreme events with dramatic consequences.
-One of the most striking black swans in ecology was the asteroid marking the mass extinctions of the K-T boundary (REF).
-Today, it is the extremeness of climate --- particularly in concert with shifts in mean temperature --- that is expected to cause the greatest societal damage (REF).
+Black swans are unexpected extreme events with dramatic consequences. One of
+the most striking black swans in ecology was the asteroid marking the mass
+extinctions of the K-T boundary (REF). Today, it is the extremeness of climate
+--- particularly in concert with shifts in mean temperature --- that is
+expected to cause the greatest societal damage (REF).
 
-However, while extremes in the physical environment are present and widely accepted, it remains unclear the extent to which ecological systems buffer or suffer from black swans.
+However, while extremes in the physical environment are present and widely
+accepted, it remains unclear the extent to which ecological systems buffer or
+suffer from black swans.
 
-Existing evidence for black swans or heavy tails in population dynamics is thus far limited.
-Considerable work has focussed on heavy tails in physical environmental processes in, for example, rain fall (REF), wave height (REF), earth quake magnitude, climate temperature (REF).
-time to extinction, plankton time series, GPDD work found little evidence
-subsequent work suggested, but...
+Existing evidence for black swans or heavy tails in population dynamics is thus
+far limited. Considerable work has focussed on heavy tails in physical
+environmental processes in, for example, rain fall (REF), wave height (REF),
+earth quake magnitude, climate temperature (REF). time to extinction, plankton
+time series, GPDD work found little evidence subsequent work suggested, but...
 Ward applied a data-intense method and found evidence for...
 
-There are two key reasons why we may find little evidence of ecological black swans.
-First, they might not exist.
-Indeed, the vast majority of population dynamics model fitting and risk forecasting assumes that population dynamics are not heavy tailed (e.g.~REF, REF, REF).
-Alternatively, population dynamics might be heavy tailed, but a paucity of statistical methods and shortness of available time series has limited our ability to detect them.
-For example, a commonly applied probability distribution to detect the shape of time series tails (the generalized extreme value distribution) requires lumping data into the most extreme event per unit time (e.g.~heaviest rainfall per year) and therefore requires a considerable quantity of data (REF).
-Previous work into the shape of population dynamic tails, has focused on model comparison to identify parsimonious explanations from a set of particular hypotheses, but does not quantify the magnitude of heavy tails on a continuous scale or put the conclusions in a probabilistic framework (REF).
-Furthermore, there is evidence that information criteria are poor at distinguishing subtle differences in population dynamics for the relatively short time series that are most readily available (REF).
+There are two key reasons why we may find little evidence of ecological black
+swans. First, they might not exist. Indeed, the majority of population
+dynamics model fitting and risk forecasting assumes that population dynamics
+are not heavy tailed (e.g.~REF, REF, REF). Alternatively, population dynamics
+might be heavy tailed, but a paucity of statistical methods and shortness of
+available time series has limited our ability to detect them. For example,
+a commonly applied probability distribution to detect the shape of time series
+tails (the generalized extreme value distribution) requires lumping data into
+the most extreme event per unit time (e.g.~heaviest rainfall per year) and
+therefore requires a considerable quantity of data (REF). Previous work into
+the shape of population dynamic tails, has focused on model comparison to
+identify parsimonious explanations from a set of particular hypotheses, but
+does not quantify the magnitude of heavy tails on a continuous scale or put the
+conclusions in a probabilistic framework (REF). Furthermore, information
+criteria are poor at distinguishing subtle differences in population dynamics
+for the relatively short time series that are most readily available (REF CHANGE
+THIS?).
 
-Or bring back in Ward - data intensive, can't be applied to most individual time series
+Or bring back in Ward - data intensive, can't be applied to most individual
+time series
 
-plankton --- hideous distribution to fit, even harder in population dynamics framework -and what about seasonal cycles, density dependence?
+plankton --- hideous distribution to fit, even harder in population dynamics
+framework -and what about seasonal cycles, density dependence?
 
-GPDD - model competition and finding most parsimonious explanation from set of hypotheses --- but only \emph{very} heavy from not heavy
+GPDD - model competition and finding most parsimonious explanation from set of
+hypotheses --- but only \emph{very} heavy from not heavy
 
@@
 
-tipping point and regime shift methods --- say something\ldots{} not appropriate for most of the data available
+tipping point and regime shift methods --- say something\ldots{} not
+appropriate for most of the data available
 
-Here we develop a probabilist framework for identifying ecological black swans in population dynamic process noise --- the stochastic jumps from time step to time step.
-Our framework quantifies both the magnitude and probability of heavy-tailed dynamics, allows for a range of population dynamic models, can incorporate observation uncertainty, and is applicable to commonly available lengths of abundance time series.
-We apply our model to hundreds of populations from around the world across four taxonomic classes to address three questions: (1) how frequent and strong are black swans in population dynamics across taxonomy, (2) what are some verified causes of black swans in ecological time series, and (3) are there characteristics of time series or intrinsic life-history characteristics associated with heavy-tailed dynamics.
-We find a high probability of ecological black swans across a wide range of taxa but in a limited number of populations.
-When they occurred, extreme climate, predation, parasites, and their interactions were common documented causes.
-We found little evidence that intrinsic life-history characteristics were related to the probability of black swans.
+Here we develop a probabilist framework for identifying ecological black swans
+in population dynamic process noise --- the stochastic jumps from time step to
+time step. Our framework quantifies both the magnitude and probability of
+heavy-tailed dynamics, allows for a range of population dynamic models, can
+incorporate observation uncertainty, and is applicable to commonly available
+lengths of abundance time series. We apply our model to hundreds of populations
+from around the world across four taxonomic classes to address three questions:
+(1) how frequent and strong are black swans in population dynamics across
+taxonomy, (2) what are some verified causes of black swans in ecological time
+series, and (3) are there characteristics of time series or intrinsic
+life-history characteristics associated with heavy-tailed dynamics. We find
+a high probability of ecological black swans across a wide range of taxa but in
+a limited number of populations. When they occurred, extreme climate,
+predation, parasites, and their interactions were common documented causes. We
+found little evidence that intrinsic life-history characteristics were related
+to the probability of black swans.
 
-Papers to work in: \citep{inchausti2002, halley2002, inchausti2001} \citep{jentsch2007} \citep{ward2007} \citep{garcia-carreras2011} \citep{sornette2009} \citep{nunez2012} \citep{thompson2013} \citep{beaugrand2012} \citep{pine-iii2009} \citep{doak2008} Extreme rain and monsoons over India increasing \citep{goswami2006} \citep{smale2013} \citep{easterling2000} \citep{scheffer2003} \citep{katz2005} \citep{taleb2007} \citep{vasseur2014} \citep{vert-pre2013} \citep{lindenmayer2010} \citep{valpine2002} \citep{gregory2010} \citet{garcia-carreras2011} \citeauthor{brook2006} \citetext{\citeyear{brook2006}; \citealp{herrandoprez2014}} \citep{sibly2005, ziebarth2010}
+Papers to work in: \citep{inchausti2002, halley2002, inchausti2001}
+\citep{jentsch2007} \citep{ward2007} \citep{garcia-carreras2011}
+\citep{sornette2009} \citep{nunez2012} \citep{thompson2013}
+\citep{beaugrand2012} \citep{pine-iii2009} \citep{doak2008} Extreme rain and
+monsoons over India increasing \citep{goswami2006} \citep{smale2013}
+\citep{easterling2000} \citep{scheffer2003} \citep{katz2005} \citep{taleb2007}
+\citep{vasseur2014} \citep{vert-pre2013} \citep{lindenmayer2010}
+\citep{valpine2002} \citep{gregory2010} \citet{garcia-carreras2011}
+\citeauthor{brook2006} \citetext{\citeyear{brook2006};
+  \citealp{herrandoprez2014}} \citep{sibly2005, ziebarth2010}
 
@@
 
-To obtain estimates of the probability and magnitude of black swans, we fitted population dynamic models to hundreds of abundance time series from around the world.
-The key addition of our models, compared to typical population models, was estimating the shape of the process noise tails.
-We then compared how these estimates related to taxonomy, time series properties, and life-history characteristics, and for a subset of populations, we documented the established causes of ecological black swans.
+To obtain estimates of the probability and magnitude of black swans, we fitted
+population dynamic models to hundreds of abundance time series from around the
+world. The key addition of our models, compared to typical population models,
+was estimating the shape of the process noise tails. We then compared how these
+estimates related to taxonomy, time series properties, and life-history
+characteristics, and for a subset of populations, we documented the established
+causes of ecological black swans.
 
@@
 
-We obtained abundance time series from the Global Population Dynamics Database (GPDD) \citep{gpdd2010}.
-The GPDD contains nearly 5000 time series of abundance across $\sim$1000 species and $\sim$100 taxonomic orders.
-To derive a high-quality subset of populations that were suitable for our analyses, we filtered the data (Supporting Material).
-Our filtering removed populations from less reliable data sources, removed those without sufficient data for our models, and interpolated some missing values.
-Our final dataset contained \NPops~populations across \NOrders~taxonomic orders and seven taxonomic classes with a median of XX time steps (range of XX--XX) (Table S1, Fig.~\ref{fig:all-ts}).
+We obtained abundance time series from the Global Population Dynamics Database
+(GPDD) \citep{gpdd2010}. The GPDD contains nearly 5000 time series of abundance
+across $\sim$1000 species and $\sim$100 taxonomic orders. To derive
+a high-quality subset of populations that were suitable for our analyses, we
+filtered the data (Supporting Material). Our filtering removed populations from
+less reliable data sources, removed those without sufficient data for our
+models, and interpolated some missing values. Our final dataset contained
+\NPops~populations across \NOrders~taxonomic orders and seven taxonomic classes
+with a median of XX time steps (range of XX--XX) (Table S1,
+Fig.~\ref{fig:all-ts}).
 
@@
 
-Our main analysis focuses on the commonly applied phenomenological Gompertz population dynamics model.
-In previous analyses of the GPDD, the Gompertz has proved to be the most frequently parsimonious model \citep{brook2006}.
-We extended the traditional Gompertz model by allowing the process noise to be drawn from a Student-t distribution with degrees of freedom parameter $\nu$ (Fig.~\ref{fig:didactic}).
-If $\nu$ is small ($\lesssim 10$) the t distribution has much heavier tails than a normal distribution.
-For example, at $\nu = 2$, the probability of drawing a value more than five standard deviations below the mean is $1.8$\%, whereas the probability of drawing such a value from a normal distribution is nearly zero ($2.9\cdot10^{-5}$\%).
-As $\nu$ approaches infinity the distribution approaches the normal distribution.
-By estimating the value of $\nu$ we can quantity how heavy-tailed the process noise deviations are (Fig.~\ref{fig:didactic}).
+Our main analysis focuses on the commonly applied phenomenological Gompertz
+population dynamics model. In previous analyses of the GPDD, the Gompertz has
+proved to be the most frequently parsimonious model \citep{brook2006}. We
+extended the traditional Gompertz model by allowing the process noise to be
+drawn from a Student-t distribution with degrees of freedom parameter $\nu$
+(Fig.~\ref{fig:didactic}). If $\nu$ is small ($\lesssim 10$) the t distribution
+has much heavier tails than a normal distribution. For example, at $\nu = 2$,
+the probability of drawing a value more than five standard deviations below the
+mean is $1.8$\%, whereas the probability of drawing such a value from a normal
+distribution is nearly zero ($2.9\cdot10^{-5}$\%). As $\nu$ approaches infinity
+the distribution approaches the normal distribution. By estimating the value of
+$\nu$ we can quantity how heavy-tailed the process noise deviations are
+(Fig.~\ref{fig:didactic}).
 
@@
 \end{align*}
-where $x_t$ is the $\ln$ abundance ($N$) at time $t$.
-The model is density independent if $b = 1$, maximally density dependent if $b = 0$, and inversely density dependent if $b < 0$.
-The parameter $\lambda$ represents the expected $\ln$ number of new individuals at time $t + 1$ if $N_t = 1$ (per capita population growth rate).
-The process noise is modelled as a Student-t distribution with degrees of freedom parameter $\nu$, a mean of $0$, and a scale parameter of $\sigma$.
+where $x_t$ is the $\ln$ abundance ($N$) at time $t$. The model is density
+independent if $b = 1$, maximally density dependent if $b = 0$, and inversely
+density dependent if $b < 0$. The parameter $\lambda$ represents the expected
+$\ln$ number of new individuals at time $t + 1$ if $N_t = 1$ (per capita
+population growth rate). The process noise is modelled as a Student-t
+distribution with degrees of freedom parameter $\nu$, a mean of $0$, and
+a scale parameter of $\sigma$.
 
-We chose weakly informative priors to incorporate our understanding of plausible population dynamics \citep[Supporting Material]{gelman2014}.
-For $\nu$, we chose an exponential prior with rate parameter of $0.01$ truncated at values above two (Fig.~\ref{fig:priors}), a slightly less informative prior than that suggested by \citet{fernandez1998}.
-This prior gives only a 7.7\% probability that $\nu < 10$ but constrains the sampling sufficiently to avoid wandering off towards infinity --- above approximately $\nu = 20$ the t distribution is so similar to the normal distribution that time series of the length considered here are unlikely to be informative about the precise value of $\nu$ (Fig.~\ref{fig:didactic}).
+We chose weakly informative priors to incorporate our understanding of
+plausible population dynamics \citep[Supporting Material]{gelman2014}. For
+$\nu$, we chose an exponential prior with rate parameter of $0.01$ truncated at
+values above two (Fig.~\ref{fig:priors}), a slightly less informative prior
+than that suggested by \citet{fernandez1998}. This prior gives only a 7.7\%
+probability that $\nu < 10$ but constrains the sampling sufficiently to avoid
+wandering off towards infinity --- above approximately $\nu = 20$ the
+t distribution is so similar to the normal distribution that time series of the
+length considered here are unlikely to be informative about the precise value
+of $\nu$ (Fig.~\ref{fig:didactic}).
 
-We estimated all models in a Bayesian framework using Stan \citep{stan-manual2014}.
-Stan samples from the posterior distribution with a version of Hamiltonian Markov chain Monte Carlo called the No-U-Turn Sampler \citep{hoffman2014}.
-We assured that chains had sufficiently converged and the sampler had obtained sufficient independent samples from the posterior ($\widehat{R} < 1.05$, $n_\mathrm{eff} > 200$; Supporting Material).
-We used simulated data to test how easily we could detect $\nu$ given different sample sizes and to ensure we could recover unbiased parameter estimates from the Gompertz model (Supporting Material, Fig.~\ref{fig:sim-nu}, \ref{fig:sim-gompertz}, \ref{fig:sim-gompertz-boxplots}).
+We estimated all models in a Bayesian framework using Stan
+\citep{stan-manual2014}. Stan samples from the posterior distribution with
+a version of Hamiltonian Markov chain Monte Carlo called the No-U-Turn Sampler
+\citep{hoffman2014}. We assured that chains had sufficiently converged and the
+sampler had obtained sufficient independent samples from the posterior
+($\widehat{R} < 1.05$, $n_\mathrm{eff} > 200$; Supporting Material). We used
+simulated data to test how easily we could detect $\nu$ given different sample
+sizes and to ensure we could recover unbiased parameter estimates from the
+Gompertz model (Supporting Material, Fig.~\ref{fig:sim-nu},
+\ref{fig:sim-gompertz}, \ref{fig:sim-gompertz-boxplots}).
 
-We fit four alternative population models to test if they systematically changed our conclusions.
-(1) Autocorrelation has been suggested as a reason for increased observed variability of abundance time series through time, which could create apparent heavy tails \citep{inchausti2002}; therefore, we fit a model that modelled the serial correlation in the residuals.
-(2) Previous work has focussed modelled growth rates without accounting for density dependence \citep{segura2013}; therefore, we fit a simpler model in which we assumed density independence.
-(3) Observation error could bias parameter estimates \citep{knape2012} or mask our ability to detect heavy tails \citep{ward2007}.
-However, simultaneously estimating observation error and process noise in a state space framework is a challenging computational problem \citep[e.g.][]{dennis2006, knape2008} even with longer time series and without estimating the shape of the process error tails; therefore, we fit a model where we assumed a level of observation error ($0.2$ standard deviations on a log scale).
-(4) The Gompertz model assumes that population growth rate declines linearly with log abundance.
-We also fit an alternative model, the Ricker-logistic model, which assumes that population growth rate declines linearly with abundance itself (Supporting Material).
+We fit four alternative population models to test if they systematically
+changed our conclusions. (1) Autocorrelation has been suggested as a reason for
+increased observed variability of abundance time series through time, which
+could create apparent heavy tails \citep{inchausti2002}; therefore, we fit
+a model that modelled the serial correlation in the residuals. (2) Previous
+work has focussed modelled growth rates without accounting for density
+dependence \citep{segura2013}; therefore, we fit a simpler model in which we
+assumed density independence. (3) Observation error could bias parameter
+estimates \citep{knape2012} or mask our ability to detect heavy tails
+\citep{ward2007}. However, simultaneously estimating observation error and
+process noise in a state space framework is a challenging computational problem
+\citep[e.g.][]{dennis2006, knape2008} even with longer time series and without
+estimating the shape of the process error tails; therefore, we fit a model
+where we assumed a level of observation error ($0.2$ standard deviations on
+a log scale). (4) The Gompertz model assumes that population growth rate
+declines linearly with log abundance. We also fit an alternative model, the
+Ricker-logistic model, which assumes that population growth rate declines
+linearly with abundance itself (Supporting Material).
 
@@
 
-We investigated possible covariates of heavy-tailed population dynamics visually and through multilevel modelling.
-We plotted characteristics of the time series ($\sigma$, $\lambda$, $b$, and time-series length) along with two life-history characteristics (body length and maximum lifespan obtained from \citet{brook2006a}) against our estimates of $\nu$.
-We formally investigated these relationships by fitting beta regression multilevel models to the probability that $\nu$ was less than $10$ (i.e.~approximately the probability of heavy tails).
-For covariates that were derived from Gompertz model parameter estimates, we incorporated standard deviations around the means.
-To account for broad patterns of phylogenetic relatedness, we allowed for varying intercepts at the taxonomic class, order, and species level (Supporting Material).
+We investigated possible covariates of heavy-tailed population dynamics
+visually and through multilevel modelling. We plotted characteristics of the
+time series ($\sigma$, $\lambda$, $b$, and time-series length) along with two
+life-history characteristics (body length and maximum lifespan obtained from
+\citet{brook2006a}) against our estimates of $\nu$. We formally investigated
+these relationships by fitting beta regression multilevel models to the
+probability that $\nu$ was less than $10$ (i.e.~approximately the probability
+of heavy tails). For covariates that were derived from Gompertz model parameter
+estimates, we incorporated standard deviations around the means. To account for
+broad patterns of phylogenetic relatedness, we allowed for varying intercepts
+at the taxonomic class, order, and species level (Supporting Material).
 
-Finally, we investigated a sample of populations that our method categorized as having a high probability of heavy tails (Pr$(\nu < 10) > 0.5$).
-Where possible, we found the documented causes of ecological black swans in the primary data source cited in the GPDD or in other literature describing the population.
-These populations were chosen haphazardly with the purpose of generating hypotheses about the causes of population dynamic black swans.
+Finally, we investigated a sample of populations that our method categorized as
+having a high probability of heavy tails (Pr$(\nu < 10) > 0.5$). Where
+possible, we found the documented causes of ecological black swans in the
+primary data source cited in the GPDD or in other literature describing the
+population. These populations were chosen haphazardly with the purpose of
+generating hypotheses about the causes of population dynamic black swans.
 
@@
 
-We found strong but rare evidence for black-swan dynamics.
-Defining black-swan dynamics as a greater than $0.5$ probability that $\nu < 10$, we observed black swans most frequently for birds (X\%) followed by mammals (X\%), insects (X\%), and fishes (X\%) (Fig.~\ref{fig:nu-coefs}, \ref{fig:heavy-ts}).
-Black swans were taxonomically widespread, occurring for at least one population in X\% of the taxonomic orders recorded.
-Accounting for time series length and partially pooling inference across taxonomic class and order with a multilevel model, there was stronger evidence for black swans in insect populations than is apparent in Fig.~\ref{fig:nu-coefs} --- four of 10 orders with the highest median probability of heavy tails were insect orders --- however, there was considerable uncertainty in these estimates (Fig.~\ref{fig:order-estimates}).
+We found strong but rare evidence for black-swan dynamics. Defining black-swan
+dynamics as a greater than $0.5$ probability that $\nu < 10$, we observed black
+swans most frequently for birds (X\%) followed by mammals (X\%), insects (X\%),
+and fishes (X\%) (Fig.~\ref{fig:nu-coefs}, \ref{fig:heavy-ts}). Black swans
+were taxonomically widespread, occurring for at least one population in X\% of
+the taxonomic orders recorded. Accounting for time series length and partially
+pooling inference across taxonomic class and order with a multilevel model,
+there was stronger evidence for black swans in insect populations than is
+apparent in Fig.~\ref{fig:nu-coefs} --- four of 10 orders with the highest
+median probability of heavy tails were insect orders --- however, there was
+considerable uncertainty in these estimates (Fig.~\ref{fig:order-estimates}).
 
-The majority of our heavy-tailed estimates were robust to alternative population models and observation error (Fig.~\ref{fig:alt}).
-Including an autocorrelation structure in the residuals, modelling population growth rates as a random walk with drift, or modelling the population dynamics as Ricker-logistic did not systematically alter our conclusions.
-Allowing for a fixed quantity of observation error decreased our estimates of black-swan dynamics, increasing the median estimate of $\nu$ from $<10$ to $>10$ in X of XX populations although the majority of $\nu$ estimates remained qualitatively similar (Fig.~\ref{fig:alt})
+The majority of our heavy-tailed estimates were robust to alternative
+population models and observation error (Fig.~\ref{fig:alt}). Including an
+autocorrelation structure in the residuals, modelling population growth rates
+as a random walk with drift, or modelling the population dynamics as
+Ricker-logistic did not systematically alter our conclusions. Allowing for
+a fixed quantity of observation error decreased our estimates of black-swan
+dynamics, increasing the median estimate of $\nu$ from $<10$ to $>10$ in X of
+XX populations although the majority of $\nu$ estimates remained qualitatively
+similar (Fig.~\ref{fig:alt})
 
-Across populations, the probability of observing black swan dynamics was positively related to time-series length and negatively related to magnitude of process noise ($\sigma$) but not clearly related to population growth rate ($\lambda$), density dependence ($b$), or maximum lifespan (Fig.~\ref{fig:correlates}, \ref{fig:correlate-coefs}).
-Longer time-series length was the strongest covariate of observing black swan dynamics (as was similarly observed in simulation testing, Fig.~\ref{fig:sim-nu}).
-For example, we were about 15\% TODO more likely to observe heavy tails in a population with 40 time steps than in one with 30 time steps (Fig.~\ref{fig:perc-inc-p}).
-However, the absolute change in probability with increased time series length was small ($0.X$ vs.
-$0.X$ in the previous example, Fig.~\ref{fig:correlates}).
+Across populations, the probability of observing black swan dynamics was
+positively related to time-series length and negatively related to magnitude of
+process noise ($\sigma$) but not clearly related to population growth rate
+($\lambda$), density dependence ($b$), or maximum lifespan
+(Fig.~\ref{fig:correlates}, \ref{fig:correlate-coefs}). Longer time-series
+length was the strongest covariate of observing black swan dynamics (as was
+similarly observed in simulation testing, Fig.~\ref{fig:sim-nu}). For example,
+we were about 15\% TODO more likely to observe heavy tails in a population with
+40 time steps than in one with 30 time steps (Fig.~\ref{fig:perc-inc-p}).
+However, the absolute change in probability with increased time series length
+was small ($0.X$ vs. $0.X$ in the previous example, Fig.~\ref{fig:correlates}).
 
-We assembled the documented causes of black swan population dynamics in seven populations with high probabilities of heavy tails (Table 1).
-The majority of documented events were downward black swans and the majority involved a confluence of events.
-For example, a shortage of nest sites for the shag in the UK was thought to reduce population productivity and make the population vulnerable when a red-tide hit in 1968.
-The population experienced a black swan downswing, but the new availability of nest sites resulted in a rapid population upswing \citep{potts1980}.
-As another example, an interaction between population cycles caused by predation combined with cycles caused by the environment are thought to have caused a downward black swan for a water vole population \citep{saucy1994}.
-Other black swans were the result of a sequence of extreme climate events on their own.
-For example, severe winters in 1929, 1940--1942, and 1962--1963 were associated with black swan downswings in grey heron abundance in the UK \citep{stafford1971}.
-Our analysis finds that the last event was a combination of two black swan events in a row and population recovery took three times longer than predicted \citep{stafford1971}.
+We assembled the documented causes of black swan population dynamics in seven
+populations with high probabilities of heavy tails (Table 1). The majority of
+documented events were downward black swans and the majority involved
+a confluence of events. For example, a shortage of nest sites for the shag in
+the UK was thought to reduce population productivity and make the population
+vulnerable when a red-tide hit in 1968. The population experienced a black swan
+downswing, but the new availability of nest sites resulted in a rapid
+population upswing \citep{potts1980}. As another example, an interaction
+between population cycles caused by predation combined with cycles caused by
+the environment are thought to have caused a downward black swan for a water
+vole population \citep{saucy1994}. Other black swans were the result of
+a sequence of extreme climate events on their own. For example, severe winters
+in 1929, 1940--1942, and 1962--1963 were associated with black swan downswings
+in grey heron abundance in the UK \citep{stafford1971}. Our analysis finds that
+the last event was a combination of two black swan events in a row and
+population recovery took three times longer than predicted
+\citep{stafford1971}.
 
@@
 
-We found strong evidence that the multiplicative process noise jumps from time step to time step were greater than we would expect under a normal distribution and a Gompertz or Ricker-logistic model in XX--XX\% of populations assessed.
-We detected black swans more frequently for longer time series and for populations with a smaller magnitude of process error, as we might expect.
-However, we failed to find strong evidence that black swans were associated with density dependence, population growth rate, or lifespan.
-In cases that we verified, black swans were often a result of the interaction between elements of extreme climate, predators and parasite cycles, and strong changes in human pressures.
-Our empirical results, sensitivity tests, and simulation tests suggest that estimating the tail shape of process error is a viable method of detecting heavy-tailed population dynamics.
-The presence of black swans highlights the importance of developing management strategies that detect quickly, respond to, and are robust to extremes in population dynamics --- particularly as the frequency and magnitude of climatic extremes increase over the next century (REF).
+We found strong evidence that the multiplicative process noise jumps from time
+step to time step were greater than we would expect under a normal distribution
+and a Gompertz or Ricker-logistic model in XX--XX\% of populations assessed. We
+detected black swans more frequently for longer time series and for populations
+with a smaller magnitude of process error, as we might expect. However, we
+failed to find strong evidence that black swans were associated with density
+dependence, population growth rate, or lifespan. In cases that we verified,
+black swans were often a result of the interaction between elements of extreme
+climate, predators and parasite cycles, and strong changes in human pressures.
+Our empirical results, sensitivity tests, and simulation tests suggest that
+estimating the tail shape of process error is a viable method of detecting
+heavy-tailed population dynamics. The presence of black swans highlights the
+importance of developing management strategies that detect quickly, respond to,
+and are robust to extremes in population dynamics --- particularly as the
+frequency and magnitude of climatic extremes increase over the next century
+(REF).
 
 Our results clarify, but do not disagree with previous related analyses.
-Previous work with an older version of the GPDD assessed heavy-tailed behaviour but focused on identifying if the lognormal distribution was the most frequently parsimonious model of population abundance \citep{halley2002}.
-For heavy-tailed distributions, \citet{halley2002} fit the extremely heavy-tailed Cauchy distribution and the four-parameter Levy stable distribution and found little information criteria support for these distributions in longer time series.
-However, black swan events are by definition rare, and the majority of time series in the GPDD are short for these purposes.
-Therefore, we would not expect to observe black swans in a large proportion of populations.
-Furthermore, by probabilistically quantifying process noise tail behaviour on a continuous spectrum, and allowing for non-stationary time series and density dependence, our analysis allows for a more nuanced description of the evidence of ecological black swans.
-In an earlier study, \citet{keitt1998} described heavy (power law) tails in breeding bird population abundance.
-However, this finding was challenged by \citet{allen2001}, who showed that mixing data across species could create falsely generate heavy tails.
-This rebuttal supports our use of the t distribution to represent heavy tails, since a t distribution can be represented by a mixture of normal distributions \citep[with inverse-gamma-distributed variances,][]{gelman2014}.
+Previous work with an older version of the GPDD assessed heavy-tailed behaviour
+but focused on identifying if the lognormal distribution was the most
+frequently parsimonious model of population abundance \citep{halley2002}. For
+heavy-tailed distributions, \citet{halley2002} fit the extremely heavy-tailed
+Cauchy distribution and the four-parameter Levy stable distribution and found
+little information criteria support for these distributions in longer time
+series. However, black swan events are by definition rare, and the majority of
+time series in the GPDD are short for these purposes. Therefore, we would not
+expect to observe black swans in a large proportion of populations.
+Furthermore, by probabilistically quantifying process noise tail behaviour on
+a continuous spectrum, and allowing for non-stationary time series and density
+dependence, our analysis allows for a more nuanced description of the evidence
+of ecological black swans. In an earlier study, \citet{keitt1998} described
+heavy (power law) tails in breeding bird population abundance. However, this
+finding was challenged by \citet{allen2001}, who showed that mixing data across
+species could create falsely generate heavy tails. This rebuttal supports our
+use of the t distribution to represent heavy tails, since a t distribution can
+be represented by a mixture of normal distributions \citep[with
+inverse-gamma-distributed variances,][]{gelman2014}.
 
-We might expect to observe black swan dynamics in ecological time series because of unmodelled intrinsic properties of populations or extrinsic forces acting on populations.
-Since a t-distribution can be formed by a mixture of normal distributions, we could observe heavy tails if we miss some underlying mixture of intrinsic processes \citep{allen2001}.
-That process might be an aggregation of populations across space, or population diversity, or some intrinsic change in population variability through time.
-Extrinsic forces could also cause black swan dynamics (REF).
-These forces could be extreme themselves.
-For example, extreme climate, predation from or competition with other species experiencing black swans, or sharp changes in human pressure such hunting, fishing, or habitat destruction might cause black swans (REFS).
-Alternatively, the interaction of multiple ``normal'' extrinsic forces could give rise to black swan ecological dynamics.
-This could occur if the interaction nature is synergistic \citep[e.g.][]{kirby2009} or even if non-synergistic interactions experience a rare alignment \citep{denny2009}.
+We might expect to observe black swan dynamics in ecological time series
+because of unmodelled intrinsic properties of populations or extrinsic forces
+acting on populations. Since a t-distribution can be formed by a mixture of
+normal distributions, we could observe heavy tails if we miss some underlying
+mixture of intrinsic processes \citep{allen2001}. That process might be an
+aggregation of populations across space, or population diversity, or some
+intrinsic change in population variability through time. Extrinsic forces could
+also cause black swan dynamics (REF). These forces could be extreme themselves.
+For example, extreme climate, predation from or competition with other species
+experiencing black swans, or sharp changes in human pressure such hunting,
+fishing, or habitat destruction might cause black swans (REFS). Alternatively,
+the interaction of multiple ``normal'' extrinsic forces could give rise to
+black swan ecological dynamics. This could occur if the interaction nature is
+synergistic \citep[e.g.][]{kirby2009} or even if non-synergistic interactions
+experience a rare alignment \citep{denny2009}.
 
 There a number of caveats when considering the generality of our results.
-First, the GPDD data represent a taxonomically and geographically biased sample of populations.
-Also, the longer time series we focus on are dominated by commercially and recreationally important species and a disproportionate number of populations are located in the United Kingdom.
-Although we expect that we would find strong but rare evidence for black swans in other large taxonomic or geographic samples, the common forces driving those black swans (e.g.~severe winters in Table~1) likely differ.
-Second, in a large database with disparate datasources, we cannot ensure the quality of every datapoint.
-Recording mistakes undoubtedly comprise some of the black swans we detected.
-Conversely, recording mistakes could mask heavy tails if initial observations are discarded or altered as assumed recording errors.
-Third, the temporal scale of observation and population dynamics vary considerably across populations in the GPDD and this likely influences the detection of heavy tails.
-For example, if we make frequent observations relative to generation time (e.g.~for many large bodied mammals) we will average across generations and perhaps not observe the tails.
-Furthermore, if we record population censuses infrequently relative to generation time (e.g.~many insects in the GPDD) the recorded data may average across extreme and less-extreme events and dampen observed black swan dynamics.
+First, the GPDD data represent a taxonomically and geographically biased sample
+of populations. Also, the longer time series we focus on are dominated by
+commercially and recreationally important species and a disproportionate number
+of populations are located in the United Kingdom. Although we expect that we
+would find strong but rare evidence for black swans in other large taxonomic or
+geographic samples, the common forces driving those black swans (e.g.~severe
+winters in Table~1) likely differ. Second, in a large database with disparate
+data sources, we cannot ensure the quality of every data point. Recording
+mistakes undoubtedly comprise some of the black swans we detected. Conversely,
+recording mistakes could mask heavy tails if initial observations are discarded
+or altered as assumed recording errors. Third, the temporal scale of observation
+and population dynamics vary considerably across populations in the GPDD and
+this likely influences the detection of heavy tails. For example, if we make
+frequent observations relative to generation time (e.g.~for many large bodied
+mammals) we will average across generations and perhaps not observe the tails.
+Furthermore, if we record population censuses infrequently relative to
+generation time (e.g.~many insects in the GPDD) the recorded data may average
+across extreme and less-extreme events and dampen observed black swan dynamics.
 
-Our results suggest a number of important future issues to address regarding ecological black swans.
-First, given that black swans can and do occur, can we forecast the probability of black swans in space and time?
-Furthermore, what management policies allow us to detect them quickly after they happen?
-Secondly, can we isolate the components of ecological dynamics that experience black swans events by moving from phenomenological models such as the Gompertz to mechanistic models, that take into account, for example, recruitment?
-We expect that greater insight into the mechanisms and covariates of ecological black swans may be best obtained through specific geographic and taxonomic subsets of data where longer time series are available \citep[e.g.][]{segura2013}.
-Finally, if a sizeable fraction of populations experience heavy tails in population process noise, what is the impact of allowing for black swans in forecasts of ecological risk?
-Financial analysts are acutely concerned with the shape of the tail distribution of financial returns since these impact estimates of risk --- the probability of a specific magnitude of undesired event occurring (REF).
-It is possible that a comparable focus in ecology would alter our perception of ecological forecasts such as extinction risk.
+Our results suggest a number of important future issues to address regarding
+ecological black swans. First, given that black swans can and do occur, can we
+forecast the probability of black swans in space and time? Furthermore, what
+management policies allow us to detect them quickly after they happen?
+Secondly, can we isolate the components of ecological dynamics that experience
+black swans events by moving from phenomenological models such as the Gompertz
+to mechanistic models, that take into account, for example, recruitment? We
+expect that greater insight into the mechanisms and covariates of ecological
+black swans may be best obtained through specific geographic and taxonomic
+subsets of data where longer time series are available
+\citep[e.g.][]{segura2013}. Finally, if a sizeable fraction of populations
+experience heavy tails in population process noise, what is the impact of
+allowing for black swans in forecasts of ecological risk? Financial analysts
+are acutely concerned with the shape of the tail distribution of financial
+returns since these impact estimates of risk --- the probability of a specific
+magnitude of undesired event occurring (REF). It is possible that a comparable
+focus in ecology would alter our perception of ecological forecasts such as
+extinction risk.
 
@@
 
-We thank members of the Earth to Ocean research group for helpful discussions and comments.
-We are grateful to the contributors and maintainers of the Global Population Dynamics Database and to Compute Canada's WestGrid high-performance computing resources.
-Three silhouette images were obtained from \texttt{phylopic.org} under a Creative Commons Attribution 3.0 Unported license: rabbit (Sarah Werning), hoverfly (Gareth Monger), bird (Jean-Raphaël Guillaumin {[}photography{]} and T.
-Michael Keesey {[}vectorization{]}).
-Funding was provided by a Simon Fraser University Graduate Fellowship (SCA), the Natural Sciences and Engineering Research Council of Canada (NKD, ABC), the Canada Research Chairs Program (NKD), and XX-TODO (TAB).
+We thank members of the Earth to Ocean research group for helpful discussions
+and comments. We are grateful to the contributors and maintainers of the Global
+Population Dynamics Database and to Compute Canada's WestGrid high-performance
+computing resources. Three silhouette images were obtained from
+\texttt{phylopic.org} under a Creative Commons Attribution 3.0 Unported
+license: rabbit (Sarah Werning), hoverfly (Gareth Monger), bird (Jean-Raphaël
+Guillaumin {[}photography{]} and T. Michael Keesey {[}vectorization{]}).
+Funding was provided by a Simon Fraser University Graduate Fellowship (SCA),
+the Natural Sciences and Engineering Research Council of Canada (NKD, ABC), the
+Canada Research Chairs Program (NKD), and XX-TODO (TAB).
 

### 2014-10-18 supp.tex
Wrap ms and (very) minor edits

@@
 
-We applied the following data selection and quality-control rules to the Global Population Dynamics Database (GPDD):
+We applied the following data selection and quality-control rules to the Global
+Population Dynamics Database (GPDD):
 
 \begin{enumerate}
-\item
-  To remove populations with unreliable population indices that could be strongly confounded with economics and sampling effort, we removed all populations with a sampling protocol listed as \texttt{harvest} as well populations with the words \texttt{harvest} or \texttt{fur} in the cited reference title.
-\item
-  We removed all populations with uneven sampling intervals. I.e.~we removed populations that didn't have a constant difference between the ``decimal year begin'' and ``decimal year end'' columns.
-\item
-  We removed all populations rated as $< 2$ in the GPDD quality assessment (on a scale of $1$ to $5$, with $1$ being the lowest quality data) \citep[following][]{sibly2005, ziebarth2010}
-\item
-  Populations with negative abundance values were assumed to be log values and transformed by taking $10$ to the power of the recorded abundance.
-In many cases this was noted for the population, but not in all cases.
-We inspected each of these \totalAssumedLog~time series to make sure our assumption made sense.
-\item
-  We filled in all missing time steps with \texttt{NA} values and imputed single missing values with the geometric mean of the previous and following values.
-We chose a geometric mean to be linear on the log scale that the Gompertz and Ricker-logistic models were fitted on.
-\item
-  We filled in single recorded values of zero with the lowest non-zero value in the time series \citep[following][]{brook2006a}.
-This assumes that single values of zero result from abundance being low enough that censusing missed present individuals.
-We turned multiple zero values in a row into \texttt{NA} values.
-This implies that multiple zero values were either censusing errors or caused by emigration.
-Regardless, our population models were fitted on a multiplicative (log) scale and so could not account for zero abundance.
-\item
-  We removed all populations with four or more identical values in a row since these suggest either recording error or extrapolation between two observations.
-\item
-  We removed all populations without at least four unique values \citep[following][]{brook2006a}.
-\item
-  We then wrote an algorithm to find the longest unbroken window of abundance (no \texttt{NA}s) with at least $20$ time steps in each population time series.
-If there were any populations with multiple windows of identical length, we took the most recent window.
-This is a longer window than used in some previous analyses \citep[e.g.][]{brook2006a}, but since our model attempts to capture the shape of the distribution tails, our model requires more data.
-\item
-  Finally, we removed GPDD Main ID \texttt{20531}, which we noticed was a duplicate of \texttt{10139} (a heron population).
+
+\item To remove populations with unreliable population indices that could be
+  strongly confounded with economics and sampling effort, we removed all
+  populations with a sampling protocol listed as \texttt{harvest} as well
+  populations with the words \texttt{harvest} or \texttt{fur} in the cited
+  reference title.
+
+\item We removed all populations with uneven sampling intervals. I.e.~we removed
+  populations that didn't have a constant difference between the ``decimal year
+  begin'' and ``decimal year end'' columns.
+
+\item We removed all populations rated as $< 2$ in the GPDD quality assessment
+  (on a scale of $1$ to $5$, with $1$ being the lowest quality data)
+  \citep[following][]{sibly2005, ziebarth2010}
+
+\item Populations with negative abundance values were assumed to be log values
+  and transformed by taking $10$ to the power of the recorded abundance. In many
+  cases this was noted for the population, but not in all cases. We inspected
+  each of these \totalAssumedLog~time series to make sure our assumption made
+  sense.
+
+\item We filled in all missing time steps with \texttt{NA} values and imputed
+  single missing values with the geometric mean of the previous and following
+  values. We chose a geometric mean to be linear on the log scale that the
+  Gompertz and Ricker-logistic models were fitted on.
+
+\item We filled in single recorded values of zero with the lowest non-zero value
+  in the time series \citep[following][]{brook2006a}. This assumes that single
+  values of zero result from abundance being low enough that censusing missed
+  present individuals. We turned multiple zero values in a row into \texttt{NA}
+  values. This implies that multiple zero values were either censusing errors or
+  caused by emigration. Regardless, our population models were fitted on
+  a multiplicative (log) scale and so could not account for zero abundance.
+
+\item We removed all populations with four or more identical values in a row
+  since these suggest either recording error or extrapolation between two
+  observations.
+
+\item We removed all populations without at least four unique values
+  \citep[following][]{brook2006a}.
+
+\item We then wrote an algorithm to find the longest unbroken window of
+  abundance (no \texttt{NA}s) with at least $20$ time steps in each population
+  time series. If there were any populations with multiple windows of identical
+  length, we took the most recent window. This is a longer window than used in
+  some previous analyses \citep[e.g.][]{brook2006a}, but since our model
+  attempts to capture the shape of the distribution tails, our model requires
+  more data.
+
+\item Finally, we removed GPDD Main ID \texttt{20531}, which we noticed was
+  a duplicate of \texttt{10139} (a heron population).
+
 \end{enumerate}
+
 \noindent
-We provide a supplemental figure of all the time series included in our analysis and indicate which values were interpolated (\percImputedPops\% of populations had at least one point interpolated and only \percImputedPoints\% of the total observations were interpolated) (Fig.~\ref{fig:all-ts}).
-Table S1 shows the final taxonomic breakdown and the number of populations with interpolated values.
+We provide a supplemental figure of all the time series included in our analysis
+and indicate which values were interpolated (\percImputedPops\% of populations
+had at least one point interpolated and only \percImputedPoints\% of the total
+observations were interpolated) (Fig.~\ref{fig:all-ts}). Table S1 shows the
+final taxonomic breakdown and the number of populations with interpolated
+values.
 
@@
 \nu &\sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2). \end{align*}
-Our prior on $b$ was uninformative between values of $-1$ and $2$.
-We would not expect values of $b$ with levels of inverse density dependence as low as $-1$, nor would we expect values above $1$.
-We allowed values of $b$ above $1$ to allow for somewhat non-stationary time series of growth rates.
-The estimates of $b$ were well within these bounds.
-Our prior on $\lambda$ was very weakly informative within the range of expected values for population growth and is similar to priors suggested by \citet{gelman2008d} for intercepts of regression models.
-Our prior on $\sigma$ follows \citet{gelman2006c} and \citet{gelman2008d} and is based on our expected range of the value in nature from previous studies \citep[e.g.][]{connors2014}.
-In our testing of a subsample of populations, our parameter estimates were not qualitatively changed by switching to an uninformative uniform prior on $\sigma$, but the weakly informative prior substantially sped up chain convergence.
+Our prior on $b$ was uninformative between values of $-1$ and $2$. We would not
+expect values of $b$ with levels of inverse density dependence as low as $-1$,
+nor would we expect values above $1$. We allowed values of $b$ above $1$ to
+allow for somewhat non-stationary time series of growth rates. The estimates of
+$b$ were well within these bounds. Our prior on $\lambda$ was very weakly
+informative within the range of expected values for population growth and is
+similar to priors suggested by \citet{gelman2008d} for intercepts of regression
+models. Our prior on $\sigma$ follows \citet{gelman2006c} and
+\citet{gelman2008d} and is based on our expected range of the value in nature
+from previous studies \citep[e.g.][]{connors2014}. In our testing of a subsample
+of populations, our parameter estimates were not qualitatively changed by
+switching to an uninformative uniform prior on $\sigma$, but the weakly
+informative prior substantially sped up chain convergence.
 
-Our prior on $\nu$ is based on \citet{fernandez1998}; they chose a more informative exponential rate parameter of $0.1$.
-We chose a less informative rate parameter of $0.01$ and truncated the distribution at $2$, since at $\nu < 2$ the variance of the t distribution is undefined.
-This prior gives only a $7.7$\% probability $\nu < 10$ but constrains the sampling sufficiently to avoid wandering off towards infinity --- above approximately $\nu = 20$ the t distribution is so similar to the normal distribution (Fig.~\ref{fig:didactic}) that time series of the length considered here are unlikely to be sufficiently informative about the precise value of $\nu$.
-In the scenario where the data are not heavy tailed (e.g.~Fig.~\ref{fig:didactic}e, h) the posterior will approximately match the prior (median $= 71$, mean $= 102$) and not be flagged as likely heavy tailed using the metrics we used in our paper (e.g.~Pr$(\nu < 10) > 0.5$).
+Our prior on $\nu$ was based on \citet{fernandez1998}; they chose a more
+informative exponential rate parameter of $0.1$. We chose a less informative
+rate parameter of $0.01$ and truncated the distribution at $2$, since at $\nu
+< 2$ the variance of the t distribution is undefined. This prior gives only
+a $7.7$\% probability $\nu < 10$ but constrains the sampling sufficiently to
+avoid wandering off towards infinity --- above approximately $\nu = 20$ the
+t distribution is so similar to the normal distribution
+(Fig.~\ref{fig:didactic}) that time series of the length considered here are
+unlikely to be sufficiently informative about the precise value of $\nu$. In the
+scenario where the data are not heavy tailed (e.g.~Fig.~\ref{fig:didactic}e, h)
+the posterior will approximately match the prior (median $= 71$, mean $= 102$)
+and not be flagged as likely heavy tailed using the metrics we used in our paper
+(e.g.~Pr$(\nu < 10) > 0.5$).
 
-We fitted our models with Stan 2.4.0 \citep{stan-manual2014}, and R 3.1.1 \citep{r2014}.
-We began with four chains and $2000$ iterations, discarding the first $1000$ as warm up (i.e.~4000 total samples).
-If $\hat{R}$ (a measure of chain convergence) was greater than $1.05$ for any parameter or the minimum effective sample size, $n_\mathrm{eff}$, (a measure of the effective number of uncorrelated samples) for any parameter was less than $200$, we doubled both the total iterations and warm up period and sampled from the model again.
-These thresholds are in excess of the minimums recommended by \citet{gelman2006a} of $\hat{R} < 1.1$ and effective sample size $> 100$ for reliable point estimates and confidence intervals.
-In the majority of cases our minimum thresholds were greatly exceeded.
-We continued this procedure up to $8000$ iterations ($16000$ total samples) by which all chains were deemed to have sufficiently converged.
+We fitted our models with Stan 2.4.0 \citep{stan-manual2014}, and R 3.1.1
+\citep{r2014}. We began with four chains and $2000$ iterations, discarding the
+first $1000$ as warm up (i.e.~4000 total samples). If $\hat{R}$ (a measure of
+chain convergence) was greater than $1.05$ for any parameter or the minimum
+effective sample size, $n_\mathrm{eff}$, (a measure of the effective number of
+uncorrelated samples) for any parameter was less than $200$, we doubled both the
+total iterations and warm up period and sampled from the model again. These
+thresholds are in excess of the minimums recommended by \citet{gelman2006a} of
+$\hat{R} < 1.1$ and effective sample size $> 100$ for reliable point estimates
+and confidence intervals. In the majority of cases our minimum thresholds were
+greatly exceeded. We continued this procedure up to $8000$ iterations ($16000$
+total samples) by which all chains were deemed to have sufficiently converged.
 
@@
 
-We performed two types of simulation testing.
-First, we tested how easily the Student-t distribution $\nu$ parameter could be recovered given different true values of $\nu$ and sample sizes.
-Second, we tested the ability of the heavy-tailed Gompertz model to obtain unbiased parameter estimates of $\nu$ given that a set of process deviations was provided in which the ``effective $\nu$'' value was close to the true $\nu$ value.
+We performed two types of simulation testing. First, we tested how easily the
+Student-t distribution $\nu$ parameter could be recovered given different true
+values of $\nu$ and sample sizes. Second, we tested the ability of the
+heavy-tailed Gompertz model to obtain unbiased parameter estimates of $\nu$
+given that a set of process deviations was provided in which the ``effective
+$\nu$'' value was close to the true $\nu$ value.
 
-We separated our simulation into these two components to avoid confounding two issues.
-(1) With smaller sample sizes, there may not be a stochastic draw from the tails of a distribution.
-In that case, no model, no matter how perfect a model, will be able to detect the shape of the tails.
-(2) Models (particularly complex models) may return biased parameter estimates if there are conceptual, computational, or coding errors in the models.
-Our first simulation tested the first issue and our second simulation tested the latter.
-In general, out simulations show that, if anything, our model under predicts the magnitude and frequency of heavy tailed events --- especially given the length of the time series in the GPDD.
+We separated our simulation into these two components to avoid confounding two
+issues. (1) With smaller sample sizes, there may not be a stochastic draw from
+the tails of a distribution. In that case, no model, no matter how perfect
+a model, will be able to detect the shape of the tails. (2) Models (particularly
+complex models) may return biased parameter estimates if there are conceptual,
+computational, or coding errors in the models. Our first simulation tested the
+first issue and our second simulation tested the latter. In general, out
+simulations show that, if anything, our model under predicts the magnitude and
+frequency of heavy tailed events --- especially given the length of the time
+series in the GPDD.
 
@@
 
-First, we tested the ability to estimate $\nu$ given different true values of $\nu$ and sample sizes.
-We took stochastic draws from t distributions with different $\nu$ values ($\nu = 3, 5, 10,$ and $10^6$ [$\approx$ normal]), with central tendency parameters of $0$, and scale parameters of $1$.
-We started with $1600$ stochastic draws and then fitted the models again at the first $800, 400, 200, 100, 50,$ and $25$ draws.
-Each time we recorded the posterior samples of $\nu$.
+First, we tested the ability to estimate $\nu$ given different true values of
+$\nu$ and sample sizes. We took stochastic draws from t distributions with
+different $\nu$ values ($\nu = 3, 5, 10,$ and $10^6$ [$\approx$ normal]), with
+central tendency parameters of $0$, and scale parameters of $1$. We started with
+$1600$ stochastic draws and then fitted the models again at the first $800, 400,
+200, 100, 50,$ and $25$ draws. Each time we recorded the posterior samples of
+$\nu$.
 
-We found that we could consistently and precisely recover median posterior estimates of $\nu$ near the true value of $\nu$ with large samples ($\ge 200$) (Fig.~\ref{fig:sim-gompertz} upper panels).
-At smaller samples the we could still usually distinguish ``heavy'' from ``not-heavy'' tails, but the model tended to underestimate how heavy the tails were.
-At the same time, when the model underestimated how heavy-tailed the distribution was, it tended to overestimate how large the scale parameter was (Fig.~\ref{fig:sim-gompertz} lower panels).
+We found that we could consistently and precisely recover median posterior
+estimates of $\nu$ near the true value of $\nu$ with large samples ($\ge 200$)
+(Fig.~\ref{fig:sim-gompertz} upper panels). At smaller samples the we could
+still usually distinguish ``heavy'' from ``not-heavy'' tails, but the model
+tended to underestimate how heavy the tails were. At the same time, when the
+model underestimated how heavy-tailed the distribution was, it tended to
+overestimate how large the scale parameter was (Fig.~\ref{fig:sim-gompertz}
+lower panels).
 
@@
 
-In the second part of our simulation testing, we tested the ability of the heavy-tailed Gompertz model to obtain unbiased parameter estimates given that process deviations were provided in which our simpler model (in the previous section) could accurately estimate the value of $\nu$.
-To generate these process errors, we repeatedly generated candidate process deviations and estimated the central tendency, scale, and $\nu$ values each time.
-We recorded when $\hat{\nu}$ was within $0.2$ CVs (coefficient of variations) of the true $\nu$ value and used this set of random seed values in our Gompertz simulation.
-We then fitted our Gompertz models to the simulated datasets with all parameters (except $\nu$) set near the median values estimated in the GPDD.
-We repeated this with $50$ and $100$ samples without observation error, $50$ samples with observation error ($\sigma_\mathrm{obs} = 0.2$), and $50$ samples with the same observation error and a Gompertz model that allowed for correctly specified observation error magnitude.
+In the second part of our simulation testing, we tested the ability of the
+heavy-tailed Gompertz model to obtain unbiased parameter estimates given that
+process deviations were provided in which our simpler model (in the previous
+section) could accurately estimate the value of $\nu$. To generate these process
+errors, we repeatedly generated candidate process deviations and estimated the
+central tendency, scale, and $\nu$ values each time. We recorded when
+$\hat{\nu}$ was within $0.2$ CVs (coefficient of variations) of the true $\nu$
+value and used this set of random seed values in our Gompertz simulation. We
+then fitted our Gompertz models to the simulated datasets with all parameters
+(except $\nu$) set near the median values estimated in the GPDD. We repeated
+this with $50$ and $100$ samples without observation error, $50$ samples with
+observation error ($\sigma_\mathrm{obs} = 0.2$), and $50$ samples with the same
+observation error and a Gompertz model that allowed for correctly specified
+observation error magnitude.
 
-Our results show that the more-complicated Gompertz model can still recapture the true value of $\nu$ when the process noise was chosen so that appropriate tail events were present (Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}, red and green symbols in the top rows).
-Likewise, the other Gompertz parameters were estimated without any systematic bias (Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}, red and green symbols).
-The addition of observation error caused the model to tend to underestimate the degree of heavy-tailedness, overestimate the magnitude of process noise, somewhat overestimate $\lambda$, and overestimate density dependence (blue symbols in Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}).
-The overestimation of density dependence with observation error is a known phenomenon \citep{knape2012}.
-Fitting a model with correctly specified observation error made marginal improvements to model bias (purple symbols in Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}).
+Our results show that the more-complicated Gompertz model can still recapture
+the true value of $\nu$ when the process noise was chosen so that appropriate
+tail events were present (Figs~\ref{fig:sim-gompertz} and
+\ref{fig:sim-gompertz-boxplots}, red and green symbols in the top rows).
+Likewise, the other Gompertz parameters were estimated without any systematic
+bias (Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}, red and
+green symbols). The addition of observation error caused the model to tend to
+underestimate the degree of heavy-tailedness, overestimate the magnitude of
+process noise, somewhat overestimate $\lambda$, and overestimate density
+dependence (blue symbols in Figs~\ref{fig:sim-gompertz} and
+\ref{fig:sim-gompertz-boxplots}). The overestimation of density dependence with
+observation error is a known phenomenon \citep{knape2012}. Fitting a model with
+correctly specified observation error made marginal improvements to model bias
+(purple symbols in Figs~\ref{fig:sim-gompertz} and
+\ref{fig:sim-gompertz-boxplots}).
 
-When converting the posterior distributions of $\nu$ into Pr($\nu < 10$) the models performed well.
-Without observation error, and using a probability of $0.5$ as a threshold, the model would have miscategorized only one of $40$ simulations at $\nu = 5$ across a sample size of $50$ or $100$ data points (Fig.~\ref{fig:sim-prob}, upper panels).
-The model would have correctly categorized all cases where the process noise was not heavy tailed (indicated as ``$\nu =$ infinity'' in Fig.~\ref{fig:sim-prob}) and all cases where $\nu = 3$.
-With $0.2$ standard deviations of observation error, the model still categorized XX\% of cases as heavy tailed when $\nu = 5$ and all cases where $\nu = 3$.
-Allowing for observation error made little improvement to the detection of heavy tails (Fig.~\ref{fig:sim-prob}, lower-left vs.\ lower-right panel).
-Therefore, we chose to focus on the simpler model without observation error in the main text.
+When converting the posterior distributions of $\nu$ into Pr($\nu < 10$) the
+models performed well. Without observation error, and using a probability of
+$0.5$ as a threshold, the model would have miscategorized only one of $40$
+simulations at $\nu = 5$ across a sample size of $50$ or $100$ data points
+(Fig.~\ref{fig:sim-prob}, upper panels). The model would have correctly
+categorized all cases where the process noise was not heavy tailed (indicated as
+``$\nu =$ infinity'' in Fig.~\ref{fig:sim-prob}) and all cases where $\nu = 3$.
+With $0.2$ standard deviations of observation error, the model still categorized
+XX\% of cases as heavy tailed when $\nu = 5$ and all cases where $\nu = 3$.
+Allowing for observation error made little improvement to the detection of heavy
+tails (Fig.~\ref{fig:sim-prob}, lower-left vs.\ lower-right panel). Therefore,
+we chose to focus on the simpler model without observation error in the main
+text.
 
@@
 
-We fit four alternative population models to the time-series data to check how they would influence our conclusions.
-Our alternative models allowed for autocorrelation in the residuals, assumed no density dependence, allowed for observation error, or assumed a Ricker-logistic functional form.
-The ranges of percentages of black swans by taxonomic class cited in the abstract are based on lower and upper limits across our main Gompertz model and these four alternative models.
+We fit four alternative population models to the time-series data to check how
+they would influence our conclusions. Our alternative models allowed for
+autocorrelation in the residuals, assumed no density dependence, allowed for
+observation error, or assumed a Ricker-logistic functional form. The ranges of
+percentages of black swans by taxonomic class cited in the abstract are based on
+lower and upper limits across our main Gompertz model and these four alternative
+models.
 
@@
 
-We considered a version of the Gompertz model in which an autoregressive parameter was fitted to the process noise residuals: 
+We considered a version of the Gompertz model in which an autoregressive
+parameter was fitted to the process noise residuals:
 \begin{align*}
@@
 \end{align*}
-In addition to the parameters in the original Gompertz model, we estimate an additional parameter $\phi$, which represents the relationship between subsequent residuals.
-Based on the results of previous analyses with the GPDD \citep[e.g.][]{connors2014} and the chosen priors in previous analyses \citep[e.g.][]{thorson2014a} and to greatly speed up chain convergence when running our model across all populations, we placed a weakly informative prior on $\phi$ that assumed the greatest probability density near zero with the reduced possibility of $\phi$ being near $-1$ or $1$.
-Specifically, we chose $\phi \sim \mathrm{Truncated\mhyphen Normal}(0, 1, \mathrm{min.} = -1, \mathrm{max.} = 1)$.
-The MCMC chains for a small number of these models (XX) did not converge according to our criteria ($\widehat{R} < 1.05, n_\mathrm{eff} > 200$) after 8000 iterations of four chains.
-We did not include these models in Fig.~\ref{fig:alt}.
+In addition to the parameters in the original Gompertz model, we estimate an
+additional parameter $\phi$, which represents the relationship between
+subsequent residuals. Based on the results of previous analyses with the GPDD
+\citep[e.g.][]{connors2014} and the chosen priors in previous analyses
+\citep[e.g.][]{thorson2014a} and to greatly speed up chain convergence when
+running our model across all populations, we placed a weakly informative prior
+on $\phi$ that assumed the greatest probability density near zero with the
+reduced possibility of $\phi$ being near $-1$ or $1$. Specifically, we chose
+$\phi \sim \mathrm{Truncated\mhyphen Normal}(0, 1, \mathrm{min.} = -1,
+\mathrm{max.} = 1)$. The MCMC chains for a small number of these models (XX) did
+not converge according to our criteria ($\widehat{R} < 1.05, n_\mathrm{eff}
+> 200$) after 8000 iterations of four chains. We did not include these models in
+Fig.~\ref{fig:alt}.
 
@@
 
-We fit a simplified version of the Gompertz model in which the density dependence parameter $b$ was fixed at $1$ (density independent).
-This is equivalent to fitting a random walk model (with drift) to the $\ln$ abundances or assuming the growth rates are stationary.
-The model was as follows:
+We fit a simplified version of the Gompertz model in which the density
+dependence parameter $b$ was fixed at $1$ (density independent). This is
+equivalent to fitting a random walk model (with drift) to the $\ln$ abundances
+or assuming the growth rates are stationary. The model was as follows:
 
@@
 \end{align*}
-We fit this model for three reasons: (1) it is computationally simpler and so provides a check that our more complicated full Gompertz model was obtaining ballpark reasonable estimates of $\nu$, (2) it provides a test of whether density dependence was systematically affecting our perception of heavy tails, (3) it matches how some previous authors have modelled heavy tails without accounting for density dependence \citep{segura2013}.
+We fit this model for three reasons: (1) it is computationally simpler and so
+provides a check that our more complicated full Gompertz model was obtaining
+ballpark reasonable estimates of $\nu$, (2) it provides a test of whether
+density dependence was systematically affecting our perception of heavy tails,
+(3) it matches how some previous authors have modelled heavy tails without
+accounting for density dependence \citep{segura2013}.
 
@@
 
-Observation error can bias parameter estimates \citep[e.g.][]{knape2012} and is known to affect the ability to detect extreme events \citep{ward2007}.
-However, simultaneously estimating observation and process error in state-space is a challenging problem and is known to result in identifiability issues with the Gompertz population model \citep{knape2008}.
-Furthermore, our model estimates an additional parameter --- the shape of the process error distribution tails --- potentially making identifiability and computational issues even greater.
-Therefore, we considered a version of the base Gompertz model that allowed for a fixed level of observation error:
+Observation error can bias parameter estimates \citep[e.g.][]{knape2012} and is
+known to affect the ability to detect extreme events \citep{ward2007}. However,
+simultaneously estimating observation and process error in state-space is
+a challenging problem and is known to result in identifiability issues with the
+Gompertz population model \citep{knape2008}. Furthermore, our model estimates an
+additional parameter --- the shape of the process error distribution tails ---
+potentially making identifiability and computational issues even greater.
+Therefore, we considered a version of the base Gompertz model that allowed for
+a fixed level of observation error:
 \begin{align*}
@@
 \end{align*}
-where $U$ represents the unobserved state vector, and $\sigma_\mathrm{obs}$ represents the standard deviation of observation error (on a log scale).
-We set $\sigma_\mathrm{obs}$ to $0.2$, which represents the upper end of values often used in simulation analyses \citep[e.g.][]{valpine2002, thorson2014b}.
+where $U$ represents the unobserved state vector, and $\sigma_\mathrm{obs}$
+represents the standard deviation of observation error (on a log scale). We set
+$\sigma_\mathrm{obs}$ to $0.2$, which represents the upper end of values often
+used in simulation analyses \citep[e.g.][]{valpine2002, thorson2014b}.
 
@@
 \end{align*}
-where $r_\mathrm{max}$ represents the maximum population growth rate that is obtained when $N$ (abundance) $= 0$.
-The parameter $K$ represents the carrying capacity and, as before, $x_t$ represents the $\ln$ transformed abundance at time $t$.
-The Ricker-logistic model assumes a linear decrease in population growth rate ($x_t - x_{t-1}$) with increases in abundance ($N_t$).
-In contrast, the Gompertz model assumes a linear decrease in population growth rate with increases in $\ln$ abundance ($x_t$) (REF).
+where $r_\mathrm{max}$ represents the maximum population growth rate that is
+obtained when $N$ (abundance) $= 0$. The parameter $K$ represents the carrying
+capacity and, as before, $x_t$ represents the $\ln$ transformed abundance at
+time $t$. The Ricker-logistic model assumes a linear decrease in population
+growth rate ($x_t - x_{t-1}$) with increases in abundance ($N_t$). In contrast,
+the Gompertz model assumes a linear decrease in population growth rate with
+increases in $\ln$ abundance ($x_t$) (REF).
 
-To fit the Ricker-logistic models, we chose a prior on $K$ uniform between zero and twice the maximum observed abundance (\citet{clark2010} chose uniform between zero and maximum observed, which is more informative).
-We set the prior on $r_\mathrm{max}$ as uniform between 0 and 20 as in \citet{clark2010}.
-We used the same priors on $\nu$ and $\sigma$ as in the Gompertz model.
+To fit the Ricker-logistic models, we chose a prior on $K$ uniform between zero
+and twice the maximum observed abundance (\citet{clark2010} chose uniform
+between zero and maximum observed, which is more informative). We set the prior
+on $r_\mathrm{max}$ as uniform between 0 and 20 as in \citet{clark2010}. We used
+the same priors on $\nu$ and $\sigma$ as in the Gompertz model.
 
@@
 
-We fitted a multilevel beta regression model to the predicted probability of heavy tails, Pr($\nu < 10$), to investigate potential covariates of heavy-tailed dynamics.
-The beta distribution is useful when the response data range on a continuous scale between zero and one.
-We used a logit link function as is typically used in logistic regression.
-The model was as follows:
+We fitted a multilevel beta regression model to the predicted probability of
+heavy tails, Pr($\nu < 10$), to investigate potential covariates of heavy-tailed
+dynamics. The beta distribution is useful when the response data range on
+a continuous scale between zero and one. We used a logit link function as is
+typically used in logistic regression. The model was as follows:
 \begin{align*}
@@
 \end{align*}
-where $A$ and $B$ represent the beta distribution shape parameters; $\mu_i$ represents the predicted value for population $i$, class $j$, order $k$, and species $l$; $\phi_\mathrm{disp}$ represents the dispersion parameter; and $X_i$ represents a vector of predictors for population $i$ with associated $\beta$ parameters.
-The intercepts are allowed to vary from the overall intercept $\alpha$ by taxonomic classes ($\alpha^\mathrm{class}_j$), taxonomic orders ($\alpha^\mathrm{order}_k$), and species ($\alpha^\mathrm{species}_l$) with standard deviations $\sigma_{\alpha \; \mathrm{class}}$, $\sigma_{\alpha \; \mathrm{order}}$, and $\sigma_{\alpha \; \mathrm{species}}$.
-Where possible, we also allowed for error distributions around the predictors by incorporating the standard deviation of the posterior samples for the Gompertz parameters $\lambda$, $b$, and $\sigma$ around the mean point value (not shown in the above equation).
-We log transformed $\sigma$, time-series length, and lifespan to match the way they are visually represented in Fig.~\ref{fig:correlates}.
-All input variables were standardized by subtracting their mean and dividing by two standard deviations \citep{gelman2008c} to make their coefficients comparable in magnitude.
+where $A$ and $B$ represent the beta distribution shape parameters; $\mu_i$
+represents the predicted value for population $i$, class $j$, order $k$, and
+species $l$; $\phi_\mathrm{disp}$ represents the dispersion parameter; and $X_i$
+represents a vector of predictors for population $i$ with associated $\beta$
+parameters. The intercepts are allowed to vary from the overall intercept
+$\alpha$ by taxonomic classes ($\alpha^\mathrm{class}_j$), taxonomic orders
+($\alpha^\mathrm{order}_k$), and species ($\alpha^\mathrm{species}_l$) with
+standard deviations $\sigma_{\alpha \; \mathrm{class}}$, $\sigma_{\alpha \;
+  \mathrm{order}}$, and $\sigma_{\alpha \; \mathrm{species}}$. Where possible,
+we also allowed for error distributions around the predictors by incorporating
+the standard deviation of the posterior samples for the Gompertz parameters
+$\lambda$, $b$, and $\sigma$ around the mean point value (not shown in the above
+equation). We log transformed $\sigma$, time-series length, and lifespan to
+match the way they are visually represented in Fig.~\ref{fig:correlates}. All
+input variables were standardized by subtracting their mean and dividing by two
+standard deviations \citep{gelman2008c} to make their coefficients comparable in
+magnitude.
 
-We incorporated weakly informative priors into our model: $\mathrm{Cauchy}(0, 10)$ on the global intercept $\alpha$, $\mathrm{Half\mhyphen Cauchy}(0, 2.5)$ on all standard deviation parameters, $\mathrm{Half\mhyphen Cauchy}(0, 10)$ on the dispersion parameter $\phi_\mathrm{disp}$, and $\mathrm{Cauchy}(0, 2.5)$ on all other parameters \citep{gelman2006c,gelman2008d}.
-Compared to normal priors, the Cauchy priors concentrate more probability density at reasonably parameter values while allowing for higher probability density far into the tails, thereby allowing the data to dominate the posterior more strongly if it disagrees with the prior.
-Our conclusions were not qualitatively changed by using uniform priors.
-We fitted our models with 5000 total iterations per chain, 2500 warm-up iterations, four chains, and discarding every second sample to save memory.
-We checked for chain convergence visually and with the same criteria as before ($\widehat{R} < 1.05$ and $n_\mathrm{eft} >200$ for all parameters).
+We incorporated weakly informative priors into our model: $\mathrm{Cauchy}(0,
+10)$ on the global intercept $\alpha$, $\mathrm{Half\mhyphen Cauchy}(0, 2.5)$ on
+all standard deviation parameters, $\mathrm{Half\mhyphen Cauchy}(0, 10)$ on the
+dispersion parameter $\phi_\mathrm{disp}$, and $\mathrm{Cauchy}(0, 2.5)$ on all
+other parameters \citep{gelman2006c, gelman2008d}. Compared to normal priors,
+the Cauchy priors concentrate more probability density at reasonably parameter
+values while allowing for higher probability density far into the tails, thereby
+allowing the data to dominate the posterior more strongly if it disagrees with
+the prior. Our conclusions were not qualitatively changed by using uniform
+priors. We fitted our models with 5000 total iterations per chain, 2500 warm-up
+iterations, four chains, and discarding every second sample to save memory. We
+checked for chain convergence visually and with the same criteria as before
+($\widehat{R} < 1.05$ and $n_\mathrm{eft} >200$ for all parameters).
 
-To derive taxonomic-order-level estimates of the probability of heavy tails accounting for time-series length (Fig \ref{fig:order-estimates}), we fitted a separate multilevel model with the same structure but with only time-series length as a predictor.
-(In this case, we did not want to control for intrinsic population characteristics such as density dependence.) We obtained order-level estimates by adding the posteriors for $\alpha$, $\alpha^\mathrm{class}_j$, and $\alpha^\mathrm{order}_k$.
+To derive taxonomic-order-level estimates of the probability of heavy tails
+accounting for time-series length (Fig \ref{fig:order-estimates}), we fitted
+a separate multilevel model with the same structure but with only time-series
+length as a predictor. (In this case, we did not want to control for intrinsic
+population characteristics such as density dependence.) We obtained order-level
+estimates by adding the posteriors for $\alpha$, $\alpha^\mathrm{class}_j$, and
+$\alpha^\mathrm{order}_k$.
 

### 2014-10-18 ms.tex
Edits to ms

@@
 
-There a number of caveats when considering the generality of our results.
+There are a number of caveats when considering the generality of our results.
 First, the GPDD data represent a taxonomically and geographically biased sample
-of populations. Also, the longer time series we focus on are dominated by
+of populations --- the longer time series we focus on are dominated by
 commercially and recreationally important species and a disproportionate number
@@
 mistakes undoubtedly comprise some of the black swans we detected. Conversely,
-recording mistakes could mask heavy tails if initial observations are discarded
-or altered as assumed recording errors. Third, the temporal scale of observation
-and population dynamics vary considerably across populations in the GPDD and
-this likely influences the detection of heavy tails. For example, if we make
-frequent observations relative to generation time (e.g.~for many large bodied
-mammals) we will average across generations and perhaps not observe the tails.
-Furthermore, if we record population censuses infrequently relative to
+recording mistakes could mask heavy tails if initial observations assumed to be
+recording errors and are discarded or altered. Third, the temporal scale of
+observation and population dynamics vary considerably across populations in the
+GPDD and this likely influences the detection of heavy tails. For example, if
+we make frequent observations relative to generation time (e.g.~for many large
+bodied mammals) we will average across generations and perhaps not observe
+black swans. Furthermore, if we census populations infrequently relative to
 generation time (e.g.~many insects in the GPDD) the recorded data may average
-across extreme and less-extreme events and dampen observed black swan dynamics.
+across extreme and less-extreme events and dampen black swan dynamics.
 
-Our results suggest a number of important future issues to address regarding
-ecological black swans. First, given that black swans can and do occur, can we
-forecast the probability of black swans in space and time? Furthermore, what
-management policies allow us to detect them quickly after they happen?
-Secondly, can we isolate the components of ecological dynamics that experience
-black swans events by moving from phenomenological models such as the Gompertz
-to mechanistic models, that take into account, for example, recruitment? We
-expect that greater insight into the mechanisms and covariates of ecological
-black swans may be best obtained through specific geographic and taxonomic
-subsets of data where longer time series are available
-\citep[e.g.][]{segura2013}. Finally, if a sizeable fraction of populations
-experience heavy tails in population process noise, what is the impact of
-allowing for black swans in forecasts of ecological risk? Financial analysts
-are acutely concerned with the shape of the tail distribution of financial
-returns since these impact estimates of risk --- the probability of a specific
+Our results suggest important future issues to address about ecological black
+swans. Given that black swans can and do occur, can we forecast the probability
+of black swans in space and time? Furthermore, what management policies allow
+us to detect them quickly after they happen? Secondly, can we isolate the
+components of ecological dynamics that experience black swans events by moving
+from phenomenological models such as the Gompertz to mechanistic models, that
+take into account, for example, recruitment? We expect that greater insight
+into the mechanisms and covariates of ecological black swans may be best
+obtained through specific geographic and taxonomic subsets of data where longer
+time series are available \citep[e.g.][]{segura2013}. Finally, what is the
+impact of allowing for black swans in forecasts of ecological risk? Financial
+analysts are concerned with the shape of financial-return distribution tails
+because these impact estimates of risk --- the probability of a specific
 magnitude of undesired event occurring (REF). It is possible that a comparable
@@
 \caption{
-An illustration of fitting population dynamics models that allow for heavy tails, represented by the Student-t degrees of freedom parameter $\nu$.
-(a, b) The probability density for t distributions with a scale parameter of 1 and different values of $\nu$.
-Small values of $\nu$ create heavy tails.
-As $\nu$ approaches infinity the distribution approaches the normal distribution.
-For example, at $\nu = 2$, the probability of drawing a value less than -5 is 1.8\%, whereas the probability of drawing such a value from a normal distribution is nearly zero ($2.9\cdot10^{-5}$\%).
-(c--e) Simulated population dynamics from a Gompertz model with process noise drawn from t distributions with different values of $\nu$.
-Coloured dots in panels c and d represent jumps with less than a 1 in 1000 chance of occurring in a normal distribution.
-(f--h) Estimates of $\nu$ from models fit to the times series in panels c--e.
-Shown are the posterior samples (histograms), median and interquartile range of the posterior (IQR) (dots and line segments), and the exponential prior on $\nu$ (dashed lines).
-Colour shading behind panels f--h illustrates the region of heavy tails.}
+An illustration of fitting population dynamics models that allow for heavy
+tails, represented by the Student-t degrees of freedom parameter $\nu$. (a, b)
+The probability density for t distributions with a scale parameter of 1 and
+different values of $\nu$. Small values of $\nu$ create heavy tails. As $\nu$
+approaches infinity the distribution approaches the normal distribution. For
+example, at $\nu = 2$, the probability of drawing a value less than -5 is
+1.8\%, whereas the probability of drawing such a value from a normal
+distribution is nearly zero ($2.9\cdot10^{-5}$\%). (c--e) Simulated population
+dynamics from a Gompertz model with process noise drawn from t distributions
+with different values of $\nu$. Coloured dots in panels c and d represent jumps
+with less than a 1 in 1000 chance of occurring in a normal distribution. (f--h)
+Estimates of $\nu$ from models fit to the times series in panels c--e. Shown
+are the posterior samples (histograms), median and interquartile range of the
+posterior (IQR) (dots and line segments), and the exponential prior on $\nu$
+(dashed lines). Colour shading behind panels f--h illustrates the region of
+heavy tails.}
 \label{fig:didactic}
@@
 \caption{
-Estimates of how heavy-tailed population dynamics are for \nuCoefPopN\ populations of birds, mammals, insects, and fishes.
-Small values of $\nu$ ($\lesssim 10$) suggest heavy-tailed dynamics; larger values of $\nu$ suggest approximately normal-tailed dynamics.
-Vertical points and line segments represent posterior medians and 50\% / 90\% credible intervals for individual populations.
-Inset plots show probability that $\nu < 10$ (probability of heavy tails) for populations arranged by taxonomic order.
-Taxonomic orders are sorted by decreasing mean p($\nu < 10$).
-Taxonomic orders with three or fewer populations in panel a are omitted for space.
-Red and orange colours highlight populations with a high probability of heavy-tailed dynamics.
+Estimates of how heavy-tailed population dynamics are for \nuCoefPopN\
+populations of birds, mammals, insects, and fishes. Small values of $\nu$
+($\lesssim 10$) suggest heavy-tailed dynamics; larger values of $\nu$ suggest
+approximately normal-tailed dynamics. Vertical points and line segments
+represent posterior medians and 50\% / 90\% credible intervals for individual
+populations. Inset plots show probability that $\nu < 10$ (probability of heavy
+tails) for populations arranged by taxonomic order. Taxonomic orders are sorted
+by decreasing mean p($\nu < 10$). Taxonomic orders with three or fewer
+populations in panel a are omitted for space. Red and orange colours highlight
+populations with a high probability of heavy-tailed dynamics.
 }
@@
 \caption{
-Potential correlates of heavy-tailed behaviour (indicated by a high probability that $\nu < 10$).
-Shown are (a--c) parameters from the Gompertz heavy-tailed population model, (b) number of time steps, (c) body length, and (d) lifespan.
-For the Gompertz parameters, $\sigma$ refers to the scale parameter of the Student-t process-noise distribution, $\lambda$ refers to the population growth rate at an abundance of one, $b$ refers to the density dependence parameter ($1$ is maximally density independent, $0$ is maximally density dependent, and $<0$ is inversely density dependent).
-%For parameter values, points and lines represent medians and 50\% credible intervals of the posterior.
-Circles representing a limited number of sharks, crustaceans, and gastropods are filled in white.
-Median and 90\% credible interval posterior predictions of a beta regression multilevel model are shown in panels a and d where there was a high probability the slope coefficient was different from zero (Supporting Material).
+Potential correlates of heavy-tailed behaviour (indicated by a high probability
+that $\nu < 10$). Shown are (a--c) parameters from the Gompertz heavy-tailed
+population model, (b) number of time steps, (c) body length, and (d) lifespan.
+For the Gompertz parameters, $\sigma$ refers to the scale parameter of the
+Student-t process-noise distribution, $\lambda$ refers to the population growth
+rate at an abundance of one, $b$ refers to the density dependence parameter
+($1$ is maximally density independent, $0$ is maximally density dependent, and
+$<0$ is inversely density dependent).
+%For parameter values, points and lines represent medians and 50\% credible
+%intervals of the posterior.
+Circles representing a limited number of sharks, crustaceans, and gastropods
+are filled in white. Median and 90\% credible interval posterior predictions of
+a beta regression multilevel model are shown in panels a and d where there was
+a high probability the slope coefficient was different from zero (Supporting
+Material).
 }

### 2014-10-19 ms.tex
Formatting tweaks to paper

@@
 
-Black swans are statistically improbable events that nonetheless occur --- often
-with profound implications. While extremes in the physical environment,such as
-monsoons and heat waves, are widely studied and increasing in magnitude and
-frequency, it remains unclear the extent to which ecological populations buffer
-or suffer from extremes. Here, we develop a probability model to estimate the
-degree of heavy-tailedness (presence of black swans) in ecological process
-noise. We apply our model to \NPops~time series from around the world across
-\NOrders~taxonomic orders and seven classes. We find strong evidence of black
-swans, but they are rare, occurring in \overallMinPerc--\overallMaxPerc\%of
-populations; most frequently for birds (\AvesRangePerc\%) followed by mammals
-(\MammaliaRangePerc\%), insects (\InsectaRangePerc\%), and
-fishes(\OsteichthyesRangePerc\%). When they occur, they tend to be driven by
-climate and severe winters, cycles of parasites and predators, and interactions
-between these elements. They are more frequently detected for populations with
-longer time series and lower levels of process noise. We find little evidence
-of intrinsic life-history covariates. TODO close with modelling and policy
-implications
+Black swans are statistically improbable events that nonetheless occur ---
+often with profound implications. While extremes in the physical
+environment,such as monsoons and heat waves, are widely studied and increasing
+in magnitude and frequency, it remains unclear the extent to which ecological
+populations buffer or suffer from extremes. Here, we develop a probability
+model to estimate the degree of heavy-tailedness (presence of black swans) in
+ecological process noise. We apply our model to \NPops~time series from around
+the world across \NOrders~taxonomic orders and seven classes. We find strong
+evidence of black swans, but they are rare, occurring in
+\overallMinPerc--\overallMaxPerc\%of populations; most frequently for birds
+(\AvesRangePerc\%) followed by mammals (\MammaliaRangePerc\%), insects
+(\InsectaRangePerc\%), and fishes(\OsteichthyesRangePerc\%). When they occur,
+they tend to be driven by climate and severe winters, cycles of parasites and
+predators, and interactions between these elements. They are more frequently
+detected for populations with longer time series and lower levels of process
+noise. We find little evidence of intrinsic life-history covariates. TODO close
+with modelling and policy implications
 
@@
 time series, GPDD work found little evidence subsequent work suggested, but...
-Ward applied a data-intense method and found evidence for...
+Ward applied a data-intense method and found evidence for.
 

### 2014-10-19 ms.tex
Add some notes for last paragraph

@@
 the most striking black swans in ecology was the asteroid marking the mass
-extinctions of the K-T boundary (REF). Today, it is the extremeness of climate
+extinction of the K-T boundary (REF). Today, it is the extremeness of climate
 --- particularly in concert with shifts in mean temperature --- that is
@@
 Our results clarify, but do not disagree with previous related analyses.
-Previous work with an older version of the GPDD assessed heavy-tailed behaviour
-but focused on identifying if the lognormal distribution was the most
-frequently parsimonious model of population abundance \citep{halley2002}. For
-heavy-tailed distributions, \citet{halley2002} fit the extremely heavy-tailed
-Cauchy distribution and the four-parameter Levy stable distribution and found
-little information criteria support for these distributions in longer time
-series. However, black swan events are by definition rare, and the majority of
-time series in the GPDD are short for these purposes. Therefore, we would not
-expect to observe black swans in a large proportion of populations.
-Furthermore, by probabilistically quantifying process noise tail behaviour on
-a continuous spectrum, and allowing for non-stationary time series and density
-dependence, our analysis allows for a more nuanced description of the evidence
-of ecological black swans. In an earlier study, \citet{keitt1998} described
-heavy (power law) tails in breeding bird population abundance. However, this
-finding was challenged by \citet{allen2001}, who showed that mixing data across
-species could create falsely generate heavy tails. This rebuttal supports our
-use of the t distribution to represent heavy tails, since a t distribution can
-be represented by a mixture of normal distributions \citep[with
+Previous work with an older version of the GPDD assessed the distribution of
+abundance timeseries but focused on identifying if the lognormal distribution
+was the most frequently parsimonious model \citep{halley2002}. For heavy-tailed
+distributions, \citet{halley2002} fit the extremely heavy-tailed Cauchy
+distribution and the four-parameter Levy stable distribution and found little
+information criteria support for these distributions in longer time series.
+However, black swan events are by definition rare, and the majority of time
+series in the GPDD are short for these purposes. Therefore, we would not expect
+to observe black swans in a large proportion of populations. Furthermore, by
+probabilistically quantifying process noise tail behaviour on a continuous
+spectrum, and allowing for non-stationary time series and density dependence,
+our analysis allows for a more nuanced description of the evidence of
+ecological black swans. In an earlier study, \citet{keitt1998} described heavy
+(power law) tails in breeding bird population abundance. However, this finding
+was challenged by \citet{allen2001}, who showed that mixing data across species
+could create falsely generate heavy tails. This rebuttal supports our use of
+the t distribution to represent heavy tails, since a t distribution can be
+represented by a mixture of normal distributions \citep[with
 inverse-gamma-distributed variances,][]{gelman2014}.
@@
 Policy implications
+
+unlikely to forecast black swans, at least with any useful advanced warning
+but we might be able to determine probabilities
+and develop policy that is robust to, encourages rapid detection, and that is flexible enought to respond quickly to black swans
+
+example of earch quakes - spatial planning, early warning systems, education systems, a general idea about the probability in given places, build buildings that are robust
+
+Suggests that it's important to have policy that is robust to the uncertainty
+about black swans and that can respond and adapt rapidly following black swans
+
+foot and mouth disease and earthquakes?
 
@@
 tails) for populations arranged by taxonomic order. Taxonomic orders are sorted
-by decreasing mean p($\nu < 10$). Taxonomic orders with three or fewer
+by decreasing mean Pr($\nu < 10$). Taxonomic orders with three or fewer
 populations in panel a are omitted for space. Red and orange colours highlight

### 2014-10-19 text/sparks.tex
Arrange files into analysis and text folders

--- 2014-10-02 sparks.tex
@@
 
-\caption{Example population dynamic black swans from the Global Population Dynamics Database and descriptions of their causes.
-Red and blue dots indicate downward and upward events that have a probability of occurring of 0.001\% or less if the population dynamics were explained by a Gompertz model with normally distributed process noise.
-These populations are a non-random sample that we were able to verify in the primary literature.}\\
+\caption{Example population dynamic black swans from the Global Population
+  Dynamics Database and descriptions of their causes. Red and blue dots
+  indicate downward and upward events that have a probability of occurring of
+  0.001\% or less if the population dynamics were explained by a Gompertz model
+  with normally distributed process noise. These populations are a non-random
+  sample that we were able to verify in the primary literature.}\\
 
@@
 
-\includegraphics[width=2cm]{sparks/6528} &
+\includegraphics[width=2cm]{../analysis/sparks/6528} &
 Shag,
@@
 UK &
-Shortage of nest sites reduced productivity; red-tide event in 1968 caused extreme mortality; no longer a nest shortage; population rapidly increased &
+Shortage of nest sites reduced productivity; red-tide event in 1968 caused
+extreme mortality; no longer a nest shortage; population rapidly increased &
 \citep{potts1980}\\
 
-\includegraphics[width=2cm]{sparks/10007} &
+\includegraphics[width=2cm]{../analysis/sparks/10007} &
 Water vole,
@@
 UK &
-Short-term population cycles from predator interactions combined with longterm environmental cycle caused sharp downswing  &
+Short-term population cycles from predator interactions combined with longterm
+environmental cycle caused sharp downswing  &
 \citep{saucy1994}\\
 
-\includegraphics[width=2cm]{sparks/7115} &
+\includegraphics[width=2cm]{../analysis/sparks/7115} &
 Fur seal,
@@
 South Africa &
-Strong decreases in harvesting, loss of predators, and diamond mining regulations reducing human traffic caused sharp upswings  &
+Strong decreases in harvesting, loss of predators, and diamond mining
+regulations reducing human traffic caused sharp upswings  &
 \citep{shaughnessy1982}\\
 
-\includegraphics[width=2cm]{sparks/10113} &
+\includegraphics[width=2cm]{../analysis/sparks/10113} &
 Willow grouse,
@@
 
-\includegraphics[width=2cm]{sparks/10162} &
+\includegraphics[width=2cm]{../analysis/sparks/10162} &
 Red grouse,
@@
 UK &
-Good environmental conditions produced high numbers and vulnerable populations; bad conditions and overcrowding combined to create crashes  &
+Good environmental conditions produced high numbers and vulnerable populations;
+bad conditions and overcrowding combined to create crashes  &
 \citep{mackenzie1952}\\
 
-\includegraphics[width=2cm]{sparks/1235} &
+\includegraphics[width=2cm]{../analysis/sparks/1235} &
 Wren,
@@
 
-\includegraphics[width=2cm]{sparks/20579} &
+\includegraphics[width=2cm]{../analysis/sparks/20579} &
 Grey heron,
@@
 UK &
-Severe winters in 1929, 1940--1942, and 1962--1963; 1963 event so severe that recovery took three times as long as expected &
+Severe winters in 1929, 1940--1942, and 1962--1963; 1963 event so severe that
+recovery took three times as long as expected &
 \citep{stafford1971} \\
 
-%\includegraphics[width=2cm]{sparks/20580} &
+%\includegraphics[width=2cm]{../analysis/sparks/20580} &
 %Chamois, \textit{Rupicapra rupicapra}, Switzerland &
@@
 
-%\includegraphics[width=2cm]{sparks/5019} &
+%\includegraphics[width=2cm]{../analysis/sparks/5019} &
 %Barbary macaque,  &
@@
 
-%\includegraphics[width=2cm]{sparks/9675} &
+%\includegraphics[width=2cm]{../analysis/sparks/9675} &
 %Carrot fly (\textit{Psila rosae}, Finland &
@@
 
-%\includegraphics[width=2cm]{sparks/10139} &
+%\includegraphics[width=2cm]{../analysis/sparks/10139} &
 %Grey heron (\textit{Ardea cinerea}), UK &

### 2014-10-19 text/supp.tex
Arrange files into analysis and text folders

--- 2014-10-18 supp.tex
@@
 \bibliography{/Users/seananderson/Dropbox/tex/jshort,/Users/seananderson/Dropbox/tex/ref3}
+
+% ------------------------------
+% Supplemental Tables
+% ------------------------------
+
+\clearpage
+\renewcommand{\thetable}{S\arabic{table}}
+\setcounter{table}{0}
+\input{../analysis/stat-table}
+\clearpage
+
+% ------------------------------
+% Supplemental Figures
+% ------------------------------
+
+\renewcommand{\thefigure}{S\arabic{figure}}
+\setcounter{figure}{0}
+
+%\begin{figure}[htbp]
+%\begin{center}
+%\includegraphics[width=\textwidth]{../analysis/log10-assumed.pdf}
+%\caption{Populations in which the abundance was specified as or assumed to be
+%recorded as log10 values. The numbers at the front of each panel label are the
+%GPDD IDs.}
+    %\label{fig:log10-assumed}
+%\end{center}
+%\end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=\textwidth]{../analysis/all-clean-ts-3.pdf}
+\caption{
+  All filtered times series used in our analysis. The abundances are shown on
+  a log10 vertical axis. Colours indicate taxonomic classes. Open black circles
+  indicate abundance values that were interpolated. Closed black circles
+  indicate single abundance values that were recorded as zero but were set to
+  the next lowest observed abundance.
+}
+    \label{fig:all-ts}
+\end{center}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=\textwidth]{../analysis/priors-gomp-base.pdf}
+\caption{
+  Probability density of the Bayesian priors for the Gompertz models. From left
+  to right: (1) per capita growth rate at $\ln$(abundance) = $0$: $\lambda \sim
+  \mathrm{Normal}(0, 10^2)$; (2) scale parameter of t-distribution process
+  noise: $\sigma \sim \mathrm{Half\mhyphen Cauchy} (0,
+  2.5)$; (3) t-distribution degrees of freedom parameter: $\nu \sim
+  \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2)$; (4) AR1
+  correlation coefficient of residuals: $\phi \sim \mathrm{Truncated \mhyphen
+    Normal}(0, 1, \mathrm{min.} = -1, \mathrm{max.} = 1)$. Not shown is $b$,
+  the density dependence parameter: $b \sim \mathrm{Uniform}(-1, 2)$.
+}
+  \label{fig:priors}
+\end{center}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=0.8\textwidth]{../analysis/t-dist-sampling-sim-prior-exp0point01.pdf}
+\includegraphics[width=0.8\textwidth]{../analysis/t-dist-sampling-sim-sigma-prior-exp0point01.pdf}
+\caption{
+  Testing the ability to estimate $\nu$ (top panels) and the scale parameter of
+  the process error (bottom panels) for a given number of samples (columns)
+  drawn from a distribution with a given true $\nu$ value (rows). The red lines
+  indicate the true population value. When a small number of samples are drawn
+  there may not be samples sufficiently far into the tails to recapture the
+  true $\nu$ value; however, heavy tails are still distinguished from normal
+  tails in most cases, even with only 25 or 50 samples. TODO switch to nu
+  = black swan, 3, 5, and normal.
+}
+\label{fig:sim-nu}
+\end{center}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=\textwidth]{../analysis/sim-gompertz.pdf}
+\caption{
+  Simulation testing the Gompertz estimation model when the process error draws
+  are chosen so that $\nu$ can be estimated close to the true value outside the
+  full population model (``effective $\nu$'' within a CV of 0.2 of specified
+  $\nu$). The simulation was run across population $\nu$ values (columns) and
+  different scenarios (colours): (1) 100 time steps and no observation error,
+  (2) 50 time steps and no observation error, (3) 50 time steps and observation
+  error drawn from $\mathrm{Normal} (0, 0.2^2)$ but ignored, and (4) 50 time
+  steps with observation error in which the quantity of observation error was
+  assumed known. Dashed horizontal lines show the true population values; these
+  true values were chosen to represent approximately the median values as
+  estimated from the GPDD. Individual dots and lines represent a stochastic
+  draws from the true population distribution and a model fitting (median and
+  80\% credible intervals).
+  %The panels show true values of $\nu$ =  3, 5, normal (very heavy, heavy, and
+  %not heavy tailed).
+  The panels from top to bottom show $1/\nu$, process noise scale parameter
+  $\sigma$, growth rate parameter $\lambda$, and the density dependence
+  parameter $b$.
+}
+\label{fig:sim-gompertz}
+\end{center}
+\end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=\textwidth]{../analysis/sim-gompertz-boxplots.pdf}
+\caption{
+  The distribution of median estimates from the same simulation as described in
+  Fig.~\ref{fig:sim-gompertz}. The boxes show the interquartile range. The
+  whiskers extend to $1.5$ times the interquartile range and outliers are shown
+  as dots.
+}
+\label{fig:sim-gompertz-boxplots}
+\end{center}
+\end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=\textwidth]{../analysis/check-sim-box.pdf}
+\caption{
+  The probability that $\nu < 10$ (i.e.\ the approximate probability of heavy
+  tails) for the same simulation scenarios as shown in
+  Fig.~\ref{fig:sim-gompertz} and Fig.~\ref{fig:sim-gompertz-boxplots}. Within
+  each scenario the dots represent stochastic draws from the true population
+  distributions combined with model fits.
+}
+\label{fig:sim-prob}
+\end{center}
+\end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=0.9\textwidth]{../analysis/ts-gpdd-heavy-eg-log10-base.pdf}
+\caption{
+  Time series for populations with p($\nu < 10$) $>$ 0.5 using the base
+  Gompertz population model. Panels are ordered by increasing p($\nu < 10$).
+  Vertical axes are on a log10 scale. Colours indicate taxonomic classes. TODO
+  add dots for upswings and downswings as in Table 1. The labels on each panel
+  indicate p($\nu < 10$) $>$ 0.5, the common name for the species, and the GPDD
+  ID number.
+}
+\label{fig:heavy-ts}
+\end{center}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=0.27\textwidth]{../analysis/order-level-estimates.pdf}
+\caption{
+  Taxonomic-order-level posterior estimates of probability that $\nu < 10$ from
+  a beta regression multilevel model accounting for the effect of time-series
+  length. Estimates are at the geometric mean of time series length across all
+  the data (approximately 27 time steps). Colour shading refers to the
+  taxonomic classes as indicated in Fig.~\ref{fig:nu-coefs} and
+  Fig.~\ref{fig:correlates} (yellow for fishes, green for insects, purple for
+  birds, and red for mammals). Short vertical line segments indicate median
+  posterior estimates. Long dotted vertical line throughout the figure
+  indicates the median expected Pr($\nu < 10$) from the prior distribution.
+}
+\label{fig:order-estimates}
+\end{center}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=\textwidth]{../analysis/gomp-comparison.pdf}
+\caption{
+  Estimates of $\nu$ from alternative models plotted against the base Gompertz
+  model estimates of $\nu$. Shown are medians of the posterior (dots) and 50\%
+  credible intervals (segments). The diagonal line indicates a one-to-one
+  relationship. Different colours indicate various taxonomic classes. The
+  grey-shaded regions indicate regions of disagreement if $\nu = 10$ is taken
+  as a threshold of heavy-tailed dynamics. The Gompertz observation error model
+  assumes a fixed standard deviation of observation error of $0.2$ on the log
+  scale.
+}
+\label{fig:alt}
+\end{center}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[htbp]
+\begin{center}
+%\includegraphics[width=0.7\textwidth]{../analysis/admb-coefs}
+%\includegraphics[width=0.8\textwidth]{../analysis/length-vs-prob-tails}
+\includegraphics[width=0.5\textwidth]{../analysis/stan-beta-correlates}
+\caption{
+  %Potential correlates of heavy-tailed dynamics.
+  %(Top panels) Dots, thick lines, and thin lines represent fixed-effect estimates and their 50\% and 95\% confidence intervals for the covariates listed on the left: $\log \sigma$, $b$, $\lambda$, $\log$(lifespan), $\log$(data set length).
+  Main effect posterior densities for potential correlates of heavy-tailed
+  dynamics. The response variable was Pr($\nu < 10$) (approximately the
+  probability of heavy tails). The beta regression model was fit on a logit
+  scale with random intercepts for taxonomic class, taxonomic order, and
+  species. All covariates were standardized by subtracting their mean and
+  dividing by twice their standard deviation. Therefore, the coefficient values
+  represent the expected effect of a change in two standard deviations of the
+  raw input variable on the logit-transformed response variable.
+}
+    \label{fig:correlate-coefs}
+\end{center}
+\end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=0.70\textwidth]{../analysis/perc-prob-increase-with-n}
+\caption{
+Expected percent increase in the probability of detecting heavy tails, Pr($\nu
+< 10$), with an additional ten time steps of data. These values are based on
+the curve shown in Fig.~\ref{fig:correlates}d, which is based on the
+coefficient in Fig.~\ref{fig:correlate-coefs}. For example, we have about
+a 12\% increased probability of observing heavy tails in a time series that has
+60 time steps vs.\ a time series with 50 time steps.
+}
+\label{fig:perc-inc-p}
+\end{center}
+\end{figure}
+
+% \clearpage
+%
+% \begin{figure}[htbp]
+% \begin{center}
+% \includegraphics[width=0.8\textwidth]{../analysis/length-vs-prob-tails}
+% \caption{Model predictions. TODO}
+%     \label{fig:length-predictions}
+% \end{center}
+% \end{figure}
+
+\clearpage
+
+\noindent
+Example Stan code for a heavy-tailed Gompertz model with AR1 correlated
+residuals and a specified level of observation error:
+
+\begin{spacing}{1.15}
+\begin{footnotesize}
+\begin{verbatim}
+data {
+  int<lower=3> N;              // number of observations
+  vector[N] y;                 // vector to hold ln abundance observations
+  real<lower=0> nu_rate;       // rate parameter for nu exponential prior
+}
+parameters {
+  real lambda;                 // Gompertz growth rate parameter
+  real<lower=-1, upper=2> b;   // Gompertz density dependence parameter
+  real<lower=0> sigma_proc;    // process noise scale parameter
+  real<lower=2> nu;            // t-distribution degrees of freedom
+  real<lower=-1, upper=1> phi; // AR1 parameter
+  vector[N] U;                 // unobserved states
+  real<lower=0> sigma_obs;     // specified observation error SD
+}
+transformed parameters {
+  vector[N] epsilon;           // error terms
+  epsilon[1] <- 0;
+  for (i in 2:N) {
+    epsilon[i] <- U[i] - (lambda + b * U[i - 1])
+                       - (phi * epsilon[i - 1]);
+  }
+}
+model {
+  // priors:
+  nu ~ exponential(nu_rate);
+  lambda ~ normal(0, 10);
+  sigma_proc ~ cauchy(0, 2.5);
+  phi ~ normal(0, 1);
+  // data model:
+  for (i in 2:N) {
+    U[i] ~ student_t(nu,
+                     lambda + b * U[i - 1]
+                     + phi * epsilon[i - 1],
+                     sigma_proc);
+  }
+  y ~ normal(U, sigma_obs);
+}
+\end{verbatim}
+\end{footnotesize}
+
+\clearpage
+\noindent
+Stan code for the multilevel beta regression:
+\begin{footnotesize}
+\verbatiminput{../analysis/betareg4.stan}
+\end{footnotesize}
+
+\clearpage
+
+\noindent
+The GPDD IDs used in our analysis:
+
+\begin{footnotesize}
+\noindent
+{\tt
+\input{../analysis/mainids}
+}
+\end{footnotesize}
+\end{spacing}

### 2014-10-19 text/supp.tex
Clean up supp.tex
Edits to SOM

@@
 (Fig.~\ref{fig:didactic}) that time series of the length considered here are
-unlikely to be sufficiently informative about the precise value of $\nu$. In the
+unlikely to be informative about the precise value of $\nu$. In the
 scenario where the data are not heavy tailed (e.g.~Fig.~\ref{fig:didactic}e, h)
@@
 issues. (1) With smaller sample sizes, there may not be a stochastic draw from
-the tails of a distribution. In that case, no model, no matter how perfect
-a model, will be able to detect the shape of the tails. (2) Models (particularly
-complex models) may return biased parameter estimates if there are conceptual,
-computational, or coding errors in the models. Our first simulation tested the
-first issue and our second simulation tested the latter. In general, out
-simulations show that, if anything, our model under predicts the magnitude and
-frequency of heavy tailed events --- especially given the length of the time
-series in the GPDD.
+the tails of a distribution. In that case, no model, no matter how perfect,
+will be able to detect the shape of the tails. (2) Complex models may return
+biased parameter estimates if there are conceptual, computational, or coding
+errors. Our first simulation tested the first issue and our second simulation
+tested the latter. In general, out simulations show that, if anything, our
+model under predicts the magnitude and frequency of heavy tailed events ---
+especially given the length of the time series in the GPDD.
 
@@
 estimates of $\nu$ near the true value of $\nu$ with large samples ($\ge 200$)
-(Fig.~\ref{fig:sim-gompertz} upper panels). At smaller samples the we could
-still usually distinguish ``heavy'' from ``not-heavy'' tails, but the model
-tended to underestimate how heavy the tails were. At the same time, when the
-model underestimated how heavy-tailed the distribution was, it tended to
+(Fig.~\ref{fig:sim-gompertz} upper panels). At smaller samples we could still
+usually distinguish ``heavy'' from ``not-heavy'' tails, but the model tended to
+underestimate how heavy the tails were. At the same time, the model tended to
 overestimate how large the scale parameter was (Fig.~\ref{fig:sim-gompertz}
@@
 heavy-tailed Gompertz model to obtain unbiased parameter estimates given that
-process deviations were provided in which our simpler model (in the previous
+process deviations were provided in which our simpler model (previous
 section) could accurately estimate the value of $\nu$. To generate these process
-errors, we repeatedly generated candidate process deviations and estimated the
+deviations, we repeatedly generated candidate process deviations and estimated the
 central tendency, scale, and $\nu$ values each time. We recorded when
@@
 known to affect the ability to detect extreme events \citep{ward2007}. However,
-simultaneously estimating observation and process error in state-space is
-a challenging problem and is known to result in identifiability issues with the
-Gompertz population model \citep{knape2008}. Furthermore, our model estimates an
-additional parameter --- the shape of the process error distribution tails ---
+simultaneously estimating observation and process noise is a challenging
+problem and is known to result in identifiability issues with the Gompertz
+population model \citep{knape2008}. Furthermore, our model estimates an
+additional parameter --- the shape of the process deviation tails ---
 potentially making identifiability and computational issues even greater.
@@
 
-%\begin{figure}[htbp]
-%\begin{center}
-%\includegraphics[width=\textwidth]{../analysis/log10-assumed.pdf}
-%\caption{Populations in which the abundance was specified as or assumed to be
-%recorded as log10 values. The numbers at the front of each panel label are the
-%GPDD IDs.}
-    %\label{fig:log10-assumed}
-%\end{center}
-%\end{figure}
-
 \begin{figure}[htbp]
@@
   Testing the ability to estimate $\nu$ (top panels) and the scale parameter of
-  the process error (bottom panels) for a given number of samples (columns)
+  the process deviations (bottom panels) for a given number of samples (columns)
   drawn from a distribution with a given true $\nu$ value (rows). The red lines
@@
 \caption{
-  Simulation testing the Gompertz estimation model when the process error draws
-  are chosen so that $\nu$ can be estimated close to the true value outside the
-  full population model (``effective $\nu$'' within a CV of 0.2 of specified
-  $\nu$). The simulation was run across population $\nu$ values (columns) and
-  different scenarios (colours): (1) 100 time steps and no observation error,
-  (2) 50 time steps and no observation error, (3) 50 time steps and observation
-  error drawn from $\mathrm{Normal} (0, 0.2^2)$ but ignored, and (4) 50 time
-  steps with observation error in which the quantity of observation error was
-  assumed known. Dashed horizontal lines show the true population values; these
-  true values were chosen to represent approximately the median values as
-  estimated from the GPDD. Individual dots and lines represent a stochastic
-  draws from the true population distribution and a model fitting (median and
-  80\% credible intervals).
-  %The panels show true values of $\nu$ =  3, 5, normal (very heavy, heavy, and
-  %not heavy tailed).
-  The panels from top to bottom show $1/\nu$, process noise scale parameter
-  $\sigma$, growth rate parameter $\lambda$, and the density dependence
-  parameter $b$.
+  Simulation testing the Gompertz estimation model when the process deviation
+  draws were chosen so that $\nu$ can be estimated close to the true value
+  outside the full population model (``effective $\nu$'' within a CV of 0.2 of
+  specified $\nu$). The simulation was run across population $\nu$ values
+  (columns) and different scenarios (colours): (1) 100 time steps and no
+  observation error, (2) 50 time steps and no observation error, (3) 50 time
+  steps and observation error drawn from $\mathrm{Normal} (0, 0.2^2)$ but
+  ignored, and (4) 50 time steps with observation error in which the quantity
+  of observation error was assumed known. Dashed horizontal lines show the true
+  population values; these true values were chosen to represent approximately
+  the median values as estimated from the GPDD. Individual dots and lines
+  represent a stochastic draws from the true population distribution and
+  a model fitting (median and 80\% credible intervals). The panels from top to
+  bottom show $1/\nu$, process noise scale parameter $\sigma$, growth rate
+  parameter $\lambda$, and the density dependence parameter $b$.
 }
@@
 \begin{center}
-%\includegraphics[width=0.7\textwidth]{../analysis/admb-coefs}
-%\includegraphics[width=0.8\textwidth]{../analysis/length-vs-prob-tails}
 \includegraphics[width=0.5\textwidth]{../analysis/stan-beta-correlates}
 \caption{
-  %Potential correlates of heavy-tailed dynamics.
-  %(Top panels) Dots, thick lines, and thin lines represent fixed-effect estimates and their 50\% and 95\% confidence intervals for the covariates listed on the left: $\log \sigma$, $b$, $\lambda$, $\log$(lifespan), $\log$(data set length).
   Main effect posterior densities for potential correlates of heavy-tailed
@@
 \end{figure}
-
-% \clearpage
-%
-% \begin{figure}[htbp]
-% \begin{center}
-% \includegraphics[width=0.8\textwidth]{../analysis/length-vs-prob-tails}
-% \caption{Model predictions. TODO}
-%     \label{fig:length-predictions}
-% \end{center}
-% \end{figure}
 

### 2014-10-20 text/ms.tex
Arrange files into analysis and text folders
Develop introduction

--- 2014-10-19 ms.tex
@@
 
-Black swans are unexpected extreme events with dramatic consequences. One of
-the most striking black swans in ecology was the asteroid marking the mass
-extinction of the K-T boundary (REF). Today, it is the extremeness of climate
---- particularly in concert with shifts in mean temperature --- that is
-expected to cause the greatest societal damage (REF).
+Black swans are unexpected extreme events with dramatic consequences
+\citep{taleb2007,sornette2009}. One of the most striking black swans in ecology
+was the asteroid marking the mass extinction of the K-T boundary (REF). Today,
+it is the extremeness of climate, in concert with shifts in mean temperature,
+that is expected to cause the greatest societal damage \citep{ipcc2012}.
+However, whereas extremes in the physical environment such as wave height, rain
+fall, and temperature are present and widely accepted
+\citep{gaines1993,katz2005}, it remains unclear the extent to which ecological
+systems buffer or suffer from black swans \citep{nunez2012}.
 
-However, while extremes in the physical environment are present and widely
-accepted, it remains unclear the extent to which ecological systems buffer or
-suffer from black swans.
+Existing evidence for black swans or heavy tails in population dynamics is
+mostly anecdotal or for particular systems. For example, \citet{doak2008}
+reviewed a collection of field experiment population dynamic surprises and
+conclude that these are more common than we typically assume. \citet{nunez2012}
+compiled a list of major events on geologic and contemporary scales that could
+be considered ecological black swans. Other work has identified heavy-tail
+dynamics for particular systems. For example, \citet{segura2013} identified
+heavy tails in time series of marine microbe populations and \citet{drake2014}
+identified heavy tails in time to extinction for experimental waterflea
+populations. However, systematic evidence has remained elusive.
+\citet{keitt1998} tested for power-law tail distributions in breeding bird
+abundance, but later worked showed the apparent heavy tails were a result of
+mixing populations of different species \citep{allen2001}. A large scale
+analysis of the Global Population Dynamics Database (GPDD) by
+\citet{halley2002} found little evidence for heavy-tail distributions in longer
+abundance time series.
 
-Existing evidence for black swans or heavy tails in population dynamics is thus
-far limited. Considerable work has focussed on heavy tails in physical
-environmental processes in, for example, rain fall (REF), wave height (REF),
-earth quake magnitude, climate temperature (REF). time to extinction, plankton
-time series, GPDD work found little evidence subsequent work suggested, but...
-Ward applied a data-intense method and found evidence for.
+%Considerable work has focussed on heavy tails in physical environmental
+%processes in, for example, rain fall (REF), wave height (REF), earth quake
+%magnitude, climate temperature (REF). time to extinction, plankton time series,
+%GPDD work found little evidence subsequent work suggested, but... Ward applied
+%a data-intense method and found evidence for.
+
+%Ecological population dynamic models typically assume 'thin tailed'
+%distributions, such as the normal distribution on a multiplicative scale (e.g.
+%REF). We define black swan dynamics as cases where jumps in population
+%abundance from time step to time step. from heavy-tailed distributions.
 
 There are two key reasons why we may find little evidence of ecological black
-swans. First, they might not exist. Indeed, the majority of population
-dynamics model fitting and risk forecasting assumes that population dynamics
-are not heavy tailed (e.g.~REF, REF, REF). Alternatively, population dynamics
-might be heavy tailed, but a paucity of statistical methods and shortness of
-available time series has limited our ability to detect them. For example,
-a commonly applied probability distribution to detect the shape of time series
-tails (the generalized extreme value distribution) requires lumping data into
-the most extreme event per unit time (e.g.~heaviest rainfall per year) and
-therefore requires a considerable quantity of data (REF). Previous work into
-the shape of population dynamic tails, has focused on model comparison to
-identify parsimonious explanations from a set of particular hypotheses, but
-does not quantify the magnitude of heavy tails on a continuous scale or put the
-conclusions in a probabilistic framework (REF). Furthermore, information
-criteria are poor at distinguishing subtle differences in population dynamics
-for the relatively short time series that are most readily available (REF CHANGE
-THIS?).
-
-Or bring back in Ward - data intensive, can't be applied to most individual
-time series
-
-plankton --- hideous distribution to fit, even harder in population dynamics
-framework -and what about seasonal cycles, density dependence?
-
-GPDD - model competition and finding most parsimonious explanation from set of
-hypotheses --- but only \emph{very} heavy from not heavy
-
-GEV --- super data intensive
-
-tipping point and regime shift methods --- say something\ldots{} not
-appropriate for most of the data available
+swans. First, they might not exist. Indeed, the majority of population dynamics
+model fitting and risk forecasting assumes that population dynamics are not
+heavy tailed (e.g.~REF, REF, REF). Alternatively, dynamics might be heavy
+tailed, but a paucity of statistical methods for the frequently available short
+time series has limited our ability to detect them. For example, a commonly
+applied probability distribution to detect the shape of time series tails (the
+generalized extreme value distribution) requires lumping data into the most
+extreme event per unit time (e.g.~heaviest rainfall per year) and therefore
+requires a considerable quantity of data (REF). Previous large-scale work on
+the distribution of abundance trends focused on model comparison to identify
+parsimonious descriptions, but has not quantified the magnitude of heavy tails
+on a continuous scale (REF). Furthermore, information criteria are poor at
+distinguishing subtle differences in population dynamics for the relatively
+short time series that are most readily available (REF CHANGE THIS?).
+%bring in Ward - great method but data intensive, can't be applied to most
+%individual time series from the GPDD
+%plankton --- hideous distribution to fit, even harder in population dynamics
+%framework -and what about seasonal cycles, density dependence?
+%GPDD - model competition and finding most parsimonious explanation from set of
+%hypotheses --- but only \emph{very} heavy from not heavy
+%GEV --- super data intensive
+%tipping point and regime shift methods --- say something\ldots{} not
+%appropriate for most of the data available
 
@@
 
-Papers to work in: \citep{inchausti2002, halley2002, inchausti2001}
-\citep{jentsch2007} \citep{ward2007} \citep{garcia-carreras2011}
-\citep{sornette2009} \citep{nunez2012} \citep{thompson2013}
-\citep{beaugrand2012} \citep{pine-iii2009} \citep{doak2008} Extreme rain and
-monsoons over India increasing \citep{goswami2006} \citep{smale2013}
-\citep{easterling2000} \citep{scheffer2003} \citep{katz2005} \citep{taleb2007}
-\citep{vasseur2014} \citep{vert-pre2013} \citep{lindenmayer2010}
-\citep{valpine2002} \citep{gregory2010} \citet{garcia-carreras2011}
-\citeauthor{brook2006} \citetext{\citeyear{brook2006};
-  \citealp{herrandoprez2014}} \citep{sibly2005, ziebarth2010}
+% Papers to work in: \citep{inchausti2002, halley2002, inchausti2001}
+% \citep{jentsch2007} \citep{ward2007} \citep{garcia-carreras2011}
+% \citep{sornette2009} \citep{nunez2012} \citep{thompson2013}
+% \citep{beaugrand2012} \citep{pine-iii2009} \citep{doak2008} Extreme rain and
+% monsoons over India increasing \citep{goswami2006} \citep{smale2013}
+% \citep{easterling2000} \citep{scheffer2003} \citep{katz2005} \citep{taleb2007}
+% \citep{vasseur2014} \citep{vert-pre2013} \citep{lindenmayer2010}
+% \citep{valpine2002} \citep{gregory2010} \citet{garcia-carreras2011}
+% \citeauthor{brook2006} \citetext{\citeyear{brook2006};
+%   \citealp{herrandoprez2014}} \citep{sibly2005, ziebarth2010}
 
@@
 \begin{center}
-\includegraphics[width=\textwidth]{t-nu-eg.pdf}
+\includegraphics[width=\textwidth]{../analysis/t-nu-eg.pdf}
 \caption{
@@
 \begin{center}
-\includegraphics[width=0.44\textwidth]{nu-coefs-2.pdf}
+\includegraphics[width=0.44\textwidth]{../analysis/nu-coefs-2.pdf}
 \caption{
@@
 \begin{center}
-\includegraphics[width=0.9\textwidth]{correlates-p10.pdf}
+\includegraphics[width=0.9\textwidth]{../analysis/correlates-p10.pdf}
 \caption{

### 2014-10-20 text/ms.tex
Tweak intro paragraph

@@
 
-Existing evidence for black swans or heavy tails in population dynamics is
-mostly anecdotal or for particular systems. For example, \citet{doak2008}
-reviewed a collection of field experiment population dynamic surprises and
-conclude that these are more common than we typically assume. \citet{nunez2012}
-compiled a list of major events on geologic and contemporary scales that could
-be considered ecological black swans. Other work has identified heavy-tail
-dynamics for particular systems. For example, \citet{segura2013} identified
-heavy tails in time series of marine microbe populations and \citet{drake2014}
-identified heavy tails in time to extinction for experimental waterflea
-populations. However, systematic evidence has remained elusive.
-\citet{keitt1998} tested for power-law tail distributions in breeding bird
-abundance, but later worked showed the apparent heavy tails were a result of
-mixing populations of different species \citep{allen2001}. A large scale
-analysis of the Global Population Dynamics Database (GPDD) by
+%Existing evidence for black swans or heavy tails in population dynamics is
+%mostly anecdotal or for particular systems.
+There is compelling anecdotal evidence for ecological black swans, and
+quantitative evidence for a limited number of systems, but systematic evidence
+has been elusive. For example, \citet{doak2008} reviewed a collection of field
+experiment population dynamic surprises and conclude that these are more common
+than we typically assume. \citet{nunez2012} compiled a list of major events on
+geologic and contemporary scales that could be considered ecological black
+swans. Other work has identified heavy-tail dynamics for particular systems.
+For example, \citet{segura2013} identified heavy tails in time series of marine
+microbe populations and \citet{drake2014} identified heavy tails in time to
+extinction for experimental waterflea populations. However, systematic evidence
+has been limited. \citet{keitt1998} tested for power-law tail distributions in
+breeding bird abundance, but later worked showed the apparent heavy tails were
+a result of mixing populations of different species \citep{allen2001}. A later
+large scale analysis of the Global Population Dynamics Database (GPDD) by
 \citet{halley2002} found little evidence for heavy-tail distributions in longer

### 2014-10-20 text/ms.tex
More intro tweaks

@@
 There are two key reasons why we may find little evidence of ecological black
-swans. First, they might not exist. Indeed, the majority of population dynamics
-model fitting and risk forecasting assumes that population dynamics are not
-heavy tailed (e.g.~REF, REF, REF). Alternatively, dynamics might be heavy
-tailed, but a paucity of statistical methods for the frequently available short
-time series has limited our ability to detect them. For example, a commonly
-applied probability distribution to detect the shape of time series tails (the
-generalized extreme value distribution) requires lumping data into the most
-extreme event per unit time (e.g.~heaviest rainfall per year) and therefore
-requires a considerable quantity of data (REF). Previous large-scale work on
-the distribution of abundance trends focused on model comparison to identify
-parsimonious descriptions, but has not quantified the magnitude of heavy tails
-on a continuous scale (REF). Furthermore, information criteria are poor at
-distinguishing subtle differences in population dynamics for the relatively
-short time series that are most readily available (REF CHANGE THIS?).
+swans. First, they might not systematically exist. Indeed, the majority of
+population dynamics model fitting and risk forecasting assumes that population
+dynamics are not heavy tailed (e.g.~REF, REF, REF). Alternatively, dynamics
+might be heavy tailed, but a paucity of statistical methods for the frequently
+available short time series has limited our ability to detect them. For
+example, a commonly applied probability distribution to detect the shape of
+time series tails (the generalized extreme value distribution) requires lumping
+data into the most extreme event per unit time (e.g.~heaviest rainfall per
+year) and therefore requires a considerable quantity of data (REF). Previous
+large-scale work on the distribution of abundance trends focused on model
+comparison to identify parsimonious descriptions, but has not quantified the
+magnitude of heavy tails on a continuous scale (REF). Furthermore, information
+criteria are poor at distinguishing subtle differences in population dynamics
+for the relatively short time series that are most readily available (REF
+CHANGE THIS?).
 %bring in Ward - great method but data intensive, can't be applied to most
@@
 a high probability of ecological black swans across a wide range of taxa but in
-a limited number of populations. When they occurred, extreme climate,
-predation, parasites, and their interactions were common documented causes. We
-found little evidence that intrinsic life-history characteristics were related
-to the probability of black swans.
+a limited number of populations. When they occur, extreme climate, predation,
+parasites, and their interactions are common documented causes. We find little
+evidence that intrinsic life-history characteristics are related to the
+probability of black swans.
 

### 2014-10-20 text/ms.tex
Remove unnecessary word

@@
 
-Here we develop a probabilist framework for identifying ecological black swans
-in population dynamic process noise --- the stochastic jumps from time step to
+Here we develop a framework for identifying ecological black swans in
+population dynamic process noise --- the stochastic jumps from time step to
 time step. Our framework quantifies both the magnitude and probability of

### 2014-10-20 text/ms.tex
Add supp material note

@@
 
+\section{Supplementary Material}
+
+The following supplementary material is available online for this article:\\
+Figures S1--S11\\
+Table S1\\
+Example Stan code for a heavy-tailed Gompertz model and multilevel beta
+regression\\
+The GPDD IDs used in our analysis\\
+All code and data to recreate our analysis is available at:\\
+\url{https://github.com/seananderson/heavy-tails}
+
 \bibliographystyle{ecologyletters}

### 2014-10-22 text/ms.tex
Edits to main ms

@@
 quantitative evidence for a limited number of systems, but systematic evidence
-has been elusive. For example, \citet{doak2008} reviewed a collection of field
-experiment population dynamic surprises and conclude that these are more common
-than we typically assume. \citet{nunez2012} compiled a list of major events on
-geologic and contemporary scales that could be considered ecological black
-swans. Other work has identified heavy-tail dynamics for particular systems.
-For example, \citet{segura2013} identified heavy tails in time series of marine
-microbe populations and \citet{drake2014} identified heavy tails in time to
-extinction for experimental waterflea populations. However, systematic evidence
-has been limited. \citet{keitt1998} tested for power-law tail distributions in
-breeding bird abundance, but later worked showed the apparent heavy tails were
-a result of mixing populations of different species \citep{allen2001}. A later
-large scale analysis of the Global Population Dynamics Database (GPDD) by
+has been elusive. \citet{doak2008} reviewed a collection of field experiment
+population dynamic surprises and conclude that these are more common than we
+typically assume. \citet{nunez2012} compiled a list of major events on geologic
+and contemporary scales that could be considered ecological black swans. Other
+work has identified heavy-tail dynamics for particular systems. For example,
+\citet{segura2013} identified heavy tails in time series of marine microbe
+populations and \citet{drake2014} identified heavy tails in time to extinction
+for experimental waterflea populations. However, systematic evidence has been
+limited. \citet{keitt1998} tested for power-law tail distributions in breeding
+bird abundance, but later worked showed the apparent heavy tails were a result
+of mixing populations of different species \citep{allen2001}. A later large
+scale analysis of the Global Population Dynamics Database (GPDD) by
 \citet{halley2002} found little evidence for heavy-tail distributions in longer
@@
 
-Our main analysis focuses on the commonly applied phenomenological Gompertz
-population dynamics model. In previous analyses of the GPDD, the Gompertz has
-proved to be the most frequently parsimonious model \citep{brook2006}. We
-extended the traditional Gompertz model by allowing the process noise to be
-drawn from a Student-t distribution with degrees of freedom parameter $\nu$
-(Fig.~\ref{fig:didactic}). If $\nu$ is small ($\lesssim 10$) the t distribution
-has much heavier tails than a normal distribution. For example, at $\nu = 2$,
-the probability of drawing a value more than five standard deviations below the
-mean is $1.8$\%, whereas the probability of drawing such a value from a normal
-distribution is nearly zero ($2.9\cdot10^{-5}$\%). As $\nu$ approaches infinity
-the distribution approaches the normal distribution. By estimating the value of
+Our main analysis focuses on the commonly applied Gompertz population dynamics
+model (REFs?). The Gompertz model represents population growth as a linear
+function in $\log$ space. If we let $x_t$ represent the $\log$ abundance ($N$)
+at time $t$, we can represent the Gompertz model as:
+\begin{align*}
+x_t &= \lambda + b x_{t-1} + \epsilon_t\\
+\epsilon_t &\sim \mathrm{Normal}(0, \sigma^2).
+\end{align*}
+The parameter $\lambda$ represents the expected $\log$ number of new
+individuals at time $t + 1$ if $N_t = 1$. The model is density independent if
+$b = 1$, maximally density dependent if $b = 0$, and inversely density
+dependent if $b < 0$. The process noise $\epsilon_t$ is modelled as normally
+distributed with standard deviation $\sigma$. In previous analyses of the GPDD,
+the Gompertz has proved to be the most frequently parsimonious model
+\citep{brook2006}.
+
+We extended the standard Gompertz model by allowing the process noise to be
+drawn from a Student-t distribution with degrees of freedom parameter $\nu$, and scale parameter $\sigma$:
+\begin{align*}
+\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma).
+\end{align*}
+If $\nu$ is small ($\lesssim 10$), the t distribution has much heavier tails
+than a normal distribution. For example, at $\nu = 2$, the probability of
+drawing a value more than five standard deviations below the mean is $1.8$\%,
+whereas the probability of drawing such a value from a normal distribution is
+nearly zero ($2.9\cdot10^{-5}$\%). As $\nu$ approaches infinity the
+distribution approaches the normal distribution. By estimating the value of
 $\nu$ we can quantity how heavy-tailed the process noise deviations are
 (Fig.~\ref{fig:didactic}).
-
-The heavy-tailed Gompertz model is:
-\begin{align*}
-x_t &= \ln N_t\\
-x_t &= \lambda + b x_{t-1} + \epsilon_t\\
-\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma),
-\end{align*}
-where $x_t$ is the $\ln$ abundance ($N$) at time $t$. The model is density
-independent if $b = 1$, maximally density dependent if $b = 0$, and inversely
-density dependent if $b < 0$. The parameter $\lambda$ represents the expected
-$\ln$ number of new individuals at time $t + 1$ if $N_t = 1$ (per capita
-population growth rate). The process noise is modelled as a Student-t
-distribution with degrees of freedom parameter $\nu$, a mean of $0$, and
-a scale parameter of $\sigma$.
 
@@
 $\nu$, we chose an exponential prior with rate parameter of $0.01$ truncated at
-values above two (Fig.~\ref{fig:priors}), a slightly less informative prior
-than that suggested by \citet{fernandez1998}. This prior gives only a 7.7\%
+values above two (Fig.~\ref{fig:priors}) --- a slightly less informative prior
+than suggested by \citet{fernandez1998}. This prior gives only a 7.7\%
 probability that $\nu < 10$ but constrains the sampling sufficiently to avoid
@@
 We estimated all models in a Bayesian framework using Stan
-\citep{stan-manual2014}. Stan samples from the posterior distribution with
-a version of Hamiltonian Markov chain Monte Carlo called the No-U-Turn Sampler
-\citep{hoffman2014}. We assured that chains had sufficiently converged and the
-sampler had obtained sufficient independent samples from the posterior
+\citep[][version 2.4.0]{stan-manual2014} via the R computing environment \citep{r2014}. Stan
+samples from the posterior distribution with an adaptive version of
+Hamiltonian Markov chain Monte Carlo called the No-U-Turn Sampler and
+generally obtains less correlated chains than algorithms such as the Gibbs
+sampler \citep{hoffman2014}. We assured that chains had sufficiently converged
+and the sampler had obtained sufficient independent samples from the posterior
 ($\widehat{R} < 1.05$, $n_\mathrm{eff} > 200$; Supporting Material). We used
@@
 a model that modelled the serial correlation in the residuals. (2) Previous
-work has focussed modelled growth rates without accounting for density
-dependence \citep{segura2013}; therefore, we fit a simpler model in which we
-assumed density independence. (3) Observation error could bias parameter
-estimates \citep{knape2012} or mask our ability to detect heavy tails
-\citep{ward2007}. However, simultaneously estimating observation error and
-process noise in a state space framework is a challenging computational problem
-\citep[e.g.][]{dennis2006, knape2008} even with longer time series and without
-estimating the shape of the process error tails; therefore, we fit a model
-where we assumed a level of observation error ($0.2$ standard deviations on
-a log scale). (4) The Gompertz model assumes that population growth rate
-declines linearly with log abundance. We also fit an alternative model, the
-Ricker-logistic model, which assumes that population growth rate declines
-linearly with abundance itself (Supporting Material).
+work has modelled abundance or growth rates without accounting for density
+dependence \citep{halley2002,segura2013}; therefore, we fit a simpler model in
+which we assumed density independence. (3) Observation error could bias
+parameter estimates \citep{knape2012} or mask our ability to detect heavy tails
+\citep{ward2007}; therefore, we fit a model where we allowed for a fixed
+quantity of observation error ($0.2$ standard deviations on a log scale). (4)
+The Gompertz model assumes that population growth rate declines linearly with
+log abundance. We also fit an alternative model, the Ricker-logistic model,
+which assumes that population growth rate declines linearly with abundance
+itself (Supporting Material).
+
+% TODO - make sure this is in the supp? However, simultaneously estimating
+% observation error
+%and process noise in a state space framework is a challenging computational
+%problem \citep[e.g.][]{dennis2006, knape2008} even with longer time series and
+%without estimating the shape of the process error tails;
 
@@
 autocorrelation structure in the residuals, modelling population growth rates
-as a random walk with drift, or modelling the population dynamics as
+without density dependence, or modelling the population dynamics as
 Ricker-logistic did not systematically alter our conclusions. Allowing for
@@
 documented events were downward black swans and the majority involved
-a confluence of events. For example, a shortage of nest sites for the shag in
-the UK was thought to reduce population productivity and make the population
-vulnerable when a red-tide hit in 1968. The population experienced a black swan
-downswing, but the new availability of nest sites resulted in a rapid
-population upswing \citep{potts1980}. As another example, an interaction
-between population cycles caused by predation combined with cycles caused by
-the environment are thought to have caused a downward black swan for a water
+a confluence of events. For example, a shortage of nest sites for the shag
+long-necked bird in the UK was thought to reduce population productivity and
+make the population vulnerable when a red-tide hit in 1968 \citep{potts1980}.
+The population experienced a black swan downswing, but the new availability of
+nest sites resulted in a rapid population upswing \citep{potts1980}. As
+another example, an interaction between environmentally and predation-mediated
+population cycles are thought to have caused a downward black swan for a water
 vole population \citep{saucy1994}. Other black swans were the result of
@@
 in 1929, 1940--1942, and 1962--1963 were associated with black swan downswings
-in grey heron abundance in the UK \citep{stafford1971}. Our analysis finds that
-the last event was a combination of two black swan events in a row and
-population recovery took three times longer than predicted
+in grey heron abundance in the UK \citep{stafford1971}. Our analysis finds
+that the last event was a combination of two black swan events in a row and it
+took the population three times longer to recover than predicted
 \citep{stafford1971}.
@@
 estimating the tail shape of process error is a viable method of detecting
-heavy-tailed population dynamics. The presence of black swans highlights the
-importance of developing management strategies that detect quickly, respond to,
-and are robust to extremes in population dynamics --- particularly as the
+heavy-tailed population dynamics and if anything will underestimate the
+probability of black swans. The presence of black swans highlights the
+importance of developing management strategies that detect quickly, respond
+to, and are robust to extremes in population dynamics --- particularly as the
 frequency and magnitude of climatic extremes increase over the next century
@@
 aggregation of populations across space, or population diversity, or some
-intrinsic change in population variability through time. Extrinsic forces could
-also cause black swan dynamics (REF). These forces could be extreme themselves.
-For example, extreme climate, predation from or competition with other species
-experiencing black swans, or sharp changes in human pressure such hunting,
-fishing, or habitat destruction might cause black swans (REFS). Alternatively,
-the interaction of multiple ``normal'' extrinsic forces could give rise to
-black swan ecological dynamics. This could occur if the interaction nature is
-synergistic \citep[e.g.][]{kirby2009} or even if non-synergistic interactions
-experience a rare alignment \citep{denny2009}.
+intrinsic change in population variability through time. Extrinsic forces
+could also cause black swan dynamics \citep[e.g.][]{nunez2012}. These forces
+could be extreme themselves. For example, extreme climate, predation from or
+competition with other species experiencing black swans, or sharp changes in
+human pressure such hunting, fishing, or habitat destruction might cause black
+swans. Alternatively, the interaction of multiple ``normal'' extrinsic
+forces could give rise to black swan ecological dynamics. This could occur if
+the interaction nature is synergistic \citep[e.g.][]{kirby2009} or even if
+non-synergistic interactions experience a rare alignment \citep{denny2009}.
 
@@
 mistakes undoubtedly comprise some of the black swans we detected. Conversely,
-recording mistakes could mask heavy tails if initial observations assumed to be
-recording errors and are discarded or altered. Third, the temporal scale of
-observation and population dynamics vary considerably across populations in the
-GPDD and this likely influences the detection of heavy tails. For example, if
-we make frequent observations relative to generation time (e.g.~for many large
-bodied mammals) we will average across generations and perhaps not observe
+recording mistakes could mask heavy tails if initial observations are assumed
+to be recording errors and are discarded or altered. Third, the temporal scale
+of observation and population dynamics vary considerably across populations in
+the GPDD and this likely influences the detection of heavy tails. For example,
+if we make frequent observations relative to generation time (e.g.~for many
+large-bodied mammals) we will average across generations and perhaps miss
 black swans. Furthermore, if we census populations infrequently relative to
@@
 because these impact estimates of risk --- the probability of a specific
-magnitude of undesired event occurring (REF). It is possible that a comparable
-focus in ecology would alter our perception of ecological forecasts such as
-extinction risk.
+magnitude of undesired event occurring \citep{rachev2008}. It is possible that
+a comparable focus in ecology would alter our perception of ecological
+forecasts such as extinction risk.
 
@@
 unlikely to forecast black swans, at least with any useful advanced warning
-but we might be able to determine probabilities
-and develop policy that is robust to, encourages rapid detection, and that is flexible enought to respond quickly to black swans
+but we might be able to determine probabilities and develop policy that is
+robust to, encourages rapid detection, and that is flexible enough to respond
+quickly to black swans
 
-example of earch quakes - spatial planning, early warning systems, education systems, a general idea about the probability in given places, build buildings that are robust
+example of earth quakes - spatial planning, early warning systems, education
+systems, a general idea about the probability in given places, build buildings
+that are robust
 
@@
 
-We thank members of the Earth to Ocean research group for helpful discussions
-and comments. We are grateful to the contributors and maintainers of the Global
-Population Dynamics Database and to Compute Canada's WestGrid high-performance
-computing resources. Three silhouette images were obtained from
-\texttt{phylopic.org} under a Creative Commons Attribution 3.0 Unported
-license: rabbit (Sarah Werning), hoverfly (Gareth Monger), bird (Jean-Raphaël
-Guillaumin {[}photography{]} and T. Michael Keesey {[}vectorization{]}).
-Funding was provided by a Simon Fraser University Graduate Fellowship (SCA),
-the Natural Sciences and Engineering Research Council of Canada (NKD, ABC), the
-Canada Research Chairs Program (NKD), and XX-TODO (TAB).
+We thank Justin D. Yeakel, TODO, and other members of the Earth to Ocean
+research group for helpful discussions and comments. We are grateful to the
+contributors and maintainers of the Global Population Dynamics Database and to
+Compute Canada's WestGrid high-performance computing resources. Three
+silhouette images were obtained from \texttt{phylopic.org} under a Creative
+Commons Attribution 3.0 Unported license: rabbit (Sarah Werning), hoverfly
+(Gareth Monger), bird (Jean-Raphaël Guillaumin {[}photography{]} and T.
+Michael Keesey {[}vectorization{]}). Funding was provided by a Simon Fraser
+University Graduate Fellowship (SCA), the Natural Sciences and Engineering
+Research Council of Canada (NKD, ABC), the Canada Research Chairs Program
+(NKD), and TODO (TAB).
 
@@
 \caption{
-An illustration of fitting population dynamics models that allow for heavy
+An illustration of fitting population dynamic models that allow for heavy
 tails, represented by the Student-t degrees of freedom parameter $\nu$. (a, b)
@@
 represent posterior medians and 50\% / 90\% credible intervals for individual
-populations. Inset plots show probability that $\nu < 10$ (probability of heavy
-tails) for populations arranged by taxonomic order. Taxonomic orders are sorted
-by decreasing mean Pr($\nu < 10$). Taxonomic orders with three or fewer
-populations in panel a are omitted for space. Red and orange colours highlight
+populations. Inset plots show probability that $\nu < 10$ (probability of
+heavy tails) for populations arranged by taxonomic order and sorted by
+decreasing mean Pr($\nu < 10$). Taxonomic orders with three or fewer
+populations in panel a are omitted for space. Red to yellow points highlight
 populations with a high probability of heavy-tailed dynamics.
-}
-\label{fig:nu-coefs}
-\end{center}
-\end{figure}
+} \label{fig:nu-coefs} \end{center} \end{figure}
 
@@
 \caption{
-Potential correlates of heavy-tailed behaviour (indicated by a high probability
-that $\nu < 10$). Shown are (a--c) parameters from the Gompertz heavy-tailed
-population model, (b) number of time steps, (c) body length, and (d) lifespan.
-For the Gompertz parameters, $\sigma$ refers to the scale parameter of the
-Student-t process-noise distribution, $\lambda$ refers to the population growth
-rate at an abundance of one, $b$ refers to the density dependence parameter
-($1$ is maximally density independent, $0$ is maximally density dependent, and
-$<0$ is inversely density dependent).
+Potential covariates of heavy-tailed population dynamics (indicated by a high
+probability that $\nu < 10$). Shown are (a--c) parameters from the Gompertz
+heavy-tailed population model, (b) number of time steps, (c) body length, and
+(d) lifespan. For the Gompertz parameters, $\sigma$ refers to the scale
+parameter of the Student-t process-noise distribution, $\lambda$ refers to the
+expected log abundance at the next time step at an abundance of one, $b$
+refers to the density dependence parameter ($1$ is maximally density
+independent, $0$ is maximally density dependent, and $<0$ is inversely density
+dependent).
 %For parameter values, points and lines represent medians and 50\% credible

### 2014-10-22 text/sparks.tex
Edits to spark lines table

@@
 \caption{Example population dynamic black swans from the Global Population
-  Dynamics Database and descriptions of their causes. Red and blue dots
-  indicate downward and upward events that have a probability of occurring of
-  0.001\% or less if the population dynamics were explained by a Gompertz model
+  Dynamics Database and a description of their causes. Red and blue dots
+  indicate downward and upward events that have a 0.001\% probability or less of
+  occurring if the population dynamics were explained by a Gompertz model
   with normally distributed process noise. These populations are a non-random
-  sample that we were able to verify in the primary literature.}\\
+  sample that we verified in the primary literature.}\\
 
@@
 Severe winters in 1929, 1940--1942, and 1962--1963; 1963 event so severe that
-recovery took three times as long as expected &
+recovery took three times loner than expected &
 \citep{stafford1971} \\

### 2014-10-22 text/supp.tex
Remove extra space in SOM
Edits to SOM

@@
   caused by emigration. Regardless, our population models were fitted on
-  a multiplicative (log) scale and so could not account for zero abundance.
+  a multiplicative (log) scale and so could not account for zero abundance. To
+  avoid distorting the original data too strongly, we removed populations in
+  which we filled in more than four zeros.
+
+\item We removed all populations without at least four unique values
+  \citep[following][]{brook2006a}.
 
@@
   observations.
-
-\item We removed all populations without at least four unique values
-  \citep[following][]{brook2006a}.
 
@@
 \item Finally, we removed GPDD Main ID \texttt{20531}, which we noticed was
-  a duplicate of \texttt{10139} (a heron population).
+  a duplicate of \texttt{10139} (a heron population). We removed a limited
+  number of populations from class Angiospermopsida and Bacillariophyceae to
+  focus the taxonomy in our analysis. We also removed any populations with an
+  \texttt{Unknown} taxonomic class.
 
@@
 Our prior on $b$ was uninformative between values of $-1$ and $2$. We would not
-expect values of $b$ with levels of inverse density dependence as low as $-1$,
-nor would we expect values above $1$. We allowed values of $b$ above $1$ to
-allow for somewhat non-stationary time series of growth rates. The estimates of
-$b$ were well within these bounds. Our prior on $\lambda$ was very weakly
-informative within the range of expected values for population growth and is
-similar to priors suggested by \citet{gelman2008d} for intercepts of regression
-models. Our prior on $\sigma$ follows \citet{gelman2006c} and
-\citet{gelman2008d} and is based on our expected range of the value in nature
-from previous studies \citep[e.g.][]{connors2014}. In our testing of a subsample
-of populations, our parameter estimates were not qualitatively changed by
-switching to an uninformative uniform prior on $\sigma$, but the weakly
-informative prior substantially sped up chain convergence.
+expect values of $b$ with levels of density dependence as low as $-1$ (very
+strong inverse density dependence), nor would we generally expect values above $1$. We
+allowed values of $b$ above $1$ to allow for somewhat non-stationary time
+series of growth rates. The estimates of $b$ were well within these bounds. Our
+prior on $\lambda$ was very weakly informative within the range of expected
+values for population growth and is similar to the default priors suggested by
+\citet{gelman2008d} for intercepts of regression models. Our prior on $\sigma$
+follows \citet{gelman2006c} and \citet{gelman2008d} and is based on our
+expected range of the value in nature from previous studies
+\citep[e.g.][]{connors2014}. In our testing of a subsample of populations, our
+parameter estimates were not qualitatively changed by switching to an
+uninformative uniform prior on $\sigma$, but the weakly informative Half-Cauchy prior
+substantially sped up chain convergence.
 
-Our prior on $\nu$ was based on \citet{fernandez1998}; they chose a more
+Our prior on $\nu$ was based on \citet{fernandez1998}. They chose a more
 informative exponential rate parameter of $0.1$. We chose a less informative
@@
 scenario where the data are not heavy tailed (e.g.~Fig.~\ref{fig:didactic}e, h)
-the posterior will approximately match the prior (median $= 71$, mean $= 102$)
+the posterior will approximately match the prior (prior median $= 71$, mean $= 102$)
 and not be flagged as likely heavy tailed using the metrics we used in our paper
@@
 \citep{r2014}. We began with four chains and $2000$ iterations, discarding the
-first $1000$ as warm up (i.e.~4000 total samples). If $\hat{R}$ (a measure of
-chain convergence) was greater than $1.05$ for any parameter or the minimum
-effective sample size, $n_\mathrm{eff}$, (a measure of the effective number of
-uncorrelated samples) for any parameter was less than $200$, we doubled both the
-total iterations and warm up period and sampled from the model again. These
-thresholds are in excess of the minimums recommended by \citet{gelman2006a} of
-$\hat{R} < 1.1$ and effective sample size $> 100$ for reliable point estimates
-and confidence intervals. In the majority of cases our minimum thresholds were
-greatly exceeded. We continued this procedure up to $8000$ iterations ($16000$
-total samples) by which all chains were deemed to have sufficiently converged.
+first $1000$ as warm up (i.e.~4000 total samples). If $\hat{R}$ (the scale
+reduction factor --- a measure of chain convergence) was greater than $1.05$
+for any parameter or the minimum effective sample size, $n_\mathrm{eff}$, (a
+measure of the effective number of uncorrelated samples) for any parameter was
+less than $200$, we doubled both the total iterations and warm up period and
+sampled from the model again. These thresholds are in excess of the minimums
+recommended by \citet{gelman2006a} of $\hat{R} < 1.1$ and effective sample size
+$> 100$ for reliable point estimates and confidence intervals. In the majority
+of cases our minimum thresholds were greatly exceeded. We continued this
+procedure up to $8000$ iterations ($16000$ total samples) by which all chains
+were deemed to have sufficiently converged. These chain lengths may seem low to
+those familiar with software such as WinBUGS or JAGS, but the No-U-Turn
+Hamiltonian Markov chain Monte Carlo Sampler in Stan generally requires far
+fewer iterations to obtain equivalent effective sample sizes \citep{stan-manual2014}.
 
@@
 Student-t distribution $\nu$ parameter could be recovered given different true
-values of $\nu$ and sample sizes. Second, we tested the ability of the
+values of $\nu$ and different sample sizes. Second, we tested the ability of the
 heavy-tailed Gompertz model to obtain unbiased parameter estimates of $\nu$
-given that a set of process deviations was provided in which the ``effective
-$\nu$'' value was close to the true $\nu$ value.
+given that a set of process deviations was provided in which the effective
+$\nu$ value was close to the true $\nu$ value.
 
@@
 First, we tested the ability to estimate $\nu$ given different true values of
-$\nu$ and sample sizes. We took stochastic draws from t distributions with
+$\nu$ and different sample sizes. We took stochastic draws from t distributions with
 different $\nu$ values ($\nu = 3, 5, 10,$ and $10^6$ [$\approx$ normal]), with
@@
 (Fig.~\ref{fig:sim-gompertz} upper panels). At smaller samples we could still
-usually distinguish ``heavy'' from ``not-heavy'' tails, but the model tended to
-underestimate how heavy the tails were. At the same time, the model tended to
-overestimate how large the scale parameter was (Fig.~\ref{fig:sim-gompertz}
-lower panels).
+usually qualitatively distinguish heavy from not-heavy tails, but the
+model tended to underestimate how heavy the tails were. At the same time, the
+model tended to overestimate how large the scale parameter was
+(Fig.~\ref{fig:sim-gompertz} lower panels).
 
@@
 section) could accurately estimate the value of $\nu$. To generate these process
-deviations, we repeatedly generated candidate process deviations and estimated the
+deviations, we repeatedly drew proposed candidate process deviations and estimated the
 central tendency, scale, and $\nu$ values each time. We recorded when
-$\hat{\nu}$ was within $0.2$ CVs (coefficient of variations) of the true $\nu$
+$\hat{\nu}$ (median of the posterior) was within $0.2$ CVs (coefficient of
+variations) of the true $\nu$
 value and used this set of random seed values in our Gompertz simulation. We
@@
 observation error is a known phenomenon \citep{knape2012}. Fitting a model with
-correctly specified observation error made marginal improvements to model bias
-(purple symbols in Figs~\ref{fig:sim-gompertz} and
+correctly specified observation error made marginal improvements to model
+bias (purple symbols in Figs~\ref{fig:sim-gompertz} and
 \ref{fig:sim-gompertz-boxplots}).
 
-When converting the posterior distributions of $\nu$ into Pr($\nu < 10$) the
-models performed well. Without observation error, and using a probability of
-$0.5$ as a threshold, the model would have miscategorized only one of $40$
-simulations at $\nu = 5$ across a sample size of $50$ or $100$ data points
-(Fig.~\ref{fig:sim-prob}, upper panels). The model would have correctly
-categorized all cases where the process noise was not heavy tailed (indicated as
-``$\nu =$ infinity'' in Fig.~\ref{fig:sim-prob}) and all cases where $\nu = 3$.
-With $0.2$ standard deviations of observation error, the model still categorized
-XX\% of cases as heavy tailed when $\nu = 5$ and all cases where $\nu = 3$.
-Allowing for observation error made little improvement to the detection of heavy
-tails (Fig.~\ref{fig:sim-prob}, lower-left vs.\ lower-right panel). Therefore,
-we chose to focus on the simpler model without observation error in the main
-text.
+When converting the posterior distributions of $\nu$ into Pr($\nu < 10$), the
+models distinguished heavy and not-heavy tails reasonably well. Without
+observation error, and using a probability of $0.5$ as a threshold, the model
+would have miscategorized only one of $40$ simulations at $\nu = 5$ across
+a sample size of $50$ or $100$ data points (Fig.~\ref{fig:sim-prob}, upper
+panels). The model would have correctly categorized all cases where the process
+noise was not heavy tailed (indicated as ``$\nu =$ infinity'' in
+Fig.~\ref{fig:sim-prob}) and all cases where $\nu = 3$. With $0.2$ standard
+deviations of observation error, the model still categorized XX\% of cases as
+heavy tailed when $\nu = 5$ and all cases where $\nu = 3$. Allowing for
+observation error made little improvement to the detection of heavy tails
+(Fig.~\ref{fig:sim-prob}, lower-left vs.\ lower-right panel). Therefore, we
+chose to focus on the simpler model without observation error in the main
+text, particularly given that the true magnitude of observation error was
+unknown in the empirical data.
 
@@
 autocorrelation in the residuals, assumed no density dependence, allowed for
-observation error, or assumed a Ricker-logistic functional form. The ranges of
-percentages of black swans by taxonomic class cited in the abstract are based on
-lower and upper limits across our main Gompertz model and these four alternative
-models.
+observation error, or assumed a Ricker-logistic functional form. The range of
+percentages of black swans by taxonomic class cited in the abstract and results
+are based on lower and upper limits across our main Gompertz model and these
+four alternative models.
 
@@
 dependence parameter $b$ was fixed at $1$ (density independent). This is
-equivalent to fitting a random walk model (with drift) to the $\ln$ abundances
-or assuming the growth rates are stationary. The model was as follows:
-
+equivalent to fitting a random walk model (with drift) to the $\log$ abundances
+or assuming the growth rates are drawn from a stationary distribution. The
+model was as follows:
 \begin{align*}
 x_t &= \lambda + x_{t-1} + \epsilon_t\\
-\epsilon &\sim \mathrm{Student\mhyphen t}(\nu, \epsilon, \sigma).
+\epsilon &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma).
 \end{align*}
@@
 U_t &= \lambda + b U_{t-1} + \epsilon_t\\
-x_t &\sim \mathrm{Normal}(U_t, \sigma_\mathrm{obs})\\
+x_t &\sim \mathrm{Normal}(U_t, \sigma_\mathrm{obs}^2)\\
 \epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma_\mathrm{proc}),
 \end{align*}
-where $U$ represents the unobserved state vector, and $\sigma_\mathrm{obs}$
-represents the standard deviation of observation error (on a log scale). We set
-$\sigma_\mathrm{obs}$ to $0.2$, which represents the upper end of values often
+where $U$ represents the unobserved state vector, $\sigma_\mathrm{obs}$
+represents the standard deviation of observation error (on a log scale), and
+$\sigma_\mathrm{proc}$ represent the process noise scale parameter. We set
+$\sigma_\mathrm{obs}$ to $0.2$, which represents the upper limit of values often
 used in simulation analyses \citep[e.g.][]{valpine2002, thorson2014b}.
@@
 \end{align*}
-where $r_\mathrm{max}$ represents the maximum population growth rate that is
-obtained when $N$ (abundance) $= 0$. The parameter $K$ represents the carrying
-capacity and, as before, $x_t$ represents the $\ln$ transformed abundance at
-time $t$. The Ricker-logistic model assumes a linear decrease in population
-growth rate ($x_t - x_{t-1}$) with increases in abundance ($N_t$). In contrast,
-the Gompertz model assumes a linear decrease in population growth rate with
-increases in $\ln$ abundance ($x_t$) (REF).
+where $r_\mathrm{max}$ represents the theoretical maximum population growth
+rate that is obtained when $N$ (abundance) $= 0$. The parameter $K$ represents
+the carrying capacity and, as before, $x_t$ represents the $\log$ transformed
+abundance at time $t$. The Ricker-logistic model assumes a linear decrease in
+population growth rate ($x_t - x_{t-1}$) with increases in abundance ($N_t$).
+In contrast, the Gompertz model assumes a linear decrease in population growth
+rate with increases in \textit{log} abundance ($x_t$) (REF).
 
@@
 heavy tails, Pr($\nu < 10$), to investigate potential covariates of heavy-tailed
-dynamics. The beta distribution is useful when the response data range on
+dynamics. The beta distribution is useful when response data range on
 a continuous scale between zero and one. We used a logit link function as is
@@
 species $l$; $\phi_\mathrm{disp}$ represents the dispersion parameter; and $X_i$
-represents a vector of predictors for population $i$ with associated $\beta$
-parameters. The intercepts are allowed to vary from the overall intercept
-$\alpha$ by taxonomic classes ($\alpha^\mathrm{class}_j$), taxonomic orders
-($\alpha^\mathrm{order}_k$), and species ($\alpha^\mathrm{species}_l$) with
-standard deviations $\sigma_{\alpha \; \mathrm{class}}$, $\sigma_{\alpha \;
-  \mathrm{order}}$, and $\sigma_{\alpha \; \mathrm{species}}$. Where possible,
-we also allowed for error distributions around the predictors by incorporating
-the standard deviation of the posterior samples for the Gompertz parameters
-$\lambda$, $b$, and $\sigma$ around the mean point value (not shown in the above
-equation). We log transformed $\sigma$, time-series length, and lifespan to
-match the way they are visually represented in Fig.~\ref{fig:correlates}. All
+represents a vector of predictors (such as lifespan) for population $i$ with
+associated $\beta$ coefficients. The intercepts are allowed to vary from the
+overall intercept $\alpha$ by taxonomic class ($\alpha^\mathrm{class}_j$),
+taxonomic order ($\alpha^\mathrm{order}_k$), and species
+($\alpha^\mathrm{species}_l$) with standard deviations $\sigma_{\alpha \;
+  \mathrm{class}}$, $\sigma_{\alpha \; \mathrm{order}}$, and $\sigma_{\alpha \;
+  \mathrm{species}}$. Where possible, we also allowed for error distributions
+around the predictors by incorporating the standard deviation of the posterior
+samples for the Gompertz parameters $\lambda$, $b$, and $\sigma$ around the
+mean point value as normal distributions (not shown in the above equation). We
+log transformed $\sigma$, time-series length, and lifespan to match the way
+they are visually represented in Fig.~\ref{fig:correlates} and to make the
+relationship approximately linear on the logit-transformed response scale. All
 input variables were standardized by subtracting their mean and dividing by two
-standard deviations \citep{gelman2008c} to make their coefficients comparable in
-magnitude.
+standard deviations \citep{gelman2008c} to make their coefficients comparable
+in magnitude.
 
@@
 length as a predictor. (In this case, we did not want to control for intrinsic
-population characteristics such as density dependence.) We obtained order-level
-estimates by adding the posteriors for $\alpha$, $\alpha^\mathrm{class}_j$, and
+population characteristics such as density dependence.) Since our predictors were
+centered by subtracting their mean value, we obtained order-level
+estimates of the probability of heavy tails at mean time-series length by
+adding the posteriors for $\alpha$, $\alpha^\mathrm{class}_j$, and
 $\alpha^\mathrm{order}_k$.

### 2014-10-23 text/ms.tex
Edit and write introduction and discussion

@@
 that is expected to cause the greatest societal damage \citep{ipcc2012}.
-However, whereas extremes in the physical environment such as wave height, rain
+But, while extremes in the physical environment such as wave height, rain
 fall, and temperature are present and widely accepted
@@
 %mostly anecdotal or for particular systems.
-There is compelling anecdotal evidence for ecological black swans, and
-quantitative evidence for a limited number of systems, but systematic evidence
-has been elusive. \citet{doak2008} reviewed a collection of field experiment
-population dynamic surprises and conclude that these are more common than we
-typically assume. \citet{nunez2012} compiled a list of major events on geologic
-and contemporary scales that could be considered ecological black swans. Other
-work has identified heavy-tail dynamics for particular systems. For example,
-\citet{segura2013} identified heavy tails in time series of marine microbe
-populations and \citet{drake2014} identified heavy tails in time to extinction
-for experimental waterflea populations. However, systematic evidence has been
-limited. \citet{keitt1998} tested for power-law tail distributions in breeding
-bird abundance, but later worked showed the apparent heavy tails were a result
-of mixing populations of different species \citep{allen2001}. A later large
-scale analysis of the Global Population Dynamics Database (GPDD) by
-\citet{halley2002} found little evidence for heavy-tail distributions in longer
-abundance time series.
+There is compelling anecdotal evidence for ecological black swans, but
+systematic evidence has been elusive. \citet{doak2008} reviewed a collection of
+field experiment population dynamic surprises and concluded that surprises are
+more common than we typically assume. \citet{nunez2012} compiled a list of
+major events on geologic and contemporary scales that could be considered
+ecological black swans. Other work has identified heavy-tail dynamics for
+particular systems. For example, \citet{segura2013} identified heavy tails in
+time series of marine microbe populations and \citet{drake2014} identified
+heavy tails in time to extinction for experimental waterflea populations.
+However, systematic evidence is limited. \citet{keitt1998} tested for power-law
+tail distributions in breeding bird abundance, but later work showed the
+apparent heavy tails were a result of mixing populations of different species
+\citep{allen2001}. A subsequent large scale analysis of the Global Population
+Dynamics Database (GPDD) by \citet{halley2002} found little evidence for
+heavy-tail distributions in longer abundance time series.
 
@@
 swans. First, they might not systematically exist. Indeed, the majority of
-population dynamics model fitting and risk forecasting assumes that population
-dynamics are not heavy tailed (e.g.~REF, REF, REF). Alternatively, dynamics
-might be heavy tailed, but a paucity of statistical methods for the frequently
-available short time series has limited our ability to detect them. For
-example, a commonly applied probability distribution to detect the shape of
-time series tails (the generalized extreme value distribution) requires lumping
-data into the most extreme event per unit time (e.g.~heaviest rainfall per
-year) and therefore requires a considerable quantity of data (REF). Previous
-large-scale work on the distribution of abundance trends focused on model
-comparison to identify parsimonious descriptions, but has not quantified the
-magnitude of heavy tails on a continuous scale (REF). Furthermore, information
-criteria are poor at distinguishing subtle differences in population dynamics
-for the relatively short time series that are most readily available (REF
-CHANGE THIS?).
+model fitting and risk forecasting assumes that population
+dynamics are not heavy tailed (e.g.~REF, REF, REF).
+Alternatively, dynamics might be heavy tailed, but a paucity of statistical
+methods appropriate for short time series has limited our
+ability to detect them.
+For example, (1) the generalized extreme value (GEV) distribution has been
+applied to environmental data (REF).
+The GEV describes the distribution of the most extreme
+event per time interval (e.g.~heaviest rainfall per year), but this approach
+requires long time series that can be condensed into time intervals (REF).
+(2) Heavy- and thin-tailed distributions have been competed with each
+other (REF), but information criteria likely have
+trouble accurately distinguishing such subtle differences in population
+dynamics for short time series (REF).
+(3) A catastrophic mixture distribution model allows one to elegantly quantify
+the probability that population events are catastrophic is also data intensive
+\citep{ward2007}.
+(4) Other research has focussed on tipping points and regime shifts, but such
+methods are usually focussed on entire communities or ecosystems (e.g.\ REF).
+Furthermore, black swans are more general than tipping points --- they don't
+necessarily imply a permanent change of state.
+
+% For example, a commonly applied probability distribution to detect the shape of
+% time series tails (the generalized extreme value distribution) requires lumping
+% data into the most extreme event per unit time (e.g.~heaviest rainfall per
+% year) and therefore requires a considerable quantity of data (REF). Previous
+% large-scale work on the distribution of abundance trends focused on model
+% comparison to identify parsimonious descriptions, but has not quantified the
+% magnitude of heavy tails on a continuous scale (REF). Furthermore, information
+% criteria are poor at distinguishing subtle differences in population dynamics
+% for the relatively short time series that are most readily available (REF
+% CHANGE THIS?).
 %bring in Ward - great method but data intensive, can't be applied to most
-%individual time series from the GPDD
-%plankton --- hideous distribution to fit, even harder in population dynamics
-%framework -and what about seasonal cycles, density dependence?
-%GPDD - model competition and finding most parsimonious explanation from set of
-%hypotheses --- but only \emph{very} heavy from not heavy
-%GEV --- super data intensive
-%tipping point and regime shift methods --- say something\ldots{} not
-%appropriate for most of the data available
+%individual time series from the GPDD plankton --- hideous distribution to fit,
+%even harder in population dynamics framework -and what about seasonal cycles,
+%density dependence? GPDD - model competition and finding most parsimonious
+%explanation from set of hypotheses --- but only \emph{very} heavy from not
+%heavy GEV --- super data intensive tipping point and regime shift methods ---
+%say something\ldots{} not appropriate for most of the data available
 
@@
 We extended the standard Gompertz model by allowing the process noise to be
-drawn from a Student-t distribution with degrees of freedom parameter $\nu$, and scale parameter $\sigma$:
+drawn from a Student-t distribution with degrees of freedom parameter $\nu$,
+and scale parameter $\sigma$:
 \begin{align*}
@@
 
-We estimated all models in a Bayesian framework using Stan
-\citep[][version 2.4.0]{stan-manual2014} via the R computing environment \citep{r2014}. Stan
-samples from the posterior distribution with an adaptive version of
-Hamiltonian Markov chain Monte Carlo called the No-U-Turn Sampler and
-generally obtains less correlated chains than algorithms such as the Gibbs
-sampler \citep{hoffman2014}. We assured that chains had sufficiently converged
-and the sampler had obtained sufficient independent samples from the posterior
+We estimated all models in a Bayesian framework using Stan \citep[][version
+2.4.0]{stan-manual2014} via the R computing environment \citep{r2014}. Stan
+samples from the posterior distribution with an adaptive version of Hamiltonian
+Markov chain Monte Carlo called the No-U-Turn Sampler and generally obtains
+less correlated chains than algorithms such as the Gibbs sampler
+\citep{hoffman2014}. We assured that chains had sufficiently converged and the
+sampler had obtained sufficient independent samples from the posterior
 ($\widehat{R} < 1.05$, $n_\mathrm{eff} > 200$; Supporting Material). We used
@@
 long-necked bird in the UK was thought to reduce population productivity and
-make the population vulnerable when a red-tide hit in 1968 \citep{potts1980}.
+make the population vulnerable when a red tide hit in 1968 \citep{potts1980}.
 The population experienced a black swan downswing, but the new availability of
@@
 
-Our results clarify, but do not disagree with previous related analyses.
-Previous work with an older version of the GPDD assessed the distribution of
-abundance timeseries but focused on identifying if the lognormal distribution
-was the most frequently parsimonious model \citep{halley2002}. For heavy-tailed
-distributions, \citet{halley2002} fit the extremely heavy-tailed Cauchy
-distribution and the four-parameter Levy stable distribution and found little
-information criteria support for these distributions in longer time series.
-However, black swan events are by definition rare, and the majority of time
-series in the GPDD are short for these purposes. Therefore, we would not expect
-to observe black swans in a large proportion of populations. Furthermore, by
-probabilistically quantifying process noise tail behaviour on a continuous
-spectrum, and allowing for non-stationary time series and density dependence,
-our analysis allows for a more nuanced description of the evidence of
-ecological black swans. In an earlier study, \citet{keitt1998} described heavy
-(power law) tails in breeding bird population abundance. However, this finding
-was challenged by \citet{allen2001}, who showed that mixing data across species
-could create falsely generate heavy tails. This rebuttal supports our use of
-the t distribution to represent heavy tails, since a t distribution can be
+Our results clarify previous related analyses. Previous work with an older
+version of the GPDD assessed the distribution of abundance time series but
+focused on identifying if the lognormal distribution was the most frequently
+parsimonious model \citep{halley2002}. For heavy-tailed distributions,
+\citet{halley2002} fit the extremely heavy-tailed Cauchy distribution and the
+four-parameter Levy stable distribution and found little information criteria
+support for these distributions in longer time series. However, black swan
+events are by definition rare, and the majority of time series in the GPDD are
+short for these purposes. Therefore, we would not expect to observe black swans
+in a large proportion of populations. Furthermore, by probabilistically
+quantifying process noise tail behaviour on a continuous spectrum, and allowing
+for non-stationary time series and density dependence, our analysis allows for
+a more nuanced description of the evidence of ecological black swans. In an
+earlier study, \citet{keitt1998} described heavy (power law) tails in breeding
+bird population abundance. However, this finding was challenged by
+\citet{allen2001}, who showed that mixing data across species could create
+falsely generate heavy tails. This rebuttal supports our use of the
+t distribution to represent heavy tails, since a t distribution can be
 represented by a mixture of normal distributions \citep[with
@@
 We might expect to observe black swan dynamics in ecological time series
-because of unmodelled intrinsic properties of populations or extrinsic forces
-acting on populations. Since a t-distribution can be formed by a mixture of
-normal distributions, we could observe heavy tails if we miss some underlying
-mixture of intrinsic processes \citep{allen2001}. That process might be an
-aggregation of populations across space, or population diversity, or some
-intrinsic change in population variability through time. Extrinsic forces
-could also cause black swan dynamics \citep[e.g.][]{nunez2012}. These forces
-could be extreme themselves. For example, extreme climate, predation from or
-competition with other species experiencing black swans, or sharp changes in
-human pressure such hunting, fishing, or habitat destruction might cause black
-swans. Alternatively, the interaction of multiple ``normal'' extrinsic
-forces could give rise to black swan ecological dynamics. This could occur if
-the interaction nature is synergistic \citep[e.g.][]{kirby2009} or even if
-non-synergistic interactions experience a rare alignment \citep{denny2009}.
+because of either unmodelled intrinsic properties of populations or extrinsic
+forces acting on populations. Since a t-distribution can be formed by a mixture
+of normal distributions \citep{gelman2014}, we could observe heavy tails if we
+miss some underlying mixture of intrinsic processes \citep{allen2001}. That
+process might be an aggregation of populations across space, or population
+diversity, or some intrinsic change in population variability through time.
+Extrinsic forces could also cause black swan dynamics
+\citep[e.g.][]{nunez2012}. These forces could be extreme themselves. For
+example, extreme climate, predation from or competition with other species
+experiencing black swans, or sharp changes in human pressure such hunting,
+fishing, or habitat destruction might cause black swans. Alternatively, the
+interaction of multiple ``normal'' extrinsic forces could give rise to black
+swan ecological dynamics. This could occur if the interaction nature is
+synergistic \citep[e.g.][]{kirby2009} or even if non-synergistic interactions
+experience a rare alignment \citep{denny2009}.
 
@@
 recording mistakes could mask heavy tails if initial observations are assumed
-to be recording errors and are discarded or altered. Third, the temporal scale
+to be recording errors and are discarded or altered. Third, the temporal scales
 of observation and population dynamics vary considerably across populations in
@@
 Our results suggest important future issues to address about ecological black
-swans. Given that black swans can and do occur, can we forecast the probability
-of black swans in space and time? Furthermore, what management policies allow
-us to detect them quickly after they happen? Secondly, can we isolate the
+swans. Given that black swans do occur, can we forecast the probability of
+black swans in space and time? Furthermore, what management policies allow us
+to detect them quickly after they happen? Secondly, can we isolate the
 components of ecological dynamics that experience black swans events by moving
@@
 
-Policy implications
+Recognizing the prevalence of heavy-tailed dynamics suggests a number of policy
+directions. First, ecological resource management can draw from other
+disciplines focussed on heavy tails. For example, earthquake preparedness and
+response is focussed on black swan events. Similarly to ecological black swans,
+we can rarely predict the specific location and timing of large earthquakes.
+But, earthquake preparedness involves spatial planning based on forecast
+probabilities to focus early detection efforts and develop disaster response
+plans. A similar focus could benefit resource management once our ability to
+predict the spatial probability and covariates of ecological black swans
+improves. Second, we can choose management policy that is robust to heavy
+tails. For example, setting target population abundances that are appropriately
+far from critical limits may buffer black swan events (REF), and
+maintaining genetic, phenotypic, and behavioural diversity may allow some
+components of populations persist when others are affected by extreme disease
+or environmental forces (REF). Finally, extreme and `unexpected', `surprising', or
+`counterintuitive' ecological dynamics offer a tremendous opportunity to learn
+about ecological systems, evaluate when our models break down, and tune future
+management policy (REF, REF). In recent decades ecology has moved toward
+focussing on aspects of variance in addition to mean responses (REF). Our
+results and other recent work suggests that an added focus on extremes
+represents the next frontier, particularly in the face expected increases in
+extreme climate (REF).
 
-unlikely to forecast black swans, at least with any useful advanced warning
-but we might be able to determine probabilities and develop policy that is
-robust to, encourages rapid detection, and that is flexible enough to respond
-quickly to black swans
-
-example of earth quakes - spatial planning, early warning systems, education
-systems, a general idea about the probability in given places, build buildings
-that are robust
-
-Suggests that it's important to have policy that is robust to the uncertainty
-about black swans and that can respond and adapt rapidly following black swans
-
-foot and mouth disease and earthquakes?
+%ecological resource management can draw from other disciplines focussed on
+%heavy tails, design policy that is robust to heavy tails, and take black swan
+%events as opportunities for insight.
+%The 'portfolio effect' is often used in the context of dampening variability
+%(REFs), but the ultimate portfolio effect is maintaining some stability in the
+%face of black swan events.
+%unlikely to forecast black swans, at least with any useful advanced warning
+%but we might be able to determine probabilities and develop policy that is
+%robust to, encourages rapid detection, and that is flexible enough to respond
+%quickly to black swans
+%example of earth quakes - spatial planning, early warning systems, education
+%systems, a general idea about the probability in given places, build buildings
+%that are robust
+%Suggests that it's important to have policy that is robust to the uncertainty
+%about black swans and that can respond and adapt rapidly following black swans
+%foot and mouth disease and earthquakes?
 

### 2014-10-23 text/ms.tex
Edit an intro paragraph

@@
 
-Black swans are unexpected extreme events with dramatic consequences
+Black swans are unexpected extreme events with potentially dramatic consequences
 \citep{taleb2007,sornette2009}. One of the most striking black swans in ecology
@@
 There is compelling anecdotal evidence for ecological black swans, but
-systematic evidence has been elusive. \citet{doak2008} reviewed a collection of
-field experiment population dynamic surprises and concluded that surprises are
-more common than we typically assume. \citet{nunez2012} compiled a list of
-major events on geologic and contemporary scales that could be considered
-ecological black swans. Other work has identified heavy-tail dynamics for
+systematic evidence has been elusive. A survey of ecologists indicated that
+surprising results in field experiments may be far more common than we assume
+\citep{doak2008}. And, a number of major environmental events on geologic and
+contemporary time scales could be considered to have caused ecological black
+swans \citep{nunez2012}. Other work has detected heavy-tail dynamics for
 particular systems. For example, \citet{segura2013} identified heavy tails in
@@
 heavy tails in time to extinction for experimental waterflea populations.
-However, systematic evidence is limited. \citet{keitt1998} tested for power-law
-tail distributions in breeding bird abundance, but later work showed the
-apparent heavy tails were a result of mixing populations of different species
-\citep{allen2001}. A subsequent large scale analysis of the Global Population
-Dynamics Database (GPDD) by \citet{halley2002} found little evidence for
-heavy-tail distributions in longer abundance time series.
+However, systematic evidence has been limited. What appeared to be heavy tails
+in breeding bird abundance \citep{keitt1998}, could also be explained by the
+mixing of species in the original dataset \citep{allen2001}. A subsequent
+large scale analysis of the Global Population Dynamics Database (GPDD) found
+little evidence for heavy-tail distributions in longer abundance time series
+\citet{halley2002}.
 

### 2014-10-23 text/ms.tex
More intro tweaks

@@
 surprising results in field experiments may be far more common than we assume
-\citep{doak2008}. And, a number of major environmental events on geologic and
+\citep{doak2008}, and a number of major environmental events on geologic and
 contemporary time scales could be considered to have caused ecological black
@@
 ability to detect them.
-For example, (1) the generalized extreme value (GEV) distribution has been
+As four examples: (1) The generalized extreme value (GEV) distribution has been
 applied to environmental data (REF).
@@
 dynamics for short time series (REF).
-(3) A catastrophic mixture distribution model allows one to elegantly quantify
-the probability that population events are catastrophic is also data intensive
-\citep{ward2007}.
+(3) A catastrophic mixture distribution model allows one to quantify the
+probability that population events are catastrophic but is also relatively data
+intensive \citep{ward2007}.
 (4) Other research has focussed on tipping points and regime shifts, but such
@@
 model (REFs?). The Gompertz model represents population growth as a linear
-function in $\log$ space. If we let $x_t$ represent the $\log$ abundance ($N$)
-at time $t$, we can represent the Gompertz model as:
+function in $\log$ space. If we let $x_t$ represent the $\log$ abundance at
+time $t$, we can represent the Gompertz model as:
 \begin{align*}

### 2014-10-23 text/supp.tex
Clarify why we left body size out of model in SOM
Clean up paragraph in SOM

@@
 samples for the Gompertz parameters $\lambda$, $b$, and $\sigma$ around the
-mean point value as normal distributions (not shown in the above equation). We
-log transformed $\sigma$, time-series length, and lifespan to match the way
+mean point value as normal distributions (not shown in the above equation).
+
+We log transformed $\sigma$, time-series length, and lifespan to match the way
 they are visually represented in Fig.~\ref{fig:correlates} and to make the
@@
 input variables were standardized by subtracting their mean and dividing by two
-standard deviations \citep{gelman2008c} to make their coefficients comparable
-in magnitude.
+standard deviations to make their coefficients comparable in magnitude
+\citep{gelman2008c}. We excluded body length as a covariate because it was
+highly correlated with lifespan, and lifespan exhibited more overlap across
+taxonomy than body length. Lifespan is also more directly related to time and
+potential mechanisms driving black swan dynamics.
 

### 2014-10-24 text/ms.tex
Add macro magic numbers to main text

@@
 often with profound implications. While extremes in the physical
-environment,such as monsoons and heat waves, are widely studied and increasing
+environment, such as monsoons and heat waves, are widely studied and increasing
 in magnitude and frequency, it remains unclear the extent to which ecological
@@
 \NPops~populations across \NOrders~taxonomic orders and seven taxonomic classes
-with a median of XX time steps (range of XX--XX) (Table S1,
-Fig.~\ref{fig:all-ts}).
+with a median of \medianTimeSteps~time steps (range of
+\minTimeSteps--\maxTimeSteps) (Table S1, Fig.~\ref{fig:all-ts}).
 
@@
 dynamics as a greater than $0.5$ probability that $\nu < 10$, we observed black
-swans most frequently for birds (X\%) followed by mammals (X\%), insects (X\%),
-and fishes (X\%) (Fig.~\ref{fig:nu-coefs}, \ref{fig:heavy-ts}). Black swans
-were taxonomically widespread, occurring for at least one population in X\% of
-the taxonomic orders recorded. Accounting for time series length and partially
-pooling inference across taxonomic class and order with a multilevel model,
-there was stronger evidence for black swans in insect populations than is
-apparent in Fig.~\ref{fig:nu-coefs} --- four of 10 orders with the highest
-median probability of heavy tails were insect orders --- however, there was
+swans most frequently for birds (\birdPH\%) followed by mammals (\mammalsPH\%),
+insects (\insectsPH\%), and fishes (\fishPH\%) (Fig.~\ref{fig:nu-coefs},
+\ref{fig:heavy-ts}). Black swans were taxonomically widespread, occurring for
+at least one population in \POrdersHeavy\% of the taxonomic orders recorded.
+Accounting for time series length and partially pooling inference across
+taxonomic class and order with a multilevel model, there was stronger evidence
+for black swans in insect populations than is apparent in
+Fig.~\ref{fig:nu-coefs} --- four of 10 orders with the highest median
+probability of heavy tails were insect orders --- however, there was
 considerable uncertainty in these estimates (Fig.~\ref{fig:order-estimates}).
@@
 a fixed quantity of observation error decreased our estimates of black-swan
-dynamics, increasing the median estimate of $\nu$ from $<10$ to $>10$ in X of
-XX populations although the majority of $\nu$ estimates remained qualitatively
-similar (Fig.~\ref{fig:alt})
+dynamics, increasing the median estimate of $\nu$ from $<10$ to $>10$ in
+$\baseNuTenObsTenSwitch$ of $\baseNuTen$ populations although the majority of
+$\nu$ estimates remained qualitatively similar (Fig.~\ref{fig:alt})
 
@@
 similarly observed in simulation testing, Fig.~\ref{fig:sim-nu}). For example,
-we were about 15\% TODO more likely to observe heavy tails in a population with
-40 time steps than in one with 30 time steps (Fig.~\ref{fig:perc-inc-p}).
-However, the absolute change in probability with increased time series length
-was small ($0.X$ vs. $0.X$ in the previous example, Fig.~\ref{fig:correlates}).
+we were about \pIncHeavyNThirtyNSixty~times more likely to observe heavy tails
+in a population with 60 time steps than in one with 30 time steps
+(Fig.~\ref{fig:perc-inc-p}). However, the absolute change in probability with
+increased time series length was small ($0.\pHeavyNSixty$ vs.\
+$0.\pHeavyNThirty$ in the previous example, Fig.~\ref{fig:correlates}).
 
@@
 step to time step were greater than we would expect under a normal distribution
-and a Gompertz or Ricker-logistic model in XX--XX\% of populations assessed. We
-detected black swans more frequently for longer time series and for populations
-with a smaller magnitude of process error, as we might expect. However, we
-failed to find strong evidence that black swans were associated with density
-dependence, population growth rate, or lifespan. In cases that we verified,
-black swans were often a result of the interaction between elements of extreme
-climate, predators and parasite cycles, and strong changes in human pressures.
-Our empirical results, sensitivity tests, and simulation tests suggest that
-estimating the tail shape of process error is a viable method of detecting
-heavy-tailed population dynamics and if anything will underestimate the
-probability of black swans. The presence of black swans highlights the
-importance of developing management strategies that detect quickly, respond
-to, and are robust to extremes in population dynamics --- particularly as the
-frequency and magnitude of climatic extremes increase over the next century
-(REF).
+and a Gompertz or Ricker-logistic model in \overallMinPerc--\overallMaxPerc\%
+of populations assessed. We detected black swans more frequently for longer
+time series and for populations with a smaller magnitude of process error, as
+we might expect. However, we failed to find strong evidence that black swans
+were associated with density dependence, population growth rate, or lifespan.
+In cases that we verified, black swans were often a result of the interaction
+between elements of extreme climate, predators and parasite cycles, and strong
+changes in human pressures. Our empirical results, sensitivity tests, and
+simulation tests suggest that estimating the tail shape of process error is
+a viable method of detecting heavy-tailed population dynamics and if anything
+will underestimate the probability of black swans. The presence of black swans
+highlights the importance of developing management strategies that detect
+quickly, respond to, and are robust to extremes in population dynamics ---
+particularly as the frequency and magnitude of climatic extremes increase over
+the next century (REF).
 

### 2014-10-24 text/ms.tex
Edit paper, add references, and clean out comments

@@
 
-Black swans are unexpected extreme events with potentially dramatic consequences
-\citep{taleb2007,sornette2009}. One of the most striking black swans in ecology
-was the asteroid marking the mass extinction of the K-T boundary (REF). Today,
-it is the extremeness of climate, in concert with shifts in mean temperature,
-that is expected to cause the greatest societal damage \citep{ipcc2012}.
-But, while extremes in the physical environment such as wave height, rain
-fall, and temperature are present and widely accepted
-\citep{gaines1993,katz2005}, it remains unclear the extent to which ecological
-systems buffer or suffer from black swans \citep{nunez2012}.
+Black swans are unexpected extreme events with potentially dramatic
+consequences \citep{taleb2007,sornette2009}. One of the most striking black
+swans in ecology was the asteroid marking the mass extinction of the K-T
+boundary \citep{harnik2012}. Today, it is the extremeness of climate, in
+concert with shifts in mean temperature, that is expected to cause the greatest
+societal damage \citep{ipcc2012}. But, while extremes in the physical
+environment such as wave height, rain fall, and temperature are present and
+widely accepted \citep{gaines1993,katz2005}, it remains unclear the extent to
+which ecological systems buffer or suffer from black swans \citep{nunez2012}.
 
@@
 
-%Considerable work has focussed on heavy tails in physical environmental
-%processes in, for example, rain fall (REF), wave height (REF), earth quake
-%magnitude, climate temperature (REF). time to extinction, plankton time series,
-%GPDD work found little evidence subsequent work suggested, but... Ward applied
-%a data-intense method and found evidence for.
-
-%Ecological population dynamic models typically assume 'thin tailed'
-%distributions, such as the normal distribution on a multiplicative scale (e.g.
-%REF). We define black swan dynamics as cases where jumps in population
-%abundance from time step to time step. from heavy-tailed distributions.
-
 There are two key reasons why we may find little evidence of ecological black
 swans. First, they might not systematically exist. Indeed, the majority of
-model fitting and risk forecasting assumes that population
-dynamics are not heavy tailed (e.g.~REF, REF, REF).
-Alternatively, dynamics might be heavy tailed, but a paucity of statistical
-methods appropriate for short time series has limited our
-ability to detect them.
-As four examples: (1) The generalized extreme value (GEV) distribution has been
-applied to environmental data (REF).
-The GEV describes the distribution of the most extreme
-event per time interval (e.g.~heaviest rainfall per year), but this approach
-requires long time series that can be condensed into time intervals (REF).
-(2) Heavy- and thin-tailed distributions have been competed with each
-other (REF), but information criteria likely have
-trouble accurately distinguishing such subtle differences in population
-dynamics for short time series (REF).
-(3) A catastrophic mixture distribution model allows one to quantify the
-probability that population events are catastrophic but is also relatively data
+model fitting and risk forecasting assumes that population dynamics are not
+heavy tailed \citep[e.g.][]{brook2006a,dennis2006,knape2012}. Alternatively,
+dynamics might be heavy tailed, but a paucity of statistical methods
+appropriate for commonly available series has limited our ability to detect
+them. For example: (1) The generalized extreme value (GEV) distribution has
+been applied to environmental data \citep[e.g.][]{katz2005}. The GEV describes
+the distribution of the most extreme event per time interval (e.g.~heaviest
+rainfall per year), but this approach requires long time series that can be
+condensed into time intervals. (2) Thin- and heavy-tailed distributions have
+been competed with each other in an information theoretic framework
+\citep{halley2002}, but this analysis does not quantify the probability of
+black swans on a continuous scale or allow for population dynamics. (3) Fitting
+a catastrophic mixture distribution state space model allows one to quantify
+the probability that population events are catastrophic but is also data
 intensive \citep{ward2007}.
-(4) Other research has focussed on tipping points and regime shifts, but such
-methods are usually focussed on entire communities or ecosystems (e.g.\ REF).
-Furthermore, black swans are more general than tipping points --- they don't
-necessarily imply a permanent change of state.
-
-% For example, a commonly applied probability distribution to detect the shape of
-% time series tails (the generalized extreme value distribution) requires lumping
-% data into the most extreme event per unit time (e.g.~heaviest rainfall per
-% year) and therefore requires a considerable quantity of data (REF). Previous
-% large-scale work on the distribution of abundance trends focused on model
-% comparison to identify parsimonious descriptions, but has not quantified the
-% magnitude of heavy tails on a continuous scale (REF). Furthermore, information
-% criteria are poor at distinguishing subtle differences in population dynamics
-% for the relatively short time series that are most readily available (REF
-% CHANGE THIS?).
-%bring in Ward - great method but data intensive, can't be applied to most
-%individual time series from the GPDD plankton --- hideous distribution to fit,
-%even harder in population dynamics framework -and what about seasonal cycles,
-%density dependence? GPDD - model competition and finding most parsimonious
-%explanation from set of hypotheses --- but only \emph{very} heavy from not
-%heavy GEV --- super data intensive tipping point and regime shift methods ---
-%say something\ldots{} not appropriate for most of the data available
 
@@
 probability of black swans.
-
-% Papers to work in: \citep{inchausti2002, halley2002, inchausti2001}
-% \citep{jentsch2007} \citep{ward2007} \citep{garcia-carreras2011}
-% \citep{sornette2009} \citep{nunez2012} \citep{thompson2013}
-% \citep{beaugrand2012} \citep{pine-iii2009} \citep{doak2008} Extreme rain and
-% monsoons over India increasing \citep{goswami2006} \citep{smale2013}
-% \citep{easterling2000} \citep{scheffer2003} \citep{katz2005} \citep{taleb2007}
-% \citep{vasseur2014} \citep{vert-pre2013} \citep{lindenmayer2010}
-% \citep{valpine2002} \citep{gregory2010} \citet{garcia-carreras2011}
-% \citeauthor{brook2006} \citetext{\citeyear{brook2006};
-%   \citealp{herrandoprez2014}} \citep{sibly2005, ziebarth2010}
 
@@
 Our main analysis focuses on the commonly applied Gompertz population dynamics
-model (REFs?). The Gompertz model represents population growth as a linear
-function in $\log$ space. If we let $x_t$ represent the $\log$ abundance at
-time $t$, we can represent the Gompertz model as:
+model \citep[e.g.][]{knape2012,dennis2014,connors2014}. The Gompertz model
+represents population growth as a linear function in $\log$ space. If we let
+$x_t$ represent the $\log$ abundance at time $t$, we can represent the Gompertz
+model as:
 \begin{align*}
@@
 distributed with standard deviation $\sigma$. In previous analyses of the GPDD,
-the Gompertz has proved to be the most frequently parsimonious model
+the Gompertz has proved to be the most frequently parsimonious population model
 \citep{brook2006}.
@@
 Markov chain Monte Carlo called the No-U-Turn Sampler and generally obtains
-less correlated chains than algorithms such as the Gibbs sampler
+less correlated samples than algorithms such as the Gibbs sampler
 \citep{hoffman2014}. We assured that chains had sufficiently converged and the
@@
 itself (Supporting Material).
-
-% TODO - make sure this is in the supp? However, simultaneously estimating
-% observation error
-%and process noise in a state space framework is a challenging computational
-%problem \citep[e.g.][]{dennis2006, knape2008} even with longer time series and
-%without estimating the shape of the process error tails;
 
@@
 particularly as the frequency and magnitude of climatic extremes increase over
-the next century (REF).
+the next century \citep{easterling2000,ipcc2012}.
 
@@
 tails. For example, setting target population abundances that are appropriately
-far from critical limits may buffer black swan events (REF), and
-maintaining genetic, phenotypic, and behavioural diversity may allow some
-components of populations persist when others are affected by extreme disease
-or environmental forces (REF). Finally, extreme and `unexpected', `surprising', or
-`counterintuitive' ecological dynamics offer a tremendous opportunity to learn
-about ecological systems, evaluate when our models break down, and tune future
-management policy (REF, REF). In recent decades ecology has moved toward
-focussing on aspects of variance in addition to mean responses (REF). Our
-results and other recent work suggests that an added focus on extremes
+far from critical limits may buffer black swan events
+\citep[e.g.][]{caddy1996}, and maintaining genetic, phenotypic, and behavioural
+diversity may allow some components of populations persist when others are
+affected by extreme disease or environmental forces
+\citep[e.g.][]{schindler2010, anderson2014}. Finally, extreme and `unexpected',
+`surprising', or `counterintuitive' ecological dynamics offer a tremendous
+opportunity to learn about ecological systems, evaluate when our models break
+down, and tune future management policy \citep{doak2008, pine-iii2009,
+  lindenmayer2010}. In recent decades ecology has moved toward focussing on
+aspects of variance in addition to mean responses \citep[e.g.][]{loreau2010a,
+  thompson2013}. Our results suggests that an added focus on extremes
 represents the next frontier, particularly in the face expected increases in
-extreme climate (REF).
-
-%ecological resource management can draw from other disciplines focussed on
-%heavy tails, design policy that is robust to heavy tails, and take black swan
-%events as opportunities for insight.
-%The 'portfolio effect' is often used in the context of dampening variability
-%(REFs), but the ultimate portfolio effect is maintaining some stability in the
-%face of black swan events.
-%unlikely to forecast black swans, at least with any useful advanced warning
-%but we might be able to determine probabilities and develop policy that is
-%robust to, encourages rapid detection, and that is flexible enough to respond
-%quickly to black swans
-%example of earth quakes - spatial planning, early warning systems, education
-%systems, a general idea about the probability in given places, build buildings
-%that are robust
-%Suggests that it's important to have policy that is robust to the uncertainty
-%about black swans and that can respond and adapt rapidly following black swans
-%foot and mouth disease and earthquakes?
+extreme climate \citep{meehl2004,ipcc2012}.
 

### 2014-10-29 text/ms.tex
Edit discussion

@@
 
-We found strong evidence that the multiplicative process noise jumps from time
-step to time step were greater than we would expect under a normal distribution
-and a Gompertz or Ricker-logistic model in \overallMinPerc--\overallMaxPerc\%
+We found strong evidence that the process noise in ecological time series was heavy-tailed under the Gompertz or Ricker-logistic models in \overallMinPerc--\overallMaxPerc\%
 of populations assessed. We detected black swans more frequently for longer
@@
 between elements of extreme climate, predators and parasite cycles, and strong
-changes in human pressures. Our empirical results, sensitivity tests, and
-simulation tests suggest that estimating the tail shape of process error is
-a viable method of detecting heavy-tailed population dynamics and if anything
+changes in human pressures. Our empirical results, sensitivity analyses, and
+simulation tests suggest that estimating the tail shape of process noise is
+a viable method of detecting black swan population dynamics and if anything
 will underestimate the probability of black swans. The presence of black swans
@@
 in a large proportion of populations. Furthermore, by probabilistically
-quantifying process noise tail behaviour on a continuous spectrum, and allowing
+quantifying process noise tails on a continuous spectrum, and allowing
 for non-stationary time series and density dependence, our analysis allows for
@@
 winters in Table~1) likely differ. Second, in a large database with disparate
-data sources, we cannot ensure the quality of every data point. Recording
-mistakes undoubtedly comprise some of the black swans we detected. Conversely,
-recording mistakes could mask heavy tails if initial observations are assumed
-to be recording errors and are discarded or altered. Third, the temporal scales
+data sources, we cannot ensure the quality of every data point;
+it is likely that some of the black swans in our dataset are 
+recording mistakes mistakes.
+Conversely, recording mistakes could mask heavy tails if initial observations are discarded or altered as assumed errors. Third, the temporal scales
 of observation and population dynamics vary considerably across populations in
@@
 black swans in space and time? Furthermore, what management policies allow us
-to detect them quickly after they happen? Secondly, can we isolate the
+to detect them quickly after they happen? Can we isolate the
 components of ecological dynamics that experience black swans events by moving
@@
 impact of allowing for black swans in forecasts of ecological risk? Financial
-analysts are concerned with the shape of financial-return distribution tails
+analysts are concerned with the shape of financial-return-distribution tails
 because these impact estimates of risk --- the probability of a specific
@@
 probabilities to focus early detection efforts and develop disaster response
-plans. A similar focus could benefit resource management once our ability to
+plans (e.g.~REF).
+A similar focus could benefit resource management once our ability to
 predict the spatial probability and covariates of ecological black swans
@@
 diversity may allow some components of populations persist when others are
-affected by extreme disease or environmental forces
-\citep[e.g.][]{schindler2010, anderson2014}. Finally, extreme and `unexpected',
-`surprising', or `counterintuitive' ecological dynamics offer a tremendous
+affected by disease or extreme environmental forces
+\citep[e.g.][]{schindler2010, anderson2014}. Finally, extreme and unexpected,
+surprising, or counterintuitive ecological dynamics offer a tremendous
 opportunity to learn about ecological systems, evaluate when our models break
@@
 represents the next frontier, particularly in the face expected increases in
-extreme climate \citep{meehl2004,ipcc2012}.
+climate extremes \citep{meehl2004,ipcc2012}.
 
@@
 The GPDD IDs used in our analysis\\
-All code and data to recreate our analysis is available at:\\
+All code and data to recreate our analysis are available at:\\
 \url{https://github.com/seananderson/heavy-tails}

### 2014-10-29 text/ms.tex
Edit introduction

@@
 societal damage \citep{ipcc2012}. But, while extremes in the physical
-environment such as wave height, rain fall, and temperature are present and
+environment such as wave height, storm severity, and temperature are present and
 widely accepted \citep{gaines1993,katz2005}, it remains unclear the extent to
@@
 However, systematic evidence has been limited. What appeared to be heavy tails
-in breeding bird abundance \citep{keitt1998}, could also be explained by the
+in breeding bird abundance \citep{keitt1998}, could also be explained by 
 mixing of species in the original dataset \citep{allen2001}. A subsequent
@@
 been competed with each other in an information theoretic framework
-\citep{halley2002}, but this analysis does not quantify the probability of
+\citep{halley2002}, but this analysis did not quantify the probability of
 black swans on a continuous scale or allow for population dynamics. (3) Fitting
 a catastrophic mixture distribution state space model allows one to quantify
-the probability that population events are catastrophic but is also data
-intensive \citep{ward2007}.
+the probability that population events are extreme on a continuous scale
+but is data intensive \citep{ward2007}.
 

### 2014-10-29 text/ms.tex
Add missing ($N$)
Re-wrap in the ol' Vim

@@
 However, systematic evidence has been limited. What appeared to be heavy tails
-in breeding bird abundance \citep{keitt1998}, could also be explained by 
-mixing of species in the original dataset \citep{allen2001}. A subsequent
-large scale analysis of the Global Population Dynamics Database (GPDD) found
-little evidence for heavy-tail distributions in longer abundance time series
+in breeding bird abundance \citep{keitt1998}, could also be explained by mixing
+of species in the original dataset \citep{allen2001}. A subsequent large scale
+analysis of the Global Population Dynamics Database (GPDD) found little
+evidence for heavy-tail distributions in longer abundance time series
 \citet{halley2002}.
@@
 been competed with each other in an information theoretic framework
-\citep{halley2002}, but this analysis did not quantify the probability of
-black swans on a continuous scale or allow for population dynamics. (3) Fitting
+\citep{halley2002}, but this analysis did not quantify the probability of black
+swans on a continuous scale or allow for population dynamics. (3) Fitting
 a catastrophic mixture distribution state space model allows one to quantify
-the probability that population events are extreme on a continuous scale
-but is data intensive \citep{ward2007}.
+the probability that population events are extreme on a continuous scale but is
+data intensive \citep{ward2007}.
 
@@
 represents population growth as a linear function in $\log$ space. If we let
-$x_t$ represent the $\log$ abundance at time $t$, we can represent the Gompertz
+$x_t$ represent the $\log$ abundance ($N$) at time $t$, we can represent the Gompertz
 model as:
@@
 
-We found strong evidence that the process noise in ecological time series was heavy-tailed under the Gompertz or Ricker-logistic models in \overallMinPerc--\overallMaxPerc\%
-of populations assessed. We detected black swans more frequently for longer
-time series and for populations with a smaller magnitude of process error, as
-we might expect. However, we failed to find strong evidence that black swans
-were associated with density dependence, population growth rate, or lifespan.
-In cases that we verified, black swans were often a result of the interaction
-between elements of extreme climate, predators and parasite cycles, and strong
-changes in human pressures. Our empirical results, sensitivity analyses, and
-simulation tests suggest that estimating the tail shape of process noise is
-a viable method of detecting black swan population dynamics and if anything
-will underestimate the probability of black swans. The presence of black swans
-highlights the importance of developing management strategies that detect
-quickly, respond to, and are robust to extremes in population dynamics ---
-particularly as the frequency and magnitude of climatic extremes increase over
-the next century \citep{easterling2000,ipcc2012}.
+We found strong evidence that the process noise in ecological time series was
+heavy-tailed under the Gompertz or Ricker-logistic models in
+\overallMinPerc--\overallMaxPerc\% of populations assessed. We detected black
+swans more frequently for longer time series and for populations with a smaller
+magnitude of process error, as we might expect. However, we failed to find
+strong evidence that black swans were associated with density dependence,
+population growth rate, or lifespan. In cases that we verified, black swans
+were often a result of the interaction between elements of extreme climate,
+predators and parasite cycles, and strong changes in human pressures. Our
+empirical results, sensitivity analyses, and simulation tests suggest that
+estimating the tail shape of process noise is a viable method of detecting
+black swan population dynamics and if anything will underestimate the
+probability of black swans. The presence of black swans highlights the
+importance of developing management strategies that detect quickly, respond to,
+and are robust to extremes in population dynamics --- particularly as the
+frequency and magnitude of climatic extremes increase over the next century
+\citep{easterling2000,ipcc2012}.
 
@@
 in a large proportion of populations. Furthermore, by probabilistically
-quantifying process noise tails on a continuous spectrum, and allowing
-for non-stationary time series and density dependence, our analysis allows for
+quantifying process noise tails on a continuous spectrum, and allowing for
+non-stationary time series and density dependence, our analysis allows for
 a more nuanced description of the evidence of ecological black swans. In an
@@
 winters in Table~1) likely differ. Second, in a large database with disparate
-data sources, we cannot ensure the quality of every data point;
-it is likely that some of the black swans in our dataset are 
-recording mistakes mistakes.
-Conversely, recording mistakes could mask heavy tails if initial observations are discarded or altered as assumed errors. Third, the temporal scales
-of observation and population dynamics vary considerably across populations in
-the GPDD and this likely influences the detection of heavy tails. For example,
-if we make frequent observations relative to generation time (e.g.~for many
-large-bodied mammals) we will average across generations and perhaps miss
-black swans. Furthermore, if we census populations infrequently relative to
+data sources, we cannot ensure the quality of every data point; it is likely
+that some of the black swans in our dataset are recording mistakes mistakes.
+Conversely, recording mistakes could mask heavy tails if initial observations
+are discarded or altered as assumed errors. Third, the temporal scales of
+observation and population dynamics vary considerably across populations in the
+GPDD and this likely influences the detection of heavy tails. For example, if
+we make frequent observations relative to generation time (e.g.~for many
+large-bodied mammals) we will average across generations and perhaps miss black
+swans. Furthermore, if we census populations infrequently relative to
 generation time (e.g.~many insects in the GPDD) the recorded data may average
@@
 black swans in space and time? Furthermore, what management policies allow us
-to detect them quickly after they happen? Can we isolate the
-components of ecological dynamics that experience black swans events by moving
-from phenomenological models such as the Gompertz to mechanistic models, that
-take into account, for example, recruitment? We expect that greater insight
-into the mechanisms and covariates of ecological black swans may be best
-obtained through specific geographic and taxonomic subsets of data where longer
-time series are available \citep[e.g.][]{segura2013}. Finally, what is the
-impact of allowing for black swans in forecasts of ecological risk? Financial
-analysts are concerned with the shape of financial-return-distribution tails
-because these impact estimates of risk --- the probability of a specific
-magnitude of undesired event occurring \citep{rachev2008}. It is possible that
-a comparable focus in ecology would alter our perception of ecological
-forecasts such as extinction risk.
+to detect them quickly after they happen? Can we isolate the components of
+ecological dynamics that experience black swans events by moving from
+phenomenological models such as the Gompertz to mechanistic models, that take
+into account, for example, recruitment? We expect that greater insight into the
+mechanisms and covariates of ecological black swans may be best obtained
+through specific geographic and taxonomic subsets of data where longer time
+series are available \citep[e.g.][]{segura2013}. Finally, what is the impact of
+allowing for black swans in forecasts of ecological risk? Financial analysts
+are concerned with the shape of financial-return-distribution tails because
+these impact estimates of risk --- the probability of a specific magnitude of
+undesired event occurring \citep{rachev2008}. It is possible that a comparable
+focus in ecology would alter our perception of ecological forecasts such as
+extinction risk.
 
@@
 probabilities to focus early detection efforts and develop disaster response
-plans (e.g.~REF).
-A similar focus could benefit resource management once our ability to
-predict the spatial probability and covariates of ecological black swans
-improves. Second, we can choose management policy that is robust to heavy
+plans (e.g.~REF). A similar focus could benefit resource management once our
+ability to predict the spatial probability and covariates of ecological black
+swans improves. Second, we can choose management policy that is robust to heavy
 tails. For example, setting target population abundances that are appropriately
@@
 Commons Attribution 3.0 Unported license: rabbit (Sarah Werning), hoverfly
-(Gareth Monger), bird (Jean-Raphaël Guillaumin {[}photography{]} and T.
-Michael Keesey {[}vectorization{]}). Funding was provided by a Simon Fraser
-University Graduate Fellowship (SCA), the Natural Sciences and Engineering
-Research Council of Canada (NKD, ABC), the Canada Research Chairs Program
-(NKD), and TODO (TAB).
+(Gareth Monger), bird (Jean-Raphaël Guillaumin {[}photography{]} and T. Michael
+Keesey {[}vectorization{]}). Funding was provided by a Simon Fraser University
+Graduate Fellowship (SCA), the Natural Sciences and Engineering Research
+Council of Canada (NKD, ABC), the Canada Research Chairs Program (NKD), and
+TODO (TAB).
 

### 2014-10-29 text/ms.tex
Edit edit edit

@@
 Black swans are statistically improbable events that nonetheless occur ---
-often with profound implications. While extremes in the physical
-environment, such as monsoons and heat waves, are widely studied and increasing
-in magnitude and frequency, it remains unclear the extent to which ecological
-populations buffer or suffer from extremes. Here, we develop a probability
-model to estimate the degree of heavy-tailedness (presence of black swans) in
-ecological process noise. We apply our model to \NPops~time series from around
-the world across \NOrders~taxonomic orders and seven classes. We find strong
-evidence of black swans, but they are rare, occurring in
+often with profound implications. While extremes in the physical environment,
+such as monsoons and heat waves, are widely studied and increasing in magnitude
+and frequency, it remains unclear the extent to which ecological populations
+buffer or suffer from extremes. Here, we develop a probability model to
+estimate the degree of heavy-tailedness (presence of black swans) in ecological
+process noise. We apply our model to \NPops~time series from around the world
+across \NOrders~taxonomic orders and seven classes. We find strong evidence of
+black swans, but they are rare, occurring in
 \overallMinPerc--\overallMaxPerc\%of populations; most frequently for birds
@@
 consequences \citep{taleb2007,sornette2009}. One of the most striking black
-swans in ecology was the asteroid marking the mass extinction of the K-T
+swans in ecology is the asteroid marking the mass extinctions of the K-T
 boundary \citep{harnik2012}. Today, it is the extremeness of climate, in
@@
 appropriate for commonly available series has limited our ability to detect
-them. For example: (1) The generalized extreme value (GEV) distribution has
-been applied to environmental data \citep[e.g.][]{katz2005}. The GEV describes
-the distribution of the most extreme event per time interval (e.g.~heaviest
-rainfall per year), but this approach requires long time series that can be
-condensed into time intervals. (2) Thin- and heavy-tailed distributions have
+them. For example: (1) The generalized extreme value distribution has been
+applied to environmental data \citep[e.g.][]{katz2005}. This distribution
+describes the most extreme event per time interval (e.g.~heaviest rainfall per
+year), but requires long time series that can be condensed into time intervals.
+(2) Fitting a catastrophic mixture distribution in a state space model allows
+one to quantify the probability that population events are extreme but is also
+data intensive \citep{ward2007} (3) Thin- and heavy-tailed distributions have
 been competed with each other in an information theoretic framework
 \citep{halley2002}, but this analysis did not quantify the probability of black
-swans on a continuous scale or allow for population dynamics. (3) Fitting
-a catastrophic mixture distribution state space model allows one to quantify
-the probability that population events are extreme on a continuous scale but is
-data intensive \citep{ward2007}.
+swans on a continuous scale or allow for standard population dynamics.
 
@@
 represents population growth as a linear function in $\log$ space. If we let
-$x_t$ represent the $\log$ abundance ($N$) at time $t$, we can represent the Gompertz
-model as:
+$x_t$ represent the $\log$ abundance ($N$) at time $t$, we can represent the
+Gompertz model as:
+
+\vspace{-6mm}
 \begin{align*}
@@
 
-We extended the standard Gompertz model by allowing the process noise to be
+We extended the Gompertz model by allowing the process noise to be
 drawn from a Student-t distribution with degrees of freedom parameter $\nu$,
 and scale parameter $\sigma$:
+
+\vspace{-6mm}
 \begin{align*}
@@
 (Fig.~\ref{fig:didactic}).
-
-We chose weakly informative priors to incorporate our understanding of
-plausible population dynamics \citep[Supporting Material]{gelman2014}. For
-$\nu$, we chose an exponential prior with rate parameter of $0.01$ truncated at
-values above two (Fig.~\ref{fig:priors}) --- a slightly less informative prior
-than suggested by \citet{fernandez1998}. This prior gives only a 7.7\%
-probability that $\nu < 10$ but constrains the sampling sufficiently to avoid
-wandering off towards infinity --- above approximately $\nu = 20$ the
-t distribution is so similar to the normal distribution that time series of the
-length considered here are unlikely to be informative about the precise value
-of $\nu$ (Fig.~\ref{fig:didactic}).
 
@@
 \ref{fig:sim-gompertz}, \ref{fig:sim-gompertz-boxplots}).
+
+We chose weakly informative priors to incorporate our understanding of
+plausible population dynamics \citep[Supporting Material]{gelman2014}. For
+$\nu$, we chose an exponential prior with rate parameter of $0.01$ truncated at
+values above two (Fig.~\ref{fig:priors}) --- a slightly less informative prior
+than suggested by \citet{fernandez1998}. This prior gives only a 7.7\%
+probability that $\nu < 10$ but constrains the sampling sufficiently to avoid
+wandering off towards infinity --- above approximately $\nu = 20$ the
+t distribution is so similar to the normal distribution that time series of the
+length considered here are unlikely to be informative about the precise value
+of $\nu$ (Fig.~\ref{fig:didactic}).
 
@@
 life-history characteristics (body length and maximum lifespan obtained from
-\citet{brook2006a}) against our estimates of $\nu$. We formally investigated
-these relationships by fitting beta regression multilevel models to the
-probability that $\nu$ was less than $10$ (i.e.~approximately the probability
-of heavy tails). For covariates that were derived from Gompertz model parameter
-estimates, we incorporated standard deviations around the means. To account for
-broad patterns of phylogenetic relatedness, we allowed for varying intercepts
-at the taxonomic class, order, and species level (Supporting Material).
+\citet{brook2006a}) against our estimated probability of heavy tails, Pr$(\nu
+< 10) > 0.5$. We formally investigated these relationships by fitting beta
+regression multilevel models. For covariates that were derived from Gompertz
+model parameter estimates, we incorporated standard deviations around the
+means. To account for broad patterns of phylogenetic relatedness, we allowed
+for varying intercepts at the taxonomic class, order, and species level
+(Supporting Material).
 
@@
 
-We found strong but rare evidence for black-swan dynamics. Defining black-swan
-dynamics as a greater than $0.5$ probability that $\nu < 10$, we observed black
-swans most frequently for birds (\birdPH\%) followed by mammals (\mammalsPH\%),
-insects (\insectsPH\%), and fishes (\fishPH\%) (Fig.~\ref{fig:nu-coefs},
-\ref{fig:heavy-ts}). Black swans were taxonomically widespread, occurring for
-at least one population in \POrdersHeavy\% of the taxonomic orders recorded.
-Accounting for time series length and partially pooling inference across
-taxonomic class and order with a multilevel model, there was stronger evidence
-for black swans in insect populations than is apparent in
-Fig.~\ref{fig:nu-coefs} --- four of 10 orders with the highest median
-probability of heavy tails were insect orders --- however, there was
-considerable uncertainty in these estimates (Fig.~\ref{fig:order-estimates}).
+We found strong but rare evidence for black-swan population dynamics. Defining
+black-swan dynamics as a greater than $0.5$ probability that $\nu < 10$, our
+main Gompertz model found evidence for heavy tails most frequently for birds
+(\birdPH\%) followed by mammals (\mammalsPH\%), insects (\insectsPH\%), and
+fishes (\fishPH\%) (Figs~\ref{fig:nu-coefs}, \ref{fig:heavy-ts}). Black swans
+were taxonomically widespread, occurring for at least one population in
+\POrdersHeavy\% of the taxonomic orders recorded. Accounting for time series
+length and partially pooling inference across taxonomic class and order with
+a multilevel model, there was stronger evidence for black swans in insect
+populations than is apparent in Fig.~\ref{fig:nu-coefs} --- four of 10 orders
+with the highest median probability of heavy tails were insect orders ---
+however, there was considerable uncertainty in these estimates
+(Fig.~\ref{fig:order-estimates}).
 
@@
 a fixed quantity of observation error decreased our estimates of black-swan
-dynamics, increasing the median estimate of $\nu$ from $<10$ to $>10$ in
-$\baseNuTenObsTenSwitch$ of $\baseNuTen$ populations although the majority of
+dynamics, increasing the median estimate of $\nu$ from $<10$ to $\ge 10$ in
+$\baseNuTenObsTenSwitch$ of $\baseNuTen$ populations, although the majority of
 $\nu$ estimates remained qualitatively similar (Fig.~\ref{fig:alt})
@@
 length was the strongest covariate of observing black swan dynamics (as was
-similarly observed in simulation testing, Fig.~\ref{fig:sim-nu}). For example,
+also observed in simulation testing, Fig.~\ref{fig:sim-nu}). For example,
 we were about \pIncHeavyNThirtyNSixty~times more likely to observe heavy tails
@@
 a confluence of events. For example, a shortage of nest sites for the shag
-long-necked bird in the UK was thought to reduce population productivity and
-make the population vulnerable when a red tide hit in 1968 \citep{potts1980}.
-The population experienced a black swan downswing, but the new availability of
-nest sites resulted in a rapid population upswing \citep{potts1980}. As
-another example, an interaction between environmentally and predation-mediated
-population cycles are thought to have caused a downward black swan for a water
-vole population \citep{saucy1994}. Other black swans were the result of
-a sequence of extreme climate events on their own. For example, severe winters
-in 1929, 1940--1942, and 1962--1963 were associated with black swan downswings
-in grey heron abundance in the UK \citep{stafford1971}. Our analysis finds
-that the last event was a combination of two black swan events in a row and it
-took the population three times longer to recover than predicted
-\citep{stafford1971}.
+long-necked bird in the United Kingdom was thought to reduce population
+productivity and make the population vulnerable when a red tide hit in 1968
+\citep{potts1980}. The population experienced a black swan downswing, but the
+new availability of nest sites resulted in a rapid population upswing
+\citep{potts1980}. As another example, an interaction between environmental-
+and predation-mediated population cycles are thought to have caused a downward
+black swan for a water vole population \citep{saucy1994}. Other black swans
+were the result of a sequence of extreme climate events on their own. For
+example, severe winters in 1929, 1940--1942, and 1962--1963 were associated
+with black swan downswings in grey heron abundance in the United Kingdom
+\citep{stafford1971}. Our analysis finds that the last event was a combination
+of two black swan events in a row and it took the population three times longer
+to recover than predicted \citep{stafford1971}.
 
@@
 We found strong evidence that the process noise in ecological time series was
-heavy-tailed under the Gompertz or Ricker-logistic models in
+heavy-tailed under  Gompertz or Ricker-logistic models in
 \overallMinPerc--\overallMaxPerc\% of populations assessed. We detected black
@@
 version of the GPDD assessed the distribution of abundance time series but
-focused on identifying if the lognormal distribution was the most frequently
+focused on identifying if the log-normal distribution was the most frequently
 parsimonious model \citep{halley2002}. For heavy-tailed distributions,
@@
 bird population abundance. However, this finding was challenged by
-\citet{allen2001}, who showed that mixing data across species could create
+\citet{allen2001}, who showed that mixing data across species could
 falsely generate heavy tails. This rebuttal supports our use of the
 t distribution to represent heavy tails, since a t distribution can be
-represented by a mixture of normal distributions \citep[with
+represented by a mixture of normal distributions \citep[with the same mean and
 inverse-gamma-distributed variances,][]{gelman2014}.
@@
 of populations are located in the United Kingdom. Although we expect that we
-would find strong but rare evidence for black swans in other large taxonomic or
-geographic samples, the common forces driving those black swans (e.g.~severe
-winters in Table~1) likely differ. Second, in a large database with disparate
+would find qualitatively similar evidence for black swans in many other large
+taxonomic or
+geographic samples of populations, the common forces driving
+those black swans likely differ (e.g.~severe
+winters in Table~1). Second, in a large database with disparate
 data sources, we cannot ensure the quality of every data point; it is likely
-that some of the black swans in our dataset are recording mistakes mistakes.
-Conversely, recording mistakes could mask heavy tails if initial observations
+that some of the black swans in our dataset are recording mistakes.
+Conversely, recording mistakes could mask heavy tails if extreme observations
 are discarded or altered as assumed errors. Third, the temporal scales of
@@
 to detect them quickly after they happen? Can we isolate the components of
-ecological dynamics that experience black swans events by moving from
+ecological dynamics that experience black swans by moving from
 phenomenological models such as the Gompertz to mechanistic models, that take
@@
 through specific geographic and taxonomic subsets of data where longer time
-series are available \citep[e.g.][]{segura2013}. Finally, what is the impact of
+series with low levels of observation error are available
+\citep[e.g.][]{segura2013}. Finally, what is the impact of
 allowing for black swans in forecasts of ecological risk? Financial analysts
@@
 probabilities to focus early detection efforts and develop disaster response
-plans (e.g.~REF). A similar focus could benefit resource management once our
+plans (e.g.~REF-TODO). A similar focus might benefit resource management once our
 ability to predict the spatial probability and covariates of ecological black
@@
 opportunity to learn about ecological systems, evaluate when our models break
-down, and tune future management policy \citep{doak2008, pine-iii2009,
+down, and adjust future management policy \citep{doak2008, pine-iii2009,
   lindenmayer2010}. In recent decades ecology has moved toward focussing on
@@
 approaches infinity the distribution approaches the normal distribution. For
-example, at $\nu = 2$, the probability of drawing a value less than -5 is
+example, at $\nu = 2$, the probability of drawing a value more than five
+standard deviations below the mean is
 1.8\%, whereas the probability of drawing such a value from a normal
-distribution is nearly zero ($2.9\cdot10^{-5}$\%). (c--e) Simulated population
-dynamics from a Gompertz model with process noise drawn from t distributions
-with different values of $\nu$. Coloured dots in panels c and d represent jumps
-with less than a 1 in 1000 chance of occurring in a normal distribution. (f--h)
-Estimates of $\nu$ from models fit to the times series in panels c--e. Shown
-are the posterior samples (histograms), median and interquartile range of the
-posterior (IQR) (dots and line segments), and the exponential prior on $\nu$
-(dashed lines). Colour shading behind panels f--h illustrates the region of
-heavy tails.}
+distribution is nearly zero. (c--e) Simulated population dynamics from
+a Gompertz model with process noise drawn from t distributions with different
+values of $\nu$. Coloured dots in panels c and d represent jumps with less than
+a 1 in 1000 chance of occurring in a normal distribution. (f--h) Estimates of
+$\nu$ from models fit to the times series in panels c--e. Shown are the
+posterior samples (histograms), median and interquartile range of the posterior
+(IQR) (dots and line segments), and the exponential prior on $\nu$ (dashed
+lines). Colour shading behind panels f--h illustrates the region of heavy
+tails.}
 \label{fig:didactic}
@@
 populations in panel a are omitted for space. Red to yellow points highlight
-populations with a high probability of heavy-tailed dynamics.
+populations with a high to moderately high probability of heavy-tailed dynamics.
 } \label{fig:nu-coefs} \end{center} \end{figure}

### 2014-10-29 text/supp.tex
Edit edit edit

@@
 \begin{align*}
-b &\sim \mathrm{Uniform}(-1, 2)\\
-\lambda &\sim \mathrm{Normal}(0, 10^2)\\
-\sigma &\sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)\\
-\nu &\sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2). \end{align*}
+b &\sim \mathrm{Uniform}(-1, 2)\\ \lambda &\sim \mathrm{Normal}(0, 10^2)\\
+\sigma &\sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)\\ \nu &\sim
+\mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2). \end{align*}
 Our prior on $b$ was uninformative between values of $-1$ and $2$. We would not
 expect values of $b$ with levels of density dependence as low as $-1$ (very
-strong inverse density dependence), nor would we generally expect values above $1$. We
-allowed values of $b$ above $1$ to allow for somewhat non-stationary time
-series of growth rates. The estimates of $b$ were well within these bounds. Our
-prior on $\lambda$ was very weakly informative within the range of expected
-values for population growth and is similar to the default priors suggested by
-\citet{gelman2008d} for intercepts of regression models. Our prior on $\sigma$
-follows \citet{gelman2006c} and \citet{gelman2008d} and is based on our
-expected range of the value in nature from previous studies
-\citep[e.g.][]{connors2014}. In our testing of a subsample of populations, our
-parameter estimates were not qualitatively changed by switching to an
-uninformative uniform prior on $\sigma$, but the weakly informative Half-Cauchy prior
-substantially sped up chain convergence.
+strong inverse density dependence), nor would we generally expect values above
+$1$. We allowed values of $b$ above $1$ to allow for non-stationary time series
+of growth rates. The estimates of $b$ were well within these bounds. Our prior
+on $\lambda$ was very weakly informative within the range of expected values
+for population growth and is similar to the default priors suggested by
+\citet{gelman2008d} for intercepts of regression models. Our Half-Cauchy prior
+on $\sigma$ follows \citet{gelman2006c} and \citet{gelman2008d} and the scale
+parameter of $2.5$ is based on our expected range of the value in nature from
+previous studies \citep[e.g.][]{connors2014}. In our testing of a subsample of
+populations, our parameter estimates were not qualitatively changed by
+switching to an uninformative uniform prior on $\sigma$, but the weakly
+informative Half-Cauchy prior substantially sped up chain convergence.
 
-Our prior on $\nu$ was based on \citet{fernandez1998}. They chose a more
-informative exponential rate parameter of $0.1$. We chose a less informative
-rate parameter of $0.01$ and truncated the distribution at $2$, since at $\nu
-< 2$ the variance of the t distribution is undefined. This prior gives only
-a $7.7$\% probability $\nu < 10$ but constrains the sampling sufficiently to
-avoid wandering off towards infinity --- above approximately $\nu = 20$ the
+Our prior on $\nu$ was based on \citet{fernandez1998}. They chose an
+exponential rate parameter of $0.1$. We chose a less informative rate
+parameter of $0.01$ and truncated the distribution at $2$, since at $\nu < 2$
+the variance of the t distribution is undefined. This prior gives only
+a $7.7$\% probability that $\nu < 10$ but constrains the sampling sufficiently
+to avoid wandering off towards infinity --- above approximately $\nu = 20$ the
 t distribution is so similar to the normal distribution
 (Fig.~\ref{fig:didactic}) that time series of the length considered here are
-unlikely to be informative about the precise value of $\nu$. In the
-scenario where the data are not heavy tailed (e.g.~Fig.~\ref{fig:didactic}e, h)
-the posterior will approximately match the prior (prior median $= 71$, mean $= 102$)
-and not be flagged as likely heavy tailed using the metrics we used in our paper
-(e.g.~Pr$(\nu < 10) > 0.5$).
+unlikely to be informative about the precise value of $\nu$. In the scenario
+where the data are informative about heavy tails
+(e.g.~Fig.~\ref{fig:didactic}e,~h), the posterior will approximately match the
+prior (prior median $= 71$, mean $= 102$) and the metrics used in our paper
+(e.g.~Pr$(\nu < 10) > 0.5$) are unlikely to flag the population as heavy
+tailed.
 
@@
 \citep{r2014}. We began with four chains and $2000$ iterations, discarding the
-first $1000$ as warm up (i.e.~4000 total samples). If $\hat{R}$ (the scale
-reduction factor --- a measure of chain convergence) was greater than $1.05$
-for any parameter or the minimum effective sample size, $n_\mathrm{eff}$, (a
-measure of the effective number of uncorrelated samples) for any parameter was
-less than $200$, we doubled both the total iterations and warm up period and
-sampled from the model again. These thresholds are in excess of the minimums
-recommended by \citet{gelman2006a} of $\hat{R} < 1.1$ and effective sample size
-$> 100$ for reliable point estimates and confidence intervals. In the majority
-of cases our minimum thresholds were greatly exceeded. We continued this
-procedure up to $8000$ iterations ($16000$ total samples) by which all chains
-were deemed to have sufficiently converged. These chain lengths may seem low to
-those familiar with software such as WinBUGS or JAGS, but the No-U-Turn
-Hamiltonian Markov chain Monte Carlo Sampler in Stan generally requires far
-fewer iterations to obtain equivalent effective sample sizes \citep{stan-manual2014}.
-
-\section{Simulation testing the model}
-
-We performed two types of simulation testing. First, we tested how easily the
-Student-t distribution $\nu$ parameter could be recovered given different true
-values of $\nu$ and different sample sizes. Second, we tested the ability of the
-heavy-tailed Gompertz model to obtain unbiased parameter estimates of $\nu$
-given that a set of process deviations was provided in which the effective
-$\nu$ value was close to the true $\nu$ value.
-
-We separated our simulation into these two components to avoid confounding two
-issues. (1) With smaller sample sizes, there may not be a stochastic draw from
-the tails of a distribution. In that case, no model, no matter how perfect,
-will be able to detect the shape of the tails. (2) Complex models may return
-biased parameter estimates if there are conceptual, computational, or coding
-errors. Our first simulation tested the first issue and our second simulation
-tested the latter. In general, out simulations show that, if anything, our
-model under predicts the magnitude and frequency of heavy tailed events ---
-especially given the length of the time series in the GPDD.
-
-\subsection{Estimating $\nu$ simulations}
-
-First, we tested the ability to estimate $\nu$ given different true values of
-$\nu$ and different sample sizes. We took stochastic draws from t distributions with
-different $\nu$ values ($\nu = 3, 5, 10,$ and $10^6$ [$\approx$ normal]), with
-central tendency parameters of $0$, and scale parameters of $1$. We started with
-$1600$ stochastic draws and then fitted the models again at the first $800, 400,
-200, 100, 50,$ and $25$ draws. Each time we recorded the posterior samples of
-$\nu$.
-
-We found that we could consistently and precisely recover median posterior
-estimates of $\nu$ near the true value of $\nu$ with large samples ($\ge 200$)
-(Fig.~\ref{fig:sim-gompertz} upper panels). At smaller samples we could still
-usually qualitatively distinguish heavy from not-heavy tails, but the
-model tended to underestimate how heavy the tails were. At the same time, the
-model tended to overestimate how large the scale parameter was
-(Fig.~\ref{fig:sim-gompertz} lower panels).
-
-\subsection{Heavy-tailed Gompertz model simulations}
-
-In the second part of our simulation testing, we tested the ability of the
-heavy-tailed Gompertz model to obtain unbiased parameter estimates given that
-process deviations were provided in which our simpler model (previous
-section) could accurately estimate the value of $\nu$. To generate these process
-deviations, we repeatedly drew proposed candidate process deviations and estimated the
-central tendency, scale, and $\nu$ values each time. We recorded when
-$\hat{\nu}$ (median of the posterior) was within $0.2$ CVs (coefficient of
-variations) of the true $\nu$
-value and used this set of random seed values in our Gompertz simulation. We
-then fitted our Gompertz models to the simulated datasets with all parameters
-(except $\nu$) set near the median values estimated in the GPDD. We repeated
-this with $50$ and $100$ samples without observation error, $50$ samples with
-observation error ($\sigma_\mathrm{obs} = 0.2$), and $50$ samples with the same
-observation error and a Gompertz model that allowed for correctly specified
-observation error magnitude.
-
-Our results show that the more-complicated Gompertz model can still recapture
-the true value of $\nu$ when the process noise was chosen so that appropriate
-tail events were present (Figs~\ref{fig:sim-gompertz} and
-\ref{fig:sim-gompertz-boxplots}, red and green symbols in the top rows).
-Likewise, the other Gompertz parameters were estimated without any systematic
-bias (Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}, red and
-green symbols). The addition of observation error caused the model to tend to
-underestimate the degree of heavy-tailedness, overestimate the magnitude of
-process noise, somewhat overestimate $\lambda$, and overestimate density
-dependence (blue symbols in Figs~\ref{fig:sim-gompertz} and
-\ref{fig:sim-gompertz-boxplots}). The overestimation of density dependence with
-observation error is a known phenomenon \citep{knape2012}. Fitting a model with
-correctly specified observation error made marginal improvements to model
-bias (purple symbols in Figs~\ref{fig:sim-gompertz} and
-\ref{fig:sim-gompertz-boxplots}).
-
-When converting the posterior distributions of $\nu$ into Pr($\nu < 10$), the
-models distinguished heavy and not-heavy tails reasonably well. Without
-observation error, and using a probability of $0.5$ as a threshold, the model
-would have miscategorized only one of $40$ simulations at $\nu = 5$ across
-a sample size of $50$ or $100$ data points (Fig.~\ref{fig:sim-prob}, upper
-panels). The model would have correctly categorized all cases where the process
-noise was not heavy tailed (indicated as ``$\nu =$ infinity'' in
-Fig.~\ref{fig:sim-prob}) and all cases where $\nu = 3$. With $0.2$ standard
-deviations of observation error, the model still categorized XX\% of cases as
-heavy tailed when $\nu = 5$ and all cases where $\nu = 3$. Allowing for
-observation error made little improvement to the detection of heavy tails
-(Fig.~\ref{fig:sim-prob}, lower-left vs.\ lower-right panel). Therefore, we
-chose to focus on the simpler model without observation error in the main
-text, particularly given that the true magnitude of observation error was
-unknown in the empirical data.
+first $1000$ as warm up (i.e.~4000 total samples). If $\hat{R}$ (the potential
+scale reduction factor --- a measure of chain convergence) was greater than
+$1.05$ for any parameter or the minimum effective sample size,
+$n_\mathrm{eff}$, (a measure of the effective number of uncorrelated samples)
+for any parameter was less than $200$, we doubled both the total iterations
+and warm up period and sampled from the model again. These thresholds are in
+excess of the minimums recommended by \citet{gelman2006a} of $\hat{R} < 1.1$
+and effective sample size $> 100$ for reliable point estimates and confidence
+intervals. In the majority of cases our minimum thresholds were greatly
+exceeded. We continued this procedure up to $8000$ iterations ($16000$ total
+samples) by which all chains were deemed to have sufficiently converged. These
+chain lengths may seem low to those familiar with software such as WinBUGS or
+JAGS, but the No-U-Turn Hamiltonian Markov chain Monte Carlo Sampler in Stan
+generally requires far fewer iterations to obtain equivalent effective sample
+sizes \citep{stan-manual2014}.
 
@@
 In addition to the parameters in the original Gompertz model, we estimate an
-additional parameter $\phi$, which represents the relationship between
+additional parameter $\phi$, which represents the dependence of
 subsequent residuals. Based on the results of previous analyses with the GPDD
@@
 $\phi \sim \mathrm{Truncated\mhyphen Normal}(0, 1, \mathrm{min.} = -1,
-\mathrm{max.} = 1)$. The MCMC chains for a small number of these models (XX) did
-not converge according to our criteria ($\widehat{R} < 1.05, n_\mathrm{eff}
-> 200$) after 8000 iterations of four chains. We did not include these models in
-Fig.~\ref{fig:alt}.
+\mathrm{max.} = 1)$. The MCMC chains did
+not converge for \modelsNoConvergeAROne\ populations according to
+our criteria ($\widehat{R} < 1.05, n_\mathrm{eff}
+> 200$) after 8000 iterations of four chains. This included only
+\modelsNoConvergeAROneHeavyBase\ populations in which
+Pr($\nu < 10$) $> 0.5$ categorized them as heavy in the main Gompertz model.
+We did not include these models in Fig.~\ref{fig:alt}.
 
@@
 
+\section{Simulation testing the model}
+
+We performed two types of simulation testing. First, we tested how easily the
+Student-t distribution $\nu$ parameter could be recovered given different true
+values of $\nu$ and different sample sizes. Second, we tested the ability of the
+heavy-tailed Gompertz model to obtain unbiased parameter estimates of $\nu$
+given that a set of process deviations was provided in which the effective
+$\nu$ value was close to the true $\nu$ value.
+
+We separated our simulation into these two components to avoid confounding two
+issues. (1) With smaller sample sizes, there may not be a stochastic draw from
+the tails of a distribution. In that case, no model, no matter how perfect,
+will be able to detect the shape of the tails. (2) Complex models may return
+biased parameter estimates if there are conceptual, computational, or coding
+errors. Our first simulation tested the first issue and our second simulation
+tested the latter. In general, our simulations show that, if anything, our
+model under predicts the magnitude and probability of heavy tailed events ---
+especially given the length of the time series in the GPDD.
+
+\subsection{Estimating $\nu$ from a stationary t distribution}
+
+First, we tested the ability to estimate $\nu$ given different true values of
+$\nu$ and different sample sizes. We took stochastic draws from t distributions with
+different $\nu$ values ($\nu = 3, 5, 10,$ and $10^6$ [$\approx$ normal]), with
+central tendency parameters of $0$, and scale parameters of $1$. We started with
+$1600$ stochastic draws and then fitted the models again at the first $800, 400,
+200, 100, 50,$ and $25$ draws. Each time we recorded the posterior samples of
+$\nu$.
+
+We found that we could consistently and precisely recover median posterior
+estimates of $\nu$ near the true value of $\nu$ with large samples ($\ge 200$)
+(Fig.~\ref{fig:sim-gompertz} upper panels). At smaller samples we could still
+usually qualitatively distinguish heavy from not-heavy tails, but the model
+tended to underestimate how heavy the tails were. At the same time, at smaller
+sample sizes, the model tended to overestimate how large the scale parameter
+was (Fig.~\ref{fig:sim-gompertz} lower panels).
+
+\subsection{Heavy-tailed Gompertz model simulations}
+
+In the second part of our simulation testing, we tested the ability of the
+heavy-tailed Gompertz model to obtain unbiased parameter estimates given that
+process deviations were provided in which our simpler model (previous section)
+could accurately estimate the value of $\nu$. To generate these process
+deviations, we repeatedly drew proposed candidate process deviations and
+estimated the central tendency, scale, and $\nu$ values each time. We recorded
+when $\hat{\nu}$ (median of the posterior) was within $0.2$ CVs (coefficient of
+variations) of the true $\nu$ value and used this set of random seed values in
+our Gompertz simulation. We then fitted our Gompertz models to the simulated
+datasets with all parameters (except $\nu$) set near the median values
+estimated in the GPDD. We repeated this with $50$ and $100$ samples without
+observation error, $50$ samples with observation error ($\sigma_\mathrm{obs}
+= 0.2$), and $50$ samples with the same observation error and a Gompertz model
+that allowed for correctly specified observation error magnitude.
+
+Our results show that the Gompertz model can recapture the true value of $\nu$
+when the process noise was chosen so that appropriate tail events were present
+(Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}, red and green
+symbols in the top rows). Likewise, the other Gompertz parameters were
+estimated without any systematic bias (Figs~\ref{fig:sim-gompertz} and
+\ref{fig:sim-gompertz-boxplots}, red and green symbols). The addition of
+observation error caused the model to tend to underestimate the degree of
+heavy-tailedness, overestimate the magnitude of process noise, somewhat
+overestimate $\lambda$, and overestimate density dependence (blue symbols in
+Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}). The
+overestimation of density dependence with observation error is a known
+phenomenon \citep{knape2012}. Fitting a model with correctly specified
+observation error made marginal improvements to model bias (purple symbols in
+Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}).
+
+When converting the posterior distributions of $\nu$ into Pr($\nu < 10$), the
+models distinguished heavy and not-heavy tails reasonably well. Without
+observation error, and using a probability of $0.5$ as a threshold, the model
+correctly classified all simulated systems with normally distributed process
+noise as not heavy tailed. The model would have miscategorized only one of $40$
+simulations at $\nu = 5$ across a sample size of $50$ or $100$ data points
+(Fig.~\ref{fig:sim-prob}, upper panels). The model would have correctly
+categorized all cases where the process noise was not heavy tailed (indicated
+as ``$\nu =$ infinity'' in Fig.~\ref{fig:sim-prob}) and all cases where $\nu
+= 3$. With $0.2$ standard deviations of observation error, the model still
+categorized \obsErrorNuFivePerc\% of cases as heavy tailed when $\nu = 5$ and
+all cases where $\nu = 3$. Allowing for observation error made little
+improvement to the detection of heavy tails (Fig.~\ref{fig:sim-prob},
+lower-left vs.\ lower-right panel). Therefore, we chose to focus on the simpler
+model without observation error in the main text, particularly given that the
+true magnitude of observation error was unknown in the empirical data.
+
 \section{Modelling covariates of heavy-tailed dynamics}
@@
 around the predictors by incorporating the standard deviation of the posterior
-samples for the Gompertz parameters $\lambda$, $b$, and $\sigma$ around the
+samples for the Gompertz parameters $\lambda$, $b$, and $\log \sigma$ around the
 mean point value as normal distributions (not shown in the above equation).
@@
 other parameters \citep{gelman2006c, gelman2008d}. Compared to normal priors,
-the Cauchy priors concentrate more probability density at reasonably parameter
-values while allowing for higher probability density far into the tails, thereby
+the Cauchy priors concentrate more probability density around expected parameter
+values while allowing for a higher probability density far into the tails, thereby
 allowing the data to dominate the posterior more strongly if it disagrees with
@@
 checked for chain convergence visually and with the same criteria as before
-($\widehat{R} < 1.05$ and $n_\mathrm{eft} >200$ for all parameters).
+($\widehat{R} < 1.05$ and $n_\mathrm{eff} >200$ for all parameters).
 
@@
 accounting for time-series length (Fig \ref{fig:order-estimates}), we fitted
-a separate multilevel model with the same structure but with only time-series
-length as a predictor. (In this case, we did not want to control for intrinsic
-population characteristics such as density dependence.) Since our predictors were
-centered by subtracting their mean value, we obtained order-level
-estimates of the probability of heavy tails at mean time-series length by
-adding the posteriors for $\alpha$, $\alpha^\mathrm{class}_j$, and
+a separate multilevel model with the same structure but with only $\log$
+time-series length as a predictor. (In this case, we did not want to control
+for intrinsic population characteristics such as density dependence.) Since our
+predictors were centered by subtracting their mean value, we obtained
+order-level estimates of the probability of heavy tails at mean log time-series
+length by adding the posteriors for $\alpha$, $\alpha^\mathrm{class}_j$, and
 $\alpha^\mathrm{order}_k$.
@@
   Probability density of the Bayesian priors for the Gompertz models. From left
-  to right: (1) per capita growth rate at $\ln$(abundance) = $0$: $\lambda \sim
+  to right: (1) per capita growth rate at $\log$(abundance) = $0$: $\lambda \sim
   \mathrm{Normal}(0, 10^2)$; (2) scale parameter of t-distribution process
@@
   true $\nu$ value; however, heavy tails are still distinguished from normal
-  tails in most cases, even with only 25 or 50 samples. TODO switch to nu
-  = black swan, 3, 5, and normal.
+  tails in most cases, even with only 25 or 50 samples.
 }
@@
   Simulation testing the Gompertz estimation model when the process deviation
-  draws were chosen so that $\nu$ can be estimated close to the true value
+  draws were chosen so that $\nu$ could be estimated close to the true value
   outside the full population model (``effective $\nu$'' within a CV of 0.2 of
@@
   population values; these true values were chosen to represent approximately
-  the median values as estimated from the GPDD. Individual dots and lines
+  the median values as estimated from the Global Population Dynamics Database.
+  Individual dots and lines
   represent a stochastic draws from the true population distribution and
@@
   whiskers extend to $1.5$ times the interquartile range and outliers are shown
-  as dots.
+  as dots. Note that these distributions represent the output from only 20
+  simulation iterations, so there is some level of remaining sampling noise.
 }
@@
 \caption{
-  Time series for populations with p($\nu < 10$) $>$ 0.5 using the base
+  Time series for populations with Pr($\nu < 10$) $>$ 0.5 using the base
   Gompertz population model. Panels are ordered by increasing p($\nu < 10$).
-  Vertical axes are on a log10 scale. Colours indicate taxonomic classes. TODO
-  add dots for upswings and downswings as in Table 1. The labels on each panel
-  indicate p($\nu < 10$) $>$ 0.5, the common name for the species, and the GPDD
-  ID number.
+  Vertical axes are on a log10 scale. Colours indicate taxonomic classes. The
+  labels on each panel indicate p($\nu < 10$) $>$ 0.5, the common name for the
+  species, and the Global Population Dynamics Database ID number.
 }
@@
   as a threshold of heavy-tailed dynamics. The Gompertz observation error model
-  assumes a fixed standard deviation of observation error of $0.2$ on the log
+  assumes a fixed standard deviation of observation error of $0.2$ on a log
   scale.
@@
 Example Stan code for a heavy-tailed Gompertz model with AR1 correlated
-residuals and a specified level of observation error:
+residuals and a specified level of observation error. The specific code for used for the various models in our analysis is available at \url{https://github.com/seananderson/heavy-tails}.
 
@@
 \noindent
-The GPDD IDs used in our analysis:
+The GPDD IDs used in our analysis. See \url{https://github.com/seananderson/heavy-tails}.
 

### 2014-10-30 text/ms.tex
Fix line number - align conflict in preamble
Minor edits to ms

@@
 
-\noindent
-\emph{Main message, to be deleted (25 words)}: Black swans are present \&
-taxonomically widespread but rare in population dynamics. Extreme climate,
-predation, parasites, and their interactions are common causes; intrinsic
-drivers are unclear.
+%\noindent
+%\emph{Main message, to be deleted (25 words)}: Black swans are present \&
+%taxonomically widespread but rare in population dynamics. Extreme climate,
+%predation, parasites, and their interactions are common causes; intrinsic
+%drivers are unclear.
 
@@
 detected for populations with longer time series and lower levels of process
-noise. We find little evidence of intrinsic life-history covariates. TODO close
-with modelling and policy implications
+noise. We find little evidence of intrinsic life-history covariates. \textit{TODO close
+with modelling and policy implications. Something like:
+The presence of black swans highlights the importance of developing management strategies that detect quickly, respond to, and are robust to extremes in population dynamics --- particularly as the frequency and magnitude of climatic extremes increase over the next century.}
 
@@
 Gompertz model as:
-
-\vspace{-6mm}
 \begin{align*}
@@
 and scale parameter $\sigma$:
-
-\vspace{-6mm}
 \begin{align*}
@@
 aspects of variance in addition to mean responses \citep[e.g.][]{loreau2010a,
-  thompson2013}. Our results suggests that an added focus on extremes
+  thompson2013}. Our results suggest that an added focus on extremes
 represents the next frontier, particularly in the face expected increases in

### 2014-11-01 text/ms.tex
Major edits to ms
After meeting with Nick, also reorder and incorporate
the supplementary figs

@@
 Black swans are statistically improbable events that nonetheless occur ---
-often with profound implications. While extremes in the physical environment,
+often with profound consequences. While extremes in the physical environment,
 such as monsoons and heat waves, are widely studied and increasing in magnitude
 and frequency, it remains unclear the extent to which ecological populations
-buffer or suffer from extremes. Here, we develop a probability model to
+buffer or suffer from such extremes. Here, we develop a probability model to
 estimate the degree of heavy-tailedness (presence of black swans) in ecological
@@
 across \NOrders~taxonomic orders and seven classes. We find strong evidence of
-black swans, but they are rare, occurring in
-\overallMinPerc--\overallMaxPerc\%of populations; most frequently for birds
+black swans, but they are rare occurring in
+\overallMinPerc--\overallMaxPerc\% of populations: most frequently for birds
 (\AvesRangePerc\%) followed by mammals (\MammaliaRangePerc\%), insects
-(\InsectaRangePerc\%), and fishes(\OsteichthyesRangePerc\%). When they occur,
-they tend to be driven by climate and severe winters, cycles of parasites and
-predators, and interactions between these elements. They are more frequently
-detected for populations with longer time series and lower levels of process
-noise. We find little evidence of intrinsic life-history covariates. \textit{TODO close
-with modelling and policy implications. Something like:
-The presence of black swans highlights the importance of developing management strategies that detect quickly, respond to, and are robust to extremes in population dynamics --- particularly as the frequency and magnitude of climatic extremes increase over the next century.}
+(\InsectaRangePerc\%), and fishes (\OsteichthyesRangePerc\%). We find little
+evidence of intrinsic life-history covariates, but when they occur, they tend
+to be extrinsically driven by climate and severe winters, cycles of parasites
+and predators, and interactions between these elements. Black swans were more
+frequently detected for populations with longer time series and lower levels of
+process noise. The mere presence of black swans in population dynamics
+highlights the importance of developing robust conservation and management
+strategies --- particularly as the frequency and magnitude of climatic extremes
+increase over the next century.
 
@@
 swans in ecology is the asteroid marking the mass extinctions of the K-T
-boundary \citep{harnik2012}. Today, it is the extremeness of climate, in
-concert with shifts in mean temperature, that is expected to cause the greatest
+boundary TODO-REF \citep{harnik2012}. Today, climate extremes, in concert with
+shifts in mean temperature, are expected to cause the greatest ecological and
 societal damage \citep{ipcc2012}. But, while extremes in the physical
-environment such as wave height, storm severity, and temperature are present and
-widely accepted \citep{gaines1993,katz2005}, it remains unclear the extent to
-which ecological systems buffer or suffer from black swans \citep{nunez2012}.
+environment such as wave height, storm severity, and temperature are are
+frequent experiences \citep{gaines1993,katz2005}, it remains unclear the extent
+to which ecological systems buffer or suffer from black swans
+\citep{nunez2012}.
 
-%Existing evidence for black swans or heavy tails in population dynamics is
-%mostly anecdotal or for particular systems.
 There is compelling anecdotal evidence for ecological black swans, but
 systematic evidence has been elusive. A survey of ecologists indicated that
-surprising results in field experiments may be far more common than we assume
+surprising outcomes of field experiments are far more common than we assume
 \citep{doak2008}, and a number of major environmental events on geologic and
 contemporary time scales could be considered to have caused ecological black
-swans \citep{nunez2012}. Other work has detected heavy-tail dynamics for
-particular systems. For example, \citet{segura2013} identified heavy tails in
-time series of marine microbe populations and \citet{drake2014} identified
-heavy tails in time to extinction for experimental waterflea populations.
-However, systematic evidence has been limited. What appeared to be heavy tails
-in breeding bird abundance \citep{keitt1998}, could also be explained by mixing
-of species in the original dataset \citep{allen2001}. A subsequent large scale
-analysis of the Global Population Dynamics Database (GPDD) found little
-evidence for heavy-tail distributions in longer abundance time series
-\citet{halley2002}.
+swans \citep{nunez2012}.
+% TODO: pull anecdote from that study
+Black swans, or heavy tailed distributions, have been discovered in time-series
+of marine microbe populations \citet{segura2013} and time to extinction for
+experimental waterflea populations \citet{drake2014}. As far as we can tell
+there have been two systematic surveys for ecological black swans, one in North
+American breeding birds and the Global Population Dynamics Database, that found
+little clear evidence for them \citep{keitt1998,allen2001,halley2002}.
+Nevertheless there are methodological challenges to their detection
+\citep{allen2001,ward2007}.
 
 There are two key reasons why we may find little evidence of ecological black
-swans. First, they might not systematically exist. Indeed, the majority of
+swans. First, they might not exist in higher taxa. Indeed, the majority of
 model fitting and risk forecasting assumes that population dynamics are not
 heavy tailed \citep[e.g.][]{brook2006a,dennis2006,knape2012}. Alternatively,
-dynamics might be heavy tailed, but a paucity of statistical methods
-appropriate for commonly available series has limited our ability to detect
-them. For example: (1) The generalized extreme value distribution has been
-applied to environmental data \citep[e.g.][]{katz2005}. This distribution
-describes the most extreme event per time interval (e.g.~heaviest rainfall per
-year), but requires long time series that can be condensed into time intervals.
-(2) Fitting a catastrophic mixture distribution in a state space model allows
-one to quantify the probability that population events are extreme but is also
-data intensive \citep{ward2007} (3) Thin- and heavy-tailed distributions have
-been competed with each other in an information theoretic framework
+heavy-tailed dynamics might exist, but our ability to detect them requires
+further development of statistical tools. For example: (1) The generalized
+extreme value distribution has been applied to environmental data
+\citep[e.g.][]{katz2005}. This distribution describes the most extreme event
+per time interval (e.g.~heaviest rainfall per year), but requires time series
+that are sufficiently long to be condensed into time intervals. (2) Fitting
+a catastrophic mixture distribution in a state space model allows one to
+quantify the probability that population events are extreme but is also data
+intensive \citep{ward2007} (3) Thin- and heavy-tailed distributions have been
+competed with each other in an information theoretic framework
 \citep{halley2002}, but this analysis did not quantify the probability of black
@@
 lengths of abundance time series. We apply our model to hundreds of populations
-from around the world across four taxonomic classes to address three questions:
-(1) how frequent and strong are black swans in population dynamics across
-taxonomy, (2) what are some verified causes of black swans in ecological time
-series, and (3) are there characteristics of time series or intrinsic
-life-history characteristics associated with heavy-tailed dynamics. We find
-a high probability of ecological black swans across a wide range of taxa but in
-a limited number of populations. When they occur, extreme climate, predation,
-parasites, and their interactions are common documented causes. We find little
-evidence that intrinsic life-history characteristics are related to the
-probability of black swans.
+from across four taxonomic classes to address three questions: (1) how frequent
+and strong are black swans in population dynamics across taxonomy, (2) what are
+some verified causes of black swans in ecological time series, and (3) are
+there characteristics of time series or intrinsic life-history characteristics
+associated with heavy-tailed dynamics. We find a high probability of ecological
+black swans across a wide range of taxa but in a limited number of populations.
+When they occur, extreme climate, predation, parasites, and their interactions
+are common documented causes. We find little evidence that intrinsic
+life-history characteristics are related to the probability of black swans.
 
@@
 To obtain estimates of the probability and magnitude of black swans, we fitted
-population dynamic models to hundreds of abundance time series from around the
-world. The key addition of our models, compared to typical population models,
-was estimating the shape of the process noise tails. We then compared how these
-estimates related to taxonomy, time series properties, and life-history
-characteristics, and for a subset of populations, we documented the established
-causes of ecological black swans.
+population dynamic models to abundance time series from around the world. The
+key novelty of our approach, compared to typical population models, was
+estimating the shape of the process noise tails by measuring the degrees of
+freedom ($\nu$) of a Student t-distribution.
 
@@
 less reliable data sources, removed those without sufficient data for our
-models, and interpolated some missing values. Our final dataset contained
-\NPops~populations across \NOrders~taxonomic orders and seven taxonomic classes
-with a median of \medianTimeSteps~time steps (range of
-\minTimeSteps--\maxTimeSteps) (Table S1, Fig.~\ref{fig:all-ts}).
+models, and interpolated some missing values \citep[\textit{sensu}][]{brook2006a}.
+Our final dataset contained \NPops~populations across \NOrders~taxonomic orders
+and seven taxonomic classes with a median of \medianTimeSteps~time steps (range
+of \minTimeSteps--\maxTimeSteps) (Table S1, Fig.~\ref{fig:all-ts}).
 
@@
 \end{align*}
-The parameter $\lambda$ represents the expected $\log$ number of new
-individuals at time $t + 1$ if $N_t = 1$. The model is density independent if
-$b = 1$, maximally density dependent if $b = 0$, and inversely density
-dependent if $b < 0$. The process noise $\epsilon_t$ is modelled as normally
-distributed with standard deviation $\sigma$. In previous analyses of the GPDD,
-the Gompertz has proved to be the most frequently parsimonious population model
-\citep{brook2006}.
+The growth parameter $\lambda$ represents the expected growth rate if $N_t
+= 1$. The model is density independent if $b = 1$, maximally density dependent
+if $b = 0$, and inversely density dependent if $b < 0$. The process noise
+$\epsilon_t$ is usually modelled as normally distributed with standard
+deviation $\sigma$. In previous analyses of the GPDD, the Gompertz has proved
+to be the most frequently parsimonious population model \citep{brook2006}.
 
@@
 If $\nu$ is small ($\lesssim 10$), the t distribution has much heavier tails
-than a normal distribution. For example, at $\nu = 2$, the probability of
-drawing a value more than five standard deviations below the mean is $1.8$\%,
-whereas the probability of drawing such a value from a normal distribution is
-nearly zero ($2.9\cdot10^{-5}$\%). As $\nu$ approaches infinity the
-distribution approaches the normal distribution. By estimating the value of
-$\nu$ we can quantity how heavy-tailed the process noise deviations are
-(Fig.~\ref{fig:didactic}).
+than a normal distribution (Fig.~\ref{fig:didactic}a, b). For example, at $\nu
+= 2$, the probability of drawing a value more than five standard deviations
+below the mean is $1.8$\%, whereas the probability of drawing such a value from
+a normal distribution is nearly zero. As $\nu$ approaches infinity the
+distribution approaches the normal distribution (Fig.~\ref{fig:didactic}a, b).
+By estimating the value of $\nu$ we can quantity how heavy-tailed the process
+noise deviations are.
+
+% TODO: cite panels c, d in Fig 1
 
@@
 sizes and to ensure we could recover unbiased parameter estimates from the
-Gompertz model (Supporting Material, Fig.~\ref{fig:sim-nu},
-\ref{fig:sim-gompertz}, \ref{fig:sim-gompertz-boxplots}).
+Gompertz model (Supporting Material).
 
 We chose weakly informative priors to incorporate our understanding of
-plausible population dynamics \citep[Supporting Material]{gelman2014}. For
-$\nu$, we chose an exponential prior with rate parameter of $0.01$ truncated at
+plausible population dynamics 
+(\citeauthor{gelman2014} \citeyear{gelman2014}; Supporting Material). 
+For $\nu$, we chose an exponential prior with rate parameter of $0.01$ truncated at
 values above two (Fig.~\ref{fig:priors}) --- a slightly less informative prior
-than suggested by \citet{fernandez1998}. This prior gives only a 7.7\%
+than suggested by \citet{fernandez1998}. This prior gives only a $7.7$\%
 probability that $\nu < 10$ but constrains the sampling sufficiently to avoid
@@
 
-We fit four alternative population models to test if they systematically
-changed our conclusions. (1) Autocorrelation has been suggested as a reason for
-increased observed variability of abundance time series through time, which
-could create apparent heavy tails \citep{inchausti2002}; therefore, we fit
-a model that modelled the serial correlation in the residuals. (2) Previous
-work has modelled abundance or growth rates without accounting for density
-dependence \citep{halley2002,segura2013}; therefore, we fit a simpler model in
-which we assumed density independence. (3) Observation error could bias
-parameter estimates \citep{knape2012} or mask our ability to detect heavy tails
-\citep{ward2007}; therefore, we fit a model where we allowed for a fixed
+We fit alternative population models to test if four key phenomena
+systematically changed our conclusions. (1) Autocorrelation has been suggested
+as a reason for increased observed variability of abundance time series through
+time, which could create apparent heavy tails \citep{inchausti2002}; therefore,
+we fit a model that modelled the serial correlation in the residuals. (2)
+Previous work has modelled abundance or growth rates without accounting for
+density dependence \citep{halley2002,segura2013}; therefore, we fit a simpler
+model in which we assumed density independence. (3) Observation error could
+bias parameter estimates \citep{knape2012} or mask our ability to detect heavy
+tails \citep{ward2007}; therefore, we fit a model where we allowed for a fixed
 quantity of observation error ($0.2$ standard deviations on a log scale). (4)
@@
 primary data source cited in the GPDD or in other literature describing the
-population. These populations were chosen haphazardly with the purpose of
-generating hypotheses about the causes of population dynamic black swans.
+population. These exemplar populations were chosen haphazardly with the purpose
+of generating hypotheses about the causes of population dynamic black swans.
 
@@
 
-We found strong but rare evidence for black-swan population dynamics. Defining
-black-swan dynamics as a greater than $0.5$ probability that $\nu < 10$, our
-main Gompertz model found evidence for heavy tails most frequently for birds
-(\birdPH\%) followed by mammals (\mammalsPH\%), insects (\insectsPH\%), and
-fishes (\fishPH\%) (Figs~\ref{fig:nu-coefs}, \ref{fig:heavy-ts}). Black swans
-were taxonomically widespread, occurring for at least one population in
-\POrdersHeavy\% of the taxonomic orders recorded. Accounting for time series
-length and partially pooling inference across taxonomic class and order with
-a multilevel model, there was stronger evidence for black swans in insect
-populations than is apparent in Fig.~\ref{fig:nu-coefs} --- four of 10 orders
-with the highest median probability of heavy tails were insect orders ---
-however, there was considerable uncertainty in these estimates
+We found strong, but rare, evidence for black-swan population dynamics. By
+defining black-swan dynamics as a greater than $0.5$ probability that $\nu
+< 10$, our main Gompertz model found evidence for heavy tails most frequently
+for birds (\birdPH\%) followed by mammals (\mammalsPH\%), insects
+(\insectsPH\%), and fishes (\fishPH\%) (Figs~\ref{fig:nu-coefs},
+\ref{fig:heavy-ts}). Black swans were taxonomically widespread, occurring in
+\POrdersHeavy\% of taxonomic orders. Accounting for time series length and
+partially pooling inference across taxonomic class and order with a multilevel
+model, there was stronger evidence for black swans in insect populations than
+is visually apparent in Fig.~\ref{fig:nu-coefs} --- four of 10 orders with the
+highest median probability of heavy tails were insect orders --- however, there
+was considerable uncertainty in these estimates
 (Fig.~\ref{fig:order-estimates}).
@@
 The majority of our heavy-tailed estimates were robust to alternative
-population models and observation error (Fig.~\ref{fig:alt}). Including an
-autocorrelation structure in the residuals, modelling population growth rates
-without density dependence, or modelling the population dynamics as
-Ricker-logistic did not systematically alter our conclusions. Allowing for
-a fixed quantity of observation error decreased our estimates of black-swan
-dynamics, increasing the median estimate of $\nu$ from $<10$ to $\ge 10$ in
-$\baseNuTenObsTenSwitch$ of $\baseNuTen$ populations, although the majority of
-$\nu$ estimates remained qualitatively similar (Fig.~\ref{fig:alt})
+population models and observation error (Fig.~\ref{fig:alt}). Three of the four
+population dynamic phenomena were unimportant:
+% TODO: I don't like "unimportant"
+including an autocorrelation structure in the residuals, modelling population
+growth rates without density dependence, or modelling the population dynamics
+as Ricker-logistic did not systematically alter our conclusions. However,
+allowing for a fixed quantity of observation error decreased our estimates of
+black-swan dynamics slightly, increasing the median estimate of $\nu$ from
+$<10$ to $\ge 10$ in $\baseNuTenObsTenSwitch$ of $\baseNuTen$ populations,
+although the majority of $\nu$ estimates remained qualitatively similar
+(Fig.~\ref{fig:alt}).
+Our simulation testing show that, if anything, our models underpredict the true
+magnitude and probability of heavy tailed events --- especially given the
+length of the time series in the GPDD (Figs~\ref{fig:sim-nu},
+\ref{fig:sim-gompertz}, \ref{fig:sim-gompertz-boxplots}, \ref{fig:sim-prob})
 
@@
 ($\lambda$), density dependence ($b$), or maximum lifespan
-(Fig.~\ref{fig:correlates}, \ref{fig:correlate-coefs}). Longer time-series
+(Figs~\ref{fig:correlates}, \ref{fig:correlate-coefs}). Longer time-series
 length was the strongest covariate of observing black swan dynamics (as was
@@
 We assembled the documented causes of black swan population dynamics in seven
-populations with high probabilities of heavy tails (Table 1). The majority of
-documented events were downward black swans and the majority involved
-a confluence of events. For example, a shortage of nest sites for the shag
-long-necked bird in the United Kingdom was thought to reduce population
+exemplar populations with high probabilities of heavy tails (Table 1). The
+majority of documented events were downward black swans and the majority
+involved a confluence of events. For example, a shortage of nest sites for the
+shag long-necked bird in the United Kingdom was thought to reduce population
 productivity and make the population vulnerable when a red tide hit in 1968
@@
 magnitude of process error, as we might expect. However, we failed to find
-strong evidence that black swans were associated with density dependence,
-population growth rate, or lifespan. In cases that we verified, black swans
-were often a result of the interaction between elements of extreme climate,
-predators and parasite cycles, and strong changes in human pressures. Our
-empirical results, sensitivity analyses, and simulation tests suggest that
-estimating the tail shape of process noise is a viable method of detecting
-black swan population dynamics and if anything will underestimate the
-probability of black swans. The presence of black swans highlights the
-importance of developing management strategies that detect quickly, respond to,
-and are robust to extremes in population dynamics --- particularly as the
-frequency and magnitude of climatic extremes increase over the next century
-\citep{easterling2000,ipcc2012}.
+strong evidence that black swans were associated with the intrinsic ecological
+phenomena of density dependence, population growth rate, or lifespan. In cases
+that we verified, black swans were often a result of the interaction between
+elements of extreme climate, predators and parasite cycles, and strong changes
+in human pressures. Our empirical results, sensitivity analyses, and simulation
+tests suggest that estimating the tail shape of process noise is a viable
+method of detecting black swan population dynamics and if anything will
+underestimate the probability of black swans. The presence of black swans
+highlights the importance of developing management strategies that detect
+quickly, respond to, and are robust to extremes in population dynamics ---
+particularly as the frequency and magnitude of climatic extremes increase over
+the next century \citep{easterling2000,ipcc2012}.
 
@@
 \citep[e.g.][]{nunez2012}. These forces could be extreme themselves. For
-example, extreme climate, predation from or competition with other species
+example, extreme climate, predation from (or competition with) other species
 experiencing black swans, or sharp changes in human pressure such hunting,
@@
 interaction of multiple ``normal'' extrinsic forces could give rise to black
-swan ecological dynamics. This could occur if the interaction nature is
-synergistic \citep[e.g.][]{kirby2009} or even if non-synergistic interactions
-experience a rare alignment \citep{denny2009}.
+swan ecological dynamics. This could occur if the interaction is synergistic
+\citep[e.g.][]{kirby2009} or even if non-synergistic interactions experience
+a rare alignment \citep{denny2009}.
 
@@
 would find qualitatively similar evidence for black swans in many other large
-taxonomic or
-geographic samples of populations, the common forces driving
-those black swans likely differ (e.g.~severe
-winters in Table~1). Second, in a large database with disparate
-data sources, we cannot ensure the quality of every data point; it is likely
-that some of the black swans in our dataset are recording mistakes.
-Conversely, recording mistakes could mask heavy tails if extreme observations
-are discarded or altered as assumed errors. Third, the temporal scales of
-observation and population dynamics vary considerably across populations in the
-GPDD and this likely influences the detection of heavy tails. For example, if
-we make frequent observations relative to generation time (e.g.~for many
-large-bodied mammals) we will average across generations and perhaps miss black
-swans. Furthermore, if we census populations infrequently relative to
-generation time (e.g.~many insects in the GPDD) the recorded data may average
-across extreme and less-extreme events and dampen black swan dynamics.
+taxonomic or geographic samples of populations, the common forces driving those
+black swans likely differ (e.g.~severe winters in Table~1). Second, in a large
+database with disparate data sources, we cannot ensure the quality of every
+data point; it is likely that some of the black swans in our dataset are
+recording mistakes. Conversely, recording mistakes could mask heavy tails if
+extreme observations are discarded or altered as assumed errors. Third, the
+temporal scales of observation and population dynamics vary considerably across
+populations in the GPDD and this likely influences the detection of heavy
+tails. For example, if we make frequent observations relative to generation
+time (e.g.~for many large-bodied mammals) we will average across generations
+and perhaps miss black swans. Furthermore, if we census populations
+infrequently relative to generation time (e.g.~many insects in the GPDD) the
+recorded data may average across extreme and less-extreme events and dampen
+black swan dynamics.
 
@@
 to detect them quickly after they happen? Can we isolate the components of
-ecological dynamics that experience black swans by moving from
-phenomenological models such as the Gompertz to mechanistic models, that take
-into account, for example, recruitment? We expect that greater insight into the
-mechanisms and covariates of ecological black swans may be best obtained
-through specific geographic and taxonomic subsets of data where longer time
-series with low levels of observation error are available
-\citep[e.g.][]{segura2013}. Finally, what is the impact of
-allowing for black swans in forecasts of ecological risk? Financial analysts
-are concerned with the shape of financial-return-distribution tails because
-these impact estimates of risk --- the probability of a specific magnitude of
-undesired event occurring \citep{rachev2008}. It is possible that a comparable
-focus in ecology would alter our perception of ecological forecasts such as
-extinction risk.
+ecological dynamics that experience black swans by moving from phenomenological
+models such as the Gompertz to mechanistic models, that take into account, for
+example, recruitment? We expect that greater insight into the mechanisms and
+covariates of ecological black swans may be best obtained through specific
+geographic and taxonomic subsets of data where longer time series with low
+levels of observation error are available \citep[e.g.][]{segura2013}. Finally,
+what is the impact of allowing for black swans in forecasts of ecological risk?
+Financial analysts are concerned with the shape of
+financial-return-distribution tails because these impact estimates of risk ---
+the probability of a specific magnitude of undesired event occurring
+\citep{rachev2008}. It is possible that a comparable focus in ecology would
+alter our perception of ecological forecasts such as extinction risk.
 
@@
 directions. First, ecological resource management can draw from other
-disciplines focussed on heavy tails. For example, earthquake preparedness and
+disciplines that focus on heavy tails. For example, earthquake preparedness and
 response is focussed on black swan events. Similarly to ecological black swans,
@@
 probabilities to focus early detection efforts and develop disaster response
-plans (e.g.~REF-TODO). A similar focus might benefit resource management once our
-ability to predict the spatial probability and covariates of ecological black
-swans improves. Second, we can choose management policy that is robust to heavy
-tails. For example, setting target population abundances that are appropriately
-far from critical limits may buffer black swan events
+plans (e.g.~REF-TODO). A similar focus might benefit resource management once
+our ability to predict the spatial probability and covariates of ecological
+black swans improves. Second, we can develop management policy that is robust
+to heavy tails. For example, setting target population abundances that are
+appropriately set back from critical limits may buffer black swan events
 \citep[e.g.][]{caddy1996}, and maintaining genetic, phenotypic, and behavioural
@@
 aspects of variance in addition to mean responses \citep[e.g.][]{loreau2010a,
-  thompson2013}. Our results suggest that an added focus on extremes
-represents the next frontier, particularly in the face expected increases in
-climate extremes \citep{meehl2004,ipcc2012}.
+  thompson2013}. Our results suggest that an added focus on ecological extremes
+represents the next frontier, particularly in the face of increased climate
+extremes \citep{meehl2004,ipcc2012}.
 
@@
 example, at $\nu = 2$, the probability of drawing a value more than five
-standard deviations below the mean is
-1.8\%, whereas the probability of drawing such a value from a normal
-distribution is nearly zero. (c--e) Simulated population dynamics from
-a Gompertz model with process noise drawn from t distributions with different
-values of $\nu$. Coloured dots in panels c and d represent jumps with less than
-a 1 in 1000 chance of occurring in a normal distribution. (f--h) Estimates of
-$\nu$ from models fit to the times series in panels c--e. Shown are the
-posterior samples (histograms), median and interquartile range of the posterior
-(IQR) (dots and line segments), and the exponential prior on $\nu$ (dashed
-lines). Colour shading behind panels f--h illustrates the region of heavy
-tails.}
+standard deviations below the mean is 1.8\%, whereas the probability of drawing
+such a value from a normal distribution is nearly zero ($2.9\cdot10^{-5}$\%).
+(c--e) Simulated population dynamics from a Gompertz model with process noise
+drawn from t distributions with three different values of $\nu$. Coloured dots
+in panels c and d represent jumps with less than a 1 in 1000 chance of
+occurring in a normal distribution. (f--h) Estimates of $\nu$ from models fit
+to the times series in panels c--e. Shown are the posterior samples
+(histograms), median and interquartile range of the posterior (IQR) (dots and
+line segments), and the exponential prior on $\nu$ (dashed lines). Colour
+shading behind panels f--h illustrates the region of heavy tails.}
 \label{fig:didactic}
@@
 populations of birds, mammals, insects, and fishes. Small values of $\nu$
-($\lesssim 10$) suggest heavy-tailed dynamics; larger values of $\nu$ suggest
-approximately normal-tailed dynamics. Vertical points and line segments
-represent posterior medians and 50\% / 90\% credible intervals for individual
-populations. Inset plots show probability that $\nu < 10$ (probability of
-heavy tails) for populations arranged by taxonomic order and sorted by
-decreasing mean Pr($\nu < 10$). Taxonomic orders with three or fewer
+($\lesssim 10$) suggest heavy-tailed black swan dynamics; larger values of
+$\nu$ suggest approximately normal-tailed dynamics. Vertical points and line
+segments represent posterior medians and 50\% / 90\% credible intervals for
+individual populations. Inset plots show probability that $\nu < 10$
+(probability of heavy tails) for populations arranged by taxonomic order and
+sorted by decreasing mean Pr($\nu < 10$). Taxonomic orders with three or fewer
 populations in panel a are omitted for space. Red to yellow points highlight
-populations with a high to moderately high probability of heavy-tailed dynamics.
-} \label{fig:nu-coefs} \end{center} \end{figure}
-
+populations with a high to moderately high probability of heavy-tailed black
+swan dynamics.
+}
+\label{fig:nu-coefs}
+\end{center}
+\end{figure}
 \clearpage
@@
 probability that $\nu < 10$). Shown are (a--c) parameters from the Gompertz
-heavy-tailed population model, (b) number of time steps, (c) body length, and
-(d) lifespan. For the Gompertz parameters, $\sigma$ refers to the scale
-parameter of the Student-t process-noise distribution, $\lambda$ refers to the
-expected log abundance at the next time step at an abundance of one, $b$
-refers to the density dependence parameter ($1$ is maximally density
-independent, $0$ is maximally density dependent, and $<0$ is inversely density
-dependent).
+heavy-tailed population model ($\sigma$, $\lambda$, $b$), (b) number of time
+steps, (c) body length, and (d) lifespan. For the Gompertz parameters, $\sigma$
+refers to the scale parameter of the Student-t process-noise distribution,
+$\lambda$ refers to the expected log abundance at the next time step at an
+abundance of one, $b$ refers to the density dependence parameter ($1$ is
+maximally density independent, $0$ is maximally density dependent, and $<0$ is
+inversely density dependent).
 %For parameter values, points and lines represent medians and 50\% credible
 %intervals of the posterior.
-Circles representing a limited number of sharks, crustaceans, and gastropods
-are filled in white. Median and 90\% credible interval posterior predictions of
-a beta regression multilevel model are shown in panels a and d where there was
-a high probability the slope coefficient was different from zero (Supporting
+Circles representing a few sharks, crustaceans, and gastropods are filled in
+white. Median and 90\% credible interval posterior predictions of a beta
+regression multilevel model are shown in panels a and d where there was a high
+probability the slope coefficient was different from zero (Supporting
 Material).

### 2014-11-01 text/supp.tex
Major edits to ms
After meeting with Nick, also reorder and incorporate
the supplementary figs

@@
 
-\item We removed all populations with uneven sampling intervals. I.e.~we removed
+\item We removed all populations with uneven sampling intervals, i.e.\ we removed
   populations that didn't have a constant difference between the ``decimal year
@@
 
-\item Populations with negative abundance values were assumed to be log values
-  and transformed by taking $10$ to the power of the recorded abundance. In many
-  cases this was noted for the population, but not in all cases. We inspected
-  each of these \totalAssumedLog~time series to make sure our assumption made
-  sense.
+\item Populations with negative abundance values were assumed to be log10
+  values and transformed by taking $10$ to the power of the recorded abundance.
+  In many cases this was noted for the population, but not in all cases. We
+  inspected each of these \totalAssumedLog~time series to make sure our
+  assumption made sense.
 
@@
 
-\item We filled in single recorded values of zero with the lowest non-zero value
-  in the time series \citep[following][]{brook2006a}. This assumes that single
-  values of zero result from abundance being low enough that censusing missed
-  present individuals. We turned multiple zero values in a row into \texttt{NA}
-  values. This implies that multiple zero values were either censusing errors or
-  caused by emigration. Regardless, our population models were fitted on
-  a multiplicative (log) scale and so could not account for zero abundance. To
-  avoid distorting the original data too strongly, we removed populations in
-  which we filled in more than four zeros.
+\item We filled in single recorded values of zero with the lowest non-zero
+  value in the time series \citep[following][]{brook2006a}. This assumes that
+  single values of zero result from abundance being low enough that censusing
+  overlooked individuals that were actually present. We turned multiple zero
+  values in a row into \texttt{NA} values. This implies that multiple zero
+  values were either censusing errors or caused by emigration. Regardless, our
+  population models were fitted on a multiplicative (log) scale and so could
+  not account for zero abundance. To avoid distorting the original data too
+  strongly, we removed populations in which we filled in more than four zeros.
 
@@
 Our prior on $\nu$ was based on \citet{fernandez1998}. They chose an
-exponential rate parameter of $0.1$. We chose a less informative rate
-parameter of $0.01$ and truncated the distribution at $2$, since at $\nu < 2$
-the variance of the t distribution is undefined. This prior gives only
-a $7.7$\% probability that $\nu < 10$ but constrains the sampling sufficiently
-to avoid wandering off towards infinity --- above approximately $\nu = 20$ the
+exponential rate parameter of $0.1$. We chose a less informative rate parameter
+of $0.01$ and truncated the distribution at $2$, since at $\nu < 2$ the
+variance of the t distribution is undefined. This prior gives only a $7.7$\%
+probability that $\nu < 10$ but constrains the sampling sufficiently to avoid
+wandering off towards infinity --- above approximately $\nu = 20$ the
 t distribution is so similar to the normal distribution
@@
 unlikely to be informative about the precise value of $\nu$. In the scenario
-where the data are informative about heavy tails
+where the data are uninformative about heavy tails
 (e.g.~Fig.~\ref{fig:didactic}e,~h), the posterior will approximately match the
@@
 We did not include these models in Fig.~\ref{fig:alt}.
+\textit{TODO perhaps I should go up to 16,000 so that more converge.}
 
@@
 provides a check that our more complicated full Gompertz model was obtaining
-ballpark reasonable estimates of $\nu$, (2) it provides a test of whether
-density dependence was systematically affecting our perception of heavy tails,
-(3) it matches how some previous authors have modelled heavy tails without
-accounting for density dependence \citep{segura2013}.
+reasonable estimates of $\nu$, (2) it provides a test of whether density
+dependence was systematically affecting our perception of heavy tails, (3) it
+matches how some previous authors have modelled heavy tails without accounting
+for density dependence \citep{segura2013}.
 
@@
 Observation error can bias parameter estimates \citep[e.g.][]{knape2012} and is
-known to affect the ability to detect extreme events \citep{ward2007}. However,
-simultaneously estimating observation and process noise is a challenging
-problem and is known to result in identifiability issues with the Gompertz
-population model \citep{knape2008}. Furthermore, our model estimates an
-additional parameter --- the shape of the process deviation tails ---
+known to affect the ability to detect extreme events \citep{ward2007}. In our
+main analysis, we fit a model that ignored observation error. One way to
+account for observation error would be to fit a full state-space model that
+simultaneously estimates the magnitude of process noise and observation error.
+However, simultaneously estimating observation and process noise is
+a challenging problem (e.g.\ because the observation and process noise
+parameters tend to negatively covary in model fitting) and is known to
+sometimes result in identifiability issues with the Gompertz population model
+\citep{knape2008}. Furthermore, our model was applied to hundreds of time
+series, often of short length (as few as 20 time steps) and our model estimates
+an additional parameter --- the shape of the process deviation tails ---
 potentially making identifiability and computational issues even greater.
@@
 In the second part of our simulation testing, we tested the ability of the
-heavy-tailed Gompertz model to obtain unbiased parameter estimates given that
-process deviations were provided in which our simpler model (previous section)
-could accurately estimate the value of $\nu$. To generate these process
-deviations, we repeatedly drew proposed candidate process deviations and
-estimated the central tendency, scale, and $\nu$ values each time. We recorded
-when $\hat{\nu}$ (median of the posterior) was within $0.2$ CVs (coefficient of
-variations) of the true $\nu$ value and used this set of random seed values in
-our Gompertz simulation. We then fitted our Gompertz models to the simulated
-datasets with all parameters (except $\nu$) set near the median values
-estimated in the GPDD. We repeated this with $50$ and $100$ samples without
-observation error, $50$ samples with observation error ($\sigma_\mathrm{obs}
-= 0.2$), and $50$ samples with the same observation error and a Gompertz model
-that allowed for correctly specified observation error magnitude.
+heavy-tailed Gompertz model to obtain unbiased parameter estimates 
+when the process noise was chosen so that appropriate tail events were present.
+To generate these process deviations for the $\nu = 3$ and $\nu = 5$ scenarios, we
+repeatedly drew proposed candidate process deviations and estimated the central
+tendency, scale, and $\nu$ values each time. We recorded when $\hat{\nu}$
+(median of the posterior) was within $0.2$ CVs (coefficient of variations) of
+the true $\nu$ value and used this set of random seed values in our Gompertz
+simulation. The following simplified R code illustrates this procedure 
+(the actual code is available at \url{https://github.com/seananderson/heavy-tails}):
+
+\begin{footnotesize}
+\begin{verbatim}
+get_effective_nu_seeds <- function(nu_true = 5, cv = 0.2, N = 50, seed_N = 20) {
+  # nu_true: The true nu value
+  # cv:      The permitted effective nu coefficient of variation
+  # N:       The length of time series
+  # seed_N:  The number of seed values to generate
+  seeds <- numeric(length = seed_N)
+  seed_value <- 0
+  for (i in seq_len(seed_N)) {
+    nu_close <- FALSE
+    while (!nu_close) {
+      seed_value <- seed_value + 1
+      set.seed(i)
+      y <- rt(N, df = nu_true)
+      sm <- rstan::stan(... # fit the Stan model here
+      med_nu_hat <- median(rstan::extract(sm, pars = "nu")[[1]])
+      if (med_nu_hat > (nu_true - cv) & med_nu_hat < (nu_true + cv)) {
+        nu_close <- TRUE
+        seeds[i] <- seed_value
+      }
+    }
+  }
+  seeds
+}
+nu_3_seeds_N50 <- get_effective_nu_seeds(nu_true = 3)
+nu_5_seeds_N50 <- get_effective_nu_seeds(nu_true = 5)
+\end{verbatim}
+\end{footnotesize}
+
+We then fitted our Gompertz models to the simulated datasets with all
+parameters (except $\nu$) set near the median values estimated in the GPDD. We
+repeated this with $50$ and $100$ samples without observation error, $50$
+samples with observation error ($\sigma_\mathrm{obs} = 0.2$), and $50$ samples
+with the same observation error and a Gompertz model that allowed for correctly
+specified observation error magnitude.
 
@@
   \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{species}}),
-  \: \text{for } l = 1, \dots, 301,\\
+  \: \text{for } l = 1, \dots, 301,
 \end{align*}
@@
 \begin{center}
-\includegraphics[width=0.8\textwidth]{../analysis/t-dist-sampling-sim-prior-exp0point01.pdf}
-\includegraphics[width=0.8\textwidth]{../analysis/t-dist-sampling-sim-sigma-prior-exp0point01.pdf}
-\caption{
-  Testing the ability to estimate $\nu$ (top panels) and the scale parameter of
-  the process deviations (bottom panels) for a given number of samples (columns)
-  drawn from a distribution with a given true $\nu$ value (rows). The red lines
-  indicate the true population value. When a small number of samples are drawn
-  there may not be samples sufficiently far into the tails to recapture the
-  true $\nu$ value; however, heavy tails are still distinguished from normal
-  tails in most cases, even with only 25 or 50 samples.
-}
-\label{fig:sim-nu}
-\end{center}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=\textwidth]{../analysis/sim-gompertz.pdf}
-\caption{
-  Simulation testing the Gompertz estimation model when the process deviation
-  draws were chosen so that $\nu$ could be estimated close to the true value
-  outside the full population model (``effective $\nu$'' within a CV of 0.2 of
-  specified $\nu$). The simulation was run across population $\nu$ values
-  (columns) and different scenarios (colours): (1) 100 time steps and no
-  observation error, (2) 50 time steps and no observation error, (3) 50 time
-  steps and observation error drawn from $\mathrm{Normal} (0, 0.2^2)$ but
-  ignored, and (4) 50 time steps with observation error in which the quantity
-  of observation error was assumed known. Dashed horizontal lines show the true
-  population values; these true values were chosen to represent approximately
-  the median values as estimated from the Global Population Dynamics Database.
-  Individual dots and lines
-  represent a stochastic draws from the true population distribution and
-  a model fitting (median and 80\% credible intervals). The panels from top to
-  bottom show $1/\nu$, process noise scale parameter $\sigma$, growth rate
-  parameter $\lambda$, and the density dependence parameter $b$.
-}
-\label{fig:sim-gompertz}
-\end{center}
-\end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=\textwidth]{../analysis/sim-gompertz-boxplots.pdf}
-\caption{
-  The distribution of median estimates from the same simulation as described in
-  Fig.~\ref{fig:sim-gompertz}. The boxes show the interquartile range. The
-  whiskers extend to $1.5$ times the interquartile range and outliers are shown
-  as dots. Note that these distributions represent the output from only 20
-  simulation iterations, so there is some level of remaining sampling noise.
-}
-\label{fig:sim-gompertz-boxplots}
-\end{center}
-\end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=\textwidth]{../analysis/check-sim-box.pdf}
-\caption{
-  The probability that $\nu < 10$ (i.e.\ the approximate probability of heavy
-  tails) for the same simulation scenarios as shown in
-  Fig.~\ref{fig:sim-gompertz} and Fig.~\ref{fig:sim-gompertz-boxplots}. Within
-  each scenario the dots represent stochastic draws from the true population
-  distributions combined with model fits.
-}
-\label{fig:sim-prob}
-\end{center}
-\end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
 \includegraphics[width=0.9\textwidth]{../analysis/ts-gpdd-heavy-eg-log10-base.pdf}
@@
   Time series for populations with Pr($\nu < 10$) $>$ 0.5 using the base
-  Gompertz population model. Panels are ordered by increasing p($\nu < 10$).
+  Gompertz population model. Panels are ordered by increasing Pr($\nu < 10$).
   Vertical axes are on a log10 scale. Colours indicate taxonomic classes. The
-  labels on each panel indicate p($\nu < 10$) $>$ 0.5, the common name for the
+  labels on each panel indicate Pr($\nu < 10$) $>$ 0.5, the common name for the
   species, and the Global Population Dynamics Database ID number.
@@
 \begin{center}
+\includegraphics[width=0.8\textwidth]{../analysis/t-dist-sampling-sim-prior-exp0point01.pdf}
+\includegraphics[width=0.8\textwidth]{../analysis/t-dist-sampling-sim-sigma-prior-exp0point01.pdf}
+\caption{
+  Testing the ability to estimate $\nu$ (top panels) and the scale parameter of
+  the process deviations (bottom panels) for a given number of samples (columns)
+  drawn from a distribution with a given true $\nu$ value (rows). The red lines
+  indicate the true population value. When a small number of samples are drawn
+  there may not be samples sufficiently far into the tails to recapture the
+  true $\nu$ value; however, heavy tails are still distinguished from normal
+  tails in most cases, even with only 25 or 50 samples.
+}
+\label{fig:sim-nu}
+\end{center}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=\textwidth]{../analysis/sim-gompertz.pdf}
+\caption{
+  Simulation testing the Gompertz estimation model when the process deviation
+  draws were chosen so that $\nu$ could be estimated close to the true value
+  outside the full population model (``effective $\nu$'' within a CV of 0.2 of
+  specified $\nu$). The simulation was run across population $\nu$ values
+  (columns) and different scenarios (colours): (1) 100 time steps and no
+  observation error, (2) 50 time steps and no observation error, (3) 50 time
+  steps and observation error drawn from $\mathrm{Normal} (0, 0.2^2)$ but
+  ignored, and (4) 50 time steps with observation error in which the quantity
+  of observation error was assumed known. Dashed horizontal lines show the true
+  population values; these true values were chosen to represent approximately
+  the median values as estimated from the Global Population Dynamics Database.
+  Individual dots and lines
+  represent a stochastic draws from the true population distribution and
+  a model fitting (median and 80\% credible intervals). The panels from top to
+  bottom show $1/\nu$, process noise scale parameter $\sigma$, growth rate
+  parameter $\lambda$, and the density dependence parameter $b$.
+}
+\label{fig:sim-gompertz}
+\end{center}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=\textwidth]{../analysis/sim-gompertz-boxplots.pdf}
+\caption{
+  The distribution of median estimates from the same simulation as described in
+  Fig.~\ref{fig:sim-gompertz}. The boxes show the interquartile range. The
+  whiskers extend to $1.5$ times the interquartile range and outliers are shown
+  as dots. Note that these distributions represent the output from only 20
+  simulation iterations, so there is some level of remaining sampling noise.
+  \textit{TODO wondering now if I should run these for more iterations to make
+    these less noisy.}
+}
+\label{fig:sim-gompertz-boxplots}
+\end{center}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=\textwidth]{../analysis/check-sim-box.pdf}
+\caption{
+  The probability that $\nu < 10$ (i.e.\ the approximate probability of heavy
+  tails) for the same simulation scenarios as shown in
+  Fig.~\ref{fig:sim-gompertz} and Fig.~\ref{fig:sim-gompertz-boxplots}. Within
+  each scenario the dots represent stochastic draws from the true population
+  distributions combined with model fits.
+}
+\label{fig:sim-prob}
+\end{center}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[htbp]
+\begin{center}
 \includegraphics[width=0.5\textwidth]{../analysis/stan-beta-correlates}
@@
 \end{figure}
+
+\clearpage
 

### 2014-11-01 text/ms.tex
Minor edits

@@
 consequences \citep{taleb2007,sornette2009}. One of the most striking black
-swans in ecology is the asteroid marking the mass extinctions of the K-T
-boundary TODO-REF \citep{harnik2012}. Today, climate extremes, in concert with
+swans in ecology is the asteroid marking the end-Cretaceous mass extinction 65 million
+years ago \citep{alvarez1980,harnik2012}. Today, climate extremes, in concert with
 shifts in mean temperature, are expected to cause the greatest ecological and
@@
 a high-quality subset of populations that were suitable for our analyses, we
-filtered the data (Supporting Material). Our filtering removed populations from
+filtered the data (Supporting Information). Our filtering removed populations from
 less reliable data sources, removed those without sufficient data for our
@@
 sampler had obtained sufficient independent samples from the posterior
-($\widehat{R} < 1.05$, $n_\mathrm{eff} > 200$; Supporting Material). We used
+($\widehat{R} < 1.05$, $n_\mathrm{eff} > 200$; Supporting Information). We used
 simulated data to test how easily we could detect $\nu$ given different sample
 sizes and to ensure we could recover unbiased parameter estimates from the
-Gompertz model (Supporting Material).
+Gompertz model (Supporting Information).
 
@@
 plausible population dynamics 
-(\citeauthor{gelman2014} \citeyear{gelman2014}; Supporting Material). 
+(\citeauthor{gelman2014} \citeyear{gelman2014}; Supporting Information). 
 For $\nu$, we chose an exponential prior with rate parameter of $0.01$ truncated at
@@
 which assumes that population growth rate declines linearly with abundance
-itself (Supporting Material).
+itself (Supporting Information).
 
@@
 for varying intercepts at the taxonomic class, order, and species level
-(Supporting Material).
+(Supporting Information).
 
@@
 probabilities to focus early detection efforts and develop disaster response
-plans (e.g.~REF-TODO). A similar focus might benefit resource management once
+plans (e.g.~REFERENCE?). A similar focus might benefit resource management once
 our ability to predict the spatial probability and covariates of ecological
@@
 
-\section{Supplementary Material}
+\section{Supporting Information}
 
-The following supplementary material is available online for this article:\\
+The following supporting information is available online for this article:\\
 Figures S1--S11\\
@@
 probability the slope coefficient was different from zero (Supporting
-Material).
+Information).
 }

### 2014-11-01 text/ms.tex
Minor edits

@@
 experimental waterflea populations \citet{drake2014}. As far as we can tell
-there have been two systematic surveys for ecological black swans, one in North
-American breeding birds and the Global Population Dynamics Database, that found
-little clear evidence for them \citep{keitt1998,allen2001,halley2002}.
+there have been two systematic surveys for ecological black swans: one with North
+American breeding birds and the other with the Global Population Dynamics Database, which 
+ found little clear evidence for them \citep{keitt1998,allen2001,halley2002}.
 Nevertheless there are methodological challenges to their detection

### 2014-11-01 text/supp.tex
Minor edits

@@
 \LARGE
-Supporting Material\\[1.0em]
+Supporting Information\\[1.0em]
 \end{centering}
@@
 where $r_\mathrm{max}$ represents the theoretical maximum population growth
-rate that is obtained when $N$ (abundance) $= 0$. The parameter $K$ represents
+rate that is obtained when $N_t$ (abundance at time $t$) $= 0$. The parameter $K$ represents
 the carrying capacity and, as before, $x_t$ represents the $\log$ transformed
 abundance at time $t$. The Ricker-logistic model assumes a linear decrease in
-population growth rate ($x_t - x_{t-1}$) with increases in abundance ($N_t$).
+population growth rate with increases in abundance.
 In contrast, the Gompertz model assumes a linear decrease in population growth
-rate with increases in \textit{log} abundance ($x_t$) (REF).
+rate with increases in \textit{log} abundance ($x_t$) \citep[e.g.][]{thibaut2012}.
 
@@
 \noindent
-The GPDD IDs used in our analysis. See \url{https://github.com/seananderson/heavy-tails}.
+The GPDD IDs used in our analysis.
 

### 2014-11-04 text/ms.tex
Integrate Trevor's comments and unwrap text

@@
 
+%TODO look up doi or volume for Drake 2014 Ecology
+
 %\noindent
-%\emph{Main message, to be deleted (25 words)}: Black swans are present \&
-%taxonomically widespread but rare in population dynamics. Extreme climate,
-%predation, parasites, and their interactions are common causes; intrinsic
-%drivers are unclear.
+%\emph{Main message, to be deleted (25 words)}: Black swans are present \& taxonomically widespread but rare in population dynamics.
+Extreme climate, predation, parasites, and their interactions are common causes; intrinsic drivers are unclear.
 
@@
 
-Black swans are statistically improbable events that nonetheless occur ---
-often with profound consequences. While extremes in the physical environment,
-such as monsoons and heat waves, are widely studied and increasing in magnitude
-and frequency, it remains unclear the extent to which ecological populations
-buffer or suffer from such extremes. Here, we develop a probability model to
-estimate the degree of heavy-tailedness (presence of black swans) in ecological
-process noise. We apply our model to \NPops~time series from around the world
-across \NOrders~taxonomic orders and seven classes. We find strong evidence of
-black swans, but they are rare occurring in
-\overallMinPerc--\overallMaxPerc\% of populations: most frequently for birds
-(\AvesRangePerc\%) followed by mammals (\MammaliaRangePerc\%), insects
-(\InsectaRangePerc\%), and fishes (\OsteichthyesRangePerc\%). We find little
-evidence of intrinsic life-history covariates, but when they occur, they tend
-to be extrinsically driven by climate and severe winters, cycles of parasites
-and predators, and interactions between these elements. Black swans were more
-frequently detected for populations with longer time series and lower levels of
-process noise. The mere presence of black swans in population dynamics
-highlights the importance of developing robust conservation and management
-strategies --- particularly as the frequency and magnitude of climatic extremes
-increase over the next century.
+Black swans are statistically improbable events that nonetheless occur --- often with profound consequences.
+While extremes in the physical environment, such as monsoons and heat waves, are widely studied and increasing in magnitude and frequency, it remains unclear the extent to which ecological populations buffer or suffer from such extremes.
+Here, we develop a probability model to estimate the degree of heavy-tailedness (presence of black swans) in ecological process noise.
+We apply our model to \NPops~time series from around the world across \NOrders~taxonomic orders and seven classes.
+We find strong evidence of black swans, but they are rare occurring in \overallMinPerc--\overallMaxPerc\% of populations: most frequently for birds (\AvesRangePerc\%) followed by mammals (\MammaliaRangePerc\%), insects (\InsectaRangePerc\%), and fishes (\OsteichthyesRangePerc\%).
+Black swans are not explained by any life-history covariates, but tend to be driven by external perturbations such as climate, severe winters, predator or parasite cycles, and the combined effects of multiple factors .
+Black swans were more frequently detected for populations with longer time series and lower levels of process noise; for shorter and noisier time series, our simulations suggested black-swan dynamics are often not identifiable as such.
+The presence of black swans in population dynamics highlights the importance of developing robust conservation and management strategies --- particularly as the frequency and magnitude of climatic extremes increase over the next century.
 
@@
 
-Black swans are unexpected extreme events with potentially dramatic
-consequences \citep{taleb2007,sornette2009}. One of the most striking black
-swans in ecology is the asteroid marking the end-Cretaceous mass extinction 65 million
-years ago \citep{alvarez1980,harnik2012}. Today, climate extremes, in concert with
-shifts in mean temperature, are expected to cause the greatest ecological and
-societal damage \citep{ipcc2012}. But, while extremes in the physical
-environment such as wave height, storm severity, and temperature are are
-frequent experiences \citep{gaines1993,katz2005}, it remains unclear the extent
-to which ecological systems buffer or suffer from black swans
-\citep{nunez2012}.
+Black swans are unexpected extreme events with potentially dramatic consequences \citep{taleb2007,sornette2009}.
+One of the most striking black swans in ecology is the asteroid collision marking the end-Cretaceous mass extinction 65 million years ago \citep{alvarez1980,harnik2012}.
+Today, climate extremes, in concert with shifts in mean temperature, are expected to cause the greatest ecological and societal damage \citep{ipcc2012}.
+But, while extremes in the physical environment such as wave height, storm severity, and temperature are frequent experiences \citep{gaines1993,katz2005}, it remains unclear the extent to which ecological systems buffer or suffer from black swans \citep{nunez2012}.
 
-There is compelling anecdotal evidence for ecological black swans, but
-systematic evidence has been elusive. A survey of ecologists indicated that
-surprising outcomes of field experiments are far more common than we assume
-\citep{doak2008}, and a number of major environmental events on geologic and
-contemporary time scales could be considered to have caused ecological black
-swans \citep{nunez2012}.
-% TODO: pull anecdote from that study
-Black swans, or heavy tailed distributions, have been discovered in time-series
-of marine microbe populations \citet{segura2013} and time to extinction for
-experimental waterflea populations \citet{drake2014}. As far as we can tell
-there have been two systematic surveys for ecological black swans: one with North
-American breeding birds and the other with the Global Population Dynamics Database, which 
- found little clear evidence for them \citep{keitt1998,allen2001,halley2002}.
-Nevertheless there are methodological challenges to their detection
-\citep{allen2001,ward2007}.
+There is compelling anecdotal evidence for ecological black swans, but systematic evidence has been elusive.
+A survey of ecologists indicated that surprising outcomes of field experiments are far more common than we assume \citep{doak2008}, and a number of major environmental events on geologic and contemporary time scales could be considered to have caused ecological black swans \citep{nunez2012}.
+Black swans, or heavy tailed distributions, have been discovered in time-series of marine microbe populations \citet{segura2013} and time to extinction for experimental waterflea populations \citet{drake2014}.
+As far as we can tell there have been two systematic surveys for ecological black swans: one with North American breeding birds and the other with the Global Population Dynamics Database, which found little clear evidence for them \citep{keitt1998,allen2001,halley2002}.
+Nevertheless there are methodological challenges to their detection \citep{allen2001,ward2007}.
+% TODO: pull anecdote from that Nunez study
 
-There are two key reasons why we may find little evidence of ecological black
-swans. First, they might not exist in higher taxa. Indeed, the majority of
-model fitting and risk forecasting assumes that population dynamics are not
-heavy tailed \citep[e.g.][]{brook2006a,dennis2006,knape2012}. Alternatively,
-heavy-tailed dynamics might exist, but our ability to detect them requires
-further development of statistical tools. For example: (1) The generalized
-extreme value distribution has been applied to environmental data
-\citep[e.g.][]{katz2005}. This distribution describes the most extreme event
-per time interval (e.g.~heaviest rainfall per year), but requires time series
-that are sufficiently long to be condensed into time intervals. (2) Fitting
-a catastrophic mixture distribution in a state space model allows one to
-quantify the probability that population events are extreme but is also data
-intensive \citep{ward2007} (3) Thin- and heavy-tailed distributions have been
-competed with each other in an information theoretic framework
-\citep{halley2002}, but this analysis did not quantify the probability of black
-swans on a continuous scale or allow for standard population dynamics.
+There are two key reasons why we may find little evidence of ecological black swans.
+First, they might not exist in higher taxa.
+Indeed, the majority of model fitting and risk forecasting assumes that population dynamics are not heavy tailed \citep[e.g.][]{brook2006a,dennis2006,knape2012}.
+Alternatively, heavy-tailed dynamics might exist, but our ability to detect them requires further development of statistical tools.
+One such tool is the generalized extreme value distribution, which has been applied to environmental data \citep[e.g.][]{katz2005}.
+This distribution describes the most extreme event per time interval (e.g.~heaviest rainfall per year), but requires time series that are sufficiently long to be condensed into time intervals.
+Another statistical tool involves fitting a catastrophic mixture distribution in a state space model to quantify the probability that population events are extreme, although this is also data intensive \citep{ward2007} A third tool is to compare the fits of thin- and heavy-tailed distributions with the Akaike Information Critera (AIC) TODO CHECK \citep{halley2002}, but this analysis did not quantify the probability of black swans or allow for population dynamics.
 
-Here we develop a framework for identifying ecological black swans in
-population dynamic process noise --- the stochastic jumps from time step to
-time step. Our framework quantifies both the magnitude and probability of
-heavy-tailed dynamics, allows for a range of population dynamic models, can
-incorporate observation uncertainty, and is applicable to commonly available
-lengths of abundance time series. We apply our model to hundreds of populations
-from across four taxonomic classes to address three questions: (1) how frequent
-and strong are black swans in population dynamics across taxonomy, (2) what are
-some verified causes of black swans in ecological time series, and (3) are
-there characteristics of time series or intrinsic life-history characteristics
-associated with heavy-tailed dynamics. We find a high probability of ecological
-black swans across a wide range of taxa but in a limited number of populations.
-When they occur, extreme climate, predation, parasites, and their interactions
-are common documented causes. We find little evidence that intrinsic
-life-history characteristics are related to the probability of black swans.
+Here we develop a framework for identifying ecological black swans in the process noise of population dynamics, i.e.\ the stochastic jumps in abundance from one time step to the next.
+Our framework quantifies both the magnitude and probability of heavy-tailed dynamics, allows for a range of population dynamic models, can incorporate observation uncertainty, and can be easily applied to abundance time series.
+We apply our model to \NPops\ of populations from four taxonomic classes.
+We estimate the frequency and strength of black swans in population dynamics across taxonomy; verify some causes of black swan events; and attempt to identify characteristics of time series or intrinsic life-history characteristics that are associated with heavy-tailed dynamics.
+%We find a high probability of ecological
+%black swans across a wide range of taxa but in a limited number of populations.
+%When they occur, extreme climate, predation, parasites, and their interactions
+%are common documented causes. We find little evidence that intrinsic
+%life-history characteristics are related to the probability of black swans.
 
@@
 
-To obtain estimates of the probability and magnitude of black swans, we fitted
-population dynamic models to abundance time series from around the world. The
-key novelty of our approach, compared to typical population models, was
-estimating the shape of the process noise tails by measuring the degrees of
-freedom ($\nu$) of a Student t-distribution.
+To obtain estimates of the probability and magnitude of black swans, we fitted population dynamic models to abundance time series from around the world.
+For each population, we estimated the shape of the process noise tails by measuring the degrees of freedom ($\nu$) of a Student t-distribution.
 
@@
 
-We obtained abundance time series from the Global Population Dynamics Database
-(GPDD) \citep{gpdd2010}. The GPDD contains nearly 5000 time series of abundance
-across $\sim$1000 species and $\sim$100 taxonomic orders. To derive
-a high-quality subset of populations that were suitable for our analyses, we
-filtered the data (Supporting Information). Our filtering removed populations from
-less reliable data sources, removed those without sufficient data for our
-models, and interpolated some missing values \citep[\textit{sensu}][]{brook2006a}.
-Our final dataset contained \NPops~populations across \NOrders~taxonomic orders
-and seven taxonomic classes with a median of \medianTimeSteps~time steps (range
-of \minTimeSteps--\maxTimeSteps) (Table S1, Fig.~\ref{fig:all-ts}).
+We selected abundance time series from the Global Population Dynamics Database (GPDD) \citep{gpdd2010}, which contains nearly 5000 time series of abundance from $\sim$1000 species and $\sim$100 taxonomic orders.
+We filtered the data (Supporting Information) to remove populations from less reliable data sources, and those without sufficient data for our models, and then interpolated some missing values \citep[\textit{sensu}][]{brook2006a}.
+TODO CONTEXTUALIZE THE QUANTITY OF INTERPOLATION Our final dataset contained \NPops~populations across \NOrders~taxonomic orders and seven taxonomic classes, with a median of \medianTimeSteps~time steps (range of \minTimeSteps--\maxTimeSteps) (Table S1, Fig.~\ref{fig:all-ts}).
 
@@
 
-Our main analysis focuses on the commonly applied Gompertz population dynamics
-model \citep[e.g.][]{knape2012,dennis2014,connors2014}. The Gompertz model
-represents population growth as a linear function in $\log$ space. If we let
-$x_t$ represent the $\log$ abundance ($N$) at time $t$, we can represent the
-Gompertz model as:
+Our main analysis focuses on the commonly applied Gompertz population dynamics model \citep[e.g.][]{knape2012,dennis2014,connors2014}.
+The Gompertz model represents population growth as a linear function in $\log$ space.
+If we let $x_t$ represent the $\log$ abundance ($N$) at time $t$, we can represent the Gompertz model as:
 \begin{align*}
 x_t &= \lambda + b x_{t-1} + \epsilon_t\\
-\epsilon_t &\sim \mathrm{Normal}(0, \sigma^2).
-\end{align*}
-The growth parameter $\lambda$ represents the expected growth rate if $N_t
-= 1$. The model is density independent if $b = 1$, maximally density dependent
-if $b = 0$, and inversely density dependent if $b < 0$. The process noise
-$\epsilon_t$ is usually modelled as normally distributed with standard
-deviation $\sigma$. In previous analyses of the GPDD, the Gompertz has proved
-to be the most frequently parsimonious population model \citep{brook2006}.
-
-We extended the Gompertz model by allowing the process noise to be
-drawn from a Student-t distribution with degrees of freedom parameter $\nu$,
-and scale parameter $\sigma$:
-\begin{align*}
 \epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma).
 \end{align*}
-If $\nu$ is small ($\lesssim 10$), the t distribution has much heavier tails
-than a normal distribution (Fig.~\ref{fig:didactic}a, b). For example, at $\nu
-= 2$, the probability of drawing a value more than five standard deviations
-below the mean is $1.8$\%, whereas the probability of drawing such a value from
-a normal distribution is nearly zero. As $\nu$ approaches infinity the
-distribution approaches the normal distribution (Fig.~\ref{fig:didactic}a, b).
-By estimating the value of $\nu$ we can quantity how heavy-tailed the process
-noise deviations are.
+The growth parameter $\lambda$ represents the expected growth rate if $N_t = 1$.
+The model is density independent if $b = 1$, maximally density dependent if $b = 0$, and inversely density dependent if $b < 0$.
+Usually, the process noise $\epsilon_t$ is modelled as normally distributed, but in our paper we assume it is drawn from a t distribution with scale parameter $\sigma$ and degrees of freedom $\nu$.
+In previous analyses of the GPDD, the Gompertz is most often identified as the most parsimonious population model that can be fitted to these data \citep{brook2006}.
+
+By allowing the process noise to be drawn from a Student-t distribution we can estimate the degree to which the process errors have heavy tails and are thereof evidence of black swan events (Fig.~\ref{fig:didactic}a, b).
+For example, at $\nu = 2$, the probability of drawing a value more than five standard deviations below the mean is $0.018$, whereas the probability of drawing such a value from a normal distribution is tiny ($2.9\cdot10^{-7}$) TODO CHECK.
+As the value of $\nu$ approaches infinity, the distribution approaches the normal distribution (Fig.~\ref{fig:didactic}a, b).
 
@@
 
-We estimated all models in a Bayesian framework using Stan \citep[][version
-2.4.0]{stan-manual2014} via the R computing environment \citep{r2014}. Stan
-samples from the posterior distribution with an adaptive version of Hamiltonian
-Markov chain Monte Carlo called the No-U-Turn Sampler and generally obtains
-less correlated samples than algorithms such as the Gibbs sampler
-\citep{hoffman2014}. We assured that chains had sufficiently converged and the
-sampler had obtained sufficient independent samples from the posterior
-($\widehat{R} < 1.05$, $n_\mathrm{eff} > 200$; Supporting Information). We used
-simulated data to test how easily we could detect $\nu$ given different sample
-sizes and to ensure we could recover unbiased parameter estimates from the
-Gompertz model (Supporting Information).
+We fitted all models in a Bayesian framework using Stan \citep{stan-manual2014} via the R computing environment \citep{r2014}.
+Stan samples from the posterior distribution with an adaptive version of Hamiltonian Markov chain Monte Carlo called the No-U-Turn Sampler and generally obtains less correlated samples than algorithms such as the Gibbs sampler \citep{hoffman2014}.
+We tested to ensure that the chains had sufficiently converged and that the sampler had obtained sufficient independent samples from the posterior ($\widehat{R} < 1.05$, $n_\mathrm{eff} > 200$; Supporting Information).
 
-We chose weakly informative priors to incorporate our understanding of
-plausible population dynamics 
-(\citeauthor{gelman2014} \citeyear{gelman2014}; Supporting Information). 
-For $\nu$, we chose an exponential prior with rate parameter of $0.01$ truncated at
-values above two (Fig.~\ref{fig:priors}) --- a slightly less informative prior
-than suggested by \citet{fernandez1998}. This prior gives only a $7.7$\%
-probability that $\nu < 10$ but constrains the sampling sufficiently to avoid
-wandering off towards infinity --- above approximately $\nu = 20$ the
-t distribution is so similar to the normal distribution that time series of the
-length considered here are unlikely to be informative about the precise value
-of $\nu$ (Fig.~\ref{fig:didactic}).
+We chose weakly informative priors to incorporate our understanding of plausible population dynamics (\citeauthor{gelman2014} \citeyear{gelman2014}; Supporting Information).
+For $\nu$, we chose an exponential prior with rate parameter of $0.01$ truncated at values above two (Fig.~\ref{fig:priors}) --- a slightly less informative prior than suggested by \citet{fernandez1998}.
+This prior gives only a $7.7$\% probability that $\nu < 10$ but constrains the sampling sufficiently to avoid wandering off towards infinity.
+In any case,  for $\nu > 20$ the t distribution is almost indistinguishable from the normal distribution (Fig.~\ref{fig:didactic}).
+TODO ADD ALTERNATIVE PRIOR NOTE.
 
-We fit alternative population models to test if four key phenomena
-systematically changed our conclusions. (1) Autocorrelation has been suggested
-as a reason for increased observed variability of abundance time series through
-time, which could create apparent heavy tails \citep{inchausti2002}; therefore,
-we fit a model that modelled the serial correlation in the residuals. (2)
-Previous work has modelled abundance or growth rates without accounting for
-density dependence \citep{halley2002,segura2013}; therefore, we fit a simpler
-model in which we assumed density independence. (3) Observation error could
-bias parameter estimates \citep{knape2012} or mask our ability to detect heavy
-tails \citep{ward2007}; therefore, we fit a model where we allowed for a fixed
-quantity of observation error ($0.2$ standard deviations on a log scale). (4)
-The Gompertz model assumes that population growth rate declines linearly with
-log abundance. We also fit an alternative model, the Ricker-logistic model,
-which assumes that population growth rate declines linearly with abundance
-itself (Supporting Information).
+TODO We used simulated data to test how easily we could detect $\nu$ given different sample sizes and to ensure we could recover unbiased parameter estimates from the Gompertz model (Supporting Information).
+
+% TODO fitted vs. fit
+We fitted alternative population models to test if four key phenomena systematically changed our conclusions.
+Autocorrelation has been suggested as a reason for increased observed variability of abundance time series through time, which could create apparent heavy tails \citep{inchausti2002}; therefore, we fitted a model that included serial correlation in the residuals.
+Additionally, previous work has modelled abundance or growth rates without accounting for density dependence \citep{halley2002,segura2013}; therefore, we fitted a simpler model in which we assumed density independence.
+Third, observation error could bias parameter estimates \citep{knape2012} or mask our ability to detect heavy tails \citep{ward2007}; therefore, we fit a model where we allowed for a fixed quantity of observation error ($0.2$ standard deviations on a log scale).
+Finally, the Gompertz model assumes that population growth rate declines linearly with log abundance.
+We also fit an alternative model, the Ricker-logistic model, which assumes that population growth rate declines linearly with abundance itself (Supporting Information).
 
@@
 
-We investigated possible covariates of heavy-tailed population dynamics
-visually and through multilevel modelling. We plotted characteristics of the
-time series ($\sigma$, $\lambda$, $b$, and time-series length) along with two
-life-history characteristics (body length and maximum lifespan obtained from
-\citet{brook2006a}) against our estimated probability of heavy tails, Pr$(\nu
-< 10) > 0.5$. We formally investigated these relationships by fitting beta
-regression multilevel models. For covariates that were derived from Gompertz
-model parameter estimates, we incorporated standard deviations around the
-means. To account for broad patterns of phylogenetic relatedness, we allowed
-for varying intercepts at the taxonomic class, order, and species level
-(Supporting Information).
+We investigated possible covariates of heavy-tailed population dynamics visually and through multilevel modelling.
+We plotted characteristics of the time series ($\sigma$, $\lambda$, $b$, and time-series length) along with two life-history characteristics (body length and maximum lifespan obtained from \citet{brook2006a}) against our estimated probability of heavy tails, Pr$(\nu < 10) > 0.5$.
+We formally investigated these relationships by fitting beta regression multilevel models TODO REF AND MORE DETAILS?.
+For covariates that were derived from Gompertz model parameter estimates, we incorporated standard deviations around the means.
+To account for broad patterns of phylogenetic relatedness, we allowed for varying intercepts at the taxonomic class, order, and species level (Supporting Information).
 
-Finally, we investigated a sample of populations that our method categorized as
-having a high probability of heavy tails (Pr$(\nu < 10) > 0.5$). Where
-possible, we found the documented causes of ecological black swans in the
-primary data source cited in the GPDD or in other literature describing the
-population. These exemplar populations were chosen haphazardly with the purpose
-of generating hypotheses about the causes of population dynamic black swans.
+Finally, we investigated a sample of populations that our method categorized as having a high probability of heavy tails (Pr$(\nu < 10) > 0.5$).
+Where possible, we found the documented causes of ecological black swans in the primary data source cited in the GPDD or in other literature describing the population.
+These exemplar populations were chosen haphazardly with the purpose of generating hypotheses about the causes of population dynamic black swans.
 
@@
 
-We found strong, but rare, evidence for black-swan population dynamics. By
-defining black-swan dynamics as a greater than $0.5$ probability that $\nu
-< 10$, our main Gompertz model found evidence for heavy tails most frequently
-for birds (\birdPH\%) followed by mammals (\mammalsPH\%), insects
-(\insectsPH\%), and fishes (\fishPH\%) (Figs~\ref{fig:nu-coefs},
-\ref{fig:heavy-ts}). Black swans were taxonomically widespread, occurring in
-\POrdersHeavy\% of taxonomic orders. Accounting for time series length and
-partially pooling inference across taxonomic class and order with a multilevel
-model, there was stronger evidence for black swans in insect populations than
-is visually apparent in Fig.~\ref{fig:nu-coefs} --- four of 10 orders with the
-highest median probability of heavy tails were insect orders --- however, there
-was considerable uncertainty in these estimates
-(Fig.~\ref{fig:order-estimates}).
+We found strong, but rare, evidence for black-swan population dynamics.
+By defining black-swan dynamics as a greater than $0.5$ probability that $\nu < 10$, our main Gompertz model found evidence for heavy tails most frequently for birds (\birdPH\%) followed by mammals (\mammalsPH\%), insects (\insectsPH\%), and fishes (\fishPH\%) (Figs~\ref{fig:nu-coefs}, \ref{fig:heavy-ts}).
+Black swans were taxonomically widespread, occurring in \POrdersHeavy\% of taxonomic orders.
+Accounting for time series length and partially pooling inference across taxonomic class and order with a multilevel model, there was stronger evidence for black swans in insect populations than is visually apparent in Fig.~\ref{fig:nu-coefs} --- four of 10 orders with the highest median probability of heavy tails were insect orders --- however, there was considerable uncertainty in these estimates (Fig.~\ref{fig:order-estimates}).
 
-The majority of our heavy-tailed estimates were robust to alternative
-population models and observation error (Fig.~\ref{fig:alt}). Three of the four
-population dynamic phenomena were unimportant:
-% TODO: I don't like "unimportant"
-including an autocorrelation structure in the residuals, modelling population
-growth rates without density dependence, or modelling the population dynamics
-as Ricker-logistic did not systematically alter our conclusions. However,
-allowing for a fixed quantity of observation error decreased our estimates of
-black-swan dynamics slightly, increasing the median estimate of $\nu$ from
-$<10$ to $\ge 10$ in $\baseNuTenObsTenSwitch$ of $\baseNuTen$ populations,
-although the majority of $\nu$ estimates remained qualitatively similar
-(Fig.~\ref{fig:alt}).
-Our simulation testing show that, if anything, our models underpredict the true
-magnitude and probability of heavy tailed events --- especially given the
-length of the time series in the GPDD (Figs~\ref{fig:sim-nu},
-\ref{fig:sim-gompertz}, \ref{fig:sim-gompertz-boxplots}, \ref{fig:sim-prob})
+The majority of our heavy-tailed estimates were robust to alternative population models and observation error (Fig.~\ref{fig:alt}).
+Our conclusions were not systematically altered when we included an autocorrelation structure in the residuals, modelled population growth rates without density dependence, or modelled the population dynamics as Ricker-logistic.
+However, setting observation error standard deviation to $0.2$ increased the median estimate of $\nu$ from $<10$ to $\ge 10$ in $\baseNuTenObsTenSwitch$ of $\baseNuTen$ populations, although the majority of $\nu$ estimates remained qualitatively similar (Fig.~\ref{fig:alt}).
 
-Across populations, the probability of observing black swan dynamics was
-positively related to time-series length and negatively related to magnitude of
-process noise ($\sigma$) but not clearly related to population growth rate
-($\lambda$), density dependence ($b$), or maximum lifespan
-(Figs~\ref{fig:correlates}, \ref{fig:correlate-coefs}). Longer time-series
-length was the strongest covariate of observing black swan dynamics (as was
-also observed in simulation testing, Fig.~\ref{fig:sim-nu}). For example,
-we were about \pIncHeavyNThirtyNSixty~times more likely to observe heavy tails
-in a population with 60 time steps than in one with 30 time steps
-(Fig.~\ref{fig:perc-inc-p}). However, the absolute change in probability with
-increased time series length was small ($0.\pHeavyNSixty$ vs.\
-$0.\pHeavyNThirty$ in the previous example, Fig.~\ref{fig:correlates}).
+TODO ADD TO THIS Our simulation testing show that, if anything, our models underpredict the true magnitude and probability of heavy tailed events --- especially given the length of the time series in the GPDD (Figs~\ref{fig:sim-nu}, \ref{fig:sim-gompertz}, \ref{fig:sim-gompertz-boxplots}, \ref{fig:sim-prob}) AND ``(as was also observed in simulation testing, Fig.~\ref{fig:sim-nu})''
 
-We assembled the documented causes of black swan population dynamics in seven
-exemplar populations with high probabilities of heavy tails (Table 1). The
-majority of documented events were downward black swans and the majority
-involved a confluence of events. For example, a shortage of nest sites for the
-shag long-necked bird in the United Kingdom was thought to reduce population
-productivity and make the population vulnerable when a red tide hit in 1968
-\citep{potts1980}. The population experienced a black swan downswing, but the
-new availability of nest sites resulted in a rapid population upswing
-\citep{potts1980}. As another example, an interaction between environmental-
-and predation-mediated population cycles are thought to have caused a downward
-black swan for a water vole population \citep{saucy1994}. Other black swans
-were the result of a sequence of extreme climate events on their own. For
-example, severe winters in 1929, 1940--1942, and 1962--1963 were associated
-with black swan downswings in grey heron abundance in the United Kingdom
-\citep{stafford1971}. Our analysis finds that the last event was a combination
-of two black swan events in a row and it took the population three times longer
-to recover than predicted \citep{stafford1971}.
+Across populations, the probability of observing black swan dynamics was positively related to time-series length and negatively related to magnitude of process noise ($\sigma$) but not clearly related to population growth rate ($\lambda$), density dependence ($b$), or maximum lifespan (Figs~\ref{fig:correlates}, \ref{fig:correlate-coefs}).
+Longer time-series length was the strongest covariate of observing black swan dynamics.
+For example, we were about \pIncHeavyNThirtyNSixty~times more likely to observe heavy tails in a population with 60 time steps than in one with 30 time steps (Fig.~\ref{fig:perc-inc-p}).
+However, the absolute change in probability with increased time series length was small ($0.\pHeavyNSixty$ vs.\ $0.\pHeavyNThirty$ in the previous example, Fig.~\ref{fig:correlates}).
+
+% "Perhaps then you could say "we examined all time series with published explanations of why the black swan events occurred"
+TODO STATE HOW THESE 7 WERE CHOSEN
+TODO QUANTIFY UPS AND DOWNS --- SKEW
+We assembled the documented causes of black swan population dynamics in seven exemplar populations with high probabilities of heavy tails (Table 1).
+The majority of documented events were downward black swans and involved a series of unfortunate events.
+For example, a shortage of nest sites for the shag long-necked bird (TODO SCI.\ NAME) in the United Kingdom reduced population productivity and made the population more vulnerable when a red tide hit in 1968 \citep{potts1980}.
+The population experienced a black swan downswing, but the new availability of nest sites resulted in a rapid population upswing \citep{potts1980}.
+As another example, an interaction between environmental- and predation-mediated population cycles are thought to have caused a downward black swan for a water vole population \citep{saucy1994}.
+Other black swans were the result of a sequence of extreme climate events on their own.
+For example, severe winters in 1929, 1940--1942, and 1962--1963 were associated with black swan downswings in grey heron abundance in the United Kingdom \citep{stafford1971}.
+Our analysis finds that the last event was a combination of two black swan events in a row and it took the population three times longer to recover than predicted \citep{stafford1971}.
+
+%TODO IMPORTANT: you need to quantify across all identified black swan events how many were downward and how many upward.
+The assumption of your model is a symmetric distribution, but if in fact downward events are more common, perhaps a skewed distribution would be more realistic.
+Furthermore, if there is skew this would *increase* the probability of higher extinction risk.
+Something to definitely highlight in the results and abstract.
 
@@
 
-We found strong evidence that the process noise in ecological time series was
-heavy-tailed under  Gompertz or Ricker-logistic models in
-\overallMinPerc--\overallMaxPerc\% of populations assessed. We detected black
-swans more frequently for longer time series and for populations with a smaller
-magnitude of process error, as we might expect. However, we failed to find
-strong evidence that black swans were associated with the intrinsic ecological
-phenomena of density dependence, population growth rate, or lifespan. In cases
-that we verified, black swans were often a result of the interaction between
-elements of extreme climate, predators and parasite cycles, and strong changes
-in human pressures. Our empirical results, sensitivity analyses, and simulation
-tests suggest that estimating the tail shape of process noise is a viable
-method of detecting black swan population dynamics and if anything will
-underestimate the probability of black swans. The presence of black swans
-highlights the importance of developing management strategies that detect
-quickly, respond to, and are robust to extremes in population dynamics ---
-particularly as the frequency and magnitude of climatic extremes increase over
-the next century \citep{easterling2000,ipcc2012}.
+We found strong evidence for black swans (heavy-tailed process noise) in 3--5\% of ecological time series.
+Black swans are detected more frequently in longer time series and in populations with a smaller magnitude of process error.
+However, black swans were not associated with density dependence, population growth rate, or lifespan.
+In verified cases, black swans were often a result of the interaction between elements of extreme climate, predators and parasite cycles, and strong changes in human pressures.
+Our empirical results, sensitivity analyses, and simulation tests suggest that estimating the tail shape of process noise is a viable method of detecting black swan population dynamics and if anything will underestimate the probability of black swans.
+The presence of black swans highlights the importance of developing management strategies that detect quickly, respond to, and are robust to extremes in population dynamics --- particularly as the frequency and magnitude of climatic extremes increase over the next century \citep{easterling2000,ipcc2012}.
 
-Our results clarify previous related analyses. Previous work with an older
-version of the GPDD assessed the distribution of abundance time series but
-focused on identifying if the log-normal distribution was the most frequently
-parsimonious model \citep{halley2002}. For heavy-tailed distributions,
-\citet{halley2002} fit the extremely heavy-tailed Cauchy distribution and the
-four-parameter Levy stable distribution and found little information criteria
-support for these distributions in longer time series. However, black swan
-events are by definition rare, and the majority of time series in the GPDD are
-short for these purposes. Therefore, we would not expect to observe black swans
-in a large proportion of populations. Furthermore, by probabilistically
-quantifying process noise tails on a continuous spectrum, and allowing for
-non-stationary time series and density dependence, our analysis allows for
-a more nuanced description of the evidence of ecological black swans. In an
-earlier study, \citet{keitt1998} described heavy (power law) tails in breeding
-bird population abundance. However, this finding was challenged by
-\citet{allen2001}, who showed that mixing data across species could
-falsely generate heavy tails. This rebuttal supports our use of the
-t distribution to represent heavy tails, since a t distribution can be
-represented by a mixture of normal distributions \citep[with the same mean and
-inverse-gamma-distributed variances,][]{gelman2014}.
+Our results clarify previous related analyses.
+Previous work with an older version of the GPDD assessed the distribution of abundance time series but focused on identifying if the log-normal distribution was the most frequently parsimonious model \citep{halley2002}.
+For heavy-tailed distributions, \citet{halley2002} fit the extremely heavy-tailed Cauchy distribution and the four-parameter Levy stable distribution and found little information criteria support for these distributions in longer time series.
+However, black swan events are by definition rare, and the majority of time series in the GPDD are short for these purposes.
+Therefore, we would not expect to observe black swans in a large proportion of populations.
+By probabilistically quantifying process noise tails on a continuous spectrum, and allowing for non-stationary time series and density dependence, our analysis allows for a more nuanced description of the evidence of ecological black swans.
+In an earlier study, \citet{keitt1998} described heavy (power law) tails in breeding bird population abundance.
+However, this finding was challenged by \citet{allen2001}, who showed that mixing data across species could falsely generate heavy tails.
+This rebuttal supports our use of the t distribution to represent heavy tails, since a t distribution can be represented by a mixture of normal distributions \citep[with the same mean and inverse-gamma-distributed variances,][]{gelman2014}.
 
-We might expect to observe black swan dynamics in ecological time series
-because of either unmodelled intrinsic properties of populations or extrinsic
-forces acting on populations. Since a t-distribution can be formed by a mixture
-of normal distributions \citep{gelman2014}, we could observe heavy tails if we
-miss some underlying mixture of intrinsic processes \citep{allen2001}. That
-process might be an aggregation of populations across space, or population
-diversity, or some intrinsic change in population variability through time.
-Extrinsic forces could also cause black swan dynamics
-\citep[e.g.][]{nunez2012}. These forces could be extreme themselves. For
-example, extreme climate, predation from (or competition with) other species
-experiencing black swans, or sharp changes in human pressure such hunting,
-fishing, or habitat destruction might cause black swans. Alternatively, the
-interaction of multiple ``normal'' extrinsic forces could give rise to black
-swan ecological dynamics. This could occur if the interaction is synergistic
-\citep[e.g.][]{kirby2009} or even if non-synergistic interactions experience
-a rare alignment \citep{denny2009}.
+We might expect to observe black swan dynamics in ecological time series because of either unmodelled intrinsic properties of populations or extrinsic forces acting on populations.
+Since a t-distribution can be formed by a mixture of normal distributions \citep{gelman2014}, we could observe heavy tails if we miss some underlying mixture of intrinsic processes \citep{allen2001}.
+That process might be an aggregation of populations across space, or population diversity, or some intrinsic change in population variability through time.
+Extrinsic forces could also cause black swan dynamics \citep[e.g.][]{nunez2012}.
+These forces could be extreme themselves.
+For example, extreme climate, predation from (or competition with) other species experiencing black swans, or sharp changes in human pressure such hunting, fishing, or habitat destruction might cause black swans.
+Alternatively, the interaction of multiple ``normal'' extrinsic forces could give rise to black swan ecological dynamics.
+This could occur if the interaction is synergistic \citep[e.g.][]{kirby2009} or even if non-synergistic interactions experience a rare alignment \citep{denny2009}.
 
 There are a number of caveats when considering the generality of our results.
-First, the GPDD data represent a taxonomically and geographically biased sample
-of populations --- the longer time series we focus on are dominated by
-commercially and recreationally important species and a disproportionate number
-of populations are located in the United Kingdom. Although we expect that we
-would find qualitatively similar evidence for black swans in many other large
-taxonomic or geographic samples of populations, the common forces driving those
-black swans likely differ (e.g.~severe winters in Table~1). Second, in a large
-database with disparate data sources, we cannot ensure the quality of every
-data point; it is likely that some of the black swans in our dataset are
-recording mistakes. Conversely, recording mistakes could mask heavy tails if
-extreme observations are discarded or altered as assumed errors. Third, the
-temporal scales of observation and population dynamics vary considerably across
-populations in the GPDD and this likely influences the detection of heavy
-tails. For example, if we make frequent observations relative to generation
-time (e.g.~for many large-bodied mammals) we will average across generations
-and perhaps miss black swans. Furthermore, if we census populations
-infrequently relative to generation time (e.g.~many insects in the GPDD) the
-recorded data may average across extreme and less-extreme events and dampen
-black swan dynamics.
+The GPDD data represent a taxonomically and geographically biased sample of populations --- the longer time series we focus on are dominated by commercially and recreationally important species and a disproportionate number of populations are located in the United Kingdom.
+Although we expect that we would find qualitatively similar evidence for black swans in many other large taxonomic or geographic samples of populations, the common forces driving those black swans likely differ (e.g.~severe winters in Table~1).
+In addition to a possibly biased sample of populations, some black swan detections could just be recording mistakes, or conversely, some extreme observations may have been discarded or altered if they were erroneously suspected of being recording mistakes.
+A final caveat is that the temporal scales of observation and population dynamics vary considerably across populations in the GPDD and this likely influences the detection of heavy tails.
+For example, if we make frequent observations relative to generation time (e.g.~for many large-bodied mammals) we will average across generations and perhaps miss black swans.
+Conversely, if we census populations infrequently relative to generation time (e.g.~many insects in the GPDD) the recorded data may average across extreme and less-extreme events and also dampen black swan dynamics.
 
-Our results suggest important future issues to address about ecological black
-swans. Given that black swans do occur, can we forecast the probability of
-black swans in space and time? Furthermore, what management policies allow us
-to detect them quickly after they happen? Can we isolate the components of
-ecological dynamics that experience black swans by moving from phenomenological
-models such as the Gompertz to mechanistic models, that take into account, for
-example, recruitment? We expect that greater insight into the mechanisms and
-covariates of ecological black swans may be best obtained through specific
-geographic and taxonomic subsets of data where longer time series with low
-levels of observation error are available \citep[e.g.][]{segura2013}. Finally,
-what is the impact of allowing for black swans in forecasts of ecological risk?
-Financial analysts are concerned with the shape of
-financial-return-distribution tails because these impact estimates of risk ---
-the probability of a specific magnitude of undesired event occurring
-\citep{rachev2008}. It is possible that a comparable focus in ecology would
-alter our perception of ecological forecasts such as extinction risk.
+Our results suggest important future issues to address about ecological black swans.
+Given that black swans do occur, can we forecast the probability of black swans in space and time?
+Furthermore, what management policies allow us to detect them quickly after they happen?
+Can we isolate the components of ecological dynamics that experience black swans by moving from phenomenological models such as the Gompertz to mechanistic models, that take into account, for example, recruitment?
+We expect that greater insight into the mechanisms and covariates of ecological black swans may be best obtained through specific geographic and taxonomic subsets of data where longer time series with low levels of observation error are available \citep[e.g.][]{segura2013}.
 
-Recognizing the prevalence of heavy-tailed dynamics suggests a number of policy
-directions. First, ecological resource management can draw from other
-disciplines that focus on heavy tails. For example, earthquake preparedness and
-response is focussed on black swan events. Similarly to ecological black swans,
-we can rarely predict the specific location and timing of large earthquakes.
-But, earthquake preparedness involves spatial planning based on forecast
-probabilities to focus early detection efforts and develop disaster response
-plans (e.g.~REFERENCE?). A similar focus might benefit resource management once
-our ability to predict the spatial probability and covariates of ecological
-black swans improves. Second, we can develop management policy that is robust
-to heavy tails. For example, setting target population abundances that are
-appropriately set back from critical limits may buffer black swan events
-\citep[e.g.][]{caddy1996}, and maintaining genetic, phenotypic, and behavioural
-diversity may allow some components of populations persist when others are
-affected by disease or extreme environmental forces
-\citep[e.g.][]{schindler2010, anderson2014}. Finally, extreme and unexpected,
-surprising, or counterintuitive ecological dynamics offer a tremendous
-opportunity to learn about ecological systems, evaluate when our models break
-down, and adjust future management policy \citep{doak2008, pine-iii2009,
-  lindenmayer2010}. In recent decades ecology has moved toward focussing on
-aspects of variance in addition to mean responses \citep[e.g.][]{loreau2010a,
-  thompson2013}. Our results suggest that an added focus on ecological extremes
-represents the next frontier, particularly in the face of increased climate
-extremes \citep{meehl2004,ipcc2012}.
+Most importantly, what is the impact of allowing for black swans in forecasts of ecological risk?
+Financial analysts are concerned with the shape of financial-return-distribution tails because these impact estimates of risk --- the probability of a specific magnitude of undesired event occurring \citep{rachev2008}.
+A comparable focus in ecology would increase our estimates of extinction risk, since these would be disproportionately impacted by black swan events.
+TODO END ON THIS
+
+Recognizing the prevalence of heavy-tailed dynamics suggests a number of policy directions.
+Ecological resource management can draw from other disciplines that focus on heavy tails.
+For example, earthquake preparedness and response is focussed on black swan events.
+Similarly to ecological black swans, we can rarely predict the specific location and timing of large earthquakes.
+But, earthquake preparedness involves spatial planning based on forecast probabilities to focus early detection efforts and develop disaster response plans (e.g.~REFERENCE?).
+A similar focus might benefit resource management once our ability to predict the spatial probability and covariates of ecological black swans improves.
+The presence of black swans also suggests that we develop management policy that is robust to heavy tails.
+For example, setting target population abundances that are appropriately set back from critical limits may buffer black swan events \citep[e.g.][]{caddy1996}, and maintaining genetic, phenotypic, and behavioural diversity may allow some components of populations persist when others are affected by disease or extreme environmental forces \citep[e.g.][]{hilborn2003, schindler2010, anderson2014}.
+Finally, extreme and unexpected, surprising, or counterintuitive ecological dynamics offer a tremendous opportunity to learn about ecological systems, evaluate when our models break down, and adjust future management policy \citep{doak2008, pine-iii2009, lindenmayer2010}.
+In recent decades ecology has moved toward focussing on aspects of variance in addition to mean responses \citep[e.g.][]{loreau2010a, thompson2013}.
+Our results suggest that an added focus on ecological extremes represents the next frontier, particularly in the face of increased climate extremes \citep{meehl2004,ipcc2012}.
+
+%TODO WORK IN Gerber L.R. & Hilborn R. (2001). Catastrophic events and recovery from low densities in populations of otariids: implications for risk of extinction. Mammal Rev., 31, 131-150.
 
@@
 
-We thank Justin D. Yeakel, TODO, and other members of the Earth to Ocean
-research group for helpful discussions and comments. We are grateful to the
-contributors and maintainers of the Global Population Dynamics Database and to
-Compute Canada's WestGrid high-performance computing resources. Three
-silhouette images were obtained from \texttt{phylopic.org} under a Creative
-Commons Attribution 3.0 Unported license: rabbit (Sarah Werning), hoverfly
-(Gareth Monger), bird (Jean-Raphaël Guillaumin {[}photography{]} and T. Michael
-Keesey {[}vectorization{]}). Funding was provided by a Simon Fraser University
-Graduate Fellowship (SCA), the Natural Sciences and Engineering Research
-Council of Canada (NKD, ABC), the Canada Research Chairs Program (NKD), and
-TODO (TAB).
+We thank Justin D.
+Yeakel, TODO, and other members of the Earth to Ocean research group for helpful discussions and comments.
+We are grateful to the contributors and maintainers of the Global Population Dynamics Database and to Compute Canada's WestGrid high-performance computing resources.
+Three silhouette images were obtained from \texttt{phylopic.org} under a Creative Commons Attribution 3.0 Unported license: rabbit (Sarah Werning), hoverfly (Gareth Monger), bird (Jean-Raphaël Guillaumin {[}photography{]} and T.
+Michael Keesey {[}vectorization{]}).
+Funding was provided by a Simon Fraser University Graduate Fellowship (SCA), the Natural Sciences and Engineering Research Council of Canada (NKD, ABC), the Canada Research Chairs Program (NKD).
 
@@
 \caption{
-An illustration of fitting population dynamic models that allow for heavy
-tails, represented by the Student-t degrees of freedom parameter $\nu$. (a, b)
-The probability density for t distributions with a scale parameter of 1 and
-different values of $\nu$. Small values of $\nu$ create heavy tails. As $\nu$
-approaches infinity the distribution approaches the normal distribution. For
-example, at $\nu = 2$, the probability of drawing a value more than five
-standard deviations below the mean is 1.8\%, whereas the probability of drawing
-such a value from a normal distribution is nearly zero ($2.9\cdot10^{-5}$\%).
-(c--e) Simulated population dynamics from a Gompertz model with process noise
-drawn from t distributions with three different values of $\nu$. Coloured dots
-in panels c and d represent jumps with less than a 1 in 1000 chance of
-occurring in a normal distribution. (f--h) Estimates of $\nu$ from models fit
-to the times series in panels c--e. Shown are the posterior samples
-(histograms), median and interquartile range of the posterior (IQR) (dots and
-line segments), and the exponential prior on $\nu$ (dashed lines). Colour
-shading behind panels f--h illustrates the region of heavy tails.}
+An illustration of fitting population dynamic models that allow for heavy tails, represented by the Student-t degrees of freedom parameter $\nu$.
+(a, b) The probability density for t distributions with a scale parameter of 1 and different values of $\nu$.
+Small values of $\nu$ create heavy tails.
+As $\nu$ approaches infinity the distribution approaches the normal distribution.
+For example, at $\nu = 2$, the probability of drawing a value more than five standard deviations below the mean is 1.8\%, whereas the probability of drawing such a value from a normal distribution is nearly zero ($2.9\cdot10^{-5}$\%).
+(c--e) Simulated population dynamics from a Gompertz model with process noise drawn from t distributions with three different values of $\nu$.
+Coloured dots in panels c and d represent jumps with less than a 1 in 1000 chance of occurring in a normal distribution.
+(f--h) Estimates of $\nu$ from models fit to the times series in panels c--e.
+Shown are the posterior samples (histograms), median and interquartile range of the posterior (IQR) (dots and line segments), and the exponential prior on $\nu$ (dashed lines).
+Colour shading behind panels f--h illustrates the region of heavy tails.
+}
 \label{fig:didactic}
@@
 \caption{
-Estimates of how heavy-tailed population dynamics are for \nuCoefPopN\
-populations of birds, mammals, insects, and fishes. Small values of $\nu$
-($\lesssim 10$) suggest heavy-tailed black swan dynamics; larger values of
-$\nu$ suggest approximately normal-tailed dynamics. Vertical points and line
-segments represent posterior medians and 50\% / 90\% credible intervals for
-individual populations. Inset plots show probability that $\nu < 10$
-(probability of heavy tails) for populations arranged by taxonomic order and
-sorted by decreasing mean Pr($\nu < 10$). Taxonomic orders with three or fewer
-populations in panel a are omitted for space. Red to yellow points highlight
-populations with a high to moderately high probability of heavy-tailed black
-swan dynamics.
+Estimates of how heavy-tailed population dynamics are for \nuCoefPopN\ populations of birds, mammals, insects, and fishes.
+Small values of $\nu$ ($\lesssim 10$) suggest heavy-tailed black swan dynamics; larger values of $\nu$ suggest approximately normal-tailed dynamics.
+Vertical points and line segments represent posterior medians and 50\% / 90\% credible intervals for individual populations.
+Inset plots show probability that $\nu < 10$ (probability of heavy tails) for populations arranged by taxonomic order and sorted by decreasing mean Pr($\nu < 10$).
+Taxonomic orders with three or fewer populations in panel a are omitted for space.
+Red to yellow points highlight populations with a high to moderately high probability of heavy-tailed black swan dynamics.
 }
@@
 \caption{
-Potential covariates of heavy-tailed population dynamics (indicated by a high
-probability that $\nu < 10$). Shown are (a--c) parameters from the Gompertz
-heavy-tailed population model ($\sigma$, $\lambda$, $b$), (b) number of time
-steps, (c) body length, and (d) lifespan. For the Gompertz parameters, $\sigma$
-refers to the scale parameter of the Student-t process-noise distribution,
-$\lambda$ refers to the expected log abundance at the next time step at an
-abundance of one, $b$ refers to the density dependence parameter ($1$ is
-maximally density independent, $0$ is maximally density dependent, and $<0$ is
-inversely density dependent).
-%For parameter values, points and lines represent medians and 50\% credible
-%intervals of the posterior.
-Circles representing a few sharks, crustaceans, and gastropods are filled in
-white. Median and 90\% credible interval posterior predictions of a beta
-regression multilevel model are shown in panels a and d where there was a high
-probability the slope coefficient was different from zero (Supporting
-Information).
+Potential covariates of heavy-tailed population dynamics (indicated by a high probability that $\nu < 10$).
+Shown are (a--c) parameters from the Gompertz heavy-tailed population model ($\sigma$, $\lambda$, $b$), (b) number of time steps, (c) body length, and (d) lifespan.
+For the Gompertz parameters, $\sigma$ refers to the scale parameter of the Student-t process-noise distribution, $\lambda$ refers to the expected log abundance at the next time step at an abundance of one, $b$ refers to the density dependence parameter ($1$ is maximally density independent, $0$ is maximally density dependent, and $<0$ is inversely density dependent).
+Circles representing a few sharks, crustaceans, and gastropods are filled in white.
+Median and 90\% credible interval posterior predictions of a beta regression multilevel model are shown in panels a and d where there was a high probability the slope coefficient was different from zero (Supporting Information).
 }

### 2014-11-04 text/supp.tex
Integrate Trevor's comments and unwrap text

@@
 
-We applied the following data selection and quality-control rules to the Global
-Population Dynamics Database (GPDD):
+We applied the following data selection and quality-control rules to the Global Population Dynamics Database (GPDD):
 
@@
 
-\item To remove populations with unreliable population indices that could be
-  strongly confounded with economics and sampling effort, we removed all
-  populations with a sampling protocol listed as \texttt{harvest} as well
-  populations with the words \texttt{harvest} or \texttt{fur} in the cited
-  reference title.
+\item To remove populations with unreliable population indices that could be strongly confounded with economics and sampling effort, we removed all populations with a sampling protocol listed as \texttt{harvest} as well populations with the words \texttt{harvest} or \texttt{fur} in the cited reference title.
 
-\item We removed all populations with uneven sampling intervals, i.e.\ we removed
-  populations that didn't have a constant difference between the ``decimal year
-  begin'' and ``decimal year end'' columns.
+\item We removed all populations with uneven sampling intervals, i.e.\ we removed populations that didn't have a constant difference between the ``decimal year begin'' and ``decimal year end'' columns.
 
-\item We removed all populations rated as $< 2$ in the GPDD quality assessment
-  (on a scale of $1$ to $5$, with $1$ being the lowest quality data)
-  \citep[following][]{sibly2005, ziebarth2010}
+\item We removed all populations rated as $< 2$ in the GPDD quality assessment (on a scale of $1$ to $5$, with $1$ being the lowest quality data) \citep[following][]{sibly2005, ziebarth2010}.
 
-\item Populations with negative abundance values were assumed to be log10
-  values and transformed by taking $10$ to the power of the recorded abundance.
-  In many cases this was noted for the population, but not in all cases. We
-  inspected each of these \totalAssumedLog~time series to make sure our
-  assumption made sense.
+\item Populations with negative abundance values were assumed to be log10 values and transformed by taking $10$ to the power of the recorded abundance.
+In many cases this was noted for the population, but not in all cases.
+We inspected each of these \totalAssumedLog~time series to make sure our assumption made sense.
 
-\item We filled in all missing time steps with \texttt{NA} values and imputed
-  single missing values with the geometric mean of the previous and following
-  values. We chose a geometric mean to be linear on the log scale that the
-  Gompertz and Ricker-logistic models were fitted on.
+\item We filled in all missing time steps with \texttt{NA} values and imputed single missing values with the geometric mean of the previous and following values.
+We chose a geometric mean to be linear on the log scale that the Gompertz and Ricker-logistic models were fitted on.
 
-\item We filled in single recorded values of zero with the lowest non-zero
-  value in the time series \citep[following][]{brook2006a}. This assumes that
-  single values of zero result from abundance being low enough that censusing
-  overlooked individuals that were actually present. We turned multiple zero
-  values in a row into \texttt{NA} values. This implies that multiple zero
-  values were either censusing errors or caused by emigration. Regardless, our
-  population models were fitted on a multiplicative (log) scale and so could
-  not account for zero abundance. To avoid distorting the original data too
-  strongly, we removed populations in which we filled in more than four zeros.
+\item We filled in single recorded values of zero with the lowest non-zero value in the time series \citep[following][]{brook2006a}.
+This assumes that single values of zero result from abundance being low enough that censusing overlooked individuals that were actually present.
+We turned multiple zero values in a row into \texttt{NA} values.
+This implies that multiple zero values were either censusing errors or caused by emigration.
+Regardless, our population models were fitted on a multiplicative (log) scale and so could not account for zero abundance.
+To avoid distorting the original data too strongly, we removed populations in which we filled in more than four zeros.
 
-\item We removed all populations without at least four unique values
-  \citep[following][]{brook2006a}.
+\item We removed all populations without at least four unique values \citep[following][]{brook2006a}.
 
-\item We removed all populations with four or more identical values in a row
-  since these suggest either recording error or extrapolation between two
-  observations.
+\item We removed all populations with four or more identical values in a row since these suggest either recording error or extrapolation between two observations.
 
-\item We then wrote an algorithm to find the longest unbroken window of
-  abundance (no \texttt{NA}s) with at least $20$ time steps in each population
-  time series. If there were any populations with multiple windows of identical
-  length, we took the most recent window. This is a longer window than used in
-  some previous analyses \citep[e.g.][]{brook2006a}, but since our model
-  attempts to capture the shape of the distribution tails, our model requires
-  more data.
+\item We then wrote an algorithm to find the longest unbroken window of abundance (no \texttt{NA}s) with at least $20$ time steps in each population time series.
+If there were any populations with multiple windows of identical length, we took the most recent window.
+This is a longer window than used in some previous analyses \citep[e.g.][]{brook2006a}, but since our model attempts to capture the shape of the distribution tails, our model requires more data.
 
-\item Finally, we removed GPDD Main ID \texttt{20531}, which we noticed was
-  a duplicate of \texttt{10139} (a heron population). We removed a limited
-  number of populations from class Angiospermopsida and Bacillariophyceae to
-  focus the taxonomy in our analysis. We also removed any populations with an
-  \texttt{Unknown} taxonomic class.
+\item Finally, we removed GPDD Main ID \texttt{20531}, which we noticed was a duplicate of \texttt{10139} (a heron population).
+We removed a limited number of populations from class Angiospermopsida and Bacillariophyceae to focus the taxonomy in our analysis.
+We also removed any populations with an \texttt{Unknown} taxonomic class.
 
@@
 \noindent
-We provide a supplemental figure of all the time series included in our analysis
-and indicate which values were interpolated (\percImputedPops\% of populations
-had at least one point interpolated and only \percImputedPoints\% of the total
-observations were interpolated) (Fig.~\ref{fig:all-ts}). Table S1 shows the
-final taxonomic breakdown and the number of populations with interpolated
-values.
+We provide a supplemental figure of all the time series included in our analysis and indicate which values were interpolated (\percImputedPops\% of populations had at least one point interpolated also only \percImputedPoints\% of the total observations were interpolated) (Fig.~\ref{fig:all-ts}).
+Note that interpolation is highly unlikely to lead to black swan detections, since black swans involve extreme increases or decreases.
+Table S1 shows the final taxonomic breakdown and the number of populations with interpolated values.
 
@@
 \sigma &\sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)\\ \nu &\sim
-\mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2). \end{align*}
-Our prior on $b$ was uninformative between values of $-1$ and $2$. We would not
-expect values of $b$ with levels of density dependence as low as $-1$ (very
-strong inverse density dependence), nor would we generally expect values above
-$1$. We allowed values of $b$ above $1$ to allow for non-stationary time series
-of growth rates. The estimates of $b$ were well within these bounds. Our prior
-on $\lambda$ was very weakly informative within the range of expected values
-for population growth and is similar to the default priors suggested by
-\citet{gelman2008d} for intercepts of regression models. Our Half-Cauchy prior
-on $\sigma$ follows \citet{gelman2006c} and \citet{gelman2008d} and the scale
-parameter of $2.5$ is based on our expected range of the value in nature from
-previous studies \citep[e.g.][]{connors2014}. In our testing of a subsample of
-populations, our parameter estimates were not qualitatively changed by
-switching to an uninformative uniform prior on $\sigma$, but the weakly
-informative Half-Cauchy prior substantially sped up chain convergence.
+\mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2).
+\end{align*}
+Our prior on $b$ was uninformative between values of $-1$ and $2$.
+We would not expect values of $b$ with levels of density dependence as low as $-1$ (very strong inverse density dependence), nor would we generally expect values above $1$.
+We allowed values of $b$ above $1$ to allow for non-stationary time series of growth rates.
+The estimates of $b$ were well within these bounds.
+Our prior on $\lambda$ was very weakly informative within the range of expected values for population growth and is similar to the default priors suggested by \citet{gelman2008d} for intercepts of regression models.
+Our Half-Cauchy prior on $\sigma$ follows \citet{gelman2006c} and \citet{gelman2008d} and the scale parameter of $2.5$ is based on our expected range of the value in nature from previous studies \citep[e.g.][]{connors2014}.
+In our testing of a subsample of populations, our parameter estimates were not qualitatively changed by switching to an uninformative uniform prior on $\sigma$, but the weakly informative Half-Cauchy prior substantially sped up chain convergence.
 
-Our prior on $\nu$ was based on \citet{fernandez1998}. They chose an
-exponential rate parameter of $0.1$. We chose a less informative rate parameter
-of $0.01$ and truncated the distribution at $2$, since at $\nu < 2$ the
-variance of the t distribution is undefined. This prior gives only a $7.7$\%
-probability that $\nu < 10$ but constrains the sampling sufficiently to avoid
-wandering off towards infinity --- above approximately $\nu = 20$ the
-t distribution is so similar to the normal distribution
-(Fig.~\ref{fig:didactic}) that time series of the length considered here are
-unlikely to be informative about the precise value of $\nu$. In the scenario
-where the data are uninformative about heavy tails
-(e.g.~Fig.~\ref{fig:didactic}e,~h), the posterior will approximately match the
-prior (prior median $= 71$, mean $= 102$) and the metrics used in our paper
-(e.g.~Pr$(\nu < 10) > 0.5$) are unlikely to flag the population as heavy
-tailed.
+Our prior on $\nu$ was based on \citet{fernandez1998}.
+They chose an exponential rate parameter of $0.1$.
+We chose a less informative rate parameter of $0.01$ and truncated the distribution at $2$, since at $\nu < 2$ the variance of the t distribution is undefined.
+This prior gives only a $7.7$\% probability that $\nu < 10$ but constrains the sampling sufficiently to avoid wandering off towards infinity --- above approximately $\nu = 20$ the t distribution is so similar to the normal distribution (Fig.~\ref{fig:didactic}) that time series of the length considered here are unlikely to be informative about the precise value of $\nu$.
+In the scenario where the data are uninformative about heavy tails (e.g.~Fig.~\ref{fig:didactic}e,~h), the posterior will approximately match the prior (prior median $= 71$, mean $= 102$) and the metrics used in our paper (e.g.~Pr$(\nu < 10) > 0.5$) are unlikely to flag the population as heavy tailed.
 
-We fitted our models with Stan 2.4.0 \citep{stan-manual2014}, and R 3.1.1
-\citep{r2014}. We began with four chains and $2000$ iterations, discarding the
-first $1000$ as warm up (i.e.~4000 total samples). If $\hat{R}$ (the potential
-scale reduction factor --- a measure of chain convergence) was greater than
-$1.05$ for any parameter or the minimum effective sample size,
-$n_\mathrm{eff}$, (a measure of the effective number of uncorrelated samples)
-for any parameter was less than $200$, we doubled both the total iterations
-and warm up period and sampled from the model again. These thresholds are in
-excess of the minimums recommended by \citet{gelman2006a} of $\hat{R} < 1.1$
-and effective sample size $> 100$ for reliable point estimates and confidence
-intervals. In the majority of cases our minimum thresholds were greatly
-exceeded. We continued this procedure up to $8000$ iterations ($16000$ total
-samples) by which all chains were deemed to have sufficiently converged. These
-chain lengths may seem low to those familiar with software such as WinBUGS or
-JAGS, but the No-U-Turn Hamiltonian Markov chain Monte Carlo Sampler in Stan
-generally requires far fewer iterations to obtain equivalent effective sample
-sizes \citep{stan-manual2014}.
+We fitted our models with Stan 2.4.0 \citep{stan-manual2014}, and R 3.1.1 \citep{r2014}.
+We began with four chains and $2000$ iterations, discarding the first $1000$ as warm up (i.e.~4000 total samples).
+If $\hat{R}$ (the potential scale reduction factor --- a measure of chain convergence) was greater than $1.05$ for any parameter or the minimum effective sample size, $n_\mathrm{eff}$, (a measure of the effective number of uncorrelated samples) for any parameter was less than $200$, we doubled both the total iterations and warm up period and sampled from the model again.
+These thresholds are in excess of the minimums recommended by \citet{gelman2006a} of $\hat{R} < 1.1$ and effective sample size $> 100$ for reliable point estimates and confidence intervals.
+In the majority of cases our minimum thresholds were greatly exceeded.
+We continued this procedure up to $8000$ iterations ($16000$ total samples) by which all chains were deemed to have sufficiently converged.
+These chain lengths may seem low to those familiar with software such as WinBUGS or JAGS, but the No-U-Turn Hamiltonian Markov chain Monte Carlo Sampler in Stan generally requires far fewer iterations to obtain equivalent effective sample sizes \citep{stan-manual2014}.
 
@@
 
-We fit four alternative population models to the time-series data to check how
-they would influence our conclusions. Our alternative models allowed for
-autocorrelation in the residuals, assumed no density dependence, allowed for
-observation error, or assumed a Ricker-logistic functional form. The range of
-percentages of black swans by taxonomic class cited in the abstract and results
-are based on lower and upper limits across our main Gompertz model and these
-four alternative models.
+We fit four alternative population models to the time-series data to check how they would influence our conclusions.
+Our alternative models allowed for autocorrelation in the residuals, assumed no density dependence, allowed for observation error, or assumed a Ricker-logistic functional form.
+The range of percentages of black swans by taxonomic class cited in the abstract and results are based on lower and upper limits across our main Gompertz model and these four alternative models.
 
@@
 \end{align*}
-In addition to the parameters in the original Gompertz model, we estimate an
-additional parameter $\phi$, which represents the dependence of
-subsequent residuals. Based on the results of previous analyses with the GPDD
-\citep[e.g.][]{connors2014} and the chosen priors in previous analyses
-\citep[e.g.][]{thorson2014a} and to greatly speed up chain convergence when
-running our model across all populations, we placed a weakly informative prior
-on $\phi$ that assumed the greatest probability density near zero with the
-reduced possibility of $\phi$ being near $-1$ or $1$. Specifically, we chose
-$\phi \sim \mathrm{Truncated\mhyphen Normal}(0, 1, \mathrm{min.} = -1,
-\mathrm{max.} = 1)$. The MCMC chains did
-not converge for \modelsNoConvergeAROne\ populations according to
-our criteria ($\widehat{R} < 1.05, n_\mathrm{eff}
-> 200$) after 8000 iterations of four chains. This included only
-\modelsNoConvergeAROneHeavyBase\ populations in which
-Pr($\nu < 10$) $> 0.5$ categorized them as heavy in the main Gompertz model.
+In addition to the parameters in the original Gompertz model, we estimate an additional parameter $\phi$, which represents the dependence of subsequent residuals.
+Based on the results of previous analyses with the GPDD \citep[e.g.][]{connors2014} and the chosen priors in previous analyses \citep[e.g.][]{thorson2014a} and to greatly speed up chain convergence when running our model across all populations, we placed a weakly informative prior on $\phi$ that assumed the greatest probability density near zero with the reduced possibility of $\phi$ being near $-1$ or $1$.
+Specifically, we chose $\phi \sim \mathrm{Truncated\mhyphen Normal}(0, 1, \mathrm{min.} = -1, \mathrm{max.} = 1)$.
+The MCMC chains did not converge for \modelsNoConvergeAROne\ populations according to our criteria ($\widehat{R} < 1.05, n_\mathrm{eff} > 200$) after 8000 iterations of four chains.
+This included only \modelsNoConvergeAROneHeavyBase\ populations in which Pr($\nu < 10$) $> 0.5$ categorized them as heavy in the main Gompertz model.
 We did not include these models in Fig.~\ref{fig:alt}.
@@
 
-We fit a simplified version of the Gompertz model in which the density
-dependence parameter $b$ was fixed at $1$ (density independent). This is
-equivalent to fitting a random walk model (with drift) to the $\log$ abundances
-or assuming the growth rates are drawn from a stationary distribution. The
-model was as follows:
+We fit a simplified version of the Gompertz model in which the density dependence parameter $b$ was fixed at $1$ (density independent).
+This is equivalent to fitting a random walk model (with drift) to the $\log$ abundances or assuming the growth rates are drawn from a stationary distribution.
+The model was as follows:
 \begin{align*}
@@
 \end{align*}
-We fit this model for three reasons: (1) it is computationally simpler and so
-provides a check that our more complicated full Gompertz model was obtaining
-reasonable estimates of $\nu$, (2) it provides a test of whether density
-dependence was systematically affecting our perception of heavy tails, (3) it
-matches how some previous authors have modelled heavy tails without accounting
-for density dependence \citep{segura2013}.
+We fit this model for three reasons: (1) it is computationally simpler and so provides a check that our more complicated full Gompertz model was obtaining reasonable estimates of $\nu$, (2) it provides a test of whether density dependence was systematically affecting our perception of heavy tails, (3) it matches how some previous authors have modelled heavy tails without accounting for density dependence \citep{segura2013}.
 
@@
 
-Observation error can bias parameter estimates \citep[e.g.][]{knape2012} and is
-known to affect the ability to detect extreme events \citep{ward2007}. In our
-main analysis, we fit a model that ignored observation error. One way to
-account for observation error would be to fit a full state-space model that
-simultaneously estimates the magnitude of process noise and observation error.
-However, simultaneously estimating observation and process noise is
-a challenging problem (e.g.\ because the observation and process noise
-parameters tend to negatively covary in model fitting) and is known to
-sometimes result in identifiability issues with the Gompertz population model
-\citep{knape2008}. Furthermore, our model was applied to hundreds of time
-series, often of short length (as few as 20 time steps) and our model estimates
-an additional parameter --- the shape of the process deviation tails ---
-potentially making identifiability and computational issues even greater.
-Therefore, we considered a version of the base Gompertz model that allowed for
-a fixed level of observation error:
+Observation error can bias parameter estimates \citep[e.g.][]{knape2012} and is known to affect the ability to detect extreme events \citep{ward2007}.
+In our main analysis, we fit a model that ignored observation error.
+One way to account for observation error would be to fit a full state-space model that simultaneously estimates the magnitude of process noise and observation error.
+However, simultaneously estimating observation and process noise is a challenging problem (e.g.\ because the observation and process noise parameters tend to negatively covary in model fitting) and is known to sometimes result in identifiability issues with the Gompertz population model \citep{knape2008}.
+Furthermore, our model was applied to hundreds of time series, often of short length (as few as 20 time steps) and our model estimates an additional parameter --- the shape of the process deviation tails --- potentially making identifiability and computational issues even greater.
+Therefore, we considered a version of the base Gompertz model that allowed for a fixed level of observation error:
 \begin{align*}
@@
 \end{align*}
-where $U$ represents the unobserved state vector, $\sigma_\mathrm{obs}$
-represents the standard deviation of observation error (on a log scale), and
-$\sigma_\mathrm{proc}$ represent the process noise scale parameter. We set
-$\sigma_\mathrm{obs}$ to $0.2$, which represents the upper limit of values often
-used in simulation analyses \citep[e.g.][]{valpine2002, thorson2014b}.
+where $U$ represents the unobserved state vector, $\sigma_\mathrm{obs}$ represents the standard deviation of observation error (on a log scale), and $\sigma_\mathrm{proc}$ represent the process noise scale parameter.
+We set $\sigma_\mathrm{obs}$ to $0.2$, which represents the upper limit of values often used in simulation analyses \citep[e.g.][]{valpine2002, thorson2014b}.
 
@@
 \end{align*}
-where $r_\mathrm{max}$ represents the theoretical maximum population growth
-rate that is obtained when $N_t$ (abundance at time $t$) $= 0$. The parameter $K$ represents
-the carrying capacity and, as before, $x_t$ represents the $\log$ transformed
-abundance at time $t$. The Ricker-logistic model assumes a linear decrease in
-population growth rate with increases in abundance.
-In contrast, the Gompertz model assumes a linear decrease in population growth
-rate with increases in \textit{log} abundance ($x_t$) \citep[e.g.][]{thibaut2012}.
+where $r_\mathrm{max}$ represents the theoretical maximum population growth rate that is obtained when $N_t$ (abundance at time $t$) $= 0$.
+The parameter $K$ represents the carrying capacity and, as before, $x_t$ represents the $\log$ transformed abundance at time $t$.
+The Ricker-logistic model assumes a linear decrease in population growth rate with increases in abundance.
+In contrast, the Gompertz model assumes a linear decrease in population growth rate with increases in \textit{log} abundance ($x_t$) \citep[e.g.][]{thibaut2012}.
 
-To fit the Ricker-logistic models, we chose a prior on $K$ uniform between zero
-and twice the maximum observed abundance (\citet{clark2010} chose uniform
-between zero and maximum observed, which is more informative). We set the prior
-on $r_\mathrm{max}$ as uniform between 0 and 20 as in \citet{clark2010}. We used
-the same priors on $\nu$ and $\sigma$ as in the Gompertz model.
+To fit the Ricker-logistic models, we chose a prior on $K$ uniform between zero and twice the maximum observed abundance (\citet{clark2010} chose uniform between zero and maximum observed, which is more informative).
+We set the prior on $r_\mathrm{max}$ as uniform between 0 and 20 as in \citet{clark2010}.
+We used the same priors on $\nu$ and $\sigma$ as in the Gompertz model.
 
@@
 
-We performed two types of simulation testing. First, we tested how easily the
-Student-t distribution $\nu$ parameter could be recovered given different true
-values of $\nu$ and different sample sizes. Second, we tested the ability of the
-heavy-tailed Gompertz model to obtain unbiased parameter estimates of $\nu$
-given that a set of process deviations was provided in which the effective
-$\nu$ value was close to the true $\nu$ value.
+We performed two types of simulation testing.
+First, we tested how easily the Student-t distribution $\nu$ parameter could be recovered given different true values of $\nu$ and different sample sizes.
+Second, we tested the ability of the heavy-tailed Gompertz model to obtain unbiased parameter estimates of $\nu$ given that a set of process deviations was provided in which the effective $\nu$ value was close to the true $\nu$ value.
 
-We separated our simulation into these two components to avoid confounding two
-issues. (1) With smaller sample sizes, there may not be a stochastic draw from
-the tails of a distribution. In that case, no model, no matter how perfect,
-will be able to detect the shape of the tails. (2) Complex models may return
-biased parameter estimates if there are conceptual, computational, or coding
-errors. Our first simulation tested the first issue and our second simulation
-tested the latter. In general, our simulations show that, if anything, our
-model under predicts the magnitude and probability of heavy tailed events ---
-especially given the length of the time series in the GPDD.
+We separated our simulation into these two components to avoid confounding two issues.
+(1) With smaller sample sizes, there may not be a stochastic draw from the tails of a distribution.
+In that case, no model, no matter how perfect, will be able to detect the shape of the tails.
+(2) Complex models may return biased parameter estimates if there are conceptual, computational, or coding errors.
+Our first simulation tested the first issue and our second simulation tested the latter.
+In general, our simulations show that, if anything, our model under predicts the magnitude and probability of heavy tailed events --- especially given the length of the time series in the GPDD.
 
@@
 
-First, we tested the ability to estimate $\nu$ given different true values of
-$\nu$ and different sample sizes. We took stochastic draws from t distributions with
-different $\nu$ values ($\nu = 3, 5, 10,$ and $10^6$ [$\approx$ normal]), with
-central tendency parameters of $0$, and scale parameters of $1$. We started with
-$1600$ stochastic draws and then fitted the models again at the first $800, 400,
-200, 100, 50,$ and $25$ draws. Each time we recorded the posterior samples of
-$\nu$.
+First, we tested the ability to estimate $\nu$ given different true values of $\nu$ and different sample sizes.
+We took stochastic draws from t distributions with different $\nu$ values ($\nu = 3, 5, 10,$ and $10^6$ [$\approx$ normal]), with central tendency parameters of $0$, and scale parameters of $1$.
+We started with $1600$ stochastic draws and then fitted the models again at the first $800, 400, 200, 100, 50,$ and $25$ draws.
+Each time we recorded the posterior samples of $\nu$.
 
-We found that we could consistently and precisely recover median posterior
-estimates of $\nu$ near the true value of $\nu$ with large samples ($\ge 200$)
-(Fig.~\ref{fig:sim-gompertz} upper panels). At smaller samples we could still
-usually qualitatively distinguish heavy from not-heavy tails, but the model
-tended to underestimate how heavy the tails were. At the same time, at smaller
-sample sizes, the model tended to overestimate how large the scale parameter
-was (Fig.~\ref{fig:sim-gompertz} lower panels).
+We found that we could consistently and precisely recover median posterior estimates of $\nu$ near the true value of $\nu$ with large samples ($\ge 200$) (Fig.~\ref{fig:sim-gompertz} upper panels).
+At smaller samples we could still usually qualitatively distinguish heavy from not-heavy tails, but the model tended to underestimate how heavy the tails were.
+At the same time, at smaller sample sizes, the model tended to overestimate how large the scale parameter was (Fig.~\ref{fig:sim-gompertz} lower panels).
 
@@
 
-In the second part of our simulation testing, we tested the ability of the
-heavy-tailed Gompertz model to obtain unbiased parameter estimates 
-when the process noise was chosen so that appropriate tail events were present.
-To generate these process deviations for the $\nu = 3$ and $\nu = 5$ scenarios, we
-repeatedly drew proposed candidate process deviations and estimated the central
-tendency, scale, and $\nu$ values each time. We recorded when $\hat{\nu}$
-(median of the posterior) was within $0.2$ CVs (coefficient of variations) of
-the true $\nu$ value and used this set of random seed values in our Gompertz
-simulation. The following simplified R code illustrates this procedure 
-(the actual code is available at \url{https://github.com/seananderson/heavy-tails}):
+In the second part of our simulation testing, we tested the ability of the heavy-tailed Gompertz model to obtain unbiased parameter estimates when the process noise was chosen so that appropriate tail events were present.
+To generate these process deviations for the $\nu = 3$ and $\nu = 5$ scenarios, we repeatedly drew proposed candidate process deviations and estimated the central tendency, scale, and $\nu$ values each time.
+We recorded when $\hat{\nu}$ (median of the posterior) was within $0.2$ CVs (coefficient of variations) of the true $\nu$ value and used this set of random seed values in our Gompertz simulation.
+The following simplified R code illustrates this procedure (the actual code is available at \url{https://github.com/seananderson/heavy-tails}):
 

### 2014-11-04 text/sparks.tex
Fix typo in spark table

@@
 Severe winters in 1929, 1940--1942, and 1962--1963; 1963 event so severe that
-recovery took three times loner than expected &
+recovery took three times longer than expected &
 \citep{stafford1971} \\

### 2014-11-04 text/supp.tex
Add TODOs to supp. figures

@@
 \setcounter{table}{0}
+TODO ``Chondrichtyhes'' IS A TYPO IN THE GPDD DATA\\
+TODO ADD TOTAL NUMBER OF POINTS ON RIGHT COLUMN
 \input{../analysis/stat-table}
@@
 \caption{
+TODO WILL SPLIT OVER PAGES BY CLASS
   All filtered times series used in our analysis. The abundances are shown on
@@
 \caption{
+TODO WILL ADD ALTERNATIVE $\nu$ PRIORS
   Probability density of the Bayesian priors for the Gompertz models. From left
@@
 \caption{
+TODO WILL INDICATE UPS AND DOWNS AND MATCH STYLE TO FIG S1
   Time series for populations with Pr($\nu < 10$) $>$ 0.5 using the base
@@
 \caption{
+TODO WILL COMBINE WITH COEFFICIENT POSTERIORS IN MAIN TEXT FIG 4
   Taxonomic-order-level posterior estimates of probability that $\nu < 10$ from
@@
 \caption{
+TODO WILL REMOVE ALL OR MOST OF THIS
   Simulation testing the Gompertz estimation model when the process deviation
@@
 \caption{
+TODO WILL REMOVE ALL OR MOST OF THIS
   The distribution of median estimates from the same simulation as described in
@@
   simulation iterations, so there is some level of remaining sampling noise.
-  \textit{TODO wondering now if I should run these for more iterations to make
-    these less noisy.}
 }
@@
 \caption{
+TODO WILL REMOVE ALL OR MOST OF THIS
   The probability that $\nu < 10$ (i.e.\ the approximate probability of heavy
@@
 \caption{
+TODO WILL MOVE TO MAIN TEXT WITH ORDER-LEVEL ESTIMATES
   Main effect posterior densities for potential correlates of heavy-tailed
@@
 \caption{
+TODO WILL DELETE THIS
 Expected percent increase in the probability of detecting heavy tails, Pr($\nu

### 2014-11-04 text/ms.tex
Tidy up some TODOs
Black swan-> black-swan when an adjective

@@
 
-%TODO look up doi or volume for Drake 2014 Ecology
-
-%\noindent
-%\emph{Main message, to be deleted (25 words)}: Black swans are present \& taxonomically widespread but rare in population dynamics.
-Extreme climate, predation, parasites, and their interactions are common causes; intrinsic drivers are unclear.
-
+% TODO: hyphenate the adjective 'black swan'? E.g. black-swan event, black swans.
 \section{Abstract}
@@
 A survey of ecologists indicated that surprising outcomes of field experiments are far more common than we assume \citep{doak2008}, and a number of major environmental events on geologic and contemporary time scales could be considered to have caused ecological black swans \citep{nunez2012}.
-Black swans, or heavy tailed distributions, have been discovered in time-series of marine microbe populations \citet{segura2013} and time to extinction for experimental waterflea populations \citet{drake2014}.
+Black swans, or heavy-tailed distributions, have been discovered in time-series of marine microbe populations \citet{segura2013} and time to extinction for experimental waterflea populations \citet{drake2014}.
 As far as we can tell there have been two systematic surveys for ecological black swans: one with North American breeding birds and the other with the Global Population Dynamics Database, which found little clear evidence for them \citep{keitt1998,allen2001,halley2002}.
@@
 We apply our model to \NPops\ of populations from four taxonomic classes.
-We estimate the frequency and strength of black swans in population dynamics across taxonomy; verify some causes of black swan events; and attempt to identify characteristics of time series or intrinsic life-history characteristics that are associated with heavy-tailed dynamics.
+We estimate the frequency and strength of black swans in population dynamics across taxonomy; verify some causes of black-swan events; and attempt to identify characteristics of time series or intrinsic life-history characteristics that are associated with heavy-tailed dynamics.
 %We find a high probability of ecological
@@
 
-By allowing the process noise to be drawn from a Student-t distribution we can estimate the degree to which the process errors have heavy tails and are thereof evidence of black swan events (Fig.~\ref{fig:didactic}a, b).
+By allowing the process noise to be drawn from a Student-t distribution we can estimate the degree to which the process errors have heavy tails and are thereof evidence of black-swan events (Fig.~\ref{fig:didactic}a, b).
 For example, at $\nu = 2$, the probability of drawing a value more than five standard deviations below the mean is $0.018$, whereas the probability of drawing such a value from a normal distribution is tiny ($2.9\cdot10^{-7}$) TODO CHECK.
@@
 
-Across populations, the probability of observing black swan dynamics was positively related to time-series length and negatively related to magnitude of process noise ($\sigma$) but not clearly related to population growth rate ($\lambda$), density dependence ($b$), or maximum lifespan (Figs~\ref{fig:correlates}, \ref{fig:correlate-coefs}).
-Longer time-series length was the strongest covariate of observing black swan dynamics.
+Across populations, the probability of observing black-swan dynamics was positively related to time-series length and negatively related to magnitude of process noise ($\sigma$) but not clearly related to population growth rate ($\lambda$), density dependence ($b$), or maximum lifespan (Figs~\ref{fig:correlates}, \ref{fig:correlate-coefs}).
+Longer time-series length was the strongest covariate of observing black-swan dynamics.
 For example, we were about \pIncHeavyNThirtyNSixty~times more likely to observe heavy tails in a population with 60 time steps than in one with 30 time steps (Fig.~\ref{fig:perc-inc-p}).
@@
 TODO QUANTIFY UPS AND DOWNS --- SKEW
-We assembled the documented causes of black swan population dynamics in seven exemplar populations with high probabilities of heavy tails (Table 1).
+We assembled the documented causes of black-swan population dynamics in seven exemplar populations with high probabilities of heavy tails (Table 1).
 The majority of documented events were downward black swans and involved a series of unfortunate events.
 For example, a shortage of nest sites for the shag long-necked bird (TODO SCI.\ NAME) in the United Kingdom reduced population productivity and made the population more vulnerable when a red tide hit in 1968 \citep{potts1980}.
-The population experienced a black swan downswing, but the new availability of nest sites resulted in a rapid population upswing \citep{potts1980}.
+The population experienced a black-swan downswing, but the new availability of nest sites resulted in a rapid population upswing \citep{potts1980}.
 As another example, an interaction between environmental- and predation-mediated population cycles are thought to have caused a downward black swan for a water vole population \citep{saucy1994}.
 Other black swans were the result of a sequence of extreme climate events on their own.
-For example, severe winters in 1929, 1940--1942, and 1962--1963 were associated with black swan downswings in grey heron abundance in the United Kingdom \citep{stafford1971}.
-Our analysis finds that the last event was a combination of two black swan events in a row and it took the population three times longer to recover than predicted \citep{stafford1971}.
+For example, severe winters in 1929, 1940--1942, and 1962--1963 were associated with black-swan downswings in grey heron abundance in the United Kingdom \citep{stafford1971}.
+Our analysis finds that the last event was a combination of two black-swan events in a row and it took the population three times longer to recover than predicted \citep{stafford1971}.
 
@@
 In verified cases, black swans were often a result of the interaction between elements of extreme climate, predators and parasite cycles, and strong changes in human pressures.
-Our empirical results, sensitivity analyses, and simulation tests suggest that estimating the tail shape of process noise is a viable method of detecting black swan population dynamics and if anything will underestimate the probability of black swans.
+Our empirical results, sensitivity analyses, and simulation tests suggest that estimating the tail shape of process noise is a viable method of detecting black-swan population dynamics and if anything will underestimate the probability of black swans.
 The presence of black swans highlights the importance of developing management strategies that detect quickly, respond to, and are robust to extremes in population dynamics --- particularly as the frequency and magnitude of climatic extremes increase over the next century \citep{easterling2000,ipcc2012}.
@@
 For heavy-tailed distributions, \citet{halley2002} fit the extremely heavy-tailed Cauchy distribution and the four-parameter Levy stable distribution and found little information criteria support for these distributions in longer time series.
-However, black swan events are by definition rare, and the majority of time series in the GPDD are short for these purposes.
+However, black-swan events are by definition rare, and the majority of time series in the GPDD are short for these purposes.
 Therefore, we would not expect to observe black swans in a large proportion of populations.
@@
 
-We might expect to observe black swan dynamics in ecological time series because of either unmodelled intrinsic properties of populations or extrinsic forces acting on populations.
+We might expect to observe black-swan dynamics in ecological time series because of either unmodelled intrinsic properties of populations or extrinsic forces acting on populations.
 Since a t-distribution can be formed by a mixture of normal distributions \citep{gelman2014}, we could observe heavy tails if we miss some underlying mixture of intrinsic processes \citep{allen2001}.
 That process might be an aggregation of populations across space, or population diversity, or some intrinsic change in population variability through time.
-Extrinsic forces could also cause black swan dynamics \citep[e.g.][]{nunez2012}.
+Extrinsic forces could also cause black-swan dynamics \citep[e.g.][]{nunez2012}.
 These forces could be extreme themselves.
 For example, extreme climate, predation from (or competition with) other species experiencing black swans, or sharp changes in human pressure such hunting, fishing, or habitat destruction might cause black swans.
-Alternatively, the interaction of multiple ``normal'' extrinsic forces could give rise to black swan ecological dynamics.
+Alternatively, the interaction of multiple ``normal'' extrinsic forces could give rise to black-swan ecological dynamics.
 This could occur if the interaction is synergistic \citep[e.g.][]{kirby2009} or even if non-synergistic interactions experience a rare alignment \citep{denny2009}.
@@
 Although we expect that we would find qualitatively similar evidence for black swans in many other large taxonomic or geographic samples of populations, the common forces driving those black swans likely differ (e.g.~severe winters in Table~1).
-In addition to a possibly biased sample of populations, some black swan detections could just be recording mistakes, or conversely, some extreme observations may have been discarded or altered if they were erroneously suspected of being recording mistakes.
+In addition to a possibly biased sample of populations, some black-swan detections could just be recording mistakes, or conversely, some extreme observations may have been discarded or altered if they were erroneously suspected of being recording mistakes.
 A final caveat is that the temporal scales of observation and population dynamics vary considerably across populations in the GPDD and this likely influences the detection of heavy tails.
 For example, if we make frequent observations relative to generation time (e.g.~for many large-bodied mammals) we will average across generations and perhaps miss black swans.
-Conversely, if we census populations infrequently relative to generation time (e.g.~many insects in the GPDD) the recorded data may average across extreme and less-extreme events and also dampen black swan dynamics.
+Conversely, if we census populations infrequently relative to generation time (e.g.~many insects in the GPDD) the recorded data may average across extreme and less-extreme events and also dampen black-swan dynamics.
 
@@
 Financial analysts are concerned with the shape of financial-return-distribution tails because these impact estimates of risk --- the probability of a specific magnitude of undesired event occurring \citep{rachev2008}.
-A comparable focus in ecology would increase our estimates of extinction risk, since these would be disproportionately impacted by black swan events.
+A comparable focus in ecology would increase our estimates of extinction risk, since these would be disproportionately impacted by black-swan events.
 TODO END ON THIS
@@
 Ecological resource management can draw from other disciplines that focus on heavy tails.
-For example, earthquake preparedness and response is focussed on black swan events.
+For example, earthquake preparedness and response is focussed on black-swan events.
 Similarly to ecological black swans, we can rarely predict the specific location and timing of large earthquakes.
@@
 The presence of black swans also suggests that we develop management policy that is robust to heavy tails.
-For example, setting target population abundances that are appropriately set back from critical limits may buffer black swan events \citep[e.g.][]{caddy1996}, and maintaining genetic, phenotypic, and behavioural diversity may allow some components of populations persist when others are affected by disease or extreme environmental forces \citep[e.g.][]{hilborn2003, schindler2010, anderson2014}.
+For example, setting target population abundances that are appropriately set back from critical limits may buffer black-swan events \citep[e.g.][]{caddy1996}, and maintaining genetic, phenotypic, and behavioural diversity may allow some components of populations persist when others are affected by disease or extreme environmental forces \citep[e.g.][]{hilborn2003, schindler2010, anderson2014}.
 Finally, extreme and unexpected, surprising, or counterintuitive ecological dynamics offer a tremendous opportunity to learn about ecological systems, evaluate when our models break down, and adjust future management policy \citep{doak2008, pine-iii2009, lindenmayer2010}.
@@
 Estimates of how heavy-tailed population dynamics are for \nuCoefPopN\ populations of birds, mammals, insects, and fishes.
-Small values of $\nu$ ($\lesssim 10$) suggest heavy-tailed black swan dynamics; larger values of $\nu$ suggest approximately normal-tailed dynamics.
+Small values of $\nu$ ($\lesssim 10$) suggest heavy-tailed black-swan dynamics; larger values of $\nu$ suggest approximately normal-tailed dynamics.
 Vertical points and line segments represent posterior medians and 50\% / 90\% credible intervals for individual populations.
@@
 Taxonomic orders with three or fewer populations in panel a are omitted for space.
-Red to yellow points highlight populations with a high to moderately high probability of heavy-tailed black swan dynamics.
+Red to yellow points highlight populations with a high to moderately high probability of heavy-tailed black-swan dynamics.
 }

### 2014-11-04 text/supp.tex
Black swan-> black-swan when an adjective

@@
 We provide a supplemental figure of all the time series included in our analysis and indicate which values were interpolated (\percImputedPops\% of populations had at least one point interpolated also only \percImputedPoints\% of the total observations were interpolated) (Fig.~\ref{fig:all-ts}).
-Note that interpolation is highly unlikely to lead to black swan detections, since black swans involve extreme increases or decreases.
+Note that interpolation is highly unlikely to lead to black-swan detections, since black swans involve extreme increases or decreases.
 Table S1 shows the final taxonomic breakdown and the number of populations with interpolated values.
@@
 taxonomy than body length. Lifespan is also more directly related to time and
-potential mechanisms driving black swan dynamics.
+potential mechanisms driving black-swan dynamics.
 

### 2014-11-04 text/supp.tex
Unwrap more supp.tex, bump version

@@
 
-We then fitted our Gompertz models to the simulated datasets with all
-parameters (except $\nu$) set near the median values estimated in the GPDD. We
-repeated this with $50$ and $100$ samples without observation error, $50$
-samples with observation error ($\sigma_\mathrm{obs} = 0.2$), and $50$ samples
-with the same observation error and a Gompertz model that allowed for correctly
-specified observation error magnitude.
+We then fitted our Gompertz models to the simulated datasets with all parameters (except $\nu$) set near the median values estimated in the GPDD.
+We repeated this with $50$ and $100$ samples without observation error, $50$ samples with observation error ($\sigma_\mathrm{obs} = 0.2$), and $50$ samples with the same observation error and a Gompertz model that allowed for correctly specified observation error magnitude.
 
-Our results show that the Gompertz model can recapture the true value of $\nu$
-when the process noise was chosen so that appropriate tail events were present
-(Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}, red and green
-symbols in the top rows). Likewise, the other Gompertz parameters were
-estimated without any systematic bias (Figs~\ref{fig:sim-gompertz} and
-\ref{fig:sim-gompertz-boxplots}, red and green symbols). The addition of
-observation error caused the model to tend to underestimate the degree of
-heavy-tailedness, overestimate the magnitude of process noise, somewhat
-overestimate $\lambda$, and overestimate density dependence (blue symbols in
-Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}). The
-overestimation of density dependence with observation error is a known
-phenomenon \citep{knape2012}. Fitting a model with correctly specified
-observation error made marginal improvements to model bias (purple symbols in
-Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}).
+Our results show that the Gompertz model can recapture the true value of $\nu$ when the process noise was chosen so that appropriate tail events were present (Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}, red and green symbols in the top rows).
+Likewise, the other Gompertz parameters were estimated without any systematic bias (Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}, red and green symbols).
+The addition of observation error caused the model to tend to underestimate the degree of heavy-tailedness, overestimate the magnitude of process noise, somewhat overestimate $\lambda$, and overestimate density dependence (blue symbols in Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}).
+The overestimation of density dependence with observation error is a known phenomenon \citep{knape2012}.
+Fitting a model with correctly specified observation error made marginal improvements to model bias (purple symbols in Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}).
 
-When converting the posterior distributions of $\nu$ into Pr($\nu < 10$), the
-models distinguished heavy and not-heavy tails reasonably well. Without
-observation error, and using a probability of $0.5$ as a threshold, the model
-correctly classified all simulated systems with normally distributed process
-noise as not heavy tailed. The model would have miscategorized only one of $40$
-simulations at $\nu = 5$ across a sample size of $50$ or $100$ data points
-(Fig.~\ref{fig:sim-prob}, upper panels). The model would have correctly
-categorized all cases where the process noise was not heavy tailed (indicated
-as ``$\nu =$ infinity'' in Fig.~\ref{fig:sim-prob}) and all cases where $\nu
-= 3$. With $0.2$ standard deviations of observation error, the model still
-categorized \obsErrorNuFivePerc\% of cases as heavy tailed when $\nu = 5$ and
-all cases where $\nu = 3$. Allowing for observation error made little
-improvement to the detection of heavy tails (Fig.~\ref{fig:sim-prob},
-lower-left vs.\ lower-right panel). Therefore, we chose to focus on the simpler
-model without observation error in the main text, particularly given that the
-true magnitude of observation error was unknown in the empirical data.
+When converting the posterior distributions of $\nu$ into Pr($\nu < 10$), the models distinguished heavy and not-heavy tails reasonably well.
+Without observation error, and using a probability of $0.5$ as a threshold, the model correctly classified all simulated systems with normally distributed process noise as not heavy tailed.
+The model would have miscategorized only one of $40$ simulations at $\nu = 5$ across a sample size of $50$ or $100$ data points (Fig.~\ref{fig:sim-prob}, upper panels).
+The model would have correctly categorized all cases where the process noise was not heavy tailed (indicated as ``$\nu =$ infinity'' in Fig.~\ref{fig:sim-prob}) and all cases where $\nu = 3$.
+With $0.2$ standard deviations of observation error, the model still categorized \obsErrorNuFivePerc\% of cases as heavy tailed when $\nu = 5$ and all cases where $\nu = 3$.
+Allowing for observation error made little improvement to the detection of heavy tails (Fig.~\ref{fig:sim-prob}, lower-left vs.\ lower-right panel).
+Therefore, we chose to focus on the simpler model without observation error in the main text, particularly given that the true magnitude of observation error was unknown in the empirical data.
 
@@
 
-We fitted a multilevel beta regression model to the predicted probability of
-heavy tails, Pr($\nu < 10$), to investigate potential covariates of heavy-tailed
-dynamics. The beta distribution is useful when response data range on
-a continuous scale between zero and one. We used a logit link function as is
-typically used in logistic regression. The model was as follows:
+We fitted a multilevel beta regression model to the predicted probability of heavy tails, Pr($\nu < 10$), to investigate potential covariates of heavy-tailed dynamics.
+The beta distribution is useful when response data range on a continuous scale between zero and one.
+We used a logit link function as is typically used in logistic regression.
+The model was as follows:
 \begin{align*}
@@
 \end{align*}
-where $A$ and $B$ represent the beta distribution shape parameters; $\mu_i$
-represents the predicted value for population $i$, class $j$, order $k$, and
-species $l$; $\phi_\mathrm{disp}$ represents the dispersion parameter; and $X_i$
-represents a vector of predictors (such as lifespan) for population $i$ with
-associated $\beta$ coefficients. The intercepts are allowed to vary from the
-overall intercept $\alpha$ by taxonomic class ($\alpha^\mathrm{class}_j$),
-taxonomic order ($\alpha^\mathrm{order}_k$), and species
-($\alpha^\mathrm{species}_l$) with standard deviations $\sigma_{\alpha \;
-  \mathrm{class}}$, $\sigma_{\alpha \; \mathrm{order}}$, and $\sigma_{\alpha \;
-  \mathrm{species}}$. Where possible, we also allowed for error distributions
-around the predictors by incorporating the standard deviation of the posterior
-samples for the Gompertz parameters $\lambda$, $b$, and $\log \sigma$ around the
-mean point value as normal distributions (not shown in the above equation).
+where $A$ and $B$ represent the beta distribution shape parameters; $\mu_i$ represents the predicted value for population $i$, class $j$, order $k$, and species $l$; $\phi_\mathrm{disp}$ represents the dispersion parameter; and $X_i$ represents a vector of predictors (such as lifespan) for population $i$ with associated $\beta$ coefficients.
+The intercepts are allowed to vary from the overall intercept $\alpha$ by taxonomic class ($\alpha^\mathrm{class}_j$), taxonomic order ($\alpha^\mathrm{order}_k$), and species ($\alpha^\mathrm{species}_l$) with standard deviations $\sigma_{\alpha \; \mathrm{class}}$, $\sigma_{\alpha \; \mathrm{order}}$, and $\sigma_{\alpha \; \mathrm{species}}$.
+Where possible, we also allowed for error distributions around the predictors by incorporating the standard deviation of the posterior samples for the Gompertz parameters $\lambda$, $b$, and $\log \sigma$ around the mean point value as normal distributions (not shown in the above equation).
 
-We log transformed $\sigma$, time-series length, and lifespan to match the way
-they are visually represented in Fig.~\ref{fig:correlates} and to make the
-relationship approximately linear on the logit-transformed response scale. All
-input variables were standardized by subtracting their mean and dividing by two
-standard deviations to make their coefficients comparable in magnitude
-\citep{gelman2008c}. We excluded body length as a covariate because it was
-highly correlated with lifespan, and lifespan exhibited more overlap across
-taxonomy than body length. Lifespan is also more directly related to time and
-potential mechanisms driving black-swan dynamics.
+We log transformed $\sigma$, time-series length, and lifespan to match the way they are visually represented in Fig.~\ref{fig:correlates} and to make the relationship approximately linear on the logit-transformed response scale.
+All input variables were standardized by subtracting their mean and dividing by two standard deviations to make their coefficients comparable in magnitude \citep{gelman2008c}.
+We excluded body length as a covariate because it was highly correlated with lifespan, and lifespan exhibited more overlap across taxonomy than body length.
+Lifespan is also more directly related to time and potential mechanisms driving black-swan dynamics.
 
 We incorporated weakly informative priors into our model: $\mathrm{Cauchy}(0,
-10)$ on the global intercept $\alpha$, $\mathrm{Half\mhyphen Cauchy}(0, 2.5)$ on
-all standard deviation parameters, $\mathrm{Half\mhyphen Cauchy}(0, 10)$ on the
-dispersion parameter $\phi_\mathrm{disp}$, and $\mathrm{Cauchy}(0, 2.5)$ on all
-other parameters \citep{gelman2006c, gelman2008d}. Compared to normal priors,
-the Cauchy priors concentrate more probability density around expected parameter
-values while allowing for a higher probability density far into the tails, thereby
-allowing the data to dominate the posterior more strongly if it disagrees with
-the prior. Our conclusions were not qualitatively changed by using uniform
-priors. We fitted our models with 5000 total iterations per chain, 2500 warm-up
-iterations, four chains, and discarding every second sample to save memory. We
-checked for chain convergence visually and with the same criteria as before
-($\widehat{R} < 1.05$ and $n_\mathrm{eff} >200$ for all parameters).
+10)$ on the global intercept $\alpha$, $\mathrm{Half\mhyphen Cauchy}(0, 2.5)$ on all standard deviation parameters, $\mathrm{Half\mhyphen Cauchy}(0, 10)$ on the dispersion parameter $\phi_\mathrm{disp}$, and $\mathrm{Cauchy}(0, 2.5)$ on all other parameters \citep{gelman2006c, gelman2008d}.
+Compared to normal priors, the Cauchy priors concentrate more probability density around expected parameter values while allowing for a higher probability density far into the tails, thereby allowing the data to dominate the posterior more strongly if it disagrees with the prior.
+Our conclusions were not qualitatively changed by using uniform priors.
+We fitted our models with 5000 total iterations per chain, 2500 warm-up iterations, four chains, and discarding every second sample to save memory.
+We checked for chain convergence visually and with the same criteria as before ($\widehat{R} < 1.05$ and $n_\mathrm{eff} >200$ for all parameters).
 
-To derive taxonomic-order-level estimates of the probability of heavy tails
-accounting for time-series length (Fig \ref{fig:order-estimates}), we fitted
-a separate multilevel model with the same structure but with only $\log$
-time-series length as a predictor. (In this case, we did not want to control
-for intrinsic population characteristics such as density dependence.) Since our
-predictors were centered by subtracting their mean value, we obtained
-order-level estimates of the probability of heavy tails at mean log time-series
-length by adding the posteriors for $\alpha$, $\alpha^\mathrm{class}_j$, and
-$\alpha^\mathrm{order}_k$.
+To derive taxonomic-order-level estimates of the probability of heavy tails accounting for time-series length (Fig \ref{fig:order-estimates}), we fitted a separate multilevel model with the same structure but with only $\log$ time-series length as a predictor.
+(In this case, we did not want to control for intrinsic population characteristics such as density dependence.) Since our predictors were centered by subtracting their mean value, we obtained order-level estimates of the probability of heavy tails at mean log time-series length by adding the posteriors for $\alpha$, $\alpha^\mathrm{class}_j$, and $\alpha^\mathrm{order}_k$.
 

### 2014-11-06 text/supp.tex
Add TODO to SOM fig

@@
 TODO WILL COMBINE WITH COEFFICIENT POSTERIORS IN MAIN TEXT FIG 4
+TODO note that this isn't all orders!
   Taxonomic-order-level posterior estimates of probability that $\nu < 10$ from

### 2014-11-06 text/ms.tex
Incorporate new fig 4 into main text

@@
 \end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=0.45\textwidth]{../analysis/order-posteriors-covariates.pdf}
+\caption{
+Posterior probability distributions from beta regression multilevel models.
+(a) Main effect posterior densities for potential covariates of heavy-tailed dynamics.
+The response variable was Pr($\nu < 10$) (approximately the probability of heavy tails).
+The beta regression model was fit on a logit scale with hierarchical intercepts for taxonomic class, taxonomic order, and species.
+All covariates were standardized by subtracting their mean and dividing by twice their standard deviation.
+(b) Taxonomic-order-level posterior densities of Pr($\nu < 10$) after accounting for time-series length.
+Estimates are at the geometric mean of time series length across all the data (approximately 27 time steps).
+Colour shading refers to taxonomic class (yellow: fishes, green: insects, purple: birds, and red: mammals).
+Short vertical line segments indicate median posterior estimates.
+Long dotted vertical line in panel b indicates the median expected Pr($\nu < 10$) from the prior distribution.
+}
+\label{fig:posteriors}
+\end{center}
+\end{figure}

### 2014-11-06 text/supp.tex
Update priors captions in SOM

--- 2014-11-04 text/supp.tex
@@
 \begin{center}
-\includegraphics[width=\textwidth]{../analysis/priors-gomp-base.pdf}
+\includegraphics[width=0.8\textwidth]{../analysis/priors-gomp-base.pdf}
 \caption{
-TODO WILL ADD ALTERNATIVE $\nu$ PRIORS
-  Probability density of the Bayesian priors for the Gompertz models. From left
-  to right: (1) per capita growth rate at $\log$(abundance) = $0$: $\lambda \sim
-  \mathrm{Normal}(0, 10^2)$; (2) scale parameter of t-distribution process
-  noise: $\sigma \sim \mathrm{Half\mhyphen Cauchy} (0,
-  2.5)$; (3) t-distribution degrees of freedom parameter: $\nu \sim
-  \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2)$; (4) AR1
-  correlation coefficient of residuals: $\phi \sim \mathrm{Truncated \mhyphen
-    Normal}(0, 1, \mathrm{min.} = -1, \mathrm{max.} = 1)$. Not shown is $b$,
-  the density dependence parameter: $b \sim \mathrm{Uniform}(-1, 2)$.
+  Probability density of the Bayesian priors for the Gompertz models.
+From left to right and then top to bottom: (1) per capita growth rate at $\log$(abundance) = $0$: $\lambda \sim \mathrm{Normal}(0, 10^2)$; (2) scale parameter of t-distribution process noise: $\sigma \sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)$; (3) t-distribution degrees of freedom parameter: $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2)$; (4) AR1 correlation coefficient of residuals: $\phi \sim \mathrm{Truncated \mhyphen Normal}(0, 1, \mathrm{min.} = -1, \mathrm{max.} = 1)$.
+Not shown is $b$, the density dependence parameter: $b \sim \mathrm{Uniform}(-1, 2)$.
+The $\nu$ panel also shows two alternative priors: a weaker prior $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.005, \mathrm{min.} = 2)$, and a stronger prior $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.02, \mathrm{min.} = 2)$.
 }
@@
 \label{fig:heavy-ts}
-\end{center}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=0.27\textwidth]{../analysis/order-level-estimates.pdf}
-\caption{
-TODO WILL COMBINE WITH COEFFICIENT POSTERIORS IN MAIN TEXT FIG 4
-  Taxonomic-order-level posterior estimates of probability that $\nu < 10$ from
-  a beta regression multilevel model accounting for the effect of time-series
-  length. Estimates are at the geometric mean of time series length across all
-  the data (approximately 27 time steps). Colour shading refers to the
-  taxonomic classes as indicated in Fig.~\ref{fig:nu-coefs} and
-  Fig.~\ref{fig:correlates} (yellow for fishes, green for insects, purple for
-  birds, and red for mammals). Short vertical line segments indicate median
-  posterior estimates. Long dotted vertical line throughout the figure
-  indicates the median expected Pr($\nu < 10$) from the prior distribution.
-}
-\label{fig:order-estimates}
 \end{center}
@@
 
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=0.5\textwidth]{../analysis/stan-beta-correlates}
-\caption{
-TODO WILL MOVE TO MAIN TEXT WITH ORDER-LEVEL ESTIMATES
-  Main effect posterior densities for potential correlates of heavy-tailed
-  dynamics. The response variable was Pr($\nu < 10$) (approximately the
-  probability of heavy tails). The beta regression model was fit on a logit
-  scale with random intercepts for taxonomic class, taxonomic order, and
-  species. All covariates were standardized by subtracting their mean and
-  dividing by twice their standard deviation. Therefore, the coefficient values
-  represent the expected effect of a change in two standard deviations of the
-  raw input variable on the logit-transformed response variable.
-}
-    \label{fig:correlate-coefs}
-\end{center}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=0.70\textwidth]{../analysis/perc-prob-increase-with-n}
-\caption{
-TODO WILL DELETE THIS
-Expected percent increase in the probability of detecting heavy tails, Pr($\nu
-< 10$), with an additional ten time steps of data. These values are based on
-the curve shown in Fig.~\ref{fig:correlates}d, which is based on the
-coefficient in Fig.~\ref{fig:correlate-coefs}. For example, we have about
-a 12\% increased probability of observing heavy tails in a time series that has
-60 time steps vs.\ a time series with 50 time steps.
-}
-\label{fig:perc-inc-p}
-\end{center}
-\end{figure}
-
-\clearpage
-
 \noindent

### 2014-11-06 text/supp.tex
Add alternative prior plot to SOM

@@
 \begin{center}
+\includegraphics[width=\textwidth]{../analysis/gomp-prior-comparison.pdf}
+\caption{
+Estimates of $\nu$ from Gompertz models with alternative priors on $\nu$.
+Shown are medians of the posterior (dots) and 50\% credible intervals (segments).
+The diagonal line indicates a one-to-one relationship.
+Different colours indicate various taxonomic classes.
+The grey-shaded regions indicate regions of disagreement if $\nu = 10$ is taken as a threshold of heavy-tailed dynamics.
+The base, weaker, and stronger priors on $\nu$ are illustrated in Fig.~\ref{fig:priors}.
+In general, the estimates are nearly identical in cases where the data are informative about low values of $\nu$.
+When the data are less informative about low values of $\nu$, the prior can slightly pull the estimates of $\nu$ towards higher or lower values.
+}
+\label{fig:alt}
+\end{center}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[htbp]
+\begin{center}
 \includegraphics[width=0.8\textwidth]{../analysis/t-dist-sampling-sim-prior-exp0point01.pdf}
@@
 
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=\textwidth]{../analysis/sim-gompertz.pdf}
-\caption{
-TODO WILL REMOVE ALL OR MOST OF THIS
-  Simulation testing the Gompertz estimation model when the process deviation
-  draws were chosen so that $\nu$ could be estimated close to the true value
-  outside the full population model (``effective $\nu$'' within a CV of 0.2 of
-  specified $\nu$). The simulation was run across population $\nu$ values
-  (columns) and different scenarios (colours): (1) 100 time steps and no
-  observation error, (2) 50 time steps and no observation error, (3) 50 time
-  steps and observation error drawn from $\mathrm{Normal} (0, 0.2^2)$ but
-  ignored, and (4) 50 time steps with observation error in which the quantity
-  of observation error was assumed known. Dashed horizontal lines show the true
-  population values; these true values were chosen to represent approximately
-  the median values as estimated from the Global Population Dynamics Database.
-  Individual dots and lines
-  represent a stochastic draws from the true population distribution and
-  a model fitting (median and 80\% credible intervals). The panels from top to
-  bottom show $1/\nu$, process noise scale parameter $\sigma$, growth rate
-  parameter $\lambda$, and the density dependence parameter $b$.
-}
-\label{fig:sim-gompertz}
-\end{center}
-\end{figure}
+% \begin{figure}[htbp]
+% \begin{center}
+% \includegraphics[width=\textwidth]{../analysis/sim-gompertz.pdf}
+% \caption{
+% TODO WILL REMOVE ALL OR MOST OF THIS
+%   Simulation testing the Gompertz estimation model when the process deviation
+%   draws were chosen so that $\nu$ could be estimated close to the true value
+%   outside the full population model (``effective $\nu$'' within a CV of 0.2 of
+%   specified $\nu$). The simulation was run across population $\nu$ values
+%   (columns) and different scenarios (colours): (1) 100 time steps and no
+%   observation error, (2) 50 time steps and no observation error, (3) 50 time
+%   steps and observation error drawn from $\mathrm{Normal} (0, 0.2^2)$ but
+%   ignored, and (4) 50 time steps with observation error in which the quantity
+%   of observation error was assumed known. Dashed horizontal lines show the true
+%   population values; these true values were chosen to represent approximately
+%   the median values as estimated from the Global Population Dynamics Database.
+%   Individual dots and lines
+%   represent a stochastic draws from the true population distribution and
+%   a model fitting (median and 80\% credible intervals). The panels from top to
+%   bottom show $1/\nu$, process noise scale parameter $\sigma$, growth rate
+%   parameter $\lambda$, and the density dependence parameter $b$.
+% }
+% \label{fig:sim-gompertz}
+% \end{center}
+% \end{figure}
 
@@
 
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=\textwidth]{../analysis/sim-gompertz-boxplots.pdf}
-\caption{
-TODO WILL REMOVE ALL OR MOST OF THIS
-  The distribution of median estimates from the same simulation as described in
-  Fig.~\ref{fig:sim-gompertz}. The boxes show the interquartile range. The
-  whiskers extend to $1.5$ times the interquartile range and outliers are shown
-  as dots. Note that these distributions represent the output from only 20
-  simulation iterations, so there is some level of remaining sampling noise.
-}
-\label{fig:sim-gompertz-boxplots}
-\end{center}
-\end{figure}
+% \begin{figure}[htbp]
+% \begin{center}
+% \includegraphics[width=\textwidth]{../analysis/sim-gompertz-boxplots.pdf}
+% \caption{
+% TODO WILL REMOVE ALL OR MOST OF THIS
+%   The distribution of median estimates from the same simulation as described in
+%   Fig.~\ref{fig:sim-gompertz}. The boxes show the interquartile range. The
+%   whiskers extend to $1.5$ times the interquartile range and outliers are shown
+%   as dots. Note that these distributions represent the output from only 20
+%   simulation iterations, so there is some level of remaining sampling noise.
+% }
+% \label{fig:sim-gompertz-boxplots}
+% \end{center}
+% \end{figure}
 

### 2014-11-07 text/supp.tex
Catch up on edits to the text
Bringing in figure changes and captions

@@
 \setcounter{table}{0}
-TODO ``Chondrichtyhes'' IS A TYPO IN THE GPDD DATA\\
 TODO ADD TOTAL NUMBER OF POINTS ON RIGHT COLUMN
@@
 The $\nu$ panel also shows two alternative priors: a weaker prior $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.005, \mathrm{min.} = 2)$, and a stronger prior $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.02, \mathrm{min.} = 2)$.
+The inset panel shows the same data but with a log-transformed x axis.
 }
@@
 % \caption{
-% TODO WILL REMOVE ALL OR MOST OF THIS
 % Simulation testing the Gompertz estimation model when the process deviation
@@
 % \caption{
-% TODO WILL REMOVE ALL OR MOST OF THIS
 %   The distribution of median estimates from the same simulation as described in
@@
 \caption{
-TODO WILL REMOVE ALL OR MOST OF THIS
-  The probability that $\nu < 10$ (i.e.\ the approximate probability of heavy
-  tails) for the same simulation scenarios as shown in
-  Fig.~\ref{fig:sim-gompertz} and Fig.~\ref{fig:sim-gompertz-boxplots}. Within
+Simulation testing the Gompertz estimation model when the process deviation
+draws were chosen so that $\nu$ could be estimated close to the true value
+outside the full population model (``effective $\nu$'' within a CV of 0.2 of
+specified $\nu$).
+Plots show the probability that $\nu < 10$ (i.e.\ the approximate probability
+of heavy tails) across three population $\nu$ values
+(3, 5, and normal) and different scenarios panels: (1) 100 time steps and no
+observation error, (2) 50 time steps and no observation error, (3) 50 time
+steps and observation error drawn from $\mathrm{Normal} (0, 0.2^2)$ but
+ignored, and (4) 50 time steps with observation error in which the quantity
+of observation error was assumed known.
+Within
 each scenario the dots represent stochastic draws from the true population

### 2014-11-07 text/supp.tex
Split massive ts plot across pages (#15)

@@
 
+% \begin{figure}[htbp]
+% \begin{center}
+% \includegraphics[width=\textwidth]{../analysis/all-clean-ts-3.pdf}
+% \caption{
+% TODO WILL SPLIT OVER PAGES BY CLASS
+%   All filtered times series used in our analysis. The abundances are shown on
+%   a log10 vertical axis. Colours indicate taxonomic classes. Open black circles
+%   indicate abundance values that were interpolated. Closed black circles
+%   indicate single abundance values that were recorded as zero but were set to
+%   the next lowest observed abundance.
+% }
+%     \label{fig:all-ts}
+% \end{center}
+% \end{figure}
+
+\begin{centering}
+\clearpage
+\includegraphics[width=\textwidth]{../analysis/all-clean-ts-mammals.pdf}\\
+Figure~\ref{fig:all-ts} (mammals) continued on next page \ldots
+
+\clearpage
+\includegraphics[width=\textwidth]{../analysis/all-clean-ts-birds.pdf}\\
+Figure~\ref{fig:all-ts} (birds) continued on next page \ldots
+
+\clearpage
+\includegraphics[width=\textwidth]{../analysis/all-clean-ts-insects.pdf}\\
+Figure~\ref{fig:all-ts} (insects) continued on next page \ldots
+
+\end{centering}
+
 \begin{figure}[htbp]
 \begin{center}
-\includegraphics[width=\textwidth]{../analysis/all-clean-ts-3.pdf}
+\includegraphics[width=\textwidth]{../analysis/all-clean-ts-fishes-others.pdf}
 \caption{
-TODO WILL SPLIT OVER PAGES BY CLASS
-  All filtered times series used in our analysis. The abundances are shown on
-  a log10 vertical axis. Colours indicate taxonomic classes. Open black circles
-  indicate abundance values that were interpolated. Closed black circles
-  indicate single abundance values that were recorded as zero but were set to
-  the next lowest observed abundance.
+  (fishes, crustaceans, gastropods, sharks).
+All filtered time series used in our analysis.
+The abundances are shown on a log10 vertical axis.
+Throughout this figure, red dots indicate values that were interpolated and red circles indicate values that were recorded as zero but were set to the next lowest observed abundance.
+Numbers before each species name are the GPDD main identification numbers.
 }

### 2014-11-07 analysis/0-make-data.R

@@ -296,20 +296,36 @@ saveRDS(gpdd, file = "gpdd-clean.rds")
 
 # massive ts plot:
 library(ggplot2)
-p <- ggplot(gpdd, aes(series_step, log10(population_untransformed),
-  colour = taxonomic_class)) +
-  geom_point() + geom_line() +
+
+make_big_ts_plot <- function(dat_, id, width, height) {
+  p <- ggplot(dat_, aes(series_step, log10(population_untransformed))) +
+      #colour = taxonomic_class)) +
+  geom_point(pch = 20, col = "grey70") + geom_line() +
   facet_wrap(~label, scales = "free") +
-  geom_point(data = subset(gpdd, interpolated == TRUE),
-    aes(series_step, log10(population_untransformed)), colour = "black", pch = 21) +
-  geom_point(data = subset(gpdd, zero_sub == TRUE),
-    aes(series_step, log10(population_untransformed)), colour = "black", pch = 20) +
+  geom_point(data = subset(dat_, interpolated == TRUE),
+    aes(series_step, log10(population_untransformed)), colour = "red", pch = 20, cex = 3.5) +
+  geom_point(data = subset(dat_, zero_sub == TRUE),
+    aes(series_step, log10(population_untransformed)), colour = "blue", pch = 20, cex = 3.5) +
   theme(panel.grid.major = element_blank(),
     panel.grid.minor = element_blank(),
     panel.background = element_blank(),
     strip.background = element_blank()) +
   xlab("Time series step") + ylab("log10(Abundance)")
-ggsave("all-clean-ts-3.pdf", width = 50, height = 50, limitsize = FALSE)
+  ggsave(paste0("all-clean-ts-", id, ".pdf"), width = width, height = height,
+    limitsize = FALSE)
+}
+make_big_ts_plot(gpdd[gpdd$taxonomic_class == "Insecta", ], id = "insects",
+  width = 24, height = 20)
+make_big_ts_plot(gpdd[gpdd$taxonomic_class == "Mammalia", ], id = "mammals",
+  width = 24, height = 20)
+make_big_ts_plot(gpdd[gpdd$taxonomic_class == "Aves", ], id = "birds",
+  width = 24, height = 20)
+make_big_ts_plot(gpdd[gpdd$taxonomic_class %in%
+  c("Gastropoda", "Crustacea", "Chondrichtyhes", "Osteichthyes"), ],
[…]

### 2014-11-07 analysis/6-plot-eg-ts-gpdd.R
Tweak style of heavy ts plot

@@ -8,7 +8,7 @@ heavy <- filter(gomp_hat_base, p10 > .5) %>%
 
 
 heavy <-
-  plyr::join(heavy, gpdd) %>%
+  plyr::join(heavy, as.data.frame(gpdd)) %>%
   arrange(main_id, p10) %>%
   mutate(id_name = paste(round(p10, 2), common_name, main_id))
 
@@ -18,7 +18,11 @@ theme_set(theme_bw())
 p <- ggplot(heavy, aes(series_step, population_untransformed, colour = taxonomic_class)) + geom_line() + facet_wrap(~id_name, scales = "free")
 ggsave("ts-gpdd-heavy-eg-base.pdf", width = 15, height = 12)
 
-p <- ggplot(heavy, aes(series_step, log10(population_untransformed), colour = taxonomic_class)) + geom_line() + facet_wrap(~id_name, scales = "free", ncol = 4) + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + geom_point()
+p <- ggplot(heavy, aes(decimal_year_begin, log10(population_untransformed), colour = taxonomic_class)) + geom_line() + facet_wrap(~id_name, scales = "free", ncol = 4) + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + geom_point() +
+  theme(panel.grid.major = element_blank(),
+    panel.grid.minor = element_blank(),
+    panel.background = element_blank(),
+    strip.background = element_blank())
 ggsave("ts-gpdd-heavy-eg-log10-base.pdf", width = 15, height = 17)
 
 heavy_table <- plyr::ddply(heavy, "main_id", function(x) x[1,]) %>%

### 2014-11-07 text/supp.tex
Tweak style of heavy ts plot
Remove a todo

@@
 \setcounter{table}{0}
-TODO ADD TOTAL NUMBER OF POINTS ON RIGHT COLUMN
 \input{../analysis/stat-table}
@@
 \caption{
-TODO WILL INDICATE UPS AND DOWNS AND MATCH STYLE TO FIG S1
   Time series for populations with Pr($\nu < 10$) $>$ 0.5 using the base

### 2014-11-07 text/ms.tex
Catch up on edits to the text
Bringing in figure changes and captions
Switch fitted to fit

@@
 
-To obtain estimates of the probability and magnitude of black swans, we fitted population dynamic models to abundance time series from around the world.
+To obtain estimates of the probability and magnitude of black swans, we fit population dynamic models to abundance time series from around the world.
 For each population, we estimated the shape of the process noise tails by measuring the degrees of freedom ($\nu$) of a Student t-distribution.
@@
 Usually, the process noise $\epsilon_t$ is modelled as normally distributed, but in our paper we assume it is drawn from a t distribution with scale parameter $\sigma$ and degrees of freedom $\nu$.
-In previous analyses of the GPDD, the Gompertz is most often identified as the most parsimonious population model that can be fitted to these data \citep{brook2006}.
+In previous analyses of the GPDD, the Gompertz is most often identified as the most parsimonious population model that can be fit to these data \citep{brook2006}.
 
@@
 
-We fitted all models in a Bayesian framework using Stan \citep{stan-manual2014} via the R computing environment \citep{r2014}.
+We fit all models in a Bayesian framework using Stan \citep{stan-manual2014} via the R computing environment \citep{r2014}.
 Stan samples from the posterior distribution with an adaptive version of Hamiltonian Markov chain Monte Carlo called the No-U-Turn Sampler and generally obtains less correlated samples than algorithms such as the Gibbs sampler \citep{hoffman2014}.
@@
 
-% TODO fitted vs. fit
-We fitted alternative population models to test if four key phenomena systematically changed our conclusions.
-Autocorrelation has been suggested as a reason for increased observed variability of abundance time series through time, which could create apparent heavy tails \citep{inchausti2002}; therefore, we fitted a model that included serial correlation in the residuals.
-Additionally, previous work has modelled abundance or growth rates without accounting for density dependence \citep{halley2002,segura2013}; therefore, we fitted a simpler model in which we assumed density independence.
+We fit alternative population models to test if four key phenomena systematically changed our conclusions.
+Autocorrelation has been suggested as a reason for increased observed variability of abundance time series through time, which could create apparent heavy tails \citep{inchausti2002}; therefore, we fit a model that included serial correlation in the residuals.
+Additionally, previous work has modelled abundance or growth rates without accounting for density dependence \citep{halley2002,segura2013}; therefore, we fit a simpler model in which we assumed density independence.
 Third, observation error could bias parameter estimates \citep{knape2012} or mask our ability to detect heavy tails \citep{ward2007}; therefore, we fit a model where we allowed for a fixed quantity of observation error ($0.2$ standard deviations on a log scale).
@@
 
+\clearpage
+
 \begin{figure}[htbp]
 \begin{center}
-\includegraphics[width=\textwidth]{../analysis/t-nu-eg.pdf}
+\includegraphics[width=\textwidth]{../analysis/t-nu-eg2.pdf}
 \caption{

### 2014-11-07 text/supp.tex
Switch fitted to fit

@@
 \item We filled in all missing time steps with \texttt{NA} values and imputed single missing values with the geometric mean of the previous and following values.
-We chose a geometric mean to be linear on the log scale that the Gompertz and Ricker-logistic models were fitted on.
+We chose a geometric mean to be linear on the log scale that the Gompertz and Ricker-logistic models were fit on.
 
@@
 This implies that multiple zero values were either censusing errors or caused by emigration.
-Regardless, our population models were fitted on a multiplicative (log) scale and so could not account for zero abundance.
+Regardless, our population models were fit on a multiplicative (log) scale and so could not account for zero abundance.
 To avoid distorting the original data too strongly, we removed populations in which we filled in more than four zeros.
@@
 
-We fitted our models with Stan 2.4.0 \citep{stan-manual2014}, and R 3.1.1 \citep{r2014}.
+We fit our models with Stan 2.4.0 \citep{stan-manual2014}, and R 3.1.1 \citep{r2014}.
 We began with four chains and $2000$ iterations, discarding the first $1000$ as warm up (i.e.~4000 total samples).
@@
 We considered a version of the Gompertz model in which an autoregressive
-parameter was fitted to the process noise residuals:
+parameter was fit to the process noise residuals:
 \begin{align*}
@@
 
-We also fitted a Ricker-logistic model:
+We also fit a Ricker-logistic model:
 \begin{align*}
@@
 We took stochastic draws from t distributions with different $\nu$ values ($\nu = 3, 5, 10,$ and $10^6$ [$\approx$ normal]), with central tendency parameters of $0$, and scale parameters of $1$.
-We started with $1600$ stochastic draws and then fitted the models again at the first $800, 400, 200, 100, 50,$ and $25$ draws.
+We started with $1600$ stochastic draws and then fit the models again at the first $800, 400, 200, 100, 50,$ and $25$ draws.
 Each time we recorded the posterior samples of $\nu$.
@@
 
-We then fitted our Gompertz models to the simulated datasets with all parameters (except $\nu$) set near the median values estimated in the GPDD.
+We then fit our Gompertz models to the simulated datasets with all parameters (except $\nu$) set near the median values estimated in the GPDD.
 We repeated this with $50$ and $100$ samples without observation error, $50$ samples with observation error ($\sigma_\mathrm{obs} = 0.2$), and $50$ samples with the same observation error and a Gompertz model that allowed for correctly specified observation error magnitude.
@@
 
-We fitted a multilevel beta regression model to the predicted probability of heavy tails, Pr($\nu < 10$), to investigate potential covariates of heavy-tailed dynamics.
+We fit a multilevel beta regression model to the predicted probability of heavy tails, Pr($\nu < 10$), to investigate potential covariates of heavy-tailed dynamics.
 The beta distribution is useful when response data range on a continuous scale between zero and one.
@@
 Our conclusions were not qualitatively changed by using uniform priors.
-We fitted our models with 5000 total iterations per chain, 2500 warm-up iterations, four chains, and discarding every second sample to save memory.
+We fit our models with 5000 total iterations per chain, 2500 warm-up iterations, four chains, and discarding every second sample to save memory.
 We checked for chain convergence visually and with the same criteria as before ($\widehat{R} < 1.05$ and $n_\mathrm{eff} >200$ for all parameters).
 
-To derive taxonomic-order-level estimates of the probability of heavy tails accounting for time-series length (Fig \ref{fig:order-estimates}), we fitted a separate multilevel model with the same structure but with only $\log$ time-series length as a predictor.
+To derive taxonomic-order-level estimates of the probability of heavy tails accounting for time-series length (Fig \ref{fig:order-estimates}), we fit a separate multilevel model with the same structure but with only $\log$ time-series length as a predictor.
 (In this case, we did not want to control for intrinsic population characteristics such as density dependence.) Since our predictors were centered by subtracting their mean value, we obtained order-level estimates of the probability of heavy tails at mean log time-series length by adding the posteriors for $\alpha$, $\alpha^\mathrm{class}_j$, and $\alpha^\mathrm{order}_k$.

### 2014-11-07 text/supp.tex
Update a data filtering rule in the supp.

@@
 
-\item Populations with negative abundance values were assumed to be log10 values and transformed by taking $10$ to the power of the recorded abundance.
-In many cases this was noted for the population, but not in all cases.
-We inspected each of these \totalAssumedLog~time series to make sure our assumption made sense.
+\item Populations with negative abundance values were removed.
+Of the populations that remained at the end of our other filtering rules, the remaining populations with negative abundances listed were all from time series that had been standardized by subtracting the mean and dividing by the standard deviation.
+We verified this by locating the original papers the datasets were extracted from: \citet{colebrook1978} for zooplankton and \citet{lindstrom1995} for grouse.
+Since the papers did not include the original mean time-series values we could not back transform these data points.
 

### 2014-11-08 analysis/cause-table.tex

% latex table generated in R 3.1.2 by xtable 1.7-4 package
% Fri Nov  7 16:47:38 2014
Time series & Population & ID & citation & Verified & Description & Pr($\nu < 10$ & $\widehat{\nu}$ \\ 
  \midrule
\includegraphics[width=1.7cm]{../analysis/sparks/6528.pdf} & Shag, \textit{Phalacrocorax aristotelis}, Farne Islands, Northumberland & 6528 & \citep{potts1980} & yes & red tide & 1.00 & 2.4 \\ 
  \includegraphics[width=1.7cm]{../analysis/sparks/7115.pdf} & South African fur seal, \textit{Arctocephalus pusillus}, South Africa & 7115 & \citep{shaughnessy1982} & yes? & harvesting and predation changes & 1.00 & 2.5 \\ 
  \includegraphics[width=1.7cm]{../analysis/sparks/10128.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Scotland - un-named area & 10128 & \citep{potts1984} & yes & environmental- and parisitic-caused cycles & 1.00 & 2.6 \\ 
  \includegraphics[width=1.7cm]{../analysis/sparks/9382.pdf} & Pine looper or Bordered white, \textit{Bupalus piniaria}, Kessock & 9382 & \citep{broekhuizen1993} & yes &  & 1.00 & 2.7 \\ 
  \includegraphics[width=1.7cm]{../analysis/sparks/10127.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Scotland - un-named area & 10127 & \citep{potts1984} & yes & environmental- and parisitic-caused cycles & 0.99 & 2.7 \\ 
  \includegraphics[width=1.7cm]{../analysis/sparks/10007.pdf} & Water vole, \textit{Arvicola terrestris}, Le Pont & 10007 & \citep{saucy1994} & yes & predator-environment cycle interactions & 1.00 & 2.7 \\ 
  \includegraphics[width=1.7cm]{../analysis/sparks/20579.pdf} & Grey heron, \textit{Ardea cinerea}, Southern Britain & 20579 & \citep{stafford1971} & yes? & severe winter & 0.98 & 3.0 \\ 
  \includegraphics[width=1.7cm]{../analysis/sparks/9655.pdf} & Flea beetle, \textit{Chaetocnoma concinna}, Finland & 9655 & \citep{markkula1965} &  &  & 0.99 & 3.0 \\ 
  \includegraphics[width=1.7cm]{../analysis/sparks/1235.pdf} & Wren, \textit{Troglodytes troglodytes}, Eastern Wood, Bookham Common & 1235 & \citep{newton1998} & yes? & severe winter & 0.98 & 3.0 \\ 
  \includegraphics[width=1.7cm]{../analysis/sparks/10113.pdf} & Willow grouse, \textit{Lagopus lagopus}, Northern England & 10113 & \citep{dobson1995} & yes? & parasites and predators & 0.99 & 3.1 \\ 
  \includegraphics[width=1.7cm]{../analysis/sparks/9667.pdf} & Gooseberry sawfly, \textit{Nemastus ribesii}, Finland & 9667 & \citep{markkula1965} &  &  & 0.97 & 3.2 \\ 
  \includegraphics[width=1.7cm]{../analysis/sparks/9679.pdf} & Unknown, \textit{Trioza apicalis}, Finland & 9679 & \citep{markkula1965} &  &  & 0.83 & 3.4 \\ 
  \includegraphics[width=1.7cm]{../analysis/sparks/20527.pdf} & Wandering albatross, \textit{Diomedea exulans}, Taiaroa & 20527 & \citep{robertson1998} & IN LIBRARY &  & 0.99 & 3.5 \\ 
  \includegraphics[width=1.7cm]{../analysis/sparks/10039.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Northern Scotland & 10039 & \citep{dobson1995} & yes & parasites and predators & 0.92 & 3.9 \\ 
  \includegraphics[width=1.7cm]{../analysis/sparks/10162.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Atholl Estate & 10162 & \citet{mackenzie1952} & yes & overcrowding and environemnt & 0.91 & 4.6 \\ 
  \includegraphics[width=1.7cm]{../analysis/sparks/9503.pdf} & Fisher or  Pekan, \textit{Martes pennanti}, Manitoba & 9503 & \citep{keith1963} & IN LIBRARY &  & 0.76 & 4.8 \\ 
  \includegraphics[width=1.7cm]{../analysis/sparks/7099.pdf} & European rabbit, \textit{Oryctolagus cuniculus}, Estate 2, East Anglia & 7099 & \citep{barnes1986} & yes & disease outbreak followed by years of good weather & 0.78 & 4.8 \\ 
  \includegraphics[width=1.7cm]{../analysis/sparks/2778.pdf} & Wheatear, \textit{Oenanthe oenanthe}, Skokholm Island & 2778 & \citep{lack1969} & yes & unknown, but decline noted specifically, cold winters later & 0.67 & 5.3 \\ 
  \includegraphics[width=1.7cm]{../analysis/sparks/9659.pdf} & Cabbage root fly or maggot, \textit{Delia radicum}, Finland & 9659 & \citep{markkula1965} &  &  & 0.59 & 6.8 \\ 
  \includegraphics[width=1.7cm]{../analysis/sparks/1195.pdf} & Blue jay, \textit{Cyanocitta cristata}, Robert Allerton Park & 1195 & \citep{kendeigh1982} & yes & unknown & 0.60 & 6.9 \\ 
  \includegraphics[width=1.7cm]{../analysis/sparks/5019.pdf} & Barbary macaque, \textit{Macaca sylvanus}, Queens Gate & 5019 & \citep{fa1984} &  &  & 0.61 & 6.9 \\ 
  \includegraphics[width=1.7cm]{../analysis/sparks/9953.pdf} & Rock ptarmigan, \textit{Lagopus mutus}, Iceland & 9953 & \citep{clarke1885,williams1954} &  & severe winters & 0.58 & 7.2 \\ 
  \includegraphics[width=1.7cm]{../analysis/sparks/6548.pdf} & Lesser-spotted dogfish, \textit{Scyliorhinus caniculus}, North Sea & 6548 & \citep{heessen1996} & yes & unknown & 0.53 & 8.5 \\ 
  \includegraphics[width=1.7cm]{../analysis/sparks/20546.pdf} & American red fox, \textit{Vulpes fulva}, Labrador & 20546 & \citep{dancona1954,lindstrom1994} & yes & predator-prey cycles & 0.57 & 8.7 \\ 
  \includegraphics[width=1.7cm]{../analysis/sparks/9470.pdf} & Ruffed grouse, \textit{Bonasa umbellus}, Connecticut & 9470 & \citep{keith1963} & IN LIBRARY &  & 0.53 & 8.9 \\ 
   \bottomrule

### 2014-11-08 text/supp.tex

@@
 
+\LTcapwidth=\textwidth
+\bibpunct{}{}{;}{a}{}{;}
+
+\singlespacing
+\begin{footnotesize}
+\begin{longtable}{>{\RaggedRight}m{1.5cm}>{\RaggedRight}p{4.3cm}>{\RaggedRight}p{0.8cm}>{\RaggedRight}p{1.7cm}>{\RaggedRight}p{1.0cm}>{\RaggedRight}p{3.0cm}>{\RaggedRight}p{1.7cm}>{\RaggedRight}p{0.8cm}}
+\caption{Populations with Pr$(\nu < 10) > 0.5$. TODO}\\
+\toprule
+\input{../analysis/cause-table.tex}
+\label{tab:causes-supp}
+\end{longtable}
+\end{footnotesize}
+\onehalfspacing
+
 % ------------------------------
@@
 
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=0.9\textwidth]{../analysis/ts-gpdd-heavy-eg-log10-base.pdf}
-\caption{
-  Time series for populations with Pr($\nu < 10$) $>$ 0.5 using the base
-  Gompertz population model. Panels are ordered by increasing Pr($\nu < 10$).
-  Vertical axes are on a log10 scale. Colours indicate taxonomic classes. The
-  labels on each panel indicate Pr($\nu < 10$) $>$ 0.5, the common name for the
-  species, and the Global Population Dynamics Database ID number.
-}
-\label{fig:heavy-ts}
-\end{center}
-\end{figure}
-
-\clearpage
+% \begin{figure}[htbp]
+% \begin{center}
+% \includegraphics[width=0.9\textwidth]{../analysis/ts-gpdd-heavy-eg-log10-base.pdf}
+% \caption{
+%   Time series for populations with Pr($\nu < 10$) $>$ 0.5 using the base
+%   Gompertz population model. Panels are ordered by increasing Pr($\nu < 10$).
+%   Vertical axes are on a log10 scale. Colours indicate taxonomic classes. The
+%   labels on each panel indicate Pr($\nu < 10$) $>$ 0.5, the common name for the
+%   species, and the Global Population Dynamics Database ID number.
+% }
+% \label{fig:heavy-ts}
+% \end{center}
+% \end{figure}
+%
+% \clearpage
 

### 2014-11-08 analysis/cause-table.tex
Recompile causes table

@@
 % latex table generated in R 3.1.2 by xtable 1.7-4 package
-% Fri Nov  7 16:47:38 2014
+% Fri Nov  7 19:07:39 2014
 Time series & Population & ID & citation & Verified & Description & Pr($\nu < 10$ & $\widehat{\nu}$ \\ 
   \midrule
-\includegraphics[width=1.7cm]{../analysis/sparks/6528.pdf} & Shag, \textit{Phalacrocorax aristotelis}, Farne Islands, Northumberland & 6528 & \citep{potts1980} & yes & red tide & 1.00 & 2.4 \\ 
+\includegraphics[width=1.7cm]{../analysis/sparks/6528.pdf} & Shag, \textit{Phalacrocorax aristotelis}, Farne Islands, Northumberland & 6528 & \citep{potts1980} & yes & red tide & 1.00 & 2 (2--4) \\ 
   \includegraphics[width=1.7cm]{../analysis/sparks/7115.pdf} & South African fur seal, \textit{Arctocephalus pusillus}, South Africa & 7115 & \citep{shaughnessy1982} & yes?
-& harvesting and predation changes & 1.00 & 2.5 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/10128.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Scotland - un-named area & 10128 & \citep{potts1984} & yes & environmental- and parisitic-caused cycles & 1.00 & 2.6 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/9382.pdf} & Pine looper or Bordered white, \textit{Bupalus piniaria}, Kessock & 9382 & \citep{broekhuizen1993} & yes &  & 1.00 & 2.7 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/10127.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Scotland - un-named area & 10127 & \citep{potts1984} & yes & environmental- and parisitic-caused cycles & 0.99 & 2.7 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/10007.pdf} & Water vole, \textit{Arvicola terrestris}, Le Pont & 10007 & \citep{saucy1994} & yes & predator-environment cycle interactions & 1.00 & 2.7 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/20579.pdf} & Grey heron, \textit{Ardea cinerea}, Southern Britain & 20579 & \citep{stafford1971} & yes? & severe winter & 0.98 & 3.0 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/9655.pdf} & Flea beetle, \textit{Chaetocnoma concinna}, Finland & 9655 & \citep{markkula1965} &  &  & 0.99 & 3.0 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/1235.pdf} & Wren, \textit{Troglodytes troglodytes}, Eastern Wood, Bookham Common & 1235 & \citep{newton1998} & yes? & severe winter & 0.98 & 3.0 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/10113.pdf} & Willow grouse, \textit{Lagopus lagopus}, Northern England & 10113 & \citep{dobson1995} & yes? & parasites and predators & 0.99 & 3.1 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/9667.pdf} & Gooseberry sawfly, \textit{Nemastus ribesii}, Finland & 9667 & \citep{markkula1965} &  &  & 0.97 & 3.2 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/9679.pdf} & Unknown, \textit{Trioza apicalis}, Finland & 9679 & \citep{markkula1965} &  &  & 0.83 & 3.4 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/20527.pdf} & Wandering albatross, \textit{Diomedea exulans}, Taiaroa & 20527 & \citep{robertson1998} & IN LIBRARY &  & 0.99 & 3.5 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/10039.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Northern Scotland & 10039 & \citep{dobson1995} & yes & parasites and predators & 0.92 & 3.9 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/10162.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Atholl Estate & 10162 & \citet{mackenzie1952} & yes & overcrowding and environemnt & 0.91 & 4.6 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/9503.pdf} & Fisher or  Pekan, \textit{Martes pennanti}, Manitoba & 9503 & \citep{keith1963} & IN LIBRARY &  & 0.76 & 4.8 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/7099.pdf} & European rabbit, \textit{Oryctolagus cuniculus}, Estate 2, East Anglia & 7099 & \citep{barnes1986} & yes & disease outbreak followed by years of good weather & 0.78 & 4.8 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/2778.pdf} & Wheatear, \textit{Oenanthe oenanthe}, Skokholm Island & 2778 & \citep{lack1969} & yes & unknown, but decline noted specifically, cold winters later & 0.67 & 5.3 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/9659.pdf} & Cabbage root fly or maggot, \textit{Delia radicum}, Finland & 9659 & \citep{markkula1965} &  &  & 0.59 & 6.8 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/1195.pdf} & Blue jay, \textit{Cyanocitta cristata}, Robert Allerton Park & 1195 & \citep{kendeigh1982} & yes & unknown & 0.60 & 6.9 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/5019.pdf} & Barbary macaque, \textit{Macaca sylvanus}, Queens Gate & 5019 & \citep{fa1984} &  &  & 0.61 & 6.9 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/9953.pdf} & Rock ptarmigan, \textit{Lagopus mutus}, Iceland & 9953 & \citep{clarke1885,williams1954} &  & severe winters & 0.58 & 7.2 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/6548.pdf} & Lesser-spotted dogfish, \textit{Scyliorhinus caniculus}, North Sea & 6548 & \citep{heessen1996} & yes & unknown & 0.53 & 8.5 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/20546.pdf} & American red fox, \textit{Vulpes fulva}, Labrador & 20546 & \citep{dancona1954,lindstrom1994} & yes & predator-prey cycles & 0.57 & 8.7 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/9470.pdf} & Ruffed grouse, \textit{Bonasa umbellus}, Connecticut & 9470 & \citep{keith1963} & IN LIBRARY &  & 0.53 & 8.9 \\ 
+& harvesting and predation changes & 1.00 & 2 (2--4) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/10128.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Scotland - un-named area & 10128 & \citep{potts1984} & yes & environmental- and parisitic-caused cycles & 1.00 & 3 (2--4) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/9382.pdf} & Pine looper or Bordered white, \textit{Bupalus piniaria}, Kessock & 9382 & \citep{broekhuizen1993} & yes &  & 1.00 & 3 (2--5) \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/10127.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Scotland - un-named area & 10127 & \citep{potts1984} & yes & environmental- and parisitic-caused cycles & 0.99 & 3 (2--5) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/10007.pdf} & Water vole, \textit{Arvicola terrestris}, Le Pont & 10007 & \citep{saucy1994} & yes & predator-environment cycle interactions & 1.00 & 3 (2--5) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/20579.pdf} & Grey heron, \textit{Ardea cinerea}, Southern Britain & 20579 & \citep{stafford1971} & yes? & severe winter & 0.98 & 3 (2--7) \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/9655.pdf} & Flea beetle, \textit{Chaetocnoma concinna}, Finland & 9655 & \citep{markkula1965} &  &  & 0.99 & 3 (2--6) \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/1235.pdf} & Wren, \textit{Troglodytes troglodytes}, Eastern Wood, Bookham Common & 1235 & \citep{newton1998} & yes?
+& severe winter & 0.98 & 3 (2--8) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/10113.pdf} & Willow grouse, \textit{Lagopus lagopus}, Northern England & 10113 & \citep{dobson1995} & yes?
+& parasites and predators & 0.99 & 3 (2--6) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/9667.pdf} & Gooseberry sawfly, \textit{Nemastus ribesii}, Finland & 9667 & \citep{markkula1965} &  &  & 0.97 & 3 (2--8) \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/9679.pdf} & Unknown, \textit{Trioza apicalis}, Finland & 9679 & \citep{markkula1965} &  &  & 0.83 & 3 (2--88) \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/20527.pdf} & Wandering albatross, \textit{Diomedea exulans}, Taiaroa & 20527 & \citep{robertson1998} & IN LIBRARY & IN LIBRARY & 0.99 & 4 (2--7) \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/10039.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Northern Scotland & 10039 & \citep{dobson1995} & yes & parasites and predators & 0.92 & 4 (2--12) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/10162.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Atholl Estate & 10162 & \citet{mackenzie1952} & yes & overcrowding and environemnt & 0.91 & 5 (2--12) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/9503.pdf} & Fisher or  Pekan, \textit{Martes pennanti}, Manitoba & 9503 & \citep{keith1963} & IN LIBRARY &  & 0.76 & 5 (2--87) \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/7099.pdf} & European rabbit, \textit{Oryctolagus cuniculus}, Estate 2, East Anglia & 7099 & \citep{barnes1986} & yes & disease outbreak followed by years of good weather & 0.78 & 5 (2--67) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/2778.pdf} & Wheatear, \textit{Oenanthe oenanthe}, Skokholm Island & 2778 & \citep{lack1969} & yes & unknown, but decline noted specifically, cold winters later & 0.67 & 5 (2--141) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/9659.pdf} & Cabbage root fly or maggot, \textit{Delia radicum}, Finland & 9659 & \citep{markkula1965} &  &  & 0.59 & 7 (2--161) \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/1195.pdf} & Blue jay, \textit{Cyanocitta cristata}, Robert Allerton Park & 1195 & \citep{kendeigh1982} & yes & unknown & 0.60 & 7 (2--149) \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/5019.pdf} & Barbary macaque, \textit{Macaca sylvanus}, Queens Gate & 5019 & \citep{fa1984} &  &  & 0.61 & 7 (2--132) \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/9953.pdf} & Rock ptarmigan, \textit{Lagopus mutus}, Iceland & 9953 & \citep{clarke1885,williams1954} &  & severe winters & 0.58 & 7 (2--172) \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/6548.pdf} & Lesser-spotted dogfish, \textit{Scyliorhinus caniculus}, North Sea & 6548 & \citep{heessen1996} & yes & unknown & 0.53 & 8 (2--190) \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/20546.pdf} & American red fox, \textit{Vulpes fulva}, Labrador & 20546 & \citep{dancona1954,lindstrom1994} & yes & predator-prey cycles & 0.57 & 9 (3--65) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/9470.pdf} & Ruffed grouse, \textit{Bonasa umbellus}, Connecticut & 9470 & \citep{keith1963} & IN LIBRARY & IN LIBRARY & 0.53 & 9 (3--158) \\ 
    \bottomrule

### 2014-11-08 text/ms.tex
Edit ms and supp to bring in new figs
A start

@@
 We find strong evidence of black swans, but they are rare occurring in \overallMinPerc--\overallMaxPerc\% of populations: most frequently for birds (\AvesRangePerc\%) followed by mammals (\MammaliaRangePerc\%), insects (\InsectaRangePerc\%), and fishes (\OsteichthyesRangePerc\%).
-Black swans are not explained by any life-history covariates, but tend to be driven by external perturbations such as climate, severe winters, predator or parasite cycles, and the combined effects of multiple factors .
+Black swans are not explained by any life-history covariates, but tend to be driven by external perturbations such as climate, severe winters, predator or parasite cycles, and the combined effects of multiple factors.
 Black swans were more frequently detected for populations with longer time series and lower levels of process noise; for shorter and noisier time series, our simulations suggested black-swan dynamics are often not identifiable as such.
@@
 In any case, for $\nu > 20$ the t distribution is almost indistinguishable from the normal distribution (Fig.~\ref{fig:didactic}).
-TODO ADD ALTERNATIVE PRIOR NOTE.
-
-TODO We used simulated data to test how easily we could detect $\nu$ given different sample sizes and to ensure we could recover unbiased parameter estimates from the Gompertz model (Supporting Information).
+We investigated the sensitivity of our results to weaker and stronger priors (Fig.~\ref{fig:priors}, Supporting Information).
+We used simulated data to test how easily we could detect $\nu$ given different sample sizes and to ensure we could recover unbiased parameter estimates from the Gompertz model (Supporting Information).
 
@@
 We found strong, but rare, evidence for black-swan population dynamics.
-By defining black-swan dynamics as a greater than $0.5$ probability that $\nu < 10$, our main Gompertz model found evidence for heavy tails most frequently for birds (\birdPH\%) followed by mammals (\mammalsPH\%), insects (\insectsPH\%), and fishes (\fishPH\%) (Figs~\ref{fig:nu-coefs}, \ref{fig:heavy-ts}).
+By defining black-swan dynamics as a greater than $0.5$ probability that $\nu < 10$, our main Gompertz model found evidence for heavy tails most frequently for birds (\birdPH\%) followed by mammals (\mammalsPH\%), insects (\insectsPH\%), and fishes (\fishPH\%) (Fig.~\ref{fig:nu-coefs}, Table~\ref{tab:causes-supp}).
 Black swans were taxonomically widespread, occurring in \POrdersHeavy\% of taxonomic orders.
-Accounting for time series length and partially pooling inference across taxonomic class and order with a multilevel model, there was stronger evidence for black swans in insect populations than is visually apparent in Fig.~\ref{fig:nu-coefs} --- four of 10 orders with the highest median probability of heavy tails were insect orders --- however, there was considerable uncertainty in these estimates (Fig.~\ref{fig:order-estimates}).
+Accounting for time series length and partially pooling inference across taxonomic class and order with a multilevel model, there was stronger evidence for black swans in insect populations than is visually apparent in Fig.~\ref{fig:nu-coefs} --- four of 10 orders with the highest median probability of heavy tails were insect orders --- however, there was considerable uncertainty in these estimates (Fig.~\ref{fig:posteriors}b).
 
@@
 
-TODO ADD TO THIS Our simulation testing show that, if anything, our models underpredict the true magnitude and probability of heavy tailed events --- especially given the length of the time series in the GPDD (Figs~\ref{fig:sim-nu}, \ref{fig:sim-gompertz}, \ref{fig:sim-gompertz-boxplots}, \ref{fig:sim-prob}) AND ``(as was also observed in simulation testing, Fig.~\ref{fig:sim-nu})''
+TODO sensitivity to priors Fig.~\ref{fig:alt-priors}.
 
-Across populations, the probability of observing black-swan dynamics was positively related to time-series length and negatively related to magnitude of process noise ($\sigma$) but not clearly related to population growth rate ($\lambda$), density dependence ($b$), or maximum lifespan (Figs~\ref{fig:correlates}, \ref{fig:correlate-coefs}).
+TODO ADD TO THIS Our simulation testing show that, if anything, our models underpredict the true magnitude and probability of heavy tailed events --- especially given the length of the time series in the GPDD (Figs~\ref{fig:sim-nu}, \ref{fig:sim-prob}) AND ``(as was also observed in simulation testing, Fig.~\ref{fig:sim-nu})''
+
+Across populations, the probability of observing black-swan dynamics was positively related to time-series length and negatively related to magnitude of process noise ($\sigma$) but not clearly related to population growth rate ($\lambda$), density dependence ($b$), or maximum lifespan (Figs~\ref{fig:correlates}, \ref{fig:posteriors}a).
 Longer time-series length was the strongest covariate of observing black-swan dynamics.
-For example, we were about \pIncHeavyNThirtyNSixty~times more likely to observe heavy tails in a population with 60 time steps than in one with 30 time steps (Fig.~\ref{fig:perc-inc-p}).
+For example, we were about \pIncHeavyNThirtyNSixty~times more likely to observe heavy tails in a population with 60 time steps than in one with 30 time steps.
 However, the absolute change in probability with increased time series length was small ($0.\pHeavyNSixty$ vs.\ $0.\pHeavyNThirty$ in the previous example, Fig.~\ref{fig:correlates}).

### 2014-11-08 text/supp.tex
Edit ms and supp to bring in new figs
A start

@@
 
-\item Finally, we removed GPDD Main ID \texttt{20531}, which we noticed was a duplicate of \texttt{10139} (a heron population).
+\item We removed GPDD Main IDs \texttt{20531} and \texttt{10139}, which we noticed were duplicates of \texttt{20579} (a heron population).
+\texttt{20579} contained additional years of data not present in \texttt{10139}.
 We removed a limited number of populations from class Angiospermopsida and Bacillariophyceae to focus the taxonomy in our analysis.
 We also removed any populations with an \texttt{Unknown} taxonomic class.
+
+\item Finally, we removed populations with the following GPDD Main IDs, which we discovered data entry errors when verifying the populations with suspected black swans: \texttt{1207} because 1957 was entered as 2 but should have been 27 \citep{kendeigh1982}, \texttt{6531} because 1978 was entered as 7 should have been 47 \citep{minot1986}, and \texttt{6566} because the data did not match the graph \citep{heessen1996}.
 
@@
 These chain lengths may seem low to those familiar with software such as WinBUGS or JAGS, but the No-U-Turn Hamiltonian Markov chain Monte Carlo Sampler in Stan generally requires far fewer iterations to obtain equivalent effective sample sizes \citep{stan-manual2014}.
+
+\section{Alternative priors}
+
+TODO
+
+Fig.~\ref{fig:priors}
+
+Fig.~\ref{fig:alt-priors}
 
@@
 
-We found that we could consistently and precisely recover median posterior estimates of $\nu$ near the true value of $\nu$ with large samples ($\ge 200$) (Fig.~\ref{fig:sim-gompertz} upper panels).
+We found that we could consistently and precisely recover median posterior estimates of $\nu$ near the true value of $\nu$ with large samples ($\ge 200$) (Fig.~\ref{fig:sim-nu} upper panels).
 At smaller samples we could still usually qualitatively distinguish heavy from not-heavy tails, but the model tended to underestimate how heavy the tails were.
-At the same time, at smaller sample sizes, the model tended to overestimate how large the scale parameter was (Fig.~\ref{fig:sim-gompertz} lower panels).
+At the same time, at smaller sample sizes, the model tended to overestimate how large the scale parameter was (Fig.~\ref{fig:sim-nu} lower panels).
 
@@
 
+TODO, rework this paragraph:
 Our results show that the Gompertz model can recapture the true value of $\nu$ when the process noise was chosen so that appropriate tail events were present (Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}, red and green symbols in the top rows).
@@
 
-We incorporated weakly informative priors into our model: $\mathrm{Cauchy}(0,
-10)$ on the global intercept $\alpha$, $\mathrm{Half\mhyphen Cauchy}(0, 2.5)$ on all standard deviation parameters, $\mathrm{Half\mhyphen Cauchy}(0, 10)$ on the dispersion parameter $\phi_\mathrm{disp}$, and $\mathrm{Cauchy}(0, 2.5)$ on all other parameters \citep{gelman2006c, gelman2008d}.
+We incorporated weakly informative priors into our model: $\mathrm{Cauchy}(0, 10)$ on the global intercept $\alpha$, $\mathrm{Half\mhyphen Cauchy}(0, 2.5)$ on all standard deviation parameters, $\mathrm{Half\mhyphen Cauchy}(0, 10)$ on the dispersion parameter $\phi_\mathrm{disp}$, and $\mathrm{Cauchy}(0, 2.5)$ on all other parameters \citep{gelman2006c, gelman2008d}.
 Compared to normal priors, the Cauchy priors concentrate more probability density around expected parameter values while allowing for a higher probability density far into the tails, thereby allowing the data to dominate the posterior more strongly if it disagrees with the prior.
@@
 
-To derive taxonomic-order-level estimates of the probability of heavy tails accounting for time-series length (Fig \ref{fig:order-estimates}), we fit a separate multilevel model with the same structure but with only $\log$ time-series length as a predictor.
+To derive taxonomic-order-level estimates of the probability of heavy tails accounting for time-series length (Fig.~\ref{fig:posteriors}b), we fit a separate multilevel model with the same structure but with only $\log$ time-series length as a predictor.
 (In this case, we did not want to control for intrinsic population characteristics such as density dependence.) Since our predictors were centered by subtracting their mean value, we obtained order-level estimates of the probability of heavy tails at mean log time-series length by adding the posteriors for $\alpha$, $\alpha^\mathrm{class}_j$, and $\alpha^\mathrm{order}_k$.
@@
 \begin{footnotesize}
-\begin{longtable}{>{\RaggedRight}m{1.5cm}>{\RaggedRight}p{4.3cm}>{\RaggedRight}p{0.8cm}>{\RaggedRight}p{1.7cm}>{\RaggedRight}p{1.0cm}>{\RaggedRight}p{3.0cm}>{\RaggedRight}p{1.7cm}>{\RaggedRight}p{0.8cm}}
+\begin{longtable}{>{\RaggedRight}m{1.5cm}>{\RaggedRight}p{4.3cm}>{\RaggedRight}p{0.8cm}>{\RaggedRight}p{1.7cm}>{\RaggedRight}p{1.0cm}>{\RaggedRight}p{3.0cm}>{\RaggedRight}p{1.7cm}>{\RaggedRight}p{1.3cm}}
 \caption{Populations with Pr$(\nu < 10) > 0.5$. TODO}\\
@@
 }
-\label{fig:alt}
+\label{fig:alt-priors}
 \end{center}

### 2014-11-08 text/ms.tex
Edits to text

@@
 We find strong evidence of black swans, but they are rare occurring in \overallMinPerc--\overallMaxPerc\% of populations: most frequently for birds (\AvesRangePerc\%) followed by mammals (\MammaliaRangePerc\%), insects (\InsectaRangePerc\%), and fishes (\OsteichthyesRangePerc\%).
-Black swans are not explained by any life-history covariates, but tend to be driven by external perturbations such as climate, severe winters, predator or parasite cycles, and the combined effects of multiple factors.
+Black swans were not explained by any life-history covariates, but tended to be driven by external perturbations such as climate, severe winters, predator or parasite cycles, and the combined effects of multiple factors.
 Black swans were more frequently detected for populations with longer time series and lower levels of process noise; for shorter and noisier time series, our simulations suggested black-swan dynamics are often not identifiable as such.
@@
 Black swans are unexpected extreme events with potentially dramatic consequences \citep{taleb2007,sornette2009}.
-One of the most striking black swans in ecology is the asteroid collision marking the end-Cretaceous mass extinction 65 million years ago \citep{alvarez1980,harnik2012}.
+One of the most striking black swans in ecology is the possible asteroid collision marking the end-Cretaceous mass extinction 65 million years ago \citep{alvarez1980,harnik2012}.
 Today, climate extremes, in concert with shifts in mean temperature, are expected to cause the greatest ecological and societal damage \citep{ipcc2012}.
@@
 This distribution describes the most extreme event per time interval (e.g.~heaviest rainfall per year), but requires time series that are sufficiently long to be condensed into time intervals.
-Another statistical tool involves fitting a catastrophic mixture distribution in a state space model to quantify the probability that population events are extreme, although this is also data intensive \citep{ward2007} A third tool is to compare the fits of thin- and heavy-tailed distributions with the Akaike Information Critera (AIC) TODO CHECK \citep{halley2002}, but this analysis did not quantify the probability of black swans or allow for population dynamics.
+Another statistical tool involves fitting a catastrophic mixture distribution in a state space model to quantify the probability that population events are extreme, although this is also data intensive \citep{ward2007}.
+A third tool is to compare the support for fits of thin- and heavy-tailed distributions \citep{halley2002}, but this analysis did not quantify the probability of black swans or allow for population dynamics.
 
@@
 Our framework quantifies both the magnitude and probability of heavy-tailed dynamics, allows for a range of population dynamic models, can incorporate observation uncertainty, and can be easily applied to abundance time series.
-We apply our model to \NPops\ of populations from four taxonomic classes.
-We estimate the frequency and strength of black swans in population dynamics across taxonomy; verify some causes of black-swan events; and attempt to identify characteristics of time series or intrinsic life-history characteristics that are associated with heavy-tailed dynamics.
+We apply our model to \NPops\ populations and estimate the frequency and strength of black-swan events in population dynamics across taxonomy; verify some causes of black-swan events; and attempt to identify characteristics of time series or intrinsic life-history characteristics that are associated with heavy-tailed dynamics.
 %We find a high probability of ecological
@@
 
-To obtain estimates of the probability and magnitude of black swans, we fit population dynamic models to abundance time series from around the world.
+To obtain estimates of the probability and magnitude of black-swan events, we fit population dynamic models to abundance time series from around the world.
 For each population, we estimated the shape of the process noise tails by measuring the degrees of freedom ($\nu$) of a Student t-distribution.
@@
 
-We selected abundance time series from the Global Population Dynamics Database (GPDD) \citep{gpdd2010}, which contains nearly 5000 time series of abundance from $\sim$1000 species and $\sim$100 taxonomic orders.
+We selected abundance time series from the Global Population Dynamics Database (GPDD; \citeauthor{gpdd2010} \citeyear{gpdd2010}), which contains nearly 5000 time series of abundance from $\sim$1000 species and $\sim$100 taxonomic orders.
 We filtered the data (Supporting Information) to remove populations from less reliable data sources, and those without sufficient data for our models, and then interpolated some missing values \citep[\textit{sensu}][]{brook2006a}.
-TODO CONTEXTUALIZE THE QUANTITY OF INTERPOLATION Our final dataset contained \NPops~populations across \NOrders~taxonomic orders and seven taxonomic classes, with a median of \medianTimeSteps~time steps (range of \minTimeSteps--\maxTimeSteps) (Table S1, Fig.~\ref{fig:all-ts}).
+Our interpolation affected only $\sim$\interpPointsPerc \% of the final data points (Table~\ref{tab:stats}) and none of the data points that were later considered black-swan events.
+Our final dataset contained \NPops~populations across \NOrders~taxonomic orders and seven taxonomic classes, with a median of \medianTimeSteps~time steps (range of \minTimeSteps--\maxTimeSteps) (Table~\ref{tab:stats}, Fig.~\ref{fig:all-ts}).
 
@@
 By allowing the process noise to be drawn from a Student-t distribution we can estimate the degree to which the process errors have heavy tails and are thereof evidence of black-swan events (Fig.~\ref{fig:didactic}a, b).
-For example, at $\nu = 2$, the probability of drawing a value more than five standard deviations below the mean is $0.018$, whereas the probability of drawing such a value from a normal distribution is tiny ($2.9\cdot10^{-7}$) TODO CHECK.
+For example, at $\nu = 3$, the probability of drawing a value more than five standard deviations below the mean is $0.02$, whereas the probability of drawing such a value from a normal distribution is tiny ($2.9\cdot10^{-7}$).
 As the value of $\nu$ approaches infinity, the distribution approaches the normal distribution (Fig.~\ref{fig:didactic}a, b).
@@
 For $\nu$, we chose an exponential prior with rate parameter of $0.01$ truncated at values above two (Fig.~\ref{fig:priors}) --- a slightly less informative prior than suggested by \citet{fernandez1998}.
-This prior gives only a $7.7$\% probability that $\nu < 10$ but constrains the sampling sufficiently to avoid wandering off towards infinity.
+This prior gives only a \basePriorProbHeavy \% probability that $\nu < 10$ but constrains the sampling sufficiently to avoid wandering off towards infinity.
 In any case, for $\nu > 20$ the t distribution is almost indistinguishable from the normal distribution (Fig.~\ref{fig:didactic}).
@@
 TODO QUANTIFY UPS AND DOWNS --- SKEW
-We assembled the documented causes of black-swan population dynamics in seven exemplar populations with high probabilities of heavy tails (Table 1).
+We assembled the documented causes of black-swan population dynamics in seven exemplar populations with high probabilities of heavy tails (Table~\ref{tab:sparks}).
 The majority of documented events were downward black swans and involved a series of unfortunate events.
@@
 The GPDD data represent a taxonomically and geographically biased sample of populations --- the longer time series we focus on are dominated by commercially and recreationally important species and a disproportionate number of populations are located in the United Kingdom.
-Although we expect that we would find qualitatively similar evidence for black swans in many other large taxonomic or geographic samples of populations, the common forces driving those black swans likely differ (e.g.~severe winters in Table~1).
+Although we expect that we would find qualitatively similar evidence for black swans in many other large taxonomic or geographic samples of populations, the common forces driving those black swans likely differ (e.g.~severe winters in Table~\ref{tab:sparks}).
 In addition to a possibly biased sample of populations, some black-swan detections could just be recording mistakes, or conversely, some extreme observations may have been discarded or altered if they were erroneously suspected of being recording mistakes.
@@
 The following supporting information is available online for this article:\\
-Figures S1--S11\\
-Table S1\\
+Tables S1 and S2\\
+Figures S1--S6\\
 Example Stan code for a heavy-tailed Gompertz model and multilevel beta

### 2014-11-08 text/supp.tex
Edits to text

@@
 Note that interpolation is highly unlikely to lead to black-swan detections, since black swans involve extreme increases or decreases.
-Table S1 shows the final taxonomic breakdown and the number of populations with interpolated values.
+Table~\ref{tab:stats} shows the final taxonomic breakdown and the number of populations with interpolated values.
 
@@
 \setcounter{table}{0}
-\input{../analysis/stat-table}
+
+\begin{table}
+\begin{footnotesize}
+\caption{Summary statistics for the filtered Global Population Dynamics Database time series arranged by taxonomic class.
+Columns are: number of populations, number of taxonomic orders, numbers of species, median time series length, total number of interpolated time steps, total number of substituted zeros, and total number of time steps.}
+\smallskip
+\begin{tabular}{lrrrrrrrr}
+\toprule
+\input{../analysis/stat-table.tex}
+\label{tab:stats}
+\end{tabular}
+\end{footnotesize}
+\end{table}
+
 \clearpage

### 2014-11-08 analysis/7-values-for-paper.R
Fix decimals in some sunk values

@@ -151,12 +151,12 @@ write_tex(baseNuFiveObsTenSwitch, "baseNuFiveObsTenSwitch")
 
 ## - how much more probable were heavy tails with 60 time steps. vs. 30 time
 ##   steps? (percentage and absolute)
-pHeavyN30 <- round(100 * p_inc[p_inc$N == 30, "p"])
-pHeavyN60 <- round(100 * p_inc[p_inc$N == 60, "p"])
+pHeavyN30 <- p_inc[p_inc$N == 30, "p"]
+pHeavyN60 <- p_inc[p_inc$N == 60, "p"]
 pIncHeavyN30N60 <- pHeavyN60 / pHeavyN30
-write_tex(pHeavyN30, "pHeavyNThirty")
-write_tex(pHeavyN60, "pHeavyNSixty")
-write_tex(pIncHeavyN30N60, "pIncHeavyNThirtyNSixty")
+write_tex(sprintf("%.2f", round(pHeavyN30, 2)), "pHeavyNThirty")
+write_tex(sprintf("%.2f", round(pHeavyN60, 2)), "pHeavyNSixty")
+write_tex(sprintf("%.1f", round(pIncHeavyN30N60, 1)), "pIncHeavyNThirtyNSixty")
 
 ## Discussion:
 ## - Ricker / gompertz range of pops with black swans as percent

### 2014-11-08 text/ms.tex
Edit, edit, edit

@@
 There is compelling anecdotal evidence for ecological black swans, but systematic evidence has been elusive.
-A survey of ecologists indicated that surprising outcomes of field experiments are far more common than we assume \citep{doak2008}, and a number of major environmental events on geologic and contemporary time scales could be considered to have caused ecological black swans \citep{nunez2012}.
-Black swans, or heavy-tailed distributions, have been discovered in time-series of marine microbe populations \citet{segura2013} and time to extinction for experimental waterflea populations \citet{drake2014}.
-As far as we can tell there have been two systematic surveys for ecological black swans: one with North American breeding birds and the other with the Global Population Dynamics Database, which found little clear evidence for them \citep{keitt1998,allen2001,halley2002}.
+A survey of ecologists indicated that surprising outcomes of field experiments are far more common than we assume \citep{doak2008}, and events such as the global invasion of  Argentine ants and the mutation of viruses to infect new hosts could be considered black swans \citep{nunez2012}.
+In fact, anecdotes of population catastrophes are numerous and catastrophes may be the most important element affecting population persistence \citep{mangel1994}.
+For marine mammal populations, we have compelling evidence of catastrophes \citep{gerber2001, ward2007}, and recently, heavy-tailed distributions have been discovered in time-series of marine microbe abundance \citep{segura2013} and time to extinction for experimental waterflea populations \citep{drake2014}.
+As far as we can tell there have been two systematic surveys for ecological black swans: one with North American breeding birds and the other with the Global Population Dynamics Database, which uncovered little clear evidence for them \citep{keitt1998,allen2001,halley2002}.
 Nevertheless there are methodological challenges to their detection \citep{allen2001,ward2007}.
 % TODO: pull anecdote from that Nunez study
+
+% TODO \citep{mangel1994} catastrophes may be the most important element affecting species persistence times and more common than we think
+% TODO \citep{gerber2001} and \citep{ward2007} evidence for marine mammal populations
 
@@
 In any case, for $\nu > 20$ the t distribution is almost indistinguishable from the normal distribution (Fig.~\ref{fig:didactic}).
-We investigated the sensitivity of our results to weaker and stronger priors (Fig.~\ref{fig:priors}, Supporting Information).
+We investigated the sensitivity of our results to weaker and stronger priors (exponential rate parameter $= 0.005, 0.02$; Fig.~\ref{fig:priors}, Supporting Information).
 We used simulated data to test how easily we could detect $\nu$ given different sample sizes and to ensure we could recover unbiased parameter estimates from the Gompertz model (Supporting Information).
@@
 We plotted characteristics of the time series ($\sigma$, $\lambda$, $b$, and time-series length) along with two life-history characteristics (body length and maximum lifespan obtained from \citet{brook2006a}) against our estimated probability of heavy tails, Pr$(\nu < 10) > 0.5$.
-We formally investigated these relationships by fitting beta regression multilevel models TODO REF AND MORE DETAILS?.
+We formally investigated these relationships by fitting beta regression multilevel models \citep{ferrari2004}.
 For covariates that were derived from Gompertz model parameter estimates, we incorporated standard deviations around the means.
-To account for broad patterns of phylogenetic relatedness, we allowed for varying intercepts at the taxonomic class, order, and species level (Supporting Information).
+To account for broad patterns of phylogenetic relatedness, we allowed for varying intercepts (``random effects'') at the taxonomic class, order, and species level (Supporting Information).
+We fit our model in Stan with weakly informative priors on the coefficients \citep{gelman2008d} and variance parameters (\citeauthor{gelman2006c} \citeyear{gelman2006c} \citeauthor{gelman2014} \citeyear{gelman2014}; Supporting Information).
 
@@
 Where possible, we found the documented causes of ecological black swans in the primary data source cited in the GPDD or in other literature describing the population.
-These exemplar populations were chosen haphazardly with the purpose of generating hypotheses about the causes of population dynamic black swans.
+%These exemplar populations were chosen haphazardly with the purpose of generating hypotheses about the causes of population dynamic black swans.
 
@@
 
-The majority of our heavy-tailed estimates were robust to alternative population models and observation error (Fig.~\ref{fig:alt}).
-Our conclusions were not systematically altered when we included an autocorrelation structure in the residuals, modelled population growth rates without density dependence, or modelled the population dynamics as Ricker-logistic.
+The majority of our heavy-tailed estimates were robust to alternative population models, observation error, and choice or priors.
+Our conclusions were not systematically altered when we included an autocorrelation structure in the residuals, modelled population growth rates without density dependence, or modelled the population dynamics as Ricker-logistic (Fig.~\ref{fig:alt}).
 However, setting observation error standard deviation to $0.2$ increased the median estimate of $\nu$ from $<10$ to $\ge 10$ in $\baseNuTenObsTenSwitch$ of $\baseNuTen$ populations, although the majority of $\nu$ estimates remained qualitatively similar (Fig.~\ref{fig:alt}).
-
-TODO sensitivity to priors Fig.~\ref{fig:alt-priors}.
-
-TODO ADD TO THIS Our simulation testing show that, if anything, our models underpredict the true magnitude and probability of heavy tailed events --- especially given the length of the time series in the GPDD (Figs~\ref{fig:sim-nu}, \ref{fig:sim-prob}) AND ``(as was also observed in simulation testing, Fig.~\ref{fig:sim-nu})''
+The strength of the prior on $\nu$ had little influence on estimates of black-swan dynamics (Fig.~\ref{fig:alt-priors}).
+Our simulation testing shows that, if anything, our models underpredict the true magnitude and probability of heavy tailed events --- especially given the length of the time series in the GPDD (Figs~\ref{fig:sim-nu}, \ref{fig:sim-prob}).
 
@@
 For example, we were about \pIncHeavyNThirtyNSixty~times more likely to observe heavy tails in a population with 60 time steps than in one with 30 time steps.
-However, the absolute change in probability with increased time series length was small ($0.\pHeavyNSixty$ vs.\ $0.\pHeavyNThirty$ in the previous example, Fig.~\ref{fig:correlates}).
+However, the absolute change in probability with increased time series length was small ($\pHeavyNSixty$ vs.\ $\pHeavyNThirty$ in the previous example, Fig.~\ref{fig:correlates}).
 
-% "Perhaps then you could say "we examined all time series with published explanations of why the black swan events occurred"
-TODO STATE HOW THESE 7 WERE CHOSEN
-TODO QUANTIFY UPS AND DOWNS --- SKEW
-We assembled the documented causes of black-swan population dynamics in seven exemplar populations with high probabilities of heavy tails (Table~\ref{tab:sparks}).
-The majority of documented events were downward black swans and involved a series of unfortunate events.
-For example, a shortage of nest sites for the shag long-necked bird (TODO SCI.\ NAME) in the United Kingdom reduced population productivity and made the population more vulnerable when a red tide hit in 1968 \citep{potts1980}.
+We examined all time series with published explanations of why the black-swan events occurred (Table~\ref{tab:causes-supp}) and focus on seven exemplar populations with high probabilities of heavy tails (Table~\ref{tab:sparks}).
+The majority (TODO QUANTIFY) of documented events were downward black swans and involved a series of unfortunate events.
+For example, a shortage of nest sites for the shag long-necked bird (\textit{Phalacrocorax aristotelis}) in the United Kingdom reduced population productivity and made the population more vulnerable when a red tide hit in 1968 \citep{potts1980}.
 The population experienced a black-swan downswing, but the new availability of nest sites resulted in a rapid population upswing \citep{potts1980}.
-As another example, an interaction between environmental- and predation-mediated population cycles are thought to have caused a downward black swan for a water vole population \citep{saucy1994}.
+As another example, an interaction between environmental- and predation-mediated population cycles are thought to have caused a downward black-swan event for a water vole (\textit{Arvicola terrestris}) population \citep{saucy1994}.
 Other black swans were the result of a sequence of extreme climate events on their own.
-For example, severe winters in 1929, 1940--1942, and 1962--1963 were associated with black-swan downswings in grey heron abundance in the United Kingdom \citep{stafford1971}.
+For example, severe winters in 1929, 1940--1942, and 1962--1963 were associated with black-swan downswings in grey heron (\textit{Ardea cinerea}) abundance in the United Kingdom \citep{stafford1971}.
 Our analysis finds that the last event was a combination of two black-swan events in a row and it took the population three times longer to recover than predicted \citep{stafford1971}.
+
+(TODO: substitute a parasite example.)
 
@@
 
-Our results suggest important future issues to address about ecological black swans.
-Given that black swans do occur, can we forecast the probability of black swans in space and time?
-Furthermore, what management policies allow us to detect them quickly after they happen?
-Can we isolate the components of ecological dynamics that experience black swans by moving from phenomenological models such as the Gompertz to mechanistic models, that take into account, for example, recruitment?
-We expect that greater insight into the mechanisms and covariates of ecological black swans may be best obtained through specific geographic and taxonomic subsets of data where longer time series with low levels of observation error are available \citep[e.g.][]{segura2013}.
-
-Most importantly, what is the impact of allowing for black swans in forecasts of ecological risk?
-Financial analysts are concerned with the shape of financial-return-distribution tails because these impact estimates of risk --- the probability of a specific magnitude of undesired event occurring \citep{rachev2008}.
-A comparable focus in ecology would increase our estimates of extinction risk, since these would be disproportionately impacted by black-swan events.
-TODO END ON THIS
-
 Recognizing the prevalence of heavy-tailed dynamics suggests a number of policy directions.
@@
 Finally, extreme and unexpected, surprising, or counterintuitive ecological dynamics offer a tremendous opportunity to learn about ecological systems, evaluate when our models break down, and adjust future management policy \citep{doak2008, pine-iii2009, lindenmayer2010}.
+
+Our results suggest important future issues to address about ecological black swans.
+Given that black swans do occur, can we forecast the probability of black swans in space and time?
+Furthermore, what management policies allow us to detect them quickly after they happen?
+Can we isolate the components of ecological dynamics that experience black swans by moving from phenomenological models such as the Gompertz to mechanistic models, that take into account, for example, recruitment?
+We expect that greater insight into the mechanisms and covariates of ecological black swans may be best obtained through specific geographic and taxonomic subsets of data where longer time series with low levels of observation error are available \citep[e.g.][]{segura2013}.
+
+Most importantly, what is the impact of allowing for black swans in forecasts of ecological risk? 
 In recent decades ecology has moved toward focussing on aspects of variance in addition to mean responses \citep[e.g.][]{loreau2010a, thompson2013}.
 Our results suggest that an added focus on ecological extremes represents the next frontier, particularly in the face of increased climate extremes \citep{meehl2004,ipcc2012}.
-
-%TODO WORK IN Gerber L.R. & Hilborn R. (2001). Catastrophic events and recovery from low densities in populations of otariids: implications for risk of extinction. Mammal Rev., 31, 131-150.
+Financial analysts are concerned with the shape of financial-return-distribution tails because these impact estimates of risk --- the probability of a specific magnitude of undesired event occurring \citep{rachev2008}.
+A comparable focus in ecology would increase our estimates of extinction risk, since these would be disproportionately impacted by downward black-swan events.
 

### 2014-11-08 text/supp.tex
Edit, edit, edit

@@
 
-TODO
+To test if the prior on $\nu$ influenced our estimate of black-swan dynamics, we refit our models with weaker and stronger priors.
+Our base model used a prior on $\nu$ of Truncated-Exponential(0.02, min.\ = 2).
+For a weaker prior we used Truncated-Exponential(0.005, min.\ = 2) and for a stronger prior we used Truncated-Exponential(0.02, min.\ = 2) (Fig.~\ref{fig:priors}).
+Note that the base and weaker priors are relatively flat within the region of $\nu < 20$---the region we are mostly concerned about when categorizing populations as heavy- or thin-tailed.
 
-Fig.~\ref{fig:priors}
-
-Fig.~\ref{fig:alt-priors}
+Our results show that these weaker and stronger priors would have little influence on our conclusions about heavy-tailed dynamics (Fig.~\ref{fig:alt-priors}).
+When the data are informative about tail behaviour (i.e.\ when there is strong evidence of low $\nu$ values, upper-right of Fig.~\ref{fig:alt-priors}), the prior has little impact on the estimate of $\nu$.
+When the data are less informative about $\nu$ (i.e.\ when there are no or few tail events and time series are short or noisy), the prior can pull the estimate of $\nu$ towards larger or smaller values (Fig.~\ref{fig:alt-priors}).
+The vast majority of the populations with Pr$(\nu < 10)$ in the base prior are not altered qualitatively by this range of prior strength.
 
@@
 We did not include these models in Fig.~\ref{fig:alt}.
-\textit{TODO perhaps I should go up to 16,000 so that more converge.}
 
@@
 
-TODO, rework this paragraph:
-Our results show that the Gompertz model can recapture the true value of $\nu$ when the process noise was chosen so that appropriate tail events were present (Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}, red and green symbols in the top rows).
+Our results indicate that the Gompertz model can recapture the true value of $\nu$ when the process noise was chosen so that appropriate tail events were present (TODO FIG).
+The addition of observation error caused the model to tend to underestimate the degree of heavy-tailedness.
+Fitting a model with correctly specified observation error did not make substantial improvements to model bias.
+
+%(Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}, red and green symbols in the top rows).
 Likewise, the other Gompertz parameters were estimated without any systematic bias (Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}, red and green symbols).
-The addition of observation error caused the model to tend to underestimate the degree of heavy-tailedness, overestimate the magnitude of process noise, somewhat overestimate $\lambda$, and overestimate density dependence (blue symbols in Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}).
+%, overestimate the magnitude of process noise, somewhat overestimate $\lambda$, and overestimate density dependence (blue symbols in Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}).
 The overestimation of density dependence with observation error is a known phenomenon \citep{knape2012}.
@@
 \begin{longtable}{>{\RaggedRight}m{1.5cm}>{\RaggedRight}p{4.3cm}>{\RaggedRight}p{0.8cm}>{\RaggedRight}p{1.7cm}>{\RaggedRight}p{1.0cm}>{\RaggedRight}p{3.0cm}>{\RaggedRight}p{1.7cm}>{\RaggedRight}p{1.3cm}}
-\caption{Populations with Pr$(\nu < 10) > 0.5$. TODO}\\
+\caption{All populations with Pr$(\nu < 10) > 0.5$ in the base heavy-tailed Gompertz population dynamics model.
+Shown are the log abundance time series, population descriptions, Global Population Dynamics Database Main IDs, citation for the data source or separate verification literature, a description of the cause of the black swan events (if known), the probability of heavy tails as calculated by our model, and median estimate of $\nu$ from our model with 90\% quantile credible intervals indicated in parentheses.
+Red dots on the time series indicate downward black-swan events and blue values indicate upward black-swan events.
+(TODO add what probability these are.)}\\
 \toprule
@@
 \setcounter{figure}{0}
-
-% \begin{figure}[htbp]
-% \begin{center}
-% \includegraphics[width=\textwidth]{../analysis/all-clean-ts-3.pdf}
-% \caption{
-% TODO WILL SPLIT OVER PAGES BY CLASS
-%   All filtered times series used in our analysis. The abundances are shown on
-%   a log10 vertical axis. Colours indicate taxonomic classes. Open black circles
-%   indicate abundance values that were interpolated. Closed black circles
-%   indicate single abundance values that were recorded as zero but were set to
-%   the next lowest observed abundance.
-% }
-%     \label{fig:all-ts}
-% \end{center}
-% \end{figure}
 
@@
 The inset panel shows the same data but with a log-transformed x axis.
+Note that the base and weaker priors are relatively flat within the region of $\nu < 20$ that we are concerned with.
 }
@@
 \clearpage
-
-% \begin{figure}[htbp]
-% \begin{center}
-% \includegraphics[width=0.9\textwidth]{../analysis/ts-gpdd-heavy-eg-log10-base.pdf}
-% \caption{
-%   Time series for populations with Pr($\nu < 10$) $>$ 0.5 using the base
-%   Gompertz population model. Panels are ordered by increasing Pr($\nu < 10$).
-%   Vertical axes are on a log10 scale. Colours indicate taxonomic classes. The
-%   labels on each panel indicate Pr($\nu < 10$) $>$ 0.5, the common name for the
-%   species, and the Global Population Dynamics Database ID number.
-% }
-% \label{fig:heavy-ts}
-% \end{center}
-% \end{figure}
-%
-% \clearpage
 
@@
 \end{figure}
-
-\clearpage
-
-% \begin{figure}[htbp]
-% \begin{center}
-% \includegraphics[width=\textwidth]{../analysis/sim-gompertz.pdf}
-% \caption{
-% Simulation testing the Gompertz estimation model when the process deviation
-% draws were chosen so that $\nu$ could be estimated close to the true value
-% outside the full population model (``effective $\nu$'' within a CV of 0.2 of
-% specified $\nu$). The simulation was run across population $\nu$ values
-% (columns) and different scenarios (colours): (1) 100 time steps and no
-% observation error, (2) 50 time steps and no observation error, (3) 50 time
-% steps and observation error drawn from $\mathrm{Normal} (0, 0.2^2)$ but
-% ignored, and (4) 50 time steps with observation error in which the quantity
-% of observation error was assumed known. Dashed horizontal lines show the true
-% population values; these true values were chosen to represent approximately
-% the median values as estimated from the Global Population Dynamics Database.
-% Individual dots and lines
-% represent a stochastic draws from the true population distribution and
-% a model fitting (median and 80\% credible intervals). The panels from top to
-% bottom show $1/\nu$, process noise scale parameter $\sigma$, growth rate
-% parameter $\lambda$, and the density dependence parameter $b$.
-% }
-% \label{fig:sim-gompertz}
-% \end{center}
-% \end{figure}
-
-\clearpage
-
-% \begin{figure}[htbp]
-% \begin{center}
-% \includegraphics[width=\textwidth]{../analysis/sim-gompertz-boxplots.pdf}
-% \caption{
-%   The distribution of median estimates from the same simulation as described in
-%   Fig.~\ref{fig:sim-gompertz}. The boxes show the interquartile range. The
-%   whiskers extend to $1.5$ times the interquartile range and outliers are shown
-%   as dots. Note that these distributions represent the output from only 20
-%   simulation iterations, so there is some level of remaining sampling noise.
-% }
-% \label{fig:sim-gompertz-boxplots}
-% \end{center}
-% \end{figure}
 

### 2014-11-08 analysis/phylopic.tex
Output a table of phylopic citations

% latex table generated in R 3.1.2 by xtable 1.7-4 package
% Sat Nov  8 14:57:15 2014
Taxonomic order & Credit & License URL \\ 
  \midrule
Salmoniformes & Servien (vectorized by T. Michael Keesey) & \url{http://creativecommons.org/licenses/by-sa/3.0/} \\ 
  Gadiformes &  & \url{http://creativecommons.org/publicdomain/mark/1.0/} \\ 
  Perciformes & Ellen Edmonson and Hugh Chrisp (vectorized by T. Michael Keesey) & \url{http://creativecommons.org/publicdomain/mark/1.0/} \\ 
  Pleuronectiformes & Tony Ayling (vectorized by T. Michael Keesey) & \url{http://creativecommons.org/licenses/by-sa/3.0/} \\ 
  Lepidoptera & Curtis (modified by T. Michael Keesey) & \url{http://creativecommons.org/publicdomain/mark/1.0/} \\ 
  Rodentia & Mattia Menchetti & \url{http://creativecommons.org/publicdomain/zero/1.0/} \\ 
  Carnivora & Brian Gratwicke (photo) and T. Michael Keesey (vectorization) & \url{http://creativecommons.org/licenses/by/3.0/} \\ 
  Lagomorpha & Sarah Werning & \url{http://creativecommons.org/licenses/by/3.0/} \\ 
  Coleoptera & Crystal Maier & \url{http://creativecommons.org/licenses/by/3.0/} \\ 
  Odonata & Gareth Monger & \url{http://creativecommons.org/licenses/by/3.0/} \\ 
  Passeriformes & Michael Scroggie & \url{http://creativecommons.org/publicdomain/zero/1.0/} \\ 
  Anseriformes & Sharon Wegner-Larsen & \url{http://creativecommons.org/publicdomain/zero/1.0/} \\ 
  Artiodactyla & Jan A. Venter, Herbert H. T. Prins, David A. Balfour and Rob Slotow (vectorized by T. Michael Keesey) & \url{http://creativecommons.org/licenses/by/3.0/} \\ 
  Diptera & Gareth Monger & \url{http://creativecommons.org/licenses/by/3.0/} \\ 
  Charadriiformes & JJ Harrison (vectorized by T. Michael Keesey) & \url{http://creativecommons.org/licenses/by-sa/3.0/} \\ 
  Hemiptera & T. Michael Keesey & \url{http://creativecommons.org/publicdomain/zero/1.0/} \\ 
  Falconiformes & Liftarn & \url{http://creativecommons.org/licenses/by-sa/3.0/} \\ 
  Galliformes & Steven Traver & \url{http://creativecommons.org/publicdomain/zero/1.0/} \\ 
   \bottomrule

### 2014-11-08 analysis/6-plot-order-correlate-posteriors.R

@@ -26,6 +26,30 @@ or[or$taxonomic_order == "Pleuronectiformes", "scaling_factor"] <- 0.8
 # to get ordering right:
 or <- plyr::join(a_df[,c("taxonomic_order", "sorted_order")], or)
 
+
+# get citation details:
+if(!file.exists("phylopic-metadata.rds")) {
+  or_temp <- plyr::ldply(seq_len(nrow(or)), function(i) {
+    temp <- rphylopic::image_get(uuid = or$hash[i], options = c("credit", "licenseURL"))
+    temp <- lapply(temp, function(x) ifelse(is.null(x), "", x))
+    data.frame(credit = temp$credit, licenseURL = temp$licenseURL)
+  })
+  saveRDS(or_temp, file = "phylopic-metadata.rds")
+} else {
+  or_temp <- readRDS("phylopic-metadata.rds")
+}
+or <- data.frame(or, or_temp)
+
+library("xtable")
+or_table <- or[ ,c("taxonomic_order", "credit", "licenseURL")]
+or_table$licenseURL <- paste0("\\url{", or_table$licenseURL, "}")
+or_table$credit <- gsub("&", "and", or_table$credit)
+names(or_table) <- c("Taxonomic order", "Credit", "License URL")
+
+print.xtable(xtable(or_table, caption = ""),
+  include.rownames = FALSE, file = "phylopic.tex", booktabs = TRUE,
+  sanitize.text.function = identity, only.contents = TRUE)
+
 if(any(!file.exists(paste0("silhouettes/", or$taxonomic_order, ".png")))) {
   for(i in 1:nrow(or)) {
     download.file(or$png_url[i],

### 2014-11-08 text/supp.tex
Add phylopic citations to supp

@@
 
+\section{Additional acknowledgements}
+
+Many of the silhouette images used in Figs~\ref{fig:nu-coefs}, \ref{fig:correlates} and \ref{fig:posteriors} were obtained from \texttt{phylopic.org} under Creative Commons licenses.
+We vectorized the salmon in Fig.~\ref{fig:nu-coefs} and Fig.~\ref{fig:correlates} ourselves.
+The bird in these figures was obtained from \texttt{phylopic.org} under a Creative Commons Attribution 3.0 Unported license with credit to Jean-Raphaël Guillaumin {[}photography{]} and T.
+Michael Keesey {[}vectorization{]}).
+The silhouettes in Fig.~\ref{fig:posteriors} were obtained from the following sources:
+
+\LTcapwidth=\textwidth
+\singlespacing
+\begin{footnotesize}
+\begin{longtable}{>{\RaggedRight}m{3.2cm}>{\RaggedRight}p{6.5cm}>{\RaggedRight}p{5.0cm}}
+%\caption{Phylopic credits}\\
+\toprule
+\input{../analysis/phylopic.tex}
+\label{tab:phylopic}
+\end{longtable}
+\end{footnotesize}
+\onehalfspacing
+
 \bibliographystyle{ecologyletters}

### 2014-11-08 text/ms.tex
Clean up some comments
Small edits

@@
 
-% TODO: hyphenate the adjective 'black swan'? E.g. black-swan event, black swans.
 \section{Abstract}
@@
 
-There is compelling anecdotal evidence for ecological black swans, but systematic evidence has been elusive.
+There is compelling anecdotal evidence for ecological black swans, but systematic evidence across taxa has been elusive.
 A survey of ecologists indicated that surprising outcomes of field experiments are far more common than we assume \citep{doak2008}, and events such as the global invasion of  Argentine ants and the mutation of viruses to infect new hosts could be considered black swans \citep{nunez2012}.
@@
 Nevertheless there are methodological challenges to their detection \citep{allen2001,ward2007}.
-% TODO: pull anecdote from that Nunez study
-
-% TODO \citep{mangel1994} catastrophes may be the most important element affecting species persistence times and more common than we think
-% TODO \citep{gerber2001} and \citep{ward2007} evidence for marine mammal populations
 
@@
 We apply our model to \NPops\ populations and estimate the frequency and strength of black-swan events in population dynamics across taxonomy; verify some causes of black-swan events; and attempt to identify characteristics of time series or intrinsic life-history characteristics that are associated with heavy-tailed dynamics.
-%We find a high probability of ecological
-%black swans across a wide range of taxa but in a limited number of populations.
-%When they occur, extreme climate, predation, parasites, and their interactions
-%are common documented causes. We find little evidence that intrinsic
-%life-history characteristics are related to the probability of black swans.
 
@@
 Where possible, we found the documented causes of ecological black swans in the primary data source cited in the GPDD or in other literature describing the population.
-%These exemplar populations were chosen haphazardly with the purpose of generating hypotheses about the causes of population dynamic black swans.
 
@@
 
-%TODO IMPORTANT: you need to quantify across all identified black swan events how many were downward and how many upward.
+%TODO IMPORTANT quantify across all identified black swan events how many were downward and how many upward.
 The assumption of your model is a symmetric distribution, but if in fact downward events are more common, perhaps a skewed distribution would be more realistic.
@@
 Most importantly, what is the impact of allowing for black swans in forecasts of ecological risk? 
-In recent decades ecology has moved toward focussing on aspects of variance in addition to mean responses \citep[e.g.][]{loreau2010a, thompson2013}.
+In recent decades, ecology has moved toward focussing on aspects of variance in addition to mean responses \citep[e.g.][]{loreau2010a, thompson2013}.
 Our results suggest that an added focus on ecological extremes represents the next frontier, particularly in the face of increased climate extremes \citep{meehl2004,ipcc2012}.
@@
 We are grateful to the contributors and maintainers of the Global Population Dynamics Database and to Compute Canada's WestGrid high-performance computing resources.
-Three silhouette images were obtained from \texttt{phylopic.org} under a Creative Commons Attribution 3.0 Unported license: rabbit (Sarah Werning), hoverfly (Gareth Monger), bird (Jean-Raphaël Guillaumin {[}photography{]} and T.
-Michael Keesey {[}vectorization{]}).
+Silhouette images were obtained from \texttt{phylopic.org} under a Creative Commons Attribution 3.0 Unported license; sources are listed in the Supporting Information.
 Funding was provided by a Simon Fraser University Graduate Fellowship (SCA), the Natural Sciences and Engineering Research Council of Canada (NKD, ABC), the Canada Research Chairs Program (NKD).

### 2014-11-09 analysis/6-plot-sparks.R
Output and include counts of ups vs. downs

@@ -23,6 +23,12 @@ get_gomp_res <- function(id_show) {
   return(list(res = res, l = l, u = u))
 }
 
+heavy_res <- plyr::ddply(heavy, "main_id", function(x) {
+  qq <- get_gomp_res(x$main_id[1])
+  with(qq, data.frame(res, l, u))
+})
+saveRDS(heavy_res, file = "heavy_residuals.rds")
+
 make_spark <- function(x) {
   res <- get_gomp_res(x$main_id[1])
   bsw_l <- as.numeric(na.omit(seq_along(res$res)[res$res<res$l]))

### 2014-11-09 analysis/7-values-for-paper.R

@@ -213,4 +213,14 @@ interpPointsPerc <- round((sum(stat_table$n_zero_sub) + sum(stat_table$n_interpo
 
 write_tex(interpPointsPerc, "interpPointsPerc")
 
+# count ups and downs:
+heavy_res <- readRDS("heavy_residuals.rds")
+heavy_res <- heavy_res %>% mutate(bs_l = res < l, bs_u = res > u)
+n_bs_up <- sum(heavy_res$bs_u, na.rm = TRUE)
+n_bs_down <- sum(heavy_res$bs_l, na.rm = TRUE)
+write_tex(n_bs_up, "nBSUp")
+write_tex(n_bs_down, "nBSDown")
+write_tex(sprintf("%.1f", round(n_bs_down/n_bs_up, 1)), "ratioBSDownToUp")
+write_tex(round(n_bs_down / (sum(n_bs_down, n_bs_up))*100), "percBSDown")
+
 close(zz)

### 2014-11-09 text/ms.tex
Output and include counts of ups vs. downs
Minor edits

@@
 We examined all time series with published explanations of why the black-swan events occurred (Table~\ref{tab:causes-supp}) and focus on seven exemplar populations with high probabilities of heavy tails (Table~\ref{tab:sparks}).
-The majority (TODO QUANTIFY) of documented events were downward black swans and involved a series of unfortunate events.
+The majority (\percBSDown \%) of documented events were downward black swans and involved a series of unfortunate events.
 For example, a shortage of nest sites for the shag long-necked bird (\textit{Phalacrocorax aristotelis}) in the United Kingdom reduced population productivity and made the population more vulnerable when a red tide hit in 1968 \citep{potts1980}.
@@
 Our analysis finds that the last event was a combination of two black-swan events in a row and it took the population three times longer to recover than predicted \citep{stafford1971}.
-
-(TODO: substitute a parasite example.)
+\textit{(TODO: substitute a parasite example.)}
 
@@
 We are grateful to the contributors and maintainers of the Global Population Dynamics Database and to Compute Canada's WestGrid high-performance computing resources.
-Silhouette images were obtained from \texttt{phylopic.org} under a Creative Commons Attribution 3.0 Unported license; sources are listed in the Supporting Information.
+Silhouette images were obtained from \texttt{phylopic.org} under Creative Commons licenses; sources are listed in the Supporting Information.
 Funding was provided by a Simon Fraser University Graduate Fellowship (SCA), the Natural Sciences and Engineering Research Council of Canada (NKD, ABC), the Canada Research Chairs Program (NKD).

### 2014-11-09 text/supp.tex
Small edits
Add a line break to make things look better
Minor edits

@@
 We fit a multilevel beta regression model to the predicted probability of heavy tails, Pr($\nu < 10$), to investigate potential covariates of heavy-tailed dynamics.
-The beta distribution is useful when response data range on a continuous scale between zero and one.
+The beta distribution is useful when response data range on a continuous scale between zero and one \citep{ferrari2004}.
 We used a logit link function as is typically used in logistic regression.
 The model was as follows:
+
 \begin{align*}
@@
 Michael Keesey {[}vectorization{]}).
-The silhouettes in Fig.~\ref{fig:posteriors} were obtained from the following sources:
+The silhouettes in Fig.~\ref{fig:posteriors} were obtained from the following sources (metadata obtained with the help of the rphylopic R package, \url{https://github.com/sckott/rphylopic}):
 

### 2014-11-09 text/ms.tex
Add to some captions

@@
 
-We examined all time series with published explanations of why the black-swan events occurred (Table~\ref{tab:causes-supp}) and focus on seven exemplar populations with high probabilities of heavy tails (Table~\ref{tab:sparks}).
-The majority (\percBSDown \%) of documented events were downward black swans and involved a series of unfortunate events.
+We examined all time series with published explanations of why the black-swan events occurred (Tables~\ref{tab:causes-supp} and \ref{tab:sparks}).
+The majority (\percBSDown \%) of documented events were downward black swans and involved a series of unfortunate events (Table~\ref{tab:causes-supp}).
 For example, a shortage of nest sites for the shag long-necked bird (\textit{Phalacrocorax aristotelis}) in the United Kingdom reduced population productivity and made the population more vulnerable when a red tide hit in 1968 \citep{potts1980}.

### 2014-11-09 text/sparks.tex
Add to some captions

@@
   Dynamics Database and a description of their causes. Red and blue dots
-  indicate downward and upward events that have a 0.001\% probability or less of
+  indicate downward and upward events that have a $10 \cdot 10^{-4}$ probability or less of
   occurring if the population dynamics were explained by a Gompertz model

### 2014-11-09 text/supp.tex
Add to some captions

@@
 Shown are the log abundance time series, population descriptions, Global Population Dynamics Database Main IDs, citation for the data source or separate verification literature, a description of the cause of the black swan events (if known), the probability of heavy tails as calculated by our model, and median estimate of $\nu$ from our model with 90\% quantile credible intervals indicated in parentheses.
-Red dots on the time series indicate downward black-swan events and blue values indicate upward black-swan events.
-(TODO add what probability these are.)}\\
+Red dots on the time series indicate downward black-swan events and blue values indicate upward black-swan events that have a $10 \cdot 10^{-4}$ probability or less of occurring if the population dynamics were explained by a Gompertz model with normally distributed process noise with a standard deviation equal to the scale parameter in the fitted t distribution.}\\
 \toprule

### 2014-11-09 text/ms.tex
Add 'downward events' in discussion summary
Edits to ms

@@
 
-The majority of our heavy-tailed estimates were robust to alternative population models, observation error, and choice or priors.
+The majority of our heavy-tailed estimates were robust to alternative population models, observation error, and choice of priors.
 Our conclusions were not systematically altered when we included an autocorrelation structure in the residuals, modelled population growth rates without density dependence, or modelled the population dynamics as Ricker-logistic (Fig.~\ref{fig:alt}).
@@
 Longer time-series length was the strongest covariate of observing black-swan dynamics.
-For example, we were about \pIncHeavyNThirtyNSixty~times more likely to observe heavy tails in a population with 60 time steps than in one with 30 time steps.
+For example, we were about \pIncHeavyNThirtyNSixty~times as likely to observe heavy tails in a population with 60 time steps compared to one with 30 time steps.
 However, the absolute change in probability with increased time series length was small ($\pHeavyNSixty$ vs.\ $\pHeavyNThirty$ in the previous example, Fig.~\ref{fig:correlates}).
 
-We examined all time series with published explanations of why the black-swan events occurred (Tables~\ref{tab:causes-supp} and \ref{tab:sparks}).
+We examined all time series with published explanations of why the black-swan events occurred (Tables~\ref{tab:sparks} and \ref{tab:causes-supp}).
 The majority (\percBSDown \%) of documented events were downward black swans and involved a series of unfortunate events (Table~\ref{tab:causes-supp}).
@@
 We found strong evidence for black swans (heavy-tailed process noise) in 3--5\% of ecological time series.
-Black swans are detected more frequently in longer time series and in populations with a smaller magnitude of process error.
+Black swans are usually downward events and are detected more frequently in longer time series and in populations with a smaller magnitude of process error.
 However, black swans were not associated with density dependence, population growth rate, or lifespan.
@@
 In addition to a possibly biased sample of populations, some black-swan detections could just be recording mistakes, or conversely, some extreme observations may have been discarded or altered if they were erroneously suspected of being recording mistakes.
+Indeed, three of the populations our method initially identified as heavy tailed turned out to be data-entry errors (Supporting Information).
 A final caveat is that the temporal scales of observation and population dynamics vary considerably across populations in the GPDD and this likely influences the detection of heavy tails.

### 2014-11-09 text/ms.tex
Copy editing ms

@@
 
-By allowing the process noise to be drawn from a Student-t distribution we can estimate the degree to which the process errors have heavy tails and are thereof evidence of black-swan events (Fig.~\ref{fig:didactic}a, b).
+By allowing the process noise to be drawn from a Student-t distribution we can estimate the degree to which the process deviations have heavy tails and are thereof evidence of black-swan events (Fig.~\ref{fig:didactic}a, b).
 For example, at $\nu = 3$, the probability of drawing a value more than five standard deviations below the mean is $0.02$, whereas the probability of drawing such a value from a normal distribution is tiny ($2.9\cdot10^{-7}$).
@@
 Longer time-series length was the strongest covariate of observing black-swan dynamics.
-For example, we were about \pIncHeavyNThirtyNSixty~times as likely to observe heavy tails in a population with 60 time steps compared to one with 30 time steps.
+For instance, we were about \pIncHeavyNThirtyNSixty~times as likely to observe heavy tails in a population with 60 time steps compared to one with 30 time steps.
 However, the absolute change in probability with increased time series length was small ($\pHeavyNSixty$ vs.\ $\pHeavyNThirty$ in the previous example, Fig.~\ref{fig:correlates}).
@@
 Other black swans were the result of a sequence of extreme climate events on their own.
-For example, severe winters in 1929, 1940--1942, and 1962--1963 were associated with black-swan downswings in grey heron (\textit{Ardea cinerea}) abundance in the United Kingdom \citep{stafford1971}.
+For instance, severe winters in 1929, 1940--1942, and 1962--1963 were associated with black-swan downswings in grey heron (\textit{Ardea cinerea}) abundance in the United Kingdom \citep{stafford1971}.
 Our analysis finds that the last event was a combination of two black-swan events in a row and it took the population three times longer to recover than predicted \citep{stafford1971}.
@@
 We found strong evidence for black swans (heavy-tailed process noise) in 3--5\% of ecological time series.
-Black swans are usually downward events and are detected more frequently in longer time series and in populations with a smaller magnitude of process error.
-However, black swans were not associated with density dependence, population growth rate, or lifespan.
-In verified cases, black swans were often a result of the interaction between elements of extreme climate, predators and parasite cycles, and strong changes in human pressures.
-Our empirical results, sensitivity analyses, and simulation tests suggest that estimating the tail shape of process noise is a viable method of detecting black-swan population dynamics and if anything will underestimate the probability of black swans.
+Black swans were usually downward events and were detected more frequently in longer time series and in populations with a smaller magnitude of process noise.
+Black swans were not associated with density dependence, population growth rate, or lifespan.
+In verified cases, black-swan events were often a result of the interaction between elements of extreme climate, predators and parasite cycles, and strong changes in human pressures.
+Our empirical results, sensitivity analyses, and simulation tests suggest that estimating the tail shape of process noise is a viable method of detecting black-swan population dynamics and if anything will underestimate the probability of black-swan events.
 The presence of black swans highlights the importance of developing management strategies that detect quickly, respond to, and are robust to extremes in population dynamics --- particularly as the frequency and magnitude of climatic extremes increase over the next century \citep{easterling2000,ipcc2012}.
@@
 Our results clarify previous related analyses.
-Previous work with an older version of the GPDD assessed the distribution of abundance time series but focused on identifying if the log-normal distribution was the most frequently parsimonious model \citep{halley2002}.
+An analysis with an older version of the GPDD assessed the distribution of abundance time series but focused on identifying if the log-normal distribution was the most frequently parsimonious model \citep{halley2002}.
 For heavy-tailed distributions, \citet{halley2002} fit the extremely heavy-tailed Cauchy distribution and the four-parameter Levy stable distribution and found little information criteria support for these distributions in longer time series.
@@
 
-We might expect to observe black-swan dynamics in ecological time series because of either unmodelled intrinsic properties of populations or extrinsic forces acting on populations.
+We might expect to observe black-swan dynamics in ecological time series because of unmodelled intrinsic properties of populations or extrinsic forces acting on populations.
 Since a t-distribution can be formed by a mixture of normal distributions \citep{gelman2014}, we could observe heavy tails if we miss some underlying mixture of intrinsic processes \citep{allen2001}.
-That process might be an aggregation of populations across space, or population diversity, or some intrinsic change in population variability through time.
+That process might be an aggregation of populations across space, or population diversity, or an intrinsic change in population variability through time.
 Extrinsic forces could also cause black-swan dynamics \citep[e.g.][]{nunez2012}.
@@
 The GPDD data represent a taxonomically and geographically biased sample of populations --- the longer time series we focus on are dominated by commercially and recreationally important species and a disproportionate number of populations are located in the United Kingdom.
-Although we expect that we would find qualitatively similar evidence for black swans in many other large taxonomic or geographic samples of populations, the common forces driving those black swans likely differ (e.g.~severe winters in Table~\ref{tab:sparks}).
+Although we would expect to find qualitatively similar evidence for black swans in many other large taxonomic or geographic samples of populations, the common forces driving those black swans would likely differ (e.g.~severe winters in Table~\ref{tab:sparks}).
 In addition to a possibly biased sample of populations, some black-swan detections could just be recording mistakes, or conversely, some extreme observations may have been discarded or altered if they were erroneously suspected of being recording mistakes.
-Indeed, three of the populations our method initially identified as heavy tailed turned out to be data-entry errors (Supporting Information).
+Indeed, three of the populations our method initially identified as heavy tailed turned out to be data-entry errors and were discarded (Supporting Information).
 A final caveat is that the temporal scales of observation and population dynamics vary considerably across populations in the GPDD and this likely influences the detection of heavy tails.
-For example, if we make frequent observations relative to generation time (e.g.~for many large-bodied mammals) we will average across generations and perhaps miss black swans.
+As an example, if we make frequent observations relative to generation time (e.g.~for many large-bodied mammals) we will average across generations and perhaps miss black swans.
 Conversely, if we census populations infrequently relative to generation time (e.g.~many insects in the GPDD) the recorded data may average across extreme and less-extreme events and also dampen black-swan dynamics.
@@
 Similarly to ecological black swans, we can rarely predict the specific location and timing of large earthquakes.
-But, earthquake preparedness involves spatial planning based on forecast probabilities to focus early detection efforts and develop disaster response plans (e.g.~REFERENCE?).
+But, earthquake preparedness involves spatial planning based on forecast probabilities to focus early detection efforts and develop disaster response plans.
 A similar focus might benefit resource management once our ability to predict the spatial probability and covariates of ecological black swans improves.
 The presence of black swans also suggests that we develop management policy that is robust to heavy tails.
-For example, setting target population abundances that are appropriately set back from critical limits may buffer black-swan events \citep[e.g.][]{caddy1996}, and maintaining genetic, phenotypic, and behavioural diversity may allow some components of populations persist when others are affected by disease or extreme environmental forces \citep[e.g.][]{hilborn2003, schindler2010, anderson2014}.
+For instance, setting target population abundances that are appropriately set back from critical limits may buffer black-swan events \citep[e.g.][]{caddy1996}, and maintaining genetic, phenotypic, and behavioural diversity may allow some components of populations persist when others are affected by disease or extreme environmental forces \citep[e.g.][]{hilborn2003, schindler2010, anderson2014}.
 Finally, extreme and unexpected, surprising, or counterintuitive ecological dynamics offer a tremendous opportunity to learn about ecological systems, evaluate when our models break down, and adjust future management policy \citep{doak2008, pine-iii2009, lindenmayer2010}.
 
-Our results suggest important future issues to address about ecological black swans.
+Our results suggest a number of research questions related to ecological black swans.
 Given that black swans do occur, can we forecast the probability of black swans in space and time?

### 2014-11-10 text/ms.tex
Edit edit edit

@@
 
-Black swans are statistically improbable events that nonetheless occur --- often with profound consequences.
+Black swans are statistically improbable events that nonetheless occur---often with profound consequences.
 While extremes in the physical environment, such as monsoons and heat waves, are widely studied and increasing in magnitude and frequency, it remains unclear the extent to which ecological populations buffer or suffer from such extremes.
@@
 Black swans were more frequently detected for populations with longer time series and lower levels of process noise; for shorter and noisier time series, our simulations suggested black-swan dynamics are often not identifiable as such.
-The presence of black swans in population dynamics highlights the importance of developing robust conservation and management strategies --- particularly as the frequency and magnitude of climatic extremes increase over the next century.
+The presence of black swans in population dynamics highlights the importance of developing robust conservation and management strategies---particularly as the frequency and magnitude of climatic extremes increase over the next century.
 
@@
 
-% TODO: cite panels c, d in Fig 1
-
 We fit all models in a Bayesian framework using Stan \citep{stan-manual2014} via the R computing environment \citep{r2014}.
@@
 
-We chose weakly informative priors to incorporate our understanding of plausible population dynamics (\citeauthor{gelman2014} \citeyear{gelman2014}; Supporting Information).
-For $\nu$, we chose an exponential prior with rate parameter of $0.01$ truncated at values above two (Fig.~\ref{fig:priors}) --- a slightly less informative prior than suggested by \citet{fernandez1998}.
+We chose weakly informative priors to incorporate our understanding of plausible population dynamics (\citeauthor{gelman2014} \citeyear{gelman2014}; Figs~\ref{fig:didactic}f--h, \ref{fig:priors}).
+For $\nu$, we chose an exponential prior with rate parameter of $0.01$ truncated at values above two---a slightly less informative prior than suggested by \citet{fernandez1998}.
 This prior gives only a \basePriorProbHeavy \% probability that $\nu < 10$ but constrains the sampling sufficiently to avoid wandering off towards infinity.
@@
 For covariates that were derived from Gompertz model parameter estimates, we incorporated standard deviations around the means.
-To account for broad patterns of phylogenetic relatedness, we allowed for varying intercepts (``random effects'') at the taxonomic class, order, and species level (Supporting Information).
+To account for broad patterns of phylogenetic relatedness, we allowed for hierarchical intercepts at the taxonomic class, order, and species level (Supporting Information).
 We fit our model in Stan with weakly informative priors on the coefficients \citep{gelman2008d} and variance parameters (\citeauthor{gelman2006c} \citeyear{gelman2006c} \citeauthor{gelman2014} \citeyear{gelman2014}; Supporting Information).
@@
 Black swans were taxonomically widespread, occurring in \POrdersHeavy\% of taxonomic orders.
-Accounting for time series length and partially pooling inference across taxonomic class and order with a multilevel model, there was stronger evidence for black swans in insect populations than is visually apparent in Fig.~\ref{fig:nu-coefs} --- four of 10 orders with the highest median probability of heavy tails were insect orders --- however, there was considerable uncertainty in these estimates (Fig.~\ref{fig:posteriors}b).
+Accounting for time series length and partially pooling inference across taxonomic class and order with a multilevel model, there was stronger evidence for black swans in insect populations than is visually apparent in Fig.~\ref{fig:nu-coefs}---four of 10 orders with the highest median probability of heavy tails were insect orders---however, there was considerable uncertainty in these estimates (Fig.~\ref{fig:posteriors}a).
 
@@
 The strength of the prior on $\nu$ had little influence on estimates of black-swan dynamics (Fig.~\ref{fig:alt-priors}).
-Our simulation testing shows that, if anything, our models underpredict the true magnitude and probability of heavy tailed events --- especially given the length of the time series in the GPDD (Figs~\ref{fig:sim-nu}, \ref{fig:sim-prob}).
+Our simulation testing shows that, if anything, our models underpredict the true magnitude and probability of heavy tailed events---especially given the length of the time series in the GPDD (Figs~\ref{fig:sim-nu}, \ref{fig:sim-prob}).
 
-Across populations, the probability of observing black-swan dynamics was positively related to time-series length and negatively related to magnitude of process noise ($\sigma$) but not clearly related to population growth rate ($\lambda$), density dependence ($b$), or maximum lifespan (Figs~\ref{fig:correlates}, \ref{fig:posteriors}a).
+Across populations, the probability of observing black-swan dynamics was positively related to time-series length and negatively related to magnitude of process noise ($\sigma$) but not clearly related to population growth rate ($\lambda$), density dependence ($b$), or maximum lifespan (Figs~\ref{fig:correlates}, \ref{fig:posteriors}b).
 Longer time-series length was the strongest covariate of observing black-swan dynamics.
@@
 We examined all time series with published explanations of why the black-swan events occurred (Tables~\ref{tab:sparks} and \ref{tab:causes-supp}).
-The majority (\percBSDown \%) of documented events were downward black swans and involved a series of unfortunate events (Table~\ref{tab:causes-supp}).
+The majority (\percBSDown \%) of documented events were downward black swans and involved a series of unfortunate events.
 For example, a shortage of nest sites for the shag long-necked bird (\textit{Phalacrocorax aristotelis}) in the United Kingdom reduced population productivity and made the population more vulnerable when a red tide hit in 1968 \citep{potts1980}.
@@
 Our analysis finds that the last event was a combination of two black-swan events in a row and it took the population three times longer to recover than predicted \citep{stafford1971}.
-\textit{(TODO: substitute a parasite example.)}
+\textit{(TODO: substitute a parasite example?)}
 
@@
 Our empirical results, sensitivity analyses, and simulation tests suggest that estimating the tail shape of process noise is a viable method of detecting black-swan population dynamics and if anything will underestimate the probability of black-swan events.
-The presence of black swans highlights the importance of developing management strategies that detect quickly, respond to, and are robust to extremes in population dynamics --- particularly as the frequency and magnitude of climatic extremes increase over the next century \citep{easterling2000,ipcc2012}.
+The presence of black swans highlights the importance of developing management strategies that detect quickly, respond to, and are robust to extremes in population dynamics---particularly as the frequency and magnitude of climatic extremes increase over the next century \citep{easterling2000,ipcc2012}.
 
@@
 There are a number of caveats when considering the generality of our results.
-The GPDD data represent a taxonomically and geographically biased sample of populations --- the longer time series we focus on are dominated by commercially and recreationally important species and a disproportionate number of populations are located in the United Kingdom.
+The GPDD data represent a taxonomically and geographically biased sample of populations---the longer time series we focus on are dominated by commercially and recreationally important species and a disproportionate number of populations are located in the United Kingdom.
 Although we would expect to find qualitatively similar evidence for black swans in many other large taxonomic or geographic samples of populations, the common forces driving those black swans would likely differ (e.g.~severe winters in Table~\ref{tab:sparks}).
@@
 Our results suggest that an added focus on ecological extremes represents the next frontier, particularly in the face of increased climate extremes \citep{meehl2004,ipcc2012}.
-Financial analysts are concerned with the shape of financial-return-distribution tails because these impact estimates of risk --- the probability of a specific magnitude of undesired event occurring \citep{rachev2008}.
+Financial analysts are concerned with the shape of financial-return-distribution tails because these impact estimates of risk---the probability of a specific magnitude of undesired event occurring \citep{rachev2008}.
 A comparable focus in ecology would increase our estimates of extinction risk, since these would be disproportionately impacted by downward black-swan events.
@@
 Circles representing a few sharks, crustaceans, and gastropods are filled in white.
-Median and 90\% credible interval posterior predictions of a beta regression multilevel model are shown in panels a and d where there was a high probability the slope coefficient was different from zero (Supporting Information).
+Median and 90\% credible interval posterior predictions of a beta regression multilevel model are shown in panels a and d where there was a high probability the slope coefficient was different from zero (Fig.~\ref{fig:posteriors}b).
 }
@@
 Posterior probability distributions from beta regression multilevel models.
-(a) Main effect posterior densities for potential covariates of heavy-tailed dynamics.
-The response variable was Pr($\nu < 10$) (approximately the probability of heavy tails).
-The beta regression model was fit on a logit scale with hierarchical intercepts for taxonomic class, taxonomic order, and species.
-All covariates were standardized by subtracting their mean and dividing by twice their standard deviation.
-(b) Taxonomic-order-level posterior densities of Pr($\nu < 10$) after accounting for time-series length.
+(a) Taxonomic-order-level posterior densities of Pr($\nu < 10$) (approximately the probability of heavy tails) after accounting for time-series length.
 Estimates are at the geometric mean of time series length across all the data (approximately 27 time steps).
 Colour shading refers to taxonomic class (yellow: fishes, green: insects, purple: birds, and red: mammals).
-Short vertical line segments indicate median posterior estimates.
-Long dotted vertical line in panel b indicates the median expected Pr($\nu < 10$) from the prior distribution.
+Long dotted vertical line in panel a indicates the median expected Pr($\nu < 10$) from the prior distribution.
+(b) Main effect posterior densities for potential covariates of Pr($\nu < 10$).
+The beta regression models were fit on a logit scale with hierarchical intercepts for taxonomic class, taxonomic order, and species.
+All covariates were standardized by subtracting their mean and dividing by twice their standard deviation.
+In both panels, short vertical line segments indicate median posterior estimates.
 }

### 2014-11-10 text/supp.tex
Edit edit edit

@@
 We chose a less informative rate parameter of $0.01$ and truncated the distribution at $2$, since at $\nu < 2$ the variance of the t distribution is undefined.
-This prior gives only a $7.7$\% probability that $\nu < 10$ but constrains the sampling sufficiently to avoid wandering off towards infinity --- above approximately $\nu = 20$ the t distribution is so similar to the normal distribution (Fig.~\ref{fig:didactic}) that time series of the length considered here are unlikely to be informative about the precise value of $\nu$.
+This prior gives only a $7.7$\% probability that $\nu < 10$ but constrains the sampling sufficiently to avoid wandering off towards infinity---above approximately $\nu = 20$ the t distribution is so similar to the normal distribution (Fig.~\ref{fig:didactic}) that time series of the length considered here are unlikely to be informative about the precise value of $\nu$.
 In the scenario where the data are uninformative about heavy tails (e.g.~Fig.~\ref{fig:didactic}e,~h), the posterior will approximately match the prior (prior median $= 71$, mean $= 102$) and the metrics used in our paper (e.g.~Pr$(\nu < 10) > 0.5$) are unlikely to flag the population as heavy tailed.
@@
 We began with four chains and $2000$ iterations, discarding the first $1000$ as warm up (i.e.~4000 total samples).
-If $\hat{R}$ (the potential scale reduction factor --- a measure of chain convergence) was greater than $1.05$ for any parameter or the minimum effective sample size, $n_\mathrm{eff}$, (a measure of the effective number of uncorrelated samples) for any parameter was less than $200$, we doubled both the total iterations and warm up period and sampled from the model again.
+If $\hat{R}$ (the potential scale reduction factor---a measure of chain convergence) was greater than $1.05$ for any parameter or the minimum effective sample size, $n_\mathrm{eff}$, (a measure of the effective number of uncorrelated samples) for any parameter was less than $200$, we doubled both the total iterations and warm up period and sampled from the model again.
 These thresholds are in excess of the minimums recommended by \citet{gelman2006a} of $\hat{R} < 1.1$ and effective sample size $> 100$ for reliable point estimates and confidence intervals.
@@
 However, simultaneously estimating observation and process noise is a challenging problem (e.g.\ because the observation and process noise parameters tend to negatively covary in model fitting) and is known to sometimes result in identifiability issues with the Gompertz population model \citep{knape2008}.
-Furthermore, our model was applied to hundreds of time series, often of short length (as few as 20 time steps) and our model estimates an additional parameter --- the shape of the process deviation tails --- potentially making identifiability and computational issues even greater.
+Furthermore, our model was applied to hundreds of time series, often of short length (as few as 20 time steps) and our model estimates an additional parameter---the shape of the process deviation tails---potentially making identifiability and computational issues even greater.
 Therefore, we considered a version of the base Gompertz model that allowed for a fixed level of observation error:
@@
 Our first simulation tested the first issue and our second simulation tested the latter.
-In general, our simulations show that, if anything, our model under predicts the magnitude and probability of heavy tailed events --- especially given the length of the time series in the GPDD.
+In general, our simulations show that, if anything, our model under predicts the magnitude and probability of heavy tailed events---especially given the length of the time series in the GPDD.
 
@@
 
-Our results indicate that the Gompertz model can recapture the true value of $\nu$ when the process noise was chosen so that appropriate tail events were present (TODO FIG).
+Our results indicate that the Gompertz model can recapture the true value of $\nu$ when the process noise was chosen so that appropriate tail events were present (Fig.~\ref{fig:sim-prob} upper panels).
 The addition of observation error caused the model to tend to underestimate the degree of heavy-tailedness.
-Fitting a model with correctly specified observation error did not make substantial improvements to model bias.
+Fitting a model with correctly specified observation error did not make substantial improvements to model bias (Fig.~\ref{fig:sim-prob}).
 
@@
 
-When converting the posterior distributions of $\nu$ into Pr($\nu < 10$), the models distinguished heavy and not-heavy tails reasonably well.
+When converting the posterior distributions of $\nu$ into Pr($\nu < 10$), the models distinguished heavy and not-heavy tails reasonably well (Fig.~\ref{fig:sim-prob} lower panels).
 Without observation error, and using a probability of $0.5$ as a threshold, the model correctly classified all simulated systems with normally distributed process noise as not heavy tailed.
-The model would have miscategorized only one of $40$ simulations at $\nu = 5$ across a sample size of $50$ or $100$ data points (Fig.~\ref{fig:sim-prob}, upper panels).
-The model would have correctly categorized all cases where the process noise was not heavy tailed (indicated as ``$\nu =$ infinity'' in Fig.~\ref{fig:sim-prob}) and all cases where $\nu = 3$.
-With $0.2$ standard deviations of observation error, the model still categorized \obsErrorNuFivePerc\% of cases as heavy tailed when $\nu = 5$ and all cases where $\nu = 3$.
-Allowing for observation error made little improvement to the detection of heavy tails (Fig.~\ref{fig:sim-prob}, lower-left vs.\ lower-right panel).
+The model would have miscategorized only one of $40$ simulations at $\nu = 5$ across a sample size of $50$ or $100$ data points (Fig.~\ref{fig:sim-prob}, scenarios 1 and 2 in lower row, second panel from left).
+The model would have correctly categorized all cases where the process noise was not heavy tailed (Fig.~\ref{fig:sim-prob} bottom-right panel) and all cases where $\nu = 3$ and there was not observation error.
+With $0.2$ standard deviations of observation error, the model still categorized \obsErrorNuFivePerc\% of cases as heavy tailed when $\nu = 5$ and all but one case when $\nu = 3$.
+Allowing for observation error made little improvement to the detection of heavy tails.
 Therefore, we chose to focus on the simpler model without observation error in the main text, particularly given that the true magnitude of observation error was unknown in the empirical data.
@@
 
-To derive taxonomic-order-level estimates of the probability of heavy tails accounting for time-series length (Fig.~\ref{fig:posteriors}b), we fit a separate multilevel model with the same structure but with only $\log$ time-series length as a predictor.
+To derive taxonomic-order-level estimates of the probability of heavy tails accounting for time-series length (Fig.~\ref{fig:posteriors}a), we fit a separate multilevel model with the same structure but with only $\log$ time-series length as a predictor.
 (In this case, we did not want to control for intrinsic population characteristics such as density dependence.) Since our predictors were centered by subtracting their mean value, we obtained order-level estimates of the probability of heavy tails at mean log time-series length by adding the posteriors for $\alpha$, $\alpha^\mathrm{class}_j$, and $\alpha^\mathrm{order}_k$.
@@
 The abundances are shown on a log10 vertical axis.
-Throughout this figure, red dots indicate values that were interpolated and red circles indicate values that were recorded as zero but were set to the next lowest observed abundance.
-Numbers before each species name are the GPDD main identification numbers.
+Throughout this figure, red dots indicate values that were interpolated and blue dots indicate values that were recorded as zero but were set to the next lowest observed abundance.
+Numbers before each species name are the GPDD Main ID numbers.
 }
@@
 \begin{center}
-\includegraphics[width=\textwidth]{../analysis/check-sim-box.pdf}
+\includegraphics[width=\textwidth]{../analysis/sim-gompertz-median-dist.pdf}
+\includegraphics[width=\textwidth]{../analysis/sim-gompertz-p10.pdf}
 \caption{
@@
 specified $\nu$).
-Plots show the probability that $\nu < 10$ (i.e.\ the approximate probability
-of heavy tails) across three population $\nu$ values
-(3, 5, and normal) and different scenarios panels: (1) 100 time steps and no
+Upper panels show the distribution of median $\widehat{\nu}$ across 20 simulation runs.
+Lower panels show the distribution of Pr($\nu < 10$) across 20 simulation runs.
+We ran the simulations across three population (``true'') $\nu$ values
+(3, 5, and $1\cdot 10^9$, i.e.\ approximately normal) and four scenarios: (1) 100 time steps and no
 observation error, (2) 50 time steps and no observation error, (3) 50 time
@@
 of observation error was assumed known.
-Within
-each scenario the dots represent stochastic draws from the true population
+Within each scenario the dots represent stochastic draws from the true population
 distributions combined with model fits.

### 2014-11-10 analysis/cause-table.tex
Tidy up the causes in the SOM table

@@
 % latex table generated in R 3.1.2 by xtable 1.7-4 package
-% Fri Nov  7 19:07:39 2014
+% Sun Nov  9 16:06:39 2014
 Time series & Population & ID & citation & Verified & Description & Pr($\nu < 10$ & $\widehat{\nu}$ \\ 
   \midrule
-\includegraphics[width=1.7cm]{../analysis/sparks/6528.pdf} & Shag, \textit{Phalacrocorax aristotelis}, Farne Islands, Northumberland & 6528 & \citep{potts1980} & yes & red tide & 1.00 & 2 (2--4) \\ 
+\includegraphics[width=1.7cm]{../analysis/sparks/6528.pdf} & Shag, \textit{Phalacrocorax aristotelis}, Farne Islands, Northumberland & 6528 & \citep{potts1980} & yes & Red tide event combined with low productivity due to overcrowding & 1.00 & 2 (2--4) \\
   \includegraphics[width=1.7cm]{../analysis/sparks/7115.pdf} & South African fur seal, \textit{Arctocephalus pusillus}, South Africa & 7115 & \citep{shaughnessy1982} & yes?
-& harvesting and predation changes & 1.00 & 2 (2--4) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/10128.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Scotland - un-named area & 10128 & \citep{potts1984} & yes & environmental- and parisitic-caused cycles & 1.00 & 3 (2--4) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/9382.pdf} & Pine looper or Bordered white, \textit{Bupalus piniaria}, Kessock & 9382 & \citep{broekhuizen1993} & yes &  & 1.00 & 3 (2--5) \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/10127.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Scotland - un-named area & 10127 & \citep{potts1984} & yes & environmental- and parisitic-caused cycles & 0.99 & 3 (2--5) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/10007.pdf} & Water vole, \textit{Arvicola terrestris}, Le Pont & 10007 & \citep{saucy1994} & yes & predator-environment cycle interactions & 1.00 & 3 (2--5) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/20579.pdf} & Grey heron, \textit{Ardea cinerea}, Southern Britain & 20579 & \citep{stafford1971} & yes? & severe winter & 0.98 & 3 (2--7) \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/9655.pdf} & Flea beetle, \textit{Chaetocnoma concinna}, Finland & 9655 & \citep{markkula1965} &  &  & 0.99 & 3 (2--6) \\ 
+& Harvesting and predation changes & 1.00 & 2 (2--4) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/10128.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Scotland - un-named area & 10128 & \citep{potts1984} & yes & Environment- and parisite-caused cycles & 1.00 & 3 (2--4) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/9382.pdf} & Pine looper or Bordered white, \textit{Bupalus piniaria}, Kessock & 9382 & \citep{broekhuizen1993} & yes & Unknown, but sampling intensity was decreasing & 1.00 & 3 (2--5) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/10127.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Scotland - un-named area & 10127 & \citep{potts1984} & yes & Environment- and parisite-caused cycles & 0.99 & 3 (2--5) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/10007.pdf} & Water vole, \textit{Arvicola terrestris}, Le Pont & 10007 & \citep{saucy1994} & yes & Predator-environment cycle interactions & 1.00 & 3 (2--5) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/20579.pdf} & Grey heron, \textit{Ardea cinerea}, Southern Britain & 20579 & \citep{stafford1971} & yes? & Severe winter & 0.98 & 3 (2--7) \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/9655.pdf} & Flea beetle, \textit{Chaetocnoma concinna}, Finland & 9655 & \citep{markkula1965} &  & Cannot locate original source & 0.99 & 3 (2--6) \\
   \includegraphics[width=1.7cm]{../analysis/sparks/1235.pdf} & Wren, \textit{Troglodytes troglodytes}, Eastern Wood, Bookham Common & 1235 & \citep{newton1998} & yes?
-& severe winter & 0.98 & 3 (2--8) \\
+& Severe winter & 0.98 & 3 (2--8) \\
   \includegraphics[width=1.7cm]{../analysis/sparks/10113.pdf} & Willow grouse, \textit{Lagopus lagopus}, Northern England & 10113 & \citep{dobson1995} & yes?
-& parasites and predators & 0.99 & 3 (2--6) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/9667.pdf} & Gooseberry sawfly, \textit{Nemastus ribesii}, Finland & 9667 & \citep{markkula1965} &  &  & 0.97 & 3 (2--8) \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/9679.pdf} & Unknown, \textit{Trioza apicalis}, Finland & 9679 & \citep{markkula1965} &  &  & 0.83 & 3 (2--88) \\ 
+& Parasites and predators & 0.99 & 3 (2--6) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/9667.pdf} & Gooseberry sawfly, \textit{Nemastus ribesii}, Finland & 9667 & \citep{markkula1965} &  & Cannot locate original source & 0.97 & 3 (2--8) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/9679.pdf} & Unknown, \textit{Trioza apicalis}, Finland & 9679 & \citep{markkula1965} &  & Cannot locate original source & 0.83 & 3 (2--88) \\ 
   \includegraphics[width=1.7cm]{../analysis/sparks/20527.pdf} & Wandering albatross, \textit{Diomedea exulans}, Taiaroa & 20527 & \citep{robertson1998} & IN LIBRARY & IN LIBRARY & 0.99 & 4 (2--7) \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/10039.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Northern Scotland & 10039 & \citep{dobson1995} & yes & parasites and predators & 0.92 & 4 (2--12) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/10162.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Atholl Estate & 10162 & \citet{mackenzie1952} & yes & overcrowding and environemnt & 0.91 & 5 (2--12) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/9503.pdf} & Fisher or  Pekan, \textit{Martes pennanti}, Manitoba & 9503 & \citep{keith1963} & IN LIBRARY &  & 0.76 & 5 (2--87) \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/7099.pdf} & European rabbit, \textit{Oryctolagus cuniculus}, Estate 2, East Anglia & 7099 & \citep{barnes1986} & yes & disease outbreak followed by years of good weather & 0.78 & 5 (2--67) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/2778.pdf} & Wheatear, \textit{Oenanthe oenanthe}, Skokholm Island & 2778 & \citep{lack1969} & yes & unknown, but decline noted specifically, cold winters later & 0.67 & 5 (2--141) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/9659.pdf} & Cabbage root fly or maggot, \textit{Delia radicum}, Finland & 9659 & \citep{markkula1965} &  &  & 0.59 & 7 (2--161) \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/1195.pdf} & Blue jay, \textit{Cyanocitta cristata}, Robert Allerton Park & 1195 & \citep{kendeigh1982} & yes & unknown & 0.60 & 7 (2--149) \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/5019.pdf} & Barbary macaque, \textit{Macaca sylvanus}, Queens Gate & 5019 & \citep{fa1984} &  &  & 0.61 & 7 (2--132) \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/9953.pdf} & Rock ptarmigan, \textit{Lagopus mutus}, Iceland & 9953 & \citep{clarke1885,williams1954} &  & severe winters & 0.58 & 7 (2--172) \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/6548.pdf} & Lesser-spotted dogfish, \textit{Scyliorhinus caniculus}, North Sea & 6548 & \citep{heessen1996} & yes & unknown & 0.53 & 8 (2--190) \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/20546.pdf} & American red fox, \textit{Vulpes fulva}, Labrador & 20546 & \citep{dancona1954,lindstrom1994} & yes & predator-prey cycles & 0.57 & 9 (3--65) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/10039.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Northern Scotland & 10039 & \citep{dobson1995} & yes & Parasites and predators & 0.92 & 4 (2--12) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/10162.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Atholl Estate & 10162 & \citet{mackenzie1952} & yes & Bad environmental conditions and overcrowding combined to create crashes & 0.91 & 5 (2--12) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/9503.pdf} & Fisher or  Pekan, \textit{Martes pennanti}, Manitoba & 9503 & \citep{keith1963} & IN LIBRARY & IN LIBRARY & 0.76 & 5 (2--87) \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/7099.pdf} & European rabbit, \textit{Oryctolagus cuniculus}, Estate 2, East Anglia & 7099 & \citep{barnes1986} & yes & Disease outbreak followed by years of good weather & 0.78 & 5 (2--67) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/2778.pdf} & Wheatear, \textit{Oenanthe oenanthe}, Skokholm Island & 2778 & \citep{lack1969} & yes & Unknown, but decline noted specifically, cold winters caused some crashes & 0.67 & 5 (2--141) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/9659.pdf} & Cabbage root fly or maggot, \textit{Delia radicum}, Finland & 9659 & \citep{markkula1965} &  & Cannot locate original source & 0.59 & 7 (2--161) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/1195.pdf} & Blue jay, \textit{Cyanocitta cristata}, Robert Allerton Park & 1195 & \citep{kendeigh1982} & yes & Unknown & 0.60 & 7 (2--149) \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/5019.pdf} & Barbary macaque, \textit{Macaca sylvanus}, Queens Gate & 5019 & \citep{fa1984} &  & Cannot locate original source & 0.61 & 7 (2--132) \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/9953.pdf} & Rock ptarmigan, \textit{Lagopus mutus}, Iceland & 9953 & \citep{clarke1885,williams1954} &  & Severe winters & 0.58 & 7 (2--172) \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/6548.pdf} & Lesser-spotted dogfish, \textit{Scyliorhinus caniculus}, North Sea & 6548 & \citep{heessen1996} & yes & Unknown, not specifically mentioned & 0.53 & 8 (2--190) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/20546.pdf} & American red fox, \textit{Vulpes fulva}, Labrador & 20546 & \citep{dancona1954,lindstrom1994} & yes & Predator-prey cycles & 0.57 & 9 (3--65) \\
   \includegraphics[width=1.7cm]{../analysis/sparks/9470.pdf} & Ruffed grouse, \textit{Bonasa umbellus}, Connecticut & 9470 & \citep{keith1963} & IN LIBRARY & IN LIBRARY & 0.53 & 9 (3--158) \\ 

### 2014-11-10 text/supp.tex
Whoops - mistake in equation
Edits to supp.tex

@@
 \texttt{20579} contained additional years of data not present in \texttt{10139}.
-We removed a limited number of populations from class Angiospermopsida and Bacillariophyceae to focus the taxonomy in our analysis.
+We removed a limited number of populations from class Angiospermopsida and Bacillariophyceae to focus the taxonomy in our analysis on animals.
 We also removed any populations with an \texttt{Unknown} taxonomic class.
 
-\item Finally, we removed populations with the following GPDD Main IDs, which we discovered data entry errors when verifying the populations with suspected black swans: \texttt{1207} because 1957 was entered as 2 but should have been 27 \citep{kendeigh1982}, \texttt{6531} because 1978 was entered as 7 should have been 47 \citep{minot1986}, and \texttt{6566} because the data did not match the graph \citep{heessen1996}.
+\item Finally, we removed populations with the following GPDD Main IDs, which we discovered data entry errors when verifying the populations with suspected black swans: \texttt{1207} because the 1957 data point was entered as 2 but should have been 27 \citep{kendeigh1982}, \texttt{6531} because the 1978 data point was entered as 7 should have been 47 \citep{minot1986}, and \texttt{6566} because some of the data did not match the graph \citep{heessen1996}.
 
@@
 \noindent
-We provide a supplemental figure of all the time series included in our analysis and indicate which values were interpolated (\percImputedPops\% of populations had at least one point interpolated also only \percImputedPoints\% of the total observations were interpolated) (Fig.~\ref{fig:all-ts}).
+We provide a supplemental figure of all the time series included in our analysis and indicate which values were interpolated (non-zero interpolations) (\percImputedPops\% of populations had at least one point interpolated but only \percImputedPoints\% of the total observations were interpolated) (Fig.~\ref{fig:all-ts}).
 Note that interpolation is highly unlikely to lead to black-swan detections, since black swans involve extreme increases or decreases.
@@
 \begin{align*}
-\mathrm{Pr}(\nu_i < 0.5) &\sim \mathrm{Beta}(A_i, B_i)\\
+\mathrm{Pr}(\nu_i < 10) &\sim \mathrm{Beta}(A_i, B_i)\\
 \mu_i &= \mathrm{logit}^{-1}(\alpha

### 2014-11-10 text/ms.tex
Edit abstract

@@
 We apply our model to \NPops~time series from around the world across \NOrders~taxonomic orders and seven classes.
-We find strong evidence of black swans, but they are rare occurring in \overallMinPerc--\overallMaxPerc\% of populations: most frequently for birds (\AvesRangePerc\%) followed by mammals (\MammaliaRangePerc\%), insects (\InsectaRangePerc\%), and fishes (\OsteichthyesRangePerc\%).
-Black swans were not explained by any life-history covariates, but tended to be driven by external perturbations such as climate, severe winters, predator or parasite cycles, and the combined effects of multiple factors.
-Black swans were more frequently detected for populations with longer time series and lower levels of process noise; for shorter and noisier time series, our simulations suggested black-swan dynamics are often not identifiable as such.
-The presence of black swans in population dynamics highlights the importance of developing robust conservation and management strategies---particularly as the frequency and magnitude of climatic extremes increase over the next century.
+We find strong evidence of black swans, but they are rare occurring in \overallMinPerc--\overallMaxPerc\% of populations: most frequently for birds (\AvesRangePerc\%) followed by mammals (\MammaliaRangePerc\%), and insects (\InsectaRangePerc\%).
+Black swans were predominantly downward events (\percBSDown \%) and were not explained by any life-history covariates, but tended to be driven by external perturbations such as climate, severe winters, predator or parasite cycles, and the combined effects of multiple factors.
+Extreme events were more frequently detected for populations with longer time series and lower levels of process noise; for shorter and noisier time series, our simulations suggested black-swan dynamics are often not identifiable as such.
+The presence of black swans in population dynamics highlights the importance of developing robust conservation and management strategies---particularly as the frequency and magnitude of climate extremes increase over the next century.
 

### 2014-11-18 text/ms.tex
Edit a comment
Lots of edits and wrap text

@@
 
-Black swans are statistically improbable events that nonetheless occur---often with profound consequences.
-While extremes in the physical environment, such as monsoons and heat waves, are widely studied and increasing in magnitude and frequency, it remains unclear the extent to which ecological populations buffer or suffer from such extremes.
-Here, we develop a probability model to estimate the degree of heavy-tailedness (presence of black swans) in ecological process noise.
-We apply our model to \NPops~time series from around the world across \NOrders~taxonomic orders and seven classes.
-We find strong evidence of black swans, but they are rare occurring in \overallMinPerc--\overallMaxPerc\% of populations: most frequently for birds (\AvesRangePerc\%) followed by mammals (\MammaliaRangePerc\%), and insects (\InsectaRangePerc\%).
-Black swans were predominantly downward events (\percBSDown \%) and were not explained by any life-history covariates, but tended to be driven by external perturbations such as climate, severe winters, predator or parasite cycles, and the combined effects of multiple factors.
-Extreme events were more frequently detected for populations with longer time series and lower levels of process noise; for shorter and noisier time series, our simulations suggested black-swan dynamics are often not identifiable as such.
-The presence of black swans in population dynamics highlights the importance of developing robust conservation and management strategies---particularly as the frequency and magnitude of climate extremes increase over the next century.
+Black swans are statistically improbable events that nonetheless occur---often
+with profound consequences. While extremes in the physical environment, such
+as monsoons and heat waves, are widely studied and increasing in magnitude and
+frequency, it remains unclear the extent to which ecological populations
+buffer or suffer from such extremes. Here, we develop a probability model to
+estimate the degree of heavy-tailedness (presence of black swans) in
+ecological process noise. We apply our model to \NPops~time series from around
+the world across \NOrders~taxonomic orders and seven classes. We find strong
+evidence of black swans, but they are rare occurring in
+\overallMinPerc--\overallMaxPerc\% of populations: most frequently for birds
+(\AvesRangePerc\%) followed by mammals (\MammaliaRangePerc\%), and insects
+(\InsectaRangePerc\%). Black swans were predominantly (\percBSDown \%)
+downward events and were not explained by any life-history covariates, but
+tended to be driven by external perturbations such as climate, severe winters,
+predator or parasite cycles, and the combined effects of multiple factors.
+Extreme events were more frequently detected for populations with longer time
+series and lower levels of process noise; for shorter and noisier time series,
+our simulations suggested black-swan dynamics are often not identifiable as
+such. The presence of black swans in population dynamics highlights the
+importance of developing robust conservation and management
+strategies---particularly as the frequency and magnitude of climate extremes
+increase over the next century.
+
+% TODO highlight that skew would *increase* the probability of higher
+% extinction risk
 
@@
 
-Black swans are unexpected extreme events with potentially dramatic consequences \citep{taleb2007,sornette2009}.
-One of the most striking black swans in ecology is the possible asteroid collision marking the end-Cretaceous mass extinction 65 million years ago \citep{alvarez1980,harnik2012}.
-Today, climate extremes, in concert with shifts in mean temperature, are expected to cause the greatest ecological and societal damage \citep{ipcc2012}.
-But, while extremes in the physical environment such as wave height, storm severity, and temperature are frequent experiences \citep{gaines1993,katz2005}, it remains unclear the extent to which ecological systems buffer or suffer from black swans \citep{nunez2012}.
+Black swans are unexpected extreme events with potentially dramatic
+consequences \citep{taleb2007,sornette2009}. One of the most striking black
+swans in ecology is the possible asteroid collision marking the end-Cretaceous
+mass extinction 65 million years ago \citep{alvarez1980,harnik2012}. Today,
+climate extremes, in concert with shifts in mean temperature, are expected to
+cause the greatest ecological and societal damage \citep{ipcc2012}. But, while
+extremes in the physical environment such as wave height, storm severity, and
+temperature are frequent events \citep{gaines1993,katz2005}, it remains
+unclear the extent to which ecological systems buffer or suffer from black
+swans \citep{nunez2012}.
 
-There is compelling anecdotal evidence for ecological black swans, but systematic evidence across taxa has been elusive.
-A survey of ecologists indicated that surprising outcomes of field experiments are far more common than we assume \citep{doak2008}, and events such as the global invasion of  Argentine ants and the mutation of viruses to infect new hosts could be considered black swans \citep{nunez2012}.
-In fact, anecdotes of population catastrophes are numerous and catastrophes may be the most important element affecting population persistence \citep{mangel1994}.
-For marine mammal populations, we have compelling evidence of catastrophes \citep{gerber2001, ward2007}, and recently, heavy-tailed distributions have been discovered in time-series of marine microbe abundance \citep{segura2013} and time to extinction for experimental waterflea populations \citep{drake2014}.
-As far as we can tell there have been two systematic surveys for ecological black swans: one with North American breeding birds and the other with the Global Population Dynamics Database, which uncovered little clear evidence for them \citep{keitt1998,allen2001,halley2002}.
-Nevertheless there are methodological challenges to their detection \citep{allen2001,ward2007}.
+There is compelling anecdotal evidence for ecological black swans, but
+systematic evidence across taxa has been elusive. A survey of ecologists
+indicated that surprising outcomes of field experiments are far more common
+than we assume \citep{doak2008}, and events such as the global invasion of
+Argentine ants and the mutation of viruses to infect new hosts could be
+considered black swans \citep{nunez2012}. In fact, anecdotes of population
+catastrophes are numerous and catastrophes may be the most important element
+affecting population persistence \citep{mangel1994}. For marine mammal
+populations, we have compelling evidence of catastrophes \citep{gerber2001,
+  ward2007}, and recently, heavy-tailed distributions have been discovered in
+time-series of marine microbe abundance \citep{segura2013} and time to
+extinction for experimental waterflea populations \citep{drake2014}. Despite
+these anecotes, as far as we can tell there have been two systematic surveys
+for ecological black swans: one with North American breeding birds and the
+other with the Global Population Dynamics Database, which uncovered little
+clear evidence for them \citep{keitt1998,allen2001,halley2002}. Nevertheless
+there are methodological challenges to their detection
+\citep{allen2001,ward2007}.
 
-There are two key reasons why we may find little evidence of ecological black swans.
-First, they might not exist in higher taxa.
-Indeed, the majority of model fitting and risk forecasting assumes that population dynamics are not heavy tailed \citep[e.g.][]{brook2006a,dennis2006,knape2012}.
-Alternatively, heavy-tailed dynamics might exist, but our ability to detect them requires further development of statistical tools.
-One such tool is the generalized extreme value distribution, which has been applied to environmental data \citep[e.g.][]{katz2005}.
-This distribution describes the most extreme event per time interval (e.g.~heaviest rainfall per year), but requires time series that are sufficiently long to be condensed into time intervals.
-Another statistical tool involves fitting a catastrophic mixture distribution in a state space model to quantify the probability that population events are extreme, although this is also data intensive \citep{ward2007}.
-A third tool is to compare the support for fits of thin- and heavy-tailed distributions \citep{halley2002}, but this analysis did not quantify the probability of black swans or allow for population dynamics.
+There are two key reasons why we may find little evidence of ecological black
+swans. First, they might not exist in higher taxa. Indeed, the majority of
+model fitting and risk forecasting assumes that population dynamics are not
+heavy tailed \citep[e.g.][]{brook2006a,dennis2006,knape2012}. Alternatively,
+heavy-tailed dynamics might exist, but our ability to detect them requires
+further development of statistical tools. One such tool is the generalized
+extreme value distribution, which has been applied to environmental data
+\citep[e.g.][]{katz2005}. This distribution describes the most extreme event
+per time interval (e.g.~heaviest rainfall per year), but requires time series
+that are sufficiently long to be condensed into time intervals. Another
+statistical tool involves fitting a catastrophic mixture distribution in a
+state space model to quantify the probability that population events are
+extreme, although this is also data intensive \citep{ward2007}. A third tool is
+to compare the support for fits of thin- and heavy-tailed distributions
+\citep{halley2002}, but this analysis did not quantify the probability of black
+swans or allow for population dynamics.
 
-Here we develop a framework for identifying ecological black swans in the process noise of population dynamics, i.e.\ the stochastic jumps in abundance from one time step to the next.
-Our framework quantifies both the magnitude and probability of heavy-tailed dynamics, allows for a range of population dynamic models, can incorporate observation uncertainty, and can be easily applied to abundance time series.
-We apply our model to \NPops\ populations and estimate the frequency and strength of black-swan events in population dynamics across taxonomy; verify some causes of black-swan events; and attempt to identify characteristics of time series or intrinsic life-history characteristics that are associated with heavy-tailed dynamics.
+Here we develop a framework for identifying ecological black swans in the
+process noise of population dynamics, i.e.\ the stochastic jumps in abundance
+from one time step to the next. Our framework quantifies both the magnitude and
+probability of heavy-tailed dynamics, allows for a range of population dynamic
+models, can incorporate observation uncertainty, and can be easily applied to
+abundance time series. We apply our model to \NPops\ populations and estimate
+the frequency and strength of black-swan events in population dynamics across
+taxonomy; verify some causes of black-swan events; and attempt to identify
+characteristics of time series or intrinsic life-history characteristics that
+are associated with heavy-tailed dynamics.
 
@@
 
-To obtain estimates of the probability and magnitude of black-swan events, we fit population dynamic models to abundance time series from around the world.
-For each population, we estimated the shape of the process noise tails by measuring the degrees of freedom ($\nu$) of a Student t-distribution.
+To obtain estimates of the probability and magnitude of black-swan events, we
+fit population dynamic models to abundance time series from around the world.
+For each population, we estimated the shape of the process noise tails by
+measuring the degrees of freedom ($\nu$) of a Student t-distribution.
 
@@
 
-We selected abundance time series from the Global Population Dynamics Database (GPDD; \citeauthor{gpdd2010} \citeyear{gpdd2010}), which contains nearly 5000 time series of abundance from $\sim$1000 species and $\sim$100 taxonomic orders.
-We filtered the data (Supporting Information) to remove populations from less reliable data sources, and those without sufficient data for our models, and then interpolated some missing values \citep[\textit{sensu}][]{brook2006a}.
-Our interpolation affected only $\sim$\interpPointsPerc \% of the final data points (Table~\ref{tab:stats}) and none of the data points that were later considered black-swan events.
-Our final dataset contained \NPops~populations across \NOrders~taxonomic orders and seven taxonomic classes, with a median of \medianTimeSteps~time steps (range of \minTimeSteps--\maxTimeSteps) (Table~\ref{tab:stats}, Fig.~\ref{fig:all-ts}).
+We selected abundance time series from the Global Population Dynamics Database
+(GPDD; \citeauthor{gpdd2010} \citeyear{gpdd2010}), which contains nearly 5000
+time series of abundance from $\sim$1000 species and $\sim$100 taxonomic
+orders. We filtered the data (Supporting Information) to remove populations
+from less reliable data sources, and those without sufficient data for our
+models, and then interpolated some missing values
+\citep[\textit{sensu}][]{brook2006a}. Our interpolation affected only
+$\sim$\interpPointsPerc \% of the final data points (Table~\ref{tab:stats})
+and none of the data points that were later considered black-swan events. Our
+final dataset contained \NPops~populations across \NOrders~taxonomic orders
+and seven taxonomic classes, with a median of \medianTimeSteps~time steps
+(range of \minTimeSteps--\maxTimeSteps) (Table~\ref{tab:stats},
+Fig.~\ref{fig:all-ts}).
 
@@
 
-Our main analysis focuses on the commonly applied Gompertz population dynamics model \citep[e.g.][]{knape2012,dennis2014,connors2014}.
-The Gompertz model represents population growth as a linear function in $\log$ space.
-If we let $x_t$ represent the $\log$ abundance ($N$) at time $t$, we can represent the Gompertz model as:
+Our main analysis focuses on the commonly applied Gompertz population dynamics
+model \citep[e.g.][]{knape2012,dennis2014,connors2014}. The Gompertz model
+represents population growth as a linear function in $\log$ space. If we let
+$x_t$ represent the $\log$ abundance ($N$) at time $t$, we can represent the
+Gompertz model as:
 \begin{align*}
@@
 \end{align*}
-The growth parameter $\lambda$ represents the expected growth rate if $N_t = 1$.
-The model is density independent if $b = 1$, maximally density dependent if $b = 0$, and inversely density dependent if $b < 0$.
-Usually, the process noise $\epsilon_t$ is modelled as normally distributed, but in our paper we assume it is drawn from a t distribution with scale parameter $\sigma$ and degrees of freedom $\nu$.
-In previous analyses of the GPDD, the Gompertz is most often identified as the most parsimonious population model that can be fit to these data \citep{brook2006}.
+The growth parameter $\lambda$ represents the expected growth rate if $N_t =
+1$. The model is density independent if $b = 1$, maximally density dependent if
+$b = 0$, and inversely density dependent if $b < 0$. Usually, the process noise
+$\epsilon_t$ is modelled as normally distributed, but in our paper we assume it
+is drawn from a t distribution with scale parameter $\sigma$ and degrees of
+freedom $\nu$. In previous analyses of the GPDD, the Gompertz was most often
+identified as the most parsimonious population model that can be fit to these
+data \citep{brook2006}.
 
-By allowing the process noise to be drawn from a Student-t distribution we can estimate the degree to which the process deviations have heavy tails and are thereof evidence of black-swan events (Fig.~\ref{fig:didactic}a, b).
-For example, at $\nu = 3$, the probability of drawing a value more than five standard deviations below the mean is $0.02$, whereas the probability of drawing such a value from a normal distribution is tiny ($2.9\cdot10^{-7}$).
-As the value of $\nu$ approaches infinity, the distribution approaches the normal distribution (Fig.~\ref{fig:didactic}a, b).
+By allowing the process noise to be drawn from a Student-t distribution we can
+estimate the degree to which the process deviations have heavy tails and are
+thereof evidence of black-swan events (Fig.~\ref{fig:didactic}a, b). For
+example, at $\nu = 3$, the probability of drawing a value more than five
+standard deviations below the mean is $0.02$, whereas the probability of
+drawing such a value from a normal distribution is tiny ($2.9\cdot10^{-7}$). As
+the value of $\nu$ approaches infinity, the distribution approaches the normal
+distribution (Fig.~\ref{fig:didactic}a, b).
 
-We fit all models in a Bayesian framework using Stan \citep{stan-manual2014} via the R computing environment \citep{r2014}.
-Stan samples from the posterior distribution with an adaptive version of Hamiltonian Markov chain Monte Carlo called the No-U-Turn Sampler and generally obtains less correlated samples than algorithms such as the Gibbs sampler \citep{hoffman2014}.
-We tested to ensure that the chains had sufficiently converged and that the sampler had obtained sufficient independent samples from the posterior ($\widehat{R} < 1.05$, $n_\mathrm{eff} > 200$; Supporting Information).
+We fit all models in a Bayesian framework using Stan \citep{stan-manual2014}
+via the R computing environment \citep{r2014}. Stan samples from the posterior
+distribution with an adaptive version of Hamiltonian Markov chain Monte Carlo
+called the No-U-Turn Sampler and generally obtains less correlated samples
+than algorithms such as the Gibbs sampler \citep{hoffman2014}. We tested to
+ensure that the chains had sufficiently converged and that the sampler had
+obtained sufficient independent samples from the posterior ($\widehat{R} <
+1.05$, $n_\mathrm{eff} > 200$; Supporting Information).
 
-We chose weakly informative priors to incorporate our understanding of plausible population dynamics (\citeauthor{gelman2014} \citeyear{gelman2014}; Figs~\ref{fig:didactic}f--h, \ref{fig:priors}).
-For $\nu$, we chose an exponential prior with rate parameter of $0.01$ truncated at values above two---a slightly less informative prior than suggested by \citet{fernandez1998}.
-This prior gives only a \basePriorProbHeavy \% probability that $\nu < 10$ but constrains the sampling sufficiently to avoid wandering off towards infinity.
-In any case, for $\nu > 20$ the t distribution is almost indistinguishable from the normal distribution (Fig.~\ref{fig:didactic}).
-We investigated the sensitivity of our results to weaker and stronger priors (exponential rate parameter $= 0.005, 0.02$; Fig.~\ref{fig:priors}, Supporting Information).
-We used simulated data to test how easily we could detect $\nu$ given different sample sizes and to ensure we could recover unbiased parameter estimates from the Gompertz model (Supporting Information).
+We chose weakly informative priors to incorporate our understanding of
+plausible population dynamics (\citeauthor{gelman2014} \citeyear{gelman2014};
+Figs~\ref{fig:didactic}f--h, \ref{fig:priors}). For $\nu$, we chose an
+exponential prior with rate parameter of $0.01$ truncated at values above
+two---a slightly less informative prior than suggested by
+\citet{fernandez1998}. This prior gives only a \basePriorProbHeavy \%
+probability that $\nu < 10$ but constrains the sampling sufficiently to avoid
+wandering off towards infinity. In any case, for $\nu > 20$ the t distribution
+is almost indistinguishable from the normal distribution
+(Fig.~\ref{fig:didactic}). We investigated the sensitivity of our results to
+weaker and stronger priors (exponential rate parameter $= 0.005, 0.02$;
+Fig.~\ref{fig:priors}, Supporting Information). We used simulated data to test
+how easily we could detect $\nu$ given different sample sizes and to ensure we
+could recover unbiased parameter estimates from the Gompertz model (Supporting
+Information).
 
-We fit alternative population models to test if four key phenomena systematically changed our conclusions.
-Autocorrelation has been suggested as a reason for increased observed variability of abundance time series through time, which could create apparent heavy tails \citep{inchausti2002}; therefore, we fit a model that included serial correlation in the residuals.
-Additionally, previous work has modelled abundance or growth rates without accounting for density dependence \citep{halley2002,segura2013}; therefore, we fit a simpler model in which we assumed density independence.
-Third, observation error could bias parameter estimates \citep{knape2012} or mask our ability to detect heavy tails \citep{ward2007}; therefore, we fit a model where we allowed for a fixed quantity of observation error ($0.2$ standard deviations on a log scale).
-Finally, the Gompertz model assumes that population growth rate declines linearly with log abundance.
-We also fit an alternative model, the Ricker-logistic model, which assumes that population growth rate declines linearly with abundance itself (Supporting Information).
+We fit alternative population models to test if four key phenomena
+systematically changed our conclusions. Autocorrelation has been suggested as
+a reason for increased observed variability of abundance time series through
+time, which could create apparent heavy tails \citep{inchausti2002};
+therefore, we fit a model that included serial correlation in the residuals.
+Additionally, previous work has modelled abundance or growth rates without
+accounting for density dependence \citep{halley2002,segura2013}; therefore, we
+fit a simpler model in which we assumed density independence. Third,
+observation error could bias parameter estimates \citep{knape2012} or mask our
+ability to detect heavy tails \citep{ward2007}; therefore, we fit a model
+where we allowed for a fixed quantity of observation error ($0.2$ standard
+deviations on a log scale). Finally, the Gompertz model assumes that
+population growth rate declines linearly with log abundance. We also fit an
+alternative model, the Ricker-logistic model, which assumes that population
+growth rate declines linearly with abundance itself (Supporting Information).
 
@@
 
-We investigated possible covariates of heavy-tailed population dynamics visually and through multilevel modelling.
-We plotted characteristics of the time series ($\sigma$, $\lambda$, $b$, and time-series length) along with two life-history characteristics (body length and maximum lifespan obtained from \citet{brook2006a}) against our estimated probability of heavy tails, Pr$(\nu < 10) > 0.5$.
-We formally investigated these relationships by fitting beta regression multilevel models \citep{ferrari2004}.
-For covariates that were derived from Gompertz model parameter estimates, we incorporated standard deviations around the means.
-To account for broad patterns of phylogenetic relatedness, we allowed for hierarchical intercepts at the taxonomic class, order, and species level (Supporting Information).
-We fit our model in Stan with weakly informative priors on the coefficients \citep{gelman2008d} and variance parameters (\citeauthor{gelman2006c} \citeyear{gelman2006c} \citeauthor{gelman2014} \citeyear{gelman2014}; Supporting Information).
+We investigated possible covariates of heavy-tailed population dynamics
+visually and through multilevel modelling. We plotted characteristics of the
+time series ($\sigma$, $\lambda$, $b$, and time-series length) along with two
+life-history characteristics (body length and maximum lifespan obtained from
+\citet{brook2006a}) against our estimated probability of heavy tails, Pr$(\nu <
+10)$. We formally investigated these relationships by fitting beta
+regression multilevel models \citep{ferrari2004}. Beta regression is useful
+for continuous response data that range between zero and one (e.g.\
+probabilities); we fit our models with a logit link as is common for logistic
+regression \citep{ferrari2004}. We incorporated standard deviations around the
+means for covariates that were derived from Gompertz model parameter
+estimates. To account for broad patterns of phylogenetic relatedness, we
+allowed for hierarchical intercepts at the taxonomic class, order, and species
+level (Supporting Information). We fit our model in Stan with weakly
+informative priors on the coefficients \citep{gelman2008d} and variance
+parameters (\citeauthor{gelman2006c} \citeyear{gelman2006c}
+\citeauthor{gelman2014} \citeyear{gelman2014}; Supporting Information).
 
-Finally, we investigated a sample of populations that our method categorized as having a high probability of heavy tails (Pr$(\nu < 10) > 0.5$).
-Where possible, we found the documented causes of ecological black swans in the primary data source cited in the GPDD or in other literature describing the population.
+Finally, we investigated a sample of populations that our method categorized as
+having a high probability of heavy tails (Pr$(\nu < 10) > 0.5$). Where
+possible, we found the documented causes of ecological black swans in the
+primary data source cited in the GPDD or in other literature describing the
+population.
 
@@
 
-We found strong, but rare, evidence for black-swan population dynamics.
-By defining black-swan dynamics as a greater than $0.5$ probability that $\nu < 10$, our main Gompertz model found evidence for heavy tails most frequently for birds (\birdPH\%) followed by mammals (\mammalsPH\%), insects (\insectsPH\%), and fishes (\fishPH\%) (Fig.~\ref{fig:nu-coefs}, Table~\ref{tab:causes-supp}).
-Black swans were taxonomically widespread, occurring in \POrdersHeavy\% of taxonomic orders.
-Accounting for time series length and partially pooling inference across taxonomic class and order with a multilevel model, there was stronger evidence for black swans in insect populations than is visually apparent in Fig.~\ref{fig:nu-coefs}---four of 10 orders with the highest median probability of heavy tails were insect orders---however, there was considerable uncertainty in these estimates (Fig.~\ref{fig:posteriors}a).
+We found strong, but rare, evidence for black-swan population dynamics. By
+defining black-swan dynamics as a greater than $0.5$ probability that $\nu <
+10$, our main Gompertz model found evidence for heavy tails most frequently
+for birds (\birdPH\%) followed by mammals (\mammalsPH\%), and insects
+(\insectsPH\%) (Fig.~\ref{fig:nu-coefs}, Table~\ref{tab:causes-supp}). Black
+swans were taxonomically widespread, occurring in \POrdersHeavy\% of taxonomic
+orders. Accounting for time series length and partially pooling inference
+across taxonomic class and order with a multilevel model, there was stronger
+evidence for black swans in insect populations than is visually apparent in
+Fig.~\ref{fig:nu-coefs}---four of 10 orders with the highest median
+probability of heavy tails were insect orders---however, there was
+considerable uncertainty in these estimates (Fig.~\ref{fig:posteriors}a).
 
-The majority of our heavy-tailed estimates were robust to alternative population models, observation error, and choice of priors.
-Our conclusions were not systematically altered when we included an autocorrelation structure in the residuals, modelled population growth rates without density dependence, or modelled the population dynamics as Ricker-logistic (Fig.~\ref{fig:alt}).
-However, setting observation error standard deviation to $0.2$ increased the median estimate of $\nu$ from $<10$ to $\ge 10$ in $\baseNuTenObsTenSwitch$ of $\baseNuTen$ populations, although the majority of $\nu$ estimates remained qualitatively similar (Fig.~\ref{fig:alt}).
-The strength of the prior on $\nu$ had little influence on estimates of black-swan dynamics (Fig.~\ref{fig:alt-priors}).
-Our simulation testing shows that, if anything, our models underpredict the true magnitude and probability of heavy tailed events---especially given the length of the time series in the GPDD (Figs~\ref{fig:sim-nu}, \ref{fig:sim-prob}).
+The majority of our heavy-tailed estimates were robust to alternative
+population models, observation error, and choice of priors. Our conclusions
+were not systematically altered when we included an autocorrelation structure
+in the residuals, modelled population growth rates without density dependence,
+or modelled the population dynamics as Ricker-logistic (Fig.~\ref{fig:alt}).
+However, setting observation error standard deviation to $0.2$ increased the
+median estimate of $\nu$ from $<10$ to $\ge 10$ in $\baseNuTenObsTenSwitch$ of
+$\baseNuTen$ populations, although the majority of $\nu$ estimates remained
+qualitatively similar (Fig.~\ref{fig:alt}). The strength of the prior on $\nu$
+had little influence on estimates of black-swan dynamics
+(Fig.~\ref{fig:alt-priors}). Our simulation testing shows that, if anything,
+our models underpredict the true magnitude and probability of heavy tailed
+events---especially given the length of the time series in the GPDD
+(Figs~\ref{fig:sim-nu}, \ref{fig:sim-prob}).
 
-Across populations, the probability of observing black-swan dynamics was positively related to time-series length and negatively related to magnitude of process noise ($\sigma$) but not clearly related to population growth rate ($\lambda$), density dependence ($b$), or maximum lifespan (Figs~\ref{fig:correlates}, \ref{fig:posteriors}b).
-Longer time-series length was the strongest covariate of observing black-swan dynamics.
-For instance, we were about \pIncHeavyNThirtyNSixty~times as likely to observe heavy tails in a population with 60 time steps compared to one with 30 time steps.
-However, the absolute change in probability with increased time series length was small ($\pHeavyNSixty$ vs.\ $\pHeavyNThirty$ in the previous example, Fig.~\ref{fig:correlates}).
+Across populations, the probability of observing black-swan dynamics was
+positively related to time-series length and negatively related to magnitude of
+process noise ($\sigma$) but not clearly related to population growth rate
+($\lambda$), density dependence ($b$), or maximum lifespan
+(Figs~\ref{fig:correlates}, \ref{fig:posteriors}b). Longer time-series length
+was the strongest covariate of observing black-swan dynamics. For instance, we
+were about \pIncHeavyNThirtyNSixty~times as likely to observe heavy tails in a
+population with 60 time steps compared to one with 30 time steps. However, the
+absolute change in probability with increased time series length was small
+($\pHeavyNSixty$ vs.\ $\pHeavyNThirty$ in the previous example,
+Fig.~\ref{fig:correlates}).
 
-We examined all time series with published explanations of why the black-swan events occurred (Tables~\ref{tab:sparks} and \ref{tab:causes-supp}).
-The majority (\percBSDown \%) of documented events were downward black swans and involved a series of unfortunate events.
-For example, a shortage of nest sites for the shag long-necked bird (\textit{Phalacrocorax aristotelis}) in the United Kingdom reduced population productivity and made the population more vulnerable when a red tide hit in 1968 \citep{potts1980}.
-The population experienced a black-swan downswing, but the new availability of nest sites resulted in a rapid population upswing \citep{potts1980}.
-As another example, an interaction between environmental- and predation-mediated population cycles are thought to have caused a downward black-swan event for a water vole (\textit{Arvicola terrestris}) population \citep{saucy1994}.
-Other black swans were the result of a sequence of extreme climate events on their own.
-For instance, severe winters in 1929, 1940--1942, and 1962--1963 were associated with black-swan downswings in grey heron (\textit{Ardea cinerea}) abundance in the United Kingdom \citep{stafford1971}.
-Our analysis finds that the last event was a combination of two black-swan events in a row and it took the population three times longer to recover than predicted \citep{stafford1971}.
+We examined all time series with published explanations of why the black-swan
+events occurred (Tables~\ref{tab:sparks} and \ref{tab:causes-supp}). The
+majority (\percBSDown \%) of documented events were downward black swans and
+involved a series of unfortunate events. For example, a shortage of nest sites
+for the shag long-necked bird (\textit{Phalacrocorax aristotelis}) in the
+United Kingdom reduced population productivity and made the population more
+vulnerable when a red tide hit in 1968 \citep{potts1980}. The population
+experienced a black-swan downswing, but the new availability of nest sites
+resulted in a rapid population upswing \citep{potts1980}. As another example,
+an interaction between environmental- and predation-mediated population cycles
+are thought to have caused a downward black-swan event for a water vole
+(\textit{Arvicola terrestris}) population \citep{saucy1994}. Other black swans
+were the result of a sequence of extreme climate events on their own. For
+instance, severe winters in 1929, 1940--1942, and 1962--1963 were associated
+with black-swan downswings in grey heron (\textit{Ardea cinerea}) abundance in
+the United Kingdom \citep{stafford1971}. Our analysis finds that the last event
+was a combination of two black-swan events in a row and it took the population
+three times longer to recover than predicted \citep{stafford1971}.
 \textit{(TODO: substitute a parasite example?)}
-
-%TODO IMPORTANT quantify across all identified black swan events how many were downward and how many upward.
-The assumption of your model is a symmetric distribution, but if in fact downward events are more common, perhaps a skewed distribution would be more realistic.
-Furthermore, if there is skew this would *increase* the probability of higher extinction risk.
-Something to definitely highlight in the results and abstract.
 
@@
 
-We found strong evidence for black swans (heavy-tailed process noise) in 3--5\% of ecological time series.
-Black swans were usually downward events and were detected more frequently in longer time series and in populations with a smaller magnitude of process noise.
-Black swans were not associated with density dependence, population growth rate, or lifespan.
-In verified cases, black-swan events were often a result of the interaction between elements of extreme climate, predators and parasite cycles, and strong changes in human pressures.
-Our empirical results, sensitivity analyses, and simulation tests suggest that estimating the tail shape of process noise is a viable method of detecting black-swan population dynamics and if anything will underestimate the probability of black-swan events.
-The presence of black swans highlights the importance of developing management strategies that detect quickly, respond to, and are robust to extremes in population dynamics---particularly as the frequency and magnitude of climatic extremes increase over the next century \citep{easterling2000,ipcc2012}.
+We found strong evidence for black swans (heavy-tailed process noise) in 3--5\%
+of ecological time series. Black swans were usually downward events and were
+detected more frequently in longer time series and in populations with a
+smaller magnitude of process noise. Black swans were not associated with
+density dependence, population growth rate, or lifespan. In verified cases,
+black-swan events were often a result of the interaction between elements of
+extreme climate, predators and parasite cycles, and strong changes in human
+pressures. Our empirical results, sensitivity analyses, and simulation tests
+suggest that estimating the tail shape of process noise is a viable method of
+detecting black-swan population dynamics and if anything will underestimate the
+probability of black-swan events. The presence of black swans highlights the
+importance of developing management strategies that detect quickly, respond to,
+and are robust to extremes in population dynamics---particularly as the
+frequency and magnitude of climatic extremes increase over the next century
+\citep{easterling2000,ipcc2012}.
 
-Our results clarify previous related analyses.
-An analysis with an older version of the GPDD assessed the distribution of abundance time series but focused on identifying if the log-normal distribution was the most frequently parsimonious model \citep{halley2002}.
-For heavy-tailed distributions, \citet{halley2002} fit the extremely heavy-tailed Cauchy distribution and the four-parameter Levy stable distribution and found little information criteria support for these distributions in longer time series.
-However, black-swan events are by definition rare, and the majority of time series in the GPDD are short for these purposes.
-Therefore, we would not expect to observe black swans in a large proportion of populations.
-By probabilistically quantifying process noise tails on a continuous spectrum, and allowing for non-stationary time series and density dependence, our analysis allows for a more nuanced description of the evidence of ecological black swans.
-In an earlier study, \citet{keitt1998} described heavy (power law) tails in breeding bird population abundance.
-However, this finding was challenged by \citet{allen2001}, who showed that mixing data across species could falsely generate heavy tails.
-This rebuttal supports our use of the t distribution to represent heavy tails, since a t distribution can be represented by a mixture of normal distributions \citep[with the same mean and inverse-gamma-distributed variances,][]{gelman2014}.
+Our results clarify previous related analyses. An analysis with an older
+version of the GPDD assessed the distribution of abundance time series but
+focused on identifying if the log-normal distribution was the most frequently
+parsimonious model \citep{halley2002}. For heavy-tailed distributions,
+\citet{halley2002} fit the extremely heavy-tailed Cauchy distribution and the
+four-parameter Levy stable distribution and found little information criteria
+support for these distributions in longer time series. However, black-swan
+events are by definition rare, and the majority of time series in the GPDD are
+short for these purposes. Therefore, we would not expect to observe black
+swans in a large proportion of populations. By quantifying the probability of
+heavy-tails and allowing for non-stationary time series and density
+dependence, our analysis allows for a more nuanced description of the evidence
+of ecological black swans. In an earlier study, \citet{keitt1998} described
+heavy (power law) tails in breeding bird population abundance. However, this
+finding was challenged by \citet{allen2001}, who showed that mixing data
+across species could falsely generate heavy tails.
+%This rebuttal supports our
+%use of the t distribution to represent heavy tails, since a t distribution can
+%be represented by a mixture of normal distributions \citep[with the same mean
+%and inverse-gamma-distributed variances,][]{gelman2014}.
 
-We might expect to observe black-swan dynamics in ecological time series because of unmodelled intrinsic properties of populations or extrinsic forces acting on populations.
-Since a t-distribution can be formed by a mixture of normal distributions \citep{gelman2014}, we could observe heavy tails if we miss some underlying mixture of intrinsic processes \citep{allen2001}.
-That process might be an aggregation of populations across space, or population diversity, or an intrinsic change in population variability through time.
-Extrinsic forces could also cause black-swan dynamics \citep[e.g.][]{nunez2012}.
-These forces could be extreme themselves.
-For example, extreme climate, predation from (or competition with) other species experiencing black swans, or sharp changes in human pressure such hunting, fishing, or habitat destruction might cause black swans.
-Alternatively, the interaction of multiple ``normal'' extrinsic forces could give rise to black-swan ecological dynamics.
-This could occur if the interaction is synergistic \citep[e.g.][]{kirby2009} or even if non-synergistic interactions experience a rare alignment \citep{denny2009}.
+We might expect to observe black-swan dynamics in ecological time series
+because of unmodelled intrinsic properties of populations or extrinsic forces
+acting on populations. Since a t-distribution can be formed by a mixture of
+normal distributions \citep[with the same mean and inverse-gamma-distributed
+variances,][]{gelman2014}, we could observe heavy tails if we miss some
+underlying mixture of intrinsic processes \citep{allen2001}. That process
+might be an aggregation of populations across space, or population diversity,
+or an intrinsic change in population variability through time. Extrinsic
+forces could also cause black-swan dynamics \citep[e.g.][]{nunez2012}. These
+forces could be extreme themselves. For example, extreme climate, predation
+from (or competition with) other species experiencing black swans, or sharp
+changes in human pressure such hunting, fishing, or habitat destruction might
+cause black swans. Alternatively, the interaction of multiple ``normal''
+extrinsic forces could give rise to black-swan ecological dynamics. This could
+occur if the interaction is synergistic \citep[e.g.][]{kirby2009} or even if
+non-synergistic interactions experience a rare alignment \citep{denny2009}.
 
 There are a number of caveats when considering the generality of our results.
-The GPDD data represent a taxonomically and geographically biased sample of populations---the longer time series we focus on are dominated by commercially and recreationally important species and a disproportionate number of populations are located in the United Kingdom.
-Although we would expect to find qualitatively similar evidence for black swans in many other large taxonomic or geographic samples of populations, the common forces driving those black swans would likely differ (e.g.~severe winters in Table~\ref{tab:sparks}).
-In addition to a possibly biased sample of populations, some black-swan detections could just be recording mistakes, or conversely, some extreme observations may have been discarded or altered if they were erroneously suspected of being recording mistakes.
-Indeed, three of the populations our method initially identified as heavy tailed turned out to be data-entry errors and were discarded (Supporting Information).
-A final caveat is that the temporal scales of observation and population dynamics vary considerably across populations in the GPDD and this likely influences the detection of heavy tails.
-As an example, if we make frequent observations relative to generation time (e.g.~for many large-bodied mammals) we will average across generations and perhaps miss black swans.
-Conversely, if we census populations infrequently relative to generation time (e.g.~many insects in the GPDD) the recorded data may average across extreme and less-extreme events and also dampen black-swan dynamics.
+The GPDD data represent a taxonomically and geographically biased sample of
+populations---the longer time series we focus on are dominated by commercially
+and recreationally important species and a disproportionate number of
+populations are located in the United Kingdom. Although we would expect to find
+qualitatively similar evidence for black swans in many other large taxonomic or
+geographic samples of populations, the common forces driving those black swans
+would likely differ (e.g.~severe winters in Table~\ref{tab:sparks}). In
+addition to a possibly biased sample of populations, some black-swan detections
+could just be recording mistakes, or conversely, some extreme observations may
+have been discarded or altered if they were erroneously suspected of being
+recording mistakes. Indeed, three of the populations our method initially
+identified as heavy tailed turned out to be data-entry errors and were
+discarded (Supporting Information). A final caveat is that the temporal scales
+of observation and population dynamics vary considerably across populations in
+the GPDD and this likely influences the detection of heavy tails. As an
+example, if we make frequent observations relative to generation time (e.g.~for
+many large-bodied mammals) we will average across generations and perhaps miss
+black swans. Conversely, if we census populations infrequently relative to
+generation time (e.g.~many insects in the GPDD) the recorded data may average
+across extreme and less-extreme events and also dampen black-swan dynamics.
 
-Recognizing the prevalence of heavy-tailed dynamics suggests a number of policy directions.
-Ecological resource management can draw from other disciplines that focus on heavy tails.
-For example, earthquake preparedness and response is focussed on black-swan events.
-Similarly to ecological black swans, we can rarely predict the specific location and timing of large earthquakes.
-But, earthquake preparedness involves spatial planning based on forecast probabilities to focus early detection efforts and develop disaster response plans.
-A similar focus might benefit resource management once our ability to predict the spatial probability and covariates of ecological black swans improves.
-The presence of black swans also suggests that we develop management policy that is robust to heavy tails.
-For instance, setting target population abundances that are appropriately set back from critical limits may buffer black-swan events \citep[e.g.][]{caddy1996}, and maintaining genetic, phenotypic, and behavioural diversity may allow some components of populations persist when others are affected by disease or extreme environmental forces \citep[e.g.][]{hilborn2003, schindler2010, anderson2014}.
-Finally, extreme and unexpected, surprising, or counterintuitive ecological dynamics offer a tremendous opportunity to learn about ecological systems, evaluate when our models break down, and adjust future management policy \citep{doak2008, pine-iii2009, lindenmayer2010}.
+Recognizing the prevalence of heavy-tailed dynamics suggests a number of
+policy directions. Ecological resource management can draw from other
+disciplines that focus on heavy tails. For example, earthquake preparedness
+and response is focussed on black-swan events. Similarly to ecological black
+swans, we can rarely predict the specific location and timing of large
+earthquakes. But, earthquake preparedness involves spatial planning based on
+forecast probabilities to focus early detection efforts and develop disaster
+response plans. A similar focus might benefit resource management once our
+ability to predict the spatial probability and covariates of ecological black
+swans improves. The presence of black swans also suggests that we develop
+management policy that is robust to heavy tails. For instance, setting target
+population abundances that are appropriately set back from critical limits may
+buffer black-swan events \citep[e.g.][]{caddy1996}, and maintaining genetic,
+phenotypic, and behavioural diversity may allow some components of populations
+to persist when others are affected by disease or extreme environmental forces
+\citep[e.g.][]{hilborn2003, schindler2010, anderson2014}. Finally, extreme and
+unexpected, surprising, or counterintuitive ecological dynamics offer a
+tremendous opportunity to learn about ecological systems, evaluate when our
+models break down, and adjust future management policy \citep{doak2008,
+  pine-iii2009, lindenmayer2010}.
 
-Our results suggest a number of research questions related to ecological black swans.
-Given that black swans do occur, can we forecast the probability of black swans in space and time?
-Furthermore, what management policies allow us to detect them quickly after they happen?
-Can we isolate the components of ecological dynamics that experience black swans by moving from phenomenological models such as the Gompertz to mechanistic models, that take into account, for example, recruitment?
-We expect that greater insight into the mechanisms and covariates of ecological black swans may be best obtained through specific geographic and taxonomic subsets of data where longer time series with low levels of observation error are available \citep[e.g.][]{segura2013}.
+Our results suggest a number of research questions related to ecological black
+swans. Given that black swans do occur, can we forecast the probability of
+black swans in space and time? Furthermore, what management policies allow us
+to detect them quickly after they happen? Can we isolate the components of
+ecological dynamics that experience black swans by moving from
+phenomenological models such as the Gompertz to mechanistic models that, for
+example, take into account recruitment dynamics? We expect that greater
+insight into the mechanisms and covariates of ecological black swans may be
+best obtained through specific geographic and taxonomic subsets of data where
+longer time series with low levels of observation error are available
+\citep[e.g.][]{segura2013}.
 
-Most importantly, what is the impact of allowing for black swans in forecasts of ecological risk?
-In recent decades, ecology has moved toward focussing on aspects of variance in addition to mean responses \citep[e.g.][]{loreau2010a, thompson2013}.
-Our results suggest that an added focus on ecological extremes represents the next frontier, particularly in the face of increased climate extremes \citep{meehl2004,ipcc2012}.
-Financial analysts are concerned with the shape of financial-return-distribution tails because these impact estimates of risk---the probability of a specific magnitude of undesired event occurring \citep{rachev2008}.
-A comparable focus in ecology would increase our estimates of extinction risk, since these would be disproportionately impacted by downward black-swan events.
+Most importantly, what is the impact of allowing for black swans in forecasts
+of ecological risk? In recent decades, ecology has moved toward focussing on
+aspects of variance in addition to mean responses \citep[e.g.][]{loreau2010a,
+  thompson2013}. Our results suggest that an added focus on ecological
+extremes represents the next frontier, particularly in the face of increased
+climate extremes \citep{meehl2004,thompson2013,ipcc2012}. Financial analysts
+are concerned with the shape of the tails of financial returns because these
+directly impact estimates of risk---the probability of a specific magnitude of
+undesired event occurring \citep{rachev2008}. A comparable focus in ecology
+would increase our estimates of extinction risk, since these would be
+disproportionately impacted by downward black-swan events.
 
@@
 
-We thank Justin D.
-Yeakel, TODO, and other members of the Earth to Ocean research group for helpful discussions and comments.
-We are grateful to the contributors and maintainers of the Global Population Dynamics Database and to Compute Canada's WestGrid high-performance computing resources.
-Silhouette images were obtained from \texttt{phylopic.org} under Creative Commons licenses; sources are listed in the Supporting Information.
-Funding was provided by a Simon Fraser University Graduate Fellowship (SCA), the Natural Sciences and Engineering Research Council of Canada (NKD, ABC), the Canada Research Chairs Program (NKD).
+We thank J.D. Yeakel and other members of the Earth to Ocean research
+group for helpful discussions and comments. We are grateful to the
+contributors and maintainers of the Global Population Dynamics Database and to
+Compute Canada's WestGrid high-performance computing resources. Silhouette
+images were obtained from \texttt{phylopic.org} under Creative Commons
+licenses; sources are listed in the Supporting Information. Funding was
+provided by a Simon Fraser University Graduate Fellowship (SCA), the Natural
+Sciences and Engineering Research Council of Canada (NKD, ABC), the Canada
+Research Chairs Program (NKD).
 
@@
 \bibliography{/Users/seananderson/Dropbox/tex/jshort,/Users/seananderson/Dropbox/tex/ref3}
+%\bibliography{ms}
 
@@
 \caption{
-An illustration of fitting population dynamic models that allow for heavy tails, represented by the Student-t degrees of freedom parameter $\nu$.
-(a, b) The probability density for t distributions with a scale parameter of 1 and different values of $\nu$.
-Small values of $\nu$ create heavy tails.
-As $\nu$ approaches infinity the distribution approaches the normal distribution.
-For example, at $\nu = 2$, the probability of drawing a value more than five standard deviations below the mean is 1.8\%, whereas the probability of drawing such a value from a normal distribution is nearly zero ($2.9\cdot10^{-5}$\%).
-(c--e) Simulated population dynamics from a Gompertz model with process noise drawn from t distributions with three different values of $\nu$.
-Coloured dots in panels c and d represent jumps with less than a 1 in 1000 chance of occurring in a normal distribution.
-(f--h) Estimates of $\nu$ from models fit to the times series in panels c--e.
-Shown are the posterior samples (histograms), median and interquartile range of the posterior (IQR) (dots and line segments), and the exponential prior on $\nu$ (dashed lines).
-Colour shading behind panels f--h illustrates the region of heavy tails.
+An illustration of fitting population dynamic models that allow for heavy
+tails, represented by the Student-t degrees of freedom parameter $\nu$. (a, b)
+The probability density for t distributions with a scale parameter of 1 and
+different values of $\nu$. Small values of $\nu$ create heavy tails. As $\nu$
+approaches infinity the distribution approaches the normal distribution. For
+example, at $\nu = 2$, the probability of drawing a value more than five
+standard deviations below the mean is 1.8\%, whereas the probability of drawing
+such a value from a normal distribution is nearly zero ($2.9\cdot10^{-5}$\%).
+(c--e) Simulated population dynamics from a Gompertz model with process noise
+drawn from t distributions with three different values of $\nu$. Coloured dots
+in panels c and d represent jumps with less than a 1 in 1000 chance of
+occurring in a normal distribution. (f--h) Estimates of $\nu$ from models fit
+to the times series in panels c--e. Shown are the posterior samples
+(histograms), median and interquartile range of the posterior (IQR) (dots and
+line segments), and the exponential prior on $\nu$ (dashed lines). Colour
+shading behind panels f--h illustrates the region of heavy tails.
 }
@@
 \includegraphics[width=0.44\textwidth]{../analysis/nu-coefs-2.pdf}
-\caption{
-Estimates of how heavy-tailed population dynamics are for \nuCoefPopN\ populations of birds, mammals, insects, and fishes.
-Small values of $\nu$ ($\lesssim 10$) suggest heavy-tailed black-swan dynamics; larger values of $\nu$ suggest approximately normal-tailed dynamics.
-Vertical points and line segments represent posterior medians and 50\% / 90\% credible intervals for individual populations.
-Inset plots show probability that $\nu < 10$ (probability of heavy tails) for populations arranged by taxonomic order and sorted by decreasing mean Pr($\nu < 10$).
-Taxonomic orders with three or fewer populations in panel a are omitted for space.
-Red to yellow points highlight populations with a high to moderately high probability of heavy-tailed black-swan dynamics.
-}
+
+\caption{Estimates of population dynamics heavy-tailedness for \nuCoefPopN\
+  populations of birds, mammals, insects, and fishes. Small values of $\nu$
+  ($\lesssim 10$) suggest heavy-tailed black-swan dynamics; larger values of
+  $\nu$ suggest approximately normal-tailed dynamics. Vertical points and line
+  segments represent posterior medians and 50\% / 90\% credible intervals for
+  individual populations. Inset plots show probability that $\nu < 10$
+  (probability of heavy tails) for populations arranged by taxonomic order and
+  sorted by decreasing mean Pr($\nu < 10$). Taxonomic orders with three or
+  fewer populations in panel a are omitted for space. Red to yellow points
+  highlight populations with a high to moderately high probability of
+  heavy-tailed black-swan dynamics.}
+
 \label{fig:nu-coefs}
@@
 \includegraphics[width=0.9\textwidth]{../analysis/correlates-p10.pdf}
-\caption{
-Potential covariates of heavy-tailed population dynamics (indicated by a high probability that $\nu < 10$).
-Shown are (a--c) parameters from the Gompertz heavy-tailed population model ($\sigma$, $\lambda$, $b$), (b) number of time steps, (c) body length, and (d) lifespan.
-For the Gompertz parameters, $\sigma$ refers to the scale parameter of the Student-t process-noise distribution, $\lambda$ refers to the expected log abundance at the next time step at an abundance of one, $b$ refers to the density dependence parameter ($1$ is maximally density independent, $0$ is maximally density dependent, and $<0$ is inversely density dependent).
-Circles representing a few sharks, crustaceans, and gastropods are filled in white.
-Median and 90\% credible interval posterior predictions of a beta regression multilevel model are shown in panels a and d where there was a high probability the slope coefficient was different from zero (Fig.~\ref{fig:posteriors}b).
-}
+
+\caption{Potential covariates of heavy-tailed population dynamics (indicated
+  by a high probability that $\nu < 10$). Shown are (a--c) parameters from the
+  Gompertz heavy-tailed population model ($\sigma$, $\lambda$, $b$), (b)
+  number of time steps, (c) body length, and (d) lifespan. For the Gompertz
+  parameters, $\sigma$ refers to the scale parameter of the Student-t
+  process-noise distribution, $\lambda$ refers to the expected log abundance
+  at the next time step at an abundance of one, $b$ refers to the density
+  dependence parameter ($1$ is maximally density independent, $0$ is maximally
+  density dependent, and $<0$ is inversely density dependent). Circles
+  representing a few sharks, crustaceans, and gastropods are filled in white.
+  Median and 90\% credible interval posterior predictions of a beta regression
+  multilevel model are shown in panels a and d where there was a high
+  probability the slope coefficient was different from zero
+  (Fig.~\ref{fig:posteriors}b).}
+
 \label{fig:correlates}
@@
 \includegraphics[width=0.45\textwidth]{../analysis/order-posteriors-covariates.pdf}
-\caption{
-Posterior probability distributions from beta regression multilevel models.
-(a) Taxonomic-order-level posterior densities of Pr($\nu < 10$) (approximately the probability of heavy tails) after accounting for time-series length.
-Estimates are at the geometric mean of time series length across all the data (approximately 27 time steps).
-Colour shading refers to taxonomic class (yellow: fishes, green: insects, purple: birds, and red: mammals).
-Long dotted vertical line in panel a indicates the median expected Pr($\nu < 10$) from the prior distribution.
-(b) Main effect posterior densities for potential covariates of Pr($\nu < 10$).
-The beta regression models were fit on a logit scale with hierarchical intercepts for taxonomic class, taxonomic order, and species.
-All covariates were standardized by subtracting their mean and dividing by twice their standard deviation.
-In both panels, short vertical line segments indicate median posterior estimates.
-}
+
+\caption{Posterior probability distributions from beta regression multilevel
+  models. (a) Taxonomic-order-level posterior densities of Pr($\nu < 10$)
+  (approximately the probability of heavy tails) after accounting for
+  time-series length. Estimates are at the geometric mean of time series
+  length across all the data (approximately 27 time steps). Colour shading
+  refers to taxonomic class (yellow: fishes, green: insects, purple: birds,
+  and red: mammals). Dotted vertical line in panel a indicates the median
+  expected Pr($\nu < 10$) from the prior distribution. (b) Main effect
+  posterior densities for potential covariates of Pr($\nu < 10$). The beta
+  regression models were fit on a logit scale with hierarchical intercepts for
+  taxonomic class, taxonomic order, and species. All covariates were
+  standardized by subtracting their mean and dividing by twice their standard
+  deviation. In both panels, short vertical line segments within the density
+  polygons indicate median posterior estimates.}
+
 \label{fig:posteriors}

### 2014-11-18 text/supp.tex
Add to supp caption
Lots of edits and wrap text

@@
 
-We applied the following data selection and quality-control rules to the Global Population Dynamics Database (GPDD):
+We applied the following data selection and quality-control rules to the
+Global Population Dynamics Database (GPDD):
 
@@
 
-\item To remove populations with unreliable population indices that could be strongly confounded with economics and sampling effort, we removed all populations with a sampling protocol listed as \texttt{harvest} as well populations with the words \texttt{harvest} or \texttt{fur} in the cited reference title.
+\item To remove populations with unreliable population indices that could be
+  strongly confounded with economics and sampling effort, we removed all
+  populations with a sampling protocol listed as \texttt{harvest} as well
+  populations with the words \texttt{harvest} or \texttt{fur} in the cited
+  reference title.
 
-\item We removed all populations with uneven sampling intervals, i.e.\ we removed populations that didn't have a constant difference between the ``decimal year begin'' and ``decimal year end'' columns.
+\item We removed all populations with uneven sampling intervals, i.e.\ we
+  removed populations that did not have a constant difference between the
+  ``decimal year begin'' and ``decimal year end'' columns.
 
-\item We removed all populations rated as $< 2$ in the GPDD quality assessment (on a scale of $1$ to $5$, with $1$ being the lowest quality data) \citep[following][]{sibly2005, ziebarth2010}.
+\item We removed all populations rated as $< 2$ in the GPDD quality assessment
+  (on a scale of $1$ to $5$, with $1$ being the lowest quality data)
+  \citep[following][]{sibly2005, ziebarth2010}.
 
-\item Populations with negative abundance values were removed.
-Of the populations that remained at the end of our other filtering rules, the remaining populations with negative abundances listed were all from time series that had been standardized by subtracting the mean and dividing by the standard deviation.
-We verified this by locating the original papers the datasets were extracted from: \citet{colebrook1978} for zooplankton and \citet{lindstrom1995} for grouse.
-Since the papers did not include the original mean time-series values we could not back transform these data points.
+\item Populations with negative abundance values were removed. Of the
+  populations that remained at the end of our other filtering rules, the
+  remaining populations with negative abundances listed were all from time
+  series that had been standardized by subtracting the mean and dividing by the
+  standard deviation. We verified this by locating the original papers the
+  datasets were extracted from: \citet{colebrook1978} for zooplankton and
+  \citet{lindstrom1995} for grouse. Since the papers did not include the
+  original mean time-series values we could not back transform these data
+  points.
 
-\item We filled in all missing time steps with \texttt{NA} values and imputed single missing values with the geometric mean of the previous and following values.
-We chose a geometric mean to be linear on the log scale that the Gompertz and Ricker-logistic models were fit on.
+\item We filled in all missing time steps with \texttt{NA} values and imputed
+  single missing values with the geometric mean of the previous and following
+  values. We chose a geometric mean to be linear on the log scale that the
+  Gompertz and Ricker-logistic models were fit on.
 
-\item We filled in single recorded values of zero with the lowest non-zero value in the time series \citep[following][]{brook2006a}.
-This assumes that single values of zero result from abundance being low enough that censusing overlooked individuals that were actually present.
-We turned multiple zero values in a row into \texttt{NA} values.
-This implies that multiple zero values were either censusing errors or caused by emigration.
-Regardless, our population models were fit on a multiplicative (log) scale and so could not account for zero abundance.
-To avoid distorting the original data too strongly, we removed populations in which we filled in more than four zeros.
+\item We filled in single recorded values of zero with the lowest non-zero
+  value in the time series \citep[following][]{brook2006a}. This assumes that
+  single values of zero result from abundance being low enough that censusing
+  overlooked individuals that were actually present. We turned multiple zero
+  values in a row into \texttt{NA} values. This implies that multiple zero
+  values were either censusing errors or caused by emigration. Regardless, our
+  population models were fit on a multiplicative (log) scale and so could not
+  account for zero abundance. To avoid distorting the original data too
+  strongly, we removed populations in which we filled in more than four zeros.
 
-\item We removed all populations without at least four unique values \citep[following][]{brook2006a}.
+\item We removed all populations without at least four unique values
+  \citep[following][]{brook2006a}.
 
-\item We removed all populations with four or more identical values in a row since these suggest either recording error or extrapolation between two observations.
+\item We removed all populations with four or more identical values in a row
+  since these suggest either recording error or extrapolation between two
+  observations.
 
-\item We then wrote an algorithm to find the longest unbroken window of abundance (no \texttt{NA}s) with at least $20$ time steps in each population time series.
-If there were any populations with multiple windows of identical length, we took the most recent window.
-This is a longer window than used in some previous analyses \citep[e.g.][]{brook2006a}, but since our model attempts to capture the shape of the distribution tails, our model requires more data.
+\item We then wrote an algorithm to find the longest unbroken window of
+  abundance (no \texttt{NA}s) with at least $20$ time steps in each population
+  time series. If there were any populations with multiple windows of identical
+  length, we took the most recent window. This is a longer window than used in
+  some previous analyses \citep[e.g.][]{brook2006a}, but since our model
+  attempts to capture the shape of the distribution tails, our model requires
+  more data.
 
-\item We removed GPDD Main IDs \texttt{20531} and \texttt{10139}, which we noticed were duplicates of \texttt{20579} (a heron population).
-\texttt{20579} contained additional years of data not present in \texttt{10139}.
-We removed a limited number of populations from class Angiospermopsida and Bacillariophyceae to focus the taxonomy in our analysis on animals.
-We also removed any populations with an \texttt{Unknown} taxonomic class.
+\item We removed GPDD Main IDs \texttt{20531} and \texttt{10139}, which we
+  noticed were duplicates of \texttt{20579} (a heron population).
+  \texttt{20579} contained additional years of data not present in
+  \texttt{10139}. We removed a limited number of populations from class
+  Angiospermopsida and Bacillariophyceae to focus the taxonomy in our analysis
+  on animals. We also removed any populations with an \texttt{Unknown}
+  taxonomic class.
 
-\item Finally, we removed populations with the following GPDD Main IDs, which we discovered data entry errors when verifying the populations with suspected black swans: \texttt{1207} because the 1957 data point was entered as 2 but should have been 27 \citep{kendeigh1982}, \texttt{6531} because the 1978 data point was entered as 7 should have been 47 \citep{minot1986}, and \texttt{6566} because some of the data did not match the graph \citep{heessen1996}.
+\item Finally, we removed populations with the following GPDD Main IDs, which
+  we discovered were data entry errors when verifying the populations with
+  suspected black swans: \texttt{1207} because the 1957 data point was entered
+  as 2 but should have been 27 \citep{kendeigh1982}, \texttt{6531} because the
+  1978 data point was entered as 7 but should have been 47 \citep{minot1986},
+  and \texttt{6566} because some of the data did not match the graph
+  \citep{heessen1996}.
 
@@
 \noindent
-We provide a supplemental figure of all the time series included in our analysis and indicate which values were interpolated (non-zero interpolations) (\percImputedPops\% of populations had at least one point interpolated but only \percImputedPoints\% of the total observations were interpolated) (Fig.~\ref{fig:all-ts}).
-Note that interpolation is highly unlikely to lead to black-swan detections, since black swans involve extreme increases or decreases.
-Table~\ref{tab:stats} shows the final taxonomic breakdown and the number of populations with interpolated values.
+We provide a supplemental figure of all the time series included in our
+analysis and indicate which values were interpolated (non-zero interpolations)
+(\percImputedPops\% of populations had at least one point interpolated but
+only \percImputedPoints\% of the total observations were interpolated)
+(Fig.~\ref{fig:all-ts}). Note that interpolation is highly unlikely to lead to
+black-swan detections, since black swans involve extreme increases or
+decreases. Table~\ref{tab:stats} shows the final taxonomic breakdown and the
+number of populations with interpolated values.
 
@@
 
-For the Gompertz model, our weakly-informative priors (Fig.~\ref{fig:priors}) were:
+For the Gompertz model, our weakly-informative priors (Fig.~\ref{fig:priors})
+were:
 \begin{align*}
@@
 \end{align*}
-Our prior on $b$ was uninformative between values of $-1$ and $2$.
-We would not expect values of $b$ with levels of density dependence as low as $-1$ (very strong inverse density dependence), nor would we generally expect values above $1$.
-We allowed values of $b$ above $1$ to allow for non-stationary time series of growth rates.
-The estimates of $b$ were well within these bounds.
-Our prior on $\lambda$ was very weakly informative within the range of expected values for population growth and is similar to the default priors suggested by \citet{gelman2008d} for intercepts of regression models.
-Our Half-Cauchy prior on $\sigma$ follows \citet{gelman2006c} and \citet{gelman2008d} and the scale parameter of $2.5$ is based on our expected range of the value in nature from previous studies \citep[e.g.][]{connors2014}.
-In our testing of a subsample of populations, our parameter estimates were not qualitatively changed by switching to an uninformative uniform prior on $\sigma$, but the weakly informative Half-Cauchy prior substantially sped up chain convergence.
+Our prior on $b$ was uninformative between values of $-1$ and $2$. We would
+not expect values of $b$ with levels of density dependence as low as $-1$
+(very strong inverse density dependence), nor would we generally expect values
+above $1$. We allowed values of $b$ above $1$ to allow for non-stationary time
+series of growth rates. The estimates of $b$ were well within these bounds.
+Our prior on $\lambda$ was very weakly informative within the range of
+expected values for population growth and is similar to the default priors
+suggested by \citet{gelman2008d} for intercepts of regression models. Our
+Half-Cauchy prior on $\sigma$ follows \citet{gelman2006c} and
+\citet{gelman2008d} and the specific scale parameter of $2.5$ is based on our
+expected range of the value in nature from previous studies
+\citep[e.g.][]{connors2014}. In our testing of a subsample of populations, our
+parameter estimates were not qualitatively changed by switching to an
+uninformative uniform prior on $\sigma$, but the weakly informative
+Half-Cauchy prior substantially sped up chain convergence.
 
-Our prior on $\nu$ was based on \citet{fernandez1998}.
-They chose an exponential rate parameter of $0.1$.
-We chose a less informative rate parameter of $0.01$ and truncated the distribution at $2$, since at $\nu < 2$ the variance of the t distribution is undefined.
-This prior gives only a $7.7$\% probability that $\nu < 10$ but constrains the sampling sufficiently to avoid wandering off towards infinity---above approximately $\nu = 20$ the t distribution is so similar to the normal distribution (Fig.~\ref{fig:didactic}) that time series of the length considered here are unlikely to be informative about the precise value of $\nu$.
-In the scenario where the data are uninformative about heavy tails (e.g.~Fig.~\ref{fig:didactic}e,~h), the posterior will approximately match the prior (prior median $= 71$, mean $= 102$) and the metrics used in our paper (e.g.~Pr$(\nu < 10) > 0.5$) are unlikely to flag the population as heavy tailed.
+Our prior on $\nu$ was based on \citet{fernandez1998}. They chose an
+exponential rate parameter of $0.1$. We chose a less informative rate
+parameter of $0.01$ and truncated the distribution at $2$, since at $\nu < 2$
+the variance of the t distribution is undefined. This prior gives only a
+$7.7$\% probability that $\nu < 10$ but constrains the sampling sufficiently
+to avoid wandering off towards infinity---above approximately $\nu = 20$ the t
+distribution is so similar to the normal distribution
+(Fig.~\ref{fig:didactic}) that time series of the length considered here are
+unlikely to be informative about the precise value of $\nu$. In the scenario
+where the data are uninformative about heavy tails
+(e.g.~Fig.~\ref{fig:didactic}e,~h), the posterior will approximately match the
+prior (prior median $= 71$, mean $= 102$) and the metrics used in our paper
+(e.g.~Pr$(\nu < 10) > 0.5$) are unlikely to flag the population as heavy
+tailed.
 
-We fit our models with Stan 2.4.0 \citep{stan-manual2014}, and R 3.1.1 \citep{r2014}.
-We began with four chains and $2000$ iterations, discarding the first $1000$ as warm up (i.e.~4000 total samples).
-If $\hat{R}$ (the potential scale reduction factor---a measure of chain convergence) was greater than $1.05$ for any parameter or the minimum effective sample size, $n_\mathrm{eff}$, (a measure of the effective number of uncorrelated samples) for any parameter was less than $200$, we doubled both the total iterations and warm up period and sampled from the model again.
-These thresholds are in excess of the minimums recommended by \citet{gelman2006a} of $\hat{R} < 1.1$ and effective sample size $> 100$ for reliable point estimates and confidence intervals.
-In the majority of cases our minimum thresholds were greatly exceeded.
-We continued this procedure up to $8000$ iterations ($16000$ total samples) by which all chains were deemed to have sufficiently converged.
-These chain lengths may seem low to those familiar with software such as WinBUGS or JAGS, but the No-U-Turn Hamiltonian Markov chain Monte Carlo Sampler in Stan generally requires far fewer iterations to obtain equivalent effective sample sizes \citep{stan-manual2014}.
+We fit our models with Stan 2.4.0 \citep{stan-manual2014}, and R 3.1.1
+\citep{r2014}. We began with four chains and $2000$ iterations, discarding the
+first $1000$ as warm up (i.e.~4000 total samples). If $\hat{R}$ (the potential
+scale reduction factor---a measure of chain convergence) was greater than
+$1.05$ for any parameter or the minimum effective sample size,
+$n_\mathrm{eff}$, (a measure of the effective number of uncorrelated samples)
+for any parameter was less than $200$, we doubled both the total iterations
+and warm up period and sampled from the model again. These thresholds are in
+excess of the minimums recommended by \citet{gelman2006a} of $\hat{R} < 1.1$
+and effective sample size $> 100$ for reliable point estimates and confidence
+intervals. In the majority of cases our minimum thresholds were greatly
+exceeded. We continued this procedure up to $8000$ iterations ($16000$ total
+samples) by which all chains were deemed to have sufficiently converged. These
+chain lengths may seem low to those familiar with software such as WinBUGS or
+JAGS, but the No-U-Turn Hamiltonian Markov chain Monte Carlo Sampler in Stan
+generally requires far fewer iterations to obtain equivalent effective sample
+sizes \citep{stan-manual2014}.
 
@@
 
-To test if the prior on $\nu$ influenced our estimate of black-swan dynamics, we refit our models with weaker and stronger priors.
-Our base model used a prior on $\nu$ of Truncated-Exponential(0.02, min.\ = 2).
-For a weaker prior we used Truncated-Exponential(0.005, min.\ = 2) and for a stronger prior we used Truncated-Exponential(0.02, min.\ = 2) (Fig.~\ref{fig:priors}).
-Note that the base and weaker priors are relatively flat within the region of $\nu < 20$---the region we are mostly concerned about when categorizing populations as heavy- or thin-tailed.
+To test if the prior on $\nu$ influenced our estimate of black-swan dynamics,
+we refit our models with weaker and stronger priors. Our base model used a
+prior on $\nu$ of Truncated-Exponential(0.02, min.\ = 2). For a weaker prior
+we used Truncated-Exponential(0.005, min.\ = 2) and for a stronger prior we
+used Truncated-Exponential(0.02, min.\ = 2) (Fig.~\ref{fig:priors}). Note that
+the base and weaker priors are relatively flat within the region of $\nu <
+20$, which is the region we are mostly concerned about when categorizing populations
+as heavy- or thin-tailed.
 
-Our results show that these weaker and stronger priors would have little influence on our conclusions about heavy-tailed dynamics (Fig.~\ref{fig:alt-priors}).
-When the data are informative about tail behaviour (i.e.\ when there is strong evidence of low $\nu$ values, upper-right of Fig.~\ref{fig:alt-priors}), the prior has little impact on the estimate of $\nu$.
-When the data are less informative about $\nu$ (i.e.\ when there are no or few tail events and time series are short or noisy), the prior can pull the estimate of $\nu$ towards larger or smaller values (Fig.~\ref{fig:alt-priors}).
-The vast majority of the populations with Pr$(\nu < 10)$ in the base prior are not altered qualitatively by this range of prior strength.
+Our results show that these weaker and stronger priors would have little
+influence on our conclusions about heavy-tailed dynamics
+(Fig.~\ref{fig:alt-priors}). When the data are informative about tail
+behaviour (i.e.\ when there is strong evidence of low $\nu$ values,
+upper-right of Fig.~\ref{fig:alt-priors}), the prior has little impact on the
+estimate of $\nu$. When the data are less informative about $\nu$ (i.e.\ when
+there are no or few tail events and time series are short or noisy), the prior
+can pull the estimate of $\nu$ towards larger or smaller values
+(Fig.~\ref{fig:alt-priors}). The vast majority of the populations with Pr$(\nu
+< 10)$ in the base prior were not altered qualitatively by this range of prior
+strength.
 
@@
 
-We fit four alternative population models to the time-series data to check how they would influence our conclusions.
-Our alternative models allowed for autocorrelation in the residuals, assumed no density dependence, allowed for observation error, or assumed a Ricker-logistic functional form.
-The range of percentages of black swans by taxonomic class cited in the abstract and results are based on lower and upper limits across our main Gompertz model and these four alternative models.
+We fit four alternative population models to the time-series data to check how
+they would influence our conclusions. Our alternative models allowed for
+autocorrelation in the residuals, assumed no density dependence, allowed for
+observation error, or assumed a Ricker-logistic functional form. The range of
+percentages of black swans by taxonomic class cited in the abstract and
+results are based on lower and upper limits across our main Gompertz model and
+these four alternative models.
 
@@
 \end{align*}
-In addition to the parameters in the original Gompertz model, we estimate an additional parameter $\phi$, which represents the dependence of subsequent residuals.
-Based on the results of previous analyses with the GPDD \citep[e.g.][]{connors2014} and the chosen priors in previous analyses \citep[e.g.][]{thorson2014a} and to greatly speed up chain convergence when running our model across all populations, we placed a weakly informative prior on $\phi$ that assumed the greatest probability density near zero with the reduced possibility of $\phi$ being near $-1$ or $1$.
-Specifically, we chose $\phi \sim \mathrm{Truncated\mhyphen Normal}(0, 1, \mathrm{min.} = -1, \mathrm{max.} = 1)$.
-The MCMC chains did not converge for \modelsNoConvergeAROne\ populations according to our criteria ($\widehat{R} < 1.05, n_\mathrm{eff} > 200$) after 8000 iterations of four chains.
-This included only \modelsNoConvergeAROneHeavyBase\ populations in which Pr($\nu < 10$) $> 0.5$ categorized them as heavy in the main Gompertz model.
-We did not include these models in Fig.~\ref{fig:alt}.
+In addition to the parameters in the original Gompertz model, this model
+estimates an additional parameter $\phi$, which represents the correlation of
+subsequent residuals. Based on the results of previous analyses with the GPDD
+\citep[e.g.][]{connors2014} and the chosen priors in previous analyses
+\citep[e.g.][]{thorson2014a} and to greatly speed up chain convergence when
+running our model across all populations, we placed a weakly informative prior
+on $\phi$ that assumed the greatest probability density near zero with the
+reduced possibility of $\phi$ being near $-1$ or $1$. Specifically, we chose
+$\phi \sim \mathrm{Truncated\mhyphen Normal}(0, 1, \mathrm{min.} = -1,
+\mathrm{max.} = 1)$. The MCMC chains did not converge for
+\modelsNoConvergeAROne\ populations according to our criteria ($\widehat{R} <
+1.05, n_\mathrm{eff} > 200$) after 8000 iterations of four chains. This
+  included only \modelsNoConvergeAROneHeavyBase\ populations in which Pr($\nu
+  < 10$) $> 0.5$ categorized them as heavy in the main Gompertz model. We did
+  not include these models in Fig.~\ref{fig:alt}.
 
@@
 
-We fit a simplified version of the Gompertz model in which the density dependence parameter $b$ was fixed at $1$ (density independent).
-This is equivalent to fitting a random walk model (with drift) to the $\log$ abundances or assuming the growth rates are drawn from a stationary distribution.
-The model was as follows:
+We fit a simplified version of the Gompertz model in which the density
+dependence parameter $b$ was fixed at $1$ (density independent). This is
+equivalent to fitting a random walk model (with drift) to the $\log$
+abundances or assuming the growth rates are drawn from a stationary
+distribution. The model was as follows:
 \begin{align*}
@@
 \end{align*}
-We fit this model for three reasons: (1) it is computationally simpler and so provides a check that our more complicated full Gompertz model was obtaining reasonable estimates of $\nu$, (2) it provides a test of whether density dependence was systematically affecting our perception of heavy tails, (3) it matches how some previous authors have modelled heavy tails without accounting for density dependence \citep{segura2013}.
+We fit this model for three reasons: (1) it is computationally simpler and so
+provides a check that our more complicated full Gompertz model was obtaining
+reasonable estimates of $\nu$, (2) it provides a test of whether density
+dependence was systematically affecting our perception of heavy tails, (3) it
+matches how some previous authors have modelled heavy tails without accounting
+for density dependence \citep{segura2013}.
 
@@
 
-Observation error can bias parameter estimates \citep[e.g.][]{knape2012} and is known to affect the ability to detect extreme events \citep{ward2007}.
-In our main analysis, we fit a model that ignored observation error.
-One way to account for observation error would be to fit a full state-space model that simultaneously estimates the magnitude of process noise and observation error.
-However, simultaneously estimating observation and process noise is a challenging problem (e.g.\ because the observation and process noise parameters tend to negatively covary in model fitting) and is known to sometimes result in identifiability issues with the Gompertz population model \citep{knape2008}.
-Furthermore, our model was applied to hundreds of time series, often of short length (as few as 20 time steps) and our model estimates an additional parameter---the shape of the process deviation tails---potentially making identifiability and computational issues even greater.
-Therefore, we considered a version of the base Gompertz model that allowed for a fixed level of observation error:
+Observation error can bias parameter estimates \citep[e.g.][]{knape2012} and
+is known to affect the ability to detect extreme events \citep{ward2007}. In
+our main analysis, we fit a model that ignored observation error. One way to
+account for observation error would be to fit a full state-space model that
+simultaneously estimates the magnitude of process noise and observation error.
+However, simultaneously estimating observation and process noise is a
+challenging problem (e.g.\ because the observation and process noise
+parameters tend to negatively covary in model fitting) and is known to
+sometimes result in identifiability issues with the Gompertz population model
+\citep{knape2008}. Furthermore, our model was applied to hundreds of time
+series, often of short length (as few as 20 time steps) and our model
+estimates an additional parameter---the shape of the process deviation
+tails---potentially making identifiability and computational issues even
+greater. Therefore, we considered a version of the base Gompertz model that
+allowed for a fixed level of observation error:
 \begin{align*}
@@
 \end{align*}
-where $U$ represents the unobserved state vector, $\sigma_\mathrm{obs}$ represents the standard deviation of observation error (on a log scale), and $\sigma_\mathrm{proc}$ represent the process noise scale parameter.
-We set $\sigma_\mathrm{obs}$ to $0.2$, which represents the upper limit of values often used in simulation analyses \citep[e.g.][]{valpine2002, thorson2014b}.
+where $U$ represents the unobserved state vector, $\sigma_\mathrm{obs}$
+represents the standard deviation of observation error (on a log scale), and
+$\sigma_\mathrm{proc}$ represent the process noise scale parameter. We set
+$\sigma_\mathrm{obs}$ to $0.2$, which represents the upper limit of values
+often used in simulation analyses \citep[e.g.][]{valpine2002, thorson2014b}.
 
@@
 \end{align*}
-where $r_\mathrm{max}$ represents the theoretical maximum population growth rate that is obtained when $N_t$ (abundance at time $t$) $= 0$.
-The parameter $K$ represents the carrying capacity and, as before, $x_t$ represents the $\log$ transformed abundance at time $t$.
-The Ricker-logistic model assumes a linear decrease in population growth rate with increases in abundance.
-In contrast, the Gompertz model assumes a linear decrease in population growth rate with increases in \textit{log} abundance ($x_t$) \citep[e.g.][]{thibaut2012}.
+where $r_\mathrm{max}$ represents the theoretical maximum population growth
+rate that is obtained when $N_t$ (abundance at time $t$) $= 0$. The parameter
+$K$ represents the carrying capacity and, as before, $x_t$ represents the
+$\log$ transformed abundance at time $t$. The Ricker-logistic model assumes a
+linear decrease in population growth rate with increases in abundance. In
+contrast, the Gompertz model assumes a linear decrease in population growth
+rate with increases in \textit{log} abundance ($x_t$)
+\citep[e.g.][]{thibaut2012}.
 
-To fit the Ricker-logistic models, we chose a prior on $K$ uniform between zero and twice the maximum observed abundance (\citet{clark2010} chose uniform between zero and maximum observed, which is more informative).
-We set the prior on $r_\mathrm{max}$ as uniform between 0 and 20 as in \citet{clark2010}.
+To fit the Ricker-logistic models, we chose a prior on $K$ uniform between
+zero and twice the maximum observed abundance (\citet{clark2010} chose uniform
+between zero and maximum observed, which is more informative). We set the
+prior on $r_\mathrm{max}$ as uniform between 0 and 20 as in \citet{clark2010}.
 We used the same priors on $\nu$ and $\sigma$ as in the Gompertz model.
@@
 
-We performed two types of simulation testing.
-First, we tested how easily the Student-t distribution $\nu$ parameter could be recovered given different true values of $\nu$ and different sample sizes.
-Second, we tested the ability of the heavy-tailed Gompertz model to obtain unbiased parameter estimates of $\nu$ given that a set of process deviations was provided in which the effective $\nu$ value was close to the true $\nu$ value.
+We performed two types of simulation testing. First, we tested how easily the
+Student-t distribution $\nu$ parameter could be recovered given different true
+values of $\nu$ and different sample sizes. Second, we tested the ability of
+the heavy-tailed Gompertz model to obtain unbiased parameter estimates of
+$\nu$ given that a set of process deviations was provided in which the
+effective $\nu$ value was close to the true $\nu$ value.
 
-We separated our simulation into these two components to avoid confounding two issues.
-(1) With smaller sample sizes, there may not be a stochastic draw from the tails of a distribution.
-In that case, no model, no matter how perfect, will be able to detect the shape of the tails.
-(2) Complex models may return biased parameter estimates if there are conceptual, computational, or coding errors.
-Our first simulation tested the first issue and our second simulation tested the latter.
-In general, our simulations show that, if anything, our model under predicts the magnitude and probability of heavy tailed events---especially given the length of the time series in the GPDD.
+We separated our simulation into these two components to avoid confounding two
+issues. (1) With smaller sample sizes, there may not be a stochastic draw from
+the tails of a distribution. In that case, no model, no matter how perfect,
+will be able to detect the shape of the tails. (2) Complex models may return
+biased parameter estimates if there are conceptual, computational, or coding
+errors. Our first simulation tested the first issue and our second simulation
+tested the latter. In general, our simulations show that, if anything, our
+model under predicts the magnitude and probability of heavy tailed
+events---especially given the length of the time series in the GPDD.
 
@@
 
-First, we tested the ability to estimate $\nu$ given different true values of $\nu$ and different sample sizes.
-We took stochastic draws from t distributions with different $\nu$ values ($\nu = 3, 5, 10,$ and $10^6$ [$\approx$ normal]), with central tendency parameters of $0$, and scale parameters of $1$.
-We started with $1600$ stochastic draws and then fit the models again at the first $800, 400, 200, 100, 50,$ and $25$ draws.
-Each time we recorded the posterior samples of $\nu$.
+First, we tested the ability to estimate $\nu$ given different true values of
+$\nu$ and different sample sizes. We took stochastic draws from t
+distributions with different $\nu$ values ($\nu = 3, 5, 10,$ and $10^6$
+[$\approx$ normal]), with central tendency parameters of $0$, and scale
+parameters of $1$. We started with $1600$ stochastic draws and then fit the
+models again at the first $800, 400, 200, 100, 50,$ and $25$ draws. Each time
+we recorded the posterior samples of $\nu$.
 
-We found that we could consistently and precisely recover median posterior estimates of $\nu$ near the true value of $\nu$ with large samples ($\ge 200$) (Fig.~\ref{fig:sim-nu} upper panels).
-At smaller samples we could still usually qualitatively distinguish heavy from not-heavy tails, but the model tended to underestimate how heavy the tails were.
-At the same time, at smaller sample sizes, the model tended to overestimate how large the scale parameter was (Fig.~\ref{fig:sim-nu} lower panels).
+We found that we could consistently and precisely recover median posterior
+estimates of $\nu$ near the true value of $\nu$ with large samples ($\ge 200$)
+(Fig.~\ref{fig:sim-nu} upper panels). At smaller samples we could still
+usually qualitatively distinguish heavy from not-heavy tails, but the model
+tended to underestimate how heavy the tails were. At the same time, at smaller
+sample sizes, the model tended to overestimate how large the scale parameter
+was (Fig.~\ref{fig:sim-nu} lower panels).
 
@@
 
-In the second part of our simulation testing, we tested the ability of the heavy-tailed Gompertz model to obtain unbiased parameter estimates when the process noise was chosen so that appropriate tail events were present.
-To generate these process deviations for the $\nu = 3$ and $\nu = 5$ scenarios, we repeatedly drew proposed candidate process deviations and estimated the central tendency, scale, and $\nu$ values each time.
-We recorded when $\hat{\nu}$ (median of the posterior) was within $0.2$ CVs (coefficient of variations) of the true $\nu$ value and used this set of random seed values in our Gompertz simulation.
-The following simplified R code illustrates this procedure (the actual code is available at \url{https://github.com/seananderson/heavy-tails}):
+In the second part of our simulation testing, we tested the ability of the
+heavy-tailed Gompertz model to obtain unbiased parameter estimates when the
+process noise was chosen so that appropriate tail events were present. To
+generate these process deviations for the $\nu = 3$ and $\nu = 5$ scenarios,
+we repeatedly drew proposed candidate process deviations and estimated the
+central tendency, scale, and $\nu$ values each time. We recorded when
+$\hat{\nu}$ (median of the posterior) was within $0.2$ CVs (coefficient of
+variations) of the true $\nu$ value and used this set of random seed values in
+our Gompertz simulation. The following simplified R code illustrates this
+procedure (the actual code is available at
+\url{https://github.com/seananderson/heavy-tails}):
 
@@
 
-We then fit our Gompertz models to the simulated datasets with all parameters (except $\nu$) set near the median values estimated in the GPDD.
-We repeated this with $50$ and $100$ samples without observation error, $50$ samples with observation error ($\sigma_\mathrm{obs} = 0.2$), and $50$ samples with the same observation error and a Gompertz model that allowed for correctly specified observation error magnitude.
+We then fit our Gompertz models to the simulated datasets with all parameters
+(except $\nu$) set near the median values estimated in the GPDD. We repeated
+this with $50$ and $100$ samples without observation error, $50$ samples with
+observation error ($\sigma_\mathrm{obs} = 0.2$), and $50$ samples with the
+same observation error and a Gompertz model that allowed for correctly
+specified observation error magnitude. Our results indicate that the Gompertz
+model can recapture the true value of $\nu$ when the process noise was chosen
+so that appropriate tail events were present (Fig.~\ref{fig:sim-prob} upper
+panels). The addition of observation error caused the model to tend to
+underestimate the degree of heavy-tailedness. Fitting a model with correctly
+specified observation error did not make substantial improvements to model
+bias (Fig.~\ref{fig:sim-prob}).
 
-Our results indicate that the Gompertz model can recapture the true value of $\nu$ when the process noise was chosen so that appropriate tail events were present (Fig.~\ref{fig:sim-prob} upper panels).
-The addition of observation error caused the model to tend to underestimate the degree of heavy-tailedness.
-Fitting a model with correctly specified observation error did not make substantial improvements to model bias (Fig.~\ref{fig:sim-prob}).
+%(Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}, red and
+%green symbols in the top rows). Likewise, the other Gompertz parameters were
+%estimated without any systematic bias (Figs~\ref{fig:sim-gompertz} and
+%\ref{fig:sim-gompertz-boxplots}, red and green symbols). , overestimate the
+%magnitude of process noise, somewhat overestimate $\lambda$, and overestimate
+%density dependence (blue symbols in Figs~\ref{fig:sim-gompertz} and
+%\ref{fig:sim-gompertz-boxplots}). The overestimation of density dependence
+%with observation error is a known phenomenon \citep{knape2012}. Fitting a
+%model with correctly specified observation error made marginal improvements
+%to model bias (purple symbols in Figs~\ref{fig:sim-gompertz} and
+%\ref{fig:sim-gompertz-boxplots}).
 
-%(Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}, red and green symbols in the top rows).
-Likewise, the other Gompertz parameters were estimated without any systematic bias (Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}, red and green symbols).
-%, overestimate the magnitude of process noise, somewhat overestimate $\lambda$, and overestimate density dependence (blue symbols in Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}).
-The overestimation of density dependence with observation error is a known phenomenon \citep{knape2012}.
-Fitting a model with correctly specified observation error made marginal improvements to model bias (purple symbols in Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}).
-
-When converting the posterior distributions of $\nu$ into Pr($\nu < 10$), the models distinguished heavy and not-heavy tails reasonably well (Fig.~\ref{fig:sim-prob} lower panels).
-Without observation error, and using a probability of $0.5$ as a threshold, the model correctly classified all simulated systems with normally distributed process noise as not heavy tailed.
-The model would have miscategorized only one of $40$ simulations at $\nu = 5$ across a sample size of $50$ or $100$ data points (Fig.~\ref{fig:sim-prob}, scenarios 1 and 2 in lower row, second panel from left).
-The model would have correctly categorized all cases where the process noise was not heavy tailed (Fig.~\ref{fig:sim-prob} bottom-right panel) and all cases where $\nu = 3$ and there was not observation error.
-With $0.2$ standard deviations of observation error, the model still categorized \obsErrorNuFivePerc\% of cases as heavy tailed when $\nu = 5$ and all but one case when $\nu = 3$.
-Allowing for observation error made little improvement to the detection of heavy tails.
-Therefore, we chose to focus on the simpler model without observation error in the main text, particularly given that the true magnitude of observation error was unknown in the empirical data.
+When converting the posterior distributions of $\nu$ into Pr($\nu < 10$), the
+models distinguished heavy and not-heavy tails reasonably well
+(Fig.~\ref{fig:sim-prob} lower panels). Without observation error, and using a
+probability of $0.5$ as a threshold, the model correctly classified all
+simulated systems with normally distributed process noise as not heavy tailed.
+The model would have miscategorized only one of $40$ simulations at $\nu = 5$
+across simulated populations with $50$ or $100$ time steps
+(Fig.~\ref{fig:sim-prob}, scenarios 1 and 2 in lower row, second panel from
+left). The model would have correctly categorized all cases where the process
+noise was not heavy tailed (Fig.~\ref{fig:sim-prob} bottom-right panel) and
+all cases where $\nu = 3$ and there was not observation error. With $0.2$
+standard deviations of observation error, the model still categorized
+\obsErrorNuFivePerc\% of cases as heavy tailed when $\nu = 5$ and all but one
+case when $\nu = 3$. Allowing for observation error made little improvement to
+the detection of heavy tails. Therefore, we chose to focus on the simpler
+model without observation error in the main text, particularly given that the
+true magnitude of observation error was unknown in the empirical data.
 
@@
 
-We fit a multilevel beta regression model to the predicted probability of heavy tails, Pr($\nu < 10$), to investigate potential covariates of heavy-tailed dynamics.
-The beta distribution is useful when response data range on a continuous scale between zero and one \citep{ferrari2004}.
-We used a logit link function as is typically used in logistic regression.
-The model was as follows:
+We fit a multilevel beta regression model to the predicted probability of
+heavy tails, Pr($\nu < 10$), to investigate potential covariates of
+heavy-tailed dynamics. The beta distribution is useful when response data
+range on a continuous scale between zero and one \citep{ferrari2004}. We used
+a logit link function as is typically used in logistic regression. The model
+was as follows:
 
@@
 \end{align*}
-where $A$ and $B$ represent the beta distribution shape parameters; $\mu_i$ represents the predicted value for population $i$, class $j$, order $k$, and species $l$; $\phi_\mathrm{disp}$ represents the dispersion parameter; and $X_i$ represents a vector of predictors (such as lifespan) for population $i$ with associated $\beta$ coefficients.
-The intercepts are allowed to vary from the overall intercept $\alpha$ by taxonomic class ($\alpha^\mathrm{class}_j$), taxonomic order ($\alpha^\mathrm{order}_k$), and species ($\alpha^\mathrm{species}_l$) with standard deviations $\sigma_{\alpha \; \mathrm{class}}$, $\sigma_{\alpha \; \mathrm{order}}$, and $\sigma_{\alpha \; \mathrm{species}}$.
-Where possible, we also allowed for error distributions around the predictors by incorporating the standard deviation of the posterior samples for the Gompertz parameters $\lambda$, $b$, and $\log \sigma$ around the mean point value as normal distributions (not shown in the above equation).
+where $A$ and $B$ represent the beta distribution shape parameters; $\mu_i$
+represents the predicted value for population $i$, class $j$, order $k$, and
+species $l$; $\phi_\mathrm{disp}$ represents the dispersion parameter; and
+$X_i$ represents a vector of predictors (such as lifespan) for population $i$
+with associated $\beta$ coefficients. The intercepts are allowed to vary from
+the overall intercept $\alpha$ by taxonomic class ($\alpha^\mathrm{class}_j$),
+taxonomic order ($\alpha^\mathrm{order}_k$), and species
+($\alpha^\mathrm{species}_l$) with standard deviations $\sigma_{\alpha \;
+  \mathrm{class}}$, $\sigma_{\alpha \; \mathrm{order}}$, and $\sigma_{\alpha
+  \; \mathrm{species}}$. Where possible, we also allowed for error
+distributions around the predictors by incorporating the standard deviation of
+the posterior samples for the Gompertz parameters $\lambda$, $b$, and $\log
+\sigma$ around the mean point value as normal distributions (not shown in the
+above equation).
 
-We log transformed $\sigma$, time-series length, and lifespan to match the way they are visually represented in Fig.~\ref{fig:correlates} and to make the relationship approximately linear on the logit-transformed response scale.
-All input variables were standardized by subtracting their mean and dividing by two standard deviations to make their coefficients comparable in magnitude \citep{gelman2008c}.
-We excluded body length as a covariate because it was highly correlated with lifespan, and lifespan exhibited more overlap across taxonomy than body length.
-Lifespan is also more directly related to time and potential mechanisms driving black-swan dynamics.
+We log transformed $\sigma$, time-series length, and lifespan to match the way
+they are visually represented in Fig.~\ref{fig:correlates} and to make the
+relationship approximately linear on the logit-transformed response scale. All
+input variables were standardized by subtracting their mean and dividing by
+two standard deviations to make their coefficients comparable in magnitude
+\citep{gelman2008c}. We excluded body length as a covariate because it was
+highly correlated with lifespan, and lifespan exhibited more overlap across
+taxonomy than body length. Lifespan is also more directly related to time and
+potential mechanisms driving black-swan dynamics.
 
-We incorporated weakly informative priors into our model: $\mathrm{Cauchy}(0, 10)$ on the global intercept $\alpha$, $\mathrm{Half\mhyphen Cauchy}(0, 2.5)$ on all standard deviation parameters, $\mathrm{Half\mhyphen Cauchy}(0, 10)$ on the dispersion parameter $\phi_\mathrm{disp}$, and $\mathrm{Cauchy}(0, 2.5)$ on all other parameters \citep{gelman2006c, gelman2008d}.
-Compared to normal priors, the Cauchy priors concentrate more probability density around expected parameter values while allowing for a higher probability density far into the tails, thereby allowing the data to dominate the posterior more strongly if it disagrees with the prior.
-Our conclusions were not qualitatively changed by using uniform priors.
-We fit our models with 5000 total iterations per chain, 2500 warm-up iterations, four chains, and discarding every second sample to save memory.
-We checked for chain convergence visually and with the same criteria as before ($\widehat{R} < 1.05$ and $n_\mathrm{eff} >200$ for all parameters).
+We incorporated weakly informative priors into our model: $\mathrm{Cauchy}(0,
+10)$ on the global intercept $\alpha$, $\mathrm{Half\mhyphen Cauchy}(0, 2.5)$
+on all standard deviation parameters, $\mathrm{Half\mhyphen Cauchy}(0, 10)$ on
+the dispersion parameter $\phi_\mathrm{disp}$, and $\mathrm{Cauchy}(0, 2.5)$
+on all other parameters \citep{gelman2006c, gelman2008d}. Compared to normal
+priors, the Cauchy priors concentrate more probability density around expected
+parameter values while allowing for a higher probability density far into the
+tails, thereby allowing the data to dominate the posterior more strongly if it
+disagrees with the prior. Our conclusions were not qualitatively changed by
+using uniform priors. We fit our models with 5000 total iterations per chain,
+2500 warm-up iterations, four chains, and discarding every second sample to
+save memory. We checked for chain convergence visually and with the same
+criteria as before ($\widehat{R} < 1.05$ and $n_\mathrm{eff} >200$ for all
+parameters).
 
-To derive taxonomic-order-level estimates of the probability of heavy tails accounting for time-series length (Fig.~\ref{fig:posteriors}a), we fit a separate multilevel model with the same structure but with only $\log$ time-series length as a predictor.
-(In this case, we did not want to control for intrinsic population characteristics such as density dependence.) Since our predictors were centered by subtracting their mean value, we obtained order-level estimates of the probability of heavy tails at mean log time-series length by adding the posteriors for $\alpha$, $\alpha^\mathrm{class}_j$, and $\alpha^\mathrm{order}_k$.
+To derive taxonomic-order-level estimates of the probability of heavy tails
+accounting for time-series length (Fig.~\ref{fig:posteriors}a), we fit a
+separate multilevel model with the same structure but with only $\log$
+time-series length as a predictor. (In this case, we did not want to control
+for intrinsic population characteristics such as density dependence.) Since
+our predictors were centered by subtracting their mean value, we obtained
+order-level estimates of the probability of heavy tails at mean log
+time-series length by adding the posteriors for $\alpha$,
+$\alpha^\mathrm{class}_j$, and $\alpha^\mathrm{order}_k$.
 
@@
 
-Many of the silhouette images used in Figs~\ref{fig:nu-coefs}, \ref{fig:correlates} and \ref{fig:posteriors} were obtained from \texttt{phylopic.org} under Creative Commons licenses.
-We vectorized the salmon in Fig.~\ref{fig:nu-coefs} and Fig.~\ref{fig:correlates} ourselves.
-The bird in these figures was obtained from \texttt{phylopic.org} under a Creative Commons Attribution 3.0 Unported license with credit to Jean-Raphaël Guillaumin {[}photography{]} and T.
-Michael Keesey {[}vectorization{]}).
-The silhouettes in Fig.~\ref{fig:posteriors} were obtained from the following sources (metadata obtained with the help of the rphylopic R package, \url{https://github.com/sckott/rphylopic}):
+Many of the silhouette images used in Figs~\ref{fig:nu-coefs},
+\ref{fig:correlates} and \ref{fig:posteriors} were obtained from
+\texttt{phylopic.org} under Creative Commons licenses. We vectorized the
+salmon in Fig.~\ref{fig:nu-coefs} and Fig.~\ref{fig:correlates} ourselves. The
+bird in these figures was obtained from \texttt{phylopic.org} under a Creative
+Commons Attribution 3.0 Unported license with credit to Jean-Raphaël
+Guillaumin {[}photography{]} and T. Michael Keesey {[}vectorization{]}). The
+silhouettes in Fig.~\ref{fig:posteriors} were obtained from the following
+sources (metadata obtained with the help of the rphylopic R package,
+\url{https://github.com/sckott/rphylopic}):
 
@@
 \begin{footnotesize}
-\caption{Summary statistics for the filtered Global Population Dynamics Database time series arranged by taxonomic class.
-Columns are: number of populations, number of taxonomic orders, numbers of species, median time series length, total number of interpolated time steps, total number of substituted zeros, and total number of time steps.}
+
+\caption{Summary statistics for the filtered Global Population Dynamics
+  Database time series arranged by taxonomic class. Columns are: number of
+  populations, number of taxonomic orders, numbers of species, median time
+  series length, total number of interpolated time steps, total number of
+  substituted zeros, and total number of time steps.}
+
 \smallskip
@@
 \begin{longtable}{>{\RaggedRight}m{1.5cm}>{\RaggedRight}p{4.3cm}>{\RaggedRight}p{0.8cm}>{\RaggedRight}p{1.7cm}>{\RaggedRight}p{1.0cm}>{\RaggedRight}p{3.0cm}>{\RaggedRight}p{1.7cm}>{\RaggedRight}p{1.3cm}}
-\caption{All populations with Pr$(\nu < 10) > 0.5$ in the base heavy-tailed Gompertz population dynamics model.
-Shown are the log abundance time series, population descriptions, Global Population Dynamics Database Main IDs, citation for the data source or separate verification literature, a description of the cause of the black swan events (if known), the probability of heavy tails as calculated by our model, and median estimate of $\nu$ from our model with 90\% quantile credible intervals indicated in parentheses.
-Red dots on the time series indicate downward black-swan events and blue values indicate upward black-swan events that have a $10 \cdot 10^{-4}$ probability or less of occurring if the population dynamics were explained by a Gompertz model with normally distributed process noise with a standard deviation equal to the scale parameter in the fitted t distribution.}\\
+
+\caption{All populations with Pr$(\nu < 10) > 0.5$ in the base heavy-tailed
+  Gompertz population dynamics model. Shown are the log abundance time series,
+  population descriptions, Global Population Dynamics Database Main IDs,
+  citation for the data source or separate verification literature, a
+  description of the cause of the black swan events (if known), the
+  probability of heavy tails as calculated by our model, and median estimate
+  of $\nu$ from our model with 90\% quantile credible intervals indicated in
+  parentheses. Red dots on the time series indicate downward black-swan events
+  and blue values indicate upward black-swan events that have a $10 \cdot
+  10^{-4}$ probability or less of occurring if the population dynamics were
+  explained by a Gompertz model with normally distributed process noise with a
+  standard deviation equal to the scale parameter in the fitted t
+  distribution.}\\
+
 \toprule
@@
 \includegraphics[width=\textwidth]{../analysis/all-clean-ts-fishes-others.pdf}
-\caption{
-  (fishes, crustaceans, gastropods, sharks).
-All filtered time series used in our analysis.
-The abundances are shown on a log10 vertical axis.
-Throughout this figure, red dots indicate values that were interpolated and blue dots indicate values that were recorded as zero but were set to the next lowest observed abundance.
-Numbers before each species name are the GPDD Main ID numbers.
-}
+
+\caption{(fishes, crustaceans, gastropods, sharks). All filtered time series
+  used in our analysis. The abundances are shown on a log10 vertical axis.
+  Throughout this figure, red dots indicate values that were interpolated and
+  blue dots indicate values that were recorded as zero but were set to the
+  next lowest observed abundance. Numbers before each species name are the
+  GPDD Main ID numbers.}
+
 \label{fig:all-ts}
@@
 \includegraphics[width=0.8\textwidth]{../analysis/priors-gomp-base.pdf}
-\caption{
-Probability density of the Bayesian priors for the Gompertz models.
-From left to right and then top to bottom: (1) per capita growth rate at $\log$(abundance) = $0$: $\lambda \sim \mathrm{Normal}(0, 10^2)$; (2) scale parameter of t-distribution process noise: $\sigma \sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)$; (3) t-distribution degrees of freedom parameter: $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2)$; (4) AR1 correlation coefficient of residuals: $\phi \sim \mathrm{Truncated \mhyphen Normal}(0, 1, \mathrm{min.} = -1, \mathrm{max.} = 1)$.
-Not shown is $b$, the density dependence parameter: $b \sim \mathrm{Uniform}(-1, 2)$.
-The $\nu$ panel also shows two alternative priors: a weaker prior $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.005, \mathrm{min.} = 2)$, and a stronger prior $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.02, \mathrm{min.} = 2)$.
-The inset panel shows the same data but with a log-transformed x axis.
-Note that the base and weaker priors are relatively flat within the region of $\nu < 20$ that we are concerned with.
-}
+
+\caption{ Probability density of the Bayesian priors for the Gompertz models.
+  From left to right and then top to bottom: (1) per capita growth rate at
+  $\log$(abundance) = $0$: $\lambda \sim \mathrm{Normal}(0, 10^2)$; (2) scale
+  parameter of t-distribution process noise: $\sigma \sim \mathrm{Half\mhyphen
+    Cauchy} (0, 2.5)$; (3) t-distribution degrees of freedom parameter: $\nu
+  \sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2)$; (4)
+  AR1 correlation coefficient of residuals: $\phi \sim \mathrm{Truncated
+    \mhyphen Normal}(0, 1, \mathrm{min.} = -1, \mathrm{max.} = 1)$. Not shown
+  is $b$, the density dependence parameter: $b \sim \mathrm{Uniform}(-1, 2)$.
+  The $\nu$ panel also shows two alternative priors: a weaker prior $\nu \sim
+  \mathrm{Truncated\mhyphen Exponential}(0.005, \mathrm{min.} = 2)$, and a
+  stronger prior $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.02,
+  \mathrm{min.} = 2)$. The inset panel shows the same data but with a
+  log-transformed x axis. Note that the base and weaker priors are relatively
+  flat within the region of $\nu < 20$ that we are concerned with. }
+
 \label{fig:priors}
@@
 \includegraphics[width=\textwidth]{../analysis/gomp-comparison.pdf}
-\caption{
-  Estimates of $\nu$ from alternative models plotted against the base Gompertz
-  model estimates of $\nu$. Shown are medians of the posterior (dots) and 50\%
-  credible intervals (segments). The diagonal line indicates a one-to-one
-  relationship. Different colours indicate various taxonomic classes. The
-  grey-shaded regions indicate regions of disagreement if $\nu = 10$ is taken
-  as a threshold of heavy-tailed dynamics. The Gompertz observation error model
-  assumes a fixed standard deviation of observation error of $0.2$ on a log
-  scale.
-}
+
+\caption{Estimates of $\nu$ from alternative models plotted against the base
+  Gompertz model estimates of $\nu$. Shown are medians of the posterior (dots)
+  and 50\% credible intervals (segments). The diagonal line indicates a
+  one-to-one relationship. Different colours indicate various taxonomic
+  classes. The grey-shaded regions indicate regions of disagreement if $\nu =
+  10$ is taken as a threshold of heavy-tailed dynamics. The Gompertz
+  observation error model assumes a fixed standard deviation of observation
+  error of $0.2$ on a log scale.}
+
 \label{fig:alt}
@@
 \includegraphics[width=\textwidth]{../analysis/gomp-prior-comparison.pdf}
-\caption{
-Estimates of $\nu$ from Gompertz models with alternative priors on $\nu$.
-Shown are medians of the posterior (dots) and 50\% credible intervals (segments).
-The diagonal line indicates a one-to-one relationship.
-Different colours indicate various taxonomic classes.
-The grey-shaded regions indicate regions of disagreement if $\nu = 10$ is taken as a threshold of heavy-tailed dynamics.
-The base, weaker, and stronger priors on $\nu$ are illustrated in Fig.~\ref{fig:priors}.
-In general, the estimates are nearly identical in cases where the data are informative about low values of $\nu$.
-When the data are less informative about low values of $\nu$, the prior can slightly pull the estimates of $\nu$ towards higher or lower values.
-}
+
+\caption{Estimates of $\nu$ from Gompertz models with alternative priors on
+  $\nu$. Shown are medians of the posterior (dots) and 50\% credible intervals
+  (segments). The diagonal line indicates a one-to-one relationship. Different
+  colours indicate various taxonomic classes. The grey-shaded regions indicate
+  regions of disagreement if $\nu = 10$ is taken as a threshold of
+  heavy-tailed dynamics. The base, weaker, and stronger priors on $\nu$ are
+  illustrated in Fig.~\ref{fig:priors}. In general, the estimates are nearly
+  identical in cases where the data are informative about low values of $\nu$.
+  When the data are less informative about low values of $\nu$, the prior can
+  slightly pull the estimates of $\nu$ towards higher or lower values.}
+
 \label{fig:alt-priors}
@@
 \includegraphics[width=0.8\textwidth]{../analysis/t-dist-sampling-sim-sigma-prior-exp0point01.pdf}
-\caption{
-  Testing the ability to estimate $\nu$ (top panels) and the scale parameter of
-  the process deviations (bottom panels) for a given number of samples (columns)
-  drawn from a distribution with a given true $\nu$ value (rows). The red lines
-  indicate the true population value. When a small number of samples are drawn
-  there may not be samples sufficiently far into the tails to recapture the
-  true $\nu$ value; however, heavy tails are still distinguished from normal
-  tails in most cases, even with only 25 or 50 samples.
-}
+
+\caption{Testing the ability to estimate $\nu$ (top panels) and the scale
+  parameter of the process deviations (bottom panels) for a given number of
+  samples (columns) drawn from a distribution with a given true $\nu$ value
+  (rows). The red lines indicate the true population value. When a small
+  number of samples are drawn there may not be samples sufficiently far into
+  the tails to recapture the true $\nu$ value; however, heavy tails are still
+  distinguished from normal tails in most cases, even with only 25 or 50
+  samples.}
+
 \label{fig:sim-nu}
@@
 \includegraphics[width=\textwidth]{../analysis/sim-gompertz-p10.pdf}
-\caption{
-Simulation testing the Gompertz estimation model when the process deviation
-draws were chosen so that $\nu$ could be estimated close to the true value
-outside the full population model (``effective $\nu$'' within a CV of 0.2 of
-specified $\nu$).
-Upper panels show the distribution of median $\widehat{\nu}$ across 20 simulation runs.
-Lower panels show the distribution of Pr($\nu < 10$) across 20 simulation runs.
-We ran the simulations across three population (``true'') $\nu$ values
-(3, 5, and $1\cdot 10^9$, i.e.\ approximately normal) and four scenarios: (1) 100 time steps and no
-observation error, (2) 50 time steps and no observation error, (3) 50 time
-steps and observation error drawn from $\mathrm{Normal} (0, 0.2^2)$ but
-ignored, and (4) 50 time steps with observation error in which the quantity
-of observation error was assumed known.
-Within each scenario the dots represent stochastic draws from the true population
-distributions combined with model fits.
-}
+
+\caption{ Simulation testing the Gompertz estimation model when the process
+  deviation draws were chosen so that $\nu$ could be estimated close to the
+  true value outside the full population model (``effective $\nu$'' within a
+  CV of 0.2 of specified $\nu$). Upper panels show the distribution of median
+  $\widehat{\nu}$ across 20 simulation runs. Lower panels show the
+  distribution of Pr($\nu < 10$) across 20 simulation runs. We ran the
+  simulations across three population (``true'') $\nu$ values (3, 5, and
+  $1\cdot 10^9$, i.e.\ approximately normal) and four scenarios: (1) 100 time
+  steps and no observation error, (2) 50 time steps and no observation error,
+  (3) 50 time steps and observation error drawn from $\mathrm{Normal} (0,
+  0.2^2)$ but ignored, and (4) 50 time steps with observation error in which
+  the quantity of observation error was assumed known. Within each scenario
+  the dots represent stochastic draws from the true population distributions
+  combined with model fits. Underlayed boxplots show the median, interquartile
+  range, and $1.5$ times the interquartile range. }
+
 \label{fig:sim-prob}
@@
 Example Stan code for a heavy-tailed Gompertz model with AR1 correlated
-residuals and a specified level of observation error. The specific code for used for the various models in our analysis is available at \url{https://github.com/seananderson/heavy-tails}.
+residuals and a specified level of observation error. The specific code for
+used for the various models in our analysis is available at
+\url{https://github.com/seananderson/heavy-tails}.
 

### 2014-11-18 text/ms.tex
Incorporate some of Jon's comments and add TODOs

@@
 the world across \NOrders~taxonomic orders and seven classes. We find strong
-evidence of black swans, but they are rare occurring in
+evidence of black swans, but they are rare, occurring in
 \overallMinPerc--\overallMaxPerc\% of populations: most frequently for birds
@@
 populations, we have compelling evidence of catastrophes \citep{gerber2001,
-  ward2007}, and recently, heavy-tailed distributions have been discovered in
-time-series of marine microbe abundance \citep{segura2013} and time to
-extinction for experimental waterflea populations \citep{drake2014}. Despite
-these anecotes, as far as we can tell there have been two systematic surveys
-for ecological black swans: one with North American breeding birds and the
-other with the Global Population Dynamics Database, which uncovered little
-clear evidence for them \citep{keitt1998,allen2001,halley2002}. Nevertheless
-there are methodological challenges to their detection
-\citep{allen2001,ward2007}.
+  ward2007}, and recently, time-series of marine microbe abundance
+\citep{segura2013} and time to extinction for experimental waterflea
+populations \citep{drake2014} have been found to have heavier tailed
+distributions than the normal. Despite these anecdotes, as far as we can tell
+there have been two systematic surveys for ecological black swans: one with
+North American breeding birds and the other with the Global Population
+Dynamics Database, which uncovered little clear evidence for them
+\citep{keitt1998,allen2001,halley2002}. Nevertheless there are methodological
+challenges to their detection \citep{allen2001,ward2007}.
 
@@
 
+%TODO: this paragraph is written as if geared towards a methods journal. I suggest starting with a statement of general aim rather than method.
 Here we develop a framework for identifying ecological black swans in the
@@
 
+%TODO: where did this threshold come from?  (nu < 10)
 We found strong, but rare, evidence for black-swan population dynamics. By
@@
 considerable uncertainty in these estimates (Fig.~\ref{fig:posteriors}a).
+%TODO: these multilevel modelling results (and fig 3, 4) are out of order
 
@@
 majority (\percBSDown \%) of documented events were downward black swans and
+%TODO: what else would cause a black swan event? How defined "unfortunate"?
 involved a series of unfortunate events. For example, a shortage of nest sites
+%TODO how about just calling it a piscivorous waterfowl or something like that
 for the shag long-necked bird (\textit{Phalacrocorax aristotelis}) in the
+%TODO why did this make the population more vulnerable?
+Seems like it wouldn't make it more vulnerable I guess I am trying to figure out if these are additive or synergistic.
+I am guessing you want to avoid those terms, but I might just be more straight with language and avoid the "more vulnerable" phrase.
 United Kingdom reduced population productivity and made the population more
@@
 resulted in a rapid population upswing \citep{potts1980}. As another example,
+% TODO true interaction term or just synchronous events?
 an interaction between environmental- and predation-mediated population cycles
@@
 We found strong evidence for black swans (heavy-tailed process noise) in 3--5\%
+% TODO but the students t is a symmetric distribution is that cool?
 of ecological time series. Black swans were usually downward events and were
@@
 density dependence, population growth rate, or lifespan. In verified cases,
+% TODO 'interaction' word
 black-swan events were often a result of the interaction between elements of

### 2014-11-18 text/sparks.tex
Tweak sparks table caption
Incorporate some of Jon's comments and add TODOs

@@
   occurring if the population dynamics were explained by a Gompertz model
-  with normally distributed process noise. These populations are a non-random
-  sample that we verified in the primary literature.}\\
+  with normally distributed process noise. These populations are
+  a sample from the heavy-tailed populations we could verify
+  (Table~\ref{tab:causes-supp}).}\\
 
@@
 UK &
-Short-term population cycles from predator interactions combined with longterm
+Short-term population cycles from predator interactions combined with long-term
 environmental cycle caused sharp downswing  &
@@
 UK &
-Severe winters where food was submerged in snow caused population crash &
+Severe winters where food was buried under snow caused population crash &
 \citep{newton1998} \\

### 2014-11-24 analysis/8-sim-extinction-risk.Rmd
Add possible simulation of extinction probability

This file looks at quasi-extinction probability with heavy vs. normal tails

Do downwards heavy tails really matter for extinction risk?

First, we'll figure out approximately what skewness values to use to get ballpark 85% downward deviations. This is to approximately match the skewness we see in the GPDD:

```{r}
library("skewt")
perc_downs <- function(df, gamma) {
  N <- 1e6
  proc_error <- rskt(N, df = df, gamma = gamma)
  # hist(proc_error)
  l <- qnorm(0.0001, 0, sd = 1)
  u <- qnorm(0.9999, 0, sd = 1)
  downs <- sum(proc_error < l) / N
  ups <- sum(proc_error > u) / N
  downs / (downs + ups)
}
perc_downs(df = 2, gamma = 0.75)
perc_downs(df = 3, gamma = 0.75)
```

So, 0.75 gets us in the ballpark of 85% percent downs.

Now, we'll write a little C++ function to simulate from a Gompertz model. We'll do this in C++ just to make everything fast after this.

```{r}
library("Rcpp")
cppFunction("
  NumericVector gompertz(double n, double lambda, double b,
    NumericVector proc_error, double y1) {
    NumericVector y(n);
    y(0) = y1;
    for (int i = 1; i < n; ++i) {
      y(i) = lambda + b * y(i - 1) + proc_error(i - 1);
    }
    return y;
}")
```

And we'll write a function to simulate a time series, check how many generations go below some threshold, and return the data in a form we can use. I've picked parameter values to be approximately at the mean values in the GPDD (I'll check these more precisely). I'm running the simulation, by default, for 200 generations and discarding the first 50 as burn in.

```{r}
sim_gomp_extinction <- function(lambda = 1, phi = 0.3, b = 0.5, N = 200,
  gamma = 0.75, df = 3, thresh = 3, make_plot = FALSE, sigma = 0.2, burnin = 50,
  return_dat = FALSE, id = 1) {
  y1 <- lambda / b
  proc_error_raw <- rskt(N, df = df, gamma = gamma) / (1 / sigma)
  proc_error <- vector(mode = "numeric", length = N)
  proc_error[1] <- proc_error_raw[1]
  for (i in 2:N) {
    proc_error[i] <- proc_error_raw[i] + phi * proc_error[i - 1]
  }
  y <- gompertz(N, lambda, b, proc_error, y1)
  y <- y[(burnin + 1):N]
  if(make_plot) {
    par(mfrow = c(2, 1))
    plot(y, type = "o")
    abline(h = lambda / b, col = "red")
    abline(h = mean(y), col = "blue")
    plot(exp(y), type = "o")
    abline(h = thresh, col = "green", lwd = 2)
  }
  if(return_dat) {
    y_return <- y
    generation <- seq_along(y)
  } else {
    y_return <- NA
    generation <- NA
  }
    data.frame(phi = phi, b = b, N = N, N_burned = N - burnin, gamma = gamma,
      df = df, sigma = sigma, thresh = thresh, lambda = lambda,
      n_ext = sum(exp(y) < thresh), mean_log_abund = mean(y),
      dat = y_return, generation = generation, id = id)
}
```

Let's figure out what a good threshold for 10% of longterm mean is with our parameter values. We'll run a really long time series and get the mean abundance at nu = 3:

```{r}
expectation <- sim_gomp_extinction(N = 1e6, df = 3)
expect_abund <- exp(expectation$mean_log_abund)
expect_abund
thresh <- 0.1
```

Set up parallel processing:

```{r}
library("doParallel")
library("foreach")
registerDoParallel(cores = 4)
```

And run the simulation for 2000 repetitions across nu values of 2, 3, and 1e6 (i.e. normal):

```{r}
set.seed(123)
reps <- 2000
out <- list(length = 3)
out[[1]] <- plyr::ldply(seq_len(reps), function(i)
  sim_gomp_extinction(df = 2, thresh = expect_abund*thresh), .parallel = TRUE)
out[[2]] <- plyr::ldply(seq_len(reps), function(i)
  sim_gomp_extinction(df = 3, thresh = expect_abund*thresh), .parallel = TRUE)
out[[3]] <- plyr::ldply(seq_len(reps), function(i)
  sim_gomp_extinction(df = 1e6, thresh = expect_abund*thresh), .parallel = TRUE)
out <- do.call("rbind", out)
```

Plot the output:

```{r}
library("ggplot2")
ggplot(out, aes(as.factor(df), n_ext/N_burned)) + geom_boxplot() +
  ylab("Probability of abundance falling below 10% of longterm mean") +
  xlab(expression(nu~(t-distribution~degrees~of~freedom)))
```

Now we'll get some time series to plot as examples:

```{r}
set.seed(111)
out_plot <- list(length = 3)
reps <- 3
out_plot[[1]] <- plyr::ldply(seq_len(reps), function(i)
  sim_gomp_extinction(N = 200, df = 2, thresh = expect_abund*thresh, return_dat = TRUE, id = i))
out_plot[[2]] <- plyr::ldply(seq_len(reps), function(i)
  sim_gomp_extinction(N = 200, df = 3, thresh = expect_abund*thresh, return_dat = TRUE, id = i))
out_plot[[3]] <- plyr::ldply(seq_len(reps), function(i)
  sim_gomp_extinction(N = 200, df = 1e6, thresh = expect_abund*thresh, return_dat = TRUE, id = i))
out_plot <- do.call("rbind", out_plot)
```

Columns are repetitions. Rows are different degrees of heavy tails (nu values):

```{r}
ggplot(out_plot, aes(generation, dat)) + geom_point() + 
  geom_line() + facet_grid(df~id) +
  geom_hline(yintercept = log(expect_abund * thresh)) + 
  ylab("Scaled abundance") + xlab("Generation")

ggplot(out_plot, aes(generation, exp(dat))) + geom_point() + 
  geom_line() + facet_grid(df~id) +
  geom_hline(yintercept = expect_abund * thresh) + 
  ylab("log(Scaled abundance)") + xlab("Generation")
```

### 2014-11-24 text/ms.tex
Edits based on Jon's comments

@@
 increase over the next century.
-
-% TODO highlight that skew would *increase* the probability of higher
-% extinction risk
 
@@
 swans. First, they might not exist in higher taxa. Indeed, the majority of
-model fitting and risk forecasting assumes that population dynamics are not
-heavy tailed \citep[e.g.][]{brook2006a,dennis2006,knape2012}. Alternatively,
-heavy-tailed dynamics might exist, but our ability to detect them requires
-further development of statistical tools. One such tool is the generalized
-extreme value distribution, which has been applied to environmental data
-\citep[e.g.][]{katz2005}. This distribution describes the most extreme event
-per time interval (e.g.~heaviest rainfall per year), but requires time series
-that are sufficiently long to be condensed into time intervals. Another
-statistical tool involves fitting a catastrophic mixture distribution in a
-state space model to quantify the probability that population events are
-extreme, although this is also data intensive \citep{ward2007}. A third tool is
-to compare the support for fits of thin- and heavy-tailed distributions
-\citep{halley2002}, but this analysis did not quantify the probability of black
-swans or allow for population dynamics.
+model fitting and risk forecasting assumes that population dynamics are normal
+tailed on a log scale \citep[e.g.][]{brook2006a,dennis2006,knape2012}.
+Alternatively, black-swan dynamics might exist, but our ability to detect
+them requires further development of statistical tools. One such tool is the
+generalized extreme value distribution, which has been applied to
+environmental data \citep[e.g.][]{katz2005}. This distribution describes the
+most extreme event per time interval (e.g.~heaviest rainfall per year), but
+requires time series that are sufficiently long to be condensed into time
+intervals. Another statistical tool involves fitting a catastrophic mixture
+distribution in a state space model to quantify the probability that
+population events are extreme, although this is also data intensive
+\citep{ward2007}. A third tool is to compare the support for fits of thin- and
+heavy-tailed distributions \citep{halley2002}, but this analysis did not
+quantify the probability of black swans or allow for population dynamics.
 
-%TODO: this paragraph is written as if geared towards a methods journal. I suggest starting with a statement of general aim rather than method.
-Here we develop a framework for identifying ecological black swans in the
-process noise of population dynamics, i.e.\ the stochastic jumps in abundance
-from one time step to the next. Our framework quantifies both the magnitude and
-probability of heavy-tailed dynamics, allows for a range of population dynamic
-models, can incorporate observation uncertainty, and can be easily applied to
-abundance time series. We apply our model to \NPops\ populations and estimate
-the frequency and strength of black-swan events in population dynamics across
-taxonomy; verify some causes of black-swan events; and attempt to identify
-characteristics of time series or intrinsic life-history characteristics that
-are associated with heavy-tailed dynamics.
+Here, we assess the frequency and magnitude of black-swan dynamics across
+\NPops\ populations from a wide array of taxonomic groups---mostly birds,
+mammals, insects, and fishes. We then identify characteristics of time series
+or intrinsic life-history characteristics that are associated with the
+detection of black-swan events and attempt to verify known causes. To
+accomplish this, we develop a framework for identifying heavy-tailed process
+noise in population dynamics, i.e.\ whether the largest stochastic jumps in
+abundance from one time step to the next are more extreme than typically seen
+with a log-normal distribution. Our framework allows for a range of population
+dynamic models, can incorporate observation uncertainty, and can be easily
+applied to abundance time series.
 
@@
 standard deviations below the mean is $0.02$, whereas the probability of
-drawing such a value from a normal distribution is tiny ($2.9\cdot10^{-7}$). As
-the value of $\nu$ approaches infinity, the distribution approaches the normal
-distribution (Fig.~\ref{fig:didactic}a, b).
+drawing such a value from a normal distribution is tiny ($2.9\cdot10^{-7}$).
+As the value of $\nu$ approaches infinity, the distribution approaches the
+normal distribution (Fig.~\ref{fig:didactic}a, b). While, across populations,
+the extreme process deviations might tend to be more frequently upwards or
+downwards events, a small random sample from a heavy-tailed distribution can
+have many outliers on one side and appear asymmetric \citep{gelman2013a}.
+Therefore, fitting a symmetric t distribution is appropriate for the
+relatively short population time series in our dataset and is agnostic towards
+the detection of upwards or downwards black swans.
 
@@
 is almost indistinguishable from the normal distribution
-(Fig.~\ref{fig:didactic}). We investigated the sensitivity of our results to
-weaker and stronger priors (exponential rate parameter $= 0.005, 0.02$;
-Fig.~\ref{fig:priors}, Supporting Information). We used simulated data to test
-how easily we could detect $\nu$ given different sample sizes and to ensure we
-could recover unbiased parameter estimates from the Gompertz model (Supporting
-Information).
+(Fig.~\ref{fig:didactic}). Based on the shape of the t distribution, we chose
+the probability of $\nu < 10$, Pr($\nu < 10$), to define the probability of
+heavy-tailed dynamics. When categorizing a population as heavy or normal tailed,
+we used a threshold of 0.5 probability.
 
@@
 
+In addition to alternative population models, we investigated the sensitivity
+of our results to weaker and stronger priors (exponential rate parameter $=
+0.005, 0.02$; Fig.~\ref{fig:priors}, Supporting Information). We used
+simulated data to test how easily we could detect $\nu$ given different sample
+sizes and to ensure we could recover unbiased parameter estimates from the
+Gompertz model (Supporting Information).
+
 \subsection{Covariates of population dynamic black swans}
@@
 10)$. We formally investigated these relationships by fitting beta
-regression multilevel models \citep{ferrari2004}. Beta regression is useful
-for continuous response data that range between zero and one (e.g.\
-probabilities); we fit our models with a logit link as is common for logistic
+regression multilevel models \citep{ferrari2004}. The beta probability
+distribution can represent continuous response data that range between zero
+and one; we fit our models with a logit link as is common for beta and logistic
 regression \citep{ferrari2004}. We incorporated standard deviations around the
@@
 
-%TODO: where did this threshold come from?  (nu < 10)
 We found strong, but rare, evidence for black-swan population dynamics. By
@@
 orders. Accounting for time series length and partially pooling inference
-across taxonomic class and order with a multilevel model, there was stronger
-evidence for black swans in insect populations than is visually apparent in
-Fig.~\ref{fig:nu-coefs}---four of 10 orders with the highest median
-probability of heavy tails were insect orders---however, there was
-considerable uncertainty in these estimates (Fig.~\ref{fig:posteriors}a).
-%TODO: these multilevel modelling results (and fig 3, 4) are out of order
+across taxonomic class and order with a multilevel model (Supporting
+Information), there was stronger evidence for black swans in insect
+populations than is visually apparent in Fig.~\ref{fig:nu-coefs}---four of 10
+orders with the highest median probability of heavy tails were insect
+orders---however, there was considerable uncertainty in these estimates
+(Fig.~\ref{fig:posteriors}a).
 
@@
 events occurred (Tables~\ref{tab:sparks} and \ref{tab:causes-supp}). The
-majority (\percBSDown \%) of documented events were downward black swans and
-%TODO: what else would cause a black swan event? How defined "unfortunate"?
-involved a series of unfortunate events. For example, a shortage of nest sites
-%TODO how about just calling it a piscivorous waterfowl or something like that
-for the shag long-necked bird (\textit{Phalacrocorax aristotelis}) in the
-%TODO why did this make the population more vulnerable?
-Seems like it wouldn't make it more vulnerable I guess I am trying to figure out if these are additive or synergistic.
-I am guessing you want to avoid those terms, but I might just be more straight with language and avoid the "more vulnerable" phrase.
-United Kingdom reduced population productivity and made the population more
-vulnerable when a red tide hit in 1968 \citep{potts1980}. The population
-experienced a black-swan downswing, but the new availability of nest sites
-resulted in a rapid population upswing \citep{potts1980}. As another example,
-% TODO true interaction term or just synchronous events?
-an interaction between environmental- and predation-mediated population cycles
-are thought to have caused a downward black-swan event for a water vole
-(\textit{Arvicola terrestris}) population \citep{saucy1994}. Other black swans
-were the result of a sequence of extreme climate events on their own. For
-instance, severe winters in 1929, 1940--1942, and 1962--1963 were associated
-with black-swan downswings in grey heron (\textit{Ardea cinerea}) abundance in
-the United Kingdom \citep{stafford1971}. Our analysis finds that the last event
-was a combination of two black-swan events in a row and it took the population
-three times longer to recover than predicted \citep{stafford1971}.
-\textit{(TODO: substitute a parasite example?)}
+majority of documented events (\percBSDown \%) were downward black swans and
+involved a combination of multiple factors. For example, a synchronization of
+environmental- and predation-mediated population cycles are thought to have
+caused a downward black-swan event for a water vole (\textit{Arvicola
+  terrestris}) population \citep{saucy1994}. Other black swans were the result
+of a sequence of extreme climate events on their own. For instance, severe
+winters in 1929, 1940--1942, and 1962--1963 were associated with black-swan
+downswings in grey heron (\textit{Ardea cinerea}) abundance in the United
+Kingdom \citep{stafford1971}. Our analysis finds that the last event was a
+combination of two black-swan events in a row and it took the population three
+times longer to recover than predicted \citep{stafford1971}. Downwards black
+swans were sometimes followed by upwards black swans. For example, during a
+period of population crowding and nest shortages, a population of European
+shag cormorants (\textit{Phalacrocorax aristotelis}) on the Farne Islands, UK,
+declined suddenly following a red tide event in 1968 \citep{potts1980}. This
+freed quality nest sites for first-time breeders, productivity rapidly
+increased, and the population experienced a rapid upswing in abundance.
+\citep{potts1980}.
 
@@
 
-We found strong evidence for black swans (heavy-tailed process noise) in 3--5\%
-% TODO but the students t is a symmetric distribution is that cool?
-of ecological time series. Black swans were usually downward events and were
-detected more frequently in longer time series and in populations with a
-smaller magnitude of process noise. Black swans were not associated with
-density dependence, population growth rate, or lifespan. In verified cases,
-% TODO 'interaction' word
-black-swan events were often a result of the interaction between elements of
-extreme climate, predators and parasite cycles, and strong changes in human
-pressures. Our empirical results, sensitivity analyses, and simulation tests
-suggest that estimating the tail shape of process noise is a viable method of
-detecting black-swan population dynamics and if anything will underestimate the
-probability of black-swan events. The presence of black swans highlights the
-importance of developing management strategies that detect quickly, respond to,
-and are robust to extremes in population dynamics---particularly as the
-frequency and magnitude of climatic extremes increase over the next century
-\citep{easterling2000,ipcc2012}.
+We found strong evidence for black swans (heavy-tailed process noise) in
+\overallMinPerc--\overallMaxPerc\% of ecological time series. Black swans were
+usually (\percBSDown \%) downward events and were detected more frequently in
+longer time series and in populations with a smaller magnitude of process
+noise. Black swans were not associated with density dependence, population
+growth rate, or lifespan. In verified cases, black-swan events were often a
+result of the synchrony of extreme climate, predators and parasite cycles, and
+strong changes in human pressures. Our empirical results, sensitivity
+analyses, and simulation tests suggest that estimating the tail shape of
+process noise is a viable method of detecting black-swan population dynamics
+and if anything will underestimate the probability of black-swan events. The
+presence of black swans highlights the importance of developing management
+strategies that detect quickly, respond to, and are robust to extremes in
+population dynamics---particularly as the frequency and magnitude of climatic
+extremes increase over the next century \citep{easterling2000,ipcc2012}.
 
@@
 changes in human pressure such hunting, fishing, or habitat destruction might
-cause black swans. Alternatively, the interaction of multiple ``normal''
+cause black swans. Alternatively, the synchrony of multiple ``normal''
 extrinsic forces could give rise to black-swan ecological dynamics. This could
-occur if the interaction is synergistic \citep[e.g.][]{kirby2009} or even if
-non-synergistic interactions experience a rare alignment \citep{denny2009}.
+occur if with a synergistic interaction \citep[e.g.][]{kirby2009} or even if
+non-synergistic forces experience a rare alignment \citep{denny2009}.
 
@@
 \bibliography{/Users/seananderson/Dropbox/tex/jshort,/Users/seananderson/Dropbox/tex/ref3}
-%\bibliography{ms}
 
@@
 \includegraphics[width=\textwidth]{../analysis/t-nu-eg2.pdf}
-\caption{
+\caption[An illustration of fitting population dynamic models that allow for heavy
+tails, represented by the Student-t degrees of freedom parameter $\nu$.]{
 An illustration of fitting population dynamic models that allow for heavy
@@
 
-\caption{Estimates of population dynamics heavy-tailedness for \nuCoefPopN\
+\caption[Estimates of population dynamics heavy-tailedness for \nuCoefPopN\
+  populations of birds, mammals, insects, and fishes.]{Estimates of population dynamics heavy-tailedness for \nuCoefPopN\
   populations of birds, mammals, insects, and fishes. Small values of $\nu$
@@
 \begin{center}
+\includegraphics[width=0.45\textwidth]{../analysis/order-posteriors-covariates.pdf}
+
+\caption[Posterior probability distributions from beta regression multilevel
+  models.]{Posterior probability distributions from beta regression multilevel
+  models. (a) Taxonomic-order-level posterior densities of Pr($\nu < 10$)
+  (approximately the probability of heavy tails) after accounting for
+  time-series length. Estimates are at the geometric mean of time series
+  length across all the data (approximately 27 time steps). Colour shading
+  refers to taxonomic class (yellow: fishes, green: insects, purple: birds,
+  and red: mammals). Dotted vertical line in panel a indicates the median
+  expected Pr($\nu < 10$) from the prior distribution. (b) Main effect
+  posterior densities for potential covariates of Pr($\nu < 10$). The beta
+  regression models were fit on a logit scale with hierarchical intercepts for
+  taxonomic class, taxonomic order, and species. All covariates were
+  standardized by subtracting their mean and dividing by twice their standard
+  deviation. In both panels, short vertical line segments within the density
+  polygons indicate median posterior estimates.}
+
+\label{fig:posteriors}
+\end{center}
+\end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
 \includegraphics[width=0.9\textwidth]{../analysis/correlates-p10.pdf}
 
-\caption{Potential covariates of heavy-tailed population dynamics (indicated
+\caption[Potential covariates of heavy-tailed population dynamics (indicated
+  by a high probability that $\nu < 10$).]{Potential covariates of heavy-tailed population dynamics (indicated
   by a high probability that $\nu < 10$). Shown are (a--c) parameters from the
@@
 \end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=0.45\textwidth]{../analysis/order-posteriors-covariates.pdf}
-
-\caption{Posterior probability distributions from beta regression multilevel
-  models. (a) Taxonomic-order-level posterior densities of Pr($\nu < 10$)
-  (approximately the probability of heavy tails) after accounting for
-  time-series length. Estimates are at the geometric mean of time series
-  length across all the data (approximately 27 time steps). Colour shading
-  refers to taxonomic class (yellow: fishes, green: insects, purple: birds,
-  and red: mammals). Dotted vertical line in panel a indicates the median
-  expected Pr($\nu < 10$) from the prior distribution. (b) Main effect
-  posterior densities for potential covariates of Pr($\nu < 10$). The beta
-  regression models were fit on a logit scale with hierarchical intercepts for
-  taxonomic class, taxonomic order, and species. All covariates were
-  standardized by subtracting their mean and dividing by twice their standard
-  deviation. In both panels, short vertical line segments within the density
-  polygons indicate median posterior estimates.}
-
-\label{fig:posteriors}
-\end{center}
-\end{figure}

### 2014-11-24 text/sparks.tex
Edits based on Jon's comments

@@
 
-\caption{Example population dynamic black swans from the Global Population
+\caption[Example population dynamic black swans from the Global Population
+  Dynamics Database and a description of their causes.]{Example population dynamic black swans from the Global Population
   Dynamics Database and a description of their causes. Red and blue dots

### 2014-11-24 text/supp.tex
Edits based on Jon's comments

@@
 
-\caption{Summary statistics for the filtered Global Population Dynamics
+\caption[Summary statistics for the filtered Global Population Dynamics
+  Database time series arranged by taxonomic class.]{Summary statistics for the filtered Global Population Dynamics
   Database time series arranged by taxonomic class. Columns are: number of
@@
 
-\caption{All populations with Pr$(\nu < 10) > 0.5$ in the base heavy-tailed
+\caption[All populations with Pr$(\nu < 10) > 0.5$ in the base heavy-tailed
+  Gompertz population dynamics model.]{All populations with Pr$(\nu < 10) > 0.5$ in the base heavy-tailed
   Gompertz population dynamics model. Shown are the log abundance time series,
@@
 
-\caption{(fishes, crustaceans, gastropods, sharks). All filtered time series
-  used in our analysis. The abundances are shown on a log10 vertical axis.
-  Throughout this figure, red dots indicate values that were interpolated and
-  blue dots indicate values that were recorded as zero but were set to the
-  next lowest observed abundance. Numbers before each species name are the
-  GPDD Main ID numbers.}
+\caption[All filtered time series used in our analysis.]{(fishes, crustaceans,
+  gastropods, sharks). All filtered time series used in our analysis. The
+  abundances are shown on a log10 vertical axis. Throughout this figure, red
+  dots indicate values that were interpolated and blue dots indicate values
+  that were recorded as zero but were set to the next lowest observed
+  abundance. Numbers before each species name are the GPDD Main ID numbers.}
 
@@
 
-\caption{ Probability density of the Bayesian priors for the Gompertz models.
-  From left to right and then top to bottom: (1) per capita growth rate at
+\caption[Probability density of the Bayesian priors for the Gompertz models.]{
+  Probability density of the Bayesian priors for the Gompertz models. From
+  left to right and then top to bottom: (1) per capita growth rate at
   $\log$(abundance) = $0$: $\lambda \sim \mathrm{Normal}(0, 10^2)$; (2) scale
@@
 
-\caption{Estimates of $\nu$ from alternative models plotted against the base
-  Gompertz model estimates of $\nu$. Shown are medians of the posterior (dots)
-  and 50\% credible intervals (segments). The diagonal line indicates a
-  one-to-one relationship. Different colours indicate various taxonomic
-  classes. The grey-shaded regions indicate regions of disagreement if $\nu =
-  10$ is taken as a threshold of heavy-tailed dynamics. The Gompertz
-  observation error model assumes a fixed standard deviation of observation
-  error of $0.2$ on a log scale.}
+\caption[Estimates of $\nu$ from alternative models plotted against the base
+Gompertz model estimates of $\nu$.]{Estimates of $\nu$ from alternative models
+  plotted against the base Gompertz model estimates of $\nu$. Shown are
+  medians of the posterior (dots) and 50\% credible intervals (segments). The
+  diagonal line indicates a one-to-one relationship. Different colours
+  indicate various taxonomic classes. The grey-shaded regions indicate regions
+  of disagreement if $\nu = 10$ is taken as a threshold of heavy-tailed
+  dynamics. The Gompertz observation error model assumes a fixed standard
+  deviation of observation error of $0.2$ on a log scale.}
 
@@
 
-\caption{Estimates of $\nu$ from Gompertz models with alternative priors on
+\caption[Estimates of $\nu$ from Gompertz models with alternative priors on
+$\nu$.]{Estimates of $\nu$ from Gompertz models with alternative priors on
   $\nu$. Shown are medians of the posterior (dots) and 50\% credible intervals
@@
 
-\caption{Testing the ability to estimate $\nu$ (top panels) and the scale
-  parameter of the process deviations (bottom panels) for a given number of
-  samples (columns) drawn from a distribution with a given true $\nu$ value
-  (rows). The red lines indicate the true population value. When a small
-  number of samples are drawn there may not be samples sufficiently far into
-  the tails to recapture the true $\nu$ value; however, heavy tails are still
-  distinguished from normal tails in most cases, even with only 25 or 50
-  samples.}
+\caption[Testing the ability to estimate $\nu$  and the scale parameter of the
+process deviations for a given number of samples drawn from a distribution
+with a given true $\nu$ value.]{Testing the ability to estimate $\nu$ (top
+  panels) and the scale parameter of the process deviations (bottom panels)
+  for a given number of samples (columns) drawn from a distribution with a
+  given true $\nu$ value (rows). The red lines indicate the true population
+  value. When a small number of samples are drawn there may not be samples
+  sufficiently far into the tails to recapture the true $\nu$ value; however,
+  heavy tails are still distinguished from normal tails in most cases, even
+  with only 25 or 50 samples.}
 
@@
 
-\caption{ Simulation testing the Gompertz estimation model when the process
-  deviation draws were chosen so that $\nu$ could be estimated close to the
-  true value outside the full population model (``effective $\nu$'' within a
-  CV of 0.2 of specified $\nu$). Upper panels show the distribution of median
-  $\widehat{\nu}$ across 20 simulation runs. Lower panels show the
-  distribution of Pr($\nu < 10$) across 20 simulation runs. We ran the
-  simulations across three population (``true'') $\nu$ values (3, 5, and
-  $1\cdot 10^9$, i.e.\ approximately normal) and four scenarios: (1) 100 time
-  steps and no observation error, (2) 50 time steps and no observation error,
-  (3) 50 time steps and observation error drawn from $\mathrm{Normal} (0,
+\caption[Simulation testing the Gompertz estimation model when the process
+deviation draws were chosen so that $\nu$ could be estimated close to the true
+value outside the full population model (``effective $\nu$'' within a CV of
+0.2 of specified $\nu$).]{Simulation testing the Gompertz estimation model
+  when the process deviation draws were chosen so that $\nu$ could be
+  estimated close to the true value outside the full population model
+  (``effective $\nu$'' within a CV of 0.2 of specified $\nu$). Upper panels
+  show the distribution of median $\widehat{\nu}$ across 20 simulation runs.
+  Lower panels show the distribution of Pr($\nu < 10$) across 20 simulation
+  runs. We ran the simulations across three population (``true'') $\nu$ values
+  (3, 5, and $1\cdot 10^9$, i.e.\ approximately normal) and four scenarios:
+  (1) 100 time steps and no observation error, (2) 50 time steps and no
+  observation error, (3) 50 time steps and observation error drawn from
+  $\mathrm{Normal} (0,
   0.2^2)$ but ignored, and (4) 50 time steps with observation error in which
@@
     the dots represent stochastic draws from the true population distributions
-  combined with model fits. Underlayed boxplots show the median, interquartile
-  range, and $1.5$ times the interquartile range. }
+    combined with model fits. Underlayed boxplots show the median,
+    interquartile range, and $1.5$ times the interquartile range. }
 

### 2014-11-25 text/ms.tex
Switch back bib style and bump version
A round of minor edits

@@
 frequency, it remains unclear the extent to which ecological populations
-buffer or suffer from such extremes. Here, we develop a probability model to
-estimate the degree of heavy-tailedness (presence of black swans) in
-ecological process noise. We apply our model to \NPops~time series from around
-the world across \NOrders~taxonomic orders and seven classes. We find strong
-evidence of black swans, but they are rare, occurring in
+buffer or suffer from such extremes. Here, estimate the degree of
+heavy-tailedness (presence of black swans) in ecological process noise by
+applying a probability model to \NPops~time series from around the world
+across \NOrders~taxonomic orders and seven classes. We find strong evidence of
+black swans, but they are rare, occurring in
 \overallMinPerc--\overallMaxPerc\% of populations: most frequently for birds
@@
 consequences \citep{taleb2007,sornette2009}. One of the most striking black
-swans in ecology is the possible asteroid collision marking the end-Cretaceous
-mass extinction 65 million years ago \citep{alvarez1980,harnik2012}. Today,
-climate extremes, in concert with shifts in mean temperature, are expected to
-cause the greatest ecological and societal damage \citep{ipcc2012}. But, while
-extremes in the physical environment such as wave height, storm severity, and
-temperature are frequent events \citep{gaines1993,katz2005}, it remains
-unclear the extent to which ecological systems buffer or suffer from black
-swans \citep{nunez2012}.
+swans in ecology is the asteroid collision that may have marked the
+end-Cretaceous mass extinction 65 million years ago
+\citep{alvarez1980,harnik2012}. Today, climate extremes, in concert with
+shifts in mean temperature, are expected to cause the greatest ecological and
+societal damage \citep{ipcc2012}. But, while extremes in the physical
+environment such as wave height, storm severity, and temperature are frequent
+events \citep{gaines1993,katz2005}, it remains unclear the extent to which
+ecological systems buffer or suffer from black swans \citep{nunez2012}.
 
@@
 \citep{segura2013} and time to extinction for experimental waterflea
-populations \citep{drake2014} have been found to have heavier tailed
-distributions than the normal. Despite these anecdotes, as far as we can tell
+populations \citep{drake2014} have been found to follow heavier tailed
+distributions than the normal. Despite these examples, as far as we can tell
 there have been two systematic surveys for ecological black swans: one with
@@
 The growth parameter $\lambda$ represents the expected growth rate if $N_t =
-1$. The model is density independent if $b = 1$, maximally density dependent if
-$b = 0$, and inversely density dependent if $b < 0$. Usually, the process noise
-$\epsilon_t$ is modelled as normally distributed, but in our paper we assume it
-is drawn from a t distribution with scale parameter $\sigma$ and degrees of
-freedom $\nu$. In previous analyses of the GPDD, the Gompertz was most often
-identified as the most parsimonious population model that can be fit to these
+1$. The model is density independent if $b = 1$, maximally density dependent
+if $b = 0$, and inversely density dependent if $b < 0$. Usually, the process
+noise $\epsilon_t$ is modelled as normally distributed, but in our paper we
+assume it is drawn from a t distribution with scale parameter $\sigma$ and
+degrees of freedom $\nu$. In previous analyses of the GPDD, the Gompertz was
+most often identified as the most parsimonious population model fit to these
 data \citep{brook2006}.
@@
 (Fig.~\ref{fig:didactic}). Based on the shape of the t distribution, we chose
-the probability of $\nu < 10$, Pr($\nu < 10$), to define the probability of
+the probability that $\nu < 10$, Pr($\nu < 10$), to define the probability of
 heavy-tailed dynamics. When categorizing a population as heavy or normal tailed,
@@
 deviations on a log scale). Finally, the Gompertz model assumes that
-population growth rate declines linearly with log abundance. We also fit an
-alternative model, the Ricker-logistic model, which assumes that population
-growth rate declines linearly with abundance itself (Supporting Information).
+population growth rate declines linearly with log abundance. Therefore, we
+also fit an alternative model, the Ricker-logistic model, which assumes that
+population growth rate declines linearly with abundance itself (Supporting
+Information).
 
@@
 period of population crowding and nest shortages, a population of European
-shag cormorants (\textit{Phalacrocorax aristotelis}) on the Farne Islands, UK,
-declined suddenly following a red tide event in 1968 \citep{potts1980}. This
-freed quality nest sites for first-time breeders, productivity rapidly
-increased, and the population experienced a rapid upswing in abundance.
-\citep{potts1980}.
+shag cormorants (\textit{Phalacrocorax aristotelis}) on the Farne Islands,
+United Kingdom, declined suddenly following a red tide event in 1968
+\citep{potts1980}. This freed quality nest sites for first-time breeders,
+productivity rapidly increased, and the population experienced a rapid upswing
+in abundance. \citep{potts1980}.
 
@@
 growth rate, or lifespan. In verified cases, black-swan events were often a
-result of the synchrony of extreme climate, predators and parasite cycles, and
-strong changes in human pressures. Our empirical results, sensitivity
+result of the combination of extreme climate, predators and parasite cycles,
+and strong changes in human pressures. Our empirical results, sensitivity
 analyses, and simulation tests suggest that estimating the tail shape of
@@
 and recreationally important species and a disproportionate number of
-populations are located in the United Kingdom. Although we would expect to find
-qualitatively similar evidence for black swans in many other large taxonomic or
-geographic samples of populations, the common forces driving those black swans
-would likely differ (e.g.~severe winters in Table~\ref{tab:sparks}). In
-addition to a possibly biased sample of populations, some black-swan detections
-could just be recording mistakes, or conversely, some extreme observations may
-have been discarded or altered if they were erroneously suspected of being
-recording mistakes. Indeed, three of the populations our method initially
-identified as heavy tailed turned out to be data-entry errors and were
-discarded (Supporting Information). A final caveat is that the temporal scales
-of observation and population dynamics vary considerably across populations in
-the GPDD and this likely influences the detection of heavy tails. As an
-example, if we make frequent observations relative to generation time (e.g.~for
-many large-bodied mammals) we will average across generations and perhaps miss
-black swans. Conversely, if we census populations infrequently relative to
-generation time (e.g.~many insects in the GPDD) the recorded data may average
-across extreme and less-extreme events and also dampen black-swan dynamics.
+populations are located in the United Kingdom. Although we would expect to
+find qualitatively similar evidence for black swans in many other large
+taxonomic or geographic samples of populations, the common forces driving
+those black swans would likely differ (e.g.~severe winters in
+Table~\ref{tab:sparks}). In addition to a possibly biased sample of
+populations, some black-swan detections could just be recording mistakes, or
+conversely, some extreme observations may have been discarded or altered if
+they were erroneously suspected of being recording mistakes. Indeed, we
+discarded three of the populations that our method initially identified as
+heavy tailed because they turned out to be data-entry errors (Supporting
+Information). A final caveat is that the temporal scales of observation and
+population dynamics vary considerably across populations in the GPDD and this
+likely influences the detection of heavy tails. As an example, if we make
+frequent observations relative to generation time (e.g.~for many large-bodied
+mammals) we will average across generations and perhaps miss black swans.
+Conversely, if we census populations infrequently relative to generation time
+(e.g.~many insects in the GPDD) the recorded data may average across extreme
+and less-extreme events and also dampen black-swan dynamics.
 
@@
 Most importantly, what is the impact of allowing for black swans in forecasts
-of ecological risk? In recent decades, ecology has moved toward focussing on
-aspects of variance in addition to mean responses \citep[e.g.][]{loreau2010a,
-  thompson2013}. Our results suggest that an added focus on ecological
-extremes represents the next frontier, particularly in the face of increased
-climate extremes \citep{meehl2004,thompson2013,ipcc2012}. Financial analysts
-are concerned with the shape of the tails of financial returns because these
-directly impact estimates of risk---the probability of a specific magnitude of
-undesired event occurring \citep{rachev2008}. A comparable focus in ecology
-would increase our estimates of extinction risk, since these would be
-disproportionately impacted by downward black-swan events.
+of ecological risk? Even extremely rare catastrophes can have a profound
+influence on population persistence \citep{mangel1994}. In recent decades,
+ecology has moved toward focussing on aspects of variance in addition to mean
+responses \citep[e.g.][]{loreau2010a, thompson2013}. Our results suggest that
+an added focus on ecological extremes represents the next frontier,
+particularly in the face of increased climate extremes \citep{meehl2004,
+  thompson2013, ipcc2012}. Financial analysts are concerned with the shape of
+the downward tails of financial returns because these directly impact
+estimates of risk---the probability of a specific magnitude of undesired event
+occurring \citep{rachev2008}. A comparable focus in ecology would increase our
+estimates of extinction risk, since these would be disproportionately impacted
+by downward black-swan events.
 
@@
 
-\bibliographystyle{ecologyletters}
+\bibliographystyle{apalike}
 \bibliography{/Users/seananderson/Dropbox/tex/jshort,/Users/seananderson/Dropbox/tex/ref3}

### 2014-12-04 analysis/8-sim-extinction-risk.Rmd
Stop dividing by N where we shouldn't

@@
   u <- qnorm(0.9999, 0, sd = 1)
-  downs <- sum(proc_error < l) / N
-  ups <- sum(proc_error > u) / N
+  downs <- sum(proc_error < l)
+  ups <- sum(proc_error > u)
   downs / (downs + ups)

### 2014-12-04 text/ms.tex
Tweak text to work with thesis

@@
 
+\renewcommand{\baselinestretch}{\tighttextstretch}
+\normalsize
 \bibliographystyle{apalike}
@@
 \clearpage
+\renewcommand{\baselinestretch}{\textstretch}
+\normalsize
 
@@
   populations of birds, mammals, insects, and fishes. Small values of $\nu$
-  ($\lesssim 10$) suggest heavy-tailed black-swan dynamics; larger values of
+  approximately ($< 10$) suggest heavy-tailed black-swan dynamics; larger values of
   $\nu$ suggest approximately normal-tailed dynamics. Vertical points and line

### 2014-12-11 text/supp.tex
Switch back bib style and bump version
Tweak text to work with thesis
Remove a causes table column and reduce width

@@
 \LTcapwidth=\textwidth
-\singlespacing
+%\singlespacing
 \begin{footnotesize}
@@
 \end{footnotesize}
-\onehalfspacing
+%\onehalfspacing
 
-\bibliographystyle{ecologyletters}
+\renewcommand{\baselinestretch}{\tighttextstretch}
+\normalsize
+\bibliographystyle{apalike}
 \bibliography{/Users/seananderson/Dropbox/tex/jshort,/Users/seananderson/Dropbox/tex/ref3}
+\renewcommand{\baselinestretch}{\textstretch}
+\normalsize
 
@@
 
-\singlespacing
+%\singlespacing
 \begin{footnotesize}
-\begin{longtable}{>{\RaggedRight}m{1.5cm}>{\RaggedRight}p{4.3cm}>{\RaggedRight}p{0.8cm}>{\RaggedRight}p{1.7cm}>{\RaggedRight}p{1.0cm}>{\RaggedRight}p{3.0cm}>{\RaggedRight}p{1.7cm}>{\RaggedRight}p{1.3cm}}
+\begin{longtable}{>{\RaggedRight}m{1.4cm}>{\RaggedRight}p{3.3cm}>{\RaggedRight}p{0.7cm}>{\RaggedRight}p{1.5cm}>{\RaggedRight}p{3.0cm}>{\RaggedRight}p{1.4cm}>{\RaggedRight}p{1.25cm}}
 
@@
   parentheses. Red dots on the time series indicate downward black-swan events
-  and blue values indicate upward black-swan events that have a $10 \cdot
-  10^{-4}$ probability or less of occurring if the population dynamics were
+  and blue values indicate upward black-swan events that have a $10^{-4}$ probability or less of occurring if the population dynamics were
   explained by a Gompertz model with normally distributed process noise with a
@@
 \end{footnotesize}
-\onehalfspacing
+%\onehalfspacing
 
@@
 
-\begin{spacing}{1.15}
+%\begin{spacing}{1.15}
 \begin{footnotesize}
@@
 
+\renewcommand{\baselinestretch}{\tighttextstretch}
+\normalsize
 \begin{footnotesize}
@@
 \end{footnotesize}
-\end{spacing}
+\renewcommand{\baselinestretch}{\textstretch}
+\normalsize
+%\end{spacing}

### 2014-12-11 analysis/6-causes-table.R

@@ -25,9 +25,10 @@ heavy <- mutate(heavy,
   mutate(nu_95 = sprintf("%.0f", round(nu_95, 0))) %>%
   mutate(spark = paste0("\\includegraphics[width=1.7cm]{../analysis/sparks/", main_id, ".pdf}")) %>%
   mutate(nu_hat = paste0(nu_50, " (", nu_5, "--", nu_95, ")")) %>%
-  select(spark, population, main_id, citation, data_correct, reasons, p10, nu_hat) %>%
-  rename("Time series" = spark, "Population" = population, "ID" = main_id, "Verified" = data_correct,
-    "Description" = reasons, "Pr($\\nu < 10$" = p10, "$\\widehat{\\nu}$" = nu_hat)
+  select(spark, population, main_id, citation, reasons, p10, nu_hat) %>%
+  rename("Time series" = spark, "Population" = population, "ID" = main_id,
+    "Citation" = citation,
+    "Description" = reasons, "Pr($\\nu < 10$)" = p10, "$\\widehat{\\nu}$" = nu_hat)
 
 print.xtable(xtable(heavy,
     caption = ""),

### 2014-12-11 text/supp.tex
Remove a comment in a Stan model

@@
 
-\section{Data selection}
+\subsection{Data selection}
 
@@
 
-\section{Details on the heavy-tailed Gompertz probability model}
+\subsection{Details on the heavy-tailed Gompertz probability model}
 
@@
 
-\section{Alternative priors}
+\subsection{Alternative priors}
 
@@
 
-\section{Alternative population models}
+\subsection{Alternative population models}
 
@@
 
-\subsection{Autocorrelated residuals}
+\subsubsection{Autocorrelated residuals}
 
@@
 
-\subsection{Assumed density independence}\label{assumed-density-independence}
+\subsubsection{Assumed density independence}\label{assumed-density-independence}
 
@@
 
-\subsection{Assumed observation error}
+\subsubsection{Assumed observation error}
 
@@
 
-\subsection{Ricker-logistic}
+\subsubsection{Ricker-logistic}
 
@@
 
-\section{Simulation testing the model}
+\subsection{Simulation testing the model}
 
@@
 
-\subsection{Estimating $\nu$ from a stationary t distribution}
+\subsubsection{Estimating $\nu$ from a stationary t distribution}
 
@@
 
-\subsection{Heavy-tailed Gompertz model simulations}
+\subsubsection{Heavy-tailed Gompertz model simulations}
 
@@
 
-\section{Modelling covariates of heavy-tailed dynamics}
+\subsection{Modelling covariates of heavy-tailed dynamics}
 
@@
 
-\section{Additional acknowledgements}
+\subsection{Additional acknowledgements}
 
@@
 
-\renewcommand{\baselinestretch}{\tighttextstretch}
-\normalsize
-\bibliographystyle{apalike}
-\bibliography{/Users/seananderson/Dropbox/tex/jshort,/Users/seananderson/Dropbox/tex/ref3}
-\renewcommand{\baselinestretch}{\textstretch}
-\normalsize
-
+% \renewcommand{\baselinestretch}{\tighttextstretch}
+% \normalsize
+% \bibliographystyle{apalike}
+% \bibliography{/Users/seananderson/Dropbox/tex/jshort,/Users/seananderson/Dropbox/tex/ref3}
+% \renewcommand{\baselinestretch}{\textstretch}
+% \normalsize
+%
 % ------------------------------
@@
 
-\renewcommand{\baselinestretch}{\tighttextstretch}
-\normalsize
-\begin{footnotesize}
+\begin{scriptsize}
+%\renewcommand{\baselinestretch}{1.11}
 \noindent
@@
 }
-\end{footnotesize}
-\renewcommand{\baselinestretch}{\textstretch}
+%\renewcommand{\baselinestretch}{\textstretch}
+\end{scriptsize}
 \normalsize

### 2014-12-11 analysis/cause-table.tex
Add the rendered causes table with column removed

--- 2014-11-08 analysis/cause-table.tex
@@
 % latex table generated in R 3.1.2 by xtable 1.7-4 package
-% Fri Nov  7 16:47:38 2014
-Time series & Population & ID & citation & Verified & Description & Pr($\nu < 10$ & $\widehat{\nu}$ \\ 
+Time series & Population & ID & Citation & Description & Pr($\nu < 10$) & $\widehat{\nu}$ \\ 
   \midrule
-\includegraphics[width=1.7cm]{../analysis/sparks/6528.pdf} & Shag, \textit{Phalacrocorax aristotelis}, Farne Islands, Northumberland & 6528 & \citep{potts1980} & yes & red tide & 1.00 & 2.4 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/7115.pdf} & South African fur seal, \textit{Arctocephalus pusillus}, South Africa & 7115 & \citep{shaughnessy1982} & yes?
-& harvesting and predation changes & 1.00 & 2.5 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/10128.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Scotland - un-named area & 10128 & \citep{potts1984} & yes & environmental- and parisitic-caused cycles & 1.00 & 2.6 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/9382.pdf} & Pine looper or Bordered white, \textit{Bupalus piniaria}, Kessock & 9382 & \citep{broekhuizen1993} & yes &  & 1.00 & 2.7 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/10127.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Scotland - un-named area & 10127 & \citep{potts1984} & yes & environmental- and parisitic-caused cycles & 0.99 & 2.7 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/10007.pdf} & Water vole, \textit{Arvicola terrestris}, Le Pont & 10007 & \citep{saucy1994} & yes & predator-environment cycle interactions & 1.00 & 2.7 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/20579.pdf} & Grey heron, \textit{Ardea cinerea}, Southern Britain & 20579 & \citep{stafford1971} & yes? & severe winter & 0.98 & 3.0 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/9655.pdf} & Flea beetle, \textit{Chaetocnoma concinna}, Finland & 9655 & \citep{markkula1965} &  &  & 0.99 & 3.0 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/1235.pdf} & Wren, \textit{Troglodytes troglodytes}, Eastern Wood, Bookham Common & 1235 & \citep{newton1998} & yes? & severe winter & 0.98 & 3.0 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/10113.pdf} & Willow grouse, \textit{Lagopus lagopus}, Northern England & 10113 & \citep{dobson1995} & yes? & parasites and predators & 0.99 & 3.1 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/9667.pdf} & Gooseberry sawfly, \textit{Nemastus ribesii}, Finland & 9667 & \citep{markkula1965} &  &  & 0.97 & 3.2 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/9679.pdf} & Unknown, \textit{Trioza apicalis}, Finland & 9679 & \citep{markkula1965} &  &  & 0.83 & 3.4 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/20527.pdf} & Wandering albatross, \textit{Diomedea exulans}, Taiaroa & 20527 & \citep{robertson1998} & IN LIBRARY &  & 0.99 & 3.5 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/10039.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Northern Scotland & 10039 & \citep{dobson1995} & yes & parasites and predators & 0.92 & 3.9 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/10162.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Atholl Estate & 10162 & \citet{mackenzie1952} & yes & overcrowding and environemnt & 0.91 & 4.6 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/9503.pdf} & Fisher or  Pekan, \textit{Martes pennanti}, Manitoba & 9503 & \citep{keith1963} & IN LIBRARY &  & 0.76 & 4.8 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/7099.pdf} & European rabbit, \textit{Oryctolagus cuniculus}, Estate 2, East Anglia & 7099 & \citep{barnes1986} & yes & disease outbreak followed by years of good weather & 0.78 & 4.8 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/2778.pdf} & Wheatear, \textit{Oenanthe oenanthe}, Skokholm Island & 2778 & \citep{lack1969} & yes & unknown, but decline noted specifically, cold winters later & 0.67 & 5.3 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/9659.pdf} & Cabbage root fly or maggot, \textit{Delia radicum}, Finland & 9659 & \citep{markkula1965} &  &  & 0.59 & 6.8 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/1195.pdf} & Blue jay, \textit{Cyanocitta cristata}, Robert Allerton Park & 1195 & \citep{kendeigh1982} & yes & unknown & 0.60 & 6.9 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/5019.pdf} & Barbary macaque, \textit{Macaca sylvanus}, Queens Gate & 5019 & \citep{fa1984} &  &  & 0.61 & 6.9 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/9953.pdf} & Rock ptarmigan, \textit{Lagopus mutus}, Iceland & 9953 & \citep{clarke1885,williams1954} &  & severe winters & 0.58 & 7.2 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/6548.pdf} & Lesser-spotted dogfish, \textit{Scyliorhinus caniculus}, North Sea & 6548 & \citep{heessen1996} & yes & unknown & 0.53 & 8.5 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/20546.pdf} & American red fox, \textit{Vulpes fulva}, Labrador & 20546 & \citep{dancona1954,lindstrom1994} & yes & predator-prey cycles & 0.57 & 8.7 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/9470.pdf} & Ruffed grouse, \textit{Bonasa umbellus}, Connecticut & 9470 & \citep{keith1963} & IN LIBRARY &  & 0.53 & 8.9 \\ 
+\includegraphics[width=1.7cm]{../analysis/sparks/6528.pdf} & Shag, \textit{Phalacrocorax aristotelis}, Farne Islands, Northumberland & 6528 & \citep{potts1980} & Red tide event combined with low productivity due to overcrowding & 1.00 & 2 (2--4) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/7115.pdf} & South African fur seal, \textit{Arctocephalus pusillus}, South Africa & 7115 & \citep{shaughnessy1982} & Harvesting and predation changes & 1.00 & 2 (2--4) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/10128.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Scotland - un-named area & 10128 & \citep{potts1984} & Environment- and parisite-caused cycles & 1.00 & 3 (2--4) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/9382.pdf} & Pine looper or Bordered white, \textit{Bupalus piniaria}, Kessock & 9382 & \citep{broekhuizen1993} & Unknown, but sampling intensity was decreasing & 1.00 & 3 (2--5) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/10127.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Scotland - un-named area & 10127 & \citep{potts1984} & Environment- and parisite-caused cycles & 0.99 & 3 (2--5) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/10007.pdf} & Water vole, \textit{Arvicola terrestris}, Le Pont & 10007 & \citep{saucy1994} & Predator-environment cycle interactions & 1.00 & 3 (2--5) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/20579.pdf} & Grey heron, \textit{Ardea cinerea}, Southern Britain & 20579 & \citep{stafford1971} & Severe winter & 0.98 & 3 (2--7) \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/9655.pdf} & Flea beetle, \textit{Chaetocnoma concinna}, Finland & 9655 & \citep{markkula1965} & Cannot locate original source & 0.99 & 3 (2--6) \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/1235.pdf} & Wren, \textit{Troglodytes troglodytes}, Eastern Wood, Bookham Common & 1235 & \citep{newton1998} & Severe winter & 0.98 & 3 (2--8) \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/10113.pdf} & Willow grouse, \textit{Lagopus lagopus}, Northern England & 10113 & \citep{dobson1995} & Parasites and predators & 0.99 & 3 (2--6) \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/9667.pdf} & Gooseberry sawfly, \textit{Nemastus ribesii}, Finland & 9667 & \citep{markkula1965} & Cannot locate original source & 0.97 & 3 (2--8) \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/9679.pdf} & Unknown, \textit{Trioza apicalis}, Finland & 9679 & \citep{markkula1965} & Cannot locate original source & 0.83 & 3 (2--88) \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/20527.pdf} & Wandering albatross, \textit{Diomedea exulans}, Taiaroa & 20527 & \citep{robertson1998} & Unknown & 0.99 & 4 (2--7) \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/10039.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Northern Scotland & 10039 & \citep{dobson1995} & Parasites and predators & 0.92 & 4 (2--12) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/10162.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Atholl Estate & 10162 & \citet{mackenzie1952} & Bad environmental conditions and overcrowding combined to create crashes & 0.91 & 5 (2--12) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/9503.pdf} & Fisher or  Pekan, \textit{Martes pennanti}, Manitoba & 9503 & \citep{keith1963} & Unknown & 0.76 & 5 (2--87) \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/7099.pdf} & European rabbit, \textit{Oryctolagus cuniculus}, Estate 2, East Anglia & 7099 & \citep{barnes1986} & Disease outbreak followed by years of good weather & 0.78 & 5 (2--67) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/2778.pdf} & Wheatear, \textit{Oenanthe oenanthe}, Skokholm Island & 2778 & \citep{lack1969} & Unknown, but decline noted specifically, cold winters caused some crashes & 0.67 & 5 (2--141) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/9659.pdf} & Cabbage root fly or maggot, \textit{Delia radicum}, Finland & 9659 & \citep{markkula1965} & Cannot locate original source & 0.59 & 7 (2--161) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/1195.pdf} & Blue jay, \textit{Cyanocitta cristata}, Robert Allerton Park & 1195 & \citep{kendeigh1982} & Unknown & 0.60 & 7 (2--149) \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/5019.pdf} & Barbary macaque, \textit{Macaca sylvanus}, Queens Gate & 5019 & \citep{fa1984} & Cannot locate original source & 0.61 & 7 (2--132) \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/9953.pdf} & Rock ptarmigan, \textit{Lagopus mutus}, Iceland & 9953 & \citep{clarke1885,williams1954} & Severe winters & 0.58 & 7 (2--172) \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/6548.pdf} & Lesser-spotted dogfish, \textit{Scyliorhinus caniculus}, North Sea & 6548 & \citep{heessen1996} & Unknown, not specifically mentioned & 0.53 & 8 (2--190) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/20546.pdf} & American red fox, \textit{Vulpes fulva}, Labrador & 20546 & \citep{dancona1954,lindstrom1994} & Predator-prey cycles & 0.57 & 9 (3--65) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/9470.pdf} & Ruffed grouse, \textit{Bonasa umbellus}, Connecticut & 9470 & \citep{keith1963} & Unknown & 0.53 & 9 (3--158) \\ 
    \bottomrule

### 2014-12-11 text/ms.tex
Minor edits to main text

@@
 frequency, it remains unclear the extent to which ecological populations
-buffer or suffer from such extremes. Here, estimate the degree of
+buffer or suffer from such extremes. Here, we estimate the degree of
 heavy-tailedness (presence of black swans) in ecological process noise by
@@
 \overallMinPerc--\overallMaxPerc\% of populations: most frequently for birds
-(\AvesRangePerc\%) followed by mammals (\MammaliaRangePerc\%), and insects
+(\AvesRangePerc\%) followed by mammals (\MammaliaRangePerc\%) and insects
 (\InsectaRangePerc\%). Black swans were predominantly (\percBSDown \%)
@@
 Across populations, the probability of observing black-swan dynamics was
-positively related to time-series length and negatively related to magnitude of
-process noise ($\sigma$) but not clearly related to population growth rate
+positively related to time-series length and negatively related to magnitude
+of process noise ($\sigma$) but not clearly related to population growth rate
 ($\lambda$), density dependence ($b$), or maximum lifespan
-(Figs~\ref{fig:correlates}, \ref{fig:posteriors}b). Longer time-series length
-was the strongest covariate of observing black-swan dynamics. For instance, we
-were about \pIncHeavyNThirtyNSixty~times as likely to observe heavy tails in a
-population with 60 time steps compared to one with 30 time steps. However, the
-absolute change in probability with increased time series length was small
-($\pHeavyNSixty$ vs.\ $\pHeavyNThirty$ in the previous example,
-Fig.~\ref{fig:correlates}).
+(Figs~\ref{fig:posteriors}b, \ref{fig:correlates}). Longer time-series length
+was the strongest covariate of observing black-swan dynamics. For instance,
+the expected probability density below $\nu = 10$ was about \pIncHeavyNThirtyNSixty~times
+greater for a population with 60 time steps compared
+to one with 30 time steps. However, the absolute change in probability with
+increased time series length was small ($\pHeavyNSixty$ vs.\ $\pHeavyNThirty$
+in the previous example, Fig.~\ref{fig:correlates}).
 
@@
 extrinsic forces could give rise to black-swan ecological dynamics. This could
-occur if with a synergistic interaction \citep[e.g.][]{kirby2009} or even if
+occur with a synergistic interaction \citep[e.g.][]{kirby2009} or even if
 non-synergistic forces experience a rare alignment \citep{denny2009}.
@@
 particularly in the face of increased climate extremes \citep{meehl2004,
-  thompson2013, ipcc2012}. Financial analysts are concerned with the shape of
+  ipcc2012, thompson2013}. Financial analysts are concerned with the shape of
 the downward tails of financial returns because these directly impact
@@
 
-We thank J.D. Yeakel and other members of the Earth to Ocean research
-group for helpful discussions and comments. We are grateful to the
+We thank J.W. Moore, J.D. Yeakel, and other members of the Earth to Ocean
+research group for helpful discussions and comments. We are grateful to the
 contributors and maintainers of the Global Population Dynamics Database and to

### 2014-12-11 text/sparks.tex
Fix 10*10^ should just be 10^

@@
   Dynamics Database and a description of their causes. Red and blue dots
-  indicate downward and upward events that have a $10 \cdot 10^{-4}$ probability or less of
+  indicate downward and upward events that have a $10^{-4}$ probability or less of
   occurring if the population dynamics were explained by a Gompertz model

### 2014-12-11 text/ms.tex
Tweak text for thesis formatting, see notes
- Supporting Information -> Supporting materials
- Figures weaved throughout
- Embed som.tex at end

@@
 time series of abundance from $\sim$1000 species and $\sim$100 taxonomic
-orders. We filtered the data (Supporting Information) to remove populations
+orders. We filtered the data (Supporting materials) to remove populations
 from less reliable data sources, and those without sufficient data for our
@@
 \citep[\textit{sensu}][]{brook2006a}. Our interpolation affected only
-$\sim$\interpPointsPerc \% of the final data points (Table~\ref{tab:stats})
+$\sim$\interpPointsPerc \% of the final data points (Supporting materials~Table~\ref{tab:stats})
 and none of the data points that were later considered black-swan events. Our
@@
 and seven taxonomic classes, with a median of \medianTimeSteps~time steps
-(range of \minTimeSteps--\maxTimeSteps) (Table~\ref{tab:stats},
+(range of \minTimeSteps--\maxTimeSteps) (Supporting materials Table~\ref{tab:stats},
 Fig.~\ref{fig:all-ts}).
@@
 
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=\textwidth]{../analysis/t-nu-eg2.pdf}
+\caption[An illustration of fitting population dynamic models that allow for heavy
+tails, represented by the Student-t degrees of freedom parameter $\nu$.]{
+An illustration of fitting population dynamic models that allow for heavy
+tails, represented by the Student-t degrees of freedom parameter $\nu$. (a, b)
+The probability density for t distributions with a scale parameter of 1 and
+different values of $\nu$. Small values of $\nu$ create heavy tails. As $\nu$
+approaches infinity the distribution approaches the normal distribution. For
+example, at $\nu = 2$, the probability of drawing a value more than five
+standard deviations below the mean is $0.02$, whereas the probability of drawing
+such a value from a normal distribution is nearly zero ($2.9\cdot10^{-7}$).
+(c--e) Simulated population dynamics from a Gompertz model with process noise
+drawn from t distributions with three different values of $\nu$. Coloured dots
+in panels c and d represent jumps with less than a 1 in 1000 chance of
+occurring in a normal distribution. (f--h) Estimates of $\nu$ from models fit
+to the times series in panels c--e. Shown are the posterior samples
+(histograms), median and interquartile range of the posterior (IQR) (dots and
+line segments), and the exponential prior on $\nu$ (dashed lines). Colour
+shading behind panels f--h illustrates the region of heavy tails.
+}
+\label{fig:didactic}
+\end{center}
+\end{figure}
+
 We fit all models in a Bayesian framework using Stan \citep{stan-manual2014}
@@
 obtained sufficient independent samples from the posterior ($\widehat{R} <
-1.05$, $n_\mathrm{eff} > 200$; Supporting Information).
+1.05$, $n_\mathrm{eff} > 200$; Supporting materials).
 
@@
 population growth rate declines linearly with abundance itself (Supporting
-Information).
+materials).
 
@@
 of our results to weaker and stronger priors (exponential rate parameter $=
-0.005, 0.02$; Fig.~\ref{fig:priors}, Supporting Information). We used
+0.005, 0.02$; Supporting materials Fig.~\ref{fig:priors}, Supporting materials). We used
 simulated data to test how easily we could detect $\nu$ given different sample
 sizes and to ensure we could recover unbiased parameter estimates from the
-Gompertz model (Supporting Information).
+Gompertz model (Supporting materials).
 
@@
 allowed for hierarchical intercepts at the taxonomic class, order, and species
-level (Supporting Information). We fit our model in Stan with weakly
+level (Supporting materials). We fit our model in Stan with weakly
 informative priors on the coefficients \citep{gelman2008d} and variance
-parameters (\citeauthor{gelman2006c} \citeyear{gelman2006c}
-\citeauthor{gelman2014} \citeyear{gelman2014}; Supporting Information).
+parameters (\citeauthor{gelman2006c} \citeyear{gelman2006c};
+\citeauthor{gelman2014} \citeyear{gelman2014}; Supporting materials).
 
@@
 for birds (\birdPH\%) followed by mammals (\mammalsPH\%), and insects
-(\insectsPH\%) (Fig.~\ref{fig:nu-coefs}, Table~\ref{tab:causes-supp}). Black
+(\insectsPH\%) (Fig.~\ref{fig:nu-coefs}, Supporting materials Table~\ref{tab:causes-supp}). Black
 swans were taxonomically widespread, occurring in \POrdersHeavy\% of taxonomic
@@
 across taxonomic class and order with a multilevel model (Supporting
-Information), there was stronger evidence for black swans in insect
+materials), there was stronger evidence for black swans in insect
 populations than is visually apparent in Fig.~\ref{fig:nu-coefs}---four of 10
@@
 
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=0.44\textwidth]{../analysis/nu-coefs-2.pdf}
+
+\caption[Estimates of population dynamics heavy-tailedness for \nuCoefPopN\
+  populations of birds, mammals, insects, and fishes.]{Estimates of population dynamics heavy-tailedness for \nuCoefPopN\
+  populations of birds, mammals, insects, and fishes. Small values of $\nu$
+  approximately ($< 10$) suggest heavy-tailed black-swan dynamics; larger values of
+  $\nu$ suggest approximately normal-tailed dynamics. Vertical points and line
+  segments represent posterior medians and 50\% / 90\% credible intervals for
+  individual populations. Inset plots show probability that $\nu < 10$
+  (probability of heavy tails) for populations arranged by taxonomic order and
+  sorted by decreasing mean Pr($\nu < 10$). Taxonomic orders with three or
+  fewer populations in panel a are omitted for space. Red to yellow points
+  highlight populations with a high to moderately high probability of
+  heavy-tailed black-swan dynamics.}
+
+\label{fig:nu-coefs}
+\end{center}
+\end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=0.45\textwidth]{../analysis/order-posteriors-covariates.pdf}
+
+\caption[Posterior probability distributions from beta regression multilevel
+  models.]{Posterior probability distributions from beta regression multilevel
+  models. (a) Taxonomic-order-level posterior densities of Pr($\nu < 10$)
+  (approximately the probability of heavy tails) after accounting for
+  time-series length. Estimates are at the geometric mean of time series
+  length across all the data (approximately 27 time steps). Colour shading
+  refers to taxonomic class (yellow: fishes, green: insects, purple: birds,
+  and red: mammals). Dotted vertical line in panel a indicates the
+  Pr($\nu < 10$) from the prior distribution. (b) Main effect
+  posterior densities for potential covariates of Pr($\nu < 10$). The beta
+  regression models were fit on a logit scale with hierarchical intercepts for
+  taxonomic class, taxonomic order, and species. All covariates were
+  standardized by subtracting their mean and dividing by twice their standard
+  deviation. In both panels, short vertical line segments within the density
+  polygons indicate median posterior estimates.}
+
+\label{fig:posteriors}
+\end{center}
+\end{figure}
+
 The majority of our heavy-tailed estimates were robust to alternative
@@
 in the residuals, modelled population growth rates without density dependence,
-or modelled the population dynamics as Ricker-logistic (Fig.~\ref{fig:alt}).
+or modelled the population dynamics as Ricker-logistic (Supporting materials Fig.~\ref{fig:alt}).
 However, setting observation error standard deviation to $0.2$ increased the
@@
 $\baseNuTen$ populations, although the majority of $\nu$ estimates remained
-qualitatively similar (Fig.~\ref{fig:alt}). The strength of the prior on $\nu$
+qualitatively similar (Supporting materials Fig.~\ref{fig:alt}). The strength of the prior on $\nu$
 had little influence on estimates of black-swan dynamics
-(Fig.~\ref{fig:alt-priors}). Our simulation testing shows that, if anything,
+(Supporting materials~Fig.~\ref{fig:alt-priors}). Our simulation testing shows that, if anything,
 our models underpredict the true magnitude and probability of heavy tailed
 events---especially given the length of the time series in the GPDD
-(Figs~\ref{fig:sim-nu}, \ref{fig:sim-prob}).
+(Supporting materials Figs~\ref{fig:sim-nu}, \ref{fig:sim-prob}).
 
@@
 
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=0.9\textwidth]{../analysis/correlates-p10.pdf}
+
+\caption[Potential covariates of heavy-tailed population dynamics.]{Potential covariates of heavy-tailed population dynamics (indicated
+  by a high probability that $\nu < 10$). Shown are (a--c) parameters from the
+  Gompertz heavy-tailed population model ($\sigma$, $\lambda$, $b$), (b)
+  number of time steps, (c) body length, and (d) lifespan. For the Gompertz
+  parameters, $\sigma$ refers to the scale parameter of the Student-t
+  process-noise distribution, $\lambda$ refers to the expected log abundance
+  at the next time step at an abundance of one, $b$ refers to the density
+  dependence parameter ($1$ is maximally density independent, $0$ is maximally
+  density dependent, and $<0$ is inversely density dependent). Circles
+  representing a few sharks, crustaceans, and gastropods are filled in white.
+  Median and 90\% credible interval posterior predictions of a beta regression
+  multilevel model are shown in panels a and d where there was a high
+  probability the slope coefficient was different from zero
+  (Fig.~\ref{fig:posteriors}b).}
+
+\label{fig:correlates}
+\end{center}
+\end{figure}
+
+\input{sparks.tex}
+
 We examined all time series with published explanations of why the black-swan
-events occurred (Tables~\ref{tab:sparks} and \ref{tab:causes-supp}). The
+events occurred (Table~\ref{tab:sparks} and Supporting materials~Table~\ref{tab:causes-supp}). The
 majority of documented events (\percBSDown \%) were downward black swans and
@@
 productivity rapidly increased, and the population experienced a rapid upswing
-in abundance. \citep{potts1980}.
+in abundance \citep{potts1980}.
 
@@
 heavy tailed because they turned out to be data-entry errors (Supporting
-Information). A final caveat is that the temporal scales of observation and
+materials). A final caveat is that the temporal scales of observation and
 population dynamics vary considerably across populations in the GPDD and this
@@
 images were obtained from \texttt{phylopic.org} under Creative Commons
-licenses; sources are listed in the Supporting Information. Funding was
-provided by a Simon Fraser University Graduate Fellowship (SCA), the Natural
-Sciences and Engineering Research Council of Canada (NKD, ABC), the Canada
-Research Chairs Program (NKD).
+licenses; sources are listed in the Supporting materials. Funding was provided
+by a Simon Fraser University Graduate Fellowship (SCA), the Natural Sciences
+and Engineering Research Council of Canada (NKD, ABC), the Canada Research
+Chairs Program (NKD).
 
-\section{Supporting Information}
+% \section{Supporting Information}
+%
+% The following supporting information is available online for this article:\\
+% Tables S1 and S2\\
+% Figures S1--S6\\
+% Example Stan code for a heavy-tailed Gompertz model and multilevel beta
+% regression\\
+% The GPDD IDs used in our analysis\\
+% All code and data to recreate our analysis are available at:\\
+% \url{https://github.com/seananderson/heavy-tails}
+%
+% \renewcommand{\baselinestretch}{\tighttextstretch}
+% \normalsize
+% \bibliographystyle{apalike}
+% \bibliography{/Users/seananderson/Dropbox/tex/jshort,/Users/seananderson/Dropbox/tex/ref3}
+%
+% \clearpage
+% \renewcommand{\baselinestretch}{\textstretch}
+% \normalsize
 
-The following supporting information is available online for this article:\\
-Tables S1 and S2\\
-Figures S1--S6\\
-Example Stan code for a heavy-tailed Gompertz model and multilevel beta
-regression\\
-The GPDD IDs used in our analysis\\
-All code and data to recreate our analysis are available at:\\
-\url{https://github.com/seananderson/heavy-tails}
+%\section{Tables}
 
-\renewcommand{\baselinestretch}{\tighttextstretch}
-\normalsize
-\bibliographystyle{apalike}
-\bibliography{/Users/seananderson/Dropbox/tex/jshort,/Users/seananderson/Dropbox/tex/ref3}
+%\clearpage
 
-\clearpage
-\renewcommand{\baselinestretch}{\textstretch}
-\normalsize
+%\clearpage
 
-\section{Tables}
+%\clearpage
 
-\input{sparks.tex}
-
-\clearpage
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=\textwidth]{../analysis/t-nu-eg2.pdf}
-\caption[An illustration of fitting population dynamic models that allow for heavy
-tails, represented by the Student-t degrees of freedom parameter $\nu$.]{
-An illustration of fitting population dynamic models that allow for heavy
-tails, represented by the Student-t degrees of freedom parameter $\nu$. (a, b)
-The probability density for t distributions with a scale parameter of 1 and
-different values of $\nu$. Small values of $\nu$ create heavy tails. As $\nu$
-approaches infinity the distribution approaches the normal distribution. For
-example, at $\nu = 2$, the probability of drawing a value more than five
-standard deviations below the mean is 1.8\%, whereas the probability of drawing
-such a value from a normal distribution is nearly zero ($2.9\cdot10^{-5}$\%).
-(c--e) Simulated population dynamics from a Gompertz model with process noise
-drawn from t distributions with three different values of $\nu$. Coloured dots
-in panels c and d represent jumps with less than a 1 in 1000 chance of
-occurring in a normal distribution. (f--h) Estimates of $\nu$ from models fit
-to the times series in panels c--e. Shown are the posterior samples
-(histograms), median and interquartile range of the posterior (IQR) (dots and
-line segments), and the exponential prior on $\nu$ (dashed lines). Colour
-shading behind panels f--h illustrates the region of heavy tails.
-}
-\label{fig:didactic}
-\end{center}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=0.44\textwidth]{../analysis/nu-coefs-2.pdf}
-
-\caption[Estimates of population dynamics heavy-tailedness for \nuCoefPopN\
-  populations of birds, mammals, insects, and fishes.]{Estimates of population dynamics heavy-tailedness for \nuCoefPopN\
-  populations of birds, mammals, insects, and fishes. Small values of $\nu$
-  approximately ($< 10$) suggest heavy-tailed black-swan dynamics; larger values of
-  $\nu$ suggest approximately normal-tailed dynamics. Vertical points and line
-  segments represent posterior medians and 50\% / 90\% credible intervals for
-  individual populations. Inset plots show probability that $\nu < 10$
-  (probability of heavy tails) for populations arranged by taxonomic order and
-  sorted by decreasing mean Pr($\nu < 10$). Taxonomic orders with three or
-  fewer populations in panel a are omitted for space. Red to yellow points
-  highlight populations with a high to moderately high probability of
-  heavy-tailed black-swan dynamics.}
-
-\label{fig:nu-coefs}
-\end{center}
-\end{figure}
-\clearpage
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=0.45\textwidth]{../analysis/order-posteriors-covariates.pdf}
-
-\caption[Posterior probability distributions from beta regression multilevel
-  models.]{Posterior probability distributions from beta regression multilevel
-  models. (a) Taxonomic-order-level posterior densities of Pr($\nu < 10$)
-  (approximately the probability of heavy tails) after accounting for
-  time-series length. Estimates are at the geometric mean of time series
-  length across all the data (approximately 27 time steps). Colour shading
-  refers to taxonomic class (yellow: fishes, green: insects, purple: birds,
-  and red: mammals). Dotted vertical line in panel a indicates the median
-  expected Pr($\nu < 10$) from the prior distribution. (b) Main effect
-  posterior densities for potential covariates of Pr($\nu < 10$). The beta
-  regression models were fit on a logit scale with hierarchical intercepts for
-  taxonomic class, taxonomic order, and species. All covariates were
-  standardized by subtracting their mean and dividing by twice their standard
-  deviation. In both panels, short vertical line segments within the density
-  polygons indicate median posterior estimates.}
-
-\label{fig:posteriors}
-\end{center}
-\end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=0.9\textwidth]{../analysis/correlates-p10.pdf}
-
-\caption[Potential covariates of heavy-tailed population dynamics (indicated
-  by a high probability that $\nu < 10$).]{Potential covariates of heavy-tailed population dynamics (indicated
-  by a high probability that $\nu < 10$). Shown are (a--c) parameters from the
-  Gompertz heavy-tailed population model ($\sigma$, $\lambda$, $b$), (b)
-  number of time steps, (c) body length, and (d) lifespan. For the Gompertz
-  parameters, $\sigma$ refers to the scale parameter of the Student-t
-  process-noise distribution, $\lambda$ refers to the expected log abundance
-  at the next time step at an abundance of one, $b$ refers to the density
-  dependence parameter ($1$ is maximally density independent, $0$ is maximally
-  density dependent, and $<0$ is inversely density dependent). Circles
-  representing a few sharks, crustaceans, and gastropods are filled in white.
-  Median and 90\% credible interval posterior predictions of a beta regression
-  multilevel model are shown in panels a and d where there was a high
-  probability the slope coefficient was different from zero
-  (Fig.~\ref{fig:posteriors}b).}
-
-\label{fig:correlates}
-\end{center}
-\end{figure}
+%\clearpage
+\input{blackswans/som}

### 2015-06-16 analysis/8-sim-extinction-risk.Rmd
Start adapting projected time series to use posterior samples

@@
-This file looks at quasi-extinction probability with heavy vs. normal tails
+This file looks at quasi-extinction probability with heavy vs. normal tails.
+
+**I'm currently adapting this script to bring in the parameter values from the model fitting.**
+
+Bring in the fitted data:
+
+```{r}
+source("5-shape-data.R")
+```
+
+Now write a function that crawls through `gomp_hat_base` row by row and repeatedly draws all needed parameter values stochastically and feeds these into a forward projection of the Gompertz model.
 
@@
 
-First, we'll figure out approximately what skewness values to use to get ballpark 85% downward deviations. This is to approximately match the skewness we see in the GPDD:
+Two ways to go about this: repeatedly bootstrap from the observed process deviations or draw the deviations from a t distribution with given sigma and nu.
+The problem with the later is that the deviations won't be appropriately downwardly skewed.
+But the problem with the former is that I'm not sure this gets what we want out of the exercise.
+The assumed normal distributions might not in fact be normal.
+The model might just have fit poorly.
+Therefore, I'm going to start with the first option.
+
+<!--
+I'll get those deviations the same way I do for the spark line plots.
+The idea is this: for a given population, draw lambda and draw b stochastically.
+Now, find what the residuals would have been for those
+
+```{r}
+get_gomp_res <- function(id) {
+  p <- subset(gomp_hat_base, main_id == id)
+  pop <- subset(gpdd, main_id == id_show)$population_untransformed
+  res <- rep(NA, length(pop))
+  for(i in 2:length(pop)) {
+    res[i] <- log(pop)[i] - (p$lambda_50 + p$b_50 * log(pop[i-1]))
+  }
+  return(list(res = res, l = l, u = u))
+}
+```
+
+First, we'll figure out approximately what skewness values to use to get ballpark 85% downward deviations.
+This is to approximately match the skewness we see in the GPDD: (TODO come back to this skewness issue)
 
@@
 So, 0.75 gets us in the ballpark of 85% percent downs.
+-->
 
-Now, we'll write a little C++ function to simulate from a Gompertz model. We'll do this in C++ just to make everything fast after this.
+We'll write a little C++ function to simulate from a Gompertz model. We'll do this in C++ just to make everything fast after this.
 
@@
 And we'll write a function to simulate a time series, check how many generations go below some threshold, and return the data in a form we can use.
+
+For each, we'll be drawing all parameter values from the posterior and starting at the state the time series ended. Then we'll continue on for some given number of generations.
+
+<!--
 I've picked parameter values to be approximately at the mean values in the GPDD (I'll check these more precisely).
 I'm running the simulation, by default, for 200 generations and discarding the first 50 as burn in.
+-->
 
 ```{r}
-sim_gomp_extinction <- function(lambda = 1, phi = 0.3, b = 0.5, N = 200,
-  gamma = 0.75, df = 3, thresh = 3, make_plot = FALSE, sigma = 0.2, burnin = 50,
+library("skewt")
+sim_gomp_extinction <- function(abundance_ts, lambda = 1, b = 0.5, N = 200,
+  gamma = 0.75, df = 3, thresh = 3, make_plot = FALSE, sigma = 0.2,
   return_dat = FALSE, id = 1) {
-  y1 <- lambda / b
-  proc_error_raw <- rskt(N, df = df, gamma = gamma) / (1 / sigma)
-  proc_error <- vector(mode = "numeric", length = N)
-  proc_error[1] <- proc_error_raw[1]
-  for (i in 2:N) {
-    proc_error[i] <- proc_error_raw[i] + phi * proc_error[i - 1]
-  }
+
+  y1 <- log(abundance_ts[length(abundance_ts)])
+
+  proc_error <- rskt(N, df = df, gamma = gamma) / (1 / sigma)
   y <- gompertz(N, lambda, b, proc_error, y1)
-  y <- y[(burnin + 1):N]
   if(make_plot) {
@@
   }
-  if(return_dat) {
-    y_return <- y
-    generation <- seq_along(y)
-  } else {
-    y_return <- NA
-    generation <- NA
-  }
-    data.frame(phi = phi, b = b, N = N, N_burned = N - burnin, gamma = gamma,
+    y_return <- c(log(abundance_ts), y)
+    generation <- c(seq(-length(abundance_ts) + 1, 0), seq_along(y))
+    data.frame(b = b, N = N, gamma = gamma,
       df = df, sigma = sigma, thresh = thresh, lambda = lambda,
       n_ext = sum(exp(y) < thresh), mean_log_abund = mean(y),
-      dat = y_return, generation = generation, id = id)
+      dat = y_return, generation = generation, main_id = id)
 }
 ```
+
+OK, let's try applying this to some sample populations:
+
+```{r}
+ids <- c(6528, 10127, 10007, 20579)
+out_ids <- lapply(ids, function(id_i) {
+samples <- 80L
+posterior_length <- 2000L
+posterior_ids <- round(seq(1, posterior_length, length.out = samples))
+out <- lapply(posterior_ids, function(i) {
+library("rstan")
+id <- id_i
+x <- readRDS(paste0("../scratch/base/sm-", id, ".rds"))
+x <- rstan::extract(x)
+gpdd <- readRDS("gpdd-clean.rds") # to get the last abundance
+this_dat <- filter(gpdd, main_id == id)
+
+out_gomp <- sim_gomp_extinction(abundance_ts = this_dat$population_untransformed, 
+  lambda = x$lambda[i], 
+  b = x$b[i], 
+  N = 100,
+  gamma = 0.75, 
+  df = x$nu[i], 
+  thresh = 3, 
+  make_plot = FALSE, 
+  sigma = x$sigma[i], 
+  return_dat = FALSE, 
+  id = id)
+out_gomp$posterior_sample <- i
+out_gomp
+ })
+out_df <- do.call("rbind", out)
+out_df
+})
+out_df2 <- do.call("rbind", out_ids)
+
+out_df2 <- mutate(out_df2, projection = ifelse(generation > 0, TRUE, FALSE))
+p <- ggplot(out_df2, 
+  aes(generation, exp(dat), colour = projection, group = posterior_sample)) + 
+  geom_line(alpha = 0.2) + 
+  geom_vline(xintercept = 0, lty = 2) + 
+  theme_bw() + ylab("Abundance") + xlab("Year") + 
+  scale_y_log10(breaks = scales::trans_breaks("log10", function(x) 10^x),
+                labels = scales::trans_format("log10", scales::math_format(10^.x))) + 
+  annotation_logticks(sides = "l") + 
+  guides(colour = guide_legend(override.aes= list(alpha = 1))) +
+  facet_wrap(~main_id, scales = "free_y") 
+print(p)
+
+```
+
+<!--#############################-->
+
+<!--
+```{r}
+gpdd <- readRDS("gpdd-clean.rds") # to get the last abundance
+dat <- readRDS("gomp-base-mean-sd.rds") # means and sds instead of medians
+x <- gomp_hat_base[1,]
+y1 <- filter(gpdd, main_id == x$main_id) %>% 
+  filter(sample_year == max(sample_year))
+y1 <- y1$population_untransformed
+sim_gomp_extinction(
+  y1 = y1,
+  lambda = 
+```
+-->
 

### 2015-06-19 analysis/8-sim-extinction-risk.Rmd
Continue developing projections code

@@
 ```{r}
-library("skewt")
-sim_gomp_extinction <- function(abundance_ts, lambda = 1, b = 0.5, N = 200,
-  gamma = 0.75, df = 3, thresh = 3, make_plot = FALSE, sigma = 0.2,
-  return_dat = FALSE, id = 1) {
+# library("skewt")
+sim_gomp_extinction <- function(abundance_ts, lambda = 1, b = 0.5, N = 100L,
+  df = 3, thresh = 3, sigma = 0.2, id = 1, bootstrap_residuals = FALSE, 
+  model = "heavy") {
 
-  y1 <- log(abundance_ts[length(abundance_ts)])
+  if (is.null(df)) df <- NA
 
-  proc_error <- rskt(N, df = df, gamma = gamma) / (1 / sigma)
+  log_abundance_ts <- log(abundance_ts)
+  y1 <- log_abundance_ts[length(abundance_ts)]
+
+  if (bootstrap_residuals) {
+    res <- log_abundance_ts[-1] - (lambda + b * log_abundance_ts[-length(abundance_ts)])
+    proc_error <- sample(res, size = N, replace = TRUE) # bootstrap the process residuals
+  } else {
+    if (model == "heavy") {
+      # proc_error <- rskt(N, df = df, gamma = gamma) / (1 / sigma)
+      proc_error <- rt(N, df = df) * sigma
+    } else {
+      proc_error <- rnorm(N, sd = sigma)
+    }
+  }
+
   y <- gompertz(N, lambda, b, proc_error, y1)
-  if(make_plot) {
-    par(mfrow = c(2, 1))
-    plot(y, type = "o")
-    abline(h = lambda / b, col = "red")
-    abline(h = mean(y), col = "blue")
-    plot(exp(y), type = "o")
-    abline(h = thresh, col = "green", lwd = 2)
-  }
-    y_return <- c(log(abundance_ts), y)
+
+    y_return <- c(log_abundance_ts, y)
     generation <- c(seq(-length(abundance_ts) + 1, 0), seq_along(y))
-    data.frame(b = b, N = N, gamma = gamma,
+    data.frame(b = b, N = N,
       df = df, sigma = sigma, thresh = thresh, lambda = lambda,
@@
 ```{r}
-ids <- c(6528, 10127, 10007, 20579)
-out_ids <- lapply(ids, function(id_i) {
-samples <- 80L
+ids <- c(6528, 10127, 20579)
+library("rstan")
+gpdd <- readRDS("gpdd-clean.rds") # to get the last abundance
+samples <- 2000L
 posterior_length <- 2000L
 posterior_ids <- round(seq(1, posterior_length, length.out = samples))
+
+out_ids <- lapply(ids, function(id_i) {
+  this_dat <- dplyr::filter(gpdd, main_id == id_i)
+  message(paste("main_id:", id_i))
+  # x <- readRDS(paste0("gomp-base-normal/sm-", id_i, ".rds"))
+  x <- readRDS(paste0("../scratch/gomp-base/sm-", id_i, ".rds"))
+  x <- rstan::extract(x)
+
   out <- lapply(posterior_ids, function(i) {
-library("rstan")
-id <- id_i
-x <- readRDS(paste0("../scratch/base/sm-", id, ".rds"))
-x <- rstan::extract(x)
-gpdd <- readRDS("gpdd-clean.rds") # to get the last abundance
-this_dat <- filter(gpdd, main_id == id)
-
     out_gomp <- sim_gomp_extinction(abundance_ts = this_dat$population_untransformed, 
-  lambda = x$lambda[i], 
-  b = x$b[i], 
-  N = 100,
-  gamma = 0.75, 
-  df = x$nu[i], 
-  thresh = 3, 
-  make_plot = FALSE, 
-  sigma = x$sigma[i], 
-  return_dat = FALSE, 
-  id = id)
+      lambda = x$lambda[i], b = x$b[i], N = 10L, df = x$nu[i], 
+      thresh = 3, sigma = x$sigma[i], id = id_i, bootstrap_residuals = FALSE, model = "heavy")
     out_gomp$posterior_sample <- i
@@
 out_df2 <- do.call("rbind", out_ids)
+out_df2 <- mutate(out_df2, projection = ifelse(generation > 0, TRUE, FALSE))
 
-out_df2 <- mutate(out_df2, projection = ifelse(generation > 0, TRUE, FALSE))
-p <- ggplot(out_df2, 
-  aes(generation, exp(dat), colour = projection, group = posterior_sample)) + 
-  geom_line(alpha = 0.2) + 
+ # out_df2 <- filter(out_df2, exp(dat) < 10e5)
+library(ggplot2)
+# p <- ggplot(out_df2, 
+#   aes(generation, exp(dat), colour = projection, group = posterior_sample)) + 
+#   geom_line(alpha = 0.2) + 
+#   xlim(-30, 10) +
+#   geom_vline(xintercept = 0, lty = 2) + 
+#   geom_hline(yintercept = 10, lty = 2) + 
+#   theme_bw() + ylab("Abundance") + xlab("Year") + 
+#   scale_y_log10(breaks = scales::trans_breaks("log10", function(x) 10^x),
+#     labels = scales::trans_format("log10", scales::math_format(10^.x)),
+#     limits = c(1, 1e5)) + 
+#   annotation_logticks(sides = "l") + 
+#   guides(colour = guide_legend(override.aes= list(alpha = 1))) +
+#   facet_wrap(~main_id) 
+# print(p)
+# ggsave("heavy-tailed-projections.png", width = 10, height = 6)
+cvar <- function(x) mean(x[x<=quantile(x, probs = 0.01)])
+group_by(out_df2, main_id) %>% summarise(cvar_ = cvar(exp(dat)))
+
+out_df3 <- out_df2 %>% group_by(main_id, generation, projection) %>%
+  summarise(
+    min_dat = min(dat, na.rm = TRUE),
+    q99 = quantile(dat, probs = 0.01, na.rm = TRUE)
+    )
+
+p <- ggplot(out_df3, 
+  aes(generation, exp(min_dat), colour = projection)) + 
+  geom_line() + 
+  geom_line(aes(y = exp(q99)), lty = 2) + 
   geom_vline(xintercept = 0, lty = 2) + 
+  geom_hline(yintercept = 10, lty = 2) + 
+  xlim(-30, 10) +
   theme_bw() + ylab("Abundance") + xlab("Year") + 
   scale_y_log10(breaks = scales::trans_breaks("log10", function(x) 10^x),
-                labels = scales::trans_format("log10", scales::math_format(10^.x))) + 
+    labels = scales::trans_format("log10", scales::math_format(10^.x)),
+    limits = c(1, 1e4)) + 
   annotation_logticks(sides = "l") + 
   guides(colour = guide_legend(override.aes= list(alpha = 1))) +
-  facet_wrap(~main_id, scales = "free_y") 
-print(p)
+  facet_wrap(~main_id) 
+p
 

### 2015-06-23 text/supp.tex
Revert some thesis-specific modifications
Switch to local .bib files
Add draft section to SOM on skew t

@@
 sizes \citep{stan-manual2014}.
+
+\subsection{The skewed Student's t distribution}
+
+We fit Gompertz models with skewed Student's t distributed process noise based
+on the probability density for the skewed t distribution described in
+\citet{fernandez1998}. Their definition equates to the following probability density as implemented in \citet{king2012}:
+
+\[
+  \mathrm{Skewed\,Student\mhyphen t}(x) =
+\begin{dcases}
+\frac{2}{(\gamma + 1/\gamma)}
+         \mathrm{Student\mhyphen t}(x \cdot \gamma, \mu \cdot \gamma, \nu, \sigma),& \text{for } x < 0\\
+       \frac{2}{(\gamma + 1/\gamma)}
+         \mathrm{Student\mhyphen t}(x / \gamma, \mu / \gamma, \nu, \sigma),& \text{for } x \ge 0
+\end{dcases}
+\]
+
+\smallskip
+\noindent
+Where $\gamma$ represents the skewness parameter. The distribution is
+symmetrical if $\gamma = 1$, left skewed (skewed downwards in abundance) if
+$0 < \gamma < 1$, and right skewed (skewed upwards in abundance) if $1 <
+\gamma < \infty$. This translates to the log probability:
+
+\[
+\mathrm{Skewed\,Student\mhyphen t}(x) =
+\begin{dcases}
+  \log (2 \gamma) - \log(\gamma^2 + 1) +
+  \log \mathrm{Student\mhyphen t}(x \cdot \gamma, \mu \cdot \gamma, \nu, \sigma),& \text{for } x < 0\\
+  \log (2 \gamma) - \log(\gamma^2 + 1) +
+  \log \mathrm{Student\mhyphen t}(x / \gamma, \mu / \gamma, \nu, \sigma),& \text{for } x \ge 0
+\end{dcases}
+\]
+
+\smallskip
+\noindent
+Since Stan only needs the probability density up to an additive constant, we
+can replace $\log(2 \gamma)$ in the above for $\log(\gamma)$. The final log
+probability density for the skewed Student's t distribution implemented in
+Stan\footnote{See the Stan users mailing list: \url{https://groups.google.com/d/msg/stan-users/jeZQnQRUsDs/BeqhicWm1GEJ}} looks like the following:
+
+\begin{verbatim}
+real skew_student_t_log(real y, real nu, real mu, real sigma, real skew) {
+  real lp;
+  lp <- log(skew) - log1p(square(skew));
+  if (y < mu)
+    return lp + student_t_log(y * skew, nu, mu * skew, sigma);
+  else
+    return lp + student_t_log(y / skew, nu, mu / skew, sigma);
+}
+\end{verbatim}
+
+% \[
+%   \log p(x) =
+% \begin{dcases}
+%   \log(2 \gamma) - log(\gamma^2 + 1) +
+%     \log \mathrm{dt}(x \gamma, \nu) ,& \text{for } x < 0\\
+%   \log(2 \gamma) - log(\gamma^2 + 1) +
+%     \log \mathrm{dt}(x / \gamma, \nu) ,& \text{for } x \ge 0
+% \end{dcases}
+% \]
 
@@
 
-% \renewcommand{\baselinestretch}{\tighttextstretch}
-% \normalsize
-% \bibliographystyle{apalike}
-% \bibliography{/Users/seananderson/Dropbox/tex/jshort,/Users/seananderson/Dropbox/tex/ref3}
-% \renewcommand{\baselinestretch}{\textstretch}
-% \normalsize
+\bibliographystyle{apalike}
+\bibliography{/Users/seananderson/Dropbox/tex/jshort,supp}
 %
@@
 
-\begin{scriptsize}
+\begin{footnotesize}
 %\renewcommand{\baselinestretch}{1.11}
@@
 %\renewcommand{\baselinestretch}{\textstretch}
-\end{scriptsize}
+\end{footnotesize}
 \normalsize

### 2015-06-27 analysis/8-sim-extinction-risk.Rmd
Remove testing subset on filenames

@@
 
-<!--
-I've picked parameter values to be approximately at the mean values in the GPDD (I'll check these more precisely).
-I'm running the simulation, by default, for 200 generations and discarding the first 50 as burn in.
--->
-
 ```{r}
@@
 sim_gomp_extinction <- function(abundance_ts, lambda = 1, b = 0.5, N = 100L,
-  df = 3, thresh = 3, sigma = 0.2, id = 1, bootstrap_residuals = FALSE, 
-  model = "heavy") {
+  nu = 3, log_skew = 1, sigma_proc = 0.2, id = 1, 
+  bootstrap_residuals = FALSE, model = "heavy") {
 
@@
     if (model == "heavy") {
-      # proc_error <- rskt(N, df = df, gamma = gamma) / (1 / sigma)
-      proc_error <- rt(N, df = df) * sigma
+      proc_error <- rskt(N, df = nu, gamma = exp(log_skew)) * sigma_proc
     } else {
-      proc_error <- rnorm(N, sd = sigma)
+      proc_error <- rnorm(N, sd = sigma_proc)
     }
@@
     data.frame(b = b, N = N,
-      df = df, sigma = sigma, thresh = thresh, lambda = lambda,
-      n_ext = sum(exp(y) < thresh), mean_log_abund = mean(y),
-      dat = y_return, generation = generation, main_id = id)
+      nu = nu, sigma = sigma_proc, thresh = thresh, lambda = lambda,
+      skew = exp(log_skew), dat = y_return, generation = generation, 
+      main_id = id)
 }
@@
 gpdd <- readRDS("gpdd-clean.rds") # to get the last abundance
-samples <- 2000L
-posterior_length <- 2000L
-posterior_ids <- round(seq(1, posterior_length, length.out = samples))
 

### 2015-06-27 analysis/5.10-extract-skew-samples.R

@@ -1,6 +1,6 @@
 sample_posterior <- function(path = ".", N = 1000L) {
   f <- list.files(path, pattern = "*.rds")
-  samples <- plyr::ldply(f[1:3], function(x) {
+  samples <- plyr::ldply(f, function(x) {
 
     message(x)
     e <- rstan::extract(readRDS(paste0(path, x)))

### 2015-07-21 analysis/cause-table.tex
Bankruptcy commit
Major restructuring of text

@@
   \midrule
-\includegraphics[width=1.7cm]{../analysis/sparks/6528.pdf} & Shag, \textit{Phalacrocorax aristotelis}, Farne Islands, Northumberland & 6528 & \citep{potts1980} & Red tide event combined with low productivity due to overcrowding & 1.00 & 2 (2--4) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/7115.pdf} & South African fur seal, \textit{Arctocephalus pusillus}, South Africa & 7115 & \citep{shaughnessy1982} & Harvesting and predation changes & 1.00 & 2 (2--4) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/10128.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Scotland - un-named area & 10128 & \citep{potts1984} & Environment- and parisite-caused cycles & 1.00 & 3 (2--4) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/9382.pdf} & Pine looper or Bordered white, \textit{Bupalus piniaria}, Kessock & 9382 & \citep{broekhuizen1993} & Unknown, but sampling intensity was decreasing & 1.00 & 3 (2--5) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/10127.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Scotland - un-named area & 10127 & \citep{potts1984} & Environment- and parisite-caused cycles & 0.99 & 3 (2--5) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/10007.pdf} & Water vole, \textit{Arvicola terrestris}, Le Pont & 10007 & \citep{saucy1994} & Predator-environment cycle interactions & 1.00 & 3 (2--5) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/20579.pdf} & Grey heron, \textit{Ardea cinerea}, Southern Britain & 20579 & \citep{stafford1971} & Severe winter & 0.98 & 3 (2--7) \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/9655.pdf} & Flea beetle, \textit{Chaetocnoma concinna}, Finland & 9655 & \citep{markkula1965} & Cannot locate original source & 0.99 & 3 (2--6) \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/1235.pdf} & Wren, \textit{Troglodytes troglodytes}, Eastern Wood, Bookham Common & 1235 & \citep{newton1998} & Severe winter & 0.98 & 3 (2--8) \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/10113.pdf} & Willow grouse, \textit{Lagopus lagopus}, Northern England & 10113 & \citep{dobson1995} & Parasites and predators & 0.99 & 3 (2--6) \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/9667.pdf} & Gooseberry sawfly, \textit{Nemastus ribesii}, Finland & 9667 & \citep{markkula1965} & Cannot locate original source & 0.97 & 3 (2--8) \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/9679.pdf} & Unknown, \textit{Trioza apicalis}, Finland & 9679 & \citep{markkula1965} & Cannot locate original source & 0.83 & 3 (2--88) \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/20527.pdf} & Wandering albatross, \textit{Diomedea exulans}, Taiaroa & 20527 & \citep{robertson1998} & Unknown & 0.99 & 4 (2--7) \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/10039.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Northern Scotland & 10039 & \citep{dobson1995} & Parasites and predators & 0.92 & 4 (2--12) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/10162.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Atholl Estate & 10162 & \citet{mackenzie1952} & Bad environmental conditions and overcrowding combined to create crashes & 0.91 & 5 (2--12) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/9503.pdf} & Fisher or  Pekan, \textit{Martes pennanti}, Manitoba & 9503 & \citep{keith1963} & Unknown & 0.76 & 5 (2--87) \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/7099.pdf} & European rabbit, \textit{Oryctolagus cuniculus}, Estate 2, East Anglia & 7099 & \citep{barnes1986} & Disease outbreak followed by years of good weather & 0.78 & 5 (2--67) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/2778.pdf} & Wheatear, \textit{Oenanthe oenanthe}, Skokholm Island & 2778 & \citep{lack1969} & Unknown, but decline noted specifically, cold winters caused some crashes & 0.67 & 5 (2--141) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/9659.pdf} & Cabbage root fly or maggot, \textit{Delia radicum}, Finland & 9659 & \citep{markkula1965} & Cannot locate original source & 0.59 & 7 (2--161) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/1195.pdf} & Blue jay, \textit{Cyanocitta cristata}, Robert Allerton Park & 1195 & \citep{kendeigh1982} & Unknown & 0.60 & 7 (2--149) \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/5019.pdf} & Barbary macaque, \textit{Macaca sylvanus}, Queens Gate & 5019 & \citep{fa1984} & Cannot locate original source & 0.61 & 7 (2--132) \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/9953.pdf} & Rock ptarmigan, \textit{Lagopus mutus}, Iceland & 9953 & \citep{clarke1885,williams1954} & Severe winters & 0.58 & 7 (2--172) \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/6548.pdf} & Lesser-spotted dogfish, \textit{Scyliorhinus caniculus}, North Sea & 6548 & \citep{heessen1996} & Unknown, not specifically mentioned & 0.53 & 8 (2--190) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/20546.pdf} & American red fox, \textit{Vulpes fulva}, Labrador & 20546 & \citep{dancona1954,lindstrom1994} & Predator-prey cycles & 0.57 & 9 (3--65) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/9470.pdf} & Ruffed grouse, \textit{Bonasa umbellus}, Connecticut & 9470 & \citep{keith1963} & Unknown & 0.53 & 9 (3--158) \\ 
+\includegraphics[width=1.7cm]{../analysis/sparks/6528.pdf} & Shag, \textit{Phalacrocorax aristotelis}, Farne Islands, Northumberland & 6528 & \cite{potts1980} & Red tide event combined with low productivity due to overcrowding & 1.00 & 2 (2--4) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/7115.pdf} & South African fur seal, \textit{Arctocephalus pusillus}, South Africa & 7115 & \cite{shaughnessy1982} & Harvesting and predation changes & 1.00 & 2 (2--4) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/10128.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Scotland - un-named area & 10128 & \cite{potts1984} & Environment- and parisite-caused cycles & 1.00 & 3 (2--4) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/9382.pdf} & Pine looper or Bordered white, \textit{Bupalus piniaria}, Kessock & 9382 & \cite{broekhuizen1993} & Unknown, but sampling intensity was decreasing & 1.00 & 3 (2--5) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/10127.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Scotland - un-named area & 10127 & \cite{potts1984} & Environment- and parisite-caused cycles & 0.99 & 3 (2--5) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/10007.pdf} & Water vole, \textit{Arvicola terrestris}, Le Pont & 10007 & \cite{saucy1994} & Predator-environment cycle interactions & 1.00 & 3 (2--5) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/20579.pdf} & Grey heron, \textit{Ardea cinerea}, Southern Britain & 20579 & \cite{stafford1971} & Severe winter & 0.98 & 3 (2--7) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/9655.pdf} & Flea beetle, \textit{Chaetocnoma concinna}, Finland & 9655 & \cite{markkula1965} & Cannot locate original source & 0.99 & 3 (2--6) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/1235.pdf} & Wren, \textit{Troglodytes troglodytes}, Eastern Wood, Bookham Common & 1235 & \cite{newton1998} & Severe winter & 0.98 & 3 (2--8) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/10113.pdf} & Willow grouse, \textit{Lagopus lagopus}, Northern England & 10113 & \cite{dobson1995} & Parasites and predators & 0.99 & 3 (2--6) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/9667.pdf} & Gooseberry sawfly, \textit{Nemastus ribesii}, Finland & 9667 & \cite{markkula1965} & Cannot locate original source & 0.97 & 3 (2--8) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/9679.pdf} & Unknown, \textit{Trioza apicalis}, Finland & 9679 & \cite{markkula1965} & Cannot locate original source & 0.83 & 3 (2--88) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/20527.pdf} & Wandering albatross, \textit{Diomedea exulans}, Taiaroa & 20527 & \cite{robertson1998} & Unknown & 0.99 & 4 (2--7) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/10039.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Northern Scotland & 10039 & \cite{dobson1995} & Parasites and predators & 0.92 & 4 (2--12) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/10162.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Atholl Estate & 10162 & \cite{mackenzie1952} & Bad environmental conditions and overcrowding combined to create crashes & 0.91 & 5 (2--12) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/9503.pdf} & Fisher or  Pekan, \textit{Martes pennanti}, Manitoba & 9503 & \cite{keith1963} & Unknown & 0.76 & 5 (2--87) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/7099.pdf} & European rabbit, \textit{Oryctolagus cuniculus}, Estate 2, East Anglia & 7099 & \cite{barnes1986} & Disease outbreak followed by years of good weather & 0.78 & 5 (2--67) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/2778.pdf} & Wheatear, \textit{Oenanthe oenanthe}, Skokholm Island & 2778 & \cite{lack1969} & Unknown, but decline noted specifically, cold winters caused some crashes & 0.67 & 5 (2--141) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/9659.pdf} & Cabbage root fly or maggot, \textit{Delia radicum}, Finland & 9659 & \cite{markkula1965} & Cannot locate original source & 0.59 & 7 (2--161) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/1195.pdf} & Blue jay, \textit{Cyanocitta cristata}, Robert Allerton Park & 1195 & \cite{kendeigh1982} & Unknown & 0.60 & 7 (2--149) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/5019.pdf} & Barbary macaque, \textit{Macaca sylvanus}, Queens Gate & 5019 & \cite{fa1984} & Cannot locate original source & 0.61 & 7 (2--132) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/9953.pdf} & Rock ptarmigan, \textit{Lagopus mutus}, Iceland & 9953 & \cite{clarke1885,williams1954} & Severe winters & 0.58 & 7 (2--172) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/6548.pdf} & Lesser-spotted dogfish, \textit{Scyliorhinus caniculus}, North Sea & 6548 & \cite{heessen1996} & Unknown, not specifically mentioned & 0.53 & 8 (2--190) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/20546.pdf} & American red fox, \textit{Vulpes fulva}, Labrador & 20546 & \cite{dancona1954,lindstrom1994} & Predator-prey cycles & 0.57 & 9 (3--65) \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/9470.pdf} & Ruffed grouse, \textit{Bonasa umbellus}, Connecticut & 9470 & \cite{keith1963} & Unknown & 0.53 & 9 (3--158) \\
    \bottomrule

### 2015-07-21 text/ms.tex
Switch to local .bib files
Bankruptcy commit
Major restructuring of text

--- 2014-12-04 text/ms.tex
@@
-%!TEX root = anderson-etal-blackswan-timeseries.tex
+\textbf{Title: Evidence for black-swan events in animal populations}
 
-\section{Abstract}
+\textbf{Authors}: Sean C. Anderson\textsuperscript{1,2*}, Trevor A. Branch\textsuperscript{3}, Andrew B. Cooper\textsuperscript{2}, Nicholas K. Dulvy\textsuperscript{1}
 
-Black swans are statistically improbable events that nonetheless occur---often
-with profound consequences. While extremes in the physical environment, such
-as monsoons and heat waves, are widely studied and increasing in magnitude and
-frequency, it remains unclear the extent to which ecological populations
-buffer or suffer from such extremes. Here, estimate the degree of
-heavy-tailedness (presence of black swans) in ecological process noise by
-applying a probability model to \NPops~time series from around the world
-across \NOrders~taxonomic orders and seven classes. We find strong evidence of
-black swans, but they are rare, occurring in
-\overallMinPerc--\overallMaxPerc\% of populations: most frequently for birds
-(\AvesRangePerc\%) followed by mammals (\MammaliaRangePerc\%), and insects
-(\InsectaRangePerc\%). Black swans were predominantly (\percBSDown \%)
-downward events and were not explained by any life-history covariates, but
-tended to be driven by external perturbations such as climate, severe winters,
-predator or parasite cycles, and the combined effects of multiple factors.
-Extreme events were more frequently detected for populations with longer time
-series and lower levels of process noise; for shorter and noisier time series,
-our simulations suggested black-swan dynamics are often not identifiable as
-such. The presence of black swans in population dynamics highlights the
-importance of developing robust conservation and management
-strategies---particularly as the frequency and magnitude of climate extremes
-increase over the next century.
+\textsuperscript{1}Earth to Ocean Research Group, Department of Biological Sciences, Simon Fraser University, Burnaby BC, V5A 1S6, Canada
 
-\section{Introduction}
+\textsuperscript{2}School of Resource and Environmental Management, Simon Fraser University, Burnaby, BC, V5A 1S6, Canada
 
-Black swans are unexpected extreme events with potentially dramatic
-consequences \citep{taleb2007,sornette2009}. One of the most striking black
-swans in ecology is the asteroid collision that may have marked the
-end-Cretaceous mass extinction 65 million years ago
-\citep{alvarez1980,harnik2012}. Today, climate extremes, in concert with
-shifts in mean temperature, are expected to cause the greatest ecological and
-societal damage \citep{ipcc2012}. But, while extremes in the physical
-environment such as wave height, storm severity, and temperature are frequent
-events \citep{gaines1993,katz2005}, it remains unclear the extent to which
-ecological systems buffer or suffer from black swans \citep{nunez2012}.
+\textsuperscript{3}School of Aquatic and Fishery Sciences, University of Washington, Box 355020, Seattle, WA 98195, USA
 
-There is compelling anecdotal evidence for ecological black swans, but
-systematic evidence across taxa has been elusive. A survey of ecologists
-indicated that surprising outcomes of field experiments are far more common
-than we assume \citep{doak2008}, and events such as the global invasion of
-Argentine ants and the mutation of viruses to infect new hosts could be
-considered black swans \citep{nunez2012}. In fact, anecdotes of population
-catastrophes are numerous and catastrophes may be the most important element
-affecting population persistence \citep{mangel1994}. For marine mammal
-populations, we have compelling evidence of catastrophes \citep{gerber2001,
-  ward2007}, and recently, time-series of marine microbe abundance
-\citep{segura2013} and time to extinction for experimental waterflea
-populations \citep{drake2014} have been found to follow heavier tailed
-distributions than the normal. Despite these examples, as far as we can tell
-there have been two systematic surveys for ecological black swans: one with
-North American breeding birds and the other with the Global Population
-Dynamics Database, which uncovered little clear evidence for them
-\citep{keitt1998,allen2001,halley2002}. Nevertheless there are methodological
-challenges to their detection \citep{allen2001,ward2007}.
+\textsuperscript{*}Corresponding author: Sean C. Anderson; E-mail: sean\_anderson@sfu.ca
 
-There are two key reasons why we may find little evidence of ecological black
-swans. First, they might not exist in higher taxa. Indeed, the majority of
-model fitting and risk forecasting assumes that population dynamics are normal
-tailed on a log scale \citep[e.g.][]{brook2006a,dennis2006,knape2012}.
-Alternatively, black-swan dynamics might exist, but our ability to detect
-them requires further development of statistical tools. One such tool is the
-generalized extreme value distribution, which has been applied to
-environmental data \citep[e.g.][]{katz2005}. This distribution describes the
-most extreme event per time interval (e.g.~heaviest rainfall per year), but
-requires time series that are sufficiently long to be condensed into time
-intervals. Another statistical tool involves fitting a catastrophic mixture
-distribution in a state space model to quantify the probability that
-population events are extreme, although this is also data intensive
-\citep{ward2007}. A third tool is to compare the support for fits of thin- and
-heavy-tailed distributions \citep{halley2002}, but this analysis did not
-quantify the probability of black swans or allow for population dynamics.
+\textbf{Abstract}:
 
-Here, we assess the frequency and magnitude of black-swan dynamics across
-\NPops\ populations from a wide array of taxonomic groups---mostly birds,
-mammals, insects, and fishes. We then identify characteristics of time series
-or intrinsic life-history characteristics that are associated with the
-detection of black-swan events and attempt to verify known causes. To
-accomplish this, we develop a framework for identifying heavy-tailed process
-noise in population dynamics, i.e.\ whether the largest stochastic jumps in
-abundance from one time step to the next are more extreme than typically seen
-with a log-normal distribution. Our framework allows for a range of population
-dynamic models, can incorporate observation uncertainty, and can be easily
-applied to abundance time series.
-
-\section{Methods}
-
-To obtain estimates of the probability and magnitude of black-swan events, we
-fit population dynamic models to abundance time series from around the world.
-For each population, we estimated the shape of the process noise tails by
-measuring the degrees of freedom ($\nu$) of a Student t-distribution.
-
-\subsection{Time-series data}
-
-We selected abundance time series from the Global Population Dynamics Database
-(GPDD; \citeauthor{gpdd2010} \citeyear{gpdd2010}), which contains nearly 5000
-time series of abundance from $\sim$1000 species and $\sim$100 taxonomic
-orders. We filtered the data (Supporting Information) to remove populations
-from less reliable data sources, and those without sufficient data for our
-models, and then interpolated some missing values
-\citep[\textit{sensu}][]{brook2006a}. Our interpolation affected only
-$\sim$\interpPointsPerc \% of the final data points (Table~\ref{tab:stats})
-and none of the data points that were later considered black-swan events. Our
-final dataset contained \NPops~populations across \NOrders~taxonomic orders
-and seven taxonomic classes, with a median of \medianTimeSteps~time steps
-(range of \minTimeSteps--\maxTimeSteps) (Table~\ref{tab:stats},
-Fig.~\ref{fig:all-ts}).
-
-\subsection{Population models}
-
-Our main analysis focuses on the commonly applied Gompertz population dynamics
-model \citep[e.g.][]{knape2012,dennis2014,connors2014}. The Gompertz model
-represents population growth as a linear function in $\log$ space. If we let
-$x_t$ represent the $\log$ abundance ($N$) at time $t$, we can represent the
-Gompertz model as:
-\begin{align*}
-x_t &= \lambda + b x_{t-1} + \epsilon_t\\
-\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma).
-\end{align*}
-The growth parameter $\lambda$ represents the expected growth rate if $N_t =
-1$. The model is density independent if $b = 1$, maximally density dependent
-if $b = 0$, and inversely density dependent if $b < 0$. Usually, the process
-noise $\epsilon_t$ is modelled as normally distributed, but in our paper we
-assume it is drawn from a t distribution with scale parameter $\sigma$ and
-degrees of freedom $\nu$. In previous analyses of the GPDD, the Gompertz was
-most often identified as the most parsimonious population model fit to these
-data \citep{brook2006}.
-
-By allowing the process noise to be drawn from a Student-t distribution we can
-estimate the degree to which the process deviations have heavy tails and are
-thereof evidence of black-swan events (Fig.~\ref{fig:didactic}a, b). For
-example, at $\nu = 3$, the probability of drawing a value more than five
-standard deviations below the mean is $0.02$, whereas the probability of
-drawing such a value from a normal distribution is tiny ($2.9\cdot10^{-7}$).
-As the value of $\nu$ approaches infinity, the distribution approaches the
-normal distribution (Fig.~\ref{fig:didactic}a, b). While, across populations,
-the extreme process deviations might tend to be more frequently upwards or
-downwards events, a small random sample from a heavy-tailed distribution can
-have many outliers on one side and appear asymmetric \citep{gelman2013a}.
-Therefore, fitting a symmetric t distribution is appropriate for the
-relatively short population time series in our dataset and is agnostic towards
-the detection of upwards or downwards black swans.
-
-We fit all models in a Bayesian framework using Stan \citep{stan-manual2014}
-via the R computing environment \citep{r2014}. Stan samples from the posterior
-distribution with an adaptive version of Hamiltonian Markov chain Monte Carlo
-called the No-U-Turn Sampler and generally obtains less correlated samples
-than algorithms such as the Gibbs sampler \citep{hoffman2014}. We tested to
-ensure that the chains had sufficiently converged and that the sampler had
-obtained sufficient independent samples from the posterior ($\widehat{R} <
-1.05$, $n_\mathrm{eff} > 200$; Supporting Information).
-
-We chose weakly informative priors to incorporate our understanding of
-plausible population dynamics (\citeauthor{gelman2014} \citeyear{gelman2014};
-Figs~\ref{fig:didactic}f--h, \ref{fig:priors}). For $\nu$, we chose an
-exponential prior with rate parameter of $0.01$ truncated at values above
-two---a slightly less informative prior than suggested by
-\citet{fernandez1998}. This prior gives only a \basePriorProbHeavy \%
-probability that $\nu < 10$ but constrains the sampling sufficiently to avoid
-wandering off towards infinity. In any case, for $\nu > 20$ the t distribution
-is almost indistinguishable from the normal distribution
-(Fig.~\ref{fig:didactic}). Based on the shape of the t distribution, we chose
-the probability that $\nu < 10$, Pr($\nu < 10$), to define the probability of
-heavy-tailed dynamics. When categorizing a population as heavy or normal tailed,
-we used a threshold of 0.5 probability.
-
-We fit alternative population models to test if four key phenomena
-systematically changed our conclusions. Autocorrelation has been suggested as
-a reason for increased observed variability of abundance time series through
-time, which could create apparent heavy tails \citep{inchausti2002};
-therefore, we fit a model that included serial correlation in the residuals.
-Additionally, previous work has modelled abundance or growth rates without
-accounting for density dependence \citep{halley2002,segura2013}; therefore, we
-fit a simpler model in which we assumed density independence. Third,
-observation error could bias parameter estimates \citep{knape2012} or mask our
-ability to detect heavy tails \citep{ward2007}; therefore, we fit a model
-where we allowed for a fixed quantity of observation error ($0.2$ standard
-deviations on a log scale). Finally, the Gompertz model assumes that
-population growth rate declines linearly with log abundance. Therefore, we
-also fit an alternative model, the Ricker-logistic model, which assumes that
-population growth rate declines linearly with abundance itself (Supporting
-Information).
-
-In addition to alternative population models, we investigated the sensitivity
-of our results to weaker and stronger priors (exponential rate parameter $=
-0.005, 0.02$; Fig.~\ref{fig:priors}, Supporting Information). We used
-simulated data to test how easily we could detect $\nu$ given different sample
-sizes and to ensure we could recover unbiased parameter estimates from the
-Gompertz model (Supporting Information).
-
-\subsection{Covariates of population dynamic black swans}
-
-We investigated possible covariates of heavy-tailed population dynamics
-visually and through multilevel modelling. We plotted characteristics of the
-time series ($\sigma$, $\lambda$, $b$, and time-series length) along with two
-life-history characteristics (body length and maximum lifespan obtained from
-\citet{brook2006a}) against our estimated probability of heavy tails, Pr$(\nu <
-10)$. We formally investigated these relationships by fitting beta
-regression multilevel models \citep{ferrari2004}. The beta probability
-distribution can represent continuous response data that range between zero
-and one; we fit our models with a logit link as is common for beta and logistic
-regression \citep{ferrari2004}. We incorporated standard deviations around the
-means for covariates that were derived from Gompertz model parameter
-estimates. To account for broad patterns of phylogenetic relatedness, we
-allowed for hierarchical intercepts at the taxonomic class, order, and species
-level (Supporting Information). We fit our model in Stan with weakly
-informative priors on the coefficients \citep{gelman2008d} and variance
-parameters (\citeauthor{gelman2006c} \citeyear{gelman2006c}
-\citeauthor{gelman2014} \citeyear{gelman2014}; Supporting Information).
-
-Finally, we investigated a sample of populations that our method categorized as
-having a high probability of heavy tails (Pr$(\nu < 10) > 0.5$). Where
-possible, we found the documented causes of ecological black swans in the
-primary data source cited in the GPDD or in other literature describing the
-population.
-
-\section{Results}
-
-We found strong, but rare, evidence for black-swan population dynamics. By
-defining black-swan dynamics as a greater than $0.5$ probability that $\nu <
-10$, our main Gompertz model found evidence for heavy tails most frequently
-for birds (\birdPH\%) followed by mammals (\mammalsPH\%), and insects
-(\insectsPH\%) (Fig.~\ref{fig:nu-coefs}, Table~\ref{tab:causes-supp}). Black
-swans were taxonomically widespread, occurring in \POrdersHeavy\% of taxonomic
-orders. Accounting for time series length and partially pooling inference
-across taxonomic class and order with a multilevel model (Supporting
-Information), there was stronger evidence for black swans in insect
-populations than is visually apparent in Fig.~\ref{fig:nu-coefs}---four of 10
-orders with the highest median probability of heavy tails were insect
-orders---however, there was considerable uncertainty in these estimates
-(Fig.~\ref{fig:posteriors}a).
-
-The majority of our heavy-tailed estimates were robust to alternative
-population models, observation error, and choice of priors. Our conclusions
-were not systematically altered when we included an autocorrelation structure
-in the residuals, modelled population growth rates without density dependence,
-or modelled the population dynamics as Ricker-logistic (Fig.~\ref{fig:alt}).
-However, setting observation error standard deviation to $0.2$ increased the
-median estimate of $\nu$ from $<10$ to $\ge 10$ in $\baseNuTenObsTenSwitch$ of
-$\baseNuTen$ populations, although the majority of $\nu$ estimates remained
-qualitatively similar (Fig.~\ref{fig:alt}). The strength of the prior on $\nu$
-had little influence on estimates of black-swan dynamics
-(Fig.~\ref{fig:alt-priors}). Our simulation testing shows that, if anything,
-our models underpredict the true magnitude and probability of heavy tailed
-events---especially given the length of the time series in the GPDD
-(Figs~\ref{fig:sim-nu}, \ref{fig:sim-prob}).
-
-Across populations, the probability of observing black-swan dynamics was
-positively related to time-series length and negatively related to magnitude of
-process noise ($\sigma$) but not clearly related to population growth rate
-($\lambda$), density dependence ($b$), or maximum lifespan
-(Figs~\ref{fig:correlates}, \ref{fig:posteriors}b). Longer time-series length
-was the strongest covariate of observing black-swan dynamics. For instance, we
-were about \pIncHeavyNThirtyNSixty~times as likely to observe heavy tails in a
-population with 60 time steps compared to one with 30 time steps. However, the
-absolute change in probability with increased time series length was small
-($\pHeavyNSixty$ vs.\ $\pHeavyNThirty$ in the previous example,
-Fig.~\ref{fig:correlates}).
-
-We examined all time series with published explanations of why the black-swan
-events occurred (Tables~\ref{tab:sparks} and \ref{tab:causes-supp}). The
-majority of documented events (\percBSDown \%) were downward black swans and
-involved a combination of multiple factors. For example, a synchronization of
-environmental- and predation-mediated population cycles are thought to have
-caused a downward black-swan event for a water vole (\textit{Arvicola
-  terrestris}) population \citep{saucy1994}. Other black swans were the result
-of a sequence of extreme climate events on their own. For instance, severe
-winters in 1929, 1940--1942, and 1962--1963 were associated with black-swan
-downswings in grey heron (\textit{Ardea cinerea}) abundance in the United
-Kingdom \citep{stafford1971}. Our analysis finds that the last event was a
-combination of two black-swan events in a row and it took the population three
-times longer to recover than predicted \citep{stafford1971}. Downwards black
-swans were sometimes followed by upwards black swans. For example, during a
-period of population crowding and nest shortages, a population of European
-shag cormorants (\textit{Phalacrocorax aristotelis}) on the Farne Islands,
-United Kingdom, declined suddenly following a red tide event in 1968
-\citep{potts1980}. This freed quality nest sites for first-time breeders,
-productivity rapidly increased, and the population experienced a rapid upswing
-in abundance. \citep{potts1980}.
-
-\section{Discussion}
-
-We found strong evidence for black swans (heavy-tailed process noise) in
-\overallMinPerc--\overallMaxPerc\% of ecological time series. Black swans were
-usually (\percBSDown \%) downward events and were detected more frequently in
-longer time series and in populations with a smaller magnitude of process
-noise. Black swans were not associated with density dependence, population
-growth rate, or lifespan. In verified cases, black-swan events were often a
-result of the combination of extreme climate, predators and parasite cycles,
-and strong changes in human pressures. Our empirical results, sensitivity
-analyses, and simulation tests suggest that estimating the tail shape of
-process noise is a viable method of detecting black-swan population dynamics
-and if anything will underestimate the probability of black-swan events. The
-presence of black swans highlights the importance of developing management
-strategies that detect quickly, respond to, and are robust to extremes in
-population dynamics---particularly as the frequency and magnitude of climatic
-extremes increase over the next century \citep{easterling2000,ipcc2012}.
-
-Our results clarify previous related analyses. An analysis with an older
-version of the GPDD assessed the distribution of abundance time series but
-focused on identifying if the log-normal distribution was the most frequently
-parsimonious model \citep{halley2002}. For heavy-tailed distributions,
-\citet{halley2002} fit the extremely heavy-tailed Cauchy distribution and the
-four-parameter Levy stable distribution and found little information criteria
-support for these distributions in longer time series. However, black-swan
-events are by definition rare, and the majority of time series in the GPDD are
-short for these purposes. Therefore, we would not expect to observe black
-swans in a large proportion of populations. By quantifying the probability of
-heavy-tails and allowing for non-stationary time series and density
-dependence, our analysis allows for a more nuanced description of the evidence
-of ecological black swans. In an earlier study, \citet{keitt1998} described
-heavy (power law) tails in breeding bird population abundance. However, this
-finding was challenged by \citet{allen2001}, who showed that mixing data
-across species could falsely generate heavy tails.
-%This rebuttal supports our
-%use of the t distribution to represent heavy tails, since a t distribution can
-%be represented by a mixture of normal distributions \citep[with the same mean
-%and inverse-gamma-distributed variances,][]{gelman2014}.
-
-We might expect to observe black-swan dynamics in ecological time series
-because of unmodelled intrinsic properties of populations or extrinsic forces
-acting on populations. Since a t-distribution can be formed by a mixture of
-normal distributions \citep[with the same mean and inverse-gamma-distributed
-variances,][]{gelman2014}, we could observe heavy tails if we miss some
-underlying mixture of intrinsic processes \citep{allen2001}. That process
-might be an aggregation of populations across space, or population diversity,
-or an intrinsic change in population variability through time. Extrinsic
-forces could also cause black-swan dynamics \citep[e.g.][]{nunez2012}. These
-forces could be extreme themselves. For example, extreme climate, predation
-from (or competition with) other species experiencing black swans, or sharp
-changes in human pressure such hunting, fishing, or habitat destruction might
-cause black swans. Alternatively, the synchrony of multiple ``normal''
-extrinsic forces could give rise to black-swan ecological dynamics. This could
-occur if with a synergistic interaction \citep[e.g.][]{kirby2009} or even if
-non-synergistic forces experience a rare alignment \citep{denny2009}.
-
-There are a number of caveats when considering the generality of our results.
-The GPDD data represent a taxonomically and geographically biased sample of
-populations---the longer time series we focus on are dominated by commercially
-and recreationally important species and a disproportionate number of
-populations are located in the United Kingdom. Although we would expect to
-find qualitatively similar evidence for black swans in many other large
-taxonomic or geographic samples of populations, the common forces driving
-those black swans would likely differ (e.g.~severe winters in
-Table~\ref{tab:sparks}). In addition to a possibly biased sample of
-populations, some black-swan detections could just be recording mistakes, or
-conversely, some extreme observations may have been discarded or altered if
-they were erroneously suspected of being recording mistakes. Indeed, we
-discarded three of the populations that our method initially identified as
-heavy tailed because they turned out to be data-entry errors (Supporting
-Information). A final caveat is that the temporal scales of observation and
-population dynamics vary considerably across populations in the GPDD and this
-likely influences the detection of heavy tails. As an example, if we make
-frequent observations relative to generation time (e.g.~for many large-bodied
-mammals) we will average across generations and perhaps miss black swans.
-Conversely, if we census populations infrequently relative to generation time
-(e.g.~many insects in the GPDD) the recorded data may average across extreme
-and less-extreme events and also dampen black-swan dynamics.
-
-Recognizing the prevalence of heavy-tailed dynamics suggests a number of
-policy directions. Ecological resource management can draw from other
-disciplines that focus on heavy tails. For example, earthquake preparedness
-and response is focussed on black-swan events. Similarly to ecological black
-swans, we can rarely predict the specific location and timing of large
-earthquakes. But, earthquake preparedness involves spatial planning based on
-forecast probabilities to focus early detection efforts and develop disaster
-response plans. A similar focus might benefit resource management once our
-ability to predict the spatial probability and covariates of ecological black
-swans improves. The presence of black swans also suggests that we develop
-management policy that is robust to heavy tails. For instance, setting target
-population abundances that are appropriately set back from critical limits may
-buffer black-swan events \citep[e.g.][]{caddy1996}, and maintaining genetic,
-phenotypic, and behavioural diversity may allow some components of populations
-to persist when others are affected by disease or extreme environmental forces
-\citep[e.g.][]{hilborn2003, schindler2010, anderson2014}. Finally, extreme and
-unexpected, surprising, or counterintuitive ecological dynamics offer a
-tremendous opportunity to learn about ecological systems, evaluate when our
-models break down, and adjust future management policy \citep{doak2008,
-  pine-iii2009, lindenmayer2010}.
-
-Our results suggest a number of research questions related to ecological black
-swans. Given that black swans do occur, can we forecast the probability of
-black swans in space and time? Furthermore, what management policies allow us
-to detect them quickly after they happen? Can we isolate the components of
-ecological dynamics that experience black swans by moving from
-phenomenological models such as the Gompertz to mechanistic models that, for
-example, take into account recruitment dynamics? We expect that greater
-insight into the mechanisms and covariates of ecological black swans may be
-best obtained through specific geographic and taxonomic subsets of data where
-longer time series with low levels of observation error are available
-\citep[e.g.][]{segura2013}.
-
-Most importantly, what is the impact of allowing for black swans in forecasts
-of ecological risk? Even extremely rare catastrophes can have a profound
-influence on population persistence \citep{mangel1994}. In recent decades,
-ecology has moved toward focussing on aspects of variance in addition to mean
-responses \citep[e.g.][]{loreau2010a, thompson2013}. Our results suggest that
-an added focus on ecological extremes represents the next frontier,
-particularly in the face of increased climate extremes \citep{meehl2004,
-  thompson2013, ipcc2012}. Financial analysts are concerned with the shape of
-the downward tails of financial returns because these directly impact
-estimates of risk---the probability of a specific magnitude of undesired event
-occurring \citep{rachev2008}. A comparable focus in ecology would increase our
-estimates of extinction risk, since these would be disproportionately impacted
-by downward black-swan events.
-
-\section{Acknowledgements}
-
-We thank J.D. Yeakel and other members of the Earth to Ocean research
-group for helpful discussions and comments. We are grateful to the
-contributors and maintainers of the Global Population Dynamics Database and to
-Compute Canada's WestGrid high-performance computing resources. Silhouette
-images were obtained from \texttt{phylopic.org} under Creative Commons
-licenses; sources are listed in the Supporting Information. Funding was
-provided by a Simon Fraser University Graduate Fellowship (SCA), the Natural
-Sciences and Engineering Research Council of Canada (NKD, ABC), the Canada
-Research Chairs Program (NKD).
-
-\section{Supporting Information}
-
-The following supporting information is available online for this article:\\
-Tables S1 and S2\\
-Figures S1--S6\\
-Example Stan code for a heavy-tailed Gompertz model and multilevel beta
-regression\\
-The GPDD IDs used in our analysis\\
-All code and data to recreate our analysis are available at:\\
-\url{https://github.com/seananderson/heavy-tails}
-
-\renewcommand{\baselinestretch}{\tighttextstretch}
-\normalsize
-\bibliographystyle{apalike}
-\bibliography{/Users/seananderson/Dropbox/tex/jshort,/Users/seananderson/Dropbox/tex/ref3}
-
-\clearpage
-\renewcommand{\baselinestretch}{\textstretch}
-\normalsize
-
-\section{Tables}
-
-\input{sparks.tex}
+Black swans are statistically improbable events that nonetheless occur---often with profound consequences\cite{taleb2007, sornette2009}.
+While extremes in the physical environment, such as monsoons and heat waves, are widely studied and increasing in magnitude and frequency\cite{meehl2004, katz2005, ipcc2012}, it remains unclear the extent to which ecological populations buffer or suffer from such extremes.
+We derived a model for estimating the frequency of black-swan events and applied it to abundance time series of 609 animal populations.
+We find strong evidence for black-swan events, but they are rare, occurring in 3--5\% of populations: most frequently for birds (4--8\%), mammals (4--6\%), and insects (2--3\%).
+Black swan events were predominantly (87\%) downward, implying that unexpected population crashes occur far more frequently than unexpected population increases.
+Black swan events are not explained by any life-history covariates, but tend to be driven by external perturbations such as climate or severe winters.
+Data and simulations reveal that black swan-events are more likely to be detected in longer and less noisy time series.
+Our results show that accurate forecasts of extinction risk require accounting for the possibility of black swans and the predominance of crashes over increases, or predicted abundance will underestimate the probability of crashes by a factor of XX--XX.
+In addition, the presence of black swans highlights the importance of developing flexible and robust conservation and management strategies, particularly as the frequency and magnitude of climate extremes increase over the next century.
 
@@
 
+Major surprises happen more often than expected in financial, social, and natural systems\cite{taleb2007, may2008}.
+Massive unpredictable market swings are responsible for the majority of financial gains and losses\cite{taleb2007}, fatalities from the largest wars dwarf those from other wars\cite{newman2005}, and the most damaging earthquakes greatly exceed their expected frequency\cite{sornette2009}.
+In ecology, background rates of global extinction are punctuated by mass extinction\cite{harnik2012}, viruses can mutate suddenly to infect new hosts, and billions of animals can die at once in mass mortality events\cite{fey2015}.
+Indeed, catastrophes may be the most important element affecting population persistence\cite{mangel1994} and their importance is likely to increase given projected increases in the frequency and magnitude of climate-related extremes\cite{ipcc2012}.
+Despite this compelling anecdotal evidence for ecological black swans, systematic evidence in population time series of higher taxa has been elusive\cite{keitt1998, allen2001, halley2002} and the vast majority of model fitting and risk forecasting assumes that assumes that uncertainty can be represented by a normal distribution\cite{brook2006a, dennis2006, knape2012}.
+If black swans occur, though, a normal distribution would under-estimate the probability of extreme events occurring.
+
+Here, we assess the frequency and magnitude of black-swan dynamics across time series of 609 populations from a wide array of taxonomic groups---mostly birds, mammals, insects, and fishes.
+We then identify characteristics of time series or life-history traits associated with the detection of black-swan events and verify known causes.
+To accomplish this, we develop a framework for identifying heavy-tailed (black-swan) process noise in population dynamics, i.e.~whether the largest stochastic jumps in abundance from one time step to the next are more extreme than typically seen with a lognormal distribution.
+Our framework allows for a range of population dynamic models, can incorporate observation uncertainty and skewness in process noise, and can be readily applied to abundance time series.
+
+Specifically, we fit population dynamics models in which the process noise is drawn from a Student-t distribution\emph{.} By estimating the degrees of freedom parameter, \(\nu\), we can estimate the degree to which the process deviations have heavy tails and are thereof evidence of black-swan events (Fig.~1).
+For example, at \(\nu = 3\), the probability of drawing a value more than five standard deviations below the mean is \(0.02\), whereas the probability of drawing such a value from a normal distribution is tiny (\(2.9\cdot10^-7\)).
+As \(\nu\) approaches infinity, the distribution approaches the normal distribution (Fig.~1).
+
+We found strong, but rare, evidence for black-swan dynamics: most frequently for birds (7\%) followed by mammals (5\%), and insects (3\%) (Fig.~2, Table~S2).
+Black-swan events were taxonomically widespread, occurring in 38\% of taxonomic orders.
+Accounting for time series length and partially pooling inference across taxonomic class and order with a multilevel model, we found stronger evidence for black swans in insect populations than is apparent in Fig.~2---four of eight orders with the highest median probability of heavy tails were insect orders---however, there was considerable uncertainty in these estimates (Fig.~3a).
+
+The majority of our heavy-tailed estimates were robust to alternative population models, observation error, and choice of Bayesian priors.
+Our conclusions were not systematically altered when we included an autocorrelation structure in the residuals, modelled population growth rates with or without density dependence, or modelled the population dynamics as Ricker-logistic instead of Gompertz (Fig.~S3).
+Introducing moderate observation error (CV = 0.2) slightly decreased the estimated prevalence of black-swan events (Fig.~S3).
+The strength of the prior on ν had little influence on estimates of black-swan dynamics (Fig.~S4).
+Our simulation testing shows that, if anything, our models underpredict the true magnitude and probability of heavy tailed events---especially given the length of the time series in our analysis (Figs S5, S6).
+
+Across populations, the probability of observing black-swan dynamics was positively related to time-series length and negatively related to magnitude of process noise but not clearly related to population growth rate, density dependence, or maximum lifespan (Figs 3b, 4).
+Longer time-series length was the strongest covariate of observing black-swan dynamics.
+For instance, the expected probability density suggesting black-swan dynamics was about 1.6 times greater for a population with 60 time steps compared to one with 30 time steps.
+However, the absolute change in probability with increased time series length was small (0.20 vs.~0.13 in this example, Fig.~4).
+
+We examined all time series with published explanations of why black-swan events occurred (Table~S2).
+The majority (87\%) were downward black swans and involved a combination of multiple factors.
+For example, a synchronization of environmental- and predation-mediated population cycles is thought to have caused a downward black-swan event for a water vole (\emph{Arvicola terrestris}) population\cite{saucy1994}.
+Other black swans were the result of a sequence of extreme climate events on their own.
+For instance, severe winters in 1929, 1940--1942, and 1962--1963 were associated with black-swan downswings in grey heron (\emph{Ardea cinerea}) abundance in the United Kingdom\cite{stafford1971} (Fig.~4c).
+Our analysis finds that the last event was a combination of two black-swan events in a row and it took the population three times longer to recover than predicted\cite{stafford1971}.
+Downward black swans were sometimes followed by upward black swans.
+For example, during a period of population crowding and nest shortages, a population of European shag cormorants (\emph{Phalacrocorax aristotelis}) on the Farne Islands, United Kingdom, declined suddenly following a red tide event in 1968\cite{potts1980}.
+This freed quality nest sites for first-time breeders, productivity rapidly increased, and the population experienced a rapid upswing in abundance\cite{potts1980}.
+
+Given the prevalence of downwards events, we refit our heavy-tailed models estimating the skewness of process noise (\(\gamma\); Figs 1a, SX) and used these models to make near-term risk forecasts.
+Aggregated across populations with strong evidence of heavy tails (median \(\nu < 10\)), 85\% of the \(\gamma\) probability density was below 1, indicating strong evidence of downwardly skewed process noise (Fig.~4b, SX).
+In contrast, populations with little evidence of heavy tails (median \(\nu \geq 70\)) also had little evidence of skewed process noise (Fig.~4b; 90\% of 95\% credible intervals overlapped 1).
+Projecting these heavy-tailed populations forward five years revealed that assuming lognormal process noise underestimated risk by X--X fold (ratio of interquartile range of 99\% lower credible interval of abundance).
+\emph{TODO: closing sentence emphasizing the implications of this?}
+
+Thus our analysis provides strong evidence for the presence of downwardly skewed black-swan events in ecological process noise of higher taxa.
+Unmodelled intrinsic properties of populations or extrinsic forces acting on populations could generate black-swan dynamics.
+Given that a mixture of normal distributions with different variances can generate a Student-t distribution with heavy tails\cite{allen2001}, we could observe black-swan dynamics if we miss an underlying mixture of processes.
+Those processes might be an aggregation of populations across space, population diversity\cite{anderson2015}, or an intrinsic change in population variability through time.
+Extrinsic ecological forces could also cause black-swan dynamics\cite{nunez2012}.
+These forces could be extreme themselves.
+For example, extreme climate, predation from (or competition with) other species experiencing black swans, or sharp changes in human pressures such hunting, fishing, or habitat destruction might cause black-swan events.
+Alternatively, the synchrony of multiple extrinsic forces could give rise to black-swan ecological dynamics.
+This could occur with synergistic interactions\cite{kirby2009} or even if non-synergistic forces experience a rare alignment\cite{denny2009}.
+
+There are a number of caveats when considering the generality of our results.
+The GPDD data represent a taxonomically and geographically biased sample of populations---the longer time series we focus on are dominated by commercially and recreationally important species and a disproportionate number of populations are located in the United Kingdom.
+Although we would expect to find qualitatively similar evidence for black swans in other large taxonomic or geographic samples of populations, the common forces driving those black swans would likely differ.
+Additionally, some apparent black-swan events could be recording mistakes, or conversely, some extreme observations may have been discarded or altered if they were erroneously suspected of being recording mistakes.
+Indeed, we discarded three of the populations that our method initially identified as heavy tailed because they turned out to be data-entry errors (Supporting Information).
+Finally, the temporal scales of observation and population dynamics vary considerably across populations in the GPDD and these likely influence the detection of heavy tails.
+For example, if we make frequent observations relative to generation time (e.g.~for many large-bodied mammals) we will average across generations and perhaps miss black-swan events.
+Conversely, if we census populations infrequently relative to generation time (e.g.~many insects in the GPDD) the recorded data may average across extreme and less-extreme events and also dampen black-swan dynamics.
+
+Recognizing the prevalence of heavy-tailed dynamics suggests a number of policy directions.
+Ecological resource management can draw from other disciplines that focus on heavy tails.
+For example, earthquake preparedness and response is focussed on black-swan events.
+Similarly to ecological black swans, we can rarely predict the specific location and timing of large earthquakes.
+But, earthquake preparedness involves spatial planning based on forecast probabilities to focus early detection efforts and develop disaster response plans\cite{nrc2007}.
+A similar focus might benefit ecological resource management once our ability to predict the spatial probability and covariates of ecological black swans improves.
+The presence of black swans also suggests that we develop management policy that is robust to heavy tails.
+For instance, setting target population abundances that are appropriately set back from critical limits may buffer black-swan events\cite{caddy1996}, and maintaining genetic, phenotypic, and behavioural diversity may allow some components of populations to persist when others are affected by disease or extreme environmental forces\cite{hilborn2003, schindler2010, anderson2015}.
+Finally, extreme and unexpected, surprising, or counterintuitive ecological dynamics offer a tremendous opportunity to learn about ecological systems, evaluate when models break down, and adjust future management policy\cite{doak2008, pine-iii2009, lindenmayer2010}
+
+Rare catastrophes can have a profound influence on population persistence\cite{mangel1994}.
+In recent decades, ecology has moved toward focussing on aspects of variance in addition to mean responses\cite{loreau2010a, thompson2013}.
+Our results suggest that an added focus on ecological extremes represents the next frontier, particularly in the face of increased climate extremes\cite{meehl2004, ipcc2012, thompson2013}.
+Financial analysts are concerned with the shape of the downward tails of financial returns because these directly impact estimates of risk---the probability of a specific magnitude of undesired event occurring\cite{rachev2008}.
+A comparable focus in ecology would increase our estimates of extinction risk, since these would be disproportionately impacted by downward black-swan events.
+Similarly, ecologists should focus more on estimating and predicting downward tails of population abundance, since these increase true extinction risk.
+
+\textbf{References}
+
+\renewcommand{\section}[2]{}%
+
+\bibliographystyle{nature}\bibliography{jshort,ms}
+
+\textbf{Online Content:} Methods, along with any Extended Data display items are available in the online version of the paper; references unique to these sections appear only in the online paper.
+
+\textbf{Acknowledgements}: We thank J.W.
+Moore, J.D.
+Yeakel, C.
+Minto and members of the Earth to Ocean Group for helpful discussions and comments.
+We are grateful to the contributors and maintainers of the Global Population Dynamics Database and to Compute Canada's WestGrid high-performance computing resources.
+Silhouette images were obtained from phylopic.org under Creative Commons licenses; sources are listed in the Supporting Information.
+Funding was provided by a Simon Fraser University Graduate Fellowship (S.C.A.), the Natural Sciences and Engineering Research Council of Canada (N.K.D., A.B.C.), the Canada Research Chairs Program (N.K.D.).
+
+\textbf{Author Contributions}: S.C.A.
+and T.A.B.
+conceived the project; S.C.A., T.A.B., A.B.C., and N.K.D.
+designed the project; S.C.A.
+analyzed the data and wrote the paper with input from all authors.
+
+\textbf{Author information}: The authors declare no competing financial interests.
+Reprints and permissions information is available at .
+Correspondence and requests for materials should be addressed to S.C.A.
+(sean\_anderson@sfu.ca).
+
+\clearpage
+
+\textbf{Figures}
+
+\begin{center}
+\includegraphics[width=\textwidth]{../analysis/t-nu-eg2.pdf}
+\end{center}
+
+\textbf{Figure~1: An illustration of fitting population dynamic models that allow for heavy tails, represented by the Student-t degrees of freedom parameter} \(\mathbf{\nu}\)\textbf{.
+a, b,} The probability density for Student-t distributions with a scale parameter of 1 and different values of \(\nu\).
+Small values of \(\nu\) create heavy tails.
+As \(\nu\) approaches infinity the distribution approaches the normal distribution.
+For example, at \(\nu = 2\), the probability of drawing a value more than five standard deviations below the mean is 0.02, whereas the probability of drawing such a value from a normal distribution is nearly zero (\(2.9 \cdot 10^{- 7}\)).
+\textbf{c--e,} Simulated population dynamics from a Gompertz model with process noise drawn from t distributions with three different values of \(\nu\).
+Coloured dots in panels \textbf{c} and \textbf{d} represent jumps with less than a 1 in 1000 chance of occurring in a normal distribution.
+\textbf{f--h,} Estimates of \(\nu\) from models fit to the times series in \textbf{c--e}.
+Shown are the posterior samples (histograms), median and interquartile range of the posterior (IQR, dots and line segments), and the exponential prior on \(\nu\) (dashed lines).
+Colour shading behind \textbf{f--h} illustrates the region of heavy tails.
+
+\clearpage
+
+\begin{center}
+\includegraphics[width=0.36\textwidth]{../analysis/nu-coefs-2.pdf}
+\end{center}
+
+\textbf{Figure~2: Estimates of population dynamic heavy-tailedness for populations of birds, mammals, insects, and fishes.} Small values of \(\nu\) approximately (\(< 10\)) suggest heavy-tailed black-swan dynamics; larger values of \(\nu\) suggest approximately normal-tailed dynamics.
+Vertical points and line segments represent posterior medians and 50\% / 90\% credible intervals for individual populations.
+Inset plots show probability that \(\nu < 10\) (probability of heavy tails) for populations arranged by taxonomic order and sorted by decreasing mean Pr(\(\nu < 10\)).
+Taxonomic orders with three or fewer populations in \textbf{a} are omitted for space.
+Red to yellow points highlight populations with a high to moderately high probability of heavy-tailed black-swan dynamics.
+
+\clearpage
+
+\begin{center}
+\includegraphics[width=0.35\textwidth]{../analysis/order-posteriors-covariates.pdf}
+\end{center}
+
+\textbf{Figure~3: Standardized probabilities and covariates of heavy-tailed dynamics.} \textbf{a,} Taxonomic-order-level posterior densities of Pr(\(\nu < 10\)) after accounting for time-series length in a multilevel model.
+Estimates are at the geometric mean of time series length across all the data (approximately 27 time steps).
+Colour shading refers to taxonomic class (yellow: fishes, green: insects, purple: birds, and red: mammals).
+Dotted vertical line in \textbf{a} indicates the median expected Pr(\(\nu < 10\)) from the prior distribution.
+\textbf{b,} Posterior densities for potential covariates of Pr(\(\nu < 10\)).
+The beta regression models were fit on a logit scale with hierarchical intercepts for taxonomic class, taxonomic order, and species.
+All covariates were standardized by subtracting their mean and dividing by twice their standard deviation.
+In both panels, short vertical line segments within the density polygons indicate medians.
+
+\clearpage
+
+\begin{center}
+\includegraphics[width=0.55\textwidth]{../analysis/skew-fig.pdf}
+\end{center}
+
+\textbf{Figure~4: Evidence for downward skewness in process noise for heavy-tailed populations.
+a,} Illustrations of Student-t distributions with 3 levels of skewness (\(\gamma\)) and heavy-tails (\(\nu\)).
+\textbf{b,} Posterior density of the skewness parameters aggregated across all populations grouped into heavy-tailed populations (\(\hat{\nu} < 10\)), slightly heavy-tailed populations (\(10 \leq \hat{\nu} \leq 70\)), and normal tails (\(\hat{\nu} > 70\)).
+The approximate median posterior density of \(\gamma\) and a mid-value of \(\nu\) from distribution in \textbf{b} are shown in \textbf{a}.
+\textbf{c,} Example time series of heavy-tailed populations with different levels of skewness.
+Red dots highlight likely heavy-tailed events.
+Forecasts (grey regions) show median (solid lines) and lower 99\% credible intervals (dotted lines) of abundance.
+Black lines indicate forecasts from a standard Gompertz model with lognormal process noise.
+Red lines indicate forecasts from a Gompertz model with estimated tail shape and skewness.
+
+\clearpage
+
+\textbf{Methods}
+
+We selected abundance time series from the Global Population Dynamics Database (GPDD;\cite{gpdd2010}), which contains nearly 5000 time series of abundance from \(\sim 1000\) species and \(\sim 100\) taxonomic orders.
+We filtered the data (Supporting materials) to remove populations from less reliable data sources, and those without sufficient data for our models, and then interpolated some missing values (sensu Brook \textit{et al.}\cite{brook2006a}).
+Our interpolation affected only \(\sim\) \interpPointsPerc \% of the final data points (Supporting materials Table~\ref{tab:stats}) and none of the data points that were later considered black-swan events.
+Our final dataset contained \NPops\ populations across \NOrders\ taxonomic orders and seven taxonomic classes, with a median of \medianTimeSteps\ time steps (range of \minTimeSteps--\maxTimeSteps) (Extended Data Table~\ref{tab:stats}).
+
+We fit heavy-tailed Gompertz population dynamics models to data from the Global Population Dynamics Database (GPDD).
+\emph{{[}TODO something about extensive quality control filtering and interpolation of the time series{]}} The Gompertz model represents population growth as a linear function in log space.
+If \(x_{t}\) represents the log abundance (\(N\)) at time \emph{t} then \(x_t = \lambda + bx_{t - 1} + \epsilon_t,\ \epsilon_t \sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma)\).
+The growth parameter \(\lambda\) represents the expected growth rate if \(N_{t} = 1\).
+The model is density independent if \(b = 1\), maximally density dependent if \(b = 0\), and inversely density dependent if \(b < 0\).
+
+Usually, the process noise \(\epsilon_{t}\) is modelled as normally distributed{[}REFs{]}, but we allow it to be drawn from a Student-t distribution with scale parameter \(\sigma\) and degrees of freedom \(\nu\).
+We can then estimate the degree to which the process deviations have heavy tails and are thereof evidence of black-swan events (Fig.~1a, b).
+
+We chose a weakly informative exponential prior on \(\nu\) with rate parameter of \(0.01\) truncated at values above two---a slightly less informative prior than suggested by Fernandez and Steel\cite{fernandez1998}.
+This prior gives only a \(0.08\) probability that \(\nu < 10\) but constrains sampling sufficiently to avoid drifting towards infinity.
+In any case, for \(\nu > 20\) the t distribution is almost indistinguishable from the normal distribution (Fig.~1).
+Based on the shape of the t distribution, we chose the probability that \(\nu < 10\), Pr(\(\nu < 10\)), to define the probability of heavy-tailed dynamics.
+When categorizing a population as heavy or normal tailed, we applied a 0.5 probability threshold.
+
+\emph{Details on the heavy-tailed Gompertz probability model}
+
+For the Gompertz model, our weakly-informative priors (Extended Data Fig.~\ref{fig:priors}) were: \(b \sim \mathrm{Uniform}(-1, 2)\), \(\lambda \sim \mathrm{Normal}(0, 10^2)\), \(\sigma \sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)\), \(\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2)\).
+Our prior on \(b\) was uninformative between values of \(-1\) and \(2\).
+We would not expect values of \(b\) with levels of density dependence as low as \(-1\) (very strong inverse density dependence), nor would we generally expect values above \(1\).
+We allowed values of \(b\) above \(1\) to allow for non-stationary time series of growth rates.
+The estimates of \(b\) were well within these bounds.
+Our prior on \(\lambda\) was very weakly informative within the range of expected values for population growth and is similar to the default priors suggested by Gelman \emph{et al.}\cite{gelman2008d} for intercepts of regression models.
+Our Half-Cauchy prior on \(\sigma\) follows Gelman \emph{et al.}\cite{gelman2006c} and Gelman \emph{et al.}\cite{gelman2008d} and the specific scale parameter of \(2.5\) is based on our expected range of the value in nature from previous studies\cite{connors2014}.
+In our testing of a subsample of populations, our parameter estimates were not qualitatively changed by switching to an uninformative uniform prior on \(\sigma\), but the weakly informative Half-Cauchy prior sped up chain convergence.
+
+Our prior on \(\nu\) was based on Fernandez and Steel\cite{fernandez1998}.
+They chose an exponential rate parameter of \(0.1\).
+We chose a less informative rate parameter of \(0.01\) and truncated the distribution at \(2\), since at \(\nu < 2\) the variance of the t distribution is undefined.
+This prior gives only a \(7.7\)\% probability that \(\nu < 10\) but constrains the sampling sufficiently to avoid wandering off towards infinity---above approximately \(\nu = 20\) the t distribution is so similar to the normal distribution (Fig.~1) that time series of the length considered here are unlikely to be informative about the precise value of \(\nu\).
+In the scenario where the data are uninformative about heavy tails (e.g.~Fig.~1e, h), the posterior will approximately match the prior (prior median \(= 71\), mean \(= 102\)) and the metrics used in our paper (e.g.
+\(Pr(\nu < 10) > 0.5\)) are unlikely to flag the population as heavy tailed.
+
+We fit our models with Stan 2.4.0\cite{stan-manual2014}, and R 3.1.1\cite{r2014}.
+We began with four chains and \(2000\) iterations, discarding the first \(1000\) as warm up (i.e.~4000 total samples).
+If \(\hat{R}\) (the potential scale reduction factor---a measure of chain convergence) was greater than \(1.05\) for any parameter or the minimum effective sample size, \(n_\mathrm{eff}\), (a measure of the effective number of uncorrelated samples) for any parameter was less than \(200\), we doubled both the total iterations and warm up period and sampled from the model again.
+These thresholds are in excess of the minimums recommended by Gelman \emph{et al.}\cite{gelman2006a} of \(\hat{R} < 1.1\) and effective sample size \(> 100\) for reliable point estimates and confidence intervals.
+In the majority of cases our minimum thresholds were greatly exceeded.
+We continued this procedure up to \(8000\) iterations (\(16000\) total samples) by which all chains were deemed to have sufficiently converged.
+The No-U-Turn Hamiltonian Markov chain Monte Carlo Sampler in Stan generally requires far fewer iterations to obtain equivalent effective sample sizes than other software such as WinBUGS or JAGS\cite{stan-manual2014}.
+
+\emph{The skewed Student's t distribution}
+
+We fit Gompertz models with skewed Student's t distributed process noise based on the probability density for the skewed t distribution described in Fernandez and Steel\cite{fernandez1998}.
+Their definition equates to the following probability density as implemented in King\cite{king2012}:
+
+\begin{equation*}
+  \mathrm{Skewed\,Student\mhyphen t}(x) =
+\begin{dcases}
+\frac{2}{(\gamma + 1/\gamma)}
+         \mathrm{Student\mhyphen t}(x \cdot \gamma, \nu, \mu \cdot \gamma, \sigma),& \text{for } x < 0\\
+       \frac{2}{(\gamma + 1/\gamma)}
+         \mathrm{Student\mhyphen t}(x / \gamma, \nu, \mu / \gamma, \sigma),& \text{for } x \ge 0
+\end{dcases}
+\end{equation*}
+
+Here \(\gamma\) represents the skewness parameter.
+The Student-t(\(\cdot\)) portion represents the data point of interest, the heavy-tailedness parameter, the location parameter, and the scale parameter as fed to the standard Student-t probability density.
+The distribution is symmetrical if \(\gamma = 1\), left skewed (skewed downwards in abundance) if \(0 < \gamma < 1\), and right skewed (skewed upwards in abundance) if \(1 < \gamma < \infty\).
+
+\emph{Alternative priors}
+
+To test if the prior on \(\nu\) influenced our estimate of black-swan dynamics, we refit our models with weaker and stronger priors.
+Our base model used a prior on \(\nu\) of Truncated-Exponential(0.02, min.\ = 2).
+For a weaker prior we used Truncated-Exponential(0.005, min.\ = 2) and for a stronger prior we used Truncated-Exponential(0.02, min.\ = 2) (Extended Data Fig.~\ref{fig:priors}).
+Note that the base and weaker priors are relatively flat within the region of \(\nu < 20\), which is the region we are mostly concerned about when categorizing populations as heavy- or thin-tailed.
+
+Our results show that these weaker and stronger priors would have little influence on our conclusions about heavy-tailed dynamics (Extended Data Fig.~\ref{fig:alt-priors}).
+When the data are informative about tail behaviour (i.e.
+when there is strong evidence of low \(\nu\) values, upper-right of Extended Data Fig.~\ref{fig:alt-priors}), the prior has little impact on the estimate of \(\nu\).
+When the data are less informative about \(\nu\) (i.e.\ when there are no or few tail events and time series are short or noisy), the prior can pull the estimate of \(\nu\) towards larger or smaller values (Extended Data Fig.~\ref{fig:alt-priors}).
+The vast majority of the populations with Pr\((\nu < 10)\) in the base prior were not altered qualitatively by this range of prior strength.
+
+\emph{Alternative population models}
+
+We fit alternative population models to test if four key phenomena systematically changed our conclusions.
+These models allowed for AR1 autocorrelation of the residuals, density independence, observation error, and Ricker-logistic population dynamics.
+Furthermore, we tested the sensitivity of our results to weaker and stronger priors on \(\nu\) and used simulated data to test how easily we could detect \(\nu\) given different sample sizes and to ensure we could recover unbiased parameter estimates.
+The range of percentages of black swans by taxonomic class cited in the abstract and results are based on lower and upper limits across our main Gompertz model and these four alternative models.
+
+\emph{Autocorrelated residuals}
+
+We considered a version of the Gompertz model in which an autoregressive parameter was fit to the process noise residuals: \(x_t = \lambda + b x_{t-1} + \epsilon_t,\, \epsilon_t \sim \mathrm{Student\mhyphen t}(\nu, \phi \epsilon_{t-1}, \sigma)\).
+In addition to the parameters in the original Gompertz model, this model estimates an additional parameter \(\phi\), which represents the correlation of subsequent residuals.
+Based on the results of previous analyses with the GPDD\cite{connors2014} and the chosen priors in previous analyses\cite{thorson2014a} and to greatly speed up chain convergence when running our model across all populations, we placed a weakly informative prior on \(\phi\) that assumed the greatest probability density near zero with the reduced possibility of \(\phi\) being near \(-1\) or \(1\).
+Specifically, we chose \(\phi \sim \mathrm{Truncated\mhyphen Normal}(0, 1, \mathrm{min.} = -1, \mathrm{max.} = 1)\).
+The MCMC chains did not converge for \modelsNoConvergeAROne\ populations according to our criteria (\(\widehat{R} < 1.05, n_\mathrm{eff} > 200\)) after 8000 iterations of four chains.
+This included only \modelsNoConvergeAROneHeavyBase\ populations in which Pr(\(\nu < 10\)) \(> 0.5\) categorized them as heavy in the main Gompertz model.
+We did not include these models in Extended Data Fig.~\ref{fig:alt}.
+
+\emph{Assumed density independence}
+
+We fit a simplified version of the Gompertz model in which the density dependence parameter \(b\) was fixed at \(1\) (density independent).
+This is equivalent to fitting a random walk model (with drift) to the \(\log\) abundances or assuming the growth rates are drawn from a stationary distribution.
+The model was \(x_t = \lambda + x_{t-1} + \epsilon_t,\, \epsilon \sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma)\).
+We fit this model for three reasons: (1) it is computationally simpler and so provides a check that our more complicated full Gompertz model was obtaining reasonable estimates of \(\nu\), (2) it provides a test of whether density dependence was systematically affecting our perception of heavy tails, (3) it matches how some previous authors have modelled heavy tails without accounting for density dependence\cite{segura2013}.
+
+\emph{Assumed observation error}
+
+Observation error can bias parameter estimates\cite{knape2012} and is known to affect the ability to detect extreme events\cite{ward2007}.
+In our main analysis, we fit a model that ignored observation error.
+One way to account for observation error would be to fit a full state-space model that simultaneously estimates the magnitude of process noise and observation error.
+However, simultaneously estimating observation and process noise is a challenging problem (e.g.\ because the observation and process noise parameters tend to negatively covary in model fitting) and is known to sometimes result in identifiability issues with the Gompertz population model\cite{knape2008}.
+Furthermore, our model was applied to hundreds of time series, often of short length (as few as 20 time steps) and our model estimates an additional parameter---the shape of the process deviation tails---potentially making identifiability and computational issues even greater.
+Therefore, we considered a version of the base Gompertz model that allowed for a fixed level of observation error:
+
+\begin{align*}
+U_t &= \lambda + b U_{t-1} + \epsilon_t\\
+x_t &\sim \mathrm{Normal}(U_t, \sigma_\mathrm{obs}^2)\\
+\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma_\mathrm{proc}),
+\end{align*}
+
+where \(U\) represents the unobserved state vector, \(\sigma_\mathrm{obs}\) represents the standard deviation of observation error (on a log scale), and \(\sigma_\mathrm{proc}\) represent the process noise scale parameter.
+We set \(\sigma_\mathrm{obs}\) to \(0.2\), which represents the upper limit of values often used in simulation analyses\cite{valpine2002, thorson2014b}.
+
+\emph{Ricker-logistic}
+
+We also fit a Ricker-logistic model: \(x_t = x_{t-1} + r_{\mathrm{max}}\left(1 - N_{t-1}/K\right) + \epsilon_t,\, \epsilon_t \sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma)\), where \(r_\mathrm{max}\) represents the theoretical maximum population growth rate that is obtained when \(N_t\) (abundance at time \(t\)) \(= 0\).
+The parameter \(K\) represents the carrying capacity and, as before, \(x_t\) represents the \(\log\) transformed abundance at time \(t\).
+The Ricker-logistic model assumes a linear decrease in population growth rate with increases in abundance.
+In contrast, the Gompertz model assumes a linear decrease in population growth rate with increases in \textit{log} abundance (\(x_t\))\cite{thibaut2012}.
+
+To fit the Ricker-logistic models, we chose a prior on \(K\) uniform between zero and twice the maximum observed abundance Clark \emph{et al.}\cite{clark2010} chose uniform between zero and maximum observed, which is more informative).
+We set the prior on \(r_\mathrm{max}\) as uniform between 0 and 20 as in Clark \emph{et al.}\cite{clark2010}.
+We used the same priors on \(\nu\) and \(\sigma\) as in the Gompertz model.
+
+\emph{Modelling covariates of heavy-tailed dynamics}
+
+We fit a multilevel beta regression model to the predicted probability of heavy tails, Pr(\(\nu < 10\)), to investigate potential covariates of heavy-tailed dynamics.
+The beta distribution is useful when response data range on a continuous scale between zero and one\cite{ferrari2004}.
+We used a logit link function as is typically used in logistic regression.
+The model was as follows:
+
+\begin{align*}
+\mathrm{Pr}(\nu_i < 10) &\sim \mathrm{Beta}(A_i, B_i)\\
+\mu_i &= \mathrm{logit}^{-1}(\alpha
+  + \alpha^\mathrm{class}_{j[i]}
+  + \alpha^\mathrm{order}_{k[i]}
+  + \alpha^\mathrm{species}_{l[i]}
+  + X_i \beta),
+  \: \text{for } i = 1, \dots, 617\\
+A_i &= \phi_\mathrm{disp} \mu_i\\
+B_i &= \phi_\mathrm{disp} (1 - \mu_i)\\
+\alpha^\mathrm{class}_j &\sim
+  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{class}}),
+  \: \text{for } j = 1, \dots, 6\\
+\alpha^\mathrm{order}_k &\sim
+  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{order}}),
+  \: \text{for } k = 1, \dots, 38\\
+\alpha^\mathrm{species}_l &\sim
+  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{species}}),
+  \: \text{for } l = 1, \dots, 301,
+\end{align*}
+
+where \(A\) and \(B\) represent the beta distribution shape parameters; \(\mu_i\) represents the predicted value for population \(i\), class \(j\), order \(k\), and species \(l\); \(\phi_\mathrm{disp}\) represents the dispersion parameter; and \(X_i\) represents a vector of predictors (such as lifespan) for population \(i\) with associated \(\beta\) coefficients.
+The intercepts are allowed to vary from the overall intercept \(\alpha\) by taxonomic class (\(\alpha^\mathrm{class}_j\)), taxonomic order (\(\alpha^\mathrm{order}_k\)), and species (\(\alpha^\mathrm{species}_l\)) with standard deviations \(\sigma_{\alpha \;  \mathrm{class}}\), \(\sigma_{\alpha \; \mathrm{order}}\), and \(\sigma_{\alpha  \; \mathrm{species}}\).
+Where possible, we also allowed for error distributions around the predictors by incorporating the standard deviation of the posterior samples for the Gompertz parameters \(\lambda\), \(b\), and \(\log \sigma\) around the mean point value as normal distributions (not shown in the above equation).
+
+We log transformed \(\sigma\), time-series length, and lifespan to match the way they are visually represented in Extended Data Fig.~\ref{fig:correlates} and to make the relationship approximately linear on the logit-transformed response scale.
+All input variables were standardized by subtracting their mean and dividing by two standard deviations to make their coefficients comparable in magnitude \cite{gelman2008c}.
+We excluded body length as a covariate because it was highly correlated with lifespan, and lifespan exhibited more overlap across taxonomy than body length.
+Lifespan is also more directly related to time and potential mechanisms driving black-swan dynamics.
+
+We incorporated weakly informative priors into our model: \(\mathrm{Cauchy}(0, 10)\) on the global intercept \(\alpha\), \(\mathrm{Half\mhyphen Cauchy}(0, 2.5)\) on all standard deviation parameters, \(\mathrm{Half\mhyphen Cauchy}(0, 10)\) on the dispersion parameter \(\phi_\mathrm{disp}\), and \(\mathrm{Cauchy}(0, 2.5)\) on all other parameters \cite{gelman2006c, gelman2008d}.
+Compared to normal priors, the Cauchy priors concentrate more probability density around expected parameter values while allowing for a higher probability density far into the tails, thereby allowing the data to dominate the posterior more strongly if it disagrees with the prior.
+Our conclusions were not qualitatively changed by using uniform priors.
+We fit our models with 5000 total iterations per chain, 2500 warm-up iterations, four chains, and discarding every second sample to save memory.
+We checked for chain convergence visually and with the same criteria as before (\(\widehat{R} < 1.05\) and \(n_\mathrm{eff} >200\) for all parameters).
+
+To derive taxonomic-order-level estimates of the probability of heavy tails accounting for time-series length (Fig.~3a), we fit a separate multilevel model with the same structure but with only \(\log\) time-series length as a predictor.
+(In this case, we did not want to control for intrinsic population characteristics such as density dependence.) Since our predictors were centered by subtracting their mean value, we obtained order-level estimates of the probability of heavy tails at mean log time-series length by adding the posteriors for \(\alpha\), \(\alpha^\mathrm{class}_j\), and \(\alpha^\mathrm{order}_k\).
+
+\clearpage
+
+\textbf{Extended Data Tables}
+
+\newenvironment{helvetica}{\fontfamily{phv}\selectfont}{\par}
+
+\singlespacing
+
+\textbf{Extended Data Table~1: Summary statistics for the filtered Global Population Dynamics Database
+time series arranged by taxonomic class.} Columns are: number of populations, number of taxonomic orders, numbers of species, median time series length, total number of interpolated time steps, total number of substituted zeros, and total number of time steps.
+
+\onehalfspacing
+
+\begin{helvetica}
+\smallskip
+\begin{scriptsize}
+\begin{tabular}{lrrrrrrrr}
+\toprule
+\input{../analysis/stat-table.tex}
+\label{tab:stats}
+\end{tabular}
+\end{scriptsize}
+\end{helvetica}
+
+\clearpage
+
+\LTcapwidth=\textwidth
+
+\renewcommand{\tablename}{\textbf{Extended Data Table}}
+
+\setcounter{table}{1}
+
+\begin{helvetica}
+\singlespacing
+\begin{scriptsize}
+\begin{longtable}{>{\RaggedRight}m{1.4cm}>{\RaggedRight}p{3.3cm}>{\RaggedRight}p{0.7cm}>{\RaggedRight}p{1.5cm}>{\RaggedRight}p{3.0cm}>{\RaggedRight}p{1.4cm}>{\RaggedRight}p{1.25cm}}
+
+\caption{\textbf{All populations with Pr$(\nu < 10) > 0.5$ in the base heavy-tailed
+  Gompertz population dynamics model.} Shown are the log abundance time series,
+  population descriptions, Global Population Dynamics Database Main IDs,
+  citation for the data source or separate verification literature, a
+  description of the cause of the black swan events (if known), the
+  probability of heavy tails as calculated by our model, and median estimate
+  of $\nu$ from our model with 90\% quantile credible intervals indicated in
+  parentheses. Red dots on the time series indicate downward black-swan events
+  and blue values indicate upward black-swan events that have a $10^{-4}$ probability or less of occurring if the population dynamics were
+  explained by a Gompertz model with normally distributed process noise with a
+  standard deviation equal to the scale parameter in the fitted t
+  distribution.}\\
+
+\toprule
+\input{../analysis/cause-table.tex}
+\label{tab:causes-supp}
+\end{longtable}
+\end{scriptsize}
+\onehalfspacing
+\end{helvetica}
+
+\clearpage
+
+\textbf{Extended Data Figures}
+
+\renewcommand{\figurename}{\textbf{Extended Data Fig.}}
+
 \begin{figure}[htbp]
 \begin{center}
-\includegraphics[width=\textwidth]{../analysis/t-nu-eg2.pdf}
-\caption[An illustration of fitting population dynamic models that allow for heavy
-tails, represented by the Student-t degrees of freedom parameter $\nu$.]{
-An illustration of fitting population dynamic models that allow for heavy
-tails, represented by the Student-t degrees of freedom parameter $\nu$. (a, b)
-The probability density for t distributions with a scale parameter of 1 and
-different values of $\nu$. Small values of $\nu$ create heavy tails. As $\nu$
-approaches infinity the distribution approaches the normal distribution. For
-example, at $\nu = 2$, the probability of drawing a value more than five
-standard deviations below the mean is 1.8\%, whereas the probability of drawing
-such a value from a normal distribution is nearly zero ($2.9\cdot10^{-5}$\%).
-(c--e) Simulated population dynamics from a Gompertz model with process noise
-drawn from t distributions with three different values of $\nu$. Coloured dots
-in panels c and d represent jumps with less than a 1 in 1000 chance of
-occurring in a normal distribution. (f--h) Estimates of $\nu$ from models fit
-to the times series in panels c--e. Shown are the posterior samples
-(histograms), median and interquartile range of the posterior (IQR) (dots and
-line segments), and the exponential prior on $\nu$ (dashed lines). Colour
-shading behind panels f--h illustrates the region of heavy tails.
-}
-\label{fig:didactic}
+\includegraphics[width=0.6\textwidth]{../analysis/priors-gomp-base.pdf}
+
+\caption{
+  \textbf{Probability density of the Bayesian priors for the Gompertz models.} From
+  left to right and then top to bottom: (1) per capita growth rate at
+  $\log$(abundance) = $0$: $\lambda \sim \mathrm{Normal}(0, 10^2)$; (2) scale
+  parameter of t-distribution process noise: $\sigma \sim \mathrm{Half\mhyphen
+    Cauchy} (0, 2.5)$; (3) t-distribution degrees of freedom parameter: $\nu
+  \sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2)$; (4)
+  AR1 correlation coefficient of residuals: $\phi \sim \mathrm{Truncated
+    \mhyphen Normal}(0, 1, \mathrm{min.} = -1, \mathrm{max.} = 1)$. Not shown
+  is $b$, the density dependence parameter: $b \sim \mathrm{Uniform}(-1, 2)$.
+  The $\nu$ panel also shows two alternative priors: a weaker prior $\nu \sim
+  \mathrm{Truncated\mhyphen Exponential}(0.005, \mathrm{min.} = 2)$, and a
+  stronger prior $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.02,
+  \mathrm{min.} = 2)$. The inset panel shows the same data but with a
+  log-transformed x axis. Note that the base and weaker priors are relatively
+  flat within the region of $\nu < 20$ that we are concerned with. }
+
+\label{fig:priors}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=0.44\textwidth]{../analysis/nu-coefs-2.pdf}
+\includegraphics[width=0.8\textwidth]{../analysis/gomp-comparison.pdf}
 
-\caption[Estimates of population dynamics heavy-tailedness for \nuCoefPopN\
-  populations of birds, mammals, insects, and fishes.]{Estimates of population dynamics heavy-tailedness for \nuCoefPopN\
-  populations of birds, mammals, insects, and fishes. Small values of $\nu$
-  approximately ($< 10$) suggest heavy-tailed black-swan dynamics; larger values of
-  $\nu$ suggest approximately normal-tailed dynamics. Vertical points and line
-  segments represent posterior medians and 50\% / 90\% credible intervals for
-  individual populations. Inset plots show probability that $\nu < 10$
-  (probability of heavy tails) for populations arranged by taxonomic order and
-  sorted by decreasing mean Pr($\nu < 10$). Taxonomic orders with three or
-  fewer populations in panel a are omitted for space. Red to yellow points
-  highlight populations with a high to moderately high probability of
-  heavy-tailed black-swan dynamics.}
+\caption{\textbf{Estimates of $\nu$ from alternative models
+  plotted against the base Gompertz model estimates of $\nu$.} Shown are
+  medians of the posterior (dots) and 50\% credible intervals (segments). The
+  diagonal line indicates a one-to-one relationship. Different colours
+  indicate various taxonomic classes. The grey-shaded regions indicate regions
+  of disagreement if $\nu = 10$ is taken as a threshold of heavy-tailed
+  dynamics. The Gompertz observation error model assumes a fixed standard
+  deviation of observation error of $0.2$ on a log scale.}
 
-\label{fig:nu-coefs}
+\label{fig:alt}
 \end{center}
 \end{figure}
+
 \clearpage
@@
 \begin{center}
-\includegraphics[width=0.45\textwidth]{../analysis/order-posteriors-covariates.pdf}
+\includegraphics[width=0.8\textwidth]{../analysis/gomp-prior-comparison.pdf}
 
-\caption[Posterior probability distributions from beta regression multilevel
-  models.]{Posterior probability distributions from beta regression multilevel
-  models. (a) Taxonomic-order-level posterior densities of Pr($\nu < 10$)
-  (approximately the probability of heavy tails) after accounting for
-  time-series length. Estimates are at the geometric mean of time series
-  length across all the data (approximately 27 time steps). Colour shading
-  refers to taxonomic class (yellow: fishes, green: insects, purple: birds,
-  and red: mammals). Dotted vertical line in panel a indicates the median
-  expected Pr($\nu < 10$) from the prior distribution. (b) Main effect
-  posterior densities for potential covariates of Pr($\nu < 10$). The beta
-  regression models were fit on a logit scale with hierarchical intercepts for
-  taxonomic class, taxonomic order, and species. All covariates were
-  standardized by subtracting their mean and dividing by twice their standard
-  deviation. In both panels, short vertical line segments within the density
-  polygons indicate median posterior estimates.}
+\caption{\textbf{Estimates of $\nu$ from Gompertz models with alternative priors on
+  $\nu$.} Shown are medians of the posterior (dots) and 50\% credible intervals
+  (segments). The diagonal line indicates a one-to-one relationship. Different
+  colours indicate various taxonomic classes. The grey-shaded regions indicate
+  regions of disagreement if $\nu = 10$ is taken as a threshold of
+  heavy-tailed dynamics. The base, weaker, and stronger priors on $\nu$ are
+  illustrated in Fig.~\ref{fig:priors}. In general, the estimates are nearly
+  identical in cases where the data are informative about low values of $\nu$.
+  When the data are less informative about low values of $\nu$, the prior can
+  slightly pull the estimates of $\nu$ towards higher or lower values.}
 
-\label{fig:posteriors}
+\label{fig:alt-priors}
 \end{center}
@@
 
+\clearpage
+
 \begin{figure}[htbp]
 \begin{center}
-\includegraphics[width=0.9\textwidth]{../analysis/correlates-p10.pdf}
+\includegraphics[width=0.8\textwidth]{../analysis/correlates-p10.pdf}
 
-\caption[Potential covariates of heavy-tailed population dynamics (indicated
-  by a high probability that $\nu < 10$).]{Potential covariates of heavy-tailed population dynamics (indicated
-  by a high probability that $\nu < 10$). Shown are (a--c) parameters from the
+\caption{\textbf{Potential covariates of heavy-tailed population dynamics (indicated
+  by a high probability that $\nu < 10$).} Shown are (a--c) parameters from the
   Gompertz heavy-tailed population model ($\sigma$, $\lambda$, $b$), (b)
@@
   probability the slope coefficient was different from zero
-  (Fig.~\ref{fig:posteriors}b).}
+  (Fig.~3b).}
 

### 2015-07-21 text/supp.tex
Add missing text in caption
Bankruptcy commit
Major restructuring of text

--- 2014-12-11 text/supp.tex
@@
-%!TEX root = anderson-etal-blackswan-timeseries.tex
+\documentclass[12pt]{article}
+\usepackage{geometry}
+\geometry{verbose, letterpaper, tmargin = 2.54cm, bmargin = 2.54cm,
+  lmargin = 2.54cm, rmargin = 2.54cm}
+\geometry{letterpaper}
+\usepackage{graphicx}
+\usepackage{amssymb}
+\usepackage{amsmath}
+\usepackage{amsfonts}
+\usepackage{setspace}
+\usepackage{booktabs}
+\usepackage{ragged2e}
+\usepackage{verbatim}
+\usepackage[sectionbib]{chapterbib}
+\usepackage{amsmath}
+\usepackage{longtable}
+\usepackage{multirow}
+\usepackage{array}
+\usepackage{tabu} % for spacing between rows in longtable
+\usepackage{titlesec}
+\usepackage{url}
+\usepackage{mathtools}
+\usepackage{cite}
+\usepackage{citesupernumber}
 
-\begin{centering}
-\LARGE
-Supporting Information\\[1.0em]
-\end{centering}
+\usepackage{lineno}
+\usepackage{xcolor}
+\renewcommand\linenumberfont{\normalfont\tiny\sffamily\color{gray}}
+\modulolinenumbers[2]
+
+% Linux Libertine:
+\usepackage{textcomp}
+\usepackage[sb]{libertine}
+\usepackage[varqu,varl]{inconsolata}% sans serif typewriter
+\usepackage[libertine,bigdelims,vvarbb]{newtxmath} % bb from STIX
+\usepackage[cal=boondoxo]{mathalfa} % mathcal
+\useosf % osf for text, not math
+\usepackage[supstfm=libertinesups,%
+  supscaled=1.2,%
+  raised=-.13em]{superiors}
+
+\mathchardef\mhyphen="2D % math hyphen
+
+\textheight 22.0cm
+
+\usepackage[round,sectionbib]{natbib}
+\bibpunct{(}{)}{;}{a}{}{;}
+
+\setlength\parskip{0.10in}
+\setlength\parindent{0in}
+
+% Fix line numbering and align environment
+% http://phaseportrait.blogspot.ca/2007/08/lineno-and-amsmath-compatibility.html
+\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
+  \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
+  \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
+  \renewenvironment{#1}%
+     {\linenomath\csname old#1\endcsname}%
+     {\csname oldend#1\endcsname\endlinenomath}}%
+\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
+  \patchAmsMathEnvironmentForLineno{#1}%
+  \patchAmsMathEnvironmentForLineno{#1*}}%
+\AtBeginDocument{%
+\patchBothAmsMathEnvironmentsForLineno{equation}%
+\patchBothAmsMathEnvironmentsForLineno{align}%
+\patchBothAmsMathEnvironmentsForLineno{flalign}%
+\patchBothAmsMathEnvironmentsForLineno{alignat}%
+\patchBothAmsMathEnvironmentsForLineno{gather}%
+\patchBothAmsMathEnvironmentsForLineno{multline}%
+}
+
+% remove numbers in front of sections:
+\makeatletter
+\renewcommand\@seccntformat[1]{}
+\makeatother
+
+\title{Supporting Information:\\Evidence for black-swan events in animal
+  populations}
+
+\author{
+Sean C. Anderson$^{1,2}$ \and
+Trevor A. Branch$^2$ \and
+Andrew B. Cooper$^3$ \and
+Nicholas K. Dulvy$^1$
+}
+\date{}
+
+\begin{document}
+
+\input{../analysis/values} % R output
+
+\maketitle
+
+\textsuperscript{1}Earth to Ocean Research Group, Department of Biological
+Sciences, Simon Fraser University, Burnaby BC, V5A 1S6, Canada
+
+\textsuperscript{2}School of Resource and Environmental Management, Simon
+Fraser University, Burnaby, BC, V5A 1S6, Canada
+
+\textsuperscript{3}School of Aquatic and Fishery Sciences, University of
+Washington, Box 355020, Seattle, WA 98195, USA
+
+\linenumbers
+\onehalfspacing
+
+%\begin{centering}
+%\LARGE
+%Supporting Information\\[1.0em]
+%\end{centering}
 
@@
 
-\section{Details on the heavy-tailed Gompertz probability model}
-
-For the Gompertz model, our weakly-informative priors (Fig.~\ref{fig:priors})
-were:
-\begin{align*}
-b &\sim \mathrm{Uniform}(-1, 2)\\ \lambda &\sim \mathrm{Normal}(0, 10^2)\\
-\sigma &\sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)\\ \nu &\sim
-\mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2).
-\end{align*}
-Our prior on $b$ was uninformative between values of $-1$ and $2$. We would
-not expect values of $b$ with levels of density dependence as low as $-1$
-(very strong inverse density dependence), nor would we generally expect values
-above $1$. We allowed values of $b$ above $1$ to allow for non-stationary time
-series of growth rates. The estimates of $b$ were well within these bounds.
-Our prior on $\lambda$ was very weakly informative within the range of
-expected values for population growth and is similar to the default priors
-suggested by \citet{gelman2008d} for intercepts of regression models. Our
-Half-Cauchy prior on $\sigma$ follows \citet{gelman2006c} and
-\citet{gelman2008d} and the specific scale parameter of $2.5$ is based on our
-expected range of the value in nature from previous studies
-\citep[e.g.][]{connors2014}. In our testing of a subsample of populations, our
-parameter estimates were not qualitatively changed by switching to an
-uninformative uniform prior on $\sigma$, but the weakly informative
-Half-Cauchy prior substantially sped up chain convergence.
-
-Our prior on $\nu$ was based on \citet{fernandez1998}. They chose an
-exponential rate parameter of $0.1$. We chose a less informative rate
-parameter of $0.01$ and truncated the distribution at $2$, since at $\nu < 2$
-the variance of the t distribution is undefined. This prior gives only a
-$7.7$\% probability that $\nu < 10$ but constrains the sampling sufficiently
-to avoid wandering off towards infinity---above approximately $\nu = 20$ the t
-distribution is so similar to the normal distribution
-(Fig.~\ref{fig:didactic}) that time series of the length considered here are
-unlikely to be informative about the precise value of $\nu$. In the scenario
-where the data are uninformative about heavy tails
-(e.g.~Fig.~\ref{fig:didactic}e,~h), the posterior will approximately match the
-prior (prior median $= 71$, mean $= 102$) and the metrics used in our paper
-(e.g.~Pr$(\nu < 10) > 0.5$) are unlikely to flag the population as heavy
-tailed.
-
-We fit our models with Stan 2.4.0 \citep{stan-manual2014}, and R 3.1.1
-\citep{r2014}. We began with four chains and $2000$ iterations, discarding the
-first $1000$ as warm up (i.e.~4000 total samples). If $\hat{R}$ (the potential
-scale reduction factor---a measure of chain convergence) was greater than
-$1.05$ for any parameter or the minimum effective sample size,
-$n_\mathrm{eff}$, (a measure of the effective number of uncorrelated samples)
-for any parameter was less than $200$, we doubled both the total iterations
-and warm up period and sampled from the model again. These thresholds are in
-excess of the minimums recommended by \citet{gelman2006a} of $\hat{R} < 1.1$
-and effective sample size $> 100$ for reliable point estimates and confidence
-intervals. In the majority of cases our minimum thresholds were greatly
-exceeded. We continued this procedure up to $8000$ iterations ($16000$ total
-samples) by which all chains were deemed to have sufficiently converged. These
-chain lengths may seem low to those familiar with software such as WinBUGS or
-JAGS, but the No-U-Turn Hamiltonian Markov chain Monte Carlo Sampler in Stan
-generally requires far fewer iterations to obtain equivalent effective sample
-sizes \citep{stan-manual2014}.
-
-\section{Alternative priors}
-
-To test if the prior on $\nu$ influenced our estimate of black-swan dynamics,
-we refit our models with weaker and stronger priors. Our base model used a
-prior on $\nu$ of Truncated-Exponential(0.02, min.\ = 2). For a weaker prior
-we used Truncated-Exponential(0.005, min.\ = 2) and for a stronger prior we
-used Truncated-Exponential(0.02, min.\ = 2) (Fig.~\ref{fig:priors}). Note that
-the base and weaker priors are relatively flat within the region of $\nu <
-20$, which is the region we are mostly concerned about when categorizing populations
-as heavy- or thin-tailed.
-
-Our results show that these weaker and stronger priors would have little
-influence on our conclusions about heavy-tailed dynamics
-(Fig.~\ref{fig:alt-priors}). When the data are informative about tail
-behaviour (i.e.\ when there is strong evidence of low $\nu$ values,
-upper-right of Fig.~\ref{fig:alt-priors}), the prior has little impact on the
-estimate of $\nu$. When the data are less informative about $\nu$ (i.e.\ when
-there are no or few tail events and time series are short or noisy), the prior
-can pull the estimate of $\nu$ towards larger or smaller values
-(Fig.~\ref{fig:alt-priors}). The vast majority of the populations with Pr$(\nu
-< 10)$ in the base prior were not altered qualitatively by this range of prior
-strength.
-
 \section{Alternative population models}
@@
 autocorrelation in the residuals, assumed no density dependence, allowed for
-observation error, or assumed a Ricker-logistic functional form. The range of
-percentages of black swans by taxonomic class cited in the abstract and
-results are based on lower and upper limits across our main Gompertz model and
-these four alternative models.
+observation error, or assumed a Ricker-logistic functional form.
 
@@
 
+\section{Implementation of skewed Student t distribution in Stan}
+
+The probability density for the skewed Student t distribution described in the
+methods section translates to the log probability:
+
+\[
+\mathrm{Skewed\,Student\mhyphen t}(x) =
+\begin{dcases}
+  \log (2 \gamma) - \log(\gamma^2 + 1) +
+  \log \mathrm{Student\mhyphen t}(x \cdot \gamma, \nu, \mu \cdot \gamma, \sigma),& \text{for } x < 0\\
+  \log (2 \gamma) - \log(\gamma^2 + 1) +
+  \log \mathrm{Student\mhyphen t}(x / \gamma, \nu, \mu / \gamma, \sigma),& \text{for } x \ge 0
+\end{dcases}
+\]
+
+Since Stan only needs the probability density up to an additive constant, we
+can replace $\log(2 \gamma)$ in the above for $\log(\gamma)$. The final log
+probability density for the skewed Student's t distribution implemented in
+Stan\footnote{See the Stan users mailing list:
+\url{https://groups.google.com/d/msg/stan-users/jeZQnQRUsDs/BeqhicWm1GEJ}}
+looks like the following:
+
+\begin{verbatim}
+real skew_student_t_log(real y, real nu, real mu, real sigma, real skew) {
+  real lp;
+  lp <- log(skew) - log1p(square(skew));
+  if (y < mu)
+    return lp + student_t_log(y * skew, nu, mu * skew, sigma);
+  else
+    return lp + student_t_log(y / skew, nu, mu / skew, sigma);
+}
+\end{verbatim}
+
 \section{Simulation testing the model}
@@
 
+In the following image, the red lines indicate the true population value. Dots
+represent the median estimate the lines represent the 80\% credible interval.
+When a small number of samples are drawn there may not be samples sufficiently
+far into the tails to recapture the true $\nu$ value; however, heavy tails are
+still distinguished from normal tails in most cases, even with only 25 or 50
+samples:
+
+\begin{center}
+\includegraphics[width=0.8\textwidth]{../analysis/t-dist-sampling-sim-prior-exp0point01.pdf}
+\includegraphics[width=0.8\textwidth]{../analysis/t-dist-sampling-sim-sigma-prior-exp0point01.pdf}
+\end{center}
+
 We found that we could consistently and precisely recover median posterior
 estimates of $\nu$ near the true value of $\nu$ with large samples ($\ge 200$)
-(Fig.~\ref{fig:sim-nu} upper panels). At smaller samples we could still
-usually qualitatively distinguish heavy from not-heavy tails, but the model
-tended to underestimate how heavy the tails were. At the same time, at smaller
-sample sizes, the model tended to overestimate how large the scale parameter
-was (Fig.~\ref{fig:sim-nu} lower panels).
+(upper panels). At smaller samples we could still usually qualitatively
+distinguish heavy from not-heavy tails, but the model tended to underestimate
+how heavy the tails were. At the same time, at smaller sample sizes, the model
+tended to overestimate how large the scale parameter was (lower panels).
 
@@
 
-%(Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}, red and
-%green symbols in the top rows). Likewise, the other Gompertz parameters were
-%estimated without any systematic bias (Figs~\ref{fig:sim-gompertz} and
-%\ref{fig:sim-gompertz-boxplots}, red and green symbols). , overestimate the
-%magnitude of process noise, somewhat overestimate $\lambda$, and overestimate
-%density dependence (blue symbols in Figs~\ref{fig:sim-gompertz} and
-%\ref{fig:sim-gompertz-boxplots}). The overestimation of density dependence
-%with observation error is a known phenomenon \citep{knape2012}. Fitting a
-%model with correctly specified observation error made marginal improvements
-%to model bias (purple symbols in Figs~\ref{fig:sim-gompertz} and
-%\ref{fig:sim-gompertz-boxplots}).
-
 When converting the posterior distributions of $\nu$ into Pr($\nu < 10$), the
@@
 
-\section{Modelling covariates of heavy-tailed dynamics}
+In the following image, upper panels show the distribution of median
+$\widehat{\nu}$ across 20 simulation runs. Lower panels show the distribution
+of Pr($\nu < 10$) across 20 simulation runs. We ran the simulations across
+three population (``true'') $\nu$ values (3, 5, and $1\cdot 10^9$, i.e.\
+approximately normal) and four scenarios: (1) 100 time steps and no
+observation error, (2) 50 time steps and no observation error, (3) 50 time
+steps and observation error drawn from $\mathrm{Normal} (0, 0.2^2)$ but
+ignored, and (4) 50 time steps with observation error in which the quantity of
+observation error was assumed known. Within each scenario the dots represent
+stochastic draws from the true population distributions combined with model
+fits. Underlayed boxplots show the median, interquartile range, and $1.5$
+times the interquartile range:
 
-We fit a multilevel beta regression model to the predicted probability of
-heavy tails, Pr($\nu < 10$), to investigate potential covariates of
-heavy-tailed dynamics. The beta distribution is useful when response data
-range on a continuous scale between zero and one \citep{ferrari2004}. We used
-a logit link function as is typically used in logistic regression. The model
-was as follows:
-
-\begin{align*}
-\mathrm{Pr}(\nu_i < 10) &\sim \mathrm{Beta}(A_i, B_i)\\
-\mu_i &= \mathrm{logit}^{-1}(\alpha
-  + \alpha^\mathrm{class}_{j[i]}
-  + \alpha^\mathrm{order}_{k[i]}
-  + \alpha^\mathrm{species}_{l[i]}
-  + X_i \beta),
-  \: \text{for } i = 1, \dots, 617\\
-A_i &= \phi_\mathrm{disp} \mu_i\\
-B_i &= \phi_\mathrm{disp} (1 - \mu_i)\\
-\alpha^\mathrm{class}_j &\sim
-  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{class}}),
-  \: \text{for } j = 1, \dots, 6\\
-\alpha^\mathrm{order}_k &\sim
-  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{order}}),
-  \: \text{for } k = 1, \dots, 38\\
-\alpha^\mathrm{species}_l &\sim
-  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{species}}),
-  \: \text{for } l = 1, \dots, 301,
-\end{align*}
-where $A$ and $B$ represent the beta distribution shape parameters; $\mu_i$
-represents the predicted value for population $i$, class $j$, order $k$, and
-species $l$; $\phi_\mathrm{disp}$ represents the dispersion parameter; and
-$X_i$ represents a vector of predictors (such as lifespan) for population $i$
-with associated $\beta$ coefficients. The intercepts are allowed to vary from
-the overall intercept $\alpha$ by taxonomic class ($\alpha^\mathrm{class}_j$),
-taxonomic order ($\alpha^\mathrm{order}_k$), and species
-($\alpha^\mathrm{species}_l$) with standard deviations $\sigma_{\alpha \;
-  \mathrm{class}}$, $\sigma_{\alpha \; \mathrm{order}}$, and $\sigma_{\alpha
-  \; \mathrm{species}}$. Where possible, we also allowed for error
-distributions around the predictors by incorporating the standard deviation of
-the posterior samples for the Gompertz parameters $\lambda$, $b$, and $\log
-\sigma$ around the mean point value as normal distributions (not shown in the
-above equation).
-
-We log transformed $\sigma$, time-series length, and lifespan to match the way
-they are visually represented in Fig.~\ref{fig:correlates} and to make the
-relationship approximately linear on the logit-transformed response scale. All
-input variables were standardized by subtracting their mean and dividing by
-two standard deviations to make their coefficients comparable in magnitude
-\citep{gelman2008c}. We excluded body length as a covariate because it was
-highly correlated with lifespan, and lifespan exhibited more overlap across
-taxonomy than body length. Lifespan is also more directly related to time and
-potential mechanisms driving black-swan dynamics.
-
-We incorporated weakly informative priors into our model: $\mathrm{Cauchy}(0,
-10)$ on the global intercept $\alpha$, $\mathrm{Half\mhyphen Cauchy}(0, 2.5)$
-on all standard deviation parameters, $\mathrm{Half\mhyphen Cauchy}(0, 10)$ on
-the dispersion parameter $\phi_\mathrm{disp}$, and $\mathrm{Cauchy}(0, 2.5)$
-on all other parameters \citep{gelman2006c, gelman2008d}. Compared to normal
-priors, the Cauchy priors concentrate more probability density around expected
-parameter values while allowing for a higher probability density far into the
-tails, thereby allowing the data to dominate the posterior more strongly if it
-disagrees with the prior. Our conclusions were not qualitatively changed by
-using uniform priors. We fit our models with 5000 total iterations per chain,
-2500 warm-up iterations, four chains, and discarding every second sample to
-save memory. We checked for chain convergence visually and with the same
-criteria as before ($\widehat{R} < 1.05$ and $n_\mathrm{eff} >200$ for all
-parameters).
-
-To derive taxonomic-order-level estimates of the probability of heavy tails
-accounting for time-series length (Fig.~\ref{fig:posteriors}a), we fit a
-separate multilevel model with the same structure but with only $\log$
-time-series length as a predictor. (In this case, we did not want to control
-for intrinsic population characteristics such as density dependence.) Since
-our predictors were centered by subtracting their mean value, we obtained
-order-level estimates of the probability of heavy tails at mean log
-time-series length by adding the posteriors for $\alpha$,
-$\alpha^\mathrm{class}_j$, and $\alpha^\mathrm{order}_k$.
+%\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=\textwidth]{../analysis/sim-gompertz-median-dist.pdf}
+\includegraphics[width=\textwidth]{../analysis/sim-gompertz-p10.pdf}
+%\label{fig:sim-prob}
+\end{center}
+%\end{figure}
 
@@
 
-Many of the silhouette images used in Figs~\ref{fig:nu-coefs},
-\ref{fig:correlates} and \ref{fig:posteriors} were obtained from
+Many of the silhouette images used in Figs 2, and 3 were obtained from
 \texttt{phylopic.org} under Creative Commons licenses. We vectorized the
-salmon in Fig.~\ref{fig:nu-coefs} and Fig.~\ref{fig:correlates} ourselves. The
-bird in these figures was obtained from \texttt{phylopic.org} under a Creative
-Commons Attribution 3.0 Unported license with credit to Jean-Raphaël
-Guillaumin {[}photography{]} and T. Michael Keesey {[}vectorization{]}). The
-silhouettes in Fig.~\ref{fig:posteriors} were obtained from the following
-sources (metadata obtained with the help of the rphylopic R package,
+salmon in Fig.~2. The bird in these figures was obtained
+from \texttt{phylopic.org} under a Creative Commons Attribution 3.0 Unported
+license with credit to Jean-Raphaël Guillaumin {[}photography{]} and T.
+Michael Keesey {[}vectorization{]}). The silhouettes in
+Fig.~3 were obtained from the following sources (metadata
+obtained with the help of the rphylopic R package,
 \url{https://github.com/sckott/rphylopic}):
@@
 
-\renewcommand{\baselinestretch}{\tighttextstretch}
-\normalsize
 \bibliographystyle{apalike}
-\bibliography{/Users/seananderson/Dropbox/tex/jshort,/Users/seananderson/Dropbox/tex/ref3}
-\renewcommand{\baselinestretch}{\textstretch}
-\normalsize
-
+\bibliography{/Users/seananderson/Dropbox/tex/jshort,supp}
+%
 % ------------------------------
@@
 
-\clearpage
-\renewcommand{\thetable}{S\arabic{table}}
-\setcounter{table}{0}
-
-\begin{table}
-\begin{footnotesize}
-
-\caption[Summary statistics for the filtered Global Population Dynamics
-  Database time series arranged by taxonomic class.]{Summary statistics for the filtered Global Population Dynamics
-  Database time series arranged by taxonomic class. Columns are: number of
-  populations, number of taxonomic orders, numbers of species, median time
-  series length, total number of interpolated time steps, total number of
-  substituted zeros, and total number of time steps.}
-
-\smallskip
-\begin{tabular}{lrrrrrrrr}
-\toprule
-\input{../analysis/stat-table.tex}
-\label{tab:stats}
-\end{tabular}
-\end{footnotesize}
-\end{table}
-
-\clearpage
-
-\LTcapwidth=\textwidth
-\bibpunct{}{}{;}{a}{}{;}
-
-%\singlespacing
-\begin{footnotesize}
-\begin{longtable}{>{\RaggedRight}m{1.4cm}>{\RaggedRight}p{3.3cm}>{\RaggedRight}p{0.7cm}>{\RaggedRight}p{1.5cm}>{\RaggedRight}p{3.0cm}>{\RaggedRight}p{1.4cm}>{\RaggedRight}p{1.25cm}}
-
-\caption[All populations with Pr$(\nu < 10) > 0.5$ in the base heavy-tailed
-  Gompertz population dynamics model.]{All populations with Pr$(\nu < 10) > 0.5$ in the base heavy-tailed
-  Gompertz population dynamics model. Shown are the log abundance time series,
-  population descriptions, Global Population Dynamics Database Main IDs,
-  citation for the data source or separate verification literature, a
-  description of the cause of the black swan events (if known), the
-  probability of heavy tails as calculated by our model, and median estimate
-  of $\nu$ from our model with 90\% quantile credible intervals indicated in
-  parentheses. Red dots on the time series indicate downward black-swan events
-  and blue values indicate upward black-swan events that have a $10^{-4}$ probability or less of occurring if the population dynamics were
-  explained by a Gompertz model with normally distributed process noise with a
-  standard deviation equal to the scale parameter in the fitted t
-  distribution.}\\
-
-\toprule
-\input{../analysis/cause-table.tex}
-\label{tab:causes-supp}
-\end{longtable}
-\end{footnotesize}
-%\onehalfspacing
+%\clearpage
+%\renewcommand{\thetable}{S\arabic{table}}
+%\setcounter{table}{0}
 
@@
 
-\renewcommand{\thefigure}{S\arabic{figure}}
-\setcounter{figure}{0}
+%\renewcommand{\thefigure}{S\arabic{figure}}
+%\setcounter{figure}{0}
 
-\begin{centering}
-\clearpage
-\includegraphics[width=\textwidth]{../analysis/all-clean-ts-mammals.pdf}\\
-Figure~\ref{fig:all-ts} (mammals) continued on next page \ldots
+% \begin{centering}
+% \clearpage
+% \includegraphics[width=\textwidth]{../analysis/all-clean-ts-mammals.pdf}\\
+% Figure~\ref{fig:all-ts} (mammals) continued on next page \ldots
+%
+% \clearpage
+% \includegraphics[width=\textwidth]{../analysis/all-clean-ts-birds.pdf}\\
+% Figure~\ref{fig:all-ts} (birds) continued on next page \ldots
+%
+% \clearpage
+% \includegraphics[width=\textwidth]{../analysis/all-clean-ts-insects.pdf}\\
+% Figure~\ref{fig:all-ts} (insects) continued on next page \ldots
+%
+% \end{centering}
 
-\clearpage
-\includegraphics[width=\textwidth]{../analysis/all-clean-ts-birds.pdf}\\
-Figure~\ref{fig:all-ts} (birds) continued on next page \ldots
+% \begin{figure}[htbp]
+% \begin{center}
+% \includegraphics[width=\textwidth]{../analysis/all-clean-ts-fishes-others.pdf}
+%
+% \caption[All filtered time series used in our analysis.]{(fishes, crustaceans,
+%   gastropods, sharks). All filtered time series used in our analysis. The
+%   abundances are shown on a log10 vertical axis. Throughout this figure, red
+%   dots indicate values that were interpolated and blue dots indicate values
+%   that were recorded as zero but were set to the next lowest observed
+%   abundance. Numbers before each species name are the GPDD Main ID numbers.}
+%
+% \label{fig:all-ts}
+% \end{center}
+% \end{figure}
 
-\clearpage
-\includegraphics[width=\textwidth]{../analysis/all-clean-ts-insects.pdf}\\
-Figure~\ref{fig:all-ts} (insects) continued on next page \ldots
-
-\end{centering}
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=\textwidth]{../analysis/all-clean-ts-fishes-others.pdf}
-
-\caption[All filtered time series used in our analysis.]{(fishes, crustaceans,
-  gastropods, sharks). All filtered time series used in our analysis. The
-  abundances are shown on a log10 vertical axis. Throughout this figure, red
-  dots indicate values that were interpolated and blue dots indicate values
-  that were recorded as zero but were set to the next lowest observed
-  abundance. Numbers before each species name are the GPDD Main ID numbers.}
-
-\label{fig:all-ts}
-\end{center}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=0.8\textwidth]{../analysis/priors-gomp-base.pdf}
-
-\caption[Probability density of the Bayesian priors for the Gompertz models.]{
-  Probability density of the Bayesian priors for the Gompertz models. From
-  left to right and then top to bottom: (1) per capita growth rate at
-  $\log$(abundance) = $0$: $\lambda \sim \mathrm{Normal}(0, 10^2)$; (2) scale
-  parameter of t-distribution process noise: $\sigma \sim \mathrm{Half\mhyphen
-    Cauchy} (0, 2.5)$; (3) t-distribution degrees of freedom parameter: $\nu
-  \sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2)$; (4)
-  AR1 correlation coefficient of residuals: $\phi \sim \mathrm{Truncated
-    \mhyphen Normal}(0, 1, \mathrm{min.} = -1, \mathrm{max.} = 1)$. Not shown
-  is $b$, the density dependence parameter: $b \sim \mathrm{Uniform}(-1, 2)$.
-  The $\nu$ panel also shows two alternative priors: a weaker prior $\nu \sim
-  \mathrm{Truncated\mhyphen Exponential}(0.005, \mathrm{min.} = 2)$, and a
-  stronger prior $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.02,
-  \mathrm{min.} = 2)$. The inset panel shows the same data but with a
-  log-transformed x axis. Note that the base and weaker priors are relatively
-  flat within the region of $\nu < 20$ that we are concerned with. }
-
-\label{fig:priors}
-\end{center}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=\textwidth]{../analysis/gomp-comparison.pdf}
-
-\caption[Estimates of $\nu$ from alternative models plotted against the base
-Gompertz model estimates of $\nu$.]{Estimates of $\nu$ from alternative models
-  plotted against the base Gompertz model estimates of $\nu$. Shown are
-  medians of the posterior (dots) and 50\% credible intervals (segments). The
-  diagonal line indicates a one-to-one relationship. Different colours
-  indicate various taxonomic classes. The grey-shaded regions indicate regions
-  of disagreement if $\nu = 10$ is taken as a threshold of heavy-tailed
-  dynamics. The Gompertz observation error model assumes a fixed standard
-  deviation of observation error of $0.2$ on a log scale.}
-
-\label{fig:alt}
-\end{center}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=\textwidth]{../analysis/gomp-prior-comparison.pdf}
-
-\caption[Estimates of $\nu$ from Gompertz models with alternative priors on
-$\nu$.]{Estimates of $\nu$ from Gompertz models with alternative priors on
-  $\nu$. Shown are medians of the posterior (dots) and 50\% credible intervals
-  (segments). The diagonal line indicates a one-to-one relationship. Different
-  colours indicate various taxonomic classes. The grey-shaded regions indicate
-  regions of disagreement if $\nu = 10$ is taken as a threshold of
-  heavy-tailed dynamics. The base, weaker, and stronger priors on $\nu$ are
-  illustrated in Fig.~\ref{fig:priors}. In general, the estimates are nearly
-  identical in cases where the data are informative about low values of $\nu$.
-  When the data are less informative about low values of $\nu$, the prior can
-  slightly pull the estimates of $\nu$ towards higher or lower values.}
-
-\label{fig:alt-priors}
-\end{center}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=0.8\textwidth]{../analysis/t-dist-sampling-sim-prior-exp0point01.pdf}
-\includegraphics[width=0.8\textwidth]{../analysis/t-dist-sampling-sim-sigma-prior-exp0point01.pdf}
-
-\caption[Testing the ability to estimate $\nu$  and the scale parameter of the
-process deviations for a given number of samples drawn from a distribution
-with a given true $\nu$ value.]{Testing the ability to estimate $\nu$ (top
-  panels) and the scale parameter of the process deviations (bottom panels)
-  for a given number of samples (columns) drawn from a distribution with a
-  given true $\nu$ value (rows). The red lines indicate the true population
-  value. When a small number of samples are drawn there may not be samples
-  sufficiently far into the tails to recapture the true $\nu$ value; however,
-  heavy tails are still distinguished from normal tails in most cases, even
-  with only 25 or 50 samples.}
-
-\label{fig:sim-nu}
-\end{center}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=\textwidth]{../analysis/sim-gompertz-median-dist.pdf}
-\includegraphics[width=\textwidth]{../analysis/sim-gompertz-p10.pdf}
-
-\caption[Simulation testing the Gompertz estimation model when the process
-deviation draws were chosen so that $\nu$ could be estimated close to the true
-value outside the full population model (``effective $\nu$'' within a CV of
-0.2 of specified $\nu$).]{Simulation testing the Gompertz estimation model
-  when the process deviation draws were chosen so that $\nu$ could be
-  estimated close to the true value outside the full population model
-  (``effective $\nu$'' within a CV of 0.2 of specified $\nu$). Upper panels
-  show the distribution of median $\widehat{\nu}$ across 20 simulation runs.
-  Lower panels show the distribution of Pr($\nu < 10$) across 20 simulation
-  runs. We ran the simulations across three population (``true'') $\nu$ values
-  (3, 5, and $1\cdot 10^9$, i.e.\ approximately normal) and four scenarios:
-  (1) 100 time steps and no observation error, (2) 50 time steps and no
-  observation error, (3) 50 time steps and observation error drawn from
-  $\mathrm{Normal} (0,
-  0.2^2)$ but ignored, and (4) 50 time steps with observation error in which
-    the quantity of observation error was assumed known. Within each scenario
-    the dots represent stochastic draws from the true population distributions
-    combined with model fits. Underlayed boxplots show the median,
-    interquartile range, and $1.5$ times the interquartile range. }
-
-\label{fig:sim-prob}
-\end{center}
-\end{figure}
+%\clearpage
 
@@
 residuals and a specified level of observation error. The specific code for
-used for the various models in our analysis is available at
+the various models in our analysis is available at
 \url{https://github.com/seananderson/heavy-tails}.
@@
 
-\renewcommand{\baselinestretch}{\tighttextstretch}
-\normalsize
 \begin{footnotesize}
+%\renewcommand{\baselinestretch}{1.11}
 \noindent
@@
 }
+%\renewcommand{\baselinestretch}{\textstretch}
 \end{footnotesize}
-\renewcommand{\baselinestretch}{\textstretch}
 \normalsize
 %\end{spacing}
+
+\end{document}

### 2015-07-21 analysis/2-run-models-base-skew.R

@@ -10,7 +10,7 @@ id <- "gomp-base-skew"
 stan_pars <- c("lambda", "sigma_proc", "nu", "b", "log_skew")
 
 fit_gpdd_model(gpdd_dat = gpdd, model = model, sub_folder = id,
-  pars = stan_pars)
+  pars = stan_pars, iterations = 20000, max_iterations = 20000, warmup = 10000)
 
 out <- plyr::ldply(unique(gpdd$main_id), extract_model,
   sub_folder = id, get_phi = FALSE, type = "gompertz",

### 2015-07-21 analysis/2-run-models-normal.R

@@ -10,7 +10,7 @@ id <- "gomp-base-normal"
 stan_pars <- c("lambda", "sigma_proc", "b")
 
 fit_gpdd_model(gpdd_dat = gpdd, model = model, sub_folder = id,
-  pars = stan_pars)
+  pars = stan_pars, iterations = 20000, max_iterations = 20000, warmup = 10000)
 
 out <- plyr::ldply(unique(gpdd$main_id), extract_model,
   sub_folder = id, get_phi = FALSE, type = "gompertz", get_nu = FALSE)

### 2015-07-21 analysis/5-shape-data.R

@@ -6,6 +6,7 @@ library("dplyr")
 
 gpdd <- readRDS("gpdd-clean.rds")
 
+gomp_hat_normal <- readRDS("gomp-base-normal-hat.rds")
 gomp_hat_skew <- readRDS("gomp-base-skew-hat.rds")
 gomp_hat_base <- readRDS("gomp-base-hat.rds")
 gomp_hat_logistic <- readRDS("logistic-hat.rds")
@@ -19,6 +20,7 @@ gomp_hat_stronger <- readRDS("gomp-base-stronger-hat.rds")
 # but, some of these have already been run in my locally cached files
 # therefore, I will remove all GPDD main_ids not in `gpdd`:
 clean_main_ids <- unique(gpdd$main_id)
+gomp_hat_normal <- filter(gomp_hat_normal, main_id %in% clean_main_ids)
 gomp_hat_skew <- filter(gomp_hat_skew, main_id %in% clean_main_ids)
 gomp_hat_base <- filter(gomp_hat_base, main_id %in% clean_main_ids)
 gomp_hat_logistic <- filter(gomp_hat_logistic, main_id %in% clean_main_ids)
@@ -52,6 +54,7 @@ lookup <- gpdd[,c("main_id", "common_name", "taxonomic_class",
   "taxon_name", "taxonomic_order", "taxonomic_family", "dataset_length")]
 lookup <- lookup[!duplicated(lookup), ]
 
+gomp_hat_normal <- inner_join(gomp_hat_normal, lookup, by = "main_id")
 gomp_hat_base <- inner_join(gomp_hat_base, lookup, by = "main_id")
 gomp_hat_ar1 <- inner_join(gomp_hat_ar1, lookup, by = "main_id")
 gomp_hat_obs_0.2 <- inner_join(gomp_hat_obs_0.2, lookup, by = "main_id")

### 2015-07-21 analysis/8-sim-extinction-risk.R

@@ -16,6 +16,7 @@
 # We'll write a little C++ function to simulate from a Gompertz model. We'll do
 # this in C++ just to make everything fast:
 
+library("dplyr")
 library("Rcpp")
 cppFunction("
   NumericVector gompertz(double n, double lambda, double b,
@@ -35,21 +36,22 @@ cppFunction("
 # starting at the state the time series ended. Then we'll continue on for some
 # given number of generations.
 
-sim_gomp_extinction <- function(abundance_ts, lambda = 1, b = 0.5, N = 10L,
+sim_gomp_extinction <- function(y1, abundance_ts = NULL, lambda = 1, b = 0.5, N = 10L,
   nu = 3, log_skew = 1, sigma_proc = 0.2, bootstrap_residuals = FALSE,
   model = c("heavy-skewd", "normal")) {
 
-  log_abundance_ts <- log(abundance_ts)
-  y1 <- log_abundance_ts[length(abundance_ts)]
+  y1 <- log(y1)
 
   if (bootstrap_residuals) {
+    log_abundance_ts <- log(abundance_ts)
     res <- log_abundance_ts[-1] - (lambda + b * log_abundance_ts[-length(abundance_ts)])
     # bootstrap the process residuals:
     proc_error <- sample(res, size = N, replace = TRUE)
   } else {
     if (model[1] == "heavy-skewed") {
       proc_error <- skewt::rskt(N, df = nu, gamma = exp(log_skew)) * sigma_proc
-    } else {
+    }
+    if (model[1] == "normal") {
       proc_error <- rnorm(N, sd = sigma_proc)
     }
   }
@@ -67,57 +69,80 @@ gpdd <- readRDS("gpdd-clean.rds") # to get the last abundance
 skew_samples <- readRDS("skew_samples.rds")
 normal_samples <- readRDS("normal_samples.rds")
 
-out_skew <- plyr::ddply(skew_samples, "main_id", function(x) {
-  id_i <- x$main_id[1L]
-  message(paste("main_id:", id_i))
-  this_dat <- dplyr::filter(gpdd, main_id == id_i)
-  out_inner <- plyr::ddply(x, c("main_id", "sample_ids"), function(xx) {
-    sim_gomp_extinction(abundance_ts = this_dat$population_untransformed,
-      lambda = xx$lambda[1L], b = xx$b[1L], N = 10L, nu = xx$nu[1L],
[…]

### 2015-07-21 analysis/8.1-plot-skewness.R

@@ -1,0 +1,329 @@
+## plot various skewness figures
+
+# look at the skew-t distribution:
+library(skewt)
+library(ggplot2)
+params <- expand.grid(x = seq(-7, 7, length.out = 100), df = c(2, 5, 1e6),
+  gamma = round(exp(c(-0.6, -0.3, 0, 0.3, 0.6)), 1))
+o <- plyr::mdply(params, dskt)
+
+o$nu_label <- paste("nu =", o$df)
+o$skew_label <- paste("skew =", o$gamma)
+
+o$nu_label <- reorder(o$nu_label, o$df)
+
+plain_theme <- theme_bw() + theme(
+  panel.grid.major = element_blank(),
+  panel.grid.minor = element_blank(),
+  panel.background = element_blank(),
+  strip.background = element_blank())
+
+p <- ggplot(o, aes(x, V1, colour = nu_label)) + geom_line() +
+  facet_wrap(~skew_label, nrow = 1) + ylab("Probability density") +
+  xlab("x") + plain_theme + labs(colour=expression(nu))
+  print(p)
+ggsave("skew-t-illustration.pdf", width = 9, height = 2)
+
+source("5-shape-data.R")
+
+gomp_hat_skew <- gomp_hat_skew %>% group_by(taxonomic_class) %>%
+  arrange(log_skew_50) %>%
+  mutate(plot_id = 1:n()) %>% as.data.frame()
+
+p <- ggplot(gomp_hat_skew, aes(1/nu_50, log_skew_50)) +
+  geom_segment(aes(y = log_skew_25, yend = log_skew_75, x = 1/nu_50, xend = 1/nu_50),
+    alpha = 0.08, lwd = 0.3) +
+  #geom_segment(aes(y = log_skew_5, yend = log_skew_95, x = nu_50, xend = nu_50),
+    #alpha = 0.06, lwd = 0.2) +
+  geom_segment(aes(y = log_skew_50, yend = log_skew_50, x = 1/nu_25, xend = 1/nu_75),
+    alpha = 0.09, lwd = 0.3) +
+  geom_point(alpha = 0.3, pch = 21) +
+  theme_bw() + geom_hline(yintercept = 0, lty = 2) +
+  scale_x_continuous(breaks = 1/c(100, 10, 5, 3, 2), labels = c(100, 10, 5, 3, 2)) +
[…]

### 2015-07-22 analysis/cause-table.tex
Shrink sparkline table to one page

@@
-% latex table generated in R 3.1.2 by xtable 1.7-4 package
-Time series & Population & ID & Citation & Description & Pr($\nu < 10$) & $\widehat{\nu}$ \\
+% latex table generated in R 3.2.0 by xtable 1.7-4 package
+Time series & Population & ID & Ref & Description & Pr($\nu < 10$) \\ 
   \midrule
-\includegraphics[width=1.7cm]{../analysis/sparks/6528.pdf} & Shag, \textit{Phalacrocorax aristotelis}, Farne Islands, Northumberland & 6528 & \cite{potts1980} & Red tide event combined with low productivity due to overcrowding & 1.00 & 2 (2--4) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/7115.pdf} & South African fur seal, \textit{Arctocephalus pusillus}, South Africa & 7115 & \cite{shaughnessy1982} & Harvesting and predation changes & 1.00 & 2 (2--4) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/10128.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Scotland - un-named area & 10128 & \cite{potts1984} & Environment- and parisite-caused cycles & 1.00 & 3 (2--4) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/9382.pdf} & Pine looper or Bordered white, \textit{Bupalus piniaria}, Kessock & 9382 & \cite{broekhuizen1993} & Unknown, but sampling intensity was decreasing & 1.00 & 3 (2--5) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/10127.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Scotland - un-named area & 10127 & \cite{potts1984} & Environment- and parisite-caused cycles & 0.99 & 3 (2--5) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/10007.pdf} & Water vole, \textit{Arvicola terrestris}, Le Pont & 10007 & \cite{saucy1994} & Predator-environment cycle interactions & 1.00 & 3 (2--5) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/20579.pdf} & Grey heron, \textit{Ardea cinerea}, Southern Britain & 20579 & \cite{stafford1971} & Severe winter & 0.98 & 3 (2--7) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/9655.pdf} & Flea beetle, \textit{Chaetocnoma concinna}, Finland & 9655 & \cite{markkula1965} & Cannot locate original source & 0.99 & 3 (2--6) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/1235.pdf} & Wren, \textit{Troglodytes troglodytes}, Eastern Wood, Bookham Common & 1235 & \cite{newton1998} & Severe winter & 0.98 & 3 (2--8) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/10113.pdf} & Willow grouse, \textit{Lagopus lagopus}, Northern England & 10113 & \cite{dobson1995} & Parasites and predators & 0.99 & 3 (2--6) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/9667.pdf} & Gooseberry sawfly, \textit{Nemastus ribesii}, Finland & 9667 & \cite{markkula1965} & Cannot locate original source & 0.97 & 3 (2--8) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/9679.pdf} & Unknown, \textit{Trioza apicalis}, Finland & 9679 & \cite{markkula1965} & Cannot locate original source & 0.83 & 3 (2--88) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/20527.pdf} & Wandering albatross, \textit{Diomedea exulans}, Taiaroa & 20527 & \cite{robertson1998} & Unknown & 0.99 & 4 (2--7) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/10039.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Northern Scotland & 10039 & \cite{dobson1995} & Parasites and predators & 0.92 & 4 (2--12) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/10162.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Atholl Estate & 10162 & \cite{mackenzie1952} & Bad environmental conditions and overcrowding combined to create crashes & 0.91 & 5 (2--12) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/9503.pdf} & Fisher or  Pekan, \textit{Martes pennanti}, Manitoba & 9503 & \cite{keith1963} & Unknown & 0.76 & 5 (2--87) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/7099.pdf} & European rabbit, \textit{Oryctolagus cuniculus}, Estate 2, East Anglia & 7099 & \cite{barnes1986} & Disease outbreak followed by years of good weather & 0.78 & 5 (2--67) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/2778.pdf} & Wheatear, \textit{Oenanthe oenanthe}, Skokholm Island & 2778 & \cite{lack1969} & Unknown, but decline noted specifically, cold winters caused some crashes & 0.67 & 5 (2--141) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/9659.pdf} & Cabbage root fly or maggot, \textit{Delia radicum}, Finland & 9659 & \cite{markkula1965} & Cannot locate original source & 0.59 & 7 (2--161) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/1195.pdf} & Blue jay, \textit{Cyanocitta cristata}, Robert Allerton Park & 1195 & \cite{kendeigh1982} & Unknown & 0.60 & 7 (2--149) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/5019.pdf} & Barbary macaque, \textit{Macaca sylvanus}, Queens Gate & 5019 & \cite{fa1984} & Cannot locate original source & 0.61 & 7 (2--132) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/9953.pdf} & Rock ptarmigan, \textit{Lagopus mutus}, Iceland & 9953 & \cite{clarke1885,williams1954} & Severe winters & 0.58 & 7 (2--172) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/6548.pdf} & Lesser-spotted dogfish, \textit{Scyliorhinus caniculus}, North Sea & 6548 & \cite{heessen1996} & Unknown, not specifically mentioned & 0.53 & 8 (2--190) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/20546.pdf} & American red fox, \textit{Vulpes fulva}, Labrador & 20546 & \cite{dancona1954,lindstrom1994} & Predator-prey cycles & 0.57 & 9 (3--65) \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/9470.pdf} & Ruffed grouse, \textit{Bonasa umbellus}, Connecticut & 9470 & \cite{keith1963} & Unknown & 0.53 & 9 (3--158) \\
+\includegraphics[width=1.7cm]{../analysis/sparks/6528.pdf} & Shag, \textit{Phalacrocorax aristotelis}, Farne Islands, Northumberland & 6528 & \cite{potts1980} & Red tide event combined with low productivity due to overcrowding & 1.00 \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/7115.pdf} & South African fur seal, \textit{Arctocephalus pusillus}, South Africa & 7115 & \cite{shaughnessy1982} & Harvesting and predation changes & 1.00 \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/10128.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Scotland - un-named area & 10128 & \cite{potts1984} & Environment- and parisite-caused cycles & 1.00 \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/9382.pdf} & Pine looper or Bordered white, \textit{Bupalus piniaria}, Kessock & 9382 & \cite{broekhuizen1993} & Unknown, but sampling intensity was decreasing & 1.00 \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/10127.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Scotland - un-named area & 10127 & \cite{potts1984} & Environment- and parisite-caused cycles & 0.99 \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/10007.pdf} & Water vole, \textit{Arvicola terrestris}, Le Pont & 10007 & \cite{saucy1994} & Predator-environment cycle interactions & 1.00 \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/20579.pdf} & Grey heron, \textit{Ardea cinerea}, Southern Britain & 20579 & \cite{stafford1971} & Severe winter & 0.98 \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/9655.pdf} & Flea beetle, \textit{Chaetocnoma concinna}, Finland & 9655 & \cite{markkula1965} & Cannot locate original source & 0.99 \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/1235.pdf} & Wren, \textit{Troglodytes troglodytes}, Eastern Wood, Bookham Common & 1235 & \cite{newton1998} & Severe winter & 0.98 \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/10113.pdf} & Willow grouse, \textit{Lagopus lagopus}, Northern England & 10113 & \cite{dobson1995} & Parasites and predators & 0.99 \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/9667.pdf} & Gooseberry sawfly, \textit{Nemastus ribesii}, Finland & 9667 & \cite{markkula1965} & Cannot locate original source & 0.97 \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/9679.pdf} & Unknown, \textit{Trioza apicalis}, Finland & 9679 & \cite{markkula1965} & Cannot locate original source & 0.83 \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/20527.pdf} & Wandering albatross, \textit{Diomedea exulans}, Taiaroa & 20527 & \cite{robertson1998} & Unknown & 0.99 \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/10039.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Northern Scotland & 10039 & \cite{dobson1995} & Parasites and predators & 0.92 \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/10162.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Atholl Estate & 10162 & \cite{mackenzie1952} & Bad environmental conditions and overcrowding combined to create crashes & 0.91 \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/9503.pdf} & Fisher or  Pekan, \textit{Martes pennanti}, Manitoba & 9503 & \cite{keith1963} & Unknown & 0.76 \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/7099.pdf} & European rabbit, \textit{Oryctolagus cuniculus}, Estate 2, East Anglia & 7099 & \cite{barnes1986} & Disease outbreak followed by years of good weather & 0.78 \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/2778.pdf} & Wheatear, \textit{Oenanthe oenanthe}, Skokholm Island & 2778 & \cite{lack1969} & Unknown, but decline noted specifically, cold winters caused some crashes & 0.67 \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/1195.pdf} & Blue jay, \textit{Cyanocitta cristata}, Robert Allerton Park & 1195 & \cite{kendeigh1982} & Unknown & 0.60 \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/5019.pdf} & Barbary macaque, \textit{Macaca sylvanus}, Queens Gate & 5019 & \cite{fa1984} & Cannot locate original source & 0.61 \\ 
    \bottomrule

### 2015-07-22 analysis/6-causes-table.R

@@ -9,7 +9,7 @@ heavy_refs <- read.csv("heavy-filled.csv", stringsAsFactors = FALSE,
 
 source("5-shape-data.R")
 
-heavy <- filter(gomp_hat_base, p10 > .5) %>%
+heavy <- filter(gomp_hat_base, p10 > .6) %>%
   select(main_id, common_name, p10, nu_50, nu_5, nu_95, taxon_name)
 
 heavy <- inner_join(select(heavy, -taxon_name, -common_name), heavy_refs)
@@ -25,10 +25,10 @@ heavy <- mutate(heavy,
   mutate(nu_95 = sprintf("%.0f", round(nu_95, 0))) %>%
   mutate(spark = paste0("\\includegraphics[width=1.7cm]{../analysis/sparks/", main_id, ".pdf}")) %>%
   mutate(nu_hat = paste0(nu_50, " (", nu_5, "--", nu_95, ")")) %>%
-  select(spark, population, main_id, citation, reasons, p10, nu_hat) %>%
+  select(spark, population, main_id, citation, reasons, p10) %>%
   rename("Time series" = spark, "Population" = population, "ID" = main_id,
-    "Citation" = citation,
-    "Description" = reasons, "Pr($\\nu < 10$)" = p10, "$\\widehat{\\nu}$" = nu_hat)
+    "Ref" = citation,
+    "Description" = reasons, "Pr($\\nu < 10$)" = p10)#, "$\\widehat{\\nu}$" = nu_hat)
 
 print.xtable(xtable(heavy,
     caption = ""),

### 2015-07-22 analysis/6-plot-sparks.R

@@ -33,8 +33,8 @@ make_spark <- function(x) {
   res <- get_gomp_res(x$main_id[1])
   bsw_l <- as.numeric(na.omit(seq_along(res$res)[res$res<res$l]))
   bsw_u <- as.numeric(na.omit(seq_along(res$res)[res$res>res$u]))
-  pdf(paste0("sparks/", x$main_id[1], ".pdf"), width = 2.5, height = 1.9)
-  par(mar = c(0.5,0.2,0.5,0.2), oma = c(0,0,6,0), cex = 0.8)
+  pdf(paste0("sparks/", x$main_id[1], ".pdf"), width = 2.5, height = 1.2)
+  par(mar = c(0.5,0.2,0.5,0.2), oma = c(0,0,0.1,0), cex = 0.8)
   par(xpd = NA)
   with(x, plot(seq_along(population_untransformed), population_untransformed, type = "l", axes = FALSE,
       xlab = "", ylab = "", yaxs = "i", xaxs = "i", log = "y"))

### 2015-07-22 text/ms.tex
Substantial restructure of methods

@@
 \begin{center}
-\includegraphics[width=0.55\textwidth]{../analysis/skew-fig.pdf}
+\includegraphics[width=0.65\textwidth]{../analysis/skew-fig.pdf}
 \end{center}
@@
 
-We selected abundance time series from the Global Population Dynamics Database (GPDD;\cite{gpdd2010}), which contains nearly 5000 time series of abundance from \(\sim 1000\) species and \(\sim 100\) taxonomic orders.
-We filtered the data (Supporting materials) to remove populations from less reliable data sources, and those without sufficient data for our models, and then interpolated some missing values (sensu Brook \textit{et al.}\cite{brook2006a}).
-Our interpolation affected only \(\sim\) \interpPointsPerc \% of the final data points (Supporting materials Table~\ref{tab:stats}) and none of the data points that were later considered black-swan events.
-Our final dataset contained \NPops\ populations across \NOrders\ taxonomic orders and seven taxonomic classes, with a median of \medianTimeSteps\ time steps (range of \minTimeSteps--\maxTimeSteps) (Extended Data Table~\ref{tab:stats}).
+We selected abundance time series from the Global Population Dynamics Database\cite{gpdd2010} (GPDD), which contains nearly 5000 time series of abundance from \(\sim 1000\) species and \(\sim 100\) taxonomic orders.
+We filtered the data (Supporting Information) to remove populations from less reliable data sources, and those without sufficient data for our models, and then interpolated some missing values (sensu Brook \textit{et al.}\cite{brook2006a}).
+Our interpolation affected only \(\sim\) \interpPointsPerc \% of the final data points (Extended Data Table~1) and none of the data points that were later considered black-swan events.
+Our final dataset contained \NPops\ populations across \NOrders\
+taxonomic orders and seven taxonomic classes, with a median of \medianTimeSteps\ time steps (range of \minTimeSteps--\maxTimeSteps) (Extended Data Table~1).
 
-We fit heavy-tailed Gompertz population dynamics models to data from the Global Population Dynamics Database (GPDD).
-\emph{{[}TODO something about extensive quality control filtering and interpolation of the time series{]}} The Gompertz model represents population growth as a linear function in log space.
-If \(x_{t}\) represents the log abundance (\(N\)) at time \emph{t} then \(x_t = \lambda + bx_{t - 1} + \epsilon_t,\ \epsilon_t \sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma)\).
+We fit heavy-tailed Gompertz population dynamics models to data from the GPDD.
+The Gompertz model represents population growth as a linear function in log space.
+If \(x_{t}\) represents the log abundance (\(N\)) at time \emph{t} then
+\begin{align}
+x_t &= \lambda + bx_{t - 1} + \epsilon_t\\
+\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma)
+\end{align}
 The growth parameter \(\lambda\) represents the expected growth rate if \(N_{t} = 1\).
 The model is density independent if \(b = 1\), maximally density dependent if \(b = 0\), and inversely density dependent if \(b < 0\).
-
-Usually, the process noise \(\epsilon_{t}\) is modelled as normally distributed{[}REFs{]}, but we allow it to be drawn from a Student-t distribution with scale parameter \(\sigma\) and degrees of freedom \(\nu\).
+Usually, the process noise \(\epsilon_{t}\) is modelled as normally distributed, but we allow it to be drawn from a Student-t distribution with scale parameter \(\sigma\) and degrees of freedom \(\nu\).
 We can then estimate the degree to which the process deviations have heavy tails and are thereof evidence of black-swan events (Fig.~1a, b).
 
-We chose a weakly informative exponential prior on \(\nu\) with rate parameter of \(0.01\) truncated at values above two---a slightly less informative prior than suggested by Fernandez and Steel\cite{fernandez1998}.
-This prior gives only a \(0.08\) probability that \(\nu < 10\) but constrains sampling sufficiently to avoid drifting towards infinity.
-In any case, for \(\nu > 20\) the t distribution is almost indistinguishable from the normal distribution (Fig.~1).
-Based on the shape of the t distribution, we chose the probability that \(\nu < 10\), Pr(\(\nu < 10\)), to define the probability of heavy-tailed dynamics.
-When categorizing a population as heavy or normal tailed, we applied a 0.5 probability threshold.
-
-\emph{Details on the heavy-tailed Gompertz probability model}
-
-For the Gompertz model, our weakly-informative priors (Extended Data Fig.~\ref{fig:priors}) were: \(b \sim \mathrm{Uniform}(-1, 2)\), \(\lambda \sim \mathrm{Normal}(0, 10^2)\), \(\sigma \sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)\), \(\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2)\).
+For the Gompertz model, our weakly-informative priors (Extended Data Fig.~\ref{fig:priors}) were:
+\begin{align}
+b &\sim \mathrm{Uniform}(-1, 2)\\ \lambda &\sim \mathrm{Normal}(0, 10^2)\\
+\sigma &\sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)\\ \nu &\sim
+\mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2).
+\end{align}
 Our prior on \(b\) was uninformative between values of \(-1\) and \(2\).
@@
 
-Our prior on \(\nu\) was based on Fernandez and Steel\cite{fernandez1998}.
-They chose an exponential rate parameter of \(0.1\).
-We chose a less informative rate parameter of \(0.01\) and truncated the distribution at \(2\), since at \(\nu < 2\) the variance of the t distribution is undefined.
-This prior gives only a \(7.7\)\% probability that \(\nu < 10\) but constrains the sampling sufficiently to avoid wandering off towards infinity---above approximately \(\nu = 20\) the t distribution is so similar to the normal distribution (Fig.~1) that time series of the length considered here are unlikely to be informative about the precise value of \(\nu\).
-In the scenario where the data are uninformative about heavy tails (e.g.~Fig.~1e, h), the posterior will approximately match the prior (prior median \(= 71\), mean \(= 102\)) and the metrics used in our paper (e.g.
-\(Pr(\nu < 10) > 0.5\)) are unlikely to flag the population as heavy tailed.
+We chose a weakly informative exponential prior on \(\nu\) with rate parameter of \(0.01\) truncated at values above two---a slightly less informative prior than suggested by Fernandez and Steel\cite{fernandez1998}.
+We truncated the distribution at \(2\), since at \(\nu < 2\) the variance of the t distribution is undefined.
+This prior gives only a \(7.7\)\% probability that \(\nu < 10\) but constrains the sampling sufficiently to avoid drifting towards infinity.
+In any case, for \(\nu > 20\) the t distribution is almost indistinguishable from the normal distribution (Fig.~1).
+Based on the shape of the t distribution, we chose the probability that \(\nu < 10\), Pr(\(\nu < 10\)), to define the probability of heavy-tailed dynamics.
+When categorizing a population as heavy or normal tailed, we applied a 0.5 probability threshold.
+In the scenario where the data are uninformative about heavy tails (e.g.~Fig.~1e, h), the posterior will approximately match the prior (prior median \(= 71\), mean \(= 102\)) and a metric of \(Pr(\nu < 10) > 0.5\)) is unlikely to flag the population as heavy tailed.
 
 We fit our models with Stan 2.4.0\cite{stan-manual2014}, and R 3.1.1\cite{r2014}.
-We began with four chains and \(2000\) iterations, discarding the first \(1000\) as warm up (i.e.~4000 total samples).
+We began with four chains and \(2000\) iterations, discarding the first \(1000\) as warm up (i.e.~4000 final samples).
 If \(\hat{R}\) (the potential scale reduction factor---a measure of chain convergence) was greater than \(1.05\) for any parameter or the minimum effective sample size, \(n_\mathrm{eff}\), (a measure of the effective number of uncorrelated samples) for any parameter was less than \(200\), we doubled both the total iterations and warm up period and sampled from the model again.
@@
 We continued this procedure up to \(8000\) iterations (\(16000\) total samples) by which all chains were deemed to have sufficiently converged.
-The No-U-Turn Hamiltonian Markov chain Monte Carlo Sampler in Stan generally requires far fewer iterations to obtain equivalent effective sample sizes than other software such as WinBUGS or JAGS\cite{stan-manual2014}.
+The No-U-Turn Hamiltonian Markov chain Monte Carlo sampler in Stan generally requires far fewer iterations to obtain equivalent effective sample sizes than Gibbs or Metropolis--Hastings algorithms\cite{stan-manual2014}.
 
-\emph{The skewed Student's t distribution}
-
-We fit Gompertz models with skewed Student's t distributed process noise based on the probability density for the skewed t distribution described in Fernandez and Steel\cite{fernandez1998}.
+\textbf{Skewed Student-t distribution} We fit Gompertz models with skewed Student-t distributed process noise based on the probability density for the skewed t distribution described in Fernandez and Steel\cite{fernandez1998}.
 Their definition equates to the following probability density as implemented in King\cite{king2012}:
 
-\begin{equation*}
+\begin{equation}
   \mathrm{Skewed\,Student\mhyphen t}(x) =
@@
 \end{dcases}
-\end{equation*}
+\end{equation}
 
 Here \(\gamma\) represents the skewness parameter.
-The Student-t(\(\cdot\)) portion represents the data point of interest, the heavy-tailedness parameter, the location parameter, and the scale parameter as fed to the standard Student-t probability density.
+The Student-t(\(\cdot\)) portion represents in order: the data point of interest, the heavy-tailedness parameter, the location parameter, and the scale parameter as passed to the standard Student-t probability density.
 The distribution is symmetrical if \(\gamma = 1\), left skewed (skewed downwards in abundance) if \(0 < \gamma < 1\), and right skewed (skewed upwards in abundance) if \(1 < \gamma < \infty\).
+See the Supporting Information for details on the implementation of the skewed Student-t distribution in Stan.
 
-\emph{Alternative priors}
-
-To test if the prior on \(\nu\) influenced our estimate of black-swan dynamics, we refit our models with weaker and stronger priors.
+\textbf{Alternative priors} To test if the prior on \(\nu\) influenced our estimate of black-swan dynamics, we refit our models with weaker and stronger priors.
 Our base model used a prior on \(\nu\) of Truncated-Exponential(0.02, min.\ = 2).
@@
 Note that the base and weaker priors are relatively flat within the region of \(\nu < 20\), which is the region we are mostly concerned about when categorizing populations as heavy- or thin-tailed.
+TODO - note Gelman 1/x prior.
 
@@
 
-\emph{Alternative population models}
+\textbf{Alternative population models} We fit alternative population models to test if four key phenomena systematically changed our conclusions The range of percentages of black swans by taxonomic class cited in the abstract are based on lower and upper limits across our main Gompertz model and these four alternative models.
 
-We fit alternative population models to test if four key phenomena systematically changed our conclusions.
-These models allowed for AR1 autocorrelation of the residuals, density independence, observation error, and Ricker-logistic population dynamics.
-Furthermore, we tested the sensitivity of our results to weaker and stronger priors on \(\nu\) and used simulated data to test how easily we could detect \(\nu\) given different sample sizes and to ensure we could recover unbiased parameter estimates.
-The range of percentages of black swans by taxonomic class cited in the abstract and results are based on lower and upper limits across our main Gompertz model and these four alternative models.
-
-\emph{Autocorrelated residuals}
-
-We considered a version of the Gompertz model in which an autoregressive parameter was fit to the process noise residuals: \(x_t = \lambda + b x_{t-1} + \epsilon_t,\, \epsilon_t \sim \mathrm{Student\mhyphen t}(\nu, \phi \epsilon_{t-1}, \sigma)\).
+\emph{Autocorrelated residuals} We considered a version of the Gompertz model in which an autoregressive parameter was fit to the process noise residuals:
+\begin{align}
+x_t &= \lambda + b x_{t-1} + \epsilon_t\\
+\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, \phi \epsilon_{t-1}, \sigma).
+\end{align}
 In addition to the parameters in the original Gompertz model, this model estimates an additional parameter \(\phi\), which represents the correlation of subsequent residuals.
-Based on the results of previous analyses with the GPDD\cite{connors2014} and the chosen priors in previous analyses\cite{thorson2014a} and to greatly speed up chain convergence when running our model across all populations, we placed a weakly informative prior on \(\phi\) that assumed the greatest probability density near zero with the reduced possibility of \(\phi\) being near \(-1\) or \(1\).
+Based on the results of previous analyses with the GPDD\cite{connors2014} and the chosen priors in previous analyses\cite{thorson2014a} and to speed up chain convergence when running our model across all populations, we placed a weakly informative prior on \(\phi\) that assumed the greatest probability density near zero with the reduced possibility of \(\phi\) being near \(-1\) or \(1\).
 Specifically, we chose \(\phi \sim \mathrm{Truncated\mhyphen Normal}(0, 1, \mathrm{min.} = -1, \mathrm{max.} = 1)\).
 The MCMC chains did not converge for \modelsNoConvergeAROne\ populations according to our criteria (\(\widehat{R} < 1.05, n_\mathrm{eff} > 200\)) after 8000 iterations of four chains.
-This included only \modelsNoConvergeAROneHeavyBase\ populations in which Pr(\(\nu < 10\)) \(> 0.5\) categorized them as heavy in the main Gompertz model.
-We did not include these models in Extended Data Fig.~\ref{fig:alt}.
+This included only \modelsNoConvergeAROneHeavyBase\ population in which Pr(\(\nu < 10\)) \(> 0.5\) categorized it as heavy in the main Gompertz model.
+We did not include this population in Extended Data Fig.~\ref{fig:alt}.
 
-\emph{Assumed density independence}
-
-We fit a simplified version of the Gompertz model in which the density dependence parameter \(b\) was fixed at \(1\) (density independent).
+\emph{Assumed density independence} We fit a simplified version of the Gompertz model in which the density dependence parameter \(b\) was fixed at \(1\) (density independent).
 This is equivalent to fitting a random walk model (with drift) to the \(\log\) abundances or assuming the growth rates are drawn from a stationary distribution.
-The model was \(x_t = \lambda + x_{t-1} + \epsilon_t,\, \epsilon \sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma)\).
+The model was as follows:
+\begin{align}
+x_t &= \lambda + x_{t-1} + \epsilon_t\\
+\epsilon &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma).
+\end{align}
 We fit this model for three reasons: (1) it is computationally simpler and so provides a check that our more complicated full Gompertz model was obtaining reasonable estimates of \(\nu\), (2) it provides a test of whether density dependence was systematically affecting our perception of heavy tails, (3) it matches how some previous authors have modelled heavy tails without accounting for density dependence\cite{segura2013}.
 
-\emph{Assumed observation error}
-
-Observation error can bias parameter estimates\cite{knape2012} and is known to affect the ability to detect extreme events\cite{ward2007}.
+\emph{Assumed observation error} Observation error can bias parameter estimates\cite{knape2012} and is known to affect the ability to detect extreme events\cite{ward2007}.
 In our main analysis, we fit a model that ignored observation error.
@@
 However, simultaneously estimating observation and process noise is a challenging problem (e.g.\ because the observation and process noise parameters tend to negatively covary in model fitting) and is known to sometimes result in identifiability issues with the Gompertz population model\cite{knape2008}.
-Furthermore, our model was applied to hundreds of time series, often of short length (as few as 20 time steps) and our model estimates an additional parameter---the shape of the process deviation tails---potentially making identifiability and computational issues even greater.
+Furthermore, our model was applied to hundreds of time series, often of short length (as few as \minTimeSteps\ time steps) and our model estimates an additional parameter---the shape of the process deviation tails---potentially making identifiability and computational issues even greater.
 Therefore, we considered a version of the base Gompertz model that allowed for a fixed level of observation error:
-
-\begin{align*}
+\begin{align}
 U_t &= \lambda + b U_{t-1} + \epsilon_t\\
@@
 \epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma_\mathrm{proc}),
-\end{align*}
-
+\end{align}
 where \(U\) represents the unobserved state vector, \(\sigma_\mathrm{obs}\) represents the standard deviation of observation error (on a log scale), and \(\sigma_\mathrm{proc}\) represent the process noise scale parameter.
@@
 
-\emph{Ricker-logistic}
-
-We also fit a Ricker-logistic model: \(x_t = x_{t-1} + r_{\mathrm{max}}\left(1 - N_{t-1}/K\right) + \epsilon_t,\, \epsilon_t \sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma)\), where \(r_\mathrm{max}\) represents the theoretical maximum population growth rate that is obtained when \(N_t\) (abundance at time \(t\)) \(= 0\).
+\emph{Ricker-logistic} We also fit a Ricker-logistic model:
+\begin{align}
+x_t &= x_{t-1} + r_{\mathrm{max}}\left(1 - \frac{N_{t-1}}{K}\right) + \epsilon_t\\
+\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma),
+\end{align}
+where \(r_\mathrm{max}\) represents the theoretical maximum population growth rate that is obtained when \(N_t\) (abundance at time \(t\)) \(= 0\).
 The parameter \(K\) represents the carrying capacity and, as before, \(x_t\) represents the \(\log\) transformed abundance at time \(t\).
 The Ricker-logistic model assumes a linear decrease in population growth rate with increases in abundance.
-In contrast, the Gompertz model assumes a linear decrease in population growth rate with increases in \textit{log} abundance (\(x_t\))\cite{thibaut2012}.
+In contrast, the Gompertz model assumes a linear decrease in population growth rate with increases in \textit{log} abundance (\(x_t\)).
 
-To fit the Ricker-logistic models, we chose a prior on \(K\) uniform between zero and twice the maximum observed abundance Clark \emph{et al.}\cite{clark2010} chose uniform between zero and maximum observed, which is more informative).
+To fit the Ricker-logistic models, we chose a prior on \(K\) uniform between zero and twice the maximum observed abundance (similar and less informative than in Clark \emph{et al.}\cite{clark2010}).
 We set the prior on \(r_\mathrm{max}\) as uniform between 0 and 20 as in Clark \emph{et al.}\cite{clark2010}.
@@
 
-\emph{Modelling covariates of heavy-tailed dynamics}
+\textbf{Simulation testing} We performed two types of simulation testing.
+First, we tested how easily the Student-t distribution \(\nu\) parameter could be recovered given different true values of \(\nu\) and different sample sizes.
+Second, we tested the ability of the heavy-tailed Gompertz model to obtain unbiased parameter estimates of \(\nu\) given that a set of process deviations was provided in which the effective \(\nu\) value was close to the true \(\nu\) value.
 
-We fit a multilevel beta regression model to the predicted probability of heavy tails, Pr(\(\nu < 10\)), to investigate potential covariates of heavy-tailed dynamics.
+We separated our simulation into these two components to avoid confounding two issues.
+(1) With smaller sample sizes, there may not be a stochastic draw from the tails of a distribution.
+In that case, no model, no matter how perfect, will be able to detect the shape of the tails.
+(2) Complex models may return biased parameter estimates if there are conceptual, computational, or coding errors.
+Our first simulation tested the first issue and our second simulation tested the latter.
+In general, our simulations show that, if anything, our model under predicts the magnitude and probability of heavy tailed events---especially given the length of the time series in the GPDD.
+
+First, we tested the ability to estimate \(\nu\) given different true values of \(\nu\) and different sample sizes.
+We took stochastic draws from t distributions with different \(\nu\) values (\(\nu = 3, 5, 10,\) and \(10^6\) {[}\(\approx\) normal{]}), with central tendency parameters of \(0\), and scale parameters of \(1\).
+We started with \(1600\) stochastic draws and then fit the models again at the first \(800, 400, 200, 100, 50,\) and \(25\) draws.
+Each time we recorded the posterior samples of \(\nu\).
+
+We found that we could consistently and precisely recover median posterior estimates of \(\nu\) near the true value of \(\nu\) with large samples (\(\ge 200\)) (Extended Data Fig.~\ref{fig:sim-nu} upper panels).
+At smaller samples we could still usually qualitatively distinguish heavy from not-heavy tails, but the model tended to underestimate how heavy the tails were.
+At the same time, at smaller sample sizes, the model tended to overestimate how large the scale parameter was (Extended Data Fig.~\ref{fig:sim-nu} lower panels).
+
+Secondly, we tested the ability of the heavy-tailed Gompertz model to obtain unbiased parameter estimates when the process noise was chosen so that appropriate tail events were present.
+To generate these process deviations for the \(\nu = 3\) and \(\nu = 5\) scenarios, we repeatedly drew proposed candidate process deviations and estimated the central tendency, scale, and \(\nu\) values each time.
+We recorded when \(\hat{\nu}\) (median of the posterior) was within \(0.2\) CVs (coefficient of variations) of the true \(\nu\) value and used this set of random seed values in our Gompertz simulation (Supporting Information).
+
+We then fit our Gompertz models to the simulated datasets with all parameters (except \(\nu\)) set near the median values estimated in the GPDD.
+We repeated this with \(50\) and \(100\) samples without observation error, \(50\) samples with observation error (\(\sigma_\mathrm{obs} = 0.2\)), and \(50\) samples with the same observation error and a Gompertz model that allowed for correctly specified observation error magnitude.
+Our results indicate that the Gompertz model can recapture the true value of \(\nu\) when the process noise was chosen so that appropriate tail events were present (Extended Data Fig.~\ref{fig:sim-prob} upper panels).
+The addition of observation error caused the model to tend to underestimate the degree of heavy-tailedness.
+Fitting a model with correctly specified observation error did not make substantial improvements to model bias (Extended Data Fig.~\ref{fig:sim-prob}).
+
+When converting the posterior distributions of \(\nu\) into Pr(\(\nu < 10\)), the models distinguished heavy and not-heavy tails reasonably well (Extended Data Fig.~\ref{fig:sim-prob} lower panels).
+Without observation error, and using a probability of \(0.5\) as a threshold, the model correctly classified all simulated systems with normally distributed process noise as not heavy tailed.
+The model would have miscategorized only one of \(40\) simulations at \(\nu = 5\) across simulated populations with \(50\) or \(100\) time steps (Extended Data Fig.~\ref{fig:sim-prob}, scenarios 1 and 2 in lower row, second panel from left).
+The model would have correctly categorized all cases where the process noise was not heavy tailed (Extended Data Fig.~\ref{fig:sim-prob} bottom-right panel) and all cases where \(\nu = 3\) and there was not observation error.
+With \(0.2\) standard deviations of observation error, the model still categorized \obsErrorNuFivePerc\% of cases as heavy tailed when \(\nu = 5\) and all but one case when \(\nu = 3\).
+Allowing for observation error made little improvement to the detection of heavy tails.
+Therefore, we chose to focus on the simpler model without observation error in the main text, particularly given that the true magnitude of observation error was unknown in the empirical data.
+
+\textbf{Covariates of heavy-tailed dynamics} We fit a multilevel beta regression model to the predicted probability of heavy tails, Pr(\(\nu < 10\)), to investigate potential covariates of heavy-tailed dynamics.
 The beta distribution is useful when response data range on a continuous scale between zero and one\cite{ferrari2004}.
@@
 The model was as follows:
-
-\begin{align*}
+\begin{align}
 \mathrm{Pr}(\nu_i < 10) &\sim \mathrm{Beta}(A_i, B_i)\\
@@
   \: \text{for } l = 1, \dots, 301,
-\end{align*}
-
+\end{align}
 where \(A\) and \(B\) represent the beta distribution shape parameters; \(\mu_i\) represents the predicted value for population \(i\), class \(j\), order \(k\), and species \(l\); \(\phi_\mathrm{disp}\) represents the dispersion parameter; and \(X_i\) represents a vector of predictors (such as lifespan) for population \(i\) with associated \(\beta\) coefficients.
@@
 We log transformed \(\sigma\), time-series length, and lifespan to match the way they are visually represented in Extended Data Fig.~\ref{fig:correlates} and to make the relationship approximately linear on the logit-transformed response scale.
-All input variables were standardized by subtracting their mean and dividing by two standard deviations to make their coefficients comparable in magnitude \cite{gelman2008c}.
+All input variables were standardized by subtracting their mean and dividing by two standard deviations to make their coefficients comparable in magnitude\cite{gelman2008c}.
 We excluded body length as a covariate because it was highly correlated with lifespan, and lifespan exhibited more overlap across taxonomy than body length.
@@
 
-We incorporated weakly informative priors into our model: \(\mathrm{Cauchy}(0, 10)\) on the global intercept \(\alpha\), \(\mathrm{Half\mhyphen Cauchy}(0, 2.5)\) on all standard deviation parameters, \(\mathrm{Half\mhyphen Cauchy}(0, 10)\) on the dispersion parameter \(\phi_\mathrm{disp}\), and \(\mathrm{Cauchy}(0, 2.5)\) on all other parameters \cite{gelman2006c, gelman2008d}.
+We incorporated weakly informative priors into our model: \(\mathrm{Cauchy}(0, 10)\) on the global intercept \(\alpha\), \(\mathrm{Half\mhyphen Cauchy}(0, 2.5)\) on all standard deviation parameters, \(\mathrm{Half\mhyphen Cauchy}(0, 10)\) on the dispersion parameter \(\phi_\mathrm{disp}\), and \(\mathrm{Cauchy}(0, 2.5)\) on all other parameters\cite{gelman2006c, gelman2008d}.
 Compared to normal priors, the Cauchy priors concentrate more probability density around expected parameter values while allowing for a higher probability density far into the tails, thereby allowing the data to dominate the posterior more strongly if it disagrees with the prior.
@@
 
+\renewcommand{\arraystretch}{0.1}% Tighter
+
 \begin{helvetica}
@@
 \begin{scriptsize}
-\begin{longtable}{>{\RaggedRight}m{1.4cm}>{\RaggedRight}p{3.3cm}>{\RaggedRight}p{0.7cm}>{\RaggedRight}p{1.5cm}>{\RaggedRight}p{3.0cm}>{\RaggedRight}p{1.4cm}>{\RaggedRight}p{1.25cm}}
-
-\caption{\textbf{All populations with Pr$(\nu < 10) > 0.5$ in the base heavy-tailed
-  Gompertz population dynamics model.} Shown are the log abundance time series,
-  population descriptions, Global Population Dynamics Database Main IDs,
-  citation for the data source or separate verification literature, a
-  description of the cause of the black swan events (if known), the
-  probability of heavy tails as calculated by our model, and median estimate
-  of $\nu$ from our model with 90\% quantile credible intervals indicated in
-  parentheses. Red dots on the time series indicate downward black-swan events
-  and blue values indicate upward black-swan events that have a $10^{-4}$ probability or less of occurring if the population dynamics were
+\begin{longtable}{>{\RaggedRight}m{1.4cm}>{\RaggedRight}p{6.2cm}>{\RaggedRight}p{0.7cm}>{\RaggedRight}p{0.4cm}>{\RaggedRight}p{4.4cm}>{\RaggedRight}p{1.4cm}}
+\caption{\textbf{Populations with a high probability of heavy-tailed
+dynamics in the base heavy-tailed Gompertz population dynamics model.} Shown
+are the log abundance time series, population descriptions, Global Population
+Dynamics Database Main IDs, citation for the data source or separate
+verification literature, a description of the cause of the black swan events
+(if known), the probability of heavy tails as calculated by our model, and
+median estimate of $\nu$ from our model with 90\% quantile credible intervals
+indicated in parentheses. Red dots on the time series indicate downward
+black-swan events and blue values indicate upward black-swan events that have a
+$10^{-4}$ probability or less of occurring if the population dynamics were
 explained by a Gompertz model with normally distributed process noise with a
@@
 \end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=0.8\textwidth]{../analysis/t-dist-sampling-sim-prior-exp0point01.pdf}
+\includegraphics[width=0.8\textwidth]{../analysis/t-dist-sampling-sim-sigma-prior-exp0point01.pdf}
+
+\caption{Testing the ability to estimate $\nu$ (top
+  panels) and the scale parameter of the process deviations (bottom panels)
+  for a given number of samples (columns) drawn from a distribution with a
+  given true $\nu$ value (rows). The red lines indicate the true population
+  value. When a small number of samples are drawn there may not be samples
+  sufficiently far into the tails to recapture the true $\nu$ value; however,
+  heavy tails are still distinguished from normal tails in most cases, even
+  with only 25 or 50 samples.}
+
+\label{fig:sim-nu}
+\end{center}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=\textwidth]{../analysis/sim-gompertz-median-dist.pdf}
+\includegraphics[width=\textwidth]{../analysis/sim-gompertz-p10.pdf}
+
+\caption{Simulation testing the Gompertz estimation model when the process
+deviation draws were chosen so that $\nu$ could be estimated close to the true
+value outside the full population model (``effective $\nu$'' within a CV of 0.2
+of specified $\nu$). Upper panels show the distribution of median
+$\widehat{\nu}$ across 20 simulation runs. Lower panels show the distribution
+of Pr($\nu < 10$) across 20 simulation runs. We ran the simulations across
+three population (``true'') $\nu$ values (3, 5, and $1\cdot 10^9$, i.e.\
+approximately normal) and four scenarios: (1) 100 time steps and no observation
+error, (2) 50 time steps and no observation error, (3) 50 time steps and
+observation error drawn from $\mathrm{Normal} (0, 0.2^2)$ but ignored, and (4)
+50 time steps with observation error in which the quantity of observation error
+was assumed known. Within each scenario the dots represent stochastic draws
+from the true population distributions combined with model fits. Underlayed
+boxplots show the median, interquartile range, and $1.5$ times the
+interquartile range. }
+
+\label{fig:sim-prob}
+\end{center}
+\end{figure}

### 2015-07-22 text/supp.tex
Substantial restructure of methods

@@
 \noindent
-We provide a supplemental figure of all the time series included in our
-analysis and indicate which values were interpolated (non-zero interpolations)
-(\percImputedPops\% of populations had at least one point interpolated but
-only \percImputedPoints\% of the total observations were interpolated)
-(Fig.~\ref{fig:all-ts}). Note that interpolation is highly unlikely to lead to
+%We provide a supplemental figure of all the time series included in our
+%analysis and indicate which values were interpolated (non-zero interpolations)
+\percImputedPops\% of populations had at least one point interpolated but
+only \percImputedPoints\% of the total observations were interpolated.
+We note that interpolation is highly unlikely to lead to
 black-swan detections, since black swans involve extreme increases or
-decreases. Table~\ref{tab:stats} shows the final taxonomic breakdown and the
-number of populations with interpolated values.
+decreases.
 
-\section{Alternative population models}
-
-We fit four alternative population models to the time-series data to check how
-they would influence our conclusions. Our alternative models allowed for
-autocorrelation in the residuals, assumed no density dependence, allowed for
-observation error, or assumed a Ricker-logistic functional form.
-
-\subsection{Autocorrelated residuals}
-
-We considered a version of the Gompertz model in which an autoregressive
-parameter was fit to the process noise residuals:
-\begin{align*}
-x_t &= \lambda + b x_{t-1} + \epsilon_t\\
-\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, \phi \epsilon_{t-1}, \sigma).
-\end{align*}
-In addition to the parameters in the original Gompertz model, this model
-estimates an additional parameter $\phi$, which represents the correlation of
-subsequent residuals. Based on the results of previous analyses with the GPDD
-\citep[e.g.][]{connors2014} and the chosen priors in previous analyses
-\citep[e.g.][]{thorson2014a} and to greatly speed up chain convergence when
-running our model across all populations, we placed a weakly informative prior
-on $\phi$ that assumed the greatest probability density near zero with the
-reduced possibility of $\phi$ being near $-1$ or $1$. Specifically, we chose
-$\phi \sim \mathrm{Truncated\mhyphen Normal}(0, 1, \mathrm{min.} = -1,
-\mathrm{max.} = 1)$. The MCMC chains did not converge for
-\modelsNoConvergeAROne\ populations according to our criteria ($\widehat{R} <
-1.05, n_\mathrm{eff} > 200$) after 8000 iterations of four chains. This
-  included only \modelsNoConvergeAROneHeavyBase\ populations in which Pr($\nu
-  < 10$) $> 0.5$ categorized them as heavy in the main Gompertz model. We did
-  not include these models in Fig.~\ref{fig:alt}.
-
-\subsection{Assumed density independence}\label{assumed-density-independence}
-
-We fit a simplified version of the Gompertz model in which the density
-dependence parameter $b$ was fixed at $1$ (density independent). This is
-equivalent to fitting a random walk model (with drift) to the $\log$
-abundances or assuming the growth rates are drawn from a stationary
-distribution. The model was as follows:
-\begin{align*}
-x_t &= \lambda + x_{t-1} + \epsilon_t\\
-\epsilon &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma).
-\end{align*}
-We fit this model for three reasons: (1) it is computationally simpler and so
-provides a check that our more complicated full Gompertz model was obtaining
-reasonable estimates of $\nu$, (2) it provides a test of whether density
-dependence was systematically affecting our perception of heavy tails, (3) it
-matches how some previous authors have modelled heavy tails without accounting
-for density dependence \citep{segura2013}.
-
-\subsection{Assumed observation error}
-
-Observation error can bias parameter estimates \citep[e.g.][]{knape2012} and
-is known to affect the ability to detect extreme events \citep{ward2007}. In
-our main analysis, we fit a model that ignored observation error. One way to
-account for observation error would be to fit a full state-space model that
-simultaneously estimates the magnitude of process noise and observation error.
-However, simultaneously estimating observation and process noise is a
-challenging problem (e.g.\ because the observation and process noise
-parameters tend to negatively covary in model fitting) and is known to
-sometimes result in identifiability issues with the Gompertz population model
-\citep{knape2008}. Furthermore, our model was applied to hundreds of time
-series, often of short length (as few as 20 time steps) and our model
-estimates an additional parameter---the shape of the process deviation
-tails---potentially making identifiability and computational issues even
-greater. Therefore, we considered a version of the base Gompertz model that
-allowed for a fixed level of observation error:
-\begin{align*}
-U_t &= \lambda + b U_{t-1} + \epsilon_t\\
-x_t &\sim \mathrm{Normal}(U_t, \sigma_\mathrm{obs}^2)\\
-\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma_\mathrm{proc}),
-\end{align*}
-where $U$ represents the unobserved state vector, $\sigma_\mathrm{obs}$
-represents the standard deviation of observation error (on a log scale), and
-$\sigma_\mathrm{proc}$ represent the process noise scale parameter. We set
-$\sigma_\mathrm{obs}$ to $0.2$, which represents the upper limit of values
-often used in simulation analyses \citep[e.g.][]{valpine2002, thorson2014b}.
-
-\subsection{Ricker-logistic}
-
-We also fit a Ricker-logistic model:
-\begin{align*}
-x_t &= x_{t-1} + r_{\mathrm{max}}\left(1 - \frac{N_{t-1}}{K}\right) + \epsilon_t\\
-\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma),
-\end{align*}
-where $r_\mathrm{max}$ represents the theoretical maximum population growth
-rate that is obtained when $N_t$ (abundance at time $t$) $= 0$. The parameter
-$K$ represents the carrying capacity and, as before, $x_t$ represents the
-$\log$ transformed abundance at time $t$. The Ricker-logistic model assumes a
-linear decrease in population growth rate with increases in abundance. In
-contrast, the Gompertz model assumes a linear decrease in population growth
-rate with increases in \textit{log} abundance ($x_t$)
-\citep[e.g.][]{thibaut2012}.
-
-To fit the Ricker-logistic models, we chose a prior on $K$ uniform between
-zero and twice the maximum observed abundance (\citet{clark2010} chose uniform
-between zero and maximum observed, which is more informative). We set the
-prior on $r_\mathrm{max}$ as uniform between 0 and 20 as in \citet{clark2010}.
-We used the same priors on $\nu$ and $\sigma$ as in the Gompertz model.
-
-\section{Implementation of skewed Student t distribution in Stan}
+\section{Implementation of the skewed Student-t distribution in Stan}
 
 The probability density for the skewed Student t distribution described in the
-methods section translates to the log probability:
+Methods section translates to the log probability:
 
@@
 
-Since Stan only needs the probability density up to an additive constant, we
+Since Stan only needs the probability density up to an additive
+constant \citep{stan-manual2014}, we
 can replace $\log(2 \gamma)$ in the above for $\log(\gamma)$. The final log
 probability density for the skewed Student's t distribution implemented in
-Stan\footnote{See the Stan users mailing list:
+Stan\footnote{Credit to Stan developer Bob Carpenter:
 \url{https://groups.google.com/d/msg/stan-users/jeZQnQRUsDs/BeqhicWm1GEJ}}
@@
 
-\section{Simulation testing the model}
+\section{Heavy-tailed Gompertz model simulations}
 
-We performed two types of simulation testing. First, we tested how easily the
-Student-t distribution $\nu$ parameter could be recovered given different true
-values of $\nu$ and different sample sizes. Second, we tested the ability of
-the heavy-tailed Gompertz model to obtain unbiased parameter estimates of
-$\nu$ given that a set of process deviations was provided in which the
-effective $\nu$ value was close to the true $\nu$ value.
-
-We separated our simulation into these two components to avoid confounding two
-issues. (1) With smaller sample sizes, there may not be a stochastic draw from
-the tails of a distribution. In that case, no model, no matter how perfect,
-will be able to detect the shape of the tails. (2) Complex models may return
-biased parameter estimates if there are conceptual, computational, or coding
-errors. Our first simulation tested the first issue and our second simulation
-tested the latter. In general, our simulations show that, if anything, our
-model under predicts the magnitude and probability of heavy tailed
-events---especially given the length of the time series in the GPDD.
-
-\subsection{Estimating $\nu$ from a stationary t distribution}
-
-First, we tested the ability to estimate $\nu$ given different true values of
-$\nu$ and different sample sizes. We took stochastic draws from t
-distributions with different $\nu$ values ($\nu = 3, 5, 10,$ and $10^6$
-[$\approx$ normal]), with central tendency parameters of $0$, and scale
-parameters of $1$. We started with $1600$ stochastic draws and then fit the
-models again at the first $800, 400, 200, 100, 50,$ and $25$ draws. Each time
-we recorded the posterior samples of $\nu$.
-
-In the following image, the red lines indicate the true population value. Dots
-represent the median estimate the lines represent the 80\% credible interval.
-When a small number of samples are drawn there may not be samples sufficiently
-far into the tails to recapture the true $\nu$ value; however, heavy tails are
-still distinguished from normal tails in most cases, even with only 25 or 50
-samples:
-
-\begin{center}
-\includegraphics[width=0.8\textwidth]{../analysis/t-dist-sampling-sim-prior-exp0point01.pdf}
-\includegraphics[width=0.8\textwidth]{../analysis/t-dist-sampling-sim-sigma-prior-exp0point01.pdf}
-\end{center}
-
-We found that we could consistently and precisely recover median posterior
-estimates of $\nu$ near the true value of $\nu$ with large samples ($\ge 200$)
-(upper panels). At smaller samples we could still usually qualitatively
-distinguish heavy from not-heavy tails, but the model tended to underestimate
-how heavy the tails were. At the same time, at smaller sample sizes, the model
-tended to overestimate how large the scale parameter was (lower panels).
-
-\subsection{Heavy-tailed Gompertz model simulations}
-
-In the second part of our simulation testing, we tested the ability of the
-heavy-tailed Gompertz model to obtain unbiased parameter estimates when the
-process noise was chosen so that appropriate tail events were present. To
-generate these process deviations for the $\nu = 3$ and $\nu = 5$ scenarios,
-we repeatedly drew proposed candidate process deviations and estimated the
-central tendency, scale, and $\nu$ values each time. We recorded when
-$\hat{\nu}$ (median of the posterior) was within $0.2$ CVs (coefficient of
-variations) of the true $\nu$ value and used this set of random seed values in
-our Gompertz simulation. The following simplified R code illustrates this
-procedure (the actual code is available at
+The following pseudo R code illustrates the procedure to generate process
+noise with effective $\nu$ values within a CV of 0.2 of the specified value
+(the actual code is available at
 \url{https://github.com/seananderson/heavy-tails}):
@@
 
-We then fit our Gompertz models to the simulated datasets with all parameters
-(except $\nu$) set near the median values estimated in the GPDD. We repeated
-this with $50$ and $100$ samples without observation error, $50$ samples with
-observation error ($\sigma_\mathrm{obs} = 0.2$), and $50$ samples with the
-same observation error and a Gompertz model that allowed for correctly
-specified observation error magnitude. Our results indicate that the Gompertz
-model can recapture the true value of $\nu$ when the process noise was chosen
-so that appropriate tail events were present (Fig.~\ref{fig:sim-prob} upper
-panels). The addition of observation error caused the model to tend to
-underestimate the degree of heavy-tailedness. Fitting a model with correctly
-specified observation error did not make substantial improvements to model
-bias (Fig.~\ref{fig:sim-prob}).
-
-When converting the posterior distributions of $\nu$ into Pr($\nu < 10$), the
-models distinguished heavy and not-heavy tails reasonably well
-(Fig.~\ref{fig:sim-prob} lower panels). Without observation error, and using a
-probability of $0.5$ as a threshold, the model correctly classified all
-simulated systems with normally distributed process noise as not heavy tailed.
-The model would have miscategorized only one of $40$ simulations at $\nu = 5$
-across simulated populations with $50$ or $100$ time steps
-(Fig.~\ref{fig:sim-prob}, scenarios 1 and 2 in lower row, second panel from
-left). The model would have correctly categorized all cases where the process
-noise was not heavy tailed (Fig.~\ref{fig:sim-prob} bottom-right panel) and
-all cases where $\nu = 3$ and there was not observation error. With $0.2$
-standard deviations of observation error, the model still categorized
-\obsErrorNuFivePerc\% of cases as heavy tailed when $\nu = 5$ and all but one
-case when $\nu = 3$. Allowing for observation error made little improvement to
-the detection of heavy tails. Therefore, we chose to focus on the simpler
-model without observation error in the main text, particularly given that the
-true magnitude of observation error was unknown in the empirical data.
-
-In the following image, upper panels show the distribution of median
-$\widehat{\nu}$ across 20 simulation runs. Lower panels show the distribution
-of Pr($\nu < 10$) across 20 simulation runs. We ran the simulations across
-three population (``true'') $\nu$ values (3, 5, and $1\cdot 10^9$, i.e.\
-approximately normal) and four scenarios: (1) 100 time steps and no
-observation error, (2) 50 time steps and no observation error, (3) 50 time
-steps and observation error drawn from $\mathrm{Normal} (0, 0.2^2)$ but
-ignored, and (4) 50 time steps with observation error in which the quantity of
-observation error was assumed known. Within each scenario the dots represent
-stochastic draws from the true population distributions combined with model
-fits. Underlayed boxplots show the median, interquartile range, and $1.5$
-times the interquartile range:
-
-%\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=\textwidth]{../analysis/sim-gompertz-median-dist.pdf}
-\includegraphics[width=\textwidth]{../analysis/sim-gompertz-p10.pdf}
-%\label{fig:sim-prob}
-\end{center}
-%\end{figure}
-
+%\clearpage
 \section{Additional acknowledgements}
@@
 \LTcapwidth=\textwidth
-%\singlespacing
+\singlespacing
 \begin{footnotesize}
@@
 \end{footnotesize}
-%\onehalfspacing
+\onehalfspacing
 

### 2015-07-22 text/ms.tex
Text edits (and unwrap)

@@
 We find strong evidence for black-swan events, but they are rare, occurring in 3--5\% of populations: most frequently for birds (4--8\%), mammals (4--6\%), and insects (2--3\%).
-Black swan events were predominantly (87\%) downward, implying that unexpected population crashes occur far more frequently than unexpected population increases.
+Black swan events were predominantly (\percBSDown \%) downward, implying that unexpected population crashes occur far more frequently than unexpected population increases.
 Black swan events are not explained by any life-history covariates, but tend to be driven by external perturbations such as climate or severe winters.
@@
 Our results show that accurate forecasts of extinction risk require accounting for the possibility of black swans and the predominance of crashes over increases, or predicted abundance will underestimate the probability of crashes by a factor of XX--XX.
-In addition, the presence of black swans highlights the importance of developing flexible and robust conservation and management strategies, particularly as the frequency and magnitude of climate extremes increase over the next century.
+The presence of black swans highlights the importance of developing flexible and robust conservation and management strategies, particularly as the frequency and magnitude of climate extremes increase over the next century.
 
@@
 
-Here, we assess the frequency and magnitude of black-swan dynamics across time series of 609 populations from a wide array of taxonomic groups---mostly birds, mammals, insects, and fishes.
+Here, we assess the frequency and magnitude of black-swan dynamics across time series of 609 populations from a wide array of taxonomic groups---mostly birds, mammals, insects, and fishes (Extended Data Table~1).
 We then identify characteristics of time series or life-history traits associated with the detection of black-swan events and verify known causes.
@@
 
-Specifically, we fit population dynamics models in which the process noise is drawn from a Student-t distribution\emph{.} By estimating the degrees of freedom parameter, \(\nu\), we can estimate the degree to which the process deviations have heavy tails and are thereof evidence of black-swan events (Fig.~1).
+Specifically, we fit population dynamics models in which the process noise is drawn from a Student-t distribution.
+By estimating the degrees of freedom parameter, \(\nu\), we can estimate the degree to which the process deviations have heavy tails and are thereof evidence of black-swan events (Fig.~1).
 For example, at \(\nu = 3\), the probability of drawing a value more than five standard deviations below the mean is \(0.02\), whereas the probability of drawing such a value from a normal distribution is tiny (\(2.9\cdot10^-7\)).
 As \(\nu\) approaches infinity, the distribution approaches the normal distribution (Fig.~1).
+TODO CITE Extended Data Fig.~\ref{fig:priors}
 
-We found strong, but rare, evidence for black-swan dynamics: most frequently for birds (7\%) followed by mammals (5\%), and insects (3\%) (Fig.~2, Table~S2).
+We found strong, but rare, evidence for black-swan dynamics: most frequently for birds (7\%) followed by mammals (5\%), and insects (3\%) (Fig.~2).
 Black-swan events were taxonomically widespread, occurring in 38\% of taxonomic orders.
@@
 The majority of our heavy-tailed estimates were robust to alternative population models, observation error, and choice of Bayesian priors.
-Our conclusions were not systematically altered when we included an autocorrelation structure in the residuals, modelled population growth rates with or without density dependence, or modelled the population dynamics as Ricker-logistic instead of Gompertz (Fig.~S3).
-Introducing moderate observation error (CV = 0.2) slightly decreased the estimated prevalence of black-swan events (Fig.~S3).
-The strength of the prior on ν had little influence on estimates of black-swan dynamics (Fig.~S4).
-Our simulation testing shows that, if anything, our models underpredict the true magnitude and probability of heavy tailed events---especially given the length of the time series in our analysis (Figs S5, S6).
+Our conclusions were not systematically altered when we included an autocorrelation structure in the residuals, modelled population growth rates with or without density dependence, or modelled the population dynamics as Ricker-logistic instead of Gompertz (Extended Data Fig.~\ref{fig:alt}).
+Introducing moderate observation error (CV = 0.2) slightly decreased the estimated prevalence of black-swan events (Extended Data Fig.~\ref{fig:alt}).
+The strength of the prior on \(\nu\) had little influence on estimates of black-swan dynamics (Extended Data Fig.~\ref{fig:alt-priors}).
+Our simulation testing shows that, if anything, our models underpredict the true magnitude and probability of heavy tailed events---especially given the length of the time series in our analysis (Extended Data Figs \ref{fig:sim-nu}, \ref{fig:sim-prob}).
 
-Across populations, the probability of observing black-swan dynamics was positively related to time-series length and negatively related to magnitude of process noise but not clearly related to population growth rate, density dependence, or maximum lifespan (Figs 3b, 4).
+Across populations, the probability of observing black-swan dynamics was positively related to time-series length and negatively related to magnitude of process noise but not clearly related to population growth rate, density dependence, or maximum lifespan (Fig.~3b, Extended Data Fig.~\ref{fig:correlates}).
 Longer time-series length was the strongest covariate of observing black-swan dynamics.
 For instance, the expected probability density suggesting black-swan dynamics was about 1.6 times greater for a population with 60 time steps compared to one with 30 time steps.
-However, the absolute change in probability with increased time series length was small (0.20 vs.~0.13 in this example, Fig.~4).
+However, the absolute change in probability with increased time series length was small (0.20 vs.~0.13 in this example, Extended Data Fig.~\ref{fig:correlates}).
 
-We examined all time series with published explanations of why black-swan events occurred (Table~S2).
-The majority (87\%) were downward black swans and involved a combination of multiple factors.
+We examined all time series with published explanations of why black-swan events occurred (Extended Data Table~2).
+The majority (\percBSDown \%) were downward black swans and involved a combination of multiple factors.
 For example, a synchronization of environmental- and predation-mediated population cycles is thought to have caused a downward black-swan event for a water vole (\emph{Arvicola terrestris}) population\cite{saucy1994}.
@@
 This could occur with synergistic interactions\cite{kirby2009} or even if non-synergistic forces experience a rare alignment\cite{denny2009}.
-
-There are a number of caveats when considering the generality of our results.
-The GPDD data represent a taxonomically and geographically biased sample of populations---the longer time series we focus on are dominated by commercially and recreationally important species and a disproportionate number of populations are located in the United Kingdom.
-Although we would expect to find qualitatively similar evidence for black swans in other large taxonomic or geographic samples of populations, the common forces driving those black swans would likely differ.
-Additionally, some apparent black-swan events could be recording mistakes, or conversely, some extreme observations may have been discarded or altered if they were erroneously suspected of being recording mistakes.
-Indeed, we discarded three of the populations that our method initially identified as heavy tailed because they turned out to be data-entry errors (Supporting Information).
-Finally, the temporal scales of observation and population dynamics vary considerably across populations in the GPDD and these likely influence the detection of heavy tails.
-For example, if we make frequent observations relative to generation time (e.g.~for many large-bodied mammals) we will average across generations and perhaps miss black-swan events.
-Conversely, if we census populations infrequently relative to generation time (e.g.~many insects in the GPDD) the recorded data may average across extreme and less-extreme events and also dampen black-swan dynamics.
 
@@
 Our interpolation affected only \(\sim\) \interpPointsPerc \% of the final data points (Extended Data Table~1) and none of the data points that were later considered black-swan events.
-Our final dataset contained \NPops\ populations across \NOrders\
-taxonomic orders and seven taxonomic classes, with a median of \medianTimeSteps\ time steps (range of \minTimeSteps--\maxTimeSteps) (Extended Data Table~1).
+Our final dataset contained \NPops\ populations across \NOrders\ taxonomic orders and seven taxonomic classes, with a median of \medianTimeSteps\ time steps (range of \minTimeSteps--\maxTimeSteps) (Extended Data Table~1).
 
@@
 When categorizing a population as heavy or normal tailed, we applied a 0.5 probability threshold.
-In the scenario where the data are uninformative about heavy tails (e.g.~Fig.~1e, h), the posterior will approximately match the prior (prior median \(= 71\), mean \(= 102\)) and a metric of \(Pr(\nu < 10) > 0.5\)) is unlikely to flag the population as heavy tailed.
+In the scenario where the data are uninformative about heavy tails (e.g.~Fig.~1e, h), the posterior will approximately match the prior (prior median \(= 71\), mean \(= 102\)) and a metric of Pr\((\nu < 10) > 0.5\)) is unlikely to flag the population as heavy tailed.
 
@@
 
-\textbf{Skewed Student-t distribution} We fit Gompertz models with skewed Student-t distributed process noise based on the probability density for the skewed t distribution described in Fernandez and Steel\cite{fernandez1998}.
-Their definition equates to the following probability density as implemented in King\cite{king2012}:
-
-\begin{equation}
-  \mathrm{Skewed\,Student\mhyphen t}(x) =
-\begin{dcases}
-\frac{2}{(\gamma + 1/\gamma)}
-         \mathrm{Student\mhyphen t}(x \cdot \gamma, \nu, \mu \cdot \gamma, \sigma),& \text{for } x < 0\\
-       \frac{2}{(\gamma + 1/\gamma)}
-         \mathrm{Student\mhyphen t}(x / \gamma, \nu, \mu / \gamma, \sigma),& \text{for } x \ge 0
-\end{dcases}
-\end{equation}
-
-Here \(\gamma\) represents the skewness parameter.
-The Student-t(\(\cdot\)) portion represents in order: the data point of interest, the heavy-tailedness parameter, the location parameter, and the scale parameter as passed to the standard Student-t probability density.
-The distribution is symmetrical if \(\gamma = 1\), left skewed (skewed downwards in abundance) if \(0 < \gamma < 1\), and right skewed (skewed upwards in abundance) if \(1 < \gamma < \infty\).
-See the Supporting Information for details on the implementation of the skewed Student-t distribution in Stan.
+There are a number of caveats when considering the generality of our results.
+The GPDD data represent a taxonomically and geographically biased sample of populations---the longer time series we focus on are dominated by commercially and recreationally important species and a disproportionate number of populations are located in the United Kingdom.
+Although we would expect to find qualitatively similar evidence for black swans in other large taxonomic or geographic samples of populations, the common forces driving those black swans would likely differ.
+Additionally, some apparent black-swan events could be recording mistakes, or conversely, some extreme observations may have been discarded or altered if they were erroneously suspected of being recording mistakes.
+Indeed, we discarded three of the populations that our method initially identified as heavy tailed because they turned out to be data-entry errors (Supporting Information).
+Finally, the temporal scales of observation and population dynamics vary considerably across populations in the GPDD and these likely influence the detection of heavy tails.
+For example, if we make frequent observations relative to generation time (e.g.~for many large-bodied mammals) we will average across generations and perhaps miss black-swan events.
+Conversely, if we census populations infrequently relative to generation time (e.g.~many insects in the GPDD) the recorded data may average across extreme and less-extreme events and also dampen black-swan dynamics.
 
@@
 Note that the base and weaker priors are relatively flat within the region of \(\nu < 20\), which is the region we are mostly concerned about when categorizing populations as heavy- or thin-tailed.
-TODO - note Gelman 1/x prior.
+We furthermore considered a Uniform(\(0, 1\)) prior on \(1/\nu\) as used in Gelman \emph{et al.}\cite{gelman2014}, but this prior puts considerably more weight on lower (heavy-tailed) values of \(\nu\) than the truncated exponential priors considered here (not shown).
 
@@
 
-\textbf{Alternative population models} We fit alternative population models to test if four key phenomena systematically changed our conclusions The range of percentages of black swans by taxonomic class cited in the abstract are based on lower and upper limits across our main Gompertz model and these four alternative models.
+\textbf{Alternative population models} We fit alternative population models to test if four key phenomena systematically changed our conclusions.
+The range of percentages of black swans by taxonomic class cited in the abstract are based on lower and upper limits across our main Gompertz model and these four alternative models.
 
@@
 We found that we could consistently and precisely recover median posterior estimates of \(\nu\) near the true value of \(\nu\) with large samples (\(\ge 200\)) (Extended Data Fig.~\ref{fig:sim-nu} upper panels).
-At smaller samples we could still usually qualitatively distinguish heavy from not-heavy tails, but the model tended to underestimate how heavy the tails were.
+At smaller samples we could usually qualitatively distinguish heavy from not-heavy tails, but the model tended to underestimate how heavy the tails were.
 At the same time, at smaller sample sizes, the model tended to overestimate how large the scale parameter was (Extended Data Fig.~\ref{fig:sim-nu} lower panels).
@@
 \textbf{Covariates of heavy-tailed dynamics} We fit a multilevel beta regression model to the predicted probability of heavy tails, Pr(\(\nu < 10\)), to investigate potential covariates of heavy-tailed dynamics.
+We obtained maximum lifespan and body-size data from Brook \emph{et al.}\cite{brook2006a}.
 The beta distribution is useful when response data range on a continuous scale between zero and one\cite{ferrari2004}.
@@
 Compared to normal priors, the Cauchy priors concentrate more probability density around expected parameter values while allowing for a higher probability density far into the tails, thereby allowing the data to dominate the posterior more strongly if it disagrees with the prior.
-Our conclusions were not qualitatively changed by using uniform priors.
-We fit our models with 5000 total iterations per chain, 2500 warm-up iterations, four chains, and discarding every second sample to save memory.
+We fit our models with 5000 total iterations per chain, 2500 warm-up iterations, four chains, and discarding every second sample.
 We checked for chain convergence visually and with the same criteria as before (\(\widehat{R} < 1.05\) and \(n_\mathrm{eff} >200\) for all parameters).
 
-To derive taxonomic-order-level estimates of the probability of heavy tails accounting for time-series length (Fig.~3a), we fit a separate multilevel model with the same structure but with only \(\log\) time-series length as a predictor.
-(In this case, we did not want to control for intrinsic population characteristics such as density dependence.) Since our predictors were centered by subtracting their mean value, we obtained order-level estimates of the probability of heavy tails at mean log time-series length by adding the posteriors for \(\alpha\), \(\alpha^\mathrm{class}_j\), and \(\alpha^\mathrm{order}_k\).
+To derive taxonomic-order-level estimates of the probability of heavy tails accounting for time-series length (Fig.~3a), we fit a separate multilevel model with the same structure but with only \(\log\) time-series length as a predictor---in this case, we did not want to control for intrinsic population characteristics such as density dependence.
+Since our predictors were centered by subtracting their mean value, we obtained order-level estimates of the probability of heavy tails at mean log time-series length by adding the posteriors for \(\alpha\), \(\alpha^\mathrm{class}_j\), and \(\alpha^\mathrm{order}_k\).
+
+\textbf{Skewed Student-t forecasts} To formally evaluate the apparent skewness of heavy-tailed process noise we fit Gompertz models with skewed Student-t distributed process noise based on the probability density for the skewed t distribution described in Fernandez and Steel\cite{fernandez1998}.
+Their definition equates to the following probability density as implemented in King\cite{king2012}:
+
+\begin{equation}
+  \mathrm{Skewed\,Student\mhyphen t}(x) =
+\begin{dcases}
+\frac{2}{(\gamma + 1/\gamma)}
+         \mathrm{Student\mhyphen t}(x \cdot \gamma, \nu, \mu \cdot \gamma, \sigma),& \text{for } x < 0\\
+       \frac{2}{(\gamma + 1/\gamma)}
+         \mathrm{Student\mhyphen t}(x / \gamma, \nu, \mu / \gamma, \sigma),& \text{for } x \ge 0
+\end{dcases}
+\end{equation}
+
+Here \(\gamma\) represents the skewness parameter.
+The Student-t(\(\cdot\)) portion represents in order: the data point of interest, the heavy-tailedness parameter, the location parameter, and the scale parameter as passed to the standard Student-t probability density.
+The distribution is symmetrical if \(\gamma = 1\), left skewed (skewed downwards in abundance) if \(0 < \gamma < 1\), and right skewed (skewed upwards in abundance) if \(1 < \gamma < \infty\).
+See the Supporting Information for details on the implementation of the skewed Student-t distribution in Stan.
+
+TODO \ldots{} we fit these\ldots{} and compare 5-year forecasts to standard Gompertz with normal error.
+We project all with median nu \textless{} X and calculate the ratio of the lower 99\% credible interval between the two models..
+increased number of samples because getting far into tails, aggregate posterior.
 
@@
 
-\textbf{Extended Data Table~1: Summary statistics for the filtered Global Population Dynamics Database
-time series arranged by taxonomic class.} Columns are: number of populations, number of taxonomic orders, numbers of species, median time series length, total number of interpolated time steps, total number of substituted zeros, and total number of time steps.
+\textbf{Extended Data Table~1: Summary statistics for the filtered Global Population Dynamics Database time series arranged by taxonomic class.} Columns are: number of populations, number of taxonomic orders, numbers of species, median time series length, total number of interpolated time steps, total number of substituted zeros, and total number of time steps.
 
@@
 \begin{longtable}{>{\RaggedRight}m{1.4cm}>{\RaggedRight}p{6.2cm}>{\RaggedRight}p{0.7cm}>{\RaggedRight}p{0.4cm}>{\RaggedRight}p{4.4cm}>{\RaggedRight}p{1.4cm}}
-\caption{\textbf{Populations with a high probability of heavy-tailed
-dynamics in the base heavy-tailed Gompertz population dynamics model.} Shown
-are the log abundance time series, population descriptions, Global Population
-Dynamics Database Main IDs, citation for the data source or separate
-verification literature, a description of the cause of the black swan events
-(if known), the probability of heavy tails as calculated by our model, and
-median estimate of $\nu$ from our model with 90\% quantile credible intervals
-indicated in parentheses. Red dots on the time series indicate downward
-black-swan events and blue values indicate upward black-swan events that have a
-$10^{-4}$ probability or less of occurring if the population dynamics were
-explained by a Gompertz model with normally distributed process noise with a
-standard deviation equal to the scale parameter in the fitted t
-distribution.}\\
+\caption{\textbf{Populations with a high probability of heavy-tailed dynamics in the base heavy-tailed Gompertz population dynamics model.} Shown are the log abundance time series, population descriptions, Global Population Dynamics Database Main IDs, citation for the data source or separate verification literature, a description of the cause of the black swan events (if known), the probability of heavy tails as calculated by our model, and median estimate of $\nu$ from our model with 90\% quantile credible intervals indicated in parentheses.
+Red dots on the time series indicate downward black-swan events and blue values indicate upward black-swan events that have a $10^{-4}$ probability or less of occurring if the population dynamics were explained by a Gompertz model with normally distributed process noise with a standard deviation equal to the scale parameter in the fitted t distribution.}\\
 
@@
 
-\caption{
-  \textbf{Probability density of the Bayesian priors for the Gompertz models.} From
-  left to right and then top to bottom: (1) per capita growth rate at
-  $\log$(abundance) = $0$: $\lambda \sim \mathrm{Normal}(0, 10^2)$; (2) scale
-  parameter of t-distribution process noise: $\sigma \sim \mathrm{Half\mhyphen
-    Cauchy} (0, 2.5)$; (3) t-distribution degrees of freedom parameter: $\nu
-  \sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2)$; (4)
-  AR1 correlation coefficient of residuals: $\phi \sim \mathrm{Truncated
-    \mhyphen Normal}(0, 1, \mathrm{min.} = -1, \mathrm{max.} = 1)$. Not shown
-  is $b$, the density dependence parameter: $b \sim \mathrm{Uniform}(-1, 2)$.
-  The $\nu$ panel also shows two alternative priors: a weaker prior $\nu \sim
-  \mathrm{Truncated\mhyphen Exponential}(0.005, \mathrm{min.} = 2)$, and a
-  stronger prior $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.02,
-  \mathrm{min.} = 2)$. The inset panel shows the same data but with a
-  log-transformed x axis. Note that the base and weaker priors are relatively
-  flat within the region of $\nu < 20$ that we are concerned with. }
+\caption{\textbf{Probability density of the Bayesian priors for the Gompertz models.} From left to right and then top to bottom: (1) per capita growth rate at $\log$(abundance) = $0$: $\lambda \sim \mathrm{Normal}(0, 10^2)$; (2) scale parameter of t-distribution process noise: $\sigma \sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)$; (3) t-distribution degrees of freedom parameter: $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2)$; (4) AR1 correlation coefficient of residuals: $\phi \sim \mathrm{Truncated \mhyphen Normal}(0, 1, \mathrm{min.} = -1, \mathrm{max.} = 1)$.
+Not shown is $b$, the density dependence parameter: $b \sim \mathrm{Uniform}(-1, 2)$.
+The $\nu$ panel also shows two alternative priors: a weaker prior $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.005, \mathrm{min.} = 2)$, and a stronger prior $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.02, \mathrm{min.} = 2)$.
+The inset panel shows the same data but with a log-transformed x axis.
+Note that the base and weaker priors are relatively flat within the region of $\nu < 20$ that we are concerned with.
+}
 
@@
 
-\caption{\textbf{Estimates of $\nu$ from alternative models
-  plotted against the base Gompertz model estimates of $\nu$.} Shown are
-  medians of the posterior (dots) and 50\% credible intervals (segments). The
-  diagonal line indicates a one-to-one relationship. Different colours
-  indicate various taxonomic classes. The grey-shaded regions indicate regions
-  of disagreement if $\nu = 10$ is taken as a threshold of heavy-tailed
-  dynamics. The Gompertz observation error model assumes a fixed standard
-  deviation of observation error of $0.2$ on a log scale.}
+\caption{\textbf{Estimates of $\nu$ from alternative models plotted against the base Gompertz model estimates of $\nu$.} Shown are medians of the posterior (dots) and 50\% credible intervals (segments).
+The diagonal line indicates a one-to-one relationship.
+Different colours indicate various taxonomic classes.
+The grey-shaded regions indicate regions of disagreement if $\nu = 10$ is taken as a threshold of heavy-tailed dynamics.
+The Gompertz observation error model assumes a fixed standard deviation of observation error of $0.2$ on a log scale.}
 
@@
 
-\caption{\textbf{Estimates of $\nu$ from Gompertz models with alternative priors on
-  $\nu$.} Shown are medians of the posterior (dots) and 50\% credible intervals
-  (segments). The diagonal line indicates a one-to-one relationship. Different
-  colours indicate various taxonomic classes. The grey-shaded regions indicate
-  regions of disagreement if $\nu = 10$ is taken as a threshold of
-  heavy-tailed dynamics. The base, weaker, and stronger priors on $\nu$ are
-  illustrated in Fig.~\ref{fig:priors}. In general, the estimates are nearly
-  identical in cases where the data are informative about low values of $\nu$.
-  When the data are less informative about low values of $\nu$, the prior can
-  slightly pull the estimates of $\nu$ towards higher or lower values.}
+\caption{\textbf{Estimates of $\nu$ from Gompertz models with alternative priors on $\nu$.} Shown are medians of the posterior (dots) and 50\% credible intervals (segments).
+The diagonal line indicates a one-to-one relationship.
+Different colours indicate various taxonomic classes.
+The grey-shaded regions indicate regions of disagreement if $\nu = 10$ is taken as a threshold of heavy-tailed dynamics.
+The base, weaker, and stronger priors on $\nu$ are illustrated in Fig.~\ref{fig:priors}.
+In general, the estimates are nearly identical in cases where the data are informative about low values of $\nu$.
+When the data are less informative about low values of $\nu$, the prior can slightly pull the estimates of $\nu$ towards higher or lower values.}
 
@@
 
-\caption{\textbf{Potential covariates of heavy-tailed population dynamics (indicated
-  by a high probability that $\nu < 10$).} Shown are (a--c) parameters from the
-  Gompertz heavy-tailed population model ($\sigma$, $\lambda$, $b$), (b)
-  number of time steps, (c) body length, and (d) lifespan. For the Gompertz
-  parameters, $\sigma$ refers to the scale parameter of the Student-t
-  process-noise distribution, $\lambda$ refers to the expected log abundance
-  at the next time step at an abundance of one, $b$ refers to the density
-  dependence parameter ($1$ is maximally density independent, $0$ is maximally
-  density dependent, and $<0$ is inversely density dependent). Circles
-  representing a few sharks, crustaceans, and gastropods are filled in white.
-  Median and 90\% credible interval posterior predictions of a beta regression
-  multilevel model are shown in panels a and d where there was a high
-  probability the slope coefficient was different from zero
-  (Fig.~3b).}
+\caption{\textbf{Potential covariates of heavy-tailed population dynamics (indicated by a high probability that $\nu < 10$).} Shown are (a--c) parameters from the Gompertz heavy-tailed population model ($\sigma$, $\lambda$, $b$), (b) number of time steps, (c) body length, and (d) lifespan.
+For the Gompertz parameters, $\sigma$ refers to the scale parameter of the Student-t process-noise distribution, $\lambda$ refers to the expected log abundance at the next time step at an abundance of one, $b$ refers to the density dependence parameter ($1$ is maximally density independent, $0$ is maximally density dependent, and $<0$ is inversely density dependent).
+Circles representing a few sharks, crustaceans, and gastropods are filled in white.
+Median and 90\% credible interval posterior predictions of a beta regression multilevel model are shown in panels a and d where there was a high probability the slope coefficient was different from zero (Fig.~3b).}
 
@@
 
-\caption{Testing the ability to estimate $\nu$ (top
-  panels) and the scale parameter of the process deviations (bottom panels)
-  for a given number of samples (columns) drawn from a distribution with a
-  given true $\nu$ value (rows). The red lines indicate the true population
-  value. When a small number of samples are drawn there may not be samples
-  sufficiently far into the tails to recapture the true $\nu$ value; however,
-  heavy tails are still distinguished from normal tails in most cases, even
-  with only 25 or 50 samples.}
+\caption{\textbf{Testing the ability to estimate $\nu$ (top panels) and the scale parameter of the process deviations (bottom panels) for a given number of samples (columns) drawn from a distribution with a given true $\nu$ value (rows).} The red lines indicate the true population value.
+When a small number of samples are drawn there may not be samples sufficiently far into the tails to recapture the true $\nu$ value; however, heavy tails are still distinguished from normal tails in most cases, even with only 25 or 50 samples.}
 
@@
 
-\caption{Simulation testing the Gompertz estimation model when the process
-deviation draws were chosen so that $\nu$ could be estimated close to the true
-value outside the full population model (``effective $\nu$'' within a CV of 0.2
-of specified $\nu$). Upper panels show the distribution of median
-$\widehat{\nu}$ across 20 simulation runs. Lower panels show the distribution
-of Pr($\nu < 10$) across 20 simulation runs. We ran the simulations across
-three population (``true'') $\nu$ values (3, 5, and $1\cdot 10^9$, i.e.\
-approximately normal) and four scenarios: (1) 100 time steps and no observation
-error, (2) 50 time steps and no observation error, (3) 50 time steps and
-observation error drawn from $\mathrm{Normal} (0, 0.2^2)$ but ignored, and (4)
-50 time steps with observation error in which the quantity of observation error
-was assumed known. Within each scenario the dots represent stochastic draws
-from the true population distributions combined with model fits. Underlayed
-boxplots show the median, interquartile range, and $1.5$ times the
-interquartile range. }
+\caption{\textbf{Simulation testing the Gompertz estimation model when the process deviation draws were chosen so that $\nu$ could be estimated close to the true value outside the full population model} (``effective $\nu$'' within a CV of 0.2 of specified $\nu$).
+Upper panels show the distribution of median $\widehat{\nu}$ across 20 simulation runs.
+Lower panels show the distribution of Pr($\nu < 10$) across 20 simulation runs.
+We ran the simulations across three population (``true'') $\nu$ values (3, 5, and $1\cdot 10^9$, i.e.\ approximately normal) and four scenarios: (1) 100 time steps and no observation error, (2) 50 time steps and no observation error, (3) 50 time steps and observation error drawn from $\mathrm{Normal} (0, 0.2^2)$ but ignored, and (4) 50 time steps with observation error in which the quantity of observation error was assumed known.
+Within each scenario the dots represent stochastic draws from the true population distributions combined with model fits.
+Underlayed boxplots show the median, interquartile range, and $1.5$ times the interquartile range.}
 

### 2015-07-25 text/supp.tex
Edits to paper

@@
 
-The probability density for the skewed Student t distribution described in the
-Methods section translates to the log probability:
+The probability density for the skewed Student-t distribution \citep{fernandez1998} equates to the following probability density as implemented in \citet{king2012}:
 
-\[
+\begin{equation}
+  \mathrm{Skewed\,Student\mhyphen t}(x) =
+\begin{dcases}
+\frac{2}{(\gamma + 1/\gamma)}
+         \mathrm{Student\mhyphen t}(x \cdot \gamma, \nu, \mu \cdot \gamma, \sigma),& \text{for } x < 0\\
+       \frac{2}{(\gamma + 1/\gamma)}
+         \mathrm{Student\mhyphen t}(x / \gamma, \nu, \mu / \gamma, \sigma),& \text{for } x \ge 0
+\end{dcases}
+\end{equation}
+
+Here $\gamma$ represents the skewness parameter.
+The Student-t($\cdot$) portion represents in order: the data point of interest, the heavy-tailedness parameter, the location parameter, and the scale parameter as passed to the standard Student-t probability density.
+This translates to the log probability density:
+
+\begin{equation}
 \mathrm{Skewed\,Student\mhyphen t}(x) =
@@
 \end{dcases}
-\]
+\end{equation}
 

### 2015-11-24 text/supp.tex
Update before reformat

@@
 
-\title{Supporting Information:\\Evidence for black-swan events in animal
+\title{Supplementary Information:\\Evidence for black-swan events in animal
   populations}
@@
 Sean C. Anderson$^{1,2}$ \and
-Trevor A. Branch$^2$ \and
-Andrew B. Cooper$^3$ \and
+Trevor A. Branch$^3$ \and
+Andrew B. Cooper$^2$ \and
 Nicholas K. Dulvy$^1$
@@
 
+We separated our simulation into these two components to avoid confounding two
+issues. (1) With smaller sample sizes, there may not be a stochastic draw from
+the tails of a distribution. In that case, no model, no matter how perfect,
+will be able to detect the shape of the tails. (2) Complex models may return
+biased parameter estimates if there are conceptual, computational, or coding
+errors. Our first simulation tested the first issue and our second simulation
+tested the latter. In general, our simulations show that, if anything, our
+model under predicts the magnitude and probability of heavy tailed
+events---especially given the length of the time series in the GPDD.
+
+First, we tested the ability to estimate $\nu$ given different true values of
+$\nu$ and different sample sizes. We took stochastic draws from t
+distributions with different $\nu$ values ($\nu = 3, 5, 10,$ and $10^6$
+[$\approx$ normal]), with central tendency parameters of $0$, and scale
+parameters of $1$. We started with $1600$ stochastic draws and then fit the
+models again at the first $800, 400, 200, 100, 50,$ and $25$ draws. Each time
+we recorded the posterior samples of $\nu$.
+
+Second, we tested the ability of the heavy-tailed Gompertz model to obtain
+unbiased parameter estimates when the process noise was chosen so that
+appropriate tail events were present. To generate these process deviations for
+the $\nu = 3$ and $\nu = 5$ scenarios, we repeatedly drew proposed candidate
+process deviations and estimated the central tendency, scale, and $\nu$ values
+each time. We recorded when $\hat{\nu}$ (median of the posterior) was within
+$0.2$ CVs (coefficient of variations) of the true $\nu$ value and used this
+set of random seed values in our Gompertz simulation.
+
 The following pseudo R code illustrates the procedure to generate process
@@
 \end{footnotesize}
+
+We then fit our Gompertz models to the simulated datasets with all parameters
+(except $\nu$) set near the median values estimated in the GPDD. We repeated
+this with $50$ and $100$ samples without observation error, $50$ samples with
+observation error ($\sigma_\mathrm{obs} = 0.2$), and $50$ samples with the
+same observation error and a Gompertz model that allowed for correctly
+specified observation error magnitude. Our results indicate that the Gompertz
+model can recapture the true value of $\nu$ when the process noise was chosen
+so that appropriate tail events were present (Extended Data Fig.
+5 upper panels). The addition of observation error caused the
+model to tend to underestimate the degree of heavy-tailedness. Fitting a model
+with correctly specified observation error did not make substantial
+improvements to model bias (Extended Data Fig. 5).
+
+When converting the posterior distributions of $\nu$ into Pr($\nu < 10$), the
+models distinguished heavy and not-heavy tails reasonably well (Extended Data
+Fig. 5 lower panels). Without observation error, and using a
+probability of $0.5$ as a threshold, the model correctly classified all
+simulated systems with normally distributed process noise as not heavy tailed.
+The model would have miscategorized only one of $40$ simulations at $\nu = 5$
+across simulated populations with $50$ or $100$ time steps (Extended Data Fig.
+5, scenarios 1 and 2 in lower row, second panel from left).
+The model would have correctly categorized all cases where the process noise
+was not heavy tailed (Extended Data Fig. 5 bottom-right
+panel) and all cases where $\nu = 3$ and there was not observation error. With
+$0.2$ standard deviations of observation error, the model still categorized
+\obsErrorNuFivePerc\% of cases as heavy tailed when $\nu = 5$ and all but one
+case when $\nu = 3$. Allowing for observation error made little improvement to
+the detection of heavy tails. Therefore, we chose to focus on the simpler
+model without observation error in the main text, particularly given that the
+true magnitude of observation error was unknown in the empirical data.
 
@@
 \noindent
-Example Stan code for a heavy-tailed Gompertz model with AR1 correlated
-residuals and a specified level of observation error. The specific code for
+Example Stan code for a heavy-tailed Gompertz model. The specific code for
 the various models in our analysis is available at
-\url{https://github.com/seananderson/heavy-tails}.
+\url{https://github.com/seananderson/heavy-tails} (currently private).
 
 %\begin{spacing}{1.15}
+%% \begin{footnotesize}
+%% \begin{verbatim}
+%% data {
+%%   int<lower=3> N;              // number of observations
+%%   vector[N] y;                 // vector to hold ln abundance observations
+%%   real<lower=0> nu_rate;       // rate parameter for nu exponential prior
+%% }
+%% parameters {
+%%   real lambda;                 // Gompertz growth rate parameter
+%%   real<lower=-1, upper=2> b;   // Gompertz density dependence parameter
+%%   real<lower=0> sigma_proc;    // process noise scale parameter
+%%   real<lower=2> nu;            // t-distribution degrees of freedom
+%%   real<lower=-1, upper=1> phi; // AR1 parameter
+%%   vector[N] U;                 // unobserved states
+%%   real<lower=0> sigma_obs;     // specified observation error SD
+%% }
+%% transformed parameters {
+%%   vector[N] epsilon;           // error terms
+%%   epsilon[1] <- 0;
+%%   for (i in 2:N) {
+%%     epsilon[i] <- U[i] - (lambda + b * U[i - 1])
+%%                        - (phi * epsilon[i - 1]);
+%%   }
+%% }
+%% model {
+%%   // priors:
+%%   nu ~ exponential(nu_rate);
+%%   lambda ~ normal(0, 10);
+%%   sigma_proc ~ cauchy(0, 2.5);
+%%   phi ~ normal(0, 1);
+%%   // data model:
+%%   for (i in 2:N) {
+%%     U[i] ~ student_t(nu,
+%%                      lambda + b * U[i - 1]
+%%                      + phi * epsilon[i - 1],
+%%                      sigma_proc);
+%%   }
+%%   y ~ normal(U, sigma_obs);
+%% }
+%% \end{verbatim}
+%% \end{footnotesize}
+%%
+
 \begin{footnotesize}
@@
 data {
-  int<lower=3> N;              // number of observations
-  vector[N] y;                 // vector to hold ln abundance observations
+  int<lower=1> N;              // number of observations
+  vector[N] y;                 // vector to hold log abundance observations
   real<lower=0> nu_rate;       // rate parameter for nu exponential prior
@@
   real<lower=2> nu;            // t-distribution degrees of freedom
-  real<lower=-1, upper=1> phi; // AR1 parameter
-  vector[N] U;                 // unobserved states
-  real<lower=0> sigma_obs;     // specified observation error SD
-}
-transformed parameters {
-  vector[N] epsilon;           // error terms
-  epsilon[1] <- 0;
-  for (i in 2:N) {
-    epsilon[i] <- U[i] - (lambda + b * U[i - 1])
-                       - (phi * epsilon[i - 1]);
-  }
 }
 model {
-  // priors:
   nu ~ exponential(nu_rate);
@@
   sigma_proc ~ cauchy(0, 2.5);
-  phi ~ normal(0, 1);
-  // data model:
   for (i in 2:N) {
-    U[i] ~ student_t(nu,
-                     lambda + b * U[i - 1]
-                     + phi * epsilon[i - 1],
-                     sigma_proc);
+    y[i] ~ student_t(nu, lambda + b * y[i-1], sigma_proc);
   }
-  y ~ normal(U, sigma_obs);
 }

### 2015-11-24 analysis/1-compile-models.R

@@ -225,6 +225,45 @@ model {
 stan_logistic <- stan_model(model_code = stan_model)
 saveRDS(stan_logistic, file = "stan-logistic.rds")
 
+# Logistic-ricker model, alternative parameterization
+stan_model <-
+'data {
+  int<lower=0> N; // rows of data
+  vector[N] y; // log(numbers) at time t
+  real<lower=0> K_upper;
+  real<lower=0> r_upper;
+  real<lower=0> nu_rate; // rate parameter for nu exponential prior
+}
+parameters {
+  real<lower=0, upper=K_upper> K;
+  real<lower=0, upper=r_upper> r;
+  real<lower=0> sigma_proc;
+  real<lower=2> nu;
+}
+model {
+  sigma_proc ~ cauchy(0, 2.5);
+  nu ~ exponential(nu_rate);
+  for (i in 2:N) {
+    y[i] ~ student_t(nu, y[i-1] + r * (1 - exp(y[i-1]) / K), sigma_proc);
+  }
+}
+'
+stan_logistic2 <- stan_model(model_code = stan_model)
+saveRDS(stan_logistic2, file = "stan-logistic2.rds")
+
+## test
+# x <- vector(length = 40)
+# x[1] <- 2
+# r <- 0.6
+# K <- 20
+# for (i in 2:40) {
+#   x[i] <- x[i-1] + r * (1 - exp(x[i-1]) / K) + rnorm(1, 0, 0.2)
+# }
+# model <- readRDS("stan-logistic2.rds")
+# sm <- sampling(model, data = list(N = length(x), y = x, K_upper = exp(max(x)) * 2,
+#   r_upper = 10, nu_rate = 0.01))
+##
+
 # AR1 Gompertz that has been extensively tested
 # coded similarly to the Stan 2.4.0 manual
 stan_model <-
@@ -234,6 +273,7 @@ stan_model <-
   real<lower=0> nu_rate; // rate parameter for nu exponential prior
   real b_lower;
   real b_upper;
+  real phi_sd;
 }
 parameters {
   real lambda;
@@ -254,7 +294,7 @@ model {
   nu ~ exponential(nu_rate);
   lambda ~ normal(0, 10);
   sigma_proc ~ cauchy(0, 2.5);
-  phi ~ normal(0, 1);
+  phi ~ normal(0, phi_sd);
   for (i in 2:N) {
     y[i] ~ student_t(nu,
                      lambda + b * y[i - 1]

### 2015-11-24 analysis/1.5-compile-fit-function.R

@@ -5,7 +5,7 @@ fit_gpdd_model <- function(gpdd_dat, model, sub_folder,
   pars = c("lambda", "sigma_proc", "nu", "b", "phi"), max_rhat_allowed = 1.05,
   min_neff_allowed = 200, iterations = 2000, max_iterations = 8000,
   iteration_increment = 2, warmup = 1000, chains = 4, overwrite = FALSE,
-  .parallel = FALSE, refresh = -1, cores = 2) {
+  .parallel = FALSE, refresh = -1, cores = 2, thin = 1) {
 
   library("rstan")
 
@@ -38,7 +38,7 @@ fit_gpdd_model <- function(gpdd_dat, model, sub_folder,
           iterations <= max_iterations) {
         d <- eval(parse(text = stan_dat))
         # refresh = -1 supresses the running messages
-        sm <- sampling(model, data = d, refresh = refresh,
+        sm <- sampling(model, data = d, thin = thin,
           pars = pars, iter = iterations, chains = chains, warmup = warmup)
         max_rhat <- max(summary(sm)$summary[, "Rhat"])
         min_neff <- min(summary(sm)$summary[, "n_eff"])

### 2015-11-24 analysis/2-run-model-logistic.R

@@ -3,17 +3,20 @@
 source("1.5-compile-fit-function.R")
 source("1.6-extract-function.R")
 
-model <- readRDS("stan-logistic.rds")
+model <- readRDS("stan-logistic2.rds")
 gpdd <- readRDS("gpdd-clean.rds")
 id <- "logistic"
 
-stan_dat <- "list(N = nrow(x) - 1, r_obs = x$r_obs[-1], nu_rate = 0.01, K_upper = 2, r_upper = 10, Nt = x$population_untrans_scaled[-N])"
+stan_dat <- paste("list(N = nrow(x), y = log(x$population_untransformed),",
+  "nu_rate = 0.01, K_upper = 2 * max(x$population_untransformed), r_upper = 10)")
 stan_pars <- c("r", "K", "nu", "sigma_proc")
 
+library("rstan")
+options(mc.cores = 1)
 fit_gpdd_model(gpdd_dat = gpdd, model = model,
-  sub_folder = id, stan_dat = stan_dat, pars = stan_pars)
+  sub_folder = id, stan_dat = stan_dat, pars = stan_pars, root_folder = ".", .parallel = TRUE)
 
 out <- plyr::ldply(unique(gpdd$main_id), extract_model,
-  sub_folder = id, get_phi = FALSE, type = "logistic")
+  sub_folder = id, get_phi = FALSE, type = "logistic", root_folder = ".")
 
 saveRDS(out, file = paste0(id, "-hat.rds"))

### 2015-11-24 analysis/2-run-models-ar1.R

@@ -7,9 +7,27 @@ model <- readRDS("stan-gomp2-ar1.rds")
 gpdd <- readRDS("gpdd-clean.rds")
 id <- "gomp-ar1"
 
-fit_gpdd_model(gpdd_dat = gpdd, model = model, sub_folder = id)
+stan_dat <- "list(N = nrow(x), y = log(x$population_untransformed), nu_rate = 0.01, b_lower = -1, b_upper = 2, phi_sd = 0.5)"
+
+fit_gpdd_model(gpdd_dat = gpdd, model = model, sub_folder = id,
+  iterations = 2000, max_iterations = 128000*2, warmup = 1000, chains = 4, root_folder = ".",
+  stan_dat = stan_dat)
 
 out <- plyr::ldply(unique(gpdd$main_id), extract_model,
-  sub_folder = id, get_phi = TRUE, type = "gompertz")
+  sub_folder = id, get_phi = TRUE, type = "gompertz", root_folder = ".")
+
+library(dplyr)
+ids_non_converged <- filter(out,  max_rhat > 1.05 | min_neff < 200)$main_id
+ids_non_converged
+# gpdd_non_converged <- filter(gpdd, main_id %in% ids_non_converged)
+# gpdd_non_converged <- filter(gpdd, main_id %in% 10144) # example problem population
 
 saveRDS(out, file = paste0(id, "-hat.rds"))
+
+# out <- readRDS("gomp-ar1-hat.rds")
+# out2 <- readRDS("gomp-ar12-hat.rds")
+# ids_non_converged <- filter(out2,  max_rhat > 1.05 | min_neff < 200)$main_id
+# out <- out[!out$main_id %in% out2$main_id, ]
+# out <- bind_rows(out, out2) %>% as.data.frame()
+
+# saveRDS(out, file = "gomp-ar1-extra-iterations-hat.rds")

### 2015-11-24 analysis/2-run-models-obs-0.2.R

@@ -13,10 +13,29 @@ stan_dat <- "list(N = nrow(x), y = log(x$population_untransformed), nu_rate = 0.
 stan_pars <- c("lambda", "sigma_proc", "nu", "b")
 
 fit_gpdd_model(gpdd_dat = gpdd, model = model,
-  sub_folder = id, stan_dat = stan_dat, iterations = 2000, warmup = 1000,
-  max_iterations = 4000, pars = stan_pars)
-
+ sub_folder = id, stan_dat = stan_dat, iterations = 2000, warmup = 1000,
+ max_iterations = 4000, pars = stan_pars, root_folder = ".")
 out <- plyr::ldply(unique(gpdd$main_id), extract_model,
-  sub_folder = id, get_phi = FALSE, type = "gompertz")
-
+ sub_folder = id, get_phi = FALSE, type = "gompertz")
 saveRDS(out, file = paste0(id, "-hat.rds"))
+
+out <- readRDS(paste0(id, "-hat.rds"))
+nc <- subset(out, max_rhat > 1.05 | min_neff < 200)$main_id
+
+id <- "gomp-obs-0.2-2"
+# didn't converge:
+fit_gpdd_model(gpdd_dat = subset(gpdd, main_id %in% nc), model = model,
+  sub_folder = id, stan_dat = stan_dat, iterations = 8000, warmup = 8000/2,
+  max_iterations = 256000, pars = stan_pars, thin = 10, root_folder = ".")
+
+out2 <- plyr::ldply(nc, extract_model,
+  sub_folder = id, get_phi = FALSE, type = "gompertz", root_folder = ".")
+saveRDS(out2, file = paste0(id, "-hat.rds"))
+
+library(dplyr)
+out <- readRDS("gomp-obs-0.2-hat.rds")
+out2 <- readRDS("gomp-obs-0.2-2-hat.rds")
+out <- out[!out$main_id %in% out2$main_id, ]
+out <- bind_rows(out, out2) %>% as.data.frame()
+
+saveRDS(out, file = "gomp-obs-0.2-extra-iterations-hat.rds")

### 2015-11-24 analysis/5-shape-data.R

@@ -11,7 +11,7 @@ gomp_hat_skew <- readRDS("gomp-base-skew-hat.rds")
 gomp_hat_base <- readRDS("gomp-base-hat.rds")
 gomp_hat_logistic <- readRDS("logistic-hat.rds")
 gomp_hat_ar1 <- readRDS("gomp-ar1-hat.rds")
-gomp_hat_obs_0.2 <- readRDS("gomp-obs-0.2-hat.rds")
+gomp_hat_obs_0.2 <- readRDS("gomp-obs-0.2-extra-iterations-hat.rds")
 gomp_hat_rate <- readRDS("rate-hat.rds")
 gomp_hat_weaker <- readRDS("gomp-base-weaker-hat.rds")
 gomp_hat_stronger <- readRDS("gomp-base-stronger-hat.rds")

### 2015-11-24 analysis/5.10-extract-skew-samples.R

@@ -1,5 +1,9 @@
-sample_posterior <- function(path = ".", N = 1000L) {
-  f <- list.files(path, pattern = "*.rds")
+sample_posterior <- function(path = ".", N = 5000L, files = NULL) {
+  if (is.null(files)) {
+    f <- list.files(path, pattern = "*.rds")
+  } else {
+    f <- files
+  }
   samples <- plyr::ldply(f, function(x) {
 
     message(x)
@@ -32,7 +36,22 @@ sample_posterior <- function(path = ".", N = 1000L) {
 }
 
 set.seed(123)
-skew_samples   <- sample_posterior("~/scratch/heavy/gomp-base-skew/")
-normal_samples <- sample_posterior("~/scratch/heavy/gomp-base-normal/")
+skew_samples   <- sample_posterior("~/scratch/heavy/gomp-base-skew/", N = 5000L)
+normal_samples <- sample_posterior("~/scratch/heavy/gomp-base-normal/", N = 5000L)
 saveRDS(skew_samples,   file = "skew_samples.rds")
 saveRDS(normal_samples, file = "normal_samples.rds")
+
+source("5-shape-data.R")
+h1 <- dplyr::filter(gomp_hat_skew, nu_50 < 30)
+h2 <- dplyr::filter(gomp_hat_base, p10 >= 0.5)
+heavy_ids <- union(h1$main_id, h2$main_id)
+write.csv(heavy_ids, file = "heavy_ids.csv", row.names = FALSE)
+
+heavy_ids <- read.csv("heavy_ids.csv")$x
+set.seed(123)
+skew_samples   <- sample_posterior("~/scratch/heavy/gomp-base-skew/", N = 40000L,
+  files = paste0("sm-", heavy_ids, ".rds"))
+normal_samples <- sample_posterior("~/scratch/heavy/gomp-base-normal/", N = 40000L,
+  files = paste0("sm-", heavy_ids, ".rds"))
+saveRDS(skew_samples,   file = "skew_samples_heavy40000.rds")
+saveRDS(normal_samples, file = "normal_samples_heavy40000.rds")

### 2015-11-24 analysis/6-plot-alt-models.R

@@ -4,6 +4,19 @@ gomp_hat_base <- arrange(gomp_hat_base, main_id)
 gomp_hat_obs_0.2 <- arrange(gomp_hat_obs_0.2, main_id)
 gomp_hat_logistic <- arrange(gomp_hat_logistic, main_id)
 gomp_hat_rate <- arrange(gomp_hat_rate, main_id)
+gomp_hat_ar1 <- arrange(gomp_hat_ar1, main_id)
+
+heavy_ids <- filter(gomp_hat_base, p10 >= 0.50)$main_id
+
+check_nc_heavy <- function(dat, heavy_ids) {
+  nc <- filter(dat, max_rhat > 1.05 | min_neff < 200)$main_id
+  x <- nc[nc %in% heavy_ids]
+  if (length(x) > 0) warning(paste(x, collapse = ", "))
+}
+check_nc_heavy(gomp_hat_obs_0.2, heavy_ids)
+check_nc_heavy(gomp_hat_logistic, heavy_ids)
+check_nc_heavy(gomp_hat_rate, heavy_ids)
+check_nc_heavy(gomp_hat_ar1, heavy_ids)
 
 cols_df <- data.frame(col =
     c(RColorBrewer::brewer.pal(4, "Set3")[c(3, 4, 1, 2)],
@@ -15,6 +28,10 @@ cols_df <- data.frame(col =
 comp_panel <- function(xdat, ydat, xlab, ylab, label = "") {
 
   not_converged <- which(ydat$max_rhat > 1.05)
+  if (length(not_converged) > 0) {
+    message(paste(paste(not_converged, collapse = ", "),
+      "didn't converge and will not be plotted"))
+  }
   if(length(not_converged) > 0) {
     xdat <- xdat[-not_converged, ]
     ydat <- ydat[-not_converged, ]

### 2015-11-24 analysis/6-plot-order-correlate-posteriors.R

@@ -86,10 +86,10 @@ x <- rexp(2e6, 0.01)
 x <- x[x > 2]
 prior_p10 <- length(x[x < 10])/length(x)
 
-pdf("order-posteriors-covariates.pdf", width = 3, height = 5.65)
+pdf("order-posteriors-covariates.pdf", width = 3.1, height = 5.65)
 #layout(mat =c(rep(1, 4), rep(2, 9)))
 layout(mat =c(rep(1, 9), rep(2, 4)))
-par(mar = c(4.1,2,1.4,0), oma = c(0.2, 5.2, 0, 0.8),
+par(mar = c(4.1,2,1.4,0), oma = c(0.2, 5.9, 0, 0.8),
   tck = -0.04, mgp = c(2, 0.5, 0), col.axis = "grey25", col = "grey25")
 par(cex = 0.8)
 
@@ -138,14 +138,15 @@ m <- readRDS("beta-stan-samples.rds") # or reload this
 means <- plyr::laply(extract(m), mean)[1:5]
 ord <- order(means)
 
-coefs <- c(expression(lambda), expression(b), expression(log(sigma)),
-  expression(log(Time~steps)), expression(log(Lifespan)))
+coefs <- c(expression(Productivity~(lambda)), expression(Density~dep.~(b)),
+  expression(Process~noise~ln(sigma)),
+  expression(ln(Time~steps)), expression(ln(Lifespan)))
 
 plot(1, 1, xlim = c(-1, 1), ylim = c(0.5, 5.5), main = "", axes = FALSE, yaxs = "i",
   type = "n", yaxs = "i", yaxt = "n", xlab = "", ylab = "")
 scaling_factor <- 8
 
-abline(v = 0, lty = 1, col = "grey60")
+abline(v = 0, lty = 2, col = "grey60")
 j <- 0
 for(i in ord) {
   j <- j + 1

### 2015-11-24 analysis/6-plot-prior.R

@@ -63,7 +63,7 @@ TeachingDemos::subplot({
   y = c(0.01, 0.02))
 
 x <- seq(-1.1, 1.1, length.out = 200)
-plot_prior(x, dtrunc(x, "norm", a = -1, b = 1, mean = 0, sd = 1),
+plot_prior(x, dtrunc(x, "norm", a = -1, b = 1, mean = 0, sd = 0.5),
   expression(phi), xlim = c(-1.1, 1.1), label = "d")
 mtext("Probability density", side = 2, outer = TRUE, line = 1, las = 0, cex = 0.9)
 dev.off()

### 2015-11-24 analysis/6-plot-sparks.R

@@ -13,6 +13,7 @@ heavy <- plyr::join(heavy, as.data.frame(gpdd)) %>%
 get_gomp_res <- function(id_show) {
   p <- subset(gomp_hat_base, main_id == id_show)
   pop <- subset(gpdd, main_id == id_show)$population_untransformed
+  yr <- subset(gpdd, main_id == id_show)$sample_year
   res <- rep(NA, length(pop))
   for(i in 2:length(pop)) {
     res[i] <- log(pop)[i] - (p$lambda_50 + p$b_50 * log(pop[i-1]))
@@ -20,12 +21,12 @@ get_gomp_res <- function(id_show) {
   }
   l <- qnorm(0.0001, 0, sd = p$sigma_proc_50)
   u <- qnorm(0.9999, 0, sd = p$sigma_proc_50)
-  return(list(res = res, l = l, u = u))
+  return(list(res = res, l = l, u = u, yr = yr))
 }
 
 heavy_res <- plyr::ddply(heavy, "main_id", function(x) {
   qq <- get_gomp_res(x$main_id[1])
-  with(qq, data.frame(res, l, u))
+  with(qq, data.frame(res, l, u, yr))
 })
 saveRDS(heavy_res, file = "heavy_residuals.rds")
 
@@ -59,3 +60,17 @@ h <- h[,c("main_id", "p10", "common_name", "taxon_name", "exact_name", "ref", "n
 
 
 write.csv(h, file = "heavy.csv")
+
+heavy_res$bs <- ifelse(heavy_res$res < heavy_res$l | heavy_res$res > heavy_res$u,
+  TRUE, FALSE)
+b <- filter(heavy_res, bs == TRUE, yr > 0)
+b <- b %>% filter(res < l) %>% mutate(bs_frac = res / l)
+ggplot(b, aes(yr, bs_frac)) + geom_point()
+
+yrs <- gpdd %>% group_by(sample_year) %>% summarise(n = n()) %>%
+  filter(sample_year > 0)
+b_yrs <- b %>% group_by(yr) %>% summarise(nbs = n()) %>%
+  rename(sample_year = yr) %>% inner_join(yrs) %>%
+  mutate(perc_bs = nbs / n)
+ggplot(b_yrs, aes(sample_year, perc_bs)) + geom_point()
+

### 2015-11-24 analysis/6-plot-t-nu-eg.R

@@ -89,7 +89,7 @@ text(6, 7.1, quote(nu==3), pos = 4)
 mtext("Simulated population dynamics", side = 3, line = 0, cex = 0.9)
 text(13, -3.9, "Black swans", pos = 4)
 y2 <- make_ts(seed = seed, nu = 5, bs_col = bs_cols[3])
-mtext("log Abundance", side = 2, line = 0.5, las = 0, cex = 0.9)
+mtext("ln Abundance", side = 2, line = 0.5, las = 0, cex = 0.9)
 #text(24, -2.5, "Gompertz\npopulation\nmodel", pos = 4)
 text(1, 7.1, "d", pos = 4, font = 2, cex = 1.4)
 text(6, 7.1, quote(nu==5), pos = 4)

### 2015-11-24 analysis/7-values-for-paper.R

@@ -1,219 +1,0 @@
-# this file produces some values for inclusion of the text of the associated paper
-
-library("dplyr")
-source("5-shape-data.R")
-p_inc <- readRDS("prob_inc_heavy_with_n.rds")
-
-write_tex <- function(x, macro, ...) {
-  out <- paste0("\\newcommand{\\", macro, "}{", x, "}")
-  cat(out, file = zz)
-  cat("\n", file = zz)
-}
-zz <- file("values.tex", "w") # open .tex file to write to throughout
-
-# # what's the median and mean nu in the exponential prior?
-set.seed(1)
-x <- rexp(3e6, 0.01)
-x <- x[x > 2]
-write_tex(round(mean(x), 0) , "basePriorMean")
-write_tex(round(median(x), 0) , "basePriorMedian")
-write_tex(round(100*length(x[x < 10])/length(x), 1) , "basePriorProbHeavy")
-
-# how much was imputed?
-gpdd <- readRDS("gpdd-clean.rds")
-perc_imputed <- gpdd %>% group_by(main_id) %>%
-  summarise(has_imputed = ifelse(max(interpolated) == 1, TRUE, FALSE)) %>%
-  summarise(percent_imputed = sum(has_imputed)/length(has_imputed)) %>%
-  round(2) * 100
-perc_imputed_pops <- as.character(perc_imputed$percent_imputed)
-
-perc_imputed_points <- round(sum(gpdd$interpolated)/nrow(gpdd) * 100, 1)
-
-fig2_pop_n <- filter(gpdd, taxonomic_class %in%
-    c("Aves", "Mammalia", "Insecta", "Osteichthyes")) %>%
-  summarise(n = length(unique(main_id))) %>% summarise(N = sum(n))
-fig2_pop_n <- fig2_pop_n$N
-
-NPops <- length(unique(gpdd$main_id))
-NOrders <- length(unique(gpdd$taxonomic_order))
-NClasses <- length(unique(gpdd$taxonomic_class))
-
-cols_include <- c("taxonomic_class", "main_id", "p10", "nu_50", "type",
-  "max_rhat", "min_neff")
-gomp_hat_base$ type <- "base"
-gomp_hat_ar1$type <- "ar1"
-gomp_hat_logistic$type <- "logistic"
-gomp_hat_obs_0.2$type <- "obs"
-
-gtemp <- rbind(
[…]

### 2015-11-24 analysis/8-sim-extinction-risk.R

@@ -41,7 +41,6 @@ sim_gomp_extinction <- function(y1, abundance_ts = NULL, lambda = 1, b = 0.5, N 
   model = c("heavy-skewd", "normal")) {
 
   y1 <- log(y1)
-
   if (bootstrap_residuals) {
     log_abundance_ts <- log(abundance_ts)
     res <- log_abundance_ts[-1] - (lambda + b * log_abundance_ts[-length(abundance_ts)])
@@ -65,9 +64,13 @@ sim_gomp_extinction <- function(y1, abundance_ts = NULL, lambda = 1, b = 0.5, N 
 # some heavy tail IDs to look at:
 # ids <- c(6528, 10127, 20579)
 
+source("5-shape-data.R")
+# heavy_pops <- dplyr::filter(gomp_hat_skew, log_skew_50 < 0.7, nu_50 < 30)
+heavy_pops <- dplyr::filter(gomp_hat_base, p10 >= 0.5)
+
 gpdd <- readRDS("gpdd-clean.rds") # to get the last abundance
-skew_samples <- readRDS("skew_samples.rds")
-normal_samples <- readRDS("normal_samples.rds")
+skew_samples <- readRDS("skew_samples_heavy40000.rds")
+normal_samples <- readRDS("normal_samples_heavy40000.rds")
 
 library("doParallel")
 registerDoParallel(cores = 4)
@@ -78,33 +81,38 @@ last_abund <- group_by(gpdd, main_id) %>%
 skew_samples <- inner_join(skew_samples, last_abund, by = "main_id")
 normal_samples <- inner_join(normal_samples, last_abund, by = "main_id")
 
-ww <- skew_samples %>% group_by(main_id) %>% mutate(heavyness = ifelse(median(nu) < 10, "heavy", ifelse(median(nu) < 70, "moderate", "normal"))) %>% group_by(main_id, heavyness) %>% summarise(ci_low_skew = quantile(log_skew, probs = 0.025), ci_up_skew = quantile(log_skew, probs = 0.975)) %>% mutate(skew_ci_contains_zero = ifelse(ci_low_skew < 0 & ci_up_skew > 0, TRUE, FALSE)) %>% as.data.frame()
+ww <- skew_samples %>% group_by(main_id) %>%
+  mutate(heavyness = ifelse(median(nu) < 10, "heavy",
+    ifelse(median(nu) < 70, "moderate", "normal"))) %>%
+  group_by(main_id, heavyness) %>%
[…]

### 2015-11-24 analysis/8.1-plot-skewness.R

@@ -21,7 +21,7 @@ plain_theme <- theme_bw() + theme(
 p <- ggplot(o, aes(x, V1, colour = nu_label)) + geom_line() +
   facet_wrap(~skew_label, nrow = 1) + ylab("Probability density") +
   xlab("x") + plain_theme + labs(colour=expression(nu))
-  print(p)
+  # print(p)
 ggsave("skew-t-illustration.pdf", width = 9, height = 2)
 
 source("5-shape-data.R")
@@ -191,8 +191,8 @@ par(xpd = FALSE)
 # 7099 hare - but re-run with - nu 27, skew 0.5
 # red grouse - 10039 parasites and predators
 
-ids <- c(10113, 7099, 10039, 20579)
-n_project <- 5L - 1L
+ids <- c(10113, 7099, 1235, 20579)
+n_project <- 6L - 1L
 
 get_gomp_res <- function(pop, sigma_proc, lambda, b) {
   res <- rep(NA, length(pop))
@@ -235,10 +235,10 @@ par(mgp = c(2, 0.35, 0))
 
 spark <- function(dat) {
   ylim <- log(range(na.omit(c(dat$q_low3, dat$med))))
-  plot(1, 1, xlim = c(-54, 5), ylim = ylim,
+  plot(1, 1, xlim = c(-54, 6), ylim = ylim,
     type = "n", xlab = "", ylab = "", axes = FALSE)
 
-  rect(1, min(ylim), 5, max(ylim), col = "grey90", border = NA)
+  rect(1, min(ylim), 6, max(ylim), col = "grey90", border = NA)
 
   dat_d <- filter(dat, year <= 1, model == "normal")
   dat_pr_n <- filter(dat, year >= 1, model == "normal")

### 2015-11-24 analysis/9-values-for-paper.R

@@ -1,0 +1,224 @@
+# this file produces some values for inclusion of the text of the associated paper
+
+library("dplyr")
+source("5-shape-data.R")
+p_inc <- readRDS("prob_inc_heavy_with_n.rds")
+
+write_tex <- function(x, macro, ...) {
+  out <- paste0("\\newcommand{\\", macro, "}{", x, "}")
+  cat(out, file = zz)
+  cat("\n", file = zz)
+}
+zz <- file("values.tex", "w") # open .tex file to write to throughout
+
+# # what's the median and mean nu in the exponential prior?
+set.seed(1)
+x <- rexp(3e6, 0.01)
+x <- x[x > 2]
+write_tex(round(mean(x), 0) , "basePriorMean")
+write_tex(round(median(x), 0) , "basePriorMedian")
+write_tex(round(100*length(x[x < 10])/length(x), 1) , "basePriorProbHeavy")
+
+# how much was imputed?
+gpdd <- readRDS("gpdd-clean.rds")
+perc_imputed <- gpdd %>% group_by(main_id) %>%
+  summarise(has_imputed = ifelse(max(interpolated) == 1, TRUE, FALSE)) %>%
+  summarise(percent_imputed = sum(has_imputed)/length(has_imputed)) %>%
+  round(2) * 100
+perc_imputed_pops <- as.character(perc_imputed$percent_imputed)
+
+perc_imputed_points <- round(sum(gpdd$interpolated)/nrow(gpdd) * 100, 1)
+
+fig2_pop_n <- filter(gpdd, taxonomic_class %in%
+    c("Aves", "Mammalia", "Insecta", "Osteichthyes")) %>%
+  summarise(n = length(unique(main_id))) %>% summarise(N = sum(n))
+fig2_pop_n <- fig2_pop_n$N
+
+NPops <- length(unique(gpdd$main_id))
+NOrders <- length(unique(gpdd$taxonomic_order))
+NClasses <- length(unique(gpdd$taxonomic_class))
+
+cols_include <- c("taxonomic_class", "main_id", "p10", "nu_50", "type",
+  "max_rhat", "min_neff")
+gomp_hat_base$ type <- "base"
+gomp_hat_ar1$type <- "ar1"
+gomp_hat_logistic$type <- "logistic"
+gomp_hat_obs_0.2$type <- "obs"
+
+gtemp <- rbind(
[…]

### 2015-11-24 text/supp.tex
Update journal format

@@
 \usepackage{mathtools}
-\usepackage{cite}
+%\usepackage{cite}
 \usepackage{citesupernumber}
+\usepackage{scicite}
 
@@
 \usepackage[cal=boondoxo]{mathalfa} % mathcal
-\useosf % osf for text, not math
+%\useosf % osf for text, not math
 \usepackage[supstfm=libertinesups,%
@@
 
-\usepackage[round,sectionbib]{natbib}
-\bibpunct{(}{)}{;}{a}{}{;}
+%\usepackage[round,sectionbib]{natbib}
+%\bibpunct{(}{)}{;}{a}{}{;}
 
@@
 
+\nocite{taleb2007,sornette2009}
+\nocite{meehl2004,katz2005,ipcc2012}
+\nocite{taleb2007,sornette2009,may2008}
+\nocite{taleb2007}
+\nocite{newman2005}
+\nocite{sornette2009}
+\nocite{harnik2012}
+\nocite{fey2015}
+\nocite{mangel1994}
+\nocite{ipcc2012}
+\nocite{keitt1998,allen2001,halley2002}
+\nocite{brook2006a,knape2012}
+\nocite{SOM}
+\nocite{saucy1994}
+\nocite{stafford1971}
+\nocite{stafford1971}
+\nocite{potts1980}
+\nocite{potts1980}
+\nocite{allen2001}
+\nocite{schindler2010}
+\nocite{carpenter2006}
+\nocite{nunez2012}
+\nocite{meehl2004,katz2005,ipcc2012}
+\nocite{kirby2009}
+\nocite{denny2009}
+\nocite{nrc2007}
+\nocite{caddy1996}
+\nocite{schindler2010}
+\nocite{doak2008,lindenmayer2010}
+\nocite{mangel1994}
+\nocite{loreau2010a,thompson2013}
+\nocite{meehl2004,ipcc2012,thompson2013}
+\nocite{rachev2008}
+
 \input{../analysis/values} % R output
@@
 
+We selected abundance time series from the Global Population Dynamics Database\cite{gpdd2010} (GPDD), which contains nearly 5,000 time series of abundance from \(\sim\) 1,000 species and \(\sim\) 100 taxonomic orders.
+We filtered the data (Supplementary Information) to remove populations from less reliable data sources, and those without sufficient data for our models, and then interpolated some missing values (sensu Brook \textit{et al.}\cite{brook2006a}).
+Our interpolation affected only \(\sim\) \interpPointsPerc \% of the final data points (table~S1) and none of the data points that were later considered black-swan events.
+Our final dataset contained \NPops\ populations across \NOrders\ taxonomic orders and seven taxonomic classes, with a median of \medianTimeSteps\ time steps (range \minTimeSteps--\maxTimeSteps) (table~S1).
+
 We applied the following data selection and quality-control rules to the
@@
   (on a scale of $1$ to $5$, with $1$ being the lowest quality data)
-  \citep[following][]{sibly2005, ziebarth2010}.
+  \cite{sibly2005, ziebarth2010}.
 
@@
   standard deviation. We verified this by locating the original papers the
-  datasets were extracted from: \citet{colebrook1978} for zooplankton and
-  \citet{lindstrom1995} for grouse. Since the papers did not include the
+  datasets were extracted from: \cite{colebrook1978} for zooplankton and
+  \cite{lindstrom1995} for grouse. Since the papers did not include the
   original mean time-series values we could not back transform these data
@@
 \item We filled in single recorded values of zero with the lowest non-zero
-  value in the time series \citep[following][]{brook2006a}. This assumes that
+  value in the time series \cite{brook2006a}. This assumes that
   single values of zero result from abundance being low enough that censusing
@@
 \item We removed all populations without at least four unique values
-  \citep[following][]{brook2006a}.
+  \cite{brook2006a}.
 
@@
   length, we took the most recent window. This is a longer window than used in
-  some previous analyses \citep[e.g.][]{brook2006a}, but since our model
+  some previous analyses \cite{brook2006a}, but since our model
   attempts to capture the shape of the distribution tails, our model requires
@@
   suspected black swans: \texttt{1207} because the 1957 data point was entered
-  as 2 but should have been 27 \citep{kendeigh1982}, \texttt{6531} because the
-  1978 data point was entered as 7 but should have been 47 \citep{minot1986},
+  as 2 but should have been 27 \cite{kendeigh1982}, \texttt{6531} because the
+  1978 data point was entered as 7 but should have been 47 \cite{minot1986},
   and \texttt{6566} because some of the data did not match the graph
-  \citep{heessen1996}.
+  \cite{heessen1996}.
 
@@
 
-\section{Implementation of the skewed Student-t distribution in Stan}
+\section{Details on the heavy-tailed Gompertz probability model} We fit heavy-tailed Gompertz population dynamics models to data from the GPDD.
+The Gompertz model represents population growth as a linear function in log space.
+If \(x_{t}\) represents the ln abundance (\(N\)) at time \emph{t} then
+\begin{align}
+x_t &= \lambda + bx_{t - 1} + \epsilon_t\\
+\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma)
+\end{align}
+The growth parameter \(\lambda\) represents the expected growth rate if \(N_{t} = 1\).
+The model is density independent if \(b = 1\), maximally density dependent if \(b = 0\), and inversely density dependent if \(b < 0\).
+Usually, the process noise \(\epsilon_{t}\) is modelled as normally distributed, but we allow it to be drawn from a Student-t distribution with scale parameter \(\sigma\) and degrees of freedom \(\nu\).
+We can then estimate the degree to which the process deviations have heavy tails and are therefore evidence of black-swan events (Fig.~1a, b).
 
-The probability density for the skewed Student-t distribution \citep{fernandez1998} equates to the following probability density as implemented in \citet{king2012}:
+For the Gompertz model, our weakly-informative priors (fig.~\ref{fig:priors}) were:
+\begin{align}
+b &\sim \mathrm{Uniform}(-1, 2)\\ \lambda &\sim \mathrm{Normal}(0, 10^2)\\
+\sigma &\sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)\\ \nu &\sim
+\mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2).
+\end{align}
+Our prior on \(b\) was uninformative between values of \(-1\) and \(2\).
+We would not expect values of \(b\) with levels of density dependence as low as \(-1\) (very strong inverse density dependence), nor would we generally expect values above \(1\).
+We allowed values of \(b\) above \(1\) to allow for non-stationary time series of growth rates.
+The estimates of \(b\) were well within these bounds.
+Our prior on \(\lambda\) was weakly informative within the range of expected values for population growth and is similar to the default priors suggested by Gelman \emph{et al.}\cite{gelman2008d} for intercepts of regression models.
+Our Half-Cauchy prior on \(\sigma\) follows Gelman \emph{et al.}\cite{gelman2006c} and Gelman \emph{et al.}\cite{gelman2008d} and the specific scale parameter of \(2.5\) is based on our expected range of the value in nature from previous studies\cite{connors2014}.
+In our testing of a subsample of populations, our parameter estimates were not qualitatively changed by switching to a uniform prior on \(\sigma\), but the weakly informative Half-Cauchy prior sped up chain convergence.
 
-\begin{equation}
-  \mathrm{Skewed\,Student\mhyphen t}(x) =
-\begin{dcases}
-\frac{2}{(\gamma + 1/\gamma)}
-         \mathrm{Student\mhyphen t}(x \cdot \gamma, \nu, \mu \cdot \gamma, \sigma),& \text{for } x < 0\\
-       \frac{2}{(\gamma + 1/\gamma)}
-         \mathrm{Student\mhyphen t}(x / \gamma, \nu, \mu / \gamma, \sigma),& \text{for } x \ge 0
-\end{dcases}
-\end{equation}
+We chose a weakly informative exponential prior on \(\nu\) with rate parameter of \(0.01\) truncated at values above two---a slightly less informative prior than suggested by Fernandez and Steel\cite{fernandez1998}.
+We truncated the distribution at \(2\), since at \(\nu < 2\) the variance of the t distribution is undefined.
+This prior gives only a \(7.7\)\% probability that \(\nu < 10\) but constrains the sampling sufficiently to avoid drifting towards infinity.
+In any case, for \(\nu > 20\) the t distribution is almost indistinguishable from the normal distribution (Fig.~1).
+Based on the shape of the t distribution, we chose the probability that \(\nu < 10\), Pr(\(\nu < 10\)), to define the probability of heavy-tailed dynamics.
+When categorizing a population as heavy tailed, we applied a 0.5 probability threshold.
+In the scenario where the data are uninformative about heavy tails (e.g.~Fig.~1e, h), the posterior for \(\nu\) will approximately match the prior (prior median \(= 71\), mean \(= 102\)) and a metric of Pr\((\nu < 10) > 0.5\) is unlikely to flag the population as heavy tailed.
 
-Here $\gamma$ represents the skewness parameter.
-The Student-t($\cdot$) portion represents in order: the data point of interest, the heavy-tailedness parameter, the location parameter, and the scale parameter as passed to the standard Student-t probability density.
-This translates to the log probability density:
+We fit our models with Stan 2.6.0\cite{stan-manual2015, hoffman2014}, and R 3.2.0\cite{r2015}.
+We began with four chains and 2,000 iterations, discarding the first 1,000 as warm up (i.e.~4,000 final samples).
+If \(\hat{R}\) (the potential scale reduction factor---a measure of chain convergence) was greater than \(1.05\) for any parameter, or if the minimum effective sample size, \(n_\mathrm{eff}\) (a measure of the effective number of uncorrelated samples), for any parameter was less than \(200\), we doubled both the total iterations and warm up period and sampled from the model again.
+These thresholds are in excess of the minimums recommended by Gelman \emph{et al.}\cite{gelman2006a} of \(\hat{R} < 1.1\) and \(n_\mathrm{eff} > 100\) for reliable point estimates and confidence intervals.
+In the majority of cases our minimum thresholds were greatly exceeded.
+We continued this procedure up to 8,000 iterations (16,000 total samples) by which all chains were deemed to have sufficiently converged.
+The No-U-Turn Hamiltonian Markov chain Monte Carlo sampler in Stan generally requires fewer iterations to obtain equivalent effective sample sizes than the Gibbs or Metropolis--Hastings algorithms\cite{stan-manual2015}.
 
-\begin{equation}
-\mathrm{Skewed\,Student\mhyphen t}(x) =
-\begin{dcases}
-  \log (2 \gamma) - \log(\gamma^2 + 1) +
-  \log \mathrm{Student\mhyphen t}(x \cdot \gamma, \nu, \mu \cdot \gamma, \sigma),& \text{for } x < 0\\
-  \log (2 \gamma) - \log(\gamma^2 + 1) +
-  \log \mathrm{Student\mhyphen t}(x / \gamma, \nu, \mu / \gamma, \sigma),& \text{for } x \ge 0
-\end{dcases}
-\end{equation}
+There are a number of caveats when considering the generality of our results.
+The GPDD data represent a taxonomically and geographically biased sample of populations---the longer time series we focus on are dominated by commercially and recreationally important species and a disproportionate number of populations are located in the United Kingdom.
+Although we would expect to find qualitatively similar evidence for black swans in other large taxonomic or geographic samples of populations, the common forces driving those black swans would likely differ.
+Additionally, some apparent black-swan events could be recording mistakes, or conversely, some extreme observations may have been discarded or altered if they were erroneously suspected of being recording mistakes.
+Indeed, we discarded three of the populations that our method initially identified as heavy tailed because they turned out to be data-entry errors (Supplementary Information).
+Finally, the temporal scales of observation and population dynamics vary considerably across populations in the GPDD and these likely influence the detection of heavy tails.
+For example, if we make frequent observations relative to generation time (e.g.~for many large-bodied mammals) we will average across generations and perhaps miss black-swan events.
+Conversely, if we census populations infrequently relative to generation time (e.g.~many insects in the GPDD) the recorded data may average across extreme and less-extreme events and also dampen black-swan dynamics.
 
-Since Stan only needs the probability density up to an additive
-constant \citep{stan-manual2014}, we
-can replace $\log(2 \gamma)$ in the above for $\log(\gamma)$. The final log
-probability density for the skewed Student's t distribution implemented in
-Stan\footnote{Credit to Stan developer Bob Carpenter:
-\url{https://groups.google.com/d/msg/stan-users/jeZQnQRUsDs/BeqhicWm1GEJ}}
-looks like the following:
+\section{Alternative priors} To test if the prior on \(\nu\) influenced our estimate of black-swan dynamics, we refit our models with weaker and stronger priors.
+Our base model used a prior on \(\nu\) of Truncated-Exponential(0.01, min.\ = 2).
+For a weaker prior we used Truncated-Exponential(0.005, min.\ = 2) and for a stronger prior we used Truncated-Exponential(0.02, min.\ = 2) (fig.~\ref{fig:priors}c).
+Note that the base and weaker priors are relatively flat within the region of \(\nu < 20\), which is the region we are mostly concerned about when categorizing populations as heavy tailed.
+We furthermore considered a Uniform(\(0, 1\)) prior on \(1/\nu\) as used in Gelman \emph{et al.}\cite{gelman2014}, but this prior puts considerably more weight on lower (heavy-tailed) values of \(\nu\) than the truncated exponential priors considered here (not shown).
 
-\begin{verbatim}
-real skew_student_t_log(real y, real nu, real mu, real sigma, real skew) {
-  real lp;
-  lp <- log(skew) - log1p(square(skew));
-  if (y < mu)
-    return lp + student_t_log(y * skew, nu, mu * skew, sigma);
-  else
-    return lp + student_t_log(y / skew, nu, mu / skew, sigma);
-}
-\end{verbatim}
+Our results show that weaker and stronger priors have little influence on our conclusions about heavy-tailed dynamics (fig.~\ref{fig:alt-priors}).
+When the data are informative about tail behaviour (i.e.
+when there is strong evidence of low \(\nu\) values, upper-right of fig.~\ref{fig:alt-priors}), the prior has little impact on the estimate of \(\nu\).
+When the data are less informative about \(\nu\) (i.e.\ when there are no or few tail events and time series are short or noisy), the prior can pull the estimate of \(\nu\) towards larger or smaller values (fig.~\ref{fig:alt-priors}).
+The vast majority of the populations with Pr\((\nu < 10)\) in the base prior were not altered qualitatively by this range of prior strength.
+
+\section{Alternative population models} We fit alternative population models to test if four key phenomena systematically changed our conclusions.
+The range of percentages of black swans by taxonomic class cited in the abstract are based on lower and upper limits across our main Gompertz model and these four alternative models.
+
+\subsection{Autocorrelated residuals} We considered a version of the Gompertz model in which an autoregressive parameter was fit to the process-noise residuals:
+\begin{align}
+x_t &= \lambda + b x_{t-1} + \epsilon_t\\
+\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, \phi \epsilon_{t-1}, \sigma).
+\end{align}
+In addition to the parameters in the original Gompertz model, this model estimates an additional parameter \(\phi\), which represents the correlation of subsequent residuals.
+Based on the results of previous analyses with the GPDD\cite{connors2014} and the chosen prior in previous analyses\cite{thorson2014a}, we placed a moderately informative prior on \(\phi\) that assumed the greatest probability density near zero with the reduced possibility of \(\phi\) being near \(-1\) or \(1\).
+Specifically, we chose \(\phi \sim \mathrm{Truncated\mhyphen Normal}(0, 0.5, \mathrm{min.} = -1, \mathrm{max.} = 1)\).
+
+\subsection{Assumed density independence} We fit a simplified version of the Gompertz model in which the density dependence parameter \(b\) was fixed at \(1\) (density independent).
+This is equivalent to fitting a random walk model (with drift) to the \(\ln\) abundances or assuming the growth rates are drawn from a stationary distribution.
+The model was as follows:
+\begin{align}
+x_t &= \lambda + x_{t-1} + \epsilon_t\\
+\epsilon &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma).
+\end{align}
+We fit this model for three reasons: (1) it is computationally simpler and so provides a check that our more complicated full Gompertz model was obtaining reasonable estimates of \(\nu\), (2) it provides a test of whether density dependence was systematically affecting our perception of heavy tails, (3) it matches how some previous authors have modelled heavy tails without accounting for density dependence\cite{segura2013}.
+
+\subsection{Assumed observation error} Observation error can bias parameter estimates\cite{knape2012} and is known to affect the ability to detect extreme events\cite{ward2007}.
+In our main analysis, we fit a model that ignored observation error.
+One way to account for observation error would be to fit a full state-space model that simultaneously estimates the magnitude of process noise and observation error.
+However, simultaneously estimating observation and process noise is a challenging problem (e.g.\ because the observation and process noise parameters tend to negatively covary in model fitting) and is known to sometimes result in identifiability issues with the Gompertz population model\cite{knape2008}.
+Furthermore, our model was applied to hundreds of time series, often of short length (as few as \minTimeSteps\ time steps) and our model estimates an additional parameter---the shape of the process deviation tails---potentially making identifiability and computational issues even greater.
+Therefore, we considered a version of the base Gompertz model that allowed for a fixed level of observation error:
+\begin{align}
+U_t &= \lambda + b U_{t-1} + \epsilon_t\\
+x_t &\sim \mathrm{Normal}(U_t, \sigma_\mathrm{obs}^2)\\
+\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma_\mathrm{proc}),
+\end{align}
+where \(U\) represents the unobserved state vector, \(\sigma_\mathrm{obs}\) represents the standard deviation of observation error (on a log scale), and \(\sigma_\mathrm{proc}\) represent the process noise scale parameter.
+We set \(\sigma_\mathrm{obs}\) to \(0.2\), which represents a typical value used in simulation analyses\cite{valpine2002, thorson2014b}.
+
+\subsection{Ricker-logistic} We also fit a Ricker-logistic model:
+\begin{align}
+x_t &= x_{t-1} + r_{\mathrm{max}}\left(1 - \frac{N_{t-1}}{K}\right) + \epsilon_t\\
+\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma),
+\end{align}
+where \(r_\mathrm{max}\) represents the theoretical maximum population growth rate that is obtained when \(N_t = 0\).
+The parameter \(K\) represents the carrying capacity and, as before, \(x_t\) represents the \(\ln\) transformed abundance at time \(t\).
+The Ricker-logistic model assumes a linear decrease in population growth rate with increases in abundance.
+In contrast, the Gompertz model assumes a linear decrease in population growth rate with increases in \textit{log} abundance.
+
+To fit the Ricker-logistic models, we chose a prior on \(K\) uniform between zero and twice the maximum observed abundance (similar and less informative than in Delean \emph{et al.}\cite{delean2013}).
+We set the prior on \(r_\mathrm{max}\) as uniform between 0 and 20 as in Delean \emph{et al.}\cite{delean2013}.
+We used the same priors on \(\nu\) and \(\sigma\) as in the Gompertz model.
+
+\section{Simulation testing} We performed two types of simulation testing (Supporting Information).
+First, we tested how easily the Student-t distribution \(\nu\) parameter could be recovered given different true values of \(\nu\) and different sample sizes.
+Second, we tested the ability of the heavy-tailed Gompertz model to obtain unbiased parameter estimates of \(\nu\) given that a set of process deviations was provided in which the effective \(\nu\) value was close to the true \(\nu\) value.
+
+We found that we could consistently and precisely recover median posterior estimates of \(\nu\) near the true value of \(\nu\) with large samples (\(\ge 200\)) (fig.~\ref{fig:sim-nu} upper panels).
+At smaller samples we could usually qualitatively distinguish heavy from not-heavy tails, but the model tended to underestimate how heavy the tails were.
+At the same time, at smaller sample sizes, the model tended to overestimate how large the scale parameter was (fig.~\ref{fig:sim-nu} lower panels).
+
+Our simulation-testing results indicate that the Gompertz model can recapture the true value of \(\nu\) when the process noise was chosen so that appropriate tail events were present (fig.~\ref{fig:sim-prob} upper panels).
+The addition of observation error caused the model to tend to underestimate the degree of heavy-tailedness.
+Fitting a model with correctly specified observation error did not make substantial improvements to model bias (fig.~\ref{fig:sim-prob}).
 
@@
 \url{https://github.com/seananderson/heavy-tails}):
+
+\clearpage
 
@@
 model can recapture the true value of $\nu$ when the process noise was chosen
-so that appropriate tail events were present (Extended Data Fig.
-5 upper panels). The addition of observation error caused the
+so that appropriate tail events were present (fig.
+\ref{fig:sim-prob} upper panels). The addition of observation error caused the
 model to tend to underestimate the degree of heavy-tailedness. Fitting a model
 with correctly specified observation error did not make substantial
-improvements to model bias (Extended Data Fig. 5).
+improvements to model bias (fig. \ref{fig:sim-prob}).
 
-When converting the posterior distributions of $\nu$ into Pr($\nu < 10$), the
-models distinguished heavy and not-heavy tails reasonably well (Extended Data
-Fig. 5 lower panels). Without observation error, and using a
-probability of $0.5$ as a threshold, the model correctly classified all
-simulated systems with normally distributed process noise as not heavy tailed.
-The model would have miscategorized only one of $40$ simulations at $\nu = 5$
-across simulated populations with $50$ or $100$ time steps (Extended Data Fig.
-5, scenarios 1 and 2 in lower row, second panel from left).
-The model would have correctly categorized all cases where the process noise
-was not heavy tailed (Extended Data Fig. 5 bottom-right
-panel) and all cases where $\nu = 3$ and there was not observation error. With
-$0.2$ standard deviations of observation error, the model still categorized
-\obsErrorNuFivePerc\% of cases as heavy tailed when $\nu = 5$ and all but one
-case when $\nu = 3$. Allowing for observation error made little improvement to
-the detection of heavy tails. Therefore, we chose to focus on the simpler
-model without observation error in the main text, particularly given that the
-true magnitude of observation error was unknown in the empirical data.
+When converting the posterior distributions of $\nu$ into Pr($\nu < 10$), the models distinguished heavy and not-heavy tails reasonably well (fig.
+\ref{fig:sim-prob} lower panels).
+Without observation error, and using a probability of $0.5$ as a threshold, the model correctly classified all simulated systems with normally distributed process noise as not heavy tailed.
+The model would have miscategorized only one of $40$ simulations at $\nu = 5$ across simulated populations with $50$ or $100$ time steps (fig.
+\ref{fig:sim-prob}, scenarios 1 and 2 in lower row, second panel from left).
+The model would have correctly categorized all cases where the process noise was not heavy tailed (fig.
+\ref{fig:sim-prob} bottom-right panel) and all cases where $\nu = 3$ and there was not observation error.
+With $0.2$ standard deviations of observation error, the model still categorized \obsErrorNuFivePerc\% of cases as heavy tailed when $\nu = 5$ and all but one case when $\nu = 3$.
+Allowing for observation error made little improvement to the detection of heavy tails.
+Therefore, we chose to focus on the simpler model without observation error in the main text, particularly given that the true magnitude of observation error was unknown in the empirical data.
+
+\section{Covariates of heavy-tailed dynamics} We fit a hierarchical beta regression model to the predicted probability of heavy tails, Pr(\(\nu < 10\)), to investigate potential covariates of heavy-tailed dynamics.
+We obtained maximum lifespan and body-size data from Brook \emph{et al.}\cite{brook2006a}.
+The beta distribution is useful when response data are on a continuous scale between zero and one\cite{ferrari2004}.
+The model was as follows:
+\begin{align}
+\mathrm{Pr}(\nu_i < 10) &\sim \mathrm{Beta}(A_i, B_i)\\
+\mu_i &= \mathrm{logit}^{-1}(\alpha
+  + \alpha^\mathrm{class}_{j[i]}
+  + \alpha^\mathrm{order}_{k[i]}
+  + \alpha^\mathrm{species}_{l[i]}
+  + X_i \beta),
+  \: \text{for } i = 1, \dots, 617\\
+A_i &= \phi_\mathrm{disp} \mu_i\\
+B_i &= \phi_\mathrm{disp} (1 - \mu_i)\\
+\alpha^\mathrm{class}_j &\sim
+  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{class}}),
+  \: \text{for } j = 1, \dots, 6\\
+\alpha^\mathrm{order}_k &\sim
+  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{order}}),
+  \: \text{for } k = 1, \dots, 38\\
+\alpha^\mathrm{species}_l &\sim
+  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{species}}),
+  \: \text{for } l = 1, \dots, 301,
+\end{align}
+where \(A\) and \(B\) represent the beta distribution shape parameters; \(\mu_i\) represents the predicted value for population \(i\), class \(j\), order \(k\), and species \(l\); \(\phi_\mathrm{disp}\) represents the dispersion parameter; and \(X_i\) represents a vector of predictors (such as lifespan) for population \(i\) with associated \(\beta\) coefficients.
+The intercepts are allowed to vary from the overall intercept \(\alpha\) by taxonomic class (\(\alpha^\mathrm{class}_j\)), taxonomic order (\(\alpha^\mathrm{order}_k\)), and species (\(\alpha^\mathrm{species}_l\)) with standard deviations \(\sigma_{\alpha \; \mathrm{class}}\), \(\sigma_{\alpha \; \mathrm{order}}\), and \(\sigma_{\alpha \; \mathrm{species}}\).
+Where possible, we also allowed for error distributions around the predictors by incorporating the standard deviation of the posterior samples for the Gompertz parameters \(\lambda\), \(b\), and \(\log \sigma\) around the mean point value as normal distributions (not shown in the above equation).
+
+We log transformed \(\sigma\), time-series length, and lifespan to match the way they are visually represented in fig.~\ref{fig:correlates} and to make the relationship approximately linear on the logit-transformed response scale.
+All input variables were standardized by subtracting their mean and dividing by two standard deviations to make their coefficients comparable in magnitude\cite{gelman2008c}.
+We excluded body length as a covariate because it was highly correlated with lifespan, and lifespan exhibited more overlap across taxonomy than body length.
+Lifespan is also more directly related to time and potential mechanisms driving black-swan dynamics.
+
+We incorporated weakly informative priors into our model: \(\mathrm{Cauchy}(0, 10)\) on the global intercept \(\alpha\), \(\mathrm{Half\mhyphen Cauchy}(0, 2.5)\) on all standard deviation parameters, \(\mathrm{Half\mhyphen Cauchy}(0, 10)\) on the dispersion parameter \(\phi_\mathrm{disp}\), and \(\mathrm{Cauchy}(0, 2.5)\) on all other parameters\cite{gelman2006c, gelman2008d}.
+Compared to normal priors, the Cauchy priors concentrate more probability density around expected parameter values while allowing for a higher probability density far into the tails, thereby allowing the data to dominate the posterior more strongly if it disagrees with the prior.
+We fit our models with 5,000 total iterations per chain, 2500 warm-up iterations, four chains, and discarding every second sample.
+We checked for chain convergence visually and with the same criteria as before (\(\widehat{R} < 1.05\) and \(n_\mathrm{eff} >200\) for all parameters).
+We further checked our model against a similar model with the glmmADMB \textsf{R} package\cite{glmmadmb} without uncertainty around the predictors.
+
+To derive taxonomic-order-level estimates of the probability of heavy tails accounting for time-series length (Fig.~3a), we fit a separate hierarchical model with the same structure but with only \(\log\) time-series length as a predictor---in this case, we did not want to control for intrinsic population characteristics such as density dependence.
+Since our predictors were centered by subtracting their mean value, we obtained order-level estimates of the probability of heavy tails at mean log time-series length by adding the posteriors for \(\alpha\), \(\alpha^\mathrm{class}_j\), and \(\alpha^\mathrm{order}_k\).
+
+\section{Skewed Student-t forecasts} To formally evaluate the apparent skewness of heavy-tailed process noise we fit Gompertz models with skewed Student-t distributed process noise\cite{fernandez1998} (Supplementary Information).
+The distribution adds one parameter to the Student-t distribution, \(\gamma\), which controls the skewness.
+The distribution is symmetrical if \(\gamma = 1\), left skewed if \(0 < \gamma < 1\), and right skewed if \(1 < \gamma < \infty\).
+We placed a weakly informative prior of \(\mathrm{Cauchy}(0, 2.5)\) on \(\log \gamma\).
+We aggregated 5,000 randomly selected posterior samples from the \(\gamma\) parameter of each model at three levels of evidence for heavy-tails: \emph{heavy tailed} median \(\nu < 10\); \emph{slightly heavy tailed} \(10 \le \mathrm{median}\, \nu \le 70\), and \emph{normal tails}, median \(\nu > 70\) (Fig.~4).
+
+To generate five-year forecasts of abundance, we combined the posterior parameter samples from the skewed-t Gompertz models with stochastically generated process noise.
+We compared these forecasts to those generated by a standard Gompertz model with normally distributed process noise fit to the same data.
+We calculated the ratio of the lower 99\% quantile credible interval between the two model projections for all populations in which Pr(\(\nu < 10\)) \(> 0.5\).
+To ensure the results had stabilized in the tails of the forecast posterior, we increased the number of posterior samples in Stan.
+We ran four chains of 20,000 iterations and discarded the first 10,000 as warmup for a total of 40,000 samples.
+
+\section{Implementation of the skewed Student-t distribution in Stan}
+
+The probability density for the skewed Student-t distribution \cite{fernandez1998} equates to the following probability density as implemented in Ref.~\cite{king2012}:
+
+\begin{equation}
+  \mathrm{Skewed\,Student\mhyphen t}(x) =
+\begin{dcases}
+\frac{2}{(\gamma + 1/\gamma)}
+         \mathrm{Student\mhyphen t}(x \cdot \gamma, \nu, \mu \cdot \gamma, \sigma),& \text{for } x < 0\\
+       \frac{2}{(\gamma + 1/\gamma)}
+         \mathrm{Student\mhyphen t}(x / \gamma, \nu, \mu / \gamma, \sigma),& \text{for } x \ge 0
+\end{dcases}
+\end{equation}
+
+Here $\gamma$ represents the skewness parameter.
+The Student-t($\cdot$) portion represents in order: the data point of interest, the heavy-tailedness parameter, the location parameter, and the scale parameter as passed to the standard Student-t probability density.
+This translates to the log probability density:
+
+\begin{equation}
+\mathrm{Skewed\,Student\mhyphen t}(x) =
+\begin{dcases}
+  \log (2 \gamma) - \log(\gamma^2 + 1) +
+  \log \mathrm{Student\mhyphen t}(x \cdot \gamma, \nu, \mu \cdot \gamma, \sigma),& \text{for } x < 0\\
+  \log (2 \gamma) - \log(\gamma^2 + 1) +
+  \log \mathrm{Student\mhyphen t}(x / \gamma, \nu, \mu / \gamma, \sigma),& \text{for } x \ge 0
+\end{dcases}
+\end{equation}
+
+Since Stan only needs the probability density up to an additive
+constant \cite{stan-manual2014}, we
+can replace $\log(2 \gamma)$ in the above for $\log(\gamma)$. The final log
+probability density for the skewed Student's t distribution implemented in
+Stan\footnote{Credit to Stan developer Bob Carpenter:
+\url{https://groups.google.com/d/msg/stan-users/jeZQnQRUsDs/BeqhicWm1GEJ}}
+looks like the following:
+
+\begin{verbatim}
+real skew_student_t_log(real y, real nu, real mu, real sigma, real skew) {
+  real lp;
+  lp <- log(skew) - log1p(square(skew));
+  if (y < mu)
+    return lp + student_t_log(y * skew, nu, mu * skew, sigma);
+  else
+    return lp + student_t_log(y / skew, nu, mu / skew, sigma);
+}
+\end{verbatim}
 
@@
 
-\bibliographystyle{apalike}
-\bibliography{/Users/seananderson/Dropbox/tex/jshort,supp}
+\section{Code availability.} Data and code to reproduce our analysis are available at \url{https://github.com/seananderson/heavy-tails} (currently private).
+
+\bibliographystyle{Science}
+%\bibliography{/Users/seananderson/Dropbox/tex/jshort,supp}
+\bibliography{jshort,newbib}
+
+\section{Supplementary tables and figures}
+
+%\newenvironment{helvetica}{\fontfamily{phv}\selectfont}{\par}
+
+\singlespacing
+
+\textbf{Table~S1: Summary statistics for the filtered Global Population Dynamics Database time series arranged by taxonomic class.} Columns are number of populations, number of taxonomic orders, numbers of species, median time series length, total number of interpolated time steps, total number of substituted zeros, and total number of time steps.
+
+\onehalfspacing
+
+%\begin{helvetica}
+\smallskip
+\begin{scriptsize}
+\begin{tabular}{lrrrrrrrr}
+\toprule
+\input{../analysis/stat-table.tex}
+\label{tab:stats}
+\end{tabular}
+\end{scriptsize}
+%\end{helvetica}
+
+\clearpage
+
+\LTcapwidth=\textwidth
+
+\renewcommand{\thefigure}{S\arabic{figure}}
+\renewcommand{\thetable}{S\arabic{table}}
+
+\renewcommand{\tablename}{Table}
+
+\setcounter{table}{1}
+
+\renewcommand{\arraystretch}{0.1}% Tighter
+
+%\begin{helvetica}
+\singlespacing
+\begin{scriptsize}
+\begin{longtable}{>{\RaggedRight}m{1.4cm}>{\RaggedRight}p{6.2cm}>{\RaggedRight}p{0.7cm}>{\RaggedRight}p{0.4cm}>{\RaggedRight}p{4.4cm}>{\RaggedRight}p{1.4cm}}
+\caption{\textbf{Populations with a high probability of heavy-tailed dynamics in the base heavy-tailed Gompertz population dynamics model.} Shown are the log abundance time series, population descriptions, Global Population Dynamics Database Main IDs, citation for the data source or separate verification literature, a description of the cause of the black-swan events (if known), the probability of heavy tails as calculated by our model, and median estimate of $\nu$ from our model with 90\% quantile credible intervals indicated in parentheses.
+Red dots on the time series indicate downward black-swan events and blue values indicate upward black-swan events that have a $10^{-4}$ probability or less of occurring if the population dynamics were explained by a Gompertz model with normally distributed process noise with a standard deviation equal to the scale parameter in the fitted t distribution.}\\
+
+\toprule
+\input{../analysis/cause-table.tex}
+\label{tab:causes-supp}
+\end{longtable}
+\end{scriptsize}
+\onehalfspacing
+%\end{helvetica}
+
+\clearpage
+
+\renewcommand{\figurename}{Figure}
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=0.6\textwidth]{../analysis/priors-gomp-base.pdf}
+
+\caption{\textbf{Probability density of the Bayesian priors for the Gompertz models.} (\textbf{A}) per capita growth rate at $\log$(abundance) = $0$: $\lambda \sim \mathrm{Normal}(0, 10^2)$.
+(\textbf{B}) scale parameter of t-distribution process noise: $\sigma \sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)$ (\textbf{C}) Student-t distribution degrees of freedom parameter: $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2)$.
+(\textbf{D}) AR1 correlation coefficient of residuals: $\phi \sim \mathrm{Truncated \mhyphen Normal}(0, 0.5, \mathrm{min.} = -1, \mathrm{max.} = 1)$.
+Not shown is $b$, the density dependence parameter: $b \sim \mathrm{Uniform}(-1, 2)$.
+Panel \textbf{C} also shows two alternative priors: a weaker prior $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.005, \mathrm{min.} = 2)$, and a stronger prior $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.02, \mathrm{min.} = 2)$.
+The inset panel shows the same data but with a log-transformed x axis.
+Note that the base and weaker priors are relatively flat within the region of $\nu < 20$ with which we are concerned.}
+
+\label{fig:priors}
+\end{center}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=0.8\textwidth]{../analysis/gomp-comparison.pdf}
+
+\caption{\textbf{Estimates of $\nu$ from alternative models plotted against the
+    base Gompertz model estimates of $\nu$.} Alternative models are (\textbf{A})
+    Ricker-logistic; (\textbf{B}) a Gompertz model with AR1 residual correlation;
+    (\textbf{C}) modelling population growth rate with no density dependence; and
+    (\textbf{D}) a Gompertz model with fixed magnitude of observation error. Shown
+are medians of the posterior (dots) and 50\% credible intervals (segments). The
+diagonal line indicates a one-to-one relationship. Different colours indicate
+various taxonomic classes. The grey-shaded regions indicate regions of
+disagreement if $\nu = 10$ is taken as a threshold of heavy-tailed dynamics.
+The Gompertz observation error model assumes a fixed standard deviation of
+observation error of $0.2$ on a log scale. Models in \textbf{B} and \textbf{D}
+that had not met convergence thresholds were run for four chains, 256,000 iterations (128,000 warm up), and retaining every 10th sample, after which the MCMC chains for one population in \textbf{B} and four populations in \textbf{D} had still not met convergence thresholds.
+These populations are excluded from their respective panels, but all populations with Pr($\nu < 10$) $> 0.5$ in the base model had converged and are shown.}
+
+\label{fig:alt}
+\end{center}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=0.8\textwidth]{../analysis/gomp-prior-comparison.pdf}
+
+\caption{\textbf{Estimates of $\nu$ from Gompertz models with alternative priors on $\nu$.} \textbf{A} shows a weaker prior on $\nu$, \textbf{B} shows a stronger prior on $\nu$, as illustrated in Fig.~1.
+Shown are medians of the posterior (dots) and 50\% credible intervals (segments).
+The diagonal line indicates a one-to-one relationship.
+Different colours indicate various taxonomic classes.
+The grey-shaded regions indicate regions of disagreement if $\nu = 10$ is taken as a threshold of heavy-tailed dynamics.
+In general, the estimates are nearly identical in cases where the data are informative about low values of $\nu$.
+When the data are less informative about low values of $\nu$, the prior can slightly pull the estimates of $\nu$ towards higher or lower values.}
+
+\label{fig:alt-priors}
+\end{center}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=0.8\textwidth]{../analysis/t-dist-sampling-sim-prior-exp0point01.pdf}
+\includegraphics[width=0.8\textwidth]{../analysis/t-dist-sampling-sim-sigma-prior-exp0point01.pdf}
+
+\caption{\textbf{Testing the ability to estimate $\nu$ (top panels) and the scale parameter of the process deviations (bottom panels) for a given number of samples (columns) drawn from a distribution with a given true $\nu$ value (rows).} The red lines indicate the true population value.
+When a small number of samples are drawn there may not be samples sufficiently far into the tails to recapture the true $\nu$ value; however, heavy tails are still distinguished from normal tails in most cases, even with only 25 or 50 samples.}
+
+\label{fig:sim-nu}
+\end{center}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=\textwidth]{../analysis/sim-gompertz-median-dist.pdf}
+\includegraphics[width=\textwidth]{../analysis/sim-gompertz-p10.pdf}
+
+\caption{\textbf{Simulation testing the Gompertz estimation model when the process deviation draws were chosen so that $\nu$ could be estimated close to the true value outside the full population model} (``effective $\nu$'' within a CV of 0.2 of specified $\nu$).
+Upper panels show the distribution of median $\widehat{\nu}$ across 20 simulation runs.
+Lower panels show the distribution of Pr($\nu < 10$) across 20 simulation runs.
+We ran the simulations across three population (``true'') $\nu$ values (3, 5, and $1\cdot 10^9$, i.e.\ approximately normal) and four scenarios: (1) 100 time steps and no observation error, (2) 50 time steps and no observation error, (3) 50 time steps and observation error drawn from $\mathrm{Normal} (0, 0.2^2)$ but ignored, and (4) 50 time steps with observation error in which the quantity of observation error was assumed known.
+Within each scenario the dots represent stochastic draws from the true population distributions combined with model fits.
+Underlayed boxplots show the median, interquartile range, and $1.5$ times the interquartile range.}
+
+\label{fig:sim-prob}
+\end{center}
+\end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=0.8\textwidth]{../analysis/correlates-p10.pdf}
+
+\caption{\textbf{Potential covariates of heavy-tailed population dynamics (indicated by a high probability that $\nu < 10$).} Shown are (\textbf{A} to \textbf{C}) parameters from the Gompertz heavy-tailed population model ($\sigma$, $\lambda$, $b$), (\textbf{D}) number of time steps, (\textbf{E}) body length, and (\textbf{F}) lifespan.
+For the Gompertz parameters, $\sigma$ refers to the scale parameter of the Student-t process-noise distribution, $\lambda$ refers to the expected log abundance at the next time step at an abundance of one, $b$ refers to the density dependence parameter.
+Circles representing a few sharks, crustaceans, and gastropods are filled in white.
+Median and 90\% credible interval posterior predictions of a beta regression hierarchical model are shown in panels \textbf{A} and \textbf{D} where there was a high probability the slope coefficient was different from zero (Fig.~3b).}
+
+\label{fig:correlates}
+\end{center}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=6in]{../analysis/skew-t-illustration}
+\includegraphics[width=4in]{../analysis/skewness-vs-nu}
+
+\caption{\textbf{Skewness and heavy-tailed parameter posteriors.} Top panel shows the shape of the skew-t distribution at different combinations of skewness ($\gamma$) and heavy-tailedness ($\nu$).
+Lower panel shows the posterior medians and 50\% credible intervals for the two parameters.
+The horizontal red line indicates the median ln(skewness) parameter across all populations.}
+
+\label{fig:skew-nu}
+\end{center}
+\end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=1.1\textwidth]{../analysis/heavy-skew-projections-ggplot}
+\caption{\textbf{Projections of population abundance under skew-t (red) and
+normally distributed process error models (blue) for populations with
+Pr($\nu<10$) $>0.5$.} Numbers above panels indicate GPDD population IDs as
+shown in table~S2. Horizontal axes indicates year before last
+recorded abundance and a projection five years into the future. For the
+projected region, solid lines represents the median projection and
+dashed lines represent lower 99\% credible intervals. Vertical axes are
+log-distributed. Assuming normal process error tends to underestimate risk of
+low abundance if the process error is in fact heavy-tailed.}
+
+\label{fig:skew-projections}
+\end{center}
+\end{figure}
+
 %

### 2015-11-24 analysis/6-plot-alt-models.R

@@ -73,11 +73,11 @@ pdf("gomp-comparison.pdf", width = 7, height = 6.8)
 par(mfrow = c(2, 2), mar = c(3,3,0,0), oma = c(.5, .5, 3.5, .5),
   tck = -0.02, mgp = c(1.5, 0.4, 0), col.axis = "grey25", col = "grey25", las = 1)
 par(cex = 0.9)
-comp_panel(gomp_hat_base, gomp_hat_logistic, quote(Gompertz~widehat(nu)), quote(Ricker-logistic~widehat(nu)), label = "a")
+comp_panel(gomp_hat_base, gomp_hat_logistic, quote(Gompertz~widehat(nu)), quote(Ricker-logistic~widehat(nu)), label = "A")
 legend(0.3, 0.63, legend = cols_df$taxonomic_class[1:4], fill = cols_df$col[1:4], bty = "n", ncol = 2)
-comp_panel(gomp_hat_base, gomp_hat_ar1, quote(Gompertz~widehat(nu)), quote(Gompertz~AR1~widehat(nu)), label = "b")
-comp_panel(gomp_hat_base, gomp_hat_rate, quote(Gompertz~widehat(nu)), quote(Rate~only~widehat(nu)), label = "c")
-comp_panel(gomp_hat_base, gomp_hat_obs_0.2, quote(Gompertz~widehat(nu)), quote(Gompertz~obs.~error~widehat(nu)), label = "d")
+comp_panel(gomp_hat_base, gomp_hat_ar1, quote(Gompertz~widehat(nu)), quote(Gompertz~AR1~widehat(nu)), label = "B")
+comp_panel(gomp_hat_base, gomp_hat_rate, quote(Gompertz~widehat(nu)), quote(Rate~only~widehat(nu)), label = "C")
+comp_panel(gomp_hat_base, gomp_hat_obs_0.2, quote(Gompertz~widehat(nu)), quote(Gompertz~obs.~error~widehat(nu)), label = "D")
 dev.off()
 
 # pdf("gomp-prior-comparison.pdf", width = 7, height = 6.8)
@@ -85,8 +85,8 @@ pdf("gomp-prior-comparison.pdf", width = 7.2, height = 4)
 par(mfrow = c(1, 2), mar = c(3,3,0,0), oma = c(.5, .5, 3.5, .5),
   tck = -0.02, mgp = c(1.5, 0.4, 0), col.axis = "grey25", col = "grey25", las = 1)
 par(cex = 0.9)
[…]

### 2015-11-24 analysis/6-plot-nu-coefs.R

@@ -33,7 +33,7 @@ for(i in 1:length(classes)) {
   plot(1, 1, type = "n", xlim = c(-0.2, 100.2), ylim = c(0, 0.5), xlab = "", ylab = "", yaxt = "n", xaxs = "i", yaxs = "i", xaxt = xaxt, las = 1)
   #axis(2, at = seq(0, 100, 50), las = 1)
   par(xpd = NA)
-  add_label(label = letters[i], font = 2, cex = 1.2)
+  add_label(label = LETTERS[i], font = 2, cex = 1.2)
   add_label(label = classes[i], xfrac = 0.03)
   if(i == 1) {
     mtext("Heavy", side = 2, line = 2.5, adj = 0.90, col = "grey55")

### 2015-11-24 analysis/6-plot-order-correlate-posteriors.R

@@ -129,7 +129,7 @@ axis(2, at = seq_along(op),
 axis(1, at = seq(0, 0.3, 0.1), mgp = c(2, 0.3, 0))
 mtext("Probability of black swans", side = 1, line = 1.8, cex = 0.8)
 mtext(quote(Pr(nu<10)), side = 1, line = 3, cex = 0.8, col = subtext_col)
-mtext("a", side = 3, line = 0, cex = 1.2, adj = -0.6, font = 2)
+mtext("A", side = 3, line = 0, cex = 1.2, adj = -0.6, font = 2)
 
 ################
 # the main coefficients:
@@ -180,7 +180,7 @@ mtext("Coefficient value", side = 1, line = 1.55,
   cex = 0.8, outer = FALSE)
 mtext("(per 2 SDs of predictor)", side = 1, line = 2.6,
   cex = 0.8, outer = FALSE, col = subtext_col)
-mtext("b", side = 3, line = 0.5, cex = 1.2, adj = -0.6, font = 2)
+mtext("B", side = 3, line = 0.5, cex = 1.2, adj = -0.6, font = 2)
 
 
 dev.off()

### 2015-11-24 analysis/6-plot-prior.R

@@ -32,12 +32,12 @@ pdf("priors-gomp-base.pdf", width = 6, height = 5.2)
 par(mfrow = c(2, 2), mar = c(3,3,0,0), oma = c(.5, 3, .5, .5),
   tck = -0.02, mgp = c(1.5, 0.4, 0), col.axis = "grey25", col = "grey25", las = 1)
 x <- seq(-9, 9, length.out = 200)
-plot_prior(x, dnorm(x, 0, 10), expression(lambda), label = "a")
+plot_prior(x, dnorm(x, 0, 10), expression(lambda), label = "A")
 x <- seq(0, 6, length.out = 200)
-plot_prior(x, dcauchy(x, 0, 2.5), expression(sigma[proc]), ylim = c(0, 0.16), label = "b")
+plot_prior(x, dcauchy(x, 0, 2.5), expression(sigma[proc]), ylim = c(0, 0.16), label = "B")
 x <- seq(2, 500, length.out = 2000)
 plot_prior(x, dtrunc(x, "exp", a = 2, b = Inf, rate = 0.02), expression(nu),
-  col = "grey70", log = "", ylim = NULL, lty = "93", label = "c")
+  col = "grey70", log = "", ylim = NULL, lty = "93", label = "C")
 plot_prior(x, dtrunc(x, "exp", a = 2, b = Inf, rate = 0.01),
   expression(nu), add = TRUE, lty = 1, col = "black")
 plot_prior(x, dtrunc(x, "exp", a = 2, b = Inf, rate = 0.005),
@@ -64,6 +64,6 @@ TeachingDemos::subplot({
 
 x <- seq(-1.1, 1.1, length.out = 200)
 plot_prior(x, dtrunc(x, "norm", a = -1, b = 1, mean = 0, sd = 0.5),
-  expression(phi), xlim = c(-1.1, 1.1), label = "d")
+  expression(phi), xlim = c(-1.1, 1.1), label = "D")
 mtext("Probability density", side = 2, outer = TRUE, line = 1, las = 0, cex = 0.9)
 dev.off()

### 2015-11-24 analysis/6-plot-t-nu-eg.R

@@ -27,7 +27,7 @@ x <- seq(-9.5, 9.5, length.out = 500)
 plot(x, dt(x, 2), type = "l", xlab = "",
   ylab = "", yaxs = "i", lty = 1, , lwd = lwd,
   xlim = c(-8, 8), ylim = c(0, 0.42), col = cols[1], xaxt = "n", yaxt = "n")
-text(-7.5, 0.393, "a", font = 2, cex = 1.4)
+text(-7.5, 0.393, "A", font = 2, cex = 1.4)
 mtext("Student-t distribution", side = 3, line = 0, cex = 0.9)
 axis(2, at = c(0, 0.2, 0.4), col = box.col)
 lines(x, dt(x, 3), lty = 1, col = cols[2], lwd = lwd)
@@ -42,7 +42,7 @@ legend("topright", legend = c(expression(nu==2), expression(nu==3),
 plot(x, dt(x, 2), type = "l", xlab = "Standard deviation",
   ylab = "", yaxs = "i", lty = 1, , lwd = lwd,
   xlim = c(-8, 8), ylim = c(0, 0.025), col = cols[1], yaxt = "n", mgp = c(2, 0.4, 0))
-text(-7.5, 0.0218, "b", font = 2, cex = 1.4)
+text(-7.5, 0.0218, "B", font = 2, cex = 1.4)
 axis(2, at = c(0, 0.02), col = box.col)
 lines(x, dt(x, 3), lty = 1, col = cols[2], lwd = lwd)
 lines(x, dt(x, 5), lty = 1, col = cols[3], lwd = lwd)
@@ -84,18 +84,18 @@ make_ts <- function(seed, nu, ylim = c(-6.5, 9), bs_col = "red") {
 
 seed <- nu_5_seeds_N50$seeds[19]
 y1 <- make_ts(seed = seed, nu = 3, bs_col = bs_cols[2])
-text(1, 7.1, "c", pos = 4, font = 2, cex = 1.4)
+text(1, 7.1, "C", pos = 4, font = 2, cex = 1.4)
 text(6, 7.1, quote(nu==3), pos = 4)
 mtext("Simulated population dynamics", side = 3, line = 0, cex = 0.9)
 text(13, -3.9, "Black swans", pos = 4)
 y2 <- make_ts(seed = seed, nu = 5, bs_col = bs_cols[3])
[…]

### 2015-11-24 analysis/8.1-plot-skewness.R

@@ -149,7 +149,7 @@ layout(l)
 
 # illustrations of skew-nu vs. normal:
 pfunc2(nu = 5, skew = 0.65, col = pal[1], label1 = expression(nu == 5), label2 = expression(gamma == 0.65))
-add_label(label = "a", cex = 1.8, font = 2, yfrac = -0.64)
+add_label(label = "A", cex = 1.8, font = 2, yfrac = -0.64)
 add_label(label = "Process deviation shape", cex = 1.4, font = 1,
   xfrac = 0.2, col = "grey20", yfrac = -0.64)
 pfunc2(nu = 20, skew = 0.75, col = pal[2], label1 = expression(nu == 20), label2 = expression(gamma == 0.75))
@@ -174,7 +174,7 @@ mtext(expression(Skewness~parameter~(gamma)), side = 1, col = "grey20", line = 2
 #mtext("Skewed up", side = 1, col = "grey40", line = -6.7, adj = 0.87)
 # mtext("up", side = 1, col = "grey40", line = -5.3, adj = 0.87)
 
-add_label(label = "b", cex = 1.8, font = 2, yfrac = -0.05)
+add_label(label = "B", cex = 1.8, font = 2, yfrac = -0.05)
 add_label(label = "Skewness parameter posteriors", cex = 1.4, font = 1,
   yfrac = -0.05, xfrac = 0.05, col = "grey20")
 
@@ -268,7 +268,7 @@ for (i in ids) {
   this_dat <- filter(plot_dat2, main_id == i)
   spark(this_dat)
   if (ii == 1) {
-    add_label(label = "c", cex = 1.8, font = 2, yfrac = -0.5, xfrac = -0.16)
+    add_label(label = "C", cex = 1.8, font = 2, yfrac = -0.5, xfrac = -0.16)
     add_label(label = "Example time series", cex = 1.4, font = 1,
       xfrac = -0.08, col = "grey20", yfrac = -0.5)
   }

### 2016-01-02 text/supp.tex
Format tweaks

@@
 \author{
-Sean C. Anderson$^{1,2}$ \and
+Sean C. Anderson$^{1,2*}$ \and
 Trevor A. Branch$^3$ \and
@@
 Washington, Box 355020, Seattle, WA 98195, USA
+
+\textsuperscript{*}Corresponding author: Sean C. Anderson; E-mail: sandrsn@uw.edu; Present address: School of Aquatic and Fishery Sciences, Box 355020, University of Washington, Seattle, WA 98195, USA
 
@@
 Usually, the process noise \(\epsilon_{t}\) is modelled as normally distributed, but we allow it to be drawn from a Student-t distribution with scale parameter \(\sigma\) and degrees of freedom \(\nu\).
-We can then estimate the degree to which the process deviations have heavy tails and are therefore evidence of black-swan events (Fig.~1a, b).
+We can then estimate the degree to which the process deviations have heavy tails and are therefore evidence of black-swan events (Fig.~1A, B).
 
@@
 When categorizing a population as heavy tailed, we applied a 0.5 probability threshold.
-In the scenario where the data are uninformative about heavy tails (e.g.~Fig.~1e, h), the posterior for \(\nu\) will approximately match the prior (prior median \(= 71\), mean \(= 102\)) and a metric of Pr\((\nu < 10) > 0.5\) is unlikely to flag the population as heavy tailed.
+In the scenario where the data are uninformative about heavy tails (e.g.~Fig.~1E, H), the posterior for \(\nu\) will approximately match the prior (prior median \(= 71\), mean \(= 102\)) and a metric of Pr\((\nu < 10) > 0.5\) is unlikely to flag the population as heavy tailed.
 
@@
 
-To derive taxonomic-order-level estimates of the probability of heavy tails accounting for time-series length (Fig.~3a), we fit a separate hierarchical model with the same structure but with only \(\log\) time-series length as a predictor---in this case, we did not want to control for intrinsic population characteristics such as density dependence.
+To derive taxonomic-order-level estimates of the probability of heavy tails accounting for time-series length (Fig.~3A), we fit a separate hierarchical model with the same structure but with only \(\log\) time-series length as a predictor---in this case, we did not want to control for intrinsic population characteristics such as density dependence.
 Since our predictors were centered by subtracting their mean value, we obtained order-level estimates of the probability of heavy tails at mean log time-series length by adding the posteriors for \(\alpha\), \(\alpha^\mathrm{class}_j\), and \(\alpha^\mathrm{order}_k\).
@@
 can replace $\log(2 \gamma)$ in the above for $\log(\gamma)$. The final log
-probability density for the skewed Student's t distribution implemented in
+probability density for the skewed Student-t distribution implemented in
 Stan\footnote{Credit to Stan developer Bob Carpenter:
@@
 Circles representing a few sharks, crustaceans, and gastropods are filled in white.
-Median and 90\% credible interval posterior predictions of a beta regression hierarchical model are shown in panels \textbf{A} and \textbf{D} where there was a high probability the slope coefficient was different from zero (Fig.~3b).}
+Median and 90\% credible interval posterior predictions of a beta regression hierarchical model are shown in panels \textbf{A} and \textbf{D} where there was a high probability the slope coefficient was different from zero (Fig.~3B).}
 

### 2016-01-02 text/ms.tex
Add ms.tex

--- 2015-07-21 text/ms.tex
@@
 
-\textsuperscript{3}School of Aquatic and Fishery Sciences, University of Washington, Box 355020, Seattle, WA 98195, USA
+\textsuperscript{3}School of Aquatic and Fishery Sciences, Box 355020, University of Washington, Seattle, WA 98195, USA
 
-\textsuperscript{*}Corresponding author: Sean C. Anderson; E-mail: sean\_anderson@sfu.ca
+\textsuperscript{*}Corresponding author: Sean C. Anderson; E-mail: sandrsn@uw.edu; Present address: School of Aquatic and Fishery Sciences, Box 355020, University of Washington, Seattle, WA 98195, USA
+
+%125
 
 \textbf{Abstract}:
-
 Black swans are statistically improbable events that nonetheless occur---often with profound consequences\cite{taleb2007, sornette2009}.
 While extremes in the physical environment, such as monsoons and heat waves, are widely studied and increasing in magnitude and frequency\cite{meehl2004, katz2005, ipcc2012}, it remains unclear the extent to which ecological populations buffer or suffer from such extremes.
-We derived a model for estimating the frequency of black-swan events and applied it to abundance time series of 609 animal populations.
-We find strong evidence for black-swan events, but they are rare, occurring in 3--5\% of populations: most frequently for birds (4--8\%), mammals (4--6\%), and insects (2--3\%).
-Black swan events were predominantly (87\%) downward, implying that unexpected population crashes occur far more frequently than unexpected population increases.
-Black swan events are not explained by any life-history covariates, but tend to be driven by external perturbations such as climate or severe winters.
-Data and simulations reveal that black swan-events are more likely to be detected in longer and less noisy time series.
-Our results show that accurate forecasts of extinction risk require accounting for the possibility of black swans and the predominance of crashes over increases, or predicted abundance will underestimate the probability of crashes by a factor of XX--XX.
-In addition, the presence of black swans highlights the importance of developing flexible and robust conservation and management strategies, particularly as the frequency and magnitude of climate extremes increase over the next century.
+Here we derive a model for estimating the frequency and direction of black-swan events and apply it to abundance time series of 609 animal populations.
+We find strong evidence for black-swan events, but they are rare, occurring in \(\sim\)\overallBasePerc \% of populations: most frequently for birds (\birdPH \%), mammals (\mammalsPH \%), and insects (\insectsPH \%).
+Black-swan events are predominantly (\percBSDown \%) downward, implying that unexpected population crashes occur far more frequently than unexpected population increases.
+Black-swan events are not explained by any intrinsic life-history covariates, but tend to be driven by extrinsic perturbations such as climate or severe winters.
+Data and simulations reveal that black-swan events are more likely to be detected in longer and less noisy time series.
+Rare events tend to be crashes and ignoring them will underestimate extinction risk.
+Ignoring black-swan events leads to a \crashUnderMedian-fold (50\% interval \crashUnderRange) underestimate in the magnitude of population crashes.
+Our results demonstrate the critical importance of allowing for downward and heavy-tailed events in population modelling and developing robust conservation and management strategies, particularly in the face of increased climate extremes.
 
-\clearpage
+\bigskip
+\bigskip
 
-Major surprises happen more often than expected in financial, social, and natural systems\cite{taleb2007, may2008}.
-Massive unpredictable market swings are responsible for the majority of financial gains and losses\cite{taleb2007}, fatalities from the largest wars dwarf those from other wars\cite{newman2005}, and the most damaging earthquakes greatly exceed their expected frequency\cite{sornette2009}.
+\textbf{One Sentence Summary}:
+
+\textbf{Main text}:
+Major surprises happen more often than expected in financial, social, and natural systems\cite{taleb2007, sornette2009, may2008}.
+Massive unpredictable market swings are responsible for the majority of financial gains and losses\cite{taleb2007}, fatalities from the largest wars dwarf those from other wars\cite{newman2005}, and the frequency of the most damaging earthquakes has exceeded past expectations\cite{sornette2009}.
 In ecology, background rates of global extinction are punctuated by mass extinction\cite{harnik2012}, viruses can mutate suddenly to infect new hosts, and billions of animals can die at once in mass mortality events\cite{fey2015}.
 Indeed, catastrophes may be the most important element affecting population persistence\cite{mangel1994} and their importance is likely to increase given projected increases in the frequency and magnitude of climate-related extremes\cite{ipcc2012}.
-Despite this compelling anecdotal evidence for ecological black swans, systematic evidence in population time series of higher taxa has been elusive\cite{keitt1998, allen2001, halley2002} and the vast majority of model fitting and risk forecasting assumes that assumes that uncertainty can be represented by a normal distribution\cite{brook2006a, dennis2006, knape2012}.
+Despite this compelling anecdotal evidence for ecological black swans, systematic evidence in population time series of higher taxa has been elusive\cite{keitt1998, allen2001, halley2002} and the vast majority of model fitting and risk forecasting assumes that uncertainty can be represented by a normal distribution\cite{brook2006a, knape2012}.
 If black swans occur, though, a normal distribution would under-estimate the probability of extreme events occurring.
 
-Here, we assess the frequency and magnitude of black-swan dynamics across time series of 609 populations from a wide array of taxonomic groups---mostly birds, mammals, insects, and fishes.
+Here, we assess the frequency and magnitude of black-swan dynamics across time series of 609 populations from a wide array of taxonomic groups---mostly birds, mammals, insects, and fishes (table~S1) \cite{SOM}.
 We then identify characteristics of time series or life-history traits associated with the detection of black-swan events and verify known causes.
-To accomplish this, we develop a framework for identifying heavy-tailed (black-swan) process noise in population dynamics, i.e.~whether the largest stochastic jumps in abundance from one time step to the next are more extreme than typically seen with a lognormal distribution.
+To accomplish this, we develop a framework for identifying heavy-tailed (black-swan) process noise in population dynamics, i.e.~whether the largest stochastic jumps in log abundance from one time step to the next are more extreme than typically seen with a normal distribution.
 Our framework allows for a range of population dynamic models, can incorporate observation uncertainty and skewness in process noise, and can be readily applied to abundance time series.
 
-Specifically, we fit population dynamics models in which the process noise is drawn from a Student-t distribution\emph{.} By estimating the degrees of freedom parameter, \(\nu\), we can estimate the degree to which the process deviations have heavy tails and are thereof evidence of black-swan events (Fig.~1).
-For example, at \(\nu = 3\), the probability of drawing a value more than five standard deviations below the mean is \(0.02\), whereas the probability of drawing such a value from a normal distribution is tiny (\(2.9\cdot10^-7\)).
+Specifically, we fit population dynamics models in which the process noise is drawn from a Student-t distribution.
+By estimating the degrees of freedom parameter, \(\nu\), we can estimate the degree to which the process deviations have heavy tails and are therefore evidence of black-swan events (Fig.~1, fig.~S\ref{fig:priors}).
+Lower values of \(\nu\) result in heavy-tailed distributions.
+For example, at \(\nu = 3\), the probability of drawing a value more than five standard deviations below the mean is moderately high (\(0.02\)), whereas the probability of drawing such a value from a normal distribution is tiny (\(2.9\cdot10^{-7}\)).
 As \(\nu\) approaches infinity, the distribution approaches the normal distribution (Fig.~1).
 
-We found strong, but rare, evidence for black-swan dynamics: most frequently for birds (7\%) followed by mammals (5\%), and insects (3\%) (Fig.~2, Table~S2).
+We found strong, but rare, evidence for black-swan dynamics: most frequently for birds (7\%) followed by mammals (5\%), and insects (3\%) (Fig.~2).
 Black-swan events were taxonomically widespread, occurring in 38\% of taxonomic orders.
-Accounting for time series length and partially pooling inference across taxonomic class and order with a multilevel model, we found stronger evidence for black swans in insect populations than is apparent in Fig.~2---four of eight orders with the highest median probability of heavy tails were insect orders---however, there was considerable uncertainty in these estimates (Fig.~3a).
+Accounting for time series length and partially pooling inference across taxonomic class and order with a hierarchical model, we found stronger evidence for black swans in insect populations than these statistics suggest---four of eight orders with the highest median probability of heavy tails were insect orders (Fig.~3a).
 
 The majority of our heavy-tailed estimates were robust to alternative population models, observation error, and choice of Bayesian priors.
-Our conclusions were not systematically altered when we included an autocorrelation structure in the residuals, modelled population growth rates with or without density dependence, or modelled the population dynamics as Ricker-logistic instead of Gompertz (Fig.~S3).
-Introducing moderate observation error (CV = 0.2) slightly decreased the estimated prevalence of black-swan events (Fig.~S3).
-The strength of the prior on ν had little influence on estimates of black-swan dynamics (Fig.~S4).
-Our simulation testing shows that, if anything, our models underpredict the true magnitude and probability of heavy tailed events---especially given the length of the time series in our analysis (Figs S5, S6).
+Our conclusions were not systematically altered when we included an autocorrelation structure in the residuals, modelled population growth rates with or without density dependence, or modelled the population dynamics as Ricker-logistic instead of Gompertz (fig.~S\ref{fig:alt}).
+Similarly, introducing moderate observation error (CV = 0.2) only slightly decreased the estimated prevalence of black-swan events (fig.~S\ref{fig:alt}), and the strength of the prior on \(\nu\) had little influence on estimates of black-swan dynamics (fig.~S\ref{fig:alt-priors}).
+Finally, our simulation testing shows that, if anything, our models underpredict the true magnitude and probability of heavy-tailed events---especially given the time series have a median of only 26 years in our analysis (figs~S\ref{fig:sim-nu},~S\ref{fig:sim-prob}).
 
-Across populations, the probability of observing black-swan dynamics was positively related to time-series length and negatively related to magnitude of process noise but not clearly related to population growth rate, density dependence, or maximum lifespan (Figs 3b, 4).
-Longer time-series length was the strongest covariate of observing black-swan dynamics.
-For instance, the expected probability density suggesting black-swan dynamics was about 1.6 times greater for a population with 60 time steps compared to one with 30 time steps.
-However, the absolute change in probability with increased time series length was small (0.20 vs.~0.13 in this example, Fig.~4).
+For model fits to the population data, the probability of detecting black-swan dynamics was positively related to time-series length and negatively related to magnitude of process noise but not clearly related to population growth rate, density dependence, or maximum lifespan (Fig.~3b, fig.~S\ref{fig:correlates}).
+Longer time-series length was the strongest covariate of observing black-swan dynamics: there is a 1.6 times greater probability of detecting a black swan event in populations with 60 time steps than in one with 30 time steps (fig.~S\ref{fig:correlates}).
 
-We examined all time series with published explanations of why black-swan events occurred (Table~S2).
-The majority (87\%) were downward black swans and involved a combination of multiple factors.
+We examined all time series with published explanations of why black-swan events occurred (table~S2).
+The majority (\percBSDown \%) were downward events and involved a combination of multiple factors.
 For example, a synchronization of environmental- and predation-mediated population cycles is thought to have caused a downward black-swan event for a water vole (\emph{Arvicola terrestris}) population\cite{saucy1994}.
@@
 For instance, severe winters in 1929, 1940--1942, and 1962--1963 were associated with black-swan downswings in grey heron (\emph{Ardea cinerea}) abundance in the United Kingdom\cite{stafford1971} (Fig.~4c).
-Our analysis finds that the last event was a combination of two black-swan events in a row and it took the population three times longer to recover than predicted\cite{stafford1971}.
+Our analysis finds that the last event was a combination of two heavy-tailed events in a row and the population took three times longer to recover than predicted\cite{stafford1971}.
 Downward black swans were sometimes followed by upward black swans.
-For example, during a period of population crowding and nest shortages, a population of European shag cormorants (\emph{Phalacrocorax aristotelis}) on the Farne Islands, United Kingdom, declined suddenly following a red tide event in 1968\cite{potts1980}.
-This freed quality nest sites for first-time breeders, productivity rapidly increased, and the population experienced a rapid upswing in abundance\cite{potts1980}.
+For example, during a period of population crowding and nest shortages, a population of European shag cormorants (\emph{Phalacrocorax aristotelis}) on the Farne Islands, United Kingdom declined suddenly following a red tide event in 1968\cite{potts1980}.
+This freed quality nest sites for first-time breeders, productivity increased, and the population experienced a rapid upswing in abundance\cite{potts1980}.
 
-Given the prevalence of downwards events, we refit our heavy-tailed models estimating the skewness of process noise (\(\gamma\); Figs 1a, SX) and used these models to make near-term risk forecasts.
-Aggregated across populations with strong evidence of heavy tails (median \(\nu < 10\)), 85\% of the \(\gamma\) probability density was below 1, indicating strong evidence of downwardly skewed process noise (Fig.~4b, SX).
-In contrast, populations with little evidence of heavy tails (median \(\nu \geq 70\)) also had little evidence of skewed process noise (Fig.~4b; 90\% of 95\% credible intervals overlapped 1).
-Projecting these heavy-tailed populations forward five years revealed that assuming lognormal process noise underestimated risk by X--X fold (ratio of interquartile range of 99\% lower credible interval of abundance).
-\emph{TODO: closing sentence emphasizing the implications of this?}
+Given the prevalence of downwards events, we refit our heavy-tailed models to measure the degree of skewness of the process noise using a skew-t distribution (\(\gamma\); Fig.~4a), and used these models to make near-term risk forecasts.
+Aggregated across populations with strong evidence of heavy tails (median \(\nu < 10\)), 85\% of the \(\gamma\) probability density was below 1, indicating strong evidence of downwardly skewed process noise (Fig.~4b, fig.~S\ref{fig:skew-nu}).
+In contrast, populations that did not have heavy tails (median \(\nu \geq 70\)) had little evidence of skewed process noise (Fig.~4b; 90\% of 95\% credible intervals overlapped 1).
+Projecting these heavy-tailed populations forward five years revealed that assuming the standard normal process noise underestimated risk by \crashUnderRange-fold (interquartile range of 99\% lower credible interval of abundance; Fig.~4c, fig.~S\ref{fig:skew-projections}).
+Thus our analysis provides strong evidence for downward-skewed heavy-tailed events in abundance time series of higher taxa, and ignoring these events will tend to underestimate the risk of population declines.
 
-Thus our analysis provides strong evidence for the presence of downwardly skewed black-swan events in ecological process noise of higher taxa.
-Unmodelled intrinsic properties of populations or extrinsic forces acting on populations could generate black-swan dynamics.
-Given that a mixture of normal distributions with different variances can generate a Student-t distribution with heavy tails\cite{allen2001}, we could observe black-swan dynamics if we miss an underlying mixture of processes.
-Those processes might be an aggregation of populations across space, population diversity\cite{anderson2015}, or an intrinsic change in population variability through time.
-Extrinsic ecological forces could also cause black-swan dynamics\cite{nunez2012}.
-These forces could be extreme themselves.
-For example, extreme climate, predation from (or competition with) other species experiencing black swans, or sharp changes in human pressures such hunting, fishing, or habitat destruction might cause black-swan events.
-Alternatively, the synchrony of multiple extrinsic forces could give rise to black-swan ecological dynamics.
-This could occur with synergistic interactions\cite{kirby2009} or even if non-synergistic forces experience a rare alignment\cite{denny2009}.
+There are a number of possible causes of black swan events, including unmodelled intrinsic properties of populations or extrinsic forces acting on populations.
+For example, we could observe black-swan dynamics if we miss an underlying mixture of processes, since a mixture of normal distributions with different variances can generate a t distribution with heavy tails\cite{allen2001}.
+In ecology, these processes could arise through an aggregation of populations across space, population diversity\cite{schindler2010}, or an intrinsic change in population variability through time\cite{carpenter2006}.
+Extrinsic ecological forces could also cause black-swan dynamics\cite{nunez2012} including extreme climate\cite{meehl2004, katz2005, ipcc2012}, predation from (or competition with) other species experiencing black swans, or sharp changes in human pressures such hunting, fishing, or habitat destruction.
+Alternatively, the synchrony of multiple extrinsic forces could give rise to black-swan dynamics through synergistic interactions\cite{kirby2009} or a rare alignment of non-synergistic forces\cite{denny2009}.
 
-There are a number of caveats when considering the generality of our results.
-The GPDD data represent a taxonomically and geographically biased sample of populations---the longer time series we focus on are dominated by commercially and recreationally important species and a disproportionate number of populations are located in the United Kingdom.
-Although we would expect to find qualitatively similar evidence for black swans in other large taxonomic or geographic samples of populations, the common forces driving those black swans would likely differ.
-Additionally, some apparent black-swan events could be recording mistakes, or conversely, some extreme observations may have been discarded or altered if they were erroneously suspected of being recording mistakes.
-Indeed, we discarded three of the populations that our method initially identified as heavy tailed because they turned out to be data-entry errors (Supporting Information).
-Finally, the temporal scales of observation and population dynamics vary considerably across populations in the GPDD and these likely influence the detection of heavy tails.
-For example, if we make frequent observations relative to generation time (e.g.~for many large-bodied mammals) we will average across generations and perhaps miss black-swan events.
-Conversely, if we census populations infrequently relative to generation time (e.g.~many insects in the GPDD) the recorded data may average across extreme and less-extreme events and also dampen black-swan dynamics.
-
-Recognizing the prevalence of heavy-tailed dynamics suggests a number of policy directions.
-Ecological resource management can draw from other disciplines that focus on heavy tails.
-For example, earthquake preparedness and response is focussed on black-swan events.
-Similarly to ecological black swans, we can rarely predict the specific location and timing of large earthquakes.
+In light of our findings, we suggest that ecological resource management can learn from disciplines that focus on heavy tails.
+For example, earthquake preparedness and response is focused on black-swan events.
+Similarly to ecological black swans, we can rarely predict the specific timing of large earthquakes.
 But, earthquake preparedness involves spatial planning based on forecast probabilities to focus early detection efforts and develop disaster response plans\cite{nrc2007}.
-A similar focus might benefit ecological resource management once our ability to predict the spatial probability and covariates of ecological black swans improves.
 The presence of black swans also suggests that we develop management policy that is robust to heavy tails.
-For instance, setting target population abundances that are appropriately set back from critical limits may buffer black-swan events\cite{caddy1996}, and maintaining genetic, phenotypic, and behavioural diversity may allow some components of populations to persist when others are affected by disease or extreme environmental forces\cite{hilborn2003, schindler2010, anderson2015}.
-Finally, extreme and unexpected, surprising, or counterintuitive ecological dynamics offer a tremendous opportunity to learn about ecological systems, evaluate when models break down, and adjust future management policy\cite{doak2008, pine-iii2009, lindenmayer2010}
+For instance, setting target population abundances far from critical limits may buffer black-swan events\cite{caddy1996}, and maintaining genetic, phenotypic, and behavioural diversity may allow some components of populations to persist when others are affected by disease or extreme environmental forces\cite{schindler2010}.
+Finally, surprising, or counterintuitive ecological dynamics offer a tremendous opportunity to learn about ecological systems, evaluate when models break down, and adjust future management policy\cite{doak2008, lindenmayer2010}.
 
 Rare catastrophes can have a profound influence on population persistence\cite{mangel1994}.
-In recent decades, ecology has moved toward focussing on aspects of variance in addition to mean responses\cite{loreau2010a, thompson2013}.
+In recent decades, ecology has moved toward focusing on aspects of variance in addition to mean responses\cite{loreau2010a, thompson2013}.
 Our results suggest that an added focus on ecological extremes represents the next frontier, particularly in the face of increased climate extremes\cite{meehl2004, ipcc2012, thompson2013}.
-Financial analysts are concerned with the shape of the downward tails of financial returns because these directly impact estimates of risk---the probability of a specific magnitude of undesired event occurring\cite{rachev2008}.
-A comparable focus in ecology would increase our estimates of extinction risk, since these would be disproportionately impacted by downward black-swan events.
+Financial analysts are concerned with the shape of financial return downward tails because these directly impact estimates of risk---the probability of a specific magnitude of undesired event occurring\cite{rachev2008}.
 Similarly, ecologists should focus more on estimating and predicting downward tails of population abundance, since these increase true extinction risk.
@@
 
-\bibliographystyle{nature}\bibliography{jshort,ms}
-
-\textbf{Online Content:} Methods, along with any Extended Data display items are available in the online version of the paper; references unique to these sections appear only in the online paper.
+\bibliographystyle{Science}\bibliography{jshort,ms}
 
 \textbf{Acknowledgements}: We thank J.W.
-Moore, J.D.
+Moore, A.O.
+Mooers, L.R.
+Gerber, J.D.
 Yeakel, C.
@@
 We are grateful to the contributors and maintainers of the Global Population Dynamics Database and to Compute Canada's WestGrid high-performance computing resources.
-Silhouette images were obtained from phylopic.org under Creative Commons licenses; sources are listed in the Supporting Information.
+Silhouette images were obtained from phylopic.org under Creative Commons licenses; sources are listed in the Supplementary Information.
 Funding was provided by a Simon Fraser University Graduate Fellowship (S.C.A.), the Natural Sciences and Engineering Research Council of Canada (N.K.D., A.B.C.), the Canada Research Chairs Program (N.K.D.).
 
-\textbf{Author Contributions}: S.C.A.
+\textbf{Supplementary Materials}\\
+Materials and Methods\\
+References (32 to 72)\\
+Tables S1 to S2\\
+Figs.~S1 to S8\\
+
+%\textbf{Author Contributions}: S.C.A.
 and T.A.B.
 conceived the project; S.C.A., T.A.B., A.B.C., and N.K.D.
-designed the project; S.C.A.
+designed the study; S.C.A.
 analyzed the data and wrote the paper with input from all authors.
 
-\textbf{Author information}: The authors declare no competing financial interests.
-Reprints and permissions information is available at .
+%\textbf{Author information}: The authors declare no competing financial interests.
+Reprints and permissions information is available at www.nature.com/reprints.
 Correspondence and requests for materials should be addressed to S.C.A.
@@
 
-\textbf{Figure~1: An illustration of fitting population dynamic models that allow for heavy tails, represented by the Student-t degrees of freedom parameter} \(\mathbf{\nu}\)\textbf{.
-a, b,} The probability density for Student-t distributions with a scale parameter of 1 and different values of \(\nu\).
+\textbf{Fig.~1: Illustration of population dynamic models that allow for heavy tails.} (\textbf{A} to \textbf{B}) Probability density for the Student-t distribution with scale parameter of 1 and different values of \(\nu\).
 Small values of \(\nu\) create heavy tails.
 As \(\nu\) approaches infinity the distribution approaches the normal distribution.
-For example, at \(\nu = 2\), the probability of drawing a value more than five standard deviations below the mean is 0.02, whereas the probability of drawing such a value from a normal distribution is nearly zero (\(2.9 \cdot 10^{- 7}\)).
-\textbf{c--e,} Simulated population dynamics from a Gompertz model with process noise drawn from t distributions with three different values of \(\nu\).
-Coloured dots in panels \textbf{c} and \textbf{d} represent jumps with less than a 1 in 1000 chance of occurring in a normal distribution.
-\textbf{f--h,} Estimates of \(\nu\) from models fit to the times series in \textbf{c--e}.
-Shown are the posterior samples (histograms), median and interquartile range of the posterior (IQR, dots and line segments), and the exponential prior on \(\nu\) (dashed lines).
-Colour shading behind \textbf{f--h} illustrates the region of heavy tails.
+(\textbf{C} to \textbf{E}) Simulated population dynamics from a Gompertz model with process noise drawn from Student-t distributions with three values of \(\nu\).
+Coloured dots in panels \textbf{C} and \textbf{D} represent jumps with less than a 1 in 1000 chance of occurring in a normal distribution.
+(\textbf{F} to \textbf{H}) Estimates of \(\nu\) from models fit to the times series in \textbf{C} to \textbf{E}.
+Shown are posterior samples (histograms), median and interquartile range of the posterior (IQR, dots and line segments), and the exponential prior on \(\nu\) (dashed lines).
+Colour shading behind \textbf{F} to \textbf{H} illustrates the approximate region of heavy tails.
 
@@
 
-\textbf{Figure~2: Estimates of population dynamic heavy-tailedness for populations of birds, mammals, insects, and fishes.} Small values of \(\nu\) approximately (\(< 10\)) suggest heavy-tailed black-swan dynamics; larger values of \(\nu\) suggest approximately normal-tailed dynamics.
+\textbf{Fig.~2: Estimates of population dynamic heavy-tailedness for populations of birds, mammals, insects, and fishes.} (\textbf{A} to \textbf{D}) Small values of \(\nu\) suggest heavy-tailed black-swan dynamics.
 Vertical points and line segments represent posterior medians and 50\% / 90\% credible intervals for individual populations.
-Inset plots show probability that \(\nu < 10\) (probability of heavy tails) for populations arranged by taxonomic order and sorted by decreasing mean Pr(\(\nu < 10\)).
-Taxonomic orders with three or fewer populations in \textbf{a} are omitted for space.
-Red to yellow points highlight populations with a high to moderately high probability of heavy-tailed black-swan dynamics.
+Inset plots show probability that \(\nu < 10\) for populations arranged by taxonomic order and sorted by decreasing mean Pr(\(\nu < 10\)).
+Taxonomic orders with three or fewer populations in \textbf{A} are omitted for space.
+Red to yellow points highlight populations with a high to moderately high probability of heavy-tailed dynamics.
 
@@
 
-\textbf{Figure~3: Standardized probabilities and covariates of heavy-tailed dynamics.} \textbf{a,} Taxonomic-order-level posterior densities of Pr(\(\nu < 10\)) after accounting for time-series length in a multilevel model.
-Estimates are at the geometric mean of time series length across all the data (approximately 27 time steps).
+\textbf{Fig.~3: Standardized probabilities and covariates of heavy-tailed dynamics.} (\textbf{A}) Taxonomic-order-level posterior densities of Pr(\(\nu < 10\)) after accounting for time-series length in a hierarchical model.
+Dotted vertical line in a indicates the median expected Pr(\(\nu < 10\)) from the prior distribution.
 Colour shading refers to taxonomic class (yellow: fishes, green: insects, purple: birds, and red: mammals).
-Dotted vertical line in \textbf{a} indicates the median expected Pr(\(\nu < 10\)) from the prior distribution.
-\textbf{b,} Posterior densities for potential covariates of Pr(\(\nu < 10\)).
-The beta regression models were fit on a logit scale with hierarchical intercepts for taxonomic class, taxonomic order, and species.
-All covariates were standardized by subtracting their mean and dividing by twice their standard deviation.
+Estimates are standardized at the geometric mean of time series length across all the data (approximately 27 time steps) and shown for orders with \(\ge 5\) populations.
+(\textbf{B}) Posterior densities for potential covariates of Pr(\(\nu < 10\)).
 In both panels, short vertical line segments within the density polygons indicate medians.
@@
 \begin{center}
-\includegraphics[width=0.55\textwidth]{../analysis/skew-fig.pdf}
+\includegraphics[width=0.65\textwidth]{../analysis/skew-fig.pdf}
 \end{center}
 
-\textbf{Figure~4: Evidence for downward skewness in process noise for heavy-tailed populations.
-a,} Illustrations of Student-t distributions with 3 levels of skewness (\(\gamma\)) and heavy-tails (\(\nu\)).
-\textbf{b,} Posterior density of the skewness parameters aggregated across all populations grouped into heavy-tailed populations (\(\hat{\nu} < 10\)), slightly heavy-tailed populations (\(10 \leq \hat{\nu} \leq 70\)), and normal tails (\(\hat{\nu} > 70\)).
-The approximate median posterior density of \(\gamma\) and a mid-value of \(\nu\) from distribution in \textbf{b} are shown in \textbf{a}.
-\textbf{c,} Example time series of heavy-tailed populations with different levels of skewness.
+\textbf{Fig.~4: Heavy-tailed process noise tends to be downward skewed and ignoring this can underestimate risk.} (\textbf{A}) Illustrations of Student-t distributions with 3 levels of skewness (\(\gamma\)) and heavy-tails (\(\nu\)).
+(\textbf{B}) Posterior density of the skewness parameters aggregated across populations grouped into heavy-tailed (\(\hat{\nu} < 10\)), slightly heavy-tailed (\(10 \leq \hat{\nu} \leq 70\)), and normal-tailed populations (\(\hat{\nu} > 70\)).
+Approximate mid-values from \textbf{B} are illustrated in \textbf{A}.
+(\textbf{C}) Example time series of heavy-tailed populations with different levels of skewness.
 Red dots highlight likely heavy-tailed events.
 Forecasts (grey regions) show median (solid lines) and lower 99\% credible intervals (dotted lines) of abundance.
-Black lines indicate forecasts from a standard Gompertz model with lognormal process noise.
-Red lines indicate forecasts from a Gompertz model with estimated tail shape and skewness.
-
-\clearpage
-
-\textbf{Methods}
-
-We selected abundance time series from the Global Population Dynamics Database (GPDD;\cite{gpdd2010}), which contains nearly 5000 time series of abundance from \(\sim 1000\) species and \(\sim 100\) taxonomic orders.
-We filtered the data (Supporting materials) to remove populations from less reliable data sources, and those without sufficient data for our models, and then interpolated some missing values (sensu Brook \textit{et al.}\cite{brook2006a}).
-Our interpolation affected only \(\sim\) \interpPointsPerc \% of the final data points (Supporting materials Table~\ref{tab:stats}) and none of the data points that were later considered black-swan events.
-Our final dataset contained \NPops\ populations across \NOrders\ taxonomic orders and seven taxonomic classes, with a median of \medianTimeSteps\ time steps (range of \minTimeSteps--\maxTimeSteps) (Extended Data Table~\ref{tab:stats}).
-
-We fit heavy-tailed Gompertz population dynamics models to data from the Global Population Dynamics Database (GPDD).
-\emph{{[}TODO something about extensive quality control filtering and interpolation of the time series{]}} The Gompertz model represents population growth as a linear function in log space.
-If \(x_{t}\) represents the log abundance (\(N\)) at time \emph{t} then \(x_t = \lambda + bx_{t - 1} + \epsilon_t,\ \epsilon_t \sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma)\).
-The growth parameter \(\lambda\) represents the expected growth rate if \(N_{t} = 1\).
-The model is density independent if \(b = 1\), maximally density dependent if \(b = 0\), and inversely density dependent if \(b < 0\).
-
-Usually, the process noise \(\epsilon_{t}\) is modelled as normally distributed{[}REFs{]}, but we allow it to be drawn from a Student-t distribution with scale parameter \(\sigma\) and degrees of freedom \(\nu\).
-We can then estimate the degree to which the process deviations have heavy tails and are thereof evidence of black-swan events (Fig.~1a, b).
-
-We chose a weakly informative exponential prior on \(\nu\) with rate parameter of \(0.01\) truncated at values above two---a slightly less informative prior than suggested by Fernandez and Steel\cite{fernandez1998}.
-This prior gives only a \(0.08\) probability that \(\nu < 10\) but constrains sampling sufficiently to avoid drifting towards infinity.
-In any case, for \(\nu > 20\) the t distribution is almost indistinguishable from the normal distribution (Fig.~1).
-Based on the shape of the t distribution, we chose the probability that \(\nu < 10\), Pr(\(\nu < 10\)), to define the probability of heavy-tailed dynamics.
-When categorizing a population as heavy or normal tailed, we applied a 0.5 probability threshold.
-
-\emph{Details on the heavy-tailed Gompertz probability model}
-
-For the Gompertz model, our weakly-informative priors (Extended Data Fig.~\ref{fig:priors}) were: \(b \sim \mathrm{Uniform}(-1, 2)\), \(\lambda \sim \mathrm{Normal}(0, 10^2)\), \(\sigma \sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)\), \(\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2)\).
-Our prior on \(b\) was uninformative between values of \(-1\) and \(2\).
-We would not expect values of \(b\) with levels of density dependence as low as \(-1\) (very strong inverse density dependence), nor would we generally expect values above \(1\).
-We allowed values of \(b\) above \(1\) to allow for non-stationary time series of growth rates.
-The estimates of \(b\) were well within these bounds.
-Our prior on \(\lambda\) was very weakly informative within the range of expected values for population growth and is similar to the default priors suggested by Gelman \emph{et al.}\cite{gelman2008d} for intercepts of regression models.
-Our Half-Cauchy prior on \(\sigma\) follows Gelman \emph{et al.}\cite{gelman2006c} and Gelman \emph{et al.}\cite{gelman2008d} and the specific scale parameter of \(2.5\) is based on our expected range of the value in nature from previous studies\cite{connors2014}.
-In our testing of a subsample of populations, our parameter estimates were not qualitatively changed by switching to an uninformative uniform prior on \(\sigma\), but the weakly informative Half-Cauchy prior sped up chain convergence.
-
-Our prior on \(\nu\) was based on Fernandez and Steel\cite{fernandez1998}.
-They chose an exponential rate parameter of \(0.1\).
-We chose a less informative rate parameter of \(0.01\) and truncated the distribution at \(2\), since at \(\nu < 2\) the variance of the t distribution is undefined.
-This prior gives only a \(7.7\)\% probability that \(\nu < 10\) but constrains the sampling sufficiently to avoid wandering off towards infinity---above approximately \(\nu = 20\) the t distribution is so similar to the normal distribution (Fig.~1) that time series of the length considered here are unlikely to be informative about the precise value of \(\nu\).
-In the scenario where the data are uninformative about heavy tails (e.g.~Fig.~1e, h), the posterior will approximately match the prior (prior median \(= 71\), mean \(= 102\)) and the metrics used in our paper (e.g.
-\(Pr(\nu < 10) > 0.5\)) are unlikely to flag the population as heavy tailed.
-
-We fit our models with Stan 2.4.0\cite{stan-manual2014}, and R 3.1.1\cite{r2014}.
-We began with four chains and \(2000\) iterations, discarding the first \(1000\) as warm up (i.e.~4000 total samples).
-If \(\hat{R}\) (the potential scale reduction factor---a measure of chain convergence) was greater than \(1.05\) for any parameter or the minimum effective sample size, \(n_\mathrm{eff}\), (a measure of the effective number of uncorrelated samples) for any parameter was less than \(200\), we doubled both the total iterations and warm up period and sampled from the model again.
-These thresholds are in excess of the minimums recommended by Gelman \emph{et al.}\cite{gelman2006a} of \(\hat{R} < 1.1\) and effective sample size \(> 100\) for reliable point estimates and confidence intervals.
-In the majority of cases our minimum thresholds were greatly exceeded.
-We continued this procedure up to \(8000\) iterations (\(16000\) total samples) by which all chains were deemed to have sufficiently converged.
-The No-U-Turn Hamiltonian Markov chain Monte Carlo Sampler in Stan generally requires far fewer iterations to obtain equivalent effective sample sizes than other software such as WinBUGS or JAGS\cite{stan-manual2014}.
-
-\emph{The skewed Student's t distribution}
-
-We fit Gompertz models with skewed Student's t distributed process noise based on the probability density for the skewed t distribution described in Fernandez and Steel\cite{fernandez1998}.
-Their definition equates to the following probability density as implemented in King\cite{king2012}:
-
-\begin{equation*}
-  \mathrm{Skewed\,Student\mhyphen t}(x) =
-\begin{dcases}
-\frac{2}{(\gamma + 1/\gamma)}
-         \mathrm{Student\mhyphen t}(x \cdot \gamma, \nu, \mu \cdot \gamma, \sigma),& \text{for } x < 0\\
-       \frac{2}{(\gamma + 1/\gamma)}
-         \mathrm{Student\mhyphen t}(x / \gamma, \nu, \mu / \gamma, \sigma),& \text{for } x \ge 0
-\end{dcases}
-\end{equation*}
-
-Here \(\gamma\) represents the skewness parameter.
-The Student-t(\(\cdot\)) portion represents the data point of interest, the heavy-tailedness parameter, the location parameter, and the scale parameter as fed to the standard Student-t probability density.
-The distribution is symmetrical if \(\gamma = 1\), left skewed (skewed downwards in abundance) if \(0 < \gamma < 1\), and right skewed (skewed upwards in abundance) if \(1 < \gamma < \infty\).
-
-\emph{Alternative priors}
-
-To test if the prior on \(\nu\) influenced our estimate of black-swan dynamics, we refit our models with weaker and stronger priors.
-Our base model used a prior on \(\nu\) of Truncated-Exponential(0.02, min.\ = 2).
-For a weaker prior we used Truncated-Exponential(0.005, min.\ = 2) and for a stronger prior we used Truncated-Exponential(0.02, min.\ = 2) (Extended Data Fig.~\ref{fig:priors}).
-Note that the base and weaker priors are relatively flat within the region of \(\nu < 20\), which is the region we are mostly concerned about when categorizing populations as heavy- or thin-tailed.
-
-Our results show that these weaker and stronger priors would have little influence on our conclusions about heavy-tailed dynamics (Extended Data Fig.~\ref{fig:alt-priors}).
-When the data are informative about tail behaviour (i.e.
-when there is strong evidence of low \(\nu\) values, upper-right of Extended Data Fig.~\ref{fig:alt-priors}), the prior has little impact on the estimate of \(\nu\).
-When the data are less informative about \(\nu\) (i.e.\ when there are no or few tail events and time series are short or noisy), the prior can pull the estimate of \(\nu\) towards larger or smaller values (Extended Data Fig.~\ref{fig:alt-priors}).
-The vast majority of the populations with Pr\((\nu < 10)\) in the base prior were not altered qualitatively by this range of prior strength.
-
-\emph{Alternative population models}
-
-We fit alternative population models to test if four key phenomena systematically changed our conclusions.
-These models allowed for AR1 autocorrelation of the residuals, density independence, observation error, and Ricker-logistic population dynamics.
-Furthermore, we tested the sensitivity of our results to weaker and stronger priors on \(\nu\) and used simulated data to test how easily we could detect \(\nu\) given different sample sizes and to ensure we could recover unbiased parameter estimates.
-The range of percentages of black swans by taxonomic class cited in the abstract and results are based on lower and upper limits across our main Gompertz model and these four alternative models.
-
-\emph{Autocorrelated residuals}
-
-We considered a version of the Gompertz model in which an autoregressive parameter was fit to the process noise residuals: \(x_t = \lambda + b x_{t-1} + \epsilon_t,\, \epsilon_t \sim \mathrm{Student\mhyphen t}(\nu, \phi \epsilon_{t-1}, \sigma)\).
-In addition to the parameters in the original Gompertz model, this model estimates an additional parameter \(\phi\), which represents the correlation of subsequent residuals.
-Based on the results of previous analyses with the GPDD\cite{connors2014} and the chosen priors in previous analyses\cite{thorson2014a} and to greatly speed up chain convergence when running our model across all populations, we placed a weakly informative prior on \(\phi\) that assumed the greatest probability density near zero with the reduced possibility of \(\phi\) being near \(-1\) or \(1\).
-Specifically, we chose \(\phi \sim \mathrm{Truncated\mhyphen Normal}(0, 1, \mathrm{min.} = -1, \mathrm{max.} = 1)\).
-The MCMC chains did not converge for \modelsNoConvergeAROne\ populations according to our criteria (\(\widehat{R} < 1.05, n_\mathrm{eff} > 200\)) after 8000 iterations of four chains.
-This included only \modelsNoConvergeAROneHeavyBase\ populations in which Pr(\(\nu < 10\)) \(> 0.5\) categorized them as heavy in the main Gompertz model.
-We did not include these models in Extended Data Fig.~\ref{fig:alt}.
-
-\emph{Assumed density independence}
-
-We fit a simplified version of the Gompertz model in which the density dependence parameter \(b\) was fixed at \(1\) (density independent).
-This is equivalent to fitting a random walk model (with drift) to the \(\log\) abundances or assuming the growth rates are drawn from a stationary distribution.
-The model was \(x_t = \lambda + x_{t-1} + \epsilon_t,\, \epsilon \sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma)\).
-We fit this model for three reasons: (1) it is computationally simpler and so provides a check that our more complicated full Gompertz model was obtaining reasonable estimates of \(\nu\), (2) it provides a test of whether density dependence was systematically affecting our perception of heavy tails, (3) it matches how some previous authors have modelled heavy tails without accounting for density dependence\cite{segura2013}.
-
-\emph{Assumed observation error}
-
-Observation error can bias parameter estimates\cite{knape2012} and is known to affect the ability to detect extreme events\cite{ward2007}.
-In our main analysis, we fit a model that ignored observation error.
-One way to account for observation error would be to fit a full state-space model that simultaneously estimates the magnitude of process noise and observation error.
-However, simultaneously estimating observation and process noise is a challenging problem (e.g.\ because the observation and process noise parameters tend to negatively covary in model fitting) and is known to sometimes result in identifiability issues with the Gompertz population model\cite{knape2008}.
-Furthermore, our model was applied to hundreds of time series, often of short length (as few as 20 time steps) and our model estimates an additional parameter---the shape of the process deviation tails---potentially making identifiability and computational issues even greater.
-Therefore, we considered a version of the base Gompertz model that allowed for a fixed level of observation error:
-
-\begin{align*}
-U_t &= \lambda + b U_{t-1} + \epsilon_t\\
-x_t &\sim \mathrm{Normal}(U_t, \sigma_\mathrm{obs}^2)\\
-\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma_\mathrm{proc}),
-\end{align*}
-
-where \(U\) represents the unobserved state vector, \(\sigma_\mathrm{obs}\) represents the standard deviation of observation error (on a log scale), and \(\sigma_\mathrm{proc}\) represent the process noise scale parameter.
-We set \(\sigma_\mathrm{obs}\) to \(0.2\), which represents the upper limit of values often used in simulation analyses\cite{valpine2002, thorson2014b}.
-
-\emph{Ricker-logistic}
-
-We also fit a Ricker-logistic model: \(x_t = x_{t-1} + r_{\mathrm{max}}\left(1 - N_{t-1}/K\right) + \epsilon_t,\, \epsilon_t \sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma)\), where \(r_\mathrm{max}\) represents the theoretical maximum population growth rate that is obtained when \(N_t\) (abundance at time \(t\)) \(= 0\).
-The parameter \(K\) represents the carrying capacity and, as before, \(x_t\) represents the \(\log\) transformed abundance at time \(t\).
-The Ricker-logistic model assumes a linear decrease in population growth rate with increases in abundance.
-In contrast, the Gompertz model assumes a linear decrease in population growth rate with increases in \textit{log} abundance (\(x_t\))\cite{thibaut2012}.
-
-To fit the Ricker-logistic models, we chose a prior on \(K\) uniform between zero and twice the maximum observed abundance Clark \emph{et al.}\cite{clark2010} chose uniform between zero and maximum observed, which is more informative).
-We set the prior on \(r_\mathrm{max}\) as uniform between 0 and 20 as in Clark \emph{et al.}\cite{clark2010}.
-We used the same priors on \(\nu\) and \(\sigma\) as in the Gompertz model.
-
-\emph{Modelling covariates of heavy-tailed dynamics}
-
-We fit a multilevel beta regression model to the predicted probability of heavy tails, Pr(\(\nu < 10\)), to investigate potential covariates of heavy-tailed dynamics.
-The beta distribution is useful when response data range on a continuous scale between zero and one\cite{ferrari2004}.
-We used a logit link function as is typically used in logistic regression.
-The model was as follows:
-
-\begin{align*}
-\mathrm{Pr}(\nu_i < 10) &\sim \mathrm{Beta}(A_i, B_i)\\
-\mu_i &= \mathrm{logit}^{-1}(\alpha
-  + \alpha^\mathrm{class}_{j[i]}
-  + \alpha^\mathrm{order}_{k[i]}
-  + \alpha^\mathrm{species}_{l[i]}
-  + X_i \beta),
-  \: \text{for } i = 1, \dots, 617\\
-A_i &= \phi_\mathrm{disp} \mu_i\\
-B_i &= \phi_\mathrm{disp} (1 - \mu_i)\\
-\alpha^\mathrm{class}_j &\sim
-  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{class}}),
-  \: \text{for } j = 1, \dots, 6\\
-\alpha^\mathrm{order}_k &\sim
-  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{order}}),
-  \: \text{for } k = 1, \dots, 38\\
-\alpha^\mathrm{species}_l &\sim
-  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{species}}),
-  \: \text{for } l = 1, \dots, 301,
-\end{align*}
-
-where \(A\) and \(B\) represent the beta distribution shape parameters; \(\mu_i\) represents the predicted value for population \(i\), class \(j\), order \(k\), and species \(l\); \(\phi_\mathrm{disp}\) represents the dispersion parameter; and \(X_i\) represents a vector of predictors (such as lifespan) for population \(i\) with associated \(\beta\) coefficients.
-The intercepts are allowed to vary from the overall intercept \(\alpha\) by taxonomic class (\(\alpha^\mathrm{class}_j\)), taxonomic order (\(\alpha^\mathrm{order}_k\)), and species (\(\alpha^\mathrm{species}_l\)) with standard deviations \(\sigma_{\alpha \;  \mathrm{class}}\), \(\sigma_{\alpha \; \mathrm{order}}\), and \(\sigma_{\alpha  \; \mathrm{species}}\).
-Where possible, we also allowed for error distributions around the predictors by incorporating the standard deviation of the posterior samples for the Gompertz parameters \(\lambda\), \(b\), and \(\log \sigma\) around the mean point value as normal distributions (not shown in the above equation).
-
-We log transformed \(\sigma\), time-series length, and lifespan to match the way they are visually represented in Extended Data Fig.~\ref{fig:correlates} and to make the relationship approximately linear on the logit-transformed response scale.
-All input variables were standardized by subtracting their mean and dividing by two standard deviations to make their coefficients comparable in magnitude \cite{gelman2008c}.
-We excluded body length as a covariate because it was highly correlated with lifespan, and lifespan exhibited more overlap across taxonomy than body length.
-Lifespan is also more directly related to time and potential mechanisms driving black-swan dynamics.
-
-We incorporated weakly informative priors into our model: \(\mathrm{Cauchy}(0, 10)\) on the global intercept \(\alpha\), \(\mathrm{Half\mhyphen Cauchy}(0, 2.5)\) on all standard deviation parameters, \(\mathrm{Half\mhyphen Cauchy}(0, 10)\) on the dispersion parameter \(\phi_\mathrm{disp}\), and \(\mathrm{Cauchy}(0, 2.5)\) on all other parameters \cite{gelman2006c, gelman2008d}.
-Compared to normal priors, the Cauchy priors concentrate more probability density around expected parameter values while allowing for a higher probability density far into the tails, thereby allowing the data to dominate the posterior more strongly if it disagrees with the prior.
-Our conclusions were not qualitatively changed by using uniform priors.
-We fit our models with 5000 total iterations per chain, 2500 warm-up iterations, four chains, and discarding every second sample to save memory.
-We checked for chain convergence visually and with the same criteria as before (\(\widehat{R} < 1.05\) and \(n_\mathrm{eff} >200\) for all parameters).
-
-To derive taxonomic-order-level estimates of the probability of heavy tails accounting for time-series length (Fig.~3a), we fit a separate multilevel model with the same structure but with only \(\log\) time-series length as a predictor.
-(In this case, we did not want to control for intrinsic population characteristics such as density dependence.) Since our predictors were centered by subtracting their mean value, we obtained order-level estimates of the probability of heavy tails at mean log time-series length by adding the posteriors for \(\alpha\), \(\alpha^\mathrm{class}_j\), and \(\alpha^\mathrm{order}_k\).
-
-\clearpage
-
-\textbf{Extended Data Tables}
-
-\newenvironment{helvetica}{\fontfamily{phv}\selectfont}{\par}
-
-\singlespacing
-
-\textbf{Extended Data Table~1: Summary statistics for the filtered Global Population Dynamics Database
-time series arranged by taxonomic class.} Columns are: number of populations, number of taxonomic orders, numbers of species, median time series length, total number of interpolated time steps, total number of substituted zeros, and total number of time steps.
-
-\onehalfspacing
-
-\begin{helvetica}
-\smallskip
-\begin{scriptsize}
-\begin{tabular}{lrrrrrrrr}
-\toprule
-\input{../analysis/stat-table.tex}
-\label{tab:stats}
-\end{tabular}
-\end{scriptsize}
-\end{helvetica}
-
-\clearpage
-
-\LTcapwidth=\textwidth
-
-\renewcommand{\tablename}{\textbf{Extended Data Table}}
-
-\setcounter{table}{1}
-
-\begin{helvetica}
-\singlespacing
-\begin{scriptsize}
-\begin{longtable}{>{\RaggedRight}m{1.4cm}>{\RaggedRight}p{3.3cm}>{\RaggedRight}p{0.7cm}>{\RaggedRight}p{1.5cm}>{\RaggedRight}p{3.0cm}>{\RaggedRight}p{1.4cm}>{\RaggedRight}p{1.25cm}}
-
-\caption{\textbf{All populations with Pr$(\nu < 10) > 0.5$ in the base heavy-tailed
-  Gompertz population dynamics model.} Shown are the log abundance time series,
-  population descriptions, Global Population Dynamics Database Main IDs,
-  citation for the data source or separate verification literature, a
-  description of the cause of the black swan events (if known), the
-  probability of heavy tails as calculated by our model, and median estimate
-  of $\nu$ from our model with 90\% quantile credible intervals indicated in
-  parentheses. Red dots on the time series indicate downward black-swan events
-  and blue values indicate upward black-swan events that have a $10^{-4}$ probability or less of occurring if the population dynamics were
-  explained by a Gompertz model with normally distributed process noise with a
-  standard deviation equal to the scale parameter in the fitted t
-  distribution.}\\
-
-\toprule
-\input{../analysis/cause-table.tex}
-\label{tab:causes-supp}
-\end{longtable}
-\end{scriptsize}
-\onehalfspacing
-\end{helvetica}
-
-\clearpage
-
-\textbf{Extended Data Figures}
-
-\renewcommand{\figurename}{\textbf{Extended Data Fig.}}
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=0.6\textwidth]{../analysis/priors-gomp-base.pdf}
-
-\caption{
-  \textbf{Probability density of the Bayesian priors for the Gompertz models.} From
-  left to right and then top to bottom: (1) per capita growth rate at
-  $\log$(abundance) = $0$: $\lambda \sim \mathrm{Normal}(0, 10^2)$; (2) scale
-  parameter of t-distribution process noise: $\sigma \sim \mathrm{Half\mhyphen
-    Cauchy} (0, 2.5)$; (3) t-distribution degrees of freedom parameter: $\nu
-  \sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2)$; (4)
-  AR1 correlation coefficient of residuals: $\phi \sim \mathrm{Truncated
-    \mhyphen Normal}(0, 1, \mathrm{min.} = -1, \mathrm{max.} = 1)$. Not shown
-  is $b$, the density dependence parameter: $b \sim \mathrm{Uniform}(-1, 2)$.
-  The $\nu$ panel also shows two alternative priors: a weaker prior $\nu \sim
-  \mathrm{Truncated\mhyphen Exponential}(0.005, \mathrm{min.} = 2)$, and a
-  stronger prior $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.02,
-  \mathrm{min.} = 2)$. The inset panel shows the same data but with a
-  log-transformed x axis. Note that the base and weaker priors are relatively
-  flat within the region of $\nu < 20$ that we are concerned with. }
-
-\label{fig:priors}
-\end{center}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=0.8\textwidth]{../analysis/gomp-comparison.pdf}
-
-\caption{\textbf{Estimates of $\nu$ from alternative models
-  plotted against the base Gompertz model estimates of $\nu$.} Shown are
-  medians of the posterior (dots) and 50\% credible intervals (segments). The
-  diagonal line indicates a one-to-one relationship. Different colours
-  indicate various taxonomic classes. The grey-shaded regions indicate regions
-  of disagreement if $\nu = 10$ is taken as a threshold of heavy-tailed
-  dynamics. The Gompertz observation error model assumes a fixed standard
-  deviation of observation error of $0.2$ on a log scale.}
-
-\label{fig:alt}
-\end{center}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=0.8\textwidth]{../analysis/gomp-prior-comparison.pdf}
-
-\caption{\textbf{Estimates of $\nu$ from Gompertz models with alternative priors on
-  $\nu$.} Shown are medians of the posterior (dots) and 50\% credible intervals
-  (segments). The diagonal line indicates a one-to-one relationship. Different
-  colours indicate various taxonomic classes. The grey-shaded regions indicate
-  regions of disagreement if $\nu = 10$ is taken as a threshold of
-  heavy-tailed dynamics. The base, weaker, and stronger priors on $\nu$ are
-  illustrated in Fig.~\ref{fig:priors}. In general, the estimates are nearly
-  identical in cases where the data are informative about low values of $\nu$.
-  When the data are less informative about low values of $\nu$, the prior can
-  slightly pull the estimates of $\nu$ towards higher or lower values.}
-
-\label{fig:alt-priors}
-\end{center}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=0.8\textwidth]{../analysis/correlates-p10.pdf}
-
-\caption{\textbf{Potential covariates of heavy-tailed population dynamics (indicated
-  by a high probability that $\nu < 10$).} Shown are (a--c) parameters from the
-  Gompertz heavy-tailed population model ($\sigma$, $\lambda$, $b$), (b)
-  number of time steps, (c) body length, and (d) lifespan. For the Gompertz
-  parameters, $\sigma$ refers to the scale parameter of the Student-t
-  process-noise distribution, $\lambda$ refers to the expected log abundance
-  at the next time step at an abundance of one, $b$ refers to the density
-  dependence parameter ($1$ is maximally density independent, $0$ is maximally
-  density dependent, and $<0$ is inversely density dependent). Circles
-  representing a few sharks, crustaceans, and gastropods are filled in white.
-  Median and 90\% credible interval posterior predictions of a beta regression
-  multilevel model are shown in panels a and d where there was a high
-  probability the slope coefficient was different from zero
-  (Fig.~3b).}
-
-\label{fig:correlates}
-\end{center}
-\end{figure}
+Black and red lines indicate forecasts from Gompertz models with lognormal and skew-t process noise, respectively.

### 2016-01-02 text/ms.tex
edits

@@
 \textbf{Abstract}:
-Black swans are statistically improbable events that nonetheless occur---often with profound consequences\cite{taleb2007, sornette2009}.
-While extremes in the physical environment, such as monsoons and heat waves, are widely studied and increasing in magnitude and frequency\cite{meehl2004, katz2005, ipcc2012}, it remains unclear the extent to which ecological populations buffer or suffer from such extremes.
-Here we derive a model for estimating the frequency and direction of black-swan events and apply it to abundance time series of 609 animal populations.
-We find strong evidence for black-swan events, but they are rare, occurring in \(\sim\)\overallBasePerc \% of populations: most frequently for birds (\birdPH \%), mammals (\mammalsPH \%), and insects (\insectsPH \%).
-Black-swan events are predominantly (\percBSDown \%) downward, implying that unexpected population crashes occur far more frequently than unexpected population increases.
-Black-swan events are not explained by any intrinsic life-history covariates, but tend to be driven by extrinsic perturbations such as climate or severe winters.
-Data and simulations reveal that black-swan events are more likely to be detected in longer and less noisy time series.
-Rare events tend to be crashes and ignoring them will underestimate extinction risk.
-Ignoring black-swan events leads to a \crashUnderMedian-fold (50\% interval \crashUnderRange) underestimate in the magnitude of population crashes.
-Our results demonstrate the critical importance of allowing for downward and heavy-tailed events in population modelling and developing robust conservation and management strategies, particularly in the face of increased climate extremes.
+Black swans are improbable events that nonetheless occur---often with profound
+consequences. While physical extremes, such as monsoons and heat waves, are
+widely studied and increasing in magnitude and frequency, it remains unclear
+the extent to which ecological populations buffer or suffer from such
+extremes. Here we estimate the frequency and direction of black-swan events in
+609 animal populations. We find strong evidence for black-swan events
+occurring in \(\sim\)\overallBasePerc \% of populations: most frequently for
+birds (\birdPH \%), mammals (\mammalsPH \%), and insects (\insectsPH \%).
+Black-swan events are predominantly population crashes (\percBSDown \%),
+driven by extrinsic perturbations such as climate or severe winters, and
+ignoring them may underestimate extinction risk. Our results demonstrate the
+importance of allowing for heavy-tailed downward events in population
+modelling and developing robust conservation strategies, particularly in the
+face of increased climate extremes.
 
-\bigskip
-\bigskip
+\textbf{One Sentence Summary}: Black swans---statistically improbable events
+with profound consequences---can and do occur in animal populations and
+ignoring them may severely underestimate extinction risk.
 
-\textbf{One Sentence Summary}:
+%\clearpage
 
@@
 We are grateful to the contributors and maintainers of the Global Population Dynamics Database and to Compute Canada's WestGrid high-performance computing resources.
-Silhouette images were obtained from phylopic.org under Creative Commons licenses; sources are listed in the Supplementary Information.
+Silhouette images were obtained from phylopic.org under Creative Commons licenses; sources are listed in the Supplementary Materials.
 Funding was provided by a Simon Fraser University Graduate Fellowship (S.C.A.), the Natural Sciences and Engineering Research Council of Canada (N.K.D., A.B.C.), the Canada Research Chairs Program (N.K.D.).

### 2016-01-02 text/supp.tex
edits

@@
 
-\title{Supplementary Information:\\Evidence for black-swan events in animal
+\title{Supplementary Materials:\\Evidence for black-swan events in animal
   populations}
@@
 We selected abundance time series from the Global Population Dynamics Database\cite{gpdd2010} (GPDD), which contains nearly 5,000 time series of abundance from \(\sim\) 1,000 species and \(\sim\) 100 taxonomic orders.
-We filtered the data (Supplementary Information) to remove populations from less reliable data sources, and those without sufficient data for our models, and then interpolated some missing values (sensu Brook \textit{et al.}\cite{brook2006a}).
+We filtered the data to remove populations from less reliable data sources, and those without sufficient data for our models, and then interpolated some missing values (sensu Brook \textit{et al.}\cite{brook2006a}).
 Our interpolation affected only \(\sim\) \interpPointsPerc \% of the final data points (table~S1) and none of the data points that were later considered black-swan events.
@@
 Additionally, some apparent black-swan events could be recording mistakes, or conversely, some extreme observations may have been discarded or altered if they were erroneously suspected of being recording mistakes.
-Indeed, we discarded three of the populations that our method initially identified as heavy tailed because they turned out to be data-entry errors (Supplementary Information).
+Indeed, we discarded three of the populations that our method initially identified as heavy tailed because they turned out to be data-entry errors.
 Finally, the temporal scales of observation and population dynamics vary considerably across populations in the GPDD and these likely influence the detection of heavy tails.
@@
 
-\section{Simulation testing} We performed two types of simulation testing (Supporting Information).
+\section{Simulation testing} We performed two types of simulation testing.
 First, we tested how easily the Student-t distribution \(\nu\) parameter could be recovered given different true values of \(\nu\) and different sample sizes.
@@
 
-\section{Skewed Student-t forecasts} To formally evaluate the apparent skewness of heavy-tailed process noise we fit Gompertz models with skewed Student-t distributed process noise\cite{fernandez1998} (Supplementary Information).
+\section{Skewed Student-t forecasts} To formally evaluate the apparent skewness of heavy-tailed process noise we fit Gompertz models with skewed Student-t distributed process noise\cite{fernandez1998}.
 The distribution adds one parameter to the Student-t distribution, \(\gamma\), which controls the skewness.

### 2016-02-05 text/ms.tex
Incorporate Nick and Trevor's comments

@@
-\textbf{Title: Evidence for black-swan events in animal populations}
+\textbf{Title: Black-swan events in animal populations}
 
@@
 \textbf{Abstract}:
-Black swans are improbable events that nonetheless occur---often with profound
-consequences. While physical extremes, such as monsoons and heat waves, are
-widely studied and increasing in magnitude and frequency, it remains unclear
-the extent to which ecological populations buffer or suffer from such
-extremes. Here we estimate the frequency and direction of black-swan events in
-609 animal populations. We find strong evidence for black-swan events
-occurring in \(\sim\)\overallBasePerc \% of populations: most frequently for
-birds (\birdPH \%), mammals (\mammalsPH \%), and insects (\insectsPH \%).
-Black-swan events are predominantly population crashes (\percBSDown \%),
-driven by extrinsic perturbations such as climate or severe winters, and
-ignoring them may underestimate extinction risk. Our results demonstrate the
-importance of allowing for heavy-tailed downward events in population
-modelling and developing robust conservation strategies, particularly in the
-face of increased climate extremes.
+Black swans are improbable events that nonetheless occur---often with profound consequences.
+While physical extremes, such as monsoons and heat waves, are widely studied and increasing in magnitude and frequency, it remains unclear the extent to which ecological population numbers buffer or suffer from such extremes.
+Here we estimate the frequency and direction of black-swan events in 609 animal populations, finding strong evidence for their occurrence in \(\sim\)\overallBasePerc \% of populations.
+Black-swan events occur most frequently for birds (\birdPH \%), mammals (\mammalsPH \%), and insects (\insectsPH \%), and manifest primarily as   population die-offs and crashes (\percBSDown \%) driven by extrinsic perturbations from climate variability.
+Thus, ignoring black-swan events will under-estimate extinction risk.
+Our results demonstrate the importance of both modeling heavy-tailed downward events in populations, and developing conservation strategies that are robust to future increases in climate extremes.
 
-\textbf{One Sentence Summary}: Black swans---statistically improbable events
-with profound consequences---can and do occur in animal populations and
-ignoring them may severely underestimate extinction risk.
+\textbf{One Sentence Summary}: Black swans---statistically improbable events with profound consequences---occur in animal populations and understanding them will improve climate risk assessments.
 
@@
 Major surprises happen more often than expected in financial, social, and natural systems\cite{taleb2007, sornette2009, may2008}.
-Massive unpredictable market swings are responsible for the majority of financial gains and losses\cite{taleb2007}, fatalities from the largest wars dwarf those from other wars\cite{newman2005}, and the frequency of the most damaging earthquakes has exceeded past expectations\cite{sornette2009}.
-In ecology, background rates of global extinction are punctuated by mass extinction\cite{harnik2012}, viruses can mutate suddenly to infect new hosts, and billions of animals can die at once in mass mortality events\cite{fey2015}.
-Indeed, catastrophes may be the most important element affecting population persistence\cite{mangel1994} and their importance is likely to increase given projected increases in the frequency and magnitude of climate-related extremes\cite{ipcc2012}.
-Despite this compelling anecdotal evidence for ecological black swans, systematic evidence in population time series of higher taxa has been elusive\cite{keitt1998, allen2001, halley2002} and the vast majority of model fitting and risk forecasting assumes that uncertainty can be represented by a normal distribution\cite{brook2006a, knape2012}.
+Massive unpredictable market swings are responsible for the majority of financial gains and losses\cite{taleb2007}, fatalities from the largest wars dwarf those from all others \cite{newman2005}, and the frequency of the most damaging earthquakes has exceeded past expectations\cite{sornette2009}.
+In ecological systems, background rates of global extinction are punctuated by mass extinction\cite{harnik2012}, viruses can mutate suddenly to infect new hosts, and billions of animals can die at once in mass mortality events\cite{fey2015}.
+Indeed, such die-offs may be the most important element affecting population persistence\cite{mangel1994} and their importance is likely to increase given projected increases in the frequency and magnitude of climate-related extremes\cite{ipcc2012}.
+Despite this anecdotal evidence for ecological black swans, systemtic searches of large numbers of timeseries have yet to reveal black-swan events\cite{keitt1998, allen2001, halley2002} However, the overwhelming majority of model fitting and risk forecasting assumes that deviations from model predictions can be represented by a normal distribution\cite[e.g.][]{brook2006a, knape2012}.
 If black swans occur, though, a normal distribution would under-estimate the probability of extreme events occurring.
 
-Here, we assess the frequency and magnitude of black-swan dynamics across time series of 609 populations from a wide array of taxonomic groups---mostly birds, mammals, insects, and fishes (table~S1) \cite{SOM}.
-We then identify characteristics of time series or life-history traits associated with the detection of black-swan events and verify known causes.
+Here, we develop a new approach to estimate the frequency and magnitude of black-swan dynamics across time series of 609 populations from a wide array of taxonomic groups---including many  birds, mammals, insects, and fishes (table~S1) \cite{SOM}.
+We identify characteristics of time series or life-history traits associated with the detection of black-swan events and verify known causes.
 To accomplish this, we develop a framework for identifying heavy-tailed (black-swan) process noise in population dynamics, i.e.~whether the largest stochastic jumps in log abundance from one time step to the next are more extreme than typically seen with a normal distribution.
@@
 Specifically, we fit population dynamics models in which the process noise is drawn from a Student-t distribution.
-By estimating the degrees of freedom parameter, \(\nu\), we can estimate the degree to which the process deviations have heavy tails and are therefore evidence of black-swan events (Fig.~1, fig.~S\ref{fig:priors}).
+By estimating the degrees of freedom parameter, \(\nu\), we can estimate the degree to which the process deviations have heavy tails and are therefore diagnositc of black-swan events (Fig.~1, fig.~S\ref{fig:priors}).
 Lower values of \(\nu\) result in heavy-tailed distributions.
-For example, at \(\nu = 3\), the probability of drawing a value more than five standard deviations below the mean is moderately high (\(0.02\)), whereas the probability of drawing such a value from a normal distribution is tiny (\(2.9\cdot10^{-7}\)).
+For example, at \(\nu = 2\), the probability of drawing a value more than five standard deviations below the mean is almost 70,000 times higher than the probability of drawing such a value from a normal distribution (\(0.02\) vs.\ \(2.9\cdot10^{-7}\)).
 As \(\nu\) approaches infinity, the distribution approaches the normal distribution (Fig.~1).
 
-We found strong, but rare, evidence for black-swan dynamics: most frequently for birds (7\%) followed by mammals (5\%), and insects (3\%) (Fig.~2).
+We found evidence that black-swan dynamics are infreqent but highly influentiual: they occur most frequently for birds (7\%), mammals (5\%), and insects (3\%) and almost never in fishes for this dataset (Fig.~2).
 Black-swan events were taxonomically widespread, occurring in 38\% of taxonomic orders.
@@
 For model fits to the population data, the probability of detecting black-swan dynamics was positively related to time-series length and negatively related to magnitude of process noise but not clearly related to population growth rate, density dependence, or maximum lifespan (Fig.~3b, fig.~S\ref{fig:correlates}).
-Longer time-series length was the strongest covariate of observing black-swan dynamics: there is a 1.6 times greater probability of detecting a black swan event in populations with 60 time steps than in one with 30 time steps (fig.~S\ref{fig:correlates}).
+Longer time-series length was the strongest covariate of observing black-swan dynamics: there is a 1.6 times greater probability of detecting a black-swan event in populations with 60 time steps than in one with 30 time steps (fig.~S\ref{fig:correlates}).
 
-We examined all time series with published explanations of why black-swan events occurred (table~S2).
-The majority (\percBSDown \%) were downward events and involved a combination of multiple factors.
+The majority of black-swan events (\percBSDown \%) were downward (die-offs) rather than upward (unexpectedly rapid abundance increases).
+Of the black-swan events with published explanations (table~S2), the majority involved a combination of multiple factors.
 For example, a synchronization of environmental- and predation-mediated population cycles is thought to have caused a downward black-swan event for a water vole (\emph{Arvicola terrestris}) population\cite{saucy1994}.
@@
 For example, during a period of population crowding and nest shortages, a population of European shag cormorants (\emph{Phalacrocorax aristotelis}) on the Farne Islands, United Kingdom declined suddenly following a red tide event in 1968\cite{potts1980}.
-This freed quality nest sites for first-time breeders, productivity increased, and the population experienced a rapid upswing in abundance\cite{potts1980}.
+This freed up quality nest sites for first-time breeders, productivity increased, and the population experienced a rapid upswing in abundance\cite{potts1980}.
 
-Given the prevalence of downwards events, we refit our heavy-tailed models to measure the degree of skewness of the process noise using a skew-t distribution (\(\gamma\); Fig.~4a), and used these models to make near-term risk forecasts.
+Given the prevalence of downwards events, we refit our heavy-tailed models to measure the degree of skewness \(\gamma\) of the process noise using a skew-t distribution (Fig.~4a), and used these models to make near-term risk forecasts.
 Aggregated across populations with strong evidence of heavy tails (median \(\nu < 10\)), 85\% of the \(\gamma\) probability density was below 1, indicating strong evidence of downwardly skewed process noise (Fig.~4b, fig.~S\ref{fig:skew-nu}).
@@
 
-There are a number of possible causes of black swan events, including unmodelled intrinsic properties of populations or extrinsic forces acting on populations.
+% TODO How are you defining risk. See if you can make this more concrete to avoid IUCN people getting upset with lack of specificity (i.e. Resit akcakaya)
+
+There are a number of possible causes of black-swan events, including unmodelled intrinsic properties of populations or extrinsic forces acting on populations.
 For example, we could observe black-swan dynamics if we miss an underlying mixture of processes, since a mixture of normal distributions with different variances can generate a t distribution with heavy tails\cite{allen2001}.
 In ecology, these processes could arise through an aggregation of populations across space, population diversity\cite{schindler2010}, or an intrinsic change in population variability through time\cite{carpenter2006}.
-Extrinsic ecological forces could also cause black-swan dynamics\cite{nunez2012} including extreme climate\cite{meehl2004, katz2005, ipcc2012}, predation from (or competition with) other species experiencing black swans, or sharp changes in human pressures such hunting, fishing, or habitat destruction.
+Extrinsic ecological forces could also cause black-swan dynamics\cite{nunez2012} including extreme climate events\cite{meehl2004, katz2005, ipcc2012}, predation from (or competition with) other species experiencing black swans, or sharp changes in human pressures such hunting, fishing, or habitat destruction.
 Alternatively, the synchrony of multiple extrinsic forces could give rise to black-swan dynamics through synergistic interactions\cite{kirby2009} or a rare alignment of non-synergistic forces\cite{denny2009}.
@@
 The presence of black swans also suggests that we develop management policy that is robust to heavy tails.
-For instance, setting target population abundances far from critical limits may buffer black-swan events\cite{caddy1996}, and maintaining genetic, phenotypic, and behavioural diversity may allow some components of populations to persist when others are affected by disease or extreme environmental forces\cite{schindler2010}.
+For instance, setting target population abundances far from critical limits will buffer against black-swan events\cite{caddy1996}, and maintaining genetic, phenotypic, and behavioural diversity may allow some components of populations to persist when others are affected by disease or extreme environmental forces\cite{schindler2010}.
 Finally, surprising, or counterintuitive ecological dynamics offer a tremendous opportunity to learn about ecological systems, evaluate when models break down, and adjust future management policy\cite{doak2008, lindenmayer2010}.
@@
 \textbf{Fig.~1: Illustration of population dynamic models that allow for heavy tails.} (\textbf{A} to \textbf{B}) Probability density for the Student-t distribution with scale parameter of 1 and different values of \(\nu\).
-Small values of \(\nu\) create heavy tails.
-As \(\nu\) approaches infinity the distribution approaches the normal distribution.
+Small values of \(\nu\) create heavy tails while as \(\nu\) approaches infinity the distribution approaches the normal distribution.
 (\textbf{C} to \textbf{E}) Simulated population dynamics from a Gompertz model with process noise drawn from Student-t distributions with three values of \(\nu\).
@@
 
+% TODO The A, B, C are too close to the text in the figure.
 \textbf{Fig.~4: Heavy-tailed process noise tends to be downward skewed and ignoring this can underestimate risk.} (\textbf{A}) Illustrations of Student-t distributions with 3 levels of skewness (\(\gamma\)) and heavy-tails (\(\nu\)).

### 2016-05-24 text/ms.tex
Copyediting of main paper

@@
 
-\textbf{Abstract}:
-Black swans are improbable events that nonetheless occur---often with profound consequences.
-While physical extremes, such as monsoons and heat waves, are widely studied and increasing in magnitude and frequency, it remains unclear the extent to which ecological population numbers buffer or suffer from such extremes.
-Here we estimate the frequency and direction of black-swan events in 609 animal populations, finding strong evidence for their occurrence in \(\sim\)\overallBasePerc \% of populations.
-Black-swan events occur most frequently for birds (\birdPH \%), mammals (\mammalsPH \%), and insects (\insectsPH \%), and manifest primarily as   population die-offs and crashes (\percBSDown \%) driven by extrinsic perturbations from climate variability.
-Thus, ignoring black-swan events will under-estimate extinction risk.
-Our results demonstrate the importance of both modeling heavy-tailed downward events in populations, and developing conservation strategies that are robust to future increases in climate extremes.
+\textbf{Abstract}: Black swans are improbable events that nonetheless
+occur---often with profound consequences. While physical extremes, such as
+monsoons and heat waves, are widely studied and increasing in magnitude and
+frequency, it remains unclear the extent to which ecological population numbers
+buffer or suffer from such extremes. Here we estimate the frequency and
+direction of black-swan events in \NPops\ animal populations, finding strong
+evidence for their occurrence in \(\sim\)\overallBasePerc \% of populations.
+Black-swan events occur most frequently for birds (\birdPH \%), mammals
+(\mammalsPH \%), and insects (\insectsPH \%), and manifest primarily as
+population die-offs and crashes (\percBSDown \%) driven by extrinsic
+perturbations from climate variability. Thus, ignoring black-swan events will
+underestimate extinction risk. Our results demonstrate the importance of both
+modeling heavy-tailed downward events in populations, and developing
+conservation strategies that are robust to future increases in climate
+extremes.
 
-\textbf{One Sentence Summary}: Black swans---statistically improbable events with profound consequences---occur in animal populations and understanding them will improve climate risk assessments.
+% \textbf{One Sentence Summary}: Black swans---statistically improbable events with profound consequences---occur in animal populations and understanding them will improve climate risk assessments.
 
@@
 Indeed, such die-offs may be the most important element affecting population persistence\cite{mangel1994} and their importance is likely to increase given projected increases in the frequency and magnitude of climate-related extremes\cite{ipcc2012}.
-Despite this anecdotal evidence for ecological black swans, systemtic searches of large numbers of timeseries have yet to reveal black-swan events\cite{keitt1998, allen2001, halley2002} However, the overwhelming majority of model fitting and risk forecasting assumes that deviations from model predictions can be represented by a normal distribution\cite[e.g.][]{brook2006a, knape2012}.
+Despite this anecdotal evidence for ecological black swans, systematic searches of large numbers of timeseries have yet to reveal black-swan events\cite{keitt1998, allen2001, halley2002} However, the overwhelming majority of model fitting and risk forecasting assumes that deviations from model predictions can be represented by a normal distribution\cite{brook2006a, knape2012}.
 If black swans occur, though, a normal distribution would under-estimate the probability of extreme events occurring.
@@
 Specifically, we fit population dynamics models in which the process noise is drawn from a Student-t distribution.
-By estimating the degrees of freedom parameter, \(\nu\), we can estimate the degree to which the process deviations have heavy tails and are therefore diagnositc of black-swan events (Fig.~1, fig.~S\ref{fig:priors}).
+By estimating the degrees of freedom parameter, \(\nu\), we can estimate the degree to which the process deviations have heavy tails and are therefore diagnostic of black-swan events (Fig.~1, fig.~S\ref{fig:priors}).
 Lower values of \(\nu\) result in heavy-tailed distributions.
@@
 
-We found evidence that black-swan dynamics are infreqent but highly influentiual: they occur most frequently for birds (7\%), mammals (5\%), and insects (3\%) and almost never in fishes for this dataset (Fig.~2).
+We found evidence that black-swan dynamics are infrequent but highly influential: they occur most frequently for birds (7\%), mammals (5\%), and insects (3\%) and almost never in fishes for this dataset (Fig.~2).
 Black-swan events were taxonomically widespread, occurring in 38\% of taxonomic orders.
@@
 In contrast, populations that did not have heavy tails (median \(\nu \geq 70\)) had little evidence of skewed process noise (Fig.~4b; 90\% of 95\% credible intervals overlapped 1).
-Projecting these heavy-tailed populations forward five years revealed that assuming the standard normal process noise underestimated risk by \crashUnderRange-fold (interquartile range of 99\% lower credible interval of abundance; Fig.~4c, fig.~S\ref{fig:skew-projections}).
+Projecting these heavy-tailed populations forward five years revealed that assuming standard normal process noise underestimated risk (99\% lower credible interval of abundance) by \crashUnderRange-fold (interquartile range; Fig.~4c, fig.~S\ref{fig:skew-projections}).
 Thus our analysis provides strong evidence for downward-skewed heavy-tailed events in abundance time series of higher taxa, and ignoring these events will tend to underestimate the risk of population declines.
-
-% TODO How are you defining risk. See if you can make this more concrete to avoid IUCN people getting upset with lack of specificity (i.e. Resit akcakaya)
 
@@
 But, earthquake preparedness involves spatial planning based on forecast probabilities to focus early detection efforts and develop disaster response plans\cite{nrc2007}.
-The presence of black swans also suggests that we develop management policy that is robust to heavy tails.
+The presence of black swans also suggests that we develop management policy that is robust to heavy tails and encourages \textit{general} resilience\cite{carpenter2012}.
 For instance, setting target population abundances far from critical limits will buffer against black-swan events\cite{caddy1996}, and maintaining genetic, phenotypic, and behavioural diversity may allow some components of populations to persist when others are affected by disease or extreme environmental forces\cite{schindler2010}.
@@
 Rare catastrophes can have a profound influence on population persistence\cite{mangel1994}.
-In recent decades, ecology has moved toward focusing on aspects of variance in addition to mean responses\cite{loreau2010a, thompson2013}.
+In recent decades, ecology has moved toward focusing on aspects of variance in addition to mean responses\cite{thompson2013}.
 Our results suggest that an added focus on ecological extremes represents the next frontier, particularly in the face of increased climate extremes\cite{meehl2004, ipcc2012, thompson2013}.
-Financial analysts are concerned with the shape of financial return downward tails because these directly impact estimates of risk---the probability of a specific magnitude of undesired event occurring\cite{rachev2008}.
+Financial analysts are concerned with the shape of financial return downward tails because these directly impact estimates of risk---the probability of a specific magnitude of undesired event occurring.
 Similarly, ecologists should focus more on estimating and predicting downward tails of population abundance, since these increase true extinction risk.
@@
 Materials and Methods\\
-References (32 to 72)\\
+Supplementary References\\
 Tables S1 to S2\\
@@
 
-% TODO The A, B, C are too close to the text in the figure.
 \textbf{Fig.~4: Heavy-tailed process noise tends to be downward skewed and ignoring this can underestimate risk.} (\textbf{A}) Illustrations of Student-t distributions with 3 levels of skewness (\(\gamma\)) and heavy-tails (\(\nu\)).

### 2016-05-24 text/ms.tex
Wrap text for voice editing

@@
 
-\textbf{Authors}: Sean C. Anderson\textsuperscript{1,2*}, Trevor A. Branch\textsuperscript{3}, Andrew B. Cooper\textsuperscript{2}, Nicholas K. Dulvy\textsuperscript{1}
+\textbf{Authors}: Sean C. Anderson\textsuperscript{1,2*}, Trevor A.
+Branch\textsuperscript{3}, Andrew B. Cooper\textsuperscript{2}, Nicholas K.
+Dulvy\textsuperscript{1}
 
-\textsuperscript{1}Earth to Ocean Research Group, Department of Biological Sciences, Simon Fraser University, Burnaby BC, V5A 1S6, Canada
+\textsuperscript{1}Earth to Ocean Research Group, Department of Biological
+Sciences, Simon Fraser University, Burnaby BC, V5A 1S6, Canada
 
-\textsuperscript{2}School of Resource and Environmental Management, Simon Fraser University, Burnaby, BC, V5A 1S6, Canada
+\textsuperscript{2}School of Resource and Environmental Management, Simon
+Fraser University, Burnaby, BC, V5A 1S6, Canada
 
-\textsuperscript{3}School of Aquatic and Fishery Sciences, Box 355020, University of Washington, Seattle, WA 98195, USA
+\textsuperscript{3}School of Aquatic and Fishery Sciences, Box 355020,
+University of Washington, Seattle, WA 98195, USA
 
-\textsuperscript{*}Corresponding author: Sean C. Anderson; E-mail: sandrsn@uw.edu; Present address: School of Aquatic and Fishery Sciences, Box 355020, University of Washington, Seattle, WA 98195, USA
+\textsuperscript{*}Corresponding author: Sean C. Anderson; E-mail:
+sandrsn@uw.edu; Present address: School of Aquatic and Fishery Sciences, Box
+355020, University of Washington, Seattle, WA 98195, USA
 
@@
 
-% \textbf{One Sentence Summary}: Black swans---statistically improbable events with profound consequences---occur in animal populations and understanding them will improve climate risk assessments.
+% \textbf{One Sentence Summary}: Black swans---statistically improbable events
+% with profound consequences---occur in animal populations and understanding
+% them will improve climate risk assessments.
 
@@
 
-\textbf{Main text}:
-Major surprises happen more often than expected in financial, social, and natural systems\cite{taleb2007, sornette2009, may2008}.
-Massive unpredictable market swings are responsible for the majority of financial gains and losses\cite{taleb2007}, fatalities from the largest wars dwarf those from all others \cite{newman2005}, and the frequency of the most damaging earthquakes has exceeded past expectations\cite{sornette2009}.
-In ecological systems, background rates of global extinction are punctuated by mass extinction\cite{harnik2012}, viruses can mutate suddenly to infect new hosts, and billions of animals can die at once in mass mortality events\cite{fey2015}.
-Indeed, such die-offs may be the most important element affecting population persistence\cite{mangel1994} and their importance is likely to increase given projected increases in the frequency and magnitude of climate-related extremes\cite{ipcc2012}.
-Despite this anecdotal evidence for ecological black swans, systematic searches of large numbers of timeseries have yet to reveal black-swan events\cite{keitt1998, allen2001, halley2002} However, the overwhelming majority of model fitting and risk forecasting assumes that deviations from model predictions can be represented by a normal distribution\cite{brook2006a, knape2012}.
-If black swans occur, though, a normal distribution would under-estimate the probability of extreme events occurring.
+% \textbf{Main text}: Major surprises happen more often than expected in
+financial, social, and natural systems\cite{taleb2007, sornette2009, may2008}.
+Massive unpredictable market swings are responsible for the majority of
+financial gains and losses\cite{taleb2007}, fatalities from the largest wars
+dwarf those from all others \cite{newman2005}, and the frequency of the most
+damaging earthquakes has exceeded past expectations\cite{sornette2009}. In
+ecological systems, background rates of global extinction are punctuated by
+mass extinction\cite{harnik2012}, viruses can mutate suddenly to infect new
+hosts, and billions of animals can die at once in mass mortality
+events\cite{fey2015}. Indeed, such die-offs may be the most important element
+affecting population persistence\cite{mangel1994} and their importance is
+likely to increase given projected increases in the frequency and magnitude of
+climate-related extremes\cite{ipcc2012}. Despite this anecdotal evidence for
+ecological black swans, systematic searches of large numbers of timeseries have
+yet to reveal black-swan events\cite{keitt1998, allen2001, halley2002} However,
+the overwhelming majority of model fitting and risk forecasting assumes that
+deviations from model predictions can be represented by a normal
+distribution\cite{brook2006a, knape2012}. If black swans occur, though,
+a normal distribution would under-estimate the probability of extreme events
+occurring.
 
-Here, we develop a new approach to estimate the frequency and magnitude of black-swan dynamics across time series of 609 populations from a wide array of taxonomic groups---including many  birds, mammals, insects, and fishes (table~S1) \cite{SOM}.
-We identify characteristics of time series or life-history traits associated with the detection of black-swan events and verify known causes.
-To accomplish this, we develop a framework for identifying heavy-tailed (black-swan) process noise in population dynamics, i.e.~whether the largest stochastic jumps in log abundance from one time step to the next are more extreme than typically seen with a normal distribution.
-Our framework allows for a range of population dynamic models, can incorporate observation uncertainty and skewness in process noise, and can be readily applied to abundance time series.
+Here, we develop a new approach to estimate the frequency and magnitude of
+black-swan dynamics across time series of 609 populations from a wide array of
+taxonomic groups---including many  birds, mammals, insects, and fishes
+(table~S1) \cite{SOM}. We identify characteristics of time series or
+life-history traits associated with the detection of black-swan events and
+verify known causes. To accomplish this, we develop a framework for identifying
+heavy-tailed (black-swan) process noise in population dynamics, i.e.~whether
+the largest stochastic jumps in log abundance from one time step to the next
+are more extreme than typically seen with a normal distribution. Our framework
+allows for a range of population dynamic models, can incorporate observation
+uncertainty and skewness in process noise, and can be readily applied to
+abundance time series.
 
-Specifically, we fit population dynamics models in which the process noise is drawn from a Student-t distribution.
-By estimating the degrees of freedom parameter, \(\nu\), we can estimate the degree to which the process deviations have heavy tails and are therefore diagnostic of black-swan events (Fig.~1, fig.~S\ref{fig:priors}).
-Lower values of \(\nu\) result in heavy-tailed distributions.
-For example, at \(\nu = 2\), the probability of drawing a value more than five standard deviations below the mean is almost 70,000 times higher than the probability of drawing such a value from a normal distribution (\(0.02\) vs.\ \(2.9\cdot10^{-7}\)).
-As \(\nu\) approaches infinity, the distribution approaches the normal distribution (Fig.~1).
+Specifically, we fit population dynamics models in which the process noise is
+drawn from a Student-t distribution. By estimating the degrees of freedom
+parameter, \(\nu\), we can estimate the degree to which the process deviations
+have heavy tails and are therefore diagnostic of black-swan events (Fig.~1,
+fig.~S\ref{fig:priors}). Lower values of \(\nu\) result in heavy-tailed
+distributions. For example, at \(\nu = 2\), the probability of drawing a value
+more than five standard deviations below the mean is almost 70,000 times higher
+than the probability of drawing such a value from a normal distribution
+(\(0.02\) vs.\ \(2.9\cdot10^{-7}\)). As \(\nu\) approaches infinity, the
+distribution approaches the normal distribution (Fig.~1).
 
-We found evidence that black-swan dynamics are infrequent but highly influential: they occur most frequently for birds (7\%), mammals (5\%), and insects (3\%) and almost never in fishes for this dataset (Fig.~2).
-Black-swan events were taxonomically widespread, occurring in 38\% of taxonomic orders.
-Accounting for time series length and partially pooling inference across taxonomic class and order with a hierarchical model, we found stronger evidence for black swans in insect populations than these statistics suggest---four of eight orders with the highest median probability of heavy tails were insect orders (Fig.~3a).
+We found evidence that black-swan dynamics are infrequent but highly
+influential: they occur most frequently for birds (7\%), mammals (5\%), and
+insects (3\%) and almost never in fishes for this dataset (Fig.~2). Black-swan
+events were taxonomically widespread, occurring in 38\% of taxonomic orders.
+Accounting for time series length and partially pooling inference across
+taxonomic class and order with a hierarchical model, we found stronger evidence
+for black swans in insect populations than these statistics suggest---four of
+eight orders with the highest median probability of heavy tails were insect
+orders (Fig.~3a).
 
-The majority of our heavy-tailed estimates were robust to alternative population models, observation error, and choice of Bayesian priors.
-Our conclusions were not systematically altered when we included an autocorrelation structure in the residuals, modelled population growth rates with or without density dependence, or modelled the population dynamics as Ricker-logistic instead of Gompertz (fig.~S\ref{fig:alt}).
-Similarly, introducing moderate observation error (CV = 0.2) only slightly decreased the estimated prevalence of black-swan events (fig.~S\ref{fig:alt}), and the strength of the prior on \(\nu\) had little influence on estimates of black-swan dynamics (fig.~S\ref{fig:alt-priors}).
-Finally, our simulation testing shows that, if anything, our models underpredict the true magnitude and probability of heavy-tailed events---especially given the time series have a median of only 26 years in our analysis (figs~S\ref{fig:sim-nu},~S\ref{fig:sim-prob}).
+The majority of our heavy-tailed estimates were robust to alternative
+population models, observation error, and choice of Bayesian priors. Our
+conclusions were not systematically altered when we included an autocorrelation
+structure in the residuals, modelled population growth rates with or without
+density dependence, or modelled the population dynamics as Ricker-logistic
+instead of Gompertz (fig.~S\ref{fig:alt}). Similarly, introducing moderate
+observation error (CV = 0.2) only slightly decreased the estimated prevalence
+of black-swan events (fig.~S\ref{fig:alt}), and the strength of the prior on
+\(\nu\) had little influence on estimates of black-swan dynamics
+(fig.~S\ref{fig:alt-priors}). Finally, our simulation testing shows that, if
+anything, our models underpredict the true magnitude and probability of
+heavy-tailed events---especially given the time series have a median of only 26
+years in our analysis (figs~S\ref{fig:sim-nu},~S\ref{fig:sim-prob}).
 
-For model fits to the population data, the probability of detecting black-swan dynamics was positively related to time-series length and negatively related to magnitude of process noise but not clearly related to population growth rate, density dependence, or maximum lifespan (Fig.~3b, fig.~S\ref{fig:correlates}).
-Longer time-series length was the strongest covariate of observing black-swan dynamics: there is a 1.6 times greater probability of detecting a black-swan event in populations with 60 time steps than in one with 30 time steps (fig.~S\ref{fig:correlates}).
+For model fits to the population data, the probability of detecting black-swan
+dynamics was positively related to time-series length and negatively related to
+magnitude of process noise but not clearly related to population growth rate,
+density dependence, or maximum lifespan (Fig.~3b, fig.~S\ref{fig:correlates}).
+Longer time-series length was the strongest covariate of observing black-swan
+dynamics: there is a 1.6 times greater probability of detecting a black-swan
+event in populations with 60 time steps than in one with 30 time steps
+(fig.~S\ref{fig:correlates}).
 
-The majority of black-swan events (\percBSDown \%) were downward (die-offs) rather than upward (unexpectedly rapid abundance increases).
-Of the black-swan events with published explanations (table~S2), the majority involved a combination of multiple factors.
-For example, a synchronization of environmental- and predation-mediated population cycles is thought to have caused a downward black-swan event for a water vole (\emph{Arvicola terrestris}) population\cite{saucy1994}.
-Other black swans were the result of a sequence of extreme climate events on their own.
-For instance, severe winters in 1929, 1940--1942, and 1962--1963 were associated with black-swan downswings in grey heron (\emph{Ardea cinerea}) abundance in the United Kingdom\cite{stafford1971} (Fig.~4c).
-Our analysis finds that the last event was a combination of two heavy-tailed events in a row and the population took three times longer to recover than predicted\cite{stafford1971}.
-Downward black swans were sometimes followed by upward black swans.
-For example, during a period of population crowding and nest shortages, a population of European shag cormorants (\emph{Phalacrocorax aristotelis}) on the Farne Islands, United Kingdom declined suddenly following a red tide event in 1968\cite{potts1980}.
-This freed up quality nest sites for first-time breeders, productivity increased, and the population experienced a rapid upswing in abundance\cite{potts1980}.
+The majority of black-swan events (\percBSDown \%) were downward (die-offs)
+rather than upward (unexpectedly rapid abundance increases). Of the black-swan
+events with published explanations (table~S2), the majority involved
+a combination of multiple factors. For example, a synchronization of
+environmental- and predation-mediated population cycles is thought to have
+caused a downward black-swan event for a water vole (\emph{Arvicola
+terrestris}) population\cite{saucy1994}. Other black swans were the result of
+a sequence of extreme climate events on their own. For instance, severe winters
+in 1929, 1940--1942, and 1962--1963 were associated with black-swan downswings
+in grey heron (\emph{Ardea cinerea}) abundance in the United
+Kingdom\cite{stafford1971} (Fig.~4c). Our analysis finds that the last event
+was a combination of two heavy-tailed events in a row and the population took
+three times longer to recover than predicted\cite{stafford1971}. Downward black
+swans were sometimes followed by upward black swans. For example, during
+a period of population crowding and nest shortages, a population of European
+shag cormorants (\emph{Phalacrocorax aristotelis}) on the Farne Islands, United
+Kingdom declined suddenly following a red tide event in 1968\cite{potts1980}.
+This freed up quality nest sites for first-time breeders, productivity
+increased, and the population experienced a rapid upswing in
+abundance\cite{potts1980}.
 
-Given the prevalence of downwards events, we refit our heavy-tailed models to measure the degree of skewness \(\gamma\) of the process noise using a skew-t distribution (Fig.~4a), and used these models to make near-term risk forecasts.
-Aggregated across populations with strong evidence of heavy tails (median \(\nu < 10\)), 85\% of the \(\gamma\) probability density was below 1, indicating strong evidence of downwardly skewed process noise (Fig.~4b, fig.~S\ref{fig:skew-nu}).
-In contrast, populations that did not have heavy tails (median \(\nu \geq 70\)) had little evidence of skewed process noise (Fig.~4b; 90\% of 95\% credible intervals overlapped 1).
-Projecting these heavy-tailed populations forward five years revealed that assuming standard normal process noise underestimated risk (99\% lower credible interval of abundance) by \crashUnderRange-fold (interquartile range; Fig.~4c, fig.~S\ref{fig:skew-projections}).
-Thus our analysis provides strong evidence for downward-skewed heavy-tailed events in abundance time series of higher taxa, and ignoring these events will tend to underestimate the risk of population declines.
+Given the prevalence of downwards events, we refit our heavy-tailed models to
+measure the degree of skewness \(\gamma\) of the process noise using a skew-t
+distribution (Fig.~4a), and used these models to make near-term risk forecasts.
+Aggregated across populations with strong evidence of heavy tails (median \(\nu
+< 10\)), 85\% of the \(\gamma\) probability density was below 1, indicating
+strong evidence of downwardly skewed process noise (Fig.~4b,
+fig.~S\ref{fig:skew-nu}). In contrast, populations that did not have heavy
+tails (median \(\nu \geq 70\)) had little evidence of skewed process noise
+(Fig.~4b; 90\% of 95\% credible intervals overlapped 1). Projecting these
+heavy-tailed populations forward five years revealed that assuming standard
+normal process noise underestimated risk (99\% lower credible interval of
+abundance) by \crashUnderRange-fold (interquartile range; Fig.~4c,
+fig.~S\ref{fig:skew-projections}). Thus our analysis provides strong evidence
+for downward-skewed heavy-tailed events in abundance time series of higher
+taxa, and ignoring these events will tend to underestimate the risk of
+population declines.
 
-There are a number of possible causes of black-swan events, including unmodelled intrinsic properties of populations or extrinsic forces acting on populations.
-For example, we could observe black-swan dynamics if we miss an underlying mixture of processes, since a mixture of normal distributions with different variances can generate a t distribution with heavy tails\cite{allen2001}.
-In ecology, these processes could arise through an aggregation of populations across space, population diversity\cite{schindler2010}, or an intrinsic change in population variability through time\cite{carpenter2006}.
-Extrinsic ecological forces could also cause black-swan dynamics\cite{nunez2012} including extreme climate events\cite{meehl2004, katz2005, ipcc2012}, predation from (or competition with) other species experiencing black swans, or sharp changes in human pressures such hunting, fishing, or habitat destruction.
-Alternatively, the synchrony of multiple extrinsic forces could give rise to black-swan dynamics through synergistic interactions\cite{kirby2009} or a rare alignment of non-synergistic forces\cite{denny2009}.
+There are a number of possible causes of black-swan events, including
+unmodelled intrinsic properties of populations or extrinsic forces acting on
+populations. For example, we could observe black-swan dynamics if we miss an
+underlying mixture of processes, since a mixture of normal distributions with
+different variances can generate a t distribution with heavy
+tails\cite{allen2001}. In ecology, these processes could arise through an
+aggregation of populations across space, population
+diversity\cite{schindler2010}, or an intrinsic change in population variability
+through time\cite{carpenter2006}. Extrinsic ecological forces could also cause
+black-swan dynamics\cite{nunez2012} including extreme climate
+events\cite{meehl2004, katz2005, ipcc2012}, predation from (or competition
+with) other species experiencing black swans, or sharp changes in human
+pressures such hunting, fishing, or habitat destruction. Alternatively, the
+synchrony of multiple extrinsic forces could give rise to black-swan dynamics
+through synergistic interactions\cite{kirby2009} or a rare alignment of
+non-synergistic forces\cite{denny2009}.
 
-In light of our findings, we suggest that ecological resource management can learn from disciplines that focus on heavy tails.
-For example, earthquake preparedness and response is focused on black-swan events.
-Similarly to ecological black swans, we can rarely predict the specific timing of large earthquakes.
-But, earthquake preparedness involves spatial planning based on forecast probabilities to focus early detection efforts and develop disaster response plans\cite{nrc2007}.
-The presence of black swans also suggests that we develop management policy that is robust to heavy tails and encourages \textit{general} resilience\cite{carpenter2012}.
-For instance, setting target population abundances far from critical limits will buffer against black-swan events\cite{caddy1996}, and maintaining genetic, phenotypic, and behavioural diversity may allow some components of populations to persist when others are affected by disease or extreme environmental forces\cite{schindler2010}.
-Finally, surprising, or counterintuitive ecological dynamics offer a tremendous opportunity to learn about ecological systems, evaluate when models break down, and adjust future management policy\cite{doak2008, lindenmayer2010}.
+In light of our findings, we suggest that ecological resource management can
+learn from disciplines that focus on heavy tails. For example, earthquake
+preparedness and response is focused on black-swan events. Similarly to
+ecological black swans, we can rarely predict the specific timing of large
+earthquakes. But, earthquake preparedness involves spatial planning based on
+forecast probabilities to focus early detection efforts and develop disaster
+response plans\cite{nrc2007}. The presence of black swans also suggests that we
+develop management policy that is robust to heavy tails and encourages
+\textit{general} resilience\cite{carpenter2012}. For instance, setting target
+population abundances far from critical limits will buffer against black-swan
+events\cite{caddy1996}, and maintaining genetic, phenotypic, and behavioural
+diversity may allow some components of populations to persist when others are
+affected by disease or extreme environmental forces\cite{schindler2010}.
+Finally, surprising, or counterintuitive ecological dynamics offer a tremendous
+opportunity to learn about ecological systems, evaluate when models break down,
+and adjust future management policy\cite{doak2008, lindenmayer2010}.
 
-Rare catastrophes can have a profound influence on population persistence\cite{mangel1994}.
-In recent decades, ecology has moved toward focusing on aspects of variance in addition to mean responses\cite{thompson2013}.
-Our results suggest that an added focus on ecological extremes represents the next frontier, particularly in the face of increased climate extremes\cite{meehl2004, ipcc2012, thompson2013}.
-Financial analysts are concerned with the shape of financial return downward tails because these directly impact estimates of risk---the probability of a specific magnitude of undesired event occurring.
-Similarly, ecologists should focus more on estimating and predicting downward tails of population abundance, since these increase true extinction risk.
+Rare catastrophes can have a profound influence on population
+persistence\cite{mangel1994}. In recent decades, ecology has moved toward
+focusing on aspects of variance in addition to mean
+responses\cite{thompson2013}. Our results suggest that an added focus on
+ecological extremes represents the next frontier, particularly in the face of
+increased climate extremes\cite{meehl2004, ipcc2012, thompson2013}. Financial
+analysts are concerned with the shape of financial return downward tails
+because these directly impact estimates of risk---the probability of a specific
+magnitude of undesired event occurring. Similarly, ecologists should focus more
+on estimating and predicting downward tails of population abundance, since
+these increase true extinction risk.
 
@@
 
-\textbf{Acknowledgements}: We thank J.W.
-Moore, A.O.
-Mooers, L.R.
-Gerber, J.D.
-Yeakel, C.
-Minto and members of the Earth to Ocean Group for helpful discussions and comments.
-We are grateful to the contributors and maintainers of the Global Population Dynamics Database and to Compute Canada's WestGrid high-performance computing resources.
-Silhouette images were obtained from phylopic.org under Creative Commons licenses; sources are listed in the Supplementary Materials.
-Funding was provided by a Simon Fraser University Graduate Fellowship (S.C.A.), the Natural Sciences and Engineering Research Council of Canada (N.K.D., A.B.C.), the Canada Research Chairs Program (N.K.D.).
+\textbf{Acknowledgements}: We thank J.W. Moore, A.O. Mooers, L.R. Gerber, J.D.
+Yeakel, C. Minto and members of the Earth to Ocean Group for helpful
+discussions and comments. We are grateful to the contributors and maintainers
+of the Global Population Dynamics Database and to Compute Canada's WestGrid
+high-performance computing resources. Silhouette images were obtained from
+phylopic.org under Creative Commons licenses; sources are listed in the
+Supplementary Materials. Funding was provided by a Simon Fraser University
+Graduate Fellowship (S.C.A.), the Natural Sciences and Engineering Research
+Council of Canada (N.K.D., A.B.C.), the Canada Research Chairs Program
+(N.K.D.).
 
@@
 
-%\textbf{Author Contributions}: S.C.A.
-and T.A.B.
-conceived the project; S.C.A., T.A.B., A.B.C., and N.K.D.
-designed the study; S.C.A.
-analyzed the data and wrote the paper with input from all authors.
-
-%\textbf{Author information}: The authors declare no competing financial interests.
-Reprints and permissions information is available at www.nature.com/reprints.
-Correspondence and requests for materials should be addressed to S.C.A.
-(sean\_anderson@sfu.ca).
+%\textbf{Author Contributions}: S.C.A. and T.A.B. conceived the project;
+%S.C.A., T.A.B., A.B.C., and N.K.D. designed the study; S.C.A. analyzed the
+%data and wrote the paper with input from all authors.
 
@@
 
-\textbf{Fig.~1: Illustration of population dynamic models that allow for heavy tails.} (\textbf{A} to \textbf{B}) Probability density for the Student-t distribution with scale parameter of 1 and different values of \(\nu\).
-Small values of \(\nu\) create heavy tails while as \(\nu\) approaches infinity the distribution approaches the normal distribution.
-(\textbf{C} to \textbf{E}) Simulated population dynamics from a Gompertz model with process noise drawn from Student-t distributions with three values of \(\nu\).
-Coloured dots in panels \textbf{C} and \textbf{D} represent jumps with less than a 1 in 1000 chance of occurring in a normal distribution.
-(\textbf{F} to \textbf{H}) Estimates of \(\nu\) from models fit to the times series in \textbf{C} to \textbf{E}.
-Shown are posterior samples (histograms), median and interquartile range of the posterior (IQR, dots and line segments), and the exponential prior on \(\nu\) (dashed lines).
-Colour shading behind \textbf{F} to \textbf{H} illustrates the approximate region of heavy tails.
+\textbf{Fig.~1: Illustration of population dynamic models that allow for heavy
+tails.} (\textbf{A} to \textbf{B}) Probability density for the Student-t
+distribution with scale parameter of 1 and different values of \(\nu\). Small
+values of \(\nu\) create heavy tails while as \(\nu\) approaches infinity the
+distribution approaches the normal distribution. (\textbf{C} to \textbf{E})
+Simulated population dynamics from a Gompertz model with process noise drawn
+from Student-t distributions with three values of \(\nu\). Coloured dots in
+panels \textbf{C} and \textbf{D} represent jumps with less than a 1 in 1000
+chance of occurring in a normal distribution. (\textbf{F} to \textbf{H})
+Estimates of \(\nu\) from models fit to the times series in \textbf{C} to
+\textbf{E}. Shown are posterior samples (histograms), median and interquartile
+range of the posterior (IQR, dots and line segments), and the exponential prior
+on \(\nu\) (dashed lines). Colour shading behind \textbf{F} to \textbf{H}
+illustrates the approximate region of heavy tails.
 
@@
 
-\textbf{Fig.~2: Estimates of population dynamic heavy-tailedness for populations of birds, mammals, insects, and fishes.} (\textbf{A} to \textbf{D}) Small values of \(\nu\) suggest heavy-tailed black-swan dynamics.
-Vertical points and line segments represent posterior medians and 50\% / 90\% credible intervals for individual populations.
-Inset plots show probability that \(\nu < 10\) for populations arranged by taxonomic order and sorted by decreasing mean Pr(\(\nu < 10\)).
-Taxonomic orders with three or fewer populations in \textbf{A} are omitted for space.
-Red to yellow points highlight populations with a high to moderately high probability of heavy-tailed dynamics.
+\textbf{Fig.~2: Estimates of population dynamic heavy-tailedness for
+populations of birds, mammals, insects, and fishes.} (\textbf{A} to \textbf{D})
+Small values of \(\nu\) suggest heavy-tailed black-swan dynamics. Vertical
+points and line segments represent posterior medians and 50\% / 90\% credible
+intervals for individual populations. Inset plots show probability that \(\nu
+< 10\) for populations arranged by taxonomic order and sorted by decreasing
+mean Pr(\(\nu < 10\)). Taxonomic orders with three or fewer populations in
+\textbf{A} are omitted for space. Red to yellow points highlight populations
+with a high to moderately high probability of heavy-tailed dynamics.
 
@@
 
-\textbf{Fig.~3: Standardized probabilities and covariates of heavy-tailed dynamics.} (\textbf{A}) Taxonomic-order-level posterior densities of Pr(\(\nu < 10\)) after accounting for time-series length in a hierarchical model.
-Dotted vertical line in a indicates the median expected Pr(\(\nu < 10\)) from the prior distribution.
-Colour shading refers to taxonomic class (yellow: fishes, green: insects, purple: birds, and red: mammals).
-Estimates are standardized at the geometric mean of time series length across all the data (approximately 27 time steps) and shown for orders with \(\ge 5\) populations.
-(\textbf{B}) Posterior densities for potential covariates of Pr(\(\nu < 10\)).
-In both panels, short vertical line segments within the density polygons indicate medians.
+\textbf{Fig.~3: Standardized probabilities and covariates of heavy-tailed
+dynamics.} (\textbf{A}) Taxonomic-order-level posterior densities of Pr(\(\nu
+< 10\)) after accounting for time-series length in a hierarchical model. Dotted
+vertical line in a indicates the median expected Pr(\(\nu < 10\)) from the
+prior distribution. Colour shading refers to taxonomic class (yellow: fishes,
+green: insects, purple: birds, and red: mammals). Estimates are standardized at
+the geometric mean of time series length across all the data (approximately 27
+time steps) and shown for orders with \(\ge 5\) populations. (\textbf{B})
+Posterior densities for potential covariates of Pr(\(\nu < 10\)). In both
+panels, short vertical line segments within the density polygons indicate
+medians.
 
@@
 
-\textbf{Fig.~4: Heavy-tailed process noise tends to be downward skewed and ignoring this can underestimate risk.} (\textbf{A}) Illustrations of Student-t distributions with 3 levels of skewness (\(\gamma\)) and heavy-tails (\(\nu\)).
-(\textbf{B}) Posterior density of the skewness parameters aggregated across populations grouped into heavy-tailed (\(\hat{\nu} < 10\)), slightly heavy-tailed (\(10 \leq \hat{\nu} \leq 70\)), and normal-tailed populations (\(\hat{\nu} > 70\)).
-Approximate mid-values from \textbf{B} are illustrated in \textbf{A}.
-(\textbf{C}) Example time series of heavy-tailed populations with different levels of skewness.
-Red dots highlight likely heavy-tailed events.
-Forecasts (grey regions) show median (solid lines) and lower 99\% credible intervals (dotted lines) of abundance.
-Black and red lines indicate forecasts from Gompertz models with lognormal and skew-t process noise, respectively.
+\textbf{Fig.~4: Heavy-tailed process noise tends to be downward skewed and
+ignoring this can underestimate risk.} (\textbf{A}) Illustrations of Student-t
+distributions with 3 levels of skewness (\(\gamma\)) and heavy-tails (\(\nu\)).
+(\textbf{B}) Posterior density of the skewness parameters aggregated across
+populations grouped into heavy-tailed (\(\hat{\nu} < 10\)), slightly
+heavy-tailed (\(10 \leq \hat{\nu} \leq 70\)), and normal-tailed populations
+(\(\hat{\nu} > 70\)). Approximate mid-values from \textbf{B} are illustrated in
+\textbf{A}. (\textbf{C}) Example time series of heavy-tailed populations with
+different levels of skewness. Red dots highlight likely heavy-tailed events.
+Forecasts (grey regions) show median (solid lines) and lower 99\% credible
+intervals (dotted lines) of abundance. Black and red lines indicate forecasts
+from Gompertz models with lognormal and skew-t process noise, respectively.

### 2016-05-24 text/supp.tex
Update references in supplement
Wrap supplement file for voice editing

@@
 
-\title{Supplementary Materials:\\Evidence for black-swan events in animal
+\title{Supplementary Materials:\\Black-swan events in animal
   populations}
@@
 
-\nocite{taleb2007,sornette2009}
-\nocite{meehl2004,katz2005,ipcc2012}
-\nocite{taleb2007,sornette2009,may2008}
 \nocite{taleb2007}
+\nocite{sornette2009}
+\nocite{may2008}
 \nocite{newman2005}
-\nocite{sornette2009}
 \nocite{harnik2012}
@@
 \nocite{ipcc2012}
-\nocite{keitt1998,allen2001,halley2002}
-\nocite{brook2006a,knape2012}
+\nocite{keitt1998}
+\nocite{allen2001}
+\nocite{halley2002}
+\nocite{brook2006a}
+\nocite{knape2012}
 \nocite{SOM}
@@
 \nocite{stafford1971}
-\nocite{stafford1971}
 \nocite{potts1980}
-\nocite{potts1980}
-\nocite{allen2001}
 \nocite{schindler2010}
@@
 \nocite{nunez2012}
-\nocite{meehl2004,katz2005,ipcc2012}
+\nocite{meehl2004}
+\nocite{katz2005}
 \nocite{kirby2009}
@@
 \nocite{nrc2007}
+\nocite{carpenter2012}
 \nocite{caddy1996}
-\nocite{schindler2010}
-\nocite{doak2008,lindenmayer2010}
-\nocite{mangel1994}
-\nocite{loreau2010a,thompson2013}
-\nocite{meehl2004,ipcc2012,thompson2013}
-\nocite{rachev2008}
+\nocite{doak2008}
+\nocite{lindenmayer2010}
+\nocite{thompson2013}
 
@@
 \end{align}
-Our prior on \(b\) was uninformative between values of \(-1\) and \(2\).
-We would not expect values of \(b\) with levels of density dependence as low as \(-1\) (very strong inverse density dependence), nor would we generally expect values above \(1\).
-We allowed values of \(b\) above \(1\) to allow for non-stationary time series of growth rates.
-The estimates of \(b\) were well within these bounds.
-Our prior on \(\lambda\) was weakly informative within the range of expected values for population growth and is similar to the default priors suggested by Gelman \emph{et al.}\cite{gelman2008d} for intercepts of regression models.
-Our Half-Cauchy prior on \(\sigma\) follows Gelman \emph{et al.}\cite{gelman2006c} and Gelman \emph{et al.}\cite{gelman2008d} and the specific scale parameter of \(2.5\) is based on our expected range of the value in nature from previous studies\cite{connors2014}.
-In our testing of a subsample of populations, our parameter estimates were not qualitatively changed by switching to a uniform prior on \(\sigma\), but the weakly informative Half-Cauchy prior sped up chain convergence.
+Our prior on \(b\) was uninformative between values of \(-1\) and \(2\). We
+would not expect values of \(b\) with levels of density dependence as low as
+\(-1\) (very strong inverse density dependence), nor would we generally expect
+values above \(1\). We allowed values of \(b\) above \(1\) to allow for
+non-stationary time series of growth rates. The estimates of \(b\) were well
+within these bounds. Our prior on \(\lambda\) was weakly informative within the
+range of expected values for population growth and is similar to the default
+priors suggested by Gelman \emph{et al.}\cite{gelman2008d} for intercepts of
+regression models. Our Half-Cauchy prior on \(\sigma\) follows Gelman \emph{et
+al.}\cite{gelman2006c} and Gelman \emph{et al.}\cite{gelman2008d} and the
+specific scale parameter of \(2.5\) is based on our expected range of the value
+in nature from previous studies\cite{connors2014}. In our testing of a
+subsample of populations, our parameter estimates were not qualitatively
+changed by switching to a uniform prior on \(\sigma\), but the weakly
+informative Half-Cauchy prior sped up chain convergence.
 
-We chose a weakly informative exponential prior on \(\nu\) with rate parameter of \(0.01\) truncated at values above two---a slightly less informative prior than suggested by Fernandez and Steel\cite{fernandez1998}.
-We truncated the distribution at \(2\), since at \(\nu < 2\) the variance of the t distribution is undefined.
-This prior gives only a \(7.7\)\% probability that \(\nu < 10\) but constrains the sampling sufficiently to avoid drifting towards infinity.
-In any case, for \(\nu > 20\) the t distribution is almost indistinguishable from the normal distribution (Fig.~1).
-Based on the shape of the t distribution, we chose the probability that \(\nu < 10\), Pr(\(\nu < 10\)), to define the probability of heavy-tailed dynamics.
-When categorizing a population as heavy tailed, we applied a 0.5 probability threshold.
-In the scenario where the data are uninformative about heavy tails (e.g.~Fig.~1E, H), the posterior for \(\nu\) will approximately match the prior (prior median \(= 71\), mean \(= 102\)) and a metric of Pr\((\nu < 10) > 0.5\) is unlikely to flag the population as heavy tailed.
+We chose a weakly informative exponential prior on \(\nu\) with rate parameter
+of \(0.01\) truncated at values above two---a slightly less informative prior
+than suggested by Fernandez and Steel\cite{fernandez1998}. We truncated the
+distribution at \(2\), since at \(\nu < 2\) the variance of the t distribution
+is undefined. This prior gives only a \(7.7\)\% probability that \(\nu < 10\)
+but constrains the sampling sufficiently to avoid drifting towards infinity. In
+any case, for \(\nu > 20\) the t distribution is almost indistinguishable from
+the normal distribution (Fig.~1). Based on the shape of the t distribution, we
+chose the probability that \(\nu < 10\), Pr(\(\nu < 10\)), to define the
+probability of heavy-tailed dynamics. When categorizing a population as heavy
+tailed, we applied a 0.5 probability threshold. In the scenario where the data
+are uninformative about heavy tails (e.g.~Fig.~1E, H), the posterior for
+\(\nu\) will approximately match the prior (prior median \(= 71\), mean \(=
+102\)) and a metric of Pr\((\nu < 10) > 0.5\) is unlikely to flag the
+population as heavy tailed.
 
-We fit our models with Stan 2.6.0\cite{stan-manual2015, hoffman2014}, and R 3.2.0\cite{r2015}.
-We began with four chains and 2,000 iterations, discarding the first 1,000 as warm up (i.e.~4,000 final samples).
-If \(\hat{R}\) (the potential scale reduction factor---a measure of chain convergence) was greater than \(1.05\) for any parameter, or if the minimum effective sample size, \(n_\mathrm{eff}\) (a measure of the effective number of uncorrelated samples), for any parameter was less than \(200\), we doubled both the total iterations and warm up period and sampled from the model again.
-These thresholds are in excess of the minimums recommended by Gelman \emph{et al.}\cite{gelman2006a} of \(\hat{R} < 1.1\) and \(n_\mathrm{eff} > 100\) for reliable point estimates and confidence intervals.
-In the majority of cases our minimum thresholds were greatly exceeded.
-We continued this procedure up to 8,000 iterations (16,000 total samples) by which all chains were deemed to have sufficiently converged.
-The No-U-Turn Hamiltonian Markov chain Monte Carlo sampler in Stan generally requires fewer iterations to obtain equivalent effective sample sizes than the Gibbs or Metropolis--Hastings algorithms\cite{stan-manual2015}.
+We fit our models with Stan 2.6.0\cite{stan-manual2015, hoffman2014}, and R
+3.2.0\cite{r2015}. We began with four chains and 2,000 iterations, discarding
+the first 1,000 as warm up (i.e.~4,000 final samples). If \(\hat{R}\) (the
+potential scale reduction factor---a measure of chain convergence) was greater
+than \(1.05\) for any parameter, or if the minimum effective sample size,
+\(n_\mathrm{eff}\) (a measure of the effective number of uncorrelated samples),
+for any parameter was less than \(200\), we doubled both the total iterations
+and warm up period and sampled from the model again. These thresholds are in
+excess of the minimums recommended by Gelman \emph{et al.}\cite{gelman2006a} of
+\(\hat{R} < 1.1\) and \(n_\mathrm{eff} > 100\) for reliable point estimates and
+confidence intervals. In the majority of cases our minimum thresholds were
+greatly exceeded. We continued this procedure up to 8,000 iterations (16,000
+total samples) by which all chains were deemed to have sufficiently converged.
+The No-U-Turn Hamiltonian Markov chain Monte Carlo sampler in Stan generally
+requires fewer iterations to obtain equivalent effective sample sizes than the
+Gibbs or Metropolis--Hastings algorithms\cite{stan-manual2015}.
 
 There are a number of caveats when considering the generality of our results.
-The GPDD data represent a taxonomically and geographically biased sample of populations---the longer time series we focus on are dominated by commercially and recreationally important species and a disproportionate number of populations are located in the United Kingdom.
-Although we would expect to find qualitatively similar evidence for black swans in other large taxonomic or geographic samples of populations, the common forces driving those black swans would likely differ.
-Additionally, some apparent black-swan events could be recording mistakes, or conversely, some extreme observations may have been discarded or altered if they were erroneously suspected of being recording mistakes.
-Indeed, we discarded three of the populations that our method initially identified as heavy tailed because they turned out to be data-entry errors.
-Finally, the temporal scales of observation and population dynamics vary considerably across populations in the GPDD and these likely influence the detection of heavy tails.
-For example, if we make frequent observations relative to generation time (e.g.~for many large-bodied mammals) we will average across generations and perhaps miss black-swan events.
-Conversely, if we census populations infrequently relative to generation time (e.g.~many insects in the GPDD) the recorded data may average across extreme and less-extreme events and also dampen black-swan dynamics.
+The GPDD data represent a taxonomically and geographically biased sample of
+populations---the longer time series we focus on are dominated by commercially
+and recreationally important species and a disproportionate number of
+populations are located in the United Kingdom. Although we would expect to find
+qualitatively similar evidence for black swans in other large taxonomic or
+geographic samples of populations, the common forces driving those black swans
+would likely differ. Additionally, some apparent black-swan events could be
+recording mistakes, or conversely, some extreme observations may have been
+discarded or altered if they were erroneously suspected of being recording
+mistakes. Indeed, we discarded three of the populations that our method
+initially identified as heavy tailed because they turned out to be data-entry
+errors. Finally, the temporal scales of observation and population dynamics
+vary considerably across populations in the GPDD and these likely influence the
+detection of heavy tails. For example, if we make frequent observations
+relative to generation time (e.g.~for many large-bodied mammals) we will
+average across generations and perhaps miss black-swan events. Conversely, if
+we census populations infrequently relative to generation time (e.g.~many
+insects in the GPDD) the recorded data may average across extreme and
+less-extreme events and also dampen black-swan dynamics.
 
-\section{Alternative priors} To test if the prior on \(\nu\) influenced our estimate of black-swan dynamics, we refit our models with weaker and stronger priors.
-Our base model used a prior on \(\nu\) of Truncated-Exponential(0.01, min.\ = 2).
-For a weaker prior we used Truncated-Exponential(0.005, min.\ = 2) and for a stronger prior we used Truncated-Exponential(0.02, min.\ = 2) (fig.~\ref{fig:priors}c).
-Note that the base and weaker priors are relatively flat within the region of \(\nu < 20\), which is the region we are mostly concerned about when categorizing populations as heavy tailed.
-We furthermore considered a Uniform(\(0, 1\)) prior on \(1/\nu\) as used in Gelman \emph{et al.}\cite{gelman2014}, but this prior puts considerably more weight on lower (heavy-tailed) values of \(\nu\) than the truncated exponential priors considered here (not shown).
+\section{Alternative priors} To test if the prior on \(\nu\) influenced our
+estimate of black-swan dynamics, we refit our models with weaker and stronger
+priors. Our base model used a prior on \(\nu\) of Truncated-Exponential(0.01,
+min.\ = 2). For a weaker prior we used Truncated-Exponential(0.005, min.\ = 2)
+and for a stronger prior we used Truncated-Exponential(0.02, min.\ = 2)
+(fig.~\ref{fig:priors}c). Note that the base and weaker priors are relatively
+flat within the region of \(\nu < 20\), which is the region we are mostly
+concerned about when categorizing populations as heavy tailed. We furthermore
+considered a Uniform(\(0, 1\)) prior on \(1/\nu\) as used in Gelman \emph{et
+al.}\cite{gelman2014}, but this prior puts considerably more weight on lower
+(heavy-tailed) values of \(\nu\) than the truncated exponential priors
+considered here (not shown).
 
-Our results show that weaker and stronger priors have little influence on our conclusions about heavy-tailed dynamics (fig.~\ref{fig:alt-priors}).
-When the data are informative about tail behaviour (i.e.
-when there is strong evidence of low \(\nu\) values, upper-right of fig.~\ref{fig:alt-priors}), the prior has little impact on the estimate of \(\nu\).
-When the data are less informative about \(\nu\) (i.e.\ when there are no or few tail events and time series are short or noisy), the prior can pull the estimate of \(\nu\) towards larger or smaller values (fig.~\ref{fig:alt-priors}).
-The vast majority of the populations with Pr\((\nu < 10)\) in the base prior were not altered qualitatively by this range of prior strength.
+Our results show that weaker and stronger priors have little influence on our
+conclusions about heavy-tailed dynamics (fig.~\ref{fig:alt-priors}). When the
+data are informative about tail behaviour (i.e. when there is strong evidence
+of low \(\nu\) values, upper-right of fig.~\ref{fig:alt-priors}), the prior has
+little impact on the estimate of \(\nu\). When the data are less informative
+about \(\nu\) (i.e.\ when there are no or few tail events and time series are
+short or noisy), the prior can pull the estimate of \(\nu\) towards larger or
+smaller values (fig.~\ref{fig:alt-priors}). The vast majority of the
+populations with Pr\((\nu < 10)\) in the base prior were not altered
+qualitatively by this range of prior strength.
 
-\section{Alternative population models} We fit alternative population models to test if four key phenomena systematically changed our conclusions.
-The range of percentages of black swans by taxonomic class cited in the abstract are based on lower and upper limits across our main Gompertz model and these four alternative models.
+\section{Alternative population models} We fit alternative population models to
+test if four key phenomena systematically changed our conclusions. The range of
+percentages of black swans by taxonomic class cited in the abstract are based
+on lower and upper limits across our main Gompertz model and these four
+alternative models.
 
-\subsection{Autocorrelated residuals} We considered a version of the Gompertz model in which an autoregressive parameter was fit to the process-noise residuals:
-\begin{align}
-x_t &= \lambda + b x_{t-1} + \epsilon_t\\
-\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, \phi \epsilon_{t-1}, \sigma).
-\end{align}
-In addition to the parameters in the original Gompertz model, this model estimates an additional parameter \(\phi\), which represents the correlation of subsequent residuals.
-Based on the results of previous analyses with the GPDD\cite{connors2014} and the chosen prior in previous analyses\cite{thorson2014a}, we placed a moderately informative prior on \(\phi\) that assumed the greatest probability density near zero with the reduced possibility of \(\phi\) being near \(-1\) or \(1\).
-Specifically, we chose \(\phi \sim \mathrm{Truncated\mhyphen Normal}(0, 0.5, \mathrm{min.} = -1, \mathrm{max.} = 1)\).
+\subsection{Autocorrelated residuals} We considered a version of the Gompertz
+model in which an autoregressive parameter was fit to the process-noise
+residuals: \begin{align} x_t &= \lambda + b x_{t-1} + \epsilon_t\\ \epsilon_t
+&\sim \mathrm{Student\mhyphen t}(\nu, \phi \epsilon_{t-1}, \sigma). \end{align}
+In addition to the parameters in the original Gompertz model, this model
+estimates an additional parameter \(\phi\), which represents the correlation of
+subsequent residuals. Based on the results of previous analyses with the
+GPDD\cite{connors2014} and the chosen prior in previous
+analyses\cite{thorson2014a}, we placed a moderately informative prior on
+\(\phi\) that assumed the greatest probability density near zero with the
+reduced possibility of \(\phi\) being near \(-1\) or \(1\). Specifically, we
+chose \(\phi \sim \mathrm{Truncated\mhyphen Normal}(0, 0.5, \mathrm{min.} = -1,
+\mathrm{max.} = 1)\).
 
-\subsection{Assumed density independence} We fit a simplified version of the Gompertz model in which the density dependence parameter \(b\) was fixed at \(1\) (density independent).
-This is equivalent to fitting a random walk model (with drift) to the \(\ln\) abundances or assuming the growth rates are drawn from a stationary distribution.
-The model was as follows:
+\subsection{Assumed density independence} We fit a simplified version of the
+Gompertz model in which the density dependence parameter \(b\) was fixed at
+\(1\) (density independent). This is equivalent to fitting a random walk model
+(with drift) to the \(\ln\) abundances or assuming the growth rates are drawn
+from a stationary distribution. The model was as follows:
 \begin{align}
@@
 \end{align}
-We fit this model for three reasons: (1) it is computationally simpler and so provides a check that our more complicated full Gompertz model was obtaining reasonable estimates of \(\nu\), (2) it provides a test of whether density dependence was systematically affecting our perception of heavy tails, (3) it matches how some previous authors have modelled heavy tails without accounting for density dependence\cite{segura2013}.
+We fit this model for three reasons: (1) it is computationally simpler and so
+provides a check that our more complicated full Gompertz model was obtaining
+reasonable estimates of \(\nu\), (2) it provides a test of whether density
+dependence was systematically affecting our perception of heavy tails, (3) it
+matches how some previous authors have modelled heavy tails without accounting
+for density dependence\cite{segura2013}.
 
-\subsection{Assumed observation error} Observation error can bias parameter estimates\cite{knape2012} and is known to affect the ability to detect extreme events\cite{ward2007}.
-In our main analysis, we fit a model that ignored observation error.
-One way to account for observation error would be to fit a full state-space model that simultaneously estimates the magnitude of process noise and observation error.
-However, simultaneously estimating observation and process noise is a challenging problem (e.g.\ because the observation and process noise parameters tend to negatively covary in model fitting) and is known to sometimes result in identifiability issues with the Gompertz population model\cite{knape2008}.
-Furthermore, our model was applied to hundreds of time series, often of short length (as few as \minTimeSteps\ time steps) and our model estimates an additional parameter---the shape of the process deviation tails---potentially making identifiability and computational issues even greater.
-Therefore, we considered a version of the base Gompertz model that allowed for a fixed level of observation error:
+\subsection{Assumed observation error} Observation error can bias parameter
+estimates\cite{knape2012} and is known to affect the ability to detect extreme
+events\cite{ward2007}. In our main analysis, we fit a model that ignored
+observation error. One way to account for observation error would be to fit a
+full state-space model that simultaneously estimates the magnitude of process
+noise and observation error. However, simultaneously estimating observation and
+process noise is a challenging problem (e.g.\ because the observation and
+process noise parameters tend to negatively covary in model fitting) and is
+known to sometimes result in identifiability issues with the Gompertz
+population model\cite{knape2008}. Furthermore, our model was applied to
+hundreds of time series, often of short length (as few as \minTimeSteps\ time
+steps) and our model estimates an additional parameter---the shape of the
+process deviation tails---potentially making identifiability and computational
+issues even greater. Therefore, we considered a version of the base Gompertz
+model that allowed for a fixed level of observation error:
 \begin{align}
@@
 \end{align}
-where \(U\) represents the unobserved state vector, \(\sigma_\mathrm{obs}\) represents the standard deviation of observation error (on a log scale), and \(\sigma_\mathrm{proc}\) represent the process noise scale parameter.
-We set \(\sigma_\mathrm{obs}\) to \(0.2\), which represents a typical value used in simulation analyses\cite{valpine2002, thorson2014b}.
+where \(U\) represents the unobserved state vector, \(\sigma_\mathrm{obs}\)
+represents the standard deviation of observation error (on a log scale), and
+\(\sigma_\mathrm{proc}\) represent the process noise scale parameter. We set
+\(\sigma_\mathrm{obs}\) to \(0.2\), which represents a typical value used in
+simulation analyses\cite{valpine2002, thorson2014b}.
 
@@
 \end{align}
-where \(r_\mathrm{max}\) represents the theoretical maximum population growth rate that is obtained when \(N_t = 0\).
-The parameter \(K\) represents the carrying capacity and, as before, \(x_t\) represents the \(\ln\) transformed abundance at time \(t\).
-The Ricker-logistic model assumes a linear decrease in population growth rate with increases in abundance.
-In contrast, the Gompertz model assumes a linear decrease in population growth rate with increases in \textit{log} abundance.
+where \(r_\mathrm{max}\) represents the theoretical maximum population growth
+rate that is obtained when \(N_t = 0\). The parameter \(K\) represents the
+carrying capacity and, as before, \(x_t\) represents the \(\ln\) transformed
+abundance at time \(t\). The Ricker-logistic model assumes a linear decrease in
+population growth rate with increases in abundance. In contrast, the Gompertz
+model assumes a linear decrease in population growth rate with increases in
+\textit{log} abundance.
 
-To fit the Ricker-logistic models, we chose a prior on \(K\) uniform between zero and twice the maximum observed abundance (similar and less informative than in Delean \emph{et al.}\cite{delean2013}).
-We set the prior on \(r_\mathrm{max}\) as uniform between 0 and 20 as in Delean \emph{et al.}\cite{delean2013}.
-We used the same priors on \(\nu\) and \(\sigma\) as in the Gompertz model.
+To fit the Ricker-logistic models, we chose a prior on \(K\) uniform between
+zero and twice the maximum observed abundance (similar and less informative
+than in Delean \emph{et al.}\cite{delean2013}). We set the prior on
+\(r_\mathrm{max}\) as uniform between 0 and 20 as in Delean \emph{et
+al.}\cite{delean2013}. We used the same priors on \(\nu\) and \(\sigma\) as in
+the Gompertz model.
 
 \section{Simulation testing} We performed two types of simulation testing.
-First, we tested how easily the Student-t distribution \(\nu\) parameter could be recovered given different true values of \(\nu\) and different sample sizes.
-Second, we tested the ability of the heavy-tailed Gompertz model to obtain unbiased parameter estimates of \(\nu\) given that a set of process deviations was provided in which the effective \(\nu\) value was close to the true \(\nu\) value.
+First, we tested how easily the Student-t distribution \(\nu\) parameter could
+be recovered given different true values of \(\nu\) and different sample sizes.
+Second, we tested the ability of the heavy-tailed Gompertz model to obtain
+unbiased parameter estimates of \(\nu\) given that a set of process deviations
+was provided in which the effective \(\nu\) value was close to the true \(\nu\)
+value.
 
-We found that we could consistently and precisely recover median posterior estimates of \(\nu\) near the true value of \(\nu\) with large samples (\(\ge 200\)) (fig.~\ref{fig:sim-nu} upper panels).
-At smaller samples we could usually qualitatively distinguish heavy from not-heavy tails, but the model tended to underestimate how heavy the tails were.
-At the same time, at smaller sample sizes, the model tended to overestimate how large the scale parameter was (fig.~\ref{fig:sim-nu} lower panels).
+We found that we could consistently and precisely recover median posterior
+estimates of \(\nu\) near the true value of \(\nu\) with large samples (\(\ge
+200\)) (fig.~\ref{fig:sim-nu} upper panels). At smaller samples we could
+usually qualitatively distinguish heavy from not-heavy tails, but the model
+tended to underestimate how heavy the tails were. At the same time, at smaller
+sample sizes, the model tended to overestimate how large the scale parameter
+was (fig.~\ref{fig:sim-nu} lower panels).
 
-Our simulation-testing results indicate that the Gompertz model can recapture the true value of \(\nu\) when the process noise was chosen so that appropriate tail events were present (fig.~\ref{fig:sim-prob} upper panels).
-The addition of observation error caused the model to tend to underestimate the degree of heavy-tailedness.
-Fitting a model with correctly specified observation error did not make substantial improvements to model bias (fig.~\ref{fig:sim-prob}).
+Our simulation-testing results indicate that the Gompertz model can recapture
+the true value of \(\nu\) when the process noise was chosen so that appropriate
+tail events were present (fig.~\ref{fig:sim-prob} upper panels). The addition
+of observation error caused the model to tend to underestimate the degree of
+heavy-tailedness. Fitting a model with correctly specified observation error
+did not make substantial improvements to model bias (fig.~\ref{fig:sim-prob}).
 
@@
 
-When converting the posterior distributions of $\nu$ into Pr($\nu < 10$), the models distinguished heavy and not-heavy tails reasonably well (fig.
-\ref{fig:sim-prob} lower panels).
-Without observation error, and using a probability of $0.5$ as a threshold, the model correctly classified all simulated systems with normally distributed process noise as not heavy tailed.
-The model would have miscategorized only one of $40$ simulations at $\nu = 5$ across simulated populations with $50$ or $100$ time steps (fig.
+When converting the posterior distributions of $\nu$ into Pr($\nu < 10$), the
+models distinguished heavy and not-heavy tails reasonably well (fig.
+\ref{fig:sim-prob} lower panels). Without observation error, and using a
+probability of $0.5$ as a threshold, the model correctly classified all
+simulated systems with normally distributed process noise as not heavy tailed.
+The model would have miscategorized only one of $40$ simulations at $\nu = 5$
+across simulated populations with $50$ or $100$ time steps (fig.
 \ref{fig:sim-prob}, scenarios 1 and 2 in lower row, second panel from left).
-The model would have correctly categorized all cases where the process noise was not heavy tailed (fig.
-\ref{fig:sim-prob} bottom-right panel) and all cases where $\nu = 3$ and there was not observation error.
-With $0.2$ standard deviations of observation error, the model still categorized \obsErrorNuFivePerc\% of cases as heavy tailed when $\nu = 5$ and all but one case when $\nu = 3$.
-Allowing for observation error made little improvement to the detection of heavy tails.
-Therefore, we chose to focus on the simpler model without observation error in the main text, particularly given that the true magnitude of observation error was unknown in the empirical data.
+The model would have correctly categorized all cases where the process noise
+was not heavy tailed (fig. \ref{fig:sim-prob} bottom-right panel) and all cases
+where $\nu = 3$ and there was not observation error. With $0.2$ standard
+deviations of observation error, the model still categorized
+\obsErrorNuFivePerc\% of cases as heavy tailed when $\nu = 5$ and all but one
+case when $\nu = 3$. Allowing for observation error made little improvement to
+the detection of heavy tails. Therefore, we chose to focus on the simpler model
+without observation error in the main text, particularly given that the true
+magnitude of observation error was unknown in the empirical data.
 
-\section{Covariates of heavy-tailed dynamics} We fit a hierarchical beta regression model to the predicted probability of heavy tails, Pr(\(\nu < 10\)), to investigate potential covariates of heavy-tailed dynamics.
-We obtained maximum lifespan and body-size data from Brook \emph{et al.}\cite{brook2006a}.
-The beta distribution is useful when response data are on a continuous scale between zero and one\cite{ferrari2004}.
-The model was as follows:
+\section{Covariates of heavy-tailed dynamics} We fit a hierarchical beta
+regression model to the predicted probability of heavy tails, Pr(\(\nu < 10\)),
+to investigate potential covariates of heavy-tailed dynamics. We obtained
+maximum lifespan and body-size data from Brook \emph{et al.}\cite{brook2006a}.
+The beta distribution is useful when response data are on a continuous scale
+between zero and one\cite{ferrari2004}. The model was as follows:
 \begin{align}
@@
 \end{align}
-where \(A\) and \(B\) represent the beta distribution shape parameters; \(\mu_i\) represents the predicted value for population \(i\), class \(j\), order \(k\), and species \(l\); \(\phi_\mathrm{disp}\) represents the dispersion parameter; and \(X_i\) represents a vector of predictors (such as lifespan) for population \(i\) with associated \(\beta\) coefficients.
-The intercepts are allowed to vary from the overall intercept \(\alpha\) by taxonomic class (\(\alpha^\mathrm{class}_j\)), taxonomic order (\(\alpha^\mathrm{order}_k\)), and species (\(\alpha^\mathrm{species}_l\)) with standard deviations \(\sigma_{\alpha \; \mathrm{class}}\), \(\sigma_{\alpha \; \mathrm{order}}\), and \(\sigma_{\alpha \; \mathrm{species}}\).
-Where possible, we also allowed for error distributions around the predictors by incorporating the standard deviation of the posterior samples for the Gompertz parameters \(\lambda\), \(b\), and \(\log \sigma\) around the mean point value as normal distributions (not shown in the above equation).
+where \(A\) and \(B\) represent the beta distribution shape parameters;
+\(\mu_i\) represents the predicted value for population \(i\), class \(j\),
+order \(k\), and species \(l\); \(\phi_\mathrm{disp}\) represents the
+dispersion parameter; and \(X_i\) represents a vector of predictors (such as
+lifespan) for population \(i\) with associated \(\beta\) coefficients. The
+intercepts are allowed to vary from the overall intercept \(\alpha\) by
+taxonomic class (\(\alpha^\mathrm{class}_j\)), taxonomic order
+(\(\alpha^\mathrm{order}_k\)), and species (\(\alpha^\mathrm{species}_l\)) with
+standard deviations \(\sigma_{\alpha \; \mathrm{class}}\), \(\sigma_{\alpha \;
+\mathrm{order}}\), and \(\sigma_{\alpha \; \mathrm{species}}\). Where possible,
+we also allowed for error distributions around the predictors by incorporating
+the standard deviation of the posterior samples for the Gompertz parameters
+\(\lambda\), \(b\), and \(\log \sigma\) around the mean point value as normal
+distributions (not shown in the above equation).
 
-We log transformed \(\sigma\), time-series length, and lifespan to match the way they are visually represented in fig.~\ref{fig:correlates} and to make the relationship approximately linear on the logit-transformed response scale.
-All input variables were standardized by subtracting their mean and dividing by two standard deviations to make their coefficients comparable in magnitude\cite{gelman2008c}.
-We excluded body length as a covariate because it was highly correlated with lifespan, and lifespan exhibited more overlap across taxonomy than body length.
-Lifespan is also more directly related to time and potential mechanisms driving black-swan dynamics.
+We log transformed \(\sigma\), time-series length, and lifespan to match the
+way they are visually represented in fig.~\ref{fig:correlates} and to make the
+relationship approximately linear on the logit-transformed response scale. All
+input variables were standardized by subtracting their mean and dividing by two
+standard deviations to make their coefficients comparable in
+magnitude\cite{gelman2008c}. We excluded body length as a covariate because it
+was highly correlated with lifespan, and lifespan exhibited more overlap across
+taxonomy than body length. Lifespan is also more directly related to time and
+potential mechanisms driving black-swan dynamics.
 
-We incorporated weakly informative priors into our model: \(\mathrm{Cauchy}(0, 10)\) on the global intercept \(\alpha\), \(\mathrm{Half\mhyphen Cauchy}(0, 2.5)\) on all standard deviation parameters, \(\mathrm{Half\mhyphen Cauchy}(0, 10)\) on the dispersion parameter \(\phi_\mathrm{disp}\), and \(\mathrm{Cauchy}(0, 2.5)\) on all other parameters\cite{gelman2006c, gelman2008d}.
-Compared to normal priors, the Cauchy priors concentrate more probability density around expected parameter values while allowing for a higher probability density far into the tails, thereby allowing the data to dominate the posterior more strongly if it disagrees with the prior.
-We fit our models with 5,000 total iterations per chain, 2500 warm-up iterations, four chains, and discarding every second sample.
-We checked for chain convergence visually and with the same criteria as before (\(\widehat{R} < 1.05\) and \(n_\mathrm{eff} >200\) for all parameters).
-We further checked our model against a similar model with the glmmADMB \textsf{R} package\cite{glmmadmb} without uncertainty around the predictors.
+We incorporated weakly informative priors into our model: \(\mathrm{Cauchy}(0,
+10)\) on the global intercept \(\alpha\), \(\mathrm{Half\mhyphen Cauchy}(0,
+2.5)\) on all standard deviation parameters, \(\mathrm{Half\mhyphen Cauchy}(0,
+10)\) on the dispersion parameter \(\phi_\mathrm{disp}\), and
+\(\mathrm{Cauchy}(0, 2.5)\) on all other parameters\cite{gelman2006c,
+gelman2008d}. Compared to normal priors, the Cauchy priors concentrate more
+probability density around expected parameter values while allowing for a
+higher probability density far into the tails, thereby allowing the data to
+dominate the posterior more strongly if it disagrees with the prior. We fit our
+models with 5,000 total iterations per chain, 2500 warm-up iterations, four
+chains, and discarding every second sample. We checked for chain convergence
+visually and with the same criteria as before (\(\widehat{R} < 1.05\) and
+\(n_\mathrm{eff} >200\) for all parameters). We further checked our model
+against a similar model with the glmmADMB \textsf{R} package\cite{glmmadmb}
+without uncertainty around the predictors.
 
-To derive taxonomic-order-level estimates of the probability of heavy tails accounting for time-series length (Fig.~3A), we fit a separate hierarchical model with the same structure but with only \(\log\) time-series length as a predictor---in this case, we did not want to control for intrinsic population characteristics such as density dependence.
-Since our predictors were centered by subtracting their mean value, we obtained order-level estimates of the probability of heavy tails at mean log time-series length by adding the posteriors for \(\alpha\), \(\alpha^\mathrm{class}_j\), and \(\alpha^\mathrm{order}_k\).
+To derive taxonomic-order-level estimates of the probability of heavy tails
+accounting for time-series length (Fig.~3A), we fit a separate hierarchical
+model with the same structure but with only \(\log\) time-series length as a
+predictor---in this case, we did not want to control for intrinsic population
+characteristics such as density dependence. Since our predictors were centered
+by subtracting their mean value, we obtained order-level estimates of the
+probability of heavy tails at mean log time-series length by adding the
+posteriors for \(\alpha\), \(\alpha^\mathrm{class}_j\), and
+\(\alpha^\mathrm{order}_k\).
 
-\section{Skewed Student-t forecasts} To formally evaluate the apparent skewness of heavy-tailed process noise we fit Gompertz models with skewed Student-t distributed process noise\cite{fernandez1998}.
-The distribution adds one parameter to the Student-t distribution, \(\gamma\), which controls the skewness.
-The distribution is symmetrical if \(\gamma = 1\), left skewed if \(0 < \gamma < 1\), and right skewed if \(1 < \gamma < \infty\).
-We placed a weakly informative prior of \(\mathrm{Cauchy}(0, 2.5)\) on \(\log \gamma\).
-We aggregated 5,000 randomly selected posterior samples from the \(\gamma\) parameter of each model at three levels of evidence for heavy-tails: \emph{heavy tailed} median \(\nu < 10\); \emph{slightly heavy tailed} \(10 \le \mathrm{median}\, \nu \le 70\), and \emph{normal tails}, median \(\nu > 70\) (Fig.~4).
+\section{Skewed Student-t forecasts} To formally evaluate the apparent skewness
+of heavy-tailed process noise we fit Gompertz models with skewed Student-t
+distributed process noise\cite{fernandez1998}. The distribution adds one
+parameter to the Student-t distribution, \(\gamma\), which controls the
+skewness. The distribution is symmetrical if \(\gamma = 1\), left skewed if \(0
+< \gamma < 1\), and right skewed if \(1 < \gamma < \infty\). We placed a weakly
+informative prior of \(\mathrm{Cauchy}(0, 2.5)\) on \(\log \gamma\). We
+aggregated 5,000 randomly selected posterior samples from the \(\gamma\)
+parameter of each model at three levels of evidence for heavy-tails:
+\emph{heavy tailed} median \(\nu < 10\); \emph{slightly heavy tailed} \(10 \le
+\mathrm{median}\, \nu \le 70\), and \emph{normal tails}, median \(\nu > 70\)
+(Fig.~4).
 
-To generate five-year forecasts of abundance, we combined the posterior parameter samples from the skewed-t Gompertz models with stochastically generated process noise.
-We compared these forecasts to those generated by a standard Gompertz model with normally distributed process noise fit to the same data.
-We calculated the ratio of the lower 99\% quantile credible interval between the two model projections for all populations in which Pr(\(\nu < 10\)) \(> 0.5\).
-To ensure the results had stabilized in the tails of the forecast posterior, we increased the number of posterior samples in Stan.
-We ran four chains of 20,000 iterations and discarded the first 10,000 as warmup for a total of 40,000 samples.
+To generate five-year forecasts of abundance, we combined the posterior
+parameter samples from the skewed-t Gompertz models with stochastically
+generated process noise. We compared these forecasts to those generated by a
+standard Gompertz model with normally distributed process noise fit to the same
+data. We calculated the ratio of the lower 99\% quantile credible interval
+between the two model projections for all populations in which Pr(\(\nu < 10\))
+\(> 0.5\). To ensure the results had stabilized in the tails of the forecast
+posterior, we increased the number of posterior samples in Stan. We ran four
+chains of 20,000 iterations and discarded the first 10,000 as warmup for a
+total of 40,000 samples.
 
@@
 
-The probability density for the skewed Student-t distribution \cite{fernandez1998} equates to the following probability density as implemented in Ref.~\cite{king2012}:
+The probability density for the skewed Student-t distribution
+\cite{fernandez1998} equates to the following probability density as
+implemented in Ref.~\cite{king2012}:
 
@@
 
-Here $\gamma$ represents the skewness parameter.
-The Student-t($\cdot$) portion represents in order: the data point of interest, the heavy-tailedness parameter, the location parameter, and the scale parameter as passed to the standard Student-t probability density.
-This translates to the log probability density:
+Here $\gamma$ represents the skewness parameter. The Student-t($\cdot$) portion
+represents in order: the data point of interest, the heavy-tailedness
+parameter, the location parameter, and the scale parameter as passed to the
+standard Student-t probability density. This translates to the log probability
+density:
 
@@
 
-\section{Code availability.} Data and code to reproduce our analysis are available at \url{https://github.com/seananderson/heavy-tails} (currently private).
+\section{Code availability.} Data and code to reproduce our analysis are
+available at \url{https://github.com/seananderson/heavy-tails} (currently
+private).
 
@@
 
-\textbf{Table~S1: Summary statistics for the filtered Global Population Dynamics Database time series arranged by taxonomic class.} Columns are number of populations, number of taxonomic orders, numbers of species, median time series length, total number of interpolated time steps, total number of substituted zeros, and total number of time steps.
+\textbf{Table~S1: Summary statistics for the filtered Global Population
+Dynamics Database time series arranged by taxonomic class.} Columns are number
+of populations, number of taxonomic orders, numbers of species, median time
+series length, total number of interpolated time steps, total number of
+substituted zeros, and total number of time steps.
 

### 2016-05-26 text/supp.tex
Update citation style for supplement
Move most of the supplement into the methods section

@@
 %\usepackage{cite}
-\usepackage{citesupernumber}
-\usepackage{scicite}
+% \usepackage{citesupernumber}
+% \usepackage{scicite}
 
@@
 
-%\usepackage[round,sectionbib]{natbib}
-%\bibpunct{(}{)}{;}{a}{}{;}
+\usepackage[round,sectionbib]{natbib}
+\bibpunct{(}{)}{;}{a}{}{;}
 
@@
 
-\title{Supplementary Materials:\\Black-swan events in animal
+\title{Supporting Information:\\Black-swan events in animal
   populations}
@@
 
-\nocite{taleb2007}
-\nocite{sornette2009}
-\nocite{may2008}
-\nocite{newman2005}
-\nocite{harnik2012}
-\nocite{fey2015}
-\nocite{mangel1994}
-\nocite{ipcc2012}
-\nocite{keitt1998}
-\nocite{allen2001}
-\nocite{halley2002}
-\nocite{brook2006a}
-\nocite{knape2012}
-\nocite{SOM}
-\nocite{saucy1994}
-\nocite{stafford1971}
-\nocite{potts1980}
-\nocite{schindler2010}
-\nocite{carpenter2006}
-\nocite{nunez2012}
-\nocite{meehl2004}
-\nocite{katz2005}
-\nocite{kirby2009}
-\nocite{denny2009}
-\nocite{nrc2007}
-\nocite{carpenter2012}
-\nocite{caddy1996}
-\nocite{doak2008}
-\nocite{lindenmayer2010}
-\nocite{thompson2013}
+% \nocite{taleb2007}
+% \nocite{sornette2009}
+% \nocite{may2008}
+% \nocite{newman2005}
+% \nocite{harnik2012}
+% \nocite{fey2015}
+% \nocite{mangel1994}
+% \nocite{ipcc2012}
+% \nocite{keitt1998}
+% \nocite{allen2001}
+% \nocite{halley2002}
+% \nocite{brook2006a}
+% \nocite{knape2012}
+% \nocite{SOM}
+% \nocite{saucy1994}
+% \nocite{stafford1971}
+% \nocite{potts1980}
+% \nocite{schindler2010}
+% \nocite{carpenter2006}
+% \nocite{nunez2012}
+% \nocite{meehl2004}
+% \nocite{katz2005}
+% \nocite{kirby2009}
+% \nocite{denny2009}
+% \nocite{nrc2007}
+% \nocite{carpenter2012}
+% \nocite{caddy1996}
+% \nocite{doak2008}
+% \nocite{lindenmayer2010}
+% \nocite{thompson2013}
 
@@
 
-We selected abundance time series from the Global Population Dynamics Database\cite{gpdd2010} (GPDD), which contains nearly 5,000 time series of abundance from \(\sim\) 1,000 species and \(\sim\) 100 taxonomic orders.
-We filtered the data to remove populations from less reliable data sources, and those without sufficient data for our models, and then interpolated some missing values (sensu Brook \textit{et al.}\cite{brook2006a}).
-Our interpolation affected only \(\sim\) \interpPointsPerc \% of the final data points (table~S1) and none of the data points that were later considered black-swan events.
-Our final dataset contained \NPops\ populations across \NOrders\ taxonomic orders and seven taxonomic classes, with a median of \medianTimeSteps\ time steps (range \minTimeSteps--\maxTimeSteps) (table~S1).
-
 We applied the following data selection and quality-control rules to the
@@
   (on a scale of $1$ to $5$, with $1$ being the lowest quality data)
-  \cite{sibly2005, ziebarth2010}.
+  \citep{sibly2005, ziebarth2010}.
 
@@
   standard deviation. We verified this by locating the original papers the
-  datasets were extracted from: \cite{colebrook1978} for zooplankton and
-  \cite{lindstrom1995} for grouse. Since the papers did not include the
+  datasets were extracted from: \citet{colebrook1978} for zooplankton and
+  \citet{lindstrom1995} for grouse. Since the papers did not include the
   original mean time-series values we could not back transform these data
@@
 \item We filled in single recorded values of zero with the lowest non-zero
-  value in the time series \cite{brook2006a}. This assumes that
+  value in the time series \citep{brook2006a}. This assumes that
   single values of zero result from abundance being low enough that censusing
@@
 \item We removed all populations without at least four unique values
-  \cite{brook2006a}.
+  \citep{brook2006a}.
 
@@
   length, we took the most recent window. This is a longer window than used in
-  some previous analyses \cite{brook2006a}, but since our model
+  some previous analyses \citep{brook2006a}, but since our model
   attempts to capture the shape of the distribution tails, our model requires
@@
   suspected black swans: \texttt{1207} because the 1957 data point was entered
-  as 2 but should have been 27 \cite{kendeigh1982}, \texttt{6531} because the
-  1978 data point was entered as 7 but should have been 47 \cite{minot1986},
+  as 2 but should have been 27 \citep{kendeigh1982}, \texttt{6531} because the
+  1978 data point was entered as 7 but should have been 47 \citep{minot1986},
   and \texttt{6566} because some of the data did not match the graph
-  \cite{heessen1996}.
+  \citep{heessen1996}.
 
@@
 
-\section{Details on the heavy-tailed Gompertz probability model} We fit heavy-tailed Gompertz population dynamics models to data from the GPDD.
-The Gompertz model represents population growth as a linear function in log space.
-If \(x_{t}\) represents the ln abundance (\(N\)) at time \emph{t} then
-\begin{align}
-x_t &= \lambda + bx_{t - 1} + \epsilon_t\\
-\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma)
-\end{align}
-The growth parameter \(\lambda\) represents the expected growth rate if \(N_{t} = 1\).
-The model is density independent if \(b = 1\), maximally density dependent if \(b = 0\), and inversely density dependent if \(b < 0\).
-Usually, the process noise \(\epsilon_{t}\) is modelled as normally distributed, but we allow it to be drawn from a Student-t distribution with scale parameter \(\sigma\) and degrees of freedom \(\nu\).
-We can then estimate the degree to which the process deviations have heavy tails and are therefore evidence of black-swan events (Fig.~1A, B).
-
-For the Gompertz model, our weakly-informative priors (fig.~\ref{fig:priors}) were:
-\begin{align}
-b &\sim \mathrm{Uniform}(-1, 2)\\ \lambda &\sim \mathrm{Normal}(0, 10^2)\\
-\sigma &\sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)\\ \nu &\sim
-\mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2).
-\end{align}
-Our prior on \(b\) was uninformative between values of \(-1\) and \(2\). We
-would not expect values of \(b\) with levels of density dependence as low as
-\(-1\) (very strong inverse density dependence), nor would we generally expect
-values above \(1\). We allowed values of \(b\) above \(1\) to allow for
-non-stationary time series of growth rates. The estimates of \(b\) were well
-within these bounds. Our prior on \(\lambda\) was weakly informative within the
-range of expected values for population growth and is similar to the default
-priors suggested by Gelman \emph{et al.}\cite{gelman2008d} for intercepts of
-regression models. Our Half-Cauchy prior on \(\sigma\) follows Gelman \emph{et
-al.}\cite{gelman2006c} and Gelman \emph{et al.}\cite{gelman2008d} and the
-specific scale parameter of \(2.5\) is based on our expected range of the value
-in nature from previous studies\cite{connors2014}. In our testing of a
-subsample of populations, our parameter estimates were not qualitatively
-changed by switching to a uniform prior on \(\sigma\), but the weakly
-informative Half-Cauchy prior sped up chain convergence.
-
-We chose a weakly informative exponential prior on \(\nu\) with rate parameter
-of \(0.01\) truncated at values above two---a slightly less informative prior
-than suggested by Fernandez and Steel\cite{fernandez1998}. We truncated the
-distribution at \(2\), since at \(\nu < 2\) the variance of the t distribution
-is undefined. This prior gives only a \(7.7\)\% probability that \(\nu < 10\)
-but constrains the sampling sufficiently to avoid drifting towards infinity. In
-any case, for \(\nu > 20\) the t distribution is almost indistinguishable from
-the normal distribution (Fig.~1). Based on the shape of the t distribution, we
-chose the probability that \(\nu < 10\), Pr(\(\nu < 10\)), to define the
-probability of heavy-tailed dynamics. When categorizing a population as heavy
-tailed, we applied a 0.5 probability threshold. In the scenario where the data
-are uninformative about heavy tails (e.g.~Fig.~1E, H), the posterior for
-\(\nu\) will approximately match the prior (prior median \(= 71\), mean \(=
-102\)) and a metric of Pr\((\nu < 10) > 0.5\) is unlikely to flag the
-population as heavy tailed.
-
-We fit our models with Stan 2.6.0\cite{stan-manual2015, hoffman2014}, and R
-3.2.0\cite{r2015}. We began with four chains and 2,000 iterations, discarding
-the first 1,000 as warm up (i.e.~4,000 final samples). If \(\hat{R}\) (the
-potential scale reduction factor---a measure of chain convergence) was greater
-than \(1.05\) for any parameter, or if the minimum effective sample size,
-\(n_\mathrm{eff}\) (a measure of the effective number of uncorrelated samples),
-for any parameter was less than \(200\), we doubled both the total iterations
-and warm up period and sampled from the model again. These thresholds are in
-excess of the minimums recommended by Gelman \emph{et al.}\cite{gelman2006a} of
-\(\hat{R} < 1.1\) and \(n_\mathrm{eff} > 100\) for reliable point estimates and
-confidence intervals. In the majority of cases our minimum thresholds were
-greatly exceeded. We continued this procedure up to 8,000 iterations (16,000
-total samples) by which all chains were deemed to have sufficiently converged.
-The No-U-Turn Hamiltonian Markov chain Monte Carlo sampler in Stan generally
-requires fewer iterations to obtain equivalent effective sample sizes than the
-Gibbs or Metropolis--Hastings algorithms\cite{stan-manual2015}.
-
+% todo Add this section to the discussion :
 There are a number of caveats when considering the generality of our results.
@@
 less-extreme events and also dampen black-swan dynamics.
-
-\section{Alternative priors} To test if the prior on \(\nu\) influenced our
-estimate of black-swan dynamics, we refit our models with weaker and stronger
-priors. Our base model used a prior on \(\nu\) of Truncated-Exponential(0.01,
-min.\ = 2). For a weaker prior we used Truncated-Exponential(0.005, min.\ = 2)
-and for a stronger prior we used Truncated-Exponential(0.02, min.\ = 2)
-(fig.~\ref{fig:priors}c). Note that the base and weaker priors are relatively
-flat within the region of \(\nu < 20\), which is the region we are mostly
-concerned about when categorizing populations as heavy tailed. We furthermore
-considered a Uniform(\(0, 1\)) prior on \(1/\nu\) as used in Gelman \emph{et
-al.}\cite{gelman2014}, but this prior puts considerably more weight on lower
-(heavy-tailed) values of \(\nu\) than the truncated exponential priors
-considered here (not shown).
-
-Our results show that weaker and stronger priors have little influence on our
-conclusions about heavy-tailed dynamics (fig.~\ref{fig:alt-priors}). When the
-data are informative about tail behaviour (i.e. when there is strong evidence
-of low \(\nu\) values, upper-right of fig.~\ref{fig:alt-priors}), the prior has
-little impact on the estimate of \(\nu\). When the data are less informative
-about \(\nu\) (i.e.\ when there are no or few tail events and time series are
-short or noisy), the prior can pull the estimate of \(\nu\) towards larger or
-smaller values (fig.~\ref{fig:alt-priors}). The vast majority of the
-populations with Pr\((\nu < 10)\) in the base prior were not altered
-qualitatively by this range of prior strength.
-
-\section{Alternative population models} We fit alternative population models to
-test if four key phenomena systematically changed our conclusions. The range of
-percentages of black swans by taxonomic class cited in the abstract are based
-on lower and upper limits across our main Gompertz model and these four
-alternative models.
-
-\subsection{Autocorrelated residuals} We considered a version of the Gompertz
-model in which an autoregressive parameter was fit to the process-noise
-residuals: \begin{align} x_t &= \lambda + b x_{t-1} + \epsilon_t\\ \epsilon_t
-&\sim \mathrm{Student\mhyphen t}(\nu, \phi \epsilon_{t-1}, \sigma). \end{align}
-In addition to the parameters in the original Gompertz model, this model
-estimates an additional parameter \(\phi\), which represents the correlation of
-subsequent residuals. Based on the results of previous analyses with the
-GPDD\cite{connors2014} and the chosen prior in previous
-analyses\cite{thorson2014a}, we placed a moderately informative prior on
-\(\phi\) that assumed the greatest probability density near zero with the
-reduced possibility of \(\phi\) being near \(-1\) or \(1\). Specifically, we
-chose \(\phi \sim \mathrm{Truncated\mhyphen Normal}(0, 0.5, \mathrm{min.} = -1,
-\mathrm{max.} = 1)\).
-
-\subsection{Assumed density independence} We fit a simplified version of the
-Gompertz model in which the density dependence parameter \(b\) was fixed at
-\(1\) (density independent). This is equivalent to fitting a random walk model
-(with drift) to the \(\ln\) abundances or assuming the growth rates are drawn
-from a stationary distribution. The model was as follows:
-\begin{align}
-x_t &= \lambda + x_{t-1} + \epsilon_t\\
-\epsilon &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma).
-\end{align}
-We fit this model for three reasons: (1) it is computationally simpler and so
-provides a check that our more complicated full Gompertz model was obtaining
-reasonable estimates of \(\nu\), (2) it provides a test of whether density
-dependence was systematically affecting our perception of heavy tails, (3) it
-matches how some previous authors have modelled heavy tails without accounting
-for density dependence\cite{segura2013}.
-
-\subsection{Assumed observation error} Observation error can bias parameter
-estimates\cite{knape2012} and is known to affect the ability to detect extreme
-events\cite{ward2007}. In our main analysis, we fit a model that ignored
-observation error. One way to account for observation error would be to fit a
-full state-space model that simultaneously estimates the magnitude of process
-noise and observation error. However, simultaneously estimating observation and
-process noise is a challenging problem (e.g.\ because the observation and
-process noise parameters tend to negatively covary in model fitting) and is
-known to sometimes result in identifiability issues with the Gompertz
-population model\cite{knape2008}. Furthermore, our model was applied to
-hundreds of time series, often of short length (as few as \minTimeSteps\ time
-steps) and our model estimates an additional parameter---the shape of the
-process deviation tails---potentially making identifiability and computational
-issues even greater. Therefore, we considered a version of the base Gompertz
-model that allowed for a fixed level of observation error:
-\begin{align}
-U_t &= \lambda + b U_{t-1} + \epsilon_t\\
-x_t &\sim \mathrm{Normal}(U_t, \sigma_\mathrm{obs}^2)\\
-\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma_\mathrm{proc}),
-\end{align}
-where \(U\) represents the unobserved state vector, \(\sigma_\mathrm{obs}\)
-represents the standard deviation of observation error (on a log scale), and
-\(\sigma_\mathrm{proc}\) represent the process noise scale parameter. We set
-\(\sigma_\mathrm{obs}\) to \(0.2\), which represents a typical value used in
-simulation analyses\cite{valpine2002, thorson2014b}.
-
-\subsection{Ricker-logistic} We also fit a Ricker-logistic model:
-\begin{align}
-x_t &= x_{t-1} + r_{\mathrm{max}}\left(1 - \frac{N_{t-1}}{K}\right) + \epsilon_t\\
-\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma),
-\end{align}
-where \(r_\mathrm{max}\) represents the theoretical maximum population growth
-rate that is obtained when \(N_t = 0\). The parameter \(K\) represents the
-carrying capacity and, as before, \(x_t\) represents the \(\ln\) transformed
-abundance at time \(t\). The Ricker-logistic model assumes a linear decrease in
-population growth rate with increases in abundance. In contrast, the Gompertz
-model assumes a linear decrease in population growth rate with increases in
-\textit{log} abundance.
-
-To fit the Ricker-logistic models, we chose a prior on \(K\) uniform between
-zero and twice the maximum observed abundance (similar and less informative
-than in Delean \emph{et al.}\cite{delean2013}). We set the prior on
-\(r_\mathrm{max}\) as uniform between 0 and 20 as in Delean \emph{et
-al.}\cite{delean2013}. We used the same priors on \(\nu\) and \(\sigma\) as in
-the Gompertz model.
-
-\section{Simulation testing} We performed two types of simulation testing.
-First, we tested how easily the Student-t distribution \(\nu\) parameter could
-be recovered given different true values of \(\nu\) and different sample sizes.
-Second, we tested the ability of the heavy-tailed Gompertz model to obtain
-unbiased parameter estimates of \(\nu\) given that a set of process deviations
-was provided in which the effective \(\nu\) value was close to the true \(\nu\)
-value.
-
-We found that we could consistently and precisely recover median posterior
-estimates of \(\nu\) near the true value of \(\nu\) with large samples (\(\ge
-200\)) (fig.~\ref{fig:sim-nu} upper panels). At smaller samples we could
-usually qualitatively distinguish heavy from not-heavy tails, but the model
-tended to underestimate how heavy the tails were. At the same time, at smaller
-sample sizes, the model tended to overestimate how large the scale parameter
-was (fig.~\ref{fig:sim-nu} lower panels).
-
-Our simulation-testing results indicate that the Gompertz model can recapture
-the true value of \(\nu\) when the process noise was chosen so that appropriate
-tail events were present (fig.~\ref{fig:sim-prob} upper panels). The addition
-of observation error caused the model to tend to underestimate the degree of
-heavy-tailedness. Fitting a model with correctly specified observation error
-did not make substantial improvements to model bias (fig.~\ref{fig:sim-prob}).
 
@@
 
-\section{Covariates of heavy-tailed dynamics} We fit a hierarchical beta
-regression model to the predicted probability of heavy tails, Pr(\(\nu < 10\)),
-to investigate potential covariates of heavy-tailed dynamics. We obtained
-maximum lifespan and body-size data from Brook \emph{et al.}\cite{brook2006a}.
-The beta distribution is useful when response data are on a continuous scale
-between zero and one\cite{ferrari2004}. The model was as follows:
-\begin{align}
-\mathrm{Pr}(\nu_i < 10) &\sim \mathrm{Beta}(A_i, B_i)\\
-\mu_i &= \mathrm{logit}^{-1}(\alpha
-  + \alpha^\mathrm{class}_{j[i]}
-  + \alpha^\mathrm{order}_{k[i]}
-  + \alpha^\mathrm{species}_{l[i]}
-  + X_i \beta),
-  \: \text{for } i = 1, \dots, 617\\
-A_i &= \phi_\mathrm{disp} \mu_i\\
-B_i &= \phi_\mathrm{disp} (1 - \mu_i)\\
-\alpha^\mathrm{class}_j &\sim
-  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{class}}),
-  \: \text{for } j = 1, \dots, 6\\
-\alpha^\mathrm{order}_k &\sim
-  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{order}}),
-  \: \text{for } k = 1, \dots, 38\\
-\alpha^\mathrm{species}_l &\sim
-  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{species}}),
-  \: \text{for } l = 1, \dots, 301,
-\end{align}
-where \(A\) and \(B\) represent the beta distribution shape parameters;
-\(\mu_i\) represents the predicted value for population \(i\), class \(j\),
-order \(k\), and species \(l\); \(\phi_\mathrm{disp}\) represents the
-dispersion parameter; and \(X_i\) represents a vector of predictors (such as
-lifespan) for population \(i\) with associated \(\beta\) coefficients. The
-intercepts are allowed to vary from the overall intercept \(\alpha\) by
-taxonomic class (\(\alpha^\mathrm{class}_j\)), taxonomic order
-(\(\alpha^\mathrm{order}_k\)), and species (\(\alpha^\mathrm{species}_l\)) with
-standard deviations \(\sigma_{\alpha \; \mathrm{class}}\), \(\sigma_{\alpha \;
-\mathrm{order}}\), and \(\sigma_{\alpha \; \mathrm{species}}\). Where possible,
-we also allowed for error distributions around the predictors by incorporating
-the standard deviation of the posterior samples for the Gompertz parameters
-\(\lambda\), \(b\), and \(\log \sigma\) around the mean point value as normal
-distributions (not shown in the above equation).
-
-We log transformed \(\sigma\), time-series length, and lifespan to match the
-way they are visually represented in fig.~\ref{fig:correlates} and to make the
-relationship approximately linear on the logit-transformed response scale. All
-input variables were standardized by subtracting their mean and dividing by two
-standard deviations to make their coefficients comparable in
-magnitude\cite{gelman2008c}. We excluded body length as a covariate because it
-was highly correlated with lifespan, and lifespan exhibited more overlap across
-taxonomy than body length. Lifespan is also more directly related to time and
-potential mechanisms driving black-swan dynamics.
-
-We incorporated weakly informative priors into our model: \(\mathrm{Cauchy}(0,
-10)\) on the global intercept \(\alpha\), \(\mathrm{Half\mhyphen Cauchy}(0,
-2.5)\) on all standard deviation parameters, \(\mathrm{Half\mhyphen Cauchy}(0,
-10)\) on the dispersion parameter \(\phi_\mathrm{disp}\), and
-\(\mathrm{Cauchy}(0, 2.5)\) on all other parameters\cite{gelman2006c,
-gelman2008d}. Compared to normal priors, the Cauchy priors concentrate more
-probability density around expected parameter values while allowing for a
-higher probability density far into the tails, thereby allowing the data to
-dominate the posterior more strongly if it disagrees with the prior. We fit our
-models with 5,000 total iterations per chain, 2500 warm-up iterations, four
-chains, and discarding every second sample. We checked for chain convergence
-visually and with the same criteria as before (\(\widehat{R} < 1.05\) and
-\(n_\mathrm{eff} >200\) for all parameters). We further checked our model
-against a similar model with the glmmADMB \textsf{R} package\cite{glmmadmb}
-without uncertainty around the predictors.
-
-To derive taxonomic-order-level estimates of the probability of heavy tails
-accounting for time-series length (Fig.~3A), we fit a separate hierarchical
-model with the same structure but with only \(\log\) time-series length as a
-predictor---in this case, we did not want to control for intrinsic population
-characteristics such as density dependence. Since our predictors were centered
-by subtracting their mean value, we obtained order-level estimates of the
-probability of heavy tails at mean log time-series length by adding the
-posteriors for \(\alpha\), \(\alpha^\mathrm{class}_j\), and
-\(\alpha^\mathrm{order}_k\).
-
-\section{Skewed Student-t forecasts} To formally evaluate the apparent skewness
-of heavy-tailed process noise we fit Gompertz models with skewed Student-t
-distributed process noise\cite{fernandez1998}. The distribution adds one
-parameter to the Student-t distribution, \(\gamma\), which controls the
-skewness. The distribution is symmetrical if \(\gamma = 1\), left skewed if \(0
-< \gamma < 1\), and right skewed if \(1 < \gamma < \infty\). We placed a weakly
-informative prior of \(\mathrm{Cauchy}(0, 2.5)\) on \(\log \gamma\). We
-aggregated 5,000 randomly selected posterior samples from the \(\gamma\)
-parameter of each model at three levels of evidence for heavy-tails:
-\emph{heavy tailed} median \(\nu < 10\); \emph{slightly heavy tailed} \(10 \le
-\mathrm{median}\, \nu \le 70\), and \emph{normal tails}, median \(\nu > 70\)
-(Fig.~4).
-
-To generate five-year forecasts of abundance, we combined the posterior
-parameter samples from the skewed-t Gompertz models with stochastically
-generated process noise. We compared these forecasts to those generated by a
-standard Gompertz model with normally distributed process noise fit to the same
-data. We calculated the ratio of the lower 99\% quantile credible interval
-between the two model projections for all populations in which Pr(\(\nu < 10\))
-\(> 0.5\). To ensure the results had stabilized in the tails of the forecast
-posterior, we increased the number of posterior samples in Stan. We ran four
-chains of 20,000 iterations and discarded the first 10,000 as warmup for a
-total of 40,000 samples.
-
 \section{Implementation of the skewed Student-t distribution in Stan}
@@
 The probability density for the skewed Student-t distribution
-\cite{fernandez1998} equates to the following probability density as
-implemented in Ref.~\cite{king2012}:
+\citep{fernandez1998} equates to the following probability density as
+implemented in \citet{king2012}:
 
@@
 Since Stan only needs the probability density up to an additive
-constant \cite{stan-manual2014}, we
+constant \citep{stan-manual2014}, we
 can replace $\log(2 \gamma)$ in the above for $\log(\gamma)$. The final log
@@
 
-\bibliographystyle{Science}
 %\bibliography{/Users/seananderson/Dropbox/tex/jshort,supp}
-\bibliography{jshort,newbib}
+
+\bibliographystyle{ecologyletters}\bibliography{jshort,newbib}
 

### 2016-05-26 text/ms.tex
Update cover page

@@
-\textbf{Title: Black-swan events in animal populations}
+\section{Black-swan events in animal populations}
 
-\textbf{Authors}: Sean C. Anderson\textsuperscript{1,2*}, Trevor A.
+Sean C. Anderson\textsuperscript{1,2*}, Trevor A.
 Branch\textsuperscript{3}, Andrew B. Cooper\textsuperscript{2}, Nicholas K.
@@
 
-%125
+Short title: Black-swan events in animal populations
 
-\textbf{Abstract}: Black swans are improbable events that nonetheless
+Keywords:
+Bayesian,
+black swans,
+die-offs,
+ecological surprises,
+heavy tails,
+Gompertz,
+mass mortality,
+population dynamics,
+risk,
+Student t
+
+Words in abstract: TOdO
+
+Words in main text: todo
+
+Number of references: todo
+
+Number of figures: 4
+
+Number of tables: 0
+
+\newpage
+%125 now is 150
+
+\section{Abstract}
 occur---often with profound consequences. While physical extremes, such as

### 2016-05-26 text/ms.tex
Revise main text including adding new introduction paragraph

@@
 \section{Abstract}
+
+Black swans are improbable events that nonetheless
 occur---often with profound consequences. While physical extremes, such as
@@
 
-% \textbf{Main text}: Major surprises happen more often than expected in
-financial, social, and natural systems\cite{taleb2007, sornette2009, may2008}.
+\section{Introduction}
+
+Major surprises happen more often than expected in
+financial, social, and natural systems \citep{taleb2007, sornette2009, may2008}.
 Massive unpredictable market swings are responsible for the majority of
-financial gains and losses\cite{taleb2007}, fatalities from the largest wars
-dwarf those from all others \cite{newman2005}, and the frequency of the most
-damaging earthquakes has exceeded past expectations\cite{sornette2009}. In
+financial gains and losses \citep{taleb2007}, fatalities from the largest wars
+dwarf those from all others  \citep{newman2005}, and the frequency of the most
+damaging earthquakes has exceeded past expectations \citep{sornette2009}. In
 ecological systems, background rates of global extinction are punctuated by
-mass extinction\cite{harnik2012}, viruses can mutate suddenly to infect new
+mass extinction \citep{harnik2012}, viruses can mutate suddenly to infect new
 hosts, and billions of animals can die at once in mass mortality
-events\cite{fey2015}. Indeed, such die-offs may be the most important element
-affecting population persistence\cite{mangel1994} and their importance is
+events \citep{fey2015}.
+
+Indeed, such die-offs may be the most important element
+affecting population persistence \citep{mangel1994} and their importance is
 likely to increase given projected increases in the frequency and magnitude of
-climate-related extremes\cite{ipcc2012}. Despite this anecdotal evidence for
-ecological black swans, systematic searches of large numbers of timeseries have
-yet to reveal black-swan events\cite{keitt1998, allen2001, halley2002} However,
-the overwhelming majority of model fitting and risk forecasting assumes that
-deviations from model predictions can be represented by a normal
-distribution\cite{brook2006a, knape2012}. If black swans occur, though,
-a normal distribution would under-estimate the probability of extreme events
-occurring.
+climate-related extremes \citep{ipcc2012}.
+% Despite this anecdotal evidence for
+% ecological black swans, systematic searches of large numbers of timeseries have
+% yet to reveal black-swan events \citep{keitt1998, allen2001, halley2002}.
+However, the overwhelming majority of population model fitting and risk forecasting
+assumes that deviations from model predictions can be represented by a normal
+distribution \citep[typically on a log scale; e.g.][]{brook2006a, knape2012}.
+If black swans occur, though, a normal distribution would under-estimate the
+probability of extreme events occurring.
+
+There are many examples of black-swan events in populations
+but we lack a large-scale comparative analysis
+with a method that allows for
+population dynamics
+(e.g.\ per capita productivity, density dependence, and typical stochasticity).
+Previous comparative analyses fitting
+heavy-tailed distributions to timeseries
+(distributions with higher probabilities of extreme events
+than the normal distribution)
+have not accounted for an underlying population dynamics model
+\citep{keitt1998, allen2001, halley2002, segura2013}.
+Alternatively, most examples of population die offs come from identifying
+jumps in abundance that exceed a threshold of decline
+\citep[e.g.][]{young1994, gerber2001, reed2003, fey2015}.
+But this approach gives no indication of relative frequency
+and again does not distinguish events from
+expected dynamics.
+Methodologically, the generalized extreme value (GEV) distribution represents
+the limit distribution of a series of maxima or minima.
+Although the GEV is well-suited for physical ecological variables
+\citep[e.g.\ sedimentation rates or wind speeds;][]{gaines1993, denny2009},
+because the GEV requires maxima per time block
+or peaks beyond a chosen threshold,
+the GEV requires longer timeseries then typically available
+for populations and is not simply integrated
+into a population dynamics model.
 
@@
 taxonomic groups---including many  birds, mammals, insects, and fishes
-(table~S1) \cite{SOM}. We identify characteristics of time series or
+(Table~S1). We identify characteristics of time series or
 life-history traits associated with the detection of black-swan events and
@@
 
+\section{Methods}
+
+\subsection{Data}
+
+We selected abundance time series from the Global Population Dynamics
+Database \citep{gpdd2010} (GPDD), which contains nearly 5,000 time series of
+abundance from \(\sim\) 1,000 species and \(\sim\) 100 taxonomic orders. We
+filtered the data (Supporting Information) to remove populations from less
+reliable data sources, and those without sufficient data for our models, and
+then interpolated some missing values (sensu Brook \textit{et
+al.} \citep{brook2006a}). Our interpolation affected only \(\sim\)
+\interpPointsPerc \% of the final data points (Table~S1) and none of the data
+points that were later considered black-swan events. Our final dataset
+contained \NPops\ populations across \NOrders\ taxonomic orders and seven
+taxonomic classes, with a median of \medianTimeSteps\ time steps (range
+\minTimeSteps--\maxTimeSteps) (Table~S1).
+
+\subsection{Modeling framework}
+
+We fit heavy-tailed Gompertz population dynamics models to data from the GPDD.
+The Gompertz model represents population growth as a linear function in log space.
+If \(x_{t}\) represents the ln abundance (\(N\)) at time \emph{t} then
+\begin{align}
+x_t &= \lambda + bx_{t - 1} + \epsilon_t\\
+\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma)
+\end{align}
+The growth parameter \(\lambda\) represents the expected growth rate if \(N_{t}
+= 1\). The model is density independent if \(b = 1\), maximally density
+dependent if \(b = 0\), and inversely density dependent if \(b < 0\). Usually,
+the process noise \(\epsilon_{t}\) is modelled as normally distributed, but we
+allow it to be drawn from a Student-t distribution with scale parameter
+\(\sigma\) and degrees of freedom \(\nu\). We can then estimate the degree to
+which the process deviations have heavy tails and are therefore evidence of
+black-swan events (Fig.~1A, B).
+
+For the Gompertz model, our weakly-informative priors (Fig.~\ref{fig:priors}) were:
+\begin{align}
+b &\sim \mathrm{Uniform}(-1, 2)\\ \lambda &\sim \mathrm{Normal}(0, 10^2)\\
+\sigma &\sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)\\ \nu &\sim
+\mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2).
+\end{align}
+Our prior on \(b\) was uninformative between values of \(-1\) and \(2\). We
+would not expect values of \(b\) with levels of density dependence as low as
+\(-1\) (very strong inverse density dependence), nor would we generally expect
+values above \(1\). We allowed values of \(b\) above \(1\) to allow for
+non-stationary time series of growth rates. The estimates of \(b\) were well
+within these bounds. Our prior on \(\lambda\) was weakly informative within the
+range of expected values for population growth and is similar to the default
+priors suggested by Gelman \emph{et al.} \citep{gelman2008d} for intercepts of
+regression models. Our Half-Cauchy prior on \(\sigma\) follows Gelman \emph{et
+al.} \citep{gelman2006c} and Gelman \emph{et al.} \citep{gelman2008d} and the
+specific scale parameter of \(2.5\) is based on our expected range of the value
+in nature from previous studies \citep{connors2014}. In our testing of a
+subsample of populations, our parameter estimates were not qualitatively
+changed by switching to a uniform prior on \(\sigma\), but the weakly
+informative Half-Cauchy prior sped up chain convergence.
+
+We chose a weakly informative exponential prior on \(\nu\) with rate parameter
+of \(0.01\) truncated at values above two---a slightly less informative prior
+than suggested by Fernandez and Steel \citep{fernandez1998}. We truncated the
+distribution at \(2\), since at \(\nu < 2\) the variance of the t distribution
+is undefined. This prior gives only a \(7.7\)\% probability that \(\nu < 10\)
+but constrains the sampling sufficiently to avoid drifting towards infinity. In
+any case, for \(\nu > 20\) the t distribution is almost indistinguishable from
+the normal distribution (Fig.~1). Based on the shape of the t distribution, we
+chose the probability that \(\nu < 10\), Pr(\(\nu < 10\)), to define the
+probability of heavy-tailed dynamics. When categorizing a population as heavy
+tailed, we applied a 0.5 probability threshold. In the scenario where the data
+are uninformative about heavy tails (e.g.~Fig.~1E, H), the posterior for
+\(\nu\) will approximately match the prior (prior median \(= 71\), mean \(=
+102\)) and a metric of Pr\((\nu < 10) > 0.5\) is unlikely to flag the
+population as heavy tailed.
+
+We fit our models with Stan 2.6.0 \citep{stan-manual2015, hoffman2014}, and R
+3.2.0 \citep{r2015}. We began with four chains and 2,000 iterations, discarding
+the first 1,000 as warm up (i.e.~4,000 final samples). If \(\hat{R}\) (the
+potential scale reduction factor---a measure of chain convergence) was greater
+than \(1.05\) for any parameter, or if the minimum effective sample size,
+\(n_\mathrm{eff}\) (a measure of the effective number of uncorrelated samples),
+for any parameter was less than \(200\), we doubled both the total iterations
+and warm up period and sampled from the model again. These thresholds are in
+excess of the minimums recommended by Gelman \emph{et al.} \citep{gelman2006a} of
+\(\hat{R} < 1.1\) and \(n_\mathrm{eff} > 100\) for reliable point estimates and
+confidence intervals. In the majority of cases our minimum thresholds were
+greatly exceeded. We continued this procedure up to 8,000 iterations (16,000
+total samples) by which all chains were deemed to have sufficiently converged.
+The No-U-Turn Hamiltonian Markov chain Monte Carlo sampler in Stan generally
+requires fewer iterations to obtain equivalent effective sample sizes than the
+Gibbs or Metropolis--Hastings algorithms \citep{stan-manual2015}.
+
 Specifically, we fit population dynamics models in which the process noise is
@@
 have heavy tails and are therefore diagnostic of black-swan events (Fig.~1,
-fig.~S\ref{fig:priors}). Lower values of \(\nu\) result in heavy-tailed
+Fig.~S\ref{fig:priors}). Lower values of \(\nu\) result in heavy-tailed
 distributions. For example, at \(\nu = 2\), the probability of drawing a value
@@
 distribution approaches the normal distribution (Fig.~1).
+
+\subsection{Alternative priors}
+
+To test if the prior on \(\nu\) influenced our estimate of black-swan dynamics,
+we refit our models with weaker and stronger priors. Our base model used a
+prior on \(\nu\) of Truncated-Exponential(0.01, min.\ = 2). For a weaker prior
+we used Truncated-Exponential(0.005, min.\ = 2) and for a stronger prior we
+used Truncated-Exponential(0.02, min.\ = 2) (Fig.~\ref{fig:priors}c). Note that
+the base and weaker priors are relatively flat within the region of \(\nu <
+20\), which is the region we are mostly concerned about when categorizing
+populations as heavy tailed. We furthermore considered a Uniform(\(0, 1\))
+prior on \(1/\nu\) as used in Gelman \emph{et al.} \citep{gelman2014}, but this
+prior puts considerably more weight on lower (heavy-tailed) values of \(\nu\)
+than the truncated exponential priors considered here (not shown).
+
+\subsection{Alternative population models} We fit alternative population models to
+test if four key phenomena systematically changed our conclusions. The range of
+percentages of black swans by taxonomic class cited in the abstract are based
+on lower and upper limits across our main Gompertz model and these four
+alternative models.
+
+\subsubsection{Autocorrelated residuals} We considered a version of the Gompertz
+model in which an autoregressive parameter was fit to the process-noise
+residuals: \begin{align} x_t &= \lambda + b x_{t-1} + \epsilon_t\\ \epsilon_t
+&\sim \mathrm{Student\mhyphen t}(\nu, \phi \epsilon_{t-1}, \sigma). \end{align}
+In addition to the parameters in the original Gompertz model, this model
+estimates an additional parameter \(\phi\), which represents the correlation of
+subsequent residuals. Based on the results of previous analyses with the
+GPDD \citep{connors2014} and the chosen prior in previous
+analyses \citep{thorson2014a}, we placed a moderately informative prior on
+\(\phi\) that assumed the greatest probability density near zero with the
+reduced possibility of \(\phi\) being near \(-1\) or \(1\). Specifically, we
+chose \(\phi \sim \mathrm{Truncated\mhyphen Normal}(0, 0.5, \mathrm{min.} = -1,
+\mathrm{max.} = 1)\).
+
+\subsubsection{Density independence} We fit a simplified version of the
+Gompertz model in which the density dependence parameter \(b\) was fixed at
+\(1\) (density independent). This is equivalent to fitting a random walk model
+(with drift) to the \(\ln\) abundances or assuming the growth rates are drawn
+from a stationary distribution. The model was as follows:
+\begin{align}
+x_t &= \lambda + x_{t-1} + \epsilon_t\\
+\epsilon &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma).
+\end{align}
+We fit this model for three reasons: (1) it is computationally simpler and so
+provides a check that our more complicated full Gompertz model was obtaining
+reasonable estimates of \(\nu\), (2) it provides a test of whether density
+dependence was systematically affecting our perception of heavy tails, (3) it
+matches how some previous authors have modelled heavy tails without accounting
+for density dependence \citep{segura2013}.
+
+\subsubsection{Observation error} Observation error can bias parameter
+estimates \citep{knape2012} and is known to affect the ability to detect extreme
+events \citep{ward2007}. In our main analysis, we fit a model that ignored
+observation error. One way to account for observation error would be to fit a
+full state-space model that simultaneously estimates the magnitude of process
+noise and observation error. However, simultaneously estimating observation and
+process noise is a challenging problem (e.g.\ because the observation and
+process noise parameters tend to negatively covary in model fitting) and is
+known to sometimes result in identifiability issues with the Gompertz
+population model \citep{knape2008}. Furthermore, our model was applied to
+hundreds of time series, often of short length (as few as \minTimeSteps\ time
+steps) and our model estimates an additional parameter---the shape of the
+process deviation tails---potentially making identifiability and computational
+issues even greater. Therefore, we considered a version of the base Gompertz
+model that allowed for a fixed level of observation error:
+\begin{align}
+U_t &= \lambda + b U_{t-1} + \epsilon_t\\
+x_t &\sim \mathrm{Normal}(U_t, \sigma_\mathrm{obs}^2)\\
+\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma_\mathrm{proc}),
+\end{align}
+where \(U\) represents the unobserved state vector, \(\sigma_\mathrm{obs}\)
+represents the standard deviation of observation error (on a log scale), and
+\(\sigma_\mathrm{proc}\) represent the process noise scale parameter. We set
+\(\sigma_\mathrm{obs}\) to \(0.2\), which represents a typical value used in
+simulation analyses \citep{valpine2002, thorson2014b}.
+
+\subsubsection{Ricker-logistic} We also fit a Ricker-logistic model:
+\begin{align}
+x_t &= x_{t-1} + r_{\mathrm{max}}\left(1 - \frac{N_{t-1}}{K}\right) + \epsilon_t\\
+\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma),
+\end{align}
+where \(r_\mathrm{max}\) represents the theoretical maximum population growth
+rate that is obtained when \(N_t = 0\). The parameter \(K\) represents the
+carrying capacity and, as before, \(x_t\) represents the \(\ln\) transformed
+abundance at time \(t\). The Ricker-logistic model assumes a linear decrease in
+population growth rate with increases in abundance. In contrast, the Gompertz
+model assumes a linear decrease in population growth rate with increases in
+\textit{log} abundance.
+
+To fit the Ricker-logistic models, we chose a prior on \(K\) uniform between
+zero and twice the maximum observed abundance (similar and less informative
+than in \citet{delean2013}). We set the prior on
+\(r_\mathrm{max}\) as uniform between 0 and 20 as in \citet{delean2013}. We used the same priors on \(\nu\) and \(\sigma\) as in
+the Gompertz model.
+
+\subsection{Simulation testing}
+We performed two types of simulation testing.
+First, we tested how easily the Student-t distribution \(\nu\) parameter could
+be recovered given different true values of \(\nu\) and different sample sizes.
+Second, we tested the ability of the heavy-tailed Gompertz model to obtain
+unbiased parameter estimates of \(\nu\) given that a set of process deviations
+was provided in which the effective \(\nu\) value was close to the true \(\nu\)
+value.
+
+We found that we could consistently and precisely recover median posterior
+estimates of \(\nu\) near the true value of \(\nu\) with large samples (\(\ge
+200\)) (Fig.~\ref{fig:sim-nu} upper panels). At smaller samples we could
+usually qualitatively distinguish heavy from not-heavy tails, but the model
+tended to underestimate how heavy the tails were. At the same time, at smaller
+sample sizes, the model tended to overestimate how large the scale parameter
+was (Fig.~\ref{fig:sim-nu} lower panels).
+
+Our simulation-testing results indicate that the Gompertz model can recapture
+the true value of \(\nu\) when the process noise was chosen so that appropriate
+tail events were present (Fig.~\ref{fig:sim-prob} upper panels). The addition
+of observation error caused the model to tend to underestimate the degree of
+heavy-tailedness. Fitting a model with correctly specified observation error
+did not make substantial improvements to model bias (Fig.~\ref{fig:sim-prob}).
+
+\subsection{Covariates of heavy-tailed dynamics} We fit a hierarchical beta
+regression model to the predicted probability of heavy tails, Pr(\(\nu < 10\)),
+to investigate potential covariates of heavy-tailed dynamics. We obtained
+maximum lifespan and body-size data from \citet{brook2006a}.
+The beta distribution is useful when response data are on a continuous scale
+between zero and one \citep{ferrari2004}. The model was as follows:
+\begin{align}
+\mathrm{Pr}(\nu_i < 10) &\sim \mathrm{Beta}(A_i, B_i)\\
+\mu_i &= \mathrm{logit}^{-1}(\alpha
+  + \alpha^\mathrm{class}_{j[i]}
+  + \alpha^\mathrm{order}_{k[i]}
+  + \alpha^\mathrm{species}_{l[i]}
+  + X_i \beta),
+  \: \text{for } i = 1, \dots, 617\\
+A_i &= \phi_\mathrm{disp} \mu_i\\
+B_i &= \phi_\mathrm{disp} (1 - \mu_i)\\
+\alpha^\mathrm{class}_j &\sim
+  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{class}}),
+  \: \text{for } j = 1, \dots, 6\\
+\alpha^\mathrm{order}_k &\sim
+  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{order}}),
+  \: \text{for } k = 1, \dots, 38\\
+\alpha^\mathrm{species}_l &\sim
+  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{species}}),
+  \: \text{for } l = 1, \dots, 301,
+\end{align}
+where \(A\) and \(B\) represent the beta distribution shape parameters;
+\(\mu_i\) represents the predicted value for population \(i\), class \(j\),
+order \(k\), and species \(l\); \(\phi_\mathrm{disp}\) represents the
+dispersion parameter; and \(X_i\) represents a vector of predictors (such as
+lifespan) for population \(i\) with associated \(\beta\) coefficients. The
+intercepts are allowed to vary from the overall intercept \(\alpha\) by
+taxonomic class (\(\alpha^\mathrm{class}_j\)), taxonomic order
+(\(\alpha^\mathrm{order}_k\)), and species (\(\alpha^\mathrm{species}_l\)) with
+standard deviations \(\sigma_{\alpha \; \mathrm{class}}\), \(\sigma_{\alpha \;
+\mathrm{order}}\), and \(\sigma_{\alpha \; \mathrm{species}}\). Where possible,
+we also allowed for error distributions around the predictors by incorporating
+the standard deviation of the posterior samples for the Gompertz parameters
+\(\lambda\), \(b\), and \(\log \sigma\) around the mean point value as normal
+distributions (not shown in the above equation).
+
+We log transformed \(\sigma\), time-series length, and lifespan to match the
+way they are visually represented in Fig.~\ref{fig:correlates} and to make the
+relationship approximately linear on the logit-transformed response scale. All
+input variables were standardized by subtracting their mean and dividing by two
+standard deviations to make their coefficients comparable in
+magnitude \citep{gelman2008c}. We excluded body length as a covariate because it
+was highly correlated with lifespan, and lifespan exhibited more overlap across
+taxonomy than body length. Lifespan is also more directly related to time and
+potential mechanisms driving black-swan dynamics.
+
+We incorporated weakly informative priors into our model: \(\mathrm{Cauchy}(0,
+10)\) on the global intercept \(\alpha\), \(\mathrm{Half\mhyphen Cauchy}(0,
+2.5)\) on all standard deviation parameters, \(\mathrm{Half\mhyphen Cauchy}(0,
+10)\) on the dispersion parameter \(\phi_\mathrm{disp}\), and
+\(\mathrm{Cauchy}(0, 2.5)\) on all other parameters \citep{gelman2006c,
+gelman2008d}. Compared to normal priors, the Cauchy priors concentrate more
+probability density around expected parameter values while allowing for a
+higher probability density far into the tails, thereby allowing the data to
+dominate the posterior more strongly if it disagrees with the prior. We fit our
+models with 5,000 total iterations per chain, 2500 warm-up iterations, four
+chains, and discarding every second sample. We checked for chain convergence
+visually and with the same criteria as before (\(\widehat{R} < 1.05\) and
+\(n_\mathrm{eff} >200\) for all parameters). We further checked our model
+against a similar model with the glmmADMB \textsf{R} package \citep{glmmadmb}
+without uncertainty around the predictors.
+
+To derive taxonomic-order-level estimates of the probability of heavy tails
+accounting for time-series length (Fig.~3A), we fit a separate hierarchical
+model with the same structure but with only \(\log\) time-series length as a
+predictor---in this case, we did not want to control for intrinsic population
+characteristics such as density dependence. Since our predictors were centered
+by subtracting their mean value, we obtained order-level estimates of the
+probability of heavy tails at mean log time-series length by adding the
+posteriors for \(\alpha\), \(\alpha^\mathrm{class}_j\), and
+\(\alpha^\mathrm{order}_k\).
+
+\subsection{Skewed Student-t forecasts} To formally evaluate the apparent skewness
+of heavy-tailed process noise we fit Gompertz models with skewed Student-t
+distributed process noise \citep{fernandez1998}. The distribution adds one
+parameter to the Student-t distribution, \(\gamma\), which controls the
+skewness. The distribution is symmetrical if \(\gamma = 1\), left skewed if \(0
+< \gamma < 1\), and right skewed if \(1 < \gamma < \infty\). We placed a weakly
+informative prior of \(\mathrm{Cauchy}(0, 2.5)\) on \(\log \gamma\). We
+aggregated 5,000 randomly selected posterior samples from the \(\gamma\)
+parameter of each model at three levels of evidence for heavy-tails:
+\emph{heavy tailed} median \(\nu < 10\); \emph{slightly heavy tailed} \(10 \le
+\mathrm{median}\, \nu \le 70\), and \emph{normal tails}, median \(\nu > 70\)
+(Fig.~4).
+
+To generate five-year forecasts of abundance, we combined the posterior
+parameter samples from the skewed-t Gompertz models with stochastically
+generated process noise. We compared these forecasts to those generated by a
+standard Gompertz model with normally distributed process noise fit to the same
+data. We calculated the ratio of the lower 99\% quantile credible interval
+between the two model projections for all populations in which Pr(\(\nu < 10\))
+\(> 0.5\). To ensure the results had stabilized in the tails of the forecast
+posterior, we increased the number of posterior samples in Stan. We ran four
+chains of 20,000 iterations and discarded the first 10,000 as warmup for a
+total of 40,000 samples.
+
+\section{Results}
 
@@
 density dependence, or modelled the population dynamics as Ricker-logistic
-instead of Gompertz (fig.~S\ref{fig:alt}). Similarly, introducing moderate
+instead of Gompertz (Fig.~S\ref{fig:alt}). Similarly, introducing moderate
 observation error (CV = 0.2) only slightly decreased the estimated prevalence
-of black-swan events (fig.~S\ref{fig:alt}), and the strength of the prior on
+of black-swan events (Fig.~S\ref{fig:alt}), and the strength of the prior on
 \(\nu\) had little influence on estimates of black-swan dynamics
-(fig.~S\ref{fig:alt-priors}). Finally, our simulation testing shows that, if
+(Fig.~S\ref{fig:alt-priors}). Finally, our simulation testing shows that, if
 anything, our models underpredict the true magnitude and probability of
 heavy-tailed events---especially given the time series have a median of only 26
-years in our analysis (figs~S\ref{fig:sim-nu},~S\ref{fig:sim-prob}).
+years in our analysis (Figs~S\ref{fig:sim-nu},~S\ref{fig:sim-prob}).
+
+% Our results show that weaker and stronger priors have little influence on our
+% conclusions about heavy-tailed dynamics (Fig.~\ref{fig:alt-priors}). When the
+% data are informative about tail behaviour (i.e. when there is strong evidence
+% of low \(\nu\) values, upper-right of Fig.~\ref{fig:alt-priors}), the prior has
+% little impact on the estimate of \(\nu\). When the data are less informative
+% about \(\nu\) (i.e.\ when there are no or few tail events and time series are
+% short or noisy), the prior can pull the estimate of \(\nu\) towards larger or
+% smaller values (Fig.~\ref{fig:alt-priors}). The vast majority of the
+% populations with Pr\((\nu < 10)\) in the base prior were not altered
+% qualitatively by this range of prior strength.
 
@@
 magnitude of process noise but not clearly related to population growth rate,
-density dependence, or maximum lifespan (Fig.~3b, fig.~S\ref{fig:correlates}).
+density dependence, or maximum lifespan (Fig.~3b, Fig.~S\ref{fig:correlates}).
 Longer time-series length was the strongest covariate of observing black-swan
@@
 event in populations with 60 time steps than in one with 30 time steps
-(fig.~S\ref{fig:correlates}).
+(Fig.~S\ref{fig:correlates}).
 
@@
 rather than upward (unexpectedly rapid abundance increases). Of the black-swan
-events with published explanations (table~S2), the majority involved
+events with published explanations (Table~S2), the majority involved
 a combination of multiple factors. For example, a synchronization of
@@
 caused a downward black-swan event for a water vole (\emph{Arvicola
-terrestris}) population\cite{saucy1994}. Other black swans were the result of
+terrestris}) population \citep{saucy1994}. Other black swans were the result of
 a sequence of extreme climate events on their own. For instance, severe winters
@@
 in grey heron (\emph{Ardea cinerea}) abundance in the United
-Kingdom\cite{stafford1971} (Fig.~4c). Our analysis finds that the last event
+Kingdom \citep{stafford1971} (Fig.~4c). Our analysis finds that the last event
 was a combination of two heavy-tailed events in a row and the population took
-three times longer to recover than predicted\cite{stafford1971}. Downward black
+three times longer to recover than predicted \citep{stafford1971}. Downward black
 swans were sometimes followed by upward black swans. For example, during
@@
 shag cormorants (\emph{Phalacrocorax aristotelis}) on the Farne Islands, United
-Kingdom declined suddenly following a red tide event in 1968\cite{potts1980}.
+Kingdom declined suddenly following a red tide event in 1968 \citep{potts1980}.
 This freed up quality nest sites for first-time breeders, productivity
 increased, and the population experienced a rapid upswing in
-abundance\cite{potts1980}.
+abundance \citep{potts1980}.
 
@@
 strong evidence of downwardly skewed process noise (Fig.~4b,
-fig.~S\ref{fig:skew-nu}). In contrast, populations that did not have heavy
+Fig.~S\ref{fig:skew-nu}). In contrast, populations that did not have heavy
 tails (median \(\nu \geq 70\)) had little evidence of skewed process noise
@@
 abundance) by \crashUnderRange-fold (interquartile range; Fig.~4c,
-fig.~S\ref{fig:skew-projections}). Thus our analysis provides strong evidence
-for downward-skewed heavy-tailed events in abundance time series of higher
-taxa, and ignoring these events will tend to underestimate the risk of
-population declines.
+Fig.~S\ref{fig:skew-projections}).
+
+% todo Thus our analysis provides strong evidence
+% for downward-skewed heavy-tailed events in abundance time series of higher
+% taxa, and ignoring these events will tend to underestimate the risk of
+% population declines.
+
+\section{Discussion}
 
@@
 different variances can generate a t distribution with heavy
-tails\cite{allen2001}. In ecology, these processes could arise through an
+tails \citep{allen2001}. In ecology, these processes could arise through an
 aggregation of populations across space, population
-diversity\cite{schindler2010}, or an intrinsic change in population variability
-through time\cite{carpenter2006}. Extrinsic ecological forces could also cause
-black-swan dynamics\cite{nunez2012} including extreme climate
-events\cite{meehl2004, katz2005, ipcc2012}, predation from (or competition
+diversity \citep{schindler2010}, or an intrinsic change in population variability
+through time \citep{carpenter2006}. Extrinsic ecological forces could also cause
+black-swan dynamics \citep{nunez2012} including extreme climate
+events \citep{meehl2004, katz2005, ipcc2012}, predation from (or competition
 with) other species experiencing black swans, or sharp changes in human
@@
 synchrony of multiple extrinsic forces could give rise to black-swan dynamics
-through synergistic interactions\cite{kirby2009} or a rare alignment of
-non-synergistic forces\cite{denny2009}.
+through synergistic interactions \citep{kirby2009} or a rare alignment of
+non-synergistic forces \citep{denny2009}.
+
+Given the multitude of reasons why a population modeler might expect heavy tail process noise, what should a population modeler do?
+
+Essentially any reason for why a population modeler has incorrectly specified
+the exact process generating the abundance times there is under question could
+generate have detailed process noise.
+
+We think modelers should routinely consider substituting a heavy tailed
+distribution, such as the student t-distribution, to represent the process
+noise in population dynamic modeling. Especially when making forecasts to make
+predictions about risk of specified bad events.
+
+But what heavy tailed distribution to pick?
+
+This is an open research question. However, we think the student t-distribution
+represents a simple solution in that it can be easily substituted for the
+normal distribution and if the degrees of freedom are fixed at a moderately
+heavy level it doesn't require the estimation of any extra parameters over the
+normal distribution.
+
+And if we use a student t-distribution than what degrees of freedom parameter should be you assume?
+
+The exact shape of the tails has been shown to be important in the field of
+dispersal biology -- has a big effect on dynamic \citep{kot1996}
+\citep{clark1999}
+\citep{kot1996}
+
+The degrees of
+freedom could be estimated from meta-analysis but even here using one of the
+best large data sets of animal abundance timeseries the timeseries are
+relatively short and the taxonomy is a biased sample of species of interest to humans  economic
+interest ...
+
+for example the majority of the blocks one offense that we observed were
+population crashes This could be due to basic biology --- animal birthrates are
+constrained whereas every animal can die --- on the other hand it could be due
+to the biased sample of species in the data set which represent species of
+concern to humans and we don't have much data for highly reproductive species
+that are not threatened or of economic interest
+
+If there is any reason for a model are to expect that external forces could act
+or that there is a mixture of distributions through spatial heterogeneity we
+think it is prudent to err on the side of caution and allow for heavy tailed
+process noise in forecasts
+
+The barring a better estimate we might suggest using a degrees of freedom parameter around 5
+
+this allows for a catastrophic event of an x fold decline of abundance that
+happens every Y years compared to a normal distribution allowing only for this
+event to happen every Z years.
+
+But what was once considered a one in X million event then becomes a one in Y
+million event
+
+A t-distribution compared to a normal distribution on the process error he
+essentially lets us be right most of the time while allowing are models to be
+very wrong occasionally this better matches our current intuition that
+ecological surprises can and do occur
+
+Our hypothesis is that our first priority is to allow for tails that are
+heavier than the normal distribution and approximately in line with those
+discovered in this paper and our second priority is to determine exactly how
+heavy a default distribution we should assume.
+
+Answers might come from simulation analyses exploring the implications of
+different tail shapes and also from fitting population models to known time
+series of mass mortality events \citep[e.g.][]{fey2015}
 
@@
 forecast probabilities to focus early detection efforts and develop disaster
-response plans\cite{nrc2007}. The presence of black swans also suggests that we
+response plans \citep{nrc2007}. The presence of black swans also suggests that we
 develop management policy that is robust to heavy tails and encourages
-\textit{general} resilience\cite{carpenter2012}. For instance, setting target
+\textit{general} resilience \citep{carpenter2012}. For instance, setting target
 population abundances far from critical limits will buffer against black-swan
-events\cite{caddy1996}, and maintaining genetic, phenotypic, and behavioural
+events \citep{caddy1996}, and maintaining genetic, phenotypic, and behavioural
 diversity may allow some components of populations to persist when others are
-affected by disease or extreme environmental forces\cite{schindler2010}.
+affected by disease or extreme environmental forces \citep{schindler2010}.
 Finally, surprising, or counterintuitive ecological dynamics offer a tremendous
 opportunity to learn about ecological systems, evaluate when models break down,
-and adjust future management policy\cite{doak2008, lindenmayer2010}.
+and adjust future management policy \citep{doak2008, lindenmayer2010}.
 
 Rare catastrophes can have a profound influence on population
-persistence\cite{mangel1994}. In recent decades, ecology has moved toward
+persistence \citep{mangel1994}. In recent decades, ecology has moved toward
 focusing on aspects of variance in addition to mean
-responses\cite{thompson2013}. Our results suggest that an added focus on
+responses \citep{thompson2013}. Our results suggest that an added focus on
 ecological extremes represents the next frontier, particularly in the face of
-increased climate extremes\cite{meehl2004, ipcc2012, thompson2013}. Financial
+increased climate extremes \citep{meehl2004, ipcc2012, thompson2013}. Financial
 analysts are concerned with the shape of financial return downward tails
@@
 
-\textbf{References}
+\section{Acknowledgements}
 
-\renewcommand{\section}[2]{}%
-
-\bibliographystyle{Science}\bibliography{jshort,ms}
-
-\textbf{Acknowledgements}: We thank J.W. Moore, A.O. Mooers, L.R. Gerber, J.D.
+We thank J.W. Moore, A.O. Mooers, L.R. Gerber, J.D.
 Yeakel, C. Minto and members of the Earth to Ocean Group for helpful
@@
 phylopic.org under Creative Commons licenses; sources are listed in the
-Supplementary Materials. Funding was provided by a Simon Fraser University
+Supporting Information. Funding was provided by a Simon Fraser University
 Graduate Fellowship (S.C.A.), the Natural Sciences and Engineering Research
@@
 
-\textbf{Supplementary Materials}\\
-Materials and Methods\\
-Supplementary References\\
-Tables S1 to S2\\
-Figs.~S1 to S8\\
+\section{References}
+
+\renewcommand{\section}[2]{}%
+
+\bibliographystyle{ecologyletters}\bibliography{jshort,ms}
+
+% \textbf{Supporting Information}\\
+% Materials and Methods\\
+% Supplementary References\\
+% Tables S1 to S2\\
+% Figs.~S1 to S8\\
 
@@
 
-\textbf{Figures}
+% \section{Figures}
 

### 2016-05-27 text/supp.tex
Remove paragraph from supplement - moved to main text

@@
 decreases.
-
-% todo Add this section to the discussion :
-There are a number of caveats when considering the generality of our results.
-The GPDD data represent a taxonomically and geographically biased sample of
-populations---the longer time series we focus on are dominated by commercially
-and recreationally important species and a disproportionate number of
-populations are located in the United Kingdom. Although we would expect to find
-qualitatively similar evidence for black swans in other large taxonomic or
-geographic samples of populations, the common forces driving those black swans
-would likely differ. Additionally, some apparent black-swan events could be
-recording mistakes, or conversely, some extreme observations may have been
-discarded or altered if they were erroneously suspected of being recording
-mistakes. Indeed, we discarded three of the populations that our method
-initially identified as heavy tailed because they turned out to be data-entry
-errors. Finally, the temporal scales of observation and population dynamics
-vary considerably across populations in the GPDD and these likely influence the
-detection of heavy tails. For example, if we make frequent observations
-relative to generation time (e.g.~for many large-bodied mammals) we will
-average across generations and perhaps miss black-swan events. Conversely, if
-we census populations infrequently relative to generation time (e.g.~many
-insects in the GPDD) the recorded data may average across extreme and
-less-extreme events and also dampen black-swan dynamics.
 

### 2016-05-27 text/ms.tex
Adjust keywords
Add back overview 2tto methods

@@
 Bayesian,
-black swans,
 die-offs,
 ecological surprises,
+Gompertz,
 heavy tails,
-Gompertz,
+leptokurtic,
 mass mortality,
@@
 \section{Methods}
+
+We fit population dynamics models in which the process noise is
+drawn from a Student-t distribution. By estimating the degrees of freedom
+parameter, \(\nu\), we can estimate the degree to which the process deviations
+have heavy tails and are therefore diagnostic of black-swan events (Fig.~1,
+fig.~S\ref{fig:priors}). Lower values of \(\nu\) result in heavy-tailed
+distributions.
+% For example, at \(\nu = 2\), an event
+% more than 4 standard deviations below the mean is expected
+For example, an event four standard deviations below the mean
+is expected to happen once every 35 years with a t distribution with \(\nu = 2\)
+compared to once every 32,000 years with a normal distribution
+% more than four standard deviations below the mean is almost 70,000 times higher
+% than the probability of drawing such a value from a normal distribution
+% (\(0.02\) vs.\ \(2.9\cdot10^{-7}\)).
+As \(\nu\) approaches infinity, the t
+distribution approaches the normal distribution (Fig.~1).
 

### 2016-05-27 text/ms.tex
Add caveats and modeling suggestion paragraphs

@@
 
-Given the multitude of reasons why a population modeler might expect heavy tail process noise, what should a population modeler do?
+There are a number of caveats when considering the generality of our results.
+The GPDD data represent a taxonomically and geographically biased sample of
+populations---the longer time series we focus on are dominated by commercially
+and recreationally important species and a disproportionate number of
+populations are located in the United Kingdom. Although we would expect to find
+qualitatively similar evidence for black swans in other large taxonomic or
+geographic samples of populations, the common forces driving those black swans
+would likely differ. Additionally, some apparent black-swan events could be
+recording mistakes, or conversely, some extreme observations may have been
+discarded or altered if they were erroneously suspected of being recording
+mistakes. Indeed, we discarded three of the populations that our method
+initially identified as heavy tailed because they turned out to be data-entry
+errors. Finally, the temporal scales of observation and population dynamics
+vary considerably across populations in the GPDD and these likely influence the
+detection of heavy tails. For example, if we make frequent observations
+relative to generation time (e.g.~for many large-bodied mammals) we will
+average across generations and perhaps miss black-swan events. Conversely, if
+we census populations infrequently relative to generation time (e.g.~many
+insects in the GPDD) the recorded data may average across extreme and
+less-extreme events and also dampen black-swan dynamics.
 
-Essentially any reason for why a population modeler has incorrectly specified
-the exact process generating the abundance times there is under question could
-generate have detailed process noise.
-
-We think modelers should routinely consider substituting a heavy tailed
-distribution, such as the student t-distribution, to represent the process
-noise in population dynamic modeling. Especially when making forecasts to make
-predictions about risk of specified bad events.
-
-But what heavy tailed distribution to pick?
-
-This is an open research question. However, we think the student t-distribution
-represents a simple solution in that it can be easily substituted for the
-normal distribution and if the degrees of freedom are fixed at a moderately
-heavy level it doesn't require the estimation of any extra parameters over the
-normal distribution.
-
-And if we use a student t-distribution than what degrees of freedom parameter should be you assume?
-
-The exact shape of the tails has been shown to be important in the field of
-dispersal biology -- has a big effect on dynamic \citep{kot1996}
-\citep{clark1999}
-\citep{kot1996}
-
-The degrees of
-freedom could be estimated from meta-analysis but even here using one of the
-best large data sets of animal abundance timeseries the timeseries are
-relatively short and the taxonomy is a biased sample of species of interest to humans  economic
-interest ...
-
-for example the majority of the blocks one offense that we observed were
-population crashes This could be due to basic biology --- animal birthrates are
-constrained whereas every animal can die --- on the other hand it could be due
-to the biased sample of species in the data set which represent species of
-concern to humans and we don't have much data for highly reproductive species
-that are not threatened or of economic interest
-
-If there is any reason for a model are to expect that external forces could act
-or that there is a mixture of distributions through spatial heterogeneity we
-think it is prudent to err on the side of caution and allow for heavy tailed
-process noise in forecasts
-
-The barring a better estimate we might suggest using a degrees of freedom parameter around 5
-
-this allows for a catastrophic event of an x fold decline of abundance that
-happens every Y years compared to a normal distribution allowing only for this
-event to happen every Z years.
-
-But what was once considered a one in X million event then becomes a one in Y
-million event
-
-A t-distribution compared to a normal distribution on the process error he
-essentially lets us be right most of the time while allowing are models to be
-very wrong occasionally this better matches our current intuition that
-ecological surprises can and do occur
-
-Our hypothesis is that our first priority is to allow for tails that are
-heavier than the normal distribution and approximately in line with those
-discovered in this paper and our second priority is to determine exactly how
-heavy a default distribution we should assume.
-
-Answers might come from simulation analyses exploring the implications of
-different tail shapes and also from fitting population models to known time
-series of mass mortality events \citep[e.g.][]{fey2015}
+We think modelers should routinely consider representing population dynamic
+process noise with a (possibly downward-skewed)
+heavy-tailed distribution,
+especially when making forecasts to evaluate risk.
+But what heavy-tailed distribution should one pick
+and how heavy tailed should it be?
+% nearly any reason for why a population modeler has incorrectly specified the exact and full process generating the abundance times could generate a heavy tailed process noise.
+This is an open research question.
+The exact shape of the tails has proven important
+in the field of dispersal biology \citep{kot1996, clark1999}.
+Answers for the field of population dynamics
+might come from simulation analyses exploring
+the implications of different tail shapes and
+from comparing the relative fit of heavy-tailed
+population models to timeseries
+with suspected mass mortality events
+\citep[e.g.~the t distribution compared to a mixture
+of two normal distributions;][]{ward2007}.
+In the meantime, we think the t distribution
+represents a simple solution.
+It can be easily substituted for the normal distribution
+and if the degrees of freedom parameter, $\nu$,
+is fixed
+it requires estimating no more parameters
+then the normal distribution.
+Barring a better estimate, we suggest $\nu = 4$ or $5$ and possibly a skewness parameter ($\gamma$) of $\sim 0.7$ (Fig.~4a).
+Above $\nu  =  10$ the t distribution closely resembles
+the normal distribution,
+and $\nu  = 4$ or $5$ allows for an event
+three standard deviations below the mean to happen
+once every 50--65 years instead of once every $\sim$ 740 years.
+Adopting heavy-tailed process noise follows the intuition
+that our models are usually reasonable but occasionally very wrong.
 

### 2016-05-27 text/ms.tex
Additions to main text

@@
 distribution approaches the normal distribution (Fig.~1).
+
+We picked the t distribution mainly for pragmatic reasons:
+it requires the estimation of only one additional
+parameter over the normal distribution and reverts to the normal distribution
+if strong evidence of heavy-tailed process deviations are not present.
+In addition, the t distribution has some mechanistic justification.
+The t distribution can be formed by a mixture of normal distributions
+with the same mean and different variances \citep{gelman2014}.
+This could represent the effect of unmodeled spatial heterogeneity,
+a nonstationary environmental process, or
+varying responses to temperature among individuals,
+for example.
 
@@
 
-There are a number of possible causes of black-swan events, including
+We systematically evaluated the presence of black-swan events in hundreds of
+animal populations. We find strong evidence for their occurrence. Black-swan
+population crashes are substantially more frequent than black-swan population
+increases and are usually driven by external events such as weather and
+disease. Therefore, ignoring the presence of black-swan events, and the
+predominance of crashes over increases, may underestimate extinction risk.
+
+There are many possible causes of black-swan events, including
 unmodelled intrinsic properties of populations or extrinsic forces acting on
@@
 
-We think modelers should routinely consider representing population dynamic
-process noise with a (possibly downward-skewed)
-heavy-tailed distribution,
+% The models we fit may have misspecified
+% the dynamics of the populations,
+% but this is precisely the point of our paper.
+% Black swans represent extreme deviations from our expectations
+% and our expectations are represented by the models we fit.
+% However, rarely, if ever, are our models perfect representations
+% of the underlying truth.
+% We might misunderstand the breeding biology of a species,
+% the weather can suddenly change,
+% disease can suddenly spread,
+% and species can unexpectedly interact.
+
+% todo the strongest correlates of finding blocks one events was the timeseries length
+% ...implications
+
+Given our results, we think modelers should routinely consider representing population dynamic
+process noise with a heavy-tailed and downward-skewed distribution,
 especially when making forecasts to evaluate risk.
@@
 of two normal distributions;][]{ward2007}.
-In the meantime, we think the t distribution
+In the meantime, we think the t (or the skew-t) distribution
 represents a simple solution.

### 2016-05-27 text/ms.tex
Edits

@@
 have heavy tails and are therefore diagnostic of black-swan events (Fig.~1,
-fig.~S\ref{fig:priors}). Lower values of \(\nu\) result in heavy-tailed
+Fig.~S\ref{fig:priors}). Lower values of \(\nu\) result in heavy-tailed
 distributions.
@@
 For example, an event four standard deviations below the mean
-is expected to happen once every 35 years with a t distribution with \(\nu = 2\)
+is expected to happen once every 35 years with a t distribution at \(\nu = 2\)
 compared to once every 32,000 years with a normal distribution
@@
 reliable data sources, and those without sufficient data for our models, and
-then interpolated some missing values (sensu Brook \textit{et
-al.} \citep{brook2006a}). Our interpolation affected only \(\sim\)
+then interpolated some missing values (sensu \citet{brook2006a}). Our interpolation affected only \(\sim\)
 \interpPointsPerc \% of the final data points (Table~S1) and none of the data
@@
 range of expected values for population growth and is similar to the default
-priors suggested by Gelman \emph{et al.} \citep{gelman2008d} for intercepts of
-regression models. Our Half-Cauchy prior on \(\sigma\) follows Gelman \emph{et
-al.} \citep{gelman2006c} and Gelman \emph{et al.} \citep{gelman2008d} and the
-specific scale parameter of \(2.5\) is based on our expected range of the value
-in nature from previous studies \citep{connors2014}. In our testing of a
-subsample of populations, our parameter estimates were not qualitatively
-changed by switching to a uniform prior on \(\sigma\), but the weakly
-informative Half-Cauchy prior sped up chain convergence.
+priors suggested by \citet{gelman2008d} for intercepts of
+regression models. Our Half-Cauchy prior on \(\sigma\) follows
+\citet{gelman2006c} and \citet{gelman2008d} and the specific scale parameter of
+\(2.5\) is based on our expected range of the value in nature from previous
+studies \citep{connors2014}. In our testing of a subsample of populations, our
+parameter estimates were not qualitatively changed by switching to a uniform
+prior on \(\sigma\), but the weakly informative Half-Cauchy prior sped up chain
+convergence.
 
@@
 of \(0.01\) truncated at values above two---a slightly less informative prior
-than suggested by Fernandez and Steel \citep{fernandez1998}. We truncated the
+than suggested by \citet{fernandez1998}. We truncated the
 distribution at \(2\), since at \(\nu < 2\) the variance of the t distribution
@@
 and warm up period and sampled from the model again. These thresholds are in
-excess of the minimums recommended by Gelman \emph{et al.} \citep{gelman2006a} of
+excess of the minimums recommended by \citet{gelman2006a} of
 \(\hat{R} < 1.1\) and \(n_\mathrm{eff} > 100\) for reliable point estimates and
@@
 Gibbs or Metropolis--Hastings algorithms \citep{stan-manual2015}.
-
-Specifically, we fit population dynamics models in which the process noise is
-drawn from a Student-t distribution. By estimating the degrees of freedom
-parameter, \(\nu\), we can estimate the degree to which the process deviations
-have heavy tails and are therefore diagnostic of black-swan events (Fig.~1,
-Fig.~S\ref{fig:priors}). Lower values of \(\nu\) result in heavy-tailed
-distributions. For example, at \(\nu = 2\), the probability of drawing a value
-more than five standard deviations below the mean is almost 70,000 times higher
-than the probability of drawing such a value from a normal distribution
-(\(0.02\) vs.\ \(2.9\cdot10^{-7}\)). As \(\nu\) approaches infinity, the
-distribution approaches the normal distribution (Fig.~1).
 
@@
 populations as heavy tailed. We furthermore considered a Uniform(\(0, 1\))
-prior on \(1/\nu\) as used in Gelman \emph{et al.} \citep{gelman2014}, but this
+prior on \(1/\nu\) as used in \citet{gelman2014}, but this
 prior puts considerably more weight on lower (heavy-tailed) values of \(\nu\)
@@
 different variances can generate a t distribution with heavy
-tails \citep{allen2001}. In ecology, these processes could arise through an
+tails \citep{gelman2014}. In ecology, these processes could arise through an
 aggregation of populations across space, population

### 2016-06-23 text/ms.tex
Add my edits to manuscript before Nick/Trevor edits

@@
 
-Sean C. Anderson\textsuperscript{1,2*}, Trevor A.
-Branch\textsuperscript{3}, Andrew B. Cooper\textsuperscript{2}, Nicholas K.
-Dulvy\textsuperscript{1}
+Sean C. Anderson\textsuperscript{1*}, Trevor A.
+Branch\textsuperscript{2}, Andrew B. Cooper\textsuperscript{3}, Nicholas K.
+Dulvy\textsuperscript{4}
 
 \textsuperscript{1}Earth to Ocean Research Group, Department of Biological
-Sciences, Simon Fraser University, Burnaby BC, V5A 1S6, Canada
+Sciences, Simon Fraser University, Burnaby BC, V5A 1S6, Canada, e-mail:
+sandrsn@uw.edu
 
-\textsuperscript{2}School of Resource and Environmental Management, Simon
-Fraser University, Burnaby, BC, V5A 1S6, Canada
+\textsuperscript{2}School of Aquatic and Fishery Sciences, Box 355020,
+University of Washington, Seattle, WA 98195, USA, e-mail:
+tbranch@uw.edu
 
-\textsuperscript{3}School of Aquatic and Fishery Sciences, Box 355020,
-University of Washington, Seattle, WA 98195, USA
+\textsuperscript{3}School of Resource and Environmental Management, Simon
+Fraser University, Burnaby, BC, V5A 1S6, Canada, e-mail:
+andrew\_cooper@sfu.ca
 
-\textsuperscript{*}Corresponding author: Sean C. Anderson; E-mail:
-sandrsn@uw.edu; Present address: School of Aquatic and Fishery Sciences, Box
+\textsuperscript{4}Earth to Ocean Research Group, Department of Biological
+Sciences, Simon Fraser University, Burnaby BC, V5A 1S6, Canada, e-mail:
+nick\_dulvy@sfu.ca
+
+\textsuperscript{*}Corresponding author; present address:
+School of Aquatic and Fishery Sciences, Box
 355020, University of Washington, Seattle, WA 98195, USA
@@
 risk,
-Student t
+Student-t distribution
 
-Words in abstract: TOdO
+Words in abstract: 142
 
-Words in main text: todo
+Words in main text: 4820
 
@@
 
-Number of figures: 4
+Number of figures/tables: 4/0
 
-Number of tables: 0
+Statement of authorship:
+S.C.A. and T.A.B. conceived the project;
+S.C.A., T.A.B., A.B.C., and N.K.D. designed the study; S.C.A. analyzed the
+data and wrote the paper with input from all authors.
 
@@
 Black-swan events occur most frequently for birds (\birdPH \%), mammals
-(\mammalsPH \%), and insects (\insectsPH \%), and manifest primarily as
-population die-offs and crashes (\percBSDown \%) driven by extrinsic
-perturbations from climate variability. Thus, ignoring black-swan events will
-underestimate extinction risk. Our results demonstrate the importance of both
+(\mammalsPH \%), and insects (\insectsPH \%),
+and manifest primarily as
+population die-offs and crashes (\percBSDown \%) rather than unexpected increases
+and tend to be driven by extrinsic perturbations
+such as climate, severe winters, or disease.
+Thus, ignoring black-swan events will
+underestimate extinction risk.
+% Ignoring black-swan events leads to a \crashUnderMedian-fold (50\% interval \crashUnderRange) underestimate in the magnitude of population crashes.
+Our results demonstrate the importance of both
 modeling heavy-tailed downward events in populations, and developing
-conservation strategies that are robust to future increases in climate
-extremes.
+conservation strategies that are robust to such extremes.
 
@@
 assumes that deviations from model predictions can be represented by a normal
-distribution \citep[typically on a log scale; e.g.][]{brook2006a, knape2012}.
+distribution \citep[on a log scale; e.g.][]{brook2006a, knape2012}.
 If black swans occur, though, a normal distribution would under-estimate the
@@
 Previous comparative analyses fitting
-heavy-tailed distributions to timeseries
+heavy-tailed distributions
 (distributions with higher probabilities of extreme events
 than the normal distribution)
+to timeseries
 have not accounted for an underlying population dynamics model
@@
 \citep[e.g.][]{young1994, gerber2001, reed2003, fey2015}.
-But this approach gives no indication of relative frequency
+But this approach gives no indication of relative event frequency
 and again does not distinguish events from
@@
 the GEV requires longer timeseries then typically available
-for populations and is not simply integrated
+for wild populations and is not simply integrated
 into a population dynamics model.
@@
 verify known causes. To accomplish this, we develop a framework for identifying
-heavy-tailed (black-swan) process noise in population dynamics, i.e.~whether
-the largest stochastic jumps in log abundance from one time step to the next
+heavy-tailed (black-swan) process noise in population dynamics.
+We test if the largest stochastic jumps in log abundance from one time step to the next
 are more extreme than typically seen with a normal distribution. Our framework
@@
 parameter, \(\nu\), we can estimate the degree to which the process deviations
-have heavy tails and are therefore diagnostic of black-swan events (Fig.~1,
-Fig.~S\ref{fig:priors}). Lower values of \(\nu\) result in heavy-tailed
-distributions.
+have heavy tails and are therefore diagnostic of black-swan events (Fig.~1).
+Lower values of \(\nu\) result in heavy-tailed distributions.
 % For example, at \(\nu = 2\), an event
@@
 
-We picked the t distribution mainly for pragmatic reasons:
-it requires the estimation of only one additional
+We picked the t distribution for pragmatic and mechanistic reasons.
+Pragmatically, the t distribution requires the estimation of only one additional
 parameter over the normal distribution and reverts to the normal distribution
 if strong evidence of heavy-tailed process deviations are not present.
-In addition, the t distribution has some mechanistic justification.
-The t distribution can be formed by a mixture of normal distributions
+Mechanistically, the t distribution can be formed from a mixture of normal distributions
 with the same mean and different variances \citep{gelman2014}.
 This could represent the effect of unmodeled spatial heterogeneity,
-a nonstationary environmental process, or
-varying responses to temperature among individuals,
-for example.
+nonstationary environmental processes, or
+varying responses to temperature among individuals, for example.
 
@@
 reliable data sources, and those without sufficient data for our models, and
-then interpolated some missing values (sensu \citet{brook2006a}). Our interpolation affected only \(\sim\)
+then interpolated remaining missing values (sensu \citet{brook2006a}).
+Our interpolation affected only \(\sim\)
 \interpPointsPerc \% of the final data points (Table~S1) and none of the data
@@
 \end{align}
-The growth parameter \(\lambda\) represents the expected growth rate if \(N_{t}
+The parameter \(\lambda\) represents the expected growth rate if \(N_{t}
 = 1\). The model is density independent if \(b = 1\), maximally density
@@
 
-For the Gompertz model, our weakly-informative priors (Fig.~\ref{fig:priors}) were:
+For the Gompertz model, our weakly-informative priors (Fig.~S\ref{fig:priors}) were:
 \begin{align}
@@
 \(2.5\) is based on our expected range of the value in nature from previous
-studies \citep{connors2014}. In our testing of a subsample of populations, our
+studies \citep[e.g.][]{connors2014}. In our testing of a subsample of populations, our
 parameter estimates were not qualitatively changed by switching to a uniform
@@
 chose the probability that \(\nu < 10\), Pr(\(\nu < 10\)), to define the
-probability of heavy-tailed dynamics. When categorizing a population as heavy
+probability of heavy-tailed (black-swan) dynamics. When categorizing a population as heavy
 tailed, we applied a 0.5 probability threshold. In the scenario where the data
@@
 greatly exceeded. We continued this procedure up to 8,000 iterations (16,000
-total samples) by which all chains were deemed to have sufficiently converged.
+total samples) by which all chains had sufficiently converged.
 The No-U-Turn Hamiltonian Markov chain Monte Carlo sampler in Stan generally
@@
 we used Truncated-Exponential(0.005, min.\ = 2) and for a stronger prior we
-used Truncated-Exponential(0.02, min.\ = 2) (Fig.~\ref{fig:priors}c). Note that
+used Truncated-Exponential(0.02, min.\ = 2) (Fig.~S\ref{fig:priors}c). Note that
 the base and weaker priors are relatively flat within the region of \(\nu <
@@
 \subsection{Alternative population models} We fit alternative population models to
-test if four key phenomena systematically changed our conclusions. The range of
-percentages of black swans by taxonomic class cited in the abstract are based
-on lower and upper limits across our main Gompertz model and these four
-alternative models.
+test if four key phenomena systematically changed our conclusions.
 
@@
 \(\sigma_\mathrm{obs}\) to \(0.2\), which represents a typical value used in
-simulation analyses \citep{valpine2002, thorson2014b}.
+simulation analyses \citep[e.g.][]{valpine2002, thorson2014b}.
 
@@
 
+\textit{TODO: Should the following 2 paragraphs go in the results?}
 We found that we could consistently and precisely recover median posterior
 estimates of \(\nu\) near the true value of \(\nu\) with large samples (\(\ge
-200\)) (Fig.~\ref{fig:sim-nu} upper panels). At smaller samples we could
+200\)) (Fig.~S\ref{fig:sim-nu} upper panels). At smaller samples we could
 usually qualitatively distinguish heavy from not-heavy tails, but the model
@@
 sample sizes, the model tended to overestimate how large the scale parameter
-was (Fig.~\ref{fig:sim-nu} lower panels).
+was (Fig.~S\ref{fig:sim-nu} lower panels).
 
@@
 the true value of \(\nu\) when the process noise was chosen so that appropriate
-tail events were present (Fig.~\ref{fig:sim-prob} upper panels). The addition
+tail events were present (Fig.~S\ref{fig:sim-prob} upper panels). The addition
 of observation error caused the model to tend to underestimate the degree of
 heavy-tailedness. Fitting a model with correctly specified observation error
-did not make substantial improvements to model bias (Fig.~\ref{fig:sim-prob}).
+did not make substantial improvements to model bias (Fig.~S\ref{fig:sim-prob}).
 
@@
   + \alpha^\mathrm{species}_{l[i]}
-  + X_i \beta),
+  + \bm{X_i} \bm{\beta}),
   \: \text{for } i = 1, \dots, 617\\
@@
 order \(k\), and species \(l\); \(\phi_\mathrm{disp}\) represents the
-dispersion parameter; and \(X_i\) represents a vector of predictors (such as
-lifespan) for population \(i\) with associated \(\beta\) coefficients. The
+dispersion parameter; and \(\bm{X_i}\) represents a vector of predictors (such as
+lifespan) for population \(i\) with associated \(\bm{\beta}\) coefficients. The
 intercepts are allowed to vary from the overall intercept \(\alpha\) by
@@
 We log transformed \(\sigma\), time-series length, and lifespan to match the
-way they are visually represented in Fig.~\ref{fig:correlates} and to make the
+way they are visually represented in Fig.~S\ref{fig:correlates} and to make the
 relationship approximately linear on the logit-transformed response scale. All
@@
 visually and with the same criteria as before (\(\widehat{R} < 1.05\) and
-\(n_\mathrm{eff} >200\) for all parameters). We further checked our model
+\(n_\mathrm{eff} >199\) for all parameters). Furthermore, we checked our model
 against a similar model with the glmmADMB \textsf{R} package \citep{glmmadmb}
@@
 
-\subsection{Skewed Student-t forecasts} To formally evaluate the apparent skewness
+\subsection{Skewed Student-t forecasts} To evaluate the apparent skewness
 of heavy-tailed process noise we fit Gompertz models with skewed Student-t
-distributed process noise \citep{fernandez1998}. The distribution adds one
+distributed process noise \citep{fernandez1998}. This distribution adds one
 parameter to the Student-t distribution, \(\gamma\), which controls the
@@
 parameter of each model at three levels of evidence for heavy-tails:
-\emph{heavy tailed} median \(\nu < 10\); \emph{slightly heavy tailed} \(10 \le
-\mathrm{median}\, \nu \le 70\), and \emph{normal tails}, median \(\nu > 70\)
+\emph{heavy tailed}, median \(\nu < 10\); \emph{slightly heavy tailed}, \(10 \le
+\mathrm{median}\, \nu \le 70\); and \emph{normal tailed}, median \(\nu > 70\)
 (Fig.~4).
@@
 
-% todo Thus our analysis provides strong evidence
-% for downward-skewed heavy-tailed events in abundance time series of higher
-% taxa, and ignoring these events will tend to underestimate the risk of
-% population declines.
-
 \section{Discussion}
@@
 animal populations. We find strong evidence for their occurrence. Black-swan
-population crashes are substantially more frequent than black-swan population
+population crashes are substantially more frequent
+than black-swan population
 increases and are usually driven by external events such as weather and
-disease. Therefore, ignoring the presence of black-swan events, and the
-predominance of crashes over increases, may underestimate extinction risk.
+disease.
+Thus our analysis provides strong evidence
+for downward-skewed heavy-tailed events in abundance time series of higher
+taxa, and ignoring these events will tend to underestimate the risk of
+population declines.
+\textit{TODO: do we want to have a recap paragraph at the beginning of the discussion now?}
+
+% than black-swan population
+% increases and are usually driven by external events such as weather and
+% disease. Therefore, ignoring the presence of black-swan events, and the
+% predominance of crashes over increases, may underestimate extinction risk.
+% TODO: add numbers, add skew add projections add length and Sigma
 
@@
 events \citep{meehl2004, katz2005, ipcc2012}, predation from (or competition
-with) other species experiencing black swans, or sharp changes in human
-pressures such hunting, fishing, or habitat destruction. Alternatively, the
+with) other species experiencing black-swan events, or sharp changes in human
+pressures such as fishing or habitat destruction. Alternatively, the
 synchrony of multiple extrinsic forces could give rise to black-swan dynamics
@@
 Given our results, we think modelers should routinely consider representing population dynamic
-process noise with a heavy-tailed and downward-skewed distribution,
+process noise with a heavy-tailed, and possibly downward-skewed, distribution,
 especially when making forecasts to evaluate risk.
@@
 then the normal distribution.
-Barring a better estimate, we suggest $\nu = 4$ or $5$ and possibly a skewness parameter ($\gamma$) of $\sim 0.7$ (Fig.~4a).
+Barring a better estimate,
+we suggest $\nu = 4$ or $5$ and possibly a
+skewness parameter ($\gamma$) of $\sim 0.7$ (Fig.~4a)
+if a modeler wishes to allow for downward heavy-tailed events.
+These are midrange values from the populations identified as heavy tailed here.
 Above $\nu  =  10$ the t distribution closely resembles
@@
 
-We thank J.W. Moore, A.O. Mooers, L.R. Gerber, J.D.
-Yeakel, C. Minto and members of the Earth to Ocean Group for helpful
-discussions and comments. We are grateful to the contributors and maintainers
-of the Global Population Dynamics Database and to Compute Canada's WestGrid
-high-performance computing resources. Silhouette images were obtained from
-phylopic.org under Creative Commons licenses; sources are listed in the
-Supporting Information. Funding was provided by a Simon Fraser University
-Graduate Fellowship (S.C.A.), the Natural Sciences and Engineering Research
-Council of Canada (N.K.D., A.B.C.), the Canada Research Chairs Program
+We thank J.W. Moore, A.O. Mooers, L.R. Gerber, J.D. Yeakel, C. Minto and
+members of the Earth to Ocean Group for helpful discussions and comments. We
+are grateful to the contributors and maintainers of the Global Population
+Dynamics Database and to Compute Canada's WestGrid high-performance computing
+resources. Silhouette images were obtained from phylopic.org under Creative
+Commons licenses; sources are listed in the Supporting Information. Funding was
+provided by a Simon Fraser University Graduate Fellowship and David H. Smith
+Conservation Research Fellowship (S.C.A.), the Natural Sciences and Engineering
+Research Council of Canada (N.K.D., A.B.C.), the Canada Research Chairs Program
 (N.K.D.).
@@
 % Figs.~S1 to S8\\
-
-%\textbf{Author Contributions}: S.C.A. and T.A.B. conceived the project;
-%S.C.A., T.A.B., A.B.C., and N.K.D. designed the study; S.C.A. analyzed the
-%data and wrote the paper with input from all authors.
 
@@
 from Gompertz models with lognormal and skew-t process noise, respectively.
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: "anderson-etal-blackswan-timeseries"
+%%% End:

### 2016-06-23 text/ms.tex
Incorporate Nick and Trevor 's comments
Also fix formatting of captions

@@
 risk,
-Student-t distribution
+Student-$t$ distribution
 
@@
 (\mammalsPH \%), and insects (\insectsPH \%),
-and manifest primarily as
-population die-offs and crashes (\percBSDown \%) rather than unexpected increases
+manifest primarily as
+population die-offs and crashes (\percBSDown \%) rather than unexpected increases,
 and tend to be driven by extrinsic perturbations
@@
 modeling heavy-tailed downward events in populations, and developing
-conservation strategies that are robust to such extremes.
+conservation strategies that are robust to ecological surprises.
 
@@
 If black swans occur, though, a normal distribution would under-estimate the
-probability of extreme events occurring.
+probability of extreme events occurring \citep{taleb2007}.
 
-There are many examples of black-swan events in populations
-but we lack a large-scale comparative analysis
-with a method that allows for
-population dynamics
-(e.g.\ per capita productivity, density dependence, and typical stochasticity).
+Whereas there are many reports of black-swan events,
+only a flexible comparative approach consistently applied 
+to a large number of timeseries can yield insights into the frequency, strength,
+and correlates of black-swan events. 
+We are unaware of such a comparative analysis.
 Previous comparative analyses fitting
@@
 Alternatively, most examples of population die offs come from identifying
-jumps in abundance that exceed a threshold of decline
+sudden changes in abundance that exceed a threshold of decline
 \citep[e.g.][]{young1994, gerber2001, reed2003, fey2015}.
@@
 \citep[e.g.\ sedimentation rates or wind speeds;][]{gaines1993, denny2009},
-because the GEV requires maxima per time block
+the GEV requires maxima per time block
 or peaks beyond a chosen threshold,
-the GEV requires longer timeseries then typically available
-for wild populations and is not simply integrated
-into a population dynamics model.
+and hence requires longer timeseries then typically available
+for wild populations.
+Furthermore, the GEV approach cannot be easily integrated into population dynamics models.
 
@@
 We fit population dynamics models in which the process noise is
-drawn from a Student-t distribution. By estimating the degrees of freedom
+drawn from a Student-$t$ distribution. By estimating the degrees of freedom
 parameter, \(\nu\), we can estimate the degree to which the process deviations
@@
 For example, an event four standard deviations below the mean
-is expected to happen once every 35 years with a t distribution at \(\nu = 2\)
+is expected to happen once every 35 years with a $t$ distribution at \(\nu = 2\)
 compared to once every 32,000 years with a normal distribution
@@
 
-We picked the t distribution for pragmatic and mechanistic reasons.
-Pragmatically, the t distribution requires the estimation of only one additional
+We chose the $t$ distribution for a number of reasons. 
+Pragmatically, the $t$ distribution requires the estimation of only one additional
 parameter over the normal distribution and reverts to the normal distribution
 if strong evidence of heavy-tailed process deviations are not present.
-Mechanistically, the t distribution can be formed from a mixture of normal distributions
-with the same mean and different variances \citep{gelman2014}.
+Furthermore, the $t$ distribution arises from a mixture of normal distributions
+with the same mean and different variances \citep{gelman2014},
+which mimics the underlying basis for black-swan events.
+There are a wide range of mechanistic processes 
+that could give rise to extreme events,
+such as spatial heterogeneity, nonstationary or heavy-tailed environmental processes, 
+or varying responses to temperature among individuals. 
+While evaluating these various hypotheses 
+would be important for diagnosing the underlying causes 
+of extreme events and to develop predictive models, 
+it would be impossible to tease out each of these 
+underlying causes from population trend data alone in a comparative manner. 
+Instead we seek a simple flexible phenomenological model 
+that can be applied across the widest possible 
+range of population time series to provide a 
+comparative overview of the prevalence 
+of black-swan events.
+
 This could represent the effect of unmodeled spatial heterogeneity,
@@
 the process noise \(\epsilon_{t}\) is modelled as normally distributed, but we
-allow it to be drawn from a Student-t distribution with scale parameter
+allow it to be drawn from a Student-$t$ distribution with scale parameter
 \(\sigma\) and degrees of freedom \(\nu\). We can then estimate the degree to
@@
 priors suggested by \citet{gelman2008d} for intercepts of
-regression models. Our Half-Cauchy prior on \(\sigma\) follows
+regression models. Our half-Cauchy prior on \(\sigma\) follows
 \citet{gelman2006c} and \citet{gelman2008d} and the specific scale parameter of
@@
 parameter estimates were not qualitatively changed by switching to a uniform
-prior on \(\sigma\), but the weakly informative Half-Cauchy prior sped up chain
+prior on \(\sigma\), but the weakly informative half-Cauchy prior sped up chain
 convergence.
@@
 than suggested by \citet{fernandez1998}. We truncated the
-distribution at \(2\), since at \(\nu < 2\) the variance of the t distribution
+distribution at \(2\), since at \(\nu < 2\) the variance of the $t$ distribution
 is undefined. This prior gives only a \(7.7\)\% probability that \(\nu < 10\)
-but constrains the sampling sufficiently to avoid drifting towards infinity. In
-any case, for \(\nu > 20\) the t distribution is almost indistinguishable from
-the normal distribution (Fig.~1). Based on the shape of the t distribution, we
+but constrains the sampling sufficiently to avoid \(\nu\) drifting towards infinity. In
+any case, for \(\nu > 20\) the $t$ distribution is almost indistinguishable from
+the normal distribution (Fig.~1). Based on the shape of the $t$ distribution, we
 chose the probability that \(\nu < 10\), Pr(\(\nu < 10\)), to define the
@@
 \(\nu\) will approximately match the prior (prior median \(= 71\), mean \(=
-102\)) and a metric of Pr\((\nu < 10) > 0.5\) is unlikely to flag the
+102\)) and a metric of Pr\((\nu < 10) > 0.5\) will not flag the
 population as heavy tailed.
@@
 20\), which is the region we are mostly concerned about when categorizing
-populations as heavy tailed. We furthermore considered a Uniform(\(0, 1\))
+populations as heavy tailed. We also considered using a Uniform(\(0, 1\))
 prior on \(1/\nu\) as used in \citet{gelman2014}, but this
 prior puts considerably more weight on lower (heavy-tailed) values of \(\nu\)
-than the truncated exponential priors considered here (not shown).
+than the truncated exponential priors considered here,
+and the results are therefore not shown.
 
@@
 \end{align}
-where \(U\) represents the unobserved state vector, \(\sigma_\mathrm{obs}\)
+where \(U\) represents the served state vector, \(\sigma_\mathrm{obs}\)
 represents the standard deviation of observation error (on a log scale), and
@@
 model assumes a linear decrease in population growth rate with increases in
-\textit{log} abundance.
+\textit{ln} abundance.
 
@@
 We performed two types of simulation testing.
-First, we tested how easily the Student-t distribution \(\nu\) parameter could
+First, we tested how easily the Student-$t$ distribution \(\nu\) parameter could
 be recovered given different true values of \(\nu\) and different sample sizes.
@@
 heavy-tailedness. Fitting a model with correctly specified observation error
-did not make substantial improvements to model bias (Fig.~S\ref{fig:sim-prob}).
+did not substantially reduce model bias (Fig.~S\ref{fig:sim-prob}).
 
@@
 distributed process noise \citep{fernandez1998}. This distribution adds one
-parameter to the Student-t distribution, \(\gamma\), which controls the
+parameter to the Student-$t$ distribution, \(\gamma\), which controls the
 skewness. The distribution is symmetrical if \(\gamma = 1\), left skewed if \(0
@@
 We found evidence that black-swan dynamics are infrequent but highly
-influential: they occur most frequently for birds (7\%), mammals (5\%), and
-insects (3\%) and almost never in fishes for this dataset (Fig.~2). Black-swan
+influential: in our dataset they were detected most 
+frequently for birds (7\%), mammals (5\%), and
+insects (3\%) but almost never in fishes (Fig.~2). Black-swan
 events were taxonomically widespread, occurring in 38\% of taxonomic orders.
@@
 population models, observation error, and choice of Bayesian priors. Our
-conclusions were not systematically altered when we included an autocorrelation
-structure in the residuals, modelled population growth rates with or without
+conclusions were not systematically altered when we included autocorrelation
+in the residuals, modelled population growth rates with or without
 density dependence, or modelled the population dynamics as Ricker-logistic
@@
 anything, our models underpredict the true magnitude and probability of
-heavy-tailed events---especially given the time series have a median of only 26
+heavy-tailed events---especially given that the time series are relatively short,
+with a median of only 26
 years in our analysis (Figs~S\ref{fig:sim-nu},~S\ref{fig:sim-prob}).
@@
 in 1929, 1940--1942, and 1962--1963 were associated with black-swan downswings
-in grey heron (\emph{Ardea cinerea}) abundance in the United
+in grey herons (\emph{Ardea cinerea}) in the United
 Kingdom \citep{stafford1971} (Fig.~4c). Our analysis finds that the last event
@@
 We systematically evaluated the presence of black-swan events in hundreds of
-animal populations. We find strong evidence for their occurrence. Black-swan
+animal populations in a consistent manner. 
+We find strong evidence for their occurrence in animal populations. Black-swan
 population crashes are substantially more frequent
@@
 population declines.
-\textit{TODO: do we want to have a recap paragraph at the beginning of the discussion now?}
+Next we consider 
+(1) the possible mechanisms underlying black-swan events, 
+(2) caveats to our findings, 
+(3) the consequences of our findings for making risk forecasts, and 
+(4) for managing natural resources.
 
@@
 underlying mixture of processes, since a mixture of normal distributions with
-different variances can generate a t distribution with heavy
+different variances can generate a $t$ distribution with heavy
 tails \citep{gelman2014}. In ecology, these processes could arise through an
@@
 would likely differ. Additionally, some apparent black-swan events could be
-recording mistakes, or conversely, some extreme observations may have been
+recording mistakes, although conversely, some extreme observations may have been
 discarded or altered if they were erroneously suspected of being recording
@@
 
-Given our results, we think modelers should routinely consider representing population dynamic
+Given our results, we suggest modelers should routinely consider representing population dynamic
 process noise with a heavy-tailed, and possibly downward-skewed, distribution,
@@
 with suspected mass mortality events
-\citep[e.g.~the t distribution compared to a mixture
+\citep[e.g.~the $t$ distribution compared to a mixture
 of two normal distributions;][]{ward2007}.
-In the meantime, we think the t (or the skew-t) distribution
-represents a simple solution.
-It can be easily substituted for the normal distribution
-and if the degrees of freedom parameter, $\nu$,
+Nevertheless, our proposed $t$ (or the skew-$t$) distribution
+is a simple and elegant solution that
+can be easily substituted for the normal distribution
+Furthermore, if the degrees of freedom parameter, $\nu$,
 is fixed
 it requires estimating no more parameters
-then the normal distribution.
+than the normal distribution.
 Barring a better estimate,
-we suggest $\nu = 4$ or $5$ and possibly a
-skewness parameter ($\gamma$) of $\sim 0.7$ (Fig.~4a)
+we suggest using $\nu = 4$; and
+a skewness parameter ($\gamma$) of $\sim 0.7$ (Fig.~4a)
 if a modeler wishes to allow for downward heavy-tailed events.
 These are midrange values from the populations identified as heavy tailed here.
-Above $\nu  =  10$ the t distribution closely resembles
+Above $\nu  =  10$ the $t$ distribution closely resembles
 the normal distribution,
-and $\nu  = 4$ or $5$ allows for an event
+and $\nu  = 4$ allows for an event
 three standard deviations below the mean to happen
-once every 50--65 years instead of once every $\sim$ 740 years.
-Adopting heavy-tailed process noise follows the intuition
+once every 50 years instead of once every 740 years.
+Adopting heavy-tailed process noise by default follows the intuition
 that our models are usually reasonable but occasionally very wrong.
 
-In light of our findings, we suggest that ecological resource management can
+In light of our findings, we suggest that natural resource management can
 learn from disciplines that focus on heavy tails. For example, earthquake
@@
 forecast probabilities to focus early detection efforts and develop disaster
-response plans \citep{nrc2007}. The presence of black swans also suggests that we
+response plans \citep{nrc2007}. The presence of ecological black swans also suggests that we
 develop management policy that is robust to heavy tails and encourages
-\textit{general} resilience \citep{carpenter2012}. For instance, setting target
-population abundances far from critical limits will buffer against black-swan
+general resilience \citep{carpenter2012}. For instance, setting target
+population abundances far back from critical limits will buffer against black-swan
 events \citep{caddy1996}, and maintaining genetic, phenotypic, and behavioural
@@
 increased climate extremes \citep{meehl2004, ipcc2012, thompson2013}. Financial
-analysts are concerned with the shape of financial return downward tails
-because these directly impact estimates of risk---the probability of a specific
-magnitude of undesired event occurring. Similarly, ecologists should focus more
+analysts are concerned with the shape of downward tails in financial returns 
+because these directly impact estimates of risk---the probability of a
+market crash occurring.
+Similarly, ecologists should focus more
 on estimating and predicting downward tails of population abundance, since
@@
 
-\textbf{Fig.~1: Illustration of population dynamic models that allow for heavy
-tails.} (\textbf{A} to \textbf{B}) Probability density for the Student-t
+Fig.~1: Illustration of population dynamic models that allow for heavy
+tails. (a to b) Probability density for the Student-t
 distribution with scale parameter of 1 and different values of \(\nu\). Small
 values of \(\nu\) create heavy tails while as \(\nu\) approaches infinity the
-distribution approaches the normal distribution. (\textbf{C} to \textbf{E})
+distribution approaches the normal distribution. (c to e)
 Simulated population dynamics from a Gompertz model with process noise drawn
-from Student-t distributions with three values of \(\nu\). Coloured dots in
-panels \textbf{C} and \textbf{D} represent jumps with less than a 1 in 1000
-chance of occurring in a normal distribution. (\textbf{F} to \textbf{H})
-Estimates of \(\nu\) from models fit to the times series in \textbf{C} to
-\textbf{E}. Shown are posterior samples (histograms), median and interquartile
+from Student-$t$ distributions with three values of \(\nu\). Coloured dots in
+panels c and d represent jumps with less than a 1 in 1000
+chance of occurring in a normal distribution. (f to h)
+Estimates of \(\nu\) from models fit to the times series in c to
+e. Shown are posterior samples (histograms), median and interquartile
 range of the posterior (IQR, dots and line segments), and the exponential prior
-on \(\nu\) (dashed lines). Colour shading behind \textbf{F} to \textbf{H}
+on \(\nu\) (dashed lines). Colour shading behind f to h
 illustrates the approximate region of heavy tails.
@@
 
-\textbf{Fig.~2: Estimates of population dynamic heavy-tailedness for
-populations of birds, mammals, insects, and fishes.} (\textbf{A} to \textbf{D})
+Fig.~2: Estimates of population dynamic heavy-tailedness for
+populations of birds, mammals, insects, and fishes. (a to d)
 Small values of \(\nu\) suggest heavy-tailed black-swan dynamics. Vertical
@@
 mean Pr(\(\nu < 10\)). Taxonomic orders with three or fewer populations in
-\textbf{A} are omitted for space. Red to yellow points highlight populations
+panel a are omitted for space. Red to yellow points highlight populations
 with a high to moderately high probability of heavy-tailed dynamics.
@@
 
-\textbf{Fig.~3: Standardized probabilities and covariates of heavy-tailed
-dynamics.} (\textbf{A}) Taxonomic-order-level posterior densities of Pr(\(\nu
+Fig.~3: Standardized probabilities and covariates of heavy-tailed
+dynamics. (a) Taxonomic-order-level posterior densities of Pr(\(\nu
 < 10\)) after accounting for time-series length in a hierarchical model. Dotted
@@
 the geometric mean of time series length across all the data (approximately 27
-time steps) and shown for orders with \(\ge 5\) populations. (\textbf{B})
+time steps) and shown for orders with \(\ge 5\) populations. (b)
 Posterior densities for potential covariates of Pr(\(\nu < 10\)). In both
@@
 
-\textbf{Fig.~4: Heavy-tailed process noise tends to be downward skewed and
-ignoring this can underestimate risk.} (\textbf{A}) Illustrations of Student-t
+Fig.~4: Heavy-tailed process noise tends to be downward skewed and
+ignoring this can underestimate risk. (a) Illustrations of Student-t
 distributions with 3 levels of skewness (\(\gamma\)) and heavy-tails (\(\nu\)).
-(\textbf{B}) Posterior density of the skewness parameters aggregated across
+(b) Posterior density of the skewness parameters aggregated across
 populations grouped into heavy-tailed (\(\hat{\nu} < 10\)), slightly
 heavy-tailed (\(10 \leq \hat{\nu} \leq 70\)), and normal-tailed populations
-(\(\hat{\nu} > 70\)). Approximate mid-values from \textbf{B} are illustrated in
-\textbf{A}. (\textbf{C}) Example time series of heavy-tailed populations with
+(\(\hat{\nu} > 70\)). Approximate mid-values from b are illustrated in
+panel a. (c) Example time series of heavy-tailed populations with
 different levels of skewness. Red dots highlight likely heavy-tailed events.
@@
 intervals (dotted lines) of abundance. Black and red lines indicate forecasts
-from Gompertz models with lognormal and skew-t process noise, respectively.
+from Gompertz models with lognormal and skew-$t$ process noise, respectively.
 

### 2016-06-24 text/ms.tex
More copyediting

@@
 
-Words in abstract: 142
+Words in abstract: 145
 
@@
 
-Number of references: todo
+Number of references: 55
 
@@
 underestimate extinction risk.
-% Ignoring black-swan events leads to a \crashUnderMedian-fold (50\% interval \crashUnderRange) underestimate in the magnitude of population crashes.
 Our results demonstrate the importance of both
@@
 conservation strategies that are robust to ecological surprises.
+% Ignoring black-swan events leads to a \crashUnderMedian-fold (50\% interval \crashUnderRange) underestimate in the magnitude of population crashes.
 
@@
 only a flexible comparative approach consistently applied 
-to a large number of timeseries can yield insights into the frequency, strength,
+to a large number of time series can yield insights into the frequency, strength,
 and correlates of black-swan events. 
@@
 than the normal distribution)
-to timeseries
+to time series
 have not accounted for an underlying population dynamics model
@@
 expected dynamics.
-Methodologically, the generalized extreme value (GEV) distribution represents
-the limit distribution of a series of maxima or minima.
-Although the GEV is well-suited for physical ecological variables
-\citep[e.g.\ sedimentation rates or wind speeds;][]{gaines1993, denny2009},
-the GEV requires maxima per time block
-or peaks beyond a chosen threshold,
-and hence requires longer timeseries then typically available
-for wild populations.
-Furthermore, the GEV approach cannot be easily integrated into population dynamics models.
 
@@
 is expected to happen once every 35 years with a $t$ distribution at \(\nu = 2\)
-compared to once every 32,000 years with a normal distribution
+compared to once every 32,000 years with a normal distribution.
 % more than four standard deviations below the mean is almost 70,000 times higher
@@
 
-This could represent the effect of unmodeled spatial heterogeneity,
-nonstationary environmental processes, or
-varying responses to temperature among individuals, for example.
+One alternative, the generalized extreme value (GEV) distribution,
+represents the limit distribution of a series of maxima or minima.
+Although the GEV is well-suited for physical ecological variables
+\citep[e.g.\ sedimentation rates or wind speeds;][]{gaines1993, denny2009},
+the GEV requires maxima per time block
+or peaks beyond a chosen threshold.
+Therefore, the GEV requires longer time series then typically available
+for wild populations.
+Furthermore, the GEV approach cannot be easily integrated into population dynamics models.
+
+% This could represent the effect of unmodeled spatial heterogeneity,
+% nonstationary environmental processes, or
+% varying responses to temperature among individuals, for example.
 
@@
 We selected abundance time series from the Global Population Dynamics
-Database \citep{gpdd2010} (GPDD), which contains nearly 5,000 time series of
+Database \citep[GPDD,][]{gpdd2010}, which contains nearly 5,000 time series of
 abundance from \(\sim\) 1,000 species and \(\sim\) 100 taxonomic orders. We
@@
 x_t &= \lambda + x_{t-1} + \epsilon_t\\
-\epsilon &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma).
+\epsilon &\sim \mathrm{Student}\mhyphen t(\nu, 0, \sigma).
 \end{align}
@@
 x_t &\sim \mathrm{Normal}(U_t, \sigma_\mathrm{obs}^2)\\
-\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma_\mathrm{proc}),
+\epsilon_t &\sim \mathrm{Student}\mhyphen t(\nu, 0, \sigma_\mathrm{proc}),
 \end{align}
@@
 x_t &= x_{t-1} + r_{\mathrm{max}}\left(1 - \frac{N_{t-1}}{K}\right) + \epsilon_t\\
-\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma),
+\epsilon_t &\sim \mathrm{Student}\mhyphen t(\nu, 0, \sigma),
 \end{align}
@@
 
-\textit{TODO: Should the following 2 paragraphs go in the results?}
 We found that we could consistently and precisely recover median posterior
@@
 
-\subsection{Skewed Student-t forecasts} To evaluate the apparent skewness
-of heavy-tailed process noise we fit Gompertz models with skewed Student-t
+\subsection{Skewed Student-$t$ forecasts} To evaluate the apparent skewness
+of heavy-tailed process noise we fit Gompertz models with skewed Student-$t$
 distributed process noise \citep{fernandez1998}. This distribution adds one
@@
 from comparing the relative fit of heavy-tailed
-population models to timeseries
+population models to time series
 with suspected mass mortality events
@@
 
-Fig.~1: Illustration of population dynamic models that allow for heavy
-tails. (a to b) Probability density for the Student-t
-distribution with scale parameter of 1 and different values of \(\nu\). Small
+\textbf{Figure~1} Illustration of population dynamic models that allow for heavy
+tails. (a to b) Probability density for the Student-$t$
+distribution with scale parameter of $1$ and different values of \(\nu\). Small
 values of \(\nu\) create heavy tails while as \(\nu\) approaches infinity the
@@
 chance of occurring in a normal distribution. (f to h)
-Estimates of \(\nu\) from models fit to the times series in c to
+Estimates of \(\nu\) from models fit to the times series in panels c to
 e. Shown are posterior samples (histograms), median and interquartile
 range of the posterior (IQR, dots and line segments), and the exponential prior
-on \(\nu\) (dashed lines). Colour shading behind f to h
+on \(\nu\) (dashed lines). Colour shading behind panels f to h
 illustrates the approximate region of heavy tails.
@@
 
-Fig.~2: Estimates of population dynamic heavy-tailedness for
+\textbf{Figure~2} Estimates of population dynamic heavy-tailedness for
 populations of birds, mammals, insects, and fishes. (a to d)
@@
 
-Fig.~3: Standardized probabilities and covariates of heavy-tailed
+\textbf{Figure~3} Standardized probabilities and covariates of heavy-tailed
 dynamics. (a) Taxonomic-order-level posterior densities of Pr(\(\nu
@@
 
-Fig.~4: Heavy-tailed process noise tends to be downward skewed and
-ignoring this can underestimate risk. (a) Illustrations of Student-t
+\textbf{Figure~4} Heavy-tailed process noise tends to be downward skewed and
+ignoring this can underestimate risk. (a) Illustrations of Student-$t$
 distributions with 3 levels of skewness (\(\gamma\)) and heavy-tails (\(\nu\)).

### 2016-06-24 text/supp.tex
More copyediting

@@
 \author{
-Sean C. Anderson$^{1,2*}$ \and
-Trevor A. Branch$^3$ \and
-Andrew B. Cooper$^2$ \and
+Sean C. Anderson$^{1*}$ \and
+Trevor A. Branch$^2$ \and
+Andrew B. Cooper$^3$ \and
 Nicholas K. Dulvy$^1$
@@
 
-\textsuperscript{2}School of Resource and Environmental Management, Simon
+\textsuperscript{2}School of Aquatic and Fishery Sciences, University of
+Washington, Box 355020, Seattle, WA 98195, USA
+
+\textsuperscript{3}School of Resource and Environmental Management, Simon
 Fraser University, Burnaby, BC, V5A 1S6, Canada
-
-\textsuperscript{3}School of Aquatic and Fishery Sciences, University of
-Washington, Box 355020, Seattle, WA 98195, USA
 
@@
 
-\section{Implementation of the skewed Student-t distribution in Stan}
+\section{Implementation of the skewed Student-$t$ distribution in Stan}
 
-The probability density for the skewed Student-t distribution
+The probability density for the skewed Student-$t$ distribution
 \citep{fernandez1998} equates to the following probability density as
@@
 \begin{equation}
-  \mathrm{Skewed\,Student\mhyphen t}(x) =
+  \mathrm{Skewed\,Student}\mhyphen t(x) =
 \begin{dcases}
 \frac{2}{(\gamma + 1/\gamma)}
-         \mathrm{Student\mhyphen t}(x \cdot \gamma, \nu, \mu \cdot \gamma, \sigma),& \text{for } x < 0\\
+         \mathrm{Student}\mhyphen t(x \cdot \gamma, \nu, \mu \cdot \gamma, \sigma),& \text{for } x < 0\\
        \frac{2}{(\gamma + 1/\gamma)}
-         \mathrm{Student\mhyphen t}(x / \gamma, \nu, \mu / \gamma, \sigma),& \text{for } x \ge 0
+         \mathrm{Student}\mhyphen t(x / \gamma, \nu, \mu / \gamma, \sigma),& \text{for } x \ge 0
 \end{dcases}
@@
 
-Here $\gamma$ represents the skewness parameter. The Student-t($\cdot$) portion
+Here $\gamma$ represents the skewness parameter. The Student-$t$($\cdot$) portion
 represents in order: the data point of interest, the heavy-tailedness
 parameter, the location parameter, and the scale parameter as passed to the
-standard Student-t probability density. This translates to the log probability
+standard Student-$t$ probability density. This translates to the log probability
 density:
@@
 \begin{equation}
-\mathrm{Skewed\,Student\mhyphen t}(x) =
+\mathrm{Skewed\,Student}\mhyphen t(x) =
 \begin{dcases}
   \log (2 \gamma) - \log(\gamma^2 + 1) +
-  \log \mathrm{Student\mhyphen t}(x \cdot \gamma, \nu, \mu \cdot \gamma, \sigma),& \text{for } x < 0\\
+  \log \mathrm{Student}\mhyphen t(x \cdot \gamma, \nu, \mu \cdot \gamma, \sigma),& \text{for } x < 0\\
   \log (2 \gamma) - \log(\gamma^2 + 1) +
-  \log \mathrm{Student\mhyphen t}(x / \gamma, \nu, \mu / \gamma, \sigma),& \text{for } x \ge 0
+  \log \mathrm{Student}\mhyphen t(x / \gamma, \nu, \mu / \gamma, \sigma),& \text{for } x \ge 0
 \end{dcases}
@@
 can replace $\log(2 \gamma)$ in the above for $\log(\gamma)$. The final log
-probability density for the skewed Student-t distribution implemented in
+probability density for the skewed Student-$t$ distribution implemented in
 Stan\footnote{Credit to Stan developer Bob Carpenter:
@@
 
-\textbf{Table~S1: Summary statistics for the filtered Global Population
-Dynamics Database time series arranged by taxonomic class.} Columns are number
+Table~S1: Summary statistics for the filtered Global Population
+Dynamics Database time series arranged by taxonomic class. Columns are number
 of populations, number of taxonomic orders, numbers of species, median time
@@
 \begin{longtable}{>{\RaggedRight}m{1.4cm}>{\RaggedRight}p{6.2cm}>{\RaggedRight}p{0.7cm}>{\RaggedRight}p{0.4cm}>{\RaggedRight}p{4.4cm}>{\RaggedRight}p{1.4cm}}
-\caption{\textbf{Populations with a high probability of heavy-tailed dynamics in the base heavy-tailed Gompertz population dynamics model.} Shown are the log abundance time series, population descriptions, Global Population Dynamics Database Main IDs, citation for the data source or separate verification literature, a description of the cause of the black-swan events (if known), the probability of heavy tails as calculated by our model, and median estimate of $\nu$ from our model with 90\% quantile credible intervals indicated in parentheses.
-Red dots on the time series indicate downward black-swan events and blue values indicate upward black-swan events that have a $10^{-4}$ probability or less of occurring if the population dynamics were explained by a Gompertz model with normally distributed process noise with a standard deviation equal to the scale parameter in the fitted t distribution.}\\
+\caption{Populations with a high probability of heavy-tailed dynamics in the base heavy-tailed Gompertz population dynamics model.
+Shown are the log abundance time series, population descriptions, Global Population Dynamics Database Main IDs, citation for the data source or separate verification literature, a description of the cause of the black-swan events (if known), the probability of heavy tails as calculated by our model, and median estimate of $\nu$ from our model with 90\% quantile credible intervals indicated in parentheses.
+Red dots on the time series indicate downward black-swan events and blue values indicate upward black-swan events that have a $10^{-4}$ probability or less of occurring if the population dynamics were explained by a Gompertz model with normally distributed process noise with a standard deviation equal to the scale parameter in the fitted $t$ distribution.}\\
 
@@
 
-\caption{\textbf{Probability density of the Bayesian priors for the Gompertz models.} (\textbf{A}) per capita growth rate at $\log$(abundance) = $0$: $\lambda \sim \mathrm{Normal}(0, 10^2)$.
-(\textbf{B}) scale parameter of t-distribution process noise: $\sigma \sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)$ (\textbf{C}) Student-t distribution degrees of freedom parameter: $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2)$.
+\caption{Probability density of the Bayesian priors for the Gompertz models.
+(\textbf{A}) per capita growth rate at $\log$(abundance) = $0$: $\lambda \sim \mathrm{Normal}(0, 10^2)$.
+(\textbf{B}) scale parameter of t-distribution process noise: $\sigma \sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)$ (\textbf{C}) Student-$t$ distribution degrees of freedom parameter: $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2)$.
 (\textbf{D}) AR1 correlation coefficient of residuals: $\phi \sim \mathrm{Truncated \mhyphen Normal}(0, 0.5, \mathrm{min.} = -1, \mathrm{max.} = 1)$.
@@
 
-\caption{\textbf{Estimates of $\nu$ from alternative models plotted against the
-    base Gompertz model estimates of $\nu$.} Alternative models are (\textbf{A})
+\caption{Estimates of $\nu$ from alternative models plotted against the
+    base Gompertz model estimates of $\nu$. Alternative models are (\textbf{A})
     Ricker-logistic; (\textbf{B}) a Gompertz model with AR1 residual correlation;
@@
 
-\caption{\textbf{Estimates of $\nu$ from Gompertz models with alternative priors on $\nu$.} \textbf{A} shows a weaker prior on $\nu$, \textbf{B} shows a stronger prior on $\nu$, as illustrated in Fig.~1.
+\caption{Estimates of $\nu$ from Gompertz models with alternative priors on $\nu$.
+\textbf{A} shows a weaker prior on $\nu$, \textbf{B} shows a stronger prior on $\nu$, as illustrated in Fig.~1.
 Shown are medians of the posterior (dots) and 50\% credible intervals (segments).
@@
 
-\caption{\textbf{Testing the ability to estimate $\nu$ (top panels) and the scale parameter of the process deviations (bottom panels) for a given number of samples (columns) drawn from a distribution with a given true $\nu$ value (rows).} The red lines indicate the true population value.
+\caption{Testing the ability to estimate $\nu$ (top panels) and the scale parameter of the process deviations (bottom panels) for a given number of samples (columns) drawn from a distribution with a given true $\nu$ value (rows).
+The red lines indicate the true population value.
 When a small number of samples are drawn there may not be samples sufficiently far into the tails to recapture the true $\nu$ value; however, heavy tails are still distinguished from normal tails in most cases, even with only 25 or 50 samples.}
@@
 
-\caption{\textbf{Simulation testing the Gompertz estimation model when the process deviation draws were chosen so that $\nu$ could be estimated close to the true value outside the full population model} (``effective $\nu$'' within a CV of 0.2 of specified $\nu$).
+\caption{Simulation testing the Gompertz estimation model when the process deviation draws were chosen so that $\nu$ could be estimated close to the true value outside the full population model (``effective $\nu$'' within a CV of 0.2 of specified $\nu$).
 Upper panels show the distribution of median $\widehat{\nu}$ across 20 simulation runs.
@@
 
-\caption{\textbf{Potential covariates of heavy-tailed population dynamics (indicated by a high probability that $\nu < 10$).} Shown are (\textbf{A} to \textbf{C}) parameters from the Gompertz heavy-tailed population model ($\sigma$, $\lambda$, $b$), (\textbf{D}) number of time steps, (\textbf{E}) body length, and (\textbf{F}) lifespan.
-For the Gompertz parameters, $\sigma$ refers to the scale parameter of the Student-t process-noise distribution, $\lambda$ refers to the expected log abundance at the next time step at an abundance of one, $b$ refers to the density dependence parameter.
+\caption{Potential covariates of heavy-tailed population dynamics (indicated by a high probability that $\nu < 10$).
+Shown are (\textbf{A} to \textbf{C}) parameters from the Gompertz heavy-tailed population model ($\sigma$, $\lambda$, $b$), (\textbf{D}) number of time steps, (\textbf{E}) body length, and (\textbf{F}) lifespan.
+For the Gompertz parameters, $\sigma$ refers to the scale parameter of the Student-$t$ process-noise distribution, $\lambda$ refers to the expected log abundance at the next time step at an abundance of one, $b$ refers to the density dependence parameter.
 Circles representing a few sharks, crustaceans, and gastropods are filled in white.
@@
 
-\caption{\textbf{Skewness and heavy-tailed parameter posteriors.} Top panel shows the shape of the skew-t distribution at different combinations of skewness ($\gamma$) and heavy-tailedness ($\nu$).
+\caption{Skewness and heavy-tailed parameter posteriors.
+Top panel shows the shape of the skew-$t$ distribution at different combinations of skewness ($\gamma$) and heavy-tailedness ($\nu$).
 Lower panel shows the posterior medians and 50\% credible intervals for the two parameters.
@@
 \includegraphics[width=1.1\textwidth]{../analysis/heavy-skew-projections-ggplot}
-\caption{\textbf{Projections of population abundance under skew-t (red) and
+\caption{Projections of population abundance under skew-$t$ (red) and
 normally distributed process error models (blue) for populations with
-Pr($\nu<10$) $>0.5$.} Numbers above panels indicate GPDD population IDs as
+Pr($\nu<10$) $>0.5$. Numbers above panels indicate GPDD population IDs as
 shown in table~S2. Horizontal axes indicates year before last

### 2016-06-25 text/ms.tex

@@
+% TODO 
+% change all ln in figures 
 \section{Black-swan events in animal populations}
@@
 There are a wide range of mechanistic processes 
-that could give rise to extreme events,
-such as spatial heterogeneity, nonstationary or heavy-tailed environmental processes, 
-or varying responses to temperature among individuals. 
+that could give rise to extreme events in population dynamics,
+such as extreme climate events
+\citep{meehl2004, katz2005, ipcc2012}
+or sudden changes in predation from other species.
 While evaluating these various hypotheses 
@@
 it would be impossible to tease out each of these 
-underlying causes from population trend data alone in a comparative manner. 
+underlying causes from population trend data alone. 
 Instead we seek a simple flexible phenomenological model 
@@
 
-One alternative, the generalized extreme value (GEV) distribution,
-represents the limit distribution of a series of maxima or minima.
+One alternative approach would involve fitting 
+the generalized extreme value (GEV) distribution,
+which represents the limit distribution of a series of maxima or minima.
 Although the GEV is well-suited for physical ecological variables
@@
 Therefore, the GEV requires longer time series then typically available
-for wild populations.
+for wild animal population abundance.
 Furthermore, the GEV approach cannot be easily integrated into population dynamics models.
@@
 We fit heavy-tailed Gompertz population dynamics models to data from the GPDD.
-The Gompertz model represents population growth as a linear function in log space.
-If \(x_{t}\) represents the ln abundance (\(N\)) at time \emph{t} then
+The Gompertz model represents population growth as a linear function in 
+log space (here and throughout we use log 
+to refer to the natural logarithm log).
+If \(x_{t}\) represents the log abundance (\(N\)) at time \emph{t} then
 \begin{align}
 x_t &= \lambda + bx_{t - 1} + \epsilon_t\\
-\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma)
+\epsilon_t &\sim \mathrm{Studen}t\mhyphen t(\nu, 0, \sigma)
 \end{align}
@@
 b &\sim \mathrm{Uniform}(-1, 2)\\ \lambda &\sim \mathrm{Normal}(0, 10^2)\\
-\sigma &\sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)\\ \nu &\sim
-\mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2).
+\sigma &\sim \mathrm{half\mhyphen Cauchy} (0, 2.5)\\ \nu &\sim
+\mathrm{truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2).
 \end{align}
@@
 we refit our models with weaker and stronger priors. Our base model used a
-prior on \(\nu\) of Truncated-Exponential(0.01, min.\ = 2). For a weaker prior
-we used Truncated-Exponential(0.005, min.\ = 2) and for a stronger prior we
-used Truncated-Exponential(0.02, min.\ = 2) (Fig.~S\ref{fig:priors}c). Note that
+prior on \(\nu\) of truncated-Exponential(0.01, min.\ = 2). For a weaker prior
+we used truncated-Exponential(0.005, min.\ = 2) and for a stronger prior we
+used truncated-Exponential(0.02, min.\ = 2) (Fig.~S\ref{fig:priors}c). Note that
 the base and weaker priors are relatively flat within the region of \(\nu <
-20\), which is the region we are mostly concerned about when categorizing
+20\), which is the region we are concerned about when categorizing
 populations as heavy tailed. We also considered using a Uniform(\(0, 1\))
-prior on \(1/\nu\) as used in \citet{gelman2014}, but this
+prior on \(1/\nu\) as used in \citet[][p.~443]{gelman2014}, but this
 prior puts considerably more weight on lower (heavy-tailed) values of \(\nu\)
@@
 model in which an autoregressive parameter was fit to the process-noise
-residuals: \begin{align} x_t &= \lambda + b x_{t-1} + \epsilon_t\\ \epsilon_t
-&\sim \mathrm{Student\mhyphen t}(\nu, \phi \epsilon_{t-1}, \sigma). \end{align}
+residuals: 
+\begin{align} 
+x_t &= \lambda + b x_{t-1} + \epsilon_t\\ 
+\epsilon_t &\sim \mathrm{Student}\mhyphen t(\nu, \phi \epsilon_{t-1}, \sigma).
+\end{align}
 In addition to the parameters in the original Gompertz model, this model
@@
 reduced possibility of \(\phi\) being near \(-1\) or \(1\). Specifically, we
-chose \(\phi \sim \mathrm{Truncated\mhyphen Normal}(0, 0.5, \mathrm{min.} = -1,
+chose \(\phi \sim \mathrm{truncated\mhyphen Normal}(0, 0.5, \mathrm{min.} = -1,
 \mathrm{max.} = 1)\).
@@
 \(1\) (density independent). This is equivalent to fitting a random walk model
-(with drift) to the \(\ln\) abundances or assuming the growth rates are drawn
+(with drift) to the \(\log\) abundances or assuming the growth rates are drawn
 from a stationary distribution. The model was as follows:
@@
 rate that is obtained when \(N_t = 0\). The parameter \(K\) represents the
-carrying capacity and, as before, \(x_t\) represents the \(\ln\) transformed
+carrying capacity and, as before, \(x_t\) represents the \(\log\) transformed
 abundance at time \(t\). The Ricker-logistic model assumes a linear decrease in
@@
 model assumes a linear decrease in population growth rate with increases in
-\textit{ln} abundance.
+\textit{log} abundance.
 
@@
 maximum lifespan and body-size data from \citet{brook2006a}.
-The beta distribution is useful when response data are on a continuous scale
-between zero and one \citep{ferrari2004}. The model was as follows:
+Beta regression is useful when response data are on a continuous scale
+between zero and one \citep[e.g.][]{ferrari2004}. The model was as follows:
 \begin{align}
@@
 We incorporated weakly informative priors into our model: \(\mathrm{Cauchy}(0,
-10)\) on the global intercept \(\alpha\), \(\mathrm{Half\mhyphen Cauchy}(0,
-2.5)\) on all standard deviation parameters, \(\mathrm{Half\mhyphen Cauchy}(0,
+10)\) on the global intercept \(\alpha\), \(\mathrm{half\mhyphen Cauchy}(0,
+2.5)\) on all standard deviation parameters, \(\mathrm{half\mhyphen Cauchy}(0,
 10)\) on the dispersion parameter \(\phi_\mathrm{disp}\), and
@@
 visually and with the same criteria as before (\(\widehat{R} < 1.05\) and
-\(n_\mathrm{eff} >199\) for all parameters). Furthermore, we checked our model
+\(n_\mathrm{eff} >200\) for all parameters). Furthermore, we checked our model
 against a similar model with the glmmADMB \textsf{R} package \citep{glmmadmb}
@@
 is a simple and elegant solution that
-can be easily substituted for the normal distribution
-Furthermore, if the degrees of freedom parameter, $\nu$,
-is fixed
+can be easily substituted for the normal distribution.
+Furthermore, if the degrees of freedom parameter ($\nu$)
+is fixed,
 it requires estimating no more parameters

### 2016-06-25 text/supp.tex

@@
 
+\clearpage
+
 \section{Supplementary tables and figures}
@@
 \begin{scriptsize}
-\begin{longtable}{>{\RaggedRight}m{1.4cm}>{\RaggedRight}p{6.2cm}>{\RaggedRight}p{0.7cm}>{\RaggedRight}p{0.4cm}>{\RaggedRight}p{4.4cm}>{\RaggedRight}p{1.4cm}}
+\begin{longtable}{>{\RaggedRight}m{1.4cm}>{\RaggedRight}p{6.0cm}>{\RaggedRight}p{0.7cm}>{\RaggedRight}p{1.2cm}>{\RaggedRight}p{4.4cm}>{\RaggedRight}p{1.4cm}}
 \caption{Populations with a high probability of heavy-tailed dynamics in the base heavy-tailed Gompertz population dynamics model.
@@
 \caption{Probability density of the Bayesian priors for the Gompertz models.
-(\textbf{A}) per capita growth rate at $\log$(abundance) = $0$: $\lambda \sim \mathrm{Normal}(0, 10^2)$.
-(\textbf{B}) scale parameter of t-distribution process noise: $\sigma \sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)$ (\textbf{C}) Student-$t$ distribution degrees of freedom parameter: $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2)$.
-(\textbf{D}) AR1 correlation coefficient of residuals: $\phi \sim \mathrm{Truncated \mhyphen Normal}(0, 0.5, \mathrm{min.} = -1, \mathrm{max.} = 1)$.
+\textbf{(a)} per capita growth rate at $\log$(abundance) = $0$: $\lambda \sim \mathrm{Normal}(0, 10^2)$.
+\textbf{(b)} scale parameter of t-distribution process noise: $\sigma \sim \mathrm{half\mhyphen Cauchy} (0, 2.5)$ \textbf{(c)} Student-$t$ distribution degrees of freedom parameter: $\nu \sim \mathrm{truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2)$.
+\textbf{(d)} AR1 correlation coefficient of residuals: $\phi \sim \mathrm{truncated \mhyphen Normal}(0, 0.5, \mathrm{min.} = -1, \mathrm{max.} = 1)$.
 Not shown is $b$, the density dependence parameter: $b \sim \mathrm{Uniform}(-1, 2)$.
-Panel \textbf{C} also shows two alternative priors: a weaker prior $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.005, \mathrm{min.} = 2)$, and a stronger prior $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.02, \mathrm{min.} = 2)$.
+Panel \textbf{(c)} also shows two alternative priors: a weaker prior $\nu \sim \mathrm{truncated\mhyphen Exponential}(0.005, \mathrm{min.} = 2)$, and a stronger prior $\nu \sim \mathrm{truncated\mhyphen Exponential}(0.02, \mathrm{min.} = 2)$.
 The inset panel shows the same data but with a log-transformed x axis.
@@
 \label{fig:priors}
-\end{center}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=0.8\textwidth]{../analysis/gomp-comparison.pdf}
-
-\caption{Estimates of $\nu$ from alternative models plotted against the
-    base Gompertz model estimates of $\nu$. Alternative models are (\textbf{A})
-    Ricker-logistic; (\textbf{B}) a Gompertz model with AR1 residual correlation;
-    (\textbf{C}) modelling population growth rate with no density dependence; and
-    (\textbf{D}) a Gompertz model with fixed magnitude of observation error. Shown
-are medians of the posterior (dots) and 50\% credible intervals (segments). The
-diagonal line indicates a one-to-one relationship. Different colours indicate
-various taxonomic classes. The grey-shaded regions indicate regions of
-disagreement if $\nu = 10$ is taken as a threshold of heavy-tailed dynamics.
-The Gompertz observation error model assumes a fixed standard deviation of
-observation error of $0.2$ on a log scale. Models in \textbf{B} and \textbf{D}
-that had not met convergence thresholds were run for four chains, 256,000 iterations (128,000 warm up), and retaining every 10th sample, after which the MCMC chains for one population in \textbf{B} and four populations in \textbf{D} had still not met convergence thresholds.
-These populations are excluded from their respective panels, but all populations with Pr($\nu < 10$) $> 0.5$ in the base model had converged and are shown.}
-
-\label{fig:alt}
-\end{center}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=0.8\textwidth]{../analysis/gomp-prior-comparison.pdf}
-
-\caption{Estimates of $\nu$ from Gompertz models with alternative priors on $\nu$.
-\textbf{A} shows a weaker prior on $\nu$, \textbf{B} shows a stronger prior on $\nu$, as illustrated in Fig.~1.
-Shown are medians of the posterior (dots) and 50\% credible intervals (segments).
-The diagonal line indicates a one-to-one relationship.
-Different colours indicate various taxonomic classes.
-The grey-shaded regions indicate regions of disagreement if $\nu = 10$ is taken as a threshold of heavy-tailed dynamics.
-In general, the estimates are nearly identical in cases where the data are informative about low values of $\nu$.
-When the data are less informative about low values of $\nu$, the prior can slightly pull the estimates of $\nu$ towards higher or lower values.}
-
-\label{fig:alt-priors}
 \end{center}
@@
 \begin{center}
+\includegraphics[width=0.8\textwidth]{../analysis/gomp-comparison.pdf}
+
+\caption{Estimates of $\nu$ from alternative models plotted against the
+    base Gompertz model estimates of $\nu$. Alternative models are \textbf{(a)}
+    Ricker-logistic; \textbf{(b)} a Gompertz model with AR1 residual correlation;
+    \textbf{(c)} modelling population growth rate with no density dependence; and
+    \textbf{(d)} a Gompertz model with fixed magnitude of observation error. Shown
+are medians of the posterior (dots) and 50\% credible intervals (segments). The
+diagonal line indicates a one-to-one relationship. Different colours indicate
+various taxonomic classes. The grey-shaded regions indicate regions of
+disagreement if $\nu = 10$ is taken as a threshold of heavy-tailed dynamics.
+The Gompertz observation error model assumes a fixed standard deviation of
+observation error of $0.2$ on a log scale. Models in \textbf{(b)} and \textbf{(d)}
+that had not met convergence thresholds were run for four chains, 256,000 iterations (128,000 warm up), and retaining every 10th sample, after which the MCMC chains for one population in \textbf{(b)} and four populations in \textbf{(d)} had still not met convergence thresholds.
+These populations are excluded from their respective panels, but all populations with Pr($\nu < 10$) $> 0.5$ in the base model had converged and are shown.}
+
+\label{fig:alt}
+\end{center}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=0.8\textwidth]{../analysis/gomp-prior-comparison.pdf}
+
+\caption{Estimates of $\nu$ from Gompertz models with alternative priors on $\nu$.
+\textbf{(a)} shows a weaker prior on $\nu$, \textbf{(b)} shows a stronger prior on $\nu$, as illustrated in Fig.~1.
+Shown are medians of the posterior (dots) and 50\% credible intervals (segments).
+The diagonal line indicates a one-to-one relationship.
+Different colours indicate various taxonomic classes.
+The grey-shaded regions indicate regions of disagreement if $\nu = 10$ is taken as a threshold of heavy-tailed dynamics.
+In general, the estimates are nearly identical in cases where the data are informative about low values of $\nu$.
+When the data are less informative about low values of $\nu$, the prior can slightly pull the estimates of $\nu$ towards higher or lower values.}
+
+\label{fig:alt-priors}
+\end{center}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[htbp]
+\begin{center}
 \includegraphics[width=0.8\textwidth]{../analysis/correlates-p10.pdf}
@@
 \caption{Potential covariates of heavy-tailed population dynamics (indicated by a high probability that $\nu < 10$).
-Shown are (\textbf{A} to \textbf{C}) parameters from the Gompertz heavy-tailed population model ($\sigma$, $\lambda$, $b$), (\textbf{D}) number of time steps, (\textbf{E}) body length, and (\textbf{F}) lifespan.
+Shown are (\textbf{a} to \textbf{c}) parameters from the Gompertz heavy-tailed population model ($\sigma$, $\lambda$, $b$), \textbf{(d)} number of time steps, \textbf{(e)} body length, and \textbf{(f)} lifespan.
 For the Gompertz parameters, $\sigma$ refers to the scale parameter of the Student-$t$ process-noise distribution, $\lambda$ refers to the expected log abundance at the next time step at an abundance of one, $b$ refers to the density dependence parameter.
 Circles representing a few sharks, crustaceans, and gastropods are filled in white.
-Median and 90\% credible interval posterior predictions of a beta regression hierarchical model are shown in panels \textbf{A} and \textbf{D} where there was a high probability the slope coefficient was different from zero (Fig.~3B).}
+Median and 90\% credible interval posterior predictions of a beta regression hierarchical model are shown in panels \textbf{(a)} and \textbf{(d)} where there was a high probability the slope coefficient was different from zero (Fig.~3B).}
 
@@
 Lower panel shows the posterior medians and 50\% credible intervals for the two parameters.
-The horizontal red line indicates the median ln(skewness) parameter across all populations.}
+The horizontal red line indicates the median log(skewness) parameter across all populations.}
 

### 2016-06-27 text/ms.tex
Copyediting of text

@@
-% TODO 
-% change all ln in figures 
 \section{Black-swan events in animal populations}
@@
 
-Words in abstract: 145
+Words in abstract: 141
 
-Words in main text: 4820
+Words in main text: 4980
 
@@
 Alternatively, most examples of population die offs come from identifying
-sudden changes in abundance that exceed a threshold of decline
+sudden changes in abundance that exceed a chosen threshold of decline
 \citep[e.g.][]{young1994, gerber2001, reed2003, fey2015}.
-But this approach gives no indication of relative event frequency
-and again does not distinguish events from
+But this approach
+does not distinguish events from
 expected dynamics.
+For instance, the range of natural variability 
+can differ dramatically across taxa.
 
@@
 parameter over the normal distribution and reverts to the normal distribution
-if strong evidence of heavy-tailed process deviations are not present.
+if strong evidence of heavy-tailed process deviations is not present.
 Furthermore, the $t$ distribution arises from a mixture of normal distributions
@@
 log space (here and throughout we use log
-to refer to the natural logarithm log).
+to refer to the natural logarithm).
 If \(x_{t}\) represents the log abundance (\(N\)) at time \emph{t} then
@@
 \begin{align}
-b &\sim \mathrm{Uniform}(-1, 2)\\ \lambda &\sim \mathrm{Normal}(0, 10^2)\\
-\sigma &\sim \mathrm{half\mhyphen Cauchy} (0, 2.5)\\ \nu &\sim
-\mathrm{truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2).
+b &\sim \mathrm{Uniform}(-1, 2)\\
+\lambda &\sim \mathrm{Normal}(0, 10^2)\\
+\sigma &\sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)\\
+\nu &\sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2).
 \end{align}
@@
 we refit our models with weaker and stronger priors. Our base model used a
-prior on \(\nu\) of truncated-Exponential(0.01, min.\ = 2). For a weaker prior
-we used truncated-Exponential(0.005, min.\ = 2) and for a stronger prior we
-used truncated-Exponential(0.02, min.\ = 2) (Fig.~S\ref{fig:priors}c). Note that
+prior on \(\nu\) of Truncated-Exponential(0.01, min.\ = 2). For a weaker prior
+we used Truncated-Exponential(0.005, min.\ = 2) and for a stronger prior we
+used Truncated-Exponential(0.02, min.\ = 2) (Fig.~S\ref{fig:priors}c). Note that
 the base and weaker priors are relatively flat within the region of \(\nu <
@@
 \subsection{Alternative population models} We fit alternative population models to
-test if four key phenomena systematically changed our conclusions.
+test if four key phenomena changed our conclusions.
 
@@
 reasonable estimates of \(\nu\), (2) it provides a test of whether density
-dependence was systematically affecting our perception of heavy tails, (3) it
+dependence was affecting our perception of heavy tails, (3) it
 matches how some previous authors have modelled heavy tails without accounting
@@
   + \bm{X_i} \bm{\beta}),
-  \: \text{for } i = 1, \dots, 617\\
+  \: \text{for } i = 1, \dots, 609\\
 A_i &= \phi_\mathrm{disp} \mu_i\\
@@
 We systematically evaluated the presence of black-swan events in hundreds of
-animal populations in a consistent manner. 
+animal populations.
 We find strong evidence for their occurrence in animal populations. Black-swan
@@
 The GPDD data represent a taxonomically and geographically biased sample of
-populations---the longer time series we focus on are dominated by commercially
+populations---the longer time series we focused on are dominated by commercially
 and recreationally important species and a disproportionate number of
@@
 But what heavy-tailed distribution should one pick
-and how heavy tailed should it be?
+and how heavy-tailed should it be?
 % nearly any reason for why a population modeler has incorrectly specified the exact and full process generating the abundance times could generate a heavy tailed process noise.
@@
 The exact shape of the tails has proven important
-in the field of dispersal biology \citep{kot1996, clark1999}.
+in the field of dispersal biology \citep[e.g.][]{kot1996, clark1999}.
 Answers for the field of population dynamics
 might come from simulation analyses exploring
-the implications of different tail shapes and
+the implications of tail shapes and
 from comparing the relative fit of heavy-tailed
@@
 
-\section{Acknowledgements}
+\section{Acknowledgments}
 
@@
 Dynamics Database and to Compute Canada's WestGrid high-performance computing
-resources. Silhouette images were obtained from phylopic.org under Creative
+resources. Silhouette images were obtained from \url{phylopic.org} under Creative
 Commons licenses; sources are listed in the Supporting Information. Funding was
@@
 \textbf{Figure~1} Illustration of population dynamic models that allow for heavy
-tails. (a to b) Probability density for the Student-$t$
+tails. (a, b) Probability density for the Student-$t$
 distribution with scale parameter of $1$ and different values of \(\nu\). Small

### 2016-06-27 text/supp.tex
Copyediting of text

@@
 \caption{Probability density of the Bayesian priors for the Gompertz models.
-\textbf{(a)} per capita growth rate at $\log$(abundance) = $0$: $\lambda \sim \mathrm{Normal}(0, 10^2)$.
-\textbf{(b)} scale parameter of t-distribution process noise: $\sigma \sim \mathrm{half\mhyphen Cauchy} (0, 2.5)$ \textbf{(c)} Student-$t$ distribution degrees of freedom parameter: $\nu \sim \mathrm{truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2)$.
-\textbf{(d)} AR1 correlation coefficient of residuals: $\phi \sim \mathrm{truncated \mhyphen Normal}(0, 0.5, \mathrm{min.} = -1, \mathrm{max.} = 1)$.
+\textbf{(a)} Per capita growth rate at $\log$(abundance) = $0$: $\lambda \sim \mathrm{Normal}(0, 10^2)$.
+\textbf{(b)} Scale parameter of t-distribution process noise: $\sigma \sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)$ \textbf{(c)} Student-$t$ distribution degrees of freedom parameter: $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2)$.
+\textbf{(d)} AR1 correlation coefficient of residuals: $\phi \sim \mathrm{Truncated \mhyphen Normal}(0, 0.5, \mathrm{min.} = -1, \mathrm{max.} = 1)$.
 Not shown is $b$, the density dependence parameter: $b \sim \mathrm{Uniform}(-1, 2)$.
-Panel \textbf{(c)} also shows two alternative priors: a weaker prior $\nu \sim \mathrm{truncated\mhyphen Exponential}(0.005, \mathrm{min.} = 2)$, and a stronger prior $\nu \sim \mathrm{truncated\mhyphen Exponential}(0.02, \mathrm{min.} = 2)$.
+Panel \textbf{(c)} also shows two alternative priors: a weaker prior $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.005, \mathrm{min.} = 2)$, and a stronger prior $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.02, \mathrm{min.} = 2)$.
 The inset panel shows the same data but with a log-transformed x axis.
@@
 \caption{Estimates of $\nu$ from Gompertz models with alternative priors on $\nu$.
-\textbf{(a)} shows a weaker prior on $\nu$, \textbf{(b)} shows a stronger prior on $\nu$, as illustrated in Fig.~1.
+\textbf{(a)} A weaker prior on $\nu$; \textbf{(b)} a stronger prior on $\nu$, as illustrated in Fig.~1.
 Shown are medians of the posterior (dots) and 50\% credible intervals (segments).

### 2016-07-05 text/ms_pnas.tex
Create pnas-formatted ms

--- 2016-06-27 text/ms.tex
@@
-\section{Black-swan events in animal populations}
+\documentclass[9pt,twocolumn,twoside]{pnas-new}
+% Use the lineno option to display guide line numbers if required.
+% Note that the use of elements such as single-column equations
+% may affect the guide line number alignment.
 
-Sean C. Anderson\textsuperscript{1*}, Trevor A.
-Branch\textsuperscript{2}, Andrew B. Cooper\textsuperscript{3}, Nicholas K.
-Dulvy\textsuperscript{4}
+\templatetype{pnasresearcharticle} % Choose template
+% {pnasresearcharticle} = Template for a two-column research article
+% {pnasmathematics} = Template for a one-column mathematics article
+% {pnasinvited} = Template for a PNAS invited submission
 
-\textsuperscript{1}Earth to Ocean Research Group, Department of Biological
+\newcommand{\tabstats}{{}{28}}
+\newcommand{\tabcausessupp}{{2}}
+\newcommand{\figpriors}{{1}}
+\newcommand{\figsimnu}{{2}}
+\newcommand{\figsimprob}{{3}}
+\newcommand{\figalt}{{4}}
+\newcommand{\figaltpriors}{{5}}
+\newcommand{\figcorrelates}{{6}}
+\newcommand{\figskewnu}{{7}}
+\newcommand{\figskewprojections}{{8}}
+\mathchardef\mhyphen="2D % math hyphen
+
+\usepackage{bm}
+
+\input{../analysis/values} % R output
+
+\title{Black-swan events in animal populations}
+
+% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
+\author[a,b]{Sean C. Anderson}
+\author[b]{Trevor A. Branch}
+\author[c]{Andrew B. Cooper}
+\author[a]{Nicholas K. Dulvy}
+
+\affil[a]{Earth to Ocean Research Group, Department of Biological
 Sciences, Simon Fraser University, Burnaby BC, V5A 1S6, Canada, e-mail:
-sandrsn@uw.edu
+sandrsn@uw.edu}
+\affil[b]{School of Aquatic and Fishery Sciences, Box 355020,
+University of Washington, Seattle, WA 98195, USA}
+\affil[c]{School of Resource and Environmental Management, Simon
+Fraser University, Burnaby, BC, V5A 1S6, Canada}
 
-\textsuperscript{2}School of Aquatic and Fishery Sciences, Box 355020,
-University of Washington, Seattle, WA 98195, USA, e-mail:
-tbranch@uw.edu
+% Please give the surname of the lead author for the running footer
+\leadauthor{Anderson}
 
-\textsuperscript{3}School of Resource and Environmental Management, Simon
-Fraser University, Burnaby, BC, V5A 1S6, Canada, e-mail:
-andrew\_cooper@sfu.ca
+% Please add here a significance statement to explain the relevance of your work
+\significancestatement{Black swans---statistically improbable events
+with profound consequences---happen
+more often than expected in financial, social, and natural systems. Our work
+demonstrates the rare but systematic presence of black-swan events in animal populations
+around the world (mostly birds, mammals, insects, and fishes). These events are
+predominantly downward, implying that unexpected population crashes occur more
+frequently than increases. Black-swan events are not driven by life history (e.g. lifespan),
+but by external causes such as extreme winters and disease. Ignoring the presence of
+downward black swans may severely underestimate extinction risk in animal populations,
+particularly under a changing climate where such extreme events are expected to increase
+in frequency and magnitude.}
 
-\textsuperscript{4}Earth to Ocean Research Group, Department of Biological
-Sciences, Simon Fraser University, Burnaby BC, V5A 1S6, Canada, e-mail:
-nick\_dulvy@sfu.ca
+% Please include corresponding author, author contribution and author declaration information
+\authorcontributions{S.C.A. and T.A.B. conceived the project;
+S.C.A., T.A.B., A.B.C., and N.K.D. designed the study; S.C.A. analyzed the
+data and wrote the paper with input from all authors.}
+\authordeclaration{Please declare any conflict of interest here.}
+\correspondingauthor{\textsuperscript{2}To whom correspondence should be addressed. E-mail: sandrsn\@uw.edu}
 
-\textsuperscript{*}Corresponding author; present address:
-School of Aquatic and Fishery Sciences, Box
-355020, University of Washington, Seattle, WA 98195, USA
+% Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
+\keywords{
+Bayesian $|$
+die-offs $|$
+ecological surprises $|$
+Gompertz $|$
+heavy tails $|$
+leptokurtic $|$
+mass mortality $|$
+population dynamics $|$
+risk $|$
+Student-$t$ distribution
+}
 
-Short title: Black-swan events in animal populations
-
-Keywords:
-Bayesian,
-die-offs,
-ecological surprises,
-Gompertz,
-heavy tails,
-leptokurtic,
-mass mortality,
-population dynamics,
-risk,
-Student-$t$ distribution
-
-Words in abstract: 141
-
-Words in main text: 4980
-
-Number of references: 55
-
-Number of figures/tables: 4/0
-
-Statement of authorship:
-S.C.A. and T.A.B. conceived the project;
-S.C.A., T.A.B., A.B.C., and N.K.D. designed the study; S.C.A. analyzed the
-data and wrote the paper with input from all authors.
-
-\newpage
-%125 now is 150
-
-\section{Abstract}
-
+\begin{abstract}
 Black swans are improbable events that nonetheless
@@
 conservation strategies that are robust to ecological surprises.
-% Ignoring black-swan events leads to a \crashUnderMedian-fold (50\% interval \crashUnderRange) underestimate in the magnitude of population crashes.
+\end{abstract}
 
-% \textbf{One Sentence Summary}: Black swans---statistically improbable events
-% with profound consequences---occur in animal populations and understanding
-% them will improve climate risk assessments.
+\dates{This manuscript was compiled on \today}
+\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}
 
-%\clearpage
+\begin{document}
 
-\section{Introduction}
+% Optional adjustment to line up main text (after abstract) of first page with line numbers, when using both lineno and twocolumn options.
+% You should only change this length when you've finalised the article contents.
+\verticaladjustment{-2pt}
 
-Major surprises happen more often than expected in
-financial, social, and natural systems \citep{taleb2007, sornette2009, may2008}.
+\maketitle
+\thispagestyle{firststyle}
+\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}
+
+% If your first paragraph (i.e.
+with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation.
+If this is the case, add \parshape=0 to the end of the list environment.
+\dropcap{M}ajor surprises happen more often than expected in
+financial, social, and natural systems \cite{taleb2007, sornette2009, may2008}.
 Massive unpredictable market swings are responsible for the majority of
-financial gains and losses \citep{taleb2007}, fatalities from the largest wars
-dwarf those from all others  \citep{newman2005}, and the frequency of the most
-damaging earthquakes has exceeded past expectations \citep{sornette2009}. In
+financial gains and losses \cite{taleb2007}, fatalities from the largest wars
+dwarf those from all others  \cite{newman2005}, and the frequency of the most
+damaging earthquakes has exceeded past expectations \cite{sornette2009}. In
 ecological systems, background rates of global extinction are punctuated by
-mass extinction \citep{harnik2012}, viruses can mutate suddenly to infect new
+mass extinction \cite{harnik2012}, viruses can mutate suddenly to infect new
 hosts, and billions of animals can die at once in mass mortality
-events \citep{fey2015}.
+events \cite{fey2015}.
 
 Indeed, such die-offs may be the most important element
-affecting population persistence \citep{mangel1994} and their importance is
+affecting population persistence \cite{mangel1994} and their importance is
 likely to increase given projected increases in the frequency and magnitude of
-climate-related extremes \citep{ipcc2012}.
+climate-related extremes \cite{ipcc2012}.
 % Despite this anecdotal evidence for
 % ecological black swans, systematic searches of large numbers of timeseries have
-% yet to reveal black-swan events \citep{keitt1998, allen2001, halley2002}.
+% yet to reveal black-swan events \cite{keitt1998, allen2001, halley2002}.
 However, the overwhelming majority of population model fitting and risk forecasting
 assumes that deviations from model predictions can be represented by a normal
-distribution \citep[on a log scale; e.g.][]{brook2006a, knape2012}.
+distribution (on a log scale; e.g.~\cite{brook2006a, knape2012}).
 If black swans occur, though, a normal distribution would under-estimate the
-probability of extreme events occurring \citep{taleb2007}.
+probability of extreme events occurring \cite{taleb2007}.
 
@@
 have not accounted for an underlying population dynamics model
-\citep{keitt1998, allen2001, halley2002, segura2013}.
+\cite{keitt1998, allen2001, halley2002, segura2013}.
 Alternatively, most examples of population die offs come from identifying
 sudden changes in abundance that exceed a chosen threshold of decline
-\citep[e.g.][]{young1994, gerber2001, reed2003, fey2015}.
+\cite{young1994, gerber2001, reed2003, fey2015}.
 But this approach
@@
 
-\section{Methods}
-
 We fit population dynamics models in which the process noise is
@@
 Lower values of \(\nu\) result in heavy-tailed distributions.
-% For example, at \(\nu = 2\), an event
-% more than 4 standard deviations below the mean is expected
 For example, an event four standard deviations below the mean
@@
 compared to once every 32,000 years with a normal distribution.
-% more than four standard deviations below the mean is almost 70,000 times higher
-% than the probability of drawing such a value from a normal distribution
-% (\(0.02\) vs.\ \(2.9\cdot10^{-7}\)).
 As \(\nu\) approaches infinity, the t
@@
 
-We chose the $t$ distribution for a number of reasons.
-Pragmatically, the $t$ distribution requires the estimation of only one additional
-parameter over the normal distribution and reverts to the normal distribution
-if strong evidence of heavy-tailed process deviations is not present.
-Furthermore, the $t$ distribution arises from a mixture of normal distributions
-with the same mean and different variances \citep{gelman2014},
-which mimics the underlying basis for black-swan events.
-There are a wide range of mechanistic processes
-that could give rise to extreme events in population dynamics,
-such as extreme climate events
-\citep{meehl2004, katz2005, ipcc2012}
-or sudden changes in predation from other species.
-While evaluating these various hypotheses
-would be important for diagnosing the underlying causes
-of extreme events and to develop predictive models,
-it would be impossible to tease out each of these
-underlying causes from population trend data alone.
-Instead we seek a simple flexible phenomenological model
-that can be applied across the widest possible
-range of population time series to provide a
-comparative overview of the prevalence
-of black-swan events.
+\begin{figure*}[htb]
+\centering
+\includegraphics[width=0.65\textwidth]{../analysis/t-nu-eg2.pdf}
+\caption{ Illustration of population dynamic models that allow for heavy
+tails. (a, b) Probability density for the Student-$t$
+distribution with scale parameter of $1$ and different values of \(\nu\). Small
+values of \(\nu\) create heavy tails while as \(\nu\) approaches infinity the
+distribution approaches the normal distribution. (c to e)
+Simulated population dynamics from a Gompertz model with process noise drawn
+from Student-$t$ distributions with three values of \(\nu\). Coloured dots in
+panels c and d represent jumps with less than a 1 in 1000
+chance of occurring in a normal distribution. (f to h)
+Estimates of \(\nu\) from models fit to the times series in panels c to
+e. Shown are posterior samples (histograms), median and interquartile
+range of the posterior (IQR, dots and line segments), and the exponential prior
+on \(\nu\) (dashed lines). Colour shading behind panels f to h
+illustrates the approximate region of heavy tails.}
+\label{fig:1}
+\end{figure*}
 
-One alternative approach would involve fitting
-the generalized extreme value (GEV) distribution,
-which represents the limit distribution of a series of maxima or minima.
-Although the GEV is well-suited for physical ecological variables
-\citep[e.g.\ sedimentation rates or wind speeds;][]{gaines1993, denny2009},
-the GEV requires maxima per time block
-or peaks beyond a chosen threshold.
-Therefore, the GEV requires longer time series then typically available
-for wild animal population abundance.
-Furthermore, the GEV approach cannot be easily integrated into population dynamics models.
-
-% This could represent the effect of unmodeled spatial heterogeneity,
-% nonstationary environmental processes, or
-% varying responses to temperature among individuals, for example.
-
-\subsection{Data}
-
-We selected abundance time series from the Global Population Dynamics
-Database \citep[GPDD,][]{gpdd2010}, which contains nearly 5,000 time series of
-abundance from \(\sim\) 1,000 species and \(\sim\) 100 taxonomic orders. We
-filtered the data (Supporting Information) to remove populations from less
-reliable data sources, and those without sufficient data for our models, and
-then interpolated remaining missing values (sensu \citet{brook2006a}).
-Our interpolation affected only \(\sim\)
-\interpPointsPerc \% of the final data points (Table~S1) and none of the data
-points that were later considered black-swan events. Our final dataset
-contained \NPops\ populations across \NOrders\ taxonomic orders and seven
-taxonomic classes, with a median of \medianTimeSteps\ time steps (range
-\minTimeSteps--\maxTimeSteps) (Table~S1).
-
-\subsection{Modeling framework}
-
-We fit heavy-tailed Gompertz population dynamics models to data from the GPDD.
-The Gompertz model represents population growth as a linear function in
-log space (here and throughout we use log
-to refer to the natural logarithm).
-If \(x_{t}\) represents the log abundance (\(N\)) at time \emph{t} then
-\begin{align}
-x_t &= \lambda + bx_{t - 1} + \epsilon_t\\
-\epsilon_t &\sim \mathrm{Studen}t\mhyphen t(\nu, 0, \sigma)
-\end{align}
-The parameter \(\lambda\) represents the expected growth rate if \(N_{t}
-= 1\). The model is density independent if \(b = 1\), maximally density
-dependent if \(b = 0\), and inversely density dependent if \(b < 0\). Usually,
-the process noise \(\epsilon_{t}\) is modelled as normally distributed, but we
-allow it to be drawn from a Student-$t$ distribution with scale parameter
-\(\sigma\) and degrees of freedom \(\nu\). We can then estimate the degree to
-which the process deviations have heavy tails and are therefore evidence of
-black-swan events (Fig.~1A, B).
-
-For the Gompertz model, our weakly-informative priors (Fig.~S\ref{fig:priors}) were:
-\begin{align}
-b &\sim \mathrm{Uniform}(-1, 2)\\
-\lambda &\sim \mathrm{Normal}(0, 10^2)\\
-\sigma &\sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)\\
-\nu &\sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2).
-\end{align}
-Our prior on \(b\) was uninformative between values of \(-1\) and \(2\). We
-would not expect values of \(b\) with levels of density dependence as low as
-\(-1\) (very strong inverse density dependence), nor would we generally expect
-values above \(1\). We allowed values of \(b\) above \(1\) to allow for
-non-stationary time series of growth rates. The estimates of \(b\) were well
-within these bounds. Our prior on \(\lambda\) was weakly informative within the
-range of expected values for population growth and is similar to the default
-priors suggested by \citet{gelman2008d} for intercepts of
-regression models. Our half-Cauchy prior on \(\sigma\) follows
-\citet{gelman2006c} and \citet{gelman2008d} and the specific scale parameter of
-\(2.5\) is based on our expected range of the value in nature from previous
-studies \citep[e.g.][]{connors2014}. In our testing of a subsample of populations, our
-parameter estimates were not qualitatively changed by switching to a uniform
-prior on \(\sigma\), but the weakly informative half-Cauchy prior sped up chain
-convergence.
-
-We chose a weakly informative exponential prior on \(\nu\) with rate parameter
-of \(0.01\) truncated at values above two---a slightly less informative prior
-than suggested by \citet{fernandez1998}. We truncated the
-distribution at \(2\), since at \(\nu < 2\) the variance of the $t$ distribution
-is undefined. This prior gives only a \(7.7\)\% probability that \(\nu < 10\)
-but constrains the sampling sufficiently to avoid \(\nu\) drifting towards infinity. In
-any case, for \(\nu > 20\) the $t$ distribution is almost indistinguishable from
-the normal distribution (Fig.~1). Based on the shape of the $t$ distribution, we
-chose the probability that \(\nu < 10\), Pr(\(\nu < 10\)), to define the
-probability of heavy-tailed (black-swan) dynamics. When categorizing a population as heavy
-tailed, we applied a 0.5 probability threshold. In the scenario where the data
-are uninformative about heavy tails (e.g.~Fig.~1E, H), the posterior for
-\(\nu\) will approximately match the prior (prior median \(= 71\), mean \(=
-102\)) and a metric of Pr\((\nu < 10) > 0.5\) will not flag the
-population as heavy tailed.
-
-We fit our models with Stan 2.6.0 \citep{stan-manual2015, hoffman2014}, and R
-3.2.0 \citep{r2015}. We began with four chains and 2,000 iterations, discarding
-the first 1,000 as warm up (i.e.~4,000 final samples). If \(\hat{R}\) (the
-potential scale reduction factor---a measure of chain convergence) was greater
-than \(1.05\) for any parameter, or if the minimum effective sample size,
-\(n_\mathrm{eff}\) (a measure of the effective number of uncorrelated samples),
-for any parameter was less than \(200\), we doubled both the total iterations
-and warm up period and sampled from the model again. These thresholds are in
-excess of the minimums recommended by \citet{gelman2006a} of
-\(\hat{R} < 1.1\) and \(n_\mathrm{eff} > 100\) for reliable point estimates and
-confidence intervals. In the majority of cases our minimum thresholds were
-greatly exceeded. We continued this procedure up to 8,000 iterations (16,000
-total samples) by which all chains had sufficiently converged.
-The No-U-Turn Hamiltonian Markov chain Monte Carlo sampler in Stan generally
-requires fewer iterations to obtain equivalent effective sample sizes than the
-Gibbs or Metropolis--Hastings algorithms \citep{stan-manual2015}.
-
-\subsection{Alternative priors}
-
-To test if the prior on \(\nu\) influenced our estimate of black-swan dynamics,
-we refit our models with weaker and stronger priors. Our base model used a
-prior on \(\nu\) of Truncated-Exponential(0.01, min.\ = 2). For a weaker prior
-we used Truncated-Exponential(0.005, min.\ = 2) and for a stronger prior we
-used Truncated-Exponential(0.02, min.\ = 2) (Fig.~S\ref{fig:priors}c). Note that
-the base and weaker priors are relatively flat within the region of \(\nu <
-20\), which is the region we are concerned about when categorizing
-populations as heavy tailed. We also considered using a Uniform(\(0, 1\))
-prior on \(1/\nu\) as used in \citet[][p.~443]{gelman2014}, but this
-prior puts considerably more weight on lower (heavy-tailed) values of \(\nu\)
-than the truncated exponential priors considered here,
-and the results are therefore not shown.
-
-\subsection{Alternative population models} We fit alternative population models to
-test if four key phenomena changed our conclusions.
-
-\subsubsection{Autocorrelated residuals} We considered a version of the Gompertz
-model in which an autoregressive parameter was fit to the process-noise
-residuals:
-\begin{align}
-x_t &= \lambda + b x_{t-1} + \epsilon_t\\
-\epsilon_t &\sim \mathrm{Student}\mhyphen t(\nu, \phi \epsilon_{t-1}, \sigma).
-\end{align}
-In addition to the parameters in the original Gompertz model, this model
-estimates an additional parameter \(\phi\), which represents the correlation of
-subsequent residuals. Based on the results of previous analyses with the
-GPDD \citep{connors2014} and the chosen prior in previous
-analyses \citep{thorson2014a}, we placed a moderately informative prior on
-\(\phi\) that assumed the greatest probability density near zero with the
-reduced possibility of \(\phi\) being near \(-1\) or \(1\). Specifically, we
-chose \(\phi \sim \mathrm{truncated\mhyphen Normal}(0, 0.5, \mathrm{min.} = -1,
-\mathrm{max.} = 1)\).
-
-\subsubsection{Density independence} We fit a simplified version of the
-Gompertz model in which the density dependence parameter \(b\) was fixed at
-\(1\) (density independent). This is equivalent to fitting a random walk model
-(with drift) to the \(\log\) abundances or assuming the growth rates are drawn
-from a stationary distribution. The model was as follows:
-\begin{align}
-x_t &= \lambda + x_{t-1} + \epsilon_t\\
-\epsilon &\sim \mathrm{Student}\mhyphen t(\nu, 0, \sigma).
-\end{align}
-We fit this model for three reasons: (1) it is computationally simpler and so
-provides a check that our more complicated full Gompertz model was obtaining
-reasonable estimates of \(\nu\), (2) it provides a test of whether density
-dependence was affecting our perception of heavy tails, (3) it
-matches how some previous authors have modelled heavy tails without accounting
-for density dependence \citep{segura2013}.
-
-\subsubsection{Observation error} Observation error can bias parameter
-estimates \citep{knape2012} and is known to affect the ability to detect extreme
-events \citep{ward2007}. In our main analysis, we fit a model that ignored
-observation error. One way to account for observation error would be to fit a
-full state-space model that simultaneously estimates the magnitude of process
-noise and observation error. However, simultaneously estimating observation and
-process noise is a challenging problem (e.g.\ because the observation and
-process noise parameters tend to negatively covary in model fitting) and is
-known to sometimes result in identifiability issues with the Gompertz
-population model \citep{knape2008}. Furthermore, our model was applied to
-hundreds of time series, often of short length (as few as \minTimeSteps\ time
-steps) and our model estimates an additional parameter---the shape of the
-process deviation tails---potentially making identifiability and computational
-issues even greater. Therefore, we considered a version of the base Gompertz
-model that allowed for a fixed level of observation error:
-\begin{align}
-U_t &= \lambda + b U_{t-1} + \epsilon_t\\
-x_t &\sim \mathrm{Normal}(U_t, \sigma_\mathrm{obs}^2)\\
-\epsilon_t &\sim \mathrm{Student}\mhyphen t(\nu, 0, \sigma_\mathrm{proc}),
-\end{align}
-where \(U\) represents the served state vector, \(\sigma_\mathrm{obs}\)
-represents the standard deviation of observation error (on a log scale), and
-\(\sigma_\mathrm{proc}\) represent the process noise scale parameter. We set
-\(\sigma_\mathrm{obs}\) to \(0.2\), which represents a typical value used in
-simulation analyses \citep[e.g.][]{valpine2002, thorson2014b}.
-
-\subsubsection{Ricker-logistic} We also fit a Ricker-logistic model:
-\begin{align}
-x_t &= x_{t-1} + r_{\mathrm{max}}\left(1 - \frac{N_{t-1}}{K}\right) + \epsilon_t\\
-\epsilon_t &\sim \mathrm{Student}\mhyphen t(\nu, 0, \sigma),
-\end{align}
-where \(r_\mathrm{max}\) represents the theoretical maximum population growth
-rate that is obtained when \(N_t = 0\). The parameter \(K\) represents the
-carrying capacity and, as before, \(x_t\) represents the \(\log\) transformed
-abundance at time \(t\). The Ricker-logistic model assumes a linear decrease in
-population growth rate with increases in abundance. In contrast, the Gompertz
-model assumes a linear decrease in population growth rate with increases in
-\textit{log} abundance.
-
-To fit the Ricker-logistic models, we chose a prior on \(K\) uniform between
-zero and twice the maximum observed abundance (similar and less informative
-than in \citet{delean2013}). We set the prior on
-\(r_\mathrm{max}\) as uniform between 0 and 20 as in \citet{delean2013}. We used the same priors on \(\nu\) and \(\sigma\) as in
-the Gompertz model.
-
-\subsection{Simulation testing}
-We performed two types of simulation testing.
-First, we tested how easily the Student-$t$ distribution \(\nu\) parameter could
-be recovered given different true values of \(\nu\) and different sample sizes.
-Second, we tested the ability of the heavy-tailed Gompertz model to obtain
-unbiased parameter estimates of \(\nu\) given that a set of process deviations
-was provided in which the effective \(\nu\) value was close to the true \(\nu\)
-value.
-
-We found that we could consistently and precisely recover median posterior
-estimates of \(\nu\) near the true value of \(\nu\) with large samples (\(\ge
-200\)) (Fig.~S\ref{fig:sim-nu} upper panels). At smaller samples we could
-usually qualitatively distinguish heavy from not-heavy tails, but the model
-tended to underestimate how heavy the tails were. At the same time, at smaller
-sample sizes, the model tended to overestimate how large the scale parameter
-was (Fig.~S\ref{fig:sim-nu} lower panels).
-
-Our simulation-testing results indicate that the Gompertz model can recapture
-the true value of \(\nu\) when the process noise was chosen so that appropriate
-tail events were present (Fig.~S\ref{fig:sim-prob} upper panels). The addition
-of observation error caused the model to tend to underestimate the degree of
-heavy-tailedness. Fitting a model with correctly specified observation error
-did not substantially reduce model bias (Fig.~S\ref{fig:sim-prob}).
-
-\subsection{Covariates of heavy-tailed dynamics} We fit a hierarchical beta
-regression model to the predicted probability of heavy tails, Pr(\(\nu < 10\)),
-to investigate potential covariates of heavy-tailed dynamics. We obtained
-maximum lifespan and body-size data from \citet{brook2006a}.
-Beta regression is useful when response data are on a continuous scale
-between zero and one \citep[e.g.][]{ferrari2004}. The model was as follows:
-\begin{align}
-\mathrm{Pr}(\nu_i < 10) &\sim \mathrm{Beta}(A_i, B_i)\\
-\mu_i &= \mathrm{logit}^{-1}(\alpha
-  + \alpha^\mathrm{class}_{j[i]}
-  + \alpha^\mathrm{order}_{k[i]}
-  + \alpha^\mathrm{species}_{l[i]}
-  + \bm{X_i} \bm{\beta}),
-  \: \text{for } i = 1, \dots, 609\\
-A_i &= \phi_\mathrm{disp} \mu_i\\
-B_i &= \phi_\mathrm{disp} (1 - \mu_i)\\
-\alpha^\mathrm{class}_j &\sim
-  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{class}}),
-  \: \text{for } j = 1, \dots, 6\\
-\alpha^\mathrm{order}_k &\sim
-  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{order}}),
-  \: \text{for } k = 1, \dots, 38\\
-\alpha^\mathrm{species}_l &\sim
-  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{species}}),
-  \: \text{for } l = 1, \dots, 301,
-\end{align}
-where \(A\) and \(B\) represent the beta distribution shape parameters;
-\(\mu_i\) represents the predicted value for population \(i\), class \(j\),
-order \(k\), and species \(l\); \(\phi_\mathrm{disp}\) represents the
-dispersion parameter; and \(\bm{X_i}\) represents a vector of predictors (such as
-lifespan) for population \(i\) with associated \(\bm{\beta}\) coefficients. The
-intercepts are allowed to vary from the overall intercept \(\alpha\) by
-taxonomic class (\(\alpha^\mathrm{class}_j\)), taxonomic order
-(\(\alpha^\mathrm{order}_k\)), and species (\(\alpha^\mathrm{species}_l\)) with
-standard deviations \(\sigma_{\alpha \; \mathrm{class}}\), \(\sigma_{\alpha \;
-\mathrm{order}}\), and \(\sigma_{\alpha \; \mathrm{species}}\). Where possible,
-we also allowed for error distributions around the predictors by incorporating
-the standard deviation of the posterior samples for the Gompertz parameters
-\(\lambda\), \(b\), and \(\log \sigma\) around the mean point value as normal
-distributions (not shown in the above equation).
-
-We log transformed \(\sigma\), time-series length, and lifespan to match the
-way they are visually represented in Fig.~S\ref{fig:correlates} and to make the
-relationship approximately linear on the logit-transformed response scale. All
-input variables were standardized by subtracting their mean and dividing by two
-standard deviations to make their coefficients comparable in
-magnitude \citep{gelman2008c}. We excluded body length as a covariate because it
-was highly correlated with lifespan, and lifespan exhibited more overlap across
-taxonomy than body length. Lifespan is also more directly related to time and
-potential mechanisms driving black-swan dynamics.
-
-We incorporated weakly informative priors into our model: \(\mathrm{Cauchy}(0,
-10)\) on the global intercept \(\alpha\), \(\mathrm{half\mhyphen Cauchy}(0,
-2.5)\) on all standard deviation parameters, \(\mathrm{half\mhyphen Cauchy}(0,
-10)\) on the dispersion parameter \(\phi_\mathrm{disp}\), and
-\(\mathrm{Cauchy}(0, 2.5)\) on all other parameters \citep{gelman2006c,
-gelman2008d}. Compared to normal priors, the Cauchy priors concentrate more
-probability density around expected parameter values while allowing for a
-higher probability density far into the tails, thereby allowing the data to
-dominate the posterior more strongly if it disagrees with the prior. We fit our
-models with 5,000 total iterations per chain, 2500 warm-up iterations, four
-chains, and discarding every second sample. We checked for chain convergence
-visually and with the same criteria as before (\(\widehat{R} < 1.05\) and
-\(n_\mathrm{eff} >200\) for all parameters). Furthermore, we checked our model
-against a similar model with the glmmADMB \textsf{R} package \citep{glmmadmb}
-without uncertainty around the predictors.
-
-To derive taxonomic-order-level estimates of the probability of heavy tails
-accounting for time-series length (Fig.~3A), we fit a separate hierarchical
-model with the same structure but with only \(\log\) time-series length as a
-predictor---in this case, we did not want to control for intrinsic population
-characteristics such as density dependence. Since our predictors were centered
-by subtracting their mean value, we obtained order-level estimates of the
-probability of heavy tails at mean log time-series length by adding the
-posteriors for \(\alpha\), \(\alpha^\mathrm{class}_j\), and
-\(\alpha^\mathrm{order}_k\).
-
-\subsection{Skewed Student-$t$ forecasts} To evaluate the apparent skewness
-of heavy-tailed process noise we fit Gompertz models with skewed Student-$t$
-distributed process noise \citep{fernandez1998}. This distribution adds one
-parameter to the Student-$t$ distribution, \(\gamma\), which controls the
-skewness. The distribution is symmetrical if \(\gamma = 1\), left skewed if \(0
-< \gamma < 1\), and right skewed if \(1 < \gamma < \infty\). We placed a weakly
-informative prior of \(\mathrm{Cauchy}(0, 2.5)\) on \(\log \gamma\). We
-aggregated 5,000 randomly selected posterior samples from the \(\gamma\)
-parameter of each model at three levels of evidence for heavy-tails:
-\emph{heavy tailed}, median \(\nu < 10\); \emph{slightly heavy tailed}, \(10 \le
-\mathrm{median}\, \nu \le 70\); and \emph{normal tailed}, median \(\nu > 70\)
-(Fig.~4).
-
-To generate five-year forecasts of abundance, we combined the posterior
-parameter samples from the skewed-t Gompertz models with stochastically
-generated process noise. We compared these forecasts to those generated by a
-standard Gompertz model with normally distributed process noise fit to the same
-data. We calculated the ratio of the lower 99\% quantile credible interval
-between the two model projections for all populations in which Pr(\(\nu < 10\))
-\(> 0.5\). To ensure the results had stabilized in the tails of the forecast
-posterior, we increased the number of posterior samples in Stan. We ran four
-chains of 20,000 iterations and discarded the first 10,000 as warmup for a
-total of 40,000 samples.
-
-\section{Results}
+\section*{Results}
 
@@
 
+\begin{figure}[htb]
+\centering
+\includegraphics[width=0.3\textwidth]{../analysis/nu-coefs-2.pdf}
+\caption{Estimates of population dynamic heavy-tailedness for
+populations of birds, mammals, insects, and fishes. (a to d)
+Small values of \(\nu\) suggest heavy-tailed black-swan dynamics. Vertical
+points and line segments represent posterior medians and 50\% / 90\% credible
+intervals for individual populations. Inset plots show probability that \(\nu
+< 10\) for populations arranged by taxonomic order and sorted by decreasing
+mean Pr(\(\nu < 10\)). Taxonomic orders with three or fewer populations in
+panel a are omitted for space. Red to yellow points highlight populations
+with a high to moderately high probability of heavy-tailed dynamics.
+}
+\label{fig:2}
+\end{figure}
+
 The majority of our heavy-tailed estimates were robust to alternative
@@
 density dependence, or modelled the population dynamics as Ricker-logistic
-instead of Gompertz (Fig.~S\ref{fig:alt}). Similarly, introducing moderate
+instead of Gompertz (Fig.~S\figalt). Similarly, introducing moderate
 observation error (CV = 0.2) only slightly decreased the estimated prevalence
-of black-swan events (Fig.~S\ref{fig:alt}), and the strength of the prior on
+of black-swan events (Fig.~S\figalt), and the strength of the prior on
 \(\nu\) had little influence on estimates of black-swan dynamics
-(Fig.~S\ref{fig:alt-priors}). Finally, our simulation testing shows that, if
+(Fig.~S\figaltpriors). Finally, our simulation testing shows that, if
 anything, our models underpredict the true magnitude and probability of
@@
 with a median of only 26
-years in our analysis (Figs~S\ref{fig:sim-nu},~S\ref{fig:sim-prob}).
+years in our analysis (Figs~S\figsimnu,~S\figsimprob).
 
 % Our results show that weaker and stronger priors have little influence on our
-% conclusions about heavy-tailed dynamics (Fig.~\ref{fig:alt-priors}). When the
+% conclusions about heavy-tailed dynamics (Fig.~\figaltpriors). When the
 % data are informative about tail behaviour (i.e. when there is strong evidence
-% of low \(\nu\) values, upper-right of Fig.~\ref{fig:alt-priors}), the prior has
+% of low \(\nu\) values, upper-right of Fig.~\figaltpriors), the prior has
 % little impact on the estimate of \(\nu\). When the data are less informative
@@
 % short or noisy), the prior can pull the estimate of \(\nu\) towards larger or
-% smaller values (Fig.~\ref{fig:alt-priors}). The vast majority of the
+% smaller values (Fig.~\figaltpriors). The vast majority of the
 % populations with Pr\((\nu < 10)\) in the base prior were not altered
@@
 magnitude of process noise but not clearly related to population growth rate,
-density dependence, or maximum lifespan (Fig.~3b, Fig.~S\ref{fig:correlates}).
+density dependence, or maximum lifespan (Fig.~3b, Fig.~S\figcorrelates).
 Longer time-series length was the strongest covariate of observing black-swan
@@
 event in populations with 60 time steps than in one with 30 time steps
-(Fig.~S\ref{fig:correlates}).
+(Fig.~S\figcorrelates).
+
+\begin{figure}[htb]
+\centering
+\includegraphics[width=0.3\textwidth]{../analysis/order-posteriors-covariates.pdf}
+\caption{Standardized probabilities and covariates of heavy-tailed
+dynamics. (a) Taxonomic-order-level posterior densities of Pr(\(\nu
+< 10\)) after accounting for time-series length in a hierarchical model. Dotted
+vertical line in a indicates the median expected Pr(\(\nu < 10\)) from the
+prior distribution. Colour shading refers to taxonomic class (yellow: fishes,
+green: insects, purple: birds, and red: mammals). Estimates are standardized at
+the geometric mean of time series length across all the data (approximately 27
+time steps) and shown for orders with \(\ge 5\) populations. (b)
+Posterior densities for potential covariates of Pr(\(\nu < 10\)). In both
+panels, short vertical line segments within the density polygons indicate
+medians.
+}
+\label{fig:3}
+\end{figure}
 
@@
 caused a downward black-swan event for a water vole (\emph{Arvicola
-terrestris}) population \citep{saucy1994}. Other black swans were the result of
+terrestris}) population \cite{saucy1994}. Other black swans were the result of
 a sequence of extreme climate events on their own. For instance, severe winters
@@
 in grey herons (\emph{Ardea cinerea}) in the United
-Kingdom \citep{stafford1971} (Fig.~4c). Our analysis finds that the last event
+Kingdom \cite{stafford1971} (Fig.~4c). Our analysis finds that the last event
 was a combination of two heavy-tailed events in a row and the population took
-three times longer to recover than predicted \citep{stafford1971}. Downward black
+three times longer to recover than predicted \cite{stafford1971}. Downward black
 swans were sometimes followed by upward black swans. For example, during
@@
 shag cormorants (\emph{Phalacrocorax aristotelis}) on the Farne Islands, United
-Kingdom declined suddenly following a red tide event in 1968 \citep{potts1980}.
+Kingdom declined suddenly following a red tide event in 1968 \cite{potts1980}.
 This freed up quality nest sites for first-time breeders, productivity
 increased, and the population experienced a rapid upswing in
-abundance \citep{potts1980}.
+abundance \cite{potts1980}.
+
+\begin{figure}[htb]
+\centering
+\includegraphics[width=0.45\textwidth]{../analysis/skew-fig.pdf}
+\caption{Heavy-tailed process noise tends to be downward skewed and
+ignoring this can underestimate risk. (a) Illustrations of Student-$t$
+distributions with 3 levels of skewness (\(\gamma\)) and heavy-tails (\(\nu\)).
+(b) Posterior density of the skewness parameters aggregated across
+populations grouped into heavy-tailed (\(\hat{\nu} < 10\)), slightly
+heavy-tailed (\(10 \leq \hat{\nu} \leq 70\)), and normal-tailed populations
+(\(\hat{\nu} > 70\)). Approximate mid-values from b are illustrated in
+panel a. (c) Example time series of heavy-tailed populations with
+different levels of skewness. Red dots highlight likely heavy-tailed events.
+Forecasts (grey regions) show median (solid lines) and lower 99\% credible
+intervals (dotted lines) of abundance. Black and red lines indicate forecasts
+from Gompertz models with lognormal and skew-$t$ process noise, respectively.
+}
+\label{fig:4}
+\end{figure}
 
@@
 strong evidence of downwardly skewed process noise (Fig.~4b,
-Fig.~S\ref{fig:skew-nu}). In contrast, populations that did not have heavy
+Fig.~S\figskewnu). In contrast, populations that did not have heavy
 tails (median \(\nu \geq 70\)) had little evidence of skewed process noise
@@
 abundance) by \crashUnderRange-fold (interquartile range; Fig.~4c,
-Fig.~S\ref{fig:skew-projections}).
+Fig.~S\figskewprojections).
 
-\section{Discussion}
+\section*{Discussion}
 
@@
 different variances can generate a $t$ distribution with heavy
-tails \citep{gelman2014}. In ecology, these processes could arise through an
+tails \cite{gelman2014}. In ecology, these processes could arise through an
 aggregation of populations across space, population
-diversity \citep{schindler2010}, or an intrinsic change in population variability
-through time \citep{carpenter2006}. Extrinsic ecological forces could also cause
-black-swan dynamics \citep{nunez2012} including extreme climate
-events \citep{meehl2004, katz2005, ipcc2012}, predation from (or competition
+diversity \cite{schindler2010}, or an intrinsic change in population variability
+through time \cite{carpenter2006}. Extrinsic ecological forces could also cause
+black-swan dynamics \cite{nunez2012} including extreme climate
+events \cite{meehl2004, katz2005, ipcc2012}, predation from (or competition
 with) other species experiencing black-swan events, or sharp changes in human
@@
 synchrony of multiple extrinsic forces could give rise to black-swan dynamics
-through synergistic interactions \citep{kirby2009} or a rare alignment of
-non-synergistic forces \citep{denny2009}.
+through synergistic interactions \cite{kirby2009} or a rare alignment of
+non-synergistic forces \cite{denny2009}.
 
@@
 and how heavy-tailed should it be?
-% nearly any reason for why a population modeler has incorrectly specified the exact and full process generating the abundance times could generate a heavy tailed process noise.
+% nearly any reason for why a population modeler has incorrectly specified the exact and full ess %%% generating the abundance times could generate a heavy tailed process noise.
 This is an open research question.
 The exact shape of the tails has proven important
-in the field of dispersal biology \citep[e.g.][]{kot1996, clark1999}.
+in the field of dispersal biology \cite{kot1996, clark1999}.
 Answers for the field of population dynamics
@@
 with suspected mass mortality events
-\citep[e.g.~the $t$ distribution compared to a mixture
-of two normal distributions;][]{ward2007}.
+(e.g.~the $t$ distribution compared to a mixture
+of two normal distributions \cite{ward2007}).
 Nevertheless, our proposed $t$ (or the skew-$t$) distribution
@@
 forecast probabilities to focus early detection efforts and develop disaster
-response plans \citep{nrc2007}. The presence of ecological black swans also suggests that we
+response plans \cite{nrc2007}. The presence of ecological black swans also suggests that we
 develop management policy that is robust to heavy tails and encourages
-general resilience \citep{carpenter2012}. For instance, setting target
+general resilience \cite{carpenter2012}. For instance, setting target
 population abundances far back from critical limits will buffer against black-swan
-events \citep{caddy1996}, and maintaining genetic, phenotypic, and behavioural
+events \cite{caddy1996}, and maintaining genetic, phenotypic, and behavioural
 diversity may allow some components of populations to persist when others are
-affected by disease or extreme environmental forces \citep{schindler2010}.
+affected by disease or extreme environmental forces \cite{schindler2010}.
 Finally, surprising, or counterintuitive ecological dynamics offer a tremendous
 opportunity to learn about ecological systems, evaluate when models break down,
-and adjust future management policy \citep{doak2008, lindenmayer2010}.
+and adjust future management policy \cite{doak2008, lindenmayer2010}.
 
 Rare catastrophes can have a profound influence on population
-persistence \citep{mangel1994}. In recent decades, ecology has moved toward
+persistence \cite{mangel1994}. In recent decades, ecology has moved toward
 focusing on aspects of variance in addition to mean
-responses \citep{thompson2013}. Our results suggest that an added focus on
+responses \cite{thompson2013}. Our results suggest that an added focus on
 ecological extremes represents the next frontier, particularly in the face of
-increased climate extremes \citep{meehl2004, ipcc2012, thompson2013}. Financial
+increased climate extremes \cite{meehl2004, ipcc2012, thompson2013}. Financial
 analysts are concerned with the shape of downward tails in financial returns
@@
 
-\section{Acknowledgments}
+\matmethods{
 
-We thank J.W. Moore, A.O. Mooers, L.R. Gerber, J.D. Yeakel, C. Minto and
+\subsection{Data}
+
+We selected abundance time series from the Global Population Dynamics
+Database \cite[GPDD,][]{gpdd2010}, which contains nearly 5,000 time series of
+abundance from \(\sim\) 1,000 species and \(\sim\) 100 taxonomic orders. We
+filtered the data (Supporting Information) to remove populations from less
+reliable data sources, and those without sufficient data for our models, and
+then interpolated remaining missing values (sensu \cite{brook2006a}).
+Our interpolation affected only \(\sim\)
+\interpPointsPerc \% of the final data points (Table~S1) and none of the data
+points that were later considered black-swan events. Our final dataset
+contained \NPops\ populations across \NOrders\ taxonomic orders and seven
+taxonomic classes, with a median of \medianTimeSteps\ time steps (range
+\minTimeSteps--\maxTimeSteps) (Table~S1).
+
+% This could represent the effect of unmodeled spatial heterogeneity,
+% nonstationary environmental processes, or
+% varying responses to temperature among individuals, for example.
+
+\subsection{Modeling framework}
+
+We fit heavy-tailed Gompertz population dynamics models to data from the GPDD.
+The Gompertz model represents population growth as a linear function in
+log space (here and throughout we use log
+to refer to the natural logarithm).
+If \(x_{t}\) represents the log abundance (\(N\)) at time \emph{t} then
+\begin{align}
+x_t &= \lambda + bx_{t - 1} + \epsilon_t\\
+\epsilon_t &\sim \mathrm{Studen}t\mhyphen t(\nu, 0, \sigma)
+\end{align}
+The parameter \(\lambda\) represents the expected growth rate if \(N_{t}
+= 1\). The model is density independent if \(b = 1\), maximally density
+dependent if \(b = 0\), and inversely density dependent if \(b < 0\). Usually,
+the process noise \(\epsilon_{t}\) is modelled as normally distributed, but we
+allow it to be drawn from a Student-$t$ distribution with scale parameter
+\(\sigma\) and degrees of freedom \(\nu\). We can then estimate the degree to
+which the process deviations have heavy tails and are therefore evidence of
+black-swan events (Fig.~1A, B).
+
+We chose the $t$ distribution for a number of reasons.
+Pragmatically, the $t$ distribution requires the estimation of only one additional
+parameter over the normal distribution and reverts to the normal distribution
+if strong evidence of heavy-tailed process deviations is not present.
+Furthermore, the $t$ distribution arises from a mixture of normal distributions
+with the same mean and different variances \cite{gelman2014},
+which mimics the underlying basis for black-swan events.
+There are a wide range of mechanistic processes
+that could give rise to extreme events in population dynamics,
+such as extreme climate events
+\cite{meehl2004, katz2005, ipcc2012}
+or sudden changes in predation from other species.
+While evaluating these various hypotheses
+would be important for diagnosing the underlying causes
+of extreme events and to develop predictive models,
+it would be impossible to tease out each of these
+underlying causes from population trend data alone.
+Instead we seek a simple flexible phenomenological model
+that can be applied across the widest possible
+range of population time series to provide a
+comparative overview of the prevalence
+of black-swan events.
+
+One alternative approach would involve fitting
+the generalized extreme value (GEV) distribution,
+which represents the limit distribution of a series of maxima or minima.
+Although the GEV is well-suited for physical ecological variables
+(e.g.\ sedimentation rates or wind speeds \cite{gaines1993, denny2009}),
+the GEV requires maxima per time block
+or peaks beyond a chosen threshold.
+Therefore, the GEV requires longer time series then typically available
+for wild animal population abundance.
+Furthermore, the GEV approach cannot be easily integrated into population dynamics models.
+
+For the Gompertz model, our weakly-informative priors (Fig.~S\figpriors) were:
+\begin{align}
+b &\sim \mathrm{Uniform}(-1, 2)\\
+\lambda &\sim \mathrm{Normal}(0, 10^2)\\
+\sigma &\sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)\\
+\nu &\sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2).
+\end{align}
+Our prior on \(b\) was uninformative between values of \(-1\) and \(2\). We
+would not expect values of \(b\) with levels of density dependence as low as
+\(-1\) (very strong inverse density dependence), nor would we generally expect
+values above \(1\). We allowed values of \(b\) above \(1\) to allow for
+non-stationary time series of growth rates. The estimates of \(b\) were well
+within these bounds. Our prior on \(\lambda\) was weakly informative within the
+range of expected values for population growth and is similar to the default
+priors suggested by Ref.~\cite{gelman2008d} for intercepts of
+regression models. Our half-Cauchy prior on \(\sigma\) follows
+Ref.~\cite{gelman2006c} and Ref.~\cite{gelman2008d} and the specific scale parameter of
+\(2.5\) is based on our expected range of the value in nature from previous
+studies \cite{connors2014}. In our testing of a subsample of populations, our
+parameter estimates were not qualitatively changed by switching to a uniform
+prior on \(\sigma\), but the weakly informative half-Cauchy prior sped up chain
+convergence.
+
+We chose a weakly informative exponential prior on \(\nu\) with rate parameter
+of \(0.01\) truncated at values above two---a slightly less informative prior
+than suggested by Ref.~\cite{fernandez1998}. We truncated the
+distribution at \(2\), since at \(\nu < 2\) the variance of the $t$ distribution
+is undefined. This prior gives only a \(7.7\)\% probability that \(\nu < 10\)
+but constrains the sampling sufficiently to avoid \(\nu\) drifting towards infinity. In
+any case, for \(\nu > 20\) the $t$ distribution is almost indistinguishable from
+the normal distribution (Fig.~1). Based on the shape of the $t$ distribution, we
+chose the probability that \(\nu < 10\), Pr(\(\nu < 10\)), to define the
+probability of heavy-tailed (black-swan) dynamics. When categorizing a population as heavy
+tailed, we applied a 0.5 probability threshold. In the scenario where the data
+are uninformative about heavy tails (e.g.~Fig.~1E, H), the posterior for
+\(\nu\) will approximately match the prior (prior median \(= 71\), mean \(=
+102\)) and a metric of Pr\((\nu < 10) > 0.5\) will not flag the
+population as heavy tailed.
+
+We fit our models with Stan 2.6.0 \cite{stan-manual2015, hoffman2014}, and R
+3.2.0 \cite{r2015}. We began with four chains and 2,000 iterations, discarding
+the first 1,000 as warm up (i.e.~4,000 final samples). If \(\hat{R}\) (the
+potential scale reduction factor---a measure of chain convergence) was greater
+than \(1.05\) for any parameter, or if the minimum effective sample size,
+\(n_\mathrm{eff}\) (a measure of the effective number of uncorrelated samples),
+for any parameter was less than \(200\), we doubled both the total iterations
+and warm up period and sampled from the model again. These thresholds are in
+excess of the minimums recommended by Ref.~\cite{gelman2006a} of
+\(\hat{R} < 1.1\) and \(n_\mathrm{eff} > 100\) for reliable point estimates and
+confidence intervals. In the majority of cases our minimum thresholds were
+greatly exceeded. We continued this procedure up to 8,000 iterations (16,000
+total samples) by which all chains had sufficiently converged.
+The No-U-Turn Hamiltonian Markov chain Monte Carlo sampler in Stan generally
+requires fewer iterations to obtain equivalent effective sample sizes than the
+Gibbs or Metropolis--Hastings algorithms \cite{stan-manual2015}.
+
+\subsection{Alternative models and simulation testing}
+We fit alternative population models to test if four key phenomena
+systematically changed our conclusions. Autocorrelation has been suggested as
+a reason for increased observed variability of abundance time series through
+time, which could create apparent heavy tails \citep{inchausti2002};
+therefore, we fit a model that included serial correlation in the residuals.
+Additionally, previous work has modelled abundance or growth rates without
+accounting for density dependence \citep{halley2002,segura2013}; therefore, we
+fit a simpler model in which we assumed density independence. Third,
+observation error could bias parameter estimates \citep{knape2012} or mask our
+ability to detect heavy tails \citep{ward2007}; therefore, we fit a model
+where we allowed for a fixed quantity of observation error ($0.2$ standard
+deviations on a log scale). Finally, the Gompertz model assumes that
+population growth rate declines linearly with log abundance. Therefore, we
+also fit an alternative model, the Ricker-logistic model, which assumes that
+population growth rate declines linearly with abundance itself (Supporting Information).
+
+In addition to alternative population models, we investigated the sensitivity
+of our results to weaker and stronger priors (exponential rate parameter $=
+0.005, 0.02$; Fig.~S\figpriors, Supporting Information). We used
+simulated data to test how easily we could detect $\nu$ given different sample
+sizes and to ensure we could recover unbiased parameter estimates from the
+Gompertz model (Supporting Information).
+
+\subsection{Covariates of heavy-tailed dynamics} We fit a hierarchical beta
+regression model to the predicted probability of heavy tails, Pr(\(\nu < 10\)),
+to investigate potential covariates of heavy-tailed dynamics. We obtained
+maximum lifespan and body-size data from Ref.~\cite{brook2006a}.
+Beta regression is useful when response data are on a continuous scale
+between zero and one \cite{ferrari2004}. The model was as follows:
+\begin{align}
+\mathrm{Pr}(\nu_i < 10) &\sim \mathrm{Beta}(A_i, B_i)\\
+\mu_i &= \mathrm{logit}^{-1}(\alpha
+  + \alpha^\mathrm{class}_{j[i]}
+  + \alpha^\mathrm{order}_{k[i]}
+  + \alpha^\mathrm{species}_{l[i]}
+  + \bm{X_i} \bm{\beta}),
+  \: \text{for } i = 1, \dots, 609\\
+A_i &= \phi_\mathrm{disp} \mu_i\\
+B_i &= \phi_\mathrm{disp} (1 - \mu_i)\\
+\alpha^\mathrm{class}_j &\sim
+  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{class}}),
+  \: \text{for } j = 1, \dots, 6\\
+\alpha^\mathrm{order}_k &\sim
+  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{order}}),
+  \: \text{for } k = 1, \dots, 38\\
+\alpha^\mathrm{species}_l &\sim
+  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{species}}),
+  \: \text{for } l = 1, \dots, 301,
+\end{align}
+where \(A\) and \(B\) represent the beta distribution shape parameters;
+\(\mu_i\) represents the predicted value for population \(i\), class \(j\),
+order \(k\), and species \(l\); \(\phi_\mathrm{disp}\) represents the
+dispersion parameter; and \(\bm{X_i}\) represents a vector of predictors (such as
+lifespan) for population \(i\) with associated \(\bm{\beta}\) coefficients. The
+intercepts are allowed to vary from the overall intercept \(\alpha\) by
+taxonomic class (\(\alpha^\mathrm{class}_j\)), taxonomic order
+(\(\alpha^\mathrm{order}_k\)), and species (\(\alpha^\mathrm{species}_l\)) with
+standard deviations \(\sigma_{\alpha \; \mathrm{class}}\), \(\sigma_{\alpha \;
+\mathrm{order}}\), and \(\sigma_{\alpha \; \mathrm{species}}\). Where possible,
+we also allowed for error distributions around the predictors by incorporating
+the standard deviation of the posterior samples for the Gompertz parameters
+\(\lambda\), \(b\), and \(\log \sigma\) around the mean point value as normal
+distributions (not shown in the above equation).
+
+We log transformed \(\sigma\), time-series length, and lifespan to match the
+way they are visually represented in Fig.~S\figcorrelates\ and to make the
+relationship approximately linear on the logit-transformed response scale. All
+input variables were standardized by subtracting their mean and dividing by two
+standard deviations to make their coefficients comparable in
+magnitude \cite{gelman2008c}. We excluded body length as a covariate because it
+was highly correlated with lifespan, and lifespan exhibited more overlap across
+taxonomy than body length. Lifespan is also more directly related to time and
+potential mechanisms driving black-swan dynamics.
+
+We incorporated weakly informative priors into our model: \(\mathrm{Cauchy}(0,
+10)\) on the global intercept \(\alpha\), \(\mathrm{half\mhyphen Cauchy}(0,
+2.5)\) on all standard deviation parameters, \(\mathrm{half\mhyphen Cauchy}(0,
+10)\) on the dispersion parameter \(\phi_\mathrm{disp}\), and
+\(\mathrm{Cauchy}(0, 2.5)\) on all other parameters \cite{gelman2006c,
+gelman2008d}. Compared to normal priors, the Cauchy priors concentrate more
+probability density around expected parameter values while allowing for a
+higher probability density far into the tails, thereby allowing the data to
+dominate the posterior more strongly if it disagrees with the prior. We fit our
+models with 5,000 total iterations per chain, 2500 warm-up iterations, four
+chains, and discarding every second sample. We checked for chain convergence
+visually and with the same criteria as before (\(\widehat{R} < 1.05\) and
+\(n_\mathrm{eff} >200\) for all parameters). Furthermore, we checked our model
+against a similar model with the glmmADMB \textsf{R} package \cite{glmmadmb}
+without uncertainty around the predictors.
+
+To derive taxonomic-order-level estimates of the probability of heavy tails
+accounting for time-series length (Fig.~3A), we fit a separate hierarchical
+model with the same structure but with only \(\log\) time-series length as a
+predictor---in this case, we did not want to control for intrinsic population
+characteristics such as density dependence. Since our predictors were centered
+by subtracting their mean value, we obtained order-level estimates of the
+probability of heavy tails at mean log time-series length by adding the
+posteriors for \(\alpha\), \(\alpha^\mathrm{class}_j\), and
+\(\alpha^\mathrm{order}_k\).
+
+\subsection{Skewed Student-$t$ forecasts} To evaluate the apparent skewness
+of heavy-tailed process noise we fit Gompertz models with skewed Student-$t$
+distributed process noise \cite{fernandez1998}. This distribution adds one
+parameter to the Student-$t$ distribution, \(\gamma\), which controls the
+skewness. The distribution is symmetrical if \(\gamma = 1\), left skewed if \(0
+< \gamma < 1\), and right skewed if \(1 < \gamma < \infty\). We placed a weakly
+informative prior of \(\mathrm{Cauchy}(0, 2.5)\) on \(\log \gamma\). We
+aggregated 5,000 randomly selected posterior samples from the \(\gamma\)
+parameter of each model at three levels of evidence for heavy-tails:
+\emph{heavy tailed}, median \(\nu < 10\); \emph{slightly heavy tailed}, \(10 \le
+\mathrm{median}\, \nu \le 70\); and \emph{normal tailed}, median \(\nu > 70\)
+(Fig.~4).
+
+To generate five-year forecasts of abundance, we combined the posterior
+parameter samples from the skewed-t Gompertz models with stochastically
+generated process noise. We compared these forecasts to those generated by a
+standard Gompertz model with normally distributed process noise fit to the same
+data. We calculated the ratio of the lower 99\% quantile credible interval
+between the two model projections for all populations in which Pr(\(\nu < 10\))
+\(> 0.5\). To ensure the results had stabilized in the tails of the forecast
+posterior, we increased the number of posterior samples in Stan. We ran four
+chains of 20,000 iterations and discarded the first 10,000 as warmup for a
+total of 40,000 samples.
+
+}
+
+\showmatmethods % Display the Materials and Methods section
+
+\acknow{We thank J.W. Moore, A.O. Mooers, L.R. Gerber, J.D. Yeakel, C. Minto and
 members of the Earth to Ocean Group for helpful discussions and comments. We
@@
 (N.K.D.).
+}
 
-\section{References}
+\showacknow % Display the acknowledgments section
 
-\renewcommand{\section}[2]{}%
+% \pnasbreak splits and balances the columns before the references.
+% If you see unexpected formatting errors, try commenting out this line
+% as it can run into problems with floats and footnotes on the final page.
+\pnasbreak
 
-\bibliographystyle{ecologyletters}\bibliography{jshort,ms}
+% Bibliography
+\bibliography{jshort,ms}
 
-% \textbf{Supporting Information}\\
-% Materials and Methods\\
-% Supplementary References\\
-% Tables S1 to S2\\
-% Figs.~S1 to S8\\
+% \section*{Figures}
 
-\clearpage
+% \clearpage
 
-% \section{Figures}
+% \clearpage
 
-\begin{center}
-\includegraphics[width=\textwidth]{../analysis/t-nu-eg2.pdf}
-\end{center}
+% \clearpage
 
-\textbf{Figure~1} Illustration of population dynamic models that allow for heavy
-tails. (a, b) Probability density for the Student-$t$
-distribution with scale parameter of $1$ and different values of \(\nu\). Small
-values of \(\nu\) create heavy tails while as \(\nu\) approaches infinity the
-distribution approaches the normal distribution. (c to e)
-Simulated population dynamics from a Gompertz model with process noise drawn
-from Student-$t$ distributions with three values of \(\nu\). Coloured dots in
-panels c and d represent jumps with less than a 1 in 1000
-chance of occurring in a normal distribution. (f to h)
-Estimates of \(\nu\) from models fit to the times series in panels c to
-e. Shown are posterior samples (histograms), median and interquartile
-range of the posterior (IQR, dots and line segments), and the exponential prior
-on \(\nu\) (dashed lines). Colour shading behind panels f to h
-illustrates the approximate region of heavy tails.
-
-\clearpage
-
-\begin{center}
-\includegraphics[width=0.36\textwidth]{../analysis/nu-coefs-2.pdf}
-\end{center}
-
-\textbf{Figure~2} Estimates of population dynamic heavy-tailedness for
-populations of birds, mammals, insects, and fishes. (a to d)
-Small values of \(\nu\) suggest heavy-tailed black-swan dynamics. Vertical
-points and line segments represent posterior medians and 50\% / 90\% credible
-intervals for individual populations. Inset plots show probability that \(\nu
-< 10\) for populations arranged by taxonomic order and sorted by decreasing
-mean Pr(\(\nu < 10\)). Taxonomic orders with three or fewer populations in
-panel a are omitted for space. Red to yellow points highlight populations
-with a high to moderately high probability of heavy-tailed dynamics.
-
-\clearpage
-
-\begin{center}
-\includegraphics[width=0.35\textwidth]{../analysis/order-posteriors-covariates.pdf}
-\end{center}
-
-\textbf{Figure~3} Standardized probabilities and covariates of heavy-tailed
-dynamics. (a) Taxonomic-order-level posterior densities of Pr(\(\nu
-< 10\)) after accounting for time-series length in a hierarchical model. Dotted
-vertical line in a indicates the median expected Pr(\(\nu < 10\)) from the
-prior distribution. Colour shading refers to taxonomic class (yellow: fishes,
-green: insects, purple: birds, and red: mammals). Estimates are standardized at
-the geometric mean of time series length across all the data (approximately 27
-time steps) and shown for orders with \(\ge 5\) populations. (b)
-Posterior densities for potential covariates of Pr(\(\nu < 10\)). In both
-panels, short vertical line segments within the density polygons indicate
-medians.
-
-\clearpage
-
-\begin{center}
-\includegraphics[width=0.65\textwidth]{../analysis/skew-fig.pdf}
-\end{center}
-
-\textbf{Figure~4} Heavy-tailed process noise tends to be downward skewed and
-ignoring this can underestimate risk. (a) Illustrations of Student-$t$
-distributions with 3 levels of skewness (\(\gamma\)) and heavy-tails (\(\nu\)).
-(b) Posterior density of the skewness parameters aggregated across
-populations grouped into heavy-tailed (\(\hat{\nu} < 10\)), slightly
-heavy-tailed (\(10 \leq \hat{\nu} \leq 70\)), and normal-tailed populations
-(\(\hat{\nu} > 70\)). Approximate mid-values from b are illustrated in
-panel a. (c) Example time series of heavy-tailed populations with
-different levels of skewness. Red dots highlight likely heavy-tailed events.
-Forecasts (grey regions) show median (solid lines) and lower 99\% credible
-intervals (dotted lines) of abundance. Black and red lines indicate forecasts
-from Gompertz models with lognormal and skew-$t$ process noise, respectively.
-
-%%% Local Variables:
-%%% mode: latex
-%%% TeX-master: "anderson-etal-blackswan-timeseries"
-%%% End:
+\end{document}

### 2016-07-05 text/supp.tex
Create pnas-formatted ms

@@
 decreases.
+
+\subsubsection{Autocorrelated residuals} We considered a version of the Gompertz
+model in which an autoregressive parameter was fit to the process-noise
+residuals:
+\begin{align}
+x_t &= \lambda + b x_{t-1} + \epsilon_t\\
+\epsilon_t &\sim \mathrm{Student}\mhyphen t(\nu, \phi \epsilon_{t-1}, \sigma).
+\end{align}
+In addition to the parameters in the original Gompertz model, this model
+estimates an additional parameter \(\phi\), which represents the correlation of
+subsequent residuals. Based on the results of previous analyses with the
+GPDD \cite{connors2014} and the chosen prior in previous
+analyses \cite{thorson2014a}, we placed a moderately informative prior on
+\(\phi\) that assumed the greatest probability density near zero with the
+reduced possibility of \(\phi\) being near \(-1\) or \(1\). Specifically, we
+chose \(\phi \sim \mathrm{truncated\mhyphen Normal}(0, 0.5, \mathrm{min.} = -1,
+\mathrm{max.} = 1)\).
+
+\subsubsection{Density independence} We fit a simplified version of the
+Gompertz model in which the density dependence parameter \(b\) was fixed at
+\(1\) (density independent). This is equivalent to fitting a random walk model
+(with drift) to the \(\log\) abundances or assuming the growth rates are drawn
+from a stationary distribution. The model was as follows:
+\begin{align}
+x_t &= \lambda + x_{t-1} + \epsilon_t\\
+\epsilon &\sim \mathrm{Student}\mhyphen t(\nu, 0, \sigma).
+\end{align}
+We fit this model for three reasons: (1) it is computationally simpler and so
+provides a check that our more complicated full Gompertz model was obtaining
+reasonable estimates of \(\nu\), (2) it provides a test of whether density
+dependence was affecting our perception of heavy tails, (3) it
+matches how some previous authors have modelled heavy tails without accounting
+for density dependence \cite{segura2013}.
+
+\subsubsection{Observation error} Observation error can bias parameter
+estimates \cite{knape2012} and is known to affect the ability to detect extreme
+events \cite{ward2007}. In our main analysis, we fit a model that ignored
+observation error. One way to account for observation error would be to fit a
+full state-space model that simultaneously estimates the magnitude of process
+noise and observation error. However, simultaneously estimating observation and
+process noise is a challenging problem (e.g.\ because the observation and
+process noise parameters tend to negatively covary in model fitting) and is
+known to sometimes result in identifiability issues with the Gompertz
+population model \cite{knape2008}. Furthermore, our model was applied to
+hundreds of time series, often of short length (as few as \minTimeSteps\ time
+steps) and our model estimates an additional parameter---the shape of the
+process deviation tails---potentially making identifiability and computational
+issues even greater. Therefore, we considered a version of the base Gompertz
+model that allowed for a fixed level of observation error:
+\begin{align}
+U_t &= \lambda + b U_{t-1} + \epsilon_t\\
+x_t &\sim \mathrm{Normal}(U_t, \sigma_\mathrm{obs}^2)\\
+\epsilon_t &\sim \mathrm{Student}\mhyphen t(\nu, 0, \sigma_\mathrm{proc}),
+\end{align}
+where \(U\) represents the served state vector, \(\sigma_\mathrm{obs}\)
+represents the standard deviation of observation error (on a log scale), and
+\(\sigma_\mathrm{proc}\) represent the process noise scale parameter. We set
+\(\sigma_\mathrm{obs}\) to \(0.2\), which represents a typical value used in
+simulation analyses \cite{valpine2002, thorson2014b}.
+
+\subsubsection{Ricker-logistic} We also fit a Ricker-logistic model:
+\begin{align}
+x_t &= x_{t-1} + r_{\mathrm{max}}\left(1 - \frac{N_{t-1}}{K}\right) + \epsilon_t\\
+\epsilon_t &\sim \mathrm{Student}\mhyphen t(\nu, 0, \sigma),
+\end{align}
+where \(r_\mathrm{max}\) represents the theoretical maximum population growth
+rate that is obtained when \(N_t = 0\). The parameter \(K\) represents the
+carrying capacity and, as before, \(x_t\) represents the \(\log\) transformed
+abundance at time \(t\). The Ricker-logistic model assumes a linear decrease in
+population growth rate with increases in abundance. In contrast, the Gompertz
+model assumes a linear decrease in population growth rate with increases in
+\textit{log} abundance.
+
+To fit the Ricker-logistic models, we chose a prior on \(K\) uniform between
+zero and twice the maximum observed abundance (similar and less informative
+than in Ref.~\cite{delean2013}). We set the prior on
+\(r_\mathrm{max}\) as uniform between 0 and 20 as in Ref.~\cite{delean2013}. We used the same priors on \(\nu\) and \(\sigma\) as in
+the Gompertz model.
+
+\subsection{Alternative priors}
+
+To test if the prior on \(\nu\) influenced our estimate of black-swan dynamics,
+we refit our models with weaker and stronger priors. Our base model used a
+prior on \(\nu\) of Truncated-Exponential(0.01, min.\ = 2). For a weaker prior
+we used Truncated-Exponential(0.005, min.\ = 2) and for a stronger prior we
+used Truncated-Exponential(0.02, min.\ = 2) (Fig.~S\ref{fig:priors}c). Note that
+the base and weaker priors are relatively flat within the region of \(\nu <
+20\), which is the region we are concerned about when categorizing
+populations as heavy tailed. We also considered using a Uniform(\(0, 1\))
+prior on \(1/\nu\) as used in Ref.~\cite{gelman2014}, but this
+prior puts considerably more weight on lower (heavy-tailed) values of \(\nu\)
+than the truncated exponential priors considered here,
+and the results are therefore not shown.
+
+\subsection{Simulation testing}
+We performed two types of simulation testing.
+First, we tested how easily the Student-$t$ distribution \(\nu\) parameter could
+be recovered given different true values of \(\nu\) and different sample sizes.
+Second, we tested the ability of the heavy-tailed Gompertz model to obtain
+unbiased parameter estimates of \(\nu\) given that a set of process deviations
+was provided in which the effective \(\nu\) value was close to the true \(\nu\)
+value.
+
+We found that we could consistently and precisely recover median posterior
+estimates of \(\nu\) near the true value of \(\nu\) with large samples (\(\ge
+200\)) (Fig.~S\ref{fig:sim-nu} upper panels). At smaller samples we could
+usually qualitatively distinguish heavy from not-heavy tails, but the model
+tended to underestimate how heavy the tails were. At the same time, at smaller
+sample sizes, the model tended to overestimate how large the scale parameter
+was (Fig.~S\ref{fig:sim-nu} lower panels).
+
+Our simulation-testing results indicate that the Gompertz model can recapture
+the true value of \(\nu\) when the process noise was chosen so that appropriate
+tail events were present (Fig.~S\ref{fig:sim-prob} upper panels). The addition
+of observation error caused the model to tend to underestimate the degree of
+heavy-tailedness. Fitting a model with correctly specified observation error
+did not substantially reduce model bias (Fig.~S\ref{fig:sim-prob}).
 

### 2016-07-05 text/ms.tex
Edit original ms for draft-pnas-format

@@
-\section{Black-swan events in animal populations}
+\noindent
+Classification: BIOLOGICAL SCIENCES, Population Biology
+
+\section{Title: Black-swan events in animal populations}
 
@@
 
-Short title: Black-swan events in animal populations
-
 Keywords:
@@
 
-Words in abstract: 141
-
-Words in main text: 4980
-
-Number of references: 55
-
-Number of figures/tables: 4/0
-
 Statement of authorship:
@@
 \newpage
-%125 now is 150
+%250 TODO
 
@@
 
+\section{Significance Statement}
+
+TODO
+
+% authors must submit a 120-word-maximum statement about the significance of their research paper written at a level understandable to an undergraduate-educated scientist outside their field of specialty.
+The primary goal of the Significance Statement is to explain the relevance of the work in broad context to a broad readership.
+The Significance Statement appears in the paper itself and is required for all research papers.
+
 \section{Introduction}
@@
 Major surprises happen more often than expected in
-financial, social, and natural systems \citep{taleb2007, sornette2009, may2008}.
+financial, social, and natural systems \cite{taleb2007, sornette2009, may2008}.
 Massive unpredictable market swings are responsible for the majority of
-financial gains and losses \citep{taleb2007}, fatalities from the largest wars
-dwarf those from all others  \citep{newman2005}, and the frequency of the most
-damaging earthquakes has exceeded past expectations \citep{sornette2009}. In
+financial gains and losses \cite{taleb2007}, fatalities from the largest wars
+dwarf those from all others  \cite{newman2005}, and the frequency of the most
+damaging earthquakes has exceeded past expectations \cite{sornette2009}. In
 ecological systems, background rates of global extinction are punctuated by
-mass extinction \citep{harnik2012}, viruses can mutate suddenly to infect new
+mass extinction \cite{harnik2012}, viruses can mutate suddenly to infect new
 hosts, and billions of animals can die at once in mass mortality
-events \citep{fey2015}.
+events \cite{fey2015}.
 
 Indeed, such die-offs may be the most important element
-affecting population persistence \citep{mangel1994} and their importance is
+affecting population persistence \cite{mangel1994} and their importance is
 likely to increase given projected increases in the frequency and magnitude of
-climate-related extremes \citep{ipcc2012}.
+climate-related extremes \cite{ipcc2012}.
 % Despite this anecdotal evidence for
 % ecological black swans, systematic searches of large numbers of timeseries have
-% yet to reveal black-swan events \citep{keitt1998, allen2001, halley2002}.
+% yet to reveal black-swan events \cite{keitt1998, allen2001, halley2002}.
 However, the overwhelming majority of population model fitting and risk forecasting
 assumes that deviations from model predictions can be represented by a normal
-distribution \citep[on a log scale; e.g.][]{brook2006a, knape2012}.
+distribution (on a log scale; e.g.~\cite{brook2006a, knape2012}).
 If black swans occur, though, a normal distribution would under-estimate the
-probability of extreme events occurring \citep{taleb2007}.
+probability of extreme events occurring \cite{taleb2007}.
 
@@
 abundance time series.
-
-\section{Methods}
-
-We fit population dynamics models in which the process noise is
-drawn from a Student-$t$ distribution. By estimating the degrees of freedom
-parameter, \(\nu\), we can estimate the degree to which the process deviations
-have heavy tails and are therefore diagnostic of black-swan events (Fig.~1).
-Lower values of \(\nu\) result in heavy-tailed distributions.
-% For example, at \(\nu = 2\), an event
-% more than 4 standard deviations below the mean is expected
-For example, an event four standard deviations below the mean
-is expected to happen once every 35 years with a $t$ distribution at \(\nu = 2\)
-compared to once every 32,000 years with a normal distribution.
-% more than four standard deviations below the mean is almost 70,000 times higher
-% than the probability of drawing such a value from a normal distribution
-% (\(0.02\) vs.\ \(2.9\cdot10^{-7}\)).
-As \(\nu\) approaches infinity, the t
-distribution approaches the normal distribution (Fig.~1).
-
-We chose the $t$ distribution for a number of reasons.
-Pragmatically, the $t$ distribution requires the estimation of only one additional
-parameter over the normal distribution and reverts to the normal distribution
-if strong evidence of heavy-tailed process deviations is not present.
-Furthermore, the $t$ distribution arises from a mixture of normal distributions
-with the same mean and different variances \citep{gelman2014},
-which mimics the underlying basis for black-swan events.
-There are a wide range of mechanistic processes
-that could give rise to extreme events in population dynamics,
-such as extreme climate events
-\citep{meehl2004, katz2005, ipcc2012}
-or sudden changes in predation from other species.
-While evaluating these various hypotheses
-would be important for diagnosing the underlying causes
-of extreme events and to develop predictive models,
-it would be impossible to tease out each of these
-underlying causes from population trend data alone.
-Instead we seek a simple flexible phenomenological model
-that can be applied across the widest possible
-range of population time series to provide a
-comparative overview of the prevalence
-of black-swan events.
-
-One alternative approach would involve fitting
-the generalized extreme value (GEV) distribution,
-which represents the limit distribution of a series of maxima or minima.
-Although the GEV is well-suited for physical ecological variables
-\citep[e.g.\ sedimentation rates or wind speeds;][]{gaines1993, denny2009},
-the GEV requires maxima per time block
-or peaks beyond a chosen threshold.
-Therefore, the GEV requires longer time series then typically available
-for wild animal population abundance.
-Furthermore, the GEV approach cannot be easily integrated into population dynamics models.
-
-% This could represent the effect of unmodeled spatial heterogeneity,
-% nonstationary environmental processes, or
-% varying responses to temperature among individuals, for example.
-
-\subsection{Data}
-
-We selected abundance time series from the Global Population Dynamics
-Database \citep[GPDD,][]{gpdd2010}, which contains nearly 5,000 time series of
-abundance from \(\sim\) 1,000 species and \(\sim\) 100 taxonomic orders. We
-filtered the data (Supporting Information) to remove populations from less
-reliable data sources, and those without sufficient data for our models, and
-then interpolated remaining missing values (sensu \citet{brook2006a}).
-Our interpolation affected only \(\sim\)
-\interpPointsPerc \% of the final data points (Table~S1) and none of the data
-points that were later considered black-swan events. Our final dataset
-contained \NPops\ populations across \NOrders\ taxonomic orders and seven
-taxonomic classes, with a median of \medianTimeSteps\ time steps (range
-\minTimeSteps--\maxTimeSteps) (Table~S1).
-
-\subsection{Modeling framework}
-
-We fit heavy-tailed Gompertz population dynamics models to data from the GPDD.
-The Gompertz model represents population growth as a linear function in
-log space (here and throughout we use log
-to refer to the natural logarithm).
-If \(x_{t}\) represents the log abundance (\(N\)) at time \emph{t} then
-\begin{align}
-x_t &= \lambda + bx_{t - 1} + \epsilon_t\\
-\epsilon_t &\sim \mathrm{Studen}t\mhyphen t(\nu, 0, \sigma)
-\end{align}
-The parameter \(\lambda\) represents the expected growth rate if \(N_{t}
-= 1\). The model is density independent if \(b = 1\), maximally density
-dependent if \(b = 0\), and inversely density dependent if \(b < 0\). Usually,
-the process noise \(\epsilon_{t}\) is modelled as normally distributed, but we
-allow it to be drawn from a Student-$t$ distribution with scale parameter
-\(\sigma\) and degrees of freedom \(\nu\). We can then estimate the degree to
-which the process deviations have heavy tails and are therefore evidence of
-black-swan events (Fig.~1A, B).
-
-For the Gompertz model, our weakly-informative priors (Fig.~S\ref{fig:priors}) were:
-\begin{align}
-b &\sim \mathrm{Uniform}(-1, 2)\\
-\lambda &\sim \mathrm{Normal}(0, 10^2)\\
-\sigma &\sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)\\
-\nu &\sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2).
-\end{align}
-Our prior on \(b\) was uninformative between values of \(-1\) and \(2\). We
-would not expect values of \(b\) with levels of density dependence as low as
-\(-1\) (very strong inverse density dependence), nor would we generally expect
-values above \(1\). We allowed values of \(b\) above \(1\) to allow for
-non-stationary time series of growth rates. The estimates of \(b\) were well
-within these bounds. Our prior on \(\lambda\) was weakly informative within the
-range of expected values for population growth and is similar to the default
-priors suggested by \citet{gelman2008d} for intercepts of
-regression models. Our half-Cauchy prior on \(\sigma\) follows
-\citet{gelman2006c} and \citet{gelman2008d} and the specific scale parameter of
-\(2.5\) is based on our expected range of the value in nature from previous
-studies \citep[e.g.][]{connors2014}. In our testing of a subsample of populations, our
-parameter estimates were not qualitatively changed by switching to a uniform
-prior on \(\sigma\), but the weakly informative half-Cauchy prior sped up chain
-convergence.
-
-We chose a weakly informative exponential prior on \(\nu\) with rate parameter
-of \(0.01\) truncated at values above two---a slightly less informative prior
-than suggested by \citet{fernandez1998}. We truncated the
-distribution at \(2\), since at \(\nu < 2\) the variance of the $t$ distribution
-is undefined. This prior gives only a \(7.7\)\% probability that \(\nu < 10\)
-but constrains the sampling sufficiently to avoid \(\nu\) drifting towards infinity. In
-any case, for \(\nu > 20\) the $t$ distribution is almost indistinguishable from
-the normal distribution (Fig.~1). Based on the shape of the $t$ distribution, we
-chose the probability that \(\nu < 10\), Pr(\(\nu < 10\)), to define the
-probability of heavy-tailed (black-swan) dynamics. When categorizing a population as heavy
-tailed, we applied a 0.5 probability threshold. In the scenario where the data
-are uninformative about heavy tails (e.g.~Fig.~1E, H), the posterior for
-\(\nu\) will approximately match the prior (prior median \(= 71\), mean \(=
-102\)) and a metric of Pr\((\nu < 10) > 0.5\) will not flag the
-population as heavy tailed.
-
-We fit our models with Stan 2.6.0 \citep{stan-manual2015, hoffman2014}, and R
-3.2.0 \citep{r2015}. We began with four chains and 2,000 iterations, discarding
-the first 1,000 as warm up (i.e.~4,000 final samples). If \(\hat{R}\) (the
-potential scale reduction factor---a measure of chain convergence) was greater
-than \(1.05\) for any parameter, or if the minimum effective sample size,
-\(n_\mathrm{eff}\) (a measure of the effective number of uncorrelated samples),
-for any parameter was less than \(200\), we doubled both the total iterations
-and warm up period and sampled from the model again. These thresholds are in
-excess of the minimums recommended by \citet{gelman2006a} of
-\(\hat{R} < 1.1\) and \(n_\mathrm{eff} > 100\) for reliable point estimates and
-confidence intervals. In the majority of cases our minimum thresholds were
-greatly exceeded. We continued this procedure up to 8,000 iterations (16,000
-total samples) by which all chains had sufficiently converged.
-The No-U-Turn Hamiltonian Markov chain Monte Carlo sampler in Stan generally
-requires fewer iterations to obtain equivalent effective sample sizes than the
-Gibbs or Metropolis--Hastings algorithms \citep{stan-manual2015}.
-
-\subsection{Alternative priors}
-
-To test if the prior on \(\nu\) influenced our estimate of black-swan dynamics,
-we refit our models with weaker and stronger priors. Our base model used a
-prior on \(\nu\) of Truncated-Exponential(0.01, min.\ = 2). For a weaker prior
-we used Truncated-Exponential(0.005, min.\ = 2) and for a stronger prior we
-used Truncated-Exponential(0.02, min.\ = 2) (Fig.~S\ref{fig:priors}c). Note that
-the base and weaker priors are relatively flat within the region of \(\nu <
-20\), which is the region we are concerned about when categorizing
-populations as heavy tailed. We also considered using a Uniform(\(0, 1\))
-prior on \(1/\nu\) as used in \citet[][p.~443]{gelman2014}, but this
-prior puts considerably more weight on lower (heavy-tailed) values of \(\nu\)
-than the truncated exponential priors considered here,
-and the results are therefore not shown.
-
-\subsection{Alternative population models} We fit alternative population models to
-test if four key phenomena changed our conclusions.
-
-\subsubsection{Autocorrelated residuals} We considered a version of the Gompertz
-model in which an autoregressive parameter was fit to the process-noise
-residuals:
-\begin{align}
-x_t &= \lambda + b x_{t-1} + \epsilon_t\\
-\epsilon_t &\sim \mathrm{Student}\mhyphen t(\nu, \phi \epsilon_{t-1}, \sigma).
-\end{align}
-In addition to the parameters in the original Gompertz model, this model
-estimates an additional parameter \(\phi\), which represents the correlation of
-subsequent residuals. Based on the results of previous analyses with the
-GPDD \citep{connors2014} and the chosen prior in previous
-analyses \citep{thorson2014a}, we placed a moderately informative prior on
-\(\phi\) that assumed the greatest probability density near zero with the
-reduced possibility of \(\phi\) being near \(-1\) or \(1\). Specifically, we
-chose \(\phi \sim \mathrm{truncated\mhyphen Normal}(0, 0.5, \mathrm{min.} = -1,
-\mathrm{max.} = 1)\).
-
-\subsubsection{Density independence} We fit a simplified version of the
-Gompertz model in which the density dependence parameter \(b\) was fixed at
-\(1\) (density independent). This is equivalent to fitting a random walk model
-(with drift) to the \(\log\) abundances or assuming the growth rates are drawn
-from a stationary distribution. The model was as follows:
-\begin{align}
-x_t &= \lambda + x_{t-1} + \epsilon_t\\
-\epsilon &\sim \mathrm{Student}\mhyphen t(\nu, 0, \sigma).
-\end{align}
-We fit this model for three reasons: (1) it is computationally simpler and so
-provides a check that our more complicated full Gompertz model was obtaining
-reasonable estimates of \(\nu\), (2) it provides a test of whether density
-dependence was affecting our perception of heavy tails, (3) it
-matches how some previous authors have modelled heavy tails without accounting
-for density dependence \citep{segura2013}.
-
-\subsubsection{Observation error} Observation error can bias parameter
-estimates \citep{knape2012} and is known to affect the ability to detect extreme
-events \citep{ward2007}. In our main analysis, we fit a model that ignored
-observation error. One way to account for observation error would be to fit a
-full state-space model that simultaneously estimates the magnitude of process
-noise and observation error. However, simultaneously estimating observation and
-process noise is a challenging problem (e.g.\ because the observation and
-process noise parameters tend to negatively covary in model fitting) and is
-known to sometimes result in identifiability issues with the Gompertz
-population model \citep{knape2008}. Furthermore, our model was applied to
-hundreds of time series, often of short length (as few as \minTimeSteps\ time
-steps) and our model estimates an additional parameter---the shape of the
-process deviation tails---potentially making identifiability and computational
-issues even greater. Therefore, we considered a version of the base Gompertz
-model that allowed for a fixed level of observation error:
-\begin{align}
-U_t &= \lambda + b U_{t-1} + \epsilon_t\\
-x_t &\sim \mathrm{Normal}(U_t, \sigma_\mathrm{obs}^2)\\
-\epsilon_t &\sim \mathrm{Student}\mhyphen t(\nu, 0, \sigma_\mathrm{proc}),
-\end{align}
-where \(U\) represents the served state vector, \(\sigma_\mathrm{obs}\)
-represents the standard deviation of observation error (on a log scale), and
-\(\sigma_\mathrm{proc}\) represent the process noise scale parameter. We set
-\(\sigma_\mathrm{obs}\) to \(0.2\), which represents a typical value used in
-simulation analyses \citep[e.g.][]{valpine2002, thorson2014b}.
-
-\subsubsection{Ricker-logistic} We also fit a Ricker-logistic model:
-\begin{align}
-x_t &= x_{t-1} + r_{\mathrm{max}}\left(1 - \frac{N_{t-1}}{K}\right) + \epsilon_t\\
-\epsilon_t &\sim \mathrm{Student}\mhyphen t(\nu, 0, \sigma),
-\end{align}
-where \(r_\mathrm{max}\) represents the theoretical maximum population growth
-rate that is obtained when \(N_t = 0\). The parameter \(K\) represents the
-carrying capacity and, as before, \(x_t\) represents the \(\log\) transformed
-abundance at time \(t\). The Ricker-logistic model assumes a linear decrease in
-population growth rate with increases in abundance. In contrast, the Gompertz
-model assumes a linear decrease in population growth rate with increases in
-\textit{log} abundance.
-
-To fit the Ricker-logistic models, we chose a prior on \(K\) uniform between
-zero and twice the maximum observed abundance (similar and less informative
-than in \citet{delean2013}). We set the prior on
-\(r_\mathrm{max}\) as uniform between 0 and 20 as in \citet{delean2013}. We used the same priors on \(\nu\) and \(\sigma\) as in
-the Gompertz model.
-
-\subsection{Simulation testing}
-We performed two types of simulation testing.
-First, we tested how easily the Student-$t$ distribution \(\nu\) parameter could
-be recovered given different true values of \(\nu\) and different sample sizes.
-Second, we tested the ability of the heavy-tailed Gompertz model to obtain
-unbiased parameter estimates of \(\nu\) given that a set of process deviations
-was provided in which the effective \(\nu\) value was close to the true \(\nu\)
-value.
-
-We found that we could consistently and precisely recover median posterior
-estimates of \(\nu\) near the true value of \(\nu\) with large samples (\(\ge
-200\)) (Fig.~S\ref{fig:sim-nu} upper panels). At smaller samples we could
-usually qualitatively distinguish heavy from not-heavy tails, but the model
-tended to underestimate how heavy the tails were. At the same time, at smaller
-sample sizes, the model tended to overestimate how large the scale parameter
-was (Fig.~S\ref{fig:sim-nu} lower panels).
-
-Our simulation-testing results indicate that the Gompertz model can recapture
-the true value of \(\nu\) when the process noise was chosen so that appropriate
-tail events were present (Fig.~S\ref{fig:sim-prob} upper panels). The addition
-of observation error caused the model to tend to underestimate the degree of
-heavy-tailedness. Fitting a model with correctly specified observation error
-did not substantially reduce model bias (Fig.~S\ref{fig:sim-prob}).
-
-\subsection{Covariates of heavy-tailed dynamics} We fit a hierarchical beta
-regression model to the predicted probability of heavy tails, Pr(\(\nu < 10\)),
-to investigate potential covariates of heavy-tailed dynamics. We obtained
-maximum lifespan and body-size data from \citet{brook2006a}.
-Beta regression is useful when response data are on a continuous scale
-between zero and one \citep[e.g.][]{ferrari2004}. The model was as follows:
-\begin{align}
-\mathrm{Pr}(\nu_i < 10) &\sim \mathrm{Beta}(A_i, B_i)\\
-\mu_i &= \mathrm{logit}^{-1}(\alpha
-  + \alpha^\mathrm{class}_{j[i]}
-  + \alpha^\mathrm{order}_{k[i]}
-  + \alpha^\mathrm{species}_{l[i]}
-  + \bm{X_i} \bm{\beta}),
-  \: \text{for } i = 1, \dots, 609\\
-A_i &= \phi_\mathrm{disp} \mu_i\\
-B_i &= \phi_\mathrm{disp} (1 - \mu_i)\\
-\alpha^\mathrm{class}_j &\sim
-  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{class}}),
-  \: \text{for } j = 1, \dots, 6\\
-\alpha^\mathrm{order}_k &\sim
-  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{order}}),
-  \: \text{for } k = 1, \dots, 38\\
-\alpha^\mathrm{species}_l &\sim
-  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{species}}),
-  \: \text{for } l = 1, \dots, 301,
-\end{align}
-where \(A\) and \(B\) represent the beta distribution shape parameters;
-\(\mu_i\) represents the predicted value for population \(i\), class \(j\),
-order \(k\), and species \(l\); \(\phi_\mathrm{disp}\) represents the
-dispersion parameter; and \(\bm{X_i}\) represents a vector of predictors (such as
-lifespan) for population \(i\) with associated \(\bm{\beta}\) coefficients. The
-intercepts are allowed to vary from the overall intercept \(\alpha\) by
-taxonomic class (\(\alpha^\mathrm{class}_j\)), taxonomic order
-(\(\alpha^\mathrm{order}_k\)), and species (\(\alpha^\mathrm{species}_l\)) with
-standard deviations \(\sigma_{\alpha \; \mathrm{class}}\), \(\sigma_{\alpha \;
-\mathrm{order}}\), and \(\sigma_{\alpha \; \mathrm{species}}\). Where possible,
-we also allowed for error distributions around the predictors by incorporating
-the standard deviation of the posterior samples for the Gompertz parameters
-\(\lambda\), \(b\), and \(\log \sigma\) around the mean point value as normal
-distributions (not shown in the above equation).
-
-We log transformed \(\sigma\), time-series length, and lifespan to match the
-way they are visually represented in Fig.~S\ref{fig:correlates} and to make the
-relationship approximately linear on the logit-transformed response scale. All
-input variables were standardized by subtracting their mean and dividing by two
-standard deviations to make their coefficients comparable in
-magnitude \citep{gelman2008c}. We excluded body length as a covariate because it
-was highly correlated with lifespan, and lifespan exhibited more overlap across
-taxonomy than body length. Lifespan is also more directly related to time and
-potential mechanisms driving black-swan dynamics.
-
-We incorporated weakly informative priors into our model: \(\mathrm{Cauchy}(0,
-10)\) on the global intercept \(\alpha\), \(\mathrm{half\mhyphen Cauchy}(0,
-2.5)\) on all standard deviation parameters, \(\mathrm{half\mhyphen Cauchy}(0,
-10)\) on the dispersion parameter \(\phi_\mathrm{disp}\), and
-\(\mathrm{Cauchy}(0, 2.5)\) on all other parameters \citep{gelman2006c,
-gelman2008d}. Compared to normal priors, the Cauchy priors concentrate more
-probability density around expected parameter values while allowing for a
-higher probability density far into the tails, thereby allowing the data to
-dominate the posterior more strongly if it disagrees with the prior. We fit our
-models with 5,000 total iterations per chain, 2500 warm-up iterations, four
-chains, and discarding every second sample. We checked for chain convergence
-visually and with the same criteria as before (\(\widehat{R} < 1.05\) and
-\(n_\mathrm{eff} >200\) for all parameters). Furthermore, we checked our model
-against a similar model with the glmmADMB \textsf{R} package \citep{glmmadmb}
-without uncertainty around the predictors.
-
-To derive taxonomic-order-level estimates of the probability of heavy tails
-accounting for time-series length (Fig.~3A), we fit a separate hierarchical
-model with the same structure but with only \(\log\) time-series length as a
-predictor---in this case, we did not want to control for intrinsic population
-characteristics such as density dependence. Since our predictors were centered
-by subtracting their mean value, we obtained order-level estimates of the
-probability of heavy tails at mean log time-series length by adding the
-posteriors for \(\alpha\), \(\alpha^\mathrm{class}_j\), and
-\(\alpha^\mathrm{order}_k\).
-
-\subsection{Skewed Student-$t$ forecasts} To evaluate the apparent skewness
-of heavy-tailed process noise we fit Gompertz models with skewed Student-$t$
-distributed process noise \citep{fernandez1998}. This distribution adds one
-parameter to the Student-$t$ distribution, \(\gamma\), which controls the
-skewness. The distribution is symmetrical if \(\gamma = 1\), left skewed if \(0
-< \gamma < 1\), and right skewed if \(1 < \gamma < \infty\). We placed a weakly
-informative prior of \(\mathrm{Cauchy}(0, 2.5)\) on \(\log \gamma\). We
-aggregated 5,000 randomly selected posterior samples from the \(\gamma\)
-parameter of each model at three levels of evidence for heavy-tails:
-\emph{heavy tailed}, median \(\nu < 10\); \emph{slightly heavy tailed}, \(10 \le
-\mathrm{median}\, \nu \le 70\); and \emph{normal tailed}, median \(\nu > 70\)
-(Fig.~4).
-
-To generate five-year forecasts of abundance, we combined the posterior
-parameter samples from the skewed-t Gompertz models with stochastically
-generated process noise. We compared these forecasts to those generated by a
-standard Gompertz model with normally distributed process noise fit to the same
-data. We calculated the ratio of the lower 99\% quantile credible interval
-between the two model projections for all populations in which Pr(\(\nu < 10\))
-\(> 0.5\). To ensure the results had stabilized in the tails of the forecast
-posterior, we increased the number of posterior samples in Stan. We ran four
-chains of 20,000 iterations and discarded the first 10,000 as warmup for a
-total of 40,000 samples.
 
@@
 shag cormorants (\emph{Phalacrocorax aristotelis}) on the Farne Islands, United
-Kingdom declined suddenly following a red tide event in 1968 \citep{potts1980}.
+Kingdom declined suddenly following a red tide event in 1968 \cite{potts1980}.
 This freed up quality nest sites for first-time breeders, productivity
 increased, and the population experienced a rapid upswing in
-abundance \citep{potts1980}.
+abundance \cite{potts1980}.
 
@@
 The exact shape of the tails has proven important
-in the field of dispersal biology \citep[e.g.][]{kot1996, clark1999}.
+in the field of dispersal biology \cite{kot1996, clark1999}.
 Answers for the field of population dynamics
@@
 these increase true extinction risk.
+
+\section{Methods}
+
+We fit population dynamics models in which the process noise is
+drawn from a Student-$t$ distribution. By estimating the degrees of freedom
+parameter, \(\nu\), we can estimate the degree to which the process deviations
+have heavy tails and are therefore diagnostic of black-swan events (Fig.~1).
+Lower values of \(\nu\) result in heavy-tailed distributions.
+% For example, at \(\nu = 2\), an event
+% more than 4 standard deviations below the mean is expected
+For example, an event four standard deviations below the mean
+is expected to happen once every 35 years with a $t$ distribution at \(\nu = 2\)
+compared to once every 32,000 years with a normal distribution.
+% more than four standard deviations below the mean is almost 70,000 times higher
+% than the probability of drawing such a value from a normal distribution
+% (\(0.02\) vs.\ \(2.9\cdot10^{-7}\)).
+As \(\nu\) approaches infinity, the t
+distribution approaches the normal distribution (Fig.~1).
+
+We chose the $t$ distribution for a number of reasons.
+Pragmatically, the $t$ distribution requires the estimation of only one additional
+parameter over the normal distribution and reverts to the normal distribution
+if strong evidence of heavy-tailed process deviations is not present.
+Furthermore, the $t$ distribution arises from a mixture of normal distributions
+with the same mean and different variances \cite{gelman2014},
+which mimics the underlying basis for black-swan events.
+There are a wide range of mechanistic processes
+that could give rise to extreme events in population dynamics,
+such as extreme climate events
+\cite{meehl2004, katz2005, ipcc2012}
+or sudden changes in predation from other species.
+While evaluating these various hypotheses
+would be important for diagnosing the underlying causes
+of extreme events and to develop predictive models,
+it would be impossible to tease out each of these
+underlying causes from population trend data alone.
+Instead we seek a simple flexible phenomenological model
+that can be applied across the widest possible
+range of population time series to provide a
+comparative overview of the prevalence
+of black-swan events.
+
+One alternative approach would involve fitting
+the generalized extreme value (GEV) distribution,
+which represents the limit distribution of a series of maxima or minima.
+Although the GEV is well-suited for physical ecological variables
+(e.g.\ sedimentation rates or wind speeds \cite{gaines1993, denny2009}),
+the GEV requires maxima per time block
+or peaks beyond a chosen threshold.
+Therefore, the GEV requires longer time series then typically available
+for wild animal population abundance.
+Furthermore, the GEV approach cannot be easily integrated into population dynamics models.
+
+% This could represent the effect of unmodeled spatial heterogeneity,
+% nonstationary environmental processes, or
+% varying responses to temperature among individuals, for example.
+
+\subsection{Data}
+
+We selected abundance time series from the Global Population Dynamics
+Database \cite[GPDD,][]{gpdd2010}, which contains nearly 5,000 time series of
+abundance from \(\sim\) 1,000 species and \(\sim\) 100 taxonomic orders. We
+filtered the data (Supporting Information) to remove populations from less
+reliable data sources, and those without sufficient data for our models, and
+then interpolated remaining missing values (sensu \cite{brook2006a}).
+Our interpolation affected only \(\sim\)
+\interpPointsPerc \% of the final data points (Table~S1) and none of the data
+points that were later considered black-swan events. Our final dataset
+contained \NPops\ populations across \NOrders\ taxonomic orders and seven
+taxonomic classes, with a median of \medianTimeSteps\ time steps (range
+\minTimeSteps--\maxTimeSteps) (Table~S1).
+
+\subsection{Modeling framework}
+
+We fit heavy-tailed Gompertz population dynamics models to data from the GPDD.
+The Gompertz model represents population growth as a linear function in
+log space (here and throughout we use log
+to refer to the natural logarithm).
+If \(x_{t}\) represents the log abundance (\(N\)) at time \emph{t} then
+\begin{align}
+x_t &= \lambda + bx_{t - 1} + \epsilon_t\\
+\epsilon_t &\sim \mathrm{Studen}t\mhyphen t(\nu, 0, \sigma)
+\end{align}
+The parameter \(\lambda\) represents the expected growth rate if \(N_{t}
+= 1\). The model is density independent if \(b = 1\), maximally density
+dependent if \(b = 0\), and inversely density dependent if \(b < 0\). Usually,
+the process noise \(\epsilon_{t}\) is modelled as normally distributed, but we
+allow it to be drawn from a Student-$t$ distribution with scale parameter
+\(\sigma\) and degrees of freedom \(\nu\). We can then estimate the degree to
+which the process deviations have heavy tails and are therefore evidence of
+black-swan events (Fig.~1A, B).
+
+For the Gompertz model, our weakly-informative priors (Fig.~S\ref{fig:priors}) were:
+\begin{align}
+b &\sim \mathrm{Uniform}(-1, 2)\\
+\lambda &\sim \mathrm{Normal}(0, 10^2)\\
+\sigma &\sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)\\
+\nu &\sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2).
+\end{align}
+Our prior on \(b\) was uninformative between values of \(-1\) and \(2\). We
+would not expect values of \(b\) with levels of density dependence as low as
+\(-1\) (very strong inverse density dependence), nor would we generally expect
+values above \(1\). We allowed values of \(b\) above \(1\) to allow for
+non-stationary time series of growth rates. The estimates of \(b\) were well
+within these bounds. Our prior on \(\lambda\) was weakly informative within the
+range of expected values for population growth and is similar to the default
+priors suggested by Ref.~\cite{gelman2008d} for intercepts of
+regression models. Our half-Cauchy prior on \(\sigma\) follows
+Ref.~\cite{gelman2006c} and Ref.~\cite{gelman2008d} and the specific scale parameter of
+\(2.5\) is based on our expected range of the value in nature from previous
+studies \cite{connors2014}. In our testing of a subsample of populations, our
+parameter estimates were not qualitatively changed by switching to a uniform
+prior on \(\sigma\), but the weakly informative half-Cauchy prior sped up chain
+convergence.
+
+We chose a weakly informative exponential prior on \(\nu\) with rate parameter
+of \(0.01\) truncated at values above two---a slightly less informative prior
+than suggested by Ref.~\cite{fernandez1998}. We truncated the
+distribution at \(2\), since at \(\nu < 2\) the variance of the $t$ distribution
+is undefined. This prior gives only a \(7.7\)\% probability that \(\nu < 10\)
+but constrains the sampling sufficiently to avoid \(\nu\) drifting towards infinity. In
+any case, for \(\nu > 20\) the $t$ distribution is almost indistinguishable from
+the normal distribution (Fig.~1). Based on the shape of the $t$ distribution, we
+chose the probability that \(\nu < 10\), Pr(\(\nu < 10\)), to define the
+probability of heavy-tailed (black-swan) dynamics. When categorizing a population as heavy
+tailed, we applied a 0.5 probability threshold. In the scenario where the data
+are uninformative about heavy tails (e.g.~Fig.~1E, H), the posterior for
+\(\nu\) will approximately match the prior (prior median \(= 71\), mean \(=
+102\)) and a metric of Pr\((\nu < 10) > 0.5\) will not flag the
+population as heavy tailed.
+
+We fit our models with Stan 2.6.0 \cite{stan-manual2015, hoffman2014}, and R
+3.2.0 \cite{r2015}. We began with four chains and 2,000 iterations, discarding
+the first 1,000 as warm up (i.e.~4,000 final samples). If \(\hat{R}\) (the
+potential scale reduction factor---a measure of chain convergence) was greater
+than \(1.05\) for any parameter, or if the minimum effective sample size,
+\(n_\mathrm{eff}\) (a measure of the effective number of uncorrelated samples),
+for any parameter was less than \(200\), we doubled both the total iterations
+and warm up period and sampled from the model again. These thresholds are in
+excess of the minimums recommended by Ref.~\cite{gelman2006a} of
+\(\hat{R} < 1.1\) and \(n_\mathrm{eff} > 100\) for reliable point estimates and
+confidence intervals. In the majority of cases our minimum thresholds were
+greatly exceeded. We continued this procedure up to 8,000 iterations (16,000
+total samples) by which all chains had sufficiently converged.
+The No-U-Turn Hamiltonian Markov chain Monte Carlo sampler in Stan generally
+requires fewer iterations to obtain equivalent effective sample sizes than the
+Gibbs or Metropolis--Hastings algorithms \cite{stan-manual2015}.
+
+\subsection{Alternative priors}
+
+To test if the prior on \(\nu\) influenced our estimate of black-swan dynamics,
+we refit our models with weaker and stronger priors. Our base model used a
+prior on \(\nu\) of Truncated-Exponential(0.01, min.\ = 2). For a weaker prior
+we used Truncated-Exponential(0.005, min.\ = 2) and for a stronger prior we
+used Truncated-Exponential(0.02, min.\ = 2) (Fig.~S\ref{fig:priors}c). Note that
+the base and weaker priors are relatively flat within the region of \(\nu <
+20\), which is the region we are concerned about when categorizing
+populations as heavy tailed. We also considered using a Uniform(\(0, 1\))
+prior on \(1/\nu\) as used in Ref.~\cite{gelman2014}, but this
+prior puts considerably more weight on lower (heavy-tailed) values of \(\nu\)
+than the truncated exponential priors considered here,
+and the results are therefore not shown.
+
+\subsection{Alternative population models} We fit alternative population models to
+test if four key phenomena changed our conclusions.
+
+\subsubsection{Autocorrelated residuals} We considered a version of the Gompertz
+model in which an autoregressive parameter was fit to the process-noise
+residuals:
+\begin{align}
+x_t &= \lambda + b x_{t-1} + \epsilon_t\\
+\epsilon_t &\sim \mathrm{Student}\mhyphen t(\nu, \phi \epsilon_{t-1}, \sigma).
+\end{align}
+In addition to the parameters in the original Gompertz model, this model
+estimates an additional parameter \(\phi\), which represents the correlation of
+subsequent residuals. Based on the results of previous analyses with the
+GPDD \cite{connors2014} and the chosen prior in previous
+analyses \cite{thorson2014a}, we placed a moderately informative prior on
+\(\phi\) that assumed the greatest probability density near zero with the
+reduced possibility of \(\phi\) being near \(-1\) or \(1\). Specifically, we
+chose \(\phi \sim \mathrm{truncated\mhyphen Normal}(0, 0.5, \mathrm{min.} = -1,
+\mathrm{max.} = 1)\).
+
+\subsubsection{Density independence} We fit a simplified version of the
+Gompertz model in which the density dependence parameter \(b\) was fixed at
+\(1\) (density independent). This is equivalent to fitting a random walk model
+(with drift) to the \(\log\) abundances or assuming the growth rates are drawn
+from a stationary distribution. The model was as follows:
+\begin{align}
+x_t &= \lambda + x_{t-1} + \epsilon_t\\
+\epsilon &\sim \mathrm{Student}\mhyphen t(\nu, 0, \sigma).
+\end{align}
+We fit this model for three reasons: (1) it is computationally simpler and so
+provides a check that our more complicated full Gompertz model was obtaining
+reasonable estimates of \(\nu\), (2) it provides a test of whether density
+dependence was affecting our perception of heavy tails, (3) it
+matches how some previous authors have modelled heavy tails without accounting
+for density dependence \cite{segura2013}.
+
+\subsubsection{Observation error} Observation error can bias parameter
+estimates \cite{knape2012} and is known to affect the ability to detect extreme
+events \cite{ward2007}. In our main analysis, we fit a model that ignored
+observation error. One way to account for observation error would be to fit a
+full state-space model that simultaneously estimates the magnitude of process
+noise and observation error. However, simultaneously estimating observation and
+process noise is a challenging problem (e.g.\ because the observation and
+process noise parameters tend to negatively covary in model fitting) and is
+known to sometimes result in identifiability issues with the Gompertz
+population model \cite{knape2008}. Furthermore, our model was applied to
+hundreds of time series, often of short length (as few as \minTimeSteps\ time
+steps) and our model estimates an additional parameter---the shape of the
+process deviation tails---potentially making identifiability and computational
+issues even greater. Therefore, we considered a version of the base Gompertz
+model that allowed for a fixed level of observation error:
+\begin{align}
+U_t &= \lambda + b U_{t-1} + \epsilon_t\\
+x_t &\sim \mathrm{Normal}(U_t, \sigma_\mathrm{obs}^2)\\
+\epsilon_t &\sim \mathrm{Student}\mhyphen t(\nu, 0, \sigma_\mathrm{proc}),
+\end{align}
+where \(U\) represents the served state vector, \(\sigma_\mathrm{obs}\)
+represents the standard deviation of observation error (on a log scale), and
+\(\sigma_\mathrm{proc}\) represent the process noise scale parameter. We set
+\(\sigma_\mathrm{obs}\) to \(0.2\), which represents a typical value used in
+simulation analyses \cite{valpine2002, thorson2014b}.
+
+\subsubsection{Ricker-logistic} We also fit a Ricker-logistic model:
+\begin{align}
+x_t &= x_{t-1} + r_{\mathrm{max}}\left(1 - \frac{N_{t-1}}{K}\right) + \epsilon_t\\
+\epsilon_t &\sim \mathrm{Student}\mhyphen t(\nu, 0, \sigma),
+\end{align}
+where \(r_\mathrm{max}\) represents the theoretical maximum population growth
+rate that is obtained when \(N_t = 0\). The parameter \(K\) represents the
+carrying capacity and, as before, \(x_t\) represents the \(\log\) transformed
+abundance at time \(t\). The Ricker-logistic model assumes a linear decrease in
+population growth rate with increases in abundance. In contrast, the Gompertz
+model assumes a linear decrease in population growth rate with increases in
+\textit{log} abundance.
+
+To fit the Ricker-logistic models, we chose a prior on \(K\) uniform between
+zero and twice the maximum observed abundance (similar and less informative
+than in Ref.~\cite{delean2013}). We set the prior on
+\(r_\mathrm{max}\) as uniform between 0 and 20 as in Ref.~\cite{delean2013}. We used the same priors on \(\nu\) and \(\sigma\) as in
+the Gompertz model.
+
+\subsection{Simulation testing}
+We performed two types of simulation testing.
+First, we tested how easily the Student-$t$ distribution \(\nu\) parameter could
+be recovered given different true values of \(\nu\) and different sample sizes.
+Second, we tested the ability of the heavy-tailed Gompertz model to obtain
+unbiased parameter estimates of \(\nu\) given that a set of process deviations
+was provided in which the effective \(\nu\) value was close to the true \(\nu\)
+value.
+
+We found that we could consistently and precisely recover median posterior
+estimates of \(\nu\) near the true value of \(\nu\) with large samples (\(\ge
+200\)) (Fig.~S\ref{fig:sim-nu} upper panels). At smaller samples we could
+usually qualitatively distinguish heavy from not-heavy tails, but the model
+tended to underestimate how heavy the tails were. At the same time, at smaller
+sample sizes, the model tended to overestimate how large the scale parameter
+was (Fig.~S\ref{fig:sim-nu} lower panels).
+
+Our simulation-testing results indicate that the Gompertz model can recapture
+the true value of \(\nu\) when the process noise was chosen so that appropriate
+tail events were present (Fig.~S\ref{fig:sim-prob} upper panels). The addition
+of observation error caused the model to tend to underestimate the degree of
+heavy-tailedness. Fitting a model with correctly specified observation error
+did not substantially reduce model bias (Fig.~S\ref{fig:sim-prob}).
+
+\subsection{Covariates of heavy-tailed dynamics} We fit a hierarchical beta
+regression model to the predicted probability of heavy tails, Pr(\(\nu < 10\)),
+to investigate potential covariates of heavy-tailed dynamics. We obtained
+maximum lifespan and body-size data from Ref.~\cite{brook2006a}.
+Beta regression is useful when response data are on a continuous scale
+between zero and one \cite{ferrari2004}. The model was as follows:
+\begin{align}
+\mathrm{Pr}(\nu_i < 10) &\sim \mathrm{Beta}(A_i, B_i)\\
+\mu_i &= \mathrm{logit}^{-1}(\alpha
+  + \alpha^\mathrm{class}_{j[i]}
+  + \alpha^\mathrm{order}_{k[i]}
+  + \alpha^\mathrm{species}_{l[i]}
+  + \bm{X_i} \bm{\beta}),
+  \: \text{for } i = 1, \dots, 609\\
+A_i &= \phi_\mathrm{disp} \mu_i\\
+B_i &= \phi_\mathrm{disp} (1 - \mu_i)\\
+\alpha^\mathrm{class}_j &\sim
+  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{class}}),
+  \: \text{for } j = 1, \dots, 6\\
+\alpha^\mathrm{order}_k &\sim
+  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{order}}),
+  \: \text{for } k = 1, \dots, 38\\
+\alpha^\mathrm{species}_l &\sim
+  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{species}}),
+  \: \text{for } l = 1, \dots, 301,
+\end{align}
+where \(A\) and \(B\) represent the beta distribution shape parameters;
+\(\mu_i\) represents the predicted value for population \(i\), class \(j\),
+order \(k\), and species \(l\); \(\phi_\mathrm{disp}\) represents the
+dispersion parameter; and \(\bm{X_i}\) represents a vector of predictors (such as
+lifespan) for population \(i\) with associated \(\bm{\beta}\) coefficients. The
+intercepts are allowed to vary from the overall intercept \(\alpha\) by
+taxonomic class (\(\alpha^\mathrm{class}_j\)), taxonomic order
+(\(\alpha^\mathrm{order}_k\)), and species (\(\alpha^\mathrm{species}_l\)) with
+standard deviations \(\sigma_{\alpha \; \mathrm{class}}\), \(\sigma_{\alpha \;
+\mathrm{order}}\), and \(\sigma_{\alpha \; \mathrm{species}}\). Where possible,
+we also allowed for error distributions around the predictors by incorporating
+the standard deviation of the posterior samples for the Gompertz parameters
+\(\lambda\), \(b\), and \(\log \sigma\) around the mean point value as normal
+distributions (not shown in the above equation).
+
+We log transformed \(\sigma\), time-series length, and lifespan to match the
+way they are visually represented in Fig.~S\ref{fig:correlates} and to make the
+relationship approximately linear on the logit-transformed response scale. All
+input variables were standardized by subtracting their mean and dividing by two
+standard deviations to make their coefficients comparable in
+magnitude \cite{gelman2008c}. We excluded body length as a covariate because it
+was highly correlated with lifespan, and lifespan exhibited more overlap across
+taxonomy than body length. Lifespan is also more directly related to time and
+potential mechanisms driving black-swan dynamics.
+
+We incorporated weakly informative priors into our model: \(\mathrm{Cauchy}(0,
+10)\) on the global intercept \(\alpha\), \(\mathrm{half\mhyphen Cauchy}(0,
+2.5)\) on all standard deviation parameters, \(\mathrm{half\mhyphen Cauchy}(0,
+10)\) on the dispersion parameter \(\phi_\mathrm{disp}\), and
+\(\mathrm{Cauchy}(0, 2.5)\) on all other parameters \cite{gelman2006c,
+gelman2008d}. Compared to normal priors, the Cauchy priors concentrate more
+probability density around expected parameter values while allowing for a
+higher probability density far into the tails, thereby allowing the data to
+dominate the posterior more strongly if it disagrees with the prior. We fit our
+models with 5,000 total iterations per chain, 2500 warm-up iterations, four
+chains, and discarding every second sample. We checked for chain convergence
+visually and with the same criteria as before (\(\widehat{R} < 1.05\) and
+\(n_\mathrm{eff} >200\) for all parameters). Furthermore, we checked our model
+against a similar model with the glmmADMB \textsf{R} package \cite{glmmadmb}
+without uncertainty around the predictors.
+
+To derive taxonomic-order-level estimates of the probability of heavy tails
+accounting for time-series length (Fig.~3A), we fit a separate hierarchical
+model with the same structure but with only \(\log\) time-series length as a
+predictor---in this case, we did not want to control for intrinsic population
+characteristics such as density dependence. Since our predictors were centered
+by subtracting their mean value, we obtained order-level estimates of the
+probability of heavy tails at mean log time-series length by adding the
+posteriors for \(\alpha\), \(\alpha^\mathrm{class}_j\), and
+\(\alpha^\mathrm{order}_k\).
+
+\subsection{Skewed Student-$t$ forecasts} To evaluate the apparent skewness
+of heavy-tailed process noise we fit Gompertz models with skewed Student-$t$
+distributed process noise \cite{fernandez1998}. This distribution adds one
+parameter to the Student-$t$ distribution, \(\gamma\), which controls the
+skewness. The distribution is symmetrical if \(\gamma = 1\), left skewed if \(0
+< \gamma < 1\), and right skewed if \(1 < \gamma < \infty\). We placed a weakly
+informative prior of \(\mathrm{Cauchy}(0, 2.5)\) on \(\log \gamma\). We
+aggregated 5,000 randomly selected posterior samples from the \(\gamma\)
+parameter of each model at three levels of evidence for heavy-tails:
+\emph{heavy tailed}, median \(\nu < 10\); \emph{slightly heavy tailed}, \(10 \le
+\mathrm{median}\, \nu \le 70\); and \emph{normal tailed}, median \(\nu > 70\)
+(Fig.~4).
+
+To generate five-year forecasts of abundance, we combined the posterior
+parameter samples from the skewed-t Gompertz models with stochastically
+generated process noise. We compared these forecasts to those generated by a
+standard Gompertz model with normally distributed process noise fit to the same
+data. We calculated the ratio of the lower 99\% quantile credible interval
+between the two model projections for all populations in which Pr(\(\nu < 10\))
+\(> 0.5\). To ensure the results had stabilized in the tails of the forecast
+posterior, we increased the number of posterior samples in Stan. We ran four
+chains of 20,000 iterations and discarded the first 10,000 as warmup for a
+total of 40,000 samples.
 
@@
 
-\bibliographystyle{ecologyletters}\bibliography{jshort,ms}
+\bibliographystyle{pnas2011}\bibliography{jshort,ms}
 

### 2016-07-05 text/ms_pnas.tex
Copyediting of pnas ms

@@
 \affil[a]{Earth to Ocean Research Group, Department of Biological
-Sciences, Simon Fraser University, Burnaby BC, V5A 1S6, Canada, e-mail:
-sandrsn@uw.edu}
+Sciences, Simon Fraser University, Burnaby BC, V5A 1S6, Canada}
 \affil[b]{School of Aquatic and Fishery Sciences, Box 355020,
@@
 predominantly downward, implying that unexpected population crashes occur more
-frequently than increases. Black-swan events are not driven by life history (e.g. lifespan),
+frequently than increases. Black-swan events are not driven by life history (e.g.\ lifespan),
 but by external causes such as extreme winters and disease. Ignoring the presence of
@@
 \authordeclaration{Please declare any conflict of interest here.}
-\correspondingauthor{\textsuperscript{2}To whom correspondence should be addressed. E-mail: sandrsn\@uw.edu}
+\correspondingauthor{\textsuperscript{2}To whom correspondence should be addressed. E-mail: sandrsn@uw.edu}
 
@@
 Black swans are improbable events that nonetheless
-occur---often with profound consequences. While physical extremes, such as
-monsoons and heat waves, are widely studied and increasing in magnitude and
-frequency, it remains unclear the extent to which ecological population numbers
-buffer or suffer from such extremes. Here we estimate the frequency and
-direction of black-swan events in \NPops\ animal populations, finding strong
-evidence for their occurrence in \(\sim\)\overallBasePerc \% of populations.
-Black-swan events occur most frequently for birds (\birdPH \%), mammals
+occur---often with profound consequences. 
+% While physical extremes, such as
+% monsoons and heat waves, are widely studied and increasing in magnitude and frequency, 
+Such events drive important transitions in social systems (e.g.\ banking
+collapses) and physical systems (e.g.\ earthquakes)
+yet it remains unclear the extent to which ecological population numbers
+buffer or suffer from such extremes. 
+Here we estimate the prevalence and
+direction of black-swan events (heavy-tailed process noise)
+in \NPops\ animal populations
+after accounting for population dynamics (productivity, 
+density dependence, and typical stochasticity).
+We find strong evidence for 
+black-swan events in \(\sim\)\overallBasePerc \% of populations.
+These events occur most frequently for birds (\birdPH \%), mammals
 (\mammalsPH \%), and insects (\insectsPH \%),
-manifest primarily as
+and are not explained by any life-history covariates, but
+tend to be driven by external perturbations such as climate, severe winters,
+predators, parasites, or the combined effects of multiple factors.
+Black-swan events manifest primarily as
 population die-offs and crashes (\percBSDown \%) rather than unexpected increases,
-and tend to be driven by extrinsic perturbations
-such as climate, severe winters, or disease.
-Thus, ignoring black-swan events will
-underestimate extinction risk.
+and ignoring heavy-tailed process noise leads to 
+% a \crashUnderMedian-fold (50\% interval \crashUnderRange) 
+an underestimate in the magnitude of population crashes.
+We suggest modelers consider heavy-tailed downward-skewed 
+probability distributions, 
+such as the skewed Student-$t$ used here, 
+when making forecasts of population abundance.
 Our results demonstrate the importance of both
@@
 
+\begin{figure*}[htb]
+\centering
+\includegraphics[width=0.65\textwidth]{../analysis/t-nu-eg2.pdf}
+\caption{ Illustration of population dynamic models that allow for heavy
+tails. (a, b) Probability density for the Student-$t$
+distribution with scale parameter of $1$ and different values of \(\nu\). Small
+values of \(\nu\) create heavy tails while as \(\nu\) approaches infinity the
+distribution approaches the normal distribution. (c to e)
+Simulated population dynamics from a Gompertz model with process noise drawn
+from Student-$t$ distributions with three values of \(\nu\). Coloured dots in
+panels c and d represent jumps with less than a 1 in 1000
+chance of occurring in a normal distribution. (f to h)
+Estimates of \(\nu\) from models fit to the times series in panels c to
+e. Shown are posterior samples (histograms), median and interquartile
+range of the posterior (IQR, dots and line segments), and the exponential prior
+on \(\nu\) (dashed lines). Colour shading behind panels f to h
+illustrates the approximate region of heavy tails.}
+\label{fig:1}
+\end{figure*}
+
 Here, we develop a new approach to estimate the frequency and magnitude of
@@
 
-\begin{figure*}[htb]
-\centering
-\includegraphics[width=0.65\textwidth]{../analysis/t-nu-eg2.pdf}
-\caption{ Illustration of population dynamic models that allow for heavy
-tails. (a, b) Probability density for the Student-$t$
-distribution with scale parameter of $1$ and different values of \(\nu\). Small
-values of \(\nu\) create heavy tails while as \(\nu\) approaches infinity the
-distribution approaches the normal distribution. (c to e)
-Simulated population dynamics from a Gompertz model with process noise drawn
-from Student-$t$ distributions with three values of \(\nu\). Coloured dots in
-panels c and d represent jumps with less than a 1 in 1000
-chance of occurring in a normal distribution. (f to h)
-Estimates of \(\nu\) from models fit to the times series in panels c to
-e. Shown are posterior samples (histograms), median and interquartile
-range of the posterior (IQR, dots and line segments), and the exponential prior
-on \(\nu\) (dashed lines). Colour shading behind panels f to h
-illustrates the approximate region of heavy tails.}
-\label{fig:1}
-\end{figure*}
-
 \section*{Results}
 
-We found evidence that black-swan dynamics are infrequent but highly
+We found that black-swan dynamics are infrequent but highly
 influential: in our dataset they were detected most
@@
 \centering
-\includegraphics[width=0.3\textwidth]{../analysis/order-posteriors-covariates.pdf}
+\includegraphics[width=0.26\textwidth]{../analysis/order-posteriors-covariates.pdf}
 \caption{Standardized probabilities and covariates of heavy-tailed
@@
 Given the prevalence of downwards events, we refit our heavy-tailed models to
-measure the degree of skewness \(\gamma\) of the process noise using a skew-t
+measure the degree of skewness \(\gamma\) of the process noise using a skew-$t$
 distribution (Fig.~4a), and used these models to make near-term risk forecasts.
@@
 Barring a better estimate,
-we suggest using $\nu = 4$; and
+we suggest using $\nu = 5$; and
 a skewness parameter ($\gamma$) of $\sim 0.7$ (Fig.~4a)
@@
 the normal distribution,
-and $\nu  = 4$ allows for an event
+and $\nu  = 5$ allows for an event
 three standard deviations below the mean to happen
-once every 50 years instead of once every 740 years.
+% 1/(1-pt(3, 5))
+once every 66 years instead of once every 740 years.
 Adopting heavy-tailed process noise by default follows the intuition
@@
 x_t &= \lambda + bx_{t - 1} + \epsilon_t\\
-\epsilon_t &\sim \mathrm{Studen}t\mhyphen t(\nu, 0, \sigma)
+  \epsilon_t &\sim \mathrm{Student}\mhyphen t(\nu, 0, \sigma).
 \end{align}
@@
 parameter over the normal distribution and reverts to the normal distribution
-if strong evidence of heavy-tailed process deviations is not present.
+if strong evidence of heavy-tailed process deviations is absent.
 Furthermore, the $t$ distribution arises from a mixture of normal distributions
@@
 
-For the Gompertz model, our weakly-informative priors (Fig.~S\figpriors) were:
-\begin{align}
-b &\sim \mathrm{Uniform}(-1, 2)\\
-\lambda &\sim \mathrm{Normal}(0, 10^2)\\
-\sigma &\sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)\\
-\nu &\sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2).
-\end{align}
-Our prior on \(b\) was uninformative between values of \(-1\) and \(2\). We
-would not expect values of \(b\) with levels of density dependence as low as
-\(-1\) (very strong inverse density dependence), nor would we generally expect
-values above \(1\). We allowed values of \(b\) above \(1\) to allow for
-non-stationary time series of growth rates. The estimates of \(b\) were well
-within these bounds. Our prior on \(\lambda\) was weakly informative within the
-range of expected values for population growth and is similar to the default
-priors suggested by Ref.~\cite{gelman2008d} for intercepts of
-regression models. Our half-Cauchy prior on \(\sigma\) follows
-Ref.~\cite{gelman2006c} and Ref.~\cite{gelman2008d} and the specific scale parameter of
-\(2.5\) is based on our expected range of the value in nature from previous
-studies \cite{connors2014}. In our testing of a subsample of populations, our
-parameter estimates were not qualitatively changed by switching to a uniform
-prior on \(\sigma\), but the weakly informative half-Cauchy prior sped up chain
-convergence.
-
-We chose a weakly informative exponential prior on \(\nu\) with rate parameter
+For the Gompertz model,
+we chose weakly informative priors for 
+all parameters (Fig.~S\figpriors, Supporting Information).
+For the degrees of freedom parameter $\nu$, 
+we chose an exponential prior 
+with a rate parameter
 of \(0.01\) truncated at values above two---a slightly less informative prior
@@
 for any parameter was less than \(200\), we doubled both the total iterations
-and warm up period and sampled from the model again. These thresholds are in
-excess of the minimums recommended by Ref.~\cite{gelman2006a} of
-\(\hat{R} < 1.1\) and \(n_\mathrm{eff} > 100\) for reliable point estimates and
-confidence intervals. In the majority of cases our minimum thresholds were
+and warm up period and sampled from the model again. 
+% These thresholds are in
+% excess of the minimums recommended by Ref.~\cite{gelman2006a} of
+% \(\hat{R} < 1.1\) and \(n_\mathrm{eff} > 100\) for reliable point estimates and
+% confidence intervals.
+In the majority of cases our minimum thresholds were
 greatly exceeded. We continued this procedure up to 8,000 iterations (16,000
@@
 also fit an alternative model, the Ricker-logistic model, which assumes that
-population growth rate declines linearly with abundance itself (Supporting Information).
+population growth rate declines linearly with abundance (Supporting Information).
 
-In addition to alternative population models, we investigated the sensitivity
+We investigated the sensitivity
 of our results to weaker and stronger priors (exponential rate parameter $=
-0.005, 0.02$; Fig.~S\figpriors, Supporting Information). We used
+0.005, 0.02$; Fig.~S\figpriors, Supporting Information). Furthermore, we used
 simulated data to test how easily we could detect $\nu$ given different sample
@@
 between zero and one \cite{ferrari2004}. The model was as follows:
-\begin{align}
+\begin{align*}
 \mathrm{Pr}(\nu_i < 10) &\sim \mathrm{Beta}(A_i, B_i)\\
@@
   + \bm{X_i} \bm{\beta}),
-  \: \text{for } i = 1, \dots, 609\\
+  \: \\
+  &\quad \, \text{for } i = 1, \dots, 609\\
 A_i &= \phi_\mathrm{disp} \mu_i\\
@@
   \: \text{for } l = 1, \dots, 301,
-\end{align}
+\end{align*}
 where \(A\) and \(B\) represent the beta distribution shape parameters;
@@
 
-We log transformed \(\sigma\), time-series length, and lifespan to match the
-way they are visually represented in Fig.~S\figcorrelates\ and to make the
-relationship approximately linear on the logit-transformed response scale. All
-input variables were standardized by subtracting their mean and dividing by two
+We log transformed \(\sigma\), time-series length, and lifespan to 
+to make the
+relationship approximately linear on the logit-transformed response scale;
+this matches the way the predictors are visually presented in 
+Fig.~S\figcorrelates. 
+All input variables were standardized by subtracting their mean and dividing by two
 standard deviations to make their coefficients comparable in
@@
 \(\mathrm{Cauchy}(0, 2.5)\) on all other parameters \cite{gelman2006c,
-gelman2008d}. Compared to normal priors, the Cauchy priors concentrate more
-probability density around expected parameter values while allowing for a
-higher probability density far into the tails, thereby allowing the data to
-dominate the posterior more strongly if it disagrees with the prior. We fit our
+gelman2008d}. 
+% Compared to normal priors, the Cauchy priors concentrate more
+% probability density around expected parameter values while allowing for a
+% higher probability density far into the tails, thereby allowing the data to
+% dominate the posterior more strongly if it disagrees with the prior. 
+We fit our
 models with 5,000 total iterations per chain, 2500 warm-up iterations, four

### 2016-07-05 text/supp.tex
Copy editing and rearranging of supp for pnas

@@
 
-\subsubsection{Autocorrelated residuals} We considered a version of the Gompertz
+\section{Prior distributions for Gompertz models}
+
+For the Gompertz model, our weakly informative priors (Fig.~\ref{fig:priors}) were:
+\begin{align}
+b &\sim \mathrm{Uniform}(-1, 2)\\
+\lambda &\sim \mathrm{Normal}(0, 10^2)\\
+\sigma &\sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)\\
+\nu &\sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2).
+\end{align}
+Our prior on \(b\) was uninformative between values of \(-1\) and \(2\). We
+would not expect values of \(b\) with levels of density dependence as low as
+\(-1\) (very strong inverse density dependence), nor would we generally expect
+values above \(1\). We allowed values of \(b\) above \(1\) to allow for
+non-stationary time series of growth rates. The estimates of \(b\) were well
+within these bounds. Our prior on \(\lambda\) was weakly informative within the
+range of expected values for population growth and is similar to the default
+priors suggested by \citet{gelman2008d} for intercepts of
+regression models. Our half-Cauchy prior on \(\sigma\) follows
+\citet{gelman2006c} and \citet{gelman2008d} and the specific scale parameter of
+\(2.5\) is based on our expected range of the value in nature from previous
+studies \citep{connors2014}. In our testing of a subsample of populations, our
+parameter estimates were not qualitatively changed by switching to a uniform
+prior on \(\sigma\), but the weakly informative half-Cauchy prior sped up chain
+convergence.
+
+To test if the prior on \(\nu\) influenced our estimate of black-swan dynamics,
+we refit our models with weaker and stronger priors. Our base model used a
+prior on \(\nu\) of Truncated-Exponential(0.01, min.\ = 2). For a weaker prior
+we used Truncated-Exponential(0.005, min.\ = 2) and for a stronger prior we
+used Truncated-Exponential(0.02, min.\ = 2) (Fig.~\ref{fig:priors}c). Note that
+the base and weaker priors are relatively flat within the region of \(\nu <
+20\), which is the region we are concerned about when categorizing
+populations as heavy tailed. We also considered using a Uniform(\(0, 1\))
+prior on \(1/\nu\) as used in \citet{gelman2014}, but this
+prior puts considerably more weight on lower (heavy-tailed) values of \(\nu\)
+than the truncated exponential priors considered here,
+and the results are therefore not shown.
+
+\section{Alternative population models}
+
+\subsection{Autocorrelated residuals} We considered a version of the Gompertz
 model in which an autoregressive parameter was fit to the process-noise
@@
 subsequent residuals. Based on the results of previous analyses with the
-GPDD \cite{connors2014} and the chosen prior in previous
-analyses \cite{thorson2014a}, we placed a moderately informative prior on
+GPDD \citep{connors2014} and the chosen prior in previous
+analyses \citep{thorson2014a}, we placed a moderately informative prior on
 \(\phi\) that assumed the greatest probability density near zero with the
@@
 
-\subsubsection{Density independence} We fit a simplified version of the
+\subsection{Density independence} We fit a simplified version of the
 Gompertz model in which the density dependence parameter \(b\) was fixed at
@@
 matches how some previous authors have modelled heavy tails without accounting
-for density dependence \cite{segura2013}.
+for density dependence \citep{segura2013}.
 
-\subsubsection{Observation error} Observation error can bias parameter
-estimates \cite{knape2012} and is known to affect the ability to detect extreme
-events \cite{ward2007}. In our main analysis, we fit a model that ignored
+\subsection{Observation error} Observation error can bias parameter
+estimates \citep{knape2012} and is known to affect the ability to detect extreme
+events \citep{ward2007}. In our main analysis, we fit a model that ignored
 observation error. One way to account for observation error would be to fit a
@@
 known to sometimes result in identifiability issues with the Gompertz
-population model \cite{knape2008}. Furthermore, our model was applied to
+population model \citep{knape2008}. Furthermore, our model was applied to
 hundreds of time series, often of short length (as few as \minTimeSteps\ time
@@
 \(\sigma_\mathrm{obs}\) to \(0.2\), which represents a typical value used in
-simulation analyses \cite{valpine2002, thorson2014b}.
+simulation analyses \citep{valpine2002, thorson2014b}.
 
-\subsubsection{Ricker-logistic} We also fit a Ricker-logistic model:
+\subsection{Ricker-logistic} We also fit a Ricker-logistic model:
 \begin{align}
@@
 zero and twice the maximum observed abundance (similar and less informative
-than in Ref.~\cite{delean2013}). We set the prior on
-\(r_\mathrm{max}\) as uniform between 0 and 20 as in Ref.~\cite{delean2013}. We used the same priors on \(\nu\) and \(\sigma\) as in
+than in \citet{delean2013}). We set the prior on
+\(r_\mathrm{max}\) as uniform between 0 and 20 as in \citet{delean2013}. 
+We used the same priors on \(\nu\) and \(\sigma\) as in
 the Gompertz model.
 
-\subsection{Alternative priors}
+\section{Heavy-tailed Gompertz model simulations}
 
-To test if the prior on \(\nu\) influenced our estimate of black-swan dynamics,
-we refit our models with weaker and stronger priors. Our base model used a
-prior on \(\nu\) of Truncated-Exponential(0.01, min.\ = 2). For a weaker prior
-we used Truncated-Exponential(0.005, min.\ = 2) and for a stronger prior we
-used Truncated-Exponential(0.02, min.\ = 2) (Fig.~S\ref{fig:priors}c). Note that
-the base and weaker priors are relatively flat within the region of \(\nu <
-20\), which is the region we are concerned about when categorizing
-populations as heavy tailed. We also considered using a Uniform(\(0, 1\))
-prior on \(1/\nu\) as used in Ref.~\cite{gelman2014}, but this
-prior puts considerably more weight on lower (heavy-tailed) values of \(\nu\)
-than the truncated exponential priors considered here,
-and the results are therefore not shown.
-
-\subsection{Simulation testing}
 We performed two types of simulation testing.
@@
 value.
-
-We found that we could consistently and precisely recover median posterior
-estimates of \(\nu\) near the true value of \(\nu\) with large samples (\(\ge
-200\)) (Fig.~S\ref{fig:sim-nu} upper panels). At smaller samples we could
-usually qualitatively distinguish heavy from not-heavy tails, but the model
-tended to underestimate how heavy the tails were. At the same time, at smaller
-sample sizes, the model tended to overestimate how large the scale parameter
-was (Fig.~S\ref{fig:sim-nu} lower panels).
-
-Our simulation-testing results indicate that the Gompertz model can recapture
-the true value of \(\nu\) when the process noise was chosen so that appropriate
-tail events were present (Fig.~S\ref{fig:sim-prob} upper panels). The addition
-of observation error caused the model to tend to underestimate the degree of
-heavy-tailedness. Fitting a model with correctly specified observation error
-did not substantially reduce model bias (Fig.~S\ref{fig:sim-prob}).
-
-\section{Heavy-tailed Gompertz model simulations}
 
@@
 First, we tested the ability to estimate $\nu$ given different true values of
-$\nu$ and different sample sizes. We took stochastic draws from t
+$\nu$ and different sample sizes. We took stochastic draws from $t$
 distributions with different $\nu$ values ($\nu = 3, 5, 10,$ and $10^6$
@@
 we recorded the posterior samples of $\nu$.
+
+We found that we could consistently and precisely recover median posterior
+estimates of \(\nu\) near the true value of \(\nu\) with large samples (\(\ge
+200\)) (Fig.~\ref{fig:sim-nu} upper panels). At smaller samples we could
+usually qualitatively distinguish heavy from not-heavy tails, but the model
+tended to underestimate how heavy the tails were. At the same time, at smaller
+sample sizes, the model tended to overestimate how large the scale parameter
+was (Fig.~\ref{fig:sim-nu} lower panels).
 
@@
 
+We found that the Gompertz model can recapture
+the true value of \(\nu\) when the process noise was chosen so that appropriate
+tail events were present (Fig.~\ref{fig:sim-prob} upper panels). The addition
+of observation error caused the model to tend to underestimate the degree of
+heavy-tailedness. Fitting a model with correctly specified observation error
+did not substantially reduce model bias (Fig.~\ref{fig:sim-prob} upper panels).
+
 When converting the posterior distributions of $\nu$ into Pr($\nu < 10$), the
@@
        \frac{2}{(\gamma + 1/\gamma)}
-         \mathrm{Student}\mhyphen t(x / \gamma, \nu, \mu / \gamma, \sigma),& \text{for } x \ge 0
+         \mathrm{Student}\mhyphen t(x / \gamma, \nu, \mu / \gamma, \sigma),& \text{for } x \ge 0.
 \end{dcases}
@@
   \log (2 \gamma) - \log(\gamma^2 + 1) +
-  \log \mathrm{Student}\mhyphen t(x / \gamma, \nu, \mu / \gamma, \sigma),& \text{for } x \ge 0
+  \log \mathrm{Student}\mhyphen t(x / \gamma, \nu, \mu / \gamma, \sigma),& \text{for } x \ge 0.
 \end{dcases}
@@
 %\clearpage
-\section{Additional acknowledgements}
+\section{Silhouette image licenses}
 
 Many of the silhouette images used in Figs 2, and 3 were obtained from
-\texttt{phylopic.org} under Creative Commons licenses. We vectorized the
+\texttt{phylopic.org} under Creative Commons licenses. We drew the
 salmon in Fig.~2. The bird in these figures was obtained
@@
 Pr($\nu<10$) $>0.5$. Numbers above panels indicate GPDD population IDs as
-shown in table~S2. Horizontal axes indicates year before last
+shown in Table~S2. Horizontal axes indicates year before last
 recorded abundance and a projection five years into the future. For the

### 2016-07-13 text/ms_pnas.tex
Cleanup
Switch panel letters in all captions to uppercase

@@
 \caption{ Illustration of population dynamic models that allow for heavy
-tails. (a, b) Probability density for the Student-$t$
+tails. (A, B) Probability density for the Student-$t$
 distribution with scale parameter of $1$ and different values of \(\nu\). Small
 values of \(\nu\) create heavy tails while as \(\nu\) approaches infinity the
-distribution approaches the normal distribution. (c to e)
+distribution approaches the normal distribution. (C to E)
 Simulated population dynamics from a Gompertz model with process noise drawn
 from Student-$t$ distributions with three values of \(\nu\). Coloured dots in
-panels c and d represent jumps with less than a 1 in 1000
-chance of occurring in a normal distribution. (f to h)
-Estimates of \(\nu\) from models fit to the times series in panels c to
-e. Shown are posterior samples (histograms), median and interquartile
+panels C and D represent jumps with less than a 1 in 1000
+chance of occurring in a normal distribution. (F to H)
+Estimates of \(\nu\) from models fit to the times series in panels C to
+E. Shown are posterior samples (histograms), median and interquartile
 range of the posterior (IQR, dots and line segments), and the exponential prior
-on \(\nu\) (dashed lines). Colour shading behind panels f to h
+on \(\nu\) (dashed lines). Colour shading behind panels F to H
 illustrates the approximate region of heavy tails.}
@@
 \caption{Estimates of population dynamic heavy-tailedness for
-populations of birds, mammals, insects, and fishes. (a to d)
+populations of birds, mammals, insects, and fishes. (A to D)
 Small values of \(\nu\) suggest heavy-tailed black-swan dynamics. Vertical
@@
 \caption{Standardized probabilities and covariates of heavy-tailed
-dynamics. (a) Taxonomic-order-level posterior densities of Pr(\(\nu
+dynamics. (A) Taxonomic-order-level posterior densities of Pr(\(\nu
 < 10\)) after accounting for time-series length in a hierarchical model. Dotted
@@
 the geometric mean of time series length across all the data (approximately 27
-time steps) and shown for orders with \(\ge 5\) populations. (b)
+time steps) and shown for orders with \(\ge 5\) populations. (B)
 Posterior densities for potential covariates of Pr(\(\nu < 10\)). In both
@@
 \caption{Heavy-tailed process noise tends to be downward skewed and
-ignoring this can underestimate risk. (a) Illustrations of Student-$t$
-distributions with 3 levels of skewness (\(\gamma\)) and heavy-tails (\(\nu\)).
-(b) Posterior density of the skewness parameters aggregated across
+ignoring this can underestimate risk. (A) Illustrations of Student-$t$
+distributions with three levels of skewness (\(\gamma\)) and heavy-tails (\(\nu\)).
+(B) Posterior density of the skewness parameters aggregated across
 populations grouped into heavy-tailed (\(\hat{\nu} < 10\)), slightly
 heavy-tailed (\(10 \leq \hat{\nu} \leq 70\)), and normal-tailed populations
-(\(\hat{\nu} > 70\)). Approximate mid-values from b are illustrated in
-panel a. (c) Example time series of heavy-tailed populations with
+(\(\hat{\nu} > 70\)). Approximate mid-values from B are illustrated in
+panel A. (C) Example time series of heavy-tailed populations with
 different levels of skewness. Red dots highlight likely heavy-tailed events.
@@
 
-% \section*{Figures}
-
-% \clearpage
-
-% \clearpage
-
-% \clearpage
-
 \end{document}

### 2016-07-13 text/supp.tex
Switch panel letters in all captions to uppercase

@@
 \caption{Probability density of the Bayesian priors for the Gompertz models.
-\textbf{(a)} Per capita growth rate at $\log$(abundance) = $0$: $\lambda \sim \mathrm{Normal}(0, 10^2)$.
-\textbf{(b)} Scale parameter of t-distribution process noise: $\sigma \sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)$ \textbf{(c)} Student-$t$ distribution degrees of freedom parameter: $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2)$.
-\textbf{(d)} AR1 correlation coefficient of residuals: $\phi \sim \mathrm{Truncated \mhyphen Normal}(0, 0.5, \mathrm{min.} = -1, \mathrm{max.} = 1)$.
+\textbf{(A)} Per capita growth rate at $\log$(abundance) = $0$: $\lambda \sim \mathrm{Normal}(0, 10^2)$.
+\textbf{(B)} Scale parameter of t-distribution process noise: $\sigma \sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)$ \textbf{(C)} Student-$t$ distribution degrees of freedom parameter: $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2)$.
+\textbf{(D)} AR1 correlation coefficient of residuals: $\phi \sim \mathrm{Truncated \mhyphen Normal}(0, 0.5, \mathrm{min.} = -1, \mathrm{max.} = 1)$.
 Not shown is $b$, the density dependence parameter: $b \sim \mathrm{Uniform}(-1, 2)$.
-Panel \textbf{(c)} also shows two alternative priors: a weaker prior $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.005, \mathrm{min.} = 2)$, and a stronger prior $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.02, \mathrm{min.} = 2)$.
+Panel \textbf{(C)} also shows two alternative priors: a weaker prior $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.005, \mathrm{min.} = 2)$, and a stronger prior $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.02, \mathrm{min.} = 2)$.
 The inset panel shows the same data but with a log-transformed x axis.
@@
 \caption{Estimates of $\nu$ from alternative models plotted against the
-    base Gompertz model estimates of $\nu$. Alternative models are \textbf{(a)}
-    Ricker-logistic; \textbf{(b)} a Gompertz model with AR1 residual correlation;
-    \textbf{(c)} modelling population growth rate with no density dependence; and
-    \textbf{(d)} a Gompertz model with fixed magnitude of observation error. Shown
+    base Gompertz model estimates of $\nu$. Alternative models are \textbf{(A)}
+    Ricker-logistic; \textbf{(B)} a Gompertz model with AR1 residual correlation;
+    \textbf{(C)} modelling population growth rate with no density dependence; and
+    \textbf{(D)} a Gompertz model with fixed magnitude of observation error. Shown
 are medians of the posterior (dots) and 50\% credible intervals (segments). The
@@
 The Gompertz observation error model assumes a fixed standard deviation of
-observation error of $0.2$ on a log scale. Models in \textbf{(b)} and \textbf{(d)}
-that had not met convergence thresholds were run for four chains, 256,000 iterations (128,000 warm up), and retaining every 10th sample, after which the MCMC chains for one population in \textbf{(b)} and four populations in \textbf{(d)} had still not met convergence thresholds.
+observation error of $0.2$ on a log scale. Models in \textbf{(B)} and \textbf{(D)}
+that had not met convergence thresholds were run for four chains, 256,000 iterations (128,000 warm up), and retaining every 10th sample, after which the MCMC chains for one population in \textbf{(B)} and four populations in \textbf{(D)} had still not met convergence thresholds.
 These populations are excluded from their respective panels, but all populations with Pr($\nu < 10$) $> 0.5$ in the base model had converged and are shown.}
@@
 \caption{Estimates of $\nu$ from Gompertz models with alternative priors on $\nu$.
-\textbf{(a)} A weaker prior on $\nu$; \textbf{(b)} a stronger prior on $\nu$, as illustrated in Fig.~1.
+\textbf{(A)} A weaker prior on $\nu$; \textbf{(B)} a stronger prior on $\nu$, as illustrated in Fig.~1.
 Shown are medians of the posterior (dots) and 50\% credible intervals (segments).
@@
 \caption{Potential covariates of heavy-tailed population dynamics (indicated by a high probability that $\nu < 10$).
-Shown are (\textbf{a} to \textbf{c}) parameters from the Gompertz heavy-tailed population model ($\sigma$, $\lambda$, $b$), \textbf{(d)} number of time steps, \textbf{(e)} body length, and \textbf{(f)} lifespan.
+Shown are (\textbf{A} to \textbf{C}) parameters from the Gompertz heavy-tailed population model ($\sigma$, $\lambda$, $b$), \textbf{(D)} number of time steps, \textbf{(E)} body length, and \textbf{(F)} lifespan.
 For the Gompertz parameters, $\sigma$ refers to the scale parameter of the Student-$t$ process-noise distribution, $\lambda$ refers to the expected log abundance at the next time step at an abundance of one, $b$ refers to the density dependence parameter.
 Circles representing a few sharks, crustaceans, and gastropods are filled in white.
-Median and 90\% credible interval posterior predictions of a beta regression hierarchical model are shown in panels \textbf{(a)} and \textbf{(d)} where there was a high probability the slope coefficient was different from zero (Fig.~3B).}
+Median and 90\% credible interval posterior predictions of a beta regression hierarchical model are shown in panels \textbf{(A)} and \textbf{(D)} where there was a high probability the slope coefficient was different from zero (Fig.~3B).}
 

### 2016-07-13 text/ms_pnas.tex
Fix figure panel references in text

@@
 eight orders with the highest median probability of heavy tails were insect
-orders (Fig.~3a).
+orders (Fig.~3A).
 
@@
 magnitude of process noise but not clearly related to population growth rate,
-density dependence, or maximum lifespan (Fig.~3b, Fig.~S\figcorrelates).
+density dependence, or maximum lifespan (Fig.~3B, Fig.~S\figcorrelates).
 Longer time-series length was the strongest covariate of observing black-swan
@@
 < 10\)), 85\% of the \(\gamma\) probability density was below 1, indicating
-strong evidence of downwardly skewed process noise (Fig.~4b,
+strong evidence of downwardly skewed process noise (Fig.~4B,
 Fig.~S\figskewnu). In contrast, populations that did not have heavy
 tails (median \(\nu \geq 70\)) had little evidence of skewed process noise
-(Fig.~4b; 90\% of 95\% credible intervals overlapped 1). Projecting these
+(Fig.~4B; 90\% of 95\% credible intervals overlapped 1). Projecting these
 heavy-tailed populations forward five years revealed that assuming standard
 normal process noise underestimated risk (99\% lower credible interval of
-abundance) by \crashUnderRange-fold (interquartile range; Fig.~4c,
+abundance) by \crashUnderRange-fold (interquartile range; Fig.~4C,
 Fig.~S\figskewprojections).
@@
 we suggest using $\nu = 5$; and
-a skewness parameter ($\gamma$) of $\sim 0.7$ (Fig.~4a)
+a skewness parameter ($\gamma$) of $\sim 0.7$ (Fig.~4A)
 if a modeler wishes to allow for downward heavy-tailed events.

### 2016-07-13 text/supp.tex
Fix figure panel references in text

@@
 we used Truncated-Exponential(0.005, min.\ = 2) and for a stronger prior we
-used Truncated-Exponential(0.02, min.\ = 2) (Fig.~\ref{fig:priors}c). Note that
+used Truncated-Exponential(0.02, min.\ = 2) (Fig.~\ref{fig:priors}C). Note that
 the base and weaker priors are relatively flat within the region of \(\nu <
@@
 model can recapture the true value of $\nu$ when the process noise was chosen
-so that appropriate tail events were present (fig.
-\ref{fig:sim-prob} upper panels). The addition of observation error caused the
+so that appropriate tail events were present (Fig.~\ref{fig:sim-prob} upper panels).
+The addition of observation error caused the
 model to tend to underestimate the degree of heavy-tailedness. Fitting a model
 with correctly specified observation error did not make substantial
-improvements to model bias (fig. \ref{fig:sim-prob}).
+improvements to model bias (Fig.~\ref{fig:sim-prob}).
 
@@
 When converting the posterior distributions of $\nu$ into Pr($\nu < 10$), the
-models distinguished heavy and not-heavy tails reasonably well (fig.
-\ref{fig:sim-prob} lower panels). Without observation error, and using a
+models distinguished heavy and not-heavy tails reasonably well (Fig.~\ref{fig:sim-prob} lower panels).
+Without observation error, and using a
 probability of $0.5$ as a threshold, the model correctly classified all
@@
 The model would have miscategorized only one of $40$ simulations at $\nu = 5$
-across simulated populations with $50$ or $100$ time steps (fig.
-\ref{fig:sim-prob}, scenarios 1 and 2 in lower row, second panel from left).
+across simulated populations with $50$ or $100$ time steps 
+(Fig.~\ref{fig:sim-prob}, scenarios 1 and 2 in lower row, second panel from left).
 The model would have correctly categorized all cases where the process noise
-was not heavy tailed (fig. \ref{fig:sim-prob} bottom-right panel) and all cases
+was not heavy tailed (Fig.~\ref{fig:sim-prob} bottom-right panel) and all cases
 where $\nu = 3$ and there was not observation error. With $0.2$ standard

### 2016-07-13 text/ms_pnas.tex
Copyediting

@@
 demonstrates the rare but systematic presence of black-swan events in animal populations
-around the world (mostly birds, mammals, insects, and fishes). These events are
+around the world (mostly birds, mammals, and insects). These events are
 predominantly downward, implying that unexpected population crashes occur more
@@
 to a large number of time series can yield insights into the frequency, strength,
-and correlates of black-swan events.
+and correlates of such events.
 We are unaware of such a comparative analysis.
@@
 drawn from a Student-$t$ distribution. By estimating the degrees of freedom
-parameter, \(\nu\), we can estimate the degree to which the process deviations
+parameter, \(\nu\), we can estimate the extent to which the process deviations
 have heavy tails and are therefore diagnostic of black-swan events (Fig.~1).
@@
 
-We systematically evaluated the presence of black-swan events in hundreds of
+We systematically evaluated the prevalence of black-swan events in hundreds of
 animal populations.
@@
 the normal distribution,
-and $\nu  = 5$ allows for an event
+and compared to the normal distribution, $\nu  = 5$ allows for an event
 three standard deviations below the mean to happen

### 2016-07-13 text/ms_pnas.tex
Add nonbreaking space

@@
 tails (median \(\nu \geq 70\)) had little evidence of skewed process noise
-(Fig.~4B; 90\% of 95\% credible intervals overlapped 1). Projecting these
+(Fig.~4B; 90\% of 95\% credible intervals overlapped~1). Projecting these
 heavy-tailed populations forward five years revealed that assuming standard

### 2016-11-10 text/ms_pnas.tex
Reorder supplementary figures
Edits while responding to reviewers
Also strip spaces

@@
 \newcommand{\tabcausessupp}{{2}}
-\newcommand{\figpriors}{{1}}
-\newcommand{\figsimnu}{{2}}
-\newcommand{\figsimprob}{{3}}
-\newcommand{\figalt}{{4}}
-\newcommand{\figaltpriors}{{5}}
-\newcommand{\figcorrelates}{{6}}
-\newcommand{\figskewnu}{{7}}
-\newcommand{\figskewprojections}{{8}}
+\newcommand{\figalt}{{1}}
+\newcommand{\figaltpriors}{{2}}
+\newcommand{\figsimnu}{{3}}
+\newcommand{\figsimprob}{{4}}
+\newcommand{\figcorrelates}{{5}}
+\newcommand{\figskewnu}{{6}}
+\newcommand{\figskewprojections}{{7}}
+\newcommand{\figpriors}{{8}}
 \mathchardef\mhyphen="2D % math hyphen
@@
 data and wrote the paper with input from all authors.}
-\authordeclaration{Please declare any conflict of interest here.}
+\authordeclaration{The authors declare no conflict of interest.}
 \correspondingauthor{\textsuperscript{2}To whom correspondence should be addressed. E-mail: sandrsn@uw.edu}
@@
 with) other species experiencing black-swan events, or sharp changes in human
-pressures such as fishing or habitat destruction. Alternatively, the
+pressures such as fishing or habitat destruction.
+Alternatively, the
 synchrony of multiple extrinsic forces could give rise to black-swan dynamics
@@
 non-synergistic forces \cite{denny2009}.
+We expect that it is extrinsic forces that lead to the
+prevalence of skew-$t$ distributed process noise:
+there is no limit to how fast a population can decline due to a catastrophe,
+but population increases are constrained
+by the maximum biological rate of increase
+possible for a given set of demographic traits
+(e.g. population birth rate, mortality, age at maturity).
 
 There are a number of caveats when considering the generality of our results.
-The GPDD data represent a taxonomically and geographically biased sample of
+The Global Population Dynamics Database \cite[GPDD,][]{gpdd2010} we used
+represents a taxonomically and geographically biased sample of
 populations---the longer time series we focused on are dominated by commercially
@@
 
+TODO add assumptions here: assuming that productivity and density dependence are not time varying, this could be an important development for understanding the dynamics of longer time series
+
 We chose the $t$ distribution for a number of reasons.
@@
 (e.g.\ sedimentation rates or wind speeds \cite{gaines1993, denny2009}),
-the GEV requires maxima per time block
-or peaks beyond a chosen threshold.
-Therefore, the GEV requires longer time series then typically available
+the GEV requires maxima per time block.
+Therefore, the GEV requires longer or higher frequency time series than typically available
 for wild animal population abundance.
@@
 than suggested by Ref.~\cite{fernandez1998}. We truncated the
-distribution at \(2\), since at \(\nu < 2\) the variance of the $t$ distribution
-is undefined. This prior gives only a \(7.7\)\% probability that \(\nu < 10\)
+distribution at \(2\) since
+below $2$ there can be issues with computational stability.
+This prior gives only a \(7.7\)\% probability that \(\nu < 10\)
 but constrains the sampling sufficiently to avoid \(\nu\) drifting towards infinity. In
@@
 accounting for density dependence \citep{halley2002,segura2013}; therefore, we
-fit a simpler model in which we assumed density independence. Third,
-observation error could bias parameter estimates \citep{knape2012} or mask our
+fit a simpler model in which we assumed density independence.
+This model is equivalent to a random walk with drift
+model on log abundance, and therefore does not assume
+that the timeseries is stationary.
+Third, observation error could bias parameter estimates \citep{knape2012} or mask our
 ability to detect heavy tails \citep{ward2007}; therefore, we fit a model
@@
 of our results to weaker and stronger priors (exponential rate parameter $=
-0.005, 0.02$; Fig.~S\figpriors, Supporting Information). Furthermore, we used
+0.005, 0.02$) and alternative priors (Fig.~S\figpriors, Supporting Information).
+Furthermore, we used
 simulated data to test how easily we could detect $\nu$ given different sample

### 2016-11-10 text/supp.tex
Reorder supplementary figures

@@
 20\), which is the region we are concerned about when categorizing
-populations as heavy tailed. We also considered using a Uniform(\(0, 1\))
+populations as heavy tailed. We considered using a Uniform(\(0, 1\))
 prior on \(1/\nu\) as used in \citet{gelman2014}, but this
@@
 and the results are therefore not shown.
+Furthermore, we considered the prior suggested in \citet{juarez2010} of
+$\nu \sim \mathrm{Gamma}(2, 0.1)$, again limited to $\nu > 2$.
+Compared to our main truncated exponential prior,
+this prior places a higher probability density at $\nu < 10$ ($0.25$ vs.\ $0.08$)
+and has a mode at $\nu = 10$.
+This prior raised the estimated value of $\nu$
+for some of the populations identified as heavy-tailed
+with our default prior,
+but overall gave us similar inference
+with a Spearman correlation of 0.90 between Pr($\nu < 10$)
+for the two priors.
 
@@
 \begin{center}
-\includegraphics[width=0.6\textwidth]{../analysis/priors-gomp-base.pdf}
-
-\caption{Probability density of the Bayesian priors for the Gompertz models.
-\textbf{(A)} Per capita growth rate at $\log$(abundance) = $0$: $\lambda \sim \mathrm{Normal}(0, 10^2)$.
-\textbf{(B)} Scale parameter of t-distribution process noise: $\sigma \sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)$ \textbf{(C)} Student-$t$ distribution degrees of freedom parameter: $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2)$.
-\textbf{(D)} AR1 correlation coefficient of residuals: $\phi \sim \mathrm{Truncated \mhyphen Normal}(0, 0.5, \mathrm{min.} = -1, \mathrm{max.} = 1)$.
-Not shown is $b$, the density dependence parameter: $b \sim \mathrm{Uniform}(-1, 2)$.
-Panel \textbf{(C)} also shows two alternative priors: a weaker prior $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.005, \mathrm{min.} = 2)$, and a stronger prior $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.02, \mathrm{min.} = 2)$.
-The inset panel shows the same data but with a log-transformed x axis.
-Note that the base and weaker priors are relatively flat within the region of $\nu < 20$ with which we are concerned.}
-
-\label{fig:priors}
-\end{center}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=0.8\textwidth]{../analysis/t-dist-sampling-sim-prior-exp0point01.pdf}
-\includegraphics[width=0.8\textwidth]{../analysis/t-dist-sampling-sim-sigma-prior-exp0point01.pdf}
-
-\caption{Testing the ability to estimate $\nu$ (top panels) and the scale parameter of the process deviations (bottom panels) for a given number of samples (columns) drawn from a distribution with a given true $\nu$ value (rows).
-The red lines indicate the true population value.
-When a small number of samples are drawn there may not be samples sufficiently far into the tails to recapture the true $\nu$ value; however, heavy tails are still distinguished from normal tails in most cases, even with only 25 or 50 samples.}
-
-\label{fig:sim-nu}
-\end{center}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=\textwidth]{../analysis/sim-gompertz-median-dist.pdf}
-\includegraphics[width=\textwidth]{../analysis/sim-gompertz-p10.pdf}
-
-\caption{Simulation testing the Gompertz estimation model when the process deviation draws were chosen so that $\nu$ could be estimated close to the true value outside the full population model (``effective $\nu$'' within a CV of 0.2 of specified $\nu$).
-Upper panels show the distribution of median $\widehat{\nu}$ across 20 simulation runs.
-Lower panels show the distribution of Pr($\nu < 10$) across 20 simulation runs.
-We ran the simulations across three population (``true'') $\nu$ values (3, 5, and $1\cdot 10^9$, i.e.\ approximately normal) and four scenarios: (1) 100 time steps and no observation error, (2) 50 time steps and no observation error, (3) 50 time steps and observation error drawn from $\mathrm{Normal} (0, 0.2^2)$ but ignored, and (4) 50 time steps with observation error in which the quantity of observation error was assumed known.
-Within each scenario the dots represent stochastic draws from the true population distributions combined with model fits.
-Underlayed boxplots show the median, interquartile range, and $1.5$ times the interquartile range.}
-
-\label{fig:sim-prob}
-\end{center}
-\end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
 \includegraphics[width=0.8\textwidth]{../analysis/gomp-comparison.pdf}
@@
 \clearpage
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=0.8\textwidth]{../analysis/t-dist-sampling-sim-prior-exp0point01.pdf}
+\includegraphics[width=0.8\textwidth]{../analysis/t-dist-sampling-sim-sigma-prior-exp0point01.pdf}
+
+\caption{Testing the ability to estimate $\nu$ (top panels) and the scale parameter of the process deviations (bottom panels) for a given number of samples (columns) drawn from a distribution with a given true $\nu$ value (rows).
+The red lines indicate the true population value.
+When a small number of samples are drawn there may not be samples sufficiently far into the tails to recapture the true $\nu$ value; however, heavy tails are still distinguished from normal tails in most cases, even with only 25 or 50 samples.}
+
+\label{fig:sim-nu}
+\end{center}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=\textwidth]{../analysis/sim-gompertz-median-dist.pdf}
+\includegraphics[width=\textwidth]{../analysis/sim-gompertz-p10.pdf}
+
+\caption{Simulation testing the Gompertz estimation model when the process deviation draws were chosen so that $\nu$ could be estimated close to the true value outside the full population model (``effective $\nu$'' within a CV of 0.2 of specified $\nu$).
+Upper panels show the distribution of median $\widehat{\nu}$ across 20 simulation runs.
+Lower panels show the distribution of Pr($\nu < 10$) across 20 simulation runs.
+We ran the simulations across three population (``true'') $\nu$ values (3, 5, and $1\cdot 10^9$, i.e.\ approximately normal) and four scenarios: (1) 100 time steps and no observation error, (2) 50 time steps and no observation error, (3) 50 time steps and observation error drawn from $\mathrm{Normal} (0, 0.2^2)$ but ignored, and (4) 50 time steps with observation error in which the quantity of observation error was assumed known.
+Within each scenario the dots represent stochastic draws from the true population distributions combined with model fits.
+Underlayed boxplots show the median, interquartile range, and $1.5$ times the interquartile range.}
+
+\label{fig:sim-prob}
+\end{center}
+\end{figure}
 
@@
 \label{fig:skew-projections}
+\end{center}
+\end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=0.6\textwidth]{../analysis/priors-gomp-base.pdf}
+
+\caption{Probability density of the Bayesian priors for the Gompertz models.
+\textbf{(A)} Per capita growth rate at $\log$(abundance) = $0$: $\lambda \sim \mathrm{Normal}(0, 10^2)$.
+\textbf{(B)} Scale parameter of t-distribution process noise: $\sigma \sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)$ \textbf{(C)} Student-$t$ distribution degrees of freedom parameter: $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2)$.
+\textbf{(D)} AR1 correlation coefficient of residuals: $\phi \sim \mathrm{Truncated \mhyphen Normal}(0, 0.5, \mathrm{min.} = -1, \mathrm{max.} = 1)$.
+Not shown is $b$, the density dependence parameter: $b \sim \mathrm{Uniform}(-1, 2)$.
+Panel \textbf{(C)} also shows two alternative priors: a weaker prior $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.005, \mathrm{min.} = 2)$, and a stronger prior $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.02, \mathrm{min.} = 2)$.
+The inset panel shows the same data but with a log-transformed x axis.
+Note that the base and weaker priors are relatively flat within the region of $\nu < 20$ with which we are concerned.}
+
+\label{fig:priors}
 \end{center}

### 2016-11-10 text/ms_pnas.tex

@@
 \dropcap{M}ajor surprises happen more often than expected in
-financial, social, and natural systems \cite{taleb2007, sornette2009, may2008}.
+financial, social, and natural systems
+\cite{sornette2002, albeverio2006, taleb2007, may2008, sornette2009}.
 Massive unpredictable market swings are responsible for the majority of
@@
 damaging earthquakes has exceeded past expectations \cite{sornette2009}. In
-ecological systems, background rates of global extinction are punctuated by
-mass extinction \cite{harnik2012}, viruses can mutate suddenly to infect new
-hosts, and billions of animals can die at once in mass mortality
-events \cite{fey2015}.
+ecological systems,
+billions of animals can die at once in mass mortality
+events \cite{fey2015},
+background rates of global extinction are punctuated by
+mass extinction \cite{harnik2012},
+and evolution is characterized by
+long periods of stasis punctuated by
+bursts of speciation \cite{gould1977}.
 
@@
 \includegraphics[width=0.65\textwidth]{../analysis/t-nu-eg2.pdf}
-\caption{ Illustration of population dynamic models that allow for heavy
+\caption{Illustration of population dynamic models that allow for heavy
 tails. (A, B) Probability density for the Student-$t$
@@
 compared to once every 32,000 years with a normal distribution.
-As \(\nu\) approaches infinity, the t
+As \(\nu\) approaches infinity, the $t$
 distribution approaches the normal distribution (Fig.~1).
@@
 mean Pr(\(\nu < 10\)). Taxonomic orders with three or fewer populations in
-panel a are omitted for space. Red to yellow points highlight populations
+panel A are omitted for space. Red to yellow points highlight populations
 with a high to moderately high probability of heavy-tailed dynamics.
@@
 < 10\)) after accounting for time-series length in a hierarchical model. Dotted
-vertical line in a indicates the median expected Pr(\(\nu < 10\)) from the
+vertical line in A indicates the median expected Pr(\(\nu < 10\)) from the
 prior distribution. Colour shading refers to taxonomic class (yellow: fishes,
 green: insects, purple: birds, and red: mammals). Estimates are standardized at
-the geometric mean of time series length across all the data (approximately 27
+the geometric mean of time-series length across all the data (approximately 27
 time steps) and shown for orders with \(\ge 5\) populations. (B)
@@
 
-% than black-swan population
-% increases and are usually driven by external events such as weather and
-% disease. Therefore, ignoring the presence of black-swan events, and the
-% predominance of crashes over increases, may underestimate extinction risk.
-% TODO: add numbers, add skew add projections add length and Sigma
-
 There are many possible causes of black-swan events, including
@@
 through time \cite{carpenter2006}. Extrinsic ecological forces could also cause
-black-swan dynamics \cite{nunez2012} including extreme climate
+black-swan dynamics \cite{nunez2012}, including extreme climate
 events \cite{meehl2004, katz2005, ipcc2012}, predation from (or competition
@@
 non-synergistic forces \cite{denny2009}.
-We expect that it is extrinsic forces that lead to the
-prevalence of skew-$t$ distributed process noise:
+The prevalence of population die-offs rather than unexpected increases
+is likely due to biology:
 there is no limit to how fast a population can decline due to a catastrophe,
@@
 by the maximum biological rate of increase
-possible for a given set of demographic traits
+for a given set of demographic traits
 (e.g. population birth rate, mortality, age at maturity).
@@
 There are a number of caveats when considering the generality of our results.
-The Global Population Dynamics Database \cite[GPDD,][]{gpdd2010} we used
+First, the Global Population Dynamics Database \cite[GPDD,][]{gpdd2010} we used
 represents a taxonomically and geographically biased sample of
@@
 geographic samples of populations, the common forces driving those black swans
-would likely differ. Additionally, some apparent black-swan events could be
+would likely differ. Second, some apparent black-swan events could be
 recording mistakes, although conversely, some extreme observations may have been
@@
 initially identified as heavy tailed because they turned out to be data-entry
-errors. Finally, the temporal scales of observation and population dynamics
+errors (Supporting Information).
+Third, the temporal scales of observation and population dynamics
 vary considerably across populations in the GPDD and these likely influence the
@@
 less-extreme events and also dampen black-swan dynamics.
+Finally, our models assume that
+the fitted parameters (e.g. productivity and density dependence)
+do not vary over time.
+Allowing for time varying parameters
+could be an important development for
+understanding the dynamics of longer time series.
 
@@
 % and species can unexpectedly interact.
-
-% todo the strongest correlates of finding blocks one events was the timeseries length
-% ...implications
 
@@
 
-TODO add assumptions here: assuming that productivity and density dependence are not time varying, this could be an important development for understanding the dynamics of longer time series
-
 We chose the $t$ distribution for a number of reasons.
@@
 parameter over the normal distribution and reverts to the normal distribution
-if strong evidence of heavy-tailed process deviations is absent.
+if evidence of heavy-tailed process deviations is absent.
 Furthermore, the $t$ distribution arises from a mixture of normal distributions
@@
 
-We fit our models with Stan 2.6.0 \cite{stan-manual2015, hoffman2014}, and R
-3.2.0 \cite{r2015}. We began with four chains and 2,000 iterations, discarding
+We fit our models with Stan 2.12.1 \cite{hoffman2014, stan-manual2015, carpenter2016}, and R
+3.3.1 \cite{r2015}. We began with four chains and 2,000 iterations, discarding
 the first 1,000 as warm up (i.e.~4,000 final samples). If \(\hat{R}\) (the
@@
 and warm up period and sampled from the model again.
-% These thresholds are in
-% excess of the minimums recommended by Ref.~\cite{gelman2006a} of
-% \(\hat{R} < 1.1\) and \(n_\mathrm{eff} > 100\) for reliable point estimates and
-% confidence intervals.
 In the majority of cases our minimum thresholds were
@@
 where we allowed for a fixed quantity of observation error ($0.2$ standard
-deviations on a log scale). Finally, the Gompertz model assumes that
+deviations on a log scale). Fourth, the Gompertz model assumes that
 population growth rate declines linearly with log abundance. Therefore, we
@@
 population growth rate declines linearly with abundance (Supporting Information).
+% Our models also assume that their parameters
+% (e.g. parameters representing productivity and density dependence)
 

### 2016-11-11 text/ms_pnas.tex
Copy editing

@@
 tend to be driven by external perturbations such as climate, severe winters,
-predators, parasites, or the combined effects of multiple factors.
+predators, parasites, or the combined effect of multiple factors.
 Black-swan events manifest primarily as
@@
 on estimating and predicting downward tails of population abundance, since
-these increase true extinction risk.
+these may increase true extinction risk.
 

### 2016-11-11 text/ms_pnas.tex
Copyediting

@@
 through time \cite{carpenter2006}. Extrinsic ecological forces could also cause
-black-swan dynamics \cite{nunez2012}, including extreme climate
+black-swan dynamics \cite{nunez2012} including extreme climate
 events \cite{meehl2004, katz2005, ipcc2012}, predation from (or competition
@@
 do not vary over time.
-Allowing for time varying parameters
+Allowing for time-varying parameters
 could be an important development for

### 2016-11-17 text/ms_pnas.tex
Copyediting
Clarify the section on 0.5 threshold

@@
 probability of heavy-tailed (black-swan) dynamics. When categorizing a population as heavy
-tailed, we applied a 0.5 probability threshold. In the scenario where the data
-are uninformative about heavy tails (e.g.~Fig.~1E, H), the posterior for
+tailed, we applied a 0.5 probability threshold.
+The data must therefore alter the posterior so that
+more than 50\% of the posterior probability has $\nu < 10$ for a
+population to be declared heavy tailed. 
+If the data
+are uninformative about heavy tails (e.g.~Fig.~1E, H), 
+and the population model is otherwise a reasonable representation of the dynamics, 
+the posterior for
 \(\nu\) will approximately match the prior (prior median \(= 71\), mean \(=
-102\)) and a metric of Pr\((\nu < 10) > 0.5\) will not flag the
-population as heavy tailed.
+102\)) and a metric of Pr\((\nu < 10) > 0.5\) will not flag any
+populations as heavy tailed.
 
@@
 This model is equivalent to a random walk with drift
-model on log abundance, and therefore does not assume
+on log abundance, and therefore does not assume
 that the timeseries is stationary.

### 2016-11-17 text/ms_pnas.tex
Copyediting
Cut some text to reduce length slightly

@@
 3.3.1 \cite{r2015}. We began with four chains and 2,000 iterations, discarding
-the first 1,000 as warm up (i.e.~4,000 final samples). If \(\hat{R}\) (the
+the first 1,000 as warm up. If \(\hat{R}\) (the
 potential scale reduction factor---a measure of chain convergence) was greater
@@
 In the majority of cases our minimum thresholds were
-greatly exceeded. We continued this procedure up to 8,000 iterations (16,000
-total samples) by which all chains had sufficiently converged.
-The No-U-Turn Hamiltonian Markov chain Monte Carlo sampler in Stan generally
-requires fewer iterations to obtain equivalent effective sample sizes than the
-Gibbs or Metropolis--Hastings algorithms \cite{stan-manual2015}.
+greatly exceeded. We continued this procedure up to 8,000 iterations 
+by which all chains had sufficiently converged.
+% The No-U-Turn Hamiltonian Markov chain Monte Carlo sampler in Stan generally
+% requires fewer iterations to obtain equivalent effective sample sizes than the
+% Gibbs or Metropolis--Hastings algorithms \cite{stan-manual2015}.
 
@@
 on log abundance, and therefore does not assume
-that the timeseries is stationary.
+that the time series is stationary.
 Third, observation error could bias parameter estimates \citep{knape2012} or mask our

### 2016-11-17 analysis/cause-table.tex
Update output tables

@@
-% latex table generated in R 3.2.0 by xtable 1.7-4 package
+% latex table generated in R 3.3.1 by xtable 1.8-2 package
 Time series & Population & ID & Ref & Description & Pr($\nu < 10$) \\ 
   \midrule
-\includegraphics[width=1.7cm]{../analysis/sparks/6528.pdf} & Shag, \textit{Phalacrocorax aristotelis}, Farne Islands, Northumberland & 6528 & \cite{potts1980} & Red tide event combined with low productivity due to overcrowding & 1.00 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/7115.pdf} & South African fur seal, \textit{Arctocephalus pusillus}, South Africa & 7115 & \cite{shaughnessy1982} & Harvesting and predation changes & 1.00 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/10128.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Scotland - un-named area & 10128 & \cite{potts1984} & Environment- and parisite-caused cycles & 1.00 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/9382.pdf} & Pine looper or Bordered white, \textit{Bupalus piniaria}, Kessock & 9382 & \cite{broekhuizen1993} & Unknown, but sampling intensity was decreasing & 1.00 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/10127.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Scotland - un-named area & 10127 & \cite{potts1984} & Environment- and parisite-caused cycles & 0.99 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/10007.pdf} & Water vole, \textit{Arvicola terrestris}, Le Pont & 10007 & \cite{saucy1994} & Predator-environment cycle interactions & 1.00 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/20579.pdf} & Grey heron, \textit{Ardea cinerea}, Southern Britain & 20579 & \cite{stafford1971} & Severe winter & 0.98 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/9655.pdf} & Flea beetle, \textit{Chaetocnoma concinna}, Finland & 9655 & \cite{markkula1965} & Cannot locate original source & 0.99 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/1235.pdf} & Wren, \textit{Troglodytes troglodytes}, Eastern Wood, Bookham Common & 1235 & \cite{newton1998} & Severe winter & 0.98 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/10113.pdf} & Willow grouse, \textit{Lagopus lagopus}, Northern England & 10113 & \cite{dobson1995} & Parasites and predators & 0.99 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/9667.pdf} & Gooseberry sawfly, \textit{Nemastus ribesii}, Finland & 9667 & \cite{markkula1965} & Cannot locate original source & 0.97 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/9679.pdf} & Unknown, \textit{Trioza apicalis}, Finland & 9679 & \cite{markkula1965} & Cannot locate original source & 0.83 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/20527.pdf} & Wandering albatross, \textit{Diomedea exulans}, Taiaroa & 20527 & \cite{robertson1998} & Unknown & 0.99 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/10039.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Northern Scotland & 10039 & \cite{dobson1995} & Parasites and predators & 0.92 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/10162.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Atholl Estate & 10162 & \cite{mackenzie1952} & Bad environmental conditions and overcrowding combined to create crashes & 0.91 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/9503.pdf} & Fisher or  Pekan, \textit{Martes pennanti}, Manitoba & 9503 & \cite{keith1963} & Unknown & 0.76 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/7099.pdf} & European rabbit, \textit{Oryctolagus cuniculus}, Estate 2, East Anglia & 7099 & \cite{barnes1986} & Disease outbreak followed by years of good weather & 0.78 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/2778.pdf} & Wheatear, \textit{Oenanthe oenanthe}, Skokholm Island & 2778 & \cite{lack1969} & Unknown, but decline noted specifically, cold winters caused some crashes & 0.67 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/1195.pdf} & Blue jay, \textit{Cyanocitta cristata}, Robert Allerton Park & 1195 & \cite{kendeigh1982} & Unknown & 0.60 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/5019.pdf} & Barbary macaque, \textit{Macaca sylvanus}, Queens Gate & 5019 & \cite{fa1984} & Cannot locate original source & 0.61 \\ 
+\includegraphics[width=1.7cm]{../analysis/sparks/6528.pdf} & Shag, \textit{Phalacrocorax aristotelis}, Farne Islands, Northumberland & 6528 & \citep{potts1980} & Red tide event combined with low productivity due to overcrowding & 1.00 \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/7115.pdf} & South African fur seal, \textit{Arctocephalus pusillus}, South Africa & 7115 & \citep{shaughnessy1982} & Harvesting and predation changes & 1.00 \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/9382.pdf} & Pine looper or Bordered white, \textit{Bupalus piniaria}, Kessock & 9382 & \citep{broekhuizen1993} & Unknown, but sampling intensity was decreasing & 1.00 \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/10127.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Scotland - un-named area & 10127 & \citep{potts1984} & Environment- and parisite-caused cycles & 1.00 \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/10128.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Scotland - un-named area & 10128 & \citep{potts1984} & Environment- and parisite-caused cycles & 1.00 \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/10007.pdf} & Water vole, \textit{Arvicola terrestris}, Le Pont & 10007 & \citep{saucy1994} & Predator-environment cycle interactions & 1.00 \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/20579.pdf} & Grey heron, \textit{Ardea cinerea}, Southern Britain & 20579 & \citep{stafford1971} & Severe winter & 0.99 \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/1235.pdf} & Wren, \textit{Troglodytes troglodytes}, Eastern Wood, Bookham Common & 1235 & \citep{newton1998} & Severe winter & 0.98 \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/9655.pdf} & Flea beetle, \textit{Chaetocnoma concinna}, Finland & 9655 & \citep{markkula1965} & Cannot locate original source & 0.99 \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/10113.pdf} & Willow grouse, \textit{Lagopus lagopus}, Northern England & 10113 & \citep{dobson1995} & Parasites and predators & 0.99 \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/9667.pdf} & Gooseberry sawfly, \textit{Nemastus ribesii}, Finland & 9667 & \citep{markkula1965} & Cannot locate original source & 0.97 \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/9679.pdf} & Unknown, \textit{Trioza apicalis}, Finland & 9679 & \citep{markkula1965} & Cannot locate original source & 0.85 \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/20527.pdf} & Wandering albatross, \textit{Diomedea exulans}, Taiaroa & 20527 & \citep{robertson1998} & Unknown & 1.00 \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/10039.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Northern Scotland & 10039 & \citep{dobson1995} & Parasites and predators & 0.92 \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/10162.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Atholl Estate & 10162 & \citet{mackenzie1952} & Bad environmental conditions and overcrowding combined to create crashes & 0.92 \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/7099.pdf} & European rabbit, \textit{Oryctolagus cuniculus}, Estate 2, East Anglia & 7099 & \citep{barnes1986} & Disease outbreak followed by years of good weather & 0.81 \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/9503.pdf} & Fisher or  Pekan, \textit{Martes pennanti}, Manitoba & 9503 & \citep{keith1963} & Unknown & 0.74 \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/2778.pdf} & Wheatear, \textit{Oenanthe oenanthe}, Skokholm Island & 2778 & \citep{lack1969} & Unknown, but decline noted specifically, cold winters caused some crashes & 0.68 \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/1195.pdf} & Blue jay, \textit{Cyanocitta cristata}, Robert Allerton Park & 1195 & \citep{kendeigh1982} & Unknown & 0.62 \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/9953.pdf} & Rock ptarmigan, \textit{Lagopus mutus}, Iceland & 9953 & \citep{clarke1885,williams1954} & Severe winters & 0.61 \\ 
    \bottomrule

### 2016-11-22 analysis/8-sim-extinction-risk.R
Sample 5 times from projections
To stabilize result

@@ -73,7 +73,7 @@ skew_samples <- readRDS("skew_samples_heavy40000.rds")
 normal_samples <- readRDS("normal_samples_heavy40000.rds")
 
 library("doParallel")
-registerDoParallel(cores = 4)
+registerDoParallel(cores = parallel::detectCores())
 
 # join in the starting abundances first for speed:
 last_abund <- group_by(gpdd, main_id) %>%
@@ -97,7 +97,8 @@ ww / rowSums(ww) * 100
 skew_samples_heavy <- filter(skew_samples, main_id %in% heavy_pops$main_id)
 normal_samples_heavy <- filter(normal_samples, main_id %in% heavy_pops$main_id)
 
-out_skew <- plyr::ldply(seq_len(1L), function(i) {
+set.seed(1)
+out_skew <- plyr::ldply(seq_len(5L), function(i) {
   temp <- plyr::ddply(skew_samples_heavy, c("main_id", "sample_ids"), function(x) {
       sim_gomp_extinction(y1 = x$last_abundance, lambda = x$lambda, b = x$b,
         N = 6L, nu = x$nu, sigma_proc = x$sigma_proc, log_skew = x$log_skew,
@@ -106,7 +107,7 @@ out_skew <- plyr::ldply(seq_len(1L), function(i) {
   temp}, .parallel = FALSE)
 saveRDS(out_skew, file = "project-skew-heavy.rds")
 
-out_normal <- plyr::ldply(seq_len(1L), function(i) {
+out_normal <- plyr::ldply(seq_len(5L), function(i) {
   temp <- plyr::ddply(normal_samples_heavy, c("main_id", "sample_ids"), function(x) {
       sim_gomp_extinction(y1 = x$last_abundance, lambda = x$lambda, b = x$b,
         N = 6L, nu = NULL, sigma_proc = x$sigma_proc, log_skew = NULL,
@@ -167,7 +168,7 @@ qq <- before_and_projections %>% group_by(main_id, year) %>%
 plot(qq$r, log = "y", ylim = c(0.2, 5))
 abline(h = 1)
 median(1/qq$r)
-plot(density(log(qq$r)), xlim = c(-2.5, 2.5))
+plot(density(1/qq$r), xlim = c(-1, 4.5));abline(v = 1)
 quantile(1/qq$r, probs = c(0, 0.25, 0.5, 0.75, 1)) %>%
   round(1)
 saveRDS(qq, file = "skew-understimates.rds")

### 2016-11-23 analysis/cause-table.tex
Update values and tables

@@
-% latex table generated in R 3.3.1 by xtable 1.8-2 package
+% latex table generated in R 3.3.2 by xtable 1.8-2 package
 Time series & Population & ID & Ref & Description & Pr($\nu < 10$) \\ 
@@
   \includegraphics[width=1.7cm]{../analysis/sparks/7115.pdf} & South African fur seal, \textit{Arctocephalus pusillus}, South Africa & 7115 & \citep{shaughnessy1982} & Harvesting and predation changes & 1.00 \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/10127.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Scotland - un-named area & 10127 & \citep{potts1984} & Environment- and parisite-caused cycles & 0.99 \\
   \includegraphics[width=1.7cm]{../analysis/sparks/9382.pdf} & Pine looper or Bordered white, \textit{Bupalus piniaria}, Kessock & 9382 & \citep{broekhuizen1993} & Unknown, but sampling intensity was decreasing & 1.00 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/10127.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Scotland - un-named area & 10127 & \citep{potts1984} & Environment- and parisite-caused cycles & 1.00 \\
   \includegraphics[width=1.7cm]{../analysis/sparks/10128.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Scotland - un-named area & 10128 & \citep{potts1984} & Environment- and parisite-caused cycles & 1.00 \\
   \includegraphics[width=1.7cm]{../analysis/sparks/10007.pdf} & Water vole, \textit{Arvicola terrestris}, Le Pont & 10007 & \citep{saucy1994} & Predator-environment cycle interactions & 1.00 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/20579.pdf} & Grey heron, \textit{Ardea cinerea}, Southern Britain & 20579 & \citep{stafford1971} & Severe winter & 0.99 \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/20579.pdf} & Grey heron, \textit{Ardea cinerea}, Southern Britain & 20579 & \citep{stafford1971} & Severe winter & 0.98 \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/9655.pdf} & Flea beetle, \textit{Chaetocnoma concinna}, Finland & 9655 & \citep{markkula1965} & Cannot locate original source & 0.99 \\ 
   \includegraphics[width=1.7cm]{../analysis/sparks/1235.pdf} & Wren, \textit{Troglodytes troglodytes}, Eastern Wood, Bookham Common & 1235 & \citep{newton1998} & Severe winter & 0.98 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/9655.pdf} & Flea beetle, \textit{Chaetocnoma concinna}, Finland & 9655 & \citep{markkula1965} & Cannot locate original source & 0.99 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/10113.pdf} & Willow grouse, \textit{Lagopus lagopus}, Northern England & 10113 & \citep{dobson1995} & Parasites and predators & 0.99 \\ 
   \includegraphics[width=1.7cm]{../analysis/sparks/9667.pdf} & Gooseberry sawfly, \textit{Nemastus ribesii}, Finland & 9667 & \citep{markkula1965} & Cannot locate original source & 0.97 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/9679.pdf} & Unknown, \textit{Trioza apicalis}, Finland & 9679 & \citep{markkula1965} & Cannot locate original source & 0.85 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/20527.pdf} & Wandering albatross, \textit{Diomedea exulans}, Taiaroa & 20527 & \citep{robertson1998} & Unknown & 1.00 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/10039.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Northern Scotland & 10039 & \citep{dobson1995} & Parasites and predators & 0.92 \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/10113.pdf} & Willow grouse, \textit{Lagopus lagopus}, Northern England & 10113 & \citep{dobson1995} & Parasites and predators & 0.98 \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/9679.pdf} & Unknown, \textit{Trioza apicalis}, Finland & 9679 & \citep{markkula1965} & Cannot locate original source & 0.86 \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/20527.pdf} & Wandering albatross, \textit{Diomedea exulans}, Taiaroa & 20527 & \citep{robertson1998} & Unknown & 0.99 \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/10039.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Northern Scotland & 10039 & \citep{dobson1995} & Parasites and predators & 0.93 \\ 
   \includegraphics[width=1.7cm]{../analysis/sparks/10162.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Atholl Estate & 10162 & \citet{mackenzie1952} & Bad environmental conditions and overcrowding combined to create crashes & 0.92 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/7099.pdf} & European rabbit, \textit{Oryctolagus cuniculus}, Estate 2, East Anglia & 7099 & \citep{barnes1986} & Disease outbreak followed by years of good weather & 0.81 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/9503.pdf} & Fisher or  Pekan, \textit{Martes pennanti}, Manitoba & 9503 & \citep{keith1963} & Unknown & 0.74 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/2778.pdf} & Wheatear, \textit{Oenanthe oenanthe}, Skokholm Island & 2778 & \citep{lack1969} & Unknown, but decline noted specifically, cold winters caused some crashes & 0.68 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/1195.pdf} & Blue jay, \textit{Cyanocitta cristata}, Robert Allerton Park & 1195 & \citep{kendeigh1982} & Unknown & 0.62 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/9953.pdf} & Rock ptarmigan, \textit{Lagopus mutus}, Iceland & 9953 & \citep{clarke1885,williams1954} & Severe winters & 0.61 \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/7099.pdf} & European rabbit, \textit{Oryctolagus cuniculus}, Estate 2, East Anglia & 7099 & \citep{barnes1986} & Disease outbreak followed by years of good weather & 0.78 \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/9503.pdf} & Fisher or  Pekan, \textit{Martes pennanti}, Manitoba & 9503 & \citep{keith1963} & Unknown & 0.76 \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/2778.pdf} & Wheatear, \textit{Oenanthe oenanthe}, Skokholm Island & 2778 & \citep{lack1969} & Unknown, but decline noted specifically, cold winters caused some crashes & 0.66 \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/9659.pdf} & Cabbage root fly or maggot, \textit{Delia radicum}, Finland & 9659 & \citep{markkula1965} & Cannot locate original source & 0.60 \\ 
    \bottomrule

### 2016-11-23 analysis/9-values-for-paper.R
Calculate more magic numbers for ms

@@ -221,4 +221,26 @@ qq <- readRDS("skew-understimates.rds")
 q <- quantile(1/qq$r, probs = c(0.25, 0.5, 0.75)) %>% round(1)
 write_tex(paste0(q[[1]], "--", q[[3]]), "crashUnderRange")
 write_tex(q[[2]], "crashUnderMedian")
+
+sk <- readRDS("skew_samples.rds")
+sk <- sk %>% group_by(main_id) %>%
+  mutate(
+    median_nu = median(nu),
+    h1 = ifelse(median_nu <= 10, "1 heavy",
+      ifelse(median_nu <= 70, "2 moderate", "3 normal"))) %>%
+  ungroup()
+
+normal <- filter(sk, h1 == "3 normal")$log_skew %>% sort
+moderate <- filter(sk, h1 == "2 moderate")$log_skew %>% sort
+heavy <- filter(sk, h1 == "1 heavy")$log_skew %>% sort
+write_tex(round(mean(heavy < 0), 2) * 100, "probDensSkewedForHeavyPops")
+
+overlap <- sk %>% filter(h1 == "3 normal") %>%
+  group_by(main_id) %>%
+    summarise(
+      l = exp(quantile(log_skew, prob = 0.025)),
+      u = exp(quantile(log_skew, prob = 0.975))) %>%
+  mutate(overlap1 = l <= 1 & u >= 1)
+write_tex(round(mean(overlap$overlap1), 2) * 100, "percNormPopsNotSkewed")
+
 close(zz)

### 2016-11-23 text/ms_pnas.tex
Copyediting

@@
 If this is the case, add \parshape=0 to the end of the list environment.
-\dropcap{M}ajor surprises happen more often than expected in
+\dropcap{M}ajor surprises (black-swan events) happen more often than expected in
 financial, social, and natural systems
@@
 % Despite this anecdotal evidence for
-% ecological black swans, systematic searches of large numbers of timeseries have
+% ecological black swans, systematic searches of large numbers of time series have
 % yet to reveal black-swan events \cite{keitt1998, allen2001, halley2002}.
@@
 Here, we develop a new approach to estimate the frequency and magnitude of
-black-swan dynamics across time series of 609 populations from a wide array of
+black-swan dynamics across time series of \NPops\ populations from a wide array of
 taxonomic groups---including many  birds, mammals, insects, and fishes
@@
 influential: in our dataset they were detected most
-frequently for birds (7\%), mammals (5\%), and
-insects (3\%) but almost never in fishes (Fig.~2). Black-swan
-events were taxonomically widespread, occurring in 38\% of taxonomic orders.
+frequently for birds (\birdPH\%), mammals (\mammalsPH\%), and
+insects (\insectsPH\%) but almost never in fishes (Fig.~2). Black-swan
+events were taxonomically widespread, occurring in \POrdersHeavy\% of taxonomic orders.
 Accounting for time series length and partially pooling inference across
 taxonomic class and order with a hierarchical model, we found stronger evidence
-for black swans in insect populations than these statistics suggest---four of
-eight orders with the highest median probability of heavy tails were insect
+for black swans in insect populations than these statistics suggest---two of
+five orders with the highest median probability of heavy tails were insect
 orders (Fig.~3A).
@@
 The majority of our heavy-tailed estimates were robust to alternative
-population models, observation error, and choice of Bayesian priors. Our
-conclusions were not systematically altered when we included autocorrelation
+population models, observation error, and choice of Bayesian priors (Supporting Information). 
+Our conclusions were not systematically altered when we included autocorrelation
 in the residuals, modelled population growth rates with or without
@@
 Longer time-series length was the strongest covariate of observing black-swan
-dynamics: there is a 1.6 times greater probability of detecting a black-swan
+dynamics: there is a \pIncHeavyNThirtyNSixty\ times greater probability of detecting a black-swan
 event in populations with 60 time steps than in one with 30 time steps
@@
 Aggregated across populations with strong evidence of heavy tails (median \(\nu
-< 10\)), 85\% of the \(\gamma\) probability density was below 1, indicating
+< 10\)), \probDensSkewedForHeavyPops\% of the \(\gamma\)
+probability density was below 1, indicating
 strong evidence of downwardly skewed process noise (Fig.~4B,
@@
 tails (median \(\nu \geq 70\)) had little evidence of skewed process noise
-(Fig.~4B; 90\% of 95\% credible intervals overlapped~1). Projecting these
+(Fig.~4B; \percNormPopsNotSkewed\% of 95\% credible intervals overlapped~1). Projecting these
 heavy-tailed populations forward five years revealed that assuming standard
@@
 unmodelled intrinsic properties of populations or extrinsic forces acting on
-populations. For example, we could observe black-swan dynamics if we miss an
+populations. For example, we could have observed black-swan dynamics if we missed an
 underlying mixture of processes, since a mixture of normal distributions with
@@
 we suggest using $\nu = 5$; and
-a skewness parameter ($\gamma$) of $\sim 0.7$ (Fig.~4A)
+a skewness parameter ($\gamma$) of $0.7$ (Fig.~4A)
 if a modeler wishes to allow for downward heavy-tailed events.
@@
 general resilience \cite{carpenter2012}. For instance, setting target
-population abundances far back from critical limits will buffer against black-swan
+population abundances far back from critical limits may buffer against black-swan
 events \cite{caddy1996}, and maintaining genetic, phenotypic, and behavioural
@@
 greatly exceeded. We continued this procedure up to 8,000 iterations
-by which all chains had sufficiently converged.
+by which point all chains had sufficiently converged.
 % The No-U-Turn Hamiltonian Markov chain Monte Carlo sampler in Stan generally
@@
 % Gibbs or Metropolis--Hastings algorithms \cite{stan-manual2015}.
+% \cite{monnahan2016}
 
@@
 We fit our
-models with 5,000 total iterations per chain, 2500 warm-up iterations, four
-chains, and discarding every second sample. We checked for chain convergence
+model with 2,000 total iterations per chain, 1,000 warm-up iterations,
+and four chains. We checked for chain convergence
 visually and with the same criteria as before (\(\widehat{R} < 1.05\) and
 \(n_\mathrm{eff} >200\) for all parameters). Furthermore, we checked our model
-against a similar model with the glmmADMB \textsf{R} package \cite{glmmadmb}
+against a similar maximum likelihood model fit with
+the glmmADMB \textsf{R} package \cite{glmmadmb}
 without uncertainty around the predictors.
@@
 chains of 20,000 iterations and discarded the first 10,000 as warmup for a
-total of 40,000 samples.
+total of 40,000 samples. We then generated five sets of future stochastic
+process deviations for each for each sample
+for a total of 200,000 forecast samples per population.
 
@@
 
-\acknow{We thank J.W. Moore, A.O. Mooers, L.R. Gerber, J.D. Yeakel, C. Minto and
+\acknow{We thank J.W. Moore, A.O. Mooers, L.R. Gerber, J.D. Yeakel, C. Minto,
+  two anonymous reviewers and
 members of the Earth to Ocean Group for helpful discussions and comments. We

### 2016-11-23 analysis/phylopic.tex
Update header in xtable output
Update journal format
Update R version in xtable output
Update phylopic table

@@
-% latex table generated in R 3.1.2 by xtable 1.7-4 package
-% Sat Nov  8 14:57:15 2014
+% latex table generated in R 3.3.2 by xtable 1.8-2 package
 Taxonomic order & Credit & License URL \\ 
   \midrule
-Salmoniformes & Servien (vectorized by T. Michael Keesey) & \url{http://creativecommons.org/licenses/by-sa/3.0/} \\ 
 Gadiformes &  & \url{http://creativecommons.org/publicdomain/mark/1.0/} \\
-  Perciformes & Ellen Edmonson and Hugh Chrisp (vectorized by T. Michael Keesey) & \url{http://creativecommons.org/publicdomain/mark/1.0/} \\ 
+  Salmoniformes & Ellen Edmonson and Hugh Chrisp (vectorized by T. Michael Keesey) & \url{http://creativecommons.org/publicdomain/mark/1.0/} \\
+  Perciformes & Servien (vectorized by T. Michael Keesey) & \url{http://creativecommons.org/licenses/by-sa/3.0/} \\
   Pleuronectiformes & Tony Ayling (vectorized by T. Michael Keesey) & \url{http://creativecommons.org/licenses/by-sa/3.0/} \\ 
   Lepidoptera & Curtis (modified by T. Michael Keesey) & \url{http://creativecommons.org/publicdomain/mark/1.0/} \\ 
+  Odonata & Gareth Monger & \url{http://creativecommons.org/licenses/by/3.0/} \\
+  Coleoptera & Crystal Maier & \url{http://creativecommons.org/licenses/by/3.0/} \\
+  Carnivora & Brian Gratwicke (photo) and T. Michael Keesey (vectorization) & \url{http://creativecommons.org/licenses/by/3.0/} \\ 
   Rodentia & Mattia Menchetti & \url{http://creativecommons.org/publicdomain/zero/1.0/} \\
-  Carnivora & Brian Gratwicke (photo) and T. Michael Keesey (vectorization) & \url{http://creativecommons.org/licenses/by/3.0/} \\ 
   Lagomorpha & Sarah Werning & \url{http://creativecommons.org/licenses/by/3.0/} \\ 
-  Coleoptera & Crystal Maier & \url{http://creativecommons.org/licenses/by/3.0/} \\ 
-  Odonata & Gareth Monger & \url{http://creativecommons.org/licenses/by/3.0/} \\ 
   Passeriformes & Michael Scroggie & \url{http://creativecommons.org/publicdomain/zero/1.0/} \\ 
-  Anseriformes & Sharon Wegner-Larsen & \url{http://creativecommons.org/publicdomain/zero/1.0/} \\ 
-  Artiodactyla & Jan A. Venter, Herbert H. T. Prins, David A. Balfour and Rob Slotow (vectorized by T. Michael Keesey) & \url{http://creativecommons.org/licenses/by/3.0/} \\ 
+  Anseriformes & Jan A. Venter, Herbert H. T. Prins, David A. Balfour and Rob Slotow (vectorized by T. Michael Keesey) & \url{http://creativecommons.org/licenses/by/3.0/} \\
+  Artiodactyla & Sharon Wegner-Larsen & \url{http://creativecommons.org/publicdomain/zero/1.0/} \\
+  Hemiptera & T. Michael Keesey & \url{http://creativecommons.org/publicdomain/zero/1.0/} \\ 
   Diptera & Gareth Monger & \url{http://creativecommons.org/licenses/by/3.0/} \\
+  Falconiformes & Liftarn & \url{http://creativecommons.org/licenses/by-sa/3.0/} \\ 
   Charadriiformes & JJ Harrison (vectorized by T. Michael Keesey) & \url{http://creativecommons.org/licenses/by-sa/3.0/} \\
-  Hemiptera & T. Michael Keesey & \url{http://creativecommons.org/publicdomain/zero/1.0/} \\ 
-  Falconiformes & Liftarn & \url{http://creativecommons.org/licenses/by-sa/3.0/} \\ 
   Galliformes & Steven Traver & \url{http://creativecommons.org/publicdomain/zero/1.0/} \\ 

### 2016-11-24 text/ms_pnas.tex
Tiny edits

@@
 \centering
-\includegraphics[width=0.65\textwidth]{../analysis/t-nu-eg2.pdf}
+\includegraphics[width=0.63\textwidth]{../analysis/t-nu-eg2.pdf}
 \caption{Illustration of population dynamic models that allow for heavy
@@
 \centering
-\includegraphics[width=0.3\textwidth]{../analysis/nu-coefs-2.pdf}
+\includegraphics[width=0.28\textwidth]{../analysis/nu-coefs-2.pdf}
 \caption{Estimates of population dynamic heavy-tailedness for
@@
 which represents the limit distribution of a series of maxima or minima.
-Although the GEV is well-suited for physical ecological variables
+Although the GEV is well-suited for environmental variables
 (e.g.\ sedimentation rates or wind speeds \cite{gaines1993, denny2009}),
@@
 for wild animal population abundance.
-Furthermore, the GEV approach cannot be easily integrated into population dynamics models.
+Furthermore, to our knowledge, the GEV cannot
+be easily integrated into population dynamics models.
 
@@
 posterior, we increased the number of posterior samples in Stan. We ran four
-chains of 20,000 iterations and discarded the first 10,000 as warmup for a
-total of 40,000 samples. We then generated five sets of future stochastic
+chains of 20,000 iterations and discarded the first 10,000 as warmup.
+We then generated five sets of future stochastic
 process deviations for each for each sample
@@
 \acknow{We thank J.W. Moore, A.O. Mooers, L.R. Gerber, J.D. Yeakel, C. Minto,
-  two anonymous reviewers and
+  two anonymous reviewers, and
 members of the Earth to Ocean Group for helpful discussions and comments. We

### 2016-11-24 analysis/6-plot-t-nu-eg.R

@@ -28,7 +28,7 @@ plot(x, dt(x, 2), type = "l", xlab = "",
   ylab = "", yaxs = "i", lty = 1, , lwd = lwd,
   xlim = c(-8, 8), ylim = c(0, 0.42), col = cols[1], xaxt = "n", yaxt = "n")
 text(-7.5, 0.393, "A", font = 2, cex = 1.4)
-mtext("Student-t distribution", side = 3, line = 0, cex = 0.9)
+mtext("Student-t distribution", side = 3, line = 0.1, cex = 0.9)
 axis(2, at = c(0, 0.2, 0.4), col = box.col)
 lines(x, dt(x, 3), lty = 1, col = cols[2], lwd = lwd)
 lines(x, dt(x, 5), lty = 1, col = cols[3], lwd = lwd)
@@ -86,7 +86,7 @@ seed <- nu_5_seeds_N50$seeds[15] # good seed for illustration
 y1 <- make_ts(seed = seed, nu = 3, bs_col = bs_cols[2])
 text(1, 7.1, "C", pos = 4, font = 2, cex = 1.4)
 text(6, 7.1, quote(nu==3), pos = 4)
-mtext("Simulated population dynamics", side = 3, line = 0, cex = 0.9)
+mtext("Simulated population dynamics", side = 3, line = 0.1, cex = 0.9)
 text(13, -3.9, "Black swans", pos = 4)
 seed <- nu_5_seeds_N50$seeds[15]
 y2 <- make_ts(seed = seed, nu = 5, bs_col = bs_cols[3])
@@ -106,19 +106,19 @@ if(!file.exists("illustrate-nu-sampling.rda")) {
     data = list(N = 50, y = y1$y, nu_rate = 0.01,
       b_lower = -1, b_upper = 2),
     pars = c("lambda", "sigma_proc", "nu", "b"), iter = 2000,
-    chains = 4, warmup = 1000)
+    chains = 4, warmup = 1000, control = list(adapt_delta = 0.95))
 
   sm2 <- sampling(stan_gomp,
     data = list(N = 50, y = y2$y, nu_rate = 0.01,
       b_lower = -1, b_upper = 2),
     pars = c("lambda", "sigma_proc", "nu", "b"), iter = 2000,
-    chains = 4, warmup = 1000)
+    chains = 4, warmup = 1000, control = list(adapt_delta = 0.95))
 
   sm3 <- sampling(stan_gomp,
[…]

### 2016-11-24 text/supp.tex
Add random walk model to supplement and add missing bst file
Simplify supplement for submission

@@
 \usepackage{verbatim}
-\usepackage[sectionbib]{chapterbib}
-\usepackage{amsmath}
+% \usepackage[sectionbib]{chapterbib}
 \usepackage{longtable}
@@
 \usepackage{array}
-\usepackage{tabu} % for spacing between rows in longtable
-\usepackage{titlesec}
+% \usepackage{tabu} % for spacing between rows in longtable
+% \usepackage{titlesec}
 \usepackage{url}
@@
 
-\usepackage{lineno}
-\usepackage{xcolor}
-\renewcommand\linenumberfont{\normalfont\tiny\sffamily\color{gray}}
-\modulolinenumbers[2]
+% \usepackage{lineno}
+% \usepackage{xcolor}
+% \renewcommand\linenumberfont{\normalfont\tiny\sffamily\color{gray}}
+% \modulolinenumbers[2]
 
 % Linux Libertine:
-\usepackage{textcomp}
-\usepackage[sb]{libertine}
-\usepackage[varqu,varl]{inconsolata}% sans serif typewriter
-\usepackage[libertine,bigdelims,vvarbb]{newtxmath} % bb from STIX
-\usepackage[cal=boondoxo]{mathalfa} % mathcal
-%\useosf % osf for text, not math
-\usepackage[supstfm=libertinesups,%
-  supscaled=1.2,%
-  raised=-.13em]{superiors}
+% \usepackage{textcomp}
+% \usepackage[sb]{libertine}
+% \usepackage[varqu,varl]{inconsolata}% sans serif typewriter
+% \usepackage[libertine,bigdelims,vvarbb]{newtxmath} % bb from STIX
+% \usepackage[cal=boondoxo]{mathalfa} % mathcal
+% %\useosf % osf for text, not math
+% \usepackage[supstfm=libertinesups,%
+%   supscaled=1.2,%
+%   raised=-.13em]{superiors}
 
@@
 
-% Fix line numbering and align environment
-% http://phaseportrait.blogspot.ca/2007/08/lineno-and-amsmath-compatibility.html
-\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
-  \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
-  \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
-  \renewenvironment{#1}%
-     {\linenomath\csname old#1\endcsname}%
-     {\csname oldend#1\endcsname\endlinenomath}}%
-\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
-  \patchAmsMathEnvironmentForLineno{#1}%
-  \patchAmsMathEnvironmentForLineno{#1*}}%
-\AtBeginDocument{%
-\patchBothAmsMathEnvironmentsForLineno{equation}%
-\patchBothAmsMathEnvironmentsForLineno{align}%
-\patchBothAmsMathEnvironmentsForLineno{flalign}%
-\patchBothAmsMathEnvironmentsForLineno{alignat}%
-\patchBothAmsMathEnvironmentsForLineno{gather}%
-\patchBothAmsMathEnvironmentsForLineno{multline}%
-}
+% % Fix line numbering and align environment
+% % http://phaseportrait.blogspot.ca/2007/08/lineno-and-amsmath-compatibility.html
+% \newcommand*\patchAmsMathEnvironmentForLineno[1]{%
+%   \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
+%   \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
+%   \renewenvironment{#1}%
+%      {\linenomath\csname old#1\endcsname}%
+%      {\csname oldend#1\endcsname\endlinenomath}}%
+% \newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
+%   \patchAmsMathEnvironmentForLineno{#1}%
+%   \patchAmsMathEnvironmentForLineno{#1*}}%
+% \AtBeginDocument{%
+% \patchBothAmsMathEnvironmentsForLineno{equation}%
+% \patchBothAmsMathEnvironmentsForLineno{align}%
+% \patchBothAmsMathEnvironmentsForLineno{flalign}%
+% \patchBothAmsMathEnvironmentsForLineno{alignat}%
+% \patchBothAmsMathEnvironmentsForLineno{gather}%
+% \patchBothAmsMathEnvironmentsForLineno{multline}%
+% }
 
@@
 \author{
-Sean C. Anderson$^{1*}$ \and
+Sean C. Anderson$^{1,2*}$ \and
 Trevor A. Branch$^2$ \and
@@
 
-\textsuperscript{*}Corresponding author: Sean C. Anderson; E-mail: sandrsn@uw.edu; Present address: School of Aquatic and Fishery Sciences, Box 355020, University of Washington, Seattle, WA 98195, USA
+\textsuperscript{*}Corresponding author: Sean C. Anderson; E-mail: sandrsn@uw.edu
 
-\linenumbers
+% \linenumbers
 \onehalfspacing
@@
 \begin{align}
+  \label{eq:rw-drift}
 x_t &= \lambda + x_{t-1} + \epsilon_t\\
@@
 for density dependence \citep{segura2013}.
+
+We also considered a simpler model: a random walk on log abundance without drift,
+\begin{align}
+x_t &= x_{t-1} + \epsilon_t\\
+\epsilon &\sim \mathrm{Student}\mhyphen t(\nu, 0, \sigma).
+\end{align}
+However, the estimates of $\nu$ were qualitatively similar to those from
+Equation \ref{eq:rw-drift} (Fig.~\ref{fig:alt}C) and are therefore not shown.
 
@@
 \section{Code availability.} Data and code to reproduce our analysis are
-available at \url{https://github.com/seananderson/heavy-tails} (currently
-private).
+available at \url{https://github.com/seananderson/heavy-tails}.
 
@@
 
-\bibliographystyle{ecologyletters}\bibliography{jshort,newbib}
+\bibliographystyle{apalike}\bibliography{jshort,newbib}
 
@@
 \begin{scriptsize}
-\begin{longtable}{>{\RaggedRight}m{1.4cm}>{\RaggedRight}p{6.0cm}>{\RaggedRight}p{0.7cm}>{\RaggedRight}p{1.2cm}>{\RaggedRight}p{4.4cm}>{\RaggedRight}p{1.4cm}}
+\begin{longtable}{>{\RaggedRight}m{1.4cm}>{\RaggedRight}p{5.0cm}>{\RaggedRight}p{0.65cm}>{\RaggedRight}p{2.1cm}>{\RaggedRight}p{4.2cm}>{\RaggedRight}p{1.6cm}}
 \caption{Populations with a high probability of heavy-tailed dynamics in the base heavy-tailed Gompertz population dynamics model.
@@
 
-\clearpage
+% \clearpage
 
@@
 the various models in our analysis is available at
-\url{https://github.com/seananderson/heavy-tails} (currently private).
+\url{https://github.com/seananderson/heavy-tails}.
 

### 2016-11-30 text/ms/ms_pnas.tex
Add PNAS ms subfolder

--- 2016-11-24 text/ms_pnas.tex
@@
 
-\usepackage{bm}
+% \usepackage{bm}
 
-\input{../analysis/values} % R output
+% \input{../analysis/values} % R output
+\newcommand{\basePriorMean}{102}
+\newcommand{\basePriorMedian}{71}
+\newcommand{\basePriorProbHeavy}{7.7}
+\newcommand{\medianTimeSteps}{26}
+\newcommand{\meanTimeSteps}{30.2}
+\newcommand{\minTimeSteps}{20}
+\newcommand{\maxTimeSteps}{117}
+\newcommand{\birdN}{191}
+\newcommand{\insectsN}{182}
+\newcommand{\mammalsN}{125}
+\newcommand{\fishN}{108}
+\newcommand{\birdNH}{14}
+\newcommand{\insectsNH}{5}
+\newcommand{\mammalsNH}{6}
+\newcommand{\fishNH}{0}
+\newcommand{\birdPH}{7}
+\newcommand{\insectsPH}{3}
+\newcommand{\mammalsPH}{5}
+\newcommand{\fishPH}{0}
+\newcommand{\NOrdersHeavy}{16}
+\newcommand{\POrdersHeavy}{41}
+\newcommand{\baseFiftyObsFiftySwitch}{8}
+\newcommand{\baseSeventyFiveObsFiftySwitch}{2}
+\newcommand{\totalHeavyFifty}{26}
+\newcommand{\totalHeavySeventyFive}{17}
+\newcommand{\baseFiftyObsFiftySwitchPerc}{31}
+\newcommand{\baseSeventyFiveObsFiftySwitchPerc}{8}
+\newcommand{\baseNuTenObsTenSwitch}{8}
+\newcommand{\baseNuTen}{26}
+\newcommand{\baseNuFiveObsTenSwitch}{2}
+\newcommand{\pHeavyNThirty}{0.13}
+\newcommand{\pHeavyNSixty}{0.20}
+\newcommand{\pIncHeavyNThirtyNSixty}{1.6}
+\newcommand{\obsErrorNuFivePerc}{75}
+\newcommand{\modelsNoConvergeAROne}{1}
+\newcommand{\modelsNoConvergeAROneHeavyBase}{0}
+\newcommand{\percImputedPops}{17}
+\newcommand{\percImputedPoints}{0.7}
+\newcommand{\nuCoefPopN}{606}
+\newcommand{\AvesRangePerc}{4--8}
+\newcommand{\InsectaRangePerc}{2--3}
+\newcommand{\MammaliaRangePerc}{4--6}
+\newcommand{\OsteichthyesRangePerc}{0}
+\newcommand{\overallMinPerc}{3}
+\newcommand{\overallMaxPerc}{5}
+\newcommand{\overallBasePerc}{4}
+\newcommand{\NPops}{609}
+\newcommand{\NOrders}{39}
+\newcommand{\NClasses}{7}
+\newcommand{\interpPointsPerc}{1}
+\newcommand{\nBSUp}{8}
+\newcommand{\nBSDown}{51}
+\newcommand{\ratioBSDownToUp}{6.4}
+\newcommand{\percBSDown}{86}
+\newcommand{\crashUnderRange}{1.1--2}
+\newcommand{\crashUnderMedian}{1.3}
+\newcommand{\probDensSkewedForHeavyPops}{86}
+\newcommand{\percNormPopsNotSkewed}{88}
 
@@
 \keywords{
-Bayesian $|$
 die-offs $|$
 ecological surprises $|$
-Gompertz $|$
-heavy tails $|$
-leptokurtic $|$
 mass mortality $|$
 population dynamics $|$
-risk $|$
-Student-$t$ distribution
+risk
 }
@@
 \centering
-\includegraphics[width=0.63\textwidth]{../analysis/t-nu-eg2.pdf}
+\includegraphics[width=0.63\textwidth]{fig1.pdf}
 \caption{Illustration of population dynamic models that allow for heavy
@@
 \centering
-\includegraphics[width=0.28\textwidth]{../analysis/nu-coefs-2.pdf}
+\includegraphics[width=0.28\textwidth]{fig2.pdf}
 \caption{Estimates of population dynamic heavy-tailedness for
@@
 \centering
-\includegraphics[width=0.26\textwidth]{../analysis/order-posteriors-covariates.pdf}
+\includegraphics[width=0.26\textwidth]{fig3.pdf}
 \caption{Standardized probabilities and covariates of heavy-tailed
@@
 \centering
-\includegraphics[width=0.45\textwidth]{../analysis/skew-fig.pdf}
+\includegraphics[width=0.45\textwidth]{fig4.pdf}
 \caption{Heavy-tailed process noise tends to be downward skewed and
@@
 \matmethods{
-
 \subsection{Data}
@@
   + \alpha^\mathrm{species}_{l[i]}
-  + \bm{X_i} \bm{\beta}),
+  + \pmb{X_i} \pmb{\beta}),
   \: \\
@@
 order \(k\), and species \(l\); \(\phi_\mathrm{disp}\) represents the
-dispersion parameter; and \(\bm{X_i}\) represents a vector of predictors (such as
-lifespan) for population \(i\) with associated \(\bm{\beta}\) coefficients. The
+dispersion parameter; and \(\pmb{X_i}\) represents a vector of predictors (such as
+lifespan) for population \(i\) with associated \(\pmb{\beta}\) coefficients. The
 intercepts are allowed to vary from the overall intercept \(\alpha\) by
@@
 % Bibliography
-\bibliography{jshort,ms}
+% \bibliography{jshort,ms}
+\begin{thebibliography}{10}
+
+\bibitem{sornette2002}
+Sornette D (2002) Predictability of catastrophic events: Material rupture,
+  earthquakes, turbulence, financial crashes, and human birth.
+\newblock {\em Proc. Natl. Acad. Sci. U.S.A.} 99(suppl 1):2522--2529.
+
+\bibitem{albeverio2006}
+Albeverio S, Jentsch V, Kantz H, eds. (2006) {\em Extreme Events in Nature and
+  Society}, Frontiers collection.
+\newblock (Springer, Berlin).
+
+\bibitem{taleb2007}
+Taleb NN (2007) {\em The Black Swan: The Impact of the Highly Improbable}.
+\newblock (Random House).
+
+\bibitem{may2008}
+May RM, Levin SA, Sugihara G (2008) Complex systems: ecology for bankers.
+\newblock {\em Nature} 451(7181):893--895.
+
+\bibitem{sornette2009}
+Sornette D (2009) Dragon-kings, black swans and the prediction of crises.
+\newblock {\em International Journal of Terraspace Science and Engineering}
+  2(1):1--18.
+
+\bibitem{newman2005}
+Newman MEJ (2005) Power laws, {Pareto} distributions and {Zipf}'s law.
+\newblock {\em Contemporary Physics} 46(5):323--351.
+
+\bibitem{fey2015}
+Fey SB et~al. (2015) Recent shifts in the occurrence, cause, and magnitude of
+  animal mass mortality events.
+\newblock {\em Proc. Natl. Acad. Sci. U.S.A.} 112(4):1083--1088.
+
+\bibitem{harnik2012}
+Harnik PG et~al. (2012) Extinctions in ancient and modern seas.
+\newblock {\em Trends Ecol. Evolut.} 27(11):608--617.
+
+\bibitem{gould1977}
+Gould SJ, Eldredge N (1977) Punctuated equilibria: The tempo and mode of
+  evolution reconsidered.
+\newblock {\em Paleobiology} 3(02):115--151.
+
+\bibitem{mangel1994}
+Mangel M, Tier C (1994) Four facts every conservation biologists should know
+  about persistence.
+\newblock {\em Ecology} 75(3):607--614.
+
+\bibitem{ipcc2012}
+{IPCC} (2012) Managing the risks of extreme events and disasters to advance
+  climate change adaptation in {\em A Special Report of Working Groups I and II
+  of the Intergovernmental Panel on Climate Change}, eds.{} Field CB et~al.
+\newblock (Cambridge University Press, Cambridge, UK, and New York, NY, USA).
+
+\bibitem{brook2006a}
+Brook BW, Traill LW, Bradshaw CJA (2006) Minimum viable population sizes and
+  global extinction risk are unrelated.
+\newblock {\em Ecol. Lett.} 9(4):375--382.
+
+\bibitem{knape2012}
+Knape J, de~Valpine P (2012) Are patterns of density dependence in the {Global
+  Population Dynamics Database} driven by uncertainty about population
+  abundance?
+\newblock {\em Ecol. Lett.} 15(1):17--23.
+
+\bibitem{keitt1998}
+Keitt TH, Stanley HE (1998) Dynamics of {North American} breeding bird
+  populations.
+\newblock {\em Nature} 393(6682):257--260.
+
+\bibitem{allen2001}
+Allen A, Li BL, Charnov E (2001) Population fluctuations, power laws and
+  mixtures of lognormal distributions.
+\newblock {\em Ecol. Lett.} 4(1):1--3.
+
+\bibitem{halley2002}
+Halley J, Inchausti P (2002) Lognormality in ecological time series.
+\newblock {\em Oikos} 99(3):518--530.
+
+\bibitem{segura2013}
+Segura AM, Calliari D, Fort H, Lan BL (2013) Fat tails in marine microbial
+  population fluctuations.
+\newblock {\em Oikos} 122(12):1739--1745.
+
+\bibitem{young1994}
+Young TP (1994) Natural die-offs of large mammals: implications for
+  conservation.
+\newblock {\em Conserv. Biol.} 8(2):410--418.
+
+\bibitem{gerber2001}
+Gerber LR, Hilborn R (2001) Catastrophic events and recovery from low densities
+  in populations of otariids: implications for risk of extinction.
+\newblock {\em Mammal. Rev.} 31(2):131--150.
+
+\bibitem{reed2003}
+Reed DH, O'Grady JJ, Ballou JD, Frankham R (2003) The frequency and severity of
+  catastrophic die-offs in vertebrates.
+\newblock {\em Anim. Conserv.} 6(2):109--114.
+
+\bibitem{saucy1994}
+Saucy F (1994) Density dependence in time series of the fossorial form of the
+  water vole, \textit{Arvicola terrestris}.
+\newblock {\em Oikos} 71(3):381--392.
+
+\bibitem{stafford1971}
+Stafford J (1971) The heron population of {England} and {Wales}.
+\newblock {\em Bird Study} 18(4):218--221.
+
+\bibitem{potts1980}
+Potts GR, Coulson JC, Deans IR (1980) Population dynamics and breeding success
+  of the shag, \textit{Phalacrocorax aristotelis}, on the {Farne Islands},
+  {Northumberland}.
+\newblock {\em J. Anim. Ecol.} 49(2):465--484.
+
+\bibitem{gelman2014}
+Gelman A et~al. (2014) {\em Bayesian Data Analysis} ed.{} edition T.
+\newblock (Chapman {\&} Hall, Boca Raton, FL).
+
+\bibitem{schindler2010}
+Schindler DE et~al. (2010) Population diversity and the portfolio effect in an
+  exploited species.
+\newblock {\em Nature} 465(7298):609--612.
+
+\bibitem{carpenter2006}
+Carpenter SR, Brock WA (2006) Rising variance: a leading indicator of
+  ecological transition.
+\newblock {\em Ecol. Lett.} 9(3):311--318.
+
+\bibitem{nunez2012}
+Nu{\~n}ez MA, Logares R (2012) {Black Swans} in ecology and evolution: the
+  importance of improbable but highly influential events.
+\newblock {\em Ideas in Ecology and Evolution} 5:16--21.
+
+\bibitem{meehl2004}
+Meehl GA, Tebaldi C (2004) More intense, more frequent, and longer lasting heat
+  waves in the 21st century.
+\newblock {\em Science} 305(5686):994--997.
+
+\bibitem{katz2005}
+Katz RW, Brush GS, Parlange MB (2005) Statistics of extremes: modelling
+  ecological disturbances.
+\newblock {\em Ecology} 86(5):1124--1134.
+
+\bibitem{kirby2009}
+Kirby RR, Beaugrand G, Lindley JA (2009) Synergistic effects of climate and
+  fishing in a marine ecosystem.
+\newblock {\em Ecosystems} 12(4):548--561.
+
+\bibitem{denny2009}
+Denny MW, Hunt LJH, Miller LP, Harley CDG (2009) On the prediction of extreme
+  ecological events.
+\newblock {\em Ecol. Monograph.} 79(3):397--421.
+
+\bibitem{gpdd2010}
+{NERC Centre for Population Biology, Imperial College} (2010) {\em {The Global
+  Population Dynamics Database Version 2}}.
+
+\bibitem{kot1996}
+Kot M, Lewis MA, {van den Driessche} P (1996) Dispersal data and the spread of
+  invading organisms.
+\newblock {\em Ecology} 77(7):2027--2042.
+
+\bibitem{clark1999}
+Clark JS, Silman M, Kern R, Macklin E, HilleRisLambers J (1999) Seed dispersal
+  near and far: patterns across temperate and tropical forests.
+\newblock {\em Ecology} 80(5):1475--1494.
+
+\bibitem{ward2007}
+Ward EJ, Hilborn R, Towell RG, Gerber L (2007) A state--space mixture approach
+  for estimating catastrophic events in time series data.
+\newblock {\em Can. J. Fish. Aquat. Sci.} 64(6):899--910.
+
+\bibitem{nrc2007}
+{NRC} (2007) {\em Successful Response Starts with a Map: Improving Geospatial
+  Support for Disaster Management}.
+\newblock (The National Academies Press, Washington, {DC}).
+
+\bibitem{carpenter2012}
+{Carpenter} SR et~al. (2012) General resilience to cope with extreme events.
+\newblock {\em Sustainability} 4(12):3248--3259.
+
+\bibitem{caddy1996}
+Caddy JF, McGarvey R (1996) Targets or limits for management of fisheries?
+\newblock {\em N. Am. J. Fish. Manage.} 16(3):479--487.
+
+\bibitem{doak2008}
+Doak DF et~al. (2008) Understanding and predicting ecological dynamics: are
+  major surprises inevitable?
+\newblock {\em Ecology} 89(4):952--961.
+
+\bibitem{lindenmayer2010}
+Lindenmayer DB, Likens GE, Krebs CJ, Hobbs RJ (2010) Improved probability of
+  detection of ecological ``surprises''.
+\newblock {\em Proc. Natl. Acad. Sci. U.S.A.} 107(51):21957--21962.
+
+\bibitem{thompson2013}
+Thompson RM, Beardall J, Beringer J, Grace M, Sardina P (2013) Means and
+  extremes: building variability into community-level climate change
+  experiments.
+\newblock {\em Ecol. Lett.} 16(6):799--806.
+
+\bibitem{gaines1993}
+Gaines SD, Denny MW (1993) The largest, smallest, highest, lowest, longest, and
+  shortest: Extremes in ecology.
+\newblock {\em Ecology} 74(6):1677--1692.
+
+\bibitem{fernandez1998}
+Fernandez C, Steel MFJ (1998) On {Bayesian} modeling of fat tails and skewness.
+\newblock {\em J. Am. Stat. Assoc.} 93(441):359--371.
+
+\bibitem{hoffman2014}
+Hoffman MD, Gelman A (2014) The {No-U-Turn Sampler}: adaptively setting path
+  lengths in {Hamiltonian Monte Carlo}.
+\newblock {\em J. Mach. Learn. Res.} 15:1593--1623.
+
+\bibitem{stan-manual2015}
+{Stan Development Team} (2015) {\em Stan Modeling Language User's Guide and
+  Reference Manual, Version 2.12.0}.
+
+\bibitem{carpenter2016}
+Carpenter B et~al. (2016) Stan: A probabilistic programming language.
+\newblock {\em J. Stat. Soft.} In press.
+
+\bibitem{r2015}
+{R Core Team} (2016) {\em R: A Language and Environment for Statistical
+  Computing} (R Foundation for Statistical Computing, Vienna, Austria).
+
+\bibitem{inchausti2002}
+Inchausti P, Halley J (2002) The long-term temporal variability and spectral
+  colour of animal populations.
+\newblock {\em Evol. Ecol. Res.} 4:1033--1048.
+
+\bibitem{ferrari2004}
+Ferrari S, Cribari-Neto F (2004) Beta regression for modelling rates and
+  proportions.
+\newblock {\em J. Appl. Stat.} 31(7):799--815.
+
+\bibitem{gelman2008c}
+Gelman A (2008) Scaling regression inputs by dividing by two standard
+  deviations.
+\newblock {\em Statist. Med.} 27:2865--2873.
+
+\bibitem{gelman2006c}
+Gelman A (2006) Prior distributions on variance parameters in hierarchical
+  models.
+\newblock {\em Bayesian Analysis} 1(3):515--533.
+
+\bibitem{gelman2008d}
+Gelman A, Jakulin A, Pittau MG, Su YS (2008) A weakly informative default prior
+  distribution for logistic and other regression models.
+\newblock {\em Ann. Appl. Stat.} 2(4):1360--1383.
+
+\bibitem{glmmadmb}
+Skaug H, Fournier D, Bolker B, Magnusson A, Nielsen A (2015) {\em {Generalized
+  Linear Mixed Models using AD Model Builder}}.
+\newblock R package version 0.8.1.
+
+\end{thebibliography}
 

### 2016-11-30 text/copy-sfigs.sh

@@ -1,0 +1,31 @@
+cp ../analysis/gomp-comparison.pdf fig-s1.pdf
+cp ../analysis/gomp-prior-comparison.pdf fig-s2.pdf
+cp ../analysis/t-dist-sampling-sim-prior-exp0point01.pdf fig-s3a.pdf
+cp ../analysis/t-dist-sampling-sim-sigma-prior-exp0point01.pdf fig-s3b.pdf
+cp ../analysis/sim-gompertz-median-dist.pdf fig-s4a.pdf
+cp ../analysis/sim-gompertz-p10.pdf fig-s4b.pdf
+cp ../analysis/correlates-p10.pdf fig-s5.pdf
+cp ../analysis/skew-t-illustration.pdf fig-s6a.pdf
+cp ../analysis/skewness-vs-nu.pdf fig-s6b.pdf
+cp ../analysis/heavy-skew-projections-ggplot.pdf fig-s7.pdf
+cp ../analysis/priors-gomp-base.pdf fig-s8.pdf
+
+cp ../analysis/sparks/6528.pdf .
+cp ../analysis/sparks/7115.pdf .
+cp ../analysis/sparks/10127.pdf .
+cp ../analysis/sparks/9382.pdf .
+cp ../analysis/sparks/10128.pdf .
+cp ../analysis/sparks/10007.pdf .
+cp ../analysis/sparks/20579.pdf .
+cp ../analysis/sparks/9655.pdf .
+cp ../analysis/sparks/1235.pdf .
+cp ../analysis/sparks/9667.pdf .
+cp ../analysis/sparks/10113.pdf .
+cp ../analysis/sparks/9679.pdf .
+cp ../analysis/sparks/20527.pdf .
+cp ../analysis/sparks/10039.pdf .
+cp ../analysis/sparks/10162.pdf .
+cp ../analysis/sparks/7099.pdf .
+cp ../analysis/sparks/9503.pdf .
+cp ../analysis/sparks/2778.pdf .
+cp ../analysis/sparks/9659.pdf .

### 2017-01-07 text/supp.tex
Supplement tweeks

@@
 
+\widowpenalty10000
+\clubpenalty10000
+
 % Linux Libertine:
-% \usepackage{textcomp}
-% \usepackage[sb]{libertine}
-% \usepackage[varqu,varl]{inconsolata}% sans serif typewriter
-% \usepackage[libertine,bigdelims,vvarbb]{newtxmath} % bb from STIX
-% \usepackage[cal=boondoxo]{mathalfa} % mathcal
-% %\useosf % osf for text, not math
-% \usepackage[supstfm=libertinesups,%
-%   supscaled=1.2,%
-%   raised=-.13em]{superiors}
+\usepackage{textcomp}
+\usepackage[sb]{libertine}
+\usepackage[varqu,varl]{inconsolata}% sans serif typewriter
+\usepackage[libertine,bigdelims,vvarbb]{newtxmath} % bb from STIX
+\usepackage[cal=boondoxo]{mathalfa} % mathcal
+%\useosf % osf for text, not math
+\usepackage[supstfm=libertinesups,%
+  supscaled=1.2,%
+  raised=-.13em]{superiors}
 
@@
 
-\usepackage[round,sectionbib]{natbib}
-\bibpunct{(}{)}{;}{a}{}{;}
+% \usepackage[round,sectionbib]{natbib}
+% \bibpunct{(}{)}{;}{a}{}{;}
 
@@
 
-\title{Supporting Information:\\Black-swan events in animal
-  populations}
+\title{Supporting Information Appendix:\\
+  Black-swan events in animal populations}
 
@@
   (on a scale of $1$ to $5$, with $1$ being the lowest quality data)
-  \citep{sibly2005, ziebarth2010}.
+  \cite{sibly2005, ziebarth2010}.
 
@@
   standard deviation. We verified this by locating the original papers the
-  datasets were extracted from: \citet{colebrook1978} for zooplankton and
-  \citet{lindstrom1995} for grouse. Since the papers did not include the
+  datasets were extracted from: Ref.~\cite{colebrook1978} for zooplankton and
+  Ref.~\cite{lindstrom1995} for grouse. Since the papers did not include the
   original mean time-series values we could not back transform these data
@@
 \item We filled in single recorded values of zero with the lowest non-zero
-  value in the time series \citep{brook2006a}. This assumes that
+  value in the time series \cite{brook2006a}. This assumes that
   single values of zero result from abundance being low enough that censusing
@@
 \item We removed all populations without at least four unique values
-  \citep{brook2006a}.
+  \cite{brook2006a}.
 
@@
   length, we took the most recent window. This is a longer window than used in
-  some previous analyses \citep{brook2006a}, but since our model
+  some previous analyses \cite{brook2006a}, but since our model
   attempts to capture the shape of the distribution tails, our model requires
@@
 range of expected values for population growth and is similar to the default
-priors suggested by \citet{gelman2008d} for intercepts of
+priors suggested by Ref.~\cite{gelman2008d} for intercepts of
 regression models. Our half-Cauchy prior on \(\sigma\) follows
-\citet{gelman2006c} and \citet{gelman2008d} and the specific scale parameter of
+Ref.~\cite{gelman2006c} and Ref.~\cite{gelman2008d} and the specific scale parameter of
 \(2.5\) is based on our expected range of the value in nature from previous
-studies \citep{connors2014}. In our testing of a subsample of populations, our
+studies \cite{connors2014}. In our testing of a subsample of populations, our
 parameter estimates were not qualitatively changed by switching to a uniform
@@
 populations as heavy tailed. We considered using a Uniform(\(0, 1\))
-prior on \(1/\nu\) as used in \citet{gelman2014}, but this
+prior on \(1/\nu\) as used in Ref.~\cite{gelman2014}, but this
 prior puts considerably more weight on lower (heavy-tailed) values of \(\nu\)
@@
 and the results are therefore not shown.
-Furthermore, we considered the prior suggested in \citet{juarez2010} of
+Furthermore, we considered the prior suggested in Ref.~\cite{juarez2010} of
 $\nu \sim \mathrm{Gamma}(2, 0.1)$, again limited to $\nu > 2$.
@@
 subsequent residuals. Based on the results of previous analyses with the
-GPDD \citep{connors2014} and the chosen prior in previous
-analyses \citep{thorson2014a}, we placed a moderately informative prior on
+GPDD \cite{connors2014} and the chosen prior in previous
+analyses \cite{thorson2014a}, we placed a moderately informative prior on
 \(\phi\) that assumed the greatest probability density near zero with the
@@
 matches how some previous authors have modelled heavy tails without accounting
-for density dependence \citep{segura2013}.
+for density dependence \cite{segura2013}.
 
@@
 \subsection{Observation error} Observation error can bias parameter
-estimates \citep{knape2012} and is known to affect the ability to detect extreme
-events \citep{ward2007}. In our main analysis, we fit a model that ignored
+estimates \cite{knape2012} and is known to affect the ability to detect extreme
+events \cite{ward2007}. In our main analysis, we fit a model that ignored
 observation error. One way to account for observation error would be to fit a
@@
 known to sometimes result in identifiability issues with the Gompertz
-population model \citep{knape2008}. Furthermore, our model was applied to
+population model \cite{knape2008}. Furthermore, our model was applied to
 hundreds of time series, often of short length (as few as \minTimeSteps\ time
@@
 \(\sigma_\mathrm{obs}\) to \(0.2\), which represents a typical value used in
-simulation analyses \citep{valpine2002, thorson2014b}.
+simulation analyses \cite{valpine2002, thorson2014b}.
 
@@
 zero and twice the maximum observed abundance (similar and less informative
-than in \citet{delean2013}). We set the prior on
-\(r_\mathrm{max}\) as uniform between 0 and 20 as in \citet{delean2013}.
+than in Ref.~\cite{delean2013}). We set the prior on
+\(r_\mathrm{max}\) as uniform between 0 and 20 as in Ref.~\cite{delean2013}.
 We used the same priors on \(\nu\) and \(\sigma\) as in
@@
 The probability density for the skewed Student-$t$ distribution
-\citep{fernandez1998} equates to the following probability density as
-implemented in \citet{king2012}:
+\cite{fernandez1998} equates to the following probability density as
+implemented in Ref.~\cite{king2012}:
 
@@
 Since Stan only needs the probability density up to an additive
-constant \citep{stan-manual2014}, we
+constant \cite{stan-manual2014}, we
 can replace $\log(2 \gamma)$ in the above for $\log(\gamma)$. The final log
@@
 
-\section{Code availability.} Data and code to reproduce our analysis are
-available at \url{https://github.com/seananderson/heavy-tails}.
+\section{Code availability} Data and code to reproduce our analysis are
+available at\\ \url{https://github.com/seananderson/heavy-tails}
 
@@
 
-\bibliographystyle{apalike}\bibliography{jshort,newbib}
+\bibliographystyle{pnas2011}\bibliography{jshort,newbib}
 
@@
 \begin{scriptsize}
-\begin{longtable}{>{\RaggedRight}m{1.4cm}>{\RaggedRight}p{5.0cm}>{\RaggedRight}p{0.65cm}>{\RaggedRight}p{2.1cm}>{\RaggedRight}p{4.2cm}>{\RaggedRight}p{1.6cm}}
+\begin{longtable}{>{\RaggedRight}m{1.4cm}>{\RaggedRight}p{5.3cm}>{\RaggedRight}p{0.65cm}>{\RaggedRight}p{1.5cm}>{\RaggedRight}p{4.2cm}>{\RaggedRight}p{1.6cm}}
 \caption{Populations with a high probability of heavy-tailed dynamics in the base heavy-tailed Gompertz population dynamics model.
@@
 
-\noindent
-Example Stan code for a heavy-tailed Gompertz model. The specific code for
-the various models in our analysis is available at
-\url{https://github.com/seananderson/heavy-tails}.
+% \noindent
+% Example Stan code for a heavy-tailed Gompertz model. The specific code for
+% the various models in our analysis is available at
+% \url{https://github.com/seananderson/heavy-tails}.
 
@@
 %%
-
-\begin{footnotesize}
-\begin{verbatim}
-data {
-  int<lower=1> N;              // number of observations
-  vector[N] y;                 // vector to hold log abundance observations
-  real<lower=0> nu_rate;       // rate parameter for nu exponential prior
-}
-parameters {
-  real lambda;                 // Gompertz growth rate parameter
-  real<lower=-1, upper=2> b;   // Gompertz density dependence parameter
-  real<lower=0> sigma_proc;    // process noise scale parameter
-  real<lower=2> nu;            // t-distribution degrees of freedom
-}
-model {
-  nu ~ exponential(nu_rate);
-  lambda ~ normal(0, 10);
-  sigma_proc ~ cauchy(0, 2.5);
-  for (i in 2:N) {
-    y[i] ~ student_t(nu, lambda + b * y[i-1], sigma_proc);
-  }
-}
-\end{verbatim}
-\end{footnotesize}
+%%
+%% \begin{footnotesize}
+%% \begin{verbatim}
+%% data {
+%%   int<lower=1> N;              // number of observations
+%%   vector[N] y;                 // vector to hold log abundance observations
+%%   real<lower=0> nu_rate;       // rate parameter for nu exponential prior
+%% }
+%% parameters {
+%%   real lambda;                 // Gompertz growth rate parameter
+%%   real<lower=-1, upper=2> b;   // Gompertz density dependence parameter
+%%   real<lower=0> sigma_proc;    // process noise scale parameter
+%%   real<lower=2> nu;            // t-distribution degrees of freedom
+%% }
+%% model {
+%%   nu ~ exponential(nu_rate);
+%%   lambda ~ normal(0, 10);
+%%   sigma_proc ~ cauchy(0, 2.5);
+%%   for (i in 2:N) {
+%%     y[i] ~ student_t(nu, lambda + b * y[i-1], sigma_proc);
+%%   }
+%% }
+%% \end{verbatim}
+%% \end{footnotesize}
+%%
 
@@
 Stan code for the multilevel beta regression:
-\begin{footnotesize}
+\begin{scriptsize}
 \verbatiminput{../analysis/betareg4.stan}
-\end{footnotesize}
+\end{scriptsize}
 

### 2017-01-07 text/ms/ms_pnas.tex
Complete subfolder

--- 2016-11-24 text/ms_pnas.tex
@@
 ecological systems,
-billions of animals can die at once in mass mortality
-events \cite{fey2015},
 background rates of global extinction are punctuated by
 mass extinction \cite{harnik2012},
-and evolution is characterized by
-long periods of stasis punctuated by
-bursts of speciation \cite{gould1977}.
+evolution is characterized by
+long periods of stasis interrupted by
+bursts of speciation \cite{gould1977},
+and billions of animals can die at once in mass mortality
+events \cite{fey2015}.
 
@@
 
-We found that black-swan dynamics are infrequent but highly
-influential: in our dataset they were detected most
+We detected black-swan dynamics most
 frequently for birds (\birdPH\%), mammals (\mammalsPH\%), and
@@
 The majority of our heavy-tailed estimates were robust to alternative
-population models, observation error, and choice of Bayesian priors (Supporting Information).
+population models, observation error, and choice of Bayesian priors (SI Appendix).
 Our conclusions were not systematically altered when we included autocorrelation
@@
 density dependence, or modelled the population dynamics as Ricker-logistic
-instead of Gompertz (Fig.~S\figalt). Similarly, introducing moderate
+instead of Gompertz (SI Appendix, Fig.~S\figalt). Similarly, introducing moderate
 observation error (CV = 0.2) only slightly decreased the estimated prevalence
-of black-swan events (Fig.~S\figalt), and the strength of the prior on
+of black-swan events (SI Appendix, Fig.~S\figalt), and the strength of the prior on
 \(\nu\) had little influence on estimates of black-swan dynamics
-(Fig.~S\figaltpriors). Finally, our simulation testing shows that, if
+(SI Appendix, Fig.~S\figaltpriors). Finally, our simulation testing shows that, if
 anything, our models underpredict the true magnitude and probability of
@@
 with a median of only 26
-years in our analysis (Figs~S\figsimnu,~S\figsimprob).
+years in our analysis (SI Appendix, Figs~S\figsimnu,~S\figsimprob).
 
 % Our results show that weaker and stronger priors have little influence on our
-% conclusions about heavy-tailed dynamics (Fig.~\figaltpriors). When the
+% conclusions about heavy-tailed dynamics (SI Appendix, Fig.~\figaltpriors). When the
 % data are informative about tail behaviour (i.e. when there is strong evidence
-% of low \(\nu\) values, upper-right of Fig.~\figaltpriors), the prior has
+% of low \(\nu\) values, upper-right of SI Appendix, Fig.~\figaltpriors), the prior has
 % little impact on the estimate of \(\nu\). When the data are less informative
@@
 % short or noisy), the prior can pull the estimate of \(\nu\) towards larger or
-% smaller values (Fig.~\figaltpriors). The vast majority of the
+% smaller values (SI Appendix, Fig.~\figaltpriors). The vast majority of the
 % populations with Pr\((\nu < 10)\) in the base prior were not altered
@@
 magnitude of process noise but not clearly related to population growth rate,
-density dependence, or maximum lifespan (Fig.~3B, Fig.~S\figcorrelates).
+density dependence, or maximum lifespan (Fig.~3B, SI Appendix, Fig.~S\figcorrelates).
 Longer time-series length was the strongest covariate of observing black-swan
@@
 event in populations with 60 time steps than in one with 30 time steps
-(Fig.~S\figcorrelates).
+(SI Appendix, Fig.~S\figcorrelates).
 
@@
 events with published explanations (Table~S2), the majority involved
-a combination of multiple factors. For example, a synchronization of
+a combination of factors. For example, a synchronization of
 environmental- and predation-mediated population cycles is thought to have
@@
 strong evidence of downwardly skewed process noise (Fig.~4B,
-Fig.~S\figskewnu). In contrast, populations that did not have heavy
+SI Appendix, Fig.~S\figskewnu). In contrast, populations that did not have heavy
 tails (median \(\nu \geq 70\)) had little evidence of skewed process noise
@@
 abundance) by \crashUnderRange-fold (interquartile range; Fig.~4C,
-Fig.~S\figskewprojections).
+SI Appendix, Fig.~S\figskewprojections).
 
@@
 initially identified as heavy tailed because they turned out to be data-entry
-errors (Supporting Information).
+errors (SI Appendix).
 Third, the temporal scales of observation and population dynamics
@@
 abundance from \(\sim\) 1,000 species and \(\sim\) 100 taxonomic orders. We
-filtered the data (Supporting Information) to remove populations from less
+filtered the data (SI Appendix) to remove populations from less
 reliable data sources, and those without sufficient data for our models, and
@@
 we chose weakly informative priors for
-all parameters (Fig.~S\figpriors, Supporting Information).
+all parameters (SI Appendix, Fig.~S\figpriors).
 For the degrees of freedom parameter $\nu$,
@@
 also fit an alternative model, the Ricker-logistic model, which assumes that
-population growth rate declines linearly with abundance (Supporting Information).
+population growth rate declines linearly with abundance (SI Appendix).
 % Our models also assume that their parameters
@@
 of our results to weaker and stronger priors (exponential rate parameter $=
-0.005, 0.02$) and alternative priors (Fig.~S\figpriors, Supporting Information).
+0.005, 0.02$) and alternative priors (SI Appendix, Fig.~S\figpriors).
 Furthermore, we used
@@
 sizes and to ensure we could recover unbiased parameter estimates from the
-Gompertz model (Supporting Information).
+Gompertz model (SI Appendix).
 
@@
 We log transformed \(\sigma\), time-series length, and lifespan to
-to make the
+make the
 relationship approximately linear on the logit-transformed response scale;
 this matches the way the predictors are visually presented in
-Fig.~S\figcorrelates.
+SI Appendix, Fig.~S\figcorrelates.
 All input variables were standardized by subtracting their mean and dividing by two
@@
 resources. Silhouette images were obtained from \url{phylopic.org} under Creative
-Commons licenses; sources are listed in the Supporting Information. Funding was
+Commons licenses; sources are listed in the SI Appendix. Funding was
 provided by a Simon Fraser University Graduate Fellowship and David H. Smith

### 2017-01-07 analysis/cause-table.tex

--- 2015-07-22 analysis/cause-table.tex
@@
-% latex table generated in R 3.2.0 by xtable 1.7-4 package
+% latex table generated in R 3.3.2 by xtable 1.8-2 package
 Time series & Population & ID & Ref & Description & Pr($\nu < 10$) \\ 
@@
   \includegraphics[width=1.7cm]{../analysis/sparks/7115.pdf} & South African fur seal, \textit{Arctocephalus pusillus}, South Africa & 7115 & \cite{shaughnessy1982} & Harvesting and predation changes & 1.00 \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/10127.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Scotland - un-named area & 10127 & \cite{potts1984} & Environment- and parisite-caused cycles & 0.99 \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/9382.pdf} & Pine looper or Bordered white, \textit{Bupalus piniaria}, Kessock & 9382 & \cite{broekhuizen1993} & Unknown, but sampling intensity was decreasing & 1.00 \\
   \includegraphics[width=1.7cm]{../analysis/sparks/10128.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Scotland - un-named area & 10128 & \cite{potts1984} & Environment- and parisite-caused cycles & 1.00 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/9382.pdf} & Pine looper or Bordered white, \textit{Bupalus piniaria}, Kessock & 9382 & \cite{broekhuizen1993} & Unknown, but sampling intensity was decreasing & 1.00 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/10127.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Scotland - un-named area & 10127 & \cite{potts1984} & Environment- and parisite-caused cycles & 0.99 \\
   \includegraphics[width=1.7cm]{../analysis/sparks/10007.pdf} & Water vole, \textit{Arvicola terrestris}, Le Pont & 10007 & \cite{saucy1994} & Predator-environment cycle interactions & 1.00 \\
@@
   \includegraphics[width=1.7cm]{../analysis/sparks/1235.pdf} & Wren, \textit{Troglodytes troglodytes}, Eastern Wood, Bookham Common & 1235 & \cite{newton1998} & Severe winter & 0.98 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/10113.pdf} & Willow grouse, \textit{Lagopus lagopus}, Northern England & 10113 & \cite{dobson1995} & Parasites and predators & 0.99 \\ 
   \includegraphics[width=1.7cm]{../analysis/sparks/9667.pdf} & Gooseberry sawfly, \textit{Nemastus ribesii}, Finland & 9667 & \cite{markkula1965} & Cannot locate original source & 0.97 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/9679.pdf} & Unknown, \textit{Trioza apicalis}, Finland & 9679 & \cite{markkula1965} & Cannot locate original source & 0.83 \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/10113.pdf} & Willow grouse, \textit{Lagopus lagopus}, Northern England & 10113 & \cite{dobson1995} & Parasites and predators & 0.98 \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/9679.pdf} & Unknown, \textit{Trioza apicalis}, Finland & 9679 & \cite{markkula1965} & Cannot locate original source & 0.86 \\
   \includegraphics[width=1.7cm]{../analysis/sparks/20527.pdf} & Wandering albatross, \textit{Diomedea exulans}, Taiaroa & 20527 & \cite{robertson1998} & Unknown & 0.99 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/10039.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Northern Scotland & 10039 & \cite{dobson1995} & Parasites and predators & 0.92 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/10162.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Atholl Estate & 10162 & \cite{mackenzie1952} & Bad environmental conditions and overcrowding combined to create crashes & 0.91 \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/10039.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Northern Scotland & 10039 & \cite{dobson1995} & Parasites and predators & 0.93 \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/10162.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Atholl Estate & 10162 & \cite{mackenzie1952} & Bad environmental conditions and overcrowding combined to create crashes & 0.92 \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/7099.pdf} & European rabbit, \textit{Oryctolagus cuniculus}, Estate 2, East Anglia & 7099 & \cite{barnes1986} & Disease outbreak followed by years of good weather & 0.78 \\
   \includegraphics[width=1.7cm]{../analysis/sparks/9503.pdf} & Fisher or  Pekan, \textit{Martes pennanti}, Manitoba & 9503 & \cite{keith1963} & Unknown & 0.76 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/7099.pdf} & European rabbit, \textit{Oryctolagus cuniculus}, Estate 2, East Anglia & 7099 & \cite{barnes1986} & Disease outbreak followed by years of good weather & 0.78 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/2778.pdf} & Wheatear, \textit{Oenanthe oenanthe}, Skokholm Island & 2778 & \cite{lack1969} & Unknown, but decline noted specifically, cold winters caused some crashes & 0.67 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/1195.pdf} & Blue jay, \textit{Cyanocitta cristata}, Robert Allerton Park & 1195 & \cite{kendeigh1982} & Unknown & 0.60 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/5019.pdf} & Barbary macaque, \textit{Macaca sylvanus}, Queens Gate & 5019 & \cite{fa1984} & Cannot locate original source & 0.61 \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/2778.pdf} & Wheatear, \textit{Oenanthe oenanthe}, Skokholm Island & 2778 & \cite{lack1969} & Unknown, but decline noted specifically, cold winters caused some crashes & 0.66 \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/9659.pdf} & Cabbage root fly or maggot, \textit{Delia radicum}, Finland & 9659 & \cite{markkula1965} & Cannot locate original source & 0.60 \\
    \bottomrule

### 2017-01-13 analysis/phylopic.tex
Update cached files

--- 2014-11-08 analysis/phylopic.tex
@@
-% latex table generated in R 3.1.2 by xtable 1.7-4 package
-% Sat Nov  8 14:57:15 2014
+% latex table generated in R 3.3.1 by xtable 1.8-2 package
 Taxonomic order & Credit & License URL \\ 
   \midrule
+Gadiformes &  & \url{http://creativecommons.org/publicdomain/mark/1.0/} \\ 
   Salmoniformes & Servien (vectorized by T. Michael Keesey) & \url{http://creativecommons.org/licenses/by-sa/3.0/} \\ 
-  Gadiformes &  & \url{http://creativecommons.org/publicdomain/mark/1.0/} \\ 
   Perciformes & Ellen Edmonson and Hugh Chrisp (vectorized by T. Michael Keesey) & \url{http://creativecommons.org/publicdomain/mark/1.0/} \\ 
@@
   Lepidoptera & Curtis (modified by T. Michael Keesey) & \url{http://creativecommons.org/publicdomain/mark/1.0/} \\ 
+  Carnivora & Brian Gratwicke (photo) and T. Michael Keesey (vectorization) & \url{http://creativecommons.org/licenses/by/3.0/} \\ 
+  Coleoptera & Crystal Maier & \url{http://creativecommons.org/licenses/by/3.0/} \\ 
   Rodentia & Mattia Menchetti & \url{http://creativecommons.org/publicdomain/zero/1.0/} \\ 
-  Carnivora & Brian Gratwicke (photo) and T. Michael Keesey (vectorization) & \url{http://creativecommons.org/licenses/by/3.0/} \\ 
+  Odonata & Gareth Monger & \url{http://creativecommons.org/licenses/by/3.0/} \\ 
   Lagomorpha & Sarah Werning & \url{http://creativecommons.org/licenses/by/3.0/} \\ 
-  Coleoptera & Crystal Maier & \url{http://creativecommons.org/licenses/by/3.0/} \\ 
-  Odonata & Gareth Monger & \url{http://creativecommons.org/licenses/by/3.0/} \\ 
+  Anseriformes & Sharon Wegner-Larsen & \url{http://creativecommons.org/publicdomain/zero/1.0/} \\ 
   Passeriformes & Michael Scroggie & \url{http://creativecommons.org/publicdomain/zero/1.0/} \\ 
-  Anseriformes & Sharon Wegner-Larsen & \url{http://creativecommons.org/publicdomain/zero/1.0/} \\ 
   Artiodactyla & Jan A. Venter, Herbert H. T. Prins, David A. Balfour and Rob Slotow (vectorized by T. Michael Keesey) & \url{http://creativecommons.org/licenses/by/3.0/} \\ 

### 2017-01-13 analysis/cause-table.tex
Update table

--- 2016-11-17 analysis/cause-table.tex
@@
   \includegraphics[width=1.7cm]{../analysis/sparks/10128.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Scotland - un-named area & 10128 & \citep{potts1984} & Environment- and parisite-caused cycles & 1.00 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/10007.pdf} & Water vole, \textit{Arvicola terrestris}, Le Pont & 10007 & \citep{saucy1994} & Predator-environment cycle interactions & 1.00 \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/10007.pdf} & Water vole, \textit{Arvicola terrestris}, Le Pont & 10007 & \citep{saucy1994} & Predator-environment cycle interactions & 0.99 \\ 
   \includegraphics[width=1.7cm]{../analysis/sparks/20579.pdf} & Grey heron, \textit{Ardea cinerea}, Southern Britain & 20579 & \citep{stafford1971} & Severe winter & 0.99 \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/9655.pdf} & Flea beetle, \textit{Chaetocnoma concinna}, Finland & 9655 & \citep{markkula1965} & Cannot locate original source & 0.99 \\ 
   \includegraphics[width=1.7cm]{../analysis/sparks/1235.pdf} & Wren, \textit{Troglodytes troglodytes}, Eastern Wood, Bookham Common & 1235 & \citep{newton1998} & Severe winter & 0.98 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/9655.pdf} & Flea beetle, \textit{Chaetocnoma concinna}, Finland & 9655 & \citep{markkula1965} & Cannot locate original source & 0.99 \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/9667.pdf} & Gooseberry sawfly, \textit{Nemastus ribesii}, Finland & 9667 & \citep{markkula1965} & Cannot locate original source & 0.97 \\ 
   \includegraphics[width=1.7cm]{../analysis/sparks/10113.pdf} & Willow grouse, \textit{Lagopus lagopus}, Northern England & 10113 & \citep{dobson1995} & Parasites and predators & 0.99 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/9667.pdf} & Gooseberry sawfly, \textit{Nemastus ribesii}, Finland & 9667 & \citep{markkula1965} & Cannot locate original source & 0.97 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/9679.pdf} & Unknown, \textit{Trioza apicalis}, Finland & 9679 & \citep{markkula1965} & Cannot locate original source & 0.85 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/20527.pdf} & Wandering albatross, \textit{Diomedea exulans}, Taiaroa & 20527 & \citep{robertson1998} & Unknown & 1.00 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/10039.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Northern Scotland & 10039 & \citep{dobson1995} & Parasites and predators & 0.92 \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/9679.pdf} & Unknown, \textit{Trioza apicalis}, Finland & 9679 & \citep{markkula1965} & Cannot locate original source & 0.83 \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/20527.pdf} & Wandering albatross, \textit{Diomedea exulans}, Taiaroa & 20527 & \citep{robertson1998} & Unknown & 0.99 \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/10039.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Northern Scotland & 10039 & \citep{dobson1995} & Parasites and predators & 0.93 \\ 
   \includegraphics[width=1.7cm]{../analysis/sparks/10162.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Atholl Estate & 10162 & \citet{mackenzie1952} & Bad environmental conditions and overcrowding combined to create crashes & 0.92 \\
   \includegraphics[width=1.7cm]{../analysis/sparks/7099.pdf} & European rabbit, \textit{Oryctolagus cuniculus}, Estate 2, East Anglia & 7099 & \citep{barnes1986} & Disease outbreak followed by years of good weather & 0.81 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/9503.pdf} & Fisher or  Pekan, \textit{Martes pennanti}, Manitoba & 9503 & \citep{keith1963} & Unknown & 0.74 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/2778.pdf} & Wheatear, \textit{Oenanthe oenanthe}, Skokholm Island & 2778 & \citep{lack1969} & Unknown, but decline noted specifically, cold winters caused some crashes & 0.68 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/1195.pdf} & Blue jay, \textit{Cyanocitta cristata}, Robert Allerton Park & 1195 & \citep{kendeigh1982} & Unknown & 0.62 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/9953.pdf} & Rock ptarmigan, \textit{Lagopus mutus}, Iceland & 9953 & \citep{clarke1885,williams1954} & Severe winters & 0.61 \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/9503.pdf} & Fisher or  Pekan, \textit{Martes pennanti}, Manitoba & 9503 & \citep{keith1963} & Unknown & 0.75 \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/2778.pdf} & Wheatear, \textit{Oenanthe oenanthe}, Skokholm Island & 2778 & \citep{lack1969} & Unknown, but decline noted specifically, cold winters caused some crashes & 0.67 \\
    \bottomrule

### 2017-01-13 text/ms/ms_pnas.tex
Resubmit

@@
 \newcommand{\fishPH}{0}
-\newcommand{\NOrdersHeavy}{16}
-\newcommand{\POrdersHeavy}{41}
+\newcommand{\propSkewedAves}{0.21}
+\newcommand{\propSkewedInsecta}{0.18}
+\newcommand{\propSkewedMammalia}{0.31}
+\newcommand{\propSkewedOsteichthyes}{0.21}
+\newcommand{\NOrdersHeavy}{15}
+\newcommand{\POrdersHeavy}{38}
 \newcommand{\baseFiftyObsFiftySwitch}{8}
@@
 \newcommand{\pHeavyNSixty}{0.20}
-\newcommand{\pIncHeavyNThirtyNSixty}{1.6}
+\newcommand{\pIncHeavyNThirtyNSixty}{1.5}
 \newcommand{\obsErrorNuFivePerc}{75}
@@
 \newcommand{\nBSUp}{8}
-\newcommand{\nBSDown}{51}
-\newcommand{\ratioBSDownToUp}{6.4}
+\newcommand{\nBSDown}{49}
+\newcommand{\ratioBSDownToUp}{6.1}
 \newcommand{\percBSDown}{86}
-\newcommand{\crashUnderRange}{1.1--2}
-\newcommand{\crashUnderMedian}{1.3}
+\newcommand{\crashUnderRange}{1.2--1.9}
+\newcommand{\crashUnderMedian}{1.4}
 \newcommand{\probDensSkewedForHeavyPops}{86}
-\newcommand{\percNormPopsNotSkewed}{88}
+\newcommand{\percNormPopsNotSkewed}{89}
 
@@
 \(\sigma\) and degrees of freedom \(\nu\). We can then estimate the degree to
-which the process deviations have heavy tails and are therefore evidence of
-black-swan events (Fig.~1A, B).
+which the process deviations have heavy tails (i.e., heavier tails than the
+lognormal distribution) and are therefore evidence of black-swan events
+(Fig.~1A, B).
 
@@
 
-We fit our models with Stan 2.12.1 \cite{hoffman2014, stan-manual2015, carpenter2016}, and R
+We fit our models with Stan 2.14.1 \cite{hoffman2014, stan-manual2015, carpenter2016}, and R
 3.3.1 \cite{r2015}. We began with four chains and 2,000 iterations, discarding

### 2017-01-13 text/supp.tex
Resubmit

@@
 probability density for the skewed Student-$t$ distribution implemented in
-Stan\footnote{Credit to Stan developer Bob Carpenter:
+Stan\footnote{Credit to Stan developer Bob Carpenter:\\
 \url{https://groups.google.com/d/msg/stan-users/jeZQnQRUsDs/BeqhicWm1GEJ}}
@@
 \end{verbatim}
+
+In our skew-Student-$t$ models, populations tended to have downward-skewed
+process noise (Figs 4B, \ref{fig:skew-nu}). The proportions of populations with
+Pr($\gamma < 1$) $> 0.95$
+(i.e.\ strong evidence of downward-skewed process noise)
+were
+\propSkewedAves\ (birds), \propSkewedInsecta\ (insects),
+\propSkewedMammalia\ (mammals), and \propSkewedOsteichthyes\ (fishes).
 

### 2017-02-17 analysis/phylopic.tex
Switch silhouettes

@@
   \midrule
-Gadiformes &  & \url{http://creativecommons.org/publicdomain/mark/1.0/} \\ 
-  Salmoniformes & Servien (vectorized by T. Michael Keesey) & \url{http://creativecommons.org/licenses/by-sa/3.0/} \\ 
+Gadiformes & Uncredited & \url{http://creativecommons.org/publicdomain/mark/1.0/} \\
+  % Salmoniformes & Servien (vectorized by T. Michael Keesey) & \url{http://creativecommons.org/licenses/by-sa/3.0/} \\
   Perciformes & Ellen Edmonson and Hugh Chrisp (vectorized by T. Michael Keesey) & \url{http://creativecommons.org/publicdomain/mark/1.0/} \\
-  Pleuronectiformes & Tony Ayling (vectorized by T. Michael Keesey) & \url{http://creativecommons.org/licenses/by-sa/3.0/} \\ 
+  Pleuronectiformes & Birgit Lang & \url{http://creativecommons.org/licenses/by/3.0/} \\
   Lepidoptera & Curtis (modified by T. Michael Keesey) & \url{http://creativecommons.org/publicdomain/mark/1.0/} \\
@@
   Diptera & Gareth Monger & \url{http://creativecommons.org/licenses/by/3.0/} \\
-  Charadriiformes & JJ Harrison (vectorized by T. Michael Keesey) & \url{http://creativecommons.org/licenses/by-sa/3.0/} \\ 
+  Charadriiformes & Rebecca Groom & \url{http://creativecommons.org/licenses/by/3.0/} \\
   Hemiptera & T. Michael Keesey & \url{http://creativecommons.org/publicdomain/zero/1.0/} \\
-  Falconiformes & Liftarn & \url{http://creativecommons.org/licenses/by-sa/3.0/} \\ 
+  Falconiformes & Ferran Sayol & \url{http://creativecommons.org/publicdomain/zero/1.0/} \\
   Galliformes & Steven Traver & \url{http://creativecommons.org/publicdomain/zero/1.0/} \\
    \bottomrule
+
+% Salmoniformes
+% Lepidoptera http://phylopic.org/image/0fde2bb6-0472-4273-bf46-2d6073fa8fbc/
+% Pleuronectiformes http://phylopic.org/image/4c2220b9-68a8-4594-8dd8-229643e11c84/
+
+     % taxonomic_order                                                             url
+%1         Gadiformes http://phylopic.org/image/8d92b454-3131-4bbd-ac9c-e1df14c2fc5a/
+%3        Perciformes http://phylopic.org/image/426732d4-17c9-4e85-9672-b6b7f47eb99d/
+%4       Anseriformes http://phylopic.org/image/cf522e02-35cc-44f5-841c-0e642987c2e4/
+%5        Lepidoptera http://phylopic.org/image/0fde2bb6-0472-4273-bf46-2d6073fa8fbc/
+%6          Carnivora http://phylopic.org/image/20da6c7c-2584-4cee-921b-ebd09384567b/
+%7           Rodentia http://phylopic.org/image/bd3d2bc1-915c-4fa6-99b8-9a5fc6b725e0/
+%8         Lagomorpha http://phylopic.org/image/dea688b6-9168-4e79-a106-366888148eb1/
+%9         Coleoptera http://phylopic.org/image/77164e4b-8076-447b-a140-7a1f9f907da8/
+%10           Odonata http://phylopic.org/image/ce57b692-e1e0-468e-a025-8c10ceebca22/
+%11     Passeriformes http://phylopic.org/image/3d38baad-d355-432a-944c-07608828d0a4/
+%12      Artiodactyla http://phylopic.org/image/bb357997-4d4d-4e73-b853-d95a7a47ec68/
+%13 Pleuronectiformes http://phylopic.org/image/4c2220b9-68a8-4594-8dd8-229643e11c84/
+%14     Falconiformes http://phylopic.org/image/f68e0e0e-ed16-472c-9bb7-fb4b40f7fa9d/
+%15   Charadriiformes http://phylopic.org/image/6b77616d-7e99-488a-97e0-c6e919c62ec9/
+%16       Galliformes http://phylopic.org/image/2de1c95c-7e1f-429b-9c08-17f0a27d176f/
+%17           Diptera http://phylopic.org/image/fdd9c1c9-7391-4e2b-af4e-a0d289fa9466/
+%18         Hemiptera http://phylopic.org/image/3010e744-6b06-4022-9ce6-354db93e99fe/
+
+% Accipiter Brisson 1760  Accipitridae  Accipitriformes Savigny 1809  Accipitrimorphae Accipitrinae  Afroaves  Telluraves
+% http://phylopic.org/image/f52d11cf-1d3c-41ae-b37e-e46c822a71b3/
+% http://creativecommons.org/licenses/by/3.0/
+% Shyamal
+
+% http://phylopic.org/image/f68e0e0e-ed16-472c-9bb7-fb4b40f7fa9d/
+% http://creativecommons.org/publicdomain/zero/1.0/
+% Ferran Sayol

### 2017-02-17 analysis/6-plot-order-correlate-posteriors.R

@@ -20,9 +20,12 @@ or <- mutate(or,
     hash, ".256.png"),
   scaling_factor = rep(1, nrow(or)))
 or[or$taxonomic_order == "Gadiformes", "scaling_factor"] <- 0.6
-or[or$taxonomic_order == "Salmoniformes", "scaling_factor"] <- 0.6
+or[or$taxonomic_order == "Salmoniformes", "scaling_factor"] <- 0.9
 or[or$taxonomic_order == "Perciformes", "scaling_factor"] <- 0.6
 or[or$taxonomic_order == "Pleuronectiformes", "scaling_factor"] <- 0.8
+or[or$taxonomic_order == "Falconiformes", "scaling_factor"] <- 0.8
+or[or$taxonomic_order == "Charadriiformes", "scaling_factor"] <- 0.8
+
 # to get ordering right:
 or <- plyr::join(a_df[,c("taxonomic_order", "sorted_order")], or)
 
@@ -48,16 +51,16 @@ or_table$credit <- gsub("&", "and", or_table$credit)
 names(or_table) <- c("Taxonomic order", "Credit", "License URL", "URL")
 or_table$URL <- NULL # don't use after all
 
-print.xtable(xtable(or_table, caption = ""),
-  include.rownames = FALSE, file = "phylopic.tex", booktabs = TRUE,
-  sanitize.text.function = identity, only.contents = TRUE, timestamp = NULL)
-
-if(any(!file.exists(paste0("silhouettes/", or$taxonomic_order, ".png")))) {
-  for(i in 1:nrow(or)) {
-    download.file(or$png_url[i],
-      destfile = paste0("silhouettes/", or$taxonomic_order[i], ".png"))
-  }
-}
+# print.xtable(xtable(or_table, caption = ""),
+#   include.rownames = FALSE, file = "phylopic.tex", booktabs = TRUE,
+#   sanitize.text.function = identity, only.contents = TRUE, timestamp = NULL)
+#
+# if(any(!file.exists(paste0("silhouettes/", or$taxonomic_order, ".png")))) {
+#   for(i in 1:nrow(or)) {
+#     download.file(or$png_url[i],
+#       destfile = paste0("silhouettes/", or$taxonomic_order[i], ".png"))
+#   }
+# }
 
 op <- vector(mode = "list", length = nrow(a_df))
 for(i in seq_along(op)) {

### 2017-02-17 analysis/cause-table.tex
Update causes table

--- 2017-01-07 analysis/cause-table.tex
@@
   \includegraphics[width=1.7cm]{../analysis/sparks/7115.pdf} & South African fur seal, \textit{Arctocephalus pusillus}, South Africa & 7115 & \cite{shaughnessy1982} & Harvesting and predation changes & 1.00 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/10127.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Scotland - un-named area & 10127 & \cite{potts1984} & Environment- and parisite-caused cycles & 0.99 \\
   \includegraphics[width=1.7cm]{../analysis/sparks/9382.pdf} & Pine looper or Bordered white, \textit{Bupalus piniaria}, Kessock & 9382 & \cite{broekhuizen1993} & Unknown, but sampling intensity was decreasing & 1.00 \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/10127.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Scotland - un-named area & 10127 & \cite{potts1984} & Environment- and parisite-caused cycles & 1.00 \\
   \includegraphics[width=1.7cm]{../analysis/sparks/10128.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Scotland - un-named area & 10128 & \cite{potts1984} & Environment- and parisite-caused cycles & 1.00 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/10007.pdf} & Water vole, \textit{Arvicola terrestris}, Le Pont & 10007 & \cite{saucy1994} & Predator-environment cycle interactions & 1.00 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/20579.pdf} & Grey heron, \textit{Ardea cinerea}, Southern Britain & 20579 & \cite{stafford1971} & Severe winter & 0.98 \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/10007.pdf} & Water vole, \textit{Arvicola terrestris}, Le Pont & 10007 & \cite{saucy1994} & Predator-environment cycle interactions & 0.99 \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/20579.pdf} & Grey heron, \textit{Ardea cinerea}, Southern Britain & 20579 & \cite{stafford1971} & Severe winter & 0.99 \\ 
   \includegraphics[width=1.7cm]{../analysis/sparks/9655.pdf} & Flea beetle, \textit{Chaetocnoma concinna}, Finland & 9655 & \cite{markkula1965} & Cannot locate original source & 0.99 \\ 
@@
   \includegraphics[width=1.7cm]{../analysis/sparks/9667.pdf} & Gooseberry sawfly, \textit{Nemastus ribesii}, Finland & 9667 & \cite{markkula1965} & Cannot locate original source & 0.97 \\ 
-  \includegraphics[width=1.7cm]{../analysis/sparks/10113.pdf} & Willow grouse, \textit{Lagopus lagopus}, Northern England & 10113 & \cite{dobson1995} & Parasites and predators & 0.98 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/9679.pdf} & Unknown, \textit{Trioza apicalis}, Finland & 9679 & \cite{markkula1965} & Cannot locate original source & 0.86 \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/10113.pdf} & Willow grouse, \textit{Lagopus lagopus}, Northern England & 10113 & \cite{dobson1995} & Parasites and predators & 0.99 \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/9679.pdf} & Unknown, \textit{Trioza apicalis}, Finland & 9679 & \cite{markkula1965} & Cannot locate original source & 0.83 \\ 
   \includegraphics[width=1.7cm]{../analysis/sparks/20527.pdf} & Wandering albatross, \textit{Diomedea exulans}, Taiaroa & 20527 & \cite{robertson1998} & Unknown & 0.99 \\ 
@@
   \includegraphics[width=1.7cm]{../analysis/sparks/10162.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Atholl Estate & 10162 & \cite{mackenzie1952} & Bad environmental conditions and overcrowding combined to create crashes & 0.92 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/7099.pdf} & European rabbit, \textit{Oryctolagus cuniculus}, Estate 2, East Anglia & 7099 & \cite{barnes1986} & Disease outbreak followed by years of good weather & 0.78 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/9503.pdf} & Fisher or  Pekan, \textit{Martes pennanti}, Manitoba & 9503 & \cite{keith1963} & Unknown & 0.76 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/2778.pdf} & Wheatear, \textit{Oenanthe oenanthe}, Skokholm Island & 2778 & \cite{lack1969} & Unknown, but decline noted specifically, cold winters caused some crashes & 0.66 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/9659.pdf} & Cabbage root fly or maggot, \textit{Delia radicum}, Finland & 9659 & \cite{markkula1965} & Cannot locate original source & 0.60 \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/7099.pdf} & European rabbit, \textit{Oryctolagus cuniculus}, Estate 2, East Anglia & 7099 & \cite{barnes1986} & Disease outbreak followed by years of good weather & 0.81 \\
+  \includegraphics[width=1.7cm]{../analysis/sparks/9503.pdf} & Fisher or  Pekan, \textit{Martes pennanti}, Manitoba & 9503 & \cite{keith1963} & Unknown & 0.75 \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/2778.pdf} & Wheatear, \textit{Oenanthe oenanthe}, Skokholm Island & 2778 & \cite{lack1969} & Unknown, but decline noted specifically, cold winters caused some crashes & 0.67 \\
    \bottomrule

### 2017-02-21 analysis/cause-table.tex
Correct species name

@@
   \includegraphics[width=1.7cm]{../analysis/sparks/10128.pdf} & Red grouse, \textit{Lagopus lagopus scoticus}, Scotland - un-named area & 10128 & \cite{potts1984} & Environment- and parisite-caused cycles & 1.00 \\
-  \includegraphics[width=1.7cm]{../analysis/sparks/10007.pdf} & Water vole, \textit{Arvicola terrestris}, Le Pont & 10007 & \cite{saucy1994} & Predator-environment cycle interactions & 0.99 \\ 
+  \includegraphics[width=1.7cm]{../analysis/sparks/10007.pdf} & Water vole, \textit{Arvicola amphibius}, Le Pont & 10007 & \cite{saucy1994} & Predator-environment cycle interactions & 0.99 \\ 
   \includegraphics[width=1.7cm]{../analysis/sparks/20579.pdf} & Grey heron, \textit{Ardea cinerea}, Southern Britain & 20579 & \cite{stafford1971} & Severe winter & 0.99 \\ 

### 2017-02-21 text/supp.tex
Tweak formatting in supplement
Fix newer Stan syntax

@@
   if (y < mu)
-    return lp + student_t_log(y * skew, nu, mu * skew, sigma);
+    return lp + student_t_lpdf(y * skew | nu, mu * skew, sigma);
   else
-    return lp + student_t_log(y / skew, nu, mu / skew, sigma);
+    return lp + student_t_lpdf(y / skew | nu, mu / skew, sigma);
 }
@@
 \url{https://github.com/sckott/rphylopic}):
+
+\clearpage
 
@@
 \smallskip
-\begin{scriptsize}
+\begin{footnotesize}
 \begin{tabular}{lrrrrrrrr}
@@
 \end{tabular}
-\end{scriptsize}
+\end{footnotesize}
 %\end{helvetica}
@@
 \begin{longtable}{>{\RaggedRight}m{1.4cm}>{\RaggedRight}p{5.3cm}>{\RaggedRight}p{0.65cm}>{\RaggedRight}p{1.5cm}>{\RaggedRight}p{4.2cm}>{\RaggedRight}p{1.6cm}}
-\caption{Populations with a high probability of heavy-tailed dynamics in the base heavy-tailed Gompertz population dynamics model.
+\caption{Populations with a high ($> 0.6$) probability of heavy-tailed dynamics in the base heavy-tailed Gompertz population dynamics model.
 Shown are the log abundance time series, population descriptions, Global Population Dynamics Database Main IDs, citation for the data source or separate verification literature, a description of the cause of the black-swan events (if known), the probability of heavy tails as calculated by our model, and median estimate of $\nu$ from our model with 90\% quantile credible intervals indicated in parentheses.

### 2017-02-21 analysis/1-compile-models.R

@@ -72,9 +72,9 @@ functions {
     reject("Scale has to be positive.  Found sigma=", sigma);
   lp = log(skew) - log1p(square(skew));
   if (y < mu)
-    return lp + student_t_lpdf(y * skew, nu, mu * skew, sigma);
+    return lp + student_t_lpdf(y * skew | nu, mu * skew, sigma);
   else
-    return lp + student_t_lpdf(y / skew, nu, mu / skew, sigma);
+    return lp + student_t_lpdf(y / skew | nu, mu / skew, sigma);
   }
 }
 data {

### 2017-02-24 README.Rmd
Add package versions to README

---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

# Black-swan events in animal populations

This repository contains code for an analysis of heavy-tailed
population dynamics for animal populations. It accompanies the paper:

Anderson, S.C., T.B. Branch, A.B. Cooper, N.K. Dulvy. Black-swan
events in animal populations. In press at Proceedings of the 
National Academy of Sciences.

To recreate the analysis, first you will need the following R
packages installed:

```{r, eval=FALSE}
install.packages(c("rstan", "dplyr", "plyr", "reshape2", "ggplot2", "gridExtra", 
  "RColorBrewer", "grImport", "TeachingDemos", "metRology", "xtable", "devtools",
  "skewt", "foreach", "Rcpp", "png"))
devtools::install_github("sckott/rphylopic")
```

Theoretically, you should be able to run the entire analysis by sourcing the following R file:

```{r, eval=FALSE}
source("analysis/make.R")
```

However, many of the analyses will take a very long time to run (multiple days), and you might run into minor issues with changes to R packages over time. Therefore, it is probably more useful to open the file `analysis/make.R` and work through the analysis files as they are called. Alternatively, the `.R` files are named in the order they should be run. Files starting with the same number can be run in any order. Output from a number of the Stan models is cached in `.rds` files that will be available with a freshly cloned repository. Delete these files to force the models to be fit again.

## Data

The `analysis/gpdd/` folder contains data from the Global Population Dynamics
Database:

NERC Centre for Population Biology, Imperial College (2010). *The Global
Population Dynamics Database Version 2*.
<http://www3.imperial.ac.uk/cpb/databases/gpdd>

The file `analysis/brook-etal.csv` contains a copy of the data from the
supplemental Excel spreadsheet in:

Brook, B.W., Traill, L.W. & Bradshaw, C.J.A. (2006).
Minimum viable population sizes and global extinction risk are unrelated.
*Ecol. Lett.* 9, 375-382. <http://doi.org/10.1111/j.1461-0248.2006.00883.x>

## Package versions

It's possible that some portions of the analysis were run with earlier versions of R packages, but the versions of R packages installed at the time of publication were:

```{r, results='hide', echo=FALSE, message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(dplyr)
pkgs <- c("rstan", "dplyr", "plyr", "reshape2", "ggplot2", "gridExtra", 
  "RColorBrewer", "grImport", "TeachingDemos", "metRology", "xtable", "devtools",
  "skewt", "foreach", "Rcpp", "png")
x <- devtools::session_info(pkgs = pkgs)
x <- as.data.frame(x$packages)
x <- dplyr::filter(x, package %in% pkgs) %>% 
  dplyr::select(-`*`, -date, -source) %>% 
  dplyr::arrange(package)
```

```{r, echo=FALSE, results=TRUE}
x
```

### 2017-03-22 text/supp.tex
Final minor tweaks to figures and supplement

@@
 The following pseudo R code illustrates the procedure to generate process
-noise with effective $\nu$ values within a CV of 0.2 of the specified value
-(the actual code is available at
-\url{https://github.com/seananderson/heavy-tails}):
+noise with effective $\nu$ values within a CV of 0.2 of the specified value:
 
@@
 \section{Code availability} Data and code to reproduce our analysis are
-available at\\ \url{https://github.com/seananderson/heavy-tails}
+available at\\ \url{https://github.com/seananderson/heavy-tails}\\ and are archived at\\ \url{http://doi.org/10.5281/zenodo.321930}
 

### 2017-03-22 analysis/6-plot-t-nu-eg.R

@@ -48,7 +48,7 @@ lines(x, dt(x, 5), lty = 1, col = cols[3], lwd = lwd)
 lines(x, dt(x, 20), lty = 1, col = cols[4], lwd = lwd)
 lines(x, dt(x, 1e6), lty = 1, col = cols[5], lwd = lwd)
 par(xpd = NA)
-mtext("Probability density", side = 2, outer = TRUE, line = 1.2, las = 0,
+mtext("Probability density", side = 2, outer = TRUE, line = 1.1, las = 0,
   cex = 0.9)
 mtext("Standard deviations", side = 1, outer = FALSE, line = 2, las = 0,
   cex = 0.9)
@@ -173,18 +173,18 @@ for(j in 1:(nrow(bs_bg) - 1)) {
 }
 make_hist2(e1)
 text(2.5, 0.022, "F", pos = 4, font = 2, cex = 1.4)
-text(8.5, 0.022, quote(Heavy~tails), pos = 4)
+text(13, 0.022, quote(Heavy~tails), pos = 4)
 text(30, 0.009, "Prior", pos = 4)
-text(8.5, 0.0015, "Posterior samples", pos = 4)
+text(13, 0.0015, "Posterior samples", pos = 4)
 mtext("Model fits", side = 3, line = 0.1, cex = 0.9)
 make_hist2(e2)
 text(1, 0.022, "G", pos = 4, font = 2, cex = 1.4)
-text(8.5, 0.022, quote(Slightly~heavy~tails), pos = 4)
+text(13, 0.022, quote(Slightly~heavy~tails), pos = 4)
 mtext("Samples or probability", side = 2, line = 0.5, las = 0, cex = 0.9)
 make_hist2(e3)
 axis(1, at = c(2, seq(20, 80, 20)), col = box.col, mgp = c(2, 0.4, 0))
 mtext(quote(widehat(nu)), side = 1, line = 2, las = 0, cex = 0.9)
 text(37, 0.015, "Median + IQR", pos = 4)
 text(1, 0.022, "H", pos = 4, font = 2, cex = 1.4)
-text(8.5, 0.022, quote(Normal~tails), pos = 4)
+text(13, 0.022, quote(Normal~tails), pos = 4)
 dev.off()
